### 2020-10-19 inst/gitbook/01_getting_started.Rmd
gitbook INIT

---
title: "Getting start"
author: "Thomas Guillerme (guillert@tcd.ie)"
date: "`r Sys.Date()`"
bibliography: [../References.bib, ../packages.bib]
output:
  html_document: default
  pdf_document: default
---

# A simple analyses

## What you need?

 * The speciation and extinction parameter

 ```{r, eval = FALSE}
## The birth death parameters
my_params <- list(speciation = 1,
                  extinction = 1/3)
```

 * A stopping rule

 ```{r, eval = FALSE}
## Creating a stopping rule
my_stop_rule <- list(max.taxa = 15)
```

 * A "traits" object

 ```{r, eval = FALSE}
## A simple Brownian motion object
my_bm_trait <- make.traits()
```

## How to simulate disparity (traits) and diversity (taxa)?

 ```{r, eval = FALSE}
## Running a pretty basic simulation
simple_dads <- dads(bd.params = my_params,
                    stop.rule = my_stop_rule,
                    traits    = my_bm_traits)
```

> Note that you can also simply simulate diversity by setting the traits argument to `NULL`. This will simply simulate a birth death tree. If this is your objective, you'd probably be better off using one of the excellent `R` packages dedicated to this task like `TreeSim`.

### 2020-10-19 inst/gitbook/index.Rmd
gitbook INIT

--- 
title: "dads manual"
author: "Thomas Guillerme (guillert@tcd.ie)"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
output: bookdown::gitbook
documentclass: book
bibliography: [../references.bib]
link-citations: yes
github-repo: TGuillerme/dads
description: "dads R package vignette"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# `dads`

Disparity and diversity simulations.

The core of this package is based on diversitree function blabalba

## What is **not** `dads`s?

Link to RPANDA, diversitree, etc.

## What is `dads`?

Blablaba

### Modular?

Blablaba

## Installing and running the package

You can install this package easily, directly from the CRAN:

```{r, eval = FALSE}
install.packages("dads")
```

Alternatively, for the most up to data version and some functionalities not compatible with the CRAN, you can use the package through GitHub using `devtool` (see [to CRAN or not to CRAN?](#noCRAN) for more details):

```{r, eval = FALSE}
## Checking if devtools is already installed
if(!require(devtools)) install.packages("devtools")

## Installing the latest released version directly from GitHub
install_github("TGuillerme/dads", ref = "release")
```

```{r, eval = FALSE, echo = FALSE, message = FALSE}
## loading the package and setting up the start seed.
library(dads)
## Setting a random seed for repeatability
set.seed(123)
```

## Help

If you need help with the package, hopefully the following manual will be useful.
However, parts of this package are still in development and some other parts are probably not covered.
Thus if you have suggestions or comments on on what has already been developed or will be developed, please send me an email (<guillert@tcd.ie>) or if you are a GitHub user, directly create an issue on the [GitHub page](https://github.com/TGuillerme/dads).

<!-- 
## Citations

To cite the package, this manual or some specific functionalities, you can use the following references:

Furthermore, don't forget to cite `R`:

> R Core Team (2020). R: A language and environment for statistical computing. R Foundation for Statistical Computing, Vienna, Austria. URL https://www.R-project.org/.

Bonus: you can also cite `ape` since the `dispRity` package heavily relies on it:

> Paradis E. & Schliep K. 2019. ape 5.0: an environment for modern phylogenetics and evolutionary analyses in R. Bioinformatics 35: 526-528.
 -->

<!-- ### Why is it important to cite us?

Aside from how science works (if you're using a method from a specific paper, cite that specific paper to refer to that specific method), why is it important to also cite the package and the manual?

All the people involve in making the `dads` package happened to do it enthusiastically, freely and most amazingly without asking anything in return!
I created the package with this idea in mind and I am still sticking to it.
However, academia (the institutions and people producing science around the globe) is unfortunately not optimal at many level (some might even say "broken"): high impact papers attract big grants that attract high impact papers and big grants again, all this along with livelihood, permanent position and job security.
Unfortunately however, method development has a hard time to catch up with the current publish or perish system: constantly updating the `dads` package and this manual is hugely time consuming (but really fun!) and that is not even taking into account maintenance and helping users.
Although I do truly believe that this time spent doing these things modestly help the scientific endeavour, it does not contribute to our paper list!

Therefore, by citing the package and this manual, you help provide visibility to other workers and you might help them in their work!
And you directly contribute in making this project fun for all the people involved and most of all, free, updated and independent from the publish and perish system!

Thank you! -->
<!-- ## Associated papers -->

<!-- Here are some papers that have used the `dispRity` package: -->

### 2020-10-19 inst/gitbook/compiling_gitbook.sh

@@ -1,0 +1,25 @@
+#!/bin/sh
+
+rm dads_manual.Rmd
+
+## Remove files before hand
+rm -R _book/
+rm -R _bookdown_files/
+
+## Compile the html
+R -e 'bookdown::render_book("index.Rmd", "bookdown::gitbook")' # HTML version
+
+## Comment out the badges
+sed 's/<!-- badges out start -->/<!-- badges out start/g' index.Rmd > index.tmp
+sed 's/<!-- badges out end   -->/badges out end   -->/g' index.tmp > index.Rmd
+rm index.tmp
+
+## Compile the pdf
+R -e 'bookdown::render_book("index.Rmd", "bookdown::pdf_book")' # PDF version
+
+## Reactivate the badges
+sed 's/<!-- badges out start/<!-- badges out start -->/g' index.Rmd > index.tmp
+sed 's/badges out end   -->/<!-- badges out end   -->/g' index.tmp > index.Rmd
+rm index.tmp
+
+rm -R _bookdown_files/

### 2020-10-19 inst/gitbook/02_detailed_functions.Rmd
Create 02_detailed_functions.Rmd

---
title: "Getting start"
author: "Thomas Guillerme (guillert@tcd.ie)"
date: "`r Sys.Date()`"
bibliography: [../References.bib, ../packages.bib]
output:
  html_document: default
  pdf_document: default
---

# Making trait processes with `make.traits()`

Trait processes always need the following input arguments

 * `x0`: the previous trait value(s)
 * `edge.length`: the branch length value
 * `...`: a placeholder for any extra arguments

If it contains at least these arguments, it should be handled correctly by `make.traits` an `dads`

> It is highly recommended to set all arguments (but `x0`) to a default value.

### 2020-10-20 inst/gitbook/02_detailed_functions.Rmd
Some info on make traits

@@
 
-Trait processes always need the following input arguments
+## The process
+
+The function `make.traits` allows you to design the process of a trait or a set of traits.
+Here, the process of a trait designates the rules to generate the trait through time while simulating a phylogeny.
+This process can depend on the previous state in the tree (i.e. the trait of the ancestor) and the branch length to the descendant.
+One classic example is the [Brownian motion process (or Weiner process)](https://en.wikipedia.org/wiki/Brownian_motion).
+Note that it can depend from both but does not necessary needs (i.e. the process can be only based on the previous state or only on branch length or on neither).
+
+Trait processes in `dads` are functions that must always intake the following arguments by default.
 
@@
 
-If it contains at least these arguments, it should be handled correctly by `make.traits` an `dads`
+For example, the following function would be a valid process (though not dependent on either the previous state nor the branch length):
 
-> It is highly recommended to set all arguments (but `x0`) to a default value.
+```{r}
+## A valid (but useless) process
+valid.process <- function(x0, egde.length = 1, ...) {
+    return(42)
+}
+```
+
+> Note that the argument `edge.length` is set to `1` by default. In general we highly recommend to set all arguments but `x0` to a default value (this really helps the speed of the `dads` function).
+
+On the other hand, the following process (a Brownian motion) is incorrect (it's missing `edge.length` and `...`):
+
+```{r}
+## A wrongly formated process
+invalid.process <- function(x0) {
+    return(rnorm(1, mean = x0))
+}
+```
+
+The `dads` package processes inbuilt processes, namely a multidimensional Brownian motion or a a multidimensional Ornstein-Uhlenbeck process.
+You can find the list of implemented process by looking at the `?trait.process` manual page in `R`.
+
+Once a process is chosen, you can feed it to the `make.traits` function:
+
+```{r}
+## Creating a trait object
+my_trait_object <- make.trait(process = BM.process)
+```
+
+This creates `"dads"` `"traits"` objects that you can print, and visualise using the `plot` function:
+
+```{r}
+## The class of the object
+class(my_trait_object)
+## What's in it?
+my_trait_object
+## What does the process looks like
+plot(my_trait_object)
+```
+
+Note that you can see the multiple options for plotting the trait process by looking at `?plot.dads` manual. Furthermore, you can look at what's actually in the object using:
+
+```{r}
+## What's actually in that object?
+print.dads(my_trait_object, details = TRUE)
+```
+
+As traits can get more and more complex, the automatic printing of its summary allows for a easier display of what's in the traits object.
+
+## The number of traits `n` and the starting values `start`

### 2020-10-20 inst/gitbook/index.Rmd
Some info on make traits

@@
 title: "dads manual"
-author: "Thomas Guillerme (guillert@tcd.ie)"
+author: "Thomas Guillerme (guillert@tcd.ie), Alex Slavenko (email) and others"
 date: "`r Sys.Date()`"

### 2020-10-23 inst/gitbook/01_getting_started.Rmd
Update 01_getting_started.Rmd

@@
 
-# A simple analyses
+```{r, echo = FALSE}
+library(dads)
+set.seed(4242)
+```
 
-## What you need?
+# Getting started
 
- * The speciation and extinction parameter
+## The simplest of all analysis: simulating diversity only
 
- ```{r, eval = FALSE}
-## The birth death parameters
+One of the simplest things to do with the `dads` package is just to simulate a birth death tree.
+For that you can use the function `dads` and specify your stopping rule.
+The stopping rule simply tells the birth death process to step whenever it reaches one of these three conditions:
+
+ * `"max.taxa"   = n` stop when `n` taxa are generated;
+ * `"max.living" = n` stop when there is `n` co-occuring taxa of the same age (i.e. "living" taxa);
+ * `"max.time"   = n` stop when the simulated tree is `n` units of age old (these units are arbitrary);
+
+For example, we might want to generate a birth-death tree with 20 taxa:
+
+```{r}
+## Setting a stopping rule to reach a maximum of 20 taxa
+my_stop_rule <- list(max.taxa = 20)
+```
+
+We can now run the simulations using:
+
+```{r}
+## Running the birth death simulation
+my_tree <- dads(stop.rule = my_stop_rule)
+```
+
+> Note that here we could have specified more than one stopping rule, for example, we might want to run a simulation and stop it if it either reaches 10 taxa or the age 2 using `stop.rule = list(max.time = 2, max.taxa = 10)`.
+The simulation will then stop when either of these conditions are met.
+
+The resulting object is a classic `"phylo"` object that you can simply plot or visualise like so:
+
+```{r}
+## The tree object
+my_tree
+## Plotting it
+plot(my_tree)
+```
+
+### Changing the birth-death parameters
+
+People familiar with the [birth-death models](https://lukejharmon.github.io/pcm/chapter10_birthdeath/) might have noticed that we did not specify two important things here: the speciation parameter (sometimes called "lambda" or "birth") and the extinction parameter (sometimes called "mu", "death" or "background extinction").
+By default `dads` runs a pure birth model (the speciation is set to 1 and the extinction to 0).
+However, you can easily change that by specifying your new birth death parameters:
+
+```{r}
+## my birth death parameters
 my_params <- list(speciation = 1,
@@
 ```
+> Note that here it is not necessary to specify `extinction = 1` since this is the default option, you can always just change the parameter of interest (e.g.
+changing `extinciton = 0` to `extinction = 1/3`).
+However, we think it's good practice to attribute both parameters specifically to avoid any confusion.
 
- * A stopping rule
+You can then run the same birth death tree with extinction:
 
- ```{r, eval = FALSE}
-## Creating a stopping rule
-my_stop_rule <- list(max.taxa = 15)
+```{r}
+## Generating a birth death tree with extinctions:
+my_tree <- dads(bd.params = my_params, stop.rule = my_stop_rule)
+## Visualising the new tree
+plot(my_tree)
 ```
 
- * A "traits" object
+## Slightly more complex: simulating disparity and diversity
 
- ```{r, eval = FALSE}
-## A simple Brownian motion object
-my_bm_trait <- make.traits()
+Chances are that you want to also simulate traits (disparity) along with your diversity (otherwise, we suggest using the [`TreeSim`](https://github.com/tanja819/TreeSim/) package that provides many more birth death models).
+Simulating traits is not much more complicated in `dads`: you'll simply need to create a `"traits"` object using the `make.traits` function.
+These objects can have increasing complexity (see the rest of this tutorial) but we will keep it simple here.
+
+`"traits"` objects contain one or more processes which are the ways to generate the trait.
+The most common of these processes is the [Brownian Motion](https://en.wikipedia.org/wiki/Brownian_motion).
+This is used by default with the `make.traits` function:
+
+```{r}
+## Creating the traits object
+my_trait <- make.traits()
 ```
 
-## How to simulate disparity (traits) and diversity (taxa)?
+This trait object can be simply printed (to see what's in it) or plotted (to see what the process looks like in the absence of a phylogeny):
 
- ```{r, eval = FALSE}
-## Running a pretty basic simulation
-simple_dads <- dads(bd.params = my_params,
-                    stop.rule = my_stop_rule,
-                    traits    = my_bm_traits)
+```{r}
+## Which process is in here?
+my_trait
+## What does it look like?
+plot(my_trait)
 ```
 
-> Note that you can also simply simulate diversity by setting the traits argument to `NULL`.
-This will simply simulate a birth death tree.
-If this is your objective, you'd probably be better off using one of the excellent `R` packages dedicated to this task like `TreeSim`.
+By default, this trait is called "A".
+This is not a really good name but you'll see more about specifying trait names later on.
+If this is what the process should look like (theoretically) you can then add its `"traits"` object to our previous `dads` function to generate the tree and the traits:
+
+```{r}
+## Simulate disparity and diversity
+my_data <- dads(bd.params = my_params,
+                stop.rule = my_stop_rule,
+                traits    = my_trait)
+```
+
+Et voilà! We now have a simple disparity and diversity simulation.
+We can see what's in the results by simply printing it or plotting it:
+
+```{r}
+## What's in there
+my_data
+## Plotting the disparity and diversity
+plot(my_data)
+```
+
+You can then extract the components you need for your specific analysis like so:
+
+```{r}
+## Extracting the tree (a "phylo" object)
+the_generated_tree <- my_data$tree
+class(the_generated_tree)
+
+## Extracting the data (a "matrix")
+the_generated_data <- my_data$data
+class(the_generated_data)
+```
+
+## Slightly more more complex: simulating linked disparity and diversity
+
+The example above is also still pretty simple and easily done through a variety of `R` packages: here the trait and the tree are simulated at the same time but only the tree is simulating the trait (i.e.
+the trait value at a tip is affected by it's ancestor and the branch length leading to it) but not the other way around (the trait value does not affect the tree).
+It is possible to add this aspect using `"modifiers"` objects.
+`"modifiers"` are similar to `"traits"` in that you specify what should go in there and then feed it to your simulation.

### 2020-10-23 inst/gitbook/02_detailed_functions.Rmd
code running
details for make.traits up until add

@@
 
-# Making trait processes with `make.traits()`
+# Detailed functions
 
-## The process
+## Making trait processes with `make.traits()`
+
+```{r, echo = FALSE}
+set.seed(123)
+```
+
+### The process (`process`)
 
@@
 One classic example is the [Brownian motion process (or Weiner process)](https://en.wikipedia.org/wiki/Brownian_motion).
-Note that it can depend from both but does not necessary needs (i.e. the process can be only based on the previous state or only on branch length or on neither).
+Note that it _can_ depend on both the ancestor and the branch length but does _not necessary needs_ (i.e. the process can be only based on the previous state or only on branch length or on neither).
 
@@
 
-> Note that the argument `edge.length` is set to `1` by default. In general we highly recommend to set all arguments but `x0` to a default value (this really helps the speed of the `dads` function).
+> Note that the argument `edge.length` is set to `1` by default.
+In general we highly recommend to set all arguments but `x0` to a default value (this really helps the speeding up the `dads` function).
 
-On the other hand, the following process (a Brownian motion) is incorrect (it's missing `edge.length` and `...`):
+On the other hand, the following process (a unidimensional Brownian motion) is incorrect (it's missing `edge.length` and `...`):
 
@@
 
-The `dads` package processes inbuilt processes, namely a multidimensional Brownian motion or a a multidimensional Ornstein-Uhlenbeck process.
+The `dads` package proposes inbuilt processes, namely a multidimensional Brownian motion (`BM.process`) or a a multidimensional Ornstein-Uhlenbeck process (`OU.process`).
 You can find the list of implemented process by looking at the `?trait.process` manual page in `R`.
@@
 ## Creating a trait object
-my_trait_object <- make.trait(process = BM.process)
+my_trait_object <- make.traits(process = BM.process)
 ```
@@
 ## What's actually in that object?
-print.dads(my_trait_object, details = TRUE)
+print.dads(my_trait_object, all = TRUE)
 ```
@@
 
-## The number of traits `n` and the starting values `start`
+Note that it is possible to make `"traits"` objects with multiple processes (that can be the same):
+
+```{r}
+## 4 traits: two BM, one OU and one normal non process
+four_traits <- make.traits(process = c(BM.process,
+                                       BM.process,
+                                       OU.process,
+                                       no.process))
+four_traits
+```
+
+You can visualise them individually using the `trait` argument in `plot.dads`:
+
+```{r}
+## Plot options (4 plots in one window)
+par(mfrow = c(2,2))
+plot(four_traits, trait = 1)
+plot(four_traits, trait = 2)
+plot(four_traits, trait = 3)
+plot(four_traits, trait = 4)
+```
+
+### The number of traits `n` and the starting values `start`
+
+Two further important arguments are `n` the number of traits per process and `start` the starting values for all traits.
+By default they are set to `n = 1` and `start = 0`.
+This means that `make.traits` will assume that your processes are always unidimensional by default and that they always start with the value `0`.
+It is however possible to change these values.
+
+For example you can use the following to create a three dimensional Brownian motion with each dimensions starting with the value `1`:
+
+```{r}
+## Multidimensional Brownian motion
+make.traits(BM.process, n = 3, start = 1)
+```
+
+Or the following with each dimensions starting with different values (respectively `1`, `2` and `3`):
+
+```{r}
+## Multidimensional Brownian motion
+make.traits(BM.process, n = 3, start = c(1,2,3))
+```
+
+Note that the number of traits are distributed per processes.
+If the traits contains multiple process, the number of traits are distributed per processes:
+
+```{r}
+## two 3D processes (BM and OU)
+make.traits(c(BM.process, OU.process), n = 3)
+## one 1D processes (BM) and one 4D process (OU)
+make.traits(c(BM.process, OU.process), n = c(1, 4))
+```
+
+And starting values are distributed for all the traits or for the traits one by one:
+
+```{r}
+## two 3D processes (BM and OU) starting with 1
+make.traits(c(BM.process, OU.process), n = 3, start = 1)
+## two 3D processes (BM and OU) starting with values 1 to 6
+make.traits(c(BM.process, OU.process), n = 3, start = 1:6)
+## two 3D processes (BM and OU) with the two first ones starting
+## with 1 and the 4 other ones with the default (0)
+make.traits(c(BM.process, OU.process), n = 3, start = c(1,1))
+```
+
+## Extra argument for the processes with `process.args`
+
+You can also feed extra arguments to your process(es) functions.
+For example, the inbuilt process `no.process` (that is just a number generator not based on the previous value `x0` or the branch length) can intake a specific random number generator as a function:
+
+```{r}
+## no process trait using the normal distribution (default)
+make.traits(no.process, process.args = list(fun = rnorm))
+## no process trait using the uniform distribution
+## bounded between 1 and 100
+make.traits(no.process, process.args = list(fun = runif, min = 1, max = 100))
+```
+
+You can also add multiple extra arguments for multiple processes giving them as a list.
+
+```{r}
+## Two traits with no process:one normal and one uniform (1,100)
+make.traits(process      = c(no.process, no.process),
+            process.args = list(list(fun = rnorm),
+                                list(fun = runif, min = 1, max = 100)))
+```
+
+If one process do not need extra argument you must still give it and extra `NULL` process argument:
+
+```{r}
+## Three traits with no process:
+## one default, one lognormal and one uniform (1,100)
+make.traits(process      = c(no.process, no.process, no.process),
+            process.args = list(## Extra arguments for the first process (none)
+                                list(NULL),
+                                ## Extra arguments for the second process
+                                list(fun = rlnorm),
+                                ## Extra arguments for the third process
+                                list(fun = runif, min = 1, max = 100)))
+```
+
+## Combining multiple traits with `add`
+
+## Naming and testing traits (`trait.names` and `test`)

### 2020-10-23 inst/gitbook/index.Rmd
Update index.Rmd

@@
 
-The core of this package is based on diversitree function blabalba
+Allowing to simulate disparity and diversity at the same time with interaction between both.
 
-## What is **not** `dads`s?
-
-Link to RPANDA, diversitree, etc.
+The core of this package is based on the [`diversitree` birth-death algorithm](https://github.com/cran/diversitree/blob/master/R/simulate-bd.R).
 
@@
 
-Blablaba
+There are some very good packages out there to simulate birth death trees (like [TreeSim](https://github.com/tanja819/TreeSim)) or even packages simulating traits (disparity) and diversity jointly (like [RPANDA](https://github.com/cran/RPANDA) or [PETER](https://github.com/PuttickMacroevolution/PETER)).
+We strongly advice you have a look at these packages first as they might be more appropriate for your tasks.
+
+`dads` aims to be a highly modular and friendly version of all these packages: it allows to simulate disparity and diversity jointly with a vast array of options that can be easily modified by users.
+For example, you can easily generate any type of process to generate a trait (BM, OU, something else, etc...) in multiple dependent or independent dimensions through `"traits"` objects.
+You can then specify how the traits should affect disparity through `"modifiers"` objects. And finally you can create events (like mass extinctions) through `"events"` objects.
+These objects and how to modify them will be detailed throughout this manual.
+Finally we are putting an emphasise in the development of this package on the speed and reliability of the functions.
 
@@
 
-Blablaba
+Because their is an infinite way you might want to generate disparity and diversity (different traits, different modifiers and different events), the `dads` package is designed to make all these parts easy to code separately and integrate them easily in the `dads` core functions.
+This allows you to simulate finely tuned multidimensional data for your specific project!
 
@@
 
-You can install this package easily, directly from the CRAN:
-
-```{r, eval = FALSE}
-install.packages("dads")
-```
-
-Alternatively, for the most up to data version and some functionalities not compatible with the CRAN, you can use the package through GitHub using `devtool` (see [to CRAN or not to CRAN?](#noCRAN) for more details):
+You can install this package easily, directly from the github:
+<!-- TG: I do intend to release it on CRAN later on -->
 
@@
 
-## Installing the latest released version directly from GitHub
-install_github("TGuillerme/dads", ref = "release")
+## Installing the latest version directly from GitHub
+install_github("TGuillerme/dads")
 ```
@@
 Thus if you have suggestions or comments on on what has already been developed or will be developed, please send me an email (<guillert@tcd.ie>) or if you are a GitHub user, directly create an issue on the [GitHub page](https://github.com/TGuillerme/dads).
-
-<!-- 
-## Citations
-
-To cite the package, this manual or some specific functionalities, you can use the following references:
-
-Furthermore, don't forget to cite `R`:
-
-> R Core Team (2020). R: A language and environment for statistical computing. R Foundation for Statistical Computing, Vienna, Austria. URL https://www.R-project.org/.
-
-Bonus: you can also cite `ape` since the `dispRity` package heavily relies on it:
-
-> Paradis E. & Schliep K. 2019. ape 5.0: an environment for modern phylogenetics and evolutionary analyses in R. Bioinformatics 35: 526-528.
- -->
-
-<!-- ### Why is it important to cite us?
-
-Aside from how science works (if you're using a method from a specific paper, cite that specific paper to refer to that specific method), why is it important to also cite the package and the manual?
-
-All the people involve in making the `dads` package happened to do it enthusiastically, freely and most amazingly without asking anything in return!
-I created the package with this idea in mind and I am still sticking to it.
-However, academia (the institutions and people producing science around the globe) is unfortunately not optimal at many level (some might even say "broken"): high impact papers attract big grants that attract high impact papers and big grants again, all this along with livelihood, permanent position and job security.
-Unfortunately however, method development has a hard time to catch up with the current publish or perish system: constantly updating the `dads` package and this manual is hugely time consuming (but really fun!) and that is not even taking into account maintenance and helping users.
-Although I do truly believe that this time spent doing these things modestly help the scientific endeavour, it does not contribute to our paper list!
-
-Therefore, by citing the package and this manual, you help provide visibility to other workers and you might help them in their work!
-And you directly contribute in making this project fun for all the people involved and most of all, free, updated and independent from the publish and perish system!
-
-Thank you! -->
-<!-- ## Associated papers -->
-
-<!-- Here are some papers that have used the `dispRity` package: -->
+Doing so will not only hopefully help you but also other users since it will help improve this manual!

### 2020-10-26 inst/gitbook/02_detailed_functions.Rmd
Update 02_detailed_functions.Rmd

@@
 
-## Extra argument for the processes with `process.args`
+### Extra argument for the processes with `process.args`
 
@@
 
-## Combining multiple traits with `add`
+### Naming the traits with `trait.names`
 
-## Naming and testing traits (`trait.names` and `test`)
+As traits become more and more complex, it can be useful to give clearer names to each process.
+This is easily done using the `trait.names` argument that attributes one name per process:
+
+```{r}
+## A simple trait with a proper name
+simple_trait <- make.traits(trait.names = "1D Brownian Motion")
+simple_trait
+```
+
+This becomes more useful if we use the complex example above:
+
+```{r}
+## Three named traits with no process:
+## one default, one lognormal and one uniform (1,100)
+make.traits(process      = c(no.process, no.process, no.process),
+            process.args = list(## Extra arguments for the first process (none)
+                                list(NULL),
+                                ## Extra arguments for the second process
+                                list(fun = rlnorm),
+                                ## Extra arguments for the third process
+                                list(fun = runif, min = 1, max = 100)),
+            ## Naming each trait
+            trait.names  = c("Normal", "LogNormal", "Uniform(1,100)"))
+```
+
+### Combining multiple traits with `add`
+
+You can also add traits to already existing trait objects using the simple `add` option.
+This option just intakes a `"dads"` `"traits"` object and the additional process(es) will be added to it. For example:
+
+```{r}
+## Creating on simple default Brownian motion
+one_process <- make.traits(trait.names = "BM")
+
+## Creating a new trait (a 3D OU.process)
+## and adding the previous one
+two_processes <- make.traits(OU.process, n = 3, add = one_process,
+                             trait.names = "3D OU")
+
+## Only one process
+one_process
+
+## The two processes
+two_processes
+```
+
+### Testing the traits with `test`
+
+This bit is more for development.
+We highly suggest leaving `test = TRUE` so that `make.traits` returns an error if a process or its additional arguments (`process.args`) are not formatted correctly.
+`make.traits` will error if the trait cannot be directly passed to `dads`.
+However, in some specific cases (again, probably mainly for development and debugging) it could be useful to skip the tests using `test = FALSE`.

### 2020-11-09 inst/gitbook/02_make.traits.Rmd
Updated book

--- 2020-10-26 inst/gitbook/02_detailed_functions.Rmd
@@
 
-# Detailed functions
-
-## Making trait processes with `make.traits()`
+# Making trait processes with `make.traits()`
 
@@
 
-### The process (`process`)
+## The process (`process`)
 
@@
 
-### The number of traits `n` and the starting values `start`
+## The number of traits `n` and the starting values `start`
 
@@
 
-### Extra argument for the processes with `process.args`
+## Extra argument for the processes with `process.args`
 
@@
 
-### Naming the traits with `trait.names`
+## Naming the traits with `trait.names`
 
@@
 
-### Combining multiple traits with `add`
+## Combining multiple traits with `add`
 
@@
 
-### Testing the traits with `test`
+## Testing the traits with `test`
 
@@
 However, in some specific cases (again, probably mainly for development and debugging) it could be useful to skip the tests using `test = FALSE`.
+
+## Templates for making your very own process
+
+As detailed above, any process of your own design will work as long as it is a function that takes at least the arguments `x0` and `edge.length`.
+You can be imaginative and creative when designing your own process but here are two detailed example functions for a unidimensional Brownian Motion and Ornstein-Uhlenbeck process that you can use for a start (or not).
+Remember it is good practice for `dads` processes to set all the arguments but `x0` with default values (just in case).
+Also, note that the functions below are not equal to the already implemented `BM.process` and `OU.process` but are rather generalised/simplified version that you can use as a template
+
+### A simple Brownian Motion process template
+
+```{r}
+## A simple Brownian motion process
+my.BM.process <- function(x0, edge.length = 1, sd = 1, ...) {
+    ## Drawing a random number from a normal distribution
+    ## with x0 as the and a given standard deviation
+    ## and depending on branch (edge) length
+    result <- rnorm(n = 1, mean = x0, sd = sd * edge.length)
+
+    ## Return the number
+    return(result)
+}
+```
+
+### A simple Ornstein-Uhlenbeck process template
+
+```{r}
+## A simple Ornstein-Uhlenbeck motion process
+function(x0, edge.length = 1, var = 1, alpha = 1, ...) {
+    ## Calculate the mean based on alpha
+    mean <- x0 * exp(-alpha)
+    ## Calculate the standard deviation based on alpha and the variance
+    sd <- sqrt(var/(2 * alpha) * (1 - exp(-2 * alpha)))
+    ## Draw a random number from a normal distribution
+    ## using this mean and standard deviation
+    ## and depending on branch (edge) length
+    result <- rnorm(n = 1, mean = mean, sd = sd * branch.length)
+
+    ## Return the number
+    return(result)
+}
+```

### 2020-11-09 inst/gitbook/index.Rmd
Updated book

@@
 Doing so will not only hopefully help you but also other users since it will help improve this manual!
+
+## How does `dads` work?
+
+Basically, the `dads` function intakes your personalised `traits`, `modifiers` and `events` to generate your disparity and diversity.
+You will find more details about how these objects (`traits`, `modifiers` and `events`) work in the rest of the tutorial but here is a graphical representation of how `dads` work:
+
+![Schematised summary of the `dads` package architecture](dads_structure.pdf)

### 2020-11-13 inst/gitbook/01_getting_started.Rmd
gitbook up

@@
 
+You can find much more on how to design trait objects in the [`make.traits` section](#maketraits).
+
 ## Slightly more more complex: simulating linked disparity and diversity
@@
 `"modifiers"` are similar to `"traits"` in that you specify what should go in there and then feed it to your simulation.
+
+`"modifiers"` affect two key steps of the birth-death process: the calculation of the waiting time (i.e. the component generating branch lengths) and the triggering of speciation or extinction events.
+These events can be modified using some `condition` and `modify` function.
+In other words, when reaching a certain condition specified by a `condition` function, the birth-death process will modify either the branch length or the speciation (or extinction) probability by applying a `modify` function.
+
+You can use the function `make.modifiers` to design a specific `"modifiers"` object.
+By default, this function generates a `"modifiers"` object that affects branch length and speciation in the following way:
+
+ * branch length is a randomly drawn number from an exponential distribution with a rate equal to the current number of taxa multiplied by the sum of the speciation and extinction rates.
+ * speciation is triggered if a randomly drawn number (from a (0,1) uniform distribution) is smaller than the ratio between the speciation rate and the sum of the speciation and extinction rates.
+If that random number is greater, the lineage goes extinct.
+
+Note that these are default for a birth death tree and were actually already applied in the examples before (without specifying a modifier):
+
+```{r}
+## Make a default modifiers
+default_modifiers <- make.modifiers()
+## What's in it?
+default_modifiers
+```
+
+This will not do much to our simulations compared to the previous trait and tree simulation but we can provide our modifiers object to the `dads` function:
+
+```{r}
+set.seed(1)
+## Setting the simulation parameters
+extinction_02 <- list(extinction = 0.4)
+living_20     <- list(max.living = 20)
+BM_trait      <- make.traits()
+set.seed(4242)
+## Simulate disparity and diversity
+default_data <- dads(bd.params = extinction_02,
+                     stop.rule = living_20,
+                     traits    = BM_trait,
+                     modifiers = default_modifiers)
+default_data
+```
+
+Note however that the printing information is now updated to state that you've add a modifier (even though it's a default one).
+
+For more interesting simulations however, you can provide modifiers that actually modify the birth death process.
+We can create one for example that makes species go extinct if their ancestor have a negative trait value.
+For that we need to create a modifiers object that modifies the `speciation` process with a specific condition and a specific modification when that condition is met.
+First we are going to create the modification function.
+This function must intake the argument `x` and, in our case, return a logical value: `TRUE` is for speciate and `FALSE` is for go extinct.
+We want this `modify` function to always return `FALSE` (go extinct) since it will be triggered by `condition` (we will see that in a minute).
+
+```{r}
+## Going extinct
+staying.alive <- function(x) return(TRUE)
+```
+
+Now we want this function to only trigger when an ancestor has a negative trait value.
+We can do that by specifying our `condition` function (when the ancestor is trait is negative) apply our modification (here the `staying.alive` function).
+For that we can use the `parent.traits` utility function that is optimised for accessing traits in the birth death process (but you can of course write your own).
+This function intakes the `trait.values` and `parent.lineage` arguments, two arguments that you can leave named as they are to facilitate `dads`s understanding of what you which to asses:
+
+```{r}
+## Triggering a modification only if the ancestor trait is negative
+negative.ancestor <- function(trait.values, parent.lineage) {
+    return(all(parent.traits(trait.values, parent.lineage) < 0))
+}
+```
+
+Note that we use the function `all` here to evaluate all traits (if the data is multidimensional). 
+We can then provide these two functions (the condition `negative.ancestor` and how to modify the speciation event when this condition is met `staying.alive`).
+If you are an advances `dads` user, you can design your own `speciation` function but if you just want to use a normal `speciation` function, you can use the default one from `dads` called...
+`speciation`.
+
+```{r}
+## Making a modifier for species to go extinct if
+## their ancestor's trait value is (or are) negative
+negatives_extinct <- make.modifiers(
+                        ## The speciation function (default)
+                        speciation = speciation,
+                        ## What to modify
+                        modify     = staying.alive,
+                        ## When to modify it
+                        condition  = negative.ancestor)
+## What's in it?
+negatives_extinct
+```
+
+Note that the `make.modifiers` function tests whether the input is compatible with `dads` by default so unless you have an error message, your `modifiers` will work!
+We can now simulate our tree and traits with our modifier: species will go extinct if their ancestor have a negative trait value:
+
+```{r}
+set.seed(4242)
+## Simulate disparity and diversity
+biased_data <- dads(bd.params = extinction_02,
+                    stop.rule = living_20,
+                    traits    = BM_trait,
+                    modifiers = negatives_extinct)
+biased_data
+```
+
+We can now compare the two trees and their trait values.
+Note that we've used the same starting seed for both trees so the only thing differing between them is the modifier!
+Also, although species
+
+```{r, fig.height = 6, fig.width = 12}
+par(mfrow = c(1,2))
+plot(default_data, main = "default results")
+plot(biased_data, main = "results with the modifier")
+```
+
+You can find much more how to design modifiers in the [`make.modifiers` section](#makemodifiers).

### 2020-11-13 inst/gitbook/02_make.traits.Rmd
gitbook up

@@
 ---
-title: "Getting start"
+title: "`make.traits`"
 author: "Thomas Guillerme (guillert@tcd.ie)"
@@
 
-# Making trait processes with `make.traits()`
+# Making trait processes with `make.traits()` {#maketraits}
 
@@
 ## A simple Ornstein-Uhlenbeck motion process
-function(x0, edge.length = 1, var = 1, alpha = 1, ...) {
+my.OU.process <- function(x0, edge.length = 1, var = 1, alpha = 1, ...) {
     ## Calculate the mean based on alpha

### 2020-11-13 inst/gitbook/03_make.modifiers.Rmd
gitbook up

---
title: "`make.modifiers`"
author: "Thomas Guillerme (guillert@tcd.ie)"
date: "`r Sys.Date()`"
bibliography: [../References.bib, ../packages.bib]
output:
  html_document: default
  pdf_document: default
---

# Making modifiers with `make.modifiers()` {#makemodifiers}

```{r, echo = FALSE}
set.seed(123)
```

> TODO: just ideas so far

## The branch length function (`branch.length`)

A branch length function that is constant

```{r, eval = FALSE}
constant.brlen <- function(bd.params = NULL,
                         n.taxa = NULL,
                         parent.lineage = NULL,
                         trait.values = NULL,
                         modify.fun = NULL) {
    return(1)
}

constant_brlen <- dads(bd.params = list(extinction = 0.25),
                     stop.rule = list(max.living = 20),
                     traits    = make.traits(),
                     modifiers = make.modifiers(
                        branch.length = constant.brlen))

plot(constant_brlen, main = "constant branch length")
```

## The speciation function (`speciation`)

```{r, eval = FALSE}
random.brlen <- function(bd.params = NULL,
                         n.taxa = NULL,
                         parent.lineage = NULL,
                         trait.values = NULL,
                         modify.fun = NULL) {
    return(runif(1))
}

random.spec  <- function(bd.params = NULL,
                         n.taxa = NULL,
                         parent.lineage = NULL,
                         trait.values = NULL,
                         modify.fun = NULL) {
    return(sample(c(TRUE, FALSE), 1))
}

set.seed(1)
yule_tree <- dads(bd.params = list(extinction = 0),
                     stop.rule = list(max.living = 20),
                     traits    = make.traits(),
                     modifiers = make.modifiers(
                        branch.length = random.brlen,
                        speciation    = random.spec))

plot(yule_tree, main = "a (failing) yule(ish) tree")
```

## The condition function (`condition`)

## The modify function (`modify`)

## Combining and editing modifiers (`add`)

## Testing modifiers (`test`)

### 2020-11-13 inst/gitbook/03_make.modifiers.Rmd
First manual version

@@
 ## Testing modifiers (`test`)
+
+## Demo runnable
+
+```{r, fig.height = 16, fig.width = 8}
+bd.params <- list(speciation = 1, extinction = 1/3)
+traits <- make.traits()
+stop.rule <- list(max.taxa = 20, max.living = Inf, max.time = Inf)
+modifiers <- NULL
+events <- NULL
+null.error <- NULL
+
+## modifiers
+condition <- function(trait.values, parent.lineage) return(parent.traits(trait.values, parent.lineage) < 0)
+modify <- function(x) return(x * 20)
+
+## Setting up the different modifiers
+modify_speciation <- make.modifiers(speciation    = speciation.trait,
+                                    condition     = condition,
+                                    modify        = modify)
+modify_brlen <- make.modifiers(branch.length = branch.length.trait,
+                               condition     = condition,
+                               modify        = modify)
+modify_speciation_brlen <- make.modifiers(branch.length = branch.length.trait,
+                                          speciation    = speciation.trait,
+                                          condition     = condition,
+                                          modify        = modify)
+
+## Test normal (no modifiers)
+set.seed(1)
+test <- dads(bd.params, stop.rule, traits, null.error = 20)
+par(mfrow = c(4,2))
+plot(test$tree)
+plot.dads(test, main = "random tree + trait")
+
+## Test with modifiers
+set.seed(1)
+trait_table <- NULL
+test <- dads(bd.params, stop.rule, traits,
+                                modifiers = modify_speciation,
+                                null.error = 20)
+plot(test$tree)
+plot.dads(test, main = "Skewed speciation")
+
+set.seed(1)
+trait_table <- NULL
+test <- dads(bd.params, stop.rule, traits,
+                                modifiers = modify_brlen,
+                                null.error = 20)
+plot(test$tree)
+plot.dads(test, main = "Skewed branch length")
+
+set.seed(1)
+trait_table <- NULL
+test <- dads(bd.params, stop.rule, traits,
+                                modifiers = modify_speciation_brlen,
+                                null.error = 20)
+plot(test$tree)
+plot.dads(test, main = "Skewed branch length and speciation")
+
+```

### 2020-11-13 inst/gitbook/_book/dads_manual.tex
First manual version

\PassOptionsToPackage{unicode=true}{hyperref} % options for packages loaded elsewhere
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[]{book}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provides euro and other symbols
\else % if luatex or xelatex
  \usepackage{unicode-math}
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage[]{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\usepackage{hyperref}
\hypersetup{
            pdftitle={dads manual},
            pdfauthor={Thomas Guillerme (guillert@tcd.ie), Alex Slavenko (email) and others},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\usepackage{longtable,booktabs}
% Fix footnotes in tables (requires footnote package)
\IfFileExists{footnote.sty}{\usepackage{footnote}\makesavenoteenv{longtable}}{}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

% set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

\usepackage{booktabs}
\usepackage[]{natbib}
\bibliographystyle{plainnat}

\title{dads manual}
\author{Thomas Guillerme (\href{mailto:guillert@tcd.ie}{\nolinkurl{guillert@tcd.ie}}), Alex Slavenko (email) and others}
\date{2020-11-13}

\begin{document}
\maketitle

{
\setcounter{tocdepth}{1}
\tableofcontents
}
\hypertarget{dads}{%
\chapter{\texorpdfstring{\texttt{dads}}{dads}}\label{dads}}

Disparity and diversity simulations.

Allowing to simulate disparity and diversity at the same time with interaction between both.

The core of this package is based on the \href{https://github.com/cran/diversitree/blob/master/R/simulate-bd.R}{\texttt{diversitree} birth-death algorithm}.

\hypertarget{what-is-dads}{%
\section{\texorpdfstring{What is \texttt{dads}?}{What is dads?}}\label{what-is-dads}}

There are some very good packages out there to simulate birth death trees (like \href{https://github.com/tanja819/TreeSim}{TreeSim}) or even packages simulating traits (disparity) and diversity jointly (like \href{https://github.com/cran/RPANDA}{RPANDA} or \href{https://github.com/PuttickMacroevolution/PETER}{PETER}).
We strongly advice you have a look at these packages first as they might be more appropriate for your tasks.

\texttt{dads} aims to be a highly modular and friendly version of all these packages: it allows to simulate disparity and diversity jointly with a vast array of options that can be easily modified by users.
For example, you can easily generate any type of process to generate a trait (BM, OU, something else, etc\ldots{}) in multiple dependent or independent dimensions through \texttt{"traits"} objects.
You can then specify how the traits should affect disparity through \texttt{"modifiers"} objects. And finally you can create events (like mass extinctions) through \texttt{"events"} objects.
These objects and how to modify them will be detailed throughout this manual.
Finally we are putting an emphasise in the development of this package on the speed and reliability of the functions.

\hypertarget{modular}{%
\subsection{Modular?}\label{modular}}

Because their is an infinite way you might want to generate disparity and diversity (different traits, different modifiers and different events), the \texttt{dads} package is designed to make all these parts easy to code separately and integrate them easily in the \texttt{dads} core functions.
This allows you to simulate finely tuned multidimensional data for your specific project!

\hypertarget{installing-and-running-the-package}{%
\section{Installing and running the package}\label{installing-and-running-the-package}}

You can install this package easily, directly from the github:

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{## Checking if devtools is already installed}
\ControlFlowTok{if}\NormalTok{(}\OperatorTok{!}\KeywordTok{require}\NormalTok{(devtools)) }\KeywordTok{install.packages}\NormalTok{(}\StringTok{"devtools"}\NormalTok{)}

\CommentTok{## Installing the latest version directly from GitHub}
\KeywordTok{install_github}\NormalTok{(}\StringTok{"TGuillerme/dads"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\hypertarget{help}{%
\section{Help}\label{help}}

If you need help with the package, hopefully the following manual will be useful.
However, parts of this package are still in development and some other parts are probably not covered.
Thus if you have suggestions or comments on on what has already been developed or will be developed, please send me an email (\href{mailto:guillert@tcd.ie}{\nolinkurl{guillert@tcd.ie}}) or if you are a GitHub user, directly create an issue on the \href{https://github.com/TGuillerme/dads}{GitHub page}.
Doing so will not only hopefully help you but also other users since it will help improve this manual!

\hypertarget{how-does-dads-work}{%
\section{\texorpdfstring{How does \texttt{dads} work?}{How does dads work?}}\label{how-does-dads-work}}

Basically, the \texttt{dads} function intakes your personalised \texttt{traits}, \texttt{modifiers} and \texttt{events} to generate your disparity and diversity.
You will find more details about how these objects (\texttt{traits}, \texttt{modifiers} and \texttt{events}) work in the rest of the tutorial but here is a graphical representation of how \texttt{dads} work:

\begin{figure}
\centering
\includegraphics{dads_structure.pdf}
\caption{Schematised summary of the \texttt{dads} package architecture}
\end{figure}

\begin{verbatim}
## Loading required package: dispRity
\end{verbatim}

\hypertarget{getting-started}{%
\chapter{Getting started}\label{getting-started}}

\hypertarget{the-simplest-of-all-analysis-simulating-diversity-only}{%
\section{The simplest of all analysis: simulating diversity only}\label{the-simplest-of-all-analysis-simulating-diversity-only}}

One of the simplest things to do with the \texttt{dads} package is just to simulate a birth death tree.
For that you can use the function \texttt{dads} and specify your stopping rule.
The stopping rule simply tells the birth death process to step whenever it reaches one of these three conditions:

\begin{itemize}
\tightlist
\item
  \texttt{"max.taxa"\ \ \ =\ n} stop when \texttt{n} taxa are generated;
\item
  \texttt{"max.living"\ =\ n} stop when there is \texttt{n} co-occuring taxa of the same age (i.e. ``living'' taxa);
\item
  \texttt{"max.time"\ \ \ =\ n} stop when the simulated tree is \texttt{n} units of age old (these units are arbitrary);
\end{itemize}

For example, we might want to generate a birth-death tree with 20 taxa:

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{## Setting a stopping rule to reach a maximum of 20 taxa}
\NormalTok{my_stop_rule <-}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{max.taxa =} \DecValTok{20}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

We can now run the simulations using:

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{## Running the birth death simulation}
\NormalTok{my_tree <-}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{stop.rule =}\NormalTok{ my_stop_rule)}
\end{Highlighting}
\end{Shaded}

\begin{quote}
Note that here we could have specified more than one stopping rule, for example, we might want to run a simulation and stop it if it either reaches 10 taxa or the age 2 using \texttt{stop.rule\ =\ list(max.time\ =\ 2,\ max.taxa\ =\ 10)}. The simulation will then stop when either of these conditions are met.
\end{quote}

The resulting object is a classic \texttt{"phylo"} object that you can simply plot or visualise like so:

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{## The tree object}
\NormalTok{my_tree}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## 
## Phylogenetic tree with 20 tips and 19 internal nodes.
## 
## Tip labels:
##   t1, t2, t3, t4, t5, t6, ...
## Node labels:
##   n1, n2, n3, n4, n5, n6, ...
## 
## Rooted; includes branch lengths.
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{## Plotting it}
\KeywordTok{plot}\NormalTok{(my_tree)}
\end{Highlighting}
\end{Shaded}

\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-6-1.pdf}

\hypertarget{changing-the-birth-death-parameters}{%
\subsection{Changing the birth-death parameters}\label{changing-the-birth-death-parameters}}

People familiar with the \href{https://lukejharmon.github.io/pcm/chapter10_birthdeath/}{birth-death models} might have noticed that we did not specify two important things here: the speciation parameter (sometimes called ``lambda'' or ``birth'') and the extinction parameter (sometimes called ``mu'', ``death'' or ``background extinction'').
By default \texttt{dads} runs a pure birth model (the speciation is set to 1 and the extinction to 0).
However, you can easily change that by specifying your new birth death parameters:

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{## my birth death parameters}
\NormalTok{my_params <-}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{speciation =} \DecValTok{1}\NormalTok{,}
                  \DataTypeTok{extinction =} \DecValTok{1}\OperatorTok{/}\DecValTok{3}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{quote}
Note that here it is not necessary to specify \texttt{extinction\ =\ 1} since this is the default option, you can always just change the parameter of interest (e.g.~changing \texttt{extinciton\ =\ 0} to \texttt{extinction\ =\ 1/3}). However, we think it's good practice to attribute both parameters specifically to avoid any confusion.
\end{quote}

You can then run the same birth death tree with extinction:

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{## Generating a birth death tree with extinctions:}
\NormalTok{my_tree <-}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ my_params, }\DataTypeTok{stop.rule =}\NormalTok{ my_stop_rule)}
\CommentTok{## Visualising the new tree}
\KeywordTok{plot}\NormalTok{(my_tree)}
\end{Highlighting}
\end{Shaded}

\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-8-1.pdf}

\hypertarget{slightly-more-complex-simulating-disparity-and-diversity}{%
\section{Slightly more complex: simulating disparity and diversity}\label{slightly-more-complex-simulating-disparity-and-diversity}}

Chances are that you want to also simulate traits (disparity) along with your diversity (otherwise, we suggest using the \href{https://github.com/tanja819/TreeSim/}{\texttt{TreeSim}} package that provides many more birth death models).
Simulating traits is not much more complicated in \texttt{dads}: you'll simply need to create a \texttt{"traits"} object using the \texttt{make.traits} function.
These objects can have increasing complexity (see the rest of this tutorial) but we will keep it simple here.

\texttt{"traits"} objects contain one or more processes which are the ways to generate the trait.
The most common of these processes is the \href{https://en.wikipedia.org/wiki/Brownian_motion}{Brownian Motion}.
This is used by default with the \texttt{make.traits} function:

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{## Creating the traits object}
\NormalTok{my_trait <-}\StringTok{ }\KeywordTok{make.traits}\NormalTok{()}
\end{Highlighting}
\end{Shaded}

This trait object can be simply printed (to see what's in it) or plotted (to see what the process looks like in the absence of a phylogeny):

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{## Which process is in here?}
\NormalTok{my_trait}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  ---- dads traits object ---- 
## 1 trait for 1 process (A) with one starting value (0).
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{## What does it look like?}
\KeywordTok{plot}\NormalTok{(my_trait)}
\end{Highlighting}
\end{Shaded}

\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-10-1.pdf}

By default, this trait is called ``A''.
This is not a really good name but you'll see more about specifying trait names later on.
If this is what the process should look like (theoretically) you can then add its \texttt{"traits"} object to our previous \texttt{dads} function to generate the tree and the traits:

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{## Simulate disparity and diversity}
\NormalTok{my_data <-}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ my_params,}
                \DataTypeTok{stop.rule =}\NormalTok{ my_stop_rule,}
                \DataTypeTok{traits    =}\NormalTok{ my_trait)}
\end{Highlighting}
\end{Shaded}

Et voilà! We now have a simple disparity and diversity simulation.
We can see what's in the results by simply printing it or plotting it:

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{## What's in there}
\NormalTok{my_data}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  ---- dads object ---- 
## Simulated diversity data (x$tree):
## 
## Phylogenetic tree with 20 tips and 19 internal nodes.
## 
## Tip labels:
##   t1, t2, t3, t4, t5, t6, ...
## Node labels:
##   n1, n2, n3, n4, n5, n6, ...
## 
## Rooted; includes branch lengths.
## 
## Simulated disparity data (x$data):
## 1 trait for 1 process (A) with one starting value (0).
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{## Plotting the disparity and diversity}
\KeywordTok{plot}\NormalTok{(my_data)}
\end{Highlighting}
\end{Shaded}

\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-12-1.pdf}

You can then extract the components you need for your specific analysis like so:

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{## Extracting the tree (a "phylo" object)}
\NormalTok{the_generated_tree <-}\StringTok{ }\NormalTok{my_data}\OperatorTok{$}\NormalTok{tree}
\KeywordTok{class}\NormalTok{(the_generated_tree)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## [1] "phylo"
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{## Extracting the data (a "matrix")}
\NormalTok{the_generated_data <-}\StringTok{ }\NormalTok{my_data}\OperatorTok{$}\NormalTok{data}
\KeywordTok{class}\NormalTok{(the_generated_data)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## [1] "matrix" "array"
\end{verbatim}

You can find much more on how to design trait objects in the \protect\hyperlink{maketraits}{\texttt{make.traits} section}.

\hypertarget{slightly-more-more-complex-simulating-linked-disparity-and-diversity}{%
\section{Slightly more more complex: simulating linked disparity and diversity}\label{slightly-more-more-complex-simulating-linked-disparity-and-diversity}}

The example above is also still pretty simple and easily done through a variety of \texttt{R} packages: here the trait and the tree are simulated at the same time but only the tree is simulating the trait (i.e.~the trait value at a tip is affected by it's ancestor and the branch length leading to it) but not the other way around (the trait value does not affect the tree).
It is possible to add this aspect using \texttt{"modifiers"} objects.
\texttt{"modifiers"} are similar to \texttt{"traits"} in that you specify what should go in there and then feed it to your simulation.

\texttt{"modifiers"} affect two key steps of the birth-death process: the calculation of the waiting time (i.e.~the component generating branch lengths) and the triggering of speciation or extinction events.
These events can be modified using some \texttt{condition} and \texttt{modify} function.
In other words, when reaching a certain condition specified by a \texttt{condition} function, the birth-death process will modify either the branch length or the speciation (or extinction) probability by applying a \texttt{modify} function.

You can use the function \texttt{make.modifiers} to design a specific \texttt{"modifiers"} object.
By default, this function generates a \texttt{"modifiers"} object that affects branch length and speciation in the following way:

\begin{itemize}
\tightlist
\item
  branch length is a randomly drawn number from an exponential distribution with a rate equal to the current number of taxa multiplied by the sum of the speciation and extinction rates.
\item
  speciation is triggered if a randomly drawn number (from a (0,1) uniform distribution) is smaller than the ratio between the speciation rate and the sum of the speciation and extinction rates. If that random number is greater, the lineage goes extinct.
\end{itemize}

Note that these are default for a birth death tree and were actually already applied in the examples before (without specifying a modifier):

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{## Make a default modifiers}
\NormalTok{default_modifiers <-}\StringTok{ }\KeywordTok{make.modifiers}\NormalTok{()}
\CommentTok{## What's in it?}
\NormalTok{default_modifiers}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  ---- dads modifiers object ---- 
## No modifiers applied to the branch length and speciation processes (default).
\end{verbatim}

This will not do much to our simulations compared to the previous trait and tree simulation but we can provide our modifiers object to the \texttt{dads} function:

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{set.seed}\NormalTok{(}\DecValTok{1}\NormalTok{)}
\CommentTok{## Setting the simulation parameters}
\NormalTok{extinction_}\DecValTok{02}\NormalTok{ <-}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{extinction =} \FloatTok{0.4}\NormalTok{)}
\NormalTok{living_}\DecValTok{20}\NormalTok{     <-}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{max.living =} \DecValTok{20}\NormalTok{)}
\NormalTok{BM_trait      <-}\StringTok{ }\KeywordTok{make.traits}\NormalTok{()}
\KeywordTok{set.seed}\NormalTok{(}\DecValTok{4242}\NormalTok{)}
\CommentTok{## Simulate disparity and diversity}
\NormalTok{default_data <-}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ extinction_}\DecValTok{02}\NormalTok{,}
                     \DataTypeTok{stop.rule =}\NormalTok{ living_}\DecValTok{20}\NormalTok{,}
                     \DataTypeTok{traits    =}\NormalTok{ BM_trait,}
                     \DataTypeTok{modifiers =}\NormalTok{ default_modifiers)}
\NormalTok{default_data}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  ---- dads object ---- 
## Birth death process with modifiers:
## No modifiers applied to the branch length and speciation processes (default).
## 
## Simulated diversity data (x$tree):
## 
## Phylogenetic tree with 45 tips and 44 internal nodes.
## 
## Tip labels:
##   t1, t2, t3, t4, t5, t6, ...
## Node labels:
##   n1, n2, n3, n4, n5, n6, ...
## 
## Rooted; includes branch lengths.
## 
## Simulated disparity data (x$data):
## 1 trait for 1 process (A) with one starting value (0).
\end{verbatim}

Note however that the printing information is now updated to state that you've add a modifier (even though it's a default one).

For more interesting simulations however, you can provide modifiers that actually modify the birth death process.
We can create one for example that makes species go extinct if their ancestor have a negative trait value.
For that we need to create a modifiers object that modifies the \texttt{speciation} process with a specific condition and a specific modification when that condition is met.
First we are going to create the modification function.
This function must intake the argument \texttt{x} and, in our case, return a logical value: \texttt{TRUE} is for speciate and \texttt{FALSE} is for go extinct.
We want this \texttt{modify} function to always return \texttt{FALSE} (go extinct) since it will be triggered by \texttt{condition} (we will see that in a minute).

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{## Going extinct}
\NormalTok{staying.alive <-}\StringTok{ }\ControlFlowTok{function}\NormalTok{(x) }\KeywordTok{return}\NormalTok{(}\OtherTok{TRUE}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

Now we want this function to only trigger when an ancestor has a negative trait value.
We can do that by specifying our \texttt{condition} function (when the ancestor is trait is negative) apply our modification (here the \texttt{staying.alive} function).
For that we can use the \texttt{parent.traits} utility function that is optimised for accessing traits in the birth death process (but you can of course write your own).
This function intakes the \texttt{trait.values} and \texttt{parent.lineage} arguments, two arguments that you can leave named as they are to facilitate \texttt{dads}s understanding of what you which to asses:

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{## Triggering a modification only if the ancestor trait is negative}
\NormalTok{negative.ancestor <-}\StringTok{ }\ControlFlowTok{function}\NormalTok{(trait.values, parent.lineage) \{}
    \KeywordTok{return}\NormalTok{(}\KeywordTok{all}\NormalTok{(}\KeywordTok{parent.traits}\NormalTok{(trait.values, parent.lineage) }\OperatorTok{<}\StringTok{ }\DecValTok{0}\NormalTok{))}
\NormalTok{\}}
\end{Highlighting}
\end{Shaded}

Note that we use the function \texttt{all} here to evaluate all traits (if the data is multidimensional).
We can then provide these two functions (the condition \texttt{negative.ancestor} and how to modify the speciation event when this condition is met \texttt{staying.alive}).
If you are an advances \texttt{dads} user, you can design your own \texttt{speciation} function but if you just want to use a normal \texttt{speciation} function, you can use the default one from \texttt{dads} called\ldots{} \texttt{speciation}.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{## Making a modifier for species to go extinct if}
\CommentTok{## their ancestor's trait value is (or are) negative}
\NormalTok{negatives_extinct <-}\StringTok{ }\KeywordTok{make.modifiers}\NormalTok{(}
                        \CommentTok{## The speciation function (default)}
                        \DataTypeTok{speciation =}\NormalTok{ speciation,}
                        \CommentTok{## What to modify}
                        \DataTypeTok{modify     =}\NormalTok{ staying.alive,}
                        \CommentTok{## When to modify it}
                        \DataTypeTok{condition  =}\NormalTok{ negative.ancestor)}
\CommentTok{## What's in it?}
\NormalTok{negatives_extinct}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  ---- dads modifiers object ---- 
## Default branch length process.
## Speciation process is set to speciation with a condition (negative.ancestor) and a modifier (staying.alive).
\end{verbatim}

Note that the \texttt{make.modifiers} function tests whether the input is compatible with \texttt{dads} by default so unless you have an error message, your \texttt{modifiers} will work!
We can now simulate our tree and traits with our modifier: species will go extinct if their ancestor have a negative trait value:

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{set.seed}\NormalTok{(}\DecValTok{4242}\NormalTok{)}
\CommentTok{## Simulate disparity and diversity}
\NormalTok{biased_data <-}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ extinction_}\DecValTok{02}\NormalTok{,}
                    \DataTypeTok{stop.rule =}\NormalTok{ living_}\DecValTok{20}\NormalTok{,}
                    \DataTypeTok{traits    =}\NormalTok{ BM_trait,}
                    \DataTypeTok{modifiers =}\NormalTok{ negatives_extinct)}
\NormalTok{biased_data}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  ---- dads object ---- 
## Birth death process with modifiers:
## Default branch length process.
## Speciation process is set to speciation with a condition (negative.ancestor) and a modifier (staying.alive).
## 
## Simulated diversity data (x$tree):
## 
## Phylogenetic tree with 38 tips and 37 internal nodes.
## 
## Tip labels:
##   t1, t2, t3, t4, t5, t6, ...
## Node labels:
##   n1, n2, n3, n4, n5, n6, ...
## 
## Rooted; includes branch lengths.
## 
## Simulated disparity data (x$data):
## 1 trait for 1 process (A) with one starting value (0).
\end{verbatim}

We can now compare the two trees and their trait values.
Note that we've used the same starting seed for both trees so the only thing differing between them is the modifier!
Also, although species

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow =} \KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{,}\DecValTok{2}\NormalTok{))}
\KeywordTok{plot}\NormalTok{(default_data, }\DataTypeTok{main =} \StringTok{"default results"}\NormalTok{)}
\KeywordTok{plot}\NormalTok{(biased_data, }\DataTypeTok{main =} \StringTok{"results with the modifier"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-20-1.pdf}

You can find much more how to design modifiers in the \protect\hyperlink{makemodifiers}{\texttt{make.modifiers} section}.

\hypertarget{maketraits}{%
\chapter{\texorpdfstring{Making trait processes with \texttt{make.traits()}}{Making trait processes with make.traits()}}\label{maketraits}}

\hypertarget{the-process-process}{%
\section{\texorpdfstring{The process (\texttt{process})}{The process (process)}}\label{the-process-process}}

The function \texttt{make.traits} allows you to design the process of a trait or a set of traits.
Here, the process of a trait designates the rules to generate the trait through time while simulating a phylogeny.
This process can depend on the previous state in the tree (i.e.~the trait of the ancestor) and the branch length to the descendant.
One classic example is the \href{https://en.wikipedia.org/wiki/Brownian_motion}{Brownian motion process (or Weiner process)}.
Note that it \emph{can} depend on both the ancestor and the branch length but does \emph{not necessary needs} (i.e.~the process can be only based on the previous state or only on branch length or on neither).

Trait processes in \texttt{dads} are functions that must always intake the following arguments by default.

\begin{itemize}
\tightlist
\item
  \texttt{x0}: the previous trait value(s)
\item
  \texttt{edge.length}: the branch length value
\item
  \texttt{...}: a placeholder for any extra arguments
\end{itemize}

For example, the following function would be a valid process (though not dependent on either the previous state nor the branch length):

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{## A valid (but useless) process}
\NormalTok{valid.process <-}\StringTok{ }\ControlFlowTok{function}\NormalTok{(x0, }\DataTypeTok{egde.length =} \DecValTok{1}\NormalTok{, ...) \{}
    \KeywordTok{return}\NormalTok{(}\DecValTok{42}\NormalTok{)}
\NormalTok{\}}
\end{Highlighting}
\end{Shaded}

\begin{quote}
Note that the argument \texttt{edge.length} is set to \texttt{1} by default. In general we highly recommend to set all arguments but \texttt{x0} to a default value (this really helps the speeding up the \texttt{dads} function).
\end{quote}

On the other hand, the following process (a unidimensional Brownian motion) is incorrect (it's missing \texttt{edge.length} and \texttt{...}):

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{## A wrongly formated process}
\NormalTok{invalid.process <-}\StringTok{ }\ControlFlowTok{function}\NormalTok{(x0) \{}
    \KeywordTok{return}\NormalTok{(}\KeywordTok{rnorm}\NormalTok{(}\DecValTok{1}\NormalTok{, }\DataTypeTok{mean =}\NormalTok{ x0))}
\NormalTok{\}}
\end{Highlighting}
\end{Shaded}

The \texttt{dads} package proposes inbuilt processes, namely a multidimensional Brownian motion (\texttt{BM.process}) or a a multidimensional Ornstein-Uhlenbeck process (\texttt{OU.process}).
You can find the list of implemented process by looking at the \texttt{?trait.process} manual page in \texttt{R}.

Once a process is chosen, you can feed it to the \texttt{make.traits} function:

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{## Creating a trait object}
\NormalTok{my_trait_object <-}\StringTok{ }\KeywordTok{make.traits}\NormalTok{(}\DataTypeTok{process =}\NormalTok{ BM.process)}
\end{Highlighting}
\end{Shaded}

This creates \texttt{"dads"} \texttt{"traits"} objects that you can print, and visualise using the \texttt{plot} function:

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{## The class of the object}
\KeywordTok{class}\NormalTok{(my_trait_object)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## [1] "dads"   "traits"
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{## What's in it?}
\NormalTok{my_trait_object}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  ---- dads traits object ---- 
## 1 trait for 1 process (A) with one starting value (0).
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{## What does the process looks like}
\KeywordTok{plot}\NormalTok{(my_trait_object)}
\end{Highlighting}
\end{Shaded}

\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-25-1.pdf}

Note that you can see the multiple options for plotting the trait process by looking at \texttt{?plot.dads} manual. Furthermore, you can look at what's actually in the object using:

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{## What's actually in that object?}
\KeywordTok{print.dads}\NormalTok{(my_trait_object, }\DataTypeTok{all =} \OtherTok{TRUE}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## $A
## $A$process
## function (x0, edge.length = 1, Sigma = diag(length(x0)), ...) 
## {
##     return(t(MASS::mvrnorm(n = 1, mu = x0, Sigma = Sigma * edge.length, 
##         ...)))
## }
## <bytecode: 0x7fd7168d5b80>
## <environment: namespace:dads>
## 
## $A$start
## [1] 0
## 
## $A$trait_id
## [1] 1
\end{verbatim}

As traits can get more and more complex, the automatic printing of its summary allows for a easier display of what's in the traits object.

Note that it is possible to make \texttt{"traits"} objects with multiple processes (that can be the same):

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{## 4 traits: two BM, one OU and one normal non process}
\NormalTok{four_traits <-}\StringTok{ }\KeywordTok{make.traits}\NormalTok{(}\DataTypeTok{process =} \KeywordTok{c}\NormalTok{(BM.process,}
\NormalTok{                                       BM.process,}
\NormalTok{                                       OU.process,}
\NormalTok{                                       no.process))}
\NormalTok{four_traits}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  ---- dads traits object ---- 
## 4 traits for 4 processes (A, B, C, D) with one starting value (0).
\end{verbatim}

You can visualise them individually using the \texttt{trait} argument in \texttt{plot.dads}:

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{## Plot options (4 plots in one window)}
\KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow =} \KeywordTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{,}\DecValTok{2}\NormalTok{))}
\KeywordTok{plot}\NormalTok{(four_traits, }\DataTypeTok{trait =} \DecValTok{1}\NormalTok{)}
\KeywordTok{plot}\NormalTok{(four_traits, }\DataTypeTok{trait =} \DecValTok{2}\NormalTok{)}
\KeywordTok{plot}\NormalTok{(four_traits, }\DataTypeTok{trait =} \DecValTok{3}\NormalTok{)}
\KeywordTok{plot}\NormalTok{(four_traits, }\DataTypeTok{trait =} \DecValTok{4}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-28-1.pdf}

\hypertarget{the-number-of-traits-n-and-the-starting-values-start}{%
\section{\texorpdfstring{The number of traits \texttt{n} and the starting values \texttt{start}}{The number of traits n and the starting values start}}\label{the-number-of-traits-n-and-the-starting-values-start}}

Two further important arguments are \texttt{n} the number of traits per process and \texttt{start} the starting values for all traits.
By default they are set to \texttt{n\ =\ 1} and \texttt{start\ =\ 0}.
This means that \texttt{make.traits} will assume that your processes are always unidimensional by default and that they always start with the value \texttt{0}.
It is however possible to change these values.

For example you can use the following to create a three dimensional Brownian motion with each dimensions starting with the value \texttt{1}:

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{## Multidimensional Brownian motion}
\KeywordTok{make.traits}\NormalTok{(BM.process, }\DataTypeTok{n =} \DecValTok{3}\NormalTok{, }\DataTypeTok{start =} \DecValTok{1}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  ---- dads traits object ---- 
## 3 traits for 1 process (A:3) with one starting value (1).
\end{verbatim}

Or the following with each dimensions starting with different values (respectively \texttt{1}, \texttt{2} and \texttt{3}):

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{## Multidimensional Brownian motion}
\KeywordTok{make.traits}\NormalTok{(BM.process, }\DataTypeTok{n =} \DecValTok{3}\NormalTok{, }\DataTypeTok{start =} \KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{,}\DecValTok{2}\NormalTok{,}\DecValTok{3}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  ---- dads traits object ---- 
## 3 traits for 1 process (A:3) with different starting values (1,2,3).
\end{verbatim}

Note that the number of traits are distributed per processes.
If the traits contains multiple process, the number of traits are distributed per processes:

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{## two 3D processes (BM and OU)}
\KeywordTok{make.traits}\NormalTok{(}\KeywordTok{c}\NormalTok{(BM.process, OU.process), }\DataTypeTok{n =} \DecValTok{3}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  ---- dads traits object ---- 
## 6 traits for 2 processes (A:3, B:3) with one starting value (0).
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{## one 1D processes (BM) and one 4D process (OU)}
\KeywordTok{make.traits}\NormalTok{(}\KeywordTok{c}\NormalTok{(BM.process, OU.process), }\DataTypeTok{n =} \KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{, }\DecValTok{4}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  ---- dads traits object ---- 
## 5 traits for 2 processes (A:1, B:4) with one starting value (0).
\end{verbatim}

And starting values are distributed for all the traits or for the traits one by one:

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{## two 3D processes (BM and OU) starting with 1}
\KeywordTok{make.traits}\NormalTok{(}\KeywordTok{c}\NormalTok{(BM.process, OU.process), }\DataTypeTok{n =} \DecValTok{3}\NormalTok{, }\DataTypeTok{start =} \DecValTok{1}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  ---- dads traits object ---- 
## 6 traits for 2 processes (A:3, B:3) with one starting value (1).
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{## two 3D processes (BM and OU) starting with values 1 to 6}
\KeywordTok{make.traits}\NormalTok{(}\KeywordTok{c}\NormalTok{(BM.process, OU.process), }\DataTypeTok{n =} \DecValTok{3}\NormalTok{, }\DataTypeTok{start =} \DecValTok{1}\OperatorTok{:}\DecValTok{6}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  ---- dads traits object ---- 
## 6 traits for 2 processes (A:3, B:3) with different starting values (1,2,3,4,5,6).
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{## two 3D processes (BM and OU) with the two first ones starting}
\CommentTok{## with 1 and the 4 other ones with the default (0)}
\KeywordTok{make.traits}\NormalTok{(}\KeywordTok{c}\NormalTok{(BM.process, OU.process), }\DataTypeTok{n =} \DecValTok{3}\NormalTok{, }\DataTypeTok{start =} \KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{,}\DecValTok{1}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Warning in make.traits(c(BM.process, OU.process), n = 3, start = c(1, 1)): Only
## the first 2 starting values were supplied for a required 6 traits. The missing
## start values are set to 0.
\end{verbatim}

\begin{verbatim}
##  ---- dads traits object ---- 
## 6 traits for 2 processes (A:3, B:3) with different starting values (1,1,0,0,0,0).
\end{verbatim}

\hypertarget{extra-argument-for-the-processes-with-process.args}{%
\section{\texorpdfstring{Extra argument for the processes with \texttt{process.args}}{Extra argument for the processes with process.args}}\label{extra-argument-for-the-processes-with-process.args}}

You can also feed extra arguments to your process(es) functions. For example, the inbuilt process \texttt{no.process} (that is just a number generator not based on the previous value \texttt{x0} or the branch length) can intake a specific random number generator as a function:

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{## no process trait using the normal distribution (default)}
\KeywordTok{make.traits}\NormalTok{(no.process, }\DataTypeTok{process.args =} \KeywordTok{list}\NormalTok{(}\DataTypeTok{fun =}\NormalTok{ rnorm))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  ---- dads traits object ---- 
## 1 trait for 1 process (A) with one starting value (0).
## process A uses the following extra argument: fun;
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{## no process trait using the uniform distribution}
\CommentTok{## bounded between 1 and 100}
\KeywordTok{make.traits}\NormalTok{(no.process, }\DataTypeTok{process.args =} \KeywordTok{list}\NormalTok{(}\DataTypeTok{fun =}\NormalTok{ runif, }\DataTypeTok{min =} \DecValTok{1}\NormalTok{, }\DataTypeTok{max =} \DecValTok{100}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  ---- dads traits object ---- 
## 1 trait for 1 process (A) with one starting value (0).
## process A uses the following extra arguments: fun,min,max;
\end{verbatim}

You can also add multiple extra arguments for multiple processes giving them as a list.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{## Two traits with no process:one normal and one uniform (1,100)}
\KeywordTok{make.traits}\NormalTok{(}\DataTypeTok{process      =} \KeywordTok{c}\NormalTok{(no.process, no.process),}
            \DataTypeTok{process.args =} \KeywordTok{list}\NormalTok{(}\KeywordTok{list}\NormalTok{(}\DataTypeTok{fun =}\NormalTok{ rnorm),}
                                \KeywordTok{list}\NormalTok{(}\DataTypeTok{fun =}\NormalTok{ runif, }\DataTypeTok{min =} \DecValTok{1}\NormalTok{, }\DataTypeTok{max =} \DecValTok{100}\NormalTok{)))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  ---- dads traits object ---- 
## 2 traits for 2 processes (A, B) with one starting value (0).
## process A uses the following extra argument: fun;
## process B uses the following extra arguments: fun,min,max;
\end{verbatim}

If one process do not need extra argument you must still give it and extra \texttt{NULL} process argument:

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{## Three traits with no process:}
\CommentTok{## one default, one lognormal and one uniform (1,100)}
\KeywordTok{make.traits}\NormalTok{(}\DataTypeTok{process      =} \KeywordTok{c}\NormalTok{(no.process, no.process, no.process),}
            \DataTypeTok{process.args =} \KeywordTok{list}\NormalTok{(}\CommentTok{## Extra arguments for the first process (none)}
                                \KeywordTok{list}\NormalTok{(}\OtherTok{NULL}\NormalTok{),}
                                \CommentTok{## Extra arguments for the second process}
                                \KeywordTok{list}\NormalTok{(}\DataTypeTok{fun =}\NormalTok{ rlnorm),}
                                \CommentTok{## Extra arguments for the third process}
                                \KeywordTok{list}\NormalTok{(}\DataTypeTok{fun =}\NormalTok{ runif, }\DataTypeTok{min =} \DecValTok{1}\NormalTok{, }\DataTypeTok{max =} \DecValTok{100}\NormalTok{)))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  ---- dads traits object ---- 
## 3 traits for 3 processes (A, B, C) with one starting value (0).
## process B uses the following extra argument: fun;
## process C uses the following extra arguments: fun,min,max;
\end{verbatim}

\hypertarget{naming-the-traits-with-trait.names}{%
\section{\texorpdfstring{Naming the traits with \texttt{trait.names}}{Naming the traits with trait.names}}\label{naming-the-traits-with-trait.names}}

As traits become more and more complex, it can be useful to give clearer names to each process.
This is easily done using the \texttt{trait.names} argument that attributes one name per process:

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{## A simple trait with a proper name}
\NormalTok{simple_trait <-}\StringTok{ }\KeywordTok{make.traits}\NormalTok{(}\DataTypeTok{trait.names =} \StringTok{"1D Brownian Motion"}\NormalTok{)}
\NormalTok{simple_trait}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  ---- dads traits object ---- 
## 1 trait for 1 process (1D Brownian Motion) with one starting value (0).
\end{verbatim}

This becomes more useful if we use the complex example above:

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{## Three named traits with no process:}
\CommentTok{## one default, one lognormal and one uniform (1,100)}
\KeywordTok{make.traits}\NormalTok{(}\DataTypeTok{process      =} \KeywordTok{c}\NormalTok{(no.process, no.process, no.process),}
            \DataTypeTok{process.args =} \KeywordTok{list}\NormalTok{(}\CommentTok{## Extra arguments for the first process (none)}
                                \KeywordTok{list}\NormalTok{(}\OtherTok{NULL}\NormalTok{),}
                                \CommentTok{## Extra arguments for the second process}
                                \KeywordTok{list}\NormalTok{(}\DataTypeTok{fun =}\NormalTok{ rlnorm),}
                                \CommentTok{## Extra arguments for the third process}
                                \KeywordTok{list}\NormalTok{(}\DataTypeTok{fun =}\NormalTok{ runif, }\DataTypeTok{min =} \DecValTok{1}\NormalTok{, }\DataTypeTok{max =} \DecValTok{100}\NormalTok{)),}
            \CommentTok{## Naming each trait}
            \DataTypeTok{trait.names  =} \KeywordTok{c}\NormalTok{(}\StringTok{"Normal"}\NormalTok{, }\StringTok{"LogNormal"}\NormalTok{, }\StringTok{"Uniform(1,100)"}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  ---- dads traits object ---- 
## 3 traits for 3 processes (Normal, LogNormal, Uniform(1,100)) with one starting value (0).
## process LogNormal uses the following extra argument: fun;
## process Uniform(1,100) uses the following extra arguments: fun,min,max;
\end{verbatim}

\hypertarget{combining-multiple-traits-with-add}{%
\section{\texorpdfstring{Combining multiple traits with \texttt{add}}{Combining multiple traits with add}}\label{combining-multiple-traits-with-add}}

You can also add traits to already existing trait objects using the simple \texttt{add} option.
This option just intakes a \texttt{"dads"} \texttt{"traits"} object and the additional process(es) will be added to it. For example:

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{## Creating on simple default Brownian motion}
\NormalTok{one_process <-}\StringTok{ }\KeywordTok{make.traits}\NormalTok{(}\DataTypeTok{trait.names =} \StringTok{"BM"}\NormalTok{)}

\CommentTok{## Creating a new trait (a 3D OU.process)}
\CommentTok{## and adding the previous one}
\NormalTok{two_processes <-}\StringTok{ }\KeywordTok{make.traits}\NormalTok{(OU.process, }\DataTypeTok{n =} \DecValTok{3}\NormalTok{, }\DataTypeTok{add =}\NormalTok{ one_process,}
                             \DataTypeTok{trait.names =} \StringTok{"3D OU"}\NormalTok{)}

\CommentTok{## Only one process}
\NormalTok{one_process}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  ---- dads traits object ---- 
## 1 trait for 1 process (BM) with one starting value (0).
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{## The two processes}
\NormalTok{two_processes}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  ---- dads traits object ---- 
## 4 traits for 2 processes (BM:1, 3D OU:3) with one starting value (0).
\end{verbatim}

\hypertarget{testing-the-traits-with-test}{%
\section{\texorpdfstring{Testing the traits with \texttt{test}}{Testing the traits with test}}\label{testing-the-traits-with-test}}

This bit is more for development.
We highly suggest leaving \texttt{test\ =\ TRUE} so that \texttt{make.traits} returns an error if a process or its additional arguments (\texttt{process.args}) are not formatted correctly.
\texttt{make.traits} will error if the trait cannot be directly passed to \texttt{dads}.
However, in some specific cases (again, probably mainly for development and debugging) it could be useful to skip the tests using \texttt{test\ =\ FALSE}.

\hypertarget{templates-for-making-your-very-own-process}{%
\section{Templates for making your very own process}\label{templates-for-making-your-very-own-process}}

As detailed above, any process of your own design will work as long as it is a function that takes at least the arguments \texttt{x0} and \texttt{edge.length}.
You can be imaginative and creative when designing your own process but here are two detailed example functions for a unidimensional Brownian Motion and Ornstein-Uhlenbeck process that you can use for a start (or not).
Remember it is good practice for \texttt{dads} processes to set all the arguments but \texttt{x0} with default values (just in case).
Also, note that the functions below are not equal to the already implemented \texttt{BM.process} and \texttt{OU.process} but are rather generalised/simplified version that you can use as a template

\hypertarget{a-simple-brownian-motion-process-template}{%
\subsection{A simple Brownian Motion process template}\label{a-simple-brownian-motion-process-template}}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{## A simple Brownian motion process}
\NormalTok{my.BM.process <-}\StringTok{ }\ControlFlowTok{function}\NormalTok{(x0, }\DataTypeTok{edge.length =} \DecValTok{1}\NormalTok{, }\DataTypeTok{sd =} \DecValTok{1}\NormalTok{, ...) \{}
    \CommentTok{## Drawing a random number from a normal distribution}
    \CommentTok{## with x0 as the and a given standard deviation}
    \CommentTok{## and depending on branch (edge) length}
\NormalTok{    result <-}\StringTok{ }\KeywordTok{rnorm}\NormalTok{(}\DataTypeTok{n =} \DecValTok{1}\NormalTok{, }\DataTypeTok{mean =}\NormalTok{ x0, }\DataTypeTok{sd =}\NormalTok{ sd }\OperatorTok{*}\StringTok{ }\NormalTok{edge.length)}

    \CommentTok{## Return the number}
    \KeywordTok{return}\NormalTok{(result)}
\NormalTok{\}}
\end{Highlighting}
\end{Shaded}

\hypertarget{a-simple-ornstein-uhlenbeck-process-template}{%
\subsection{A simple Ornstein-Uhlenbeck process template}\label{a-simple-ornstein-uhlenbeck-process-template}}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{## A simple Ornstein-Uhlenbeck motion process}
\NormalTok{my.OU.process <-}\StringTok{ }\ControlFlowTok{function}\NormalTok{(x0, }\DataTypeTok{edge.length =} \DecValTok{1}\NormalTok{, }\DataTypeTok{var =} \DecValTok{1}\NormalTok{, }\DataTypeTok{alpha =} \DecValTok{1}\NormalTok{, ...) \{}
    \CommentTok{## Calculate the mean based on alpha}
\NormalTok{    mean <-}\StringTok{ }\NormalTok{x0 }\OperatorTok{*}\StringTok{ }\KeywordTok{exp}\NormalTok{(}\OperatorTok{-}\NormalTok{alpha)}
    \CommentTok{## Calculate the standard deviation based on alpha and the variance}
\NormalTok{    sd <-}\StringTok{ }\KeywordTok{sqrt}\NormalTok{(var}\OperatorTok{/}\NormalTok{(}\DecValTok{2} \OperatorTok{*}\StringTok{ }\NormalTok{alpha) }\OperatorTok{*}\StringTok{ }\NormalTok{(}\DecValTok{1} \OperatorTok{-}\StringTok{ }\KeywordTok{exp}\NormalTok{(}\OperatorTok{-}\DecValTok{2} \OperatorTok{*}\StringTok{ }\NormalTok{alpha)))}
    \CommentTok{## Draw a random number from a normal distribution}
    \CommentTok{## using this mean and standard deviation}
    \CommentTok{## and depending on branch (edge) length}
\NormalTok{    result <-}\StringTok{ }\KeywordTok{rnorm}\NormalTok{(}\DataTypeTok{n =} \DecValTok{1}\NormalTok{, }\DataTypeTok{mean =}\NormalTok{ mean, }\DataTypeTok{sd =}\NormalTok{ sd }\OperatorTok{*}\StringTok{ }\NormalTok{branch.length)}

    \CommentTok{## Return the number}
    \KeywordTok{return}\NormalTok{(result)}
\NormalTok{\}}
\end{Highlighting}
\end{Shaded}

\hypertarget{makemodifiers}{%
\chapter{\texorpdfstring{Making modifiers with \texttt{make.modifiers()}}{Making modifiers with make.modifiers()}}\label{makemodifiers}}

\begin{quote}
TODO: just ideas so far
\end{quote}

\hypertarget{the-branch-length-function-branch.length}{%
\section{\texorpdfstring{The branch length function (\texttt{branch.length})}{The branch length function (branch.length)}}\label{the-branch-length-function-branch.length}}

A branch length function that is constant

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{constant.brlen <-}\StringTok{ }\ControlFlowTok{function}\NormalTok{(}\DataTypeTok{bd.params =} \OtherTok{NULL}\NormalTok{,}
                         \DataTypeTok{n.taxa =} \OtherTok{NULL}\NormalTok{,}
                         \DataTypeTok{parent.lineage =} \OtherTok{NULL}\NormalTok{,}
                         \DataTypeTok{trait.values =} \OtherTok{NULL}\NormalTok{,}
                         \DataTypeTok{modify.fun =} \OtherTok{NULL}\NormalTok{) \{}
    \KeywordTok{return}\NormalTok{(}\DecValTok{1}\NormalTok{)}
\NormalTok{\}}

\NormalTok{constant_brlen <-}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =} \KeywordTok{list}\NormalTok{(}\DataTypeTok{extinction =} \FloatTok{0.25}\NormalTok{),}
                     \DataTypeTok{stop.rule =} \KeywordTok{list}\NormalTok{(}\DataTypeTok{max.living =} \DecValTok{20}\NormalTok{),}
                     \DataTypeTok{traits    =} \KeywordTok{make.traits}\NormalTok{(),}
                     \DataTypeTok{modifiers =} \KeywordTok{make.modifiers}\NormalTok{(}
                        \DataTypeTok{branch.length =}\NormalTok{ constant.brlen))}

\KeywordTok{plot}\NormalTok{(constant_brlen, }\DataTypeTok{main =} \StringTok{"constant branch length"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\hypertarget{the-speciation-function-speciation}{%
\section{\texorpdfstring{The speciation function (\texttt{speciation})}{The speciation function (speciation)}}\label{the-speciation-function-speciation}}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{random.brlen <-}\StringTok{ }\ControlFlowTok{function}\NormalTok{(}\DataTypeTok{bd.params =} \OtherTok{NULL}\NormalTok{,}
                         \DataTypeTok{n.taxa =} \OtherTok{NULL}\NormalTok{,}
                         \DataTypeTok{parent.lineage =} \OtherTok{NULL}\NormalTok{,}
                         \DataTypeTok{trait.values =} \OtherTok{NULL}\NormalTok{,}
                         \DataTypeTok{modify.fun =} \OtherTok{NULL}\NormalTok{) \{}
    \KeywordTok{return}\NormalTok{(}\KeywordTok{runif}\NormalTok{(}\DecValTok{1}\NormalTok{))}
\NormalTok{\}}

\NormalTok{random.spec  <-}\StringTok{ }\ControlFlowTok{function}\NormalTok{(}\DataTypeTok{bd.params =} \OtherTok{NULL}\NormalTok{,}
                         \DataTypeTok{n.taxa =} \OtherTok{NULL}\NormalTok{,}
                         \DataTypeTok{parent.lineage =} \OtherTok{NULL}\NormalTok{,}
                         \DataTypeTok{trait.values =} \OtherTok{NULL}\NormalTok{,}
                         \DataTypeTok{modify.fun =} \OtherTok{NULL}\NormalTok{) \{}
    \KeywordTok{return}\NormalTok{(}\KeywordTok{sample}\NormalTok{(}\KeywordTok{c}\NormalTok{(}\OtherTok{TRUE}\NormalTok{, }\OtherTok{FALSE}\NormalTok{), }\DecValTok{1}\NormalTok{))}
\NormalTok{\}}

\KeywordTok{set.seed}\NormalTok{(}\DecValTok{1}\NormalTok{)}
\NormalTok{yule_tree <-}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =} \KeywordTok{list}\NormalTok{(}\DataTypeTok{extinction =} \DecValTok{0}\NormalTok{),}
                     \DataTypeTok{stop.rule =} \KeywordTok{list}\NormalTok{(}\DataTypeTok{max.living =} \DecValTok{20}\NormalTok{),}
                     \DataTypeTok{traits    =} \KeywordTok{make.traits}\NormalTok{(),}
                     \DataTypeTok{modifiers =} \KeywordTok{make.modifiers}\NormalTok{(}
                        \DataTypeTok{branch.length =}\NormalTok{ random.brlen,}
                        \DataTypeTok{speciation    =}\NormalTok{ random.spec))}

\KeywordTok{plot}\NormalTok{(yule_tree, }\DataTypeTok{main =} \StringTok{"a (failing) yule(ish) tree"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\hypertarget{the-condition-function-condition}{%
\section{\texorpdfstring{The condition function (\texttt{condition})}{The condition function (condition)}}\label{the-condition-function-condition}}

\hypertarget{the-modify-function-modify}{%
\section{\texorpdfstring{The modify function (\texttt{modify})}{The modify function (modify)}}\label{the-modify-function-modify}}

\hypertarget{combining-and-editing-modifiers-add}{%
\section{\texorpdfstring{Combining and editing modifiers (\texttt{add})}{Combining and editing modifiers (add)}}\label{combining-and-editing-modifiers-add}}

\hypertarget{testing-modifiers-test}{%
\section{\texorpdfstring{Testing modifiers (\texttt{test})}{Testing modifiers (test)}}\label{testing-modifiers-test}}

\hypertarget{demo-runnable}{%
\section{Demo runnable}\label{demo-runnable}}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{bd.params <-}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{speciation =} \DecValTok{1}\NormalTok{, }\DataTypeTok{extinction =} \DecValTok{1}\OperatorTok{/}\DecValTok{3}\NormalTok{)}
\NormalTok{traits <-}\StringTok{ }\KeywordTok{make.traits}\NormalTok{()}
\NormalTok{stop.rule <-}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{max.taxa =} \DecValTok{20}\NormalTok{, }\DataTypeTok{max.living =} \OtherTok{Inf}\NormalTok{, }\DataTypeTok{max.time =} \OtherTok{Inf}\NormalTok{)}
\NormalTok{modifiers <-}\StringTok{ }\OtherTok{NULL}
\NormalTok{events <-}\StringTok{ }\OtherTok{NULL}
\NormalTok{null.error <-}\StringTok{ }\OtherTok{NULL}

\CommentTok{## modifiers}
\NormalTok{condition <-}\StringTok{ }\ControlFlowTok{function}\NormalTok{(trait.values, parent.lineage) }\KeywordTok{return}\NormalTok{(}\KeywordTok{parent.traits}\NormalTok{(trait.values, parent.lineage) }\OperatorTok{<}\StringTok{ }\DecValTok{0}\NormalTok{)}
\NormalTok{modify <-}\StringTok{ }\ControlFlowTok{function}\NormalTok{(x) }\KeywordTok{return}\NormalTok{(x }\OperatorTok{*}\StringTok{ }\DecValTok{20}\NormalTok{)}

\CommentTok{## Setting up the different modifiers}
\NormalTok{modify_speciation <-}\StringTok{ }\KeywordTok{make.modifiers}\NormalTok{(}\DataTypeTok{speciation    =}\NormalTok{ speciation.trait,}
                                    \DataTypeTok{condition     =}\NormalTok{ condition,}
                                    \DataTypeTok{modify        =}\NormalTok{ modify)}
\NormalTok{modify_brlen <-}\StringTok{ }\KeywordTok{make.modifiers}\NormalTok{(}\DataTypeTok{branch.length =}\NormalTok{ branch.length.trait,}
                               \DataTypeTok{condition     =}\NormalTok{ condition,}
                               \DataTypeTok{modify        =}\NormalTok{ modify)}
\NormalTok{modify_speciation_brlen <-}\StringTok{ }\KeywordTok{make.modifiers}\NormalTok{(}\DataTypeTok{branch.length =}\NormalTok{ branch.length.trait,}
                                          \DataTypeTok{speciation    =}\NormalTok{ speciation.trait,}
                                          \DataTypeTok{condition     =}\NormalTok{ condition,}
                                          \DataTypeTok{modify        =}\NormalTok{ modify)}

\CommentTok{## Test normal (no modifiers)}
\KeywordTok{set.seed}\NormalTok{(}\DecValTok{1}\NormalTok{)}
\NormalTok{test <-}\StringTok{ }\KeywordTok{dads}\NormalTok{(bd.params, stop.rule, traits, }\DataTypeTok{null.error =} \DecValTok{20}\NormalTok{)}
\KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow =} \KeywordTok{c}\NormalTok{(}\DecValTok{4}\NormalTok{,}\DecValTok{2}\NormalTok{))}
\KeywordTok{plot}\NormalTok{(test}\OperatorTok{$}\NormalTok{tree)}
\KeywordTok{plot.dads}\NormalTok{(test, }\DataTypeTok{main =} \StringTok{"random tree + trait"}\NormalTok{)}

\CommentTok{## Test with modifiers}
\KeywordTok{set.seed}\NormalTok{(}\DecValTok{1}\NormalTok{)}
\NormalTok{trait_table <-}\StringTok{ }\OtherTok{NULL}
\NormalTok{test <-}\StringTok{ }\KeywordTok{dads}\NormalTok{(bd.params, stop.rule, traits,}
                                \DataTypeTok{modifiers =}\NormalTok{ modify_speciation,}
                                \DataTypeTok{null.error =} \DecValTok{20}\NormalTok{)}
\KeywordTok{plot}\NormalTok{(test}\OperatorTok{$}\NormalTok{tree)}
\KeywordTok{plot.dads}\NormalTok{(test, }\DataTypeTok{main =} \StringTok{"Skewed speciation"}\NormalTok{)}

\KeywordTok{set.seed}\NormalTok{(}\DecValTok{1}\NormalTok{)}
\NormalTok{trait_table <-}\StringTok{ }\OtherTok{NULL}
\NormalTok{test <-}\StringTok{ }\KeywordTok{dads}\NormalTok{(bd.params, stop.rule, traits,}
                                \DataTypeTok{modifiers =}\NormalTok{ modify_brlen,}
                                \DataTypeTok{null.error =} \DecValTok{20}\NormalTok{)}
\KeywordTok{plot}\NormalTok{(test}\OperatorTok{$}\NormalTok{tree)}
\KeywordTok{plot.dads}\NormalTok{(test, }\DataTypeTok{main =} \StringTok{"Skewed branch length"}\NormalTok{)}

\KeywordTok{set.seed}\NormalTok{(}\DecValTok{1}\NormalTok{)}
\NormalTok{trait_table <-}\StringTok{ }\OtherTok{NULL}
\NormalTok{test <-}\StringTok{ }\KeywordTok{dads}\NormalTok{(bd.params, stop.rule, traits,}
                                \DataTypeTok{modifiers =}\NormalTok{ modify_speciation_brlen,}
                                \DataTypeTok{null.error =} \DecValTok{20}\NormalTok{)}
\KeywordTok{plot}\NormalTok{(test}\OperatorTok{$}\NormalTok{tree)}
\KeywordTok{plot.dads}\NormalTok{(test, }\DataTypeTok{main =} \StringTok{"Skewed branch length and speciation"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-44-1.pdf}

\bibliography{../references.bib}

\end{document}

### 2020-12-01 inst/gitbook/01_getting_started.Rmd
typos + minors

@@
 ```{r}
-set.seed(1)
 ## Setting the simulation parameters

### 2020-12-01 inst/gitbook/03_make.modifiers.Rmd
Wrapped up branch.length example

@@
 
-> TODO: just ideas so far
+`"modifiers"` have a similar structure than `"traits"` where you can design an object with increasing complexity, starting with the simplest modifiers that doesn't modify anything (using the default arguments):
+
+```{r}
+## Making a default modifier (no modification)
+my_default_modifiers <- make.modifiers()
+my_default_modifiers
+```
+
+Similarly to `"traits"` objects, `"modifiers"` are also printed by default using `print.dads`.
+You can see details about what's actually in the object using `print.dads(my_default_modifiers, all = TRUE)`.
+However, contrary to `"traits"`, you cannot plot `"modifiers"`.
 
@@
 
-A branch length function that is constant
+The first argument in `"modifiers"` is the branch length function (`branch.length`) this is the function that will be executed in `dads` to generate branch length.
+Note that in the `dads` algorithm, branch length is not generated _directly_ but actually results of the waiting time.
+In other words, the `branch.length` function just affects waiting time for all taxa present at any time in the simulation.
+These taxa can then either go extinct (stopping the "growth" of it's branch length) or survive (continuing the "growth").
+
+By default, branch length (or waiting or growth) is a randomly drawn number from an exponential distribution with the rate of the number of taxa multiplied by the speciation and extinction rate: $n \times (\lambda + \mu)$ (where $n$ is the number of taxa currently present in the simulation, $\lambda$ and $\mu$ are respectively the speciation and extinction rates).
+This default function is simply called `branch.length` in `dads` and can be used as a modifier as follows:
+
+```{r}
+## Specifying the default modifier
+default_modifiers <- make.modifiers(branch.length = branch.length)
+
+## Setting some parameters for generating trees
+bd_params  <- list(extinction = 0)
+stope_rule <- list(max.living = 20)
+
+## Generating a tree with the default branch length parameter
+set.seed(0)
+default_tree <- dads(bd.params = bd_params,
+                     stop.rule = stope_rule,
+                     modifiers = default_modifiers)
+```
+
+Of course, the point of the modularity here is that you can provide your own function for generating branch length.
+For example, we might be interested in what our tree would look like if we'd use a simple constant branch length generation (instead of randomly drawing it from an exponential distribution).
+We can do so by declaring our own `branch.length` function and adding it to a `"modifiers"` object.
+
+```{r}
+## A constant branch length generator
+## (note that the output must be numeric, not integer)
+constant.brlen <- function() {
+    return(as.numeric(1))
+}
+
+## Creating the modifiers object
+constant_modifier <- make.modifiers(branch.length = constant.brlen)
+
+## Generating a new tree with this modifier
+set.seed(0)
+modified_tree <- dads(bd.params = bd_params,
+                      stop.rule = stope_rule,
+                      modifiers = constant_modifier)
+```
+
+And we can visualise the difference between both resulting trees:
+
+```{r}
+par(mfrow = c(1,2))
+plot(default_tree,  main = "Default modifier")
+plot(modified_tree, main = "Constant branch length\nmodifier")
+```
+
+It is of course to use more complex branch length modifiers that intakes different conditions and specific modification rather than simply always output a value of one.
+
+## The condition and modify functions (`condition` and `modify`)
+
+In the examples above, we have seen how to specify modifications to the birth death process (via `branch.length`, `selection` and `speciation`), however, one might note that these modifications are not dynamic.
+In other words, throughout the process, the modifications remain constant (even if they are conditional).
+It is however possible to code the modifiers
+
+In the following section we will see how to program these functions in a `"modifiers"` before learning about other aspects of the birth death process that can be modified (the `selection` and `speciation` steps).
+
+As shown in the example with the `branch.length` modifier, it is possible to modify a specific part of the birth death process (e.g.
+the change in time) but in the previous example, the `constant.brlen` modifier only included a constant modification of the process whereas it is probably more valuable to have a variable change in branch length (e.g.
+like the default `branch.length` modifier that is dependent on the number of taxa at the time of the simulation).
+It is possible add conditions and modifications to the `"modifiers"` object through the `condition` and `modify` function.
+As stated in their names, the `condition` function is indicating in which conditions to do a special action while the `modify` function is indicating what to modify (typically when a condition is met).
+
+`condition` and `modify` are hard coded in the `branch.length` function that they concern, i.e. they are variables (functions) within the function.
 

### 2020-12-02 inst/gitbook/01_getting_started.Rmd
Update 01_getting_started.Rmd

@@
 ## Setting the simulation parameters
-extinction_02 <- list(extinction = 0.4)
+extinction_02 <- list(extinction = 0.2)
 living_20     <- list(max.living = 20)
 BM_trait      <- make.traits()
-set.seed(4242)
+set.seed(1)
 ## Simulate disparity and diversity
@@
 ## Triggering a modification only if the ancestor trait is negative
-negative.ancestor <- function(trait.values, parent.lineage) {
-    return(all(parent.traits(trait.values, parent.lineage) < 0))
+negative.ancestor <- function(trait.values, lineage) {
+    return(all(parent.traits(trait.values, lineage) < 0))
 }
@@
 ```{r}
-set.seed(4242)
+set.seed(1)
 ## Simulate disparity and diversity

### 2020-12-02 inst/gitbook/03_make.modifiers.Rmd
Update 03_make.modifiers.Rmd

@@
 
+### The allowed arguments {#allowarguments}
+
+It is of course possible to design some more advanced function to interact with the birth death process.
+You can create a function for `branch.length`, `selection` and `speciation` that involve any of the following arguments:
+
+  * `bd.params`: a named list containing `"numeric"` values that contains the birth death parameters (at least `"speciation"` and `"extinction"`);
+  * `lineage`: a named list containing the lineage data (see below).
+  * `trait.values`: a `"matrix"` containing `"numeric"` values with the trait names as column names and the lineages ID as row numbers (you can use it with the function `parent.traits` to access the trait of the previous node for example).
+  * `modify.fun`: a `"list"` of named `"function"` (usually passed through `condition` and `modify`).
+
+The `lineage` list contains the following elements (missing elements are allowed):
+
+ * `lineage$parents`: an `"integer"` vector: the list of parent lineages;
+ * `lineage$livings`: an `"integer"` vector: the list of lineages still not extinct;
+ * `lineage$drawn`: a single `"integer"`: the ID of the selected lineage;
+ * `lineage$current`: a single `"integer"`: the selected lineage (is equal to `lineage$livings[lineage$drawn]`);
+ * `lineage$n`: a single `"integer"`: the current number of non extinct lineage (is equal to `length(lineage$livings))`;
+ * `lineage$split`: a `"logical"` vector: the list of splits for each lineage (`TRUE`), the number of total tips is equal to `sum(!lineage$split)`.
+
+In general, unless you know what you're doing, you can ignore most arguments for specific modifiers since they are handled automatically within the `dads` function.
+Therefore any argument can be left undeclared or missing and is always handled internally.
+For example, if you did not declare `n.taxa` as a function argument but are using `n.taxa` in the function, `n.taxa` will be detected and treated as a current argument automatically as set accordingly within the birth death process (e.g.
+`n.taxa` will be set to the current number of taxa every iteration of the process).
+
+For example, we can create a function that increases branch length proportional to the number of species "alive" at each time of the simulation in a discrete way.
+I.e.
+for discrete numbers of taxa, the branch length increases by jumps (ten fold):
+
+```{r}
+## A more complex binned.branch.length function
+increasing.brlen <- function(bd.params, lineage) {
+
+    ## Setting the cumulated birth and death
+    birth_death <- bd.params$speciation + bd.params$extinction
+
+    ## Returning branch lengths depending on different number of taxa
+    if(lineage$n <= 5) {
+        return(1    * rexp(1, sum(5 * birth_death)))
+    }
+    if(lineage$n <= 10) {
+        return(10   * rexp(1, sum(10 * birth_death)))   
+    }
+    if(lineage$n <= 15) {
+        return(100  * rexp(1, sum(15 * birth_death)))      
+    }
+    if(lineage$n <= 20) {
+        return(1000 * rexp(1, sum(20 * birth_death)))
+    } else {
+        return(1000 * rexp(1, sum(lineage$n * birth_death)))
+    }
+}
+```
+
+We can then create it as a `"modifiers"` object and run a new simulation:
+
+```{r}
+## Creating a modifiers
+increasing_modifier <- make.modifiers(branch.length = increasing.brlen)
+
+## Generating a new tree with this modifier
+set.seed(0)
+increasing_tree <- dads(bd.params = bd_params,
+                        stop.rule = stope_rule,
+                        modifiers = increasing_modifier)
+```
+
+And we can visualise the difference between the resulting trees:
+
+```{r}
+par(mfrow = c(1,3))
+plot(default_tree,    main = "Default modifier")
+plot(modified_tree,   main = "Constant branch length\nmodifier")
+plot(increasing_tree, main = "Increasing branch length\nmodifier (binned)")
+```
+
+## The selection function (`selection`)
+
+The `selection` function is used in the birth death process to know which lineage to select when running a speciation (or extinction!) event.
+By default, this function randomly selects one taxon that is currently not extinct (using: `sample(1:n.taxa, 1))`.
+Similarly as `branch.length` it is possible to modify this part of the birth death process.
+For example, we could simply select always the last created lineage (to create a "ladder" or most asymmetric tree):
+
+```{r}
+## Our function to always select the last taxon
+## (making sure it returns an integer)
+select.last <- function(lineage) {
+    return(as.integer(lineage$n))
+}
+```
+
+> Note that here the function can only intake the [allowed arguments as described above](#allowarguments) (here `n.taxa`: the number of current living taxa).
+
+We can then create a `"modifiers"` object the same way as before using this time the `selection` argument:
+
+```{r}
+## A modifier for selection
+ladderised_modifier <- make.modifiers(selection = select.last)
+## Generating a new tree with this modifier
+set.seed(0)
+ladderised_tree <- dads(bd.params = bd_params,
+                        stop.rule = stope_rule,
+                        modifiers = ladderised_modifier)
+## Displaying the results
+par(mfrow = c(1,2))
+plot(default_tree,    main = "Default modifier")
+plot(ladderised_tree, main = "Ladderising modifier")
+```
+
+Again, it is of course possible to make the modifier more complex and in combination with other elements of the tree.
+For example, we can create a `"dads"` object that also creates a trait a add to it a `selection` modifier that only selects for tips with positive trait values.
+
+```{r}
+## Our function that only select taxa with positive trait values
+select.positive <- function(trait.values, lineage) {
+    ## Selecting the taxa with positive traits only
+    positives <- as.integer(rownames(trait.values)[which(trait.values[, 1] >= 0)])
+
+    ## Select the current taxa that descend from a node with a positive value
+    positive_living <- cbind(lineage$parents, seq_along(lineage$split)
+                             )[which(lineage$parents %in% positives), 2]
+
+    ## Select one tip randomly in the ones with descendants with positive values
+    return(sample(which(lineage$livings %in% positive_living), 1))
+}
+
+## Creating the modifier
+positive_skew <- make.modifiers(selection = select.positive)
+
+## Creating a (default) trait object
+BM_trait <- make.traits()
+
+## Simulate a tree and trait with no modifier
+set.seed(1)
+default_dads <- dads(bd.params = bd_params,
+                     stop.rule = stope_rule,
+                     traits    = BM_trait)
+
+## Simulate a tree and trait with the modifier
+set.seed(1)
+skewed_trait_dads <- dads(bd.params = bd_params,
+                          stop.rule = stope_rule,
+                          traits    = BM_trait,
+                          modifiers = positive_skew)
+
+## Plotting the differences in trees and traits
+par(mfrow = c(1, 2))
+plot(default_dads, main = "Default trait and tree")
+plot(skewed_trait_dads, main = "Skewed trait and tree")
+```
+
+## The speciation function (`speciation`)
+
+The third function that can be used to modify the birth death process is the `speciation` function.
+This one is used during the birth death process to decide whether a lineages speciates (creating a node and two new lineages) or goes extinct (creating a tip).
+
+> Note that the `speciation` function only affects tips or nodes before the simulation reaches the `stop.rule`.
+The then surviving lineages are all automatically transformed into tips.
+
+By default, the `speciation` function is trigger a speciation even if a number randomly drawn from a uniform distribution is lower than the ratio between the speciation and the speciation and extinction parameter.
+If the randomly drawn number is higher, the lineage goes extinct.
+
+```{r, eval = FALSE}
+## The speciation in pseudo-code:
+runif(1) < speciation/ (speciation + extinction)
+```
+
+Creating a `"modifiers"` with a `speciation` function works the same way as for `branch.length` and `selection` but the function that will be used needs to output a logical value (see table [below](#summarymodifiers)).
+Once the function is created simply input your function for speciation in the modifier and run the `dads` function with that modifier:
+
+```{r, eval = FALSE}
+## Speciating or going extinct randomly
+## (regardless of the extinction parameter)
+random.extinct  <- function() {
+    return(sample(c(TRUE, FALSE), 1))
+}
+
+## Creating the modifiers object
+random_extinction <- make.modifiers(speciation = random.extinct)
+
+## Generating a new tree with this modifier
+set.seed(0)
+modified_tree <- dads(bd.params = bd_params,
+                      stop.rule = stope_rule,
+                      modifiers = random_extinction)
+
+par(mfrow = c(1,2))
+plot(default_tree,  main = "Default modifier")
+plot(modified_tree, main = "Random extinction\nmodifier")
+```
+
+Note how every lineage end up going extinct!
+And again, we can make some more advanced modifiers: for example, one where a tip always goes extinct if their ancestor has a negative trait value (here we will also introduce the utility function `parent.trait` that automatically selects the trait values of the parent of the current lineage.
+
+```{r}
+## A modifier for removing tips with negative values
+bias.trait <- function(trait.values, lineage) {
+    if(parent.traits(trait.values, lineage) < 0) {
+        ## Go extinct!
+        return(FALSE)
+    } else {
+        ## Speciate
+        return(TRUE)
+    }
+}
+
+## Creating the modifier
+biased_trait <- make.modifiers(speciation = bias.trait)
+
+## Simulate a tree and trait with the modifier
+set.seed(1)
+biased_trait_dads <- dads(bd.params = bd_params,
+                          stop.rule = stope_rule,
+                          traits    = BM_trait,
+                          modifiers = biased_trait)
+
+## Plotting the differences in trees and traits
+par(mfrow = c(1, 2))
+plot(default_dads, main = "Default trait and tree")
+plot(biased_trait_dads, main = "Biased trait and tree")
+```
+
+## Summary of the inputs and outputs for the `branch.length`, `selection` and `speciation` modifiers {#summarymodifiers}
+
+modifier name   | accepted input (arguments) | required output (class) |
+----------------|----------------------------|-------------------------|
+`branch.length` | `bd.params`, `lineage`, `trait.values` | `"numeric"` |
+`selection`     | `bd.params`, `lineage`, `trait.values` | `"integer"` |
+`speciation`    | `bd.params`, `lineage`, `trait.values` | `"logical"` |
+
 ## The condition and modify functions (`condition` and `modify`)
@@
 In other words, throughout the process, the modifications remain constant (even if they are conditional).
-It is however possible to code the modifiers
+It is however possible to code the `"modifiers"` so that they can be affected by `"events"` objects (see next chapter).
 
-In the following section we will see how to program these functions in a `"modifiers"` before learning about other aspects of the birth death process that can be modified (the `selection` and `speciation` steps).
-
-As shown in the example with the `branch.length` modifier, it is possible to modify a specific part of the birth death process (e.g.
-the change in time) but in the previous example, the `constant.brlen` modifier only included a constant modification of the process whereas it is probably more valuable to have a variable change in branch length (e.g.
-like the default `branch.length` modifier that is dependent on the number of taxa at the time of the simulation).
-It is possible add conditions and modifications to the `"modifiers"` object through the `condition` and `modify` function.
-As stated in their names, the `condition` function is indicating in which conditions to do a special action while the `modify` function is indicating what to modify (typically when a condition is met).
-
+To do so, you can formally declare conditions (`condition`) and modifications (`modify`) as internal functions that can then be modified my an `"events"` object.
 `condition` and `modify` are hard coded in the `branch.length` function that they concern, i.e. they are variables (functions) within the function.
-
-```{r, eval = FALSE}
-constant.brlen <- function(bd.params      = NULL,
-                           n.taxa         = NULL,
-                           parent.lineage = NULL,
-                           trait.values   = NULL,
-                           modify.fun     = NULL) {
-    return(1)
-}
-
-constant_brlen <- dads(bd.params = list(extinction = 0.25),
-                     stop.rule = list(max.living = 20),
-                     traits    = make.traits(),
-                     modifiers = make.modifiers(
-                        branch.length = constant.brlen))
-
-plot(constant_brlen, main = "constant branch length")
-```
-
-## The speciation function (`speciation`)
-
-```{r, eval = FALSE}
-random.brlen <- function(bd.params = NULL,
-                         n.taxa = NULL,
-                         parent.lineage = NULL,
-                         trait.values = NULL,
-                         modify.fun = NULL) {
-    return(runif(1))
-}
-
-random.spec  <- function(bd.params = NULL,
-                         n.taxa = NULL,
-                         parent.lineage = NULL,
-                         trait.values = NULL,
-                         modify.fun = NULL) {
-    return(sample(c(TRUE, FALSE), 1))
-}
-
-set.seed(1)
-yule_tree <- dads(bd.params = list(extinction = 0),
-                     stop.rule = list(max.living = 20),
-                     traits    = make.traits(),
-                     modifiers = make.modifiers(
-                        branch.length = random.brlen,
-                        speciation    = random.spec))
-
-plot(yule_tree, main = "a (failing) yule(ish) tree")
-```
-
-## The condition function (`condition`)
-
-## The modify function (`modify`)
 

### 2020-12-04 inst/gitbook/03_make.modifiers.Rmd
Updated manual (running)

@@
 
-```{r, fig.height = 16, fig.width = 8}
+```{r, fig.height = 16, fig.width = 8, eval = FALSE}
 bd.params <- list(speciation = 1, extinction = 1/3)

### 2020-12-04 inst/gitbook/_book/dads_manual.tex
Updated manual (running)

@@
 \author{Thomas Guillerme (\href{mailto:guillert@tcd.ie}{\nolinkurl{guillert@tcd.ie}}), Alex Slavenko (email) and others}
-\date{2020-11-13}
+\date{2020-12-04}
 
@@
 }
-\hypertarget{dads}{%
-\chapter{\texorpdfstring{\texttt{dads}}{dads}}\label{dads}}
-
-Disparity and diversity simulations.
+\hypertarget{dads-disparity-and-diversity-simulations.}{%
+\chapter{\texorpdfstring{\texttt{dads}: disparity and diversity simulations.}{dads: disparity and diversity simulations.}}\label{dads-disparity-and-diversity-simulations.}}
 
@@
 ##  ---- dads modifiers object ---- 
-## No modifiers applied to the branch length and speciation processes (default).
+## No modifiers applied to the branch length, selection and speciation processes (default).
 \end{verbatim}
@@
 \begin{Highlighting}[]
-\KeywordTok{set.seed}\NormalTok{(}\DecValTok{1}\NormalTok{)}
 \CommentTok{## Setting the simulation parameters}
-\NormalTok{extinction_}\DecValTok{02}\NormalTok{ <-}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{extinction =} \FloatTok{0.4}\NormalTok{)}
+\NormalTok{extinction_}\DecValTok{02}\NormalTok{ <-}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{extinction =} \FloatTok{0.2}\NormalTok{)}
 \NormalTok{living_}\DecValTok{20}\NormalTok{     <-}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{max.living =} \DecValTok{20}\NormalTok{)}
 \NormalTok{BM_trait      <-}\StringTok{ }\KeywordTok{make.traits}\NormalTok{()}
-\KeywordTok{set.seed}\NormalTok{(}\DecValTok{4242}\NormalTok{)}
+\KeywordTok{set.seed}\NormalTok{(}\DecValTok{1}\NormalTok{)}
 \CommentTok{## Simulate disparity and diversity}
@@
 ## Birth death process with modifiers:
-## No modifiers applied to the branch length and speciation processes (default).
+## No modifiers applied to the branch length, selection and speciation processes (default).
 ## 
@@
 ## 
-## Phylogenetic tree with 45 tips and 44 internal nodes.
+## Phylogenetic tree with 24 tips and 23 internal nodes.
 ## 
@@
 \CommentTok{## Triggering a modification only if the ancestor trait is negative}
-\NormalTok{negative.ancestor <-}\StringTok{ }\ControlFlowTok{function}\NormalTok{(trait.values, parent.lineage) \{}
-    \KeywordTok{return}\NormalTok{(}\KeywordTok{all}\NormalTok{(}\KeywordTok{parent.traits}\NormalTok{(trait.values, parent.lineage) }\OperatorTok{<}\StringTok{ }\DecValTok{0}\NormalTok{))}
+\NormalTok{negative.ancestor <-}\StringTok{ }\ControlFlowTok{function}\NormalTok{(trait.values, lineage) \{}
+    \KeywordTok{return}\NormalTok{(}\KeywordTok{all}\NormalTok{(}\KeywordTok{parent.traits}\NormalTok{(trait.values, lineage) }\OperatorTok{<}\StringTok{ }\DecValTok{0}\NormalTok{))}
 \NormalTok{\}}
@@
 ## Default branch length process.
+## Default selection process.
 ## Speciation process is set to speciation with a condition (negative.ancestor) and a modifier (staying.alive).
@@
 \begin{Highlighting}[]
-\KeywordTok{set.seed}\NormalTok{(}\DecValTok{4242}\NormalTok{)}
+\KeywordTok{set.seed}\NormalTok{(}\DecValTok{1}\NormalTok{)}
 \CommentTok{## Simulate disparity and diversity}
@@
 ## 
-## Phylogenetic tree with 38 tips and 37 internal nodes.
+## Phylogenetic tree with 22 tips and 21 internal nodes.
 ## 
@@
 ## }
-## <bytecode: 0x7fd7168d5b80>
+## <bytecode: 0x7f7eeee57f58>
 ## <environment: namespace:dads>
@@
 
-\begin{quote}
-TODO: just ideas so far
-\end{quote}
+\texttt{"modifiers"} have a similar structure than \texttt{"traits"} where you can design an object with increasing complexity, starting with the simplest modifiers that doesn't modify anything (using the default arguments):
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{## Making a default modifier (no modification)}
+\NormalTok{my_default_modifiers <-}\StringTok{ }\KeywordTok{make.modifiers}\NormalTok{()}
+\NormalTok{my_default_modifiers}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+##  ---- dads modifiers object ---- 
+## No modifiers applied to the branch length, selection and speciation processes (default).
+\end{verbatim}
+
+Similarly to \texttt{"traits"} objects, \texttt{"modifiers"} are also printed by default using \texttt{print.dads}.
+You can see details about what's actually in the object using \texttt{print.dads(my\_default\_modifiers,\ all\ =\ TRUE)}.
+However, contrary to \texttt{"traits"}, you cannot plot \texttt{"modifiers"}.
 
@@
 
-A branch length function that is constant
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{constant.brlen <-}\StringTok{ }\ControlFlowTok{function}\NormalTok{(}\DataTypeTok{bd.params =} \OtherTok{NULL}\NormalTok{,}
-                         \DataTypeTok{n.taxa =} \OtherTok{NULL}\NormalTok{,}
-                         \DataTypeTok{parent.lineage =} \OtherTok{NULL}\NormalTok{,}
-                         \DataTypeTok{trait.values =} \OtherTok{NULL}\NormalTok{,}
-                         \DataTypeTok{modify.fun =} \OtherTok{NULL}\NormalTok{) \{}
-    \KeywordTok{return}\NormalTok{(}\DecValTok{1}\NormalTok{)}
+The first argument in \texttt{"modifiers"} is the branch length function (\texttt{branch.length}) this is the function that will be executed in \texttt{dads} to generate branch length.
+Note that in the \texttt{dads} algorithm, branch length is not generated \emph{directly} but actually results of the waiting time.
+In other words, the \texttt{branch.length} function just affects waiting time for all taxa present at any time in the simulation.
+These taxa can then either go extinct (stopping the ``growth'' of it's branch length) or survive (continuing the ``growth'').
+
+By default, branch length (or waiting or growth) is a randomly drawn number from an exponential distribution with the rate of the number of taxa multiplied by the speciation and extinction rate: \(n \times (\lambda + \mu)\) (where \(n\) is the number of taxa currently present in the simulation, \(\lambda\) and \(\mu\) are respectively the speciation and extinction rates).
+This default function is simply called \texttt{branch.length} in \texttt{dads} and can be used as a modifier as follows:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{## Specifying the default modifier}
+\NormalTok{default_modifiers <-}\StringTok{ }\KeywordTok{make.modifiers}\NormalTok{(}\DataTypeTok{branch.length =}\NormalTok{ branch.length)}
+
+\CommentTok{## Setting some parameters for generating trees}
+\NormalTok{bd_params  <-}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{extinction =} \DecValTok{0}\NormalTok{)}
+\NormalTok{stope_rule <-}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{max.living =} \DecValTok{20}\NormalTok{)}
+
+\CommentTok{## Generating a tree with the default branch length parameter}
+\KeywordTok{set.seed}\NormalTok{(}\DecValTok{0}\NormalTok{)}
+\NormalTok{default_tree <-}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ bd_params,}
+                     \DataTypeTok{stop.rule =}\NormalTok{ stope_rule,}
+                     \DataTypeTok{modifiers =}\NormalTok{ default_modifiers)}
+\end{Highlighting}
+\end{Shaded}
+
+Of course, the point of the modularity here is that you can provide your own function for generating branch length.
+For example, we might be interested in what our tree would look like if we'd use a simple constant branch length generation (instead of randomly drawing it from an exponential distribution).
+We can do so by declaring our own \texttt{branch.length} function and adding it to a \texttt{"modifiers"} object.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{## A constant branch length generator}
+\CommentTok{## (note that the output must be numeric, not integer)}
+\NormalTok{constant.brlen <-}\StringTok{ }\ControlFlowTok{function}\NormalTok{() \{}
+    \KeywordTok{return}\NormalTok{(}\KeywordTok{as.numeric}\NormalTok{(}\DecValTok{1}\NormalTok{))}
 \NormalTok{\}}
 
-\NormalTok{constant_brlen <-}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =} \KeywordTok{list}\NormalTok{(}\DataTypeTok{extinction =} \FloatTok{0.25}\NormalTok{),}
-                     \DataTypeTok{stop.rule =} \KeywordTok{list}\NormalTok{(}\DataTypeTok{max.living =} \DecValTok{20}\NormalTok{),}
-                     \DataTypeTok{traits    =} \KeywordTok{make.traits}\NormalTok{(),}
-                     \DataTypeTok{modifiers =} \KeywordTok{make.modifiers}\NormalTok{(}
-                        \DataTypeTok{branch.length =}\NormalTok{ constant.brlen))}
-
-\KeywordTok{plot}\NormalTok{(constant_brlen, }\DataTypeTok{main =} \StringTok{"constant branch length"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
+\CommentTok{## Creating the modifiers object}
+\NormalTok{constant_modifier <-}\StringTok{ }\KeywordTok{make.modifiers}\NormalTok{(}\DataTypeTok{branch.length =}\NormalTok{ constant.brlen)}
+
+\CommentTok{## Generating a new tree with this modifier}
+\KeywordTok{set.seed}\NormalTok{(}\DecValTok{0}\NormalTok{)}
+\NormalTok{modified_tree <-}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ bd_params,}
+                      \DataTypeTok{stop.rule =}\NormalTok{ stope_rule,}
+                      \DataTypeTok{modifiers =}\NormalTok{ constant_modifier)}
+\end{Highlighting}
+\end{Shaded}
+
+And we can visualise the difference between both resulting trees:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow =} \KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{,}\DecValTok{2}\NormalTok{))}
+\KeywordTok{plot}\NormalTok{(default_tree,  }\DataTypeTok{main =} \StringTok{"Default modifier"}\NormalTok{)}
+\KeywordTok{plot}\NormalTok{(modified_tree, }\DataTypeTok{main =} \StringTok{"Constant branch length}\CharTok{\textbackslash{}n}\StringTok{modifier"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-45-1.pdf}
+
+It is of course to use more complex branch length modifiers that intakes different conditions and specific modification rather than simply always output a value of one.
+
+\hypertarget{allowarguments}{%
+\subsection{The allowed arguments}\label{allowarguments}}
+
+It is of course possible to design some more advanced function to interact with the birth death process.
+You can create a function for \texttt{branch.length}, \texttt{selection} and \texttt{speciation} that involve any of the following arguments:
+
+\begin{itemize}
+\tightlist
+\item
+  \texttt{bd.params}: a named list containing \texttt{"numeric"} values that contains the birth death parameters (at least \texttt{"speciation"} and \texttt{"extinction"});
+\item
+  \texttt{lineage}: a named list containing the lineage data (see below).
+\item
+  \texttt{trait.values}: a \texttt{"matrix"} containing \texttt{"numeric"} values with the trait names as column names and the lineages ID as row numbers (you can use it with the function \texttt{parent.traits} to access the trait of the previous node for example).
+\item
+  \texttt{modify.fun}: a \texttt{"list"} of named \texttt{"function"} (usually passed through \texttt{condition} and \texttt{modify}).
+\end{itemize}
+
+The \texttt{lineage} list contains the following elements (missing elements are allowed):
+
+\begin{itemize}
+\tightlist
+\item
+  \texttt{lineage\$parents}: an \texttt{"integer"} vector: the list of parent lineages;
+\item
+  \texttt{lineage\$livings}: an \texttt{"integer"} vector: the list of lineages still not extinct;
+\item
+  \texttt{lineage\$drawn}: a single \texttt{"integer"}: the ID of the selected lineage;
+\item
+  \texttt{lineage\$current}: a single \texttt{"integer"}: the selected lineage (is equal to \texttt{lineage\$livings{[}lineage\$drawn{]}});
+\item
+  \texttt{lineage\$n}: a single \texttt{"integer"}: the current number of non extinct lineage (is equal to \texttt{length(lineage\$livings))};
+\item
+  \texttt{lineage\$split}: a \texttt{"logical"} vector: the list of splits for each lineage (\texttt{TRUE}), the number of total tips is equal to \texttt{sum(!lineage\$split)}.
+\end{itemize}
+
+In general, unless you know what you're doing, you can ignore most arguments for specific modifiers since they are handled automatically within the \texttt{dads} function.
+Therefore any argument can be left undeclared or missing and is always handled internally.
+For example, if you did not declare \texttt{n.taxa} as a function argument but are using \texttt{n.taxa} in the function, \texttt{n.taxa} will be detected and treated as a current argument automatically as set accordingly within the birth death process (e.g.
+\texttt{n.taxa} will be set to the current number of taxa every iteration of the process).
+
+For example, we can create a function that increases branch length proportional to the number of species ``alive'' at each time of the simulation in a discrete way.
+I.e.
+for discrete numbers of taxa, the branch length increases by jumps (ten fold):
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{## A more complex binned.branch.length function}
+\NormalTok{increasing.brlen <-}\StringTok{ }\ControlFlowTok{function}\NormalTok{(bd.params, lineage) \{}
+
+    \CommentTok{## Setting the cumulated birth and death}
+\NormalTok{    birth_death <-}\StringTok{ }\NormalTok{bd.params}\OperatorTok{$}\NormalTok{speciation }\OperatorTok{+}\StringTok{ }\NormalTok{bd.params}\OperatorTok{$}\NormalTok{extinction}
+
+    \CommentTok{## Returning branch lengths depending on different number of taxa}
+    \ControlFlowTok{if}\NormalTok{(lineage}\OperatorTok{$}\NormalTok{n }\OperatorTok{<=}\StringTok{ }\DecValTok{5}\NormalTok{) \{}
+        \KeywordTok{return}\NormalTok{(}\DecValTok{1}    \OperatorTok{*}\StringTok{ }\KeywordTok{rexp}\NormalTok{(}\DecValTok{1}\NormalTok{, }\KeywordTok{sum}\NormalTok{(}\DecValTok{5} \OperatorTok{*}\StringTok{ }\NormalTok{birth_death)))}
+\NormalTok{    \}}
+    \ControlFlowTok{if}\NormalTok{(lineage}\OperatorTok{$}\NormalTok{n }\OperatorTok{<=}\StringTok{ }\DecValTok{10}\NormalTok{) \{}
+        \KeywordTok{return}\NormalTok{(}\DecValTok{10}   \OperatorTok{*}\StringTok{ }\KeywordTok{rexp}\NormalTok{(}\DecValTok{1}\NormalTok{, }\KeywordTok{sum}\NormalTok{(}\DecValTok{10} \OperatorTok{*}\StringTok{ }\NormalTok{birth_death)))   }
+\NormalTok{    \}}
+    \ControlFlowTok{if}\NormalTok{(lineage}\OperatorTok{$}\NormalTok{n }\OperatorTok{<=}\StringTok{ }\DecValTok{15}\NormalTok{) \{}
+        \KeywordTok{return}\NormalTok{(}\DecValTok{100}  \OperatorTok{*}\StringTok{ }\KeywordTok{rexp}\NormalTok{(}\DecValTok{1}\NormalTok{, }\KeywordTok{sum}\NormalTok{(}\DecValTok{15} \OperatorTok{*}\StringTok{ }\NormalTok{birth_death)))      }
+\NormalTok{    \}}
+    \ControlFlowTok{if}\NormalTok{(lineage}\OperatorTok{$}\NormalTok{n }\OperatorTok{<=}\StringTok{ }\DecValTok{20}\NormalTok{) \{}
+        \KeywordTok{return}\NormalTok{(}\DecValTok{1000} \OperatorTok{*}\StringTok{ }\KeywordTok{rexp}\NormalTok{(}\DecValTok{1}\NormalTok{, }\KeywordTok{sum}\NormalTok{(}\DecValTok{20} \OperatorTok{*}\StringTok{ }\NormalTok{birth_death)))}
+\NormalTok{    \} }\ControlFlowTok{else}\NormalTok{ \{}
+        \KeywordTok{return}\NormalTok{(}\DecValTok{1000} \OperatorTok{*}\StringTok{ }\KeywordTok{rexp}\NormalTok{(}\DecValTok{1}\NormalTok{, }\KeywordTok{sum}\NormalTok{(lineage}\OperatorTok{$}\NormalTok{n }\OperatorTok{*}\StringTok{ }\NormalTok{birth_death)))}
+\NormalTok{    \}}
+\NormalTok{\}}
+\end{Highlighting}
+\end{Shaded}
+
+We can then create it as a \texttt{"modifiers"} object and run a new simulation:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{## Creating a modifiers}
+\NormalTok{increasing_modifier <-}\StringTok{ }\KeywordTok{make.modifiers}\NormalTok{(}\DataTypeTok{branch.length =}\NormalTok{ increasing.brlen)}
+
+\CommentTok{## Generating a new tree with this modifier}
+\KeywordTok{set.seed}\NormalTok{(}\DecValTok{0}\NormalTok{)}
+\NormalTok{increasing_tree <-}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ bd_params,}
+                        \DataTypeTok{stop.rule =}\NormalTok{ stope_rule,}
+                        \DataTypeTok{modifiers =}\NormalTok{ increasing_modifier)}
+\end{Highlighting}
+\end{Shaded}
+
+And we can visualise the difference between the resulting trees:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow =} \KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{,}\DecValTok{3}\NormalTok{))}
+\KeywordTok{plot}\NormalTok{(default_tree,    }\DataTypeTok{main =} \StringTok{"Default modifier"}\NormalTok{)}
+\KeywordTok{plot}\NormalTok{(modified_tree,   }\DataTypeTok{main =} \StringTok{"Constant branch length}\CharTok{\textbackslash{}n}\StringTok{modifier"}\NormalTok{)}
+\KeywordTok{plot}\NormalTok{(increasing_tree, }\DataTypeTok{main =} \StringTok{"Increasing branch length}\CharTok{\textbackslash{}n}\StringTok{modifier (binned)"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-48-1.pdf}
+
+\hypertarget{the-selection-function-selection}{%
+\section{\texorpdfstring{The selection function (\texttt{selection})}{The selection function (selection)}}\label{the-selection-function-selection}}
+
+The \texttt{selection} function is used in the birth death process to know which lineage to select when running a speciation (or extinction!) event.
+By default, this function randomly selects one taxon that is currently not extinct (using: \texttt{sample(1:n.taxa,\ 1))}.
+Similarly as \texttt{branch.length} it is possible to modify this part of the birth death process.
+For example, we could simply select always the last created lineage (to create a ``ladder'' or most asymmetric tree):
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{## Our function to always select the last taxon}
+\CommentTok{## (making sure it returns an integer)}
+\NormalTok{select.last <-}\StringTok{ }\ControlFlowTok{function}\NormalTok{(lineage) \{}
+    \KeywordTok{return}\NormalTok{(}\KeywordTok{as.integer}\NormalTok{(lineage}\OperatorTok{$}\NormalTok{n))}
+\NormalTok{\}}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{quote}
+Note that here the function can only intake the \protect\hyperlink{allowarguments}{allowed arguments as described above} (here \texttt{n.taxa}: the number of current living taxa).
+\end{quote}
+
+We can then create a \texttt{"modifiers"} object the same way as before using this time the \texttt{selection} argument:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{## A modifier for selection}
+\NormalTok{ladderised_modifier <-}\StringTok{ }\KeywordTok{make.modifiers}\NormalTok{(}\DataTypeTok{selection =}\NormalTok{ select.last)}
+\CommentTok{## Generating a new tree with this modifier}
+\KeywordTok{set.seed}\NormalTok{(}\DecValTok{0}\NormalTok{)}
+\NormalTok{ladderised_tree <-}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ bd_params,}
+                        \DataTypeTok{stop.rule =}\NormalTok{ stope_rule,}
+                        \DataTypeTok{modifiers =}\NormalTok{ ladderised_modifier)}
+\CommentTok{## Displaying the results}
+\KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow =} \KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{,}\DecValTok{2}\NormalTok{))}
+\KeywordTok{plot}\NormalTok{(default_tree,    }\DataTypeTok{main =} \StringTok{"Default modifier"}\NormalTok{)}
+\KeywordTok{plot}\NormalTok{(ladderised_tree, }\DataTypeTok{main =} \StringTok{"Ladderising modifier"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-50-1.pdf}
+
+Again, it is of course possible to make the modifier more complex and in combination with other elements of the tree.
+For example, we can create a \texttt{"dads"} object that also creates a trait a add to it a \texttt{selection} modifier that only selects for tips with positive trait values.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{## Our function that only select taxa with positive trait values}
+\NormalTok{select.positive <-}\StringTok{ }\ControlFlowTok{function}\NormalTok{(trait.values, lineage) \{}
+    \CommentTok{## Selecting the taxa with positive traits only}
+\NormalTok{    positives <-}\StringTok{ }\KeywordTok{as.integer}\NormalTok{(}\KeywordTok{rownames}\NormalTok{(trait.values)[}\KeywordTok{which}\NormalTok{(trait.values[, }\DecValTok{1}\NormalTok{] }\OperatorTok{>=}\StringTok{ }\DecValTok{0}\NormalTok{)])}
+
+    \CommentTok{## Select the current taxa that descend from a node with a positive value}
+\NormalTok{    positive_living <-}\StringTok{ }\KeywordTok{cbind}\NormalTok{(lineage}\OperatorTok{$}\NormalTok{parents, }\KeywordTok{seq_along}\NormalTok{(lineage}\OperatorTok{$}\NormalTok{split)}
+\NormalTok{                             )[}\KeywordTok{which}\NormalTok{(lineage}\OperatorTok{$}\NormalTok{parents }\OperatorTok{%in%}\StringTok{ }\NormalTok{positives), }\DecValTok{2}\NormalTok{]}
+
+    \CommentTok{## Select one tip randomly in the ones with descendants with positive values}
+    \KeywordTok{return}\NormalTok{(}\KeywordTok{sample}\NormalTok{(}\KeywordTok{which}\NormalTok{(lineage}\OperatorTok{$}\NormalTok{livings }\OperatorTok{%in%}\StringTok{ }\NormalTok{positive_living), }\DecValTok{1}\NormalTok{))}
+\NormalTok{\}}
+
+\CommentTok{## Creating the modifier}
+\NormalTok{positive_skew <-}\StringTok{ }\KeywordTok{make.modifiers}\NormalTok{(}\DataTypeTok{selection =}\NormalTok{ select.positive)}
+
+\CommentTok{## Creating a (default) trait object}
+\NormalTok{BM_trait <-}\StringTok{ }\KeywordTok{make.traits}\NormalTok{()}
+
+\CommentTok{## Simulate a tree and trait with no modifier}
+\KeywordTok{set.seed}\NormalTok{(}\DecValTok{1}\NormalTok{)}
+\NormalTok{default_dads <-}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ bd_params,}
+                     \DataTypeTok{stop.rule =}\NormalTok{ stope_rule,}
+                     \DataTypeTok{traits    =}\NormalTok{ BM_trait)}
+
+\CommentTok{## Simulate a tree and trait with the modifier}
+\KeywordTok{set.seed}\NormalTok{(}\DecValTok{1}\NormalTok{)}
+\NormalTok{skewed_trait_dads <-}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ bd_params,}
+                          \DataTypeTok{stop.rule =}\NormalTok{ stope_rule,}
+                          \DataTypeTok{traits    =}\NormalTok{ BM_trait,}
+                          \DataTypeTok{modifiers =}\NormalTok{ positive_skew)}
+
+\CommentTok{## Plotting the differences in trees and traits}
+\KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow =} \KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{, }\DecValTok{2}\NormalTok{))}
+\KeywordTok{plot}\NormalTok{(default_dads, }\DataTypeTok{main =} \StringTok{"Default trait and tree"}\NormalTok{)}
+\KeywordTok{plot}\NormalTok{(skewed_trait_dads, }\DataTypeTok{main =} \StringTok{"Skewed trait and tree"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-51-1.pdf}
 
@@
 
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{random.brlen <-}\StringTok{ }\ControlFlowTok{function}\NormalTok{(}\DataTypeTok{bd.params =} \OtherTok{NULL}\NormalTok{,}
-                         \DataTypeTok{n.taxa =} \OtherTok{NULL}\NormalTok{,}
-                         \DataTypeTok{parent.lineage =} \OtherTok{NULL}\NormalTok{,}
-                         \DataTypeTok{trait.values =} \OtherTok{NULL}\NormalTok{,}
-                         \DataTypeTok{modify.fun =} \OtherTok{NULL}\NormalTok{) \{}
-    \KeywordTok{return}\NormalTok{(}\KeywordTok{runif}\NormalTok{(}\DecValTok{1}\NormalTok{))}
-\NormalTok{\}}
-
-\NormalTok{random.spec  <-}\StringTok{ }\ControlFlowTok{function}\NormalTok{(}\DataTypeTok{bd.params =} \OtherTok{NULL}\NormalTok{,}
-                         \DataTypeTok{n.taxa =} \OtherTok{NULL}\NormalTok{,}
-                         \DataTypeTok{parent.lineage =} \OtherTok{NULL}\NormalTok{,}
-                         \DataTypeTok{trait.values =} \OtherTok{NULL}\NormalTok{,}
-                         \DataTypeTok{modify.fun =} \OtherTok{NULL}\NormalTok{) \{}
+The third function that can be used to modify the birth death process is the \texttt{speciation} function.
+This one is used during the birth death process to decide whether a lineages speciates (creating a node and two new lineages) or goes extinct (creating a tip).
+
+\begin{quote}
+Note that the \texttt{speciation} function only affects tips or nodes before the simulation reaches the \texttt{stop.rule}.
+The then surviving lineages are all automatically transformed into tips.
+\end{quote}
+
+By default, the \texttt{speciation} function is trigger a speciation even if a number randomly drawn from a uniform distribution is lower than the ratio between the speciation and the speciation and extinction parameter.
+If the randomly drawn number is higher, the lineage goes extinct.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{## The speciation in pseudo-code:}
+\KeywordTok{runif}\NormalTok{(}\DecValTok{1}\NormalTok{) }\OperatorTok{<}\StringTok{ }\NormalTok{speciation}\OperatorTok{/}\StringTok{ }\NormalTok{(speciation }\OperatorTok{+}\StringTok{ }\NormalTok{extinction)}
+\end{Highlighting}
+\end{Shaded}
+
+Creating a \texttt{"modifiers"} with a \texttt{speciation} function works the same way as for \texttt{branch.length} and \texttt{selection} but the function that will be used needs to output a logical value (see table \protect\hyperlink{summarymodifiers}{below}).
+Once the function is created simply input your function for speciation in the modifier and run the \texttt{dads} function with that modifier:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{## Speciating or going extinct randomly}
+\CommentTok{## (regardless of the extinction parameter)}
+\NormalTok{random.extinct  <-}\StringTok{ }\ControlFlowTok{function}\NormalTok{() \{}
     \KeywordTok{return}\NormalTok{(}\KeywordTok{sample}\NormalTok{(}\KeywordTok{c}\NormalTok{(}\OtherTok{TRUE}\NormalTok{, }\OtherTok{FALSE}\NormalTok{), }\DecValTok{1}\NormalTok{))}
@@
 
+\CommentTok{## Creating the modifiers object}
+\NormalTok{random_extinction <-}\StringTok{ }\KeywordTok{make.modifiers}\NormalTok{(}\DataTypeTok{speciation =}\NormalTok{ random.extinct)}
+
+\CommentTok{## Generating a new tree with this modifier}
+\KeywordTok{set.seed}\NormalTok{(}\DecValTok{0}\NormalTok{)}
+\NormalTok{modified_tree <-}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ bd_params,}
+                      \DataTypeTok{stop.rule =}\NormalTok{ stope_rule,}
+                      \DataTypeTok{modifiers =}\NormalTok{ random_extinction)}
+
+\KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow =} \KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{,}\DecValTok{2}\NormalTok{))}
+\KeywordTok{plot}\NormalTok{(default_tree,  }\DataTypeTok{main =} \StringTok{"Default modifier"}\NormalTok{)}
+\KeywordTok{plot}\NormalTok{(modified_tree, }\DataTypeTok{main =} \StringTok{"Random extinction}\CharTok{\textbackslash{}n}\StringTok{modifier"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+Note how every lineage end up going extinct!
+And again, we can make some more advanced modifiers: for example, one where a tip always goes extinct if their ancestor has a negative trait value (here we will also introduce the utility function \texttt{parent.trait} that automatically selects the trait values of the parent of the current lineage.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{## A modifier for removing tips with negative values}
+\NormalTok{bias.trait <-}\StringTok{ }\ControlFlowTok{function}\NormalTok{(trait.values, lineage) \{}
+    \ControlFlowTok{if}\NormalTok{(}\KeywordTok{parent.traits}\NormalTok{(trait.values, lineage) }\OperatorTok{<}\StringTok{ }\DecValTok{0}\NormalTok{) \{}
+        \CommentTok{## Go extinct!}
+        \KeywordTok{return}\NormalTok{(}\OtherTok{FALSE}\NormalTok{)}
+\NormalTok{    \} }\ControlFlowTok{else}\NormalTok{ \{}
+        \CommentTok{## Speciate}
+        \KeywordTok{return}\NormalTok{(}\OtherTok{TRUE}\NormalTok{)}
+\NormalTok{    \}}
+\NormalTok{\}}
+
+\CommentTok{## Creating the modifier}
+\NormalTok{biased_trait <-}\StringTok{ }\KeywordTok{make.modifiers}\NormalTok{(}\DataTypeTok{speciation =}\NormalTok{ bias.trait)}
+
+\CommentTok{## Simulate a tree and trait with the modifier}
 \KeywordTok{set.seed}\NormalTok{(}\DecValTok{1}\NormalTok{)}
-\NormalTok{yule_tree <-}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =} \KeywordTok{list}\NormalTok{(}\DataTypeTok{extinction =} \DecValTok{0}\NormalTok{),}
-                     \DataTypeTok{stop.rule =} \KeywordTok{list}\NormalTok{(}\DataTypeTok{max.living =} \DecValTok{20}\NormalTok{),}
-                     \DataTypeTok{traits    =} \KeywordTok{make.traits}\NormalTok{(),}
-                     \DataTypeTok{modifiers =} \KeywordTok{make.modifiers}\NormalTok{(}
-                        \DataTypeTok{branch.length =}\NormalTok{ random.brlen,}
-                        \DataTypeTok{speciation    =}\NormalTok{ random.spec))}
-
-\KeywordTok{plot}\NormalTok{(yule_tree, }\DataTypeTok{main =} \StringTok{"a (failing) yule(ish) tree"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\hypertarget{the-condition-function-condition}{%
-\section{\texorpdfstring{The condition function (\texttt{condition})}{The condition function (condition)}}\label{the-condition-function-condition}}
-
-\hypertarget{the-modify-function-modify}{%
-\section{\texorpdfstring{The modify function (\texttt{modify})}{The modify function (modify)}}\label{the-modify-function-modify}}
+\NormalTok{biased_trait_dads <-}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ bd_params,}
+                          \DataTypeTok{stop.rule =}\NormalTok{ stope_rule,}
+                          \DataTypeTok{traits    =}\NormalTok{ BM_trait,}
+                          \DataTypeTok{modifiers =}\NormalTok{ biased_trait)}
+
+\CommentTok{## Plotting the differences in trees and traits}
+\KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow =} \KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{, }\DecValTok{2}\NormalTok{))}
+\KeywordTok{plot}\NormalTok{(default_dads, }\DataTypeTok{main =} \StringTok{"Default trait and tree"}\NormalTok{)}
+\KeywordTok{plot}\NormalTok{(biased_trait_dads, }\DataTypeTok{main =} \StringTok{"Biased trait and tree"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-54-1.pdf}
+
+\hypertarget{summarymodifiers}{%
+\section{\texorpdfstring{Summary of the inputs and outputs for the \texttt{branch.length}, \texttt{selection} and \texttt{speciation} modifiers}{Summary of the inputs and outputs for the branch.length, selection and speciation modifiers}}\label{summarymodifiers}}
+
+\begin{longtable}[]{@{}lll@{}}
+\toprule
+modifier name & accepted input (arguments) & required output (class)\tabularnewline
+\midrule
+\endhead
+\texttt{branch.length} & \texttt{bd.params}, \texttt{lineage}, \texttt{trait.values} & \texttt{"numeric"}\tabularnewline
+\texttt{selection} & \texttt{bd.params}, \texttt{lineage}, \texttt{trait.values} & \texttt{"integer"}\tabularnewline
+\texttt{speciation} & \texttt{bd.params}, \texttt{lineage}, \texttt{trait.values} & \texttt{"logical"}\tabularnewline
+\bottomrule
+\end{longtable}
+
+\hypertarget{the-condition-and-modify-functions-condition-and-modify}{%
+\section{\texorpdfstring{The condition and modify functions (\texttt{condition} and \texttt{modify})}{The condition and modify functions (condition and modify)}}\label{the-condition-and-modify-functions-condition-and-modify}}
+
+In the examples above, we have seen how to specify modifications to the birth death process (via \texttt{branch.length}, \texttt{selection} and \texttt{speciation}), however, one might note that these modifications are not dynamic.
+In other words, throughout the process, the modifications remain constant (even if they are conditional).
+It is however possible to code the \texttt{"modifiers"} so that they can be affected by \texttt{"events"} objects (see next chapter).
+
+To do so, you can formally declare conditions (\texttt{condition}) and modifications (\texttt{modify}) as internal functions that can then be modified my an \texttt{"events"} object.
+\texttt{condition} and \texttt{modify} are hard coded in the \texttt{branch.length} function that they concern, i.e.~they are variables (functions) within the function.
 
@@
 
-\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-44-1.pdf}
-
 \bibliography{../references.bib}

### 2020-12-09 inst/gitbook/03_make.modifiers.Rmd
typos

@@
 bd_params <- list(extinction = 0)
-stope_rule <- list(max.living = 20)
+stop_rule <- list(max.living = 20)
 
@@
 default_tree <- dads(bd.params = bd_params,
-                     stop.rule = stope_rule,
+                     stop.rule = stop_rule,
                      modifiers = default_modifiers)
@@
 modified_tree <- dads(bd.params = bd_params,
-                      stop.rule = stope_rule,
+                      stop.rule = stop_rule,
                       modifiers = constant_modifier)
@@
 increasing_tree <- dads(bd.params = bd_params,
-                        stop.rule = stope_rule,
+                        stop.rule = stop_rule,
                         modifiers = increasing_modifier)
@@
 ladderised_tree <- dads(bd.params = bd_params,
-                        stop.rule = stope_rule,
+                        stop.rule = stop_rule,
                         modifiers = ladderised_modifier)
@@
 default_dads <- dads(bd.params = bd_params,
-                     stop.rule = stope_rule,
+                     stop.rule = stop_rule,
                      traits    = BM_trait)
@@
 skewed_trait_dads <- dads(bd.params = bd_params,
-                          stop.rule = stope_rule,
+                          stop.rule = stop_rule,
                           traits    = BM_trait,
@@
 modified_tree <- dads(bd.params = bd_params,
-                      stop.rule = stope_rule,
+                      stop.rule = stop_rule,
                       modifiers = random_extinction)
@@
 biased_trait_dads <- dads(bd.params = bd_params,
-                          stop.rule = stope_rule,
+                          stop.rule = stop_rule,
                           traits    = BM_trait,

### 2020-12-09 inst/gitbook/index.Rmd
typos + minors
typos

@@
 
-# `dads`
-
-Disparity and diversity simulations.
+# `dads`: disparity and diversity simulations.
 
@@
 ## Installing the latest version directly from GitHub
-install_github("TGuillerme/dads")
+devtools::install_github("TGuillerme/dads")
 ```
@@
 ```{r, eval = FALSE, echo = FALSE, message = FALSE}
-## loading the package and setting up the start seed.
+## Loading the package and setting up the start seed.
 library(dads)
-## Setting a random seed for repeatability
-set.seed(123)
 ```

### 2020-12-17 tests/testthat/test-birth.death.R
UTFs

@@ -293,7 +293,7 @@ extinction.trait <- function(bd.params, lineage, trait.values) {
     ## Select the descendants that'll go extinct
     extinct <- which(lineage$parents %in% selected_nodes)
 
-    ## Update the lineage object
+    ## Update the lineage object
     lineage$livings <- lineage$livings[!lineage$livings %in% extinct]
     lineage$n       <- length(lineage$livings)
     return(lineage)
@@ -364,7 +364,7 @@ change.death.param <- function(bd.params, lineage, trait.values) {
 
     ## Reducing speciation after reaching time t
 
-## Time condition
+## Time condition
 time.condition <- function(bd.params, lineage, trait.values, time) {
     return(time > 2)
 }
@@ -472,9 +472,28 @@ change.trait.correlation <- function(traits, bd.params, lineage, trait.values) {
     expect_lt(cor(test2$data[, 1], test2$data[, 2]), 1)
 
 
-    ## modifiers events
-    ## Adding a speciation condition after reaching time t
+#     ## modifiers events
+#     ## Adding a speciation condition after reaching time t
+#     stop.rule$max.time <- 6
+# time.condition <- function(bd.params, lineage, trait.values, time) {
+#     return(time > 4)
+# }
     
+#     ## A default modifier
+#     modifiers <- make.modifiers()
+
+# change.speciation.condition <- function(modifiers, bd.params, lineage, trait.values) {
+#     return(make.modifiers(update = modifiers,
+#                           speciation = speciation,
+#                           condition = trait.condition,
+#                           modify = trait.modify))
+# }
+
+
+
+
+
+
 
     ## Adding a branch length condition when reaching n taxa
 

### 2020-12-17 tests/testthat/test-make.modifiers.R

@@ -176,3 +176,42 @@ test_that("make.modifiers works", {
     ## Print modifiers works correctly
 
 })
+
+
+test_that("make.modifiers(update) works", {
+
+    ## Sanitizing
+    error <- capture_error(make.modifiers(update = "bob"))
+    expect_equal(error[[1]], "update must be of class dads or modifiers.")
+    test <- make.modifiers()
+    error <- capture_error(make.modifiers(update = test, branch.length = "what"))
+    expect_equal(error[[1]], "function for branch length is not a function.")
+    error <- capture_error(make.modifiers(update = test, add = test))
+    expect_equal(error[[1]], "Impossible to add and update a modifiers object at the same time.")
+    
+    # ## Nothing happens
+    test2 <- make.modifiers(update = test)
+    expect_equal(capture_output(print(test)), capture_output(print(test2)))
+
+    ## Speciation is updated
+    test1 <- make.modifiers()
+    test2 <- make.modifiers(update = test1, speciation = speciation)
+    # ## They are different processes
+    # expect_true(capture_output(print(test1[[1]]$process)) != capture_output(print(test2[[1]]$process)))
+
+    # test1 <- make.traits(process = list(BM.process, OU.process))
+    # test2 <- make.traits(process = BM.process, trait.names = "B", update = test1)
+    # ## They are different processes for test1
+    # expect_true(capture_output(print(test1[[1]]$process)) != capture_output(print(test1[[2]]$process)))
+    # ## But the same for test2
+    # expect_true(capture_output(print(test2[[1]]$process)) == capture_output(print(test2[[2]]$process)))
+
+    # my_correlations <- matrix(1, ncol = 3, nrow = 3)
+    # test <- make.traits(n = 3, start = c(0, 1, 3),
+    #                     process.args = list(Sigma = my_correlations))
+    # my_correlation2 <- matrix(2, ncol = 3, nrow = 3) 
+    # test2 <- make.traits(update = test,
+    #                      process.args = list(Sigma = my_correlation2))
+    # expect_equal(unique(c(test[[1]]$Sigma)), 1)
+    # expect_equal(unique(c(test2[[1]]$Sigma)), 2)
+})

### 2021-01-08 inst/gitbook/05_thorough_example.Rmd
drafted the demo story

---
title: "Thorough example"
author: "Thomas Guillerme (guillert@tcd.ie)"
date: "`r Sys.Date()`"
bibliography: [../References.bib, ../packages.bib]
output:
  html_document: default
  pdf_document: default
---

# Thorough example

Here we are going to look at an example on how to simulate complex evolutionary histories with `dads`.

Let say we want to simulate the following story:

 1. A lineage evolves in a normal birth-death fashion (with extinction) from a single origin and with two traits called:
    - "Distance": the physical distance between species in an arbitrary 1D space (evolving in a Brownian motion).
    - "Variable": a 2D uncorrelated Ornstein-Uhlenbeck trait that corresponds to some traits the lineages have.
 During the whole process, species that are more distant from their ancestor have a greater change of speciating.

Three events then happen during the evolutionary history of this lineage:

 2. "Correlation event": after reaching a certain number of taxa an event happens making the "Variable" trait dimensions become correlated.
 3. "Selective extinction event": after reaching a certain point in time, lineages with their "Distance" values in some interval goes extinct.
 4. "Founding event": after reaching a value for the "Distance" trait, a founding event happens with a new birth-death process with no extinction and where the "Variable" is now a 2D correlated Brownian Motion.
 5. "Mass extinction event": after reaching a certain number of taxa in the "Founding event", 80% of taxa go extinct randomly.

We are going to run each step one by one in a cumulative manner to see how they pile up:

## Step 1:

### The birth death parameters

First let's set the overall birth death parameters (a speciation of 1.2 and an extinction of 0.3):

```{r}
## The birth death parameters
bd_params <- list(speciation = 1.2, extinction = 0.3)
## The stopping rule
stop_rule <- list(max.time = 4)
```

And let's quickly test if the simulation works by running a birth death tree without traits and anything:

```{r}
## Running a birth death tree 
results <- dads(bd.params = bd_params,
                stop.rule = stop_rule,
                null.error = 100) # This option will try the requested options up until a 100 times

## And displaying the results
plot(results)
axisPhylo()
```

### The traits

Then let's create the traits:

```{r}
## The OU 2D trait correlation
correlation <- matrix(0, 2, 2)

## The trait object
two_traits <- make.traits(process = c(BM.process, OU.process), n = c(1, 2), 
                          process.args = list(list(NULL), list(Sigma = correlation)),
                          trait.names = c("Distance", "Variables"))
#TODO: fix OU with correlation
```

Note that specifying the correlation matrix is not necessary here _per se_ (since no correlation is the default) but we will modify it through an event later on so it's good practice to have it properly stated before hand (to avoid any confusion).

We can then run the tree with the traits:

```{r}
## Running a birth death tree 
results <- dads(bd.params  = bd_params,
                stop.rule  = stop_rule,
                traits     = two_traits,
                null.error = 100)

## And displaying the results
par(mfrow = c(1,2))
plot(results, trait = 1, main = "Trait 1: Distance")
plot(results, trait = c(2,3), main = "Variables correlation")

## TODO: debug correlation!
```

### The modifiers

```{r}
## Make a distance modifier
distance_speciation <- make.modifiers(speciation = NULL,
                                      condition = NULL,
                                      modify = NULL) 
```

```{r}
## Running a birth death tree 
results <- dads(bd.params  = bd_params,
                stop.rule  = stop_rule,
                traits     = two_traits,
                modifiers  = distance_speciation,
                null.error = 100)

## And displaying the results
par(mfrow = c(1,2))
plot(results, trait = 1, main = "Trait 1: Distance")
plot(results, trait = c(2,3), main = "Variables correlation")
```

## Step 2:
 2. "Correlation event": after reaching a certain number of taxa an event happens making the "Variable" trait dimensions become correlated.

```{r}
## The correlation change event
events_list <- make.events(event.name   = "Correlation",
                           target       = "traits",
                           condition    = taxa.condition(30),
                           modification = update.traits(
                    process.args = list(Sigma = matrix(c(0.7, 0, 0, 0.7), 2, 2)),
                    trait.names  = "Variables"))
#TODO: fix bug here
```

```{r}
## Running a birth death tree 
results <- dads(bd.params  = bd_params,
                stop.rule  = stop_rule,
                traits     = two_traits,
                modifiers  = distance_speciation,
                events     = events_list,
                null.error = 100)

## And displaying the results
par(mfrow = c(1,2))
plot(results, trait = 1, main = "Trait 1: Distance")
plot(results, trait = c(2,3), main = "Variables correlation")
## TODO: debug correlation!
```

## Step 3:
 3. "Selective extinction event": after reaching a certain point in time, lineages with their "Distance" values in some interval goes extinct.

```{r}
## Adding the selective event
events_list <- make.events(event.name   = "Selection",
                           add          = events_list,
                           target       = "taxa",
                           condition    = time.condition(4),
                           modification = trait.extinction(1, condition = `>`),
                           event.name   = "Selection")
```

```{r}
## Running a birth death tree 
results <- dads(bd.params  = bd_params,
                stop.rule  = stop_rule,
                traits     = two_traits,
                modifiers  = distance_speciation,
                events     = events_list,
                null.error = 100)

## And displaying the results
par(mfrow = c(1,2))
plot(results, trait = 1, main = "Trait 1: Distance")
plot(results, trait = c(2,3), main = "Variables correlation")
## TODO: debug correlation!
```

## Step 4 and 5:
 4. "Founding event": after reaching a value for the "Distance" trait, a founding event happens with a new birth-death process with no extinction and where the "Variable" is now a 2D correlated Brownian Motion.
 5. "Mass extinction event": after reaching a certain number of taxa in the "Founding event", 80% of taxa go extinct randomly.

```{r}
## The new trait for the founding event
new_traits <- make.traits(process = c(BM.process, BM.process), n = c(1, 2), 
                          process.args = list(
                              list(NULL),
                              list(Sigma = matrix(c(0.7, 0, 0, 0.7), 2, 2))),
                          trait.names = c("Distance", "Variables"))
#TODO: set starting value to be ancestors' value

## The mass extinction event after reaching 50 taxa
mass_extinction <- make.events(target       = "taxa",
                               condition    = taxa.condition(50),
                               modification = random.extinction(0.8))
```

```{r}
## The selective event
events_list <- make.events(event.name   = "Founding", 
                           add          = events_list,
                           target       = "founding",
                           condition    = trait.condition(3),
                           modification = founding.event(traits    = new_traits,
                                                         events    = mass_extinction,
                                                         bd.params = list(
                                                            speciation = 2,
                                                            extinction = 0)))
#TODO: fix bug
```

```{r}
## Running a birth death tree 
results <- dads(bd.params  = bd_params,
                stop.rule  = stop_rule,
                traits     = two_traits,
                modifiers  = distance_speciation,
                events     = events_list,
                null.error = 100)

## And displaying the results
par(mfrow = c(1,2))
plot(results, trait = 1, main = "Trait 1: Distance")
plot(results, trait = c(2,3), main = "Variables correlation")
## TODO: debug correlation!
```

### 2021-01-11 inst/gitbook/05_thorough_example.Rmd
Update 05_thorough_example.Rmd

@@
     - "Variable": a 2D uncorrelated Ornstein-Uhlenbeck trait that corresponds to some traits the lineages have.
- During the whole process, species that are more distant from their ancestor have a greater change of speciating.
+ During the whole process, lineages that are more further away from the other species in terms of distance have a higher change of speciating (proportional to that distance).
 
@@
 
- 2. "Correlation event": after reaching a certain number of taxa an event happens making the "Variable" trait dimensions become correlated.
- 3. "Selective extinction event": after reaching a certain point in time, lineages with their "Distance" values in some interval goes extinct.
+ 2. "Correlation event": after reaching a certain number of taxa an event happens making the "Variables" trait dimensions become correlated.
+ 3. "Selective extinction event": after reaching a certain point in time, lineages with their "Variables" values that are negative go extinct
  4.
@@
 ```{r}
-## The OU 2D trait correlation
-correlation <- matrix(0, 2, 2)
-
 ## The trait object
 two_traits <- make.traits(process = c(BM.process, OU.process), n = c(1, 2),
-                          process.args = list(list(NULL), list(Sigma = correlation)),
                           trait.names = c("Distance", "Variables"))
-#TODO: fix OU with correlation
+```
+
+```{r}
+## Visualise the 1D trait (BM.process)
+plot(two_traits, trait = 1)
+```
+
+```{r, eval = FALSE}
+## Visualise the 2D trait (uncorrelated OU.process)
+plot(two_traits, trait = 2, use.3D = TRUE)
 ```
@@
 plot(results, trait = c(2,3), main = "Variables correlation")
-
-## TODO: debug correlation!
 ```
@@
 ```{r}
+## Speciation event is more likely if lineage's ancestor is further away from the mean trait value
+distance.modify <- function(x, trait.values, lineage) return(x + x * abs(parent.traits(trait.values, lineage)[1] - mean(trait.values[,1])))
+
 ## Make a distance modifier
-distance_speciation <- make.modifiers(speciation = NULL,
-                                      condition = NULL,
-                                      modify = NULL) 
+distance.speciation <- make.modifiers(speciation = speciation,
+                                      modify = distance.modify)
 ```
@@
                 traits     = two_traits,
-                modifiers  = distance_speciation,
+                modifiers  = distance.speciation,
                 null.error = 100)
@@
                     trait.names  = "Variables"))
-#TODO: fix bug here
 ```
@@
                 traits     = two_traits,
-                modifiers  = distance_speciation,
+                modifiers  = distance.speciation,
                 events     = events_list,
@@
 plot(results, trait = c(2,3), main = "Variables correlation")
-## TODO: debug correlation!
 ```
@@
 ## Step 3:
- 3. "Selective extinction event": after reaching a certain point in time, lineages with their "Distance" values in some interval goes extinct.
+ 3. "Selective extinction event": after reaching a certain point in time, lineages with their "Variables" values that are negative go extinct
 
@@
                            condition    = time.condition(4),
-                           modification = trait.extinction(1, condition = `>`),
-                           event.name   = "Selection")
+                           modification = trait.extinction(0, condition = `<`, trait = 2))
 ```
@@
 new_traits <- make.traits(process = c(BM.process, BM.process), n = c(1, 2),
-                          process.args = list(
-                              list(NULL),
-                              list(Sigma = matrix(c(0.7, 0, 0, 0.7), 2, 2))),
                           trait.names = c("Distance", "Variables"))
-#TODO: set starting value to be ancestors' value
 
@@
                                                             speciation = 2,
-                                                            extinction = 0)))
-#TODO: fix bug
+                                                            extinction = 0)),
+                           additional.args = list(prefix = "found_"))
 ```

### 2021-01-11 inst/gitbook/04_make.events.Rmd
Create 04_make.events.Rmd

--- 2020-12-09 inst/gitbook/03_make.modifiers.Rmd
@@
 ---
-title: "`make.modifiers`"
+title: "`make.events`"
 author: "Thomas Guillerme (guillert@tcd.ie)"
@@
 
-# Making modifiers with `make.modifiers()` {#makemodifiers}
+# Making events with `make.events()` {#makeevents}
 
@@
 ```
+
+## Conditions
+
+## Modifications
+
+## Some events examples
 

### 2021-01-12 inst/gitbook/04_make.events.Rmd
Update 04_make.events.Rmd

@@
 
-`"modifiers"` have a similar structure than `"traits"` where you can design an object with increasing complexity, starting with the simplest modifiers that doesn't modify anything (using the default arguments):
+### Random mass extinction after some time
 
 ```{r}
-## Making a default modifier (no modification)
-my_default_modifiers <- make.modifiers()
-my_default_modifiers
+## Simulation parameters
+stop.rule <- list(max.time = 5)
+bd.params <- list(extinction = 0, speciation = 1)
+
+## 80% mass extinction at time 4
+mass_extinction <- make.events(
+                      target = "taxa",
+                      condition = time.condition(4),
+                      modification = random.extinction(0.8))
+
+## Running the simulations
+set.seed(1)
+results <- dads(bd.params = bd.params,
+                stop.rule = stop.rule,
+                events    = mass_extinction)
+## Plotting the results
+plot(results, show.tip.label = FALSE)
+axisPhylo()
 ```
 
-Similarly to `"traits"` objects, `"modifiers"` are also printed by default using `print.dads`.
-You can see details about what's actually in the object using `print.dads(my_default_modifiers, all = TRUE)`.
-However, contrary to `"traits"`, you cannot plot `"modifiers"`.
-
-## The branch length function (`branch.length`)
-
-The first argument in `"modifiers"` is the branch length function (`branch.length`) this is the function that will be executed in `dads` to generate branch length.
-Note that in the `dads` algorithm, branch length is not generated _directly_ but actually results of the waiting time.
-In other words, the `branch.length` function just affects waiting time for all taxa present at any time in the simulation.
-These taxa can then either go extinct (stopping the "growth" of it's branch length) or survive (continuing the "growth").
-
-By default, branch length (or waiting or growth) is a randomly drawn number from an exponential distribution with the rate of the number of taxa multiplied by the speciation and extinction rate: $n \times (\lambda + \mu)$ (where $n$ is the number of taxa currently present in the simulation, $\lambda$ and $\mu$ are respectively the speciation and extinction rates).
-This default function is simply called `branch.length` in `dads` and can be used as a modifier as follows:
+### Extinction based on a trait value after some time
 
 ```{r}
-## Specifying the default modifier
-default_modifiers <- make.modifiers(branch.length = branch.length)
+## Extinction of any tips with trait < 1 at time 4
+trait_extinction <- make.events(
+                      target = "taxa",
+                      condition = time.condition(4),
+                      modification = trait.extinction(1))
 
-## Setting some parameters for generating trees
-bd_params <- list(extinction = 0)
-stop_rule <- list(max.living = 20)
-
-## Generating a tree with the default branch length parameter
-set.seed(0)
-default_tree <- dads(bd.params = bd_params,
-                     stop.rule = stop_rule,
-                     modifiers = default_modifiers)
+## Running the simulations
+set.seed(7)
+results <- dads(bd.params = bd.params,
+                stop.rule = stop.rule,
+                traits    = make.traits(),
+                events    = trait_extinction)
+## Plotting the results
+plot(results)
 ```
 
-Of course, the point of the modularity here is that you can provide your own function for generating branch length.
-For example, we might be interested in what our tree would look like if we'd use a simple constant branch length generation (instead of randomly drawing it from an exponential distribution).
-We can do so by declaring our own `branch.length` function and adding it to a `"modifiers"` object.
+### Adding a background extinction after reaching a number of living taxa
 
 ```{r}
-## A constant branch length generator
-## (note that the output must be numeric, not integer)
-constant.brlen <- function() {
-    return(as.numeric(1))
+## Simulation parameters
+stop.rule <- list(max.living = 50)
+bd.params <- list(extinction = 0, speciation = 1)
+
+## Adding an extinction parameter after 30 taxa
+background_extinction <- make.events(
+                      condition    = taxa.condition(30),
+                      target       = "bd.params",
+                      modification = update.bd.params(1/3, "extinction"))
+
+## Running the simulations
+set.seed(2)
+results <- dads(bd.params = bd.params,
+                stop.rule = stop.rule,
+                events    = background_extinction)
+## Plotting the results
+plot(results)
+```
+
+### Reducing speciation after some time
+
+```{r}
+## Simulation parameters
+stop.rule <- list(max.time = 4)
+
+## Reducing speciation after reaching time 2
+reduced_speciation <- make.events(
+                      condition    = time.condition(2),
+                      target       = "bd.params",
+                      modification = update.bd.params(1/3, "speciation"))
+
+set.seed(42)
+no_event <- dads(bd.params = bd.params,
+                 stop.rule = stop.rule)
+set.seed(42)
+reduced_speciation_event <- dads(bd.params = bd.params,
+                                 stop.rule = stop.rule,
+                                 events    = reduced_speciation)
+
+## Plot both trees
+par(mfrow = c(1, 2))
+plot(no_event, main = "No event")
+axisPhylo()
+plot(reduced_speciation_event, main = "Reduced speciation after time 2")
+axisPhylo()
+```
+
+### Changing trait process after some time
+
+```{r}
+## Simulation parameters
+stop.rule <- list(max.time = 6)
+traits <- make.traits()
+
+## Create an event to change the trait process
+change_process <- make.events(
+                  condition    = time.condition(5),
+                  target       = "traits",
+                  modification = update.traits(process = OU.process))
+
+## Run the simulations
+set.seed(1)
+no_change <- dads(bd.params = bd.params,
+                  stop.rule = stop.rule,
+                  traits = traits)
+set.seed(1)
+process_change <- dads(bd.params = bd.params,
+                       stop.rule = stop.rule,
+                       traits = traits,
+                       events = events)
+## Plot the results
+par(mfrow = c(1,2))
+plot(no_change, ylim = c(-5, 8))
+plot(process_change, ylim = c(-5, 8))
+```
+
+### Changing trait correlation after reaching a trait value
+
+```{r}
+## Set the parameters
+stop.rule <- list(max.taxa = 100)
+
+## A correlated trait
+traits <- make.traits(n = 2, process.args = list(Sigma = matrix(1, 2, 2)))
+
+## Event changing a trait correlation
+change_correlation <- make.events(
+    condition    = trait.condition(3, absolute = TRUE),
+    target       = "traits",
+    modification = update.traits(process.args = list(Sigma = matrix(c(10,3,3,2),2,2))))
+
+## Run the simulations
+set.seed(8)
+no_event <- dads(bd.params = bd.params,
+                 stop.rule = stop.rule,
+                 traits    = traits)
+set.seed(8)
+change_correlation <- dads(bd.params = bd.params,
+                           stop.rule = stop.rule,
+                           traits    = traits,
+                           events    = events)
+## Visual testing
+par(mfrow = c(1,2))
+plot(no_event, trait = 2)
+plot(change_correlation, trait = 2)
+# plot(no_event, trait = c(1:2))
+# plot(change_correlation, trait = c(1:2), use.3D = TRUE)
+```
+
+### Event for changing a modifier
+
+```{r}
+## Set the parameters
+stop.rule <- list(max.time = 4)
+bd.params <- list(extinction = 0, speciation = 1)
+
+## Using a default trait and modifier object
+modifiers <- make.modifiers()
+traits <- make.traits()
+
+## New condition and new modifier (increasing speciation if trait is negative)
+new.condition <- function(trait.values, lineage) {
+    return(parent.traits(trait.values, lineage) < 0)
+}
+new.modify    <- function(x, trait.values, lineage) {
+  return(x + 1)
 }
 
-## Creating the modifiers object
-constant_modifier <- make.modifiers(branch.length = constant.brlen)
+## Update the modifier
+change_speciation <- make.events(
+    condition    = time.condition(3),
+    target       = "modifiers",
+    modification = update.modifiers(speciation = speciation,
+                                    condition  = new.condition,
+                                    modify     = new.modify))
 
-## Generating a new tree with this modifier
-set.seed(0)
-modified_tree <- dads(bd.params = bd_params,
-                      stop.rule = stop_rule,
-                      modifiers = constant_modifier)
+set.seed(4)
+no_event <- dads(bd.params = bd.params,
+                 stop.rule = stop.rule,
+                 traits    = traits,
+                 modifiers = modifiers)
+set.seed(4)
+change_spec <- dads(bd.params = bd.params,
+                    stop.rule = stop.rule,
+                    traits    = traits,
+                    modifiers = modifiers,
+                    events    = change_speciation)
+
+## Visualise the results
+par(mfrow = c(1,2))
+plot(no_event, main = "No event")
+plot(change_spec, main = "Increase extinction for negative traits\nafter time 3")
 ```
 
-And we can visualise the difference between both resulting trees:
+### Changing branch length when reaching n taxa
 
 ```{r}
+## Parameters
+stop.rule <- list(max.taxa = 100)
+modifiers <- make.modifiers()
+traits <- make.traits()
+
+## multiplying branch length 100 folds
+new.modify <- function(x, trait.values, lineage) {
+  return(x * 100)
+}
+
+## Event for increasing branch length after reaching 30 taxa
+increase_brlen <- make.events(
+                  condition    = taxa.condition(30),
+                  target       = "modifiers",
+                  modification = update.modifiers(
+                                    branch.length = branch.length,
+                                    modify = new.modify))
+
+## Run the simulations
+set.seed(5)
+no_event <- dads(bd.params = bd.params,
+                 stop.rule = stop.rule,
+                 traits    = traits,
+                 modifiers = modifiers)
+set.seed(5)
+increased_brlen <- dads(bd.params = bd.params,
+                        stop.rule = stop.rule,
+                        traits    = traits,
+                        modifiers = modifiers,
+                        events    = increase_brlen)
+
+## Visualise the results
 par(mfrow = c(1,2))
-plot(default_tree,  main = "Default modifier")
-plot(modified_tree, main = "Constant branch length\nmodifier")
+plot(no_event, main = "No event")
+plot(increased_brlen, main = "Increase branch length\nafter 30 taxa")
 ```
 
-It is of course to use more complex branch length modifiers that intakes different conditions and specific modification rather than simply always output a value of one.
-
-### The allowed arguments {#allowarguments}
-
-It is of course possible to design some more advanced function to interact with the birth death process.
-You can create a function for `branch.length`, `selection` and `speciation` that involve any of the following arguments:
-
-  * `bd.params`: a named list containing `"numeric"` values that contains the birth death parameters (at least `"speciation"` and `"extinction"`);
-  * `lineage`: a named list containing the lineage data (see below).
-  * `trait.values`: a `"matrix"` containing `"numeric"` values with the trait names as column names and the lineages ID as row numbers (you can use it with the function `parent.traits` to access the trait of the previous node for example).
-  * `modify.fun`: a `"list"` of named `"function"` (usually passed through `condition` and `modify`).
-
-The `lineage` list contains the following elements (missing elements are allowed):
-
- * `lineage$parents`: an `"integer"` vector: the list of parent lineages;
- * `lineage$livings`: an `"integer"` vector: the list of lineages still not extinct;
- * `lineage$drawn`: a single `"integer"`: the ID of the selected lineage;
- * `lineage$current`: a single `"integer"`: the selected lineage (is equal to `lineage$livings[lineage$drawn]`);
- * `lineage$n`: a single `"integer"`: the current number of non extinct lineage (is equal to `length(lineage$livings))`;
- * `lineage$split`: a `"logical"` vector: the list of splits for each lineage (`TRUE`), the number of total tips is equal to `sum(!lineage$split)`.
-
-In general, unless you know what you're doing, you can ignore most arguments for specific modifiers since they are handled automatically within the `dads` function.
-Therefore any argument can be left undeclared or missing and is always handled internally.
-For example, if you did not declare `n.taxa` as a function argument but are using `n.taxa` in the function, `n.taxa` will be detected and treated as a current argument automatically as set accordingly within the birth death process (e.g.
-`n.taxa` will be set to the current number of taxa every iteration of the process).
-
-For example, we can create a function that increases branch length proportional to the number of species "alive" at each time of the simulation in a discrete way.
-I.e.
-for discrete numbers of taxa, the branch length increases by jumps (ten fold):
+### Founding event: a subtree with no fossils
 
 ```{r}
-## A more complex binned.branch.length function
-increasing.brlen <- function(bd.params, lineage) {
+## Set up parameters
+stop.rule <- list(max.time = 4)
+bd.params <- list(speciation = 1, extinction = 0.3)
 
-    ## Setting the cumulated birth and death
-    birth_death <- bd.params$speciation + bd.params$extinction
+## Events that generate a new process (founding effects)
+founding_event <- make.events(
+                  condition    = taxa.condition(10),
+                  target       = "founding",
+                  modification = founding.event(
+                                    bd.params = list(speciation = 2,
+                                                     extinction = 0)),
+                  additional.args = list(prefix = "founding_"))
 
-    ## Returning branch lengths depending on different number of taxa
-    if(lineage$n <= 5) {
-        return(1    * rexp(1, sum(5 * birth_death)))
-    }
-    if(lineage$n <= 10) {
-        return(10   * rexp(1, sum(10 * birth_death)))   
-    }
-    if(lineage$n <= 15) {
-        return(100  * rexp(1, sum(15 * birth_death)))      
-    }
-    if(lineage$n <= 20) {
-        return(1000 * rexp(1, sum(20 * birth_death)))
-    } else {
-        return(1000 * rexp(1, sum(lineage$n * birth_death)))
-    }
-}
+## Simulations
+set.seed(11)
+founding_tree <- dads(bd.params = bd.params,
+                      stop.rule = stop.rule,
+                      events    = founding_event)
+plot(founding_tree, cex = 0.4)
 ```
 
-We can then create it as a `"modifiers"` object and run a new simulation:
+### Founding event with traits
 
 ```{r}
-## Creating a modifiers
-increasing_modifier <- make.modifiers(branch.length = increasing.brlen)
+## Set up parameters
+stop.rule <- list(max.time = 4)
+bd.params <- list(speciation = 1, extinction = 0.3)
+traits    <- make.traits()
 
-## Generating a new tree with this modifier
-set.seed(0)
-increasing_tree <- dads(bd.params = bd_params,
-                        stop.rule = stop_rule,
-                        modifiers = increasing_modifier)
+## Events that generate a new process (founding effects)
+founding_event <- make.events(
+                  condition    = taxa.condition(10),
+                  target       = "founding",
+                  modification = founding.event(
+                                    bd.params = list(speciation = 2,
+                                                     extinction = 0),
+                                    traits = make.traits(process = OU.process)),
+                  additional.args = list(prefix = "founding_"))
+
+## Simulating the tree
+set.seed(19)
+founding_tree <- dads(bd.params = bd.params,
+                      stop.rule = stop.rule,
+                      traits    = make.traits(),
+                      events    = founding_event)
+plot(founding_tree)
 ```
-
-And we can visualise the difference between the resulting trees:
-
-```{r}
-par(mfrow = c(1,3))
-plot(default_tree,    main = "Default modifier")
-plot(modified_tree,   main = "Constant branch length\nmodifier")
-plot(increasing_tree, main = "Increasing branch length\nmodifier (binned)")
-```
-
-## The selection function (`selection`)
-
-The `selection` function is used in the birth death process to know which lineage to select when running a speciation (or extinction!) event.
-By default, this function randomly selects one taxon that is currently not extinct (using: `sample(1:n.taxa, 1))`.
-Similarly as `branch.length` it is possible to modify this part of the birth death process.
-For example, we could simply select always the last created lineage (to create a "ladder" or most asymmetric tree):
-
-```{r}
-## Our function to always select the last taxon
-## (making sure it returns an integer)
-select.last <- function(lineage) {
-    return(as.integer(lineage$n))
-}
-```
-
-> Note that here the function can only intake the [allowed arguments as described above](#allowarguments) (here `n.taxa`: the number of current living taxa).
-
-We can then create a `"modifiers"` object the same way as before using this time the `selection` argument:
-
-```{r}
-## A modifier for selection
-ladderised_modifier <- make.modifiers(selection = select.last)
-## Generating a new tree with this modifier
-set.seed(0)
-ladderised_tree <- dads(bd.params = bd_params,
-                        stop.rule = stop_rule,
-                        modifiers = ladderised_modifier)
-## Displaying the results
-par(mfrow = c(1,2))
-plot(default_tree,    main = "Default modifier")
-plot(ladderised_tree, main = "Ladderising modifier")
-```
-
-Again, it is of course possible to make the modifier more complex and in combination with other elements of the tree.
-For example, we can create a `"dads"` object that also creates a trait a add to it a `selection` modifier that only selects for tips with positive trait values.
-
-```{r}
-## Our function that only select taxa with positive trait values
-select.positive <- function(trait.values, lineage) {
-    ## Selecting the taxa with positive traits only
-    positives <- as.integer(rownames(trait.values)[which(trait.values[, 1] >= 0)])
-
-    ## Select the current taxa that descend from a node with a positive value
-    positive_living <- cbind(lineage$parents, seq_along(lineage$split)
-                             )[which(lineage$parents %in% positives), 2]
-
-    ## Select one tip randomly in the ones with descendants with positive values
-    return(sample(which(lineage$livings %in% positive_living), 1))
-}
-
-## Creating the modifier
-positive_skew <- make.modifiers(selection = select.positive)
-
-## Creating a (default) trait object
-BM_trait <- make.traits()
-
-## Simulate a tree and trait with no modifier
-set.seed(1)
-default_dads <- dads(bd.params = bd_params,
-                     stop.rule = stop_rule,
-                     traits    = BM_trait)
-
-## Simulate a tree and trait with the modifier
-set.seed(1)
-skewed_trait_dads <- dads(bd.params = bd_params,
-                          stop.rule = stop_rule,
-                          traits    = BM_trait,
-                          modifiers = positive_skew)
-
-## Plotting the differences in trees and traits
-par(mfrow = c(1, 2))
-plot(default_dads, main = "Default trait and tree")
-plot(skewed_trait_dads, main = "Skewed trait and tree")
-```
-
-## The speciation function (`speciation`)
-
-The third function that can be used to modify the birth death process is the `speciation` function.
-This one is used during the birth death process to decide whether a lineages speciates (creating a node and two new lineages) or goes extinct (creating a tip).
-
-> Note that the `speciation` function only affects tips or nodes before the simulation reaches the `stop.rule`.
-The then surviving lineages are all automatically transformed into tips.
-
-By default, the `speciation` function is trigger a speciation even if a number randomly drawn from a uniform distribution is lower than the ratio between the speciation and the speciation and extinction parameter.
-If the randomly drawn number is higher, the lineage goes extinct.
-
-```{r, eval = FALSE}
-## The speciation in pseudo-code:
-runif(1) < speciation/ (speciation + extinction)
-```
-
-Creating a `"modifiers"` with a `speciation` function works the same way as for `branch.length` and `selection` but the function that will be used needs to output a logical value (see table [below](#summarymodifiers)).
-Once the function is created simply input your function for speciation in the modifier and run the `dads` function with that modifier:
-
-```{r, eval = FALSE}
-## Speciating or going extinct randomly
-## (regardless of the extinction parameter)
-random.extinct  <- function() {
-    return(sample(c(TRUE, FALSE), 1))
-}
-
-## Creating the modifiers object
-random_extinction <- make.modifiers(speciation = random.extinct)
-
-## Generating a new tree with this modifier
-set.seed(0)
-modified_tree <- dads(bd.params = bd_params,
-                      stop.rule = stop_rule,
-                      modifiers = random_extinction)
-
-par(mfrow = c(1,2))
-plot(default_tree,  main = "Default modifier")
-plot(modified_tree, main = "Random extinction\nmodifier")
-```
-
-Note how every lineage end up going extinct!
-And again, we can make some more advanced modifiers: for example, one where a tip always goes extinct if their ancestor has a negative trait value (here we will also introduce the utility function `parent.trait` that automatically selects the trait values of the parent of the current lineage.
-
-```{r}
-## A modifier for removing tips with negative values
-bias.trait <- function(trait.values, lineage) {
-    if(parent.traits(trait.values, lineage) < 0) {
-        ## Go extinct!
-        return(FALSE)
-    } else {
-        ## Speciate
-        return(TRUE)
-    }
-}
-
-## Creating the modifier
-biased_trait <- make.modifiers(speciation = bias.trait)
-
-## Simulate a tree and trait with the modifier
-set.seed(1)
-biased_trait_dads <- dads(bd.params = bd_params,
-                          stop.rule = stop_rule,
-                          traits    = BM_trait,
-                          modifiers = biased_trait)
-
-## Plotting the differences in trees and traits
-par(mfrow = c(1, 2))
-plot(default_dads, main = "Default trait and tree")
-plot(biased_trait_dads, main = "Biased trait and tree")
-```
-
-## Summary of the inputs and outputs for the `branch.length`, `selection` and `speciation` modifiers {#summarymodifiers}
-
-modifier name   | accepted input (arguments) | required output (class) |
-----------------|----------------------------|-------------------------|
-`branch.length` | `bd.params`, `lineage`, `trait.values` | `"numeric"` |
-`selection`     | `bd.params`, `lineage`, `trait.values` | `"integer"` |
-`speciation`    | `bd.params`, `lineage`, `trait.values` | `"logical"` |
-
-## The condition and modify functions (`condition` and `modify`)
-
-In the examples above, we have seen how to specify modifications to the birth death process (via `branch.length`, `selection` and `speciation`), however, one might note that these modifications are not dynamic.
-In other words, throughout the process, the modifications remain constant (even if they are conditional).
-It is however possible to code the `"modifiers"` so that they can be affected by `"events"` objects (see next chapter).
-
-To do so, you can formally declare conditions (`condition`) and modifications (`modify`) as internal functions that can then be modified my an `"events"` object.
-`condition` and `modify` are hard coded in the `branch.length` function that they concern, i.e. they are variables (functions) within the function.
-
-## Combining and editing modifiers (`add`)
-
-## Testing modifiers (`test`)
-
-## Demo runnable
-
-```{r, fig.height = 16, fig.width = 8, eval = FALSE}
-bd.params <- list(speciation = 1, extinction = 1/3)
-traits <- make.traits()
-stop.rule <- list(max.taxa = 20, max.living = Inf, max.time = Inf)
-modifiers <- NULL
-events <- NULL
-null.error <- NULL
-
-## modifiers
-condition <- function(trait.values, parent.lineage) return(parent.traits(trait.values, parent.lineage) < 0)
-modify <- function(x) return(x * 20)
-
-## Setting up the different modifiers
-modify_speciation <- make.modifiers(speciation    = speciation.trait,
-                                    condition     = condition,
-                                    modify        = modify)
-modify_brlen <- make.modifiers(branch.length = branch.length.trait,
-                               condition     = condition,
-                               modify        = modify)
-modify_speciation_brlen <- make.modifiers(branch.length = branch.length.trait,
-                                          speciation    = speciation.trait,
-                                          condition     = condition,
-                                          modify        = modify)
-
-## Test normal (no modifiers)
-set.seed(1)
-test <- dads(bd.params, stop.rule, traits, null.error = 20)
-par(mfrow = c(4,2))
-plot(test$tree)
-plot.dads(test, main = "random tree + trait")
-
-## Test with modifiers
-set.seed(1)
-trait_table <- NULL
-test <- dads(bd.params, stop.rule, traits,
-                                modifiers = modify_speciation,
-                                null.error = 20)
-plot(test$tree)
-plot.dads(test, main = "Skewed speciation")
-
-set.seed(1)
-trait_table <- NULL
-test <- dads(bd.params, stop.rule, traits,
-                                modifiers = modify_brlen,
-                                null.error = 20)
-plot(test$tree)
-plot.dads(test, main = "Skewed branch length")
-
-set.seed(1)
-trait_table <- NULL
-test <- dads(bd.params, stop.rule, traits,
-                                modifiers = modify_speciation_brlen,
-                                null.error = 20)
-plot(test$tree)
-plot.dads(test, main = "Skewed branch length and speciation")
-
-```

### 2021-01-12 inst/gitbook/03_make.modifiers.Rmd
UTFs
removed examples

@@
 ```{r, fig.height = 16, fig.width = 8, eval = FALSE}
-bd.params <- list(speciation = 1, extinction = 1/3)
-traits <- make.traits()
-stop.rule <- list(max.taxa = 20, max.living = Inf, max.time = Inf)
-modifiers <- NULL
-events <- NULL
-null.error <- NULL
-
-## modifiers
-condition <- function(trait.values, parent.lineage) return(parent.traits(trait.values, parent.lineage) < 0)
-modify <- function(x) return(x * 20)
-
-## Setting up the different modifiers
-modify_speciation <- make.modifiers(speciation    = speciation.trait,
-                                    condition     = condition,
-                                    modify        = modify)
-modify_brlen <- make.modifiers(branch.length = branch.length.trait,
-                               condition     = condition,
-                               modify        = modify)
-modify_speciation_brlen <- make.modifiers(branch.length = branch.length.trait,
-                                          speciation    = speciation.trait,
-                                          condition     = condition,
-                                          modify        = modify)
-
-## Test normal (no modifiers)
-set.seed(1)
-test <- dads(bd.params, stop.rule, traits, null.error = 20)
-par(mfrow = c(4,2))
-plot(test$tree)
-plot.dads(test, main = "random tree + trait")
-
-## Test with modifiers
-set.seed(1)
-trait_table <- NULL
-test <- dads(bd.params, stop.rule, traits,
-                                modifiers = modify_speciation,
-                                null.error = 20)
-plot(test$tree)
-plot.dads(test, main = "Skewed speciation")
-
-set.seed(1)
-trait_table <- NULL
-test <- dads(bd.params, stop.rule, traits,
-                                modifiers = modify_brlen,
-                                null.error = 20)
-plot(test$tree)
-plot.dads(test, main = "Skewed branch length")
-
-set.seed(1)
-trait_table <- NULL
-test <- dads(bd.params, stop.rule, traits,
-                                modifiers = modify_speciation_brlen,
-                                null.error = 20)
-plot(test$tree)
-plot.dads(test, main = "Skewed branch length and speciation")
 

### 2021-01-12 inst/gitbook/04_make.events.Rmd
events and thorough example INIT (working-ish examples)

@@
 ```{r}
+## Simulation parameters
+stop.rule <- list(max.time = 5)
+bd.params <- list(extinction = 0, speciation = 1)
+
 ## Extinction of any tips with trait < 1 at time 4
@@
 stop.rule <- list(max.time = 4)
+bd.params <- list(extinction = 0, speciation = 1)
 
@@
 ## Simulation parameters
+bd.params <- list(extinction = 0, speciation = 1)
 stop.rule <- list(max.time = 6)
@@
                        traits    = traits,
-                       events = events)
+                       events    = change_process)
 ## Plot the results
 par(mfrow = c(1,2))
-plot(no_change, ylim = c(-5, 8))
-plot(process_change, ylim = c(-5, 8))
+plot(no_change, ylim = c(-7, 7))
+plot(process_change, ylim = c(-7, 7))
 ```
@@
 stop.rule <- list(max.taxa = 100)
+bd.params <- list(extinction = 0, speciation = 1)
 
@@
                            traits    = traits,
-                           events    = events)
+                           events    = change_correlation)
 ## Visual testing
@@
 ## Parameters
+bd.params <- list(extinction = 0, speciation = 1)
 stop.rule <- list(max.taxa = 100)

### 2021-01-12 inst/gitbook/05_thorough_example.Rmd
events and thorough example INIT (working-ish examples)

@@
 ```{r}
+set.seed(0) # 123 also works
 ## Running a birth death tree 
 results <- dads(bd.params = bd_params,
-                stop.rule = stop_rule,
-                null.error = 100) # This option will try the requested options up until a 100 times
+                stop.rule = stop_rule)
 
@@
 
-```{r, eval = FALSE}
+```{r}
 ## Visualise the 2D trait (uncorrelated OU.process)
@@
 ```{r}
+set.seed(0)
 ## Running a birth death tree 
@@
                 stop.rule  = stop_rule,
-                traits     = two_traits,
-                null.error = 100)
+                traits     = two_traits)
 
@@
 plot(results, trait = c(2,3), main = "Variables correlation")
+```
+
+```{r}
+## Visualising the tree in 3D
+plot(results, trait = c(2,3), use.3D = TRUE)
 ```
@@
 ```{r}
+set.seed(42)
 ## Running a birth death tree 
@@
                 traits     = two_traits,
-                modifiers  = distance.speciation,
-                null.error = 100)
+                modifiers  = distance.speciation)
 
@@
 ```{r}
+set.seed(8)
+## Running a birth death tree 
+results <- dads(bd.params  = bd_params,
+                stop.rule  = stop_rule,
+                traits     = two_traits,
+                modifiers  = distance.speciation,
+                events     = events_list)
+
+## And displaying the results
+par(mfrow = c(1,2))
+plot(results, trait = 1, main = "Trait 1: Distance")
+plot(results, trait = c(2,3), main = "Variables correlation")
+```
+
+```{r, eval = FALSE}
+## Visualising the tree in 3D
+plot(results, trait = c(2,3), use.3D = TRUE)
+```
+
+## Step 3:
+ 3. "Selective extinction event": after reaching a certain point in time, lineages with their "Variables" values that are negative go extinct
+
+# BUGGED FROM HERE
+
+```{r, eval = FALSE}
+## Adding the selective event
+events_list <- make.events(event.name   = "Selection",
+                           add          = events_list,
+                           target       = "taxa",
+                           condition    = time.condition(4),
+                           modification = trait.extinction(x = 0, condition = `<`, trait = 2))
+```
+Problem with `modification` argument here
+
+```{r, eval = FALSE}
+## Running a birth death tree 
+results <- dads(bd.params  = bd_params,
+                stop.rule  = stop_rule,
+                traits     = two_traits,
+                modifiers  = distance.speciation,
+                events     = events_list)
+
+## And displaying the results
+par(mfrow = c(1,2))
+plot(results, trait = 1, main = "Trait 1: Distance")
+plot(results, trait = c(2,3), main = "Variables correlation")
+```
+
+## Step 4 and 5:
+ 4.
+"Founding event": after reaching a value for the "Distance" trait, a founding event happens with a new birth-death process with no extinction and where the "Variable" is now a 2D correlated Brownian Motion.
+ 5. "Mass extinction event": after reaching a certain number of taxa in the "Founding event", 80% of taxa go extinct randomly.
+
+```{r, eval = FALSE}
+## The new trait for the founding event
+new_traits <- make.traits(process = c(BM.process, BM.process), n = c(1, 2),
+                          trait.names = c("Distance", "Variables"))
+
+## The mass extinction event after reaching 50 taxa
+mass_extinction <- make.events(target       = "taxa",
+                               condition    = taxa.condition(50),
+                               modification = random.extinction(0.8))
+```
+
+```{r, eval = FALSE}
+## The selective event
+events_list <- make.events(event.name   = "Founding", 
+                           add          = events_list,
+                           target       = "founding",
+                           condition    = trait.condition(3),
+                           modification = founding.event(traits    = new_traits,
+                                                         events    = mass_extinction,
+                                                         bd.params = list(
+                                                            speciation = 2,
+                                                            extinction = 0)),
+                           additional.args = list(prefix = "found_"))
+```
+
+```{r, eval = FALSE}
 ## Running a birth death tree 
@@
 ```
-
-## Step 3:
- 3. "Selective extinction event": after reaching a certain point in time, lineages with their "Variables" values that are negative go extinct
-
-```{r}
-## Adding the selective event
-events_list <- make.events(event.name   = "Selection",
-                           add          = events_list,
-                           target       = "taxa",
-                           condition    = time.condition(4),
-                           modification = trait.extinction(0, condition = `<`, trait = 2))
-```
-
-```{r}
-## Running a birth death tree 
-results <- dads(bd.params  = bd_params,
-                stop.rule  = stop_rule,
-                traits     = two_traits,
-                modifiers  = distance.speciation,
-                events     = events_list,
-                null.error = 100)
-
-## And displaying the results
-par(mfrow = c(1,2))
-plot(results, trait = 1, main = "Trait 1: Distance")
-plot(results, trait = c(2,3), main = "Variables correlation")
-```
-
-## Step 4 and 5:
- 4.
-"Founding event": after reaching a value for the "Distance" trait, a founding event happens with a new birth-death process with no extinction and where the "Variable" is now a 2D correlated Brownian Motion.
- 5. "Mass extinction event": after reaching a certain number of taxa in the "Founding event", 80% of taxa go extinct randomly.
-
-```{r}
-## The new trait for the founding event
-new_traits <- make.traits(process = c(BM.process, BM.process), n = c(1, 2),
-                          trait.names = c("Distance", "Variables"))
-
-## The mass extinction event after reaching 50 taxa
-mass_extinction <- make.events(target       = "taxa",
-                               condition    = taxa.condition(50),
-                               modification = random.extinction(0.8))
-```
-
-```{r}
-## The selective event
-events_list <- make.events(event.name   = "Founding", 
-                           add          = events_list,
-                           target       = "founding",
-                           condition    = trait.condition(3),
-                           modification = founding.event(traits    = new_traits,
-                                                         events    = mass_extinction,
-                                                         bd.params = list(
-                                                            speciation = 2,
-                                                            extinction = 0)),
-                           additional.args = list(prefix = "found_"))
-```
-
-```{r}
-## Running a birth death tree 
-results <- dads(bd.params  = bd_params,
-                stop.rule  = stop_rule,
-                traits     = two_traits,
-                modifiers  = distance.speciation,
-                events     = events_list,
-                null.error = 100)
-
-## And displaying the results
-par(mfrow = c(1,2))
-plot(results, trait = 1, main = "Trait 1: Distance")
-plot(results, trait = c(2,3), main = "Variables correlation")
-```

### 2021-01-12 inst/gitbook/_book/dads_manual.tex
Updated gitbook for version 0.1.3

@@
 \author{Thomas Guillerme (\href{mailto:guillert@tcd.ie}{\nolinkurl{guillert@tcd.ie}}), Alex Slavenko (email) and others}
-\date{2020-12-04}
+\date{2021-01-12}
 
@@
 \CommentTok{## Installing the latest version directly from GitHub}
-\KeywordTok{install_github}\NormalTok{(}\StringTok{"TGuillerme/dads"}\NormalTok{)}
+\NormalTok{devtools}\OperatorTok{::}\KeywordTok{install_github}\NormalTok{(}\StringTok{"TGuillerme/dads"}\NormalTok{)}
 \end{Highlighting}
@@
 ## }
-## <bytecode: 0x7f7eeee57f58>
+## <bytecode: 0x7faff6222240>
 ## <environment: namespace:dads>
@@
 \NormalTok{bd_params <-}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{extinction =} \DecValTok{0}\NormalTok{)}
-\NormalTok{stope_rule <-}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{max.living =} \DecValTok{20}\NormalTok{)}
+\NormalTok{stop_rule <-}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{max.living =} \DecValTok{20}\NormalTok{)}
 
@@
 \NormalTok{default_tree <-}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ bd_params,}
-                     \DataTypeTok{stop.rule =}\NormalTok{ stope_rule,}
+                     \DataTypeTok{stop.rule =}\NormalTok{ stop_rule,}
                      \DataTypeTok{modifiers =}\NormalTok{ default_modifiers)}
@@
 \NormalTok{modified_tree <-}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ bd_params,}
-                      \DataTypeTok{stop.rule =}\NormalTok{ stope_rule,}
+                      \DataTypeTok{stop.rule =}\NormalTok{ stop_rule,}
                       \DataTypeTok{modifiers =}\NormalTok{ constant_modifier)}
@@
 \NormalTok{increasing_tree <-}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ bd_params,}
-                        \DataTypeTok{stop.rule =}\NormalTok{ stope_rule,}
+                        \DataTypeTok{stop.rule =}\NormalTok{ stop_rule,}
                         \DataTypeTok{modifiers =}\NormalTok{ increasing_modifier)}
@@
 \NormalTok{ladderised_tree <-}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ bd_params,}
-                        \DataTypeTok{stop.rule =}\NormalTok{ stope_rule,}
+                        \DataTypeTok{stop.rule =}\NormalTok{ stop_rule,}
                         \DataTypeTok{modifiers =}\NormalTok{ ladderised_modifier)}
@@
 \NormalTok{default_dads <-}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ bd_params,}
-                     \DataTypeTok{stop.rule =}\NormalTok{ stope_rule,}
+                     \DataTypeTok{stop.rule =}\NormalTok{ stop_rule,}
                      \DataTypeTok{traits    =}\NormalTok{ BM_trait)}
@@
 \NormalTok{skewed_trait_dads <-}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ bd_params,}
-                          \DataTypeTok{stop.rule =}\NormalTok{ stope_rule,}
+                          \DataTypeTok{stop.rule =}\NormalTok{ stop_rule,}
                           \DataTypeTok{traits    =}\NormalTok{ BM_trait,}
@@
 \NormalTok{modified_tree <-}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ bd_params,}
-                      \DataTypeTok{stop.rule =}\NormalTok{ stope_rule,}
+                      \DataTypeTok{stop.rule =}\NormalTok{ stop_rule,}
                       \DataTypeTok{modifiers =}\NormalTok{ random_extinction)}
@@
 \NormalTok{biased_trait_dads <-}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ bd_params,}
-                          \DataTypeTok{stop.rule =}\NormalTok{ stope_rule,}
+                          \DataTypeTok{stop.rule =}\NormalTok{ stop_rule,}
                           \DataTypeTok{traits    =}\NormalTok{ BM_trait,}
@@
 
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{bd.params <-}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{speciation =} \DecValTok{1}\NormalTok{, }\DataTypeTok{extinction =} \DecValTok{1}\OperatorTok{/}\DecValTok{3}\NormalTok{)}
+\hypertarget{makeevents}{%
+\chapter{\texorpdfstring{Making events with \texttt{make.events()}}{Making events with make.events()}}\label{makeevents}}
+
+\hypertarget{conditions}{%
+\section{Conditions}\label{conditions}}
+
+\hypertarget{modifications}{%
+\section{Modifications}\label{modifications}}
+
+\hypertarget{some-events-examples}{%
+\section{Some events examples}\label{some-events-examples}}
+
+\hypertarget{random-mass-extinction-after-some-time}{%
+\subsection{Random mass extinction after some time}\label{random-mass-extinction-after-some-time}}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{## Simulation parameters}
+\NormalTok{stop.rule <-}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{max.time =} \DecValTok{5}\NormalTok{)}
+\NormalTok{bd.params <-}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{extinction =} \DecValTok{0}\NormalTok{, }\DataTypeTok{speciation =} \DecValTok{1}\NormalTok{)}
+
+\CommentTok{## 80% mass extinction at time 4}
+\NormalTok{mass_extinction <-}\StringTok{ }\KeywordTok{make.events}\NormalTok{(}
+                      \DataTypeTok{target =} \StringTok{"taxa"}\NormalTok{,}
+                      \DataTypeTok{condition =} \KeywordTok{time.condition}\NormalTok{(}\DecValTok{4}\NormalTok{),}
+                      \DataTypeTok{modification =} \KeywordTok{random.extinction}\NormalTok{(}\FloatTok{0.8}\NormalTok{))}
+
+\CommentTok{## Running the simulations}
+\KeywordTok{set.seed}\NormalTok{(}\DecValTok{1}\NormalTok{)}
+\NormalTok{results <-}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ bd.params,}
+                \DataTypeTok{stop.rule =}\NormalTok{ stop.rule,}
+                \DataTypeTok{events    =}\NormalTok{ mass_extinction)}
+\CommentTok{## Plotting the results}
+\KeywordTok{plot}\NormalTok{(results, }\DataTypeTok{show.tip.label =} \OtherTok{FALSE}\NormalTok{)}
+\KeywordTok{axisPhylo}\NormalTok{()}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-57-1.pdf}
+
+\hypertarget{extinction-based-on-a-trait-value-after-some-time}{%
+\subsection{Extinction based on a trait value after some time}\label{extinction-based-on-a-trait-value-after-some-time}}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{## Simulation parameters}
+\NormalTok{stop.rule <-}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{max.time =} \DecValTok{5}\NormalTok{)}
+\NormalTok{bd.params <-}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{extinction =} \DecValTok{0}\NormalTok{, }\DataTypeTok{speciation =} \DecValTok{1}\NormalTok{)}
+
+\CommentTok{## Extinction of any tips with trait < 1 at time 4}
+\NormalTok{trait_extinction <-}\StringTok{ }\KeywordTok{make.events}\NormalTok{(}
+                      \DataTypeTok{target =} \StringTok{"taxa"}\NormalTok{,}
+                      \DataTypeTok{condition =} \KeywordTok{time.condition}\NormalTok{(}\DecValTok{4}\NormalTok{),}
+                      \DataTypeTok{modification =} \KeywordTok{trait.extinction}\NormalTok{(}\DecValTok{1}\NormalTok{))}
+
+\CommentTok{## Running the simulations}
+\KeywordTok{set.seed}\NormalTok{(}\DecValTok{7}\NormalTok{)}
+\NormalTok{results <-}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ bd.params,}
+                \DataTypeTok{stop.rule =}\NormalTok{ stop.rule,}
+                \DataTypeTok{traits    =} \KeywordTok{make.traits}\NormalTok{(),}
+                \DataTypeTok{events    =}\NormalTok{ trait_extinction)}
+\CommentTok{## Plotting the results}
+\KeywordTok{plot}\NormalTok{(results)}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-58-1.pdf}
+
+\hypertarget{adding-a-background-extinction-after-reaching-a-number-of-living-taxa}{%
+\subsection{Adding a background extinction after reaching a number of living taxa}\label{adding-a-background-extinction-after-reaching-a-number-of-living-taxa}}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{## Simulation parameters}
+\NormalTok{stop.rule <-}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{max.living =} \DecValTok{50}\NormalTok{)}
+\NormalTok{bd.params <-}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{extinction =} \DecValTok{0}\NormalTok{, }\DataTypeTok{speciation =} \DecValTok{1}\NormalTok{)}
+
+\CommentTok{## Adding an extinction parameter after 30 taxa}
+\NormalTok{background_extinction <-}\StringTok{ }\KeywordTok{make.events}\NormalTok{(}
+                      \DataTypeTok{condition    =} \KeywordTok{taxa.condition}\NormalTok{(}\DecValTok{30}\NormalTok{),}
+                      \DataTypeTok{target       =} \StringTok{"bd.params"}\NormalTok{,}
+                      \DataTypeTok{modification =} \KeywordTok{update.bd.params}\NormalTok{(}\DecValTok{1}\OperatorTok{/}\DecValTok{3}\NormalTok{, }\StringTok{"extinction"}\NormalTok{))}
+
+\CommentTok{## Running the simulations}
+\KeywordTok{set.seed}\NormalTok{(}\DecValTok{2}\NormalTok{)}
+\NormalTok{results <-}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ bd.params,}
+                \DataTypeTok{stop.rule =}\NormalTok{ stop.rule,}
+                \DataTypeTok{events    =}\NormalTok{ background_extinction)}
+\CommentTok{## Plotting the results}
+\KeywordTok{plot}\NormalTok{(results)}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-59-1.pdf}
+
+\hypertarget{reducing-speciation-after-some-time}{%
+\subsection{Reducing speciation after some time}\label{reducing-speciation-after-some-time}}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{## Simulation parameters}
+\NormalTok{stop.rule <-}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{max.time =} \DecValTok{4}\NormalTok{)}
+\NormalTok{bd.params <-}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{extinction =} \DecValTok{0}\NormalTok{, }\DataTypeTok{speciation =} \DecValTok{1}\NormalTok{)}
+
+\CommentTok{## Reducing speciation after reaching time 2}
+\NormalTok{reduced_speciation <-}\StringTok{ }\KeywordTok{make.events}\NormalTok{(}
+                      \DataTypeTok{condition    =} \KeywordTok{time.condition}\NormalTok{(}\DecValTok{2}\NormalTok{),}
+                      \DataTypeTok{target       =} \StringTok{"bd.params"}\NormalTok{,}
+                      \DataTypeTok{modification =} \KeywordTok{update.bd.params}\NormalTok{(}\DecValTok{1}\OperatorTok{/}\DecValTok{3}\NormalTok{, }\StringTok{"speciation"}\NormalTok{))}
+
+\KeywordTok{set.seed}\NormalTok{(}\DecValTok{42}\NormalTok{)}
+\NormalTok{no_event <-}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ bd.params,}
+                 \DataTypeTok{stop.rule =}\NormalTok{ stop.rule)}
+\KeywordTok{set.seed}\NormalTok{(}\DecValTok{42}\NormalTok{)}
+\NormalTok{reduced_speciation_event <-}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ bd.params,}
+                                 \DataTypeTok{stop.rule =}\NormalTok{ stop.rule,}
+                                 \DataTypeTok{events    =}\NormalTok{ reduced_speciation)}
+
+\CommentTok{## Plot both trees}
+\KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow =} \KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{, }\DecValTok{2}\NormalTok{))}
+\KeywordTok{plot}\NormalTok{(no_event, }\DataTypeTok{main =} \StringTok{"No event"}\NormalTok{)}
+\KeywordTok{axisPhylo}\NormalTok{()}
+\KeywordTok{plot}\NormalTok{(reduced_speciation_event, }\DataTypeTok{main =} \StringTok{"Reduced speciation after time 2"}\NormalTok{)}
+\KeywordTok{axisPhylo}\NormalTok{()}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-60-1.pdf}
+
+\hypertarget{changing-trait-process-after-some-time}{%
+\subsection{Changing trait process after some time}\label{changing-trait-process-after-some-time}}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{## Simulation parameters}
+\NormalTok{bd.params <-}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{extinction =} \DecValTok{0}\NormalTok{, }\DataTypeTok{speciation =} \DecValTok{1}\NormalTok{)}
+\NormalTok{stop.rule <-}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{max.time =} \DecValTok{6}\NormalTok{)}
 \NormalTok{traits    <-}\StringTok{ }\KeywordTok{make.traits}\NormalTok{()}
-\NormalTok{stop.rule <-}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{max.taxa =} \DecValTok{20}\NormalTok{, }\DataTypeTok{max.living =} \OtherTok{Inf}\NormalTok{, }\DataTypeTok{max.time =} \OtherTok{Inf}\NormalTok{)}
-\NormalTok{modifiers <-}\StringTok{ }\OtherTok{NULL}
-\NormalTok{events <-}\StringTok{ }\OtherTok{NULL}
-\NormalTok{null.error <-}\StringTok{ }\OtherTok{NULL}
-
-\CommentTok{## modifiers}
-\NormalTok{condition <-}\StringTok{ }\ControlFlowTok{function}\NormalTok{(trait.values, parent.lineage) }\KeywordTok{return}\NormalTok{(}\KeywordTok{parent.traits}\NormalTok{(trait.values, parent.lineage) }\OperatorTok{<}\StringTok{ }\DecValTok{0}\NormalTok{)}
-\NormalTok{modify <-}\StringTok{ }\ControlFlowTok{function}\NormalTok{(x) }\KeywordTok{return}\NormalTok{(x }\OperatorTok{*}\StringTok{ }\DecValTok{20}\NormalTok{)}
-
-\CommentTok{## Setting up the different modifiers}
-\NormalTok{modify_speciation <-}\StringTok{ }\KeywordTok{make.modifiers}\NormalTok{(}\DataTypeTok{speciation    =}\NormalTok{ speciation.trait,}
-                                    \DataTypeTok{condition     =}\NormalTok{ condition,}
-                                    \DataTypeTok{modify        =}\NormalTok{ modify)}
-\NormalTok{modify_brlen <-}\StringTok{ }\KeywordTok{make.modifiers}\NormalTok{(}\DataTypeTok{branch.length =}\NormalTok{ branch.length.trait,}
-                               \DataTypeTok{condition     =}\NormalTok{ condition,}
-                               \DataTypeTok{modify        =}\NormalTok{ modify)}
-\NormalTok{modify_speciation_brlen <-}\StringTok{ }\KeywordTok{make.modifiers}\NormalTok{(}\DataTypeTok{branch.length =}\NormalTok{ branch.length.trait,}
-                                          \DataTypeTok{speciation    =}\NormalTok{ speciation.trait,}
-                                          \DataTypeTok{condition     =}\NormalTok{ condition,}
-                                          \DataTypeTok{modify        =}\NormalTok{ modify)}
-
-\CommentTok{## Test normal (no modifiers)}
+
+\CommentTok{## Create an event to change the trait process}
+\NormalTok{change_process <-}\StringTok{ }\KeywordTok{make.events}\NormalTok{(}
+                  \DataTypeTok{condition    =} \KeywordTok{time.condition}\NormalTok{(}\DecValTok{5}\NormalTok{),}
+                  \DataTypeTok{target       =} \StringTok{"traits"}\NormalTok{,}
+                  \DataTypeTok{modification =} \KeywordTok{update.traits}\NormalTok{(}\DataTypeTok{process =}\NormalTok{ OU.process))}
+
+\CommentTok{## Run the simulations}
 \KeywordTok{set.seed}\NormalTok{(}\DecValTok{1}\NormalTok{)}
-\NormalTok{test <-}\StringTok{ }\KeywordTok{dads}\NormalTok{(bd.params, stop.rule, traits, }\DataTypeTok{null.error =} \DecValTok{20}\NormalTok{)}
-\KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow =} \KeywordTok{c}\NormalTok{(}\DecValTok{4}\NormalTok{,}\DecValTok{2}\NormalTok{))}
-\KeywordTok{plot}\NormalTok{(test}\OperatorTok{$}\NormalTok{tree)}
-\KeywordTok{plot.dads}\NormalTok{(test, }\DataTypeTok{main =} \StringTok{"random tree + trait"}\NormalTok{)}
-
-\CommentTok{## Test with modifiers}
+\NormalTok{no_change <-}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ bd.params,}
+                  \DataTypeTok{stop.rule =}\NormalTok{ stop.rule,}
+                  \DataTypeTok{traits    =}\NormalTok{ traits)}
 \KeywordTok{set.seed}\NormalTok{(}\DecValTok{1}\NormalTok{)}
-\NormalTok{trait_table <-}\StringTok{ }\OtherTok{NULL}
-\NormalTok{test <-}\StringTok{ }\KeywordTok{dads}\NormalTok{(bd.params, stop.rule, traits,}
-                                \DataTypeTok{modifiers =}\NormalTok{ modify_speciation,}
-                                \DataTypeTok{null.error =} \DecValTok{20}\NormalTok{)}
-\KeywordTok{plot}\NormalTok{(test}\OperatorTok{$}\NormalTok{tree)}
-\KeywordTok{plot.dads}\NormalTok{(test, }\DataTypeTok{main =} \StringTok{"Skewed speciation"}\NormalTok{)}
-
-\KeywordTok{set.seed}\NormalTok{(}\DecValTok{1}\NormalTok{)}
-\NormalTok{trait_table <-}\StringTok{ }\OtherTok{NULL}
-\NormalTok{test <-}\StringTok{ }\KeywordTok{dads}\NormalTok{(bd.params, stop.rule, traits,}
-                                \DataTypeTok{modifiers =}\NormalTok{ modify_brlen,}
-                                \DataTypeTok{null.error =} \DecValTok{20}\NormalTok{)}
-\KeywordTok{plot}\NormalTok{(test}\OperatorTok{$}\NormalTok{tree)}
-\KeywordTok{plot.dads}\NormalTok{(test, }\DataTypeTok{main =} \StringTok{"Skewed branch length"}\NormalTok{)}
-
-\KeywordTok{set.seed}\NormalTok{(}\DecValTok{1}\NormalTok{)}
-\NormalTok{trait_table <-}\StringTok{ }\OtherTok{NULL}
-\NormalTok{test <-}\StringTok{ }\KeywordTok{dads}\NormalTok{(bd.params, stop.rule, traits,}
-                                \DataTypeTok{modifiers =}\NormalTok{ modify_speciation_brlen,}
-                                \DataTypeTok{null.error =} \DecValTok{20}\NormalTok{)}
-\KeywordTok{plot}\NormalTok{(test}\OperatorTok{$}\NormalTok{tree)}
-\KeywordTok{plot.dads}\NormalTok{(test, }\DataTypeTok{main =} \StringTok{"Skewed branch length and speciation"}\NormalTok{)}
+\NormalTok{process_change <-}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ bd.params,}
+                       \DataTypeTok{stop.rule =}\NormalTok{ stop.rule,}
+                       \DataTypeTok{traits    =}\NormalTok{ traits,}
+                       \DataTypeTok{events    =}\NormalTok{ change_process)}
+\CommentTok{## Plot the results}
+\KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow =} \KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{,}\DecValTok{2}\NormalTok{))}
+\KeywordTok{plot}\NormalTok{(no_change, }\DataTypeTok{ylim =} \KeywordTok{c}\NormalTok{(}\OperatorTok{-}\DecValTok{7}\NormalTok{, }\DecValTok{7}\NormalTok{))}
+\KeywordTok{plot}\NormalTok{(process_change, }\DataTypeTok{ylim =} \KeywordTok{c}\NormalTok{(}\OperatorTok{-}\DecValTok{7}\NormalTok{, }\DecValTok{7}\NormalTok{))}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-61-1.pdf}
+
+\hypertarget{changing-trait-correlation-after-reaching-a-trait-value}{%
+\subsection{Changing trait correlation after reaching a trait value}\label{changing-trait-correlation-after-reaching-a-trait-value}}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{## Set the parameters}
+\NormalTok{stop.rule <-}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{max.taxa =} \DecValTok{100}\NormalTok{)}
+\NormalTok{bd.params <-}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{extinction =} \DecValTok{0}\NormalTok{, }\DataTypeTok{speciation =} \DecValTok{1}\NormalTok{)}
+
+\CommentTok{## A correlated trait}
+\NormalTok{traits <-}\StringTok{ }\KeywordTok{make.traits}\NormalTok{(}\DataTypeTok{n =} \DecValTok{2}\NormalTok{, }\DataTypeTok{process.args =} \KeywordTok{list}\NormalTok{(}\DataTypeTok{Sigma =} \KeywordTok{matrix}\NormalTok{(}\DecValTok{1}\NormalTok{, }\DecValTok{2}\NormalTok{, }\DecValTok{2}\NormalTok{)))}
+
+\CommentTok{## Event changing a trait correlation}
+\NormalTok{change_correlation <-}\StringTok{ }\KeywordTok{make.events}\NormalTok{(}
+    \DataTypeTok{condition    =} \KeywordTok{trait.condition}\NormalTok{(}\DecValTok{3}\NormalTok{, }\DataTypeTok{absolute =} \OtherTok{TRUE}\NormalTok{),}
+    \DataTypeTok{target       =} \StringTok{"traits"}\NormalTok{,}
+    \DataTypeTok{modification =} \KeywordTok{update.traits}\NormalTok{(}\DataTypeTok{process.args =} \KeywordTok{list}\NormalTok{(}\DataTypeTok{Sigma =} \KeywordTok{matrix}\NormalTok{(}\KeywordTok{c}\NormalTok{(}\DecValTok{10}\NormalTok{,}\DecValTok{3}\NormalTok{,}\DecValTok{3}\NormalTok{,}\DecValTok{2}\NormalTok{),}\DecValTok{2}\NormalTok{,}\DecValTok{2}\NormalTok{))))}
+
+\CommentTok{## Run the simulations}
+\KeywordTok{set.seed}\NormalTok{(}\DecValTok{8}\NormalTok{)}
+\NormalTok{no_event <-}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ bd.params,}
+                 \DataTypeTok{stop.rule =}\NormalTok{ stop.rule,}
+                 \DataTypeTok{traits    =}\NormalTok{ traits)}
+\KeywordTok{set.seed}\NormalTok{(}\DecValTok{8}\NormalTok{)}
+\NormalTok{change_correlation <-}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ bd.params,}
+                           \DataTypeTok{stop.rule =}\NormalTok{ stop.rule,}
+                           \DataTypeTok{traits    =}\NormalTok{ traits,}
+                           \DataTypeTok{events    =}\NormalTok{ change_correlation)}
+\CommentTok{## Visual testing}
+\KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow =} \KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{,}\DecValTok{2}\NormalTok{))}
+\KeywordTok{plot}\NormalTok{(no_event, }\DataTypeTok{trait =} \DecValTok{2}\NormalTok{)}
+\KeywordTok{plot}\NormalTok{(change_correlation, }\DataTypeTok{trait =} \DecValTok{2}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-62-1.pdf}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{# plot(no_event, trait = c(1:2))}
+\CommentTok{# plot(change_correlation, trait = c(1:2), use.3D = TRUE)}
+\end{Highlighting}
+\end{Shaded}
+
+\hypertarget{event-for-changing-a-modifier}{%
+\subsection{Event for changing a modifier}\label{event-for-changing-a-modifier}}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{## Set the parameters}
+\NormalTok{stop.rule <-}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{max.time =} \DecValTok{4}\NormalTok{)}
+\NormalTok{bd.params <-}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{extinction =} \DecValTok{0}\NormalTok{, }\DataTypeTok{speciation =} \DecValTok{1}\NormalTok{)}
+
+\CommentTok{## Using a default trait and modifier object}
+\NormalTok{modifiers <-}\StringTok{ }\KeywordTok{make.modifiers}\NormalTok{()}
+\NormalTok{traits <-}\StringTok{ }\KeywordTok{make.traits}\NormalTok{()}
+
+\CommentTok{## New condition and new modifier (increasing speciation if trait is negative)}
+\NormalTok{new.condition <-}\StringTok{ }\ControlFlowTok{function}\NormalTok{(trait.values, lineage) \{}
+    \KeywordTok{return}\NormalTok{(}\KeywordTok{parent.traits}\NormalTok{(trait.values, lineage) }\OperatorTok{<}\StringTok{ }\DecValTok{0}\NormalTok{)}
+\NormalTok{\}}
+\NormalTok{new.modify    <-}\StringTok{ }\ControlFlowTok{function}\NormalTok{(x, trait.values, lineage) \{}
+  \KeywordTok{return}\NormalTok{(x }\OperatorTok{+}\StringTok{ }\DecValTok{1}\NormalTok{)}
+\NormalTok{\}}
+
+\CommentTok{## Update the modifier}
+\NormalTok{change_speciation <-}\StringTok{ }\KeywordTok{make.events}\NormalTok{(}
+    \DataTypeTok{condition    =} \KeywordTok{time.condition}\NormalTok{(}\DecValTok{3}\NormalTok{),}
+    \DataTypeTok{target       =} \StringTok{"modifiers"}\NormalTok{,}
+    \DataTypeTok{modification =} \KeywordTok{update.modifiers}\NormalTok{(}\DataTypeTok{speciation =}\NormalTok{ speciation,}
+                                    \DataTypeTok{condition  =}\NormalTok{ new.condition,}
+                                    \DataTypeTok{modify     =}\NormalTok{ new.modify))}
+
+\KeywordTok{set.seed}\NormalTok{(}\DecValTok{4}\NormalTok{)}
+\NormalTok{no_event <-}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ bd.params,}
+                 \DataTypeTok{stop.rule =}\NormalTok{ stop.rule,}
+                 \DataTypeTok{traits    =}\NormalTok{ traits,}
+                 \DataTypeTok{modifiers =}\NormalTok{ modifiers)}
+\KeywordTok{set.seed}\NormalTok{(}\DecValTok{4}\NormalTok{)}
+\NormalTok{change_spec <-}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ bd.params,}
+                    \DataTypeTok{stop.rule =}\NormalTok{ stop.rule,}
+                    \DataTypeTok{traits    =}\NormalTok{ traits,}
+                    \DataTypeTok{modifiers =}\NormalTok{ modifiers,}
+                    \DataTypeTok{events    =}\NormalTok{ change_speciation)}
+
+\CommentTok{## Visualise the results}
+\KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow =} \KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{,}\DecValTok{2}\NormalTok{))}
+\KeywordTok{plot}\NormalTok{(no_event, }\DataTypeTok{main =} \StringTok{"No event"}\NormalTok{)}
+\KeywordTok{plot}\NormalTok{(change_spec, }\DataTypeTok{main =} \StringTok{"Increase extinction for negative traits}\CharTok{\textbackslash{}n}\StringTok{after time 3"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-63-1.pdf}
+
+\hypertarget{changing-branch-length-when-reaching-n-taxa}{%
+\subsection{Changing branch length when reaching n taxa}\label{changing-branch-length-when-reaching-n-taxa}}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{## Parameters}
+\NormalTok{bd.params <-}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{extinction =} \DecValTok{0}\NormalTok{, }\DataTypeTok{speciation =} \DecValTok{1}\NormalTok{)}
+\NormalTok{stop.rule <-}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{max.taxa =} \DecValTok{100}\NormalTok{)}
+\NormalTok{modifiers <-}\StringTok{ }\KeywordTok{make.modifiers}\NormalTok{()}
+\NormalTok{traits <-}\StringTok{ }\KeywordTok{make.traits}\NormalTok{()}
+
+\CommentTok{## multiplying branch length 100 folds}
+\NormalTok{new.modify <-}\StringTok{ }\ControlFlowTok{function}\NormalTok{(x, trait.values, lineage) \{}
+  \KeywordTok{return}\NormalTok{(x }\OperatorTok{*}\StringTok{ }\DecValTok{100}\NormalTok{)}
+\NormalTok{\}}
+
+\CommentTok{## Event for increasing branch length after reaching 30 taxa}
+\NormalTok{increase_brlen <-}\StringTok{ }\KeywordTok{make.events}\NormalTok{(}
+                  \DataTypeTok{condition    =} \KeywordTok{taxa.condition}\NormalTok{(}\DecValTok{30}\NormalTok{),}
+                  \DataTypeTok{target       =} \StringTok{"modifiers"}\NormalTok{,}
+                  \DataTypeTok{modification =} \KeywordTok{update.modifiers}\NormalTok{(}
+                                    \DataTypeTok{branch.length =}\NormalTok{ branch.length,}
+                                    \DataTypeTok{modify =}\NormalTok{ new.modify))}
+
+\CommentTok{## Run the simulations}
+\KeywordTok{set.seed}\NormalTok{(}\DecValTok{5}\NormalTok{)}
+\NormalTok{no_event <-}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ bd.params,}
+                 \DataTypeTok{stop.rule =}\NormalTok{ stop.rule,}
+                 \DataTypeTok{traits    =}\NormalTok{ traits,}
+                 \DataTypeTok{modifiers =}\NormalTok{ modifiers)}
+\KeywordTok{set.seed}\NormalTok{(}\DecValTok{5}\NormalTok{)}
+\NormalTok{increased_brlen <-}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ bd.params,}
+                        \DataTypeTok{stop.rule =}\NormalTok{ stop.rule,}
+                        \DataTypeTok{traits    =}\NormalTok{ traits,}
+                        \DataTypeTok{modifiers =}\NormalTok{ modifiers,}
+                        \DataTypeTok{events    =}\NormalTok{ increase_brlen)}
+
+\CommentTok{## Visualise the results}
+\KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow =} \KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{,}\DecValTok{2}\NormalTok{))}
+\KeywordTok{plot}\NormalTok{(no_event, }\DataTypeTok{main =} \StringTok{"No event"}\NormalTok{)}
+\KeywordTok{plot}\NormalTok{(increased_brlen, }\DataTypeTok{main =} \StringTok{"Increase branch length}\CharTok{\textbackslash{}n}\StringTok{after 30 taxa"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-64-1.pdf}
+
+\hypertarget{founding-event-a-subtree-with-no-fossils}{%
+\subsection{Founding event: a subtree with no fossils}\label{founding-event-a-subtree-with-no-fossils}}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{## Set up parameters}
+\NormalTok{stop.rule <-}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{max.time =} \DecValTok{4}\NormalTok{)}
+\NormalTok{bd.params <-}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{speciation =} \DecValTok{1}\NormalTok{, }\DataTypeTok{extinction =} \FloatTok{0.3}\NormalTok{)}
+
+\CommentTok{## Events that generate a new process (founding effects)}
+\NormalTok{founding_event <-}\StringTok{ }\KeywordTok{make.events}\NormalTok{(}
+                  \DataTypeTok{condition    =} \KeywordTok{taxa.condition}\NormalTok{(}\DecValTok{10}\NormalTok{),}
+                  \DataTypeTok{target       =} \StringTok{"founding"}\NormalTok{,}
+                  \DataTypeTok{modification =} \KeywordTok{founding.event}\NormalTok{(}
+                                    \DataTypeTok{bd.params =} \KeywordTok{list}\NormalTok{(}\DataTypeTok{speciation =} \DecValTok{2}\NormalTok{,}
+                                                     \DataTypeTok{extinction =} \DecValTok{0}\NormalTok{)),}
+                  \DataTypeTok{additional.args =} \KeywordTok{list}\NormalTok{(}\DataTypeTok{prefix =} \StringTok{"founding_"}\NormalTok{))}
+
+\CommentTok{## Simulations}
+\KeywordTok{set.seed}\NormalTok{(}\DecValTok{11}\NormalTok{)}
+\NormalTok{founding_tree <-}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ bd.params,}
+                      \DataTypeTok{stop.rule =}\NormalTok{ stop.rule,}
+                      \DataTypeTok{events    =}\NormalTok{ founding_event)}
+\KeywordTok{plot}\NormalTok{(founding_tree, }\DataTypeTok{cex =} \FloatTok{0.4}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-65-1.pdf}
+
+\hypertarget{founding-event-with-traits}{%
+\subsection{Founding event with traits}\label{founding-event-with-traits}}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{## Set up parameters}
+\NormalTok{stop.rule <-}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{max.time =} \DecValTok{4}\NormalTok{)}
+\NormalTok{bd.params <-}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{speciation =} \DecValTok{1}\NormalTok{, }\DataTypeTok{extinction =} \FloatTok{0.3}\NormalTok{)}
+\NormalTok{traits    <-}\StringTok{ }\KeywordTok{make.traits}\NormalTok{()}
+
+\CommentTok{## Events that generate a new process (founding effects)}
+\NormalTok{founding_event <-}\StringTok{ }\KeywordTok{make.events}\NormalTok{(}
+                  \DataTypeTok{condition    =} \KeywordTok{taxa.condition}\NormalTok{(}\DecValTok{10}\NormalTok{),}
+                  \DataTypeTok{target       =} \StringTok{"founding"}\NormalTok{,}
+                  \DataTypeTok{modification =} \KeywordTok{founding.event}\NormalTok{(}
+                                    \DataTypeTok{bd.params =} \KeywordTok{list}\NormalTok{(}\DataTypeTok{speciation =} \DecValTok{2}\NormalTok{,}
+                                                     \DataTypeTok{extinction =} \DecValTok{0}\NormalTok{),}
+                                    \DataTypeTok{traits =} \KeywordTok{make.traits}\NormalTok{(}\DataTypeTok{process =}\NormalTok{ OU.process)),}
+                  \DataTypeTok{additional.args =} \KeywordTok{list}\NormalTok{(}\DataTypeTok{prefix =} \StringTok{"founding_"}\NormalTok{))}
+
+\CommentTok{## Simulating the tree}
+\KeywordTok{set.seed}\NormalTok{(}\DecValTok{19}\NormalTok{)}
+\NormalTok{founding_tree <-}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ bd.params,}
+                      \DataTypeTok{stop.rule =}\NormalTok{ stop.rule,}
+                      \DataTypeTok{traits    =} \KeywordTok{make.traits}\NormalTok{(),}
+                      \DataTypeTok{events    =}\NormalTok{ founding_event)}
+\KeywordTok{plot}\NormalTok{(founding_tree)}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-66-1.pdf}
+
+\hypertarget{thorough-example}{%
+\chapter{Thorough example}\label{thorough-example}}
+
+Here we are going to look at an example on how to simulate complex evolutionary histories with \texttt{dads}.
+
+Let say we want to simulate the following story:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  A lineage evolves in a normal birth-death fashion (with extinction) from a single origin and with two traits called:
+
+  \begin{itemize}
+  \tightlist
+  \item
+    ``Distance'': the physical distance between species in an arbitrary 1D space (evolving in a Brownian motion).
+  \item
+    ``Variable'': a 2D uncorrelated Ornstein-Uhlenbeck trait that corresponds to some traits the lineages have.
+    During the whole process, lineages that are more further away from the other species in terms of distance have a higher change of speciating (proportional to that distance).
+  \end{itemize}
+\end{enumerate}
+
+Three events then happen during the evolutionary history of this lineage:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\setcounter{enumi}{1}
+\tightlist
+\item
+  ``Correlation event'': after reaching a certain number of taxa an event happens making the ``Variables'' trait dimensions become correlated.
+\item
+  ``Selective extinction event'': after reaching a certain point in time, lineages with their ``Variables'' values that are negative go extinct
+\item
+  ``Founding event'': after reaching a value for the ``Distance'' trait, a founding event happens with a new birth-death process with no extinction and where the ``Variable'' is now a 2D correlated Brownian Motion.
+\item
+  ``Mass extinction event'': after reaching a certain number of taxa in the ``Founding event'', 80\% of taxa go extinct randomly.
+\end{enumerate}
+
+We are going to run each step one by one in a cumulative manner to see how they pile up:
+
+\hypertarget{step-1}{%
+\section{Step 1:}\label{step-1}}
+
+\hypertarget{the-birth-death-parameters}{%
+\subsection{The birth death parameters}\label{the-birth-death-parameters}}
+
+First let's set the overall birth death parameters (a speciation of 1.2 and an extinction of 0.3):
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{## The birth death parameters}
+\NormalTok{bd_params <-}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{speciation =} \FloatTok{1.2}\NormalTok{, }\DataTypeTok{extinction =} \FloatTok{0.3}\NormalTok{)}
+\CommentTok{## The stopping rule}
+\NormalTok{stop_rule <-}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{max.time =} \DecValTok{4}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+And let's quickly test if the simulation works by running a birth death tree without traits and anything:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{set.seed}\NormalTok{(}\DecValTok{0}\NormalTok{) }\CommentTok{# 123 also works}
+\CommentTok{## Running a birth death tree }
+\NormalTok{results <-}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ bd_params,}
+                \DataTypeTok{stop.rule =}\NormalTok{ stop_rule)}
+
+\CommentTok{## And displaying the results}
+\KeywordTok{plot}\NormalTok{(results)}
+\KeywordTok{axisPhylo}\NormalTok{()}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-68-1.pdf}
+
+\hypertarget{the-traits}{%
+\subsection{The traits}\label{the-traits}}
+
+Then let's create the traits:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{## The trait object}
+\NormalTok{two_traits <-}\StringTok{ }\KeywordTok{make.traits}\NormalTok{(}\DataTypeTok{process =} \KeywordTok{c}\NormalTok{(BM.process, OU.process), }\DataTypeTok{n =} \KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{, }\DecValTok{2}\NormalTok{),}
+                          \DataTypeTok{trait.names =} \KeywordTok{c}\NormalTok{(}\StringTok{"Distance"}\NormalTok{, }\StringTok{"Variables"}\NormalTok{))}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{## Visualise the 1D trait (BM.process)}
+\KeywordTok{plot}\NormalTok{(two_traits, }\DataTypeTok{trait =} \DecValTok{1}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-70-1.pdf}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{## Visualise the 2D trait (uncorrelated OU.process)}
+\KeywordTok{plot}\NormalTok{(two_traits, }\DataTypeTok{trait =} \DecValTok{2}\NormalTok{, }\DataTypeTok{use.3D =} \OtherTok{TRUE}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+Note that specifying the correlation matrix is not necessary here \emph{per se} (since no correlation is the default) but we will modify it through an event later on so it's good practice to have it properly stated before hand (to avoid any confusion).
+
+We can then run the tree with the traits:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{set.seed}\NormalTok{(}\DecValTok{0}\NormalTok{)}
+\CommentTok{## Running a birth death tree }
+\NormalTok{results <-}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params  =}\NormalTok{ bd_params,}
+                \DataTypeTok{stop.rule  =}\NormalTok{ stop_rule,}
+                \DataTypeTok{traits     =}\NormalTok{ two_traits)}
+
+\CommentTok{## And displaying the results}
+\KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow =} \KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{,}\DecValTok{2}\NormalTok{))}
+\KeywordTok{plot}\NormalTok{(results, }\DataTypeTok{trait =} \DecValTok{1}\NormalTok{, }\DataTypeTok{main =} \StringTok{"Trait 1: Distance"}\NormalTok{)}
+\KeywordTok{plot}\NormalTok{(results, }\DataTypeTok{trait =} \KeywordTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{,}\DecValTok{3}\NormalTok{), }\DataTypeTok{main =} \StringTok{"Variables correlation"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-72-1.pdf}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{## Visualising the tree in 3D}
+\KeywordTok{plot}\NormalTok{(results, }\DataTypeTok{trait =} \KeywordTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{,}\DecValTok{3}\NormalTok{), }\DataTypeTok{use.3D =} \OtherTok{TRUE}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\hypertarget{the-modifiers}{%
+\subsection{The modifiers}\label{the-modifiers}}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{## Speciation event is more likely if lineage's ancestor is further away from the mean trait value}
+\NormalTok{distance.modify <-}\StringTok{ }\ControlFlowTok{function}\NormalTok{(x, trait.values, lineage) }\KeywordTok{return}\NormalTok{(x }\OperatorTok{+}\StringTok{ }\NormalTok{x }\OperatorTok{*}\StringTok{ }\KeywordTok{abs}\NormalTok{(}\KeywordTok{parent.traits}\NormalTok{(trait.values, lineage)[}\DecValTok{1}\NormalTok{] }\OperatorTok{-}\StringTok{ }\KeywordTok{mean}\NormalTok{(trait.values[,}\DecValTok{1}\NormalTok{])))}
+
+\CommentTok{## Make a distance modifier}
+\NormalTok{distance.speciation <-}\StringTok{ }\KeywordTok{make.modifiers}\NormalTok{(}\DataTypeTok{speciation =}\NormalTok{ speciation,}
+                                      \DataTypeTok{modify =}\NormalTok{ distance.modify)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{set.seed}\NormalTok{(}\DecValTok{42}\NormalTok{)}
+\CommentTok{## Running a birth death tree }
+\NormalTok{results <-}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params  =}\NormalTok{ bd_params,}
+                \DataTypeTok{stop.rule  =}\NormalTok{ stop_rule,}
+                \DataTypeTok{traits     =}\NormalTok{ two_traits,}
+                \DataTypeTok{modifiers  =}\NormalTok{ distance.speciation)}
+
+\CommentTok{## And displaying the results}
+\KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow =} \KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{,}\DecValTok{2}\NormalTok{))}
+\KeywordTok{plot}\NormalTok{(results, }\DataTypeTok{trait =} \DecValTok{1}\NormalTok{, }\DataTypeTok{main =} \StringTok{"Trait 1: Distance"}\NormalTok{)}
+\KeywordTok{plot}\NormalTok{(results, }\DataTypeTok{trait =} \KeywordTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{,}\DecValTok{3}\NormalTok{), }\DataTypeTok{main =} \StringTok{"Variables correlation"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-75-1.pdf}
+
+\hypertarget{step-2}{%
+\section{Step 2:}\label{step-2}}
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\setcounter{enumi}{1}
+\tightlist
+\item
+  ``Correlation event'': after reaching a certain number of taxa an event happens making the ``Variable'' trait dimensions become correlated.
+\end{enumerate}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{## The correlation change event}
+\NormalTok{events_list <-}\StringTok{ }\KeywordTok{make.events}\NormalTok{(}\DataTypeTok{event.name   =} \StringTok{"Correlation"}\NormalTok{,}
+                           \DataTypeTok{target       =} \StringTok{"traits"}\NormalTok{,}
+                           \DataTypeTok{condition    =} \KeywordTok{taxa.condition}\NormalTok{(}\DecValTok{30}\NormalTok{),}
+                           \DataTypeTok{modification =} \KeywordTok{update.traits}\NormalTok{(}
+                    \DataTypeTok{process.args =} \KeywordTok{list}\NormalTok{(}\DataTypeTok{Sigma =} \KeywordTok{matrix}\NormalTok{(}\KeywordTok{c}\NormalTok{(}\FloatTok{0.7}\NormalTok{, }\DecValTok{0}\NormalTok{, }\DecValTok{0}\NormalTok{, }\FloatTok{0.7}\NormalTok{), }\DecValTok{2}\NormalTok{, }\DecValTok{2}\NormalTok{)),}
+                    \DataTypeTok{trait.names  =} \StringTok{"Variables"}\NormalTok{))}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{set.seed}\NormalTok{(}\DecValTok{8}\NormalTok{)}
+\CommentTok{## Running a birth death tree }
+\NormalTok{results <-}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params  =}\NormalTok{ bd_params,}
+                \DataTypeTok{stop.rule  =}\NormalTok{ stop_rule,}
+                \DataTypeTok{traits     =}\NormalTok{ two_traits,}
+                \DataTypeTok{modifiers  =}\NormalTok{ distance.speciation,}
+                \DataTypeTok{events     =}\NormalTok{ events_list)}
+
+\CommentTok{## And displaying the results}
+\KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow =} \KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{,}\DecValTok{2}\NormalTok{))}
+\KeywordTok{plot}\NormalTok{(results, }\DataTypeTok{trait =} \DecValTok{1}\NormalTok{, }\DataTypeTok{main =} \StringTok{"Trait 1: Distance"}\NormalTok{)}
+\KeywordTok{plot}\NormalTok{(results, }\DataTypeTok{trait =} \KeywordTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{,}\DecValTok{3}\NormalTok{), }\DataTypeTok{main =} \StringTok{"Variables correlation"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-77-1.pdf}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{## Visualising the tree in 3D}
+\KeywordTok{plot}\NormalTok{(results, }\DataTypeTok{trait =} \KeywordTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{,}\DecValTok{3}\NormalTok{), }\DataTypeTok{use.3D =} \OtherTok{TRUE}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\hypertarget{step-3}{%
+\section{Step 3:}\label{step-3}}
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\setcounter{enumi}{2}
+\tightlist
+\item
+  ``Selective extinction event'': after reaching a certain point in time, lineages with their ``Variables'' values that are negative go extinct
+\end{enumerate}
+
+\hypertarget{bugged-from-here}{%
+\chapter{BUGGED FROM HERE}\label{bugged-from-here}}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{## Adding the selective event}
+\NormalTok{events_list <-}\StringTok{ }\KeywordTok{make.events}\NormalTok{(}\DataTypeTok{event.name   =} \StringTok{"Selection"}\NormalTok{,}
+                           \DataTypeTok{add          =}\NormalTok{ events_list,}
+                           \DataTypeTok{target       =} \StringTok{"taxa"}\NormalTok{,}
+                           \DataTypeTok{condition    =} \KeywordTok{time.condition}\NormalTok{(}\DecValTok{4}\NormalTok{),}
+                           \DataTypeTok{modification =} \KeywordTok{trait.extinction}\NormalTok{(}\DataTypeTok{x =} \DecValTok{0}\NormalTok{, }\DataTypeTok{condition =} \StringTok{`}\DataTypeTok{<}\StringTok{`}\NormalTok{, }\DataTypeTok{trait =} \DecValTok{2}\NormalTok{))}
+\end{Highlighting}
+\end{Shaded}
+
+Problem with \texttt{modification} argument here
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{## Running a birth death tree }
+\NormalTok{results <-}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params  =}\NormalTok{ bd_params,}
+                \DataTypeTok{stop.rule  =}\NormalTok{ stop_rule,}
+                \DataTypeTok{traits     =}\NormalTok{ two_traits,}
+                \DataTypeTok{modifiers  =}\NormalTok{ distance.speciation,}
+                \DataTypeTok{events     =}\NormalTok{ events_list)}
+
+\CommentTok{## And displaying the results}
+\KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow =} \KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{,}\DecValTok{2}\NormalTok{))}
+\KeywordTok{plot}\NormalTok{(results, }\DataTypeTok{trait =} \DecValTok{1}\NormalTok{, }\DataTypeTok{main =} \StringTok{"Trait 1: Distance"}\NormalTok{)}
+\KeywordTok{plot}\NormalTok{(results, }\DataTypeTok{trait =} \KeywordTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{,}\DecValTok{3}\NormalTok{), }\DataTypeTok{main =} \StringTok{"Variables correlation"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\hypertarget{step-4-and-5}{%
+\section{Step 4 and 5:}\label{step-4-and-5}}
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\setcounter{enumi}{3}
+\tightlist
+\item
+  ``Founding event'': after reaching a value for the ``Distance'' trait, a founding event happens with a new birth-death process with no extinction and where the ``Variable'' is now a 2D correlated Brownian Motion.
+\item
+  ``Mass extinction event'': after reaching a certain number of taxa in the ``Founding event'', 80\% of taxa go extinct randomly.
+\end{enumerate}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{## The new trait for the founding event}
+\NormalTok{new_traits <-}\StringTok{ }\KeywordTok{make.traits}\NormalTok{(}\DataTypeTok{process =} \KeywordTok{c}\NormalTok{(BM.process, BM.process), }\DataTypeTok{n =} \KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{, }\DecValTok{2}\NormalTok{),}
+                          \DataTypeTok{trait.names =} \KeywordTok{c}\NormalTok{(}\StringTok{"Distance"}\NormalTok{, }\StringTok{"Variables"}\NormalTok{))}
+
+\CommentTok{## The mass extinction event after reaching 50 taxa}
+\NormalTok{mass_extinction <-}\StringTok{ }\KeywordTok{make.events}\NormalTok{(}\DataTypeTok{target       =} \StringTok{"taxa"}\NormalTok{,}
+                               \DataTypeTok{condition    =} \KeywordTok{taxa.condition}\NormalTok{(}\DecValTok{50}\NormalTok{),}
+                               \DataTypeTok{modification =} \KeywordTok{random.extinction}\NormalTok{(}\FloatTok{0.8}\NormalTok{))}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{## The selective event}
+\NormalTok{events_list <-}\StringTok{ }\KeywordTok{make.events}\NormalTok{(}\DataTypeTok{event.name   =} \StringTok{"Founding"}\NormalTok{, }
+                           \DataTypeTok{add          =}\NormalTok{ events_list,}
+                           \DataTypeTok{target       =} \StringTok{"founding"}\NormalTok{,}
+                           \DataTypeTok{condition    =} \KeywordTok{trait.condition}\NormalTok{(}\DecValTok{3}\NormalTok{),}
+                           \DataTypeTok{modification =} \KeywordTok{founding.event}\NormalTok{(}\DataTypeTok{traits    =}\NormalTok{ new_traits,}
+                                                         \DataTypeTok{events    =}\NormalTok{ mass_extinction,}
+                                                         \DataTypeTok{bd.params =} \KeywordTok{list}\NormalTok{(}
+                                                            \DataTypeTok{speciation =} \DecValTok{2}\NormalTok{,}
+                                                            \DataTypeTok{extinction =} \DecValTok{0}\NormalTok{)),}
+                           \DataTypeTok{additional.args =} \KeywordTok{list}\NormalTok{(}\DataTypeTok{prefix =} \StringTok{"found_"}\NormalTok{))}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{## Running a birth death tree }
+\NormalTok{results <-}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params  =}\NormalTok{ bd_params,}
+                \DataTypeTok{stop.rule  =}\NormalTok{ stop_rule,}
+                \DataTypeTok{traits     =}\NormalTok{ two_traits,}
+                \DataTypeTok{modifiers  =}\NormalTok{ distance.speciation,}
+                \DataTypeTok{events     =}\NormalTok{ events_list,}
+                \DataTypeTok{null.error =} \DecValTok{100}\NormalTok{)}
+
+\CommentTok{## And displaying the results}
+\KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow =} \KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{,}\DecValTok{2}\NormalTok{))}
+\KeywordTok{plot}\NormalTok{(results, }\DataTypeTok{trait =} \DecValTok{1}\NormalTok{, }\DataTypeTok{main =} \StringTok{"Trait 1: Distance"}\NormalTok{)}
+\KeywordTok{plot}\NormalTok{(results, }\DataTypeTok{trait =} \KeywordTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{,}\DecValTok{3}\NormalTok{), }\DataTypeTok{main =} \StringTok{"Variables correlation"}\NormalTok{)}
 \end{Highlighting}

### 2021-01-19 inst/gitbook/04_make.events.Rmd
Added 3D example

@@
 ---
+
+```{r, echo = FALSE}
+library(rgl)
+knitr::knit_hooks$set(webgl = hook_webgl)
+```
 
@@
 ## Event changing a trait correlation
-change_correlation <- make.events(
+cor_change <- make.events(
     condition    = trait.condition(3, absolute = TRUE),
@@
 ## Run the simulations
-set.seed(8)
+set.seed(2)
 no_event <- dads(bd.params = bd.params,
@@
                  traits    = traits)
-set.seed(8)
+set.seed(2)
 change_correlation <- dads(bd.params = bd.params,
@@
                            traits    = traits,
-                           events    = change_correlation)
+                           events    = cor_change)
+
 ## Visual testing
@@
 plot(change_correlation, trait = 2)
-# plot(no_event, trait = c(1:2))
-# plot(change_correlation, trait = c(1:2), use.3D = TRUE)
+
+```
+
+```{r, webgl=TRUE}
+## 3D plot
+plot(change_correlation, trait = c(1:2), use.3D = TRUE)
 ```
@@
 plot(no_event, main = "No event")
-plot(change_spec, main = "Increase extinction for negative traits\nafter time 3")
+plot(change_spec, main = "Increase extinction for negative\ntraits after time 3")
 ```

### 2021-01-19 inst/gitbook/not_include/05_thorough_example.Rmd
Don't include thorough example yet

--- 2021-01-12 inst/gitbook/05_thorough_example.Rmd
@@
 
+```{r, echo = FALSE}
+library(rgl)
+knitr::knit_hooks$set(webgl = hook_webgl)
+```
 # Thorough example
@@
 
- 2. "Correlation event": after reaching a certain number of taxa an event happens making the "Variables" trait dimensions become correlated.
+ 2. "Correlation event": after reaching time 2 an event happens making the "Variables" trait dimensions become correlated.
  3. "Selective extinction event": after reaching a certain point in time, lineages with their "Variables" values that are negative go extinct
@@
 ## The birth death parameters
-bd_params <- list(speciation = 1.2, extinction = 0.3)
+bd_params <- list(speciation = 1.5, extinction = 0.3)
 ## The stopping rule
@@
 
-```{r}
+```{r, webgl=TRUE}
 ## Visualise the 2D trait (uncorrelated OU.process)
@@
 
-```{r}
+```{r, webgl=TRUE}
 ## Visualising the tree in 3D
@@
 ```{r}
-set.seed(42)
+set.seed(13)
 ## Running a birth death tree 
@@
 
+```{r, webgl=TRUE}
+## Visualising the tree in 3D
+plot(results, trait = c(2,3), use.3D = TRUE)
+```
+
 ## Step 2:
- 2. "Correlation event": after reaching a certain number of taxa an event happens making the "Variable" trait dimensions become correlated.
+ 2. "Correlation event": after reaching time 2 an event happens making the "Variable" trait dimensions become correlated.
 
@@
 ## The correlation change event
+update.correlation <- update.traits(process.args =list(Sigma = matrix(c(10,3,3,2),2,2)))
+
 events_list <- make.events(event.name   = "Correlation",
                            target       = "traits",
-                           condition    = taxa.condition(30),
+                           condition    = time.condition(2),
                            modification = update.traits(
-                    process.args = list(Sigma = matrix(c(0.7, 0, 0, 0.7), 2, 2)),
+                    process.args =list(Sigma = matrix(c(10,3,3,2),2,2)),
                     trait.names  = "Variables"))
@@
 ```{r}
-set.seed(8)
+results <- NULL
+set.seed(1)
 ## Running a birth death tree 
@@
 
-```{r, eval = FALSE}
+```{r, webgl=TRUE}
 ## Visualising the tree in 3D
@@
 
-# BUGGED FROM HERE
+> BUGGED FROM HERE!
 

### 2021-01-19 inst/gitbook/dads_manual.Rmd
Updated manual

--- 
title: "dads manual"
author: "Thomas Guillerme (guillert@tcd.ie) and others"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
output: bookdown::gitbook
documentclass: book
bibliography: [../references.bib]
link-citations: yes
github-repo: TGuillerme/dads
description: "dads R package vignette"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# `dads`: disparity and diversity simulations.

Allowing to simulate disparity and diversity at the same time with interaction between both.

The core of this package is based on the [`diversitree` birth-death algorithm](https://github.com/cran/diversitree/blob/master/R/simulate-bd.R).

## What is `dads`?

There are some very good packages out there to simulate birth death trees (like [TreeSim](https://github.com/tanja819/TreeSim)) or even packages simulating traits (disparity) and diversity jointly (like [RPANDA](https://github.com/cran/RPANDA) or [PETER](https://github.com/PuttickMacroevolution/PETER)).
We strongly advice you have a look at these packages first as they might be more appropriate for your tasks.

`dads` aims to be a highly modular and friendly version of all these packages: it allows to simulate disparity and diversity jointly with a vast array of options that can be easily modified by users.
For example, you can easily generate any type of process to generate a trait (BM, OU, something else, etc...) in multiple dependent or independent dimensions through `"traits"` objects.
You can then specify how the traits should affect disparity through `"modifiers"` objects. And finally you can create events (like mass extinctions) through `"events"` objects.
These objects and how to modify them will be detailed throughout this manual.
Finally we are putting an emphasise in the development of this package on the speed and reliability of the functions.

### Modular?

Because their is an infinite way you might want to generate disparity and diversity (different traits, different modifiers and different events), the `dads` package is designed to make all these parts easy to code separately and integrate them easily in the `dads` core functions.
This allows you to simulate finely tuned multidimensional data for your specific project!

## Installing and running the package

You can install this package easily, directly from the github:
<!-- TG: I do intend to release it on CRAN later on -->

```{r, eval = FALSE}
## Checking if devtools is already installed
if(!require(devtools)) install.packages("devtools")

## Installing the latest version directly from GitHub
devtools::install_github("TGuillerme/dads")
```

```{r, eval = FALSE, echo = FALSE, message = FALSE}
## Loading the package and setting up the start seed.
library(dads)
```

## Help

If you need help with the package, hopefully the following manual will be useful.
However, parts of this package are still in development and some other parts are probably not covered.
Thus if you have suggestions or comments on on what has already been developed or will be developed, please send me an email (<guillert@tcd.ie>) or if you are a GitHub user, directly create an issue on the [GitHub page](https://github.com/TGuillerme/dads).
Doing so will not only hopefully help you but also other users since it will help improve this manual!

## How does `dads` work?

Basically, the `dads` function intakes your personalised `traits`, `modifiers` and `events` to generate your disparity and diversity.
You will find more details about how these objects (`traits`, `modifiers` and `events`) work in the rest of the tutorial but here is a graphical representation of how `dads` work:

![Schematised summary of the `dads` package architecture](dads_structure.pdf)

<!--chapter:end:index.Rmd-->

---
title: "Getting start"
author: "Thomas Guillerme (guillert@tcd.ie)"
date: "`r Sys.Date()`"
bibliography: [../References.bib, ../packages.bib]
output:
  html_document: default
  pdf_document: default
---

```{r, echo = FALSE}
library(dads)
set.seed(4242)
```

# Getting started

## The simplest of all analysis: simulating diversity only

One of the simplest things to do with the `dads` package is just to simulate a birth death tree.
For that you can use the function `dads` and specify your stopping rule.
The stopping rule simply tells the birth death process to step whenever it reaches one of these three conditions:

 * `"max.taxa"   = n` stop when `n` taxa are generated;
 * `"max.living" = n` stop when there is `n` co-occuring taxa of the same age (i.e. "living" taxa);
 * `"max.time"   = n` stop when the simulated tree is `n` units of age old (these units are arbitrary);

For example, we might want to generate a birth-death tree with 20 taxa:

```{r}
## Setting a stopping rule to reach a maximum of 20 taxa
my_stop_rule <- list(max.taxa = 20)
```

We can now run the simulations using:

```{r}
## Running the birth death simulation
my_tree <- dads(stop.rule = my_stop_rule)
```

> Note that here we could have specified more than one stopping rule, for example, we might want to run a simulation and stop it if it either reaches 10 taxa or the age 2 using `stop.rule = list(max.time = 2, max.taxa = 10)`. The simulation will then stop when either of these conditions are met.

The resulting object is a classic `"phylo"` object that you can simply plot or visualise like so:

```{r}
## The tree object
my_tree
## Plotting it
plot(my_tree)
```

### Changing the birth-death parameters

People familiar with the [birth-death models](https://lukejharmon.github.io/pcm/chapter10_birthdeath/) might have noticed that we did not specify two important things here: the speciation parameter (sometimes called "lambda" or "birth") and the extinction parameter (sometimes called "mu", "death" or "background extinction").
By default `dads` runs a pure birth model (the speciation is set to 1 and the extinction to 0).
However, you can easily change that by specifying your new birth death parameters:

```{r}
## my birth death parameters
my_params <- list(speciation = 1,
                  extinction = 1/3)
```
> Note that here it is not necessary to specify `extinction = 1` since this is the default option, you can always just change the parameter of interest (e.g. changing `extinciton = 0` to `extinction = 1/3`). However, we think it's good practice to attribute both parameters specifically to avoid any confusion.

You can then run the same birth death tree with extinction:

```{r}
## Generating a birth death tree with extinctions:
my_tree <- dads(bd.params = my_params, stop.rule = my_stop_rule)
## Visualising the new tree
plot(my_tree)
```

## Slightly more complex: simulating disparity and diversity

Chances are that you want to also simulate traits (disparity) along with your diversity (otherwise, we suggest using the [`TreeSim`](https://github.com/tanja819/TreeSim/) package that provides many more birth death models).
Simulating traits is not much more complicated in `dads`: you'll simply need to create a `"traits"` object using the `make.traits` function.
These objects can have increasing complexity (see the rest of this tutorial) but we will keep it simple here.

`"traits"` objects contain one or more processes which are the ways to generate the trait.
The most common of these processes is the [Brownian Motion](https://en.wikipedia.org/wiki/Brownian_motion).
This is used by default with the `make.traits` function:

```{r}
## Creating the traits object
my_trait <- make.traits()
```

This trait object can be simply printed (to see what's in it) or plotted (to see what the process looks like in the absence of a phylogeny):

```{r}
## Which process is in here?
my_trait
## What does it look like?
plot(my_trait)
```

By default, this trait is called "A".
This is not a really good name but you'll see more about specifying trait names later on.
If this is what the process should look like (theoretically) you can then add its `"traits"` object to our previous `dads` function to generate the tree and the traits:

```{r}
## Simulate disparity and diversity
my_data <- dads(bd.params = my_params,
                stop.rule = my_stop_rule,
                traits    = my_trait)
```

Et voilà! We now have a simple disparity and diversity simulation.
We can see what's in the results by simply printing it or plotting it:

```{r}
## What's in there
my_data
## Plotting the disparity and diversity
plot(my_data)
```

You can then extract the components you need for your specific analysis like so:

```{r}
## Extracting the tree (a "phylo" object)
the_generated_tree <- my_data$tree
class(the_generated_tree)

## Extracting the data (a "matrix")
the_generated_data <- my_data$data
class(the_generated_data)
```

You can find much more on how to design trait objects in the [`make.traits` section](#maketraits).

## Slightly more more complex: simulating linked disparity and diversity

The example above is also still pretty simple and easily done through a variety of `R` packages: here the trait and the tree are simulated at the same time but only the tree is simulating the trait (i.e. the trait value at a tip is affected by it's ancestor and the branch length leading to it) but not the other way around (the trait value does not affect the tree).
It is possible to add this aspect using `"modifiers"` objects.
`"modifiers"` are similar to `"traits"` in that you specify what should go in there and then feed it to your simulation.

`"modifiers"` affect two key steps of the birth-death process: the calculation of the waiting time (i.e. the component generating branch lengths) and the triggering of speciation or extinction events.
These events can be modified using some `condition` and `modify` function.
In other words, when reaching a certain condition specified by a `condition` function, the birth-death process will modify either the branch length or the speciation (or extinction) probability by applying a `modify` function.

You can use the function `make.modifiers` to design a specific `"modifiers"` object.
By default, this function generates a `"modifiers"` object that affects branch length and speciation in the following way:

 * branch length is a randomly drawn number from an exponential distribution with a rate equal to the current number of taxa multiplied by the sum of the speciation and extinction rates.
 * speciation is triggered if a randomly drawn number (from a (0,1) uniform distribution) is smaller than the ratio between the speciation rate and the sum of the speciation and extinction rates. If that random number is greater, the lineage goes extinct.

Note that these are default for a birth death tree and were actually already applied in the examples before (without specifying a modifier):

```{r}
## Make a default modifiers
default_modifiers <- make.modifiers()
## What's in it?
default_modifiers
```

This will not do much to our simulations compared to the previous trait and tree simulation but we can provide our modifiers object to the `dads` function:

```{r}
## Setting the simulation parameters
extinction_02 <- list(extinction = 0.2)
living_20     <- list(max.living = 20)
BM_trait      <- make.traits()
set.seed(1)
## Simulate disparity and diversity
default_data <- dads(bd.params = extinction_02,
                     stop.rule = living_20,
                     traits    = BM_trait,
                     modifiers = default_modifiers)
default_data
```

Note however that the printing information is now updated to state that you've add a modifier (even though it's a default one).

For more interesting simulations however, you can provide modifiers that actually modify the birth death process.
We can create one for example that makes species go extinct if their ancestor have a negative trait value.
For that we need to create a modifiers object that modifies the `speciation` process with a specific condition and a specific modification when that condition is met.
First we are going to create the modification function.
This function must intake the argument `x` and, in our case, return a logical value: `TRUE` is for speciate and `FALSE` is for go extinct.
We want this `modify` function to always return `FALSE` (go extinct) since it will be triggered by `condition` (we will see that in a minute).

```{r}
## Going extinct
staying.alive <- function(x) return(TRUE)
```

Now we want this function to only trigger when an ancestor has a negative trait value.
We can do that by specifying our `condition` function (when the ancestor is trait is negative) apply our modification (here the `staying.alive` function).
For that we can use the `parent.traits` utility function that is optimised for accessing traits in the birth death process (but you can of course write your own).
This function intakes the `trait.values` and `parent.lineage` arguments, two arguments that you can leave named as they are to facilitate `dads`s understanding of what you which to asses:

```{r}
## Triggering a modification only if the ancestor trait is negative
negative.ancestor <- function(trait.values, lineage) {
    return(all(parent.traits(trait.values, lineage) < 0))
}
```

Note that we use the function `all` here to evaluate all traits (if the data is multidimensional). 
We can then provide these two functions (the condition `negative.ancestor` and how to modify the speciation event when this condition is met `staying.alive`).
If you are an advances `dads` user, you can design your own `speciation` function but if you just want to use a normal `speciation` function, you can use the default one from `dads` called... `speciation`.

```{r}
## Making a modifier for species to go extinct if
## their ancestor's trait value is (or are) negative
negatives_extinct <- make.modifiers(
                        ## The speciation function (default)
                        speciation = speciation,
                        ## What to modify
                        modify     = staying.alive,
                        ## When to modify it
                        condition  = negative.ancestor)
## What's in it?
negatives_extinct
```

Note that the `make.modifiers` function tests whether the input is compatible with `dads` by default so unless you have an error message, your `modifiers` will work!
We can now simulate our tree and traits with our modifier: species will go extinct if their ancestor have a negative trait value:

```{r}
set.seed(1)
## Simulate disparity and diversity
biased_data <- dads(bd.params = extinction_02,
                    stop.rule = living_20,
                    traits    = BM_trait,
                    modifiers = negatives_extinct)
biased_data
```

We can now compare the two trees and their trait values.
Note that we've used the same starting seed for both trees so the only thing differing between them is the modifier!
Also, although species

```{r, fig.height = 6, fig.width = 12}
par(mfrow = c(1,2))
plot(default_data, main = "default results")
plot(biased_data, main = "results with the modifier")
```

You can find much more how to design modifiers in the [`make.modifiers` section](#makemodifiers).

<!--chapter:end:01_getting_started.Rmd-->

---
title: "`make.traits`"
author: "Thomas Guillerme (guillert@tcd.ie)"
date: "`r Sys.Date()`"
bibliography: [../References.bib, ../packages.bib]
output:
  html_document: default
  pdf_document: default
---

# Making trait processes with `make.traits()` {#maketraits}

```{r, echo = FALSE}
set.seed(123)
```

## The process (`process`)

The function `make.traits` allows you to design the process of a trait or a set of traits.
Here, the process of a trait designates the rules to generate the trait through time while simulating a phylogeny.
This process can depend on the previous state in the tree (i.e. the trait of the ancestor) and the branch length to the descendant.
One classic example is the [Brownian motion process (or Weiner process)](https://en.wikipedia.org/wiki/Brownian_motion).
Note that it _can_ depend on both the ancestor and the branch length but does _not necessary needs_ (i.e. the process can be only based on the previous state or only on branch length or on neither).

Trait processes in `dads` are functions that must always intake the following arguments by default.

 * `x0`: the previous trait value(s)
 * `edge.length`: the branch length value
 * `...`: a placeholder for any extra arguments

For example, the following function would be a valid process (though not dependent on either the previous state nor the branch length):

```{r}
## A valid (but useless) process
valid.process <- function(x0, egde.length = 1, ...) {
    return(42)
}
```

> Note that the argument `edge.length` is set to `1` by default. In general we highly recommend to set all arguments but `x0` to a default value (this really helps the speeding up the `dads` function).

On the other hand, the following process (a unidimensional Brownian motion) is incorrect (it's missing `edge.length` and `...`):

```{r}
## A wrongly formated process
invalid.process <- function(x0) {
    return(rnorm(1, mean = x0))
}
```

The `dads` package proposes inbuilt processes, namely a multidimensional Brownian motion (`BM.process`) or a a multidimensional Ornstein-Uhlenbeck process (`OU.process`).
You can find the list of implemented process by looking at the `?trait.process` manual page in `R`.

Once a process is chosen, you can feed it to the `make.traits` function:

```{r}
## Creating a trait object
my_trait_object <- make.traits(process = BM.process)
```

This creates `"dads"` `"traits"` objects that you can print, and visualise using the `plot` function:

```{r}
## The class of the object
class(my_trait_object)
## What's in it?
my_trait_object
## What does the process looks like
plot(my_trait_object)
```

Note that you can see the multiple options for plotting the trait process by looking at `?plot.dads` manual. Furthermore, you can look at what's actually in the object using:

```{r}
## What's actually in that object?
print.dads(my_trait_object, all = TRUE)
```

As traits can get more and more complex, the automatic printing of its summary allows for a easier display of what's in the traits object.

Note that it is possible to make `"traits"` objects with multiple processes (that can be the same):

```{r}
## 4 traits: two BM, one OU and one normal non process
four_traits <- make.traits(process = c(BM.process,
                                       BM.process,
                                       OU.process,
                                       no.process))
four_traits
```

You can visualise them individually using the `trait` argument in `plot.dads`:

```{r}
## Plot options (4 plots in one window)
par(mfrow = c(2,2))
plot(four_traits, trait = 1)
plot(four_traits, trait = 2)
plot(four_traits, trait = 3)
plot(four_traits, trait = 4)
```

## The number of traits `n` and the starting values `start`

Two further important arguments are `n` the number of traits per process and `start` the starting values for all traits.
By default they are set to `n = 1` and `start = 0`.
This means that `make.traits` will assume that your processes are always unidimensional by default and that they always start with the value `0`.
It is however possible to change these values.

For example you can use the following to create a three dimensional Brownian motion with each dimensions starting with the value `1`:

```{r}
## Multidimensional Brownian motion
make.traits(BM.process, n = 3, start = 1)
```

Or the following with each dimensions starting with different values (respectively `1`, `2` and `3`):

```{r}
## Multidimensional Brownian motion
make.traits(BM.process, n = 3, start = c(1,2,3))
```

Note that the number of traits are distributed per processes.
If the traits contains multiple process, the number of traits are distributed per processes:

```{r}
## two 3D processes (BM and OU)
make.traits(c(BM.process, OU.process), n = 3)
## one 1D processes (BM) and one 4D process (OU)
make.traits(c(BM.process, OU.process), n = c(1, 4))
```

And starting values are distributed for all the traits or for the traits one by one:

```{r}
## two 3D processes (BM and OU) starting with 1
make.traits(c(BM.process, OU.process), n = 3, start = 1)
## two 3D processes (BM and OU) starting with values 1 to 6
make.traits(c(BM.process, OU.process), n = 3, start = 1:6)
## two 3D processes (BM and OU) with the two first ones starting
## with 1 and the 4 other ones with the default (0)
make.traits(c(BM.process, OU.process), n = 3, start = c(1,1))
```

## Extra argument for the processes with `process.args`

You can also feed extra arguments to your process(es) functions. For example, the inbuilt process `no.process` (that is just a number generator not based on the previous value `x0` or the branch length) can intake a specific random number generator as a function:

```{r}
## no process trait using the normal distribution (default)
make.traits(no.process, process.args = list(fun = rnorm))
## no process trait using the uniform distribution
## bounded between 1 and 100
make.traits(no.process, process.args = list(fun = runif, min = 1, max = 100))
```

You can also add multiple extra arguments for multiple processes giving them as a list.

```{r}
## Two traits with no process:one normal and one uniform (1,100)
make.traits(process      = c(no.process, no.process),
            process.args = list(list(fun = rnorm),
                                list(fun = runif, min = 1, max = 100)))
```

If one process do not need extra argument you must still give it and extra `NULL` process argument:

```{r}
## Three traits with no process:
## one default, one lognormal and one uniform (1,100)
make.traits(process      = c(no.process, no.process, no.process),
            process.args = list(## Extra arguments for the first process (none)
                                list(NULL),
                                ## Extra arguments for the second process
                                list(fun = rlnorm),
                                ## Extra arguments for the third process
                                list(fun = runif, min = 1, max = 100)))
```

## Naming the traits with `trait.names`

As traits become more and more complex, it can be useful to give clearer names to each process.
This is easily done using the `trait.names` argument that attributes one name per process:

```{r}
## A simple trait with a proper name
simple_trait <- make.traits(trait.names = "1D Brownian Motion")
simple_trait
```

This becomes more useful if we use the complex example above:

```{r}
## Three named traits with no process:
## one default, one lognormal and one uniform (1,100)
make.traits(process      = c(no.process, no.process, no.process),
            process.args = list(## Extra arguments for the first process (none)
                                list(NULL),
                                ## Extra arguments for the second process
                                list(fun = rlnorm),
                                ## Extra arguments for the third process
                                list(fun = runif, min = 1, max = 100)),
            ## Naming each trait
            trait.names  = c("Normal", "LogNormal", "Uniform(1,100)"))
```

## Combining multiple traits with `add`

You can also add traits to already existing trait objects using the simple `add` option.
This option just intakes a `"dads"` `"traits"` object and the additional process(es) will be added to it. For example:

```{r}
## Creating on simple default Brownian motion
one_process <- make.traits(trait.names = "BM")

## Creating a new trait (a 3D OU.process)
## and adding the previous one
two_processes <- make.traits(OU.process, n = 3, add = one_process,
                             trait.names = "3D OU")

## Only one process
one_process

## The two processes
two_processes
```

## Testing the traits with `test`

This bit is more for development.
We highly suggest leaving `test = TRUE` so that `make.traits` returns an error if a process or its additional arguments (`process.args`) are not formatted correctly.
`make.traits` will error if the trait cannot be directly passed to `dads`.
However, in some specific cases (again, probably mainly for development and debugging) it could be useful to skip the tests using `test = FALSE`.

## Templates for making your very own process

As detailed above, any process of your own design will work as long as it is a function that takes at least the arguments `x0` and `edge.length`.
You can be imaginative and creative when designing your own process but here are two detailed example functions for a unidimensional Brownian Motion and Ornstein-Uhlenbeck process that you can use for a start (or not).
Remember it is good practice for `dads` processes to set all the arguments but `x0` with default values (just in case).
Also, note that the functions below are not equal to the already implemented `BM.process` and `OU.process` but are rather generalised/simplified version that you can use as a template

### A simple Brownian Motion process template

```{r}
## A simple Brownian motion process
my.BM.process <- function(x0, edge.length = 1, sd = 1, ...) {
    ## Drawing a random number from a normal distribution
    ## with x0 as the and a given standard deviation
    ## and depending on branch (edge) length
    result <- rnorm(n = 1, mean = x0, sd = sd * edge.length)

    ## Return the number
    return(result)
}
```

### A simple Ornstein-Uhlenbeck process template

```{r}
## A simple Ornstein-Uhlenbeck motion process
my.OU.process <- function(x0, edge.length = 1, var = 1, alpha = 1, ...) {
    ## Calculate the mean based on alpha
    mean <- x0 * exp(-alpha)
    ## Calculate the standard deviation based on alpha and the variance
    sd <- sqrt(var/(2 * alpha) * (1 - exp(-2 * alpha)))
    ## Draw a random number from a normal distribution
    ## using this mean and standard deviation
    ## and depending on branch (edge) length
    result <- rnorm(n = 1, mean = mean, sd = sd * branch.length)

    ## Return the number
    return(result)
}
```

<!--chapter:end:02_make.traits.Rmd-->

---
title: "`make.modifiers`"
author: "Thomas Guillerme (guillert@tcd.ie)"
date: "`r Sys.Date()`"
bibliography: [../References.bib, ../packages.bib]
output:
  html_document: default
  pdf_document: default
---

# Making modifiers with `make.modifiers()` {#makemodifiers}

```{r, echo = FALSE}
set.seed(123)
```

`"modifiers"` have a similar structure than `"traits"` where you can design an object with increasing complexity, starting with the simplest modifiers that doesn't modify anything (using the default arguments):

```{r}
## Making a default modifier (no modification)
my_default_modifiers <- make.modifiers()
my_default_modifiers
```

Similarly to `"traits"` objects, `"modifiers"` are also printed by default using `print.dads`.
You can see details about what's actually in the object using `print.dads(my_default_modifiers, all = TRUE)`.
However, contrary to `"traits"`, you cannot plot `"modifiers"`.

## The branch length function (`branch.length`)

The first argument in `"modifiers"` is the branch length function (`branch.length`) this is the function that will be executed in `dads` to generate branch length.
Note that in the `dads` algorithm, branch length is not generated _directly_ but actually results of the waiting time.
In other words, the `branch.length` function just affects waiting time for all taxa present at any time in the simulation.
These taxa can then either go extinct (stopping the "growth" of it's branch length) or survive (continuing the "growth").

By default, branch length (or waiting or growth) is a randomly drawn number from an exponential distribution with the rate of the number of taxa multiplied by the speciation and extinction rate: $n \times (\lambda + \mu)$ (where $n$ is the number of taxa currently present in the simulation, $\lambda$ and $\mu$ are respectively the speciation and extinction rates).
This default function is simply called `branch.length` in `dads` and can be used as a modifier as follows:

```{r}
## Specifying the default modifier
default_modifiers <- make.modifiers(branch.length = branch.length)

## Setting some parameters for generating trees
bd_params <- list(extinction = 0)
stop_rule <- list(max.living = 20)

## Generating a tree with the default branch length parameter
set.seed(0)
default_tree <- dads(bd.params = bd_params,
                     stop.rule = stop_rule,
                     modifiers = default_modifiers)
```

Of course, the point of the modularity here is that you can provide your own function for generating branch length.
For example, we might be interested in what our tree would look like if we'd use a simple constant branch length generation (instead of randomly drawing it from an exponential distribution).
We can do so by declaring our own `branch.length` function and adding it to a `"modifiers"` object.

```{r}
## A constant branch length generator
## (note that the output must be numeric, not integer)
constant.brlen <- function() {
    return(as.numeric(1))
}

## Creating the modifiers object
constant_modifier <- make.modifiers(branch.length = constant.brlen)

## Generating a new tree with this modifier
set.seed(0)
modified_tree <- dads(bd.params = bd_params,
                      stop.rule = stop_rule,
                      modifiers = constant_modifier)
```

And we can visualise the difference between both resulting trees:

```{r}
par(mfrow = c(1,2))
plot(default_tree,  main = "Default modifier")
plot(modified_tree, main = "Constant branch length\nmodifier")
```

It is of course to use more complex branch length modifiers that intakes different conditions and specific modification rather than simply always output a value of one.

### The allowed arguments {#allowarguments}

It is of course possible to design some more advanced function to interact with the birth death process.
You can create a function for `branch.length`, `selection` and `speciation` that involve any of the following arguments:

  * `bd.params`: a named list containing `"numeric"` values that contains the birth death parameters (at least `"speciation"` and `"extinction"`);
  * `lineage`: a named list containing the lineage data (see below).
  * `trait.values`: a `"matrix"` containing `"numeric"` values with the trait names as column names and the lineages ID as row numbers (you can use it with the function `parent.traits` to access the trait of the previous node for example).
  * `modify.fun`: a `"list"` of named `"function"` (usually passed through `condition` and `modify`).

The `lineage` list contains the following elements (missing elements are allowed):

 * `lineage$parents`: an `"integer"` vector: the list of parent lineages;
 * `lineage$livings`: an `"integer"` vector: the list of lineages still not extinct;
 * `lineage$drawn`: a single `"integer"`: the ID of the selected lineage;
 * `lineage$current`: a single `"integer"`: the selected lineage (is equal to `lineage$livings[lineage$drawn]`);
 * `lineage$n`: a single `"integer"`: the current number of non extinct lineage (is equal to `length(lineage$livings))`;
 * `lineage$split`: a `"logical"` vector: the list of splits for each lineage (`TRUE`), the number of total tips is equal to `sum(!lineage$split)`.

In general, unless you know what you're doing, you can ignore most arguments for specific modifiers since they are handled automatically within the `dads` function.
Therefore any argument can be left undeclared or missing and is always handled internally. For example, if you did not declare `n.taxa` as a function argument but are using `n.taxa` in the function, `n.taxa` will be detected and treated as a current argument automatically as set accordingly within the birth death process (e.g. `n.taxa` will be set to the current number of taxa every iteration of the process).

For example, we can create a function that increases branch length proportional to the number of species "alive" at each time of the simulation in a discrete way. I.e. for discrete numbers of taxa, the branch length increases by jumps (ten fold):

```{r}
## A more complex binned.branch.length function
increasing.brlen <- function(bd.params, lineage) {

    ## Setting the cumulated birth and death
    birth_death <- bd.params$speciation + bd.params$extinction

    ## Returning branch lengths depending on different number of taxa
    if(lineage$n <= 5) {
        return(1    * rexp(1, sum(5 * birth_death)))
    }
    if(lineage$n <= 10) {
        return(10   * rexp(1, sum(10 * birth_death)))   
    }
    if(lineage$n <= 15) {
        return(100  * rexp(1, sum(15 * birth_death)))      
    }
    if(lineage$n <= 20) {
        return(1000 * rexp(1, sum(20 * birth_death)))
    } else {
        return(1000 * rexp(1, sum(lineage$n * birth_death)))
    }
}
```

We can then create it as a `"modifiers"` object and run a new simulation:

```{r}
## Creating a modifiers
increasing_modifier <- make.modifiers(branch.length = increasing.brlen)

## Generating a new tree with this modifier
set.seed(0)
increasing_tree <- dads(bd.params = bd_params,
                        stop.rule = stop_rule,
                        modifiers = increasing_modifier)
```

And we can visualise the difference between the resulting trees:

```{r}
par(mfrow = c(1,3))
plot(default_tree,    main = "Default modifier")
plot(modified_tree,   main = "Constant branch length\nmodifier")
plot(increasing_tree, main = "Increasing branch length\nmodifier (binned)")
```

## The selection function (`selection`)

The `selection` function is used in the birth death process to know which lineage to select when running a speciation (or extinction!) event.
By default, this function randomly selects one taxon that is currently not extinct (using: `sample(1:n.taxa, 1))`.
Similarly as `branch.length` it is possible to modify this part of the birth death process.
For example, we could simply select always the last created lineage (to create a "ladder" or most asymmetric tree):

```{r}
## Our function to always select the last taxon
## (making sure it returns an integer)
select.last <- function(lineage) {
    return(as.integer(lineage$n))
}
```

> Note that here the function can only intake the [allowed arguments as described above](#allowarguments) (here `n.taxa`: the number of current living taxa).

We can then create a `"modifiers"` object the same way as before using this time the `selection` argument:

```{r}
## A modifier for selection
ladderised_modifier <- make.modifiers(selection = select.last)
## Generating a new tree with this modifier
set.seed(0)
ladderised_tree <- dads(bd.params = bd_params,
                        stop.rule = stop_rule,
                        modifiers = ladderised_modifier)
## Displaying the results
par(mfrow = c(1,2))
plot(default_tree,    main = "Default modifier")
plot(ladderised_tree, main = "Ladderising modifier")
```

Again, it is of course possible to make the modifier more complex and in combination with other elements of the tree.
For example, we can create a `"dads"` object that also creates a trait a add to it a `selection` modifier that only selects for tips with positive trait values.

```{r}
## Our function that only select taxa with positive trait values
select.positive <- function(trait.values, lineage) {
    ## Selecting the taxa with positive traits only
    positives <- as.integer(rownames(trait.values)[which(trait.values[, 1] >= 0)])

    ## Select the current taxa that descend from a node with a positive value
    positive_living <- cbind(lineage$parents, seq_along(lineage$split)
                             )[which(lineage$parents %in% positives), 2]

    ## Select one tip randomly in the ones with descendants with positive values
    return(sample(which(lineage$livings %in% positive_living), 1))
}

## Creating the modifier
positive_skew <- make.modifiers(selection = select.positive)

## Creating a (default) trait object
BM_trait <- make.traits()

## Simulate a tree and trait with no modifier
set.seed(1)
default_dads <- dads(bd.params = bd_params,
                     stop.rule = stop_rule,
                     traits    = BM_trait)

## Simulate a tree and trait with the modifier
set.seed(1)
skewed_trait_dads <- dads(bd.params = bd_params,
                          stop.rule = stop_rule,
                          traits    = BM_trait,
                          modifiers = positive_skew)

## Plotting the differences in trees and traits
par(mfrow = c(1, 2))
plot(default_dads, main = "Default trait and tree")
plot(skewed_trait_dads, main = "Skewed trait and tree")
```

## The speciation function (`speciation`)

The third function that can be used to modify the birth death process is the `speciation` function.
This one is used during the birth death process to decide whether a lineages speciates (creating a node and two new lineages) or goes extinct (creating a tip).

> Note that the `speciation` function only affects tips or nodes before the simulation reaches the `stop.rule`.
The then surviving lineages are all automatically transformed into tips.

By default, the `speciation` function is trigger a speciation even if a number randomly drawn from a uniform distribution is lower than the ratio between the speciation and the speciation and extinction parameter. If the randomly drawn number is higher, the lineage goes extinct.

```{r, eval = FALSE}
## The speciation in pseudo-code:
runif(1) < speciation/ (speciation + extinction)
```

Creating a `"modifiers"` with a `speciation` function works the same way as for `branch.length` and `selection` but the function that will be used needs to output a logical value (see table [below](#summarymodifiers)).
Once the function is created simply input your function for speciation in the modifier and run the `dads` function with that modifier:

```{r, eval = FALSE}
## Speciating or going extinct randomly
## (regardless of the extinction parameter)
random.extinct  <- function() {
    return(sample(c(TRUE, FALSE), 1))
}

## Creating the modifiers object
random_extinction <- make.modifiers(speciation = random.extinct)

## Generating a new tree with this modifier
set.seed(0)
modified_tree <- dads(bd.params = bd_params,
                      stop.rule = stop_rule,
                      modifiers = random_extinction)

par(mfrow = c(1,2))
plot(default_tree,  main = "Default modifier")
plot(modified_tree, main = "Random extinction\nmodifier")
```

Note how every lineage end up going extinct!
And again, we can make some more advanced modifiers: for example, one where a tip always goes extinct if their ancestor has a negative trait value (here we will also introduce the utility function `parent.trait` that automatically selects the trait values of the parent of the current lineage.

```{r}
## A modifier for removing tips with negative values
bias.trait <- function(trait.values, lineage) {
    if(parent.traits(trait.values, lineage) < 0) {
        ## Go extinct!
        return(FALSE)
    } else {
        ## Speciate
        return(TRUE)
    }
}

## Creating the modifier
biased_trait <- make.modifiers(speciation = bias.trait)

## Simulate a tree and trait with the modifier
set.seed(1)
biased_trait_dads <- dads(bd.params = bd_params,
                          stop.rule = stop_rule,
                          traits    = BM_trait,
                          modifiers = biased_trait)

## Plotting the differences in trees and traits
par(mfrow = c(1, 2))
plot(default_dads, main = "Default trait and tree")
plot(biased_trait_dads, main = "Biased trait and tree")
```

## Summary of the inputs and outputs for the `branch.length`, `selection` and `speciation` modifiers {#summarymodifiers}

modifier name   | accepted input (arguments) | required output (class) |
----------------|----------------------------|-------------------------|
`branch.length` | `bd.params`, `lineage`, `trait.values` | `"numeric"` |
`selection`     | `bd.params`, `lineage`, `trait.values` | `"integer"` |
`speciation`    | `bd.params`, `lineage`, `trait.values` | `"logical"` |

## The condition and modify functions (`condition` and `modify`)

In the examples above, we have seen how to specify modifications to the birth death process (via `branch.length`, `selection` and `speciation`), however, one might note that these modifications are not dynamic.
In other words, throughout the process, the modifications remain constant (even if they are conditional).
It is however possible to code the `"modifiers"` so that they can be affected by `"events"` objects (see next chapter).

To do so, you can formally declare conditions (`condition`) and modifications (`modify`) as internal functions that can then be modified my an `"events"` object.
`condition` and `modify` are hard coded in the `branch.length` function that they concern, i.e. they are variables (functions) within the function.

## Combining and editing modifiers (`add`)

## Testing modifiers (`test`)

## Demo runnable

```{r, fig.height = 16, fig.width = 8, eval = FALSE}

```

<!--chapter:end:03_make.modifiers.Rmd-->

---
title: "`make.events`"
author: "Thomas Guillerme (guillert@tcd.ie)"
date: "`r Sys.Date()`"
bibliography: [../References.bib, ../packages.bib]
output:
  html_document: default
  pdf_document: default
---

```{r, echo = FALSE}
library(rgl)
knitr::knit_hooks$set(webgl = hook_webgl)
```

# Making events with `make.events()` {#makeevents}

```{r, echo = FALSE}
set.seed(123)
```

## Conditions

## Modifications

## Some events examples

### Random mass extinction after some time

```{r}
## Simulation parameters
stop.rule <- list(max.time = 5)
bd.params <- list(extinction = 0, speciation = 1)

## 80% mass extinction at time 4
mass_extinction <- make.events(
                      target = "taxa",
                      condition = time.condition(4),
                      modification = random.extinction(0.8))

## Running the simulations
set.seed(1)
results <- dads(bd.params = bd.params,
                stop.rule = stop.rule,
                events    = mass_extinction)
## Plotting the results
plot(results, show.tip.label = FALSE)
axisPhylo()
```

### Extinction based on a trait value after some time

```{r}
## Simulation parameters
stop.rule <- list(max.time = 5)
bd.params <- list(extinction = 0, speciation = 1)

## Extinction of any tips with trait < 1 at time 4
trait_extinction <- make.events(
                      target = "taxa",
                      condition = time.condition(4),
                      modification = trait.extinction(1))

## Running the simulations
set.seed(7)
results <- dads(bd.params = bd.params,
                stop.rule = stop.rule,
                traits    = make.traits(),
                events    = trait_extinction)
## Plotting the results
plot(results)
```

### Adding a background extinction after reaching a number of living taxa

```{r}
## Simulation parameters
stop.rule <- list(max.living = 50)
bd.params <- list(extinction = 0, speciation = 1)

## Adding an extinction parameter after 30 taxa
background_extinction <- make.events(
                      condition    = taxa.condition(30),
                      target       = "bd.params",
                      modification = update.bd.params(1/3, "extinction"))

## Running the simulations
set.seed(2)
results <- dads(bd.params = bd.params,
                stop.rule = stop.rule,
                events    = background_extinction)
## Plotting the results
plot(results)
```

### Reducing speciation after some time

```{r}
## Simulation parameters
stop.rule <- list(max.time = 4)
bd.params <- list(extinction = 0, speciation = 1)

## Reducing speciation after reaching time 2
reduced_speciation <- make.events(
                      condition    = time.condition(2),
                      target       = "bd.params",
                      modification = update.bd.params(1/3, "speciation"))

set.seed(42)
no_event <- dads(bd.params = bd.params,
                 stop.rule = stop.rule)
set.seed(42)
reduced_speciation_event <- dads(bd.params = bd.params,
                                 stop.rule = stop.rule,
                                 events    = reduced_speciation)

## Plot both trees
par(mfrow = c(1, 2))
plot(no_event, main = "No event")
axisPhylo()
plot(reduced_speciation_event, main = "Reduced speciation after time 2")
axisPhylo()
```

### Changing trait process after some time

```{r}
## Simulation parameters
bd.params <- list(extinction = 0, speciation = 1)
stop.rule <- list(max.time = 6)
traits    <- make.traits()

## Create an event to change the trait process
change_process <- make.events(
                  condition    = time.condition(5),
                  target       = "traits",
                  modification = update.traits(process = OU.process))

## Run the simulations
set.seed(1)
no_change <- dads(bd.params = bd.params,
                  stop.rule = stop.rule,
                  traits    = traits)
set.seed(1)
process_change <- dads(bd.params = bd.params,
                       stop.rule = stop.rule,
                       traits    = traits,
                       events    = change_process)
## Plot the results
par(mfrow = c(1,2))
plot(no_change, ylim = c(-7, 7))
plot(process_change, ylim = c(-7, 7))
```

### Changing trait correlation after reaching a trait value

```{r}
## Set the parameters
stop.rule <- list(max.taxa = 100)
bd.params <- list(extinction = 0, speciation = 1)

## A correlated trait
traits <- make.traits(n = 2, process.args = list(Sigma = matrix(1, 2, 2)))

## Event changing a trait correlation
cor_change <- make.events(
    condition    = trait.condition(3, absolute = TRUE),
    target       = "traits",
    modification = update.traits(process.args = list(Sigma = matrix(c(10,3,3,2),2,2))))

## Run the simulations
set.seed(2)
no_event <- dads(bd.params = bd.params,
                 stop.rule = stop.rule,
                 traits    = traits)
set.seed(2)
change_correlation <- dads(bd.params = bd.params,
                           stop.rule = stop.rule,
                           traits    = traits,
                           events    = cor_change)

## Visual testing
par(mfrow = c(1,2))
plot(no_event, trait = 2)
plot(change_correlation, trait = 2)

```

```{r, webgl=TRUE}
## 3D plot
plot(change_correlation, trait = c(1:2), use.3D = TRUE)
```

### Event for changing a modifier

```{r}
## Set the parameters
stop.rule <- list(max.time = 4)
bd.params <- list(extinction = 0, speciation = 1)

## Using a default trait and modifier object
modifiers <- make.modifiers()
traits <- make.traits()

## New condition and new modifier (increasing speciation if trait is negative)
new.condition <- function(trait.values, lineage) {
    return(parent.traits(trait.values, lineage) < 0)
}
new.modify    <- function(x, trait.values, lineage) {
  return(x + 1)
}

## Update the modifier
change_speciation <- make.events(
    condition    = time.condition(3),
    target       = "modifiers",
    modification = update.modifiers(speciation = speciation,
                                    condition  = new.condition,
                                    modify     = new.modify))

set.seed(4)
no_event <- dads(bd.params = bd.params,
                 stop.rule = stop.rule,
                 traits    = traits,
                 modifiers = modifiers)
set.seed(4)
change_spec <- dads(bd.params = bd.params,
                    stop.rule = stop.rule,
                    traits    = traits,
                    modifiers = modifiers,
                    events    = change_speciation)

## Visualise the results
par(mfrow = c(1,2))
plot(no_event, main = "No event")
plot(change_spec, main = "Increase extinction for negative\ntraits after time 3")
```

### Changing branch length when reaching n taxa

```{r}
## Parameters
bd.params <- list(extinction = 0, speciation = 1)
stop.rule <- list(max.taxa = 100)
modifiers <- make.modifiers()
traits <- make.traits()

## multiplying branch length 100 folds
new.modify <- function(x, trait.values, lineage) {
  return(x * 100)
}

## Event for increasing branch length after reaching 30 taxa
increase_brlen <- make.events(
                  condition    = taxa.condition(30),
                  target       = "modifiers",
                  modification = update.modifiers(
                                    branch.length = branch.length,
                                    modify = new.modify))

## Run the simulations
set.seed(5)
no_event <- dads(bd.params = bd.params,
                 stop.rule = stop.rule,
                 traits    = traits,
                 modifiers = modifiers)
set.seed(5)
increased_brlen <- dads(bd.params = bd.params,
                        stop.rule = stop.rule,
                        traits    = traits,
                        modifiers = modifiers,
                        events    = increase_brlen)

## Visualise the results
par(mfrow = c(1,2))
plot(no_event, main = "No event")
plot(increased_brlen, main = "Increase branch length\nafter 30 taxa")
```

### Founding event: a subtree with no fossils

```{r}
## Set up parameters
stop.rule <- list(max.time = 4)
bd.params <- list(speciation = 1, extinction = 0.3)

## Events that generate a new process (founding effects)
founding_event <- make.events(
                  condition    = taxa.condition(10),
                  target       = "founding",
                  modification = founding.event(
                                    bd.params = list(speciation = 2,
                                                     extinction = 0)),
                  additional.args = list(prefix = "founding_"))

## Simulations
set.seed(11)
founding_tree <- dads(bd.params = bd.params,
                      stop.rule = stop.rule,
                      events    = founding_event)
plot(founding_tree, cex = 0.4)
```

### Founding event with traits

```{r}
## Set up parameters
stop.rule <- list(max.time = 4)
bd.params <- list(speciation = 1, extinction = 0.3)
traits    <- make.traits()

## Events that generate a new process (founding effects)
founding_event <- make.events(
                  condition    = taxa.condition(10),
                  target       = "founding",
                  modification = founding.event(
                                    bd.params = list(speciation = 2,
                                                     extinction = 0),
                                    traits = make.traits(process = OU.process)),
                  additional.args = list(prefix = "founding_"))

## Simulating the tree
set.seed(19)
founding_tree <- dads(bd.params = bd.params,
                      stop.rule = stop.rule,
                      traits    = make.traits(),
                      events    = founding_event)
plot(founding_tree)
```

<!--chapter:end:04_make.events.Rmd-->

---
title: "Thorough example"
author: "Thomas Guillerme (guillert@tcd.ie)"
date: "`r Sys.Date()`"
bibliography: [../References.bib, ../packages.bib]
output:
  html_document: default
  pdf_document: default
---

```{r, echo = FALSE}
library(rgl)
knitr::knit_hooks$set(webgl = hook_webgl)
```
# Thorough example

Here we are going to look at an example on how to simulate complex evolutionary histories with `dads`.

Let say we want to simulate the following story:

 1. A lineage evolves in a normal birth-death fashion (with extinction) from a single origin and with two traits called:
    - "Distance": the physical distance between species in an arbitrary 1D space (evolving in a Brownian motion).
    - "Variable": a 2D uncorrelated Ornstein-Uhlenbeck trait that corresponds to some traits the lineages have.
 During the whole process, lineages that are more further away from the other species in terms of distance have a higher change of speciating (proportional to that distance).

Three events then happen during the evolutionary history of this lineage:

 2. "Correlation event": after reaching time 2 an event happens making the "Variables" trait dimensions become correlated.
 3. "Selective extinction event": after reaching a certain point in time, lineages with their "Variables" values that are negative go extinct
 4. "Founding event": after reaching a value for the "Distance" trait, a founding event happens with a new birth-death process with no extinction and where the "Variable" is now a 2D correlated Brownian Motion.
 5. "Mass extinction event": after reaching a certain number of taxa in the "Founding event", 80% of taxa go extinct randomly.

We are going to run each step one by one in a cumulative manner to see how they pile up:

## Step 1:

### The birth death parameters

First let's set the overall birth death parameters (a speciation of 1.2 and an extinction of 0.3):

```{r}
## The birth death parameters
bd_params <- list(speciation = 1.5, extinction = 0.3)
## The stopping rule
stop_rule <- list(max.time = 4)
```

And let's quickly test if the simulation works by running a birth death tree without traits and anything:

```{r}
set.seed(0) # 123 also works
## Running a birth death tree 
results <- dads(bd.params = bd_params,
                stop.rule = stop_rule)

## And displaying the results
plot(results)
axisPhylo()
```

### The traits

Then let's create the traits:

```{r}
## The trait object
two_traits <- make.traits(process = c(BM.process, OU.process), n = c(1, 2),
                          trait.names = c("Distance", "Variables"))
```

```{r}
## Visualise the 1D trait (BM.process)
plot(two_traits, trait = 1)
```

```{r, webgl=TRUE}
## Visualise the 2D trait (uncorrelated OU.process)
plot(two_traits, trait = 2, use.3D = TRUE)
```

Note that specifying the correlation matrix is not necessary here _per se_ (since no correlation is the default) but we will modify it through an event later on so it's good practice to have it properly stated before hand (to avoid any confusion).

We can then run the tree with the traits:

```{r}
set.seed(0)
## Running a birth death tree 
results <- dads(bd.params  = bd_params,
                stop.rule  = stop_rule,
                traits     = two_traits)

## And displaying the results
par(mfrow = c(1,2))
plot(results, trait = 1, main = "Trait 1: Distance")
plot(results, trait = c(2,3), main = "Variables correlation")
```

```{r, webgl=TRUE}
## Visualising the tree in 3D
plot(results, trait = c(2,3), use.3D = TRUE)
```

### The modifiers

```{r}
## Speciation event is more likely if lineage's ancestor is further away from the mean trait value
distance.modify <- function(x, trait.values, lineage) return(x + x * abs(parent.traits(trait.values, lineage)[1] - mean(trait.values[,1])))

## Make a distance modifier
distance.speciation <- make.modifiers(speciation = speciation,
                                      modify = distance.modify)
```

```{r}
set.seed(13)
## Running a birth death tree 
results <- dads(bd.params  = bd_params,
                stop.rule  = stop_rule,
                traits     = two_traits,
                modifiers  = distance.speciation)

## And displaying the results
par(mfrow = c(1,2))
plot(results, trait = 1, main = "Trait 1: Distance")
plot(results, trait = c(2,3), main = "Variables correlation")
```

```{r, webgl=TRUE}
## Visualising the tree in 3D
plot(results, trait = c(2,3), use.3D = TRUE)
```

> BUGGED FROM HERE!

## Step 2:
 2. "Correlation event": after reaching time 2 an event happens making the "Variable" trait dimensions become correlated.

```{r}
## The correlation change event
update.correlation <- update.traits(process.args =list(Sigma = matrix(c(10,3,3,2),2,2)))

events_list <- make.events(event.name   = "Correlation",
                           target       = "traits",
                           condition    = time.condition(2),
                           modification = update.traits(
                    process.args =list(Sigma = matrix(c(10,3,3,2),2,2)),
                    trait.names  = "Variables"))
```

```{r}
results <- NULL
set.seed(1)
## Running a birth death tree 
results <- dads(bd.params  = bd_params,
                stop.rule  = stop_rule,
                traits     = two_traits,
                modifiers  = distance.speciation,
                events     = events_list)

## And displaying the results
par(mfrow = c(1,2))
plot(results, trait = 1, main = "Trait 1: Distance")
plot(results, trait = c(2,3), main = "Variables correlation")
```

```{r, webgl=TRUE}
## Visualising the tree in 3D
plot(results, trait = c(2,3), use.3D = TRUE)
```

## Step 3:
 3. "Selective extinction event": after reaching a certain point in time, lineages with their "Variables" values that are negative go extinct

```{r, eval = FALSE}
## Adding the selective event
events_list <- make.events(event.name   = "Selection",
                           add          = events_list,
                           target       = "taxa",
                           condition    = time.condition(4),
                           modification = trait.extinction(x = 0, condition = `<`, trait = 2))
```
Problem with `modification` argument here

```{r, eval = FALSE}
## Running a birth death tree 
results <- dads(bd.params  = bd_params,
                stop.rule  = stop_rule,
                traits     = two_traits,
                modifiers  = distance.speciation,
                events     = events_list)

## And displaying the results
par(mfrow = c(1,2))
plot(results, trait = 1, main = "Trait 1: Distance")
plot(results, trait = c(2,3), main = "Variables correlation")
```

## Step 4 and 5:
 4. "Founding event": after reaching a value for the "Distance" trait, a founding event happens with a new birth-death process with no extinction and where the "Variable" is now a 2D correlated Brownian Motion.
 5. "Mass extinction event": after reaching a certain number of taxa in the "Founding event", 80% of taxa go extinct randomly.

```{r, eval = FALSE}
## The new trait for the founding event
new_traits <- make.traits(process = c(BM.process, BM.process), n = c(1, 2),
                          trait.names = c("Distance", "Variables"))

## The mass extinction event after reaching 50 taxa
mass_extinction <- make.events(target       = "taxa",
                               condition    = taxa.condition(50),
                               modification = random.extinction(0.8))
```

```{r, eval = FALSE}
## The selective event
events_list <- make.events(event.name   = "Founding", 
                           add          = events_list,
                           target       = "founding",
                           condition    = trait.condition(3),
                           modification = founding.event(traits    = new_traits,
                                                         events    = mass_extinction,
                                                         bd.params = list(
                                                            speciation = 2,
                                                            extinction = 0)),
                           additional.args = list(prefix = "found_"))
```

```{r, eval = FALSE}
## Running a birth death tree 
results <- dads(bd.params  = bd_params,
                stop.rule  = stop_rule,
                traits     = two_traits,
                modifiers  = distance.speciation,
                events     = events_list,
                null.error = 100)

## And displaying the results
par(mfrow = c(1,2))
plot(results, trait = 1, main = "Trait 1: Distance")
plot(results, trait = c(2,3), main = "Variables correlation")
```

<!--chapter:end:05_thorough_example.Rmd-->

### 2021-01-19 inst/gitbook/index.Rmd
Updated manual

@@
 title: "dads manual"
-author: "Thomas Guillerme (guillert@tcd.ie), Alex Slavenko (email) and others"
+author: "Thomas Guillerme (guillert@tcd.ie) and others"
 date: "`r Sys.Date()`"

### 2021-01-19 inst/gitbook/_book/dads_manual.tex
updated book

@@
 \author{Thomas Guillerme (\href{mailto:guillert@tcd.ie}{\nolinkurl{guillert@tcd.ie}}), Alex Slavenko (email) and others}
-\date{2021-01-12}
+\date{2021-01-19}
 
@@
 ## }
-## <bytecode: 0x7faff6222240>
+## <bytecode: 0x7fda2b5b1980>
 ## <environment: namespace:dads>
@@
 
-\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-57-1.pdf}
+\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-58-1.pdf}
 
@@
 
-\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-58-1.pdf}
+\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-59-1.pdf}
 
@@
 
-\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-59-1.pdf}
+\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-60-1.pdf}
 
@@
 
-\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-60-1.pdf}
+\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-61-1.pdf}
 
@@
 
-\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-61-1.pdf}
+\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-62-1.pdf}
 
@@
 \CommentTok{## Event changing a trait correlation}
-\NormalTok{change_correlation <-}\StringTok{ }\KeywordTok{make.events}\NormalTok{(}
+\NormalTok{cor_change <-}\StringTok{ }\KeywordTok{make.events}\NormalTok{(}
     \DataTypeTok{condition    =} \KeywordTok{trait.condition}\NormalTok{(}\DecValTok{3}\NormalTok{, }\DataTypeTok{absolute =} \OtherTok{TRUE}\NormalTok{),}
@@
 \CommentTok{## Run the simulations}
-\KeywordTok{set.seed}\NormalTok{(}\DecValTok{8}\NormalTok{)}
+\KeywordTok{set.seed}\NormalTok{(}\DecValTok{2}\NormalTok{)}
 \NormalTok{no_event <-}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ bd.params,}
@@
                  \DataTypeTok{traits    =}\NormalTok{ traits)}
-\KeywordTok{set.seed}\NormalTok{(}\DecValTok{8}\NormalTok{)}
+\KeywordTok{set.seed}\NormalTok{(}\DecValTok{2}\NormalTok{)}
 \NormalTok{change_correlation <-}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ bd.params,}
@@
                            \DataTypeTok{traits    =}\NormalTok{ traits,}
-                           \DataTypeTok{events    =}\NormalTok{ change_correlation)}
+                           \DataTypeTok{events    =}\NormalTok{ cor_change)}
+
 \CommentTok{## Visual testing}
@@
 
-\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-62-1.pdf}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{# plot(no_event, trait = c(1:2))}
-\CommentTok{# plot(change_correlation, trait = c(1:2), use.3D = TRUE)}
-\end{Highlighting}
-\end{Shaded}
+\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-63-1.pdf}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{## 3D plot}
+\KeywordTok{plot}\NormalTok{(change_correlation, }\DataTypeTok{trait =} \KeywordTok{c}\NormalTok{(}\DecValTok{1}\OperatorTok{:}\DecValTok{2}\NormalTok{), }\DataTypeTok{use.3D =} \OtherTok{TRUE}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\hypertarget{unnamed_chunk_64div}{}
+
+You must enable Javascript to view this page properly.
 
@@
 \KeywordTok{plot}\NormalTok{(no_event, }\DataTypeTok{main =} \StringTok{"No event"}\NormalTok{)}
-\KeywordTok{plot}\NormalTok{(change_spec, }\DataTypeTok{main =} \StringTok{"Increase extinction for negative traits}\CharTok{\textbackslash{}n}\StringTok{after time 3"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-63-1.pdf}
+\KeywordTok{plot}\NormalTok{(change_spec, }\DataTypeTok{main =} \StringTok{"Increase extinction for negative}\CharTok{\textbackslash{}n}\StringTok{traits after time 3"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-65-1.pdf}
 
@@
 
-\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-64-1.pdf}
+\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-66-1.pdf}
 
@@
 
-\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-65-1.pdf}
+\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-67-1.pdf}
 
@@
 
-\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-66-1.pdf}
-
-\hypertarget{thorough-example}{%
-\chapter{Thorough example}\label{thorough-example}}
-
-Here we are going to look at an example on how to simulate complex evolutionary histories with \texttt{dads}.
-
-Let say we want to simulate the following story:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  A lineage evolves in a normal birth-death fashion (with extinction) from a single origin and with two traits called:
-
-  \begin{itemize}
-  \tightlist
-  \item
-    ``Distance'': the physical distance between species in an arbitrary 1D space (evolving in a Brownian motion).
-  \item
-    ``Variable'': a 2D uncorrelated Ornstein-Uhlenbeck trait that corresponds to some traits the lineages have.
-    During the whole process, lineages that are more further away from the other species in terms of distance have a higher change of speciating (proportional to that distance).
-  \end{itemize}
-\end{enumerate}
-
-Three events then happen during the evolutionary history of this lineage:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\setcounter{enumi}{1}
-\tightlist
-\item
-  ``Correlation event'': after reaching a certain number of taxa an event happens making the ``Variables'' trait dimensions become correlated.
-\item
-  ``Selective extinction event'': after reaching a certain point in time, lineages with their ``Variables'' values that are negative go extinct
-\item
-  ``Founding event'': after reaching a value for the ``Distance'' trait, a founding event happens with a new birth-death process with no extinction and where the ``Variable'' is now a 2D correlated Brownian Motion.
-\item
-  ``Mass extinction event'': after reaching a certain number of taxa in the ``Founding event'', 80\% of taxa go extinct randomly.
-\end{enumerate}
-
-We are going to run each step one by one in a cumulative manner to see how they pile up:
-
-\hypertarget{step-1}{%
-\section{Step 1:}\label{step-1}}
-
-\hypertarget{the-birth-death-parameters}{%
-\subsection{The birth death parameters}\label{the-birth-death-parameters}}
-
-First let's set the overall birth death parameters (a speciation of 1.2 and an extinction of 0.3):
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{## The birth death parameters}
-\NormalTok{bd_params <-}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{speciation =} \FloatTok{1.2}\NormalTok{, }\DataTypeTok{extinction =} \FloatTok{0.3}\NormalTok{)}
-\CommentTok{## The stopping rule}
-\NormalTok{stop_rule <-}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{max.time =} \DecValTok{4}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-And let's quickly test if the simulation works by running a birth death tree without traits and anything:
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\KeywordTok{set.seed}\NormalTok{(}\DecValTok{0}\NormalTok{) }\CommentTok{# 123 also works}
-\CommentTok{## Running a birth death tree }
-\NormalTok{results <-}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ bd_params,}
-                \DataTypeTok{stop.rule =}\NormalTok{ stop_rule)}
-
-\CommentTok{## And displaying the results}
-\KeywordTok{plot}\NormalTok{(results)}
-\KeywordTok{axisPhylo}\NormalTok{()}
-\end{Highlighting}
-\end{Shaded}
-
 \includegraphics{dads_manual_files/figure-latex/unnamed-chunk-68-1.pdf}
 
-\hypertarget{the-traits}{%
-\subsection{The traits}\label{the-traits}}
-
-Then let's create the traits:
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{## The trait object}
-\NormalTok{two_traits <-}\StringTok{ }\KeywordTok{make.traits}\NormalTok{(}\DataTypeTok{process =} \KeywordTok{c}\NormalTok{(BM.process, OU.process), }\DataTypeTok{n =} \KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{, }\DecValTok{2}\NormalTok{),}
-                          \DataTypeTok{trait.names =} \KeywordTok{c}\NormalTok{(}\StringTok{"Distance"}\NormalTok{, }\StringTok{"Variables"}\NormalTok{))}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{## Visualise the 1D trait (BM.process)}
-\KeywordTok{plot}\NormalTok{(two_traits, }\DataTypeTok{trait =} \DecValTok{1}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-70-1.pdf}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{## Visualise the 2D trait (uncorrelated OU.process)}
-\KeywordTok{plot}\NormalTok{(two_traits, }\DataTypeTok{trait =} \DecValTok{2}\NormalTok{, }\DataTypeTok{use.3D =} \OtherTok{TRUE}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-Note that specifying the correlation matrix is not necessary here \emph{per se} (since no correlation is the default) but we will modify it through an event later on so it's good practice to have it properly stated before hand (to avoid any confusion).
-
-We can then run the tree with the traits:
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\KeywordTok{set.seed}\NormalTok{(}\DecValTok{0}\NormalTok{)}
-\CommentTok{## Running a birth death tree }
-\NormalTok{results <-}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params  =}\NormalTok{ bd_params,}
-                \DataTypeTok{stop.rule  =}\NormalTok{ stop_rule,}
-                \DataTypeTok{traits     =}\NormalTok{ two_traits)}
-
-\CommentTok{## And displaying the results}
-\KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow =} \KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{,}\DecValTok{2}\NormalTok{))}
-\KeywordTok{plot}\NormalTok{(results, }\DataTypeTok{trait =} \DecValTok{1}\NormalTok{, }\DataTypeTok{main =} \StringTok{"Trait 1: Distance"}\NormalTok{)}
-\KeywordTok{plot}\NormalTok{(results, }\DataTypeTok{trait =} \KeywordTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{,}\DecValTok{3}\NormalTok{), }\DataTypeTok{main =} \StringTok{"Variables correlation"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-72-1.pdf}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{## Visualising the tree in 3D}
-\KeywordTok{plot}\NormalTok{(results, }\DataTypeTok{trait =} \KeywordTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{,}\DecValTok{3}\NormalTok{), }\DataTypeTok{use.3D =} \OtherTok{TRUE}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\hypertarget{the-modifiers}{%
-\subsection{The modifiers}\label{the-modifiers}}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{## Speciation event is more likely if lineage's ancestor is further away from the mean trait value}
-\NormalTok{distance.modify <-}\StringTok{ }\ControlFlowTok{function}\NormalTok{(x, trait.values, lineage) }\KeywordTok{return}\NormalTok{(x }\OperatorTok{+}\StringTok{ }\NormalTok{x }\OperatorTok{*}\StringTok{ }\KeywordTok{abs}\NormalTok{(}\KeywordTok{parent.traits}\NormalTok{(trait.values, lineage)[}\DecValTok{1}\NormalTok{] }\OperatorTok{-}\StringTok{ }\KeywordTok{mean}\NormalTok{(trait.values[,}\DecValTok{1}\NormalTok{])))}
-
-\CommentTok{## Make a distance modifier}
-\NormalTok{distance.speciation <-}\StringTok{ }\KeywordTok{make.modifiers}\NormalTok{(}\DataTypeTok{speciation =}\NormalTok{ speciation,}
-                                      \DataTypeTok{modify =}\NormalTok{ distance.modify)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\KeywordTok{set.seed}\NormalTok{(}\DecValTok{42}\NormalTok{)}
-\CommentTok{## Running a birth death tree }
-\NormalTok{results <-}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params  =}\NormalTok{ bd_params,}
-                \DataTypeTok{stop.rule  =}\NormalTok{ stop_rule,}
-                \DataTypeTok{traits     =}\NormalTok{ two_traits,}
-                \DataTypeTok{modifiers  =}\NormalTok{ distance.speciation)}
-
-\CommentTok{## And displaying the results}
-\KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow =} \KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{,}\DecValTok{2}\NormalTok{))}
-\KeywordTok{plot}\NormalTok{(results, }\DataTypeTok{trait =} \DecValTok{1}\NormalTok{, }\DataTypeTok{main =} \StringTok{"Trait 1: Distance"}\NormalTok{)}
-\KeywordTok{plot}\NormalTok{(results, }\DataTypeTok{trait =} \KeywordTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{,}\DecValTok{3}\NormalTok{), }\DataTypeTok{main =} \StringTok{"Variables correlation"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-75-1.pdf}
-
-\hypertarget{step-2}{%
-\section{Step 2:}\label{step-2}}
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\setcounter{enumi}{1}
-\tightlist
-\item
-  ``Correlation event'': after reaching a certain number of taxa an event happens making the ``Variable'' trait dimensions become correlated.
-\end{enumerate}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{## The correlation change event}
-\NormalTok{events_list <-}\StringTok{ }\KeywordTok{make.events}\NormalTok{(}\DataTypeTok{event.name   =} \StringTok{"Correlation"}\NormalTok{,}
-                           \DataTypeTok{target       =} \StringTok{"traits"}\NormalTok{,}
-                           \DataTypeTok{condition    =} \KeywordTok{taxa.condition}\NormalTok{(}\DecValTok{30}\NormalTok{),}
-                           \DataTypeTok{modification =} \KeywordTok{update.traits}\NormalTok{(}
-                    \DataTypeTok{process.args =} \KeywordTok{list}\NormalTok{(}\DataTypeTok{Sigma =} \KeywordTok{matrix}\NormalTok{(}\KeywordTok{c}\NormalTok{(}\FloatTok{0.7}\NormalTok{, }\DecValTok{0}\NormalTok{, }\DecValTok{0}\NormalTok{, }\FloatTok{0.7}\NormalTok{), }\DecValTok{2}\NormalTok{, }\DecValTok{2}\NormalTok{)),}
-                    \DataTypeTok{trait.names  =} \StringTok{"Variables"}\NormalTok{))}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\KeywordTok{set.seed}\NormalTok{(}\DecValTok{8}\NormalTok{)}
-\CommentTok{## Running a birth death tree }
-\NormalTok{results <-}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params  =}\NormalTok{ bd_params,}
-                \DataTypeTok{stop.rule  =}\NormalTok{ stop_rule,}
-                \DataTypeTok{traits     =}\NormalTok{ two_traits,}
-                \DataTypeTok{modifiers  =}\NormalTok{ distance.speciation,}
-                \DataTypeTok{events     =}\NormalTok{ events_list)}
-
-\CommentTok{## And displaying the results}
-\KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow =} \KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{,}\DecValTok{2}\NormalTok{))}
-\KeywordTok{plot}\NormalTok{(results, }\DataTypeTok{trait =} \DecValTok{1}\NormalTok{, }\DataTypeTok{main =} \StringTok{"Trait 1: Distance"}\NormalTok{)}
-\KeywordTok{plot}\NormalTok{(results, }\DataTypeTok{trait =} \KeywordTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{,}\DecValTok{3}\NormalTok{), }\DataTypeTok{main =} \StringTok{"Variables correlation"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-77-1.pdf}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{## Visualising the tree in 3D}
-\KeywordTok{plot}\NormalTok{(results, }\DataTypeTok{trait =} \KeywordTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{,}\DecValTok{3}\NormalTok{), }\DataTypeTok{use.3D =} \OtherTok{TRUE}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\hypertarget{step-3}{%
-\section{Step 3:}\label{step-3}}
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\setcounter{enumi}{2}
-\tightlist
-\item
-  ``Selective extinction event'': after reaching a certain point in time, lineages with their ``Variables'' values that are negative go extinct
-\end{enumerate}
-
-\hypertarget{bugged-from-here}{%
-\chapter{BUGGED FROM HERE}\label{bugged-from-here}}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{## Adding the selective event}
-\NormalTok{events_list <-}\StringTok{ }\KeywordTok{make.events}\NormalTok{(}\DataTypeTok{event.name   =} \StringTok{"Selection"}\NormalTok{,}
-                           \DataTypeTok{add          =}\NormalTok{ events_list,}
-                           \DataTypeTok{target       =} \StringTok{"taxa"}\NormalTok{,}
-                           \DataTypeTok{condition    =} \KeywordTok{time.condition}\NormalTok{(}\DecValTok{4}\NormalTok{),}
-                           \DataTypeTok{modification =} \KeywordTok{trait.extinction}\NormalTok{(}\DataTypeTok{x =} \DecValTok{0}\NormalTok{, }\DataTypeTok{condition =} \StringTok{`}\DataTypeTok{<}\StringTok{`}\NormalTok{, }\DataTypeTok{trait =} \DecValTok{2}\NormalTok{))}
-\end{Highlighting}
-\end{Shaded}
-
-Problem with \texttt{modification} argument here
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{## Running a birth death tree }
-\NormalTok{results <-}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params  =}\NormalTok{ bd_params,}
-                \DataTypeTok{stop.rule  =}\NormalTok{ stop_rule,}
-                \DataTypeTok{traits     =}\NormalTok{ two_traits,}
-                \DataTypeTok{modifiers  =}\NormalTok{ distance.speciation,}
-                \DataTypeTok{events     =}\NormalTok{ events_list)}
-
-\CommentTok{## And displaying the results}
-\KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow =} \KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{,}\DecValTok{2}\NormalTok{))}
-\KeywordTok{plot}\NormalTok{(results, }\DataTypeTok{trait =} \DecValTok{1}\NormalTok{, }\DataTypeTok{main =} \StringTok{"Trait 1: Distance"}\NormalTok{)}
-\KeywordTok{plot}\NormalTok{(results, }\DataTypeTok{trait =} \KeywordTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{,}\DecValTok{3}\NormalTok{), }\DataTypeTok{main =} \StringTok{"Variables correlation"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\hypertarget{step-4-and-5}{%
-\section{Step 4 and 5:}\label{step-4-and-5}}
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\setcounter{enumi}{3}
-\tightlist
-\item
-  ``Founding event'': after reaching a value for the ``Distance'' trait, a founding event happens with a new birth-death process with no extinction and where the ``Variable'' is now a 2D correlated Brownian Motion.
-\item
-  ``Mass extinction event'': after reaching a certain number of taxa in the ``Founding event'', 80\% of taxa go extinct randomly.
-\end{enumerate}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{## The new trait for the founding event}
-\NormalTok{new_traits <-}\StringTok{ }\KeywordTok{make.traits}\NormalTok{(}\DataTypeTok{process =} \KeywordTok{c}\NormalTok{(BM.process, BM.process), }\DataTypeTok{n =} \KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{, }\DecValTok{2}\NormalTok{),}
-                          \DataTypeTok{trait.names =} \KeywordTok{c}\NormalTok{(}\StringTok{"Distance"}\NormalTok{, }\StringTok{"Variables"}\NormalTok{))}
-
-\CommentTok{## The mass extinction event after reaching 50 taxa}
-\NormalTok{mass_extinction <-}\StringTok{ }\KeywordTok{make.events}\NormalTok{(}\DataTypeTok{target       =} \StringTok{"taxa"}\NormalTok{,}
-                               \DataTypeTok{condition    =} \KeywordTok{taxa.condition}\NormalTok{(}\DecValTok{50}\NormalTok{),}
-                               \DataTypeTok{modification =} \KeywordTok{random.extinction}\NormalTok{(}\FloatTok{0.8}\NormalTok{))}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{## The selective event}
-\NormalTok{events_list <-}\StringTok{ }\KeywordTok{make.events}\NormalTok{(}\DataTypeTok{event.name   =} \StringTok{"Founding"}\NormalTok{, }
-                           \DataTypeTok{add          =}\NormalTok{ events_list,}
-                           \DataTypeTok{target       =} \StringTok{"founding"}\NormalTok{,}
-                           \DataTypeTok{condition    =} \KeywordTok{trait.condition}\NormalTok{(}\DecValTok{3}\NormalTok{),}
-                           \DataTypeTok{modification =} \KeywordTok{founding.event}\NormalTok{(}\DataTypeTok{traits    =}\NormalTok{ new_traits,}
-                                                         \DataTypeTok{events    =}\NormalTok{ mass_extinction,}
-                                                         \DataTypeTok{bd.params =} \KeywordTok{list}\NormalTok{(}
-                                                            \DataTypeTok{speciation =} \DecValTok{2}\NormalTok{,}
-                                                            \DataTypeTok{extinction =} \DecValTok{0}\NormalTok{)),}
-                           \DataTypeTok{additional.args =} \KeywordTok{list}\NormalTok{(}\DataTypeTok{prefix =} \StringTok{"found_"}\NormalTok{))}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{## Running a birth death tree }
-\NormalTok{results <-}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params  =}\NormalTok{ bd_params,}
-                \DataTypeTok{stop.rule  =}\NormalTok{ stop_rule,}
-                \DataTypeTok{traits     =}\NormalTok{ two_traits,}
-                \DataTypeTok{modifiers  =}\NormalTok{ distance.speciation,}
-                \DataTypeTok{events     =}\NormalTok{ events_list,}
-                \DataTypeTok{null.error =} \DecValTok{100}\NormalTok{)}
-
-\CommentTok{## And displaying the results}
-\KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow =} \KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{,}\DecValTok{2}\NormalTok{))}
-\KeywordTok{plot}\NormalTok{(results, }\DataTypeTok{trait =} \DecValTok{1}\NormalTok{, }\DataTypeTok{main =} \StringTok{"Trait 1: Distance"}\NormalTok{)}
-\KeywordTok{plot}\NormalTok{(results, }\DataTypeTok{trait =} \KeywordTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{,}\DecValTok{3}\NormalTok{), }\DataTypeTok{main =} \StringTok{"Variables correlation"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
 \bibliography{../references.bib}

### 2021-01-20 inst/gitbook/05_thorough_example.Rmd
Create 05_thorough_example.Rmd

@@
 
+```{r, webgl=TRUE}
+## Visualising the tree in 3D
+plot(results, trait = c(2,3), use.3D = TRUE)
+```
+
+> BUGGED FROM HERE!
+
 ## Step 2:
- 2. "Correlation event": after reaching a certain number of taxa an event happens making the "Variable" trait dimensions become correlated.
+ 2. "Correlation event": after reaching time 2 an event happens making the "Variable" trait dimensions become correlated.
 
@@
  3. "Selective extinction event": after reaching a certain point in time, lineages with their "Variables" values that are negative go extinct
-
-# BUGGED FROM HERE
 

### 2021-10-28 inst/gitbook/01_getting_started.Rmd
updated to linux

@@
 ```
-> Note that here it is not necessary to specify `extinction = 1` since this is the default option, you can always just change the parameter of interest (e.g.
-changing `extinciton = 0` to `extinction = 1/3`).
+> Note that here it is not necessary to specify `speciation = 1` since this is the default option, you can always just change the parameter of interest (e.g.
+changing `extinction = 0` to `extinction = 1/3`).
 However, we think it's good practice to attribute both parameters specifically to avoid any confusion.
@@
 
-Note however that the printing information is now updated to state that you've add a modifier (even though it's a default one).
+Note however that the printing information is now updated to state that you've added a modifier (even though it's a default one).
 

### 2022-03-11 inst/gitbook/02_make.traits.Rmd
typos + minors
Update 02_make.traits.Rmd

@@
     ## and depending on branch (edge) length
-    result <- rnorm(n = 1, mean = mean, sd = sd * branch.length)
+    result <- rnorm(n = 1, mean = mean, sd = sd * edge.length)
 

### 2022-05-16 inst/gitbook/02_make.traits.Rmd
sync

@@
     ## and depending on branch (edge) length
-    result <- rnorm(n = 1, mean = x0, sd = sd * edge.length)
+    result <- rnorm(n = 1, mean = x0, sd = sqrt(sd^2 * edge.length))
 
@@
     ## and depending on branch (edge) length
-    result <- rnorm(n = 1, mean = mean, sd = sd * edge.length)
+    result <- rnorm(n = 1, mean = mean, sd = sqrt(sd^2 * edge.length))
 

### 2023-01-12 inst/gitbook/02_make.traits.Rmd
edits and cleanings

--- 2022-03-11 inst/gitbook/02_make.traits.Rmd
@@
 
+### The syntax (how to code a process?)
+
 Trait processes in `dads` are functions that must always intake the following arguments by default.
@@
 
-For example, the following function would be a valid process (though not dependent on either the previous state nor the branch length):
+For example, the following function would be a valid process that always generate the _true_ trait value: 42!.
+In this example, the process is not dependent on neither the previous state (`x0`) and the branch length (`edge.length`).
 
 ```{r}
-## A valid (but useless) process
-valid.process <- function(x0, egde.length = 1, ...) {
+## A valid (but useless?) process
+valid.process <- function(x0 = 0, edge.length = 1, ...) {
     return(42)
@@
 
-> Note that the argument `edge.length` is set to `1` by default.
-In general we highly recommend to set all arguments but `x0` to a default value (this really helps the speeding up the `dads` function).
+> Note that in this function definition the arguments `x0` and `edge.length` have a default value set to `0` and `1` respectively.
+In practice, these arguments are effectively set to the correct values in the `dads` internal function (i.e.
+whatever `x0` and `edge.length` are at that specific time of the process) but providing a default can help speed up the algorithms (specifically all the internal checks).
 
@@
 ## A wrongly formated process
-invalid.process <- function(x0) {
+invalid.process <- function(x0 = 0) {
     return(rnorm(1, mean = x0))
@@
 
-The `dads` package proposes inbuilt processes, namely a multidimensional Brownian motion (`BM.process`) or a a multidimensional Ornstein-Uhlenbeck process (`OU.process`).
+### Using a `"process"` in `dads`
+
+You can design your own process as a function (as long as it has a valid syntax).
+Alternatively, the `dads` package proposes inbuilt processes, namely a multidimensional Brownian motion (`BM.process`) or a a multidimensional Ornstein-Uhlenbeck process (`OU.process`).
 You can find the list of implemented process by looking at the `?trait.process` manual page in `R`.
@@
 
-Note that you can see the multiple options for plotting the trait process by looking at `?plot.dads` manual. Furthermore, you can look at what's actually in the object using:
+Note that you can see the multiple options for plotting the trait process by looking at `?plot.dads` manual.
+Furthermore, you can look at what's actually in the object using this specific syntax (this applies to every object handled by the `dads` package):
 
@@
 make.traits(c(BM.process, OU.process), n = 3, start = c(1,1))
+```
+
+### What even is a trait?
+
+The definition of what a trait is can vary quiet a lot depending on the context of an analysis and the field.
+<!--TODO: add REF e.g.  https://onlinelibrary.wiley.com/doi/full/10.1002/ece3.8321 -->
+Because it would be impossible to accommodate all definitions in `dads` we had to go with an arbitrary one: a trait is whatever you define as a trait!
+A trait can be uni-dimensional as the measurement of a feature of an organism (e.g.
+leaf surface, femur length, etc...) but can also be multi-dimensional description of a feature, for example in 3D geometric morphometric, a trait could be defined as "position of landmark X" (which will be a trait with three dimensions, _x_, _y_ and _z_) or in ecology, the location of a plant can be a single 2D trait (expressed as latitude and longitude).
+In `dads`, the **process** corresponds to this trait definition (e.g.
+a process can be of n-dimensions and represents one organisms feature) and the **traits** represents the number of dimensions in total.
+So in the examples above, this is how the following traits are interpreted by `dads`:
+
+```{r, eval = FALSE}
+## Three traits with one process:
+make.traits(BM.process, n = 3, start = c(1,2,3))
+## Six traits with two processes:
+make.traits(c(BM.process, OU.process), n = 3)
+## Five traits with two processes
+make.traits(c(BM.process, OU.process), n = c(1, 4))
 ```
@@
 
-This bit is more for development.
+**This bit is more for development.**
 We highly suggest leaving `test = TRUE` so that `make.traits` returns an error if a process or its additional arguments (`process.args`) are not formatted correctly.
-`make.traits` will error if the trait cannot be directly passed to `dads`.
+`make.traits` will send an error if the trait cannot be directly passed to `dads`.
 However, in some specific cases (again, probably mainly for development and debugging) it could be useful to skip the tests using `test = FALSE`.
@@
 You can be imaginative and creative when designing your own process but here are two detailed example functions for a unidimensional Brownian Motion and Ornstein-Uhlenbeck process that you can use for a start (or not).
-Remember it is good practice for `dads` processes to set all the arguments but `x0` with default values (just in case).
-Also, note that the functions below are not equal to the already implemented `BM.process` and `OU.process` but are rather generalised/simplified version that you can use as a template
+Remember it is good practice for `dads` processes to set all the arguments with default values (just in case).
+
+> Note that the functions below are not equal to the already implemented `BM.process` and `OU.process` but an easier to edit version that you can use as a template:
 
@@
 ## A simple Brownian motion process
-my.BM.process <- function(x0, edge.length = 1, sd = 1, ...) {
+my.BM.process <- function(x0 = 0, edge.length = 1, sd = 1, ...) {
     ## Drawing a random number from a normal distribution
@@
 ## A simple Ornstein-Uhlenbeck motion process
-my.OU.process <- function(x0, edge.length = 1, var = 1, alpha = 1, ...) {
+my.OU.process <- function(x0 = 0, edge.length = 1, var = 1, alpha = 1, ...) {
     ## Calculate the mean based on alpha

### 2023-01-12 inst/gitbook/03_make.modifiers.Rmd
edits and cleanings

@@
 
+## The default modifier (how the process is working)
+
+The modifiers modify the core of the birth death process as implemented in `dads`.
+
+<!-- TODO: make that into a shiny app animation (or just in html?) -->
+
+By default, the birth-death process in `dads` uses this modifier:
+
+```{r}
+## What is actually in the default modifier?
+print(make.modifiers(), all = TRUE)
+```
+
+This contains a lot of information, much of it is actually not useful for using the modular aspects of `dads` at a high user level.
+I.e.
+unless you want to code very specific things, you won't need most of the information.
+The essential are these three functions in the elements named `"waiting$fun"`, `"selecting$fun"` and `"speciating$fun"`:
+
+ 1.
+`"waiting$fun"` is the branch length function (defined in details below) which returns a randomly drawn number from an exponential distribution with the rate of the number of taxa multiplied by the speciation and extinction rate: $n \times (\lambda + \mu)$ (here is uses the extra function `sum` for internal modularity reasons).
+**This function is responsible for the growth of the length/age of the tree**
+
+```{r, eval = FALSE}
+rexp(1, sum(lineage$n * (bd.params$speciation + bd.params$extinction)))
+```
+ With `lineage$n` being the number of lineages and `bd.params$speciation` and `bd.params$extinction` the speciation and extinction parameters (these specific terms are defined in details below).
+
+ 2.
+`"selecting$fun"` is the selection function (defined in details below) which, after the waiting time defined above returns a randomly selected lineage (a tip) among the existing ones.
+**This function is responsible for the branch selection**.
+
+```{r, eval = FALSE}
+sample(lineage$n, 1)
+```
+
+ 3.
+`"speciating$fun"` is the speciation function (defined in details below) which randomly draws a number (`runif(1)`) and depending on the speciation and extinction parameters makes the selected lineage (from 2.) speciate (`TRUE`) or go extinct (`FALSE`).
+**This function is responsible for the speciation or extinction of species**.
+
+```{r, eval = FALSE}
+runif(1) < (bd.params$speciation/(bd.params$speciation + bd.params$extinction)))
+```
+
+We will have a look at all these aspects below in more details which should make things clearer (but you can always refer to the default here for information).
+
 ## The branch length function (`branch.length`)
@@
 
-By default, branch length (or waiting or growth) is a randomly drawn number from an exponential distribution with the rate of the number of taxa multiplied by the speciation and extinction rate: $n \times (\lambda + \mu)$ (where $n$ is the number of taxa currently present in the simulation, $\lambda$ and $\mu$ are respectively the speciation and extinction rates).
+By default, branch length (or waiting/growth) is a randomly drawn number from an exponential distribution with the rate of the number of taxa multiplied by the speciation and extinction rate: $n \times (\lambda + \mu)$ (where $n$ is the number of taxa currently present in the simulation, $\lambda$ and $\mu$ are respectively the speciation and extinction rates).
 This default function is simply called `branch.length` in `dads` and can be used as a modifier as follows:
@@
 
-It is of course to use more complex branch length modifiers that intakes different conditions and specific modification rather than simply always output a value of one.
+It is of course possible to use more complex branch length modifiers that intakes different conditions and specific modification rather than simply always output a value of one.
 
@@
 
-It is of course possible to design some more advanced function to interact with the birth death process.
 You can create a function for `branch.length`, `selection` and `speciation` that involve any of the following arguments:
@@
 Therefore any argument can be left undeclared or missing and is always handled internally.
-For example, if you did not declare `n.taxa` as a function argument but are using `n.taxa` in the function, `n.taxa` will be detected and treated as a current argument automatically as set accordingly within the birth death process (e.g.
-`n.taxa` will be set to the current number of taxa every iteration of the process).
+For example, if you did not declare `lineage$n` as a function argument but are using `lineage$n` in the function, `lineage$n` will be detected and treated as a current argument automatically as set accordingly within the birth death process (e.g.
+`lineage$n` will be set to the current number of taxa every iteration of the process).
 
@@
 I.e.
-for discrete numbers of taxa, the branch length increases by jumps (ten fold):
+for discrete numbers of taxa, the branch length increases by jumps (ten fold) every 5 taxa:
 
@@
 The `selection` function is used in the birth death process to know which lineage to select when running a speciation (or extinction!) event.
-By default, this function randomly selects one taxon that is currently not extinct (using: `sample(1:n.taxa, 1))`.
+By default, this function randomly selects one taxon that is currently not extinct (using: `sample(1:lineage$n, 1))`.
 Similarly as `branch.length` it is possible to modify this part of the birth death process.
@@
 
-> Note that here the function can only intake the [allowed arguments as described above](#allowarguments) (here `n.taxa`: the number of current living taxa).
+> Note that here the function can only intake the [allowed arguments as described above](#allowarguments) (here `lineage$n`: the number of current living taxa).
 
@@
 Again, it is of course possible to make the modifier more complex and in combination with other elements of the tree.
-For example, we can create a `"dads"` object that also creates a trait a add to it a `selection` modifier that only selects for tips with positive trait values.
+For example, we can create a `"dads"` object that also generates a BM trait and add to it this object a `selection` modifier that only selects tips with positive trait values (only species with positive trait values will speciate).
 
@@
 select.positive <- function(trait.values, lineage) {
-    ## Selecting the taxa with positive traits only
+
+    ## Selecting the taxa names with positive values for the first trait
     positives <- as.integer(rownames(trait.values)[which(trait.values[, 1] >= 0)])
 
+    ## Combine the descendants of the current lineages (lineage$parents)
+    ## with the species that have speciated (seq_along(lineages$split))
+    ## to have a table of pairs of parents/splits
+    parents_split_table <- cbind(lineage$parents, seq_along(lineage$split))
     ## Select the current taxa that descend from a node with a positive value
-    positive_living <- cbind(lineage$parents, seq_along(lineage$split)
-                             )[which(lineage$parents %in% positives), 2]
+    positive_living <- parents_split_table[which(lineage$parents %in% positives), 2]
 
@@
 ## Generating a new tree with this modifier
-set.seed(0)
+set.seed(2)
 modified_tree <- dads(bd.params = bd_params,
@@
 
-Note how every lineage end up going extinct!
+> Note how loads of lineages end up going extinct even if the extinction parameter is set to 0!
+
 And again, we can make some more advanced modifiers: for example, one where a tip always goes extinct if their ancestor has a negative trait value (here we will also introduce the utility function `parent.trait` that automatically selects the trait values of the parent of the current lineage.

### 2023-01-12 inst/gitbook/index.Rmd
Update index.Rmd

--- 2020-12-09 inst/gitbook/index.Rmd
@@
 
+## Who is this manual for?
+
+This manual explains how `dads` work in as much details as possible and aims to give you the keys to design your own diversity and disparity simulations.
+
+The first section `Getting started` is aimed for people with a `R` beginner level:
+ * you know what a package is - and how to install it;
+ * you know what a function is - and what arguments are;
+ * you have already (vaguely) heard that they are different types of objects in R - like matrices (`"matrix"`) or trees (`"phylo"`);
+ * and you have already looked for documentation and bugs online.
+
+The sections after that are a bit more advanced and requires the following level:
+ * you have already created functions before;
+ * you know how to subset elements of a list - i.e. you understand the `list$element` syntax;
+ * you are aware that both `TRUE == 1` and `FALSE == 0` are `TRUE`;
+ * you have heard that the objects in R have classes and that it sometimes matters - for example you know that R can differentiate between `as.numeric(1)` and `as.integer(1)`;
+ * you owe part of your life to Stack Overflow.
+
 ## Installing and running the package

### 2023-01-12 inst/gitbook/01_getting_started.Rmd
clarified bits in manual

@@
 ```
+
 > Note that here it is not necessary to specify `speciation = 1` since this is the default option, you can always just change the parameter of interest (e.g.
@@
 BM_trait      <- make.traits()
+
 set.seed(1)
@@
 For that we need to create a modifiers object that modifies the `speciation` process with a specific condition and a specific modification when that condition is met.
-First we are going to create the modification function.
-This function must intake the argument `x` and, in our case, return a logical value: `TRUE` is for speciate and `FALSE` is for go extinct.
-We want this `modify` function to always return `FALSE` (go extinct) since it will be triggered by `condition` (we will see that in a minute).
+For a speciation to occur (and a species to not go extinct), the algorithm draws a random value _x_ between 0 and 1 and if the value is smaller than the speciation parameter divided by the speciation + extinction parameter, the lineage speciates ($x < (\lambda \ (\lambda + \mu))$), else it goes extinct.
 
-```{r}
-## Going extinct
-staying.alive <- function(x) return(TRUE)
-```
+Here we can create a modifier forcing the lineage to go extinct every time a specific condition is met (say the lineage trait value is negative).
+In other words, if the lineages trait value is negative, make sure that _x_ is equal to 1 (so that the speciation condition $x < (\lambda \ (\lambda + \mu))$ is never met).
 
-Now we want this function to only trigger when an ancestor has a negative trait value.
+First we need to create a condition function to trigger this (non) speciation event.
 We can do that by specifying our `condition` function (when the ancestor is trait is negative) apply our modification (here the `staying.alive` function).
 For that we can use the `parent.traits` utility function that is optimised for accessing traits in the birth death process (but you can of course write your own).
-This function intakes the `trait.values` and `parent.lineage` arguments, two arguments that you can leave named as they are to facilitate `dads`s understanding of what you which to asses:
+This function intakes the `trait.values` and `parent.lineage` arguments, two arguments that you can leave named as they are to facilitate `dads`' understanding of what you want to assess:
 
@@
 
-Note that we use the function `all` here to evaluate all traits (if the data is multidimensional). 
+> Note that we use the function `all` here to evaluate all traits: i.e. if the data has more than one trait we trigger the modification only if all the trait values are negative.
+
+The we are going to create the modification function.
+This function must intake the argument `x` and, in our case, returns the same value no matter what: `1` so that speciation never happens when the `negative.ancestor` function is triggered.
+
+```{r}
+## Always go extinct
+going.extinct <- function(x) return(1)
+```
+
 We can then provide these two functions (the condition `negative.ancestor` and how to modify the speciation event when this condition is met `staying.alive`).
@@
 negatives_extinct <- make.modifiers(
-                        ## The speciation function (default)
-                        speciation = speciation,
-                        ## What to modify
-                        modify     = staying.alive,
-                        ## When to modify it
-                        condition  = negative.ancestor)
+            ## If the following condition is met...
+            condition = negative.ancestor, # Does the lineage has a negative ancestor?
+            ## Then apply the following modifier...
+            modify = going.extinct, # The species goes extinct
+            ## To the the speciation process.
+            speciation = speciation) # Here the speciation() process is default
+
 ## What's in it?
@@
 We can now compare the two trees and their trait values.
-Note that we've used the same starting seed for both trees so the only thing differing between them is the modifier!
-Also, although species
+Note that we've used the same starting seed for both trees so the only thing differing between them is the `"modifier"` object which leads to very different trees!
 

### 2023-01-19 paper/dads_paper.tex
Papare INIT

% 2961w - 30p

\documentclass[12pt,letterpaper]{article}
\usepackage{natbib}
\usepackage{pdflscape}
\usepackage{fullpage}
\usepackage{url}
\usepackage{epsfig}
\usepackage{caption}
\usepackage{hyperref}
\usepackage{enumerate}
\usepackage{dcolumn}
\usepackage{lineno}
\usepackage[T1]{fontenc}
\usepackage{textcomp}
\usepackage{float}
\usepackage[osf]{mathpazo}
\restylefloat{table}
\newcolumntype{d}[1]{D{.}{.}{#1}}

\pagenumbering{arabic}

%Pagination style and stuff
\linespread{2}
\raggedright
\setlength{\parindent}{0.5in}
\setcounter{secnumdepth}{0} 
\renewcommand{\section}[1]{%
\bigskip
\begin{center}
\begin{Large}
\normalfont\scshape #1
\medskip
\end{Large}
\end{center}}
\renewcommand{\subsection}[1]{%
\bigskip
\begin{center}
\begin{large}
\normalfont\itshape #1
\end{large}
\end{center}}
\renewcommand{\subsubsection}[1]{%
\vspace{2ex}
\noindent
\textit{#1.}---}
\renewcommand{\tableofcontents}{}
%\bibpunct{(}{)}{;}{a}{}{,}

%---------------------------------------------
%
%       START
%
%---------------------------------------------

\newcommand{\dads}{\texttt{dads} }

\begin{document}

%Running head
\begin{flushright}
Version dated: \today
\end{flushright}
\bigskip
\noindent RH: \dads package.

\bigskip
\medskip
\begin{center}

\noindent{\Large \bf \dads: a modular \texttt{R} package for simulating diversity and traits data.} 
\bigskip

\noindent {\normalsize \sc Thomas Guillerme$^{1,*}$, Natalie Cooper$^{2}$, Andrew P. Beckerman$^{1}$, and Gavin H. Thomas$^{1,3}$}\\
\noindent {\small \it 
$^1$School of Biosciences, University of Sheffield, Sheffield, S10 2TN, United Kingdom.\\
$^2$Natural History Museum, Cromwell Road, London, SW7 5BD, United Kingdom.\\
$^3$Bird Group, Department of Life Sciences, the Natural History Museum at Tring, Tring, United Kingdom.\\}

\end{center}
\medskip
\noindent{*\bf Corresponding author.} \textit{guillert@tcd.ie}\\  
\vspace{1in}

%Line numbering
\modulolinenumbers[1]
\linenumbers

%---------------------------------------------
%
%       ABSTRACT
%
%---------------------------------------------

\newpage
\begin{abstract}

    \begin{enumerate}
        \item Simulating biological realistic data can be an important step to understand and investigate biological mechanisms.
        Like null models, base line models or something else models, they allow us to generate a pattern that arises from controlled processes and thought through mechanisms.

        \item In evolutionary biology, these simulations often involve the need of an evolutionary process where descent with modification is at the core of how the simulated data is generated.
        This can be made much more complicated with loads of different stuff needing to be taken into account to affect the simulations (e.g. traits, mechanisms such as competition or events such as mass extinctions).

        \item Here I present the \dads package, a modular \texttt{R} package for diversity and trait disparity simulations.
        This package is based on a standard birth death algorithm that can be easily tuned to the user specific needs by designing their specific trait process, birth-death modifiers and events.
        It also provides a tidy interface through the \dads object, allowing users to easily run reproducible simulations.

        \item The \dads package also comes with an extend manual regularly updated following users' questions or suggestions.

    \end{enumerate}

\end{abstract}

\noindent (Keywords: diversity, disparity, simulations, birth-death, null-models, ecology, evolution)\\

\vspace{1.5in}

\newpage

%---------------------------------------------
%
%       INTRODUCTION
%
%---------------------------------------------

\section{Introduction}

% Simulations in statistics applied to biology

% What already exists

% The problems with the absence of modularity

% The solution with dads

\section{Description}
% Very briefly how does it work

\begin{figure}[!htbp]
\centering
   \includegraphics[width=1\textwidth]{workflowsvg.eps} 
\caption{\dads package workflow: @@@.}
\label{Fig:workflow}
\end{figure}

\subsection{Simulating diversity}

\subsection{Simulating traits (disparity)}

\subsection{Using modular and dynamic modifiers}

\subsection{Introducing timed or conditional events in the simulations}

\section{Discussion}

\subsection{Modularity}

\subsection{\dads compared to other packages}

\subsection{Further directions}

\section{Conclusion}
The \dads is modular and nice to use.

\section{Package location}
The \dads package is available on the CRAN at \url{https://cran.r-project.org/web/packages/dads/index.html} or on GitHub at \url{https://github.com/TGuillerme/dads} with more associated information.
All the versions of the package are archived on ZENODO with associated DOI \url{https://zenodo.org/@@@}.

\section{Acknowledgments}
Thanks to Alex Slavenko for comments on the early stage of the development of this package. Thanks to @@@. This work was funded by UKRI-NERC Grant NE/T000139/1 and a Royal Society University Research Fellowship (URF R 180006 to GHT).

\bibliographystyle{sysbio}
\bibliography{References}

\end{document}

### 2023-01-19 paper/dads_paper.tex
Update dads_paper.tex

@@
 
+% 3000-4000 words
+
 \newpage
@@
 
-% Simulations in statistics applied to biology
+% Simulations in biology
+Comparing biological patterns is one of the key ways (and arguably the only) to understand mechanisms in evolutionary biology.
+This lead to the development of phylogenetic comparative methods as key sub-field in ecology and evolution @CITE-some-PCMreview.
+As indicated in the name, phylogenetic comparative methods rely on comparing patterns in a phylogenetical concepts to understand biological mechanisms or concepts.
+These comparisons can be done between observed patterns under different conditions suggesting different processes or mechanisms (e.g.
+trait patterns for species on an island vs.
+on the mainland @CITE-some-islandPCM-paper) or by comparing some observed pattern to one simulated under null or base conditions @CITE-paper-with-null-comparisons (but see @CITE-null-phylo-paper for distinctions or common misconceptions).
+In theory, workers can follow the research pipeline of thinking of a specific mechanism (e.g.
+mass extinction allowing the surviving species to acquire new morphologies), collect some data to test this mechanism (e.g.
+some traits of species across and extinction event) and then compare these patterns to one simulated under no specific conditions (e.g.
+a null model where the traits evolve randomly regardless of an extinction event).
+@CITE-mark-extinction
 
 % What already exists
+In practice, these evolutionary simulations can be done relatively easily multiple times on computers using a birth-death process @CITE-something-describing-BD.
+A birth-death process is @yadiyada and has been routinely implemented in \texttt{R} (e.g. @CITE-ape, @CITE-diversitree).
+Traits can be simulating using the following processes @OU, @BM, @yadidada.
+The diversity of these traits through time can be called disparity in palaeontology @CITE-disparity-review or functional diversity in ecology (for functional traits; @CITE-setafanospaper).
+In \texttt{R}, this can be done with several already well used and well documented packages.
+For example if you want to simulate diversity through time, you can use \texttt{TreeSim} to simulate diversity under a set of specific parameters (e.g.
+speciation and extinction) with some events disrupting the simulations (e.g.
+mass extinctions) @CITE-TreeSim.
+One can even improve on generating these patterns using \texttt{FossilSim} to generate a pattern that will take into account fossilization processes @CITE-FossilSim.
+On the other hand, if you need to simulate both diversity and traits through time, this can be done with specific parameters in \texttt{RPANDA} @CITE-RPANDA or in \texttt{PETER} @CITE-PETER where the traits are generated stochasticaly through time (given some process) during the birth-death process.
 
 % The problems with the absence of modularity
+Although the packages mentioned above (and more that we forgot to mention/don't know off) and are excellent (i.e.
+fast, reliable and documented) at the specific tasks they are designed for, they unfortunately don't allow much modifications beyond the parameters implemented for each specific function.
+This can be problematic for generating specific simulations to be used as a specific null model for some specific questions.
+For example, \texttt{TreeSim} can simulate a birth-death tree with some extinction event but is not designed to simulated one with an extinction event that leads to the birth-death process to be not diversity dependent anymore (e.g.
+simulating a release in selection pressure after the extinction process that leads to a different process dominating speciation).
+Or \texttt{PETER} is not designed to simulate a complex set of traits (say three correlated BM traits and two independent OU ones).
+Although we don't think this was the aim of these specific packages authors (not to create a strawman argument) we think this hampers the use of specific null model (or baseline model) to test specific dynamic mechanisms or to compare to specific datasets.
 
 % The solution with dads
+Therefore, we propose \dads a modular \texttt{R} package to simulate both diversity and disparity through time.
 
@@
 % Very briefly how does it work
+\dads is based on a core birth-death process implemented in the function \texttt{dads} that simulates diversity through time using the following steps:
+
+\begin{enumerate}
+    \item Waiting: Generating some waiting time corresponding to the growth in length of the tree (e.g. the duration of a lineage through time);
+    \item Selection: Choosing a lineage \textit{alive} at the time of the simulation;
+    \item Speciation: Choosing whether this species will go extinct or speciate leading to two new lineages.
+\end{enumerate}
+
+These three steps are repeated until the tree reaches the desired age or the desired number of species.
+If traits are simulated during the process, a fourth step is added:
+
+\indent 4. Trait: Generate some trait(s) value(s) based for the extinct species or the node leading to the two new lineages.
+This/these value(s) are generated taking into account the trait value of the extinct species/new node's ancestor and the time elapsed (generated in step 1).
+
+In \dads, these three or four steps are implemented as modular functions that the user can easily change using an internal class of objects called \texttt{``modifiers''} or \texttt{``traits''} (Fig.
+\ref{Fig:workflow}).
+The simulation then outputs a tree (of class \texttt{``phylo''} and a associated table of traits - \texttt{``matrix''}) that can be visualised using the \texttt{plot.dads} function.
 
@@
 \centering
-   \includegraphics[width=1\textwidth]{workflowsvg.eps} 
+   \includegraphics[width=1\textwidth]{../inst/gitbook/dads_structure.pdf} 
 \caption{\dads package workflow: @@@.}
@@
 \end{figure}
+% \texttt{"events"} are an additional category of modular objects discussed later.
 
 \subsection{Simulating diversity}
+
+To simulate a birth-death tree, two arguments are essentially needed: the birth-death parameters (\texttt{bd.params}) the rate of speciation (birth or \lambda), the rate of extinction (death or \mu - this parameter can be equal to 0) and \texttt{stop.rule}, a rule for when to stop the growth of the tree (e.g.
+when reaching 20 species).
+
+\texttt{## Running a simulation with a speciation rate of 1, an extinction rate of 0.1 and stopping when it reaches 20 taxa}
+\texttt{my_simulation <- dads(bd.params = list(speciation = 1, extinction = 0.1), stop.rule = list(max.taxa = 20))}
+
+By default, the three first steps are included in a default modifier that reproduces a simple birth-death process given the input birth and death parameters.
+
+\begin{enumerate}
+    \item Waiting: drawing a random number from an exponential distribution with a species dependent rate of $n \times (\lambda + \mu)$, where $n$ is the number of species currently alive in the process.
+    \item Selection: randomly choosing on of the $n$ taxa currently alive in the process;
+    \item Speciation: drawing a random number between 0 and 1; if that number is smaller than the speciation relative to the turnover ($\lambda / \lambda+\mu$), the lineage goes extinct %TODO: check!!!
+    ; else, the lineage speciates.
+\end{enumerate}
+
+These are included by default in a \texttt{dads} process but can be also set manually by creating a \texttt{``modifiers''} object with no options:
+
+\texttt{## Generating the default modifiers}
+\texttt{my_modifiers <- make.modifiers()}
+\texttt{## Running a simulation with a speciation rate of 1, an extinction rate of 0.1 and stopping when it reaches 20 taxa with the default modifiers}
+\texttt{my_simulation <- dads(bd.params = list(speciation = 1, extinction = 0), stop.rule = list(max.taxa = 20), modifiers = my_modifiers)}
+
+The modularity of the packages works with the ability to code the modifiers manually rather than using default ones.
+For example, to simulate the same process but hard coding the modifiers we can use the following
+
+\texttt{## }
 
@@
 \subsection{Using modular and dynamic modifiers}
+
+% Dynamic ones will be for bd.params = list(speciation = runif)
 
@@
 \section{Acknowledgments}
-Thanks to Alex Slavenko for comments on the early stage of the development of this package.
+Thanks to Mark Puttick and Alex Slavenko for comments on the early stage of the development of this package.
 Thanks to @@@.

### 2023-01-19 inst/gitbook/05_plotting_dads_objects.Rmd
added ploting tutorial

---
title: "`plot.dads`"
author: "Thomas Guillerme (guillert@tcd.ie)"
date: "`r Sys.Date()`"
bibliography: [../References.bib, ../packages.bib]
output:
  html_document: default
  pdf_document: default
---

```{r, echo = FALSE}
library(rgl)
knitr::knit_hooks$set(webgl = hook_webgl)
```

`"dads"` objects can be directly plotted in dads using many different options.

### 2023-01-19 inst/dads_paper.Rmd
dads paper (looks nicer in Rmd!)

---
title: "`dads`: a modular `R` package for simulating diversity and traits data."
author: "homas Guillerme, Natalie Cooper, Andrew P. Beckerman, and Gavin H. Thomas"
bibliography: references.bib
csl: mee.csl
date: "`r Sys.Date()`"
output:
  html_document:
    fig_width: 8
    fig_height: 8
    keep_tex: true
  self_contained: true
---

<!-- Keywords: diversity, disparity, simulations, birth-death, null-models, ecology, evolution -->

<!-- SET URL FOR [manual](dads_manual) -->

# Abstract

1. Simulating biological realistic data can be an important step to understand and investigate biological mechanisms.
Like null models, base line models or something else models, they allow us to generate a pattern that arises from controlled processes and thought through mechanisms.

2. In evolutionary biology, these simulations often involve the need of an evolutionary process where descent with modification is at the core of how the simulated data is generated.
This can be made much more complicated with loads of different stuff needing to be taken into account to affect the simulations (e.g. traits, mechanisms such as competition or events such as mass extinctions).

3. Here I present the `dads` package, a modular `R` package for diversity and trait disparity simulations.
This package is based on a standard birth death algorithm that can be easily tuned to the user specific needs by designing their specific trait process, birth-death modifiers and events.
It also provides a tidy interface through the `dads` object, allowing users to easily run reproducible simulations.

4. The `dads` package also comes with an extend manual regularly updated following users' questions or suggestions.

# Introduction

<!-- % Simulations in biology -->
Comparing biological patterns is one of the key ways (and arguably the only) to understand mechanisms in evolutionary biology.
This lead to the development of phylogenetic comparative methods as key sub-field in ecology and evolution @CITE-some-PCMreview.
As indicated in the name, phylogenetic comparative methods rely on comparing patterns in a phylogenetical concepts to understand biological mechanisms or concepts.
These comparisons can be done between observed patterns under different conditions suggesting different processes or mechanisms (e.g. trait patterns for species on an island vs. on the mainland @CITE-some-islandPCM-paper) or by comparing some observed pattern to one simulated under null or base conditions @CITE-paper-with-null-comparisons (but see @CITE-null-phylo-paper for distinctions or common misconceptions).
In theory, workers can follow the research pipeline of thinking of a specific mechanism (e.g. mass extinction allowing the surviving species to acquire new morphologies), collect some data to test this mechanism (e.g. some traits of species across and extinction event) and then compare these patterns to one simulated under no specific conditions (e.g. a null model where the traits evolve randomly regardless of an extinction event). @CITE-mark-extinction

<!-- % What already exists -->
In practice, these evolutionary simulations can be done relatively easily multiple times on computers using a birth-death process @CITE-something-describing-BD.
A birth-death process is @yadiyada and has been routinely implemented in `R` (e.g. @CITE-ape, @CITE-diversitree).
Traits can be simulating using the following processes @OU, @BM, @yadidada.
The diversity of these traits through time can be called disparity in palaeontology @CITE-disparity-review or functional diversity in ecology (for functional traits; @CITE-setafanospaper).
In `R`, this can be done with several already well used and well documented packages.
For example if you want to simulate diversity through time, you can use `TreeSim` to simulate diversity under a set of specific parameters (e.g. speciation and extinction) with some events disrupting the simulations (e.g. mass extinctions) @CITE-TreeSim.
One can even improve on generating these patterns using `FossilSim` to generate a pattern that will take into account fossilization processes @CITE-FossilSim.
On the other hand, if you need to simulate both diversity and traits through time, this can be done with specific parameters in `RPANDA` @CITE-RPANDA or in `PETER` @CITE-PETER where the traits are generated stochasticaly through time (given some process) during the birth-death process.

<!-- % The problems with the absence of modularity -->
Although the packages mentioned above (and more that we forgot to mention/don't know off) and are excellent (i.e. fast, reliable and documented) at the specific tasks they are designed for, they unfortunately don't allow much modifications beyond the parameters implemented for each specific function.
This can be problematic for generating specific simulations to be used as a specific null model for some specific questions.
For example, `TreeSim` can simulate a birth-death tree with some extinction event but is not designed to simulated one with an extinction event that leads to the birth-death process to be not diversity dependent anymore (e.g. simulating a release in selection pressure after the extinction process that leads to a different process dominating speciation).
Or `PETER` is not designed to simulate a complex set of traits (say three correlated BM traits and two independent OU ones).
Although we don't think this was the aim of these specific packages authors (not to create a strawman argument) we think this hampers the use of specific null model (or baseline model) to test specific dynamic mechanisms or to compare to specific datasets. 
<!-- % The solution with dads -->
Therefore, we propose `dads` a modular `R` package to simulate both diversity and disparity through time.

# Description
<!-- % Very briefly how does it work -->
`dads` is based on a core birth-death process implemented in the function `dads` that simulates diversity through time using the following steps:

1. **Branch length**: Generating some waiting time corresponding to the growth in length of the tree (e.g. the duration of a lineage through time);
2. **Selection**: Choosing a lineage _alive_ at the time of the simulation;
3. **Speciation**: Choosing whether this species will go extinct or speciate leading to two new lineages.

These three steps are repeated until the tree reaches the desired age or the desired number of species.
If traits are simulated during the process, a fourth step is added:

4. Trait: Generate some trait(s) value(s) based for the extinct species or the node leading to the two new lineages.
This/these value(s) are generated taking into account the trait value of the extinct species/new node's ancestor and the time elapsed (generated in step 1).

In `dads`, these three or four steps are implemented as modular functions that the user can easily change using an internal class of objects called `"modifiers"` or `"traits"` (Fig.  \ref{Fig_workflow}).
The simulation then outputs a tree (of class `"phylo"` and a associated table of traits - `"matrix"`) that can be visualised using the `plot.dads` function.

![Figure 1:](gitbook/dads_structure.pdf)

\label{Fig_workflow}`dads` package workflow @@@. `"events"` are an additional category of modular objects discussed later.

## Simulating diversity

To simulate a birth-death tree, two arguments are essentially needed: the birth-death parameters (`bd.params`) the rate of speciation (birth or $\lambda$), the rate of extinction (death or $\mu$ - this parameter can be equal to 0) and `stop.rule`, a rule for when to stop the growth of the tree (e.g. when reaching 20 species).

```{r, eval = FALSE, echo = TRUE}
## Running a simulation with a speciation rate of 1, an extinction rate of 0.1
## and stopping when it reaches 20 taxa
my_simulation <- dads(bd.params = list(speciation = 1, extinction = 0.1),
                      stop.rule = list(max.taxa = 20))
```

By default, the three first steps are included in a default modifier that reproduces a simple birth-death process given the input birth and death parameters.

1. **Branch length**: drawing a random number from an exponential distribution with a species dependent rate of $n \times (\lambda + \mu)$, where $n$ is the number of species currently alive in the process.
2. **Selection**: randomly choosing on of the $n$ taxa currently alive in the process;
3. **Speciation**: drawing a random number between 0 and 1; if that number is smaller than the speciation relative to the turnover ($\lambda / \lambda+\mu$), the lineage goes extinct <!-- %TODO: check!!! -->; else, the lineage speciates.

These are included by default in a `dads` process.
The modularity of the packages works with the ability to code the modifiers manually rather than using default ones.
Using a reserved syntax in this case the arguments `lineage` and `bd.params` (much more details about the syntax is available in the [manual](dads_manual)). <!-- TODO: make link -->
For example, to simulate the same process but hard coding the modifiers we can use the following functions for each process:

```{r, eval = FALSE, echo = TRUE}
## Step 1: Drawing from a diversity dependent exponential distribution
step1.branch.length <- function(lineage, bd.params) {
  rexp(1, rate = lineage$n * (bd.params$speciation + bd.params$extinction))
}

## Step 2: Selecting one of the available lineages
step2.selection <- function(lineage) {
  sample(1:lineage$n, 1)
}

## Step 3: Randomly speciating depending on the birth-death rates
step3.speciation <- function(bd.params) {
  runif(1) < (bd.params$speciation/(bd.params$speciation + bd.params$extinction))
}

## Specifying our modifiers (here they are doing the same as the default ones)
my_modifiers <- make.modifiers(branch.length = step1.branch.length,
                               selection     = step2.selection,
                               speciation    = step3.speciation)

## Running a simulation with a speciation rate of 1, an extinction rate of 0.1
## and stopping when it reaches 20 taxa with the default birth-death modifiers specified
my_simulation <- dads(bd.params = list(speciation = 1, extinction = 0.1),
                      stop.rule = list(max.taxa = 20),
                      modifiers = my_modifiers)
```

Of course these function can be greatly modified and depend on other parameters of the the simulation.
For example if traits are also simulated (see below) we can bias the branch length to be also dependent on the trait distribution, the selection to be biased towards species at the extremes of the distribution and the speciation modifier to only allow speciation when the lineage's parent trait value is positive.
More example of biased or special `"modifiers"` are available in the [manual](dads_manual).

## Simulating traits (disparity)

Another major feature of `dads` is to allow to specify the trait simulation process manually as a `"traits"` object.
This can be done in the same way as for the `"modifiers"` but with the `make.traits` function.
Here the Brownian motion syntax is also specific (especially `x0` and `edge.length`), more details are available in the [manual](dads_manual).

```{r, eval = FALSE, echo = TRUE}
## A simple Brownian motion process
my.BM.process <- function(x0, edge.length = 1, sd = 1, ...) {
  rnorm(n = 1, mean = x0, sd = sd * edge.length)
}

## Specifying our trait object
my_trait <- make.traits(process = my.BM.process)

## Running a simulation with a speciation rate of 1, an extinction rate of 0.1
## and stopping when it reaches 20 taxa with the default birth-death modifiers specified
## and a Brownian Motion trait process simulated
my_simulation <- dads(bd.params = list(speciation = 1, extinction = 0.1),
                      stop.rule = list(max.taxa = 20),
                      modifiers = my_modifiers,
                      trait     = my_trait)
```

It is of course possible to simulate multiple traits at the same time using multiple processes, whether the traits are correlated to each other, independent or even dependent on other elements of the simulations (e.g. time spent, number of currently living species, etc...).

## Plotting results

```{r, eval = TRUE, echo = FALSE}
## Simulating the tree above with a nice looking random seed
## (all the options described above, modifiers and traits are default)
## so it's pretty easy (we don't need to specify the trait process and the modifiers)
set.seed(2)
my_simulation <- dads(bd.params = list(speciation = 1, extinction = 0.1),
                      stop.rule = list(max.taxa = 20),
                      trait     = make.traits())
# plot(my_simulation)
```

The resulting object is a `"dads"` object containing a `$tree` element (a `"phylo"`) object and a `$data` object containing a `"matrix"` object:

```{r, eval = TRUE, echo = TRUE, fig.height = 4, fig.width = 4}
## Printing the dads object
my_simulation

## The generated data
head(my_simulation$data)

## The generated tree
my_simulation$tree
plot(my_simulation$tree)
```

The package also comes with it's own plotting function allowing to directly display `dads` generated tree and data in the same plot:

```{r, eval = TRUE, echo = TRUE, fig.height = 4, fig.width = 4}
## A default dads plot
plot(my_simulation, main = "Default plot")
```

Note that here too many options are available for plotting such objects. More information is again available in the [manual](dads_manual).

## Using modular and dynamic modifiers

<!-- Dynamic ones will be for bd.params = list(speciation = runif) -->
<!-- timed or conditional events in the simulations -->

# Discussion

## Modularity

## `dads` compared to other packages

## Further directions

# Conclusion
The `dads` is modular and nice to use.

# Package location
The `dads` package is available on the CRAN at \url{https://cran.r-project.org/web/packages/dads/index.html} or on GitHub at \url{https://github.com/TGuillerme/dads} with more associated information.
All the versions of the package are archived on ZENODO with associated DOI \url{https://zenodo.org/@@@}.

# Acknowledgments
Thanks to Mark Puttick and Alex Slavenko for comments on the early stage of the development of this package. Thanks to @@@. This work was funded by UKRI-NERC Grant NE/T000139/1 and a Royal Society University Research Fellowship (URF R 180006 to GHT).

# Authors contributions

TG designed the software.

# Data Availability, repeatability and reproducibility

The raw empirical data is available from the original papers [@beck2014; @jones2015, @marcy2016; @hopkins2016; @wright2017; @healy2019].
The subsets of the empirical data used in this analysis are available on figshare 
[DOI: 10.6084/m9.figshare.9943181.v1](https://doi.org/10.6084/m9.figshare.9943181.v1).
The modified empirical data are available in the package accompanying this manuscript (`data(moms::demo_data)`).
This manuscript (including the figures, tables and supplementary material) is repeatable and reproducible by compiling the vignette of the [GitHub `moms R` package](https://github/TGuillerme/moms).

# Conflict of interest
None declared.

# References

### 2023-01-19 inst/dads_paper.Rmd
Update dads_paper.Rmd

@@
 
-# Data Availability, repeatability and reproducibility
+# Repeatability and reproducibility
 
-The raw empirical data is available from the original papers [@beck2014; @jones2015, @marcy2016; @hopkins2016; @wright2017; @healy2019].
-The subsets of the empirical data used in this analysis are available on figshare 
-[DOI: 10.6084/m9.figshare.9943181.v1](https://doi.org/10.6084/m9.figshare.9943181.v1).
-The modified empirical data are available in the package accompanying this manuscript (`data(moms::demo_data)`).
-This manuscript (including the figures, tables and supplementary material) is repeatable and reproducible by compiling the vignette of the [GitHub `moms R` package](https://github/TGuillerme/moms).
+No data was used for this paper.
+However the entire paper (plots and results) is reproducible by compiling the paper in `Rmarkdown` from [github](link_to_the_paper).
+<!-- TODO: link to the Rmd file -->
 

### 2023-01-19 inst/gitbook/05_plotting_dads_objects.Rmd
Update 05_plotting_dads_objects.Rmd
draft

@@
 `"dads"` objects can be directly plotted in dads using many different options.
+<!-- 
+test_that("plot.dads works for traits", {
+
+    list_of_traits <- make.traits(process = c(no.process, BM.process, OU.process), trait.names = c("No process (normal)", "Brownian motion", "Ornstein-Uhlenbeck"))
+    ## Right output
+    expect_null(plot.dads(list_of_traits))
+    expect_null(plot.dads(list_of_traits, trait = 2))
+    expect_null(plot.dads(list_of_traits, trait = 3))
+})
+
+test_that("plot.dads works for dads", {
+
+    ## Simple example    
+    test <- dads(stop.rule = list("max.living" = 10), traits = make.traits())
+    ## Default plot
+    expect_null(plot(test))
+    ## Other options
+    expect_null(plot(test, main = "what?", xlim = c(0, test$tree$root.time), col = c(tips = "pink", nodes = "red", edges = "purple"), ylab = "data"))
+
+    ## A more complex trait
+    complex_traits <- make.traits(process = c(BM.process, BM.process), n = c(2,3), process.args = list(list(Sigma = diag(2)), list(Sigma = matrix(1/3, 3, 3))), trait.names = c("bib", "bob"))
+    test <- dads(stop.rule = list("max.living" = 10), traits = complex_traits)
+    expect_null(plot(test, trait = 3))
+
+    ## A big tree
+    set.seed(12)
+    big_test <- dads(bd.params = list(speciation = 1, extinction = 1/3), stop.rule = list(max.taxa = 1000), traits = make.traits())
+    expect_null(plot(big_test, cex = 0.5))
+
+    list_of_traits <- make.traits(process = c(no.process, BM.process, OU.process), trait.names = c("No process (normal)", "Brownian motion", "Ornstein-Uhlenbeck"))
+    ## Right output
+    expect_null(plot.dads(list_of_traits))
+    expect_null(plot.dads(list_of_traits, trait = 2))
+    expect_null(plot.dads(list_of_traits, trait = 3))
+})
+
+test_that("example works", {
+    ## Specifying a trait process
+    my_trait <- make.traits()
+    ## Plotting a trait process
+    expect_null(plot(my_trait, main = "A Brownian Motion"))
+
+    ## Simulating a tree with ten taxa
+    my_tree <- dads(stop.rule = list(max.taxa = 10))
+    ## Plotting a simple birth death tree (using ape::plot.phylo)
+    expect_is(plot(my_tree, main = "A pure birth tree"), "list")
+
+    ## Simulating a tree with traits
+    my_data <- dads(stop.rule = list(max.taxa = 10),
+                    traits    = my_trait)
+    ## Plotting the tree and traits
+    expect_null(plot(my_data))
+
+    ## Specifying a 3D trait process
+    my_3D_trait <- make.traits(n = 3)
+    ## Simulating a birth death tree with that trait
+    my_data <- dads(bd.params = list(extinction = 0.2),
+                    stop.rule = list(max.living = 50),
+                    traits    = my_3D_trait)
+
+    ## Plotting the second trait and the tree (default)
+    ## The colours are purple for nodes and blue for tips
+    expect_null(plot(my_data, trait = 2, col = c("purple", "blue"),
+         edges = "pink", tips.nodes = "black"))
+
+    ## Plotting the first and third trait correlation
+    ## The colours are a heat map based on the elements age
+    expect_null(plot(my_data, trait = c(1, 3), col = terrain.colors,
+         edges = "grey", tips.nodes = "black"))
+
+    ## Plotting the first and third trait correlation in 3D
+    expect_null(plot(my_data, trait = c(1,3), col = rainbow,
+         edges = "grey", tips.nodes = "black", use.3D = TRUE))
+
+    ## Plotting all traits in 3D (without branch lengths)
+    expect_null(plot(my_data, trait = c(1:3), col = heat.colors,
+         edges = "grey", tips.nodes = "black", use.3D = TRUE, type = "s", radius = 0.1))
+}) -->

### 2023-02-17 inst/gitbook/05_other_functionalities.Rmd
Create 05_other_functionalities.Rmd

---
title: "`make.events`"
author: "Thomas Guillerme (guillert@tcd.ie)"
date: "`r Sys.Date()`"
bibliography: [../References.bib, ../packages.bib]
output:
  html_document: default
  pdf_document: default
---

```{r, echo = FALSE}
library(rgl)
knitr::knit_hooks$set(webgl = hook_webgl)
```

# Other functionalities {#others}

## `make.bd.params` {#makebdparams}

In most examples above, birth-death parameters where set as fixed values, i.e. a speciation rate of $\lambda$ and an extinction rate of $\mu$ throughout the simulations (with maybe some events modifying these rates).
However, it is of course possible to set these rates as specific or changing distributions! 
You can do this using the `make.bd.params` function a provide either a `vector` of values:

```{r}
## An example where the speciation is randomly sampled among three values
my_bd_params <- make.bd.params(speciation = c(1/3,42))

## Building a tree using this set of parameters
set.seed(123)
plot(dads(stop.rule = list(max.taxa = 50), bd.params = my_bd_params))
## Note the regions in the tree with short branches
## (that's the speciation being 1/3 while the others are speciation = 42) 
```

Or directly a function from which to sample:

```{r}
## Another example where speciation is drawn from the interval (0, 1)
make.bd.params(speciation = runif)
```

In this example, the `"bd.params"` object passed to the `dads` function will allow the birth-death process to sample the speciation parameter each time it is called (e.g. during the speciation/extinction step, the branch length step, etc.).
If using a function, you can fine tune the arguments to be passed to that function using the `speciation.args` or the `extinction.args` arguments (as a named list matching the function's arguments):

```{r}
## Speciation is drawn from the interval (0.5, 1.5)
make.bd.params(speciation = runif,
               speciation.args = list(min = 0.5, max = 1.5))
```

When using distributions for both the speciation and extinction parameters, you can run into the undesired problem of having an extinction rate that is higher than your speciation rate (and thus your tree dying out - unless this is the desired behaviour).
You can avoid this problem by using the `joint` distribution argument this will ensure that the sampling of the extinction rate is always lower or equal to the speciation rate.

```{r}
## Joint speciation and extinction sampled from uniform distribution 
## with speciation always >= to extinction
make.bd.params(speciation = runif, extinction = runif, joint = TRUE)
```

Finally to avoid negative sampling values, you can use the argument `absolute` that will make all the sampled values positive.
This argument is set to `TRUE` by default so you shouldn't have to worry about it most of the time unless you specifically need negative sampled values for your parameters.

```{r}
## Making the speciation sampling always positive
make.bd.params(speciation = rnorm, absolute = TRUE)
## This is effectively making the speciation distribution equal to
make.bd.params(speciation = function(x) abs(rnorm(x)))
```

The two other possible arguments for this function are `test` and `update` that work the same as for [`make.traits`](#maketraits) or [`make.modifiers`](#makemodifiers)

## `drop.things`

You can use the function `drop.things` to drop specific elements of the tree and data at the same time by providing the argument `what` to be `"fossils"` for tips that went extinct, `"livings"` tips that where alive at the end of the simulation or `"singles"` for internal nodes.
Alternatively you can use the function aliases `drop.fossils`, `drop.livings` or `drop.singles` for the exact same results:

```{r}
## A random tree with fossils and traits and internal nodes every 0.5 times
set.seed(3)
my_data <- dads(stop.rule = list(max.taxa = 20),
                bd.params = list(speciation = 1, extinction = 1/3),
                traits    = make.traits(), save.steps = 0.5)

## A tree with 20 tips and 54 nodes
my_data$tree
## And a dataset with 74 rows
dim(my_data$data)

## Removing the fossil species
drop.things(my_data, what = "fossils")$tree
dim(drop.fossils(my_data)$data)

## Removing the living species
drop.things(my_data, what = "livings")$tree
dim(drop.livings(my_data)$data)

## Removing the internal nodes
drop.things(my_data, what = "singles")$tree
dim(drop.singles(my_data)$data)

## Removing the internal nodes AND the fossils
drop.singles(drop.fossils(my_data))
```

### 2023-02-21 inst/gitbook/05_other_functionalities.Rmd
done for bd.params

@@
 
+If you are a visual person and your `bd.params` objects are getting a bit too complicated to remember, you can always quickly plot them (the function will sample from the `bd.params` object and show the results):
+
+```{r}
+par(mfrow = c(2,2))
+plot(make.bd.params(), main = "Default birth-death parameters")
+plot(make.bd.params(speciation = runif,
+               speciation.args = list(min = 0.5, max = 1.5)),
+     main = "Uniform speciation between 0.5 and 1.5\n(no extinction)")
+plot(make.bd.params(speciation = runif,
+                    speciation.args = list(min = 0.5, max = 1.5)),
+     main = "Uniform speciation between 0.5 and 1.5\n(no extinction)")
+plot(make.bd.params(speciation = runif, extinction = runif, joint = TRUE),
+     main = "Joint uniform speciation and extinction")
+plot(make.bd.params(speciation = rnorm, extinction = runif,
+                    joint = FALSE, abs = FALSE),
+     main = "Disjoint normal speciation and uniform extinction")
+```
+
 ## `drop.things`

### 2023-02-21 inst/gitbook/02_make.traits.Rmd
traits done.

@@
 
+## Traits implemented in `dads`
+
+If you don't want to design your own trait process, you can use one of the following trait processes that are currently implemented in `dads`.
+You can find more information about their many options using their specific manuals in R or the generic `?trait.process`:
+
+ * `BM.process`: this is the well famous [Brownian motion process](https://en.wikipedia.org/wiki/Brownian_motion).
+
+```{r, echo = FALSE}
+plot(make.traits(process = BM.process), main = "The Brownian motion")
+```
+
+ * `OU.process`: this is the equally famous [Ornstein–Uhlenbeck process](https://en.wikipedia.org/wiki/Ornstein%E2%80%93Uhlenbeck_process).
+
+```{r, echo = FALSE}
+plot(make.traits(process = OU.process), main = "The Ornstein–Uhlenbeck process")
+```
+
+ * `no.process`: this process has...
+no process.
+In other words, this is a non time dependent process (the simulated value does not depends on the ancestors' value nor the branch length).
+It's basically a place holder for a random sampling function like `rnorm` (default), `runif`, `rlnorm`, etc...
+
+```{r, echo = FALSE}
+plot(make.traits(process = no.process, process.args = list(fun = runif)), main = "No process using runif\n(yes, it \"looks\" like an OU!)")
+```
+
+ * `multi.peak.process`: this process is a modified version of the `OU.process` that can take multiple local optima.
+The default OU process has one optimum towards which the values are drawn with the alpha parameter (the elastic band) which is usually 0.
+However, with this `multi.peak.process` we can set multiple values towards which values can be attracted with the same alpha parameter.
+
+```{r, echo = FALSE}
+plot(make.traits(multi.peak.process, process.args = list(peaks = c(1, 5, 10))), main = "A multi-peak (OU) process\nwith peaks (optima) at 1, 5 and 10")
+```
+
+ * `repulsion.process`: this is a modified version of the `BM.process` where instead of accumulating normally through time, new trait values are more likely to be different than their ancestor following a `repulsion` parameter.
+
+```{r, echo = FALSE}
+plot(make.traits(repulsion.process, process.args = list(repulsion = 5)), main = "A repulsion process\nwith a repulsion parameter of 5")
+```
+
 ## Testing the traits with `test`

### 2023-02-21 R/trait.process.R

@@ -79,6 +79,11 @@
 #' ## No process
 #' no.process()
 #' plot(make.traits(process = no.process))
+#'
+#' ## Multi peaks
+#'
+#' ## Repulsion process
+#'
 #' 
 #' @seealso
 #' 
@@ -90,6 +95,8 @@ trait.process <- function(x0 = 0, edge.length = 1, ...) {
     cat("?BM.process\n")
     cat("?OU.process\n")
     cat("?no.process\n")
+    cat("?multi.peak.process\n")
+    cat("?repulsion.process\n")
 }
 
 ## The Brownian motion

### 2023-02-21 tests/testthat/test-make.traits.R

@@ -233,3 +233,36 @@ test_that("bkg.traits works", {
 })
 
 
+test_that("trait processes work", {
+
+    ## BM.process
+    set.seed(1)
+    expect_equal(round(BM.process(), 5), matrix(round(-0.6264538, 5)))
+    expect_is(plot(BM.process), "list")
+    expect_null(plot(make.traits(BM.process)))
+
+    ## OU.process
+    set.seed(1)
+    expect_equal(round(OU.process(), 5), matrix(round(-0.4119058, 5)))
+    expect_is(plot(OU.process), "list")
+    expect_null(plot(make.traits(OU.process)))
+
+    ## no.process
+    set.seed(1)
+    expect_equal(round(no.process(), 5), c(round(-0.6264538, 5)))
+    expect_null(plot(make.traits(no.process)))
+    expect_null(plot(make.traits(no.process, process.args = list(fun = runif))))
+
+    ## multi.peak.process
+    set.seed(1)
+    expect_equal(round(multi.peak.process(), 5), matrix(round(-0.4119058, 5)))
+    expect_is(plot(multi.peak.process), "list")
+    expect_null(plot(make.traits(multi.peak.process, process.args = list(peaks = c(1, 5, 10)))))
+    expect_null(plot(make.traits(multi.peak.process, n = 3, start = c(1,2,3),process.args = list(peaks = list(1, 5, 10)))))
+
+    ## repulsion.process
+    set.seed(1)
+    expect_equal(round(repulsion.process(), 5), c(round(-0.6264538, 5)))
+    expect_null(plot(make.traits(repulsion.process)))
+    expect_null(plot(make.traits(repulsion.process, process.args = list(repulsion = 5))))
+})

### 2023-02-21 inst/gitbook/03_make.modifiers.Rmd
added modifiers and make.modifiers examples

@@
 In other words, throughout the process, the modifications remain constant (even if they are conditional).
-It is however possible to code the `"modifiers"` so that they can be affected by `"events"` objects (see next chapter).
+It is however possible to code the `"modifiers"` so that they can be affected by `"events"` objects (see next chapter [on `events`](#makeevents)).
 
@@
 
+```{r}
+    always.false <- function() return(FALSE)
+    always.one <- function(x = NULL) return(1)
+
+    test <- make.modifiers(branch.length = branch.length.trait,
+                           speciation    = speciation.trait,
+                           condition     = always.false, 
+                           modify        = always.one)
+```
+
 ## Combining and editing modifiers (`add`)
@@
 ## Testing modifiers (`test`)
-
-## Demo runnable
-
-```{r, fig.height = 16, fig.width = 8, eval = FALSE}
-
-```

### 2023-02-21 R/make.modifiers.R

@@ -50,8 +50,21 @@
 #' More details about the \code{modifiers} functions is explained in the \code{dads} manual: \url{http://tguillerme.github.io/dads}.
 #' 
 #' @examples
+#' ## These functions should be fed to the make.modifiers function to create
+#' ## modifiers for dads objects. For example, the following sets specifies that
+#' ## the branch length should be generated using the branch.length.trait function
+#' ## the selection using the selection function and the speciation using the
+#' ## speciation.trait function:
+#' my_modifiers <- make.modifiers(branch.length = branch.length.trait,
+#'                                selection     = selection,
+#'                                speciation    = speciation.trait)
 #'
-#' @seealso
+#' ## Creating a dads simulation using these modifiers
+#' dads(stop.rule = list(max.taxa = 20),
+#'      traits = make.traits(),
+#'      modifiers = my_modifiers)
+#'
+#' @seealso modifiers dads
 #' 
 #' @author Thomas Guillerme
 #' @export

### 2023-02-21 R/modifiers.R

@@ -7,6 +7,9 @@
 #' @usage branch.length(bd.params = NULL, lineage = NULL, trait.values = NULL, modify.fun = NULL)
 #' @usage selection(bd.params = NULL, lineage = NULL, trait.values = NULL, modify.fun = NULL)
 #' @usage speciation(bd.params = NULL, lineage = NULL, trait.values = NULL, modify.fun = NULL)
+#' @usage branch.length.trait(bd.params = NULL, lineage = NULL, trait.values = NULL, modify.fun = NULL)
+#' @usage speciation.trait(bd.params = NULL, lineage = NULL, trait.values = NULL, modify.fun = NULL)
+
 #'
 #' @param bd.params      A named list of birth death parameters (see details).
 #' @param lineage        A named list containing the lineage data (see details).
@@ -37,6 +40,25 @@
 #'  }
 #' 
 #' More details about the \code{modifiers} functions is explained in the \code{dads} manual: \url{http://tguillerme.github.io/dads}.
+#'
+#' @examples
+#' ## These functions should be fed to the make.modifiers function to create
+#' ## modifiers for dads objects. For example, the following sets specifies that
+#' ## the branch length should be generated using the branch.length.trait function
+#' ## the selection using the selection function and the speciation using the
+#' ## speciation.trait function:
+#' my_modifiers <- make.modifiers(branch.length = branch.length.trait,
+#'                                selection     = selection,
+#'                                speciation    = speciation.trait)
+#'
+#' ## Creating a dads simulation using these modifiers
+#' dads(stop.rule = list(max.taxa = 20),
+#'      traits = make.traits(),
+#'      modifiers = my_modifiers)
+#'
+#' @seealso make.modifiers dads
+#' @author Thomas Guillerme
+
 
 modifiers <- function(bd.params = NULL, lineage = NULL, trait.values = NULL, modify.fun = NULL) {
     cat("modifiers functions implemented in dads:\n")

### 2023-02-21 R/trait.process.R

@@ -80,10 +80,13 @@
 #' no.process()
 #' plot(make.traits(process = no.process))
 #'
-#' ## Multi peaks
-#'
+#' ## Multi peaks with peaks at the values 1, 5 and 10
+#' multi.peak.process(peaks = c(1, 5, 10))
+#' plot(make.traits(multi.peak.process, process.args = list(peaks = c(1, 5, 10)))
+#' 
 #' ## Repulsion process
-#'
+#' repulsion.process(x0 = 0, repulsion = 1)
+#' plot(make.traits(repulsion.process, process.args = list(repulsion = 5)))
 #' 
 #' @seealso
 #' 

### 2023-02-21 tests/testthat/test-make.modifiers.R

@@ -174,10 +174,9 @@ test_that("make.modifiers works", {
     expect_equal(test2[[3]]$internal$modify(42), 42)
 
     ## Print modifiers works correctly
-
 })
 
-
+## Updating
 test_that("make.modifiers(update) works", {
 
     ## Sanitizing
@@ -207,5 +206,38 @@ test_that("make.modifiers(update) works", {
     expect_true(capture_output(print(test1$waiting$internal)) != capture_output(print(test2$waiting$internal)))
     ## But the same functions
     expect_true(capture_output(print(test1$waiting$fun)) == capture_output(print(test2$waiting$fun)))
-
 })
+
+## Modifiers
+# test_that("modifiers works", {
+
+#     ## branch.length
+#     bd.params <- list(speciation = 1, extinction = 0.1)
+#     lineage <- list(n = 10)
+#     branch.length(bd.params = list(speciation = 1, extinction = 0.1))
+
+
+
+
+
+#     ## branch.length.trait
+#     ## selection
+#     ## speciation
+#     ## speciation.trait
+
+#     test2 <- NULL
+#     expect_message(test2 <- make.modifiers(branch.length = branch.length, modify = always.one, add = test))
+#     expect_is(test2, c("dads", "modifiers"))
+#     expect_equal(names(test2), c("waiting", "selecting", "speciating", "call"))
+#     expect_equal(names(test2[[1]]), c("fun", "internal"))
+#     expect_equal(names(test2[[2]]), c("fun", "internal"))
+#     expect_equal(names(test2[[3]]), c("fun", "internal"))
+#     expect_true(test2[[1]]$internal$condition())
+#     expect_equal(test2[[1]]$internal$modify(42), 1)
+#     expect_null(test[[2]]$internal$condition)
+#     expect_null(test[[2]]$internal$modify)
+#     expect_false(test2[[3]]$internal$condition())
+#     expect_equal(test2[[3]]$internal$modify(42), 42)
+
+#     ## Print modifiers works correctly
+# })

### 2023-02-22 inst/gitbook/02_make.traits.Rmd
added background and save.steps

@@
 
-# Making trait processes with `make.traits()` {#maketraits}
+# Simulating traits {#maketraits}
 
@@
 ```
+
+In `dads`, traits are simulating by providing a `traits` object to the `traits` argument.
+This object is generated using the `make.traits`.
 
@@
 
+## Using a background trait
+
+`traits` objects also allow a background trait to be used when traits are simulated ([step 3 here](#bdalgorithm)).
+This basically allows traits to be simulating for _all_ tips whenever a trait is generated for one tip.
+This can be useful for keeping track of trait values along the simulation (_cf_ just at bifurcating nodes).
+The `background` argument takes any output from the `make.traits` function in a nested way:
+
+```{r}
+## Generating a default BM trait:
+BM_trait <- make.traits()
+## Generating an OU trait with a background BM trait
+my_trait <- make.traits(process = OU.process, background = BM_trait) 
+```
+
+Note that technically you can nest as many background traits as you want (e.g. `make.traits(background = make.traits(background = make.traits(...)))` is valid).
+However, you should always make sure that the background trait has the same dimensions as the normal (main) trait.
+When using a trait with background, your tree will have internal singleton nodes (i.e. nodes linking to one ancestor and only one descendant).
+You can remove these nodes using the [`drop.things` function](#dropthings).
+
+```{r}
+set.seed(1)
+## Generating a pure birth tree with the background trait
+tree_bkg <- dads(stop.rule = list(max.taxa = 20),
+                 traits    = my_trait)
+## This tree has many internal singleton nodes
+plot(tree_bkg)
+```
+
+## Saving trait values at different time steps
+
+Alternatively to having a background trait, you can also simulate a tree by generating traits at specific time steps with the `save.steps` option in `dads`.
+This will apply the `traits` object to all lineages currently alive at the required time steps.
+These time steps can be either regular by providing a single numeric value; e.g. `save.steps = 0.1` will get a snapshot of the trait values every 0.1 units of time.
+Or specific, by providing a specific set of values; e.g. `save.steps = c(1, 1.2, 3)` will make a snapshot of the trait values at the required time steps.
+
+```{r}
+set.seed(123)
+## Generating a birth-death tree with a BM trait and saving steps at specific times
+tree_steps <- dads(stop.rule  = list(max.time = 3),
+                   bd.params  = list(speciation = 1, extinction = 0.1),
+                   traits     = make.traits(),
+                   save.steps = c(1/3, 1, 2))
+## This also creates internal singleton nodes
+plot(tree_steps)
+abline(v = 3 - c(1/3, 1, 2), lwd = 0.5, col = "grey")
+```
+
 ## Traits implemented in `dads`

### 2023-02-22 inst/gitbook/index.Rmd
added algorithm description

@@
 ![Schematised summary of the `dads` package architecture](dads_structure.pdf)
+
+# The birth-death algorithm {#bdalgorithm}
+
+> If you want to get your hands dirty, you can go straight to the [getting start section](#gettingstarted), this following section just describes the process in pseudo-code.
+
+The `dads` algorithm is based on a modular version of the birth-death process algorithm.
+The birth-death model is a continuous Markov process (i.e.
+a continuous random process where the state will change each step are linked through time and change according to some random exponential variables) which is well defined mathematically and [commonly used in biology](https://lukejharmon.github.io/pcm/chapter10_birthdeath/#section-10.2-the-birth-death-model) but also in [many other fields](https://en.wikipedia.org/wiki/Birth%E2%80%93death_process).
+
+The algorithm used in `dads` allows modularity of this process and is based on the following steps (the text in `courier font` is for the name of the process in the algorithm):
+
+ 0.
+**Starting the process**: this step is actually non modular and creates a random tree with one tip, one node and one branch length connecting both.
+This step is used to optimise the rest of the algorithm in terms of speed and memory management.
+The node, tip and branch resulting from this step are discarded at the end of the simulations.
+ 1.
+**Selecting a lineage** (`selecting`): this step selects a tip that is currently not extinct.
+In a standard birth-death process this is just done randomly, however in `dads` this can be modified based on the birth-death parameters, the currently available lineages and potential trait values.
+*For example, it is possible to put a higher probability in selecting a lineage that is next to a lineage that recently went extinct and has a positive trait value.*
+ Then go to step 2.
+ 2.
+**Growing the tree** (`waiting`): this step grows the tree by a certain amount.
+It does so by adding the same amount of branch length to all the non-extinct lineages.
+In an exact birth death process, this is done by drawing a random value from an exponential distribution with a rate of $$\text{number of living lineages} \times (\text{speciation} + \text{extinction parameters})$$.
+In `dads`, this can be modified based on the birth-death parameters, the currently available lineages and potential trait values.
+*For example, it is possible to increase branch length by the ratio of living/extinct fossils and a random number drawn from the range of current trait values.*
+ If the total tree length is less than the required tree length, go to step 3. Else go to step 6.
+ 3.
+**Simulating traits** \[optional\] (`traits`): this step allows to simulate a trait value for the selected lineage from step 1.
+This is typically not part of a standard birth-death process and is handled via the `traits` option in `dads` (see the [`make.traits`](#maketraits) chapter for more details).
+ Then go to step 4.
+ 4.
+**Speciating** (`speciating`): in this step, the selected lineage from step 1 has the option of speciating or going extinct.
+In a standard birth-death process, this happens by randomly drawing a value between 0 and 1 as a way to trigger speciation relatively to the birth-death parameters:
+```
+if
+    randomly drawn number is smaller or equal to speciation/(speciation + extinction)
+then
+    do speciate
+else
+    go extinct
+```
+ Again, in `dads`, this process can be modified based on the birth-death parameters, the currently available lineages and potential trait values.
+*For example, the lineage can only speciate if it's trait value is positive, regardless of whether speciation or extinction has been triggered*.
+ Then go to step 5.
+ 5. If the number of lineages is less than required number of lineages, then go to step 1. Else go to step 6.
+ 6. The simulation has stopped because it reached the required amount of lineages or time.

### 2023-02-22 inst/gitbook/03_make.modifiers.Rmd
added the modifiers bits

@@
 
-# Making modifiers with `make.modifiers()` {#makemodifiers}
+# Modifying the birth-death process {#makemodifiers}
 
@@
 
+For example in the `speciation` part of a modifier, the default is to trigger an event from a uniform distribution and then check if that value is smaller than ($$speciation/(speciation + extinction)$$):
+
 ```{r}
-    always.false <- function() return(FALSE)
-    always.one <- function(x = NULL) return(1)
+## The default speciation algorithm
+speciation <- function(bd.params) {
+    ## Randomly trigger an event
+    trigger_event <- runif(1)
 
-    test <- make.modifiers(branch.length = branch.length.trait,
-                           speciation    = speciation.trait,
-                           condition     = always.false, 
-                           modify        = always.one)
+    ## Speciate?
+    return(trigger_event < (bd.params$speciation/(bd.params$speciation + bd.params$extinction)))
+}
 ```
 
-## Combining and editing modifiers (`add`)
+It is possible with some specific condition to modify this trigger by providing a `condition` and `modify` function to the speciation function.
+That is, if the `condition` is met, apply the `modify` function to the algorithm.
+For example here we can edit the default speciation function to have a modification (`modify = double.the.trigger`: doubling the trigger value) happening half the time (`condition = half.the.time`):
 
-## Testing modifiers (`test`)
+```{r}
+## A conditional function that triggers half the time
+half.the.time <- function() return(sample(c(TRUE, FALSE), 1))
+## A modification that doubles the value to trigger the event
+double.the.trigger <- function(x) return(x*2)
+## A conditional modifier
+make.modifiers(speciation = speciation,
+               condition  = half.the.time, 
+               modify     = double.the.trigger)
+```
+Effectively, this will internally modify the speciation function as follows
+
+```{r, eval = FALSE}
+## The default speciation algorithm
+speciation <- function(bd.params) {
+    ## Randomly trigger an event
+    trigger_event <- runif(1)
+
+    ## Modify the triggering
+    if(half.the.time()) { ## This running the half.the.time function drawing TRUE or FALSE randomly
+        trigger_event <- double.the.trigger(trigger_event) ## This will double the random value trigger_event
+    }
+
+    ## Speciate?
+    return(trigger_event < (bd.params$speciation/(bd.params$speciation + bd.params$extinction)))
+}
+```
+
+These `condition` and `modify` functions can be applied to all the `modifiers` elements (`selection`, `branch.length` and `speciation`).
+However, they are typically used in `events` that will modify... the `modifiers` ([see the `events` section](#makeevents))!

### 2023-02-22 inst/gitbook/04_make.events.Rmd
Update 04_make.events.Rmd

@@
 
-# Making events with `make.events()` {#makeevents}
-
 ```{r, echo = FALSE}
@@
 
+# Adding events to simulations {#makeevents}
+
+One other major feature of the `dads` package is to allow simulations to run with specified events to occur during the simulation.
+These are typically events that can drastically change the course of the simulation.
+For example, you might want simulate some mass extinction at some specific point in time.
+However, these can also be more subtle, like the internal change of parameter values when reaching a specific trait value.
+These are all handled by `dads` with the `events` object that you can create using `make.events` and has the same overall logic as `make.traits` and `make.modifiers`.
+
+`events` require three main arguments:
+ * the `target` which designates what the extinction should affect (e.g. the taxa, the speciation rate, etc...);
+ * the `condition` which designates when to trigger the event;
+ * the `modification` which designates what to exactly modify when the event is triggered.
+
+There are several more arguments that can be passed to `make.events` but they are discussed later on.
+First let's focus on these three main arguments:
+
+## Target
+
+The target of the event is what the event is going to modify in the birth-death algorithm.
+You can only have one target per event (along with one condition and one modification) but you can create events that contain multiple events (i.e. multiple triplets of target/condition/modification).
+The targets that are currently available are:
+ * `"taxa"` to modify anything linked to the [`lineage` list](#allowarguments). E.g. making half of the living taxa go extinct.
+ * `"bd.params"` to modify anything linked to the [`bd.params` object](#makebdparams).
+For example you might want to change the distribution of one of the parameter after some `conditions`.
+This is typically done by updating the object using the argument `update` from `make.bd.params`.
+ * `"traits"` to modify anything linked to the [`traits` object](#maketraits).
+For example you might want to change the trait process after some `conditions`.
+This is typically done by updating the object using the argument `update` from `make.traits`.
+ * `"modifiers"` to modify anything linked to the [`modifiers` object](#makebdmodifiers).
+For example you might want to change the speciation rule after some `conditions`.
+This is typically done by updating the object using the argument `update` from `make.modifiers`.
+ * `"founding"` this target is a bit more special and allows you run a nested `dads` object in the simulation. It is cover in a [specific section below](#founding).
+
+We will see some example of these targets by illustrating the `conditions` and `modifications` below.
+
 ## Conditions
+
+`condition` is a function that returns a logical value. When a specific condition is met, it should return `TRUE` and trigger the event, else it should return `FALSE`.
+
+Currently there are three conditions functions implemented in `dads` but you can easily come up with your own version of them.
+
+All `condition` functions in `dads` take at least to arguments: `x` for the variable of interest (e.g.
+time, number of taxa, trait value, etc...) and `condition`, the relational operator to evaluate.
+A relational operator is the proper (fancy) term designating all the comparisons you're regularly using in `R` like `==` (is equal?) `<` (is smaller?) `>=` (is bigger or equal?).
+You can get the full list in the base manual (using `?Comparison` or any of the relational operator in a function form).
+
+> Note on the function form of a operator in `R`: if you weren't aware, functions in `R` are _usually_ described in the format `function(arguments)`, however many other functions you are commonly using are written as operators, for example (`1 + 1`).
+These are still functions though!
+And you can always use them in the classic function format by quoting them in back ticks: the following does the same as `(1+1)`: `` `+`(1, 1) ``.
+You can use that for parsing your relational operators as functions in `condition`.
+If you want you condition to be equal, use the syntax ``condition = `==` ``.
+More details on how functions _really_ work in the [Advanced `R` book](https://adv-r.hadley.nz/functions.html)!
+
+ * `time.condition` will trigger the `event` once a certain time is reached in the simulations and is the simplest/more basic condition with no arguments other than the time required.
+For example ``time.condition(4, condition = `==`)`` will trigger the `event` once the simulations reached 4 units of time.
+Easy.
+ * `taxa.condition` will trigger the `event` once a certain number of taxa is reached.
+This can be considered including or excluding fossil species.
+For example ``taxa.condition(42, condition = `>=`, living = TRUE)`` will trigger the `event` if there is at least 42 living species.
+ * `trait.condition` will trigger the `event` once a certain trait value is reached.
+This function allows to say which trait(s) to target (by default, the first one using `trait = 1`), what value of the trait to target (by default `what = max`) and whether to use and absolute trait value or not (`absolute = TRUE`).
+For example ``trait.condition(1/3, condition = `>`, trait = 1, what = sd)`` will trigger the condition after the standard deviation of the first trait reaches 1/3.
 
@@
 
-## Some events examples
+After defining the `event`'s `target` and `condition`, you'll need to also specify what it should modify.
+These are basically functions that should modify a specific aspect of the target
+Depending on the target you can modify the following:
 
-### Random mass extinction after some time
+ * if the target is `"taxa"` you can modify the internal [`lineage` list](#allowarguments) by removing living species using `random.extinction` or `trait.extinction`.
+ * if the target is `"taxa"` you can modify the lineage tracker by removing living species using `random.extinction` or `trait.extinction`.
+ * if the target is `"bd.params"` you can modify the [`bd.params` object](#makebdparams) using `update.bd.params`.
+ * if the target is `"traits"` you can modify the [`traits` object](#maketraits) using `update.traits`.
+ * if the target is `"modifiers"` you can modify the [`modifiers` object](#makemodifiers) using `update.modifiers`.
+ * (for the `"founding"` target [see below](#founding))
+
+The most straightforward example if for `modifications` on `bd.params`, `traits` or `modifiers` objects since they use the same syntax as for their generic `make.X` function.
+For example, for `make.traits`, you can update a `trait` using the `update` argument as follows:
 
 ```{r}
-## Simulation parameters
-stop.rule <- list(max.time = 5)
-bd.params <- list(extinction = 0, speciation = 1)
+## A BM trait in two dimensions
+(BM_2D <- make.traits(n = 2, process = BM.process))
+## Updating the 2D BM into a 2D OU
+(OU_2D <- make.traits(update = BM_2D, process = OU.process))
+```
 
+So basically the functions `update.X` apply the update to the object `X` when the event happens.
+The following `event` updates the `bd.params` object by setting the extinction parameter to 1/3 when reaching a 10 species:
+
+```{r, eval = FALSE}
+make.events(target       = "bd.params",
+            condition    = taxa.condition(10, condition = `>=`),
+            modification = update.bd.params(extinction = 1/3))
+
+```
+
+> Note that by default events are triggered only once in the whole simulation, so although the example above states that the condition is reaching at least 10 taxa, it will not trigger every times it reaches more than 10 taxa.
+You can change the number of times the events can be triggered using the argument `replications` (by default it's set to `replications = 0` for triggering it only once).
+
+## Examples
+
+Here are some example illustrating how to generate events.
+For a simple example, we can create a extinction event that will remove 80% of species after reaching time 4:
+
+```{r}
 ## 80% mass extinction at time 4
@@
 
+## Simulation parameters
+stop.rule <- list(max.time = 5)
+bd.params <- list(extinction = 0, speciation = 1)
+
 ## Running the simulations
-set.seed(1)
+set.seed(123)
 results <- dads(bd.params = bd.params,
@@
 
-### Extinction based on a trait value after some time
+Or for a slightly more complex example, we can can change the trait process from a BM to an OU when the trait values reaches an upper 95% quantile value above 2:
 
 ```{r}
-## Simulation parameters
-stop.rule <- list(max.time = 5)
-bd.params <- list(extinction = 0, speciation = 1)
+## The 95% upper quantile value of a distribution
+upper.95 <- function(x) {
+    return(quantile(x, prob = 0.95))
+} 
+## Create an event to change the trait process
+change_process <- make.events(
+                  target       = "traits",
+                  ## condition is triggered if(upper.95(x) > 3)
+                  condition    = trait.condition(3, condition = `>`, what = upper.95),
+                  modification = update.traits(process = OU.process))
 
-## Extinction of any tips with trait < 1 at time 4
-trait_extinction <- make.events(
-                      target = "taxa",
-                      condition = time.condition(4),
-                      modification = trait.extinction(1))
-
-## Running the simulations
-set.seed(7)
-results <- dads(bd.params = bd.params,
-                stop.rule = stop.rule,
-                traits    = make.traits(),
-                events    = trait_extinction)
-## Plotting the results
-plot(results)
-```
-
-### Adding a background extinction after reaching a number of living taxa
-
-```{r}
-## Simulation parameters
-stop.rule <- list(max.living = 50)
-bd.params <- list(extinction = 0, speciation = 1)
-
-## Adding an extinction parameter after 30 taxa
-background_extinction <- make.events(
-                      condition    = taxa.condition(30),
-                      target       = "bd.params",
-                      modification = update.bd.params(1/3, "extinction"))
-
-## Running the simulations
-set.seed(2)
-results <- dads(bd.params = bd.params,
-                stop.rule = stop.rule,
-                events    = background_extinction)
-## Plotting the results
-plot(results)
-```
-
-### Reducing speciation after some time
-
-```{r}
-## Simulation parameters
-stop.rule <- list(max.time = 4)
-bd.params <- list(extinction = 0, speciation = 1)
-
-## Reducing speciation after reaching time 2
-reduced_speciation <- make.events(
-                      condition    = time.condition(2),
-                      target       = "bd.params",
-                      modification = update.bd.params(1/3, "speciation"))
-
-set.seed(42)
-no_event <- dads(bd.params = bd.params,
-                 stop.rule = stop.rule)
-set.seed(42)
-reduced_speciation_event <- dads(bd.params = bd.params,
-                                 stop.rule = stop.rule,
-                                 events    = reduced_speciation)
-
-## Plot both trees
-par(mfrow = c(1, 2))
-plot(no_event, main = "No event")
-axisPhylo()
-plot(reduced_speciation_event, main = "Reduced speciation after time 2")
-axisPhylo()
-```
-
-### Changing trait process after some time
-
-```{r}
-## Simulation parameters
+## Set the simulation parameters
 bd.params <- list(extinction = 0, speciation = 1)
@@
 traits    <- make.traits()
-
-## Create an event to change the trait process
-change_process <- make.events(
-                  condition    = time.condition(5),
-                  target       = "traits",
-                  modification = update.traits(process = OU.process))
 
@@
 
-### Changing trait correlation after reaching a trait value
+## Founding events {#founding}
 
-```{r}
-## Set the parameters
-stop.rule <- list(max.taxa = 100)
-bd.params <- list(extinction = 0, speciation = 1)
+Founding events are a specific `target` for `events` that allows you to... simulate a birth-death process within the current one!
+It basically allows you to simulate a specific `dads` process i.e.
+using the `dads` function with it's own `traits`, `modifiers`, `bd.params` and `events` (and yes, that's including `events` that have their own `founding` event)...
+This is basically the ultimate nested boss of modularity!
 
-## A correlated trait
-traits <- make.traits(n = 2, process.args = list(Sigma = matrix(1, 2, 2)))
-
-## Event changing a trait correlation
-cor_change <- make.events(
-    condition    = trait.condition(3, absolute = TRUE),
-    target       = "traits",
-    modification = update.traits(process.args = list(Sigma = matrix(c(10,3,3,2),2,2))))
-
-## Run the simulations
-set.seed(2)
-no_event <- dads(bd.params = bd.params,
-                 stop.rule = stop.rule,
-                 traits    = traits)
-set.seed(2)
-change_correlation <- dads(bd.params = bd.params,
-                           stop.rule = stop.rule,
-                           traits    = traits,
-                           events    = cor_change)
-
-## Visual testing
-par(mfrow = c(1,2))
-plot(no_event, trait = 2)
-plot(change_correlation, trait = 2)
-
-```
-
-```{r, webgl=TRUE}
-## 3D plot
-plot(change_correlation, trait = c(1:2), use.3D = TRUE)
-```
-
-### Event for changing a modifier
-
-```{r}
-## Set the parameters
-stop.rule <- list(max.time = 4)
-bd.params <- list(extinction = 0, speciation = 1)
-
-## Using a default trait and modifier object
-modifiers <- make.modifiers()
-traits <- make.traits()
-
-## New condition and new modifier (increasing speciation if trait is negative)
-new.condition <- function(trait.values, lineage) {
-    return(parent.traits(trait.values, lineage) < 0)
-}
-new.modify    <- function(x, trait.values, lineage) {
-  return(x + 1)
-}
-
-## Update the modifier
-change_speciation <- make.events(
-    condition    = time.condition(3),
-    target       = "modifiers",
-    modification = update.modifiers(speciation = speciation,
-                                    condition  = new.condition,
-                                    modify     = new.modify))
-
-set.seed(4)
-no_event <- dads(bd.params = bd.params,
-                 stop.rule = stop.rule,
-                 traits    = traits,
-                 modifiers = modifiers)
-set.seed(4)
-change_spec <- dads(bd.params = bd.params,
-                    stop.rule = stop.rule,
-                    traits    = traits,
-                    modifiers = modifiers,
-                    events    = change_speciation)
-
-## Visualise the results
-par(mfrow = c(1,2))
-plot(no_event, main = "No event")
-plot(change_spec, main = "Increase extinction for negative\ntraits after time 3")
-```
-
-### Changing branch length when reaching n taxa
-
-```{r}
-## Parameters
-bd.params <- list(extinction = 0, speciation = 1)
-stop.rule <- list(max.taxa = 100)
-modifiers <- make.modifiers()
-traits <- make.traits()
-
-## multiplying branch length 100 folds
-new.modify <- function(x, trait.values, lineage) {
-  return(x * 100)
-}
-
-## Event for increasing branch length after reaching 30 taxa
-increase_brlen <- make.events(
-                  condition    = taxa.condition(30),
-                  target       = "modifiers",
-                  modification = update.modifiers(
-                                    branch.length = branch.length,
-                                    modify = new.modify))
-
-## Run the simulations
-set.seed(5)
-no_event <- dads(bd.params = bd.params,
-                 stop.rule = stop.rule,
-                 traits    = traits,
-                 modifiers = modifiers)
-set.seed(5)
-increased_brlen <- dads(bd.params = bd.params,
-                        stop.rule = stop.rule,
-                        traits    = traits,
-                        modifiers = modifiers,
-                        events    = increase_brlen)
-
-## Visualise the results
-par(mfrow = c(1,2))
-plot(no_event, main = "No event")
-plot(increased_brlen, main = "Increase branch length\nafter 30 taxa")
-```
-
-### Founding event: a subtree with no fossils
+The founding event will basically run an internal `dads` process (i.e. simulating a tree and, optionally, some data) resulting in a founding sub-tree.
+It will then branch this sub-tree to the rest of the simulation that continued normally in the mean time.
+You can specify the founding event using the inbuilt `founding.event` modification event.
+This function intakes the exact same arguments as `dads` to simulate the sub-tree with its own parameters.
+Additionally, we will use the `additional.args` argument from `make.events` to specify a prefix for the founding tree tips (to make them easier to distinguish).
 
@@
 stop.rule <- list(max.time = 4)
-bd.params <- list(speciation = 1, extinction = 0.3)
+bd.params <- make.bd.params(speciation = 1, extinction = 0.3)
 
-## Events that generate a new process (founding effects)
+## Events that generate a new process (founding tree - with no extinction)
 founding_event <- make.events(
+                  target       = "founding",
                   condition    = taxa.condition(10),
-                  target       = "founding",
                   modification = founding.event(
-                                    bd.params = list(speciation = 2,
+                                    bd.params = make.bd.params(speciation = 2,
                                                                extinction = 0)),
@@
 
-### Founding event with traits
-
-```{r}
-## Set up parameters
-stop.rule <- list(max.time = 4)
-bd.params <- list(speciation = 1, extinction = 0.3)
-traits    <- make.traits()
-
-## Events that generate a new process (founding effects)
-founding_event <- make.events(
-                  condition    = taxa.condition(10),
-                  target       = "founding",
-                  modification = founding.event(
-                                    bd.params = list(speciation = 2,
-                                                     extinction = 0),
-                                    traits = make.traits(process = OU.process)),
-                  additional.args = list(prefix = "founding_"))
-
-## Simulating the tree
-set.seed(19)
-founding_tree <- dads(bd.params = bd.params,
-                      stop.rule = stop.rule,
-                      traits    = make.traits(),
-                      events    = founding_event)
-plot(founding_tree)
-```
+Note that the nestedness here is potentially endless, for example, you can parse an `event` argument to the `founding.event` that will generate another founding event, etc...

### 2023-02-22 inst/gitbook/07_specific_examples.Rmd
Create 07_specific_examples.Rmd

--- 2021-01-19 inst/gitbook/dads_manual.Rmd
@@
 ---
-title: "dads manual"
-author: "Thomas Guillerme (guillert@tcd.ie) and others"
-date: "`r Sys.Date()`"
-site: bookdown::bookdown_site
-output: bookdown::gitbook
-documentclass: book
-bibliography: [../references.bib]
-link-citations: yes
-github-repo: TGuillerme/dads
-description: "dads R package vignette"
----
-
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
-```
-
-# `dads`: disparity and diversity simulations.
-
-Allowing to simulate disparity and diversity at the same time with interaction between both.
-
-The core of this package is based on the [`diversitree` birth-death algorithm](https://github.com/cran/diversitree/blob/master/R/simulate-bd.R).
-
-## What is `dads`?
-
-There are some very good packages out there to simulate birth death trees (like [TreeSim](https://github.com/tanja819/TreeSim)) or even packages simulating traits (disparity) and diversity jointly (like [RPANDA](https://github.com/cran/RPANDA) or [PETER](https://github.com/PuttickMacroevolution/PETER)).
-We strongly advice you have a look at these packages first as they might be more appropriate for your tasks.
-
-`dads` aims to be a highly modular and friendly version of all these packages: it allows to simulate disparity and diversity jointly with a vast array of options that can be easily modified by users.
-For example, you can easily generate any type of process to generate a trait (BM, OU, something else, etc...) in multiple dependent or independent dimensions through `"traits"` objects.
-You can then specify how the traits should affect disparity through `"modifiers"` objects. And finally you can create events (like mass extinctions) through `"events"` objects.
-These objects and how to modify them will be detailed throughout this manual.
-Finally we are putting an emphasise in the development of this package on the speed and reliability of the functions.
-
-### Modular?
-
-Because their is an infinite way you might want to generate disparity and diversity (different traits, different modifiers and different events), the `dads` package is designed to make all these parts easy to code separately and integrate them easily in the `dads` core functions.
-This allows you to simulate finely tuned multidimensional data for your specific project!
-
-## Installing and running the package
-
-You can install this package easily, directly from the github:
-<!-- TG: I do intend to release it on CRAN later on -->
-
-```{r, eval = FALSE}
-## Checking if devtools is already installed
-if(!require(devtools)) install.packages("devtools")
-
-## Installing the latest version directly from GitHub
-devtools::install_github("TGuillerme/dads")
-```
-
-```{r, eval = FALSE, echo = FALSE, message = FALSE}
-## Loading the package and setting up the start seed.
-library(dads)
-```
-
-## Help
-
-If you need help with the package, hopefully the following manual will be useful.
-However, parts of this package are still in development and some other parts are probably not covered.
-Thus if you have suggestions or comments on on what has already been developed or will be developed, please send me an email (<guillert@tcd.ie>) or if you are a GitHub user, directly create an issue on the [GitHub page](https://github.com/TGuillerme/dads).
-Doing so will not only hopefully help you but also other users since it will help improve this manual!
-
-## How does `dads` work?
-
-Basically, the `dads` function intakes your personalised `traits`, `modifiers` and `events` to generate your disparity and diversity.
-You will find more details about how these objects (`traits`, `modifiers` and `events`) work in the rest of the tutorial but here is a graphical representation of how `dads` work:
-
-![Schematised summary of the `dads` package architecture](dads_structure.pdf)
-
-<!--chapter:end:index.Rmd-->
-
----
-title: "Getting start"
-author: "Thomas Guillerme (guillert@tcd.ie)"
-date: "`r Sys.Date()`"
-bibliography: [../References.bib, ../packages.bib]
-output:
-  html_document: default
-  pdf_document: default
----
-
-```{r, echo = FALSE}
-library(dads)
-set.seed(4242)
-```
-
-# Getting started
-
-## The simplest of all analysis: simulating diversity only
-
-One of the simplest things to do with the `dads` package is just to simulate a birth death tree.
-For that you can use the function `dads` and specify your stopping rule.
-The stopping rule simply tells the birth death process to step whenever it reaches one of these three conditions:
-
- * `"max.taxa"   = n` stop when `n` taxa are generated;
- * `"max.living" = n` stop when there is `n` co-occuring taxa of the same age (i.e. "living" taxa);
- * `"max.time"   = n` stop when the simulated tree is `n` units of age old (these units are arbitrary);
-
-For example, we might want to generate a birth-death tree with 20 taxa:
-
-```{r}
-## Setting a stopping rule to reach a maximum of 20 taxa
-my_stop_rule <- list(max.taxa = 20)
-```
-
-We can now run the simulations using:
-
-```{r}
-## Running the birth death simulation
-my_tree <- dads(stop.rule = my_stop_rule)
-```
-
-> Note that here we could have specified more than one stopping rule, for example, we might want to run a simulation and stop it if it either reaches 10 taxa or the age 2 using `stop.rule = list(max.time = 2, max.taxa = 10)`.
-The simulation will then stop when either of these conditions are met.
-
-The resulting object is a classic `"phylo"` object that you can simply plot or visualise like so:
-
-```{r}
-## The tree object
-my_tree
-## Plotting it
-plot(my_tree)
-```
-
-### Changing the birth-death parameters
-
-People familiar with the [birth-death models](https://lukejharmon.github.io/pcm/chapter10_birthdeath/) might have noticed that we did not specify two important things here: the speciation parameter (sometimes called "lambda" or "birth") and the extinction parameter (sometimes called "mu", "death" or "background extinction").
-By default `dads` runs a pure birth model (the speciation is set to 1 and the extinction to 0).
-However, you can easily change that by specifying your new birth death parameters:
-
-```{r}
-## my birth death parameters
-my_params <- list(speciation = 1,
-                  extinction = 1/3)
-```
-> Note that here it is not necessary to specify `extinction = 1` since this is the default option, you can always just change the parameter of interest (e.g.
-changing `extinciton = 0` to `extinction = 1/3`).
-However, we think it's good practice to attribute both parameters specifically to avoid any confusion.
-
-You can then run the same birth death tree with extinction:
-
-```{r}
-## Generating a birth death tree with extinctions:
-my_tree <- dads(bd.params = my_params, stop.rule = my_stop_rule)
-## Visualising the new tree
-plot(my_tree)
-```
-
-## Slightly more complex: simulating disparity and diversity
-
-Chances are that you want to also simulate traits (disparity) along with your diversity (otherwise, we suggest using the [`TreeSim`](https://github.com/tanja819/TreeSim/) package that provides many more birth death models).
-Simulating traits is not much more complicated in `dads`: you'll simply need to create a `"traits"` object using the `make.traits` function.
-These objects can have increasing complexity (see the rest of this tutorial) but we will keep it simple here.
-
-`"traits"` objects contain one or more processes which are the ways to generate the trait.
-The most common of these processes is the [Brownian Motion](https://en.wikipedia.org/wiki/Brownian_motion).
-This is used by default with the `make.traits` function:
-
-```{r}
-## Creating the traits object
-my_trait <- make.traits()
-```
-
-This trait object can be simply printed (to see what's in it) or plotted (to see what the process looks like in the absence of a phylogeny):
-
-```{r}
-## Which process is in here?
-my_trait
-## What does it look like?
-plot(my_trait)
-```
-
-By default, this trait is called "A".
-This is not a really good name but you'll see more about specifying trait names later on.
-If this is what the process should look like (theoretically) you can then add its `"traits"` object to our previous `dads` function to generate the tree and the traits:
-
-```{r}
-## Simulate disparity and diversity
-my_data <- dads(bd.params = my_params,
-                stop.rule = my_stop_rule,
-                traits    = my_trait)
-```
-
-Et voilà! We now have a simple disparity and diversity simulation.
-We can see what's in the results by simply printing it or plotting it:
-
-```{r}
-## What's in there
-my_data
-## Plotting the disparity and diversity
-plot(my_data)
-```
-
-You can then extract the components you need for your specific analysis like so:
-
-```{r}
-## Extracting the tree (a "phylo" object)
-the_generated_tree <- my_data$tree
-class(the_generated_tree)
-
-## Extracting the data (a "matrix")
-the_generated_data <- my_data$data
-class(the_generated_data)
-```
-
-You can find much more on how to design trait objects in the [`make.traits` section](#maketraits).
-
-## Slightly more more complex: simulating linked disparity and diversity
-
-The example above is also still pretty simple and easily done through a variety of `R` packages: here the trait and the tree are simulated at the same time but only the tree is simulating the trait (i.e.
-the trait value at a tip is affected by it's ancestor and the branch length leading to it) but not the other way around (the trait value does not affect the tree).
-It is possible to add this aspect using `"modifiers"` objects.
-`"modifiers"` are similar to `"traits"` in that you specify what should go in there and then feed it to your simulation.
-
-`"modifiers"` affect two key steps of the birth-death process: the calculation of the waiting time (i.e. the component generating branch lengths) and the triggering of speciation or extinction events.
-These events can be modified using some `condition` and `modify` function.
-In other words, when reaching a certain condition specified by a `condition` function, the birth-death process will modify either the branch length or the speciation (or extinction) probability by applying a `modify` function.
-
-You can use the function `make.modifiers` to design a specific `"modifiers"` object.
-By default, this function generates a `"modifiers"` object that affects branch length and speciation in the following way:
-
- * branch length is a randomly drawn number from an exponential distribution with a rate equal to the current number of taxa multiplied by the sum of the speciation and extinction rates.
- * speciation is triggered if a randomly drawn number (from a (0,1) uniform distribution) is smaller than the ratio between the speciation rate and the sum of the speciation and extinction rates.
-If that random number is greater, the lineage goes extinct.
-
-Note that these are default for a birth death tree and were actually already applied in the examples before (without specifying a modifier):
-
-```{r}
-## Make a default modifiers
-default_modifiers <- make.modifiers()
-## What's in it?
-default_modifiers
-```
-
-This will not do much to our simulations compared to the previous trait and tree simulation but we can provide our modifiers object to the `dads` function:
-
-```{r}
-## Setting the simulation parameters
-extinction_02 <- list(extinction = 0.2)
-living_20     <- list(max.living = 20)
-BM_trait      <- make.traits()
-set.seed(1)
-## Simulate disparity and diversity
-default_data <- dads(bd.params = extinction_02,
-                     stop.rule = living_20,
-                     traits    = BM_trait,
-                     modifiers = default_modifiers)
-default_data
-```
-
-Note however that the printing information is now updated to state that you've add a modifier (even though it's a default one).
-
-For more interesting simulations however, you can provide modifiers that actually modify the birth death process.
-We can create one for example that makes species go extinct if their ancestor have a negative trait value.
-For that we need to create a modifiers object that modifies the `speciation` process with a specific condition and a specific modification when that condition is met.
-First we are going to create the modification function.
-This function must intake the argument `x` and, in our case, return a logical value: `TRUE` is for speciate and `FALSE` is for go extinct.
-We want this `modify` function to always return `FALSE` (go extinct) since it will be triggered by `condition` (we will see that in a minute).
-
-```{r}
-## Going extinct
-staying.alive <- function(x) return(TRUE)
-```
-
-Now we want this function to only trigger when an ancestor has a negative trait value.
-We can do that by specifying our `condition` function (when the ancestor is trait is negative) apply our modification (here the `staying.alive` function).
-For that we can use the `parent.traits` utility function that is optimised for accessing traits in the birth death process (but you can of course write your own).
-This function intakes the `trait.values` and `parent.lineage` arguments, two arguments that you can leave named as they are to facilitate `dads`s understanding of what you which to asses:
-
-```{r}
-## Triggering a modification only if the ancestor trait is negative
-negative.ancestor <- function(trait.values, lineage) {
-    return(all(parent.traits(trait.values, lineage) < 0))
-}
-```
-
-Note that we use the function `all` here to evaluate all traits (if the data is multidimensional). 
-We can then provide these two functions (the condition `negative.ancestor` and how to modify the speciation event when this condition is met `staying.alive`).
-If you are an advances `dads` user, you can design your own `speciation` function but if you just want to use a normal `speciation` function, you can use the default one from `dads` called...
-`speciation`.
-
-```{r}
-## Making a modifier for species to go extinct if
-## their ancestor's trait value is (or are) negative
-negatives_extinct <- make.modifiers(
-                        ## The speciation function (default)
-                        speciation = speciation,
-                        ## What to modify
-                        modify     = staying.alive,
-                        ## When to modify it
-                        condition  = negative.ancestor)
-## What's in it?
-negatives_extinct
-```
-
-Note that the `make.modifiers` function tests whether the input is compatible with `dads` by default so unless you have an error message, your `modifiers` will work!
-We can now simulate our tree and traits with our modifier: species will go extinct if their ancestor have a negative trait value:
-
-```{r}
-set.seed(1)
-## Simulate disparity and diversity
-biased_data <- dads(bd.params = extinction_02,
-                    stop.rule = living_20,
-                    traits    = BM_trait,
-                    modifiers = negatives_extinct)
-biased_data
-```
-
-We can now compare the two trees and their trait values.
-Note that we've used the same starting seed for both trees so the only thing differing between them is the modifier!
-Also, although species
-
-```{r, fig.height = 6, fig.width = 12}
-par(mfrow = c(1,2))
-plot(default_data, main = "default results")
-plot(biased_data, main = "results with the modifier")
-```
-
-You can find much more how to design modifiers in the [`make.modifiers` section](#makemodifiers).
-
-<!--chapter:end:01_getting_started.Rmd-->
-
----
-title: "`make.traits`"
-author: "Thomas Guillerme (guillert@tcd.ie)"
-date: "`r Sys.Date()`"
-bibliography: [../References.bib, ../packages.bib]
-output:
-  html_document: default
-  pdf_document: default
----
-
-# Making trait processes with `make.traits()` {#maketraits}
-
-```{r, echo = FALSE}
-set.seed(123)
-```
-
-## The process (`process`)
-
-The function `make.traits` allows you to design the process of a trait or a set of traits.
-Here, the process of a trait designates the rules to generate the trait through time while simulating a phylogeny.
-This process can depend on the previous state in the tree (i.e. the trait of the ancestor) and the branch length to the descendant.
-One classic example is the [Brownian motion process (or Weiner process)](https://en.wikipedia.org/wiki/Brownian_motion).
-Note that it _can_ depend on both the ancestor and the branch length but does _not necessary needs_ (i.e. the process can be only based on the previous state or only on branch length or on neither).
-
-Trait processes in `dads` are functions that must always intake the following arguments by default.
-
- * `x0`: the previous trait value(s)
- * `edge.length`: the branch length value
- * `...`: a placeholder for any extra arguments
-
-For example, the following function would be a valid process (though not dependent on either the previous state nor the branch length):
-
-```{r}
-## A valid (but useless) process
-valid.process <- function(x0, egde.length = 1, ...) {
-    return(42)
-}
-```
-
-> Note that the argument `edge.length` is set to `1` by default.
-In general we highly recommend to set all arguments but `x0` to a default value (this really helps the speeding up the `dads` function).
-
-On the other hand, the following process (a unidimensional Brownian motion) is incorrect (it's missing `edge.length` and `...`):
-
-```{r}
-## A wrongly formated process
-invalid.process <- function(x0) {
-    return(rnorm(1, mean = x0))
-}
-```
-
-The `dads` package proposes inbuilt processes, namely a multidimensional Brownian motion (`BM.process`) or a a multidimensional Ornstein-Uhlenbeck process (`OU.process`).
-You can find the list of implemented process by looking at the `?trait.process` manual page in `R`.
-
-Once a process is chosen, you can feed it to the `make.traits` function:
-
-```{r}
-## Creating a trait object
-my_trait_object <- make.traits(process = BM.process)
-```
-
-This creates `"dads"` `"traits"` objects that you can print, and visualise using the `plot` function:
-
-```{r}
-## The class of the object
-class(my_trait_object)
-## What's in it?
-my_trait_object
-## What does the process looks like
-plot(my_trait_object)
-```
-
-Note that you can see the multiple options for plotting the trait process by looking at `?plot.dads` manual. Furthermore, you can look at what's actually in the object using:
-
-```{r}
-## What's actually in that object?
-print.dads(my_trait_object, all = TRUE)
-```
-
-As traits can get more and more complex, the automatic printing of its summary allows for a easier display of what's in the traits object.
-
-Note that it is possible to make `"traits"` objects with multiple processes (that can be the same):
-
-```{r}
-## 4 traits: two BM, one OU and one normal non process
-four_traits <- make.traits(process = c(BM.process,
-                                       BM.process,
-                                       OU.process,
-                                       no.process))
-four_traits
-```
-
-You can visualise them individually using the `trait` argument in `plot.dads`:
-
-```{r}
-## Plot options (4 plots in one window)
-par(mfrow = c(2,2))
-plot(four_traits, trait = 1)
-plot(four_traits, trait = 2)
-plot(four_traits, trait = 3)
-plot(four_traits, trait = 4)
-```
-
-## The number of traits `n` and the starting values `start`
-
-Two further important arguments are `n` the number of traits per process and `start` the starting values for all traits.
-By default they are set to `n = 1` and `start = 0`.
-This means that `make.traits` will assume that your processes are always unidimensional by default and that they always start with the value `0`.
-It is however possible to change these values.
-
-For example you can use the following to create a three dimensional Brownian motion with each dimensions starting with the value `1`:
-
-```{r}
-## Multidimensional Brownian motion
-make.traits(BM.process, n = 3, start = 1)
-```
-
-Or the following with each dimensions starting with different values (respectively `1`, `2` and `3`):
-
-```{r}
-## Multidimensional Brownian motion
-make.traits(BM.process, n = 3, start = c(1,2,3))
-```
-
-Note that the number of traits are distributed per processes.
-If the traits contains multiple process, the number of traits are distributed per processes:
-
-```{r}
-## two 3D processes (BM and OU)
-make.traits(c(BM.process, OU.process), n = 3)
-## one 1D processes (BM) and one 4D process (OU)
-make.traits(c(BM.process, OU.process), n = c(1, 4))
-```
-
-And starting values are distributed for all the traits or for the traits one by one:
-
-```{r}
-## two 3D processes (BM and OU) starting with 1
-make.traits(c(BM.process, OU.process), n = 3, start = 1)
-## two 3D processes (BM and OU) starting with values 1 to 6
-make.traits(c(BM.process, OU.process), n = 3, start = 1:6)
-## two 3D processes (BM and OU) with the two first ones starting
-## with 1 and the 4 other ones with the default (0)
-make.traits(c(BM.process, OU.process), n = 3, start = c(1,1))
-```
-
-## Extra argument for the processes with `process.args`
-
-You can also feed extra arguments to your process(es) functions.
-For example, the inbuilt process `no.process` (that is just a number generator not based on the previous value `x0` or the branch length) can intake a specific random number generator as a function:
-
-```{r}
-## no process trait using the normal distribution (default)
-make.traits(no.process, process.args = list(fun = rnorm))
-## no process trait using the uniform distribution
-## bounded between 1 and 100
-make.traits(no.process, process.args = list(fun = runif, min = 1, max = 100))
-```
-
-You can also add multiple extra arguments for multiple processes giving them as a list.
-
-```{r}
-## Two traits with no process:one normal and one uniform (1,100)
-make.traits(process      = c(no.process, no.process),
-            process.args = list(list(fun = rnorm),
-                                list(fun = runif, min = 1, max = 100)))
-```
-
-If one process do not need extra argument you must still give it and extra `NULL` process argument:
-
-```{r}
-## Three traits with no process:
-## one default, one lognormal and one uniform (1,100)
-make.traits(process      = c(no.process, no.process, no.process),
-            process.args = list(## Extra arguments for the first process (none)
-                                list(NULL),
-                                ## Extra arguments for the second process
-                                list(fun = rlnorm),
-                                ## Extra arguments for the third process
-                                list(fun = runif, min = 1, max = 100)))
-```
-
-## Naming the traits with `trait.names`
-
-As traits become more and more complex, it can be useful to give clearer names to each process.
-This is easily done using the `trait.names` argument that attributes one name per process:
-
-```{r}
-## A simple trait with a proper name
-simple_trait <- make.traits(trait.names = "1D Brownian Motion")
-simple_trait
-```
-
-This becomes more useful if we use the complex example above:
-
-```{r}
-## Three named traits with no process:
-## one default, one lognormal and one uniform (1,100)
-make.traits(process      = c(no.process, no.process, no.process),
-            process.args = list(## Extra arguments for the first process (none)
-                                list(NULL),
-                                ## Extra arguments for the second process
-                                list(fun = rlnorm),
-                                ## Extra arguments for the third process
-                                list(fun = runif, min = 1, max = 100)),
-            ## Naming each trait
-            trait.names  = c("Normal", "LogNormal", "Uniform(1,100)"))
-```
-
-## Combining multiple traits with `add`
-
-You can also add traits to already existing trait objects using the simple `add` option.
-This option just intakes a `"dads"` `"traits"` object and the additional process(es) will be added to it. For example:
-
-```{r}
-## Creating on simple default Brownian motion
-one_process <- make.traits(trait.names = "BM")
-
-## Creating a new trait (a 3D OU.process)
-## and adding the previous one
-two_processes <- make.traits(OU.process, n = 3, add = one_process,
-                             trait.names = "3D OU")
-
-## Only one process
-one_process
-
-## The two processes
-two_processes
-```
-
-## Testing the traits with `test`
-
-This bit is more for development.
-We highly suggest leaving `test = TRUE` so that `make.traits` returns an error if a process or its additional arguments (`process.args`) are not formatted correctly.
-`make.traits` will error if the trait cannot be directly passed to `dads`.
-However, in some specific cases (again, probably mainly for development and debugging) it could be useful to skip the tests using `test = FALSE`.
-
-## Templates for making your very own process
-
-As detailed above, any process of your own design will work as long as it is a function that takes at least the arguments `x0` and `edge.length`.
-You can be imaginative and creative when designing your own process but here are two detailed example functions for a unidimensional Brownian Motion and Ornstein-Uhlenbeck process that you can use for a start (or not).
-Remember it is good practice for `dads` processes to set all the arguments but `x0` with default values (just in case).
-Also, note that the functions below are not equal to the already implemented `BM.process` and `OU.process` but are rather generalised/simplified version that you can use as a template
-
-### A simple Brownian Motion process template
-
-```{r}
-## A simple Brownian motion process
-my.BM.process <- function(x0, edge.length = 1, sd = 1, ...) {
-    ## Drawing a random number from a normal distribution
-    ## with x0 as the and a given standard deviation
-    ## and depending on branch (edge) length
-    result <- rnorm(n = 1, mean = x0, sd = sd * edge.length)
-
-    ## Return the number
-    return(result)
-}
-```
-
-### A simple Ornstein-Uhlenbeck process template
-
-```{r}
-## A simple Ornstein-Uhlenbeck motion process
-my.OU.process <- function(x0, edge.length = 1, var = 1, alpha = 1, ...) {
-    ## Calculate the mean based on alpha
-    mean <- x0 * exp(-alpha)
-    ## Calculate the standard deviation based on alpha and the variance
-    sd <- sqrt(var/(2 * alpha) * (1 - exp(-2 * alpha)))
-    ## Draw a random number from a normal distribution
-    ## using this mean and standard deviation
-    ## and depending on branch (edge) length
-    result <- rnorm(n = 1, mean = mean, sd = sd * branch.length)
-
-    ## Return the number
-    return(result)
-}
-```
-
-<!--chapter:end:02_make.traits.Rmd-->
-
----
-title: "`make.modifiers`"
-author: "Thomas Guillerme (guillert@tcd.ie)"
-date: "`r Sys.Date()`"
-bibliography: [../References.bib, ../packages.bib]
-output:
-  html_document: default
-  pdf_document: default
----
-
-# Making modifiers with `make.modifiers()` {#makemodifiers}
-
-```{r, echo = FALSE}
-set.seed(123)
-```
-
-`"modifiers"` have a similar structure than `"traits"` where you can design an object with increasing complexity, starting with the simplest modifiers that doesn't modify anything (using the default arguments):
-
-```{r}
-## Making a default modifier (no modification)
-my_default_modifiers <- make.modifiers()
-my_default_modifiers
-```
-
-Similarly to `"traits"` objects, `"modifiers"` are also printed by default using `print.dads`.
-You can see details about what's actually in the object using `print.dads(my_default_modifiers, all = TRUE)`.
-However, contrary to `"traits"`, you cannot plot `"modifiers"`.
-
-## The branch length function (`branch.length`)
-
-The first argument in `"modifiers"` is the branch length function (`branch.length`) this is the function that will be executed in `dads` to generate branch length.
-Note that in the `dads` algorithm, branch length is not generated _directly_ but actually results of the waiting time.
-In other words, the `branch.length` function just affects waiting time for all taxa present at any time in the simulation.
-These taxa can then either go extinct (stopping the "growth" of it's branch length) or survive (continuing the "growth").
-
-By default, branch length (or waiting or growth) is a randomly drawn number from an exponential distribution with the rate of the number of taxa multiplied by the speciation and extinction rate: $n \times (\lambda + \mu)$ (where $n$ is the number of taxa currently present in the simulation, $\lambda$ and $\mu$ are respectively the speciation and extinction rates).
-This default function is simply called `branch.length` in `dads` and can be used as a modifier as follows:
-
-```{r}
-## Specifying the default modifier
-default_modifiers <- make.modifiers(branch.length = branch.length)
-
-## Setting some parameters for generating trees
-bd_params <- list(extinction = 0)
-stop_rule <- list(max.living = 20)
-
-## Generating a tree with the default branch length parameter
-set.seed(0)
-default_tree <- dads(bd.params = bd_params,
-                     stop.rule = stop_rule,
-                     modifiers = default_modifiers)
-```
-
-Of course, the point of the modularity here is that you can provide your own function for generating branch length.
-For example, we might be interested in what our tree would look like if we'd use a simple constant branch length generation (instead of randomly drawing it from an exponential distribution).
-We can do so by declaring our own `branch.length` function and adding it to a `"modifiers"` object.
-
-```{r}
-## A constant branch length generator
-## (note that the output must be numeric, not integer)
-constant.brlen <- function() {
-    return(as.numeric(1))
-}
-
-## Creating the modifiers object
-constant_modifier <- make.modifiers(branch.length = constant.brlen)
-
-## Generating a new tree with this modifier
-set.seed(0)
-modified_tree <- dads(bd.params = bd_params,
-                      stop.rule = stop_rule,
-                      modifiers = constant_modifier)
-```
-
-And we can visualise the difference between both resulting trees:
-
-```{r}
-par(mfrow = c(1,2))
-plot(default_tree,  main = "Default modifier")
-plot(modified_tree, main = "Constant branch length\nmodifier")
-```
-
-It is of course to use more complex branch length modifiers that intakes different conditions and specific modification rather than simply always output a value of one.
-
-### The allowed arguments {#allowarguments}
-
-It is of course possible to design some more advanced function to interact with the birth death process.
-You can create a function for `branch.length`, `selection` and `speciation` that involve any of the following arguments:
-
-  * `bd.params`: a named list containing `"numeric"` values that contains the birth death parameters (at least `"speciation"` and `"extinction"`);
-  * `lineage`: a named list containing the lineage data (see below).
-  * `trait.values`: a `"matrix"` containing `"numeric"` values with the trait names as column names and the lineages ID as row numbers (you can use it with the function `parent.traits` to access the trait of the previous node for example).
-  * `modify.fun`: a `"list"` of named `"function"` (usually passed through `condition` and `modify`).
-
-The `lineage` list contains the following elements (missing elements are allowed):
-
- * `lineage$parents`: an `"integer"` vector: the list of parent lineages;
- * `lineage$livings`: an `"integer"` vector: the list of lineages still not extinct;
- * `lineage$drawn`: a single `"integer"`: the ID of the selected lineage;
- * `lineage$current`: a single `"integer"`: the selected lineage (is equal to `lineage$livings[lineage$drawn]`);
- * `lineage$n`: a single `"integer"`: the current number of non extinct lineage (is equal to `length(lineage$livings))`;
- * `lineage$split`: a `"logical"` vector: the list of splits for each lineage (`TRUE`), the number of total tips is equal to `sum(!lineage$split)`.
-
-In general, unless you know what you're doing, you can ignore most arguments for specific modifiers since they are handled automatically within the `dads` function.
-Therefore any argument can be left undeclared or missing and is always handled internally.
-For example, if you did not declare `n.taxa` as a function argument but are using `n.taxa` in the function, `n.taxa` will be detected and treated as a current argument automatically as set accordingly within the birth death process (e.g.
-`n.taxa` will be set to the current number of taxa every iteration of the process).
-
-For example, we can create a function that increases branch length proportional to the number of species "alive" at each time of the simulation in a discrete way.
-I.e.
-for discrete numbers of taxa, the branch length increases by jumps (ten fold):
-
-```{r}
-## A more complex binned.branch.length function
-increasing.brlen <- function(bd.params, lineage) {
-
-    ## Setting the cumulated birth and death
-    birth_death <- bd.params$speciation + bd.params$extinction
-
-    ## Returning branch lengths depending on different number of taxa
-    if(lineage$n <= 5) {
-        return(1    * rexp(1, sum(5 * birth_death)))
-    }
-    if(lineage$n <= 10) {
-        return(10   * rexp(1, sum(10 * birth_death)))   
-    }
-    if(lineage$n <= 15) {
-        return(100  * rexp(1, sum(15 * birth_death)))      
-    }
-    if(lineage$n <= 20) {
-        return(1000 * rexp(1, sum(20 * birth_death)))
-    } else {
-        return(1000 * rexp(1, sum(lineage$n * birth_death)))
-    }
-}
-```
-
-We can then create it as a `"modifiers"` object and run a new simulation:
-
-```{r}
-## Creating a modifiers
-increasing_modifier <- make.modifiers(branch.length = increasing.brlen)
-
-## Generating a new tree with this modifier
-set.seed(0)
-increasing_tree <- dads(bd.params = bd_params,
-                        stop.rule = stop_rule,
-                        modifiers = increasing_modifier)
-```
-
-And we can visualise the difference between the resulting trees:
-
-```{r}
-par(mfrow = c(1,3))
-plot(default_tree,    main = "Default modifier")
-plot(modified_tree,   main = "Constant branch length\nmodifier")
-plot(increasing_tree, main = "Increasing branch length\nmodifier (binned)")
-```
-
-## The selection function (`selection`)
-
-The `selection` function is used in the birth death process to know which lineage to select when running a speciation (or extinction!) event.
-By default, this function randomly selects one taxon that is currently not extinct (using: `sample(1:n.taxa, 1))`.
-Similarly as `branch.length` it is possible to modify this part of the birth death process.
-For example, we could simply select always the last created lineage (to create a "ladder" or most asymmetric tree):
-
-```{r}
-## Our function to always select the last taxon
-## (making sure it returns an integer)
-select.last <- function(lineage) {
-    return(as.integer(lineage$n))
-}
-```
-
-> Note that here the function can only intake the [allowed arguments as described above](#allowarguments) (here `n.taxa`: the number of current living taxa).
-
-We can then create a `"modifiers"` object the same way as before using this time the `selection` argument:
-
-```{r}
-## A modifier for selection
-ladderised_modifier <- make.modifiers(selection = select.last)
-## Generating a new tree with this modifier
-set.seed(0)
-ladderised_tree <- dads(bd.params = bd_params,
-                        stop.rule = stop_rule,
-                        modifiers = ladderised_modifier)
-## Displaying the results
-par(mfrow = c(1,2))
-plot(default_tree,    main = "Default modifier")
-plot(ladderised_tree, main = "Ladderising modifier")
-```
-
-Again, it is of course possible to make the modifier more complex and in combination with other elements of the tree.
-For example, we can create a `"dads"` object that also creates a trait a add to it a `selection` modifier that only selects for tips with positive trait values.
-
-```{r}
-## Our function that only select taxa with positive trait values
-select.positive <- function(trait.values, lineage) {
-    ## Selecting the taxa with positive traits only
-    positives <- as.integer(rownames(trait.values)[which(trait.values[, 1] >= 0)])
-
-    ## Select the current taxa that descend from a node with a positive value
-    positive_living <- cbind(lineage$parents, seq_along(lineage$split)
-                             )[which(lineage$parents %in% positives), 2]
-
-    ## Select one tip randomly in the ones with descendants with positive values
-    return(sample(which(lineage$livings %in% positive_living), 1))
-}
-
-## Creating the modifier
-positive_skew <- make.modifiers(selection = select.positive)
-
-## Creating a (default) trait object
-BM_trait <- make.traits()
-
-## Simulate a tree and trait with no modifier
-set.seed(1)
-default_dads <- dads(bd.params = bd_params,
-                     stop.rule = stop_rule,
-                     traits    = BM_trait)
-
-## Simulate a tree and trait with the modifier
-set.seed(1)
-skewed_trait_dads <- dads(bd.params = bd_params,
-                          stop.rule = stop_rule,
-                          traits    = BM_trait,
-                          modifiers = positive_skew)
-
-## Plotting the differences in trees and traits
-par(mfrow = c(1, 2))
-plot(default_dads, main = "Default trait and tree")
-plot(skewed_trait_dads, main = "Skewed trait and tree")
-```
-
-## The speciation function (`speciation`)
-
-The third function that can be used to modify the birth death process is the `speciation` function.
-This one is used during the birth death process to decide whether a lineages speciates (creating a node and two new lineages) or goes extinct (creating a tip).
-
-> Note that the `speciation` function only affects tips or nodes before the simulation reaches the `stop.rule`.
-The then surviving lineages are all automatically transformed into tips.
-
-By default, the `speciation` function is trigger a speciation even if a number randomly drawn from a uniform distribution is lower than the ratio between the speciation and the speciation and extinction parameter.
-If the randomly drawn number is higher, the lineage goes extinct.
-
-```{r, eval = FALSE}
-## The speciation in pseudo-code:
-runif(1) < speciation/ (speciation + extinction)
-```
-
-Creating a `"modifiers"` with a `speciation` function works the same way as for `branch.length` and `selection` but the function that will be used needs to output a logical value (see table [below](#summarymodifiers)).
-Once the function is created simply input your function for speciation in the modifier and run the `dads` function with that modifier:
-
-```{r, eval = FALSE}
-## Speciating or going extinct randomly
-## (regardless of the extinction parameter)
-random.extinct  <- function() {
-    return(sample(c(TRUE, FALSE), 1))
-}
-
-## Creating the modifiers object
-random_extinction <- make.modifiers(speciation = random.extinct)
-
-## Generating a new tree with this modifier
-set.seed(0)
-modified_tree <- dads(bd.params = bd_params,
-                      stop.rule = stop_rule,
-                      modifiers = random_extinction)
-
-par(mfrow = c(1,2))
-plot(default_tree,  main = "Default modifier")
-plot(modified_tree, main = "Random extinction\nmodifier")
-```
-
-Note how every lineage end up going extinct!
-And again, we can make some more advanced modifiers: for example, one where a tip always goes extinct if their ancestor has a negative trait value (here we will also introduce the utility function `parent.trait` that automatically selects the trait values of the parent of the current lineage.
-
-```{r}
-## A modifier for removing tips with negative values
-bias.trait <- function(trait.values, lineage) {
-    if(parent.traits(trait.values, lineage) < 0) {
-        ## Go extinct!
-        return(FALSE)
-    } else {
-        ## Speciate
-        return(TRUE)
-    }
-}
-
-## Creating the modifier
-biased_trait <- make.modifiers(speciation = bias.trait)
-
-## Simulate a tree and trait with the modifier
-set.seed(1)
-biased_trait_dads <- dads(bd.params = bd_params,
-                          stop.rule = stop_rule,
-                          traits    = BM_trait,
-                          modifiers = biased_trait)
-
-## Plotting the differences in trees and traits
-par(mfrow = c(1, 2))
-plot(default_dads, main = "Default trait and tree")
-plot(biased_trait_dads, main = "Biased trait and tree")
-```
-
-## Summary of the inputs and outputs for the `branch.length`, `selection` and `speciation` modifiers {#summarymodifiers}
-
-modifier name   | accepted input (arguments) | required output (class) |
-----------------|----------------------------|-------------------------|
-`branch.length` | `bd.params`, `lineage`, `trait.values` | `"numeric"` |
-`selection`     | `bd.params`, `lineage`, `trait.values` | `"integer"` |
-`speciation`    | `bd.params`, `lineage`, `trait.values` | `"logical"` |
-
-## The condition and modify functions (`condition` and `modify`)
-
-In the examples above, we have seen how to specify modifications to the birth death process (via `branch.length`, `selection` and `speciation`), however, one might note that these modifications are not dynamic.
-In other words, throughout the process, the modifications remain constant (even if they are conditional).
-It is however possible to code the `"modifiers"` so that they can be affected by `"events"` objects (see next chapter).
-
-To do so, you can formally declare conditions (`condition`) and modifications (`modify`) as internal functions that can then be modified my an `"events"` object.
-`condition` and `modify` are hard coded in the `branch.length` function that they concern, i.e. they are variables (functions) within the function.
-
-## Combining and editing modifiers (`add`)
-
-## Testing modifiers (`test`)
-
-## Demo runnable
-
-```{r, fig.height = 16, fig.width = 8, eval = FALSE}
-
-```
-
-<!--chapter:end:03_make.modifiers.Rmd-->
-
----
-title: "`make.events`"
+title: "Thorough example"
 author: "Thomas Guillerme (guillert@tcd.ie)"
@@
 ```
+# Thorough example
 
-# Making events with `make.events()` {#makeevents}
+Here we are going to look at an example on how to simulate complex evolutionary histories with `dads`.
 
-```{r, echo = FALSE}
-set.seed(123)
+Let say we want to simulate the following story:
+
+ 1. A lineage evolves in a normal birth-death fashion (with extinction) from a single origin and with two traits called:
+    - "Distance": the physical distance between species in an arbitrary 1D space (evolving in a Brownian motion).
+    - "Variable": a 2D uncorrelated Ornstein-Uhlenbeck trait that corresponds to some traits the lineages have.
+ During the whole process, lineages that are more further away from the other species in terms of distance have a higher change of speciating (proportional to that distance).
+
+Three events then happen during the evolutionary history of this lineage:
+
+ 2. "Correlation event": after reaching time 2 an event happens making the "Variables" trait dimensions become correlated.
+ 3. "Selective extinction event": after reaching a certain point in time, lineages with their "Variables" values that are negative go extinct
+ 4.
+"Founding event": after reaching a value for the "Distance" trait, a founding event happens with a new birth-death process with no extinction and where the "Variable" is now a 2D correlated Brownian Motion.
+ 5. "Mass extinction event": after reaching a certain number of taxa in the "Founding event", 80% of taxa go extinct randomly.
+
+We are going to run each step one by one in a cumulative manner to see how they pile up:
+
+## Step 1:
+
+### The birth death parameters
+
+First let's set the overall birth death parameters (a speciation of 1.2 and an extinction of 0.3):
+
+```{r}
+## The birth death parameters
+bd_params <- list(speciation = 1.5, extinction = 0.3)
+## The stopping rule
+stop_rule <- list(max.time = 4)
 ```
 
-## Conditions
+And let's quickly test if the simulation works by running a birth death tree without traits and anything:
 
-## Modifications
+```{r}
+set.seed(0) # 123 also works
+## Running a birth death tree 
+results <- dads(bd.params = bd_params,
+                stop.rule = stop_rule)
 
-## Some events examples
+## And displaying the results
+plot(results)
+axisPhylo()
+```
+
+### The traits
+
+Then let's create the traits:
+
+```{r}
+## The trait object
+two_traits <- make.traits(process = c(BM.process, OU.process), n = c(1, 2),
+                          trait.names = c("Distance", "Variables"))
+```
+
+```{r}
+## Visualise the 1D trait (BM.process)
+plot(two_traits, trait = 1)
+```
+
+```{r, webgl=TRUE}
+## Visualise the 2D trait (uncorrelated OU.process)
+plot(two_traits, trait = 2, use.3D = TRUE)
+```
+
+Note that specifying the correlation matrix is not necessary here _per se_ (since no correlation is the default) but we will modify it through an event later on so it's good practice to have it properly stated before hand (to avoid any confusion).
+
+We can then run the tree with the traits:
+
+```{r}
+set.seed(0)
+## Running a birth death tree 
+results <- dads(bd.params  = bd_params,
+                stop.rule  = stop_rule,
+                traits     = two_traits)
+
+## And displaying the results
+par(mfrow = c(1,2))
+plot(results, trait = 1, main = "Trait 1: Distance")
+plot(results, trait = c(2,3), main = "Variables correlation")
+```
+
+```{r, webgl=TRUE}
+## Visualising the tree in 3D
+plot(results, trait = c(2,3), use.3D = TRUE)
+```
+
+### The modifiers
+
+```{r}
+## Speciation event is more likely if lineage's ancestor is further away from the mean trait value
+distance.modify <- function(x, trait.values, lineage) return(x + x * abs(parent.traits(trait.values, lineage)[1] - mean(trait.values[,1])))
+
+## Make a distance modifier
+distance.speciation <- make.modifiers(speciation = speciation,
+                                      modify = distance.modify)
+```
+
+```{r}
+set.seed(13)
+## Running a birth death tree 
+results <- dads(bd.params  = bd_params,
+                stop.rule  = stop_rule,
+                traits     = two_traits,
+                modifiers  = distance.speciation)
+
+## And displaying the results
+par(mfrow = c(1,2))
+plot(results, trait = 1, main = "Trait 1: Distance")
+plot(results, trait = c(2,3), main = "Variables correlation")
+```
+
+```{r, webgl=TRUE}
+## Visualising the tree in 3D
+plot(results, trait = c(2,3), use.3D = TRUE)
+```
+
+> BUGGED FROM HERE!
+
+## Step 2:
+ 2. "Correlation event": after reaching time 2 an event happens making the "Variable" trait dimensions become correlated.
+
+```{r}
+## The correlation change event
+update.correlation <- update.traits(process.args =list(Sigma = matrix(c(10,3,3,2),2,2)))
+
+events_list <- make.events(event.name   = "Correlation",
+                           target       = "traits",
+                           condition    = time.condition(2),
+                           modification = update.traits(
+                    process.args =list(Sigma = matrix(c(10,3,3,2),2,2)),
+                    trait.names  = "Variables"))
+```
+
+```{r}
+results <- NULL
+set.seed(1)
+## Running a birth death tree 
+results <- dads(bd.params  = bd_params,
+                stop.rule  = stop_rule,
+                traits     = two_traits,
+                modifiers  = distance.speciation,
+                events     = events_list)
+
+## And displaying the results
+par(mfrow = c(1,2))
+plot(results, trait = 1, main = "Trait 1: Distance")
+plot(results, trait = c(2,3), main = "Variables correlation")
+```
+
+```{r, webgl=TRUE}
+## Visualising the tree in 3D
+plot(results, trait = c(2,3), use.3D = TRUE)
+```
+
+## Step 3:
+ 3. "Selective extinction event": after reaching a certain point in time, lineages with their "Variables" values that are negative go extinct
+
+```{r, eval = FALSE}
+## Adding the selective event
+events_list <- make.events(event.name   = "Selection",
+                           add          = events_list,
+                           target       = "taxa",
+                           condition    = time.condition(4),
+                           modification = trait.extinction(x = 0, condition = `<`, trait = 2))
+```
+Problem with `modification` argument here
+
+```{r, eval = FALSE}
+## Running a birth death tree 
+results <- dads(bd.params  = bd_params,
+                stop.rule  = stop_rule,
+                traits     = two_traits,
+                modifiers  = distance.speciation,
+                events     = events_list)
+
+## And displaying the results
+par(mfrow = c(1,2))
+plot(results, trait = 1, main = "Trait 1: Distance")
+plot(results, trait = c(2,3), main = "Variables correlation")
+```
+
+## Step 4 and 5:
+ 4.
+"Founding event": after reaching a value for the "Distance" trait, a founding event happens with a new birth-death process with no extinction and where the "Variable" is now a 2D correlated Brownian Motion.
+ 5. "Mass extinction event": after reaching a certain number of taxa in the "Founding event", 80% of taxa go extinct randomly.
+
+```{r, eval = FALSE}
+## The new trait for the founding event
+new_traits <- make.traits(process = c(BM.process, BM.process), n = c(1, 2),
+                          trait.names = c("Distance", "Variables"))
+
+## The mass extinction event after reaching 50 taxa
+mass_extinction <- make.events(target       = "taxa",
+                               condition    = taxa.condition(50),
+                               modification = random.extinction(0.8))
+```
+
+```{r, eval = FALSE}
+## The selective event
+events_list <- make.events(event.name   = "Founding", 
+                           add          = events_list,
+                           target       = "founding",
+                           condition    = trait.condition(3),
+                           modification = founding.event(traits    = new_traits,
+                                                         events    = mass_extinction,
+                                                         bd.params = list(
+                                                            speciation = 2,
+                                                            extinction = 0)),
+                           additional.args = list(prefix = "found_"))
+```
+
+```{r, eval = FALSE}
+## Running a birth death tree 
+results <- dads(bd.params  = bd_params,
+                stop.rule  = stop_rule,
+                traits     = two_traits,
+                modifiers  = distance.speciation,
+                events     = events_list,
+                null.error = 100)
+
+## And displaying the results
+par(mfrow = c(1,2))
+plot(results, trait = 1, main = "Trait 1: Distance")
+plot(results, trait = c(2,3), main = "Variables correlation")
+```
+
+# Examples gallery
 
@@
 ```
-
-<!--chapter:end:04_make.events.Rmd-->
-
----
-title: "Thorough example"
-author: "Thomas Guillerme (guillert@tcd.ie)"
-date: "`r Sys.Date()`"
-bibliography: [../References.bib, ../packages.bib]
-output:
-  html_document: default
-  pdf_document: default
----
-
-```{r, echo = FALSE}
-library(rgl)
-knitr::knit_hooks$set(webgl = hook_webgl)
-```
-# Thorough example
-
-Here we are going to look at an example on how to simulate complex evolutionary histories with `dads`.
-
-Let say we want to simulate the following story:
-
- 1. A lineage evolves in a normal birth-death fashion (with extinction) from a single origin and with two traits called:
-    - "Distance": the physical distance between species in an arbitrary 1D space (evolving in a Brownian motion).
-    - "Variable": a 2D uncorrelated Ornstein-Uhlenbeck trait that corresponds to some traits the lineages have.
- During the whole process, lineages that are more further away from the other species in terms of distance have a higher change of speciating (proportional to that distance).
-
-Three events then happen during the evolutionary history of this lineage:
-
- 2. "Correlation event": after reaching time 2 an event happens making the "Variables" trait dimensions become correlated.
- 3. "Selective extinction event": after reaching a certain point in time, lineages with their "Variables" values that are negative go extinct
- 4.
-"Founding event": after reaching a value for the "Distance" trait, a founding event happens with a new birth-death process with no extinction and where the "Variable" is now a 2D correlated Brownian Motion.
- 5. "Mass extinction event": after reaching a certain number of taxa in the "Founding event", 80% of taxa go extinct randomly.
-
-We are going to run each step one by one in a cumulative manner to see how they pile up:
-
-## Step 1:
-
-### The birth death parameters
-
-First let's set the overall birth death parameters (a speciation of 1.2 and an extinction of 0.3):
-
-```{r}
-## The birth death parameters
-bd_params <- list(speciation = 1.5, extinction = 0.3)
-## The stopping rule
-stop_rule <- list(max.time = 4)
-```
-
-And let's quickly test if the simulation works by running a birth death tree without traits and anything:
-
-```{r}
-set.seed(0) # 123 also works
-## Running a birth death tree 
-results <- dads(bd.params = bd_params,
-                stop.rule = stop_rule)
-
-## And displaying the results
-plot(results)
-axisPhylo()
-```
-
-### The traits
-
-Then let's create the traits:
-
-```{r}
-## The trait object
-two_traits <- make.traits(process = c(BM.process, OU.process), n = c(1, 2),
-                          trait.names = c("Distance", "Variables"))
-```
-
-```{r}
-## Visualise the 1D trait (BM.process)
-plot(two_traits, trait = 1)
-```
-
-```{r, webgl=TRUE}
-## Visualise the 2D trait (uncorrelated OU.process)
-plot(two_traits, trait = 2, use.3D = TRUE)
-```
-
-Note that specifying the correlation matrix is not necessary here _per se_ (since no correlation is the default) but we will modify it through an event later on so it's good practice to have it properly stated before hand (to avoid any confusion).
-
-We can then run the tree with the traits:
-
-```{r}
-set.seed(0)
-## Running a birth death tree 
-results <- dads(bd.params  = bd_params,
-                stop.rule  = stop_rule,
-                traits     = two_traits)
-
-## And displaying the results
-par(mfrow = c(1,2))
-plot(results, trait = 1, main = "Trait 1: Distance")
-plot(results, trait = c(2,3), main = "Variables correlation")
-```
-
-```{r, webgl=TRUE}
-## Visualising the tree in 3D
-plot(results, trait = c(2,3), use.3D = TRUE)
-```
-
-### The modifiers
-
-```{r}
-## Speciation event is more likely if lineage's ancestor is further away from the mean trait value
-distance.modify <- function(x, trait.values, lineage) return(x + x * abs(parent.traits(trait.values, lineage)[1] - mean(trait.values[,1])))
-
-## Make a distance modifier
-distance.speciation <- make.modifiers(speciation = speciation,
-                                      modify = distance.modify)
-```
-
-```{r}
-set.seed(13)
-## Running a birth death tree 
-results <- dads(bd.params  = bd_params,
-                stop.rule  = stop_rule,
-                traits     = two_traits,
-                modifiers  = distance.speciation)
-
-## And displaying the results
-par(mfrow = c(1,2))
-plot(results, trait = 1, main = "Trait 1: Distance")
-plot(results, trait = c(2,3), main = "Variables correlation")
-```
-
-```{r, webgl=TRUE}
-## Visualising the tree in 3D
-plot(results, trait = c(2,3), use.3D = TRUE)
-```
-
-> BUGGED FROM HERE!
-
-## Step 2:
- 2. "Correlation event": after reaching time 2 an event happens making the "Variable" trait dimensions become correlated.
-
-```{r}
-## The correlation change event
-update.correlation <- update.traits(process.args =list(Sigma = matrix(c(10,3,3,2),2,2)))
-
-events_list <- make.events(event.name   = "Correlation",
-                           target       = "traits",
-                           condition    = time.condition(2),
-                           modification = update.traits(
-                    process.args =list(Sigma = matrix(c(10,3,3,2),2,2)),
-                    trait.names  = "Variables"))
-```
-
-```{r}
-results <- NULL
-set.seed(1)
-## Running a birth death tree 
-results <- dads(bd.params  = bd_params,
-                stop.rule  = stop_rule,
-                traits     = two_traits,
-                modifiers  = distance.speciation,
-                events     = events_list)
-
-## And displaying the results
-par(mfrow = c(1,2))
-plot(results, trait = 1, main = "Trait 1: Distance")
-plot(results, trait = c(2,3), main = "Variables correlation")
-```
-
-```{r, webgl=TRUE}
-## Visualising the tree in 3D
-plot(results, trait = c(2,3), use.3D = TRUE)
-```
-
-## Step 3:
- 3. "Selective extinction event": after reaching a certain point in time, lineages with their "Variables" values that are negative go extinct
-
-```{r, eval = FALSE}
-## Adding the selective event
-events_list <- make.events(event.name   = "Selection",
-                           add          = events_list,
-                           target       = "taxa",
-                           condition    = time.condition(4),
-                           modification = trait.extinction(x = 0, condition = `<`, trait = 2))
-```
-Problem with `modification` argument here
-
-```{r, eval = FALSE}
-## Running a birth death tree 
-results <- dads(bd.params  = bd_params,
-                stop.rule  = stop_rule,
-                traits     = two_traits,
-                modifiers  = distance.speciation,
-                events     = events_list)
-
-## And displaying the results
-par(mfrow = c(1,2))
-plot(results, trait = 1, main = "Trait 1: Distance")
-plot(results, trait = c(2,3), main = "Variables correlation")
-```
-
-## Step 4 and 5:
- 4.
-"Founding event": after reaching a value for the "Distance" trait, a founding event happens with a new birth-death process with no extinction and where the "Variable" is now a 2D correlated Brownian Motion.
- 5. "Mass extinction event": after reaching a certain number of taxa in the "Founding event", 80% of taxa go extinct randomly.
-
-```{r, eval = FALSE}
-## The new trait for the founding event
-new_traits <- make.traits(process = c(BM.process, BM.process), n = c(1, 2),
-                          trait.names = c("Distance", "Variables"))
-
-## The mass extinction event after reaching 50 taxa
-mass_extinction <- make.events(target       = "taxa",
-                               condition    = taxa.condition(50),
-                               modification = random.extinction(0.8))
-```
-
-```{r, eval = FALSE}
-## The selective event
-events_list <- make.events(event.name   = "Founding", 
-                           add          = events_list,
-                           target       = "founding",
-                           condition    = trait.condition(3),
-                           modification = founding.event(traits    = new_traits,
-                                                         events    = mass_extinction,
-                                                         bd.params = list(
-                                                            speciation = 2,
-                                                            extinction = 0)),
-                           additional.args = list(prefix = "found_"))
-```
-
-```{r, eval = FALSE}
-## Running a birth death tree 
-results <- dads(bd.params  = bd_params,
-                stop.rule  = stop_rule,
-                traits     = two_traits,
-                modifiers  = distance.speciation,
-                events     = events_list,
-                null.error = 100)
-
-## And displaying the results
-par(mfrow = c(1,2))
-plot(results, trait = 1, main = "Trait 1: Distance")
-plot(results, trait = c(2,3), main = "Variables correlation")
-```
-
-<!--chapter:end:05_thorough_example.Rmd-->

### 2023-02-24 inst/gitbook/06_plotting_dads_objects.Rmd
Update 06_plotting_dads_objects.Rmd

--- 2023-01-19 inst/gitbook/05_plotting_dads_objects.Rmd
@@
 
-`"dads"` objects can be directly plotted in dads using many different options.
+`"dads"` objects can be directly plotted in `dads` using the S3 `plot.dads` function (or just `plot(x)` if `x` is of class `"dads"`).
+
+# Plotting traits
+
+`"dads"` `"traits"` objects are covered in the [traits section](#maketraits).
+You can use `plot.dads` to plot them by choosing which specific trait to plot using the `trait` argument (default is 1):
+
+```{r}
+## Making a list of three traits
+list_of_traits <- make.traits(process = c(no.process, BM.process, OU.process), trait.names = c("No process (normal)", "Brownian motion", "Ornstein-Uhlenbeck"))
+## Plotting each trait separately
+par(mfrow = c(2, 2))
+plot(list_of_traits, trait = 1)
+## Using different colours options
+plot(list_of_traits, trait = 2, col = c("red", "purple", "pink"))
+## Not using the default plot name
+plot(list_of_traits, trait = 3, main = "OU process")
+```
+
+<!-- TODO: add trait labels and handling colours! -->
+
+You can also control the number of replicates in the simulation by using the `simulations` option (the default is 50).
+Bigger numbers leads to more time but smoother looking plots while smaller ones are more stochastic:
+
+```{r}
+par(mfrow = c(1,2))
+plot(list_of_traits, trait = 2, simulations = 10, main = "10 BM simulations")
+plot(list_of_traits, trait = 2, simulations = 1000, main = "1k BM simulations")
+```
+
+# Plotting dads results
+
+If `dads` is used to plot only a tree (and outputs a `"phylo"` object), you can use the function `plot.phylo` from the `ape` package to plot your tree.
+You'll find all the tree plotting option in the `?plot.phylo` manual page.
+
+```{r}
+## A simple pure birth tree
+my_tree <- dads(stop.rule = list(max.taxa = 20))
+plot(my_tree, main = "Plotting a \"phylo\" object")
+```
+
+However, if you also simulated a trait along with the tree you can use the `plot.dads` function to plot both the tree and the trait:
+
+```{r}
+## A simple pure birth tree with a BM process
+my_tree <- dads(stop.rule = list(max.taxa = 20), traits = make.traits())
+## Playing with the default options
+plot(my_tree, main = "A tree and traits")
+```
+
+By default, elements are coloured as follows: nodes and tips are points and coloured in blue if they are tips (light blue if they are fossils) and nodes are in orange.
+Branches linking them are grey lines.
+You can of course change this colour palette to something of you preference by calling the normal arguments that can be passed to `points` or `lines`.
+For example `pch` for the point type or `lty` for the line type.
+If there is an ambiguity between the argument names (for example, `cex` for both lines and points), you can specify it by prefixing the argument by the function name.
+For example `cex = 2` will make both the points and lines twice bigger but `cex.points = 2` will make _only_ the points twice bigger.
+
+```{r}
+## Playing with some more options
+plot(my_tree, main = "A tree and traits", points.cex = 2, pch = 19, col = c("red", "purple", "pink"), lwd = 1/3, lty = 3)
+```
+
 <!-- 

### 2023-02-27 inst/gitbook/01_getting_started.Rmd
Update 01_getting_started.Rmd
Compiles OK

@@
 date: "`r Sys.Date()`"
-bibliography: [../References.bib, ../packages.bib]
+bibliography: [../references.bib]
 output:
@@
 
-# Getting started
+# Getting started {#gettingstarted}
 
@@
 However, we think it's good practice to attribute both parameters specifically to avoid any confusion.
+You can find more information about setting up more complex birth death parameters [in this section](#makebdparams).
 

### 2023-02-27 inst/gitbook/02_make.traits.Rmd
Compiles OK

@@
 ```{r, echo = FALSE}
+library(dads)
 set.seed(123)

### 2023-02-27 inst/gitbook/05_other_functionalities.Rmd
Update 05_other_functionalities.Rmd
Compiles OK

@@
 ```{r, echo = FALSE}
+library(dads)
 library(rgl)
@@
 make.bd.params(speciation = rnorm, absolute = TRUE)
-## This is effectively making the speciation distribution equal to
-make.bd.params(speciation = function(x) abs(rnorm(x)))
 ```
@@
 
-## `drop.things`
+## `drop.things` {#dropthings}
 

### 2023-02-27 R/plot.dads.R
Updated plot.simulations

@@ -10,6 +10,8 @@
 #' @param tips.nodes optional, a colour to circle tips and nodes (only used if \code{use.3D = FALSE}). By default \code{tips.nodes = NULL}.
 #' @param use.3D logical, whether to use a 3D plot or not (default is \code{FALSE}; see details).
 #' @param simulations if the input is a \code{dads} \code{traits} or \code{bd.params} object, how many replicates to run (default is \code{50}).
+#' @param cent.tend if the input is a \code{dads} \code{traits}, which central tendency to plot (default is \code{mean}).
+#' @param quantiles if the input is a \code{dads} \code{traits}, which quantiles to plot (default are \code{c(95, 50))}).
 #' 
 #' @details
 #' The \code{col} option can be either:
@@ -49,6 +51,7 @@
 #' 
 #' ## Plotting the second trait and the tree (default)
 #' ## The colours are purple for nodes and blue for tips
+#' ## with a black circle for highlighting the tips
 #' plot(my_data, trait = 2, col = c("purple", "blue"),
 #'      edges = "pink", tips.nodes = "black")
 #' 
@@ -70,7 +73,7 @@
 #' @author Thomas Guillerme
 #' @export
 
-plot.dads <- function(x, col, ..., trait = 1, edges = "grey", tips.nodes = NULL, use.3D = FALSE, simulations = 50) {
+plot.dads <- function(x, col, ..., trait = 1, edges = "grey", tips.nodes = NULL, use.3D = FALSE, simulations = 50, cent.tend = mean, quantiles = c(95, 50)) {
 
     match_call <- match.call()
 
@@ -101,11 +104,19 @@ plot.dads <- function(x, col, ..., trait = 1, edges = "grey", tips.nodes = NULL,
                 col <- "default"
             }
 
+            ## Set the trait name
+            dots <- list(...)
+            if(!is.null(dots$main)) {
+                trait_name <- main
+            } else {
+                trait_name <- names(data$main)[trait]
+            }
+
             ## Plotting the results
             plot.simulation(data = replicate(simulations, 
                                              sim.motion(one_trait, steps = 100),
                                              simplify = FALSE),
-                            col = col, use.3D = use.3D, trait = trait_ids, ...)
+                            col = col, use.3D = use.3D, trait = trait_ids, trait.name = trait_name, quantiles = quantiles, cent.tend = cent.tend, ...)
         }
 
         if(second_class == "bd.params") {
[…]

### 2023-02-27 inst/gitbook/06_plotting_dads_objects.Rmd
Update 06_plotting_dads_objects.Rmd

@@
 
-```{r}
+```{r, fig.height=8, fig.width=8}
 ## Making a list of three traits
@@
 ## Not using the default plot name
-plot(list_of_traits, trait = 3, main = "OU process")
+plot(list_of_traits, trait = 3, main = "OU process",
+     cent.tend = sd, quantiles = c(10, 30))
 ```
-
-<!-- TODO: add trait labels and handling colours! -->
 
@@
 
-```{r}
-par(mfrow = c(1,2))
+```{r, fig.width=6, fig.height=12}
+par(mfrow = c(2,1))
 plot(list_of_traits, trait = 2, simulations = 10, main = "10 BM simulations")
@@
 ## Playing with some more options
-plot(my_tree, main = "A tree and traits", points.cex = 2, pch = 19, col = c("red", "purple", "pink"), lwd = 1/3, lty = 3)
+plot(my_tree, main = "A tree and traits", col = "default", pch = 21, lwd = 2, lty = 3)
 ```

### 2023-03-15 inst/gitbook/06_plotting_dads_objects.Rmd
Update 06_plotting_dads_objects.Rmd

@@
 For example `pch` for the point type or `lty` for the line type.
-If there is an ambiguity between the argument names (for example, `cex` for both lines and points), you can specify it by prefixing the argument by the function name.
-For example `cex = 2` will make both the points and lines twice bigger but `cex.points = 2` will make _only_ the points twice bigger.
 
@@
 ## Playing with some more options
-plot(my_tree, main = "A tree and traits", col = "default", pch = 21, lwd = 2, lty = 3)
+plot(my_tree, main = "A tree and traits", col = c(tips = "pink", nodes = "purple"), pch = 21, lwd = 2, lty = 3, edges = "yellow")
+```
+
+> Note that you can modify colours using the `col` argument by providing either a clear indication of what you want to colour (e.g. `c(tip = "blue", `) @@@
+
+And you can add a default legend by using `legend = TRUE` (if you don't want to add the default legend you can add it after your plot using `legend(...)`).
+
+```{r}
+## Adding the default legend
+plot(my_tree, legend = TRUE)
+```
+
+```{r, eval = FALSE}
+plot(my_data, trait = 2, col = c("purple", "blue"))
+
+## Specifying a 3D trait process
+my_3D_trait <- make.traits(n = 3)
+## Simulating a birth death tree with that 3D trait
+my_data <- dads(bd.params = list(extinction = 0.2),
+                stop.rule = list(max.living = 50),
+                traits    = my_3D_trait)
+
+## Plotting the 
+plot(my_data, trait = c(1, 3), col =terrain.colors, tips.nodes = "black", legend = TRUE)
 ```

### 2023-03-27 inst/gitbook/02_make.traits.Rmd
What is a trait?

@@
 This object is generated using the `make.traits`.
+
+> **What is a trait in biology?** This is one of these epistemological question like what _is_ a species that is not covered by all means in this package.
+For the purpose of _this manual_, **biological trait** can be any coherent process of any number of dimensions which can or cannot be independent.
+Basically anything can be a trait as well as it's defined by the user (e.g. one trait can be a 20 dimensional Brownian process that's correlated to a 1D OU process - which is another trait).
+For the purpose of the syntax in _this package_ though, a **trait in `dads`** is a column in the output trait matrix.
+For example, one might be interested in a biological trait that is "beak shape" which is defined as one complex biological trait that is the Procrustes superimposition of 4 3D landmarks and 300 3D semi-landmarks.
+This single "beak shape" **biological trait** is then composed of 912 **traits in `dads`** (i.e. a matrix of $$(4+300)\times3$$ columns).
 

### 2023-03-27 inst/gitbook/06_plotting_dads_objects.Rmd
Updated plots

@@
 For example `pch` for the point type or `lty` for the line type.
+Furthermore, time is plotted conventionally from left to right (left is towards the past, right is towards the present) but you can change that by specifying `xlim`.
 
@@
 ## Playing with some more options
-plot(my_tree, main = "A tree and traits", col = c(tips = "pink", nodes = "purple"), pch = 21, lwd = 2, lty = 3, edges = "yellow")
+plot(my_tree, main = "A tree and traits",
+     ## Changing nodes colours and type
+     col = c(tips = "pink", nodes = "purple"), pch = 21,
+     ## Changing edges colour and type
+     lwd = 2, lty = 3, edges = "yellow",
+     ## Changing the x axis orientation and label
+     xlim = c(0, 2.5), xlab = "Time into the past")
 ```
 
-> Note that you can modify colours using the `col` argument by providing either a clear indication of what you want to colour (e.g. `c(tip = "blue", `) @@@
+> Note that you can modify colours using the `col` argument by providing a clear indication of what you want to colour (e.g.
+`col = c(tips = "blue", fossils = "orange")` will apply the colours to the unambiguously named elements) or by providing a function that will scale the colours with time (e.g.
+`col = grDevices::heat.colors` will use the specified function to change the elements colours with time).
 
@@
 ## Plotting the 
-plot(my_data, trait = c(1, 3), col =terrain.colors, tips.nodes = "black", legend = TRUE)
+plot(my_data, trait = c(1, 3), col = terrain.colors, tips.nodes = "black", legend = TRUE)
 ```

### 2023-03-27 R/plot.dads_fun.R

@@ -235,7 +235,7 @@ handle.colours <- function(col, points_tree_IDs, points_ages, data, legend) {
                             col_scheme["livings"] <- col["livings"]
                         } else {
                             ## Wrong names
-                            stop("col is a named vector. It must contains the names \"nodes\" and \"tips\" or \"nodes\", \"fossils\" and \"livings\".", call. = FALSE)
+                            stop("If col is a named vector, it must contains the names \"nodes\" and \"tips\" or \"nodes\", \"fossils\" and \"livings\".", call. = FALSE)
                         }
                     }                
                 }

### 2023-03-27 inst/gitbook/06_plotting_dads_objects.Rmd
Finished plotting section

@@
 date: "`r Sys.Date()`"
-bibliography: [../References.bib, ../packages.bib]
+# bibliography: [../References.bib, ../packages.bib]
 output:
@@
 ```{r, echo = FALSE}
+library(dads)
 library(rgl)
 knitr::knit_hooks$set(webgl = hook_webgl)
+options(rgl.useNULL = TRUE) # Suppress the separate window.
 ```
@@
 
-```{r}
+```{r, fig.height = 6, fig.width = 6}
 ## Playing with some more options
@@
 
-```{r}
+```{r, fig.height = 6, fig.width = 6}
 ## Adding the default legend
@@
 
-```{r, eval = FALSE}
-plot(my_data, trait = 2, col = c("purple", "blue"))
+## 3D version!
 
+Of course, if you're simulating multiple traits, you can always plot different ones.
+
+```{r, fig.height = 6, fig.width = 6}
 ## Specifying a 3D trait process
-my_3D_trait <- make.traits(n = 3)
+my_3D_trait <- make.traits(n = 3, )
 ## Simulating a birth death tree with that 3D trait
@@
                 traits    = my_3D_trait)
-
-## Plotting the 
-plot(my_data, trait = c(1, 3), col = terrain.colors, tips.nodes = "black", legend = TRUE)
 ```
 
-<!-- 
-test_that("plot.dads works for traits", {
+You can toggle which trait to plot using the `trait` option.
+Either by providing a single value to plot that specific trait against time or by providing two traits.
 
-    list_of_traits <- make.traits(process = c(no.process, BM.process, OU.process), trait.names = c("No process (normal)", "Brownian motion", "Ornstein-Uhlenbeck"))
-    ## Right output
-    expect_null(plot.dads(list_of_traits))
-    expect_null(plot.dads(list_of_traits, trait = 2))
-    expect_null(plot.dads(list_of_traits, trait = 3))
-})
+```{r, fig.height = 12, fig.width = 6}
+par(mfrow = c(2,1))
+## Plotting the second trait
+plot(my_data, trait = 2, main = "Trait 2")
+## Plotting the correlation between trait 1 and 2
+plot(my_data, trait = c(1,2), main = "Traits 1 and 2")
+```
 
-test_that("plot.dads works for dads", {
+As mentioned above, the `col` argument can take a function for scaling the elements colours with time.
+This can be useful for adding time a third dimension to these 2D plots:
 
-    ## Simple example    
-    test <- dads(stop.rule = list("max.living" = 10), traits = make.traits())
-    ## Default plot
-    expect_null(plot(test))
-    ## Other options
-    expect_null(plot(test, main = "what?", xlim = c(0, test$tree$root.time), col = c(tips = "pink", nodes = "red", edges = "purple"), ylab = "data"))
+```{r}
+## Plotting the correlation between trait 1 and 2
+## with time as a 3rd dimensions
+plot(my_data, trait = c(1,2), main = "Traits 1 and 2",
+     col = grDevices::heat.colors, legend = TRUE,
+     ## Highlighting the tips in black for visibility
+     tips.nodes = "black")
+```
 
-    ## A more complex trait
-    complex_traits <- make.traits(process = c(BM.process, BM.process), n = c(2,3), process.args = list(list(Sigma = diag(2)), list(Sigma = matrix(1/3, 3, 3))), trait.names = c("bib", "bob"))
-    test <- dads(stop.rule = list("max.living" = 10), traits = complex_traits)
-    expect_null(plot(test, trait = 3))
+But there's more! You can also plot these results using actual 3D plots that you can spin and all that!
+This is done through the `rgl` package and activated in `dads` using the option `use.3D = TRUE`.
 
-    ## A big tree
-    set.seed(12)
-    big_test <- dads(bd.params = list(speciation = 1, extinction = 1/3), stop.rule = list(max.taxa = 1000), traits = make.traits())
-    expect_null(plot(big_test, cex = 0.5))
+If two traits are provided, the 3rd dimensions is going to be time by default:
 
-    list_of_traits <- make.traits(process = c(no.process, BM.process, OU.process), trait.names = c("No process (normal)", "Brownian motion", "Ornstein-Uhlenbeck"))
-    ## Right output
-    expect_null(plot.dads(list_of_traits))
-    expect_null(plot.dads(list_of_traits, trait = 2))
-    expect_null(plot.dads(list_of_traits, trait = 3))
-})
+```{r}
+## Plotting the tree and 2 traits in 3D: woah!
+plot(my_data, trait = c(1, 2), use.3D = TRUE)
+rglwidget()
+```
 
-test_that("example works", {
-    ## Specifying a trait process
-    my_trait <- make.traits()
-    ## Plotting a trait process
-    expect_null(plot(my_trait, main = "A Brownian Motion"))
+If three traits are provide, you can set up a fourth coloured dimension as time (similarly to the 2D plots with the colour gradients).
+This uses the `rgl` functions, namely `lines3d`, `points3d` (if `type = "p"`; default) or `sphere3d` (if `type = "s"`).
+The additional arguments `...` are directly handled and attributed to the corresponding function.
 
-    ## Simulating a tree with ten taxa
-    my_tree <- dads(stop.rule = list(max.taxa = 10))
-    ## Plotting a simple birth death tree (using ape::plot.phylo)
-    expect_is(plot(my_tree, main = "A pure birth tree"), "list")
-
-    ## Simulating a tree with traits
-    my_data <- dads(stop.rule = list(max.taxa = 10),
-                    traits    = my_trait)
-    ## Plotting the tree and traits
-    expect_null(plot(my_data))
-
-    ## Specifying a 3D trait process
-    my_3D_trait <- make.traits(n = 3)
-    ## Simulating a birth death tree with that trait
-    my_data <- dads(bd.params = list(extinction = 0.2),
-                    stop.rule = list(max.living = 50),
-                    traits    = my_3D_trait)
-
-    ## Plotting the second trait and the tree (default)
-    ## The colours are purple for nodes and blue for tips
-    expect_null(plot(my_data, trait = 2, col = c("purple", "blue"),
-         edges = "pink", tips.nodes = "black"))
-
-    ## Plotting the first and third trait correlation
-    ## The colours are a heat map based on the elements age
-    expect_null(plot(my_data, trait = c(1, 3), col = terrain.colors,
-         edges = "grey", tips.nodes = "black"))
-
-    ## Plotting the first and third trait correlation in 3D
-    expect_null(plot(my_data, trait = c(1,3), col = rainbow,
-         edges = "grey", tips.nodes = "black", use.3D = TRUE))
-
-    ## Plotting all traits in 3D (without branch lengths)
-    expect_null(plot(my_data, trait = c(1:3), col = heat.colors,
-         edges = "grey", tips.nodes = "black", use.3D = TRUE, type = "s", radius = 0.1))
-}) -->
+```{r}
+## Plotting the tree and 3 traits in 3D
+plot(my_data, trait = c(1, 2, 3), use.3D = TRUE,
+     col = grDevices::heat.colors, type = "s", radius = 0.1)
+rglwidget()
+```

### 2023-03-27 inst/gitbook/05_other_functionalities.Rmd
added manual for internal utilities

@@
 ```
+
+## `"dads"` internal utilities
+
+The package also provides utilities for internal functions, namely for designing `modifiers` or `events` more easily.
+These functions don't do anything useful on their own but are optimised to be used internally in `dads`.
+For all these functions, you can look at the internal manual for an example (i.e. using `?<function_name>`)
+
+So far the package has the following internals:
+
+Function  | What it does | Where can it be used
+----------|--------------|---------------------
+`parent.traits` | selects the trait values of the current lineage's parents (i.e. direct ancestor) | in `make.modifiers`
+`taxa.condition` | provides a trigger for an `event` dependent on the number of taxa | in `make.events`
+`time.condition` | provides a trigger for an `event` dependent on time | in `make.events`
+`trait.condition` | provides a trigger for an `event` dependent on trait values | in `make.events`

### 2023-03-27 R/events.conditions.R

@@ -93,8 +93,8 @@
 ## The list of conditions
 condition <- function(x, condition) {
     cat("List of inbuilt condition functions in dads:\n")
-    cat("   ?time.condition\n")
     cat("   ?taxa.condition\n")
+    cat("   ?time.condition\n")
     cat("   ?trait.condition\n")
     return(invisible())
 } 

### 2023-03-27 R/utilities.R

@@ -12,8 +12,21 @@
 #' This function is designed to be used internally in \code{dads} to help \code{modifiers}, \code{traits} or \code{events} objects to access the parent traits of the lineages simulated through the internal birth death algorithm. 
 #' 
 #' @examples
+#' ## Speciation event is more likely if lineage's ancestor is further away from the mean trait value
+#' distance.modify <- function(x, trait.values, lineage) {
+#'      ## Distance to the parent's trait
+#'      parent_trait_val <- parent.traits(trait.values, lineage)[1]
+#'      mean_trait_val <- mean(trait.values[, 1])
+#'      distance <- abs(parent_trait_val - mean_trait_val)
+#'      ## Scales x with the distance
+#'      return(x + x * distance)
+#' }
+#' 
+#' ## Make a distance modifier (speciation more likely with distance)
+#' distance.speciation <- make.modifiers(speciation = speciation,
+#'                                       modify = distance.modify)
 #'
-#' @seealso \code{\link{dads}}
+#' @seealso \code{\link{dads}} \code{\link{make.modifiers}}
 #' 
 #' @author Thomas Guillerme
 #' @export

### 2023-03-27 inst/gitbook/index.Rmd
polished index

@@
 title: "dads manual"
-author: "Thomas Guillerme (guillert@tcd.ie), Alex Slavenko (email) and others"
+author: "Thomas Guillerme (guillert@tcd.ie)"
 date: "`r Sys.Date()`"
@@
 github-repo: TGuillerme/dads
-description: "dads R package vignette"
+description: "dads R package manual"
 ---
@@
 
-![Schematised summary of the `dads` package architecture](dads_structure.pdf)
+![Schematised summary of the `dads` package architecture](dads_structure.png)
 
@@
  3.
-**Simulating traits** \[optional\] (`traits`): this step allows to simulate a trait value for the selected lineage from step 1.
+**Simulating traits** [optional]\(`traits`): this step allows to simulate a trait value for the selected lineage from step 1.
 This is typically not part of a standard birth-death process and is handled via the `traits` option in `dads` (see the [`make.traits`](#maketraits) chapter for more details).

### 2023-03-27 inst/gitbook/07_specific_examples.Rmd
added specific examples

@@
 ---
-title: "Thorough example"
+title: "Specific examples"
 author: "Thomas Guillerme (guillert@tcd.ie)"
 date: "`r Sys.Date()`"
-bibliography: [../References.bib, ../packages.bib]
+# bibliography: [../References.bib, ../packages.bib]
 output:
@@
 ```{r, echo = FALSE}
+library(dads)
 library(rgl)
 knitr::knit_hooks$set(webgl = hook_webgl)
+options(rgl.useNULL = TRUE) # Suppress the separate window.
 ```
-# Thorough example
 
-Here we are going to look at an example on how to simulate complex evolutionary histories with `dads`.
+In this section I will illustrate a series of examples of specific more complex scenarios.
+They can be all run independently and used as a basis for your own specific scenarios.
 
-Let say we want to simulate the following story:
+Here is a table summarising which functionalities are used in which example:
 
- 1. A lineage evolves in a normal birth-death fashion (with extinction) from a single origin and with two traits called:
-    - "Distance": the physical distance between species in an arbitrary 1D space (evolving in a Brownian motion).
-    - "Variable": a 2D uncorrelated Ornstein-Uhlenbeck trait that corresponds to some traits the lineages have.
- During the whole process, lineages that are more further away from the other species in terms of distance have a higher change of speciating (proportional to that distance).
+Functionality showcase | Example
+--------------|---------
+`bd.params` | [reducing speciation rate](#EGred_spec), [generating a subtree with no extinction](#EG_founding_purebirth)
+`make.events` | [mass extinction](#EGrandom_ext), [negative trait extinction](#EGneg_ext), [background extinction](#EGbg_ext), [reducing speciation rate](#EGred_spec), [changing trait process](#EG_change_trait), [changing trait correlation](#EG_change_correlation), [changing modifiers](#EG_change_modif), [change branch length](#EG_modify_brlen), [generating a subtree with no extinction](#EG_founding_purebirth), [generating a subtree with a different process](EG_founding_traits)
+`make.traits` | [negative trait extinction](#EGneg_ext), [changing trait process](#EG_change_trait), [changing trait correlation](#EG_change_correlation), [generating a subtree with a different process](EG_founding_traits)
+`make.modifiers` | [changing modifiers](#EG_change_modif), [change branch length](#EG_modify_brlen)
 
-Three events then happen during the evolutionary history of this lineage:
+And more specifically for the events:
 
- 2. "Correlation event": after reaching time 2 an event happens making the "Variables" trait dimensions become correlated.
- 3. "Selective extinction event": after reaching a certain point in time, lineages with their "Variables" values that are negative go extinct
- 4.
-"Founding event": after reaching a value for the "Distance" trait, a founding event happens with a new birth-death process with no extinction and where the "Variable" is now a 2D correlated Brownian Motion.
- 5. "Mass extinction event": after reaching a certain number of taxa in the "Founding event", 80% of taxa go extinct randomly.
+Functionality showcase | Example
+--------------|----------
+`time.condition` | [mass extinction](#EGrandom_ext), [negative trait extinction](#EGneg_ext), [reducing speciation rate](#EGred_spec), [changing trait process](#EG_change_trait), [changing modifiers](#EG_change_modif)
+`random.extinction` | [mass extinction](#EGrandom_ext)
+`trait.extinction` | [negative trait extinction](#EGneg_ext)
+`taxa.condition` | [background extinction](#EGbg_ext), [change branch length](#EG_modify_brlen)
+`update.bd.params` | [background extinction](#EGbg_ext), [reducing speciation rate](#EGred_spec)
+`update.traits` | [changing trait process](#EG_change_trait), [changing trait correlation](#EG_change_correlation)
+`trait.condition` | [changing trait correlation](#EG_change_correlation), [generating a subtree with a different process](EG_founding_traits)
+`update.modifiers` | [changing modifiers](#EG_change_modif), [change branch length](#EG_modify_brlen)
+`parent.traits` | [changing modifiers](#EG_change_modif)
+`founding.event` | [generating a subtree with no extinction](#EG_founding_purebirth), [generating a subtree with a different process](EG_founding_traits)
 
-We are going to run each step one by one in a cumulative manner to see how they pile up:
+## Random mass extinction after some time {#EGrandom_ext}
 
-## Step 1:
+For this scenario, we want to generate a pure birth tree (no traits and no extinction) where 80% of the species go extinct two thirds into the scenario.
 
-### The birth death parameters
-
-First let's set the overall birth death parameters (a speciation of 1.2 and an extinction of 0.3):
+For that we first need to first set our simulation parameters: we will be running the simulation for 6 time units and with a speciation rate of 1 (and extinction of 0 - default).
 
 ```{r}
-## The birth death parameters
-bd_params <- list(speciation = 1.5, extinction = 0.3)
-## The stopping rule
-stop_rule <- list(max.time = 4)
+## Setting the parameters
+stop_time_6 <- list(max.time = 6)
+speciation_1 <- make.bd.params(speciation = 1)
 ```
 
-And let's quickly test if the simulation works by running a birth death tree without traits and anything:
+We will then create a event that triggers when reaching half of the simulation (using `time.condition(2.5)`).
+This event will target `"taxa"`, i.e. the number of species, and randomly remove 80% of them (using `random.extinction(0.8)`):
 
 ```{r}
-set.seed(0) # 123 also works
-## Running a birth death tree 
-results <- dads(bd.params = bd_params,
-                stop.rule = stop_rule)
-
-## And displaying the results
-plot(results)
-axisPhylo()
-```
-
-### The traits
-
-Then let's create the traits:
-
-```{r}
-## The trait object
-two_traits <- make.traits(process = c(BM.process, OU.process), n = c(1, 2),
-                          trait.names = c("Distance", "Variables"))
-```
-
-```{r}
-## Visualise the 1D trait (BM.process)
-plot(two_traits, trait = 1)
-```
-
-```{r, webgl=TRUE}
-## Visualise the 2D trait (uncorrelated OU.process)
-plot(two_traits, trait = 2, use.3D = TRUE)
-```
-
-Note that specifying the correlation matrix is not necessary here _per se_ (since no correlation is the default) but we will modify it through an event later on so it's good practice to have it properly stated before hand (to avoid any confusion).
-
-We can then run the tree with the traits:
-
-```{r}
-set.seed(0)
-## Running a birth death tree 
-results <- dads(bd.params  = bd_params,
-                stop.rule  = stop_rule,
-                traits     = two_traits)
-
-## And displaying the results
-par(mfrow = c(1,2))
-plot(results, trait = 1, main = "Trait 1: Distance")
-plot(results, trait = c(2,3), main = "Variables correlation")
-```
-
-```{r, webgl=TRUE}
-## Visualising the tree in 3D
-plot(results, trait = c(2,3), use.3D = TRUE)
-```
-
-### The modifiers
-
-```{r}
-## Speciation event is more likely if lineage's ancestor is further away from the mean trait value
-distance.modify <- function(x, trait.values, lineage) return(x + x * abs(parent.traits(trait.values, lineage)[1] - mean(trait.values[,1])))
-
-## Make a distance modifier
-distance.speciation <- make.modifiers(speciation = speciation,
-                                      modify = distance.modify)
-```
-
-```{r}
-set.seed(13)
-## Running a birth death tree 
-results <- dads(bd.params  = bd_params,
-                stop.rule  = stop_rule,
-                traits     = two_traits,
-                modifiers  = distance.speciation)
-
-## And displaying the results
-par(mfrow = c(1,2))
-plot(results, trait = 1, main = "Trait 1: Distance")
-plot(results, trait = c(2,3), main = "Variables correlation")
-```
-
-```{r, webgl=TRUE}
-## Visualising the tree in 3D
-plot(results, trait = c(2,3), use.3D = TRUE)
-```
-
-> BUGGED FROM HERE!
-
-## Step 2:
- 2. "Correlation event": after reaching time 2 an event happens making the "Variable" trait dimensions become correlated.
-
-```{r}
-## The correlation change event
-update.correlation <- update.traits(process.args =list(Sigma = matrix(c(10,3,3,2),2,2)))
-
-events_list <- make.events(event.name   = "Correlation",
-                           target       = "traits",
-                           condition    = time.condition(2),
-                           modification = update.traits(
-                    process.args =list(Sigma = matrix(c(10,3,3,2),2,2)),
-                    trait.names  = "Variables"))
-```
-
-```{r}
-results <- NULL
-set.seed(1)
-## Running a birth death tree 
-results <- dads(bd.params  = bd_params,
-                stop.rule  = stop_rule,
-                traits     = two_traits,
-                modifiers  = distance.speciation,
-                events     = events_list)
-
-## And displaying the results
-par(mfrow = c(1,2))
-plot(results, trait = 1, main = "Trait 1: Distance")
-plot(results, trait = c(2,3), main = "Variables correlation")
-```
-
-```{r, webgl=TRUE}
-## Visualising the tree in 3D
-plot(results, trait = c(2,3), use.3D = TRUE)
-```
-
-## Step 3:
- 3. "Selective extinction event": after reaching a certain point in time, lineages with their "Variables" values that are negative go extinct
-
-```{r, eval = FALSE}
-## Adding the selective event
-events_list <- make.events(event.name   = "Selection",
-                           add          = events_list,
+## 80% mass extinction at time 4
+mass.extinction <- make.events(
+                    condition = time.condition(4),
                     target = "taxa",
-                           condition    = time.condition(4),
-                           modification = trait.extinction(x = 0, condition = `<`, trait = 2))
-```
-Problem with `modification` argument here
-
-```{r, eval = FALSE}
-## Running a birth death tree 
-results <- dads(bd.params  = bd_params,
-                stop.rule  = stop_rule,
-                traits     = two_traits,
-                modifiers  = distance.speciation,
-                events     = events_list)
-
-## And displaying the results
-par(mfrow = c(1,2))
-plot(results, trait = 1, main = "Trait 1: Distance")
-plot(results, trait = c(2,3), main = "Variables correlation")
-```
-
-## Step 4 and 5:
- 4.
-"Founding event": after reaching a value for the "Distance" trait, a founding event happens with a new birth-death process with no extinction and where the "Variable" is now a 2D correlated Brownian Motion.
- 5. "Mass extinction event": after reaching a certain number of taxa in the "Founding event", 80% of taxa go extinct randomly.
-
-```{r, eval = FALSE}
-## The new trait for the founding event
-new_traits <- make.traits(process = c(BM.process, BM.process), n = c(1, 2),
-                          trait.names = c("Distance", "Variables"))
-
-## The mass extinction event after reaching 50 taxa
-mass_extinction <- make.events(target       = "taxa",
-                               condition    = taxa.condition(50),
                     modification = random.extinction(0.8))
@@
 
-```{r, eval = FALSE}
-## The selective event
-events_list <- make.events(event.name   = "Founding", 
-                           add          = events_list,
-                           target       = "founding",
-                           condition    = trait.condition(3),
-                           modification = founding.event(traits    = new_traits,
-                                                         events    = mass_extinction,
-                                                         bd.params = list(
-                                                            speciation = 2,
-                                                            extinction = 0)),
-                           additional.args = list(prefix = "found_"))
-```
-
-```{r, eval = FALSE}
-## Running a birth death tree 
-results <- dads(bd.params  = bd_params,
-                stop.rule  = stop_rule,
-                traits     = two_traits,
-                modifiers  = distance.speciation,
-                events     = events_list,
-                null.error = 100)
-
-## And displaying the results
-par(mfrow = c(1,2))
-plot(results, trait = 1, main = "Trait 1: Distance")
-plot(results, trait = c(2,3), main = "Variables correlation")
-```
-
-# Examples gallery
-
-### Random mass extinction after some time
+Once these parameters are defined, we can run the simulations and plot the results:
 
 ```{r}
-## Simulation parameters
-stop.rule <- list(max.time = 5)
-bd.params <- list(extinction = 0, speciation = 1)
-
-## 80% mass extinction at time 4
-mass_extinction <- make.events(
-                      target = "taxa",
-                      condition = time.condition(4),
-                      modification = random.extinction(0.8))
-
 ## Running the simulations
 set.seed(1)
-results <- dads(bd.params = bd.params,
-                stop.rule = stop.rule,
-                events    = mass_extinction)
+results <- dads(stop.rule = stop_time_6,
+                bd.params = speciation_1,
+                events    = mass.extinction)
 ## Plotting the results
@@
 
-### Extinction based on a trait value after some time
+## Species with negative trait values go extinct after some time {#EGneg_ext}
+
+For this scenario, we want to generate a pure birth tree with a one dimensional Brownian Motion trait for 5 time unit.
+We then want the species with a negative trait value go extinct after 4 time unit.
+
+First we need to set up the simulation parameters:
+ * The stopping rule (5 time units)
+ * The birth-death parameters (speciation rate of 1)
 
@@
 ## Simulation parameters
-stop.rule <- list(max.time = 5)
-bd.params <- list(extinction = 0, speciation = 1)
+stop_time_5 <- list(max.time = 5)
+speciation_1 <- make.bd.params(speciation = 1)
+```
 
+Then set up our trait which is a one dimensional Brownian Motion
+
+```{r}
+## Trait
+simple_bm_trait <- make.traits(n = 1, process = BM.process)
+```
+
+And our extinction event which triggers after reaching time 4 (`time.condition(4)`), targets the `"taxa"` and modifies the extinction for species with traits lower than 0
+
+```{r}
 ## Extinction of any tips with trait < 1 at time 4
-trait_extinction <- make.events(
+trait.extinction <- make.events(
                       target = "taxa",
                       condition = time.condition(4),
-                      modification = trait.extinction(1))
+                      modification = trait.extinction(x = 0,
+                                                      condition = `<`))
+```
 
+Once these parameters are defined, we can run the simulations and plot the results:
+
+```{r}
 ## Running the simulations
 set.seed(7)
-results <- dads(bd.params = bd.params,
-                stop.rule = stop.rule,
-                traits    = make.traits(),
-                events    = trait_extinction)
+results <- dads(stop.rule = stop_time_5,
+                bd.params = speciation_1,
+                traits    = simple_bm_trait,
+                events    = trait.extinction)
 ## Plotting the results
@@
 
-### Adding a background extinction after reaching a number of living taxa
+## Adding a background extinction after reaching a number of living taxa {#EGbg_ext}
+
+For this scenario, we want to generate a pure birth tree until reaching 50 living taxa but with an extinction rate appearing after reaching 30 taxa.
+
+First we need to set up the simulation parameters:
+ * The stopping rule (50 taxa max)
+ * The birth-death parameters (speciation rate of 1)
 
@@
 ## Simulation parameters
-stop.rule <- list(max.living = 50)
-bd.params <- list(extinction = 0, speciation = 1)
+stop_taxa_50 <- list(max.living = 50)
+speciation_1 <- make.bd.params(speciation = 1)
+```
 
+And our change in extinction rate event which triggers after reaching 30 taxa (`taxa.condition(30)`), targets the `"bd.params"` (birth-death parameters) sets the extinction rate to 0.5 (`update.bd.params(extinction = 0.5)`):
+
+```{r}
 ## Adding an extinction parameter after 30 taxa
-background_extinction <- make.events(
+background.extinction <- make.events(
                       condition    = taxa.condition(30),
                       target       = "bd.params",
-                      modification = update.bd.params(1/3, "extinction"))
+                      modification = update.bd.params(extinction = 0.5))
+```
 
+Once these parameters are defined, we can run the simulations and plot the results:
+
+```{r}
 ## Running the simulations
 set.seed(2)
-results <- dads(bd.params = bd.params,
-                stop.rule = stop.rule,
-                events    = background_extinction)
+results <- dads(stop.rule = stop_taxa_50,
+                bd.params = speciation_1,
+                events    = background.extinction)
 ## Plotting the results
-plot(results)
+plot(results, show.tip.label = FALSE)
 ```
 
-### Reducing speciation after some time
+## Reducing speciation rate after some time {#EGred_spec}
+
+For this scenario, we want to generate a brith tree for 6 times units with random speciation rates (i.e.
+drawn from a uniform (0.5;1) distribution) which reduces after time 4 through the simulations to a fixed value of 1/3.
+
+First we need to set up the simulation parameters:
+ * The stopping rule (6 time units)
+ * The birth-death parameters (speciation randomly drawn between 0.5 and 1)
 
@@
 ## Simulation parameters
-stop.rule <- list(max.time = 4)
-bd.params <- list(extinction = 0, speciation = 1)
+stop_time_6 <- list(max.time = 6)
+random_speciation <- make.bd.params(speciation = runif, speciation.args = list(min = 0.5, max = 1))
+```
 
-## Reducing speciation after reaching time 2
-reduced_speciation <- make.events(
-                      condition    = time.condition(2),
+And our extinction event which triggers after reaching time 4 (`time.condition(4)`), targets the `"bd.params"` and modifies the speciation rate to 1/3:
+
+```{r}
+## Reducing speciation after reaching time 4
+reduced.speciation <- make.events(
+                      condition    = time.condition(4),
                       target       = "bd.params",
-                      modification = update.bd.params(1/3, "speciation"))
+                      modification = update.bd.params(speciation = 1/3))
+```
 
+Once these parameters are defined, we can run the simulations and plot the results.
+We can contrast the results with the scenario without an event (but same random seed):
+
+```{r}
+## No event
 set.seed(42)
-no_event <- dads(bd.params = bd.params,
-                 stop.rule = stop.rule)
+no_event <- dads(stop.rule = stop_time_6,
+                 bd.params = random_speciation)
+
+## Reduced speciation event
 set.seed(42)
-reduced_speciation_event <- dads(bd.params = bd.params,
-                                 stop.rule = stop.rule,
-                                 events    = reduced_speciation)
+reduced_speciation_event <- dads(stop.rule = stop_time_6,
+                                 bd.params = random_speciation,
+                                 events    = reduced.speciation)
 
@@
 par(mfrow = c(1, 2))
-plot(no_event, main = "No event")
+plot(no_event, main = "No event", show.tip.label = FALSE)
 axisPhylo()
-plot(reduced_speciation_event, main = "Reduced speciation after time 2")
+plot(reduced_speciation_event, main = "Reduced speciation after time 5", show.tip.label = FALSE)
 axisPhylo()
@@
 
-### Changing trait process after some time
+## Changing the trait process after some time {#EG_change_trait}
+
+For this scenario, we want to generate a pure birth tree with a one dimensional Brownian Motion trait for 5 time unit which then changes to an OU process.
+
+First we need to set up the simulation parameters:
+ * The stopping rule (6 time units)
+ * The birth-death parameters (speciation rate of 1)
 
@@
 ## Simulation parameters
-bd.params <- list(extinction = 0, speciation = 1)
-stop.rule <- list(max.time = 6)
-traits    <- make.traits()
+stop_time_6 <- list(max.time = 6)
+speciation_1 <- make.bd.params(speciation = 1)
+```
 
+Then set up our trait which is a one dimensional Brownian Motion
+
+```{r}
+## Trait
+simple_bm_trait <- make.traits(n = 1, process = BM.process)
+```
+
+And our event which triggers after reaching time 5 (`time.condition(5)`), targets the `"traits"` and modifies the process to `OU.process`.
+
+```{r}
 ## Create an event to change the trait process
-change_process <- make.events(
+change.process.to.OU <- make.events(
                   condition    = time.condition(5),
@@
                   modification = update.traits(process = OU.process))
+```
 
-## Run the simulations
+Once these parameters are defined, we can run the simulations and plot the results.
+We can contrast the results with the scenario without an event (but same random seed):
+
+```{r}
+## Run the simulations without change
 set.seed(1)
-no_change <- dads(bd.params = bd.params,
-                  stop.rule = stop.rule,
-                  traits    = traits)
+no_change <- dads(stop.rule = stop_time_6,
+                  bd.params = speciation_1,
+                  traits    = simple_bm_trait)
+## Run the simulations with change
 set.seed(1)
-process_change <- dads(bd.params = bd.params,
-                       stop.rule = stop.rule,
-                       traits    = traits,
-                       events    = change_process)
+process_change <- dads(stop.rule = stop_time_6,
+                       bd.params = speciation_1,
+                       traits    = simple_bm_trait,
+                       events    = change.process.to.OU)
 ## Plot the results
@@
 
-### Changing trait correlation after reaching a trait value
+## Changing trait correlation after reaching a trait value {#EG_change_correlation}
+
+For this scenario, we want to generate a pure birth tree with a 2 dimensional Brownian Motion trait with a strict correlation between the two dimensions (1:1) that loosen up when a taxa reaches an absolute value of 2.
+
+First we need to set up the simulation parameters:
+ * The stopping rule (100 taxa)
+ * The birth-death parameters (speciation rate of 1)
 
@@
 ## Set the parameters
-stop.rule <- list(max.taxa = 100)
-bd.params <- list(extinction = 0, speciation = 1)
+stop_taxa_100 <- list(max.taxa = 100)
+speciation_1 <- make.bd.params(speciation = 1)
+```
 
-## A correlated trait
-traits <- make.traits(n = 2, process.args = list(Sigma = matrix(1, 2, 2)))
+Then set up our trait which is a 2 dimensional Brownian Motion with a correlation matrix Sigma ()
+
+```{r}
+## A 2D variance covariance matrix
+vcv_matrix <- matrix(1, 2, 2)
+
+## A correlated 2D Brownian Motion
+correlated_2D_BM <- make.traits(n = 2, process = BM.process,
+                      process.args = list(Sigma = vcv_matrix))
+```
+
+And our event which triggers after a taxa gets the trait value 3 (`trait.condition(3, absolute = TRUE)`), targets the `"traits"` and modifies the traits correlation
+
+```{r}
+## New correlation
+new_vcv <- matrix(c(10,3,3,2),2,2)
 
 ## Event changing a trait correlation
-cor_change <- make.events(
+correlation.change <- make.events(
     condition    = trait.condition(3, absolute = TRUE),
     target       = "traits",
-    modification = update.traits(process.args = list(Sigma = matrix(c(10,3,3,2),2,2))))
+    modification = update.traits(process.args = list(Sigma = new_vcv)))
+```
 
+Once these parameters are defined, we can run the simulations and plot the results.
+We can contrast the results with the scenario without an event (but same random seed):
+
+```{r}
 ## Run the simulations
 set.seed(2)
-no_event <- dads(bd.params = bd.params,
-                 stop.rule = stop.rule,
-                 traits    = traits)
+no_event <- dads(stop.rule = stop_taxa_100,
+                 bd.params = speciation_1,
+                 traits    = correlated_2D_BM)
 set.seed(2)
-change_correlation <- dads(bd.params = bd.params,
-                           stop.rule = stop.rule,
-                           traits    = traits,
-                           events    = cor_change)
+change_correlation <- dads(stop.rule = stop_taxa_100,
+                           bd.params = speciation_1,
+                           traits    = correlated_2D_BM,
+                           events    = correlation.change)
 
@@
 par(mfrow = c(1,2))
-plot(no_event, trait = 2)
-plot(change_correlation, trait = 2)
-
+plot(no_event, trait = c(1,2), main = "Strict correlation")
+plot(change_correlation, trait = c(1,2), main = "Loosened correlation")
 ```
 
-```{r, webgl=TRUE}
+And we can visualise this change through time:
+
+```{r}
 ## 3D plot
 plot(change_correlation, trait = c(1:2), use.3D = TRUE)
+rglwidget()
 ```
 
-### Event for changing a modifier
+## Event for changing a modifier: speciation event increase for species with negative values {#EG_change_modif}
+
+For this scenario, we want to generate a pure birth tree with a one dimensional Brownian Motion trait for 4 time units.
+After three time units, we want the speciation rule to increase for species which ancestors have a negative trait value.
+
+First we need to set up the simulation parameters:
+ * The stopping rule (5 time units)
+ * The birth-death parameters (speciation rate of 1)
 
@@
 ## Set the parameters
-stop.rule <- list(max.time = 4)
-bd.params <- list(extinction = 0, speciation = 1)
+stop_time_4 <- list(max.time = 4)
+speciation_1 <- make.bd.params(speciation = 1)
+```
 
-## Using a default trait and modifier object
-modifiers <- make.modifiers()
-traits <- make.traits()
+Then set up our trait which is a one dimensional Brownian Motion
 
+```{r}
+## Trait
+simple_bm_trait <- make.traits(n = 1, process = BM.process)
+```
+
+And a modifier that is the default birth-death algorithm rules
+
+```{r}
+## Birth death rules (default)
+default_modifiers <- make.modifiers()
+```
+
+And our extinction event which triggers after reaching time 3 (`time.condition(3)`), targets the `"modifiers"` and modifies birth-death rule as follows:
+ * When a species is descendant from a parent with a negative trait value (`negative.trait.condition`);
+ * Then increase your chances of going extinct by +1 (`increase.extinction.1`)
+
+```{r}
 ## New condition and new modifier (increasing speciation if trait is negative)
-new.condition <- function(trait.values, lineage) {
+negative.trait.condition <- function(trait.values, lineage) {
     return(parent.traits(trait.values, lineage) < 0)
 }
-new.modify    <- function(x, trait.values, lineage) {
+increase.extinction.1 <- function(x, trait.values, lineage) {
   return(x + 1)
@@
 ## Update the modifier
-change_speciation <- make.events(
+change.speciation <- make.events(
     condition    = time.condition(3),
@@
     modification = update.modifiers(speciation = speciation,
-                                    condition  = new.condition,
-                                    modify     = new.modify))
+                                    condition  = negative.trait.condition,
+                                    modify     = increase.extinction.1))
+```
 
+Once these parameters are defined, we can run the simulations and plot the results.
+We can contrast the results with the scenario without an event (but same random seed):
+
+```{r}
 set.seed(4)
-no_event <- dads(bd.params = bd.params,
-                 stop.rule = stop.rule,
-                 traits    = traits,
-                 modifiers = modifiers)
+no_event <- dads(stop.rule = stop_time_4,
+                 bd.params = speciation_1,
+                 traits    = simple_bm_trait,
+                 modifiers = default_modifiers)
 set.seed(4)
-change_spec <- dads(bd.params = bd.params,
-                    stop.rule = stop.rule,
-                    traits    = traits,
-                    modifiers = modifiers,
-                    events    = change_speciation)
+change_spec <- dads(stop.rule = stop_time_4,
+                    bd.params = speciation_1,
+                    traits    = simple_bm_trait,
+                    modifiers = default_modifiers,
+                    events    = change.speciation)
 
@@
 
-### Changing branch length when reaching n taxa
+# Changing branch length when reaching n taxa {#EG_modify_brlen}
+
+For this scenario, we want to generate a pure birth tree until reaching 100 taxa.
+After reaching 30 taxa we want branch length growth to increase 50 folds.
+
+First we need to set up the simulation parameters:
+ * The stopping rule (100 taxa)
+ * The birth-death parameters (speciation rate of 1)
 
 ```{r}
-## Parameters
-bd.params <- list(extinction = 0, speciation = 1)
-stop.rule <- list(max.taxa = 100)
-modifiers <- make.modifiers()
-traits <- make.traits()
+## Set the parameters
+stop_taxa_100<- list(max.taxa = 100)
+speciation_1 <- make.bd.params(speciation = 1)
+```
 
-## multiplying branch length 100 folds
-new.modify <- function(x, trait.values, lineage) {
-  return(x * 100)
+Then a modifier that is the default birth-death algorithm rules
+
+```{r}
+## Birth death rules (default)
+default_modifiers <- make.modifiers()
+```
+
+And event which triggers after reaching 30 taxa (`taxa.condition(30)`), targets the `"modifiers"` and modifies the branch length generation rule to a 50 fold increase (`increase.50.folds`)
+
+```{r}
+## Multiplying branch length 50 folds
+increase.50.folds <- function(x, trait.values, lineage) {
+  return(x * 50)
 }
-
 ## Event for increasing branch length after reaching 30 taxa
@@
                                     branch.length = branch.length,
-                                    modify = new.modify))
+                                    modify        = increase.50.folds))
+```
 
+Once these parameters are defined, we can run the simulations and plot the results.
+We can contrast the results with the scenario without an event (but same random seed):
+
+```{r}
 ## Run the simulations
 set.seed(5)
-no_event <- dads(bd.params = bd.params,
-                 stop.rule = stop.rule,
-                 traits    = traits,
-                 modifiers = modifiers)
+no_event <- dads(stop.rule = stop_taxa_100,
+                 bd.params = speciation_1,
+                 modifiers = default_modifiers)
 set.seed(5)
-increased_brlen <- dads(bd.params = bd.params,
-                        stop.rule = stop.rule,
-                        traits    = traits,
-                        modifiers = modifiers,
+increased_brlen <- dads(stop.rule = stop_taxa_100,
+                        bd.params = speciation_1,
+                        modifiers = default_modifiers,
                         events    = increase_brlen)
@@
 par(mfrow = c(1,2))
-plot(no_event, main = "No event")
-plot(increased_brlen, main = "Increase branch length\nafter 30 taxa")
+plot(no_event, main = "No event", show.tip.label = FALSE)
+plot(increased_brlen, main = "Increase branch length\nafter 30 taxa",
+    show.tip.label = FALSE)
 ```
 
-### Founding event: a subtree with no fossils
+## Founding event: a generating a subtree with no fossils {#EG_founding_purebirth}
+
+For this scenario, we want to generate a birth-death tree for 4 time units.
+After reaching 10 taxa, one random taxa will give birth to a sub-tree that is a pure birth tree (no extinction).
+
+First we need to set up the simulation parameters:
+ * The stopping rule (5 time units)
+ * The birth-death parameters (speciation rate of 1 and extinction of 0.2)
 
@@
 ## Set up parameters
-stop.rule <- list(max.time = 4)
-bd.params <- list(speciation = 1, extinction = 0.3)
+stop_time_4 <- list(max.time = 4)
+spec_1_ext_02 <- make.bd.params(speciation = 1, extinction = 0.2)
+```
+
+And our event which triggers after reaching 10 taxa (`taxa.condition(10)`), and generates a subtree ("founding") that is a pure birth tree (no extinction and speciation rate of 2).
+
+```{r}
+## Setting the pure-birth parameters
+speciation_2 <- make.bd.params(speciation = 2)
 
@@
                   modification = founding.event(
-                                    bd.params = list(speciation = 2,
-                                                     extinction = 0)),
+                                    bd.params = speciation_2),
                   additional.args = list(prefix = "founding_"))
+```
 
+> Note we are prodviding an additional argument `prefix` here so that we can track which species are part of the sub tree for colouring them down the line.
+
+Once these parameters are defined, we can run the simulations and plot the results:
+
+```{r}    
 ## Simulations
 set.seed(11)
-founding_tree <- dads(bd.params = bd.params,
-                      stop.rule = stop.rule,
+founding_tree <- dads(stop.rule = stop_time_4,
+                      bd.params = spec_1_ext_02,
                       events    = founding_event)
-plot(founding_tree, cex = 0.4)
+
+## Selecting the edges colours
+tip_values <- rep("black", Ntip(founding_tree))
+tip_values[grep("founding_", founding_tree$tip.label)] <- "orange"
+edge_colors <- match.tip.edge(tip_values, founding_tree, replace.na = "black")
+
+## Plotting the results
+plot(founding_tree, show.tip.label = FALSE, edge.color = edge_colors)
 ```
 
-### Founding event with traits
+## Founding event: a generating a subtree a different process {#EG_founding_traits}
+
+For this scenario, we want to generate a birth-death tree with a one dimensional Brownian Motion trait for 6 time unit.
+After a taxa reaches the value 3 or higher, it gives birth to a sub-tree that generates an OU trait with an optimum at the value 3.
+
+First we need to set up the simulation parameters:
+ * The stopping rule (6 time units)
+ * The birth-death parameters (speciation rate of 1 and extinction of 1/3)
 
 ```{r}
-## Set up parameters
-stop.rule <- list(max.time = 4)
-bd.params <- list(speciation = 1, extinction = 0.3)
-traits    <- make.traits()
+## The tree parameters
+stop_time_6 <- list(max.time = 6)
+speciation_1_extinction_03 <- make.bd.params(speciation = 1,
+                                             extinction = 0.3)
+```
 
-## Events that generate a new process (founding effects)
-founding_event <- make.events(
-                  condition    = taxa.condition(10),
+Then set up our trait which is a one dimensional Brownian Motion
+
+```{r}
+## Trait
+simple_bm_trait <- make.traits(n = 1, process = BM.process)
+```
+
+When a taxa reaches the value 3 `trait.condition`, it generates a pure birth tree (`speciation = 2`) with an OU trait with the optimum value 3.
+
+```{r}
+## The OU trait with an optimum value of 3
+OU_3 <- make.traits(process = OU.process,
+                    start = 3, process.args = list(optimum = 3))
+## The pure birth parameters
+speciation_2 <- make.bd.params(speciation = 2)
+
+## The founding event
+new_OU_trait <- make.events(
+                    condition    = trait.condition(3, condition = `>=`),
                     target       = "founding",
                     modification = founding.event(
-                                    bd.params = list(speciation = 2,
-                                                     extinction = 0),
-                                    traits = make.traits(process = OU.process)),
-                  additional.args = list(prefix = "founding_"))
+                                    bd.params = speciation_2,
+                                    traits    = OU_3))
+```
 
+Once these parameters are defined, we can run the simulations and plot the results:
+
+```{r}
 ## Simulating the tree
-set.seed(19)
-founding_tree <- dads(bd.params = bd.params,
-                      stop.rule = stop.rule,
-                      traits    = make.traits(),
-                      events    = founding_event)
+set.seed(1)
+founding_tree <- dads(stop.rule = stop_time_6,
+                      bd.params = speciation_1_extinction_03,
+                      traits    = simple_bm_trait,
+                      events    = new_OU_trait)
 plot(founding_tree)

### 2023-03-27 inst/gitbook/_book/dads_manual.tex
manual big update

--- 2020-12-04 inst/gitbook/_book/dads_manual.tex
@@
-\PassOptionsToPackage{unicode=true}{hyperref} % options for packages loaded elsewhere
+% Options for packages loaded elsewhere
+\PassOptionsToPackage{unicode}{hyperref}
 \PassOptionsToPackage{hyphens}{url}
 %
-\documentclass[]{book}
+\documentclass[
+]{book}
 \usepackage{lmodern}
@@
 \usepackage{ifxetex,ifluatex}
-\usepackage{fixltx2e} % provides \textsubscript
 \ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
@@
   \usepackage[utf8]{inputenc}
-  \usepackage{textcomp} % provides euro and other symbols
-\else % if luatex or xelatex
+  \usepackage{textcomp} % provide euro and other symbols
+\else % if luatex or xetex
   \usepackage{unicode-math}
-  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
+  \defaultfontfeatures{Scale=MatchLowercase}
+  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
 \fi
-% use upquote if available, for straight quotes in verbatim environments
+% Use upquote if available, for straight quotes in verbatim environments
 \IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-% use microtype if available
-\IfFileExists{microtype.sty}{%
+\IfFileExists{microtype.sty}{% use microtype if available
   \usepackage[]{microtype}
@@
 }{}
+\makeatletter
+\@ifundefined{KOMAClassName}{% if non-KOMA class
   \IfFileExists{parskip.sty}{%
@@
     \setlength{\parindent}{0pt}
-\setlength{\parskip}{6pt plus 2pt minus 1pt}
-}
-\usepackage{hyperref}
+    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
+}{% if KOMA class
+  \KOMAoptions{parskip=half}}
+\makeatother
+\usepackage{xcolor}
+\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
+\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
 \hypersetup{
-            pdftitle={dads manual},
-            pdfauthor={Thomas Guillerme (guillert@tcd.ie), Alex Slavenko (email) and others},
-            pdfborder={0 0 0},
-            breaklinks=true}
-\urlstyle{same}  % don't use monospace font for urls
+  pdftitle={Specific examples},
+  pdfauthor={Thomas Guillerme (guillert@tcd.ie)},
+  hidelinks,
+  pdfcreator={LaTeX via pandoc}}
+\urlstyle{same} % disable monospaced font for URLs
 \usepackage{color}
@@
 \usepackage{longtable,booktabs}
-% Fix footnotes in tables (requires footnote package)
-\IfFileExists{footnote.sty}{\usepackage{footnote}\makesavenoteenv{longtable}}{}
-\usepackage{graphicx,grffile}
+% Correct order of tables after \paragraph or \subparagraph
+\usepackage{etoolbox}
+\makeatletter
+\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
+\makeatother
+% Allow footnotes in longtable head/foot
+\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
+\makesavenoteenv{longtable}
+\usepackage{graphicx}
 \makeatletter
@@
 \setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
+% Set default figure placement to htbp
+\makeatletter
+\def\fps@figure{htbp}
+\makeatother
 \setlength{\emergencystretch}{3em} % prevent overfull lines
@@
 \setcounter{secnumdepth}{5}
-% Redefines (sub)paragraphs to behave more like sections
-\ifx\paragraph\undefined\else
-\let\oldparagraph\paragraph
-\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
-\fi
-\ifx\subparagraph\undefined\else
-\let\oldsubparagraph\subparagraph
-\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
-\fi
-
-% set default figure placement to htbp
-\makeatletter
-\def\fps@figure{htbp}
-\makeatother
-
 \usepackage{booktabs}
@@
 
-\title{dads manual}
-\author{Thomas Guillerme (\href{mailto:guillert@tcd.ie}{\nolinkurl{guillert@tcd.ie}}), Alex Slavenko (email) and others}
-\date{2020-12-04}
+\title{Specific examples}
+\author{Thomas Guillerme (\href{mailto:guillert@tcd.ie}{\nolinkurl{guillert@tcd.ie}})}
+\date{2023-03-27}
 
@@
 \texttt{dads} aims to be a highly modular and friendly version of all these packages: it allows to simulate disparity and diversity jointly with a vast array of options that can be easily modified by users.
-For example, you can easily generate any type of process to generate a trait (BM, OU, something else, etc\ldots{}) in multiple dependent or independent dimensions through \texttt{"traits"} objects.
+For example, you can easily generate any type of process to generate a trait (BM, OU, something else, etc\ldots) in multiple dependent or independent dimensions through \texttt{"traits"} objects.
 You can then specify how the traits should affect disparity through \texttt{"modifiers"} objects. And finally you can create events (like mass extinctions) through \texttt{"events"} objects.
@@
 
+\hypertarget{who-is-this-manual-for}{%
+\section{Who is this manual for?}\label{who-is-this-manual-for}}
+
+This manual explains how \texttt{dads} work in as much details as possible and aims to give you the keys to design your own diversity and disparity simulations.
+
+The first section \texttt{Getting\ started} is aimed for people with a \texttt{R} beginner level:
+* you know what a package is - and how to install it;
+* you know what a function is - and what arguments are;
+* you have already (vaguely) heard that they are different types of objects in R - like matrices (\texttt{"matrix"}) or trees (\texttt{"phylo"});
+* and you have already looked for documentation and bugs online.
+
+The sections after that are a bit more advanced and requires the following level:
+* you have already created functions before;
+* you know how to subset elements of a list - i.e.~you understand the \texttt{list\$element} syntax;
+* you are aware that both \texttt{TRUE\ ==\ 1} and \texttt{FALSE\ ==\ 0} are \texttt{TRUE};
+* you have heard that the objects in R have classes and that it sometimes matters - for example you know that R can differentiate between \texttt{as.numeric(1)} and \texttt{as.integer(1)};
+* you owe part of your life to Stack Overflow.
+
 \hypertarget{installing-and-running-the-package}{%
@@
 \begin{Highlighting}[]
-\CommentTok{## Checking if devtools is already installed}
+\CommentTok{\#\# Checking if devtools is already installed}
 \ControlFlowTok{if}\NormalTok{(}\OperatorTok{!}\KeywordTok{require}\NormalTok{(devtools)) }\KeywordTok{install.packages}\NormalTok{(}\StringTok{"devtools"}\NormalTok{)}
 
-\CommentTok{## Installing the latest version directly from GitHub}
-\KeywordTok{install_github}\NormalTok{(}\StringTok{"TGuillerme/dads"}\NormalTok{)}
+\CommentTok{\#\# Installing the latest version directly from GitHub}
+\NormalTok{devtools}\OperatorTok{::}\KeywordTok{install\_github}\NormalTok{(}\StringTok{"TGuillerme/dads"}\NormalTok{)}
 \end{Highlighting}
@@
 \centering
-\includegraphics{dads_structure.pdf}
+\includegraphics{dads_structure.png}
 \caption{Schematised summary of the \texttt{dads} package architecture}
@@
 
+\hypertarget{bdalgorithm}{%
+\chapter{The birth-death algorithm}\label{bdalgorithm}}
+
+\begin{quote}
+If you want to get your hands dirty, you can go straight to the \protect\hyperlink{gettingstarted}{getting start section}, this following section just describes the process in pseudo-code.
+\end{quote}
+
+The \texttt{dads} algorithm is based on a modular version of the birth-death process algorithm.
+The birth-death model is a continuous Markov process (i.e.~a continuous random process where the state will change each step are linked through time and change according to some random exponential variables) which is well defined mathematically and \href{https://lukejharmon.github.io/pcm/chapter10_birthdeath/\#section-10.2-the-birth-death-model}{commonly used in biology} but also in \href{https://en.wikipedia.org/wiki/Birth\%E2\%80\%93death_process}{many other fields}.
+
+The algorithm used in \texttt{dads} allows modularity of this process and is based on the following steps (the text in \texttt{courier\ font} is for the name of the process in the algorithm):
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\setcounter{enumi}{-1}
+\tightlist
+\item
+  \textbf{Starting the process}: this step is actually non modular and creates a random tree with one tip, one node and one branch length connecting both.
+This step is used to optimise the rest of the algorithm in terms of speed and memory management.
+The node, tip and branch resulting from this step are discarded at the end of the simulations.
+\item
+  \textbf{Selecting a lineage} (\texttt{selecting}): this step selects a tip that is currently not extinct.
+In a standard birth-death process this is just done randomly, however in \texttt{dads} this can be modified based on the birth-death parameters, the currently available lineages and potential trait values.
+\emph{For example, it is possible to put a higher probability in selecting a lineage that is next to a lineage that recently went extinct and has a positive trait value.}
+  Then go to step 2.
+\item
+  \textbf{Growing the tree} (\texttt{waiting}): this step grows the tree by a certain amount.
+It does so by adding the same amount of branch length to all the non-extinct lineages.
+In an exact birth death process, this is done by drawing a random value from an exponential distribution with a rate of \[\text{number of living lineages} \times (\text{speciation} + \text{extinction parameters})\].
+In \texttt{dads}, this can be modified based on the birth-death parameters, the currently available lineages and potential trait values.
+\emph{For example, it is possible to increase branch length by the ratio of living/extinct fossils and a random number drawn from the range of current trait values.}
+  If the total tree length is less than the required tree length, go to step 3. Else go to step 6.
+\item
+  \textbf{Simulating traits} {[}optional{]}(\texttt{traits}): this step allows to simulate a trait value for the selected lineage from step 1.
+This is typically not part of a standard birth-death process and is handled via the \texttt{traits} option in \texttt{dads} (see the \protect\hyperlink{maketraits}{\texttt{make.traits}} chapter for more details).
+  Then go to step 4.
+\item
+  \textbf{Speciating} (\texttt{speciating}): in this step, the selected lineage from step 1 has the option of speciating or going extinct.
+In a standard birth-death process, this happens by randomly drawing a value between 0 and 1 as a way to trigger speciation relatively to the birth-death parameters:
+\end{enumerate}
+
+\begin{verbatim}
+if
+    randomly drawn number is smaller or equal to speciation/(speciation + extinction)
+then
+    do speciate
+else
+    go extinct
+\end{verbatim}
+
+Again, in \texttt{dads}, this process can be modified based on the birth-death parameters, the currently available lineages and potential trait values.
+\emph{For example, the lineage can only speciate if it's trait value is positive, regardless of whether speciation or extinction has been triggered}.
+Then go to step 5.
+5. If the number of lineages is less than required number of lineages, then go to step 1. Else go to step 6.
+6. The simulation has stopped because it reached the required amount of lineages or time.
+
 \begin{verbatim}
@@
 
-\hypertarget{getting-started}{%
-\chapter{Getting started}\label{getting-started}}
+\hypertarget{gettingstarted}{%
+\chapter{Getting started}\label{gettingstarted}}
 
@@
 \item
-  \texttt{"max.living"\ =\ n} stop when there is \texttt{n} co-occuring taxa of the same age (i.e. ``living'' taxa);
+  \texttt{"max.living"\ =\ n} stop when there is \texttt{n} co-occuring taxa of the same age (i.e.~``living'' taxa);
 \item
@@
 \begin{Highlighting}[]
-\CommentTok{## Setting a stopping rule to reach a maximum of 20 taxa}
-\NormalTok{my_stop_rule <-}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{max.taxa =} \DecValTok{20}\NormalTok{)}
+\CommentTok{\#\# Setting a stopping rule to reach a maximum of 20 taxa}
+\NormalTok{my\_stop\_rule \textless{}{-}}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{max.taxa =} \DecValTok{20}\NormalTok{)}
 \end{Highlighting}
@@
 \begin{Highlighting}[]
-\CommentTok{## Running the birth death simulation}
-\NormalTok{my_tree <-}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{stop.rule =}\NormalTok{ my_stop_rule)}
+\CommentTok{\#\# Running the birth death simulation}
+\NormalTok{my\_tree \textless{}{-}}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{stop.rule =}\NormalTok{ my\_stop\_rule)}
 \end{Highlighting}
@@
 \begin{Highlighting}[]
-\CommentTok{## The tree object}
-\NormalTok{my_tree}
+\CommentTok{\#\# The tree object}
+\NormalTok{my\_tree}
 \end{Highlighting}
@@
 \begin{Highlighting}[]
-\CommentTok{## Plotting it}
-\KeywordTok{plot}\NormalTok{(my_tree)}
+\CommentTok{\#\# Plotting it}
+\KeywordTok{plot}\NormalTok{(my\_tree)}
 \end{Highlighting}
@@
 \begin{Highlighting}[]
-\CommentTok{## my birth death parameters}
-\NormalTok{my_params <-}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{speciation =} \DecValTok{1}\NormalTok{,}
+\CommentTok{\#\# my birth death parameters}
+\NormalTok{my\_params \textless{}{-}}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{speciation =} \DecValTok{1}\NormalTok{,}
                   \DataTypeTok{extinction =} \DecValTok{1}\OperatorTok{/}\DecValTok{3}\NormalTok{)}
@@
 \begin{quote}
-Note that here it is not necessary to specify \texttt{extinction\ =\ 1} since this is the default option, you can always just change the parameter of interest (e.g.~changing \texttt{extinciton\ =\ 0} to \texttt{extinction\ =\ 1/3}).
+Note that here it is not necessary to specify \texttt{speciation\ =\ 1} since this is the default option, you can always just change the parameter of interest (e.g.~changing \texttt{extinction\ =\ 0} to \texttt{extinction\ =\ 1/3}).
 However, we think it's good practice to attribute both parameters specifically to avoid any confusion.
+You can find more information about setting up more complex birth death parameters \protect\hyperlink{makebdparams}{in this section}.
 \end{quote}
@@
 \begin{Highlighting}[]
-\CommentTok{## Generating a birth death tree with extinctions:}
-\NormalTok{my_tree <-}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ my_params, }\DataTypeTok{stop.rule =}\NormalTok{ my_stop_rule)}
-\CommentTok{## Visualising the new tree}
-\KeywordTok{plot}\NormalTok{(my_tree)}
+\CommentTok{\#\# Generating a birth death tree with extinctions:}
+\NormalTok{my\_tree \textless{}{-}}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ my\_params, }\DataTypeTok{stop.rule =}\NormalTok{ my\_stop\_rule)}
+\CommentTok{\#\# Visualising the new tree}
+\KeywordTok{plot}\NormalTok{(my\_tree)}
 \end{Highlighting}
@@
 \begin{Highlighting}[]
-\CommentTok{## Creating the traits object}
-\NormalTok{my_trait <-}\StringTok{ }\KeywordTok{make.traits}\NormalTok{()}
+\CommentTok{\#\# Creating the traits object}
+\NormalTok{my\_trait \textless{}{-}}\StringTok{ }\KeywordTok{make.traits}\NormalTok{()}
 \end{Highlighting}
@@
 \begin{Highlighting}[]
-\CommentTok{## Which process is in here?}
-\NormalTok{my_trait}
+\CommentTok{\#\# Which process is in here?}
+\NormalTok{my\_trait}
 \end{Highlighting}
@@
 \begin{Highlighting}[]
-\CommentTok{## What does it look like?}
-\KeywordTok{plot}\NormalTok{(my_trait)}
+\CommentTok{\#\# What does it look like?}
+\KeywordTok{plot}\NormalTok{(my\_trait)}
 \end{Highlighting}
@@
 \begin{Highlighting}[]
-\CommentTok{## Simulate disparity and diversity}
-\NormalTok{my_data <-}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ my_params,}
-                \DataTypeTok{stop.rule =}\NormalTok{ my_stop_rule,}
-                \DataTypeTok{traits    =}\NormalTok{ my_trait)}
+\CommentTok{\#\# Simulate disparity and diversity}
+\NormalTok{my\_data \textless{}{-}}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ my\_params,}
+                \DataTypeTok{stop.rule =}\NormalTok{ my\_stop\_rule,}
+                \DataTypeTok{traits    =}\NormalTok{ my\_trait)}
 \end{Highlighting}
@@
 \begin{Highlighting}[]
-\CommentTok{## What's in there}
-\NormalTok{my_data}
+\CommentTok{\#\# What\textquotesingle{}s in there}
+\NormalTok{my\_data}
 \end{Highlighting}
@@
 \begin{Highlighting}[]
-\CommentTok{## Plotting the disparity and diversity}
-\KeywordTok{plot}\NormalTok{(my_data)}
+\CommentTok{\#\# Plotting the disparity and diversity}
+\KeywordTok{plot}\NormalTok{(my\_data)}
 \end{Highlighting}
@@
 \begin{Highlighting}[]
-\CommentTok{## Extracting the tree (a "phylo" object)}
-\NormalTok{the_generated_tree <-}\StringTok{ }\NormalTok{my_data}\OperatorTok{$}\NormalTok{tree}
-\KeywordTok{class}\NormalTok{(the_generated_tree)}
+\CommentTok{\#\# Extracting the tree (a "phylo" object)}
+\NormalTok{the\_generated\_tree \textless{}{-}}\StringTok{ }\NormalTok{my\_data}\OperatorTok{$}\NormalTok{tree}
+\KeywordTok{class}\NormalTok{(the\_generated\_tree)}
 \end{Highlighting}
@@
 \begin{Highlighting}[]
-\CommentTok{## Extracting the data (a "matrix")}
-\NormalTok{the_generated_data <-}\StringTok{ }\NormalTok{my_data}\OperatorTok{$}\NormalTok{data}
-\KeywordTok{class}\NormalTok{(the_generated_data)}
+\CommentTok{\#\# Extracting the data (a "matrix")}
+\NormalTok{the\_generated\_data \textless{}{-}}\StringTok{ }\NormalTok{my\_data}\OperatorTok{$}\NormalTok{data}
+\KeywordTok{class}\NormalTok{(the\_generated\_data)}
 \end{Highlighting}
@@
 \begin{Highlighting}[]
-\CommentTok{## Make a default modifiers}
-\NormalTok{default_modifiers <-}\StringTok{ }\KeywordTok{make.modifiers}\NormalTok{()}
-\CommentTok{## What's in it?}
-\NormalTok{default_modifiers}
+\CommentTok{\#\# Make a default modifiers}
+\NormalTok{default\_modifiers \textless{}{-}}\StringTok{ }\KeywordTok{make.modifiers}\NormalTok{()}
+\CommentTok{\#\# What\textquotesingle{}s in it?}
+\NormalTok{default\_modifiers}
 \end{Highlighting}
@@
 \begin{Highlighting}[]
-\CommentTok{## Setting the simulation parameters}
-\NormalTok{extinction_}\DecValTok{02}\NormalTok{ <-}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{extinction =} \FloatTok{0.2}\NormalTok{)}
-\NormalTok{living_}\DecValTok{20}\NormalTok{     <-}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{max.living =} \DecValTok{20}\NormalTok{)}
-\NormalTok{BM_trait      <-}\StringTok{ }\KeywordTok{make.traits}\NormalTok{()}
+\CommentTok{\#\# Setting the simulation parameters}
+\NormalTok{extinction\_}\DecValTok{02}\NormalTok{ \textless{}{-}}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{extinction =} \FloatTok{0.2}\NormalTok{)}
+\NormalTok{living\_}\DecValTok{20}\NormalTok{     \textless{}{-}}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{max.living =} \DecValTok{20}\NormalTok{)}
+\NormalTok{BM\_trait      \textless{}{-}}\StringTok{ }\KeywordTok{make.traits}\NormalTok{()}
+
 \KeywordTok{set.seed}\NormalTok{(}\DecValTok{1}\NormalTok{)}
-\CommentTok{## Simulate disparity and diversity}
-\NormalTok{default_data <-}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ extinction_}\DecValTok{02}\NormalTok{,}
-                     \DataTypeTok{stop.rule =}\NormalTok{ living_}\DecValTok{20}\NormalTok{,}
-                     \DataTypeTok{traits    =}\NormalTok{ BM_trait,}
-                     \DataTypeTok{modifiers =}\NormalTok{ default_modifiers)}
-\NormalTok{default_data}
+\CommentTok{\#\# Simulate disparity and diversity}
+\NormalTok{default\_data \textless{}{-}}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ extinction\_}\DecValTok{02}\NormalTok{,}
+                     \DataTypeTok{stop.rule =}\NormalTok{ living\_}\DecValTok{20}\NormalTok{,}
+                     \DataTypeTok{traits    =}\NormalTok{ BM\_trait,}
+                     \DataTypeTok{modifiers =}\NormalTok{ default\_modifiers)}
+\NormalTok{default\_data}
 \end{Highlighting}
@@
 ## Birth death process with modifiers:
+## speciation: 1.
+## extinction: 0.2.
 ## No modifiers applied to the branch length, selection and speciation processes (default).
@@
 For that we need to create a modifiers object that modifies the \texttt{speciation} process with a specific condition and a specific modification when that condition is met.
-First we are going to create the modification function.
-This function must intake the argument \texttt{x} and, in our case, return a logical value: \texttt{TRUE} is for speciate and \texttt{FALSE} is for go extinct.
-We want this \texttt{modify} function to always return \texttt{FALSE} (go extinct) since it will be triggered by \texttt{condition} (we will see that in a minute).
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{## Going extinct}
-\NormalTok{staying.alive <-}\StringTok{ }\ControlFlowTok{function}\NormalTok{(x) }\KeywordTok{return}\NormalTok{(}\OtherTok{TRUE}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-Now we want this function to only trigger when an ancestor has a negative trait value.
+For a speciation to occur (and a species to not go extinct), the algorithm draws a random value \emph{x} between 0 and 1 and if the value is smaller than the speciation parameter divided by the speciation + extinction parameter, the lineage speciates (\(x < (\lambda \ (\lambda + \mu))\)), else it goes extinct.
+
+Here we can create a modifier forcing the lineage to go extinct every time a specific condition is met (say the lineage trait value is negative).
+In other words, if the lineages trait value is negative, make sure that \emph{x} is equal to 1 (so that the speciation condition \(x < (\lambda \ (\lambda + \mu))\) is never met).
+
+First we need to create a condition function to trigger this (non) speciation event.
 We can do that by specifying our \texttt{condition} function (when the ancestor is trait is negative) apply our modification (here the \texttt{staying.alive} function).
 For that we can use the \texttt{parent.traits} utility function that is optimised for accessing traits in the birth death process (but you can of course write your own).
-This function intakes the \texttt{trait.values} and \texttt{parent.lineage} arguments, two arguments that you can leave named as they are to facilitate \texttt{dads}s understanding of what you which to asses:
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{## Triggering a modification only if the ancestor trait is negative}
-\NormalTok{negative.ancestor <-}\StringTok{ }\ControlFlowTok{function}\NormalTok{(trait.values, lineage) \{}
-    \KeywordTok{return}\NormalTok{(}\KeywordTok{all}\NormalTok{(}\KeywordTok{parent.traits}\NormalTok{(trait.values, lineage) }\OperatorTok{<}\StringTok{ }\DecValTok{0}\NormalTok{))}
+This function intakes the \texttt{trait.values} and \texttt{parent.lineage} arguments, two arguments that you can leave named as they are to facilitate \texttt{dads}' understanding of what you want to assess:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# Triggering a modification only if the ancestor trait is negative}
+\NormalTok{negative.ancestor \textless{}{-}}\StringTok{ }\ControlFlowTok{function}\NormalTok{(trait.values, lineage) \{}
+    \KeywordTok{return}\NormalTok{(}\KeywordTok{all}\NormalTok{(}\KeywordTok{parent.traits}\NormalTok{(trait.values, lineage) }\OperatorTok{\textless{}}\StringTok{ }\DecValTok{0}\NormalTok{))}
 \NormalTok{\}}
@@
 
-Note that we use the function \texttt{all} here to evaluate all traits (if the data is multidimensional).
+\begin{quote}
+Note that we use the function \texttt{all} here to evaluate all traits: i.e.~if the data has more than one trait we trigger the modification only if all the trait values are negative.
+\end{quote}
+
+The we are going to create the modification function.
+This function must intake the argument \texttt{x} and, in our case, returns the same value no matter what: \texttt{1} so that speciation never happens when the \texttt{negative.ancestor} function is triggered.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# Always go extinct}
+\NormalTok{going.extinct \textless{}{-}}\StringTok{ }\ControlFlowTok{function}\NormalTok{(x) }\KeywordTok{return}\NormalTok{(}\DecValTok{1}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
 We can then provide these two functions (the condition \texttt{negative.ancestor} and how to modify the speciation event when this condition is met \texttt{staying.alive}).
@@
 \begin{Highlighting}[]
-\CommentTok{## Making a modifier for species to go extinct if}
-\CommentTok{## their ancestor's trait value is (or are) negative}
-\NormalTok{negatives_extinct <-}\StringTok{ }\KeywordTok{make.modifiers}\NormalTok{(}
-                        \CommentTok{## The speciation function (default)}
-                        \DataTypeTok{speciation =}\NormalTok{ speciation,}
-                        \CommentTok{## What to modify}
-                        \DataTypeTok{modify     =}\NormalTok{ staying.alive,}
-                        \CommentTok{## When to modify it}
-                        \DataTypeTok{condition  =}\NormalTok{ negative.ancestor)}
-\CommentTok{## What's in it?}
-\NormalTok{negatives_extinct}
+\CommentTok{\#\# Making a modifier for species to go extinct if}
+\CommentTok{\#\# their ancestor\textquotesingle{}s trait value is (or are) negative}
+\NormalTok{negatives\_extinct \textless{}{-}}\StringTok{ }\KeywordTok{make.modifiers}\NormalTok{(}
+            \CommentTok{\#\# If the following condition is met...}
+            \DataTypeTok{condition =}\NormalTok{ negative.ancestor, }\CommentTok{\# Does the lineage has a negative ancestor?}
+            \CommentTok{\#\# Then apply the following modifier...}
+            \DataTypeTok{modify =}\NormalTok{ going.extinct, }\CommentTok{\# The species goes extinct}
+            \CommentTok{\#\# To the the speciation process.}
+            \DataTypeTok{speciation =}\NormalTok{ speciation) }\CommentTok{\# Here the speciation() process is default}
+
+\CommentTok{\#\# What\textquotesingle{}s in it?}
+\NormalTok{negatives\_extinct}
 \end{Highlighting}
@@
 ## Default selection process.
-## Speciation process is set to speciation with a condition (negative.ancestor) and a modifier (staying.alive).
+## Speciation process is set to speciation with a condition (negative.ancestor) and a modifier (going.extinct).
 \end{verbatim}
@@
 \KeywordTok{set.seed}\NormalTok{(}\DecValTok{1}\NormalTok{)}
-\CommentTok{## Simulate disparity and diversity}
-\NormalTok{biased_data <-}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ extinction_}\DecValTok{02}\NormalTok{,}
-                    \DataTypeTok{stop.rule =}\NormalTok{ living_}\DecValTok{20}\NormalTok{,}
-                    \DataTypeTok{traits    =}\NormalTok{ BM_trait,}
-                    \DataTypeTok{modifiers =}\NormalTok{ negatives_extinct)}
-\NormalTok{biased_data}
+\CommentTok{\#\# Simulate disparity and diversity}
+\NormalTok{biased\_data \textless{}{-}}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ extinction\_}\DecValTok{02}\NormalTok{,}
+                    \DataTypeTok{stop.rule =}\NormalTok{ living\_}\DecValTok{20}\NormalTok{,}
+                    \DataTypeTok{traits    =}\NormalTok{ BM\_trait,}
+                    \DataTypeTok{modifiers =}\NormalTok{ negatives\_extinct)}
+\NormalTok{biased\_data}
 \end{Highlighting}
@@
 ## Birth death process with modifiers:
+## speciation: 1.
+## extinction: 0.2.
 ## Default branch length process.
 ## Default selection process.
-## Speciation process is set to speciation with a condition (negative.ancestor) and a modifier (staying.alive).
+## Speciation process is set to speciation with a condition (negative.ancestor) and a modifier (going.extinct).
 ## 
@@
 We can now compare the two trees and their trait values.
-Note that we've used the same starting seed for both trees so the only thing differing between them is the modifier!
-Also, although species
+Note that we've used the same starting seed for both trees so the only thing differing between them is the \texttt{"modifier"} object which leads to very different trees!
 
@@
 \KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow =} \KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{,}\DecValTok{2}\NormalTok{))}
-\KeywordTok{plot}\NormalTok{(default_data, }\DataTypeTok{main =} \StringTok{"default results"}\NormalTok{)}
-\KeywordTok{plot}\NormalTok{(biased_data, }\DataTypeTok{main =} \StringTok{"results with the modifier"}\NormalTok{)}
+\KeywordTok{plot}\NormalTok{(default\_data, }\DataTypeTok{main =} \StringTok{"default results"}\NormalTok{)}
+\KeywordTok{plot}\NormalTok{(biased\_data, }\DataTypeTok{main =} \StringTok{"results with the modifier"}\NormalTok{)}
 \end{Highlighting}
@@
 \hypertarget{maketraits}{%
-\chapter{\texorpdfstring{Making trait processes with \texttt{make.traits()}}{Making trait processes with make.traits()}}\label{maketraits}}
+\chapter{Simulating traits}\label{maketraits}}
+
+In \texttt{dads}, traits are simulating by providing a \texttt{traits} object to the \texttt{traits} argument.
+This object is generated using the \texttt{make.traits}.
+
+\begin{quote}
+\textbf{What is a trait in biology?} This is one of these epistemological question like what \emph{is} a species that is not covered by all means in this package.
+For the purpose of \emph{this manual}, \textbf{biological trait} can be any coherent process of any number of dimensions which can or cannot be independent.
+Basically anything can be a trait as well as it's defined by the user (e.g.~one trait can be a 20 dimensional Brownian process that's correlated to a 1D OU process - which is another trait).
+For the purpose of the syntax in \emph{this package} though, a \textbf{trait in \texttt{dads}} is a column in the output trait matrix.
+For example, one might be interested in a biological trait that is ``beak shape'' which is defined as one complex biological trait that is the Procrustes superimposition of 4 3D landmarks and 300 3D semi-landmarks.
+This single ``beak shape'' \textbf{biological trait} is then composed of 912 \textbf{traits in \texttt{dads}} (i.e.~a matrix of \((4+300)\times3\) columns).
+\end{quote}
 
@@
 
+\hypertarget{the-syntax-how-to-code-a-process}{%
+\subsection{The syntax (how to code a process?)}\label{the-syntax-how-to-code-a-process}}
+
 Trait processes in \texttt{dads} are functions that must always intake the following arguments by default.
@@
 
-For example, the following function would be a valid process (though not dependent on either the previous state nor the branch length):
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{## A valid (but useless) process}
-\NormalTok{valid.process <-}\StringTok{ }\ControlFlowTok{function}\NormalTok{(x0, }\DataTypeTok{egde.length =} \DecValTok{1}\NormalTok{, ...) \{}
+For example, the following function would be a valid process that always generate the \emph{true} trait value: 42!.
+In this example, the process is not dependent on neither the previous state (\texttt{x0}) and the branch length (\texttt{edge.length}).
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# A valid (but useless?) process}
+\NormalTok{valid.process \textless{}{-}}\StringTok{ }\ControlFlowTok{function}\NormalTok{(}\DataTypeTok{x0 =} \DecValTok{0}\NormalTok{, }\DataTypeTok{edge.length =} \DecValTok{1}\NormalTok{, ...) \{}
     \KeywordTok{return}\NormalTok{(}\DecValTok{42}\NormalTok{)}
@@
 \begin{quote}
-Note that the argument \texttt{edge.length} is set to \texttt{1} by default.
-In general we highly recommend to set all arguments but \texttt{x0} to a default value (this really helps the speeding up the \texttt{dads} function).
+Note that in this function definition the arguments \texttt{x0} and \texttt{edge.length} have a default value set to \texttt{0} and \texttt{1} respectively.
+In practice, these arguments are effectively set to the correct values in the \texttt{dads} internal function (i.e.~whatever \texttt{x0} and \texttt{edge.length} are at that specific time of the process) but providing a default can help speed up the algorithms (specifically all the internal checks).
 \end{quote}
@@
 \begin{Highlighting}[]
-\CommentTok{## A wrongly formated process}
-\NormalTok{invalid.process <-}\StringTok{ }\ControlFlowTok{function}\NormalTok{(x0) \{}
+\CommentTok{\#\# A wrongly formated process}
+\NormalTok{invalid.process \textless{}{-}}\StringTok{ }\ControlFlowTok{function}\NormalTok{(}\DataTypeTok{x0 =} \DecValTok{0}\NormalTok{) \{}
     \KeywordTok{return}\NormalTok{(}\KeywordTok{rnorm}\NormalTok{(}\DecValTok{1}\NormalTok{, }\DataTypeTok{mean =}\NormalTok{ x0))}
@@
 
-The \texttt{dads} package proposes inbuilt processes, namely a multidimensional Brownian motion (\texttt{BM.process}) or a a multidimensional Ornstein-Uhlenbeck process (\texttt{OU.process}).
+\hypertarget{using-a-process-in-dads}{%
+\subsection{\texorpdfstring{Using a \texttt{"process"} in \texttt{dads}}{Using a "process" in dads}}\label{using-a-process-in-dads}}
+
+You can design your own process as a function (as long as it has a valid syntax).
+Alternatively, the \texttt{dads} package proposes inbuilt processes, namely a multidimensional Brownian motion (\texttt{BM.process}) or a a multidimensional Ornstein-Uhlenbeck process (\texttt{OU.process}).
 You can find the list of implemented process by looking at the \texttt{?trait.process} manual page in \texttt{R}.
@@
 \begin{Highlighting}[]
-\CommentTok{## Creating a trait object}
-\NormalTok{my_trait_object <-}\StringTok{ }\KeywordTok{make.traits}\NormalTok{(}\DataTypeTok{process =}\NormalTok{ BM.process)}
+\CommentTok{\#\# Creating a trait object}
+\NormalTok{my\_trait\_object \textless{}{-}}\StringTok{ }\KeywordTok{make.traits}\NormalTok{(}\DataTypeTok{process =}\NormalTok{ BM.process)}
 \end{Highlighting}
@@
 \begin{Highlighting}[]
-\CommentTok{## The class of the object}
-\KeywordTok{class}\NormalTok{(my_trait_object)}
+\CommentTok{\#\# The class of the object}
+\KeywordTok{class}\NormalTok{(my\_trait\_object)}
 \end{Highlighting}
@@
 \begin{Highlighting}[]
-\CommentTok{## What's in it?}
-\NormalTok{my_trait_object}
+\CommentTok{\#\# What\textquotesingle{}s in it?}
+\NormalTok{my\_trait\_object}
 \end{Highlighting}
@@
 \begin{Highlighting}[]
-\CommentTok{## What does the process looks like}
-\KeywordTok{plot}\NormalTok{(my_trait_object)}
+\CommentTok{\#\# What does the process looks like}
+\KeywordTok{plot}\NormalTok{(my\_trait\_object)}
 \end{Highlighting}
@@
 
-Note that you can see the multiple options for plotting the trait process by looking at \texttt{?plot.dads} manual. Furthermore, you can look at what's actually in the object using:
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{## What's actually in that object?}
-\KeywordTok{print.dads}\NormalTok{(my_trait_object, }\DataTypeTok{all =} \OtherTok{TRUE}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## $A
-## $A$process
-## function (x0, edge.length = 1, Sigma = diag(length(x0)), ...) 
+Note that you can see the multiple options for plotting the trait process by looking at \texttt{?plot.dads} manual.
+Furthermore, you can look at what's actually in the object using this specific syntax (this applies to every object handled by the \texttt{dads} package):
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# What\textquotesingle{}s actually in that object?}
+\KeywordTok{print.dads}\NormalTok{(my\_trait\_object, }\DataTypeTok{all =} \OtherTok{TRUE}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## $main
+## $main$A
+## $main$A$process
+## function (x0 = 0, edge.length = 1, Sigma = diag(length(x0)), 
+##     ...) 
 ## {
-##     return(t(MASS::mvrnorm(n = 1, mu = x0, Sigma = Sigma * edge.length, 
-##         ...)))
+##     return(t(MASS::mvrnorm(n = 1, mu = x0, Sigma = sqrt(Sigma^2 * 
+##         edge.length), ...)))
 ## }
-## <bytecode: 0x7f7eeee57f58>
+## <bytecode: 0x55a505929ed0>
 ## <environment: namespace:dads>
 ## 
-## $A$start
+## $main$A$start
 ## [1] 0
 ## 
-## $A$trait_id
+## $main$A$trait_id
 ## [1] 1
+## 
+## 
+## 
+## $background
+## NULL
 \end{verbatim}
@@
 \begin{Highlighting}[]
-\CommentTok{## 4 traits: two BM, one OU and one normal non process}
-\NormalTok{four_traits <-}\StringTok{ }\KeywordTok{make.traits}\NormalTok{(}\DataTypeTok{process =} \KeywordTok{c}\NormalTok{(BM.process,}
+\CommentTok{\#\# 4 traits: two BM, one OU and one normal non process}
+\NormalTok{four\_traits \textless{}{-}}\StringTok{ }\KeywordTok{make.traits}\NormalTok{(}\DataTypeTok{process =} \KeywordTok{c}\NormalTok{(BM.process,}
 \NormalTok{                                       BM.process,}
@@
 \NormalTok{                                       no.process))}
-\NormalTok{four_traits}
+\NormalTok{four\_traits}
 \end{Highlighting}
@@
 \begin{Highlighting}[]
-\CommentTok{## Plot options (4 plots in one window)}
+\CommentTok{\#\# Plot options (4 plots in one window)}
 \KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow =} \KeywordTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{,}\DecValTok{2}\NormalTok{))}
-\KeywordTok{plot}\NormalTok{(four_traits, }\DataTypeTok{trait =} \DecValTok{1}\NormalTok{)}
-\KeywordTok{plot}\NormalTok{(four_traits, }\DataTypeTok{trait =} \DecValTok{2}\NormalTok{)}
-\KeywordTok{plot}\NormalTok{(four_traits, }\DataTypeTok{trait =} \DecValTok{3}\NormalTok{)}
-\KeywordTok{plot}\NormalTok{(four_traits, }\DataTypeTok{trait =} \DecValTok{4}\NormalTok{)}
+\KeywordTok{plot}\NormalTok{(four\_traits, }\DataTypeTok{trait =} \DecValTok{1}\NormalTok{)}
+\KeywordTok{plot}\NormalTok{(four\_traits, }\DataTypeTok{trait =} \DecValTok{2}\NormalTok{)}
+\KeywordTok{plot}\NormalTok{(four\_traits, }\DataTypeTok{trait =} \DecValTok{3}\NormalTok{)}
+\KeywordTok{plot}\NormalTok{(four\_traits, }\DataTypeTok{trait =} \DecValTok{4}\NormalTok{)}
 \end{Highlighting}
@@
 \begin{Highlighting}[]
-\CommentTok{## Multidimensional Brownian motion}
+\CommentTok{\#\# Multidimensional Brownian motion}
 \KeywordTok{make.traits}\NormalTok{(BM.process, }\DataTypeTok{n =} \DecValTok{3}\NormalTok{, }\DataTypeTok{start =} \DecValTok{1}\NormalTok{)}
@@
 \begin{Highlighting}[]
-\CommentTok{## Multidimensional Brownian motion}
+\CommentTok{\#\# Multidimensional Brownian motion}
 \KeywordTok{make.traits}\NormalTok{(BM.process, }\DataTypeTok{n =} \DecValTok{3}\NormalTok{, }\DataTypeTok{start =} \KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{,}\DecValTok{2}\NormalTok{,}\DecValTok{3}\NormalTok{))}
@@
 \begin{Highlighting}[]
-\CommentTok{## two 3D processes (BM and OU)}
+\CommentTok{\#\# two 3D processes (BM and OU)}
 \KeywordTok{make.traits}\NormalTok{(}\KeywordTok{c}\NormalTok{(BM.process, OU.process), }\DataTypeTok{n =} \DecValTok{3}\NormalTok{)}
@@
 \begin{Highlighting}[]
-\CommentTok{## one 1D processes (BM) and one 4D process (OU)}
+\CommentTok{\#\# one 1D processes (BM) and one 4D process (OU)}
 \KeywordTok{make.traits}\NormalTok{(}\KeywordTok{c}\NormalTok{(BM.process, OU.process), }\DataTypeTok{n =} \KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{, }\DecValTok{4}\NormalTok{))}
@@
 \begin{Highlighting}[]
-\CommentTok{## two 3D processes (BM and OU) starting with 1}
+\CommentTok{\#\# two 3D processes (BM and OU) starting with 1}
 \KeywordTok{make.traits}\NormalTok{(}\KeywordTok{c}\NormalTok{(BM.process, OU.process), }\DataTypeTok{n =} \DecValTok{3}\NormalTok{, }\DataTypeTok{start =} \DecValTok{1}\NormalTok{)}
@@
 \begin{Highlighting}[]
-\CommentTok{## two 3D processes (BM and OU) starting with values 1 to 6}
+\CommentTok{\#\# two 3D processes (BM and OU) starting with values 1 to 6}
 \KeywordTok{make.traits}\NormalTok{(}\KeywordTok{c}\NormalTok{(BM.process, OU.process), }\DataTypeTok{n =} \DecValTok{3}\NormalTok{, }\DataTypeTok{start =} \DecValTok{1}\OperatorTok{:}\DecValTok{6}\NormalTok{)}
@@
 \begin{Highlighting}[]
-\CommentTok{## two 3D processes (BM and OU) with the two first ones starting}
-\CommentTok{## with 1 and the 4 other ones with the default (0)}
+\CommentTok{\#\# two 3D processes (BM and OU) with the two first ones starting}
+\CommentTok{\#\# with 1 and the 4 other ones with the default (0)}
 \KeywordTok{make.traits}\NormalTok{(}\KeywordTok{c}\NormalTok{(BM.process, OU.process), }\DataTypeTok{n =} \DecValTok{3}\NormalTok{, }\DataTypeTok{start =} \KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{,}\DecValTok{1}\NormalTok{))}
@@
 
+\hypertarget{what-even-is-a-trait}{%
+\subsection{What even is a trait?}\label{what-even-is-a-trait}}
+
+The definition of what a trait is can vary quiet a lot depending on the context of an analysis and the field.
+Because it would be impossible to accommodate all definitions in \texttt{dads} we had to go with an arbitrary one: a trait is whatever you define as a trait!
+A trait can be uni-dimensional as the measurement of a feature of an organism (e.g.~leaf surface, femur length, etc\ldots) but can also be multi-dimensional description of a feature, for example in 3D geometric morphometric, a trait could be defined as ``position of landmark X'' (which will be a trait with three dimensions, \emph{x}, \emph{y} and \emph{z}) or in ecology, the location of a plant can be a single 2D trait (expressed as latitude and longitude).
+In \texttt{dads}, the \textbf{process} corresponds to this trait definition (e.g.~a process can be of n-dimensions and represents one organisms feature) and the \textbf{traits} represents the number of dimensions in total.
+So in the examples above, this is how the following traits are interpreted by \texttt{dads}:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# Three traits with one process:}
+\KeywordTok{make.traits}\NormalTok{(BM.process, }\DataTypeTok{n =} \DecValTok{3}\NormalTok{, }\DataTypeTok{start =} \KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{,}\DecValTok{2}\NormalTok{,}\DecValTok{3}\NormalTok{))}
+\CommentTok{\#\# Six traits with two processes:}
+\KeywordTok{make.traits}\NormalTok{(}\KeywordTok{c}\NormalTok{(BM.process, OU.process), }\DataTypeTok{n =} \DecValTok{3}\NormalTok{)}
+\CommentTok{\#\# Five traits with two processes}
+\KeywordTok{make.traits}\NormalTok{(}\KeywordTok{c}\NormalTok{(BM.process, OU.process), }\DataTypeTok{n =} \KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{, }\DecValTok{4}\NormalTok{))}
+\end{Highlighting}
+\end{Shaded}
+
 \hypertarget{extra-argument-for-the-processes-with-process.args}{%
@@
 \begin{Highlighting}[]
-\CommentTok{## no process trait using the normal distribution (default)}
+\CommentTok{\#\# no process trait using the normal distribution (default)}
 \KeywordTok{make.traits}\NormalTok{(no.process, }\DataTypeTok{process.args =} \KeywordTok{list}\NormalTok{(}\DataTypeTok{fun =}\NormalTok{ rnorm))}
@@
 \begin{Highlighting}[]
-\CommentTok{## no process trait using the uniform distribution}
-\CommentTok{## bounded between 1 and 100}
+\CommentTok{\#\# no process trait using the uniform distribution}
+\CommentTok{\#\# bounded between 1 and 100}
 \KeywordTok{make.traits}\NormalTok{(no.process, }\DataTypeTok{process.args =} \KeywordTok{list}\NormalTok{(}\DataTypeTok{fun =}\NormalTok{ runif, }\DataTypeTok{min =} \DecValTok{1}\NormalTok{, }\DataTypeTok{max =} \DecValTok{100}\NormalTok{))}
@@
 \begin{Highlighting}[]
-\CommentTok{## Two traits with no process:one normal and one uniform (1,100)}
+\CommentTok{\#\# Two traits with no process:one normal and one uniform (1,100)}
 \KeywordTok{make.traits}\NormalTok{(}\DataTypeTok{process      =} \KeywordTok{c}\NormalTok{(no.process, no.process),}
@@
 \begin{Highlighting}[]
-\CommentTok{## Three traits with no process:}
-\CommentTok{## one default, one lognormal and one uniform (1,100)}
+\CommentTok{\#\# Three traits with no process:}
+\CommentTok{\#\# one default, one lognormal and one uniform (1,100)}
 \KeywordTok{make.traits}\NormalTok{(}\DataTypeTok{process      =} \KeywordTok{c}\NormalTok{(no.process, no.process, no.process),}
-            \DataTypeTok{process.args =} \KeywordTok{list}\NormalTok{(}\CommentTok{## Extra arguments for the first process (none)}
+            \DataTypeTok{process.args =} \KeywordTok{list}\NormalTok{(}\CommentTok{\#\# Extra arguments for the first process (none)}
                                 \KeywordTok{list}\NormalTok{(}\OtherTok{NULL}\NormalTok{),}
-                                \CommentTok{## Extra arguments for the second process}
+                                \CommentTok{\#\# Extra arguments for the second process}
                                 \KeywordTok{list}\NormalTok{(}\DataTypeTok{fun =}\NormalTok{ rlnorm),}
-                                \CommentTok{## Extra arguments for the third process}
+                                \CommentTok{\#\# Extra arguments for the third process}
                                 \KeywordTok{list}\NormalTok{(}\DataTypeTok{fun =}\NormalTok{ runif, }\DataTypeTok{min =} \DecValTok{1}\NormalTok{, }\DataTypeTok{max =} \DecValTok{100}\NormalTok{)))}
@@
 \begin{Highlighting}[]
-\CommentTok{## A simple trait with a proper name}
-\NormalTok{simple_trait <-}\StringTok{ }\KeywordTok{make.traits}\NormalTok{(}\DataTypeTok{trait.names =} \StringTok{"1D Brownian Motion"}\NormalTok{)}
-\NormalTok{simple_trait}
+\CommentTok{\#\# A simple trait with a proper name}
+\NormalTok{simple\_trait \textless{}{-}}\StringTok{ }\KeywordTok{make.traits}\NormalTok{(}\DataTypeTok{trait.names =} \StringTok{"1D Brownian Motion"}\NormalTok{)}
+\NormalTok{simple\_trait}
 \end{Highlighting}
@@
 \begin{Highlighting}[]
-\CommentTok{## Three named traits with no process:}
-\CommentTok{## one default, one lognormal and one uniform (1,100)}
+\CommentTok{\#\# Three named traits with no process:}
+\CommentTok{\#\# one default, one lognormal and one uniform (1,100)}
 \KeywordTok{make.traits}\NormalTok{(}\DataTypeTok{process      =} \KeywordTok{c}\NormalTok{(no.process, no.process, no.process),}
-            \DataTypeTok{process.args =} \KeywordTok{list}\NormalTok{(}\CommentTok{## Extra arguments for the first process (none)}
+            \DataTypeTok{process.args =} \KeywordTok{list}\NormalTok{(}\CommentTok{\#\# Extra arguments for the first process (none)}
                                 \KeywordTok{list}\NormalTok{(}\OtherTok{NULL}\NormalTok{),}
-                                \CommentTok{## Extra arguments for the second process}
+                                \CommentTok{\#\# Extra arguments for the second process}
                                 \KeywordTok{list}\NormalTok{(}\DataTypeTok{fun =}\NormalTok{ rlnorm),}
-                                \CommentTok{## Extra arguments for the third process}
+                                \CommentTok{\#\# Extra arguments for the third process}
                                 \KeywordTok{list}\NormalTok{(}\DataTypeTok{fun =}\NormalTok{ runif, }\DataTypeTok{min =} \DecValTok{1}\NormalTok{, }\DataTypeTok{max =} \DecValTok{100}\NormalTok{)),}
-            \CommentTok{## Naming each trait}
+            \CommentTok{\#\# Naming each trait}
             \DataTypeTok{trait.names  =} \KeywordTok{c}\NormalTok{(}\StringTok{"Normal"}\NormalTok{, }\StringTok{"LogNormal"}\NormalTok{, }\StringTok{"Uniform(1,100)"}\NormalTok{))}
@@
 \begin{Highlighting}[]
-\CommentTok{## Creating on simple default Brownian motion}
-\NormalTok{one_process <-}\StringTok{ }\KeywordTok{make.traits}\NormalTok{(}\DataTypeTok{trait.names =} \StringTok{"BM"}\NormalTok{)}
-
-\CommentTok{## Creating a new trait (a 3D OU.process)}
-\CommentTok{## and adding the previous one}
-\NormalTok{two_processes <-}\StringTok{ }\KeywordTok{make.traits}\NormalTok{(OU.process, }\DataTypeTok{n =} \DecValTok{3}\NormalTok{, }\DataTypeTok{add =}\NormalTok{ one_process,}
+\CommentTok{\#\# Creating on simple default Brownian motion}
+\NormalTok{one\_process \textless{}{-}}\StringTok{ }\KeywordTok{make.traits}\NormalTok{(}\DataTypeTok{trait.names =} \StringTok{"BM"}\NormalTok{)}
+
+\CommentTok{\#\# Creating a new trait (a 3D OU.process)}
+\CommentTok{\#\# and adding the previous one}
+\NormalTok{two\_processes \textless{}{-}}\StringTok{ }\KeywordTok{make.traits}\NormalTok{(OU.process, }\DataTypeTok{n =} \DecValTok{3}\NormalTok{, }\DataTypeTok{add =}\NormalTok{ one\_process,}
                              \DataTypeTok{trait.names =} \StringTok{"3D OU"}\NormalTok{)}
 
-\CommentTok{## Only one process}
-\NormalTok{one_process}
+\CommentTok{\#\# Only one process}
+\NormalTok{one\_process}
 \end{Highlighting}
@@
 \begin{Highlighting}[]
-\CommentTok{## The two processes}
-\NormalTok{two_processes}
+\CommentTok{\#\# The two processes}
+\NormalTok{two\_processes}
 \end{Highlighting}
@@
 
+\hypertarget{using-a-background-trait}{%
+\section{Using a background trait}\label{using-a-background-trait}}
+
+\texttt{traits} objects also allow a background trait to be used when traits are simulated (\protect\hyperlink{bdalgorithm}{step 3 here}).
+This basically allows traits to be simulating for \emph{all} tips whenever a trait is generated for one tip.
+This can be useful for keeping track of trait values along the simulation (\emph{cf} just at bifurcating nodes).
+The \texttt{background} argument takes any output from the \texttt{make.traits} function in a nested way:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# Generating a default BM trait:}
+\NormalTok{BM\_trait \textless{}{-}}\StringTok{ }\KeywordTok{make.traits}\NormalTok{()}
+\CommentTok{\#\# Generating an OU trait with a background BM trait}
+\NormalTok{my\_trait \textless{}{-}}\StringTok{ }\KeywordTok{make.traits}\NormalTok{(}\DataTypeTok{process =}\NormalTok{ OU.process, }\DataTypeTok{background =}\NormalTok{ BM\_trait) }
+\end{Highlighting}
+\end{Shaded}
+
+Note that technically you can nest as many background traits as you want (e.g.~\texttt{make.traits(background\ =\ make.traits(background\ =\ make.traits(...)))} is valid).
+However, you should always make sure that the background trait has the same dimensions as the normal (main) trait.
+When using a trait with background, your tree will have internal singleton nodes (i.e.~nodes linking to one ancestor and only one descendant).
+You can remove these nodes using the \protect\hyperlink{dropthings}{\texttt{drop.things} function}.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{set.seed}\NormalTok{(}\DecValTok{1}\NormalTok{)}
+\CommentTok{\#\# Generating a pure birth tree with the background trait}
+\NormalTok{tree\_bkg \textless{}{-}}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{stop.rule =} \KeywordTok{list}\NormalTok{(}\DataTypeTok{max.taxa =} \DecValTok{20}\NormalTok{),}
+                 \DataTypeTok{traits    =}\NormalTok{ my\_trait)}
+\CommentTok{\#\# This tree has many internal singleton nodes}
+\KeywordTok{plot}\NormalTok{(tree\_bkg)}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-41-1.pdf}
+
+\hypertarget{saving-trait-values-at-different-time-steps}{%
+\section{Saving trait values at different time steps}\label{saving-trait-values-at-different-time-steps}}
+
+Alternatively to having a background trait, you can also simulate a tree by generating traits at specific time steps with the \texttt{save.steps} option in \texttt{dads}.
+This will apply the \texttt{traits} object to all lineages currently alive at the required time steps.
+These time steps can be either regular by providing a single numeric value; e.g.~\texttt{save.steps\ =\ 0.1} will get a snapshot of the trait values every 0.1 units of time.
+Or specific, by providing a specific set of values; e.g.~\texttt{save.steps\ =\ c(1,\ 1.2,\ 3)} will make a snapshot of the trait values at the required time steps.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{set.seed}\NormalTok{(}\DecValTok{123}\NormalTok{)}
+\CommentTok{\#\# Generating a birth{-}death tree with a BM trait and saving steps at specific times}
+\NormalTok{tree\_steps \textless{}{-}}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{stop.rule  =} \KeywordTok{list}\NormalTok{(}\DataTypeTok{max.time =} \DecValTok{3}\NormalTok{),}
+                   \DataTypeTok{bd.params  =} \KeywordTok{list}\NormalTok{(}\DataTypeTok{speciation =} \DecValTok{1}\NormalTok{, }\DataTypeTok{extinction =} \FloatTok{0.1}\NormalTok{),}
+                   \DataTypeTok{traits     =} \KeywordTok{make.traits}\NormalTok{(),}
+                   \DataTypeTok{save.steps =} \KeywordTok{c}\NormalTok{(}\DecValTok{1}\OperatorTok{/}\DecValTok{3}\NormalTok{, }\DecValTok{1}\NormalTok{, }\DecValTok{2}\NormalTok{))}
+\CommentTok{\#\# This also creates internal singleton nodes}
+\KeywordTok{plot}\NormalTok{(tree\_steps)}
+\KeywordTok{abline}\NormalTok{(}\DataTypeTok{v =} \DecValTok{3} \OperatorTok{{-}}\StringTok{ }\KeywordTok{c}\NormalTok{(}\DecValTok{1}\OperatorTok{/}\DecValTok{3}\NormalTok{, }\DecValTok{1}\NormalTok{, }\DecValTok{2}\NormalTok{), }\DataTypeTok{lwd =} \FloatTok{0.5}\NormalTok{, }\DataTypeTok{col =} \StringTok{"grey"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-42-1.pdf}
+
+\hypertarget{traits-implemented-in-dads}{%
+\section{\texorpdfstring{Traits implemented in \texttt{dads}}{Traits implemented in dads}}\label{traits-implemented-in-dads}}
+
+If you don't want to design your own trait process, you can use one of the following trait processes that are currently implemented in \texttt{dads}.
+You can find more information about their many options using their specific manuals in R or the generic \texttt{?trait.process}:
+
+\begin{itemize}
+\tightlist
+\item
+  \texttt{BM.process}: this is the well famous \href{https://en.wikipedia.org/wiki/Brownian_motion}{Brownian motion process}.
+\end{itemize}
+
+\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-43-1.pdf}
+
+\begin{itemize}
+\tightlist
+\item
+  \texttt{OU.process}: this is the equally famous \href{https://en.wikipedia.org/wiki/Ornstein\%E2\%80\%93Uhlenbeck_process}{Ornstein--Uhlenbeck process}.
+\end{itemize}
+
+\begin{verbatim}
+## Warning in title(...): conversion failure on 'The Ornstein–Uhlenbeck process'
+## in 'mbcsToSbcs': dot substituted for <e2>
+\end{verbatim}
+
+\begin{verbatim}
+## Warning in title(...): conversion failure on 'The Ornstein–Uhlenbeck process'
+## in 'mbcsToSbcs': dot substituted for <80>
+\end{verbatim}
+
+\begin{verbatim}
+## Warning in title(...): conversion failure on 'The Ornstein–Uhlenbeck process'
+## in 'mbcsToSbcs': dot substituted for <93>
+\end{verbatim}
+
+\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-44-1.pdf}
+
+\begin{itemize}
+\tightlist
+\item
+  \texttt{no.process}: this process has\ldots{} no process.
+In other words, this is a non time dependent process (the simulated value does not depends on the ancestors' value nor the branch length).
+It's basically a place holder for a random sampling function like \texttt{rnorm} (default), \texttt{runif}, \texttt{rlnorm}, etc\ldots{}
+\end{itemize}
+
+\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-45-1.pdf}
+
+\begin{itemize}
+\tightlist
+\item
+  \texttt{multi.peak.process}: this process is a modified version of the \texttt{OU.process} that can take multiple local optima.
+The default OU process has one optimum towards which the values are drawn with the alpha parameter (the elastic band) which is usually 0.
+However, with this \texttt{multi.peak.process} we can set multiple values towards which values can be attracted with the same alpha parameter.
+\end{itemize}
+
+\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-46-1.pdf}
+
+\begin{itemize}
+\tightlist
+\item
+  \texttt{repulsion.process}: this is a modified version of the \texttt{BM.process} where instead of accumulating normally through time, new trait values are more likely to be different than their ancestor following a \texttt{repulsion} parameter.
+\end{itemize}
+
+\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-47-1.pdf}
+
 \hypertarget{testing-the-traits-with-test}{%
@@
 
-This bit is more for development.
+\textbf{This bit is more for development.}
 We highly suggest leaving \texttt{test\ =\ TRUE} so that \texttt{make.traits} returns an error if a process or its additional arguments (\texttt{process.args}) are not formatted correctly.
-\texttt{make.traits} will error if the trait cannot be directly passed to \texttt{dads}.
+\texttt{make.traits} will send an error if the trait cannot be directly passed to \texttt{dads}.
 However, in some specific cases (again, probably mainly for development and debugging) it could be useful to skip the tests using \texttt{test\ =\ FALSE}.
@@
 You can be imaginative and creative when designing your own process but here are two detailed example functions for a unidimensional Brownian Motion and Ornstein-Uhlenbeck process that you can use for a start (or not).
-Remember it is good practice for \texttt{dads} processes to set all the arguments but \texttt{x0} with default values (just in case).
-Also, note that the functions below are not equal to the already implemented \texttt{BM.process} and \texttt{OU.process} but are rather generalised/simplified version that you can use as a template
+Remember it is good practice for \texttt{dads} processes to set all the arguments with default values (just in case).
+
+\begin{quote}
+Note that the functions below are not equal to the already implemented \texttt{BM.process} and \texttt{OU.process} but an easier to edit version that you can use as a template:
+\end{quote}
 
@@
 \begin{Highlighting}[]
-\CommentTok{## A simple Brownian motion process}
-\NormalTok{my.BM.process <-}\StringTok{ }\ControlFlowTok{function}\NormalTok{(x0, }\DataTypeTok{edge.length =} \DecValTok{1}\NormalTok{, }\DataTypeTok{sd =} \DecValTok{1}\NormalTok{, ...) \{}
-    \CommentTok{## Drawing a random number from a normal distribution}
-    \CommentTok{## with x0 as the and a given standard deviation}
-    \CommentTok{## and depending on branch (edge) length}
-\NormalTok{    result <-}\StringTok{ }\KeywordTok{rnorm}\NormalTok{(}\DataTypeTok{n =} \DecValTok{1}\NormalTok{, }\DataTypeTok{mean =}\NormalTok{ x0, }\DataTypeTok{sd =}\NormalTok{ sd }\OperatorTok{*}\StringTok{ }\NormalTok{edge.length)}
-
-    \CommentTok{## Return the number}
+\CommentTok{\#\# A simple Brownian motion process}
+\NormalTok{my.BM.process \textless{}{-}}\StringTok{ }\ControlFlowTok{function}\NormalTok{(}\DataTypeTok{x0 =} \DecValTok{0}\NormalTok{, }\DataTypeTok{edge.length =} \DecValTok{1}\NormalTok{, }\DataTypeTok{sd =} \DecValTok{1}\NormalTok{, ...) \{}
+    \CommentTok{\#\# Drawing a random number from a normal distribution}
+    \CommentTok{\#\# with x0 as the and a given standard deviation}
+    \CommentTok{\#\# and depending on branch (edge) length}
+\NormalTok{    result \textless{}{-}}\StringTok{ }\KeywordTok{rnorm}\NormalTok{(}\DataTypeTok{n =} \DecValTok{1}\NormalTok{, }\DataTypeTok{mean =}\NormalTok{ x0, }\DataTypeTok{sd =} \KeywordTok{sqrt}\NormalTok{(sd}\OperatorTok{\^{}}\DecValTok{2} \OperatorTok{*}\StringTok{ }\NormalTok{edge.length))}
+
+    \CommentTok{\#\# Return the number}
     \KeywordTok{return}\NormalTok{(result)}
@@
 \begin{Highlighting}[]
-\CommentTok{## A simple Ornstein-Uhlenbeck motion process}
-\NormalTok{my.OU.process <-}\StringTok{ }\ControlFlowTok{function}\NormalTok{(x0, }\DataTypeTok{edge.length =} \DecValTok{1}\NormalTok{, }\DataTypeTok{var =} \DecValTok{1}\NormalTok{, }\DataTypeTok{alpha =} \DecValTok{1}\NormalTok{, ...) \{}
-    \CommentTok{## Calculate the mean based on alpha}
-\NormalTok{    mean <-}\StringTok{ }\NormalTok{x0 }\OperatorTok{*}\StringTok{ }\KeywordTok{exp}\NormalTok{(}\OperatorTok{-}\NormalTok{alpha)}
-    \CommentTok{## Calculate the standard deviation based on alpha and the variance}
-\NormalTok{    sd <-}\StringTok{ }\KeywordTok{sqrt}\NormalTok{(var}\OperatorTok{/}\NormalTok{(}\DecValTok{2} \OperatorTok{*}\StringTok{ }\NormalTok{alpha) }\OperatorTok{*}\StringTok{ }\NormalTok{(}\DecValTok{1} \OperatorTok{-}\StringTok{ }\KeywordTok{exp}\NormalTok{(}\OperatorTok{-}\DecValTok{2} \OperatorTok{*}\StringTok{ }\NormalTok{alpha)))}
-    \CommentTok{## Draw a random number from a normal distribution}
-    \CommentTok{## using this mean and standard deviation}
-    \CommentTok{## and depending on branch (edge) length}
-\NormalTok{    result <-}\StringTok{ }\KeywordTok{rnorm}\NormalTok{(}\DataTypeTok{n =} \DecValTok{1}\NormalTok{, }\DataTypeTok{mean =}\NormalTok{ mean, }\DataTypeTok{sd =}\NormalTok{ sd }\OperatorTok{*}\StringTok{ }\NormalTok{branch.length)}
-
-    \CommentTok{## Return the number}
+\CommentTok{\#\# A simple Ornstein{-}Uhlenbeck motion process}
+\NormalTok{my.OU.process \textless{}{-}}\StringTok{ }\ControlFlowTok{function}\NormalTok{(}\DataTypeTok{x0 =} \DecValTok{0}\NormalTok{, }\DataTypeTok{edge.length =} \DecValTok{1}\NormalTok{, }\DataTypeTok{var =} \DecValTok{1}\NormalTok{, }\DataTypeTok{alpha =} \DecValTok{1}\NormalTok{, ...) \{}
+    \CommentTok{\#\# Calculate the mean based on alpha}
+\NormalTok{    mean \textless{}{-}}\StringTok{ }\NormalTok{x0 }\OperatorTok{*}\StringTok{ }\KeywordTok{exp}\NormalTok{(}\OperatorTok{{-}}\NormalTok{alpha)}
+    \CommentTok{\#\# Calculate the standard deviation based on alpha and the variance}
+\NormalTok{    sd \textless{}{-}}\StringTok{ }\KeywordTok{sqrt}\NormalTok{(var}\OperatorTok{/}\NormalTok{(}\DecValTok{2} \OperatorTok{*}\StringTok{ }\NormalTok{alpha) }\OperatorTok{*}\StringTok{ }\NormalTok{(}\DecValTok{1} \OperatorTok{{-}}\StringTok{ }\KeywordTok{exp}\NormalTok{(}\OperatorTok{{-}}\DecValTok{2} \OperatorTok{*}\StringTok{ }\NormalTok{alpha)))}
+    \CommentTok{\#\# Draw a random number from a normal distribution}
+    \CommentTok{\#\# using this mean and standard deviation}
+    \CommentTok{\#\# and depending on branch (edge) length}
+\NormalTok{    result \textless{}{-}}\StringTok{ }\KeywordTok{rnorm}\NormalTok{(}\DataTypeTok{n =} \DecValTok{1}\NormalTok{, }\DataTypeTok{mean =}\NormalTok{ mean, }\DataTypeTok{sd =} \KeywordTok{sqrt}\NormalTok{(sd}\OperatorTok{\^{}}\DecValTok{2} \OperatorTok{*}\StringTok{ }\NormalTok{edge.length))}
+
+    \CommentTok{\#\# Return the number}
     \KeywordTok{return}\NormalTok{(result)}
@@
 \hypertarget{makemodifiers}{%
-\chapter{\texorpdfstring{Making modifiers with \texttt{make.modifiers()}}{Making modifiers with make.modifiers()}}\label{makemodifiers}}
+\chapter{Modifying the birth-death process}\label{makemodifiers}}
 
@@
 \begin{Highlighting}[]
-\CommentTok{## Making a default modifier (no modification)}
-\NormalTok{my_default_modifiers <-}\StringTok{ }\KeywordTok{make.modifiers}\NormalTok{()}
-\NormalTok{my_default_modifiers}
+\CommentTok{\#\# Making a default modifier (no modification)}
+\NormalTok{my\_default\_modifiers \textless{}{-}}\StringTok{ }\KeywordTok{make.modifiers}\NormalTok{()}
+\NormalTok{my\_default\_modifiers}
 \end{Highlighting}
@@
 However, contrary to \texttt{"traits"}, you cannot plot \texttt{"modifiers"}.
+
+\hypertarget{the-default-modifier-how-the-process-is-working}{%
+\section{The default modifier (how the process is working)}\label{the-default-modifier-how-the-process-is-working}}
+
+The modifiers modify the core of the birth death process as implemented in \texttt{dads}.
+
+By default, the birth-death process in \texttt{dads} uses this modifier:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# What is actually in the default modifier?}
+\KeywordTok{print}\NormalTok{(}\KeywordTok{make.modifiers}\NormalTok{(), }\DataTypeTok{all =} \OtherTok{TRUE}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## $waiting
+## $waiting$fun
+## function (bd.params, lineage = NULL, trait.values = NULL, modify.fun = NULL) 
+## {
+##     return(rexp(1, sum(lineage$n * (bd.params$speciation + bd.params$extinction))))
+## }
+## <bytecode: 0x55a507be7fe8>
+## <environment: namespace:dads>
+## 
+## $waiting$internal
+## NULL
+## 
+## 
+## $selecting
+## $selecting$fun
+## function (bd.params, lineage = NULL, trait.values = NULL, modify.fun = NULL) 
+## {
+##     return(sample(lineage$n, 1))
+## }
+## <bytecode: 0x55a507c69968>
+## <environment: namespace:dads>
+## 
+## $selecting$internal
+## NULL
+## 
+## 
+## $speciating
+## $speciating$fun
+## function (bd.params, lineage = NULL, trait.values = NULL, modify.fun = NULL) 
+## {
+##     return(runif(1) < (bd.params$speciation/(bd.params$speciation + 
+##         bd.params$extinction)))
+## }
+## <bytecode: 0x55a507c7e050>
+## <environment: namespace:dads>
+## 
+## $speciating$internal
+## NULL
+## 
+## 
+## $call
+## $call$waiting
+## $call$waiting$fun
+## [1] "default"
+## 
+## 
+## $call$selecting
+## $call$selecting$fun
+## [1] "default"
+## 
+## 
+## $call$speciating
+## $call$speciating$fun
+## [1] "default"
+\end{verbatim}
+
+This contains a lot of information, much of it is actually not useful for using the modular aspects of \texttt{dads} at a high user level.
+I.e.
+unless you want to code very specific things, you won't need most of the information.
+The essential are these three functions in the elements named \texttt{"waiting\$fun"}, \texttt{"selecting\$fun"} and \texttt{"speciating\$fun"}:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  \texttt{"waiting\$fun"} is the branch length function (defined in details below) which returns a randomly drawn number from an exponential distribution with the rate of the number of taxa multiplied by the speciation and extinction rate: \(n \times (\lambda + \mu)\) (here is uses the extra function \texttt{sum} for internal modularity reasons).
+\textbf{This function is responsible for the growth of the length/age of the tree}
+\end{enumerate}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{rexp}\NormalTok{(}\DecValTok{1}\NormalTok{, }\KeywordTok{sum}\NormalTok{(lineage}\OperatorTok{$}\NormalTok{n }\OperatorTok{*}\StringTok{ }\NormalTok{(bd.params}\OperatorTok{$}\NormalTok{speciation }\OperatorTok{+}\StringTok{ }\NormalTok{bd.params}\OperatorTok{$}\NormalTok{extinction)))}
+\end{Highlighting}
+\end{Shaded}
+
+With \texttt{lineage\$n} being the number of lineages and \texttt{bd.params\$speciation} and \texttt{bd.params\$extinction} the speciation and extinction parameters (these specific terms are defined in details below).
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\setcounter{enumi}{1}
+\tightlist
+\item
+  \texttt{"selecting\$fun"} is the selection function (defined in details below) which, after the waiting time defined above returns a randomly selected lineage (a tip) among the existing ones.
+\textbf{This function is responsible for the branch selection}.
+\end{enumerate}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{sample}\NormalTok{(lineage}\OperatorTok{$}\NormalTok{n, }\DecValTok{1}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\setcounter{enumi}{2}
+\tightlist
+\item
+  \texttt{"speciating\$fun"} is the speciation function (defined in details below) which randomly draws a number (\texttt{runif(1)}) and depending on the speciation and extinction parameters makes the selected lineage (from 2.) speciate (\texttt{TRUE}) or go extinct (\texttt{FALSE}).
+\textbf{This function is responsible for the speciation or extinction of species}.
+\end{enumerate}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{runif}\NormalTok{(}\DecValTok{1}\NormalTok{) }\OperatorTok{\textless{}}\StringTok{ }\NormalTok{(bd.params}\OperatorTok{$}\NormalTok{speciation}\OperatorTok{/}\NormalTok{(bd.params}\OperatorTok{$}\NormalTok{speciation }\OperatorTok{+}\StringTok{ }\NormalTok{bd.params}\OperatorTok{$}\NormalTok{extinction))}\ErrorTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+We will have a look at all these aspects below in more details which should make things clearer (but you can always refer to the default here for information).
 
@@
 
-By default, branch length (or waiting or growth) is a randomly drawn number from an exponential distribution with the rate of the number of taxa multiplied by the speciation and extinction rate: \(n \times (\lambda + \mu)\) (where \(n\) is the number of taxa currently present in the simulation, \(\lambda\) and \(\mu\) are respectively the speciation and extinction rates).
+By default, branch length (or waiting/growth) is a randomly drawn number from an exponential distribution with the rate of the number of taxa multiplied by the speciation and extinction rate: \(n \times (\lambda + \mu)\) (where \(n\) is the number of taxa currently present in the simulation, \(\lambda\) and \(\mu\) are respectively the speciation and extinction rates).
 This default function is simply called \texttt{branch.length} in \texttt{dads} and can be used as a modifier as follows:
@@
 \begin{Highlighting}[]
-\CommentTok{## Specifying the default modifier}
-\NormalTok{default_modifiers <-}\StringTok{ }\KeywordTok{make.modifiers}\NormalTok{(}\DataTypeTok{branch.length =}\NormalTok{ branch.length)}
-
-\CommentTok{## Setting some parameters for generating trees}
-\NormalTok{bd_params  <-}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{extinction =} \DecValTok{0}\NormalTok{)}
-\NormalTok{stope_rule <-}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{max.living =} \DecValTok{20}\NormalTok{)}
-
-\CommentTok{## Generating a tree with the default branch length parameter}
+\CommentTok{\#\# Specifying the default modifier}
+\NormalTok{default\_modifiers \textless{}{-}}\StringTok{ }\KeywordTok{make.modifiers}\NormalTok{(}\DataTypeTok{branch.length =}\NormalTok{ branch.length)}
+
+\CommentTok{\#\# Setting some parameters for generating trees}
+\NormalTok{bd\_params \textless{}{-}}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{extinction =} \DecValTok{0}\NormalTok{)}
+\NormalTok{stop\_rule \textless{}{-}}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{max.living =} \DecValTok{20}\NormalTok{)}
+
+\CommentTok{\#\# Generating a tree with the default branch length parameter}
 \KeywordTok{set.seed}\NormalTok{(}\DecValTok{0}\NormalTok{)}
-\NormalTok{default_tree <-}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ bd_params,}
-                     \DataTypeTok{stop.rule =}\NormalTok{ stope_rule,}
-                     \DataTypeTok{modifiers =}\NormalTok{ default_modifiers)}
+\NormalTok{default\_tree \textless{}{-}}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ bd\_params,}
+                     \DataTypeTok{stop.rule =}\NormalTok{ stop\_rule,}
+                     \DataTypeTok{modifiers =}\NormalTok{ default\_modifiers)}
 \end{Highlighting}
@@
 \begin{Highlighting}[]
-\CommentTok{## A constant branch length generator}
-\CommentTok{## (note that the output must be numeric, not integer)}
-\NormalTok{constant.brlen <-}\StringTok{ }\ControlFlowTok{function}\NormalTok{() \{}
+\CommentTok{\#\# A constant branch length generator}
+\CommentTok{\#\# (note that the output must be numeric, not integer)}
+\NormalTok{constant.brlen \textless{}{-}}\StringTok{ }\ControlFlowTok{function}\NormalTok{() \{}
     \KeywordTok{return}\NormalTok{(}\KeywordTok{as.numeric}\NormalTok{(}\DecValTok{1}\NormalTok{))}
@@
 
-\CommentTok{## Creating the modifiers object}
-\NormalTok{constant_modifier <-}\StringTok{ }\KeywordTok{make.modifiers}\NormalTok{(}\DataTypeTok{branch.length =}\NormalTok{ constant.brlen)}
-
-\CommentTok{## Generating a new tree with this modifier}
+\CommentTok{\#\# Creating the modifiers object}
+\NormalTok{constant\_modifier \textless{}{-}}\StringTok{ }\KeywordTok{make.modifiers}\NormalTok{(}\DataTypeTok{branch.length =}\NormalTok{ constant.brlen)}
+
+\CommentTok{\#\# Generating a new tree with this modifier}
 \KeywordTok{set.seed}\NormalTok{(}\DecValTok{0}\NormalTok{)}
-\NormalTok{modified_tree <-}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ bd_params,}
-                      \DataTypeTok{stop.rule =}\NormalTok{ stope_rule,}
-                      \DataTypeTok{modifiers =}\NormalTok{ constant_modifier)}
+\NormalTok{modified\_tree \textless{}{-}}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ bd\_params,}
+                      \DataTypeTok{stop.rule =}\NormalTok{ stop\_rule,}
+                      \DataTypeTok{modifiers =}\NormalTok{ constant\_modifier)}
 \end{Highlighting}
@@
 \KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow =} \KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{,}\DecValTok{2}\NormalTok{))}
-\KeywordTok{plot}\NormalTok{(default_tree,  }\DataTypeTok{main =} \StringTok{"Default modifier"}\NormalTok{)}
-\KeywordTok{plot}\NormalTok{(modified_tree, }\DataTypeTok{main =} \StringTok{"Constant branch length}\CharTok{\textbackslash{}n}\StringTok{modifier"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-45-1.pdf}
-
-It is of course to use more complex branch length modifiers that intakes different conditions and specific modification rather than simply always output a value of one.
+\KeywordTok{plot}\NormalTok{(default\_tree,  }\DataTypeTok{main =} \StringTok{"Default modifier"}\NormalTok{)}
+\KeywordTok{plot}\NormalTok{(modified\_tree, }\DataTypeTok{main =} \StringTok{"Constant branch length}\CharTok{\textbackslash{}n}\StringTok{modifier"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-58-1.pdf}
+
+It is of course possible to use more complex branch length modifiers that intakes different conditions and specific modification rather than simply always output a value of one.
 
@@
 
-It is of course possible to design some more advanced function to interact with the birth death process.
 You can create a function for \texttt{branch.length}, \texttt{selection} and \texttt{speciation} that involve any of the following arguments:
@@
 Therefore any argument can be left undeclared or missing and is always handled internally.
-For example, if you did not declare \texttt{n.taxa} as a function argument but are using \texttt{n.taxa} in the function, \texttt{n.taxa} will be detected and treated as a current argument automatically as set accordingly within the birth death process (e.g.
-\texttt{n.taxa} will be set to the current number of taxa every iteration of the process).
+For example, if you did not declare \texttt{lineage\$n} as a function argument but are using \texttt{lineage\$n} in the function, \texttt{lineage\$n} will be detected and treated as a current argument automatically as set accordingly within the birth death process (e.g.~\texttt{lineage\$n} will be set to the current number of taxa every iteration of the process).
 
@@
 I.e.
-for discrete numbers of taxa, the branch length increases by jumps (ten fold):
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{## A more complex binned.branch.length function}
-\NormalTok{increasing.brlen <-}\StringTok{ }\ControlFlowTok{function}\NormalTok{(bd.params, lineage) \{}
-
-    \CommentTok{## Setting the cumulated birth and death}
-\NormalTok{    birth_death <-}\StringTok{ }\NormalTok{bd.params}\OperatorTok{$}\NormalTok{speciation }\OperatorTok{+}\StringTok{ }\NormalTok{bd.params}\OperatorTok{$}\NormalTok{extinction}
-
-    \CommentTok{## Returning branch lengths depending on different number of taxa}
-    \ControlFlowTok{if}\NormalTok{(lineage}\OperatorTok{$}\NormalTok{n }\OperatorTok{<=}\StringTok{ }\DecValTok{5}\NormalTok{) \{}
-        \KeywordTok{return}\NormalTok{(}\DecValTok{1}    \OperatorTok{*}\StringTok{ }\KeywordTok{rexp}\NormalTok{(}\DecValTok{1}\NormalTok{, }\KeywordTok{sum}\NormalTok{(}\DecValTok{5} \OperatorTok{*}\StringTok{ }\NormalTok{birth_death)))}
+for discrete numbers of taxa, the branch length increases by jumps (ten fold) every 5 taxa:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# A more complex binned.branch.length function}
+\NormalTok{increasing.brlen \textless{}{-}}\StringTok{ }\ControlFlowTok{function}\NormalTok{(bd.params, lineage) \{}
+
+    \CommentTok{\#\# Setting the cumulated birth and death}
+\NormalTok{    birth\_death \textless{}{-}}\StringTok{ }\NormalTok{bd.params}\OperatorTok{$}\NormalTok{speciation }\OperatorTok{+}\StringTok{ }\NormalTok{bd.params}\OperatorTok{$}\NormalTok{extinction}
+
+    \CommentTok{\#\# Returning branch lengths depending on different number of taxa}
+    \ControlFlowTok{if}\NormalTok{(lineage}\OperatorTok{$}\NormalTok{n }\OperatorTok{\textless{}=}\StringTok{ }\DecValTok{5}\NormalTok{) \{}
+        \KeywordTok{return}\NormalTok{(}\DecValTok{1}    \OperatorTok{*}\StringTok{ }\KeywordTok{rexp}\NormalTok{(}\DecValTok{1}\NormalTok{, }\KeywordTok{sum}\NormalTok{(}\DecValTok{5} \OperatorTok{*}\StringTok{ }\NormalTok{birth\_death)))}
 \NormalTok{    \}}
-    \ControlFlowTok{if}\NormalTok{(lineage}\OperatorTok{$}\NormalTok{n }\OperatorTok{<=}\StringTok{ }\DecValTok{10}\NormalTok{) \{}
-        \KeywordTok{return}\NormalTok{(}\DecValTok{10}   \OperatorTok{*}\StringTok{ }\KeywordTok{rexp}\NormalTok{(}\DecValTok{1}\NormalTok{, }\KeywordTok{sum}\NormalTok{(}\DecValTok{10} \OperatorTok{*}\StringTok{ }\NormalTok{birth_death)))   }
+    \ControlFlowTok{if}\NormalTok{(lineage}\OperatorTok{$}\NormalTok{n }\OperatorTok{\textless{}=}\StringTok{ }\DecValTok{10}\NormalTok{) \{}
+        \KeywordTok{return}\NormalTok{(}\DecValTok{10}   \OperatorTok{*}\StringTok{ }\KeywordTok{rexp}\NormalTok{(}\DecValTok{1}\NormalTok{, }\KeywordTok{sum}\NormalTok{(}\DecValTok{10} \OperatorTok{*}\StringTok{ }\NormalTok{birth\_death)))   }
 \NormalTok{    \}}
-    \ControlFlowTok{if}\NormalTok{(lineage}\OperatorTok{$}\NormalTok{n }\OperatorTok{<=}\StringTok{ }\DecValTok{15}\NormalTok{) \{}
-        \KeywordTok{return}\NormalTok{(}\DecValTok{100}  \OperatorTok{*}\StringTok{ }\KeywordTok{rexp}\NormalTok{(}\DecValTok{1}\NormalTok{, }\KeywordTok{sum}\NormalTok{(}\DecValTok{15} \OperatorTok{*}\StringTok{ }\NormalTok{birth_death)))      }
+    \ControlFlowTok{if}\NormalTok{(lineage}\OperatorTok{$}\NormalTok{n }\OperatorTok{\textless{}=}\StringTok{ }\DecValTok{15}\NormalTok{) \{}
+        \KeywordTok{return}\NormalTok{(}\DecValTok{100}  \OperatorTok{*}\StringTok{ }\KeywordTok{rexp}\NormalTok{(}\DecValTok{1}\NormalTok{, }\KeywordTok{sum}\NormalTok{(}\DecValTok{15} \OperatorTok{*}\StringTok{ }\NormalTok{birth\_death)))      }
 \NormalTok{    \}}
-    \ControlFlowTok{if}\NormalTok{(lineage}\OperatorTok{$}\NormalTok{n }\OperatorTok{<=}\StringTok{ }\DecValTok{20}\NormalTok{) \{}
-        \KeywordTok{return}\NormalTok{(}\DecValTok{1000} \OperatorTok{*}\StringTok{ }\KeywordTok{rexp}\NormalTok{(}\DecValTok{1}\NormalTok{, }\KeywordTok{sum}\NormalTok{(}\DecValTok{20} \OperatorTok{*}\StringTok{ }\NormalTok{birth_death)))}
+    \ControlFlowTok{if}\NormalTok{(lineage}\OperatorTok{$}\NormalTok{n }\OperatorTok{\textless{}=}\StringTok{ }\DecValTok{20}\NormalTok{) \{}
+        \KeywordTok{return}\NormalTok{(}\DecValTok{1000} \OperatorTok{*}\StringTok{ }\KeywordTok{rexp}\NormalTok{(}\DecValTok{1}\NormalTok{, }\KeywordTok{sum}\NormalTok{(}\DecValTok{20} \OperatorTok{*}\StringTok{ }\NormalTok{birth\_death)))}
 \NormalTok{    \} }\ControlFlowTok{else}\NormalTok{ \{}
-        \KeywordTok{return}\NormalTok{(}\DecValTok{1000} \OperatorTok{*}\StringTok{ }\KeywordTok{rexp}\NormalTok{(}\DecValTok{1}\NormalTok{, }\KeywordTok{sum}\NormalTok{(lineage}\OperatorTok{$}\NormalTok{n }\OperatorTok{*}\StringTok{ }\NormalTok{birth_death)))}
+        \KeywordTok{return}\NormalTok{(}\DecValTok{1000} \OperatorTok{*}\StringTok{ }\KeywordTok{rexp}\NormalTok{(}\DecValTok{1}\NormalTok{, }\KeywordTok{sum}\NormalTok{(lineage}\OperatorTok{$}\NormalTok{n }\OperatorTok{*}\StringTok{ }\NormalTok{birth\_death)))}
 \NormalTok{    \}}
@@
 \begin{Highlighting}[]
-\CommentTok{## Creating a modifiers}
-\NormalTok{increasing_modifier <-}\StringTok{ }\KeywordTok{make.modifiers}\NormalTok{(}\DataTypeTok{branch.length =}\NormalTok{ increasing.brlen)}
-
-\CommentTok{## Generating a new tree with this modifier}
+\CommentTok{\#\# Creating a modifiers}
+\NormalTok{increasing\_modifier \textless{}{-}}\StringTok{ }\KeywordTok{make.modifiers}\NormalTok{(}\DataTypeTok{branch.length =}\NormalTok{ increasing.brlen)}
+
+\CommentTok{\#\# Generating a new tree with this modifier}
 \KeywordTok{set.seed}\NormalTok{(}\DecValTok{0}\NormalTok{)}
-\NormalTok{increasing_tree <-}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ bd_params,}
-                        \DataTypeTok{stop.rule =}\NormalTok{ stope_rule,}
-                        \DataTypeTok{modifiers =}\NormalTok{ increasing_modifier)}
+\NormalTok{increasing\_tree \textless{}{-}}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ bd\_params,}
+                        \DataTypeTok{stop.rule =}\NormalTok{ stop\_rule,}
+                        \DataTypeTok{modifiers =}\NormalTok{ increasing\_modifier)}
 \end{Highlighting}
@@
 \KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow =} \KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{,}\DecValTok{3}\NormalTok{))}
-\KeywordTok{plot}\NormalTok{(default_tree,    }\DataTypeTok{main =} \StringTok{"Default modifier"}\NormalTok{)}
-\KeywordTok{plot}\NormalTok{(modified_tree,   }\DataTypeTok{main =} \StringTok{"Constant branch length}\CharTok{\textbackslash{}n}\StringTok{modifier"}\NormalTok{)}
-\KeywordTok{plot}\NormalTok{(increasing_tree, }\DataTypeTok{main =} \StringTok{"Increasing branch length}\CharTok{\textbackslash{}n}\StringTok{modifier (binned)"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-48-1.pdf}
+\KeywordTok{plot}\NormalTok{(default\_tree,    }\DataTypeTok{main =} \StringTok{"Default modifier"}\NormalTok{)}
+\KeywordTok{plot}\NormalTok{(modified\_tree,   }\DataTypeTok{main =} \StringTok{"Constant branch length}\CharTok{\textbackslash{}n}\StringTok{modifier"}\NormalTok{)}
+\KeywordTok{plot}\NormalTok{(increasing\_tree, }\DataTypeTok{main =} \StringTok{"Increasing branch length}\CharTok{\textbackslash{}n}\StringTok{modifier (binned)"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-61-1.pdf}
 
@@
 The \texttt{selection} function is used in the birth death process to know which lineage to select when running a speciation (or extinction!) event.
-By default, this function randomly selects one taxon that is currently not extinct (using: \texttt{sample(1:n.taxa,\ 1))}.
+By default, this function randomly selects one taxon that is currently not extinct (using: \texttt{sample(1:lineage\$n,\ 1))}.
 Similarly as \texttt{branch.length} it is possible to modify this part of the birth death process.
@@
 \begin{Highlighting}[]
-\CommentTok{## Our function to always select the last taxon}
-\CommentTok{## (making sure it returns an integer)}
-\NormalTok{select.last <-}\StringTok{ }\ControlFlowTok{function}\NormalTok{(lineage) \{}
+\CommentTok{\#\# Our function to always select the last taxon}
+\CommentTok{\#\# (making sure it returns an integer)}
+\NormalTok{select.last \textless{}{-}}\StringTok{ }\ControlFlowTok{function}\NormalTok{(lineage) \{}
     \KeywordTok{return}\NormalTok{(}\KeywordTok{as.integer}\NormalTok{(lineage}\OperatorTok{$}\NormalTok{n))}
@@
 \begin{quote}
-Note that here the function can only intake the \protect\hyperlink{allowarguments}{allowed arguments as described above} (here \texttt{n.taxa}: the number of current living taxa).
+Note that here the function can only intake the \protect\hyperlink{allowarguments}{allowed arguments as described above} (here \texttt{lineage\$n}: the number of current living taxa).
 \end{quote}
@@
 \begin{Highlighting}[]
-\CommentTok{## A modifier for selection}
-\NormalTok{ladderised_modifier <-}\StringTok{ }\KeywordTok{make.modifiers}\NormalTok{(}\DataTypeTok{selection =}\NormalTok{ select.last)}
-\CommentTok{## Generating a new tree with this modifier}
+\CommentTok{\#\# A modifier for selection}
+\NormalTok{ladderised\_modifier \textless{}{-}}\StringTok{ }\KeywordTok{make.modifiers}\NormalTok{(}\DataTypeTok{selection =}\NormalTok{ select.last)}
+\CommentTok{\#\# Generating a new tree with this modifier}
 \KeywordTok{set.seed}\NormalTok{(}\DecValTok{0}\NormalTok{)}
-\NormalTok{ladderised_tree <-}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ bd_params,}
-                        \DataTypeTok{stop.rule =}\NormalTok{ stope_rule,}
-                        \DataTypeTok{modifiers =}\NormalTok{ ladderised_modifier)}
-\CommentTok{## Displaying the results}
+\NormalTok{ladderised\_tree \textless{}{-}}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ bd\_params,}
+                        \DataTypeTok{stop.rule =}\NormalTok{ stop\_rule,}
+                        \DataTypeTok{modifiers =}\NormalTok{ ladderised\_modifier)}
+\CommentTok{\#\# Displaying the results}
 \KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow =} \KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{,}\DecValTok{2}\NormalTok{))}
-\KeywordTok{plot}\NormalTok{(default_tree,    }\DataTypeTok{main =} \StringTok{"Default modifier"}\NormalTok{)}
-\KeywordTok{plot}\NormalTok{(ladderised_tree, }\DataTypeTok{main =} \StringTok{"Ladderising modifier"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-50-1.pdf}
+\KeywordTok{plot}\NormalTok{(default\_tree,    }\DataTypeTok{main =} \StringTok{"Default modifier"}\NormalTok{)}
+\KeywordTok{plot}\NormalTok{(ladderised\_tree, }\DataTypeTok{main =} \StringTok{"Ladderising modifier"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-63-1.pdf}
 
 Again, it is of course possible to make the modifier more complex and in combination with other elements of the tree.
-For example, we can create a \texttt{"dads"} object that also creates a trait a add to it a \texttt{selection} modifier that only selects for tips with positive trait values.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{## Our function that only select taxa with positive trait values}
-\NormalTok{select.positive <-}\StringTok{ }\ControlFlowTok{function}\NormalTok{(trait.values, lineage) \{}
-    \CommentTok{## Selecting the taxa with positive traits only}
-\NormalTok{    positives <-}\StringTok{ }\KeywordTok{as.integer}\NormalTok{(}\KeywordTok{rownames}\NormalTok{(trait.values)[}\KeywordTok{which}\NormalTok{(trait.values[, }\DecValTok{1}\NormalTok{] }\OperatorTok{>=}\StringTok{ }\DecValTok{0}\NormalTok{)])}
-
-    \CommentTok{## Select the current taxa that descend from a node with a positive value}
-\NormalTok{    positive_living <-}\StringTok{ }\KeywordTok{cbind}\NormalTok{(lineage}\OperatorTok{$}\NormalTok{parents, }\KeywordTok{seq_along}\NormalTok{(lineage}\OperatorTok{$}\NormalTok{split)}
-\NormalTok{                             )[}\KeywordTok{which}\NormalTok{(lineage}\OperatorTok{$}\NormalTok{parents }\OperatorTok{%in%}\StringTok{ }\NormalTok{positives), }\DecValTok{2}\NormalTok{]}
-
-    \CommentTok{## Select one tip randomly in the ones with descendants with positive values}
-    \KeywordTok{return}\NormalTok{(}\KeywordTok{sample}\NormalTok{(}\KeywordTok{which}\NormalTok{(lineage}\OperatorTok{$}\NormalTok{livings }\OperatorTok{%in%}\StringTok{ }\NormalTok{positive_living), }\DecValTok{1}\NormalTok{))}
+For example, we can create a \texttt{"dads"} object that also generates a BM trait and add to it this object a \texttt{selection} modifier that only selects tips with positive trait values (only species with positive trait values will speciate).
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# Our function that only select taxa with positive trait values}
+\NormalTok{select.positive \textless{}{-}}\StringTok{ }\ControlFlowTok{function}\NormalTok{(trait.values, lineage) \{}
+
+    \CommentTok{\#\# Selecting the taxa names with positive values for the first trait}
+\NormalTok{    positives \textless{}{-}}\StringTok{ }\KeywordTok{as.integer}\NormalTok{(}\KeywordTok{rownames}\NormalTok{(trait.values)[}\KeywordTok{which}\NormalTok{(trait.values[, }\DecValTok{1}\NormalTok{] }\OperatorTok{\textgreater{}=}\StringTok{ }\DecValTok{0}\NormalTok{)])}
+
+    \CommentTok{\#\# Combine the descendants of the current lineages (lineage$parents)}
+    \CommentTok{\#\# with the species that have speciated (seq\_along(lineages$split))}
+    \CommentTok{\#\# to have a table of pairs of parents/splits}
+\NormalTok{    parents\_split\_table \textless{}{-}}\StringTok{ }\KeywordTok{cbind}\NormalTok{(lineage}\OperatorTok{$}\NormalTok{parents, }\KeywordTok{seq\_along}\NormalTok{(lineage}\OperatorTok{$}\NormalTok{split))}
+    \CommentTok{\#\# Select the current taxa that descend from a node with a positive value}
+\NormalTok{    positive\_living \textless{}{-}}\StringTok{ }\NormalTok{parents\_split\_table[}\KeywordTok{which}\NormalTok{(lineage}\OperatorTok{$}\NormalTok{parents }\OperatorTok{\%in\%}\StringTok{ }\NormalTok{positives), }\DecValTok{2}\NormalTok{]}
+
+    \CommentTok{\#\# Select one tip randomly in the ones with descendants with positive values}
+    \KeywordTok{return}\NormalTok{(}\KeywordTok{sample}\NormalTok{(}\KeywordTok{which}\NormalTok{(lineage}\OperatorTok{$}\NormalTok{livings }\OperatorTok{\%in\%}\StringTok{ }\NormalTok{positive\_living), }\DecValTok{1}\NormalTok{))}
 \NormalTok{\}}
 
-\CommentTok{## Creating the modifier}
-\NormalTok{positive_skew <-}\StringTok{ }\KeywordTok{make.modifiers}\NormalTok{(}\DataTypeTok{selection =}\NormalTok{ select.positive)}
-
-\CommentTok{## Creating a (default) trait object}
-\NormalTok{BM_trait <-}\StringTok{ }\KeywordTok{make.traits}\NormalTok{()}
-
-\CommentTok{## Simulate a tree and trait with no modifier}
+\CommentTok{\#\# Creating the modifier}
+\NormalTok{positive\_skew \textless{}{-}}\StringTok{ }\KeywordTok{make.modifiers}\NormalTok{(}\DataTypeTok{selection =}\NormalTok{ select.positive)}
+
+\CommentTok{\#\# Creating a (default) trait object}
+\NormalTok{BM\_trait \textless{}{-}}\StringTok{ }\KeywordTok{make.traits}\NormalTok{()}
+
+\CommentTok{\#\# Simulate a tree and trait with no modifier}
 \KeywordTok{set.seed}\NormalTok{(}\DecValTok{1}\NormalTok{)}
-\NormalTok{default_dads <-}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ bd_params,}
-                     \DataTypeTok{stop.rule =}\NormalTok{ stope_rule,}
-                     \DataTypeTok{traits    =}\NormalTok{ BM_trait)}
-
-\CommentTok{## Simulate a tree and trait with the modifier}
+\NormalTok{default\_dads \textless{}{-}}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ bd\_params,}
+                     \DataTypeTok{stop.rule =}\NormalTok{ stop\_rule,}
+                     \DataTypeTok{traits    =}\NormalTok{ BM\_trait)}
+
+\CommentTok{\#\# Simulate a tree and trait with the modifier}
 \KeywordTok{set.seed}\NormalTok{(}\DecValTok{1}\NormalTok{)}
-\NormalTok{skewed_trait_dads <-}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ bd_params,}
-                          \DataTypeTok{stop.rule =}\NormalTok{ stope_rule,}
-                          \DataTypeTok{traits    =}\NormalTok{ BM_trait,}
-                          \DataTypeTok{modifiers =}\NormalTok{ positive_skew)}
-
-\CommentTok{## Plotting the differences in trees and traits}
+\NormalTok{skewed\_trait\_dads \textless{}{-}}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ bd\_params,}
+                          \DataTypeTok{stop.rule =}\NormalTok{ stop\_rule,}
+                          \DataTypeTok{traits    =}\NormalTok{ BM\_trait,}
+                          \DataTypeTok{modifiers =}\NormalTok{ positive\_skew)}
+
+\CommentTok{\#\# Plotting the differences in trees and traits}
 \KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow =} \KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{, }\DecValTok{2}\NormalTok{))}
-\KeywordTok{plot}\NormalTok{(default_dads, }\DataTypeTok{main =} \StringTok{"Default trait and tree"}\NormalTok{)}
-\KeywordTok{plot}\NormalTok{(skewed_trait_dads, }\DataTypeTok{main =} \StringTok{"Skewed trait and tree"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-51-1.pdf}
+\KeywordTok{plot}\NormalTok{(default\_dads, }\DataTypeTok{main =} \StringTok{"Default trait and tree"}\NormalTok{)}
+\KeywordTok{plot}\NormalTok{(skewed\_trait\_dads, }\DataTypeTok{main =} \StringTok{"Skewed trait and tree"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-64-1.pdf}
 
@@
 \begin{Highlighting}[]
-\CommentTok{## The speciation in pseudo-code:}
-\KeywordTok{runif}\NormalTok{(}\DecValTok{1}\NormalTok{) }\OperatorTok{<}\StringTok{ }\NormalTok{speciation}\OperatorTok{/}\StringTok{ }\NormalTok{(speciation }\OperatorTok{+}\StringTok{ }\NormalTok{extinction)}
+\CommentTok{\#\# The speciation in pseudo{-}code:}
+\KeywordTok{runif}\NormalTok{(}\DecValTok{1}\NormalTok{) }\OperatorTok{\textless{}}\StringTok{ }\NormalTok{speciation}\OperatorTok{/}\StringTok{ }\NormalTok{(speciation }\OperatorTok{+}\StringTok{ }\NormalTok{extinction)}
 \end{Highlighting}
@@
 \begin{Highlighting}[]
-\CommentTok{## Speciating or going extinct randomly}
-\CommentTok{## (regardless of the extinction parameter)}
-\NormalTok{random.extinct  <-}\StringTok{ }\ControlFlowTok{function}\NormalTok{() \{}
+\CommentTok{\#\# Speciating or going extinct randomly}
+\CommentTok{\#\# (regardless of the extinction parameter)}
+\NormalTok{random.extinct  \textless{}{-}}\StringTok{ }\ControlFlowTok{function}\NormalTok{() \{}
     \KeywordTok{return}\NormalTok{(}\KeywordTok{sample}\NormalTok{(}\KeywordTok{c}\NormalTok{(}\OtherTok{TRUE}\NormalTok{, }\OtherTok{FALSE}\NormalTok{), }\DecValTok{1}\NormalTok{))}
@@
 
-\CommentTok{## Creating the modifiers object}
-\NormalTok{random_extinction <-}\StringTok{ }\KeywordTok{make.modifiers}\NormalTok{(}\DataTypeTok{speciation =}\NormalTok{ random.extinct)}
-
-\CommentTok{## Generating a new tree with this modifier}
-\KeywordTok{set.seed}\NormalTok{(}\DecValTok{0}\NormalTok{)}
-\NormalTok{modified_tree <-}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ bd_params,}
-                      \DataTypeTok{stop.rule =}\NormalTok{ stope_rule,}
-                      \DataTypeTok{modifiers =}\NormalTok{ random_extinction)}
+\CommentTok{\#\# Creating the modifiers object}
+\NormalTok{random\_extinction \textless{}{-}}\StringTok{ }\KeywordTok{make.modifiers}\NormalTok{(}\DataTypeTok{speciation =}\NormalTok{ random.extinct)}
+
+\CommentTok{\#\# Generating a new tree with this modifier}
+\KeywordTok{set.seed}\NormalTok{(}\DecValTok{2}\NormalTok{)}
+\NormalTok{modified\_tree \textless{}{-}}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ bd\_params,}
+                      \DataTypeTok{stop.rule =}\NormalTok{ stop\_rule,}
+                      \DataTypeTok{modifiers =}\NormalTok{ random\_extinction)}
 
 \KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow =} \KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{,}\DecValTok{2}\NormalTok{))}
-\KeywordTok{plot}\NormalTok{(default_tree,  }\DataTypeTok{main =} \StringTok{"Default modifier"}\NormalTok{)}
-\KeywordTok{plot}\NormalTok{(modified_tree, }\DataTypeTok{main =} \StringTok{"Random extinction}\CharTok{\textbackslash{}n}\StringTok{modifier"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-Note how every lineage end up going extinct!
+\KeywordTok{plot}\NormalTok{(default\_tree,  }\DataTypeTok{main =} \StringTok{"Default modifier"}\NormalTok{)}
+\KeywordTok{plot}\NormalTok{(modified\_tree, }\DataTypeTok{main =} \StringTok{"Random extinction}\CharTok{\textbackslash{}n}\StringTok{modifier"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{quote}
+Note how loads of lineages end up going extinct even if the extinction parameter is set to 0!
+\end{quote}
+
 And again, we can make some more advanced modifiers: for example, one where a tip always goes extinct if their ancestor has a negative trait value (here we will also introduce the utility function \texttt{parent.trait} that automatically selects the trait values of the parent of the current lineage.
@@
 \begin{Highlighting}[]
-\CommentTok{## A modifier for removing tips with negative values}
-\NormalTok{bias.trait <-}\StringTok{ }\ControlFlowTok{function}\NormalTok{(trait.values, lineage) \{}
-    \ControlFlowTok{if}\NormalTok{(}\KeywordTok{parent.traits}\NormalTok{(trait.values, lineage) }\OperatorTok{<}\StringTok{ }\DecValTok{0}\NormalTok{) \{}
-        \CommentTok{## Go extinct!}
+\CommentTok{\#\# A modifier for removing tips with negative values}
+\NormalTok{bias.trait \textless{}{-}}\StringTok{ }\ControlFlowTok{function}\NormalTok{(trait.values, lineage) \{}
+    \ControlFlowTok{if}\NormalTok{(}\KeywordTok{parent.traits}\NormalTok{(trait.values, lineage) }\OperatorTok{\textless{}}\StringTok{ }\DecValTok{0}\NormalTok{) \{}
+        \CommentTok{\#\# Go extinct!}
         \KeywordTok{return}\NormalTok{(}\OtherTok{FALSE}\NormalTok{)}
 \NormalTok{    \} }\ControlFlowTok{else}\NormalTok{ \{}
-        \CommentTok{## Speciate}
+        \CommentTok{\#\# Speciate}
         \KeywordTok{return}\NormalTok{(}\OtherTok{TRUE}\NormalTok{)}
@@
 
-\CommentTok{## Creating the modifier}
-\NormalTok{biased_trait <-}\StringTok{ }\KeywordTok{make.modifiers}\NormalTok{(}\DataTypeTok{speciation =}\NormalTok{ bias.trait)}
-
-\CommentTok{## Simulate a tree and trait with the modifier}
+\CommentTok{\#\# Creating the modifier}
+\NormalTok{biased\_trait \textless{}{-}}\StringTok{ }\KeywordTok{make.modifiers}\NormalTok{(}\DataTypeTok{speciation =}\NormalTok{ bias.trait)}
+
+\CommentTok{\#\# Simulate a tree and trait with the modifier}
 \KeywordTok{set.seed}\NormalTok{(}\DecValTok{1}\NormalTok{)}
-\NormalTok{biased_trait_dads <-}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ bd_params,}
-                          \DataTypeTok{stop.rule =}\NormalTok{ stope_rule,}
-                          \DataTypeTok{traits    =}\NormalTok{ BM_trait,}
-                          \DataTypeTok{modifiers =}\NormalTok{ biased_trait)}
-
-\CommentTok{## Plotting the differences in trees and traits}
+\NormalTok{biased\_trait\_dads \textless{}{-}}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ bd\_params,}
+                          \DataTypeTok{stop.rule =}\NormalTok{ stop\_rule,}
+                          \DataTypeTok{traits    =}\NormalTok{ BM\_trait,}
+                          \DataTypeTok{modifiers =}\NormalTok{ biased\_trait)}
+
+\CommentTok{\#\# Plotting the differences in trees and traits}
 \KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow =} \KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{, }\DecValTok{2}\NormalTok{))}
-\KeywordTok{plot}\NormalTok{(default_dads, }\DataTypeTok{main =} \StringTok{"Default trait and tree"}\NormalTok{)}
-\KeywordTok{plot}\NormalTok{(biased_trait_dads, }\DataTypeTok{main =} \StringTok{"Biased trait and tree"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-54-1.pdf}
+\KeywordTok{plot}\NormalTok{(default\_dads, }\DataTypeTok{main =} \StringTok{"Default trait and tree"}\NormalTok{)}
+\KeywordTok{plot}\NormalTok{(biased\_trait\_dads, }\DataTypeTok{main =} \StringTok{"Biased trait and tree"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-67-1.pdf}
 
@@
 In other words, throughout the process, the modifications remain constant (even if they are conditional).
-It is however possible to code the \texttt{"modifiers"} so that they can be affected by \texttt{"events"} objects (see next chapter).
+It is however possible to code the \texttt{"modifiers"} so that they can be affected by \texttt{"events"} objects (see next chapter \protect\hyperlink{makeevents}{on \texttt{events}}).
 
@@
 
-\hypertarget{combining-and-editing-modifiers-add}{%
-\section{\texorpdfstring{Combining and editing modifiers (\texttt{add})}{Combining and editing modifiers (add)}}\label{combining-and-editing-modifiers-add}}
-
-\hypertarget{testing-modifiers-test}{%
-\section{\texorpdfstring{Testing modifiers (\texttt{test})}{Testing modifiers (test)}}\label{testing-modifiers-test}}
-
-\hypertarget{demo-runnable}{%
-\section{Demo runnable}\label{demo-runnable}}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{bd.params <-}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{speciation =} \DecValTok{1}\NormalTok{, }\DataTypeTok{extinction =} \DecValTok{1}\OperatorTok{/}\DecValTok{3}\NormalTok{)}
-\NormalTok{traits <-}\StringTok{ }\KeywordTok{make.traits}\NormalTok{()}
-\NormalTok{stop.rule <-}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{max.taxa =} \DecValTok{20}\NormalTok{, }\DataTypeTok{max.living =} \OtherTok{Inf}\NormalTok{, }\DataTypeTok{max.time =} \OtherTok{Inf}\NormalTok{)}
-\NormalTok{modifiers <-}\StringTok{ }\OtherTok{NULL}
-\NormalTok{events <-}\StringTok{ }\OtherTok{NULL}
-\NormalTok{null.error <-}\StringTok{ }\OtherTok{NULL}
-
-\CommentTok{## modifiers}
-\NormalTok{condition <-}\StringTok{ }\ControlFlowTok{function}\NormalTok{(trait.values, parent.lineage) }\KeywordTok{return}\NormalTok{(}\KeywordTok{parent.traits}\NormalTok{(trait.values, parent.lineage) }\OperatorTok{<}\StringTok{ }\DecValTok{0}\NormalTok{)}
-\NormalTok{modify <-}\StringTok{ }\ControlFlowTok{function}\NormalTok{(x) }\KeywordTok{return}\NormalTok{(x }\OperatorTok{*}\StringTok{ }\DecValTok{20}\NormalTok{)}
-
-\CommentTok{## Setting up the different modifiers}
-\NormalTok{modify_speciation <-}\StringTok{ }\KeywordTok{make.modifiers}\NormalTok{(}\DataTypeTok{speciation    =}\NormalTok{ speciation.trait,}
-                                    \DataTypeTok{condition     =}\NormalTok{ condition,}
-                                    \DataTypeTok{modify        =}\NormalTok{ modify)}
-\NormalTok{modify_brlen <-}\StringTok{ }\KeywordTok{make.modifiers}\NormalTok{(}\DataTypeTok{branch.length =}\NormalTok{ branch.length.trait,}
-                               \DataTypeTok{condition     =}\NormalTok{ condition,}
-                               \DataTypeTok{modify        =}\NormalTok{ modify)}
-\NormalTok{modify_speciation_brlen <-}\StringTok{ }\KeywordTok{make.modifiers}\NormalTok{(}\DataTypeTok{branch.length =}\NormalTok{ branch.length.trait,}
-                                          \DataTypeTok{speciation    =}\NormalTok{ speciation.trait,}
-                                          \DataTypeTok{condition     =}\NormalTok{ condition,}
-                                          \DataTypeTok{modify        =}\NormalTok{ modify)}
-
-\CommentTok{## Test normal (no modifiers)}
+For example in the \texttt{speciation} part of a modifier, the default is to trigger an event from a uniform distribution and then check if that value is smaller than (\[speciation/(speciation + extinction)\]):
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# The default speciation algorithm}
+\NormalTok{speciation \textless{}{-}}\StringTok{ }\ControlFlowTok{function}\NormalTok{(bd.params) \{}
+    \CommentTok{\#\# Randomly trigger an event}
+\NormalTok{    trigger\_event \textless{}{-}}\StringTok{ }\KeywordTok{runif}\NormalTok{(}\DecValTok{1}\NormalTok{)}
+
+    \CommentTok{\#\# Speciate?}
+    \KeywordTok{return}\NormalTok{(trigger\_event }\OperatorTok{\textless{}}\StringTok{ }\NormalTok{(bd.params}\OperatorTok{$}\NormalTok{speciation}\OperatorTok{/}\NormalTok{(bd.params}\OperatorTok{$}\NormalTok{speciation }\OperatorTok{+}\StringTok{ }\NormalTok{bd.params}\OperatorTok{$}\NormalTok{extinction)))}
+\NormalTok{\}}
+\end{Highlighting}
+\end{Shaded}
+
+It is possible with some specific condition to modify this trigger by providing a \texttt{condition} and \texttt{modify} function to the speciation function.
+That is, if the \texttt{condition} is met, apply the \texttt{modify} function to the algorithm.
+For example here we can edit the default speciation function to have a modification (\texttt{modify\ =\ double.the.trigger}: doubling the trigger value) happening half the time (\texttt{condition\ =\ half.the.time}):
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# A conditional function that triggers half the time}
+\NormalTok{half.the.time \textless{}{-}}\StringTok{ }\ControlFlowTok{function}\NormalTok{() }\KeywordTok{return}\NormalTok{(}\KeywordTok{sample}\NormalTok{(}\KeywordTok{c}\NormalTok{(}\OtherTok{TRUE}\NormalTok{, }\OtherTok{FALSE}\NormalTok{), }\DecValTok{1}\NormalTok{))}
+\CommentTok{\#\# A modification that doubles the value to trigger the event}
+\NormalTok{double.the.trigger \textless{}{-}}\StringTok{ }\ControlFlowTok{function}\NormalTok{(x) }\KeywordTok{return}\NormalTok{(x}\OperatorTok{*}\DecValTok{2}\NormalTok{)}
+\CommentTok{\#\# A conditional modifier}
+\KeywordTok{make.modifiers}\NormalTok{(}\DataTypeTok{speciation =}\NormalTok{ speciation,}
+               \DataTypeTok{condition  =}\NormalTok{ half.the.time, }
+               \DataTypeTok{modify     =}\NormalTok{ double.the.trigger)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+##  ---- dads modifiers object ---- 
+## Default branch length process.
+## Default selection process.
+## Speciation process is set to speciation with a condition (half.the.time) and a modifier (double.the.trigger).
+\end{verbatim}
+
+Effectively, this will internally modify the speciation function as follows
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# The default speciation algorithm}
+\NormalTok{speciation \textless{}{-}}\StringTok{ }\ControlFlowTok{function}\NormalTok{(bd.params) \{}
+    \CommentTok{\#\# Randomly trigger an event}
+\NormalTok{    trigger\_event \textless{}{-}}\StringTok{ }\KeywordTok{runif}\NormalTok{(}\DecValTok{1}\NormalTok{)}
+
+    \CommentTok{\#\# Modify the triggering}
+    \ControlFlowTok{if}\NormalTok{(}\KeywordTok{half.the.time}\NormalTok{()) \{ }\CommentTok{\#\# This running the half.the.time function drawing TRUE or FALSE randomly}
+\NormalTok{        trigger\_event \textless{}{-}}\StringTok{ }\KeywordTok{double.the.trigger}\NormalTok{(trigger\_event) }\CommentTok{\#\# This will double the random value trigger\_event}
+\NormalTok{    \}}
+
+    \CommentTok{\#\# Speciate?}
+    \KeywordTok{return}\NormalTok{(trigger\_event }\OperatorTok{\textless{}}\StringTok{ }\NormalTok{(bd.params}\OperatorTok{$}\NormalTok{speciation}\OperatorTok{/}\NormalTok{(bd.params}\OperatorTok{$}\NormalTok{speciation }\OperatorTok{+}\StringTok{ }\NormalTok{bd.params}\OperatorTok{$}\NormalTok{extinction)))}
+\NormalTok{\}}
+\end{Highlighting}
+\end{Shaded}
+
+These \texttt{condition} and \texttt{modify} functions can be applied to all the \texttt{modifiers} elements (\texttt{selection}, \texttt{branch.length} and \texttt{speciation}).
+However, they are typically used in \texttt{events} that will modify\ldots{} the \texttt{modifiers} (\protect\hyperlink{makeevents}{see the \texttt{events} section})!
+
+\begin{verbatim}
+## This build of rgl does not include OpenGL functions.  Use
+##  rglwidget() to display results, e.g. via options(rgl.printRglwidget = TRUE).
+\end{verbatim}
+
+\hypertarget{makeevents}{%
+\chapter{Adding events to simulations}\label{makeevents}}
+
+One other major feature of the \texttt{dads} package is to allow simulations to run with specified events to occur during the simulation.
+These are typically events that can drastically change the course of the simulation.
+For example, you might want simulate some mass extinction at some specific point in time.
+However, these can also be more subtle, like the internal change of parameter values when reaching a specific trait value.
+These are all handled by \texttt{dads} with the \texttt{events} object that you can create using \texttt{make.events} and has the same overall logic as \texttt{make.traits} and \texttt{make.modifiers}.
+
+\texttt{events} require three main arguments:
+* the \texttt{target} which designates what the extinction should affect (e.g.~the taxa, the speciation rate, etc\ldots);
+* the \texttt{condition} which designates when to trigger the event;
+* the \texttt{modification} which designates what to exactly modify when the event is triggered.
+
+There are several more arguments that can be passed to \texttt{make.events} but they are discussed later on.
+First let's focus on these three main arguments:
+
+\hypertarget{target}{%
+\section{Target}\label{target}}
+
+The target of the event is what the event is going to modify in the birth-death algorithm.
+You can only have one target per event (along with one condition and one modification) but you can create events that contain multiple events (i.e.~multiple triplets of target/condition/modification).
+The targets that are currently available are:
+* \texttt{"taxa"} to modify anything linked to the \protect\hyperlink{allowarguments}{\texttt{lineage} list}. E.g. making half of the living taxa go extinct.
+* \texttt{"bd.params"} to modify anything linked to the \protect\hyperlink{makebdparams}{\texttt{bd.params} object}.
+For example you might want to change the distribution of one of the parameter after some \texttt{conditions}.
+This is typically done by updating the object using the argument \texttt{update} from \texttt{make.bd.params}.
+* \texttt{"traits"} to modify anything linked to the \protect\hyperlink{maketraits}{\texttt{traits} object}.
+For example you might want to change the trait process after some \texttt{conditions}.
+This is typically done by updating the object using the argument \texttt{update} from \texttt{make.traits}.
+* \texttt{"modifiers"} to modify anything linked to the \protect\hyperlink{makebdmodifiers}{\texttt{modifiers} object}.
+For example you might want to change the speciation rule after some \texttt{conditions}.
+This is typically done by updating the object using the argument \texttt{update} from \texttt{make.modifiers}.
+* \texttt{"founding"} this target is a bit more special and allows you run a nested \texttt{dads} object in the simulation. It is cover in a \protect\hyperlink{founding}{specific section below}.
+
+We will see some example of these targets by illustrating the \texttt{conditions} and \texttt{modifications} below.
+
+\hypertarget{conditions}{%
+\section{Conditions}\label{conditions}}
+
+\texttt{condition} is a function that returns a logical value. When a specific condition is met, it should return \texttt{TRUE} and trigger the event, else it should return \texttt{FALSE}.
+
+Currently there are three conditions functions implemented in \texttt{dads} but you can easily come up with your own version of them.
+
+All \texttt{condition} functions in \texttt{dads} take at least to arguments: \texttt{x} for the variable of interest (e.g.~time, number of taxa, trait value, etc\ldots) and \texttt{condition}, the relational operator to evaluate.
+A relational operator is the proper (fancy) term designating all the comparisons you're regularly using in \texttt{R} like \texttt{==} (is equal?) \texttt{\textless{}} (is smaller?) \texttt{\textgreater{}=} (is bigger or equal?).
+You can get the full list in the base manual (using \texttt{?Comparison} or any of the relational operator in a function form).
+
+\begin{quote}
+Note on the function form of a operator in \texttt{R}: if you weren't aware, functions in \texttt{R} are \emph{usually} described in the format \texttt{function(arguments)}, however many other functions you are commonly using are written as operators, for example (\texttt{1\ +\ 1}).
+These are still functions though!
+And you can always use them in the classic function format by quoting them in back ticks: the following does the same as \texttt{(1+1)}: \texttt{\textasciigrave{}+\textasciigrave{}(1,\ 1)}.
+You can use that for parsing your relational operators as functions in \texttt{condition}.
+If you want you condition to be equal, use the syntax \texttt{condition\ =\ \textasciigrave{}==\textasciigrave{}}.
+More details on how functions \emph{really} work in the \href{https://adv-r.hadley.nz/functions.html}{Advanced \texttt{R} book}!
+\end{quote}
+
+\begin{itemize}
+\tightlist
+\item
+  \texttt{time.condition} will trigger the \texttt{event} once a certain time is reached in the simulations and is the simplest/more basic condition with no arguments other than the time required.
+For example \texttt{time.condition(4,\ condition\ =\ \textasciigrave{}==\textasciigrave{})} will trigger the \texttt{event} once the simulations reached 4 units of time.
+Easy.
+\item
+  \texttt{taxa.condition} will trigger the \texttt{event} once a certain number of taxa is reached.
+This can be considered including or excluding fossil species.
+For example \texttt{taxa.condition(42,\ condition\ =\ \textasciigrave{}\textgreater{}=\textasciigrave{},\ living\ =\ TRUE)} will trigger the \texttt{event} if there is at least 42 living species.
+\item
+  \texttt{trait.condition} will trigger the \texttt{event} once a certain trait value is reached.
+This function allows to say which trait(s) to target (by default, the first one using \texttt{trait\ =\ 1}), what value of the trait to target (by default \texttt{what\ =\ max}) and whether to use and absolute trait value or not (\texttt{absolute\ =\ TRUE}).
+For example \texttt{trait.condition(1/3,\ condition\ =\ \textasciigrave{}\textgreater{}\textasciigrave{},\ trait\ =\ 1,\ what\ =\ sd)} will trigger the condition after the standard deviation of the first trait reaches 1/3.
+\end{itemize}
+
+\hypertarget{modifications}{%
+\section{Modifications}\label{modifications}}
+
+After defining the \texttt{event}'s \texttt{target} and \texttt{condition}, you'll need to also specify what it should modify.
+These are basically functions that should modify a specific aspect of the target
+Depending on the target you can modify the following:
+
+\begin{itemize}
+\tightlist
+\item
+  if the target is \texttt{"taxa"} you can modify the internal \protect\hyperlink{allowarguments}{\texttt{lineage} list} by removing living species using \texttt{random.extinction} or \texttt{trait.extinction}.
+\item
+  if the target is \texttt{"taxa"} you can modify the lineage tracker by removing living species using \texttt{random.extinction} or \texttt{trait.extinction}.
+\item
+  if the target is \texttt{"bd.params"} you can modify the \protect\hyperlink{makebdparams}{\texttt{bd.params} object} using \texttt{update.bd.params}.
+\item
+  if the target is \texttt{"traits"} you can modify the \protect\hyperlink{maketraits}{\texttt{traits} object} using \texttt{update.traits}.
+\item
+  if the target is \texttt{"modifiers"} you can modify the \protect\hyperlink{makemodifiers}{\texttt{modifiers} object} using \texttt{update.modifiers}.
+\item
+  (for the \texttt{"founding"} target \protect\hyperlink{founding}{see below})
+\end{itemize}
+
+The most straightforward example if for \texttt{modifications} on \texttt{bd.params}, \texttt{traits} or \texttt{modifiers} objects since they use the same syntax as for their generic \texttt{make.X} function.
+For example, for \texttt{make.traits}, you can update a \texttt{trait} using the \texttt{update} argument as follows:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# A BM trait in two dimensions}
+\NormalTok{(BM\_2D \textless{}{-}}\StringTok{ }\KeywordTok{make.traits}\NormalTok{(}\DataTypeTok{n =} \DecValTok{2}\NormalTok{, }\DataTypeTok{process =}\NormalTok{ BM.process))}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+##  ---- dads traits object ---- 
+## 2 traits for 1 process (A:2) with one starting value (0).
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# Updating the 2D BM into a 2D OU}
+\NormalTok{(OU\_2D \textless{}{-}}\StringTok{ }\KeywordTok{make.traits}\NormalTok{(}\DataTypeTok{update =}\NormalTok{ BM\_2D, }\DataTypeTok{process =}\NormalTok{ OU.process))}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+##  ---- dads traits object ---- 
+## 2 traits for 1 process (A:2) with one starting value (0).
+\end{verbatim}
+
+So basically the functions \texttt{update.X} apply the update to the object \texttt{X} when the event happens.
+The following \texttt{event} updates the \texttt{bd.params} object by setting the extinction parameter to 1/3 when reaching a 10 species:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{make.events}\NormalTok{(}\DataTypeTok{target       =} \StringTok{"bd.params"}\NormalTok{,}
+            \DataTypeTok{condition    =} \KeywordTok{taxa.condition}\NormalTok{(}\DecValTok{10}\NormalTok{, }\DataTypeTok{condition =} \StringTok{\textasciigrave{}}\DataTypeTok{\textgreater{}=}\StringTok{\textasciigrave{}}\NormalTok{),}
+            \DataTypeTok{modification =} \KeywordTok{update.bd.params}\NormalTok{(}\DataTypeTok{extinction =} \DecValTok{1}\OperatorTok{/}\DecValTok{3}\NormalTok{))}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{quote}
+Note that by default events are triggered only once in the whole simulation, so although the example above states that the condition is reaching at least 10 taxa, it will not trigger every times it reaches more than 10 taxa.
+You can change the number of times the events can be triggered using the argument \texttt{replications} (by default it's set to \texttt{replications\ =\ 0} for triggering it only once).
+\end{quote}
+
+\hypertarget{examples}{%
+\section{Examples}\label{examples}}
+
+Here are some example illustrating how to generate events.
+For a simple example, we can create a extinction event that will remove 80\% of species after reaching time 4:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# 80\% mass extinction at time 4}
+\NormalTok{mass\_extinction \textless{}{-}}\StringTok{ }\KeywordTok{make.events}\NormalTok{(}
+                      \DataTypeTok{target       =} \StringTok{"taxa"}\NormalTok{,}
+                      \DataTypeTok{condition    =} \KeywordTok{time.condition}\NormalTok{(}\DecValTok{4}\NormalTok{),}
+                      \DataTypeTok{modification =} \KeywordTok{random.extinction}\NormalTok{(}\FloatTok{0.8}\NormalTok{))}
+
+\CommentTok{\#\# Simulation parameters}
+\NormalTok{stop.rule \textless{}{-}}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{max.time =} \DecValTok{5}\NormalTok{)}
+\NormalTok{bd.params \textless{}{-}}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{extinction =} \DecValTok{0}\NormalTok{, }\DataTypeTok{speciation =} \DecValTok{1}\NormalTok{)}
+
+\CommentTok{\#\# Running the simulations}
+\KeywordTok{set.seed}\NormalTok{(}\DecValTok{123}\NormalTok{)}
+\NormalTok{results \textless{}{-}}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ bd.params,}
+                \DataTypeTok{stop.rule =}\NormalTok{ stop.rule,}
+                \DataTypeTok{events    =}\NormalTok{ mass\_extinction)}
+\CommentTok{\#\# Plotting the results}
+\KeywordTok{plot}\NormalTok{(results, }\DataTypeTok{show.tip.label =} \OtherTok{FALSE}\NormalTok{)}
+\KeywordTok{axisPhylo}\NormalTok{()}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-75-1.pdf}
+
+Or for a slightly more complex example, we can can change the trait process from a BM to an OU when the trait values reaches an upper 95\% quantile value above 2:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# The 95\% upper quantile value of a distribution}
+\NormalTok{upper}\FloatTok{.95}\NormalTok{ \textless{}{-}}\StringTok{ }\ControlFlowTok{function}\NormalTok{(x) \{}
+    \KeywordTok{return}\NormalTok{(}\KeywordTok{quantile}\NormalTok{(x, }\DataTypeTok{prob =} \FloatTok{0.95}\NormalTok{))}
+\NormalTok{\} }
+\CommentTok{\#\# Create an event to change the trait process}
+\NormalTok{change\_process \textless{}{-}}\StringTok{ }\KeywordTok{make.events}\NormalTok{(}
+                  \DataTypeTok{target       =} \StringTok{"traits"}\NormalTok{,}
+                  \CommentTok{\#\# condition is triggered if(upper.95(x) \textgreater{} 3)}
+                  \DataTypeTok{condition    =} \KeywordTok{trait.condition}\NormalTok{(}\DecValTok{3}\NormalTok{, }\DataTypeTok{condition =} \StringTok{\textasciigrave{}}\DataTypeTok{\textgreater{}}\StringTok{\textasciigrave{}}\NormalTok{, }\DataTypeTok{what =}\NormalTok{ upper}\FloatTok{.95}\NormalTok{),}
+                  \DataTypeTok{modification =} \KeywordTok{update.traits}\NormalTok{(}\DataTypeTok{process =}\NormalTok{ OU.process))}
+
+\CommentTok{\#\# Set the simulation parameters}
+\NormalTok{bd.params \textless{}{-}}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{extinction =} \DecValTok{0}\NormalTok{, }\DataTypeTok{speciation =} \DecValTok{1}\NormalTok{)}
+\NormalTok{stop.rule \textless{}{-}}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{max.time =} \DecValTok{6}\NormalTok{)}
+\NormalTok{traits    \textless{}{-}}\StringTok{ }\KeywordTok{make.traits}\NormalTok{()}
+
+\CommentTok{\#\# Run the simulations}
 \KeywordTok{set.seed}\NormalTok{(}\DecValTok{1}\NormalTok{)}
-\NormalTok{test <-}\StringTok{ }\KeywordTok{dads}\NormalTok{(bd.params, stop.rule, traits, }\DataTypeTok{null.error =} \DecValTok{20}\NormalTok{)}
-\KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow =} \KeywordTok{c}\NormalTok{(}\DecValTok{4}\NormalTok{,}\DecValTok{2}\NormalTok{))}
-\KeywordTok{plot}\NormalTok{(test}\OperatorTok{$}\NormalTok{tree)}
-\KeywordTok{plot.dads}\NormalTok{(test, }\DataTypeTok{main =} \StringTok{"random tree + trait"}\NormalTok{)}
-
-\CommentTok{## Test with modifiers}
+\NormalTok{no\_change \textless{}{-}}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ bd.params,}
+                  \DataTypeTok{stop.rule =}\NormalTok{ stop.rule,}
+                  \DataTypeTok{traits    =}\NormalTok{ traits)}
 \KeywordTok{set.seed}\NormalTok{(}\DecValTok{1}\NormalTok{)}
-\NormalTok{trait_table <-}\StringTok{ }\OtherTok{NULL}
-\NormalTok{test <-}\StringTok{ }\KeywordTok{dads}\NormalTok{(bd.params, stop.rule, traits,}
-                                \DataTypeTok{modifiers =}\NormalTok{ modify_speciation,}
-                                \DataTypeTok{null.error =} \DecValTok{20}\NormalTok{)}
-\KeywordTok{plot}\NormalTok{(test}\OperatorTok{$}\NormalTok{tree)}
-\KeywordTok{plot.dads}\NormalTok{(test, }\DataTypeTok{main =} \StringTok{"Skewed speciation"}\NormalTok{)}
-
+\NormalTok{process\_change \textless{}{-}}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ bd.params,}
+                       \DataTypeTok{stop.rule =}\NormalTok{ stop.rule,}
+                       \DataTypeTok{traits    =}\NormalTok{ traits,}
+                       \DataTypeTok{events    =}\NormalTok{ change\_process)}
+\CommentTok{\#\# Plot the results}
+\KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow =} \KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{,}\DecValTok{2}\NormalTok{))}
+\KeywordTok{plot}\NormalTok{(no\_change, }\DataTypeTok{ylim =} \KeywordTok{c}\NormalTok{(}\OperatorTok{{-}}\DecValTok{7}\NormalTok{, }\DecValTok{7}\NormalTok{))}
+\KeywordTok{plot}\NormalTok{(process\_change, }\DataTypeTok{ylim =} \KeywordTok{c}\NormalTok{(}\OperatorTok{{-}}\DecValTok{7}\NormalTok{, }\DecValTok{7}\NormalTok{))}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-76-1.pdf}
+
+\hypertarget{founding}{%
+\section{Founding events}\label{founding}}
+
+Founding events are a specific \texttt{target} for \texttt{events} that allows you to\ldots{} simulate a birth-death process within the current one!
+It basically allows you to simulate a specific \texttt{dads} process i.e.~using the \texttt{dads} function with it's own \texttt{traits}, \texttt{modifiers}, \texttt{bd.params} and \texttt{events} (and yes, that's including \texttt{events} that have their own \texttt{founding} event)\ldots{}
+This is basically the ultimate nested boss of modularity!
+
+The founding event will basically run an internal \texttt{dads} process (i.e.~simulating a tree and, optionally, some data) resulting in a founding sub-tree.
+It will then branch this sub-tree to the rest of the simulation that continued normally in the mean time.
+You can specify the founding event using the inbuilt \texttt{founding.event} modification event.
+This function intakes the exact same arguments as \texttt{dads} to simulate the sub-tree with its own parameters.
+Additionally, we will use the \texttt{additional.args} argument from \texttt{make.events} to specify a prefix for the founding tree tips (to make them easier to distinguish).
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# Set up parameters}
+\NormalTok{stop.rule \textless{}{-}}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{max.time =} \DecValTok{4}\NormalTok{)}
+\NormalTok{bd.params \textless{}{-}}\StringTok{ }\KeywordTok{make.bd.params}\NormalTok{(}\DataTypeTok{speciation =} \DecValTok{1}\NormalTok{, }\DataTypeTok{extinction =} \FloatTok{0.3}\NormalTok{)}
+
+\CommentTok{\#\# Events that generate a new process (founding tree {-} with no extinction)}
+\NormalTok{founding\_event \textless{}{-}}\StringTok{ }\KeywordTok{make.events}\NormalTok{(}
+                  \DataTypeTok{target       =} \StringTok{"founding"}\NormalTok{,}
+                  \DataTypeTok{condition    =} \KeywordTok{taxa.condition}\NormalTok{(}\DecValTok{10}\NormalTok{),}
+                  \DataTypeTok{modification =} \KeywordTok{founding.event}\NormalTok{(}
+                                    \DataTypeTok{bd.params =} \KeywordTok{make.bd.params}\NormalTok{(}\DataTypeTok{speciation =} \DecValTok{2}\NormalTok{,}
+                                                               \DataTypeTok{extinction =} \DecValTok{0}\NormalTok{)),}
+                  \DataTypeTok{additional.args =} \KeywordTok{list}\NormalTok{(}\DataTypeTok{prefix =} \StringTok{"founding\_"}\NormalTok{))}
+
+\CommentTok{\#\# Simulations}
+\KeywordTok{set.seed}\NormalTok{(}\DecValTok{11}\NormalTok{)}
+\NormalTok{founding\_tree \textless{}{-}}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ bd.params,}
+                      \DataTypeTok{stop.rule =}\NormalTok{ stop.rule,}
+                      \DataTypeTok{events    =}\NormalTok{ founding\_event)}
+\KeywordTok{plot}\NormalTok{(founding\_tree, }\DataTypeTok{cex =} \FloatTok{0.4}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-77-1.pdf}
+
+Note that the nestedness here is potentially endless, for example, you can parse an \texttt{event} argument to the \texttt{founding.event} that will generate another founding event, etc\ldots{}
+
+\hypertarget{others}{%
+\chapter{Other functionalities}\label{others}}
+
+\hypertarget{makebdparams}{%
+\section{\texorpdfstring{\texttt{make.bd.params}}{make.bd.params}}\label{makebdparams}}
+
+In most examples above, birth-death parameters where set as fixed values, i.e.~a speciation rate of \(\lambda\) and an extinction rate of \(\mu\) throughout the simulations (with maybe some events modifying these rates).
+However, it is of course possible to set these rates as specific or changing distributions!
+You can do this using the \texttt{make.bd.params} function a provide either a \texttt{vector} of values:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# An example where the speciation is randomly sampled among three values}
+\NormalTok{my\_bd\_params \textless{}{-}}\StringTok{ }\KeywordTok{make.bd.params}\NormalTok{(}\DataTypeTok{speciation =} \KeywordTok{c}\NormalTok{(}\DecValTok{1}\OperatorTok{/}\DecValTok{3}\NormalTok{,}\DecValTok{42}\NormalTok{))}
+
+\CommentTok{\#\# Building a tree using this set of parameters}
+\KeywordTok{set.seed}\NormalTok{(}\DecValTok{123}\NormalTok{)}
+\KeywordTok{plot}\NormalTok{(}\KeywordTok{dads}\NormalTok{(}\DataTypeTok{stop.rule =} \KeywordTok{list}\NormalTok{(}\DataTypeTok{max.taxa =} \DecValTok{50}\NormalTok{), }\DataTypeTok{bd.params =}\NormalTok{ my\_bd\_params))}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-79-1.pdf}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# Note the regions in the tree with short branches}
+\CommentTok{\#\# (that\textquotesingle{}s the speciation being 1/3 while the others are speciation = 42) }
+\end{Highlighting}
+\end{Shaded}
+
+Or directly a function from which to sample:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# Another example where speciation is drawn from the interval (0, 1)}
+\KeywordTok{make.bd.params}\NormalTok{(}\DataTypeTok{speciation =}\NormalTok{ runif)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+##  ---- dads birth-death parameters object ---- 
+## speciation: runif.
+## extinction: 0.
+\end{verbatim}
+
+In this example, the \texttt{"bd.params"} object passed to the \texttt{dads} function will allow the birth-death process to sample the speciation parameter each time it is called (e.g.~during the speciation/extinction step, the branch length step, etc.).
+If using a function, you can fine tune the arguments to be passed to that function using the \texttt{speciation.args} or the \texttt{extinction.args} arguments (as a named list matching the function's arguments):
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# Speciation is drawn from the interval (0.5, 1.5)}
+\KeywordTok{make.bd.params}\NormalTok{(}\DataTypeTok{speciation =}\NormalTok{ runif,}
+               \DataTypeTok{speciation.args =} \KeywordTok{list}\NormalTok{(}\DataTypeTok{min =} \FloatTok{0.5}\NormalTok{, }\DataTypeTok{max =} \FloatTok{1.5}\NormalTok{))}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+##  ---- dads birth-death parameters object ---- 
+## speciation: runif (with optional arguments).
+## extinction: 0.
+\end{verbatim}
+
+When using distributions for both the speciation and extinction parameters, you can run into the undesired problem of having an extinction rate that is higher than your speciation rate (and thus your tree dying out - unless this is the desired behaviour).
+You can avoid this problem by using the \texttt{joint} distribution argument this will ensure that the sampling of the extinction rate is always lower or equal to the speciation rate.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# Joint speciation and extinction sampled from uniform distribution }
+\CommentTok{\#\# with speciation always \textgreater{}= to extinction}
+\KeywordTok{make.bd.params}\NormalTok{(}\DataTypeTok{speciation =}\NormalTok{ runif, }\DataTypeTok{extinction =}\NormalTok{ runif, }\DataTypeTok{joint =} \OtherTok{TRUE}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+##  ---- dads birth-death parameters object ---- 
+## joint sampling for:
+## speciation: runif.
+## extinction: runif.
+\end{verbatim}
+
+Finally to avoid negative sampling values, you can use the argument \texttt{absolute} that will make all the sampled values positive.
+This argument is set to \texttt{TRUE} by default so you shouldn't have to worry about it most of the time unless you specifically need negative sampled values for your parameters.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# Making the speciation sampling always positive}
+\KeywordTok{make.bd.params}\NormalTok{(}\DataTypeTok{speciation =}\NormalTok{ rnorm, }\DataTypeTok{absolute =} \OtherTok{TRUE}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+##  ---- dads birth-death parameters object ---- 
+## speciation: rnorm.
+## extinction: 0.
+## (using absolute values)
+\end{verbatim}
+
+The two other possible arguments for this function are \texttt{test} and \texttt{update} that work the same as for \protect\hyperlink{maketraits}{\texttt{make.traits}} or \protect\hyperlink{makemodifiers}{\texttt{make.modifiers}}
+
+If you are a visual person and your \texttt{bd.params} objects are getting a bit too complicated to remember, you can always quickly plot them (the function will sample from the \texttt{bd.params} object and show the results):
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow =} \KeywordTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{,}\DecValTok{2}\NormalTok{))}
+\KeywordTok{plot}\NormalTok{(}\KeywordTok{make.bd.params}\NormalTok{(), }\DataTypeTok{main =} \StringTok{"Default birth{-}death parameters"}\NormalTok{)}
+\KeywordTok{plot}\NormalTok{(}\KeywordTok{make.bd.params}\NormalTok{(}\DataTypeTok{speciation =}\NormalTok{ runif,}
+               \DataTypeTok{speciation.args =} \KeywordTok{list}\NormalTok{(}\DataTypeTok{min =} \FloatTok{0.5}\NormalTok{, }\DataTypeTok{max =} \FloatTok{1.5}\NormalTok{)),}
+     \DataTypeTok{main =} \StringTok{"Uniform speciation between 0.5 and 1.5}\CharTok{\textbackslash{}n}\StringTok{(no extinction)"}\NormalTok{)}
+\KeywordTok{plot}\NormalTok{(}\KeywordTok{make.bd.params}\NormalTok{(}\DataTypeTok{speciation =}\NormalTok{ runif,}
+                    \DataTypeTok{speciation.args =} \KeywordTok{list}\NormalTok{(}\DataTypeTok{min =} \FloatTok{0.5}\NormalTok{, }\DataTypeTok{max =} \FloatTok{1.5}\NormalTok{)),}
+     \DataTypeTok{main =} \StringTok{"Uniform speciation between 0.5 and 1.5}\CharTok{\textbackslash{}n}\StringTok{(no extinction)"}\NormalTok{)}
+\KeywordTok{plot}\NormalTok{(}\KeywordTok{make.bd.params}\NormalTok{(}\DataTypeTok{speciation =}\NormalTok{ runif, }\DataTypeTok{extinction =}\NormalTok{ runif, }\DataTypeTok{joint =} \OtherTok{TRUE}\NormalTok{),}
+     \DataTypeTok{main =} \StringTok{"Joint uniform speciation and extinction"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-84-1.pdf}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{plot}\NormalTok{(}\KeywordTok{make.bd.params}\NormalTok{(}\DataTypeTok{speciation =}\NormalTok{ rnorm, }\DataTypeTok{extinction =}\NormalTok{ runif,}
+                    \DataTypeTok{joint =} \OtherTok{FALSE}\NormalTok{, }\DataTypeTok{abs =} \OtherTok{FALSE}\NormalTok{),}
+     \DataTypeTok{main =} \StringTok{"Disjoint normal speciation and uniform extinction"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-84-2.pdf}
+
+\hypertarget{dropthings}{%
+\section{\texorpdfstring{\texttt{drop.things}}{drop.things}}\label{dropthings}}
+
+You can use the function \texttt{drop.things} to drop specific elements of the tree and data at the same time by providing the argument \texttt{what} to be \texttt{"fossils"} for tips that went extinct, \texttt{"livings"} tips that where alive at the end of the simulation or \texttt{"singles"} for internal nodes.
+Alternatively you can use the function aliases \texttt{drop.fossils}, \texttt{drop.livings} or \texttt{drop.singles} for the exact same results:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# A random tree with fossils and traits and internal nodes every 0.5 times}
+\KeywordTok{set.seed}\NormalTok{(}\DecValTok{3}\NormalTok{)}
+\NormalTok{my\_data \textless{}{-}}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{stop.rule =} \KeywordTok{list}\NormalTok{(}\DataTypeTok{max.taxa =} \DecValTok{20}\NormalTok{),}
+                \DataTypeTok{bd.params =} \KeywordTok{list}\NormalTok{(}\DataTypeTok{speciation =} \DecValTok{1}\NormalTok{, }\DataTypeTok{extinction =} \DecValTok{1}\OperatorTok{/}\DecValTok{3}\NormalTok{),}
+                \DataTypeTok{traits    =} \KeywordTok{make.traits}\NormalTok{(), }\DataTypeTok{save.steps =} \FloatTok{0.5}\NormalTok{)}
+
+\CommentTok{\#\# A tree with 20 tips and 54 nodes}
+\NormalTok{my\_data}\OperatorTok{$}\NormalTok{tree}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## 
+## Phylogenetic tree with 20 tips and 54 internal nodes.
+## 
+## Tip labels:
+##   t1, t2, t3, t4, t5, t6, ...
+## Node labels:
+##   n1, n2, n3, n4, n5, n6, ...
+## 
+## Rooted; includes branch lengths.
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# And a dataset with 74 rows}
+\KeywordTok{dim}\NormalTok{(my\_data}\OperatorTok{$}\NormalTok{data)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## [1] 74  1
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# Removing the fossil species}
+\KeywordTok{drop.things}\NormalTok{(my\_data, }\DataTypeTok{what =} \StringTok{"fossils"}\NormalTok{)}\OperatorTok{$}\NormalTok{tree}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## 
+## Phylogenetic tree with 8 tips and 31 internal nodes.
+## 
+## Tip labels:
+##   t13, t14, t15, t16, t17, t18, ...
+## Node labels:
+##   n1, n2, n7, n10, n11, n13, ...
+## 
+## Rooted; includes branch lengths.
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{dim}\NormalTok{(}\KeywordTok{drop.fossils}\NormalTok{(my\_data)}\OperatorTok{$}\NormalTok{data)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## [1] 39  1
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# Removing the living species}
+\KeywordTok{drop.things}\NormalTok{(my\_data, }\DataTypeTok{what =} \StringTok{"livings"}\NormalTok{)}\OperatorTok{$}\NormalTok{tree}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## 
+## Phylogenetic tree with 12 tips and 37 internal nodes.
+## 
+## Tip labels:
+##   t1, t2, t3, t4, t5, t6, ...
+## Node labels:
+##   n1, n2, n3, n4, n5, n6, ...
+## 
+## Rooted; includes branch lengths.
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{dim}\NormalTok{(}\KeywordTok{drop.livings}\NormalTok{(my\_data)}\OperatorTok{$}\NormalTok{data)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## [1] 49  1
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# Removing the internal nodes}
+\KeywordTok{drop.things}\NormalTok{(my\_data, }\DataTypeTok{what =} \StringTok{"singles"}\NormalTok{)}\OperatorTok{$}\NormalTok{tree}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## 
+## Phylogenetic tree with 20 tips and 19 internal nodes.
+## 
+## Tip labels:
+##   t1, t2, t3, t4, t5, t6, ...
+## Node labels:
+##   n1, n7, n33, n39, n54, n40, ...
+## 
+## Rooted; includes branch lengths.
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{dim}\NormalTok{(}\KeywordTok{drop.singles}\NormalTok{(my\_data)}\OperatorTok{$}\NormalTok{data)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## [1] 39  1
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# Removing the internal nodes AND the fossils}
+\KeywordTok{drop.singles}\NormalTok{(}\KeywordTok{drop.fossils}\NormalTok{(my\_data))}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+##  ---- dads object ---- 
+## Simulated diversity data (x$tree):
+## 
+## Phylogenetic tree with 8 tips and 7 internal nodes.
+## 
+## Tip labels:
+##   t13, t14, t15, t16, t17, t18, ...
+## Node labels:
+##   n7, n33, n54, n53, n34, n45, ...
+## 
+## Rooted; includes branch lengths.
+## 
+## Simulated disparity data (x$data):
+## 1 trait for 1 process (A) with one starting value (0).
+\end{verbatim}
+
+\hypertarget{dads-internal-utilities}{%
+\section{\texorpdfstring{\texttt{"dads"} internal utilities}{"dads" internal utilities}}\label{dads-internal-utilities}}
+
+The package also provides utilities for internal functions, namely for designing \texttt{modifiers} or \texttt{events} more easily.
+These functions don't do anything useful on their own but are optimised to be used internally in \texttt{dads}.
+For all these functions, you can look at the internal manual for an example (i.e.~using \texttt{?\textless{}function\_name\textgreater{}})
+
+So far the package has the following internals:
+
+\begin{longtable}[]{@{}lll@{}}
+\toprule
+\begin{minipage}[b]{0.20\columnwidth}\raggedright
+Function\strut
+\end{minipage} & \begin{minipage}[b]{0.28\columnwidth}\raggedright
+What it does\strut
+\end{minipage} & \begin{minipage}[b]{0.43\columnwidth}\raggedright
+Where can it be used\strut
+\end{minipage}\tabularnewline
+\midrule
+\endhead
+\begin{minipage}[t]{0.20\columnwidth}\raggedright
+\texttt{parent.traits}\strut
+\end{minipage} & \begin{minipage}[t]{0.28\columnwidth}\raggedright
+selects the trait values of the current lineage's parents (i.e.~direct ancestor)\strut
+\end{minipage} & \begin{minipage}[t]{0.43\columnwidth}\raggedright
+in \texttt{make.modifiers}\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.20\columnwidth}\raggedright
+\texttt{taxa.condition}\strut
+\end{minipage} & \begin{minipage}[t]{0.28\columnwidth}\raggedright
+provides a trigger for an \texttt{event} dependent on the number of taxa\strut
+\end{minipage} & \begin{minipage}[t]{0.43\columnwidth}\raggedright
+in \texttt{make.events}\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.20\columnwidth}\raggedright
+\texttt{time.condition}\strut
+\end{minipage} & \begin{minipage}[t]{0.28\columnwidth}\raggedright
+provides a trigger for an \texttt{event} dependent on time\strut
+\end{minipage} & \begin{minipage}[t]{0.43\columnwidth}\raggedright
+in \texttt{make.events}\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.20\columnwidth}\raggedright
+\texttt{trait.condition}\strut
+\end{minipage} & \begin{minipage}[t]{0.28\columnwidth}\raggedright
+provides a trigger for an \texttt{event} dependent on trait values\strut
+\end{minipage} & \begin{minipage}[t]{0.43\columnwidth}\raggedright
+in \texttt{make.events}\strut
+\end{minipage}\tabularnewline
+\bottomrule
+\end{longtable}
+
+\texttt{"dads"} objects can be directly plotted in \texttt{dads} using the S3 \texttt{plot.dads} function (or just \texttt{plot(x)} if \texttt{x} is of class \texttt{"dads"}).
+
+\hypertarget{plotting-traits}{%
+\chapter{Plotting traits}\label{plotting-traits}}
+
+\texttt{"dads"} \texttt{"traits"} objects are covered in the \protect\hyperlink{maketraits}{traits section}.
+You can use \texttt{plot.dads} to plot them by choosing which specific trait to plot using the \texttt{trait} argument (default is 1):
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# Making a list of three traits}
+\NormalTok{list\_of\_traits \textless{}{-}}\StringTok{ }\KeywordTok{make.traits}\NormalTok{(}\DataTypeTok{process =} \KeywordTok{c}\NormalTok{(no.process, BM.process, OU.process), }\DataTypeTok{trait.names =} \KeywordTok{c}\NormalTok{(}\StringTok{"No process (normal)"}\NormalTok{, }\StringTok{"Brownian motion"}\NormalTok{, }\StringTok{"Ornstein{-}Uhlenbeck"}\NormalTok{))}
+\CommentTok{\#\# Plotting each trait separately}
+\KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow =} \KeywordTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{, }\DecValTok{2}\NormalTok{))}
+\KeywordTok{plot}\NormalTok{(list\_of\_traits, }\DataTypeTok{trait =} \DecValTok{1}\NormalTok{)}
+\CommentTok{\#\# Using different colours options}
+\KeywordTok{plot}\NormalTok{(list\_of\_traits, }\DataTypeTok{trait =} \DecValTok{2}\NormalTok{, }\DataTypeTok{col =} \KeywordTok{c}\NormalTok{(}\StringTok{"red"}\NormalTok{, }\StringTok{"purple"}\NormalTok{, }\StringTok{"pink"}\NormalTok{))}
+\CommentTok{\#\# Not using the default plot name}
+\KeywordTok{plot}\NormalTok{(list\_of\_traits, }\DataTypeTok{trait =} \DecValTok{3}\NormalTok{, }\DataTypeTok{main =} \StringTok{"OU process"}\NormalTok{,}
+     \DataTypeTok{cent.tend =}\NormalTok{ sd, }\DataTypeTok{quantiles =} \KeywordTok{c}\NormalTok{(}\DecValTok{10}\NormalTok{, }\DecValTok{30}\NormalTok{))}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-87-1.pdf}
+
+You can also control the number of replicates in the simulation by using the \texttt{simulations} option (the default is 50).
+Bigger numbers leads to more time but smoother looking plots while smaller ones are more stochastic:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow =} \KeywordTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{,}\DecValTok{1}\NormalTok{))}
+\KeywordTok{plot}\NormalTok{(list\_of\_traits, }\DataTypeTok{trait =} \DecValTok{2}\NormalTok{, }\DataTypeTok{simulations =} \DecValTok{10}\NormalTok{, }\DataTypeTok{main =} \StringTok{"10 BM simulations"}\NormalTok{)}
+\KeywordTok{plot}\NormalTok{(list\_of\_traits, }\DataTypeTok{trait =} \DecValTok{2}\NormalTok{, }\DataTypeTok{simulations =} \DecValTok{1000}\NormalTok{, }\DataTypeTok{main =} \StringTok{"1k BM simulations"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-88-1.pdf}
+
+\hypertarget{plotting-dads-results}{%
+\chapter{Plotting dads results}\label{plotting-dads-results}}
+
+If \texttt{dads} is used to plot only a tree (and outputs a \texttt{"phylo"} object), you can use the function \texttt{plot.phylo} from the \texttt{ape} package to plot your tree.
+You'll find all the tree plotting option in the \texttt{?plot.phylo} manual page.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# A simple pure birth tree}
+\NormalTok{my\_tree \textless{}{-}}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{stop.rule =} \KeywordTok{list}\NormalTok{(}\DataTypeTok{max.taxa =} \DecValTok{20}\NormalTok{))}
+\KeywordTok{plot}\NormalTok{(my\_tree, }\DataTypeTok{main =} \StringTok{"Plotting a }\CharTok{\textbackslash{}"}\StringTok{phylo}\CharTok{\textbackslash{}"}\StringTok{ object"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-89-1.pdf}
+
+However, if you also simulated a trait along with the tree you can use the \texttt{plot.dads} function to plot both the tree and the trait:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# A simple pure birth tree with a BM process}
+\NormalTok{my\_tree \textless{}{-}}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{stop.rule =} \KeywordTok{list}\NormalTok{(}\DataTypeTok{max.taxa =} \DecValTok{20}\NormalTok{), }\DataTypeTok{traits =} \KeywordTok{make.traits}\NormalTok{())}
+\CommentTok{\#\# Playing with the default options}
+\KeywordTok{plot}\NormalTok{(my\_tree, }\DataTypeTok{main =} \StringTok{"A tree and traits"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-90-1.pdf}
+
+By default, elements are coloured as follows: nodes and tips are points and coloured in blue if they are tips (light blue if they are fossils) and nodes are in orange.
+Branches linking them are grey lines.
+You can of course change this colour palette to something of you preference by calling the normal arguments that can be passed to \texttt{points} or \texttt{lines}.
+For example \texttt{pch} for the point type or \texttt{lty} for the line type.
+Furthermore, time is plotted conventionally from left to right (left is towards the past, right is towards the present) but you can change that by specifying \texttt{xlim}.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# Playing with some more options}
+\KeywordTok{plot}\NormalTok{(my\_tree, }\DataTypeTok{main =} \StringTok{"A tree and traits"}\NormalTok{,}
+     \CommentTok{\#\# Changing nodes colours and type}
+     \DataTypeTok{col =} \KeywordTok{c}\NormalTok{(}\DataTypeTok{tips =} \StringTok{"pink"}\NormalTok{, }\DataTypeTok{nodes =} \StringTok{"purple"}\NormalTok{), }\DataTypeTok{pch =} \DecValTok{21}\NormalTok{,}
+     \CommentTok{\#\# Changing edges colour and type}
+     \DataTypeTok{lwd =} \DecValTok{2}\NormalTok{, }\DataTypeTok{lty =} \DecValTok{3}\NormalTok{, }\DataTypeTok{edges =} \StringTok{"yellow"}\NormalTok{,}
+     \CommentTok{\#\# Changing the x axis orientation and label}
+     \DataTypeTok{xlim =} \KeywordTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{, }\FloatTok{2.5}\NormalTok{), }\DataTypeTok{xlab =} \StringTok{"Time into the past"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-91-1.pdf}
+
+\begin{quote}
+Note that you can modify colours using the \texttt{col} argument by providing a clear indication of what you want to colour (e.g.~\texttt{col\ =\ c(tips\ =\ "blue",\ fossils\ =\ "orange")} will apply the colours to the unambiguously named elements) or by providing a function that will scale the colours with time (e.g.~\texttt{col\ =\ grDevices::heat.colors} will use the specified function to change the elements colours with time).
+\end{quote}
+
+And you can add a default legend by using \texttt{legend\ =\ TRUE} (if you don't want to add the default legend you can add it after your plot using \texttt{legend(...)}).
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# Adding the default legend}
+\KeywordTok{plot}\NormalTok{(my\_tree, }\DataTypeTok{legend =} \OtherTok{TRUE}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-92-1.pdf}
+
+\hypertarget{d-version}{%
+\section{3D version!}\label{d-version}}
+
+Of course, if you're simulating multiple traits, you can always plot different ones.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# Specifying a 3D trait process}
+\NormalTok{my\_3D\_trait \textless{}{-}}\StringTok{ }\KeywordTok{make.traits}\NormalTok{(}\DataTypeTok{n =} \DecValTok{3}\NormalTok{, )}
+\CommentTok{\#\# Simulating a birth death tree with that 3D trait}
+\NormalTok{my\_data \textless{}{-}}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =} \KeywordTok{list}\NormalTok{(}\DataTypeTok{extinction =} \FloatTok{0.2}\NormalTok{),}
+                \DataTypeTok{stop.rule =} \KeywordTok{list}\NormalTok{(}\DataTypeTok{max.living =} \DecValTok{50}\NormalTok{),}
+                \DataTypeTok{traits    =}\NormalTok{ my\_3D\_trait)}
+\end{Highlighting}
+\end{Shaded}
+
+You can toggle which trait to plot using the \texttt{trait} option.
+Either by providing a single value to plot that specific trait against time or by providing two traits.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow =} \KeywordTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{,}\DecValTok{1}\NormalTok{))}
+\CommentTok{\#\# Plotting the second trait}
+\KeywordTok{plot}\NormalTok{(my\_data, }\DataTypeTok{trait =} \DecValTok{2}\NormalTok{, }\DataTypeTok{main =} \StringTok{"Trait 2"}\NormalTok{)}
+\CommentTok{\#\# Plotting the correlation between trait 1 and 2}
+\KeywordTok{plot}\NormalTok{(my\_data, }\DataTypeTok{trait =} \KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{,}\DecValTok{2}\NormalTok{), }\DataTypeTok{main =} \StringTok{"Traits 1 and 2"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-94-1.pdf}
+
+As mentioned above, the \texttt{col} argument can take a function for scaling the elements colours with time.
+This can be useful for adding time a third dimension to these 2D plots:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# Plotting the correlation between trait 1 and 2}
+\CommentTok{\#\# with time as a 3rd dimensions}
+\KeywordTok{plot}\NormalTok{(my\_data, }\DataTypeTok{trait =} \KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{,}\DecValTok{2}\NormalTok{), }\DataTypeTok{main =} \StringTok{"Traits 1 and 2"}\NormalTok{,}
+     \DataTypeTok{col =}\NormalTok{ grDevices}\OperatorTok{::}\NormalTok{heat.colors, }\DataTypeTok{legend =} \OtherTok{TRUE}\NormalTok{,}
+     \CommentTok{\#\# Highlighting the tips in black for visibility}
+     \DataTypeTok{tips.nodes =} \StringTok{"black"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-95-1.pdf}
+
+But there's more! You can also plot these results using actual 3D plots that you can spin and all that!
+This is done through the \texttt{rgl} package and activated in \texttt{dads} using the option \texttt{use.3D\ =\ TRUE}.
+
+If two traits are provided, the 3rd dimensions is going to be time by default:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# Plotting the tree and 2 traits in 3D: woah!}
+\KeywordTok{plot}\NormalTok{(my\_data, }\DataTypeTok{trait =} \KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{, }\DecValTok{2}\NormalTok{), }\DataTypeTok{use.3D =} \OtherTok{TRUE}\NormalTok{)}
+\KeywordTok{rglwidget}\NormalTok{()}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## Warning in snapshot3d(scene = x, width = width, height = height): webshot =
+## TRUE requires the webshot2 package and Chrome browser; using rgl.snapshot()
+## instead
+\end{verbatim}
+
+\begin{verbatim}
+## Warning in rgl.snapshot(filename, fmt, top): this build of rgl does not support
+## snapshots
+\end{verbatim}
+
+\includegraphics{../../../../../../tmp/Rtmp8BGjPN/file97bd69ab2a6.png}
+
+If three traits are provide, you can set up a fourth coloured dimension as time (similarly to the 2D plots with the colour gradients).
+This uses the \texttt{rgl} functions, namely \texttt{lines3d}, \texttt{points3d} (if \texttt{type\ =\ "p"}; default) or \texttt{sphere3d} (if \texttt{type\ =\ "s"}).
+The additional arguments \texttt{...} are directly handled and attributed to the corresponding function.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# Plotting the tree and 3 traits in 3D}
+\KeywordTok{plot}\NormalTok{(my\_data, }\DataTypeTok{trait =} \KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{, }\DecValTok{2}\NormalTok{, }\DecValTok{3}\NormalTok{), }\DataTypeTok{use.3D =} \OtherTok{TRUE}\NormalTok{,}
+     \DataTypeTok{col =}\NormalTok{ grDevices}\OperatorTok{::}\NormalTok{heat.colors, }\DataTypeTok{type =} \StringTok{"s"}\NormalTok{, }\DataTypeTok{radius =} \FloatTok{0.1}\NormalTok{)}
+\KeywordTok{rglwidget}\NormalTok{()}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## Warning in snapshot3d(scene = x, width = width, height = height): webshot =
+## TRUE requires the webshot2 package and Chrome browser; using rgl.snapshot()
+## instead
+\end{verbatim}
+
+\begin{verbatim}
+## Warning in rgl.snapshot(filename, fmt, top): this build of rgl does not support
+## snapshots
+\end{verbatim}
+
+\includegraphics{../../../../../../tmp/Rtmp8BGjPN/file97bd7e5353fb.png}
+
+\hypertarget{specific-examples}{%
+\chapter{Specific examples}\label{specific-examples}}
+
+In this section I will illustrate a series of examples of specific more complex scenarios.
+They can be all run independently and used as a basis for your own specific scenarios.
+
+Here is a table summarising which functionalities are used in which example:
+
+\begin{longtable}[]{@{}ll@{}}
+\toprule
+\begin{minipage}[b]{0.57\columnwidth}\raggedright
+Functionality showcase\strut
+\end{minipage} & \begin{minipage}[b]{0.37\columnwidth}\raggedright
+Example\strut
+\end{minipage}\tabularnewline
+\midrule
+\endhead
+\begin{minipage}[t]{0.57\columnwidth}\raggedright
+\texttt{bd.params}\strut
+\end{minipage} & \begin{minipage}[t]{0.37\columnwidth}\raggedright
+\protect\hyperlink{EGred_spec}{reducing speciation rate}, \protect\hyperlink{EG_founding_purebirth}{generating a subtree with no extinction}\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.57\columnwidth}\raggedright
+\texttt{make.events}\strut
+\end{minipage} & \begin{minipage}[t]{0.37\columnwidth}\raggedright
+\protect\hyperlink{EGrandom_ext}{mass extinction}, \protect\hyperlink{EGneg_ext}{negative trait extinction}, \protect\hyperlink{EGbg_ext}{background extinction}, \protect\hyperlink{EGred_spec}{reducing speciation rate}, \protect\hyperlink{EG_change_trait}{changing trait process}, \protect\hyperlink{EG_change_correlation}{changing trait correlation}, \protect\hyperlink{EG_change_modif}{changing modifiers}, \protect\hyperlink{EG_modify_brlen}{change branch length}, \protect\hyperlink{EG_founding_purebirth}{generating a subtree with no extinction}, \href{EG_founding_traits}{generating a subtree with a different process}\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.57\columnwidth}\raggedright
+\texttt{make.traits}\strut
+\end{minipage} & \begin{minipage}[t]{0.37\columnwidth}\raggedright
+\protect\hyperlink{EGneg_ext}{negative trait extinction}, \protect\hyperlink{EG_change_trait}{changing trait process}, \protect\hyperlink{EG_change_correlation}{changing trait correlation}, \href{EG_founding_traits}{generating a subtree with a different process}\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.57\columnwidth}\raggedright
+\texttt{make.modifiers}\strut
+\end{minipage} & \begin{minipage}[t]{0.37\columnwidth}\raggedright
+\protect\hyperlink{EG_change_modif}{changing modifiers}, \protect\hyperlink{EG_modify_brlen}{change branch length}\strut
+\end{minipage}\tabularnewline
+\bottomrule
+\end{longtable}
+
+And more specifically for the events:
+
+\begin{longtable}[]{@{}ll@{}}
+\toprule
+\begin{minipage}[b]{0.55\columnwidth}\raggedright
+Functionality showcase\strut
+\end{minipage} & \begin{minipage}[b]{0.39\columnwidth}\raggedright
+Example\strut
+\end{minipage}\tabularnewline
+\midrule
+\endhead
+\begin{minipage}[t]{0.55\columnwidth}\raggedright
+\texttt{time.condition}\strut
+\end{minipage} & \begin{minipage}[t]{0.39\columnwidth}\raggedright
+\protect\hyperlink{EGrandom_ext}{mass extinction}, \protect\hyperlink{EGneg_ext}{negative trait extinction}, \protect\hyperlink{EGred_spec}{reducing speciation rate}, \protect\hyperlink{EG_change_trait}{changing trait process}, \protect\hyperlink{EG_change_modif}{changing modifiers}\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.55\columnwidth}\raggedright
+\texttt{random.extinction}\strut
+\end{minipage} & \begin{minipage}[t]{0.39\columnwidth}\raggedright
+\protect\hyperlink{EGrandom_ext}{mass extinction}\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.55\columnwidth}\raggedright
+\texttt{trait.extinction}\strut
+\end{minipage} & \begin{minipage}[t]{0.39\columnwidth}\raggedright
+\protect\hyperlink{EGneg_ext}{negative trait extinction}\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.55\columnwidth}\raggedright
+\texttt{taxa.condition}\strut
+\end{minipage} & \begin{minipage}[t]{0.39\columnwidth}\raggedright
+\protect\hyperlink{EGbg_ext}{background extinction}, \protect\hyperlink{EG_modify_brlen}{change branch length}\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.55\columnwidth}\raggedright
+\texttt{update.bd.params}\strut
+\end{minipage} & \begin{minipage}[t]{0.39\columnwidth}\raggedright
+\protect\hyperlink{EGbg_ext}{background extinction}, \protect\hyperlink{EGred_spec}{reducing speciation rate}\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.55\columnwidth}\raggedright
+\texttt{update.traits}\strut
+\end{minipage} & \begin{minipage}[t]{0.39\columnwidth}\raggedright
+\protect\hyperlink{EG_change_trait}{changing trait process}, \protect\hyperlink{EG_change_correlation}{changing trait correlation}\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.55\columnwidth}\raggedright
+\texttt{trait.condition}\strut
+\end{minipage} & \begin{minipage}[t]{0.39\columnwidth}\raggedright
+\protect\hyperlink{EG_change_correlation}{changing trait correlation}, \href{EG_founding_traits}{generating a subtree with a different process}\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.55\columnwidth}\raggedright
+\texttt{update.modifiers}\strut
+\end{minipage} & \begin{minipage}[t]{0.39\columnwidth}\raggedright
+\protect\hyperlink{EG_change_modif}{changing modifiers}, \protect\hyperlink{EG_modify_brlen}{change branch length}\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.55\columnwidth}\raggedright
+\texttt{parent.traits}\strut
+\end{minipage} & \begin{minipage}[t]{0.39\columnwidth}\raggedright
+\protect\hyperlink{EG_change_modif}{changing modifiers}\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.55\columnwidth}\raggedright
+\texttt{founding.event}\strut
+\end{minipage} & \begin{minipage}[t]{0.39\columnwidth}\raggedright
+\protect\hyperlink{EG_founding_purebirth}{generating a subtree with no extinction}, \href{EG_founding_traits}{generating a subtree with a different process}\strut
+\end{minipage}\tabularnewline
+\bottomrule
+\end{longtable}
+
+\hypertarget{EGrandom_ext}{%
+\section{Random mass extinction after some time}\label{EGrandom_ext}}
+
+For this scenario, we want to generate a pure birth tree (no traits and no extinction) where 80\% of the species go extinct two thirds into the scenario.
+
+For that we first need to first set our simulation parameters: we will be running the simulation for 6 time units and with a speciation rate of 1 (and extinction of 0 - default).
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# Setting the parameters}
+\NormalTok{stop\_time\_}\DecValTok{6}\NormalTok{ \textless{}{-}}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{max.time =} \DecValTok{6}\NormalTok{)}
+\NormalTok{speciation\_}\DecValTok{1}\NormalTok{ \textless{}{-}}\StringTok{ }\KeywordTok{make.bd.params}\NormalTok{(}\DataTypeTok{speciation =} \DecValTok{1}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+We will then create a event that triggers when reaching half of the simulation (using \texttt{time.condition(2.5)}).
+This event will target \texttt{"taxa"}, i.e.~the number of species, and randomly remove 80\% of them (using \texttt{random.extinction(0.8)}):
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# 80\% mass extinction at time 4}
+\NormalTok{mass.extinction \textless{}{-}}\StringTok{ }\KeywordTok{make.events}\NormalTok{(}
+                    \DataTypeTok{condition =} \KeywordTok{time.condition}\NormalTok{(}\DecValTok{4}\NormalTok{),}
+                    \DataTypeTok{target =} \StringTok{"taxa"}\NormalTok{,}
+                    \DataTypeTok{modification =} \KeywordTok{random.extinction}\NormalTok{(}\FloatTok{0.8}\NormalTok{))}
+\end{Highlighting}
+\end{Shaded}
+
+Once these parameters are defined, we can run the simulations and plot the results:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# Running the simulations}
 \KeywordTok{set.seed}\NormalTok{(}\DecValTok{1}\NormalTok{)}
-\NormalTok{trait_table <-}\StringTok{ }\OtherTok{NULL}
-\NormalTok{test <-}\StringTok{ }\KeywordTok{dads}\NormalTok{(bd.params, stop.rule, traits,}
-                                \DataTypeTok{modifiers =}\NormalTok{ modify_brlen,}
-                                \DataTypeTok{null.error =} \DecValTok{20}\NormalTok{)}
-\KeywordTok{plot}\NormalTok{(test}\OperatorTok{$}\NormalTok{tree)}
-\KeywordTok{plot.dads}\NormalTok{(test, }\DataTypeTok{main =} \StringTok{"Skewed branch length"}\NormalTok{)}
-
+\NormalTok{results \textless{}{-}}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{stop.rule =}\NormalTok{ stop\_time\_}\DecValTok{6}\NormalTok{,}
+                \DataTypeTok{bd.params =}\NormalTok{ speciation\_}\DecValTok{1}\NormalTok{,}
+                \DataTypeTok{events    =}\NormalTok{ mass.extinction)}
+\CommentTok{\#\# Plotting the results}
+\KeywordTok{plot}\NormalTok{(results, }\DataTypeTok{show.tip.label =} \OtherTok{FALSE}\NormalTok{)}
+\KeywordTok{axisPhylo}\NormalTok{()}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-101-1.pdf}
+
+\hypertarget{EGneg_ext}{%
+\section{Species with negative trait values go extinct after some time}\label{EGneg_ext}}
+
+For this scenario, we want to generate a pure birth tree with a one dimensional Brownian Motion trait for 5 time unit.
+We then want the species with a negative trait value go extinct after 4 time unit.
+
+First we need to set up the simulation parameters:
+* The stopping rule (5 time units)
+* The birth-death parameters (speciation rate of 1)
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# Simulation parameters}
+\NormalTok{stop\_time\_}\DecValTok{5}\NormalTok{ \textless{}{-}}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{max.time =} \DecValTok{5}\NormalTok{)}
+\NormalTok{speciation\_}\DecValTok{1}\NormalTok{ \textless{}{-}}\StringTok{ }\KeywordTok{make.bd.params}\NormalTok{(}\DataTypeTok{speciation =} \DecValTok{1}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+Then set up our trait which is a one dimensional Brownian Motion
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# Trait}
+\NormalTok{simple\_bm\_trait \textless{}{-}}\StringTok{ }\KeywordTok{make.traits}\NormalTok{(}\DataTypeTok{n =} \DecValTok{1}\NormalTok{, }\DataTypeTok{process =}\NormalTok{ BM.process)}
+\end{Highlighting}
+\end{Shaded}
+
+And our extinction event which triggers after reaching time 4 (\texttt{time.condition(4)}), targets the \texttt{"taxa"} and modifies the extinction for species with traits lower than 0
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# Extinction of any tips with trait \textless{} 1 at time 4}
+\NormalTok{trait.extinction \textless{}{-}}\StringTok{ }\KeywordTok{make.events}\NormalTok{(}
+                      \DataTypeTok{target =} \StringTok{"taxa"}\NormalTok{,}
+                      \DataTypeTok{condition =} \KeywordTok{time.condition}\NormalTok{(}\DecValTok{4}\NormalTok{),}
+                      \DataTypeTok{modification =} \KeywordTok{trait.extinction}\NormalTok{(}\DataTypeTok{x =} \DecValTok{0}\NormalTok{,}
+                                                      \DataTypeTok{condition =} \StringTok{\textasciigrave{}}\DataTypeTok{\textless{}}\StringTok{\textasciigrave{}}\NormalTok{))}
+\end{Highlighting}
+\end{Shaded}
+
+Once these parameters are defined, we can run the simulations and plot the results:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# Running the simulations}
+\KeywordTok{set.seed}\NormalTok{(}\DecValTok{7}\NormalTok{)}
+\NormalTok{results \textless{}{-}}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{stop.rule =}\NormalTok{ stop\_time\_}\DecValTok{5}\NormalTok{,}
+                \DataTypeTok{bd.params =}\NormalTok{ speciation\_}\DecValTok{1}\NormalTok{,}
+                \DataTypeTok{traits    =}\NormalTok{ simple\_bm\_trait,}
+                \DataTypeTok{events    =}\NormalTok{ trait.extinction)}
+\CommentTok{\#\# Plotting the results}
+\KeywordTok{plot}\NormalTok{(results)}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-105-1.pdf}
+
+\hypertarget{EGbg_ext}{%
+\section{Adding a background extinction after reaching a number of living taxa}\label{EGbg_ext}}
+
+For this scenario, we want to generate a pure birth tree until reaching 50 living taxa but with an extinction rate appearing after reaching 30 taxa.
+
+First we need to set up the simulation parameters:
+* The stopping rule (50 taxa max)
+* The birth-death parameters (speciation rate of 1)
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# Simulation parameters}
+\NormalTok{stop\_taxa\_}\DecValTok{50}\NormalTok{ \textless{}{-}}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{max.living =} \DecValTok{50}\NormalTok{)}
+\NormalTok{speciation\_}\DecValTok{1}\NormalTok{ \textless{}{-}}\StringTok{ }\KeywordTok{make.bd.params}\NormalTok{(}\DataTypeTok{speciation =} \DecValTok{1}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+And our change in extinction rate event which triggers after reaching 30 taxa (\texttt{taxa.condition(30)}), targets the \texttt{"bd.params"} (birth-death parameters) sets the extinction rate to 0.5 (\texttt{update.bd.params(extinction\ =\ 0.5)}):
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# Adding an extinction parameter after 30 taxa}
+\NormalTok{background.extinction \textless{}{-}}\StringTok{ }\KeywordTok{make.events}\NormalTok{(}
+                      \DataTypeTok{condition    =} \KeywordTok{taxa.condition}\NormalTok{(}\DecValTok{30}\NormalTok{),}
+                      \DataTypeTok{target       =} \StringTok{"bd.params"}\NormalTok{,}
+                      \DataTypeTok{modification =} \KeywordTok{update.bd.params}\NormalTok{(}\DataTypeTok{extinction =} \FloatTok{0.5}\NormalTok{))}
+\end{Highlighting}
+\end{Shaded}
+
+Once these parameters are defined, we can run the simulations and plot the results:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# Running the simulations}
+\KeywordTok{set.seed}\NormalTok{(}\DecValTok{2}\NormalTok{)}
+\NormalTok{results \textless{}{-}}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{stop.rule =}\NormalTok{ stop\_taxa\_}\DecValTok{50}\NormalTok{,}
+                \DataTypeTok{bd.params =}\NormalTok{ speciation\_}\DecValTok{1}\NormalTok{,}
+                \DataTypeTok{events    =}\NormalTok{ background.extinction)}
+\CommentTok{\#\# Plotting the results}
+\KeywordTok{plot}\NormalTok{(results, }\DataTypeTok{show.tip.label =} \OtherTok{FALSE}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-108-1.pdf}
+
+\hypertarget{EGred_spec}{%
+\section{Reducing speciation rate after some time}\label{EGred_spec}}
+
+For this scenario, we want to generate a brith tree for 6 times units with random speciation rates (i.e.~drawn from a uniform (0.5;1) distribution) which reduces after time 4 through the simulations to a fixed value of 1/3.
+
+First we need to set up the simulation parameters:
+* The stopping rule (6 time units)
+* The birth-death parameters (speciation randomly drawn between 0.5 and 1)
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# Simulation parameters}
+\NormalTok{stop\_time\_}\DecValTok{6}\NormalTok{ \textless{}{-}}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{max.time =} \DecValTok{6}\NormalTok{)}
+\NormalTok{random\_speciation \textless{}{-}}\StringTok{ }\KeywordTok{make.bd.params}\NormalTok{(}\DataTypeTok{speciation =}\NormalTok{ runif, }\DataTypeTok{speciation.args =} \KeywordTok{list}\NormalTok{(}\DataTypeTok{min =} \FloatTok{0.5}\NormalTok{, }\DataTypeTok{max =} \DecValTok{1}\NormalTok{))}
+\end{Highlighting}
+\end{Shaded}
+
+And our extinction event which triggers after reaching time 4 (\texttt{time.condition(4)}), targets the \texttt{"bd.params"} and modifies the speciation rate to 1/3:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# Reducing speciation after reaching time 4}
+\NormalTok{reduced.speciation \textless{}{-}}\StringTok{ }\KeywordTok{make.events}\NormalTok{(}
+                      \DataTypeTok{condition    =} \KeywordTok{time.condition}\NormalTok{(}\DecValTok{4}\NormalTok{),}
+                      \DataTypeTok{target       =} \StringTok{"bd.params"}\NormalTok{,}
+                      \DataTypeTok{modification =} \KeywordTok{update.bd.params}\NormalTok{(}\DataTypeTok{speciation =} \DecValTok{1}\OperatorTok{/}\DecValTok{3}\NormalTok{))}
+\end{Highlighting}
+\end{Shaded}
+
+Once these parameters are defined, we can run the simulations and plot the results.
+We can contrast the results with the scenario without an event (but same random seed):
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# No event}
+\KeywordTok{set.seed}\NormalTok{(}\DecValTok{42}\NormalTok{)}
+\NormalTok{no\_event \textless{}{-}}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{stop.rule =}\NormalTok{ stop\_time\_}\DecValTok{6}\NormalTok{,}
+                 \DataTypeTok{bd.params =}\NormalTok{ random\_speciation)}
+
+\CommentTok{\#\# Reduced speciation event}
+\KeywordTok{set.seed}\NormalTok{(}\DecValTok{42}\NormalTok{)}
+\NormalTok{reduced\_speciation\_event \textless{}{-}}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{stop.rule =}\NormalTok{ stop\_time\_}\DecValTok{6}\NormalTok{,}
+                                 \DataTypeTok{bd.params =}\NormalTok{ random\_speciation,}
+                                 \DataTypeTok{events    =}\NormalTok{ reduced.speciation)}
+
+\CommentTok{\#\# Plot both trees}
+\KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow =} \KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{, }\DecValTok{2}\NormalTok{))}
+\KeywordTok{plot}\NormalTok{(no\_event, }\DataTypeTok{main =} \StringTok{"No event"}\NormalTok{, }\DataTypeTok{show.tip.label =} \OtherTok{FALSE}\NormalTok{)}
+\KeywordTok{axisPhylo}\NormalTok{()}
+\KeywordTok{plot}\NormalTok{(reduced\_speciation\_event, }\DataTypeTok{main =} \StringTok{"Reduced speciation after time 5"}\NormalTok{, }\DataTypeTok{show.tip.label =} \OtherTok{FALSE}\NormalTok{)}
+\KeywordTok{axisPhylo}\NormalTok{()}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-111-1.pdf}
+
+\hypertarget{EG_change_trait}{%
+\section{Changing the trait process after some time}\label{EG_change_trait}}
+
+For this scenario, we want to generate a pure birth tree with a one dimensional Brownian Motion trait for 5 time unit which then changes to an OU process.
+
+First we need to set up the simulation parameters:
+* The stopping rule (6 time units)
+* The birth-death parameters (speciation rate of 1)
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# Simulation parameters}
+\NormalTok{stop\_time\_}\DecValTok{6}\NormalTok{ \textless{}{-}}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{max.time =} \DecValTok{6}\NormalTok{)}
+\NormalTok{speciation\_}\DecValTok{1}\NormalTok{ \textless{}{-}}\StringTok{ }\KeywordTok{make.bd.params}\NormalTok{(}\DataTypeTok{speciation =} \DecValTok{1}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+Then set up our trait which is a one dimensional Brownian Motion
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# Trait}
+\NormalTok{simple\_bm\_trait \textless{}{-}}\StringTok{ }\KeywordTok{make.traits}\NormalTok{(}\DataTypeTok{n =} \DecValTok{1}\NormalTok{, }\DataTypeTok{process =}\NormalTok{ BM.process)}
+\end{Highlighting}
+\end{Shaded}
+
+And our event which triggers after reaching time 5 (\texttt{time.condition(5)}), targets the \texttt{"traits"} and modifies the process to \texttt{OU.process}.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# Create an event to change the trait process}
+\NormalTok{change.process.to.OU \textless{}{-}}\StringTok{ }\KeywordTok{make.events}\NormalTok{(}
+                  \DataTypeTok{condition    =} \KeywordTok{time.condition}\NormalTok{(}\DecValTok{5}\NormalTok{),}
+                  \DataTypeTok{target       =} \StringTok{"traits"}\NormalTok{,}
+                  \DataTypeTok{modification =} \KeywordTok{update.traits}\NormalTok{(}\DataTypeTok{process =}\NormalTok{ OU.process))}
+\end{Highlighting}
+\end{Shaded}
+
+Once these parameters are defined, we can run the simulations and plot the results.
+We can contrast the results with the scenario without an event (but same random seed):
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# Run the simulations without change}
 \KeywordTok{set.seed}\NormalTok{(}\DecValTok{1}\NormalTok{)}
-\NormalTok{trait_table <-}\StringTok{ }\OtherTok{NULL}
-\NormalTok{test <-}\StringTok{ }\KeywordTok{dads}\NormalTok{(bd.params, stop.rule, traits,}
-                                \DataTypeTok{modifiers =}\NormalTok{ modify_speciation_brlen,}
-                                \DataTypeTok{null.error =} \DecValTok{20}\NormalTok{)}
-\KeywordTok{plot}\NormalTok{(test}\OperatorTok{$}\NormalTok{tree)}
-\KeywordTok{plot.dads}\NormalTok{(test, }\DataTypeTok{main =} \StringTok{"Skewed branch length and speciation"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
+\NormalTok{no\_change \textless{}{-}}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{stop.rule =}\NormalTok{ stop\_time\_}\DecValTok{6}\NormalTok{,}
+                  \DataTypeTok{bd.params =}\NormalTok{ speciation\_}\DecValTok{1}\NormalTok{,}
+                  \DataTypeTok{traits    =}\NormalTok{ simple\_bm\_trait)}
+\CommentTok{\#\# Run the simulations with change}
+\KeywordTok{set.seed}\NormalTok{(}\DecValTok{1}\NormalTok{)}
+\NormalTok{process\_change \textless{}{-}}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{stop.rule =}\NormalTok{ stop\_time\_}\DecValTok{6}\NormalTok{,}
+                       \DataTypeTok{bd.params =}\NormalTok{ speciation\_}\DecValTok{1}\NormalTok{,}
+                       \DataTypeTok{traits    =}\NormalTok{ simple\_bm\_trait,}
+                       \DataTypeTok{events    =}\NormalTok{ change.process.to.OU)}
+\CommentTok{\#\# Plot the results}
+\KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow =} \KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{,}\DecValTok{2}\NormalTok{))}
+\KeywordTok{plot}\NormalTok{(no\_change, }\DataTypeTok{ylim =} \KeywordTok{c}\NormalTok{(}\OperatorTok{{-}}\DecValTok{7}\NormalTok{, }\DecValTok{7}\NormalTok{))}
+\KeywordTok{plot}\NormalTok{(process\_change, }\DataTypeTok{ylim =} \KeywordTok{c}\NormalTok{(}\OperatorTok{{-}}\DecValTok{7}\NormalTok{, }\DecValTok{7}\NormalTok{))}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-115-1.pdf}
+
+\hypertarget{EG_change_correlation}{%
+\section{Changing trait correlation after reaching a trait value}\label{EG_change_correlation}}
+
+For this scenario, we want to generate a pure birth tree with a 2 dimensional Brownian Motion trait with a strict correlation between the two dimensions (1:1) that loosen up when a taxa reaches an absolute value of 2.
+
+First we need to set up the simulation parameters:
+* The stopping rule (100 taxa)
+* The birth-death parameters (speciation rate of 1)
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# Set the parameters}
+\NormalTok{stop\_taxa\_}\DecValTok{100}\NormalTok{ \textless{}{-}}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{max.taxa =} \DecValTok{100}\NormalTok{)}
+\NormalTok{speciation\_}\DecValTok{1}\NormalTok{ \textless{}{-}}\StringTok{ }\KeywordTok{make.bd.params}\NormalTok{(}\DataTypeTok{speciation =} \DecValTok{1}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+Then set up our trait which is a 2 dimensional Brownian Motion with a correlation matrix Sigma ()
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# A 2D variance covariance matrix}
+\NormalTok{vcv\_matrix \textless{}{-}}\StringTok{ }\KeywordTok{matrix}\NormalTok{(}\DecValTok{1}\NormalTok{, }\DecValTok{2}\NormalTok{, }\DecValTok{2}\NormalTok{)}
+
+\CommentTok{\#\# A correlated 2D Brownian Motion}
+\NormalTok{correlated\_2D\_BM \textless{}{-}}\StringTok{ }\KeywordTok{make.traits}\NormalTok{(}\DataTypeTok{n =} \DecValTok{2}\NormalTok{, }\DataTypeTok{process =}\NormalTok{ BM.process,}
+                      \DataTypeTok{process.args =} \KeywordTok{list}\NormalTok{(}\DataTypeTok{Sigma =}\NormalTok{ vcv\_matrix))}
+\end{Highlighting}
+\end{Shaded}
+
+And our event which triggers after a taxa gets the trait value 3 (\texttt{trait.condition(3,\ absolute\ =\ TRUE)}), targets the \texttt{"traits"} and modifies the traits correlation
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# New correlation}
+\NormalTok{new\_vcv \textless{}{-}}\StringTok{ }\KeywordTok{matrix}\NormalTok{(}\KeywordTok{c}\NormalTok{(}\DecValTok{10}\NormalTok{,}\DecValTok{3}\NormalTok{,}\DecValTok{3}\NormalTok{,}\DecValTok{2}\NormalTok{),}\DecValTok{2}\NormalTok{,}\DecValTok{2}\NormalTok{)}
+
+\CommentTok{\#\# Event changing a trait correlation}
+\NormalTok{correlation.change \textless{}{-}}\StringTok{ }\KeywordTok{make.events}\NormalTok{(}
+    \DataTypeTok{condition    =} \KeywordTok{trait.condition}\NormalTok{(}\DecValTok{3}\NormalTok{, }\DataTypeTok{absolute =} \OtherTok{TRUE}\NormalTok{),}
+    \DataTypeTok{target       =} \StringTok{"traits"}\NormalTok{,}
+    \DataTypeTok{modification =} \KeywordTok{update.traits}\NormalTok{(}\DataTypeTok{process.args =} \KeywordTok{list}\NormalTok{(}\DataTypeTok{Sigma =}\NormalTok{ new\_vcv)))}
+\end{Highlighting}
+\end{Shaded}
+
+Once these parameters are defined, we can run the simulations and plot the results.
+We can contrast the results with the scenario without an event (but same random seed):
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# Run the simulations}
+\KeywordTok{set.seed}\NormalTok{(}\DecValTok{2}\NormalTok{)}
+\NormalTok{no\_event \textless{}{-}}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{stop.rule =}\NormalTok{ stop\_taxa\_}\DecValTok{100}\NormalTok{,}
+                 \DataTypeTok{bd.params =}\NormalTok{ speciation\_}\DecValTok{1}\NormalTok{,}
+                 \DataTypeTok{traits    =}\NormalTok{ correlated\_2D\_BM)}
+\KeywordTok{set.seed}\NormalTok{(}\DecValTok{2}\NormalTok{)}
+\NormalTok{change\_correlation \textless{}{-}}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{stop.rule =}\NormalTok{ stop\_taxa\_}\DecValTok{100}\NormalTok{,}
+                           \DataTypeTok{bd.params =}\NormalTok{ speciation\_}\DecValTok{1}\NormalTok{,}
+                           \DataTypeTok{traits    =}\NormalTok{ correlated\_2D\_BM,}
+                           \DataTypeTok{events    =}\NormalTok{ correlation.change)}
+
+\CommentTok{\#\# Visual testing}
+\KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow =} \KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{,}\DecValTok{2}\NormalTok{))}
+\KeywordTok{plot}\NormalTok{(no\_event, }\DataTypeTok{trait =} \KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{,}\DecValTok{2}\NormalTok{), }\DataTypeTok{main =} \StringTok{"Strict correlation"}\NormalTok{)}
+\KeywordTok{plot}\NormalTok{(change\_correlation, }\DataTypeTok{trait =} \KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{,}\DecValTok{2}\NormalTok{), }\DataTypeTok{main =} \StringTok{"Loosened correlation"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-119-1.pdf}
+
+And we can visualise this change through time:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# 3D plot}
+\KeywordTok{plot}\NormalTok{(change\_correlation, }\DataTypeTok{trait =} \KeywordTok{c}\NormalTok{(}\DecValTok{1}\OperatorTok{:}\DecValTok{2}\NormalTok{), }\DataTypeTok{use.3D =} \OtherTok{TRUE}\NormalTok{)}
+\KeywordTok{rglwidget}\NormalTok{()}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## Warning in snapshot3d(scene = x, width = width, height = height): webshot =
+## TRUE requires the webshot2 package and Chrome browser; using rgl.snapshot()
+## instead
+\end{verbatim}
+
+\begin{verbatim}
+## Warning in rgl.snapshot(filename, fmt, top): this build of rgl does not support
+## snapshots
+\end{verbatim}
+
+\includegraphics{../../../../../../tmp/Rtmp8BGjPN/file97bd12f0c3a.png}
+
+\hypertarget{EG_change_modif}{%
+\section{Event for changing a modifier: speciation event increase for species with negative values}\label{EG_change_modif}}
+
+For this scenario, we want to generate a pure birth tree with a one dimensional Brownian Motion trait for 4 time units.
+After three time units, we want the speciation rule to increase for species which ancestors have a negative trait value.
+
+First we need to set up the simulation parameters:
+* The stopping rule (5 time units)
+* The birth-death parameters (speciation rate of 1)
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# Set the parameters}
+\NormalTok{stop\_time\_}\DecValTok{4}\NormalTok{ \textless{}{-}}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{max.time =} \DecValTok{4}\NormalTok{)}
+\NormalTok{speciation\_}\DecValTok{1}\NormalTok{ \textless{}{-}}\StringTok{ }\KeywordTok{make.bd.params}\NormalTok{(}\DataTypeTok{speciation =} \DecValTok{1}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+Then set up our trait which is a one dimensional Brownian Motion
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# Trait}
+\NormalTok{simple\_bm\_trait \textless{}{-}}\StringTok{ }\KeywordTok{make.traits}\NormalTok{(}\DataTypeTok{n =} \DecValTok{1}\NormalTok{, }\DataTypeTok{process =}\NormalTok{ BM.process)}
+\end{Highlighting}
+\end{Shaded}
+
+And a modifier that is the default birth-death algorithm rules
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# Birth death rules (default)}
+\NormalTok{default\_modifiers \textless{}{-}}\StringTok{ }\KeywordTok{make.modifiers}\NormalTok{()}
+\end{Highlighting}
+\end{Shaded}
+
+And our extinction event which triggers after reaching time 3 (\texttt{time.condition(3)}), targets the \texttt{"modifiers"} and modifies birth-death rule as follows:
+* When a species is descendant from a parent with a negative trait value (\texttt{negative.trait.condition});
+* Then increase your chances of going extinct by +1 (\texttt{increase.extinction.1})
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# New condition and new modifier (increasing speciation if trait is negative)}
+\NormalTok{negative.trait.condition \textless{}{-}}\StringTok{ }\ControlFlowTok{function}\NormalTok{(trait.values, lineage) \{}
+    \KeywordTok{return}\NormalTok{(}\KeywordTok{parent.traits}\NormalTok{(trait.values, lineage) }\OperatorTok{\textless{}}\StringTok{ }\DecValTok{0}\NormalTok{)}
+\NormalTok{\}}
+\NormalTok{increase.extinction}\FloatTok{.1}\NormalTok{ \textless{}{-}}\StringTok{ }\ControlFlowTok{function}\NormalTok{(x, trait.values, lineage) \{}
+  \KeywordTok{return}\NormalTok{(x }\OperatorTok{+}\StringTok{ }\DecValTok{1}\NormalTok{)}
+\NormalTok{\}}
+
+\CommentTok{\#\# Update the modifier}
+\NormalTok{change.speciation \textless{}{-}}\StringTok{ }\KeywordTok{make.events}\NormalTok{(}
+    \DataTypeTok{condition    =} \KeywordTok{time.condition}\NormalTok{(}\DecValTok{3}\NormalTok{),}
+    \DataTypeTok{target       =} \StringTok{"modifiers"}\NormalTok{,}
+    \DataTypeTok{modification =} \KeywordTok{update.modifiers}\NormalTok{(}\DataTypeTok{speciation =}\NormalTok{ speciation,}
+                                    \DataTypeTok{condition  =}\NormalTok{ negative.trait.condition,}
+                                    \DataTypeTok{modify     =}\NormalTok{ increase.extinction}\FloatTok{.1}\NormalTok{))}
+\end{Highlighting}
+\end{Shaded}
+
+Once these parameters are defined, we can run the simulations and plot the results.
+We can contrast the results with the scenario without an event (but same random seed):
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{set.seed}\NormalTok{(}\DecValTok{4}\NormalTok{)}
+\NormalTok{no\_event \textless{}{-}}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{stop.rule =}\NormalTok{ stop\_time\_}\DecValTok{4}\NormalTok{,}
+                 \DataTypeTok{bd.params =}\NormalTok{ speciation\_}\DecValTok{1}\NormalTok{,}
+                 \DataTypeTok{traits    =}\NormalTok{ simple\_bm\_trait,}
+                 \DataTypeTok{modifiers =}\NormalTok{ default\_modifiers)}
+\KeywordTok{set.seed}\NormalTok{(}\DecValTok{4}\NormalTok{)}
+\NormalTok{change\_spec \textless{}{-}}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{stop.rule =}\NormalTok{ stop\_time\_}\DecValTok{4}\NormalTok{,}
+                    \DataTypeTok{bd.params =}\NormalTok{ speciation\_}\DecValTok{1}\NormalTok{,}
+                    \DataTypeTok{traits    =}\NormalTok{ simple\_bm\_trait,}
+                    \DataTypeTok{modifiers =}\NormalTok{ default\_modifiers,}
+                    \DataTypeTok{events    =}\NormalTok{ change.speciation)}
+
+\CommentTok{\#\# Visualise the results}
+\KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow =} \KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{,}\DecValTok{2}\NormalTok{))}
+\KeywordTok{plot}\NormalTok{(no\_event, }\DataTypeTok{main =} \StringTok{"No event"}\NormalTok{)}
+\KeywordTok{plot}\NormalTok{(change\_spec, }\DataTypeTok{main =} \StringTok{"Increase extinction for negative}\CharTok{\textbackslash{}n}\StringTok{traits after time 3"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-125-1.pdf}
+
+\hypertarget{EG_modify_brlen}{%
+\chapter{Changing branch length when reaching n taxa}\label{EG_modify_brlen}}
+
+For this scenario, we want to generate a pure birth tree until reaching 100 taxa.
+After reaching 30 taxa we want branch length growth to increase 50 folds.
+
+First we need to set up the simulation parameters:
+* The stopping rule (100 taxa)
+* The birth-death parameters (speciation rate of 1)
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# Set the parameters}
+\NormalTok{stop\_taxa\_}\DecValTok{100}\NormalTok{\textless{}{-}}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{max.taxa =} \DecValTok{100}\NormalTok{)}
+\NormalTok{speciation\_}\DecValTok{1}\NormalTok{ \textless{}{-}}\StringTok{ }\KeywordTok{make.bd.params}\NormalTok{(}\DataTypeTok{speciation =} \DecValTok{1}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+Then a modifier that is the default birth-death algorithm rules
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# Birth death rules (default)}
+\NormalTok{default\_modifiers \textless{}{-}}\StringTok{ }\KeywordTok{make.modifiers}\NormalTok{()}
+\end{Highlighting}
+\end{Shaded}
+
+And event which triggers after reaching 30 taxa (\texttt{taxa.condition(30)}), targets the \texttt{"modifiers"} and modifies the branch length generation rule to a 50 fold increase (\texttt{increase.50.folds})
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# Multiplying branch length 50 folds}
+\NormalTok{increase.}\FloatTok{50.}\NormalTok{folds \textless{}{-}}\StringTok{ }\ControlFlowTok{function}\NormalTok{(x, trait.values, lineage) \{}
+  \KeywordTok{return}\NormalTok{(x }\OperatorTok{*}\StringTok{ }\DecValTok{50}\NormalTok{)}
+\NormalTok{\}}
+\CommentTok{\#\# Event for increasing branch length after reaching 30 taxa}
+\NormalTok{increase\_brlen \textless{}{-}}\StringTok{ }\KeywordTok{make.events}\NormalTok{(}
+                  \DataTypeTok{condition    =} \KeywordTok{taxa.condition}\NormalTok{(}\DecValTok{30}\NormalTok{),}
+                  \DataTypeTok{target       =} \StringTok{"modifiers"}\NormalTok{,}
+                  \DataTypeTok{modification =} \KeywordTok{update.modifiers}\NormalTok{(}
+                                    \DataTypeTok{branch.length =}\NormalTok{ branch.length,}
+                                    \DataTypeTok{modify        =}\NormalTok{ increase.}\FloatTok{50.}\NormalTok{folds))}
+\end{Highlighting}
+\end{Shaded}
+
+Once these parameters are defined, we can run the simulations and plot the results.
+We can contrast the results with the scenario without an event (but same random seed):
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# Run the simulations}
+\KeywordTok{set.seed}\NormalTok{(}\DecValTok{5}\NormalTok{)}
+\NormalTok{no\_event \textless{}{-}}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{stop.rule =}\NormalTok{ stop\_taxa\_}\DecValTok{100}\NormalTok{,}
+                 \DataTypeTok{bd.params =}\NormalTok{ speciation\_}\DecValTok{1}\NormalTok{,}
+                 \DataTypeTok{modifiers =}\NormalTok{ default\_modifiers)}
+\KeywordTok{set.seed}\NormalTok{(}\DecValTok{5}\NormalTok{)}
+\NormalTok{increased\_brlen \textless{}{-}}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{stop.rule =}\NormalTok{ stop\_taxa\_}\DecValTok{100}\NormalTok{,}
+                        \DataTypeTok{bd.params =}\NormalTok{ speciation\_}\DecValTok{1}\NormalTok{,}
+                        \DataTypeTok{modifiers =}\NormalTok{ default\_modifiers,}
+                        \DataTypeTok{events    =}\NormalTok{ increase\_brlen)}
+
+\CommentTok{\#\# Visualise the results}
+\KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow =} \KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{,}\DecValTok{2}\NormalTok{))}
+\KeywordTok{plot}\NormalTok{(no\_event, }\DataTypeTok{main =} \StringTok{"No event"}\NormalTok{, }\DataTypeTok{show.tip.label =} \OtherTok{FALSE}\NormalTok{)}
+\KeywordTok{plot}\NormalTok{(increased\_brlen, }\DataTypeTok{main =} \StringTok{"Increase branch length}\CharTok{\textbackslash{}n}\StringTok{after 30 taxa"}\NormalTok{,}
+    \DataTypeTok{show.tip.label =} \OtherTok{FALSE}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-129-1.pdf}
+
+\hypertarget{EG_founding_purebirth}{%
+\section{Founding event: a generating a subtree with no fossils}\label{EG_founding_purebirth}}
+
+For this scenario, we want to generate a birth-death tree for 4 time units.
+After reaching 10 taxa, one random taxa will give birth to a sub-tree that is a pure birth tree (no extinction).
+
+First we need to set up the simulation parameters:
+* The stopping rule (5 time units)
+* The birth-death parameters (speciation rate of 1 and extinction of 0.2)
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# Set up parameters}
+\NormalTok{stop\_time\_}\DecValTok{4}\NormalTok{ \textless{}{-}}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{max.time =} \DecValTok{4}\NormalTok{)}
+\NormalTok{spec\_}\DecValTok{1}\NormalTok{\_ext\_}\DecValTok{02}\NormalTok{ \textless{}{-}}\StringTok{ }\KeywordTok{make.bd.params}\NormalTok{(}\DataTypeTok{speciation =} \DecValTok{1}\NormalTok{, }\DataTypeTok{extinction =} \FloatTok{0.2}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+And our event which triggers after reaching 10 taxa (\texttt{taxa.condition(10)}), and generates a subtree (``founding'') that is a pure birth tree (no extinction and speciation rate of 2).
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# Setting the pure{-}birth parameters}
+\NormalTok{speciation\_}\DecValTok{2}\NormalTok{ \textless{}{-}}\StringTok{ }\KeywordTok{make.bd.params}\NormalTok{(}\DataTypeTok{speciation =} \DecValTok{2}\NormalTok{)}
+
+\CommentTok{\#\# Events that generate a new process (founding effects)}
+\NormalTok{founding\_event \textless{}{-}}\StringTok{ }\KeywordTok{make.events}\NormalTok{(}
+                  \DataTypeTok{condition    =} \KeywordTok{taxa.condition}\NormalTok{(}\DecValTok{10}\NormalTok{),}
+                  \DataTypeTok{target       =} \StringTok{"founding"}\NormalTok{,}
+                  \DataTypeTok{modification =} \KeywordTok{founding.event}\NormalTok{(}
+                                    \DataTypeTok{bd.params =}\NormalTok{ speciation\_}\DecValTok{2}\NormalTok{),}
+                  \DataTypeTok{additional.args =} \KeywordTok{list}\NormalTok{(}\DataTypeTok{prefix =} \StringTok{"founding\_"}\NormalTok{))}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{quote}
+Note we are prodviding an additional argument \texttt{prefix} here so that we can track which species are part of the sub tree for colouring them down the line.
+\end{quote}
+
+Once these parameters are defined, we can run the simulations and plot the results:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# Simulations}
+\KeywordTok{set.seed}\NormalTok{(}\DecValTok{11}\NormalTok{)}
+\NormalTok{founding\_tree \textless{}{-}}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{stop.rule =}\NormalTok{ stop\_time\_}\DecValTok{4}\NormalTok{,}
+                      \DataTypeTok{bd.params =}\NormalTok{ spec\_}\DecValTok{1}\NormalTok{\_ext\_}\DecValTok{02}\NormalTok{,}
+                      \DataTypeTok{events    =}\NormalTok{ founding\_event)}
+
+\CommentTok{\#\# Selecting the edges colours}
+\NormalTok{tip\_values \textless{}{-}}\StringTok{ }\KeywordTok{rep}\NormalTok{(}\StringTok{"black"}\NormalTok{, }\KeywordTok{Ntip}\NormalTok{(founding\_tree))}
+\NormalTok{tip\_values[}\KeywordTok{grep}\NormalTok{(}\StringTok{"founding\_"}\NormalTok{, founding\_tree}\OperatorTok{$}\NormalTok{tip.label)] \textless{}{-}}\StringTok{ "orange"}
+\NormalTok{edge\_colors \textless{}{-}}\StringTok{ }\KeywordTok{match.tip.edge}\NormalTok{(tip\_values, founding\_tree, }\DataTypeTok{replace.na =} \StringTok{"black"}\NormalTok{)}
+
+\CommentTok{\#\# Plotting the results}
+\KeywordTok{plot}\NormalTok{(founding\_tree, }\DataTypeTok{show.tip.label =} \OtherTok{FALSE}\NormalTok{, }\DataTypeTok{edge.color =}\NormalTok{ edge\_colors)}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-132-1.pdf}
+
+\hypertarget{EG_founding_traits}{%
+\section{Founding event: a generating a subtree a different process}\label{EG_founding_traits}}
+
+For this scenario, we want to generate a birth-death tree with a one dimensional Brownian Motion trait for 6 time unit.
+After a taxa reaches the value 3 or higher, it gives birth to a sub-tree that generates an OU trait with an optimum at the value 3.
+
+First we need to set up the simulation parameters:
+* The stopping rule (6 time units)
+* The birth-death parameters (speciation rate of 1 and extinction of 1/3)
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# The tree parameters}
+\NormalTok{stop\_time\_}\DecValTok{6}\NormalTok{ \textless{}{-}}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{max.time =} \DecValTok{6}\NormalTok{)}
+\NormalTok{speciation\_}\DecValTok{1}\NormalTok{\_extinction\_}\DecValTok{03}\NormalTok{ \textless{}{-}}\StringTok{ }\KeywordTok{make.bd.params}\NormalTok{(}\DataTypeTok{speciation =} \DecValTok{1}\NormalTok{,}
+                                             \DataTypeTok{extinction =} \FloatTok{0.3}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+Then set up our trait which is a one dimensional Brownian Motion
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# Trait}
+\NormalTok{simple\_bm\_trait \textless{}{-}}\StringTok{ }\KeywordTok{make.traits}\NormalTok{(}\DataTypeTok{n =} \DecValTok{1}\NormalTok{, }\DataTypeTok{process =}\NormalTok{ BM.process)}
+\end{Highlighting}
+\end{Shaded}
+
+When a taxa reaches the value 3 \texttt{trait.condition}, it generates a pure birth tree (\texttt{speciation\ =\ 2}) with an OU trait with the optimum value 3.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# The OU trait with an optimum value of 3}
+\NormalTok{OU\_}\DecValTok{3}\NormalTok{ \textless{}{-}}\StringTok{ }\KeywordTok{make.traits}\NormalTok{(}\DataTypeTok{process =}\NormalTok{ OU.process,}
+                    \DataTypeTok{start =} \DecValTok{3}\NormalTok{, }\DataTypeTok{process.args =} \KeywordTok{list}\NormalTok{(}\DataTypeTok{optimum =} \DecValTok{3}\NormalTok{))}
+\CommentTok{\#\# The pure birth parameters}
+\NormalTok{speciation\_}\DecValTok{2}\NormalTok{ \textless{}{-}}\StringTok{ }\KeywordTok{make.bd.params}\NormalTok{(}\DataTypeTok{speciation =} \DecValTok{2}\NormalTok{)}
+
+\CommentTok{\#\# The founding event}
+\NormalTok{new\_OU\_trait \textless{}{-}}\StringTok{ }\KeywordTok{make.events}\NormalTok{(}
+                    \DataTypeTok{condition    =} \KeywordTok{trait.condition}\NormalTok{(}\DecValTok{3}\NormalTok{, }\DataTypeTok{condition =} \StringTok{\textasciigrave{}}\DataTypeTok{\textgreater{}=}\StringTok{\textasciigrave{}}\NormalTok{),}
+                    \DataTypeTok{target       =} \StringTok{"founding"}\NormalTok{,}
+                    \DataTypeTok{modification =} \KeywordTok{founding.event}\NormalTok{(}
+                                    \DataTypeTok{bd.params =}\NormalTok{ speciation\_}\DecValTok{2}\NormalTok{,}
+                                    \DataTypeTok{traits    =}\NormalTok{ OU\_}\DecValTok{3}\NormalTok{))}
+\end{Highlighting}
+\end{Shaded}
+
+Once these parameters are defined, we can run the simulations and plot the results:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# Simulating the tree}
+\KeywordTok{set.seed}\NormalTok{(}\DecValTok{1}\NormalTok{)}
+\NormalTok{founding\_tree \textless{}{-}}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{stop.rule =}\NormalTok{ stop\_time\_}\DecValTok{6}\NormalTok{,}
+                      \DataTypeTok{bd.params =}\NormalTok{ speciation\_}\DecValTok{1}\NormalTok{\_extinction\_}\DecValTok{03}\NormalTok{,}
+                      \DataTypeTok{traits    =}\NormalTok{ simple\_bm\_trait,}
+                      \DataTypeTok{events    =}\NormalTok{ new\_OU\_trait)}
+\KeywordTok{plot}\NormalTok{(founding\_tree)}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-136-1.pdf}
 

### 2023-03-27 inst/gitbook/compiling_gitbook.sh

@@ -1,5 +1,3 @@
-#!/bin/sh
-
 rm dads_manual.Rmd
 
 ## Remove files before hand
@@ -9,12 +7,6 @@ rm -R _bookdown_files/
 ## Compile the html
 R -e 'bookdown::render_book("index.Rmd", "bookdown::gitbook")' # HTML version
 
-## Comment out the badges
-sed 's/<!-- badges out start -->/<!-- badges out start/g' index.Rmd > index.tmp
-sed 's/<!-- badges out end   -->/badges out end   -->/g' index.tmp > index.Rmd
-rm index.tmp
-
-## Compile the pdf
 R -e 'bookdown::render_book("index.Rmd", "bookdown::pdf_book")' # PDF version
 
 ## Reactivate the badges

### 2023-03-27 inst/gitbook/_book/dads_manual.tex
Updated numbering in gitbook

@@
 \hypertarget{bdalgorithm}{%
-\chapter{The birth-death algorithm}\label{bdalgorithm}}
+\section{The birth-death algorithm}\label{bdalgorithm}}
 
@@
 ## }
-## <bytecode: 0x55a505929ed0>
+## <bytecode: 0x55e278dc9cd8>
 ## <environment: namespace:dads>
@@
 ## }
-## <bytecode: 0x55a507be7fe8>
+## <bytecode: 0x55e27b08d1b8>
 ## <environment: namespace:dads>
@@
 ## }
-## <bytecode: 0x55a507c69968>
+## <bytecode: 0x55e27b11eb90>
 ## <environment: namespace:dads>
@@
 ## }
-## <bytecode: 0x55a507c7e050>
+## <bytecode: 0x55e27b127560>
 ## <environment: namespace:dads>
@@
 
-\includegraphics{../../../../../../tmp/Rtmp8BGjPN/file97bd69ab2a6.png}
+\includegraphics{../../../../../../tmp/RtmpmEQua5/file9cb853e0a21f.png}
 
@@
 
-\includegraphics{../../../../../../tmp/Rtmp8BGjPN/file97bd7e5353fb.png}
+\includegraphics{../../../../../../tmp/RtmpmEQua5/file9cb812c6ed9e.png}
 
@@
 
-\includegraphics{../../../../../../tmp/Rtmp8BGjPN/file97bd12f0c3a.png}
+\includegraphics{../../../../../../tmp/RtmpmEQua5/file9cb864dc16ed.png}
 
@@
 \hypertarget{EG_modify_brlen}{%
-\chapter{Changing branch length when reaching n taxa}\label{EG_modify_brlen}}
+\section{Changing branch length when reaching n taxa}\label{EG_modify_brlen}}
 

### 2023-03-29 inst/gitbook/07_specific_examples.Rmd
manual big update
Updated numbering in gitbook
typos

@@
 ---
+
+# Specific examples
 
@@
 
-# Changing branch length when reaching n taxa {#EG_modify_brlen}
+## Changing branch length when reaching n taxa {#EG_modify_brlen}
 
@@
 
-## Founding event: a generating a subtree with no fossils {#EG_founding_purebirth}
+## Founding event: generating a subtree with no fossils {#EG_founding_purebirth}
 
@@
 
-## Founding event: a generating a subtree a different process {#EG_founding_traits}
+## Founding event: generating a subtree a different process {#EG_founding_traits}
 

### 2023-03-29 inst/gitbook/index.Rmd
Updated numbering in gitbook
typos

@@
 The first section `Getting started` is aimed for people with a `R` beginner level:
+
  * you know what a package is - and how to install it;
@@
 The sections after that are a bit more advanced and requires the following level:
+
  * you have already created functions before;
@@
 
-# The birth-death algorithm {#bdalgorithm}
+## The birth-death algorithm {#bdalgorithm}
 

### 2023-04-04 inst/gitbook/01_getting_started.Rmd
Minor edits to text

@@
 
-## The simplest of all analysis: simulating diversity only
+## The simplest analysis: simulating diversity only
 
-One of the simplest things to do with the `dads` package is just to simulate a birth death tree.
+One of the simplest things to do with the `dads` package is to simulate a birth-death tree.
 For that you can use the function `dads` and specify your stopping rule.
-The stopping rule simply tells the birth death process to step whenever it reaches one of these three conditions:
+The stopping rule simply tells the birth-death process to stop whenever it reaches one of these three conditions:
 
  * `"max.taxa"   = n` stop when `n` taxa are generated;
- * `"max.living" = n` stop when there is `n` co-occuring taxa of the same age (i.e. "living" taxa);
+ * `"max.living" = n` stop when there are `n` co-occuring taxa of the same age (i.e. "living" taxa);
  * `"max.time"   = n` stop when the simulated tree is `n` units of age old (these units are arbitrary);
@@
 ```{r}
-## Running the birth death simulation
+## Running the birth-death simulation
 my_tree <- dads(stop.rule = my_stop_rule)
@@
 
-> Note that here we could have specified more than one stopping rule, for example, we might want to run a simulation and stop it if it either reaches 10 taxa or the age 2 using `stop.rule = list(max.time = 2, max.taxa = 10)`.
+> Note that here we could have specified more than one stopping rule, for example, we might want to run a simulation and stop it if it *either* reaches 10 taxa *or* the age 2 using `stop.rule = list(max.time = 2, max.taxa = 10)`.
 The simulation will then stop when either of these conditions are met.
 
-The resulting object is a classic `"phylo"` object that you can simply plot or visualise like so:
+The resulting object is a classic `"phylo"` object that you can plot like so:
 
@@
 People familiar with the [birth-death models](https://lukejharmon.github.io/pcm/chapter10_birthdeath/) might have noticed that we did not specify two important things here: the speciation parameter (sometimes called "lambda" or "birth") and the extinction parameter (sometimes called "mu", "death" or "background extinction").
-By default `dads` runs a pure birth model (the speciation is set to 1 and the extinction to 0).
-However, you can easily change that by specifying your new birth death parameters:
+By default `dads` runs a pure birth model (speciation is set to 1 and extinction to 0).
+However, you can easily change that by specifying your own birth-death parameters:
 
 ```{r}
-## my birth death parameters
+## my birth-death parameters
 my_params <- list(speciation = 1,
@@
 
-> Note that here it is not necessary to specify `speciation = 1` since this is the default option, you can always just change the parameter of interest (e.g.
-changing `extinction = 0` to `extinction = 1/3`).
-However, we think it's good practice to attribute both parameters specifically to avoid any confusion.
-You can find more information about setting up more complex birth death parameters [in this section](#makebdparams).
+> You can find more information about setting up more complex birth-death parameters [in this section](#makebdparams).
 
-You can then run the same birth death tree with extinction:
+You can then run the same birth-death tree simulation with extinction:
 
 ```{r}
-## Generating a birth death tree with extinctions:
+## Generating a birth-death tree with extinctions:
 my_tree <- dads(bd.params = my_params, stop.rule = my_stop_rule)
@@
 
-Chances are that you want to also simulate traits (disparity) along with your diversity (otherwise, we suggest using the [`TreeSim`](https://github.com/tanja819/TreeSim/) package that provides many more birth death models).
+Chances are that you are using `dads` because you also want to simulate traits (disparity) along with your diversity (otherwise, we suggest using the [`TreeSim`](https://github.com/tanja819/TreeSim/) package that provides many more birth-death models).
 Simulating traits is not much more complicated in `dads`: you'll simply need to create a `"traits"` object using the `make.traits` function.
@@
 `"traits"` objects contain one or more processes which are the ways to generate the trait.
-The most common of these processes is the [Brownian Motion](https://en.wikipedia.org/wiki/Brownian_motion).
+The most common of these processes is the [Brownian Motion](https://en.wikipedia.org/wiki/Brownian_motion) model.
 This is used by default with the `make.traits` function:
@@
 ```{r}
-## Extracting the tree (a "phylo" object)
+## Extracting the tree 
 the_generated_tree <- my_data$tree
+# Note that this is a "phylo" object
 class(the_generated_tree)
 
-## Extracting the data (a "matrix")
+## Extracting the data 
 the_generated_data <- my_data$data
+# Note that this is a "matrix" or "array"
 class(the_generated_data)
@@
 
-You can find much more on how to design trait objects in the [`make.traits` section](#maketraits).
+You can find much more about how to design trait objects in the [`make.traits` section](#maketraits).
 
-## Slightly more more complex: simulating linked disparity and diversity
+## Slightly more complex again: simulating linked disparity and diversity
 
-The example above is also still pretty simple and easily done through a variety of `R` packages: here the trait and the tree are simulated at the same time but only the tree is simulating the trait (i.e.
-the trait value at a tip is affected by it's ancestor and the branch length leading to it) but not the other way around (the trait value does not affect the tree).
+The example above is still pretty simple and easily done through a variety of `R` packages: here the trait and the tree are simulated at the same time but only the tree is simulating the trait (i.e.
+the trait value at a tip is affected by its ancestor and the branch length leading to it) but not the other way around (the trait value does not affect the tree).
 It is possible to add this aspect using `"modifiers"` objects.
@@
 `"modifiers"` affect two key steps of the birth-death process: the calculation of the waiting time (i.e. the component generating branch lengths) and the triggering of speciation or extinction events.
-These events can be modified using some `condition` and `modify` function.
+These events can be modified using `condition` and `modify` functions.
 In other words, when reaching a certain condition specified by a `condition` function, the birth-death process will modify either the branch length or the speciation (or extinction) probability by applying a `modify` function.
@@
 
-Note that these are default for a birth death tree and were actually already applied in the examples before (without specifying a modifier):
+Note that these are defaults for a birth-death tree and were already applied in the examples above without specifying a `"modifiers"` object:
 
 ```{r}
-## Make a default modifiers
+## Make a default modifiers object
 default_modifiers <- make.modifiers()
@@
 
-This will not do much to our simulations compared to the previous trait and tree simulation but we can provide our modifiers object to the `dads` function:
+This will not do anything to our simulations compared to the previous trait and tree simulation but we can provide our `"modifiers"` object to the `dads` function:
 
@@
 
+# Set random seed so we get the same results
 set.seed(1)
+
 ## Simulate disparity and diversity
@@
 
-Note however that the printing information is now updated to state that you've added a modifier (even though it's a default one).
+Note however, that the printing information is now updated to state that you've added a `"modifiers"` object (even though it's the same as the default).
 
-For more interesting simulations however, you can provide modifiers that actually modify the birth death process.
-We can create one for example that makes species go extinct if their ancestor have a negative trait value.
-For that we need to create a modifiers object that modifies the `speciation` process with a specific condition and a specific modification when that condition is met.
+For more interesting simulations however, you can provide modifiers that actually modify the birth-death process.
+We can create one for example that makes species go extinct if their ancestor has a negative trait value.
+For that we need to create a `"modifiers"` object that modifies the `speciation` process with a specific condition and a specific modification when that condition is met.
 For a speciation to occur (and a species to not go extinct), the algorithm draws a random value _x_ between 0 and 1 and if the value is smaller than the speciation parameter divided by the speciation + extinction parameter, the lineage speciates ($x < (\lambda \ (\lambda + \mu))$), else it goes extinct.
 
-Here we can create a modifier forcing the lineage to go extinct every time a specific condition is met (say the lineage trait value is negative).
-In other words, if the lineages trait value is negative, make sure that _x_ is equal to 1 (so that the speciation condition $x < (\lambda \ (\lambda + \mu))$ is never met).
+Here we can create a modifier forcing the lineage to go extinct every time a specific condition is met, e.g. if the lineage trait value is negative.
+In other words, if the lineage's trait value is negative, make sure that _x_ is equal to 1, so that the speciation condition $x < (\lambda \ (\lambda + \mu))$ is never met.
 
-First we need to create a condition function to trigger this (non) speciation event.
-We can do that by specifying our `condition` function (when the ancestor is trait is negative) apply our modification (here the `staying.alive` function).
-For that we can use the `parent.traits` utility function that is optimised for accessing traits in the birth death process (but you can of course write your own).
-This function intakes the `trait.values` and `parent.lineage` arguments, two arguments that you can leave named as they are to facilitate `dads`' understanding of what you want to assess:
+First we need to create a `condition` function to trigger this (non) speciation event.
+We can do that by specifying our `condition` function (here the `staying.alive` function) to apply our modification.
+For that we can use the `parent.traits` utility function that is optimised for accessing traits in the birth-death process (but you can of course write your own).
+This function takes the `trait.values` and `parent.lineage` arguments, two arguments that you can leave named as they are to facilitate `dads`' understanding of what you want to assess:
 
@@
 
-The we are going to create the modification function.
-This function must intake the argument `x` and, in our case, returns the same value no matter what: `1` so that speciation never happens when the `negative.ancestor` function is triggered.
+Then we create the modification function.
+This function must take the argument `x` and, in our case, returns the same value no matter what: `1` so that speciation never happens when the `negative.ancestor` function is triggered.
 
@@
 We can then provide these two functions (the condition `negative.ancestor` and how to modify the speciation event when this condition is met `staying.alive`).
-If you are an advances `dads` user, you can design your own `speciation` function but if you just want to use a normal `speciation` function, you can use the default one from `dads` called...
-`speciation`.
+If you are an advances `dads` user, you can design your own `speciation` function but if you just want to use a normal `speciation` function, you can use the default one from `dads` called `speciation`.
 
@@
             ## If the following condition is met...
-            condition = negative.ancestor, # Does the lineage has a negative ancestor?
+            condition = negative.ancestor, # Does the lineage have an ancestor with negative trait values?
             ## Then apply the following modifier...
@@
 Note that the `make.modifiers` function tests whether the input is compatible with `dads` by default so unless you have an error message, your `modifiers` will work!
-We can now simulate our tree and traits with our modifier: species will go extinct if their ancestor have a negative trait value:
+We can now simulate our tree and traits with our modifier: species will go extinct if their ancestor has a negative trait value:
 
@@
 
-You can find much more how to design modifiers in the [`make.modifiers` section](#makemodifiers).
+You can find much more about how to design modifiers in the [`make.modifiers` section](#makemodifiers).

### 2023-04-04 inst/gitbook/07_specific_examples.Rmd
Minor text edits

@@
 
-For this scenario, we want to generate a pure birth tree (no traits and no extinction) where 80% of the species go extinct two thirds into the scenario.
+For this scenario, we want to generate a pure birth tree (no traits and no extinction) where 80% of the species go extinct two thirds of the way into the scenario.
 
@@
 
-We will then create a event that triggers when reaching half of the simulation (using `time.condition(2.5)`).
+We will then create a event that triggers when reaching half of the simulation time (using `time.condition(2.5)`).
 This event will target `"taxa"`, i.e. the number of species, and randomly remove 80% of them (using `random.extinction(0.8)`):
@@
 
-## Species with negative trait values go extinct after some time {#EGneg_ext}
+## Species with negative trait values go extinct after a certain time {#EGneg_ext}
 
-For this scenario, we want to generate a pure birth tree with a one dimensional Brownian Motion trait for 5 time unit.
-We then want the species with a negative trait value go extinct after 4 time unit.
+For this scenario, we want to generate a pure birth tree with a one dimensional Brownian Motion trait for 5 time units.
+We then want species with a negative trait value go extinct after 4 time units.
 
@@
 
-And our extinction event which triggers after reaching time 4 (`time.condition(4)`), targets the `"taxa"` and modifies the extinction for species with traits lower than 0
+And our extinction event which triggers after reaching time 4 (`time.condition(4)`), targets the `"taxa"` and modifies the extinction for species with traits lower than 0.
 
@@
 
-## Reducing speciation rate after some time {#EGred_spec}
+## Reducing speciation rate after a certain time {#EGred_spec}
 
-For this scenario, we want to generate a brith tree for 6 times units with random speciation rates (i.e.
+For this scenario, we want to generate a birth tree for 6 times units with random speciation rates (i.e.
 drawn from a uniform (0.5;1) distribution) which reduces after time 4 through the simulations to a fixed value of 1/3.
@@
 stop_time_6 <- list(max.time = 6)
-random_speciation <- make.bd.params(speciation = runif, speciation.args = list(min = 0.5, max = 1))
+random_speciation <- make.bd.params(speciation = runif, 
+                                    speciation.args = list(min = 0.5, max = 1))
 ```
@@
 axisPhylo()
-plot(reduced_speciation_event, main = "Reduced speciation after time 5", show.tip.label = FALSE)
+plot(reduced_speciation_event, 
+     main = "Reduced speciation after time 5", 
+     show.tip.label = FALSE)
 axisPhylo()
@@
 
-For this scenario, we want to generate a pure birth tree with a one dimensional Brownian Motion trait for 5 time unit which then changes to an OU process.
+For this scenario, we want to generate a pure birth tree with a one dimensional Brownian Motion trait for 5 time units which then changes to an OU process.
 
@@
 For this scenario, we want to generate a pure birth tree with a one dimensional Brownian Motion trait for 4 time units.
-After three time units, we want the speciation rule to increase for species which ancestors have a negative trait value.
+After 3 time units, we want the speciation rule to increase for species which ancestors have a negative trait value.
 
@@
 ```{r}
-## Birth death rules (default)
+## birth-death rules (default)
 default_modifiers <- make.modifiers()
@@
 
-For this scenario, we want to generate a birth-death tree with a one dimensional Brownian Motion trait for 6 time unit.
-After a taxa reaches the value 3 or higher, it gives birth to a sub-tree that generates an OU trait with an optimum at the value 3.
+For this scenario, we want to generate a birth-death tree with a one dimensional Brownian Motion trait for 6 time units.
+After a taxon reaches the value 3 or higher, it gives birth to a sub-tree that generates an OU trait with a long-term mean at the value 3.
 
@@
 
-When a taxa reaches the value 3 `trait.condition`, it generates a pure birth tree (`speciation = 2`) with an OU trait with the optimum value 3.
+When a taxa reaches the value 3 `trait.condition`, it generates a pure birth tree (`speciation = 2`) with an OU trait with the long-term mean value 3.
 
 ```{r}
-## The OU trait with an optimum value of 3
+## The OU trait with a long-term mean value of 3
 OU_3 <- make.traits(process = OU.process,

### 2023-04-04 inst/gitbook/05_other_functionalities.Rmd
Minor edits to text

@@
 ---
-title: "`make.events`"
+title: "`other functionalities`"
 author: "Thomas Guillerme (guillert@tcd.ie)"
@@
 
-In most examples above, birth-death parameters where set as fixed values, i.e.
+In most examples above, birth-death parameters were set as fixed values, i.e.
 a speciation rate of $\lambda$ and an extinction rate of $\mu$ throughout the simulations (with maybe some events modifying these rates).
-However, it is of course possible to set these rates as specific or changing distributions! 
-You can do this using the `make.bd.params` function a provide either a `vector` of values:
+However, it is possible to set these rates as specific or changing distributions. 
+You can do this using the `make.bd.params` function to provide either a `vector` of values:
 
@@
 
-Or directly a function from which to sample:
+Or to directly provide a function from which to sample:
 
@@
 
-When using distributions for both the speciation and extinction parameters, you can run into the undesired problem of having an extinction rate that is higher than your speciation rate (and thus your tree dying out - unless this is the desired behaviour).
-You can avoid this problem by using the `joint` distribution argument this will ensure that the sampling of the extinction rate is always lower or equal to the speciation rate.
+When using distributions for both the speciation and extinction parameters, you can run into the (usually) undesired problem of having an extinction rate that is higher than your speciation rate and thus all the taxa in your tree dying out.
+You can avoid this problem by using the `joint` distribution argument. 
+This will ensure that the sampling of the extinction rate is always lower or equal to the speciation rate.
 
@@
 
-The two other possible arguments for this function are `test` and `update` that work the same as for [`make.traits`](#maketraits) or [`make.modifiers`](#makemodifiers)
+The two other possible arguments for this function are `test` and `update` that work the same way as [`make.traits`](#maketraits) or [`make.modifiers`](#makemodifiers)
 
-If you are a visual person and your `bd.params` objects are getting a bit too complicated to remember, you can always quickly plot them (the function will sample from the `bd.params` object and show the results):
+If you are a visual person and your `bd.params` objects are getting a bit too complicated to remember, you can always quickly plot them. 
+The function will sample from the `bd.params` object and show the results:
 
 ```{r}
-par(mfrow = c(2,2))
+par(mfrow = c(2,3))
 plot(make.bd.params(), main = "Default birth-death parameters")
@@
      main = "Disjoint normal speciation and uniform extinction")
+par(mfrow = c(1,1))
 ```
@@
 
-You can use the function `drop.things` to drop specific elements of the tree and data at the same time by providing the argument `what` to be `"fossils"` for tips that went extinct, `"livings"` tips that where alive at the end of the simulation or `"singles"` for internal nodes.
+You can use the function `drop.things` to drop specific elements of the tree and data at the same time by providing the argument `what = "fossils"` for tips that went extinct, or `what = "livings"` for tips that where alive at the end of the simulation or `what = "singles"` to drop internal singleton nodes.
 Alternatively you can use the function aliases `drop.fossils`, `drop.livings` or `drop.singles` for the exact same results:

### 2023-04-04 inst/gitbook/index.Rmd
Minor edits to text

@@
 
-Allowing to simulate disparity and diversity at the same time with interaction between both.
+Simultaneously simulate diversity (i.e. phylogenetic trees) and disparity  diversity (and the interaction between them).
 
@@
 
-There are some very good packages out there to simulate birth death trees (like [TreeSim](https://github.com/tanja819/TreeSim)) or even packages simulating traits (disparity) and diversity jointly (like [RPANDA](https://github.com/cran/RPANDA) or [PETER](https://github.com/PuttickMacroevolution/PETER)).
-We strongly advice you have a look at these packages first as they might be more appropriate for your tasks.
+%NC: I think here you need an example of use case as it's not immediately obvious what the package is for...
 
-`dads` aims to be a highly modular and friendly version of all these packages: it allows to simulate disparity and diversity jointly with a vast array of options that can be easily modified by users.
-For example, you can easily generate any type of process to generate a trait (BM, OU, something else, etc...) in multiple dependent or independent dimensions through `"traits"` objects.
-You can then specify how the traits should affect disparity through `"modifiers"` objects. And finally you can create events (like mass extinctions) through `"events"` objects.
+There are some very good R packages that simulate birth-death trees (e.g.
+[TreeSim](https://github.com/tanja819/TreeSim)) or simulate traits (disparity) and diversity jointly (e.g.
+[RPANDA](https://github.com/cran/RPANDA) or [PETER](https://github.com/PuttickMacroevolution/PETER)).
+We strongly advise you look at these packages first as they might be more appropriate for your needs.
+
+`dads` is a highly modular and user-friendly package that contains the simulations functionality of all these packages: it allows you to simulate disparity and diversity jointly with a vast array of options that can be easily modified by users.
+For example, you can easily generate any type of process to generate a trait (BM, OU, something else, etc.) in multiple dependent or independent dimensions through `"traits"` objects.
+You can then specify how the traits should affect disparity through `"modifiers"` objects. 
+And finally you can create events (like mass extinctions) through `"events"` objects.
 These objects and how to modify them will be detailed throughout this manual.
-Finally we are putting an emphasise in the development of this package on the speed and reliability of the functions.
+Finally we are putting an emphasis in the development of this package on the speed and reliability of the functions.
 
@@
 
-Because their is an infinite way you might want to generate disparity and diversity (different traits, different modifiers and different events), the `dads` package is designed to make all these parts easy to code separately and integrate them easily in the `dads` core functions.
-This allows you to simulate finely tuned multidimensional data for your specific project!
+Because there is an infinite number of ways you might want to generate disparity and diversity (different traits, different modifiers and different events), the `dads` package is designed to make all these parts easy to code separately and integrate in the `dads` core functions.
+This allows you to simulate finely-tuned multidimensional data for your specific project!
 
@@
 
-This manual explains how `dads` work in as much details as possible and aims to give you the keys to design your own diversity and disparity simulations.
+This manual explains how `dads` works in as much details as possible and aims to give you the keys to design your own diversity and disparity simulations.
 
@@
  * you know what a function is - and what arguments are;
- * you have already (vaguely) heard that they are different types of objects in R - like matrices (`"matrix"`) or trees (`"phylo"`);
- * and you have already looked for documentation and bugs online.
+ * you have already (vaguely) heard that there are different types of objects in R - like matrices (`"matrix"`) or trees (`"phylo"`);
+ * and you have already looked for documentation online.
 
-The sections after that are a bit more advanced and requires the following level:
+The sections after that are a bit more advanced and require the following level:
 
@@
  * you have heard that the objects in R have classes and that it sometimes matters - for example you know that R can differentiate between `as.numeric(1)` and `as.integer(1)`;
- * you owe part of your life to Stack Overflow.
+ * you owe part of your work to Stack Overflow.
 
@@
 
-You can install this package easily, directly from the github:
+You can install this package easily, directly from the GitHub:
 <!-- TG: I do intend to release it on CRAN later on -->
@@
 If you need help with the package, hopefully the following manual will be useful.
-However, parts of this package are still in development and some other parts are probably not covered.
-Thus if you have suggestions or comments on on what has already been developed or will be developed, please send me an email (<guillert@tcd.ie>) or if you are a GitHub user, directly create an issue on the [GitHub page](https://github.com/TGuillerme/dads).
-Doing so will not only hopefully help you but also other users since it will help improve this manual!
+However, if you have suggestions or comments on on what has already been developed or will be developed, please send me an email (<guillert@tcd.ie>) or if you are a GitHub user, directly create an issue on the [GitHub page](https://github.com/TGuillerme/dads).
+Doing so will not only help you but also other users as it will help improve this manual and the package!
 
@@
 
-Basically, the `dads` function intakes your personalised `traits`, `modifiers` and `events` to generate your disparity and diversity.
-You will find more details about how these objects (`traits`, `modifiers` and `events`) work in the rest of the tutorial but here is a graphical representation of how `dads` work:
+Basically, the `dads` function takes your `traits`, `modifiers` and `events` objects to simulate disparity and diversity.
+You will find more details about how these objects (`traits`, `modifiers` and `events`) work in the rest of the tutorial but here is a graphical representation of how `dads` works:
 
-![Schematised summary of the `dads` package architecture](dads_structure.png)
+![Schematic summary of the `dads` package architecture](dads_structure.png)
+% NC: The wonkiness of the figure is annoying me lol. Can you center the arrows etc?
 
@@
 
-> If you want to get your hands dirty, you can go straight to the [getting start section](#gettingstarted), this following section just describes the process in pseudo-code.
+> If you want to get your hands dirty, you can go straight to the [getting started section](#gettingstarted), this following section just describes the process in pseudo-code.
 
@@
 The birth-death model is a continuous Markov process (i.e.
-a continuous random process where the state will change each step are linked through time and change according to some random exponential variables) which is well defined mathematically and [commonly used in biology](https://lukejharmon.github.io/pcm/chapter10_birthdeath/#section-10.2-the-birth-death-model) but also in [many other fields](https://en.wikipedia.org/wiki/Birth%E2%80%93death_process).
+a continuous random process where future events depend only on the present, and not on past events) which is well defined mathematically and [commonly used in evolutionary biology](https://lukejharmon.github.io/pcm/chapter10_birthdeath/#section-10.2-the-birth-death-model) but also in [many other fields](https://en.wikipedia.org/wiki/Birth%E2%80%93death_process).
 
@@
  0.
-**Starting the process**: this step is actually non modular and creates a random tree with one tip, one node and one branch length connecting both.
+**Starting the process**: this step is non-modular and creates a random tree with one tip, one node and one branch connecting both.
 This step is used to optimise the rest of the algorithm in terms of speed and memory management.
@@
 **Selecting a lineage** (`selecting`): this step selects a tip that is currently not extinct.
-In a standard birth-death process this is just done randomly, however in `dads` this can be modified based on the birth-death parameters, the currently available lineages and potential trait values.
-*For example, it is possible to put a higher probability in selecting a lineage that is next to a lineage that recently went extinct and has a positive trait value.*
+In a standard birth-death process this is done randomly, however in `dads` this can be modified based on the birth-death parameters, the currently available lineages and potential trait values.
+*For example, it is possible to put a higher probability for selecting a lineage that is closely related to a lineage that recently went extinct and has a positive trait value.*
  Then go to step 2.
@@
 It does so by adding the same amount of branch length to all the non-extinct lineages.
-In an exact birth death process, this is done by drawing a random value from an exponential distribution with a rate of $$\text{number of living lineages} \times (\text{speciation} + \text{extinction parameters})$$.
+In an exact birth-death process, this is done by drawing a random value from an exponential distribution with a rate of $$\text{number of living lineages} \times (\text{speciation} + \text{extinction parameters})$$.
 In `dads`, this can be modified based on the birth-death parameters, the currently available lineages and potential trait values.
@@
  3.
-**Simulating traits** [optional]\(`traits`): this step allows to simulate a trait value for the selected lineage from step 1.
+**Simulating traits** [optional]\(`traits`): this step allows you to simulate a trait value for the selected lineage from step 1.
 This is typically not part of a standard birth-death process and is handled via the `traits` option in `dads` (see the [`make.traits`](#maketraits) chapter for more details).
@@
 **Speciating** (`speciating`): in this step, the selected lineage from step 1 has the option of speciating or going extinct.
-In a standard birth-death process, this happens by randomly drawing a value between 0 and 1 as a way to trigger speciation relatively to the birth-death parameters:
+In a standard birth-death process, this happens by randomly drawing a value between 0 and 1 as a way to trigger speciation relative to the birth-death parameters:
 ```
@@
  Again, in `dads`, this process can be modified based on the birth-death parameters, the currently available lineages and potential trait values.
-*For example, the lineage can only speciate if it's trait value is positive, regardless of whether speciation or extinction has been triggered*.
+*For example, the lineage can only speciate if its trait value is positive, regardless of whether speciation or extinction have been triggered*.
  Then go to step 5.
  5. If the number of lineages is less than required number of lineages, then go to step 1. Else go to step 6.
- 6. The simulation has stopped because it reached the required amount of lineages or time.
+ 6. The simulation stops because it has reached the required amount of lineages and/or time (i.e. branch length).

### 2023-04-05 inst/gitbook/02_make.traits.Rmd
manual big update

--- 2023-02-27 inst/gitbook/02_make.traits.Rmd
@@
 In `dads`, traits are simulating by providing a `traits` object to the `traits` argument.
-This object is generated using the `make.traits`.
+This object is generated using `make.traits`.
+
+> **What is a trait in biology?** This is one of these epistemological questions like what _is_ a species and is not covered in this package.
+For the purpose of _this manual_, a **biological trait** can be any coherent character of any number of dimensions which may or may not be independent of other traits.
+Basically anything can be a trait if it's defined by the user (e.g. one trait can be a 20 dimensional Brownian process that's correlated to a 1D OU process - which is another trait).
+For the purpose of the syntax in _this package_ a **trait in `dads`** is a column in the output trait matrix.
+For example, one might be interested in a biological trait that is "beak shape" which is defined as one complex biological trait that is the Procrustes superimposition of 4 3D landmarks and 300 3D semi-landmarks.
+This single "beak shape" **biological trait** is then composed of 912 **traits in `dads`** (i.e. a matrix of $(4+300)\times3$ columns).
 
@@
 The function `make.traits` allows you to design the process of a trait or a set of traits.
-Here, the process of a trait designates the rules to generate the trait through time while simulating a phylogeny.
+Here, the process of a trait designates the rules needed to generate the trait through time while simulating a phylogeny.
 This process can depend on the previous state in the tree (i.e. the trait of the ancestor) and the branch length to the descendant.
 One classic example is the [Brownian motion process (or Weiner process)](https://en.wikipedia.org/wiki/Brownian_motion).
-Note that it _can_ depend on both the ancestor and the branch length but does _not necessary needs_ (i.e. the process can be only based on the previous state or only on branch length or on neither).
+Note that it _can_ depend on both the ancestor and the branch length but does _not necessarily need to_, i.e.
+the process can be based only on the previous state or only on branch length or on neither.
 
@@
 
-Trait processes in `dads` are functions that must always intake the following arguments by default.
+Trait processes in `dads` are functions that must always take the following arguments by default.
 
  * `x0`: the previous trait value(s)
- * `edge.length`: the branch length value
+ * `edge.length`: the branch length
  * `...`: a placeholder for any extra arguments
@@
 For example, the following function would be a valid process that always generate the _true_ trait value: 42!.
-In this example, the process is not dependent on neither the previous state (`x0`) and the branch length (`edge.length`).
+In this example, the process is not dependent on either the previous state (`x0`) or the branch length (`edge.length`).
 
@@
 
-> Note that in this function definition the arguments `x0` and `edge.length` have a default value set to `0` and `1` respectively.
+> Note that in this function definition the arguments `x0` and `edge.length` have a default value of `0` and `1` respectively.
 In practice, these arguments are effectively set to the correct values in the `dads` internal function (i.e.
@@
 ```{r}
-## A wrongly formated process
+## A wrongly formatted process
 invalid.process <- function(x0 = 0) {
@@
 
+This will not work in `make.traits` (see below).
+
 ### Using a `"process"` in `dads`
 
-You can design your own process as a function (as long as it has a valid syntax).
-Alternatively, the `dads` package proposes inbuilt processes, namely a multidimensional Brownian motion (`BM.process`) or a a multidimensional Ornstein-Uhlenbeck process (`OU.process`).
-You can find the list of implemented process by looking at the `?trait.process` manual page in `R`.
+You can design your own process as a function, as long as it has a valid syntax.
+Alternatively, the `dads` package has several inbuilt processes, namely a multidimensional Brownian motion (`BM.process`) or a multidimensional Ornstein-Uhlenbeck process (`OU.process`).
+You can find the list of implemented processes by looking at the `?trait.process` manual page in `R`.
 
-Once a process is chosen, you can feed it to the `make.traits` function:
+Once a process is chosen, you can feed it into the `make.traits` function:
 
@@
 
-This creates `"dads"` `"traits"` objects that you can print, and visualise using the `plot` function:
+This creates `"dads"` `"traits"` objects that you can print and visualise using the `plot` function:
 
@@
 
-As traits can get more and more complex, the automatic printing of its summary allows for a easier display of what's in the traits object.
+As traits can get more and more complex, the automatic printing of its summary allows for a easier display of what's in the `"traits"` object.
 
-Note that it is possible to make `"traits"` objects with multiple processes (that can be the same):
+Note that it is possible to make `"traits"` objects with multiple traits and multiple processes (that can be the same or different for each trait):
 
 ```{r}
-## 4 traits: two BM, one OU and one normal non process
+## Four traits: two BM, one OU and one normal non process
 four_traits <- make.traits(process = c(BM.process,
@@
 plot(four_traits, trait = 4)
+par(mfrow = c(1,1))
 ```
@@
 This means that `make.traits` will assume that your processes are always unidimensional by default and that they always start with the value `0`.
-It is however possible to change these values.
+It is possible however, to change these values.
 
-For example you can use the following to create a three dimensional Brownian motion with each dimensions starting with the value `1`:
+For example you can use the following to create a three dimensional Brownian motion with each dimension starting with the value `1`:
 
@@
 Note that the number of traits are distributed per processes.
-If the traits contains multiple process, the number of traits are distributed per processes:
+If the traits contains multiple processes, the number of traits are distributed per processes:
 
@@
 
-### What even is a trait?
+### What is a trait in `dads`?
 
-The definition of what a trait is can vary quiet a lot depending on the context of an analysis and the field.
-<!--TODO: add REF e.g.  https://onlinelibrary.wiley.com/doi/full/10.1002/ece3.8321 -->
-Because it would be impossible to accommodate all definitions in `dads` we had to go with an arbitrary one: a trait is whatever you define as a trait!
+Because it would be impossible to accommodate all definitions of a trait in `dads` we chose an arbitrary one: a trait is whatever you define as a trait!
 A trait can be uni-dimensional as the measurement of a feature of an organism (e.g.
-leaf surface, femur length, etc...) but can also be multi-dimensional description of a feature, for example in 3D geometric morphometric, a trait could be defined as "position of landmark X" (which will be a trait with three dimensions, _x_, _y_ and _z_) or in ecology, the location of a plant can be a single 2D trait (expressed as latitude and longitude).
+leaf surface, femur length, etc.) but can also be a multi-dimensional description of a feature, for example in 3D geometric morphometric context, a trait could be defined as "position of landmark X" (which will be a trait with three dimensions, _x_, _y_ and _z_) or in ecology, the location of a plant can be expressed as latitude and longitude coordinates.
 In `dads`, the **process** corresponds to this trait definition (e.g.
@@
 You can also feed extra arguments to your process(es) functions. 
-For example, the inbuilt process `no.process` (that is just a number generator not based on the previous value `x0` or the branch length) can intake a specific random number generator as a function:
+For example, the inbuilt process `no.process` (that is just a number generator not based on the previous value `x0` or the branch length) can take a specific random number generator as a function:
 
@@
 
-If one process do not need extra argument you must still give it and extra `NULL` process argument:
+If one process does not need extra argument you must still give it an extra `NULL` process argument:
 
@@
 You can also add traits to already existing trait objects using the simple `add` option.
-This option just intakes a `"dads"` `"traits"` object and the additional process(es) will be added to it. For example:
+This option just takes a `"dads"` `"traits"` object and the additional process(es) will be added to it. 
+For example:
 
 ```{r}
-## Creating on simple default Brownian motion
+## Creating a simple default Brownian motion
 one_process <- make.traits(trait.names = "BM")
@@
 `traits` objects also allow a background trait to be used when traits are simulated ([step 3 here](#bdalgorithm)).
-This basically allows traits to be simulating for _all_ tips whenever a trait is generated for one tip.
+This allows traits to be simulated for _all_ tips whenever a trait is generated for one tip.
 This can be useful for keeping track of trait values along the simulation (_cf_ just at bifurcating nodes).
@@
 Note that technically you can nest as many background traits as you want (e.g. `make.traits(background = make.traits(background = make.traits(...)))` is valid).
-However, you should always make sure that the background trait has the same dimensions as the normal (main) trait.
+However, you should always make sure that the background trait has the same dimensions as the main trait.
 When using a trait with background, your tree will have internal singleton nodes (i.e. nodes linking to one ancestor and only one descendant).
@@
 
-Alternatively to having a background trait, you can also simulate a tree by generating traits at specific time steps with the `save.steps` option in `dads`.
+You can also simulate a tree by generating traits at specific time steps with the `save.steps` option in `dads`.
 This will apply the `traits` object to all lineages currently alive at the required time steps.
-These time steps can be either regular by providing a single numeric value; e.g. `save.steps = 0.1` will get a snapshot of the trait values every 0.1 units of time.
-Or specific, by providing a specific set of values; e.g. `save.steps = c(1, 1.2, 3)` will make a snapshot of the trait values at the required time steps.
+These time steps can be either regular by providing a single numeric value; e.g.
+`save.steps = 0.1` will take a snapshot of the trait values every 0.1 units of time, or specific, by providing a specific set of values; e.g.
+`save.steps = c(1, 1.2, 3)` will take a snapshot of the trait values at the required time steps.
 
@@
 
- * `BM.process`: this is the well famous [Brownian motion process](https://en.wikipedia.org/wiki/Brownian_motion).
+ * `BM.process`: this is the well known [Brownian motion process](https://en.wikipedia.org/wiki/Brownian_motion).
 
@@
 no process.
-In other words, this is a non time dependent process (the simulated value does not depends on the ancestors' value nor the branch length).
-It's basically a place holder for a random sampling function like `rnorm` (default), `runif`, `rlnorm`, etc...
+In other words, this is a non time dependent process; the simulated value does not depends on the ancestors' value nor the branch length.
+It's basically a place holder for a random sampling function like `rnorm` (default), `runif`, `rlnorm`, etc.
 
@@
 
- * `multi.peak.process`: this process is a modified version of the `OU.process` that can take multiple local optima.
-The default OU process has one optimum towards which the values are drawn with the alpha parameter (the elastic band) which is usually 0.
+ * `multi.peak.process`: this process is a modified version of the `OU.process` that can take multiple local long-term mean values.
+The default OU process has one long-term mean towards which the values are drawn with the alpha parameter (the elastic band).
+The single long-term mean is usually 0.
 However, with this `multi.peak.process` we can set multiple values towards which values can be attracted with the same alpha parameter.
@@
 
- * `repulsion.process`: this is a modified version of the `BM.process` where instead of accumulating normally through time, new trait values are more likely to be different than their ancestor following a `repulsion` parameter.
+ * `repulsion.process`: this is a modified version of the `BM.process` where instead of accumulating gradually through time, new trait values are more likely to be different to the traits of their ancestor following a `repulsion` parameter.
 
@@
 
-**This bit is more for development.**
+**This bit is still in development.**
 We highly suggest leaving `test = TRUE` so that `make.traits` returns an error if a process or its additional arguments (`process.args`) are not formatted correctly.

### 2023-04-05 inst/gitbook/03_make.modifiers.Rmd

@@
 
-The modifiers modify the core of the birth death process as implemented in `dads`.
+The modifiers modify the core of the birth-death process as implemented in `dads`.
 
@@
 
-This contains a lot of information, much of it is actually not useful for using the modular aspects of `dads` at a high user level.
-I.e.
+This contains a lot of information, much of it is actually not useful for using the modular aspects of `dads` at a high user level, i.e.
 unless you want to code very specific things, you won't need most of the information.
-The essential are these three functions in the elements named `"waiting$fun"`, `"selecting$fun"` and `"speciating$fun"`:
+The essentials are these three functions in the elements named `"waiting$fun"`, `"selecting$fun"` and `"speciating$fun"`:
 
  1.
-`"waiting$fun"` is the branch length function (defined in details below) which returns a randomly drawn number from an exponential distribution with the rate of the number of taxa multiplied by the speciation and extinction rate: $n \times (\lambda + \mu)$ (here is uses the extra function `sum` for internal modularity reasons).
+`"waiting$fun"` is the branch length function (defined in detail below) which returns a randomly drawn number from an exponential distribution with the rate of the number of taxa multiplied by the speciation and extinction rate: $n \times (\lambda + \mu)$ (here is uses the extra function `sum` for internal modularity reasons).
 **This function is responsible for the growth of the length/age of the tree**
@@
 ```
- With `lineage$n` being the number of lineages and `bd.params$speciation` and `bd.params$extinction` the speciation and extinction parameters (these specific terms are defined in details below).
+ With `lineage$n` being the number of lineages and `bd.params$speciation` and `bd.params$extinction` the speciation and extinction parameters (these specific terms are defined in detail below).
 
  2.
-`"selecting$fun"` is the selection function (defined in details below) which, after the waiting time defined above returns a randomly selected lineage (a tip) among the existing ones.
+`"selecting$fun"` is the selection function (defined in detail below) which, after the waiting time defined above returns a randomly selected lineage (a tip) among the existing ones.
 **This function is responsible for the branch selection**.
@@
  3.
-`"speciating$fun"` is the speciation function (defined in details below) which randomly draws a number (`runif(1)`) and depending on the speciation and extinction parameters makes the selected lineage (from 2.) speciate (`TRUE`) or go extinct (`FALSE`).
+`"speciating$fun"` is the speciation function (defined in detail below) which randomly draws a number (`runif(1)`) and depending on the speciation and extinction parameters makes the selected lineage (from 2.) speciate (`TRUE`) or go extinct (`FALSE`).
 **This function is responsible for the speciation or extinction of species**.
@@
 ```{r, eval = FALSE}
-runif(1) < (bd.params$speciation/(bd.params$speciation + bd.params$extinction)))
+runif(1) < (bd.params$speciation/(bd.params$speciation + bd.params$extinction))
 ```
 
-We will have a look at all these aspects below in more details which should make things clearer (but you can always refer to the default here for information).
+We will have a look at all these aspects below in more detail which should make things clearer (but you can always refer to the default here for information).
 
@@
 
-The first argument in `"modifiers"` is the branch length function (`branch.length`) this is the function that will be executed in `dads` to generate branch length.
-Note that in the `dads` algorithm, branch length is not generated _directly_ but actually results of the waiting time.
+The first argument in `"modifiers"` is the branch length function (`branch.length`) this is the function that will be executed in `dads` to generate branch lengths.
+Note that in the `dads` algorithm, branch length is not generated _directly_ but as the result of the waiting time.
 In other words, the `branch.length` function just affects waiting time for all taxa present at any time in the simulation.
-These taxa can then either go extinct (stopping the "growth" of it's branch length) or survive (continuing the "growth").
+These taxa can then either go extinct (stopping the "growth" of its branch length) or survive (continuing the "growth").
 
@@
 Of course, the point of the modularity here is that you can provide your own function for generating branch length.
-For example, we might be interested in what our tree would look like if we'd use a simple constant branch length generation (instead of randomly drawing it from an exponential distribution).
+For example, we might be interested in what our tree would look like if we use a simple constant branch length generation (instead of randomly drawing it from an exponential distribution).
 We can do so by declaring our own `branch.length` function and adding it to a `"modifiers"` object.
@@
 plot(modified_tree, main = "Constant branch length\nmodifier")
+par(mfrow = c(1,1))
 ```
 
-It is of course possible to use more complex branch length modifiers that intakes different conditions and specific modification rather than simply always output a value of one.
+It is of course possible to use more complex branch length modifiers that take different conditions and specific modification rather than simply always outputing a value of one.
 
-### The allowed arguments {#allowarguments}
+### The modifier arguments {#allowarguments}
 
@@
 
-  * `bd.params`: a named list containing `"numeric"` values that contains the birth death parameters (at least `"speciation"` and `"extinction"`);
+  * `bd.params`: a named list containing `"numeric"` values that contains the birth-death parameters (at least `"speciation"` and `"extinction"`);
   * `lineage`: a named list containing the lineage data (see below).
-  * `trait.values`: a `"matrix"` containing `"numeric"` values with the trait names as column names and the lineages ID as row numbers (you can use it with the function `parent.traits` to access the trait of the previous node for example).
+  * `trait.values`: a `"matrix"` containing `"numeric"` values with the trait names as column names and the lineage ID as row numbers (you can use it with the function `parent.traits` to access the trait of the previous node for example).
   * `modify.fun`: a `"list"` of named `"function"` (usually passed through `condition` and `modify`).
@@
 Therefore any argument can be left undeclared or missing and is always handled internally.
-For example, if you did not declare `lineage$n` as a function argument but are using `lineage$n` in the function, `lineage$n` will be detected and treated as a current argument automatically as set accordingly within the birth death process (e.g.
+For example, if you did not declare `lineage$n` as a function argument but are using `lineage$n` in the function, `lineage$n` will be detected and treated as a current argument automatically as set accordingly within the birth-death process (e.g.
 `lineage$n` will be set to the current number of taxa every iteration of the process).
 
-For example, we can create a function that increases branch length proportional to the number of species "alive" at each time of the simulation in a discrete way.
-I.e.
-for discrete numbers of taxa, the branch length increases by jumps (ten fold) every 5 taxa:
+For example, we can create a function that increases branch length proportional to the number of species "alive" at each time of the simulation in a discrete way, i.e.
+for discrete numbers of taxa, the branch length increases by jumps (ten fold) every five taxa:
 
@@
 plot(increasing_tree, main = "Increasing branch length\nmodifier (binned)")
+par(mfrow = c(1,1))
 ```
@@
 
-The `selection` function is used in the birth death process to know which lineage to select when running a speciation (or extinction!) event.
+The `selection` function is used in the birth-death process to know which lineage to select when running a speciation (or extinction!) event.
 By default, this function randomly selects one taxon that is currently not extinct (using: `sample(1:lineage$n, 1))`.
-Similarly as `branch.length` it is possible to modify this part of the birth death process.
-For example, we could simply select always the last created lineage (to create a "ladder" or most asymmetric tree):
+Similarly to `branch.length`, it is possible to modify this part of the birth-death process.
+For example, we could simply select the last created lineage to create a "ladder" or most asymmetric tree:
 
@@
 
-> Note that here the function can only intake the [allowed arguments as described above](#allowarguments) (here `lineage$n`: the number of current living taxa).
+> Note that here the function can only take the [allowed arguments as described above](#allowarguments) (here `lineage$n`: the number of current living taxa).
 
-We can then create a `"modifiers"` object the same way as before using this time the `selection` argument:
+We can then create a `"modifiers"` object the same way as before this time using the `selection` argument:
 
@@
 plot(ladderised_tree, main = "Ladderising modifier")
+par(mfrow = c(1,1))
 ```
@@
 
-The third function that can be used to modify the birth death process is the `speciation` function.
-This one is used during the birth death process to decide whether a lineages speciates (creating a node and two new lineages) or goes extinct (creating a tip).
+The third function that can be used to modify the birth-death process is the `speciation` function.
+This one is used during the birth-death process to decide whether a lineage speciates (creating a node and two new lineages) or goes extinct (creating a tip).
 
@@
 
-Creating a `"modifiers"` with a `speciation` function works the same way as for `branch.length` and `selection` but the function that will be used needs to output a logical value (see table [below](#summarymodifiers)).
+Creating `"modifiers"` with a `speciation` function works the same way as for `branch.length` and `selection` but the function that will be used needs to output a logical value (see table [below](#summarymodifiers)).
 Once the function is created simply input your function for speciation in the modifier and run the `dads` function with that modifier:
@@
 plot(modified_tree, main = "Random extinction\nmodifier")
+par(mfrow = c(1,1))
 ```
 
-> Note how loads of lineages end up going extinct even if the extinction parameter is set to 0!
+> Note how loads of lineages go extinct even if the extinction parameter is set to 0!
 
-And again, we can make some more advanced modifiers: for example, one where a tip always goes extinct if their ancestor has a negative trait value (here we will also introduce the utility function `parent.trait` that automatically selects the trait values of the parent of the current lineage.
+And again, we can make some more advanced modifiers: for example, one where a tip always goes extinct if their ancestor has a negative trait value.
+Here we will also introduce the utility function `parent.trait` that automatically selects the trait values of the parent of the current lineage.
 
@@
     } else {
-        ## Speciate
+        ## Speciate!
         return(TRUE)
@@
 
-In the examples above, we have seen how to specify modifications to the birth death process (via `branch.length`, `selection` and `speciation`), however, one might note that these modifications are not dynamic.
+In the examples above, we have seen how to specify modifications to the birth-death process (via `branch.length`, `selection` and `speciation`), however, these modifications are not dynamic.
 In other words, throughout the process, the modifications remain constant (even if they are conditional).
-It is however possible to code the `"modifiers"` so that they can be affected by `"events"` objects (see next chapter [on `events`](#makeevents)).
+It is however, possible to code the `"modifiers"` so that they can be affected by `"events"` objects (see next chapter [on `events`](#makeevents)).
 
-To do so, you can formally declare conditions (`condition`) and modifications (`modify`) as internal functions that can then be modified my an `"events"` object.
+To do so, you can formally declare conditions (`condition`) and modifications (`modify`) as internal functions that can then be modified by an `"events"` object.
 `condition` and `modify` are hard coded in the `branch.length` function that they concern, i.e. they are variables (functions) within the function.
@@
     ## Speciate?
-    return(trigger_event < (bd.params$speciation/(bd.params$speciation + bd.params$extinction)))
+    return(trigger_event < (bd.params$speciation/
+                              (bd.params$speciation + bd.params$extinction)))
 }
@@
 
-It is possible with some specific condition to modify this trigger by providing a `condition` and `modify` function to the speciation function.
+It is possible with some specific conditions to modify this trigger by providing a `condition` and `modify` function to the speciation function.
 That is, if the `condition` is met, apply the `modify` function to the algorithm.
@@
 ```
-Effectively, this will internally modify the speciation function as follows
+Effectively, this will internally modify the speciation function as follows:
 
@@
 These `condition` and `modify` functions can be applied to all the `modifiers` elements (`selection`, `branch.length` and `speciation`).
-However, they are typically used in `events` that will modify... the `modifiers` ([see the `events` section](#makeevents))!
+However, they are typically used in `events` that will modify the `modifiers` ([see the `events` section](#makeevents))!

### 2023-04-05 inst/gitbook/04_make.events.Rmd

@@
 
-One other major feature of the `dads` package is to allow simulations to run with specified events to occur during the simulation.
+One other major feature of the `dads` package is that it allows simulations to run with specified events that occur during the simulation.
 These are typically events that can drastically change the course of the simulation.
-For example, you might want simulate some mass extinction at some specific point in time.
+For example, you might want simulate a mass extinction at some specific point in time.
 However, these can also be more subtle, like the internal change of parameter values when reaching a specific trait value.
@@
  * the `condition` which designates when to trigger the event;
- * the `modification` which designates what to exactly modify when the event is triggered.
+ * the `modification` which designates what to modify when the event is triggered.
 
@@
 The targets that are currently available are:
- * `"taxa"` to modify anything linked to the [`lineage` list](#allowarguments). E.g. making half of the living taxa go extinct.
+ * `"taxa"` to modify anything linked to the [`lineage` list](#allowarguments). e.g. making half of the living taxa go extinct.
  * `"bd.params"` to modify anything linked to the [`bd.params` object](#makebdparams).
@@
 This is typically done by updating the object using the argument `update` from `make.modifiers`.
- * `"founding"` this target is a bit more special and allows you run a nested `dads` object in the simulation. It is cover in a [specific section below](#founding).
+ * `"founding"` this target is a bit more specialised and allows you run a nested `dads` object in the simulation. It is covered in a [specific section below](#founding).
 
-We will see some example of these targets by illustrating the `conditions` and `modifications` below.
+We will see some examples of these targets in the `conditions` and `modifications` described below.
 
@@
 
-All `condition` functions in `dads` take at least to arguments: `x` for the variable of interest (e.g.
-time, number of taxa, trait value, etc...) and `condition`, the relational operator to evaluate.
+All `condition` functions in `dads` take at least two arguments: `x` for the variable of interest (e.g.
+time, number of taxa, trait value, etc.) and `condition`, the relational operator to evaluate.
 A relational operator is the proper (fancy) term designating all the comparisons you're regularly using in `R` like `==` (is equal?) `<` (is smaller?) `>=` (is bigger or equal?).
-You can get the full list in the base manual (using `?Comparison` or any of the relational operator in a function form).
+You can get the full list in the R base manual (using `?Comparison` or any of the relational operator in a function form).
+For more details on how functions _really_ work in R see the [Advanced `R` book](https://adv-r.hadley.nz/functions.html)!
 
-> Note on the function form of a operator in `R`: if you weren't aware, functions in `R` are _usually_ described in the format `function(arguments)`, however many other functions you are commonly using are written as operators, for example (`1 + 1`).
-These are still functions though!
-And you can always use them in the classic function format by quoting them in back ticks: the following does the same as `(1+1)`: `` `+`(1, 1) ``.
-You can use that for parsing your relational operators as functions in `condition`.
-If you want you condition to be equal, use the syntax ``condition = `==` ``.
-More details on how functions _really_ work in the [Advanced `R` book](https://adv-r.hadley.nz/functions.html)!
-
- * `time.condition` will trigger the `event` once a certain time is reached in the simulations and is the simplest/more basic condition with no arguments other than the time required.
-For example ``time.condition(4, condition = `==`)`` will trigger the `event` once the simulations reached 4 units of time.
+ * `time.condition` will trigger the `event` once a certain time is reached in the simulations and is the simplest/most basic condition with no arguments other than the time required.
+For example ``time.condition(4, condition = `==`)`` will trigger the `event` once the simulations reaches 4 time units.
 Easy.
- * `taxa.condition` will trigger the `event` once a certain number of taxa is reached.
+ * `taxa.condition` will trigger the `event` once a certain number of taxa are reached.
 This can be considered including or excluding fossil species.
-For example ``taxa.condition(42, condition = `>=`, living = TRUE)`` will trigger the `event` if there is at least 42 living species.
+For example ``taxa.condition(42, condition = `>=`, living = TRUE)`` will trigger the `event` if there are at least 42 living taxa.
  * `trait.condition` will trigger the `event` once a certain trait value is reached.
@@
 
-After defining the `event`'s `target` and `condition`, you'll need to also specify what it should modify.
-These are basically functions that should modify a specific aspect of the target
+After defining the `event`, `target` and `condition`, you also need to specify what it should modify.
+These are functions that should modify a specific aspect of the target.
 Depending on the target you can modify the following:
@@
 
-The most straightforward example if for `modifications` on `bd.params`, `traits` or `modifiers` objects since they use the same syntax as for their generic `make.X` function.
+The most straightforward example is for `modifications` on `bd.params`, `traits` or `modifiers` objects because they use the same syntax as for their generic `make.X` function.
 For example, for `make.traits`, you can update a `trait` using the `update` argument as follows:
@@
 
-So basically the functions `update.X` apply the update to the object `X` when the event happens.
+So the function `update.X` applies the update to the object `X` when the event happens.
 The following `event` updates the `bd.params` object by setting the extinction parameter to 1/3 when reaching a 10 species:
@@
             modification = update.bd.params(extinction = 1/3))
-
 ```
 
-> Note that by default events are triggered only once in the whole simulation, so although the example above states that the condition is reaching at least 10 taxa, it will not trigger every times it reaches more than 10 taxa.
-You can change the number of times the events can be triggered using the argument `replications` (by default it's set to `replications = 0` for triggering it only once).
+> Note that by default events are triggered only once across the whole simulation, so although the example above states that the condition is reaching at least 10 taxa, it will not trigger every time it reaches more than 10 taxa, only the first time.
+You can change the number of times the events can be triggered using the argument `replications` (by default it's set to `replications = 0` for triggering the event only once).
 
@@
 
-Here are some example illustrating how to generate events.
+Here are some examples illustrating how to generate events.
 For a simple example, we can create a extinction event that will remove 80% of species after reaching time 4:
@@
 plot(process_change, ylim = c(-7, 7))
+par(mfrow = c(1,1))
 ```
@@
 
-Founding events are a specific `target` for `events` that allows you to... simulate a birth-death process within the current one!
-It basically allows you to simulate a specific `dads` process i.e.
-using the `dads` function with it's own `traits`, `modifiers`, `bd.params` and `events` (and yes, that's including `events` that have their own `founding` event)...
+Founding events are a specific `target` for `events` that allows you to simulate a birth-death process within the current one!
+It allows you to simulate a specific `dads` process using the `dads` function with its own `traits`, `modifiers`, `bd.params` and `events` (and yes, that's including `events` that have their own `founding` event).
 This is basically the ultimate nested boss of modularity!
 
-The founding event will basically run an internal `dads` process (i.e. simulating a tree and, optionally, some data) resulting in a founding sub-tree.
-It will then branch this sub-tree to the rest of the simulation that continued normally in the mean time.
+The founding event will run an internal `dads` process (i.e. simulating a tree and, optionally, some data) resulting in a founding sub-tree.
+It will then branch this sub-tree onto the rest of the simulation that continued normally in the mean time.
 You can specify the founding event using the inbuilt `founding.event` modification event.
-This function intakes the exact same arguments as `dads` to simulate the sub-tree with its own parameters.
+This function takes the exact same arguments as `dads` to simulate the sub-tree with its own parameters.
 Additionally, we will use the `additional.args` argument from `make.events` to specify a prefix for the founding tree tips (to make them easier to distinguish).
@@
 
-Note that the nestedness here is potentially endless, for example, you can parse an `event` argument to the `founding.event` that will generate another founding event, etc...
+Note that the nestedness here is potentially endless, for example, you can pass an `event` argument to the `founding.event` that will generate another founding event, etc.

### 2023-04-05 inst/gitbook/06_plotting_dads_objects.Rmd
Update 06_plotting_dads_objects.Rmd

@@
 
+# Plots
+
 `"dads"` objects can be directly plotted in `dads` using the S3 `plot.dads` function (or just `plot(x)` if `x` is of class `"dads"`).
 
-# Plotting traits
+## Plotting traits
 
@@
      cent.tend = sd, quantiles = c(10, 30))
+par(mfrow = c(1, 1))
 ```
@@
 You can also control the number of replicates in the simulation by using the `simulations` option (the default is 50).
-Bigger numbers leads to more time but smoother looking plots while smaller ones are more stochastic:
+Bigger numbers lead to more time but smoother looking plots while smaller ones are more stochastic:
 
@@
 plot(list_of_traits, trait = 2, simulations = 1000, main = "1k BM simulations")
+par(mfrow = c(1,1))
 ```
 
-# Plotting dads results
+## Plotting `dads` results
 
 If `dads` is used to plot only a tree (and outputs a `"phylo"` object), you can use the function `plot.phylo` from the `ape` package to plot your tree.
-You'll find all the tree plotting option in the `?plot.phylo` manual page.
+You'll find all the tree plotting options in the `?plot.phylo` manual page.
 
@@
 Branches linking them are grey lines.
-You can of course change this colour palette to something of you preference by calling the normal arguments that can be passed to `points` or `lines`.
+You can of course change this colour palette to whatever you prefer by calling the standard base R arguments that can be passed to `points` or `lines`.
 For example `pch` for the point type or `lty` for the line type.
@@
      col = c(tips = "pink", nodes = "purple"), pch = 21,
-     ## Changing edges colour and type
+     ## Changing edge colour and type
      lwd = 2, lty = 3, edges = "yellow",
@@
 
-> Note that you can modify colours using the `col` argument by providing a clear indication of what you want to colour (e.g.
+> Note that in `plot.dads` you can modify colours using the `col` argument by providing a clear indication of what you want to colour (e.g.
 `col = c(tips = "blue", fossils = "orange")` will apply the colours to the unambiguously named elements) or by providing a function that will scale the colours with time (e.g.
-`col = grDevices::heat.colors` will use the specified function to change the elements colours with time).
+`col = grDevices::heat.colors` will use the specified function to change the  colours of the elements through time).
 
-And you can add a default legend by using `legend = TRUE` (if you don't want to add the default legend you can add it after your plot using `legend(...)`).
+You can add a default legend by using `legend = TRUE` (if you don't want to add the default legend you can add it after your plot using `legend(...)`).
 
@@
 
-## 3D version!
+### Adding dimensions! Not necessary, but fun!
 
@@
 my_3D_trait <- make.traits(n = 3, )
-## Simulating a birth death tree with that 3D trait
+## Simulating a birth-death tree with that 3D trait
 my_data <- dads(bd.params = list(extinction = 0.2),
@@
 
-You can toggle which trait to plot using the `trait` option.
-Either by providing a single value to plot that specific trait against time or by providing two traits.
+You can toggle which trait to plot using the `trait` option, either by providing a single value to plot that specific trait against time or by providing two traits.
 
@@
 
-If two traits are provided, the 3rd dimensions is going to be time by default:
+If two traits are provided, the 3rd dimension is going to be time by default:
 

### 2023-04-24 inst/gitbook/03_make.modifiers.Rmd
pre-name change

@@
  * `lineage$livings`: an `"integer"` vector: the list of lineages still not extinct;
- * `lineage$drawn`: a single `"integer"`: the ID of the selected lineage;
+ * `lineage$drawn`: a single `"integer"`: the ID of the selected lineage; this must be a number in `1:lineage$n`;
  * `lineage$current`: a single `"integer"`: the selected lineage (is equal to `lineage$livings[lineage$drawn]`);

### 2023-04-24 inst/gitbook/verbose_example.Rmd
pre-name change

---
title: "`dads` verbose example"
author: "Thomas Guillerme"
date: "`r Sys.Date()`"
output:
  html_document:
    fig_width: 8
    fig_height: 8
---

```{r}
library(dispRity)
library(dads)
```

```{r}
## K-Pg mass extinction in mammals
data(BeckLee_tree)
data(BeckLee_mat99)
data(BeckLee_ages)
time_series <- chrono.subsets(data = BeckLee_mat99, tree = BeckLee_tree,
                              method = "continuous", model = "acctran", time = 20,
                              FADLAD = BeckLee_ages)
result <- dispRity(time_series, metric = c(sum, variances))
plot(result)
abline(v = 10)
results <- dispRity(boot.matrix(time_series), metric = c(sum, variances))
plot(results)
abline(v = 10)
```

```{r}
libary(dads)
my_params <- make.bd.params(speciation = rnorm)
my_stop_rule <- list(max.taxa = 50)
trait <- make.traits(process = BM.process, n = 10)
simulation <- dads(bd.params = my_params, stop.rule = my_stop_rule, trait = trait)

## Measure disparity through time
time_series <- chrono.subsets(data = simulation$data, tree = simulation$tree,
                              method = "continuous", model = "acctran", time = 20)
results <- dispRity(boot.matrix(time_series), metric = c(sum, variances))
plot(results)
```

```{r}
## 
library(dads)
set.seed(123)
bd.params <- make.bd.params(speciation = 1)
stop.rule <- list(max.taxa = 50, max.living = Inf, max.time = Inf)
traits <- make.traits(process = BM.process, background = make.traits())
par(mfrow = c(2,1))
plot(dads(bd.params = bd.params, stop.rule = stop.rule, traits = traits), main = "no modif")

## We want to change this in two ways:
 # 1- speciation = trait.value^2

select.scale.to.absolute.trait.value <- function(trait.values, lineage) {
    ## Sampling any of the living lineages with the probability of being selected being relative to the current lineage trait value. The bigger the value, the higher the probability

    ## Get the trait values of the descendants (as base probabilities + 0.01 to avoid 0 probabilities)
    sample_probs <- (abs(trait.values[as.character(lineage$parents[lineage$livings]), ]) + 0.01)

    return(as.integer(sample(1:lineage$n, size = 1, prob = sample_probs )))
}

modifiers <- make.modifiers(selection = select.scale.to.absolute.trait.value)

plot(dads(bd.params = bd.params, stop.rule = stop.rule, traits = traits, modifiers = modifiers), main = "Speciation increase\nwith absolute trait value")

## THIS IS BUGGED: to test in birth.death_fun.R: problem with selecting the current lineage (NA)

```

### 2023-04-24 R/birth.death_fun.R

@@ -206,10 +206,10 @@ birth.death.tree.traits <- function(stop.rule, bd.params, traits = NULL, modifie
             return(NULL)
         }
     }
-
     # ##DEBUG: creating right handed trees
     # warning("DEBUG birth.death_fun.R") ; lineage$current <- 3 ; lineage$drawn   <- 2
-
+    # warning("DEBUG"); lineage_base <- lineage; trait_values_base <- trait_values; edge_lengths_base <- edge_lengths
+    # lineage <- lineage_base ; trait_values <- trait_values_base; edge_lengths <- edge_lengths_base
     ############
     ## Rest of the tree
     ############
@@ -234,7 +234,6 @@ birth.death.tree.traits <- function(stop.rule, bd.params, traits = NULL, modifie
                             lineage      = lineage,
                             trait.values = trait_values,
                             modify.fun   = modifiers$selecting$internal)
-        
         lineage$current <- lineage$livings[lineage$drawn]
 
         ## Get the waiting time
@@ -288,6 +287,7 @@ birth.death.tree.traits <- function(stop.rule, bd.params, traits = NULL, modifie
 
 
             if(!is.null(traits$background)) {
+
                 # time.slice <- first_waiting_time + time
                 ## Update the lineage and create generate background traits
                 lineage      <- update.single.nodes(lineage)
@@ -323,7 +323,6 @@ birth.death.tree.traits <- function(stop.rule, bd.params, traits = NULL, modifie
 
         # warning("DEBUG birth.death_fun.R") ; step_counter <- step_counter + 1; record_everything[[step_counter]] <- list(lineage = lineage, edge_lengths = edge_lengths, trait_values = trait_values)
 
-
         ## Randomly triggering an event
         if(modifiers$speciating$fun(bd.params    = sample.from(bd.params),
                                     lineage      = lineage,
@@ -350,7 +349,7 @@ birth.death.tree.traits <- function(stop.rule, bd.params, traits = NULL, modifie
             ## Go extinct
             was_alive <- lineage$livings[lineage$drawn]
             lineage$livings <- lineage$livings[-lineage$drawn]
-            lineage$n <- lineage$n - 1L   
+            lineage$n <- lineage$n - 1L
         }
 
         # warning("DEBUG"); reached_event <- TRUE

### 2023-04-24 inst/dads_paper.Rmd
Update dads_paper.Rmd
Big rename

@@
 ---
-title: "`dads`: a modular `R` package for simulating diversity and traits data."
+title: "`treats`: a modular `R` package for simulating diversity and traits data."
 author: "homas Guillerme, Natalie Cooper, Andrew P. Beckerman, and Gavin H. Thomas"
@@
 
-<!-- SET URL FOR [manual](dads_manual) -->
+<!-- SET URL FOR [manual](treats_manual) -->
 
@@
 
-3. Here I present the `dads` package, a modular `R` package for diversity and trait disparity simulations.
+3. Here I present the `treats` package, a modular `R` package for diversity and trait disparity simulations.
 This package is based on a standard birth death algorithm that can be easily tuned to the user specific needs by designing their specific trait process, birth-death modifiers and events.
-It also provides a tidy interface through the `dads` object, allowing users to easily run reproducible simulations.
+It also provides a tidy interface through the `treats` object, allowing users to easily run reproducible simulations.
 
-4. The `dads` package also comes with an extend manual regularly updated following users' questions or suggestions.
+4. The `treats` package also comes with an extend manual regularly updated following users' questions or suggestions.
 
@@
 Although we don't think this was the aim of these specific packages authors (not to create a strawman argument) we think this hampers the use of specific null model (or baseline model) to test specific dynamic mechanisms or to compare to specific datasets.
-<!-- % The solution with dads -->
-Therefore, we propose `dads` a modular `R` package to simulate both diversity and disparity through time.
+<!-- % The solution with treats -->
+Therefore, we propose `treats` a modular `R` package to simulate both diversity and disparity through time.
 
@@
 <!-- % Very briefly how does it work -->
-`dads` is based on a core birth-death process implemented in the function `dads` that simulates diversity through time using the following steps:
+`treats` is based on a core birth-death process implemented in the function `treats` that simulates diversity through time using the following steps:
 
@@
 
-In `dads`, these three or four steps are implemented as modular functions that the user can easily change using an internal class of objects called `"modifiers"` or `"traits"` (Fig.
+In `treats`, these three or four steps are implemented as modular functions that the user can easily change using an internal class of objects called `"modifiers"` or `"traits"` (Fig.
 \ref{Fig_workflow}).
-The simulation then outputs a tree (of class `"phylo"` and a associated table of traits - `"matrix"`) that can be visualised using the `plot.dads` function.
+The simulation then outputs a tree (of class `"phylo"` and a associated table of traits - `"matrix"`) that can be visualised using the `plot.treats` function.
 
-![Figure 1:](gitbook/dads_structure.pdf)
+![Figure 1:](gitbook/treats_structure.pdf)
 
-\label{Fig_workflow}`dads` package workflow @@@. `"events"` are an additional category of modular objects discussed later.
+\label{Fig_workflow}`treats` package workflow @@@. `"events"` are an additional category of modular objects discussed later.
 
@@
 ## and stopping when it reaches 20 taxa
-my_simulation <- dads(bd.params = list(speciation = 1, extinction = 0.1),
+my_simulation <- treats(bd.params = list(speciation = 1, extinction = 0.1),
                       stop.rule = list(max.taxa = 20))
@@
 
-These are included by default in a `dads` process.
+These are included by default in a `treats` process.
 The modularity of the packages works with the ability to code the modifiers manually rather than using default ones.
-Using a reserved syntax in this case the arguments `lineage` and `bd.params` (much more details about the syntax is available in the [manual](dads_manual)). <!-- TODO: make link -->
+Using a reserved syntax in this case the arguments `lineage` and `bd.params` (much more details about the syntax is available in the [manual](treats_manual)). <!-- TODO: make link -->
 For example, to simulate the same process but hard coding the modifiers we can use the following functions for each process:
@@
 ## and stopping when it reaches 20 taxa with the default birth-death modifiers specified
-my_simulation <- dads(bd.params = list(speciation = 1, extinction = 0.1),
+my_simulation <- treats(bd.params = list(speciation = 1, extinction = 0.1),
                       stop.rule = list(max.taxa = 20),
@@
 For example if traits are also simulated (see below) we can bias the branch length to be also dependent on the trait distribution, the selection to be biased towards species at the extremes of the distribution and the speciation modifier to only allow speciation when the lineage's parent trait value is positive.
-More example of biased or special `"modifiers"` are available in the [manual](dads_manual).
+More example of biased or special `"modifiers"` are available in the [manual](treats_manual).
 
@@
 
-Another major feature of `dads` is to allow to specify the trait simulation process manually as a `"traits"` object.
+Another major feature of `treats` is to allow to specify the trait simulation process manually as a `"traits"` object.
 This can be done in the same way as for the `"modifiers"` but with the `make.traits` function.
-Here the Brownian motion syntax is also specific (especially `x0` and `edge.length`), more details are available in the [manual](dads_manual).
+Here the Brownian motion syntax is also specific (especially `x0` and `edge.length`), more details are available in the [manual](treats_manual).
 
@@
 ## and a Brownian Motion trait process simulated
-my_simulation <- dads(bd.params = list(speciation = 1, extinction = 0.1),
+my_simulation <- treats(bd.params = list(speciation = 1, extinction = 0.1),
                       stop.rule = list(max.taxa = 20),
@@
 set.seed(2)
-my_simulation <- dads(bd.params = list(speciation = 1, extinction = 0.1),
+my_simulation <- treats(bd.params = list(speciation = 1, extinction = 0.1),
                       stop.rule = list(max.taxa = 20),
@@
 
-The resulting object is a `"dads"` object containing a `$tree` element (a `"phylo"`) object and a `$data` object containing a `"matrix"` object:
+The resulting object is a `"treats"` object containing a `$tree` element (a `"phylo"`) object and a `$data` object containing a `"matrix"` object:
 
 ```{r, eval = TRUE, echo = TRUE, fig.height = 4, fig.width = 4}
-## Printing the dads object
+## Printing the treats object
 my_simulation
@@
 
-The package also comes with it's own plotting function allowing to directly display `dads` generated tree and data in the same plot:
+The package also comes with it's own plotting function allowing to directly display `treats` generated tree and data in the same plot:
 
 ```{r, eval = TRUE, echo = TRUE, fig.height = 4, fig.width = 4}
-## A default dads plot
+## A default treats plot
 plot(my_simulation, main = "Default plot")
@@
 
-Note that here too many options are available for plotting such objects. More information is again available in the [manual](dads_manual).
+Note that here too many options are available for plotting such objects. More information is again available in the [manual](treats_manual).
 
@@
 
-## `dads` compared to other packages
+Every bit can be coded as a function.
+Many already implemented functions (e.g. in events.modifications) but many more possibilities to modify the current ones or design completely new ones.
+
+## `treats` compared to other packages
 
@@
 # Conclusion
-The `dads` is modular and nice to use.
+The `treats` is modular and nice to use.
 
 # Package location
-The `dads` package is available on the CRAN at \url{https://cran.r-project.org/web/packages/dads/index.html} or on GitHub at \url{https://github.com/TGuillerme/dads} with more associated information.
+The `treats` package is available on the CRAN at \url{https://cran.r-project.org/web/packages/treats/index.html} or on GitHub at \url{https://github.com/TGuillerme/treats} with more associated information.
 All the versions of the package are archived on ZENODO with associated DOI \url{https://zenodo.org/@@@}.

### 2023-04-24 inst/gitbook/01_getting_started.Rmd
Big rename

@@
 ```{r, echo = FALSE}
-library(dads)
+library(treats)
 set.seed(4242)
@@
 
-One of the simplest things to do with the `dads` package is to simulate a birth-death tree.
-For that you can use the function `dads` and specify your stopping rule.
+One of the simplest things to do with the `treats` package is to simulate a birth-death tree.
+For that you can use the function `treats` and specify your stopping rule.
 The stopping rule simply tells the birth-death process to stop whenever it reaches one of these three conditions:
@@
 ## Running the birth-death simulation
-my_tree <- dads(stop.rule = my_stop_rule)
+my_tree <- treats(stop.rule = my_stop_rule)
 ```
@@
 People familiar with the [birth-death models](https://lukejharmon.github.io/pcm/chapter10_birthdeath/) might have noticed that we did not specify two important things here: the speciation parameter (sometimes called "lambda" or "birth") and the extinction parameter (sometimes called "mu", "death" or "background extinction").
-By default `dads` runs a pure birth model (speciation is set to 1 and extinction to 0).
+By default `treats` runs a pure birth model (speciation is set to 1 and extinction to 0).
 However, you can easily change that by specifying your own birth-death parameters:
@@
 ## Generating a birth-death tree with extinctions:
-my_tree <- dads(bd.params = my_params, stop.rule = my_stop_rule)
+my_tree <- treats(bd.params = my_params, stop.rule = my_stop_rule)
 ## Visualising the new tree
@@
 
-Chances are that you are using `dads` because you also want to simulate traits (disparity) along with your diversity (otherwise, we suggest using the [`TreeSim`](https://github.com/tanja819/TreeSim/) package that provides many more birth-death models).
-Simulating traits is not much more complicated in `dads`: you'll simply need to create a `"traits"` object using the `make.traits` function.
+Chances are that you are using `treats` because you also want to simulate traits (disparity) along with your diversity (otherwise, we suggest using the [`TreeSim`](https://github.com/tanja819/TreeSim/) package that provides many more birth-death models).
+Simulating traits is not much more complicated in `treats`: you'll simply need to create a `"traits"` object using the `make.traits` function.
 These objects can have increasing complexity (see the rest of this tutorial) but we will keep it simple here.
@@
 This is not a really good name but you'll see more about specifying trait names later on.
-If this is what the process should look like (theoretically) you can then add its `"traits"` object to our previous `dads` function to generate the tree and the traits:
+If this is what the process should look like (theoretically) you can then add its `"traits"` object to our previous `treats` function to generate the tree and the traits:
 
@@
 ## Simulate disparity and diversity
-my_data <- dads(bd.params = my_params,
+my_data <- treats(bd.params = my_params,
                 stop.rule = my_stop_rule,
@@
 
-This will not do anything to our simulations compared to the previous trait and tree simulation but we can provide our `"modifiers"` object to the `dads` function:
+This will not do anything to our simulations compared to the previous trait and tree simulation but we can provide our `"modifiers"` object to the `treats` function:
 
@@
 ## Simulate disparity and diversity
-default_data <- dads(bd.params = extinction_02,
+default_data <- treats(bd.params = extinction_02,
                      stop.rule = living_20,
@@
 For that we can use the `parent.traits` utility function that is optimised for accessing traits in the birth-death process (but you can of course write your own).
-This function takes the `trait.values` and `parent.lineage` arguments, two arguments that you can leave named as they are to facilitate `dads`' understanding of what you want to assess:
+This function takes the `trait.values` and `parent.lineage` arguments, two arguments that you can leave named as they are to facilitate `treats`' understanding of what you want to assess:
 
@@
 We can then provide these two functions (the condition `negative.ancestor` and how to modify the speciation event when this condition is met `staying.alive`).
-If you are an advances `dads` user, you can design your own `speciation` function but if you just want to use a normal `speciation` function, you can use the default one from `dads` called `speciation`.
+If you are an advances `treats` user, you can design your own `speciation` function but if you just want to use a normal `speciation` function, you can use the default one from `treats` called `speciation`.
 
@@
 
-Note that the `make.modifiers` function tests whether the input is compatible with `dads` by default so unless you have an error message, your `modifiers` will work!
+Note that the `make.modifiers` function tests whether the input is compatible with `treats` by default so unless you have an error message, your `modifiers` will work!
 We can now simulate our tree and traits with our modifier: species will go extinct if their ancestor has a negative trait value:
@@
 ## Simulate disparity and diversity
-biased_data <- dads(bd.params = extinction_02,
+biased_data <- treats(bd.params = extinction_02,
                     stop.rule = living_20,

### 2023-04-24 inst/gitbook/02_make.traits.Rmd
Big rename

@@
 ```{r, echo = FALSE}
-library(dads)
+library(treats)
 set.seed(123)
@@
 
-In `dads`, traits are simulating by providing a `traits` object to the `traits` argument.
+In `treats`, traits are simulating by providing a `traits` object to the `traits` argument.
 This object is generated using `make.traits`.
@@
 Basically anything can be a trait if it's defined by the user (e.g. one trait can be a 20 dimensional Brownian process that's correlated to a 1D OU process - which is another trait).
-For the purpose of the syntax in _this package_ a **trait in `dads`** is a column in the output trait matrix.
+For the purpose of the syntax in _this package_ a **trait in `treats`** is a column in the output trait matrix.
 For example, one might be interested in a biological trait that is "beak shape" which is defined as one complex biological trait that is the Procrustes superimposition of 4 3D landmarks and 300 3D semi-landmarks.
-This single "beak shape" **biological trait** is then composed of 912 **traits in `dads`** (i.e. a matrix of $(4+300)\times3$ columns).
+This single "beak shape" **biological trait** is then composed of 912 **traits in `treats`** (i.e. a matrix of $(4+300)\times3$ columns).
 
@@
 
-Trait processes in `dads` are functions that must always take the following arguments by default.
+Trait processes in `treats` are functions that must always take the following arguments by default.
 
@@
 > Note that in this function definition the arguments `x0` and `edge.length` have a default value of `0` and `1` respectively.
-In practice, these arguments are effectively set to the correct values in the `dads` internal function (i.e.
+In practice, these arguments are effectively set to the correct values in the `treats` internal function (i.e.
 whatever `x0` and `edge.length` are at that specific time of the process) but providing a default can help speed up the algorithms (specifically all the internal checks).
@@
 
-### Using a `"process"` in `dads`
+### Using a `"process"` in `treats`
 
 You can design your own process as a function, as long as it has a valid syntax.
-Alternatively, the `dads` package has several inbuilt processes, namely a multidimensional Brownian motion (`BM.process`) or a multidimensional Ornstein-Uhlenbeck process (`OU.process`).
+Alternatively, the `treats` package has several inbuilt processes, namely a multidimensional Brownian motion (`BM.process`) or a multidimensional Ornstein-Uhlenbeck process (`OU.process`).
 You can find the list of implemented processes by looking at the `?trait.process` manual page in `R`.
@@
 
-This creates `"dads"` `"traits"` objects that you can print and visualise using the `plot` function:
+This creates `"treats"` `"traits"` objects that you can print and visualise using the `plot` function:
 
@@
 
-Note that you can see the multiple options for plotting the trait process by looking at `?plot.dads` manual.
-Furthermore, you can look at what's actually in the object using this specific syntax (this applies to every object handled by the `dads` package):
+Note that you can see the multiple options for plotting the trait process by looking at `?plot.treats` manual.
+Furthermore, you can look at what's actually in the object using this specific syntax (this applies to every object handled by the `treats` package):
 
@@
 ## What's actually in that object?
-print.dads(my_trait_object, all = TRUE)
+print.treats(my_trait_object, all = TRUE)
 ```
@@
 
-You can visualise them individually using the `trait` argument in `plot.dads`:
+You can visualise them individually using the `trait` argument in `plot.treats`:
 
@@
 
-### What is a trait in `dads`?
+### What is a trait in `treats`?
 
-Because it would be impossible to accommodate all definitions of a trait in `dads` we chose an arbitrary one: a trait is whatever you define as a trait!
+Because it would be impossible to accommodate all definitions of a trait in `treats` we chose an arbitrary one: a trait is whatever you define as a trait!
 A trait can be uni-dimensional as the measurement of a feature of an organism (e.g.
 leaf surface, femur length, etc.) but can also be a multi-dimensional description of a feature, for example in 3D geometric morphometric context, a trait could be defined as "position of landmark X" (which will be a trait with three dimensions, _x_, _y_ and _z_) or in ecology, the location of a plant can be expressed as latitude and longitude coordinates.
-In `dads`, the **process** corresponds to this trait definition (e.g.
+In `treats`, the **process** corresponds to this trait definition (e.g.
 a process can be of n-dimensions and represents one organisms feature) and the **traits** represents the number of dimensions in total.
-So in the examples above, this is how the following traits are interpreted by `dads`:
+So in the examples above, this is how the following traits are interpreted by `treats`:
 
@@
 You can also add traits to already existing trait objects using the simple `add` option.
-This option just takes a `"dads"` `"traits"` object and the additional process(es) will be added to it. 
+This option just takes a `"treats"` `"traits"` object and the additional process(es) will be added to it. 
 For example:
@@
 ## Generating a pure birth tree with the background trait
-tree_bkg <- dads(stop.rule = list(max.taxa = 20),
+tree_bkg <- treats(stop.rule = list(max.taxa = 20),
                  traits = my_trait)
@@
 
-You can also simulate a tree by generating traits at specific time steps with the `save.steps` option in `dads`.
+You can also simulate a tree by generating traits at specific time steps with the `save.steps` option in `treats`.
 This will apply the `traits` object to all lineages currently alive at the required time steps.
@@
 ## Generating a birth-death tree with a BM trait and saving steps at specific times
-tree_steps <- dads(stop.rule  = list(max.time = 3),
+tree_steps <- treats(stop.rule  = list(max.time = 3),
                    bd.params  = list(speciation = 1, extinction = 0.1),
@@
 
-## Traits implemented in `dads`
+## Traits implemented in `treats`
 
-If you don't want to design your own trait process, you can use one of the following trait processes that are currently implemented in `dads`.
+If you don't want to design your own trait process, you can use one of the following trait processes that are currently implemented in `treats`.
 You can find more information about their many options using their specific manuals in R or the generic `?trait.process`:
@@
 We highly suggest leaving `test = TRUE` so that `make.traits` returns an error if a process or its additional arguments (`process.args`) are not formatted correctly.
-`make.traits` will send an error if the trait cannot be directly passed to `dads`.
+`make.traits` will send an error if the trait cannot be directly passed to `treats`.
 However, in some specific cases (again, probably mainly for development and debugging) it could be useful to skip the tests using `test = FALSE`.
@@
 You can be imaginative and creative when designing your own process but here are two detailed example functions for a unidimensional Brownian Motion and Ornstein-Uhlenbeck process that you can use for a start (or not).
-Remember it is good practice for `dads` processes to set all the arguments with default values (just in case).
+Remember it is good practice for `treats` processes to set all the arguments with default values (just in case).
 

### 2023-04-24 inst/gitbook/03_make.modifiers.Rmd
Big rename

@@
 
-Similarly to `"traits"` objects, `"modifiers"` are also printed by default using `print.dads`.
-You can see details about what's actually in the object using `print.dads(my_default_modifiers, all = TRUE)`.
+Similarly to `"traits"` objects, `"modifiers"` are also printed by default using `print.treats`.
+You can see details about what's actually in the object using `print.treats(my_default_modifiers, all = TRUE)`.
 However, contrary to `"traits"`, you cannot plot `"modifiers"`.
@@
 
-The modifiers modify the core of the birth-death process as implemented in `dads`.
+The modifiers modify the core of the birth-death process as implemented in `treats`.
 
@@
 
-By default, the birth-death process in `dads` uses this modifier:
+By default, the birth-death process in `treats` uses this modifier:
 
@@
 
-This contains a lot of information, much of it is actually not useful for using the modular aspects of `dads` at a high user level, i.e.
+This contains a lot of information, much of it is actually not useful for using the modular aspects of `treats` at a high user level, i.e.
 unless you want to code very specific things, you won't need most of the information.
@@
 
-The first argument in `"modifiers"` is the branch length function (`branch.length`) this is the function that will be executed in `dads` to generate branch lengths.
-Note that in the `dads` algorithm, branch length is not generated _directly_ but as the result of the waiting time.
+The first argument in `"modifiers"` is the branch length function (`branch.length`) this is the function that will be executed in `treats` to generate branch lengths.
+Note that in the `treats` algorithm, branch length is not generated _directly_ but as the result of the waiting time.
 In other words, the `branch.length` function just affects waiting time for all taxa present at any time in the simulation.
@@
 By default, branch length (or waiting/growth) is a randomly drawn number from an exponential distribution with the rate of the number of taxa multiplied by the speciation and extinction rate: $n \times (\lambda + \mu)$ (where $n$ is the number of taxa currently present in the simulation, $\lambda$ and $\mu$ are respectively the speciation and extinction rates).
-This default function is simply called `branch.length` in `dads` and can be used as a modifier as follows:
+This default function is simply called `branch.length` in `treats` and can be used as a modifier as follows:
 
@@
 set.seed(0)
-default_tree <- dads(bd.params = bd_params,
+default_tree <- treats(bd.params = bd_params,
                      stop.rule = stop_rule,
@@
 set.seed(0)
-modified_tree <- dads(bd.params = bd_params,
+modified_tree <- treats(bd.params = bd_params,
                       stop.rule = stop_rule,
@@
 
-In general, unless you know what you're doing, you can ignore most arguments for specific modifiers since they are handled automatically within the `dads` function.
+In general, unless you know what you're doing, you can ignore most arguments for specific modifiers since they are handled automatically within the `treats` function.
 Therefore any argument can be left undeclared or missing and is always handled internally.
@@
 set.seed(0)
-increasing_tree <- dads(bd.params = bd_params,
+increasing_tree <- treats(bd.params = bd_params,
                         stop.rule = stop_rule,
@@
 set.seed(0)
-ladderised_tree <- dads(bd.params = bd_params,
+ladderised_tree <- treats(bd.params = bd_params,
                         stop.rule = stop_rule,
@@
 Again, it is of course possible to make the modifier more complex and in combination with other elements of the tree.
-For example, we can create a `"dads"` object that also generates a BM trait and add to it this object a `selection` modifier that only selects tips with positive trait values (only species with positive trait values will speciate).
+For example, we can create a `"treats"` object that also generates a BM trait and add to it this object a `selection` modifier that only selects tips with positive trait values (only species with positive trait values will speciate).
 
@@
 set.seed(1)
-default_dads <- dads(bd.params = bd_params,
+default_treats <- treats(bd.params = bd_params,
                      stop.rule = stop_rule,
@@
 set.seed(1)
-skewed_trait_dads <- dads(bd.params = bd_params,
+skewed_trait_treats <- treats(bd.params = bd_params,
                           stop.rule = stop_rule,
@@
 par(mfrow = c(1, 2))
-plot(default_dads, main = "Default trait and tree")
-plot(skewed_trait_dads, main = "Skewed trait and tree")
+plot(default_treats, main = "Default trait and tree")
+plot(skewed_trait_treats, main = "Skewed trait and tree")
 ```
@@
 Creating `"modifiers"` with a `speciation` function works the same way as for `branch.length` and `selection` but the function that will be used needs to output a logical value (see table [below](#summarymodifiers)).
-Once the function is created simply input your function for speciation in the modifier and run the `dads` function with that modifier:
+Once the function is created simply input your function for speciation in the modifier and run the `treats` function with that modifier:
 
@@
 set.seed(2)
-modified_tree <- dads(bd.params = bd_params,
+modified_tree <- treats(bd.params = bd_params,
                       stop.rule = stop_rule,
@@
 set.seed(1)
-biased_trait_dads <- dads(bd.params = bd_params,
+biased_trait_treats <- treats(bd.params = bd_params,
                           stop.rule = stop_rule,
@@
 par(mfrow = c(1, 2))
-plot(default_dads, main = "Default trait and tree")
-plot(biased_trait_dads, main = "Biased trait and tree")
+plot(default_treats, main = "Default trait and tree")
+plot(biased_trait_treats, main = "Biased trait and tree")
 ```

### 2023-04-24 inst/gitbook/04_make.events.Rmd
Big rename

@@
 
-One other major feature of the `dads` package is that it allows simulations to run with specified events that occur during the simulation.
+One other major feature of the `treats` package is that it allows simulations to run with specified events that occur during the simulation.
 These are typically events that can drastically change the course of the simulation.
@@
 However, these can also be more subtle, like the internal change of parameter values when reaching a specific trait value.
-These are all handled by `dads` with the `events` object that you can create using `make.events` and has the same overall logic as `make.traits` and `make.modifiers`.
+These are all handled by `treats` with the `events` object that you can create using `make.events` and has the same overall logic as `make.traits` and `make.modifiers`.
 
@@
 This is typically done by updating the object using the argument `update` from `make.modifiers`.
- * `"founding"` this target is a bit more specialised and allows you run a nested `dads` object in the simulation. It is covered in a [specific section below](#founding).
+ * `"founding"` this target is a bit more specialised and allows you run a nested `treats` object in the simulation. It is covered in a [specific section below](#founding).
 
@@
 
-Currently there are three conditions functions implemented in `dads` but you can easily come up with your own version of them.
+Currently there are three conditions functions implemented in `treats` but you can easily come up with your own version of them.
 
-All `condition` functions in `dads` take at least two arguments: `x` for the variable of interest (e.g.
+All `condition` functions in `treats` take at least two arguments: `x` for the variable of interest (e.g.
 time, number of taxa, trait value, etc.) and `condition`, the relational operator to evaluate.
@@
 set.seed(123)
-results <- dads(bd.params = bd.params,
+results <- treats(bd.params = bd.params,
                 stop.rule = stop.rule,
@@
 set.seed(1)
-no_change <- dads(bd.params = bd.params,
+no_change <- treats(bd.params = bd.params,
                   stop.rule = stop.rule,
@@
 set.seed(1)
-process_change <- dads(bd.params = bd.params,
+process_change <- treats(bd.params = bd.params,
                        stop.rule = stop.rule,
@@
 Founding events are a specific `target` for `events` that allows you to simulate a birth-death process within the current one!
-It allows you to simulate a specific `dads` process using the `dads` function with its own `traits`, `modifiers`, `bd.params` and `events` (and yes, that's including `events` that have their own `founding` event).
+It allows you to simulate a specific `treats` process using the `treats` function with its own `traits`, `modifiers`, `bd.params` and `events` (and yes, that's including `events` that have their own `founding` event).
 This is basically the ultimate nested boss of modularity!
 
-The founding event will run an internal `dads` process (i.e. simulating a tree and, optionally, some data) resulting in a founding sub-tree.
+The founding event will run an internal `treats` process (i.e. simulating a tree and, optionally, some data) resulting in a founding sub-tree.
 It will then branch this sub-tree onto the rest of the simulation that continued normally in the mean time.
 You can specify the founding event using the inbuilt `founding.event` modification event.
-This function takes the exact same arguments as `dads` to simulate the sub-tree with its own parameters.
+This function takes the exact same arguments as `treats` to simulate the sub-tree with its own parameters.
 Additionally, we will use the `additional.args` argument from `make.events` to specify a prefix for the founding tree tips (to make them easier to distinguish).
@@
 set.seed(11)
-founding_tree <- dads(bd.params = bd.params,
+founding_tree <- treats(bd.params = bd.params,
                       stop.rule = stop.rule,

### 2023-04-24 inst/gitbook/05_other_functionalities.Rmd
Big rename

@@
 ```{r, echo = FALSE}
-library(dads)
+library(treats)
 library(rgl)
@@
 set.seed(123)
-plot(dads(stop.rule = list(max.taxa = 50), bd.params = my_bd_params))
+plot(treats(stop.rule = list(max.taxa = 50), bd.params = my_bd_params))
 ## Note the regions in the tree with short branches
@@
 
-In this example, the `"bd.params"` object passed to the `dads` function will allow the birth-death process to sample the speciation parameter each time it is called (e.g.
+In this example, the `"bd.params"` object passed to the `treats` function will allow the birth-death process to sample the speciation parameter each time it is called (e.g.
 during the speciation/extinction step, the branch length step, etc.).
@@
 set.seed(3)
-my_data <- dads(stop.rule = list(max.taxa = 20),
+my_data <- treats(stop.rule = list(max.taxa = 20),
                 bd.params = list(speciation = 1, extinction = 1/3),
@@
 
-## `"dads"` internal utilities
+## `"treats"` internal utilities
 
 The package also provides utilities for internal functions, namely for designing `modifiers` or `events` more easily.
-These functions don't do anything useful on their own but are optimised to be used internally in `dads`.
+These functions don't do anything useful on their own but are optimised to be used internally in `treats`.
 For all these functions, you can look at the internal manual for an example (i.e. using `?<function_name>`)

### 2023-04-24 inst/gitbook/06_plotting_dads_objects.Rmd
Big rename

@@
 ---
-title: "`plot.dads`"
+title: "`plot.treats`"
 author: "Thomas Guillerme (guillert@tcd.ie)"
@@
 
-`"dads"` objects can be directly plotted in `dads` using the S3 `plot.dads` function (or just `plot(x)` if `x` is of class `"dads"`).
+`"treats"` objects can be directly plotted in `treats` using the S3 `plot.treats` function (or just `plot(x)` if `x` is of class `"treats"`).
 
@@
 
-`"dads"` `"traits"` objects are covered in the [traits section](#maketraits).
-You can use `plot.dads` to plot them by choosing which specific trait to plot using the `trait` argument (default is 1):
+`"treats"` `"traits"` objects are covered in the [traits section](#maketraits).
+You can use `plot.treats` to plot them by choosing which specific trait to plot using the `trait` argument (default is 1):
 
@@
 
-## Plotting `dads` results
+## Plotting `treats` results
 
-If `dads` is used to plot only a tree (and outputs a `"phylo"` object), you can use the function `plot.phylo` from the `ape` package to plot your tree.
+If `treats` is used to plot only a tree (and outputs a `"phylo"` object), you can use the function `plot.phylo` from the `ape` package to plot your tree.
 You'll find all the tree plotting options in the `?plot.phylo` manual page.
@@
 ## A simple pure birth tree
-my_tree <- dads(stop.rule = list(max.taxa = 20))
+my_tree <- treats(stop.rule = list(max.taxa = 20))
 plot(my_tree, main = "Plotting a \"phylo\" object")
@@
 
-However, if you also simulated a trait along with the tree you can use the `plot.dads` function to plot both the tree and the trait:
+However, if you also simulated a trait along with the tree you can use the `plot.treats` function to plot both the tree and the trait:
 
@@
 ## A simple pure birth tree with a BM process
-my_tree <- dads(stop.rule = list(max.taxa = 20), traits = make.traits())
+my_tree <- treats(stop.rule = list(max.taxa = 20), traits = make.traits())
 ## Playing with the default options
@@
 
-> Note that in `plot.dads` you can modify colours using the `col` argument by providing a clear indication of what you want to colour (e.g.
+> Note that in `plot.treats` you can modify colours using the `col` argument by providing a clear indication of what you want to colour (e.g.
 `col = c(tips = "blue", fossils = "orange")` will apply the colours to the unambiguously named elements) or by providing a function that will scale the colours with time (e.g.
@@
 ## Simulating a birth-death tree with that 3D trait
-my_data <- dads(bd.params = list(extinction = 0.2),
+my_data <- treats(bd.params = list(extinction = 0.2),
                 stop.rule = list(max.living = 50),
@@
 But there's more! You can also plot these results using actual 3D plots that you can spin and all that!
-This is done through the `rgl` package and activated in `dads` using the option `use.3D = TRUE`.
+This is done through the `rgl` package and activated in `treats` using the option `use.3D = TRUE`.
 

### 2023-04-24 inst/gitbook/07_specific_examples.Rmd
Big rename

@@
 set.seed(1)
-results <- dads(stop.rule = stop_time_6,
+results <- treats(stop.rule = stop_time_6,
                 bd.params = speciation_1,
@@
 set.seed(7)
-results <- dads(stop.rule = stop_time_5,
+results <- treats(stop.rule = stop_time_5,
                 bd.params = speciation_1,
@@
 set.seed(2)
-results <- dads(stop.rule = stop_taxa_50,
+results <- treats(stop.rule = stop_taxa_50,
                 bd.params = speciation_1,
@@
 set.seed(42)
-no_event <- dads(stop.rule = stop_time_6,
+no_event <- treats(stop.rule = stop_time_6,
                  bd.params = random_speciation)
@@
 set.seed(42)
-reduced_speciation_event <- dads(stop.rule = stop_time_6,
+reduced_speciation_event <- treats(stop.rule = stop_time_6,
                                  bd.params = random_speciation,
@@
 set.seed(1)
-no_change <- dads(stop.rule = stop_time_6,
+no_change <- treats(stop.rule = stop_time_6,
                   bd.params = speciation_1,
@@
 set.seed(1)
-process_change <- dads(stop.rule = stop_time_6,
+process_change <- treats(stop.rule = stop_time_6,
                        bd.params = speciation_1,
@@
 set.seed(2)
-no_event <- dads(stop.rule = stop_taxa_100,
+no_event <- treats(stop.rule = stop_taxa_100,
                  bd.params = speciation_1,
@@
 set.seed(2)
-change_correlation <- dads(stop.rule = stop_taxa_100,
+change_correlation <- treats(stop.rule = stop_taxa_100,
                            bd.params = speciation_1,
@@
 set.seed(4)
-no_event <- dads(stop.rule = stop_time_4,
+no_event <- treats(stop.rule = stop_time_4,
                  bd.params = speciation_1,
@@
 set.seed(4)
-change_spec <- dads(stop.rule = stop_time_4,
+change_spec <- treats(stop.rule = stop_time_4,
                     bd.params = speciation_1,
@@
 set.seed(5)
-no_event <- dads(stop.rule = stop_taxa_100,
+no_event <- treats(stop.rule = stop_taxa_100,
                  bd.params = speciation_1,
@@
 set.seed(5)
-increased_brlen <- dads(stop.rule = stop_taxa_100,
+increased_brlen <- treats(stop.rule = stop_taxa_100,
                         bd.params = speciation_1,
@@
 set.seed(11)
-founding_tree <- dads(stop.rule = stop_time_4,
+founding_tree <- treats(stop.rule = stop_time_4,
                       bd.params = spec_1_ext_02,
@@
 set.seed(1)
-founding_tree <- dads(stop.rule = stop_time_6,
+founding_tree <- treats(stop.rule = stop_time_6,
                       bd.params = speciation_1_extinction_03,

### 2023-04-24 inst/gitbook/_book/dads_manual.tex
Big rename

@@
 }
-\hypertarget{dads-disparity-and-diversity-simulations.}{%
-\chapter{\texorpdfstring{\texttt{dads}: disparity and diversity simulations.}{dads: disparity and diversity simulations.}}\label{dads-disparity-and-diversity-simulations.}}
+\hypertarget{treats-disparity-and-diversity-simulations.}{%
+\chapter{\texorpdfstring{\texttt{treats}: disparity and diversity simulations.}{treats: disparity and diversity simulations.}}\label{treats-disparity-and-diversity-simulations.}}
 
@@
 
-\hypertarget{what-is-dads}{%
-\section{\texorpdfstring{What is \texttt{dads}?}{What is dads?}}\label{what-is-dads}}
+\hypertarget{what-is-treats}{%
+\section{\texorpdfstring{What is \texttt{treats}?}{What is treats?}}\label{what-is-treats}}
 
@@
 
-\texttt{dads} aims to be a highly modular and friendly version of all these packages: it allows to simulate disparity and diversity jointly with a vast array of options that can be easily modified by users.
+\texttt{treats} aims to be a highly modular and friendly version of all these packages: it allows to simulate disparity and diversity jointly with a vast array of options that can be easily modified by users.
 For example, you can easily generate any type of process to generate a trait (BM, OU, something else, etc\ldots) in multiple dependent or independent dimensions through \texttt{"traits"} objects.
@@
 
-Because their is an infinite way you might want to generate disparity and diversity (different traits, different modifiers and different events), the \texttt{dads} package is designed to make all these parts easy to code separately and integrate them easily in the \texttt{dads} core functions.
+Because their is an infinite way you might want to generate disparity and diversity (different traits, different modifiers and different events), the \texttt{treats} package is designed to make all these parts easy to code separately and integrate them easily in the \texttt{treats} core functions.
 This allows you to simulate finely tuned multidimensional data for your specific project!
@@
 
-This manual explains how \texttt{dads} work in as much details as possible and aims to give you the keys to design your own diversity and disparity simulations.
+This manual explains how \texttt{treats} work in as much details as possible and aims to give you the keys to design your own diversity and disparity simulations.
 
@@
 \CommentTok{\#\# Installing the latest version directly from GitHub}
-\NormalTok{devtools}\OperatorTok{::}\KeywordTok{install\_github}\NormalTok{(}\StringTok{"TGuillerme/dads"}\NormalTok{)}
+\NormalTok{devtools}\OperatorTok{::}\KeywordTok{install\_github}\NormalTok{(}\StringTok{"TGuillerme/treats"}\NormalTok{)}
 \end{Highlighting}
@@
 However, parts of this package are still in development and some other parts are probably not covered.
-Thus if you have suggestions or comments on on what has already been developed or will be developed, please send me an email (\href{mailto:guillert@tcd.ie}{\nolinkurl{guillert@tcd.ie}}) or if you are a GitHub user, directly create an issue on the \href{https://github.com/TGuillerme/dads}{GitHub page}.
+Thus if you have suggestions or comments on on what has already been developed or will be developed, please send me an email (\href{mailto:guillert@tcd.ie}{\nolinkurl{guillert@tcd.ie}}) or if you are a GitHub user, directly create an issue on the \href{https://github.com/TGuillerme/treats}{GitHub page}.
 Doing so will not only hopefully help you but also other users since it will help improve this manual!
 
-\hypertarget{how-does-dads-work}{%
-\section{\texorpdfstring{How does \texttt{dads} work?}{How does dads work?}}\label{how-does-dads-work}}
-
-Basically, the \texttt{dads} function intakes your personalised \texttt{traits}, \texttt{modifiers} and \texttt{events} to generate your disparity and diversity.
-You will find more details about how these objects (\texttt{traits}, \texttt{modifiers} and \texttt{events}) work in the rest of the tutorial but here is a graphical representation of how \texttt{dads} work:
+\hypertarget{how-does-treats-work}{%
+\section{\texorpdfstring{How does \texttt{treats} work?}{How does treats work?}}\label{how-does-treats-work}}
+
+Basically, the \texttt{treats} function intakes your personalised \texttt{traits}, \texttt{modifiers} and \texttt{events} to generate your disparity and diversity.
+You will find more details about how these objects (\texttt{traits}, \texttt{modifiers} and \texttt{events}) work in the rest of the tutorial but here is a graphical representation of how \texttt{treats} work:
 
@@
 \centering
-\includegraphics{dads_structure.png}
-\caption{Schematised summary of the \texttt{dads} package architecture}
+\includegraphics{treats_structure.png}
+\caption{Schematised summary of the \texttt{treats} package architecture}
 \end{figure}
@@
 
-The \texttt{dads} algorithm is based on a modular version of the birth-death process algorithm.
+The \texttt{treats} algorithm is based on a modular version of the birth-death process algorithm.
 The birth-death model is a continuous Markov process (i.e.~a continuous random process where the state will change each step are linked through time and change according to some random exponential variables) which is well defined mathematically and \href{https://lukejharmon.github.io/pcm/chapter10_birthdeath/\#section-10.2-the-birth-death-model}{commonly used in biology} but also in \href{https://en.wikipedia.org/wiki/Birth\%E2\%80\%93death_process}{many other fields}.
 
-The algorithm used in \texttt{dads} allows modularity of this process and is based on the following steps (the text in \texttt{courier\ font} is for the name of the process in the algorithm):
+The algorithm used in \texttt{treats} allows modularity of this process and is based on the following steps (the text in \texttt{courier\ font} is for the name of the process in the algorithm):
 
@@
   \textbf{Selecting a lineage} (\texttt{selecting}): this step selects a tip that is currently not extinct.
-In a standard birth-death process this is just done randomly, however in \texttt{dads} this can be modified based on the birth-death parameters, the currently available lineages and potential trait values.
+In a standard birth-death process this is just done randomly, however in \texttt{treats} this can be modified based on the birth-death parameters, the currently available lineages and potential trait values.
 \emph{For example, it is possible to put a higher probability in selecting a lineage that is next to a lineage that recently went extinct and has a positive trait value.}
@@
 In an exact birth death process, this is done by drawing a random value from an exponential distribution with a rate of \[\text{number of living lineages} \times (\text{speciation} + \text{extinction parameters})\].
-In \texttt{dads}, this can be modified based on the birth-death parameters, the currently available lineages and potential trait values.
+In \texttt{treats}, this can be modified based on the birth-death parameters, the currently available lineages and potential trait values.
 \emph{For example, it is possible to increase branch length by the ratio of living/extinct fossils and a random number drawn from the range of current trait values.}
@@
   \textbf{Simulating traits} {[}optional{]}(\texttt{traits}): this step allows to simulate a trait value for the selected lineage from step 1.
-This is typically not part of a standard birth-death process and is handled via the \texttt{traits} option in \texttt{dads} (see the \protect\hyperlink{maketraits}{\texttt{make.traits}} chapter for more details).
+This is typically not part of a standard birth-death process and is handled via the \texttt{traits} option in \texttt{treats} (see the \protect\hyperlink{maketraits}{\texttt{make.traits}} chapter for more details).
   Then go to step 4.
@@
 
-Again, in \texttt{dads}, this process can be modified based on the birth-death parameters, the currently available lineages and potential trait values.
+Again, in \texttt{treats}, this process can be modified based on the birth-death parameters, the currently available lineages and potential trait values.
 \emph{For example, the lineage can only speciate if it's trait value is positive, regardless of whether speciation or extinction has been triggered}.
@@
 
-One of the simplest things to do with the \texttt{dads} package is just to simulate a birth death tree.
-For that you can use the function \texttt{dads} and specify your stopping rule.
+One of the simplest things to do with the \texttt{treats} package is just to simulate a birth death tree.
+For that you can use the function \texttt{treats} and specify your stopping rule.
 The stopping rule simply tells the birth death process to step whenever it reaches one of these three conditions:
@@
 \CommentTok{\#\# Running the birth death simulation}
-\NormalTok{my\_tree \textless{}{-}}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{stop.rule =}\NormalTok{ my\_stop\_rule)}
+\NormalTok{my\_tree \textless{}{-}}\StringTok{ }\KeywordTok{treats}\NormalTok{(}\DataTypeTok{stop.rule =}\NormalTok{ my\_stop\_rule)}
 \end{Highlighting}
@@
 
-\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-6-1.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-6-1.pdf}
 
@@
 People familiar with the \href{https://lukejharmon.github.io/pcm/chapter10_birthdeath/}{birth-death models} might have noticed that we did not specify two important things here: the speciation parameter (sometimes called ``lambda'' or ``birth'') and the extinction parameter (sometimes called ``mu'', ``death'' or ``background extinction'').
-By default \texttt{dads} runs a pure birth model (the speciation is set to 1 and the extinction to 0).
+By default \texttt{treats} runs a pure birth model (the speciation is set to 1 and the extinction to 0).
 However, you can easily change that by specifying your new birth death parameters:
@@
 \CommentTok{\#\# Generating a birth death tree with extinctions:}
-\NormalTok{my\_tree \textless{}{-}}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ my\_params, }\DataTypeTok{stop.rule =}\NormalTok{ my\_stop\_rule)}
+\NormalTok{my\_tree \textless{}{-}}\StringTok{ }\KeywordTok{treats}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ my\_params, }\DataTypeTok{stop.rule =}\NormalTok{ my\_stop\_rule)}
 \CommentTok{\#\# Visualising the new tree}
@@
 
-\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-8-1.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-8-1.pdf}
 
@@
 Chances are that you want to also simulate traits (disparity) along with your diversity (otherwise, we suggest using the \href{https://github.com/tanja819/TreeSim/}{\texttt{TreeSim}} package that provides many more birth death models).
-Simulating traits is not much more complicated in \texttt{dads}: you'll simply need to create a \texttt{"traits"} object using the \texttt{make.traits} function.
+Simulating traits is not much more complicated in \texttt{treats}: you'll simply need to create a \texttt{"traits"} object using the \texttt{make.traits} function.
 These objects can have increasing complexity (see the rest of this tutorial) but we will keep it simple here.
@@
 \begin{verbatim}
-##  ---- dads traits object ---- 
+##  ---- treats traits object ---- 
 ## 1 trait for 1 process (A) with one starting value (0).
@@
 
-\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-10-1.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-10-1.pdf}
 
@@
 This is not a really good name but you'll see more about specifying trait names later on.
-If this is what the process should look like (theoretically) you can then add its \texttt{"traits"} object to our previous \texttt{dads} function to generate the tree and the traits:
+If this is what the process should look like (theoretically) you can then add its \texttt{"traits"} object to our previous \texttt{treats} function to generate the tree and the traits:
 
@@
 \CommentTok{\#\# Simulate disparity and diversity}
-\NormalTok{my\_data \textless{}{-}}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ my\_params,}
+\NormalTok{my\_data \textless{}{-}}\StringTok{ }\KeywordTok{treats}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ my\_params,}
                 \DataTypeTok{stop.rule =}\NormalTok{ my\_stop\_rule,}
@@
 \begin{verbatim}
-##  ---- dads object ---- 
+##  ---- treats object ---- 
 ## Simulated diversity data (x$tree):
@@
 
-\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-12-1.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-12-1.pdf}
 
@@
 \begin{verbatim}
-##  ---- dads modifiers object ---- 
+##  ---- treats modifiers object ---- 
 ## No modifiers applied to the branch length, selection and speciation processes (default).
@@
 
-This will not do much to our simulations compared to the previous trait and tree simulation but we can provide our modifiers object to the \texttt{dads} function:
+This will not do much to our simulations compared to the previous trait and tree simulation but we can provide our modifiers object to the \texttt{treats} function:
 
@@
 \CommentTok{\#\# Simulate disparity and diversity}
-\NormalTok{default\_data \textless{}{-}}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ extinction\_}\DecValTok{02}\NormalTok{,}
+\NormalTok{default\_data \textless{}{-}}\StringTok{ }\KeywordTok{treats}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ extinction\_}\DecValTok{02}\NormalTok{,}
                      \DataTypeTok{stop.rule =}\NormalTok{ living\_}\DecValTok{20}\NormalTok{,}
@@
 \begin{verbatim}
-##  ---- dads object ---- 
+##  ---- treats object ---- 
 ## Birth death process with modifiers:
@@
 For that we can use the \texttt{parent.traits} utility function that is optimised for accessing traits in the birth death process (but you can of course write your own).
-This function intakes the \texttt{trait.values} and \texttt{parent.lineage} arguments, two arguments that you can leave named as they are to facilitate \texttt{dads}' understanding of what you want to assess:
+This function intakes the \texttt{trait.values} and \texttt{parent.lineage} arguments, two arguments that you can leave named as they are to facilitate \texttt{treats}' understanding of what you want to assess:
 
@@
 We can then provide these two functions (the condition \texttt{negative.ancestor} and how to modify the speciation event when this condition is met \texttt{staying.alive}).
-If you are an advances \texttt{dads} user, you can design your own \texttt{speciation} function but if you just want to use a normal \texttt{speciation} function, you can use the default one from \texttt{dads} called\ldots{} \texttt{speciation}.
+If you are an advances \texttt{treats} user, you can design your own \texttt{speciation} function but if you just want to use a normal \texttt{speciation} function, you can use the default one from \texttt{treats} called\ldots{} \texttt{speciation}.
 
@@
 \begin{verbatim}
-##  ---- dads modifiers object ---- 
+##  ---- treats modifiers object ---- 
 ## Default branch length process.
@@
 
-Note that the \texttt{make.modifiers} function tests whether the input is compatible with \texttt{dads} by default so unless you have an error message, your \texttt{modifiers} will work!
+Note that the \texttt{make.modifiers} function tests whether the input is compatible with \texttt{treats} by default so unless you have an error message, your \texttt{modifiers} will work!
 We can now simulate our tree and traits with our modifier: species will go extinct if their ancestor have a negative trait value:
@@
 \CommentTok{\#\# Simulate disparity and diversity}
-\NormalTok{biased\_data \textless{}{-}}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ extinction\_}\DecValTok{02}\NormalTok{,}
+\NormalTok{biased\_data \textless{}{-}}\StringTok{ }\KeywordTok{treats}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ extinction\_}\DecValTok{02}\NormalTok{,}
                     \DataTypeTok{stop.rule =}\NormalTok{ living\_}\DecValTok{20}\NormalTok{,}
@@
 
-\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-20-1.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-20-1.pdf}
 
@@
 
-In \texttt{dads}, traits are simulating by providing a \texttt{traits} object to the \texttt{traits} argument.
+In \texttt{treats}, traits are simulating by providing a \texttt{traits} object to the \texttt{traits} argument.
 This object is generated using the \texttt{make.traits}.
@@
 Basically anything can be a trait as well as it's defined by the user (e.g.~one trait can be a 20 dimensional Brownian process that's correlated to a 1D OU process - which is another trait).
-For the purpose of the syntax in \emph{this package} though, a \textbf{trait in \texttt{dads}} is a column in the output trait matrix.
+For the purpose of the syntax in \emph{this package} though, a \textbf{trait in \texttt{treats}} is a column in the output trait matrix.
 For example, one might be interested in a biological trait that is ``beak shape'' which is defined as one complex biological trait that is the Procrustes superimposition of 4 3D landmarks and 300 3D semi-landmarks.
-This single ``beak shape'' \textbf{biological trait} is then composed of 912 \textbf{traits in \texttt{dads}} (i.e.~a matrix of \((4+300)\times3\) columns).
+This single ``beak shape'' \textbf{biological trait} is then composed of 912 \textbf{traits in \texttt{treats}} (i.e.~a matrix of \((4+300)\times3\) columns).
 \end{quote}
@@
 
-Trait processes in \texttt{dads} are functions that must always intake the following arguments by default.
+Trait processes in \texttt{treats} are functions that must always intake the following arguments by default.
 
@@
 Note that in this function definition the arguments \texttt{x0} and \texttt{edge.length} have a default value set to \texttt{0} and \texttt{1} respectively.
-In practice, these arguments are effectively set to the correct values in the \texttt{dads} internal function (i.e.~whatever \texttt{x0} and \texttt{edge.length} are at that specific time of the process) but providing a default can help speed up the algorithms (specifically all the internal checks).
+In practice, these arguments are effectively set to the correct values in the \texttt{treats} internal function (i.e.~whatever \texttt{x0} and \texttt{edge.length} are at that specific time of the process) but providing a default can help speed up the algorithms (specifically all the internal checks).
 \end{quote}
@@
 
-\hypertarget{using-a-process-in-dads}{%
-\subsection{\texorpdfstring{Using a \texttt{"process"} in \texttt{dads}}{Using a "process" in dads}}\label{using-a-process-in-dads}}
+\hypertarget{using-a-process-in-treats}{%
+\subsection{\texorpdfstring{Using a \texttt{"process"} in \texttt{treats}}{Using a "process" in treats}}\label{using-a-process-in-treats}}
 
 You can design your own process as a function (as long as it has a valid syntax).
-Alternatively, the \texttt{dads} package proposes inbuilt processes, namely a multidimensional Brownian motion (\texttt{BM.process}) or a a multidimensional Ornstein-Uhlenbeck process (\texttt{OU.process}).
+Alternatively, the \texttt{treats} package proposes inbuilt processes, namely a multidimensional Brownian motion (\texttt{BM.process}) or a a multidimensional Ornstein-Uhlenbeck process (\texttt{OU.process}).
 You can find the list of implemented process by looking at the \texttt{?trait.process} manual page in \texttt{R}.
@@
 
-This creates \texttt{"dads"} \texttt{"traits"} objects that you can print, and visualise using the \texttt{plot} function:
+This creates \texttt{"treats"} \texttt{"traits"} objects that you can print, and visualise using the \texttt{plot} function:
 
@@
 \begin{verbatim}
-## [1] "dads"   "traits"
+## [1] "treats"   "traits"
 \end{verbatim}
@@
 
-\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-25-1.pdf}
-
-Note that you can see the multiple options for plotting the trait process by looking at \texttt{?plot.dads} manual.
-Furthermore, you can look at what's actually in the object using this specific syntax (this applies to every object handled by the \texttt{dads} package):
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-25-1.pdf}
+
+Note that you can see the multiple options for plotting the trait process by looking at \texttt{?plot.treats} manual.
+Furthermore, you can look at what's actually in the object using this specific syntax (this applies to every object handled by the \texttt{treats} package):
 
@@
 \CommentTok{\#\# What\textquotesingle{}s actually in that object?}
-\KeywordTok{print.dads}\NormalTok{(my\_trait\_object, }\DataTypeTok{all =} \OtherTok{TRUE}\NormalTok{)}
+\KeywordTok{print.treats}\NormalTok{(my\_trait\_object, }\DataTypeTok{all =} \OtherTok{TRUE}\NormalTok{)}
 \end{Highlighting}
@@
 ## <bytecode: 0x55e278dc9cd8>
-## <environment: namespace:dads>
+## <environment: namespace:treats>
 ## 
@@
 \begin{verbatim}
-##  ---- dads traits object ---- 
+##  ---- treats traits object ---- 
 ## 4 traits for 4 processes (A, B, C, D) with one starting value (0).
@@
 
-You can visualise them individually using the \texttt{trait} argument in \texttt{plot.dads}:
+You can visualise them individually using the \texttt{trait} argument in \texttt{plot.treats}:
 
@@
 
-\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-28-1.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-28-1.pdf}
 
@@
 \begin{verbatim}
-##  ---- dads traits object ---- 
+##  ---- treats traits object ---- 
 ## 3 traits for 1 process (A:3) with one starting value (1).
@@
 \begin{verbatim}
-##  ---- dads traits object ---- 
+##  ---- treats traits object ---- 
 ## 3 traits for 1 process (A:3) with different starting values (1,2,3).
@@
 \begin{verbatim}
-##  ---- dads traits object ---- 
+##  ---- treats traits object ---- 
 ## 6 traits for 2 processes (A:3, B:3) with one starting value (0).
@@
 \begin{verbatim}
-##  ---- dads traits object ---- 
+##  ---- treats traits object ---- 
 ## 5 traits for 2 processes (A:1, B:4) with one starting value (0).
@@
 \begin{verbatim}
-##  ---- dads traits object ---- 
+##  ---- treats traits object ---- 
 ## 6 traits for 2 processes (A:3, B:3) with one starting value (1).
@@
 \begin{verbatim}
-##  ---- dads traits object ---- 
+##  ---- treats traits object ---- 
 ## 6 traits for 2 processes (A:3, B:3) with different starting values (1,2,3,4,5,6).
@@
 \begin{verbatim}
-##  ---- dads traits object ---- 
+##  ---- treats traits object ---- 
 ## 6 traits for 2 processes (A:3, B:3) with different starting values (1,1,0,0,0,0).
@@
 The definition of what a trait is can vary quiet a lot depending on the context of an analysis and the field.
-Because it would be impossible to accommodate all definitions in \texttt{dads} we had to go with an arbitrary one: a trait is whatever you define as a trait!
+Because it would be impossible to accommodate all definitions in \texttt{treats} we had to go with an arbitrary one: a trait is whatever you define as a trait!
 A trait can be uni-dimensional as the measurement of a feature of an organism (e.g.~leaf surface, femur length, etc\ldots) but can also be multi-dimensional description of a feature, for example in 3D geometric morphometric, a trait could be defined as ``position of landmark X'' (which will be a trait with three dimensions, \emph{x}, \emph{y} and \emph{z}) or in ecology, the location of a plant can be a single 2D trait (expressed as latitude and longitude).
-In \texttt{dads}, the \textbf{process} corresponds to this trait definition (e.g.~a process can be of n-dimensions and represents one organisms feature) and the \textbf{traits} represents the number of dimensions in total.
-So in the examples above, this is how the following traits are interpreted by \texttt{dads}:
+In \texttt{treats}, the \textbf{process} corresponds to this trait definition (e.g.~a process can be of n-dimensions and represents one organisms feature) and the \textbf{traits} represents the number of dimensions in total.
+So in the examples above, this is how the following traits are interpreted by \texttt{treats}:
 
@@
 \begin{verbatim}
-##  ---- dads traits object ---- 
+##  ---- treats traits object ---- 
 ## 2 traits for 2 processes (A, B) with one starting value (0).
@@
 \begin{verbatim}
-##  ---- dads traits object ---- 
+##  ---- treats traits object ---- 
 ## 3 traits for 3 processes (A, B, C) with one starting value (0).
@@
 \begin{verbatim}
-##  ---- dads traits object ---- 
+##  ---- treats traits object ---- 
 ## 1 trait for 1 process (1D Brownian Motion) with one starting value (0).
@@
 \begin{verbatim}
-##  ---- dads traits object ---- 
+##  ---- treats traits object ---- 
 ## 3 traits for 3 processes (Normal, LogNormal, Uniform(1,100)) with one starting value (0).
@@
 You can also add traits to already existing trait objects using the simple \texttt{add} option.
-This option just intakes a \texttt{"dads"} \texttt{"traits"} object and the additional process(es) will be added to it. For example:
+This option just intakes a \texttt{"treats"} \texttt{"traits"} object and the additional process(es) will be added to it. For example:
 
@@
 \begin{verbatim}
-##  ---- dads traits object ---- 
+##  ---- treats traits object ---- 
 ## 1 trait for 1 process (BM) with one starting value (0).
@@
 \begin{verbatim}
-##  ---- dads traits object ---- 
+##  ---- treats traits object ---- 
 ## 4 traits for 2 processes (BM:1, 3D OU:3) with one starting value (0).
@@
 \CommentTok{\#\# Generating a pure birth tree with the background trait}
-\NormalTok{tree\_bkg \textless{}{-}}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{stop.rule =} \KeywordTok{list}\NormalTok{(}\DataTypeTok{max.taxa =} \DecValTok{20}\NormalTok{),}
+\NormalTok{tree\_bkg \textless{}{-}}\StringTok{ }\KeywordTok{treats}\NormalTok{(}\DataTypeTok{stop.rule =} \KeywordTok{list}\NormalTok{(}\DataTypeTok{max.taxa =} \DecValTok{20}\NormalTok{),}
                  \DataTypeTok{traits    =}\NormalTok{ my\_trait)}
@@
 
-\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-41-1.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-41-1.pdf}
 
@@
 
-Alternatively to having a background trait, you can also simulate a tree by generating traits at specific time steps with the \texttt{save.steps} option in \texttt{dads}.
+Alternatively to having a background trait, you can also simulate a tree by generating traits at specific time steps with the \texttt{save.steps} option in \texttt{treats}.
 This will apply the \texttt{traits} object to all lineages currently alive at the required time steps.
@@
 \CommentTok{\#\# Generating a birth{-}death tree with a BM trait and saving steps at specific times}
-\NormalTok{tree\_steps \textless{}{-}}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{stop.rule  =} \KeywordTok{list}\NormalTok{(}\DataTypeTok{max.time =} \DecValTok{3}\NormalTok{),}
+\NormalTok{tree\_steps \textless{}{-}}\StringTok{ }\KeywordTok{treats}\NormalTok{(}\DataTypeTok{stop.rule  =} \KeywordTok{list}\NormalTok{(}\DataTypeTok{max.time =} \DecValTok{3}\NormalTok{),}
                    \DataTypeTok{bd.params  =} \KeywordTok{list}\NormalTok{(}\DataTypeTok{speciation =} \DecValTok{1}\NormalTok{, }\DataTypeTok{extinction =} \FloatTok{0.1}\NormalTok{),}
@@
 
-\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-42-1.pdf}
-
-\hypertarget{traits-implemented-in-dads}{%
-\section{\texorpdfstring{Traits implemented in \texttt{dads}}{Traits implemented in dads}}\label{traits-implemented-in-dads}}
-
-If you don't want to design your own trait process, you can use one of the following trait processes that are currently implemented in \texttt{dads}.
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-42-1.pdf}
+
+\hypertarget{traits-implemented-in-treats}{%
+\section{\texorpdfstring{Traits implemented in \texttt{treats}}{Traits implemented in treats}}\label{traits-implemented-in-treats}}
+
+If you don't want to design your own trait process, you can use one of the following trait processes that are currently implemented in \texttt{treats}.
 You can find more information about their many options using their specific manuals in R or the generic \texttt{?trait.process}:
@@
 
-\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-43-1.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-43-1.pdf}
 
@@
 
-\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-44-1.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-44-1.pdf}
 
@@
 
-\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-45-1.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-45-1.pdf}
 
@@
 
-\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-46-1.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-46-1.pdf}
 
@@
 
-\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-47-1.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-47-1.pdf}
 
@@
 We highly suggest leaving \texttt{test\ =\ TRUE} so that \texttt{make.traits} returns an error if a process or its additional arguments (\texttt{process.args}) are not formatted correctly.
-\texttt{make.traits} will send an error if the trait cannot be directly passed to \texttt{dads}.
+\texttt{make.traits} will send an error if the trait cannot be directly passed to \texttt{treats}.
 However, in some specific cases (again, probably mainly for development and debugging) it could be useful to skip the tests using \texttt{test\ =\ FALSE}.
@@
 You can be imaginative and creative when designing your own process but here are two detailed example functions for a unidimensional Brownian Motion and Ornstein-Uhlenbeck process that you can use for a start (or not).
-Remember it is good practice for \texttt{dads} processes to set all the arguments with default values (just in case).
+Remember it is good practice for \texttt{treats} processes to set all the arguments with default values (just in case).
 
@@
 
-Similarly to \texttt{"traits"} objects, \texttt{"modifiers"} are also printed by default using \texttt{print.dads}.
-You can see details about what's actually in the object using \texttt{print.dads(my\_default\_modifiers,\ all\ =\ TRUE)}.
+Similarly to \texttt{"traits"} objects, \texttt{"modifiers"} are also printed by default using \texttt{print.treats}.
+You can see details about what's actually in the object using \texttt{print.treats(my\_default\_modifiers,\ all\ =\ TRUE)}.
 However, contrary to \texttt{"traits"}, you cannot plot \texttt{"modifiers"}.
@@
 
-The modifiers modify the core of the birth death process as implemented in \texttt{dads}.
-
-By default, the birth-death process in \texttt{dads} uses this modifier:
+The modifiers modify the core of the birth death process as implemented in \texttt{treats}.
+
+By default, the birth-death process in \texttt{treats} uses this modifier:
 
@@
 ## <bytecode: 0x55e27b08d1b8>
-## <environment: namespace:dads>
+## <environment: namespace:treats>
 ## 
@@
 ## <bytecode: 0x55e27b11eb90>
-## <environment: namespace:dads>
+## <environment: namespace:treats>
 ## 
@@
 ## <bytecode: 0x55e27b127560>
-## <environment: namespace:dads>
+## <environment: namespace:treats>
 ## 
@@
 
-This contains a lot of information, much of it is actually not useful for using the modular aspects of \texttt{dads} at a high user level.
+This contains a lot of information, much of it is actually not useful for using the modular aspects of \texttt{treats} at a high user level.
 I.e.
@@
 
-The first argument in \texttt{"modifiers"} is the branch length function (\texttt{branch.length}) this is the function that will be executed in \texttt{dads} to generate branch length.
-Note that in the \texttt{dads} algorithm, branch length is not generated \emph{directly} but actually results of the waiting time.
+The first argument in \texttt{"modifiers"} is the branch length function (\texttt{branch.length}) this is the function that will be executed in \texttt{treats} to generate branch length.
+Note that in the \texttt{treats} algorithm, branch length is not generated \emph{directly} but actually results of the waiting time.
 In other words, the \texttt{branch.length} function just affects waiting time for all taxa present at any time in the simulation.
@@
 By default, branch length (or waiting/growth) is a randomly drawn number from an exponential distribution with the rate of the number of taxa multiplied by the speciation and extinction rate: \(n \times (\lambda + \mu)\) (where \(n\) is the number of taxa currently present in the simulation, \(\lambda\) and \(\mu\) are respectively the speciation and extinction rates).
-This default function is simply called \texttt{branch.length} in \texttt{dads} and can be used as a modifier as follows:
+This default function is simply called \texttt{branch.length} in \texttt{treats} and can be used as a modifier as follows:
 
@@
 \KeywordTok{set.seed}\NormalTok{(}\DecValTok{0}\NormalTok{)}
-\NormalTok{default\_tree \textless{}{-}}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ bd\_params,}
+\NormalTok{default\_tree \textless{}{-}}\StringTok{ }\KeywordTok{treats}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ bd\_params,}
                      \DataTypeTok{stop.rule =}\NormalTok{ stop\_rule,}
@@
 \KeywordTok{set.seed}\NormalTok{(}\DecValTok{0}\NormalTok{)}
-\NormalTok{modified\_tree \textless{}{-}}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ bd\_params,}
+\NormalTok{modified\_tree \textless{}{-}}\StringTok{ }\KeywordTok{treats}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ bd\_params,}
                       \DataTypeTok{stop.rule =}\NormalTok{ stop\_rule,}
@@
 
-\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-58-1.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-58-1.pdf}
 
@@
 
-In general, unless you know what you're doing, you can ignore most arguments for specific modifiers since they are handled automatically within the \texttt{dads} function.
+In general, unless you know what you're doing, you can ignore most arguments for specific modifiers since they are handled automatically within the \texttt{treats} function.
 Therefore any argument can be left undeclared or missing and is always handled internally.
@@
 \KeywordTok{set.seed}\NormalTok{(}\DecValTok{0}\NormalTok{)}
-\NormalTok{increasing\_tree \textless{}{-}}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ bd\_params,}
+\NormalTok{increasing\_tree \textless{}{-}}\StringTok{ }\KeywordTok{treats}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ bd\_params,}
                         \DataTypeTok{stop.rule =}\NormalTok{ stop\_rule,}
@@
 
-\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-61-1.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-61-1.pdf}
 
@@
 \KeywordTok{set.seed}\NormalTok{(}\DecValTok{0}\NormalTok{)}
-\NormalTok{ladderised\_tree \textless{}{-}}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ bd\_params,}
+\NormalTok{ladderised\_tree \textless{}{-}}\StringTok{ }\KeywordTok{treats}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ bd\_params,}
                         \DataTypeTok{stop.rule =}\NormalTok{ stop\_rule,}
@@
 
-\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-63-1.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-63-1.pdf}
 
 Again, it is of course possible to make the modifier more complex and in combination with other elements of the tree.
-For example, we can create a \texttt{"dads"} object that also generates a BM trait and add to it this object a \texttt{selection} modifier that only selects tips with positive trait values (only species with positive trait values will speciate).
+For example, we can create a \texttt{"treats"} object that also generates a BM trait and add to it this object a \texttt{selection} modifier that only selects tips with positive trait values (only species with positive trait values will speciate).
 
@@
 \KeywordTok{set.seed}\NormalTok{(}\DecValTok{1}\NormalTok{)}
-\NormalTok{default\_dads \textless{}{-}}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ bd\_params,}
+\NormalTok{default\_treats \textless{}{-}}\StringTok{ }\KeywordTok{treats}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ bd\_params,}
                      \DataTypeTok{stop.rule =}\NormalTok{ stop\_rule,}
@@
 \KeywordTok{set.seed}\NormalTok{(}\DecValTok{1}\NormalTok{)}
-\NormalTok{skewed\_trait\_dads \textless{}{-}}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ bd\_params,}
+\NormalTok{skewed\_trait\_treats \textless{}{-}}\StringTok{ }\KeywordTok{treats}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ bd\_params,}
                           \DataTypeTok{stop.rule =}\NormalTok{ stop\_rule,}
@@
 \KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow =} \KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{, }\DecValTok{2}\NormalTok{))}
-\KeywordTok{plot}\NormalTok{(default\_dads, }\DataTypeTok{main =} \StringTok{"Default trait and tree"}\NormalTok{)}
-\KeywordTok{plot}\NormalTok{(skewed\_trait\_dads, }\DataTypeTok{main =} \StringTok{"Skewed trait and tree"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-64-1.pdf}
+\KeywordTok{plot}\NormalTok{(default\_treats, }\DataTypeTok{main =} \StringTok{"Default trait and tree"}\NormalTok{)}
+\KeywordTok{plot}\NormalTok{(skewed\_trait\_treats, }\DataTypeTok{main =} \StringTok{"Skewed trait and tree"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-64-1.pdf}
 
@@
 Creating a \texttt{"modifiers"} with a \texttt{speciation} function works the same way as for \texttt{branch.length} and \texttt{selection} but the function that will be used needs to output a logical value (see table \protect\hyperlink{summarymodifiers}{below}).
-Once the function is created simply input your function for speciation in the modifier and run the \texttt{dads} function with that modifier:
+Once the function is created simply input your function for speciation in the modifier and run the \texttt{treats} function with that modifier:
 
@@
 \KeywordTok{set.seed}\NormalTok{(}\DecValTok{2}\NormalTok{)}
-\NormalTok{modified\_tree \textless{}{-}}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ bd\_params,}
+\NormalTok{modified\_tree \textless{}{-}}\StringTok{ }\KeywordTok{treats}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ bd\_params,}
                       \DataTypeTok{stop.rule =}\NormalTok{ stop\_rule,}
@@
 \KeywordTok{set.seed}\NormalTok{(}\DecValTok{1}\NormalTok{)}
-\NormalTok{biased\_trait\_dads \textless{}{-}}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ bd\_params,}
+\NormalTok{biased\_trait\_treats \textless{}{-}}\StringTok{ }\KeywordTok{treats}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ bd\_params,}
                           \DataTypeTok{stop.rule =}\NormalTok{ stop\_rule,}
@@
 \KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow =} \KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{, }\DecValTok{2}\NormalTok{))}
-\KeywordTok{plot}\NormalTok{(default\_dads, }\DataTypeTok{main =} \StringTok{"Default trait and tree"}\NormalTok{)}
-\KeywordTok{plot}\NormalTok{(biased\_trait\_dads, }\DataTypeTok{main =} \StringTok{"Biased trait and tree"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-67-1.pdf}
+\KeywordTok{plot}\NormalTok{(default\_treats, }\DataTypeTok{main =} \StringTok{"Default trait and tree"}\NormalTok{)}
+\KeywordTok{plot}\NormalTok{(biased\_trait\_treats, }\DataTypeTok{main =} \StringTok{"Biased trait and tree"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-67-1.pdf}
 
@@
 
-One other major feature of the \texttt{dads} package is to allow simulations to run with specified events to occur during the simulation.
+One other major feature of the \texttt{treats} package is to allow simulations to run with specified events to occur during the simulation.
 These are typically events that can drastically change the course of the simulation.
@@
 However, these can also be more subtle, like the internal change of parameter values when reaching a specific trait value.
-These are all handled by \texttt{dads} with the \texttt{events} object that you can create using \texttt{make.events} and has the same overall logic as \texttt{make.traits} and \texttt{make.modifiers}.
+These are all handled by \texttt{treats} with the \texttt{events} object that you can create using \texttt{make.events} and has the same overall logic as \texttt{make.traits} and \texttt{make.modifiers}.
 
@@
 This is typically done by updating the object using the argument \texttt{update} from \texttt{make.modifiers}.
-* \texttt{"founding"} this target is a bit more special and allows you run a nested \texttt{dads} object in the simulation. It is cover in a \protect\hyperlink{founding}{specific section below}.
+* \texttt{"founding"} this target is a bit more special and allows you run a nested \texttt{treats} object in the simulation. It is cover in a \protect\hyperlink{founding}{specific section below}.
 
@@
 
-Currently there are three conditions functions implemented in \texttt{dads} but you can easily come up with your own version of them.
-
-All \texttt{condition} functions in \texttt{dads} take at least to arguments: \texttt{x} for the variable of interest (e.g.~time, number of taxa, trait value, etc\ldots) and \texttt{condition}, the relational operator to evaluate.
+Currently there are three conditions functions implemented in \texttt{treats} but you can easily come up with your own version of them.
+
+All \texttt{condition} functions in \texttt{treats} take at least to arguments: \texttt{x} for the variable of interest (e.g.~time, number of taxa, trait value, etc\ldots) and \texttt{condition}, the relational operator to evaluate.
 A relational operator is the proper (fancy) term designating all the comparisons you're regularly using in \texttt{R} like \texttt{==} (is equal?) \texttt{\textless{}} (is smaller?) \texttt{\textgreater{}=} (is bigger or equal?).
@@
 \begin{verbatim}
-##  ---- dads traits object ---- 
+##  ---- treats traits object ---- 
 ## 2 traits for 1 process (A:2) with one starting value (0).
@@
 \KeywordTok{set.seed}\NormalTok{(}\DecValTok{123}\NormalTok{)}
-\NormalTok{results \textless{}{-}}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ bd.params,}
+\NormalTok{results \textless{}{-}}\StringTok{ }\KeywordTok{treats}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ bd.params,}
                 \DataTypeTok{stop.rule =}\NormalTok{ stop.rule,}
@@
 
-\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-75-1.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-75-1.pdf}
 
@@
 \KeywordTok{set.seed}\NormalTok{(}\DecValTok{1}\NormalTok{)}
-\NormalTok{no\_change \textless{}{-}}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ bd.params,}
+\NormalTok{no\_change \textless{}{-}}\StringTok{ }\KeywordTok{treats}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ bd.params,}
                   \DataTypeTok{stop.rule =}\NormalTok{ stop.rule,}
@@
 \KeywordTok{set.seed}\NormalTok{(}\DecValTok{1}\NormalTok{)}
-\NormalTok{process\_change \textless{}{-}}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ bd.params,}
+\NormalTok{process\_change \textless{}{-}}\StringTok{ }\KeywordTok{treats}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ bd.params,}
                        \DataTypeTok{stop.rule =}\NormalTok{ stop.rule,}
@@
 
-\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-76-1.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-76-1.pdf}
 
@@
 Founding events are a specific \texttt{target} for \texttt{events} that allows you to\ldots{} simulate a birth-death process within the current one!
-It basically allows you to simulate a specific \texttt{dads} process i.e.~using the \texttt{dads} function with it's own \texttt{traits}, \texttt{modifiers}, \texttt{bd.params} and \texttt{events} (and yes, that's including \texttt{events} that have their own \texttt{founding} event)\ldots{}
+It basically allows you to simulate a specific \texttt{treats} process i.e.~using the \texttt{treats} function with it's own \texttt{traits}, \texttt{modifiers}, \texttt{bd.params} and \texttt{events} (and yes, that's including \texttt{events} that have their own \texttt{founding} event)\ldots{}
 This is basically the ultimate nested boss of modularity!
 
-The founding event will basically run an internal \texttt{dads} process (i.e.~simulating a tree and, optionally, some data) resulting in a founding sub-tree.
+The founding event will basically run an internal \texttt{treats} process (i.e.~simulating a tree and, optionally, some data) resulting in a founding sub-tree.
 It will then branch this sub-tree to the rest of the simulation that continued normally in the mean time.
 You can specify the founding event using the inbuilt \texttt{founding.event} modification event.
-This function intakes the exact same arguments as \texttt{dads} to simulate the sub-tree with its own parameters.
+This function intakes the exact same arguments as \texttt{treats} to simulate the sub-tree with its own parameters.
 Additionally, we will use the \texttt{additional.args} argument from \texttt{make.events} to specify a prefix for the founding tree tips (to make them easier to distinguish).
@@
 \KeywordTok{set.seed}\NormalTok{(}\DecValTok{11}\NormalTok{)}
-\NormalTok{founding\_tree \textless{}{-}}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ bd.params,}
+\NormalTok{founding\_tree \textless{}{-}}\StringTok{ }\KeywordTok{treats}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ bd.params,}
                       \DataTypeTok{stop.rule =}\NormalTok{ stop.rule,}
@@
 
-\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-77-1.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-77-1.pdf}
 
@@
 \KeywordTok{set.seed}\NormalTok{(}\DecValTok{123}\NormalTok{)}
-\KeywordTok{plot}\NormalTok{(}\KeywordTok{dads}\NormalTok{(}\DataTypeTok{stop.rule =} \KeywordTok{list}\NormalTok{(}\DataTypeTok{max.taxa =} \DecValTok{50}\NormalTok{), }\DataTypeTok{bd.params =}\NormalTok{ my\_bd\_params))}
-\end{Highlighting}
-\end{Shaded}
-
-\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-79-1.pdf}
+\KeywordTok{plot}\NormalTok{(}\KeywordTok{treats}\NormalTok{(}\DataTypeTok{stop.rule =} \KeywordTok{list}\NormalTok{(}\DataTypeTok{max.taxa =} \DecValTok{50}\NormalTok{), }\DataTypeTok{bd.params =}\NormalTok{ my\_bd\_params))}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-79-1.pdf}
 
@@
 \begin{verbatim}
-##  ---- dads birth-death parameters object ---- 
+##  ---- treats birth-death parameters object ---- 
 ## speciation: runif.
@@
 
-In this example, the \texttt{"bd.params"} object passed to the \texttt{dads} function will allow the birth-death process to sample the speciation parameter each time it is called (e.g.~during the speciation/extinction step, the branch length step, etc.).
+In this example, the \texttt{"bd.params"} object passed to the \texttt{treats} function will allow the birth-death process to sample the speciation parameter each time it is called (e.g.~during the speciation/extinction step, the branch length step, etc.).
 If using a function, you can fine tune the arguments to be passed to that function using the \texttt{speciation.args} or the \texttt{extinction.args} arguments (as a named list matching the function's arguments):
@@
 \begin{verbatim}
-##  ---- dads birth-death parameters object ---- 
+##  ---- treats birth-death parameters object ---- 
 ## speciation: runif (with optional arguments).
@@
 \begin{verbatim}
-##  ---- dads birth-death parameters object ---- 
+##  ---- treats birth-death parameters object ---- 
 ## joint sampling for:
@@
 \begin{verbatim}
-##  ---- dads birth-death parameters object ---- 
+##  ---- treats birth-death parameters object ---- 
 ## speciation: rnorm.
@@
 
-\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-84-1.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-84-1.pdf}
 
@@
 
-\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-84-2.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-84-2.pdf}
 
@@
 \KeywordTok{set.seed}\NormalTok{(}\DecValTok{3}\NormalTok{)}
-\NormalTok{my\_data \textless{}{-}}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{stop.rule =} \KeywordTok{list}\NormalTok{(}\DataTypeTok{max.taxa =} \DecValTok{20}\NormalTok{),}
+\NormalTok{my\_data \textless{}{-}}\StringTok{ }\KeywordTok{treats}\NormalTok{(}\DataTypeTok{stop.rule =} \KeywordTok{list}\NormalTok{(}\DataTypeTok{max.taxa =} \DecValTok{20}\NormalTok{),}
                 \DataTypeTok{bd.params =} \KeywordTok{list}\NormalTok{(}\DataTypeTok{speciation =} \DecValTok{1}\NormalTok{, }\DataTypeTok{extinction =} \DecValTok{1}\OperatorTok{/}\DecValTok{3}\NormalTok{),}
@@
 
-\hypertarget{dads-internal-utilities}{%
-\section{\texorpdfstring{\texttt{"dads"} internal utilities}{"dads" internal utilities}}\label{dads-internal-utilities}}
+\hypertarget{treats-internal-utilities}{%
+\section{\texorpdfstring{\texttt{"treats"} internal utilities}{"treats" internal utilities}}\label{treats-internal-utilities}}
 
 The package also provides utilities for internal functions, namely for designing \texttt{modifiers} or \texttt{events} more easily.
-These functions don't do anything useful on their own but are optimised to be used internally in \texttt{dads}.
+These functions don't do anything useful on their own but are optimised to be used internally in \texttt{treats}.
 For all these functions, you can look at the internal manual for an example (i.e.~using \texttt{?\textless{}function\_name\textgreater{}})
@@
 
-\texttt{"dads"} objects can be directly plotted in \texttt{dads} using the S3 \texttt{plot.dads} function (or just \texttt{plot(x)} if \texttt{x} is of class \texttt{"dads"}).
+\texttt{"treats"} objects can be directly plotted in \texttt{treats} using the S3 \texttt{plot.treats} function (or just \texttt{plot(x)} if \texttt{x} is of class \texttt{"treats"}).
 
@@
 
-\texttt{"dads"} \texttt{"traits"} objects are covered in the \protect\hyperlink{maketraits}{traits section}.
-You can use \texttt{plot.dads} to plot them by choosing which specific trait to plot using the \texttt{trait} argument (default is 1):
+\texttt{"treats"} \texttt{"traits"} objects are covered in the \protect\hyperlink{maketraits}{traits section}.
+You can use \texttt{plot.treats} to plot them by choosing which specific trait to plot using the \texttt{trait} argument (default is 1):
 
@@
 
-\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-87-1.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-87-1.pdf}
 
@@
 
-\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-88-1.pdf}
-
-\hypertarget{plotting-dads-results}{%
-\chapter{Plotting dads results}\label{plotting-dads-results}}
-
-If \texttt{dads} is used to plot only a tree (and outputs a \texttt{"phylo"} object), you can use the function \texttt{plot.phylo} from the \texttt{ape} package to plot your tree.
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-88-1.pdf}
+
+\hypertarget{plotting-treats-results}{%
+\chapter{Plotting treats results}\label{plotting-treats-results}}
+
+If \texttt{treats} is used to plot only a tree (and outputs a \texttt{"phylo"} object), you can use the function \texttt{plot.phylo} from the \texttt{ape} package to plot your tree.
 You'll find all the tree plotting option in the \texttt{?plot.phylo} manual page.
@@
 \CommentTok{\#\# A simple pure birth tree}
-\NormalTok{my\_tree \textless{}{-}}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{stop.rule =} \KeywordTok{list}\NormalTok{(}\DataTypeTok{max.taxa =} \DecValTok{20}\NormalTok{))}
+\NormalTok{my\_tree \textless{}{-}}\StringTok{ }\KeywordTok{treats}\NormalTok{(}\DataTypeTok{stop.rule =} \KeywordTok{list}\NormalTok{(}\DataTypeTok{max.taxa =} \DecValTok{20}\NormalTok{))}
 \KeywordTok{plot}\NormalTok{(my\_tree, }\DataTypeTok{main =} \StringTok{"Plotting a }\CharTok{\textbackslash{}"}\StringTok{phylo}\CharTok{\textbackslash{}"}\StringTok{ object"}\NormalTok{)}
@@
 
-\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-89-1.pdf}
-
-However, if you also simulated a trait along with the tree you can use the \texttt{plot.dads} function to plot both the tree and the trait:
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-89-1.pdf}
+
+However, if you also simulated a trait along with the tree you can use the \texttt{plot.treats} function to plot both the tree and the trait:
 
@@
 \CommentTok{\#\# A simple pure birth tree with a BM process}
-\NormalTok{my\_tree \textless{}{-}}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{stop.rule =} \KeywordTok{list}\NormalTok{(}\DataTypeTok{max.taxa =} \DecValTok{20}\NormalTok{), }\DataTypeTok{traits =} \KeywordTok{make.traits}\NormalTok{())}
+\NormalTok{my\_tree \textless{}{-}}\StringTok{ }\KeywordTok{treats}\NormalTok{(}\DataTypeTok{stop.rule =} \KeywordTok{list}\NormalTok{(}\DataTypeTok{max.taxa =} \DecValTok{20}\NormalTok{), }\DataTypeTok{traits =} \KeywordTok{make.traits}\NormalTok{())}
 \CommentTok{\#\# Playing with the default options}
@@
 
-\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-90-1.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-90-1.pdf}
 
@@
 
-\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-91-1.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-91-1.pdf}
 
@@
 
-\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-92-1.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-92-1.pdf}
 
@@
 \CommentTok{\#\# Simulating a birth death tree with that 3D trait}
-\NormalTok{my\_data \textless{}{-}}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =} \KeywordTok{list}\NormalTok{(}\DataTypeTok{extinction =} \FloatTok{0.2}\NormalTok{),}
+\NormalTok{my\_data \textless{}{-}}\StringTok{ }\KeywordTok{treats}\NormalTok{(}\DataTypeTok{bd.params =} \KeywordTok{list}\NormalTok{(}\DataTypeTok{extinction =} \FloatTok{0.2}\NormalTok{),}
                 \DataTypeTok{stop.rule =} \KeywordTok{list}\NormalTok{(}\DataTypeTok{max.living =} \DecValTok{50}\NormalTok{),}
@@
 
-\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-94-1.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-94-1.pdf}
 
@@
 
-\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-95-1.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-95-1.pdf}
 
 But there's more! You can also plot these results using actual 3D plots that you can spin and all that!
-This is done through the \texttt{rgl} package and activated in \texttt{dads} using the option \texttt{use.3D\ =\ TRUE}.
+This is done through the \texttt{rgl} package and activated in \texttt{treats} using the option \texttt{use.3D\ =\ TRUE}.
 
@@
 \KeywordTok{set.seed}\NormalTok{(}\DecValTok{1}\NormalTok{)}
-\NormalTok{results \textless{}{-}}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{stop.rule =}\NormalTok{ stop\_time\_}\DecValTok{6}\NormalTok{,}
+\NormalTok{results \textless{}{-}}\StringTok{ }\KeywordTok{treats}\NormalTok{(}\DataTypeTok{stop.rule =}\NormalTok{ stop\_time\_}\DecValTok{6}\NormalTok{,}
                 \DataTypeTok{bd.params =}\NormalTok{ speciation\_}\DecValTok{1}\NormalTok{,}
@@
 
-\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-101-1.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-101-1.pdf}
 
@@
 \KeywordTok{set.seed}\NormalTok{(}\DecValTok{7}\NormalTok{)}
-\NormalTok{results \textless{}{-}}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{stop.rule =}\NormalTok{ stop\_time\_}\DecValTok{5}\NormalTok{,}
+\NormalTok{results \textless{}{-}}\StringTok{ }\KeywordTok{treats}\NormalTok{(}\DataTypeTok{stop.rule =}\NormalTok{ stop\_time\_}\DecValTok{5}\NormalTok{,}
                 \DataTypeTok{bd.params =}\NormalTok{ speciation\_}\DecValTok{1}\NormalTok{,}
@@
 
-\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-105-1.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-105-1.pdf}
 
@@
 \KeywordTok{set.seed}\NormalTok{(}\DecValTok{2}\NormalTok{)}
-\NormalTok{results \textless{}{-}}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{stop.rule =}\NormalTok{ stop\_taxa\_}\DecValTok{50}\NormalTok{,}
+\NormalTok{results \textless{}{-}}\StringTok{ }\KeywordTok{treats}\NormalTok{(}\DataTypeTok{stop.rule =}\NormalTok{ stop\_taxa\_}\DecValTok{50}\NormalTok{,}
                 \DataTypeTok{bd.params =}\NormalTok{ speciation\_}\DecValTok{1}\NormalTok{,}
@@
 
-\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-108-1.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-108-1.pdf}
 
@@
 \KeywordTok{set.seed}\NormalTok{(}\DecValTok{42}\NormalTok{)}
-\NormalTok{no\_event \textless{}{-}}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{stop.rule =}\NormalTok{ stop\_time\_}\DecValTok{6}\NormalTok{,}
+\NormalTok{no\_event \textless{}{-}}\StringTok{ }\KeywordTok{treats}\NormalTok{(}\DataTypeTok{stop.rule =}\NormalTok{ stop\_time\_}\DecValTok{6}\NormalTok{,}
                  \DataTypeTok{bd.params =}\NormalTok{ random\_speciation)}
@@
 \KeywordTok{set.seed}\NormalTok{(}\DecValTok{42}\NormalTok{)}
-\NormalTok{reduced\_speciation\_event \textless{}{-}}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{stop.rule =}\NormalTok{ stop\_time\_}\DecValTok{6}\NormalTok{,}
+\NormalTok{reduced\_speciation\_event \textless{}{-}}\StringTok{ }\KeywordTok{treats}\NormalTok{(}\DataTypeTok{stop.rule =}\NormalTok{ stop\_time\_}\DecValTok{6}\NormalTok{,}
                                  \DataTypeTok{bd.params =}\NormalTok{ random\_speciation,}
@@
 
-\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-111-1.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-111-1.pdf}
 
@@
 \KeywordTok{set.seed}\NormalTok{(}\DecValTok{1}\NormalTok{)}
-\NormalTok{no\_change \textless{}{-}}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{stop.rule =}\NormalTok{ stop\_time\_}\DecValTok{6}\NormalTok{,}
+\NormalTok{no\_change \textless{}{-}}\StringTok{ }\KeywordTok{treats}\NormalTok{(}\DataTypeTok{stop.rule =}\NormalTok{ stop\_time\_}\DecValTok{6}\NormalTok{,}
                   \DataTypeTok{bd.params =}\NormalTok{ speciation\_}\DecValTok{1}\NormalTok{,}
@@
 \KeywordTok{set.seed}\NormalTok{(}\DecValTok{1}\NormalTok{)}
-\NormalTok{process\_change \textless{}{-}}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{stop.rule =}\NormalTok{ stop\_time\_}\DecValTok{6}\NormalTok{,}
+\NormalTok{process\_change \textless{}{-}}\StringTok{ }\KeywordTok{treats}\NormalTok{(}\DataTypeTok{stop.rule =}\NormalTok{ stop\_time\_}\DecValTok{6}\NormalTok{,}
                        \DataTypeTok{bd.params =}\NormalTok{ speciation\_}\DecValTok{1}\NormalTok{,}
@@
 
-\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-115-1.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-115-1.pdf}
 
@@
 \KeywordTok{set.seed}\NormalTok{(}\DecValTok{2}\NormalTok{)}
-\NormalTok{no\_event \textless{}{-}}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{stop.rule =}\NormalTok{ stop\_taxa\_}\DecValTok{100}\NormalTok{,}
+\NormalTok{no\_event \textless{}{-}}\StringTok{ }\KeywordTok{treats}\NormalTok{(}\DataTypeTok{stop.rule =}\NormalTok{ stop\_taxa\_}\DecValTok{100}\NormalTok{,}
                  \DataTypeTok{bd.params =}\NormalTok{ speciation\_}\DecValTok{1}\NormalTok{,}
@@
 \KeywordTok{set.seed}\NormalTok{(}\DecValTok{2}\NormalTok{)}
-\NormalTok{change\_correlation \textless{}{-}}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{stop.rule =}\NormalTok{ stop\_taxa\_}\DecValTok{100}\NormalTok{,}
+\NormalTok{change\_correlation \textless{}{-}}\StringTok{ }\KeywordTok{treats}\NormalTok{(}\DataTypeTok{stop.rule =}\NormalTok{ stop\_taxa\_}\DecValTok{100}\NormalTok{,}
                            \DataTypeTok{bd.params =}\NormalTok{ speciation\_}\DecValTok{1}\NormalTok{,}
@@
 
-\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-119-1.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-119-1.pdf}
 
@@
 \KeywordTok{set.seed}\NormalTok{(}\DecValTok{4}\NormalTok{)}
-\NormalTok{no\_event \textless{}{-}}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{stop.rule =}\NormalTok{ stop\_time\_}\DecValTok{4}\NormalTok{,}
+\NormalTok{no\_event \textless{}{-}}\StringTok{ }\KeywordTok{treats}\NormalTok{(}\DataTypeTok{stop.rule =}\NormalTok{ stop\_time\_}\DecValTok{4}\NormalTok{,}
                  \DataTypeTok{bd.params =}\NormalTok{ speciation\_}\DecValTok{1}\NormalTok{,}
@@
 \KeywordTok{set.seed}\NormalTok{(}\DecValTok{4}\NormalTok{)}
-\NormalTok{change\_spec \textless{}{-}}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{stop.rule =}\NormalTok{ stop\_time\_}\DecValTok{4}\NormalTok{,}
+\NormalTok{change\_spec \textless{}{-}}\StringTok{ }\KeywordTok{treats}\NormalTok{(}\DataTypeTok{stop.rule =}\NormalTok{ stop\_time\_}\DecValTok{4}\NormalTok{,}
                     \DataTypeTok{bd.params =}\NormalTok{ speciation\_}\DecValTok{1}\NormalTok{,}
@@
 
-\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-125-1.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-125-1.pdf}
 
@@
 \KeywordTok{set.seed}\NormalTok{(}\DecValTok{5}\NormalTok{)}
-\NormalTok{no\_event \textless{}{-}}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{stop.rule =}\NormalTok{ stop\_taxa\_}\DecValTok{100}\NormalTok{,}
+\NormalTok{no\_event \textless{}{-}}\StringTok{ }\KeywordTok{treats}\NormalTok{(}\DataTypeTok{stop.rule =}\NormalTok{ stop\_taxa\_}\DecValTok{100}\NormalTok{,}
                  \DataTypeTok{bd.params =}\NormalTok{ speciation\_}\DecValTok{1}\NormalTok{,}
@@
 \KeywordTok{set.seed}\NormalTok{(}\DecValTok{5}\NormalTok{)}
-\NormalTok{increased\_brlen \textless{}{-}}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{stop.rule =}\NormalTok{ stop\_taxa\_}\DecValTok{100}\NormalTok{,}
+\NormalTok{increased\_brlen \textless{}{-}}\StringTok{ }\KeywordTok{treats}\NormalTok{(}\DataTypeTok{stop.rule =}\NormalTok{ stop\_taxa\_}\DecValTok{100}\NormalTok{,}
                         \DataTypeTok{bd.params =}\NormalTok{ speciation\_}\DecValTok{1}\NormalTok{,}
@@
 
-\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-129-1.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-129-1.pdf}
 
@@
 \KeywordTok{set.seed}\NormalTok{(}\DecValTok{11}\NormalTok{)}
-\NormalTok{founding\_tree \textless{}{-}}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{stop.rule =}\NormalTok{ stop\_time\_}\DecValTok{4}\NormalTok{,}
+\NormalTok{founding\_tree \textless{}{-}}\StringTok{ }\KeywordTok{treats}\NormalTok{(}\DataTypeTok{stop.rule =}\NormalTok{ stop\_time\_}\DecValTok{4}\NormalTok{,}
                       \DataTypeTok{bd.params =}\NormalTok{ spec\_}\DecValTok{1}\NormalTok{\_ext\_}\DecValTok{02}\NormalTok{,}
@@
 
-\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-132-1.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-132-1.pdf}
 
@@
 \KeywordTok{set.seed}\NormalTok{(}\DecValTok{1}\NormalTok{)}
-\NormalTok{founding\_tree \textless{}{-}}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{stop.rule =}\NormalTok{ stop\_time\_}\DecValTok{6}\NormalTok{,}
+\NormalTok{founding\_tree \textless{}{-}}\StringTok{ }\KeywordTok{treats}\NormalTok{(}\DataTypeTok{stop.rule =}\NormalTok{ stop\_time\_}\DecValTok{6}\NormalTok{,}
                       \DataTypeTok{bd.params =}\NormalTok{ speciation\_}\DecValTok{1}\NormalTok{\_extinction\_}\DecValTok{03}\NormalTok{,}
@@
 
-\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-136-1.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-136-1.pdf}
 

### 2023-04-24 inst/gitbook/index.Rmd
Update index.Rmd
Big rename

@@
 --- 
-title: "dads manual"
+title: "treats manual"
 author: "Thomas Guillerme (guillert@tcd.ie)"
@@
 link-citations: yes
-github-repo: TGuillerme/dads
-description: "dads R package manual"
+github-repo: TGuillerme/treats
+description: "treats R package manual"
 ---
@@
 
-# `dads`: disparity and diversity simulations.
+# `treats`: disparity and diversity simulations.
 
@@
 
-## What is `dads`?
+## What is `treats`?
 
-%NC: I think here you need an example of use case as it's not immediately obvious what the package is for...
+<!-- %NC: I think here you need an example of use case as it's not immediately obvious what the package is for... -->
 
@@
 
-`dads` is a highly modular and user-friendly package that contains the simulations functionality of all these packages: it allows you to simulate disparity and diversity jointly with a vast array of options that can be easily modified by users.
+`treats` is a highly modular and user-friendly package that contains the simulations functionality of all these packages: it allows you to simulate disparity and diversity jointly with a vast array of options that can be easily modified by users.
 For example, you can easily generate any type of process to generate a trait (BM, OU, something else, etc.) in multiple dependent or independent dimensions through `"traits"` objects.
@@
 
-Because there is an infinite number of ways you might want to generate disparity and diversity (different traits, different modifiers and different events), the `dads` package is designed to make all these parts easy to code separately and integrate in the `dads` core functions.
+Because there is an infinite number of ways you might want to generate disparity and diversity (different traits, different modifiers and different events), the `treats` package is designed to make all these parts easy to code separately and integrate in the `treats` core functions.
 This allows you to simulate finely-tuned multidimensional data for your specific project!
@@
 
-This manual explains how `dads` works in as much details as possible and aims to give you the keys to design your own diversity and disparity simulations.
+This manual explains how `treats` works in as much details as possible and aims to give you the keys to design your own diversity and disparity simulations.
 
@@
 ## Installing the latest version directly from GitHub
-devtools::install_github("TGuillerme/dads")
+devtools::install_github("TGuillerme/treats")
 ```
@@
 ## Loading the package and setting up the start seed.
-library(dads)
+library(treats)
 ```
@@
 If you need help with the package, hopefully the following manual will be useful.
-However, if you have suggestions or comments on on what has already been developed or will be developed, please send me an email (<guillert@tcd.ie>) or if you are a GitHub user, directly create an issue on the [GitHub page](https://github.com/TGuillerme/dads).
+However, if you have suggestions or comments on on what has already been developed or will be developed, please send me an email (<guillert@tcd.ie>) or if you are a GitHub user, directly create an issue on the [GitHub page](https://github.com/TGuillerme/treats).
 Doing so will not only help you but also other users as it will help improve this manual and the package!
 
-## How does `dads` work?
+## How does `treats` work?
 
-Basically, the `dads` function takes your `traits`, `modifiers` and `events` objects to simulate disparity and diversity.
-You will find more details about how these objects (`traits`, `modifiers` and `events`) work in the rest of the tutorial but here is a graphical representation of how `dads` works:
+Basically, the `treats` function takes your `traits`, `modifiers` and `events` objects to simulate disparity and diversity.
+You will find more details about how these objects (`traits`, `modifiers` and `events`) work in the rest of the tutorial but here is a graphical representation of how `treats` works:
 
-![Schematic summary of the `dads` package architecture](dads_structure.png)
+![Schematic summary of the `treats` package architecture](treats_structure.png)
 % NC: The wonkiness of the figure is annoying me lol. Can you center the arrows etc?
@@
 
-The `dads` algorithm is based on a modular version of the birth-death process algorithm.
+The `treats` algorithm is based on a modular version of the birth-death process algorithm.
 The birth-death model is a continuous Markov process (i.e.
@@
 
-The algorithm used in `dads` allows modularity of this process and is based on the following steps (the text in `courier font` is for the name of the process in the algorithm):
+The algorithm used in `treats` allows modularity of this process and is based on the following steps (the text in `courier font` is for the name of the process in the algorithm):
 
@@
 **Selecting a lineage** (`selecting`): this step selects a tip that is currently not extinct.
-In a standard birth-death process this is done randomly, however in `dads` this can be modified based on the birth-death parameters, the currently available lineages and potential trait values.
+In a standard birth-death process this is done randomly, however in `treats` this can be modified based on the birth-death parameters, the currently available lineages and potential trait values.
 *For example, it is possible to put a higher probability for selecting a lineage that is closely related to a lineage that recently went extinct and has a positive trait value.*
@@
 In an exact birth-death process, this is done by drawing a random value from an exponential distribution with a rate of $$\text{number of living lineages} \times (\text{speciation} + \text{extinction parameters})$$.
-In `dads`, this can be modified based on the birth-death parameters, the currently available lineages and potential trait values.
+In `treats`, this can be modified based on the birth-death parameters, the currently available lineages and potential trait values.
 *For example, it is possible to increase branch length by the ratio of living/extinct fossils and a random number drawn from the range of current trait values.*
@@
 **Simulating traits** [optional]\(`traits`): this step allows you to simulate a trait value for the selected lineage from step 1.
-This is typically not part of a standard birth-death process and is handled via the `traits` option in `dads` (see the [`make.traits`](#maketraits) chapter for more details).
+This is typically not part of a standard birth-death process and is handled via the `traits` option in `treats` (see the [`make.traits`](#maketraits) chapter for more details).
  Then go to step 4.
@@
 ```
- Again, in `dads`, this process can be modified based on the birth-death parameters, the currently available lineages and potential trait values.
+ Again, in `treats`, this process can be modified based on the birth-death parameters, the currently available lineages and potential trait values.
 *For example, the lineage can only speciate if its trait value is positive, regardless of whether speciation or extinction have been triggered*.

### 2023-04-24 inst/gitbook/not_include/05_thorough_example.Rmd
Big rename

@@
 
-Here we are going to look at an example on how to simulate complex evolutionary histories with `dads`.
+Here we are going to look at an example on how to simulate complex evolutionary histories with `treats`.
 
@@
 ## Running a birth death tree 
-results <- dads(bd.params = bd_params,
+results <- treats(bd.params = bd_params,
                 stop.rule = stop_rule)
@@
 ## Running a birth death tree 
-results <- dads(bd.params  = bd_params,
+results <- treats(bd.params  = bd_params,
                 stop.rule  = stop_rule,

### 2023-04-24 inst/gitbook/verbose_example.Rmd
Big rename

@@
 ---
-title: "`dads` verbose example"
+title: "`treats` verbose example"
 author: "Thomas Guillerme"
@@
 library(dispRity)
-library(dads)
+library(treats)
 ```
@@
 ```{r}
-libary(dads)
+libary(treats)
 my_params <- make.bd.params(speciation = rnorm)
@@
 trait <- make.traits(process = BM.process, n = 10)
-simulation <- dads(bd.params = my_params, stop.rule = my_stop_rule, trait = trait)
+simulation <- treats(bd.params = my_params, stop.rule = my_stop_rule, trait = trait)
 
@@
 ## 
-library(dads)
+library(treats)
 set.seed(123)
@@
 par(mfrow = c(2,1))
-plot(dads(bd.params = bd.params, stop.rule = stop.rule, traits = traits), main = "no modif")
+plot(treats(bd.params = bd.params, stop.rule = stop.rule, traits = traits), main = "no modif")
 
@@
 
-plot(dads(bd.params = bd.params, stop.rule = stop.rule, traits = traits, modifiers = modifiers), main = "Speciation increase\nwith absolute trait value")
+plot(treats(bd.params = bd.params, stop.rule = stop.rule, traits = traits, modifiers = modifiers), main = "Speciation increase\nwith absolute trait value")
 

### 2023-04-24 R/drop.things.R

@@ -1,15 +1,15 @@
-#' @title Drop things from a dads object
+#' @title Drop things from a treats object
 #' @name drop.things
 #' @aliases drop.fossils drop.livings drop.singles
 #'
-#' @usage drop.things(dads, what)
-#' @usage drop.fossils(dads)
-#' @usage drop.livings(dads)
-#' @usage drop.singles(dads)
+#' @usage drop.things(treats, what)
+#' @usage drop.fossils(treats)
+#' @usage drop.livings(treats)
+#' @usage drop.singles(treats)
 #'
-#' @description Remove fossils or living species or non-bifurcating nodes (singles) from \code{dads} objects or \code{phylo} objects.  
+#' @description Remove fossils or living species or non-bifurcating nodes (singles) from \code{treats} objects or \code{phylo} objects.  
 #'
-#' @param dads \code{dads} data.
+#' @param treats \code{treats} data.
 #' @param what what to drop. Can be \code{"fossils"}, \code{"livings"} or \code{"singles"} (non-bifurcating nodes).
 #' 
 #' @details \emph{NOTE} that dropping living or fossils species DOES NOT drop associated internal nodes and edge lengths. To drop both fossil/living taxa AND internal nodes, you can use for example: \code{drop.things(drop.things(my_data, what = "fossils"), what = "singles")}.
@@ -17,7 +17,7 @@
 #' @examples
 #' ## A random tree with fossils and traits and internal nodes every 0.5 times
 #' set.seed(3)
-#' my_data <- dads(stop.rule = list(max.taxa = 20),
+#' my_data <- treats(stop.rule = list(max.taxa = 20),
 #'                 bd.params = list(speciation = 1, extinction = 1/3),
 #'                 traits    = make.traits(), save.steps = 0.5)
 #'
@@ -41,31 +41,31 @@
 #' ## Removing the internal nodes AND the fossils
 #' drop.singles(drop.fossils(my_data))
 #'
-#' @seealso \code{\link{dads}} \code{\link{plot.dads}}
+#' @seealso \code{\link{treats}} \code{\link{plot.treats}}
 #' 
 #' @author Thomas Guillerme
 #' @export
 
-drop.singles <- function(dads) {
-    if(is(dads, "phylo")) {
+drop.singles <- function(treats) {
+    if(is(treats, "phylo")) {
         ## Add a dummy tip
[…]

### 2023-04-24 R/events.conditions.R

@@ -32,7 +32,7 @@
 #'      }
 #' }
 #' 
-#' More details about the \code{events} functions is explained in the \code{dads} manual: \url{http://tguillerme.github.io/dads}.
+#' More details about the \code{events} functions is explained in the \code{treats} manual: \url{http://tguillerme.github.io/treats}.
 #' 
 #' @examples
 
@@ -49,7 +49,7 @@
 #' 
 #' ## Run the simulations
 #' set.seed(123)
-#' results <- dads(bd.params = bd.params,
+#' results <- treats(bd.params = bd.params,
 #'                 stop.rule = stop.rule,
 #'                 events    = mass_extinction)
 #' ## Plot the results
@@ -75,11 +75,11 @@
 #' 
 #' ## Run the simulations
 #' set.seed(1)
-#' no_change <- dads(bd.params = bd.params,
+#' no_change <- treats(bd.params = bd.params,
 #'                   stop.rule = stop.rule,
 #'                   traits    = traits)
 #' set.seed(1)
-#' process_change <- dads(bd.params = bd.params,
+#' process_change <- treats(bd.params = bd.params,
 #'                        stop.rule = stop.rule,
 #'                        traits    = traits,
 #'                        events    = change_process)
@@ -89,13 +89,13 @@
 #' plot(process_change, ylim = c(-7, 7))
 #' 
 #' 
-#' @seealso \code{\link{dads}} \code{\link{make.events}} \code{\link{events.modifications}}
+#' @seealso \code{\link{treats}} \code{\link{make.events}} \code{\link{events.modifications}}
 #' 
 #' @author Thomas Guillerme
 
 ## The list of conditions
 condition <- function(x, condition) {
-    cat("List of inbuilt condition functions in dads:\n")
+    cat("List of inbuilt condition functions in treats:\n")
     cat("   ?taxa.condition\n")
     cat("   ?time.condition\n")
     cat("   ?trait.condition\n")

### 2023-04-24 R/events.modifications.R

@@ -19,10 +19,10 @@
 #'
 #' @param x   a numerical value to update.
 #' @param ... any specific argument for the modification (see details).
-#' @param bd.params a \code{"bd.params"} \code{"dads"} object for the founding event (see details for founding.event).
-#' @param traits a \code{"traits"} \code{"dads"} object for the founding event (see details for founding.event).
-#' @param modifiers a \code{"modifiers"} \code{"dads"} object for the founding event (see details for founding.event).
-#' @param events an \code{"events"} \code{"dads"} object for the founding event (see details for founding.event).
+#' @param bd.params a \code{"bd.params"} \code{"treats"} object for the founding event (see details for founding.event).
+#' @param traits a \code{"traits"} \code{"treats"} object for the founding event (see details for founding.event).
+#' @param modifiers a \code{"modifiers"} \code{"treats"} object for the founding event (see details for founding.event).
+#' @param events an \code{"events"} \code{"treats"} object for the founding event (see details for founding.event).
 #' @param condition a condition function for the trait value (see details for trait.extinction and update.modifiers).
 #' @param trait which trait to use (see details for trait.extinction).
 #' @param speciation the same argument as in \code{\link{make.bd.params}} (see details for update.bd.params).
@@ -57,22 +57,22 @@
 #'
 #' \item modifications for the target \code{"traits"} 
 #'      \itemize{
-#'          \item \code{update.traits}: this function updates a \code{"dads"} \code{"traits"} object. This function takes as arguments any arguments that can be updated in \code{\link{make.traits}}, namely \code{process}, \code{process.args} and \code{trait.names}.
+#'          \item \code{update.traits}: this function updates a \code{"treats"} \code{"traits"} object. This function takes as arguments any arguments that can be updated in \code{\link{make.traits}}, namely \code{process}, \code{process.args} and \code{trait.names}.
 #'      }
 #' 
 #' \item modifications for the target \code{"modifiers"}
 #'      \itemize{
-#'          \item \code{update.modifiers}: this function updates a \code{"dads"} \code{"modifiers"} object. This function takes as arguments any arguments that can be updated in \code{\link{make.modifiers}}, namely \code{branch.length}, \code{selection}, \code{speciation}, \code{condition} and \code{modify}.
[…]

### 2023-04-24 R/make.bd.params.R

@@ -1,6 +1,6 @@
 #' @title Make birth death parameters
 #'
-#' @description Making bd.params objects for dads.
+#' @description Making bd.params objects for treats.
 #'
 #' @param speciation The speciation parameter. Can be a single \code{numeric} value, a \code{numeric} vector or a \code{function} (default is \code{1}).
 #' @param extinction The extinction parameter. Can be a single \code{numeric} value, a \code{numeric} vector or a \code{function} (default is \code{0}).
@@ -9,7 +9,7 @@
 #' @param speciation.args If \code{speciation} is a function, any additional arguments to passed to the \code{speciation} function.
 #' @param extinction.args If \code{speciation} is a function, any additional arguments to passed to the \code{speciation} function.
 #' @param test Logical whether to test if the bd.params object will work (default is \code{TRUE}).
-#' @param update Optional, another previous \code{"dads"} \code{"bd.params"} object to update (see details).
+#' @param update Optional, another previous \code{"treats"} \code{"bd.params"} object to update (see details).
 
 #' @details
 #' When using \code{update}, the provided arguments (to \code{make.bd.params}) will be the ones updated in the \code{"bd.params"} object.
@@ -38,7 +38,7 @@
 #' plot(my_bd_params)
 #' 
 #'
-#' @seealso \code{\link{dads}}
+#' @seealso \code{\link{treats}}
 #' 
 #' @author Thomas Guillerme
 #' @export
@@ -51,8 +51,8 @@ make.bd.params <- function(speciation = NULL, extinction = NULL, joint = NULL, a
     if(is.null(update)) {
         do_update <- FALSE
     } else {
-        if(!is(update, "dads") || !is(update, "bd.params")) {
-            stop(paste0("This function can only update \"dads\", \"bd.params\" objects. The current object to update is of class: ", paste(class(update), collapse = ", "), "."), call. = FALSE)
+        if(!is(update, "treats") || !is(update, "bd.params")) {
+            stop(paste0("This function can only update \"treats\", \"bd.params\" objects. The current object to update is of class: ", paste(class(update), collapse = ", "), "."), call. = FALSE)
         } else {
             do_update <- TRUE
         }
@@ -187,6 +187,6 @@ make.bd.params <- function(speciation = NULL, extinction = NULL, joint = NULL, a
     }
 
     ## Add the classes
-    class(bd_params) <- c("dads", "bd.params")
+    class(bd_params) <- c("treats", "bd.params")
     return(bd_params)
 }

### 2023-04-24 R/make.events.R

@@ -1,6 +1,6 @@
 #' @title make.events
 #'
-#' @description Making events objects for dads
+#' @description Making events objects for treats
 #'
 #' @param target        What to modify, can be \code{"taxa"}, \code{"bd.params"}, \code{"traits"} or \code{"modifiers"} (see details).
 #' @param condition     A \code{function} returning a logical to trigger the event (see details).
@@ -17,7 +17,7 @@
 #' 
 #' \code{condition} must be a function that returns a \code{logical} value and intakes any of the following arguments: \code{bd.params}, \code{lineage}, \code{traits} and \code{time}. See \code{\link{events.conditions}} for examples.
 #'
-#' \code{modification} must be a function that intakes a first argument named \code{"x"} an returns any specific type of class that can be handled internally by dads. For example, if \code{target = "bd.params"} the \code{modification} function should typically return an updated \code{bd.params} object (see \code{\link{make.bd.params}}). See \code{\link{events.modifications}} for examples.
+#' \code{modification} must be a function that intakes a first argument named \code{"x"} an returns any specific type of class that can be handled internally by treats. For example, if \code{target = "bd.params"} the \code{modification} function should typically return an updated \code{bd.params} object (see \code{\link{make.bd.params}}). See \code{\link{events.modifications}} for examples.
 #' 
 #' 
 #' @examples
@@ -35,7 +35,7 @@
 #' 
 #' ## Run the simulations
 #' set.seed(123)
-#' results <- dads(bd.params = bd.params,
+#' results <- treats(bd.params = bd.params,
 #'                 stop.rule = stop.rule,
 #'                 events    = mass_extinction)
 #' ## Plot the results
@@ -61,11 +61,11 @@
 #' 
 #' ## Run the simulations
 #' set.seed(1)
-#' no_change <- dads(bd.params = bd.params,
+#' no_change <- treats(bd.params = bd.params,
 #'                   stop.rule = stop.rule,
 #'                   traits    = traits)
 #' set.seed(1)
-#' process_change <- dads(bd.params = bd.params,
+#' process_change <- treats(bd.params = bd.params,
 #'                        stop.rule = stop.rule,
 #'                        traits    = traits,
 #'                        events    = change_process)
@@ -75,7 +75,7 @@
 #' plot(process_change, ylim = c(-7, 7))
 #' 
 #' 
[…]

### 2023-04-24 R/make.events_fun.R

@@ -110,7 +110,7 @@ check.events <- function(events) {
 
             bd.params = {
                 bd.params_names <- names(bd.params)
-                if(!is(test_modification, "dads") && !is(test_modification, "bd.params")) {
+                if(!is(test_modification, "treats") && !is(test_modification, "bd.params")) {
                     stop(paste0("The modification function targeting \"bd.params\" must output a list of integers. Currently the modification function output is a ", paste(class(test_modification), collapse = ", "), " object."), call. = FALSE)
                 } 
                 if(!all(bd.params_names %in% names(test_modification))) {
@@ -119,17 +119,17 @@ check.events <- function(events) {
             },
 
             # traits    = {
-            #     if(!is(test_modification, "dads") && !is(test_modification, "traits")) {
+            #     if(!is(test_modification, "treats") && !is(test_modification, "traits")) {
             #         ## Check if the error comes from additional args (i.e. wrong dummy traits object)
-            #         stop(paste0("The modification function targeting \"traits\" must output a dads traits object. Currently the modification function output is a ", paste(class(test_modification), collapse = ", "), "."))
+            #         stop(paste0("The modification function targeting \"traits\" must output a treats traits object. Currently the modification function output is a ", paste(class(test_modification), collapse = ", "), "."))
             #     } else {
             #         check.traits(test_modification, events = TRUE)
             #     }
             # },
 
             modifiers = {
-                if(!is(test_modification, "dads") && !is(test_modification, "modifiers")) {
-                    stop(paste0("The modification function targeting \"modifiers\" must output a dads modifiers object. Currently the modification function output is a ", paste(class(test_modification), collapse = ", "), "."))
+                if(!is(test_modification, "treats") && !is(test_modification, "modifiers")) {
+                    stop(paste0("The modification function targeting \"modifiers\" must output a treats modifiers object. Currently the modification function output is a ", paste(class(test_modification), collapse = ", "), "."))
                 } else {
                     check.modifiers(test_modification, events = TRUE)
                 }

### 2023-04-24 R/make.modifiers.R

@@ -1,14 +1,14 @@
 #' @title make.modifiers
 #'
-#' @description Making modifiers objects for dads based on an ancestor's (parent) trait.
+#' @description Making modifiers objects for treats based on an ancestor's (parent) trait.
 #'
 #' @param branch.length A function for the waiting time generating branch length (can be left empty for the defeault branch length function; see details).
 #' @param selection     A function for selecting the lineage(s) affected by speciation (can be left empty for the default selection function; see details).
 #' @param speciation    A function for triggering the speciation events (can be left empty for the default speciation function; see details).
 #' @param condition     A function giving the condition on which to modify the output of \code{branch.length} or \code{speciation} (see details). If \code{NULL} the condition is always met.
 #' @param modify        A function giving the rule of how to modify the output of \code{branch.length} or \code{speciation} (see details). If \code{NULL} no modification is used.
-#' @param add           Whether to add this modifier to a \code{"dads"} \code{"modifier"} object.
-#' @param update        Optional, another previous \code{"dads"} modifiers object to update (see details).
+#' @param add           Whether to add this modifier to a \code{"treats"} \code{"modifier"} object.
+#' @param update        Optional, another previous \code{"treats"} modifiers object to update (see details).
 #' @param test          Logical whether to test if the modifiers object will work (default is TRUE).
 #' 
 #' @details
@@ -47,11 +47,11 @@
 #' When using \code{update}, the provided arguments (to \code{make.modifiers}) will be the ones updated in the \code{"modifiers"} object.
 #' If the \code{"modifiers"} object contains multiple modifiers (\code{branch.length}, \code{selection} or \code{speciation}), only the called arguments will be updated (e.g. \code{make.modifiers(update = previous_modifiers, speciation = new_speciation)} will only update the speciation process).
 #' 
-#' More details about the \code{modifiers} functions is explained in the \code{dads} manual: \url{http://tguillerme.github.io/dads}.
+#' More details about the \code{modifiers} functions is explained in the \code{treats} manual: \url{http://tguillerme.github.io/treats}.
 #' 
 #' @examples
 #' ## These functions should be fed to the make.modifiers function to create
-#' ## modifiers for dads objects. For example, the following sets specifies that
+#' ## modifiers for treats objects. For example, the following sets specifies that
[…]

### 2023-04-24 R/make.modifiers_fun.R

@@ -2,7 +2,7 @@
 check.modifiers <- function(modifiers, events = FALSE) {
 
     if(events) {
-        events_msg <- "The dads modifiers object returned by the modification function does not work: "
+        events_msg <- "The treats modifiers object returned by the modification function does not work: "
     } else {
         events_msg <- ""
     }

### 2023-04-24 R/make.traits.R

@@ -1,16 +1,16 @@
 #' @title make.traits
 #'
-#' @description Making traits objects for dads
+#' @description Making traits objects for treats
 #'
 #' @param process      The trait process(es) (default is \code{\link{BM.process}}).
 #' @param n            Optional, the number of traits per process (default is \code{1}).
 #' @param start        Optional, the starting values for each traits (default is {0}).
 #' @param process.args Optional, a named list of optional arguments for the trait process.
 #' @param trait.names  Optional, the name(s) of the process(s).
-#' @param add          Optional, another previous \code{"dads"} traits object to which to add the trait.
-#' @param update       Optional, another previous \code{"dads"} traits object to update (see details).
-#' @param background    Optional, another \code{"dads"} \code{"traits"} object to simulate background trait evolution (see details).
-#' @param test         Logical, whether to test if the traits object will work with \code{\link{dads}} (\code{TRUE} - default).
+#' @param add          Optional, another previous \code{"treats"} traits object to which to add the trait.
+#' @param update       Optional, another previous \code{"treats"} traits object to update (see details).
+#' @param background    Optional, another \code{"treats"} \code{"traits"} object to simulate background trait evolution (see details).
+#' @param test         Logical, whether to test if the traits object will work with \code{\link{treats}} (\code{TRUE} - default).
 #' 
 #' 
 #' @details
@@ -20,7 +20,7 @@
 #'
 #' If a \code{background} \code{"traits"} object is given, this object is then applied to all living edges at the same in the background while the main \code{"traits"} is computed. 
 #' 
-#' More details about the \code{"dads"} \code{"traits"} objects is explained in the \code{dads} manual: \url{http://tguillerme.github.io/dads}.
+#' More details about the \code{"treats"} \code{"traits"} objects is explained in the \code{treats} manual: \url{http://tguillerme.github.io/treats}.
 #' 
 #' @examples
 #' ## A simple Brownian motion trait (default)
@@ -43,7 +43,7 @@
 #' make.traits(process = OU.process, trait.names = "OU_trait",
 #'             add = my_traits)
 #'
-#' @seealso \code{\link{dads}} \code{\link{trait.process}}
+#' @seealso \code{\link{treats}} \code{\link{trait.process}}
 #' 
 #' @author Thomas Guillerme
 #' @export
@@ -56,10 +56,10 @@ make.traits <- function(process = BM.process, n = NULL, start = NULL, process.ar
     if(is.null(update)) {
         do_update <- FALSE
     } else {
[…]

### 2023-04-24 R/make.traits_fun.R

@@ -7,7 +7,7 @@ check.traits <- function(traits, events = FALSE) {
     }
 
     if(events) {
-        events_msg <- "The dads traits object returned by the modification function does not work: "
+        events_msg <- "The treats traits object returned by the modification function does not work: "
     } else {
         events_msg <- ""
     }

### 2023-04-24 R/make.treats.R

@@ -1,10 +1,10 @@
-#' @title Make a dads object
+#' @title Make a treats object
 #'
-#' @description Combines a tree and some associated data into a dads object (e.g. for plotting)
+#' @description Combines a tree and some associated data into a treats object (e.g. for plotting)
 #'
 #' @param tree a phylogenetic tree.
 #' @param data a dataset of traits, either a \code{matrix} with column names or a named \code{vector}.
-# @param ... additional \code{dads} objects to add (e.g. \code{traits}).
+# @param ... additional \code{treats} objects to add (e.g. \code{traits}).
 #'
 #' @examples
 #' ## Creating a random tree
@@ -14,16 +14,16 @@
 #' ## Creating a random dataset
 #' my_data <- matrix(rnorm(9),
 #'     dimnames = list(c(my_tree$tip.label, my_tree$node.label)))
-#' ## Creating the dads object
-#' my_dads <- make.dads(tree = my_tree, data = my_data)
-#' plot(my_dads)
+#' ## Creating the treats object
+#' my_treats <- make.treats(tree = my_tree, data = my_data)
+#' plot(my_treats)
 #'
-#' @seealso \code{\link{dads}} \code{\link{plot.dads}}
+#' @seealso \code{\link{treats}} \code{\link{plot.treats}}
 #' 
 #' @author Thomas Guillerme
 #' @export
 
-make.dads <- function(tree, data) {#, ...) {
+make.treats <- function(tree, data) {#, ...) {
     ## Sanitizing
 
     ## Check the tree
@@ -61,13 +61,13 @@ make.dads <- function(tree, data) {#, ...) {
 
     ## Catch the optional info
     # options <- list(...)
-    # valid_options <- unlist(lapply(lapply(options, class), function(x) return("dads" %in% x)))
+    # valid_options <- unlist(lapply(lapply(options, class), function(x) return("treats" %in% x)))
 
     ## Making output
     # output <- list(valid_options)
     output <- list()
     output$tree <- tree
     output$data <- data
-    class(output) <- "dads"
+    class(output) <- "treats"
     return(output)
 }

### 2023-04-24 R/modifiers.R

@@ -2,7 +2,7 @@
 #' @aliases branch.length selection speciation branch.length.trait speciation.trait
 #' @title Modifiers
 #'
-#' @description Different modifiers for the birth death process implemented in dads.
+#' @description Different modifiers for the birth death process implemented in treats.
 #'
 #' @usage branch.length(bd.params = NULL, lineage = NULL, trait.values = NULL,
 #'                      modify.fun = NULL)
@@ -21,13 +21,13 @@
 #' @param modify.fun     A list of internals functions that can modified by \code{events} (see details).
 #' 
 #' @details
-#' \code{bd.params} can be either a named list of parameter values (e.g. \code{list("extinction" = 0, "speciation" = 1)}) but it is typically handled internally from a \code{"dads"} \code{"bd.params"} object.
+#' \code{bd.params} can be either a named list of parameter values (e.g. \code{list("extinction" = 0, "speciation" = 1)}) but it is typically handled internally from a \code{"treats"} \code{"bd.params"} object.
 #' 
-#' \code{modifiers} are functions passed to the birth death process in \code{\link{dads}} to either generate the branch length (named \code{branch.length} and similar) or to decide whether to speciate or go extinct (named \code{speciation} and similar).
+#' \code{modifiers} are functions passed to the birth death process in \code{\link{treats}} to either generate the branch length (named \code{branch.length} and similar) or to decide whether to speciate or go extinct (named \code{speciation} and similar).
 #'
 #' For user defined functions, the \code{modifiers} \emph{must} have at least the arguments described above. For safety, we suggest setting these arguments to \code{NULL}.
 #'
-#' The pre-build \code{modifiers} in the \code{dads} package are (so far):
+#' The pre-build \code{modifiers} in the \code{treats} package are (so far):
 #'
 #' \itemize{
 #'
@@ -43,11 +43,11 @@
 #'
 #'  }
 #' 
-#' More details about the \code{modifiers} functions is explained in the \code{dads} manual: \url{http://tguillerme.github.io/dads}.
+#' More details about the \code{modifiers} functions is explained in the \code{treats} manual: \url{http://tguillerme.github.io/treats}.
 #'
 #' @examples
 #' ## These functions should be fed to the make.modifiers function to create
-#' ## modifiers for dads objects. For example, the following sets specifies that
+#' ## modifiers for treats objects. For example, the following sets specifies that
[…]

### 2023-04-24 R/plot.treats.R

@@ -1,24 +1,24 @@
-#' @title Plot dads objects
+#' @title Plot treats objects
 #'
-#' @description Plotting dads objects (either a simulated tree and trait(s) or a process for traits objects)
+#' @description Plotting treats objects (either a simulated tree and trait(s) or a process for traits objects)
 #'
-#' @param x \code{dads} data.
+#' @param x \code{treats} data.
 #' @param col Optional, a vector of colours that can be named (see details).
 #' @param ... Any additional options to be passed to plot functions (from \code{graphics} or \code{rgl} if \code{use.3D = TRUE}).
 #' @param trait which trait to plot (default is \code{1}; see details).
 #' @param edges either a colour name to attribute to the edges or \code{NULL} to not display the edges (default is \code{"grey"}).
 #' @param tips.nodes optional, a colour to circle tips and nodes (only used if \code{use.3D = FALSE}). By default \code{tips.nodes = NULL}.
 #' @param use.3D logical, whether to use a 3D plot or not (default is \code{FALSE}; see details).
-#' @param simulations if the input is a \code{dads} \code{traits} or \code{bd.params} object, how many replicates to run (default is \code{50}).
-#' @param cent.tend if the input is a \code{dads} \code{traits}, which central tendency to plot (default is \code{mean}).
-#' @param quantiles if the input is a \code{dads} \code{traits}, which quantiles to plot (default are \code{c(95, 50))}).
+#' @param simulations if the input is a \code{treats} \code{traits} or \code{bd.params} object, how many replicates to run (default is \code{50}).
+#' @param cent.tend if the input is a \code{treats} \code{traits}, which central tendency to plot (default is \code{mean}).
+#' @param quantiles if the input is a \code{treats} \code{traits}, which quantiles to plot (default are \code{c(95, 50))}).
 #' @param legend logical, whether to display the legend in 2D plots (\code{TRUE}) or not (\code{FALSE}; default)
 #' 
 #' @details
 #' The \code{col} option can be either:
 #' \itemize{
-#'      \item a \code{vector} of colours to be applied to \code{"dads"} \code{"traits"} objects (for respectively the median, 50% CI and 95% CI - by default this is \code{col = c("black", "grey", "lightgrey")}). This is the default option when plotting traits.
[…]

### 2023-04-24 R/print.treats.R

@@ -1,27 +1,27 @@
-#' @title Prints a \code{dads} object.
+#' @title Prints a \code{treats} object.
 #'
-#' @description Summarises the content of a \code{dads} object.
+#' @description Summarises the content of a \code{treats} object.
 #'
-#' @param x   A \code{dads} object.
+#' @param x   A \code{treats} object.
 #' @param all \code{logical}; whether to display the entire object (\code{TRUE}) or just summarise its contents (\code{FALSE} - default).
-#' @param ... further arguments to be passed to \code{print} or to \code{print.dads}.
+#' @param ... further arguments to be passed to \code{print} or to \code{print.treats}.
 #' 
 #' @examples
-#' ## A dads birth-death parameters object
+#' ## A treats birth-death parameters object
 #' make.bd.params()
-#' ## A dads traits object
+#' ## A treats traits object
 #' make.traits()
-#' ## A dads modifiers object
+#' ## A treats modifiers object
 #' make.modifiers()
-#' ## A dads object
-#' dads(stop.rule = list(max.taxa = 10), traits = make.traits())
+#' ## A treats object
+#' treats(stop.rule = list(max.taxa = 10), traits = make.traits())
 #'
-#' @seealso \code{\link{dads}}
+#' @seealso \code{\link{treats}}
 #' 
 #' @author Thomas Guillerme
 #' @export
 
-print.dads <- function(x, all = FALSE, ...) {
+print.treats <- function(x, all = FALSE, ...) {
     
     ## Get the call name
     match_call <- match.call()
@@ -33,34 +33,34 @@ print.dads <- function(x, all = FALSE, ...) {
         class(tmp) <- "list"
         print(tmp)
     } else {
-        ## Dual class dads objects
+        ## Dual class treats objects
         if(length(class(x)) > 1) {
             switch(class(x)[[2]],
                 traits = {
-                    cat(" ---- dads traits object ---- \n")
+                    cat(" ---- treats traits object ---- \n")
                     print.traits.info(x$main)
                     if(!is.null(x$background)) {
                         cat("And a background trait (see x$background for info).")
                     }
                 },
                 events = {
-                    cat(" ---- dads events object ---- \n")
+                    cat(" ---- treats events object ---- \n")
                     lapply(x, print.events.info)
                 },
                 modifiers = {
-                    cat(" ---- dads modifiers object ---- \n")
+                    cat(" ---- treats modifiers object ---- \n")
                     print.modifiers.info(x)
                 },
                 bd.params = {
[…]

### 2023-04-24 R/sanitizing.R

@@ -99,7 +99,7 @@ check.list <- function(list, check.fun, condition, ...) {
     }
 }
 
-## Checking dads specific class
+## Checking treats specific class
 check.args <- function(fun, fun_name, required_args) {
 
     ## Must be a function

### 2023-04-24 R/trait.process.R

@@ -2,7 +2,7 @@
 #' @aliases BM.process OU.process no.process multi.peak.process repulsion.process
 #' @title Trait processes
 #'
-#' @description Different trait processes implemented in dads.
+#' @description Different trait processes implemented in treats.
 #'
 #' @usage trait.process(x0, edge.length, ...)
 #'
@@ -11,7 +11,7 @@
 #' @param ... Any optional argument for the specific process (see details).
 #' 
 #' @details
-#' The different trait processes implemented in dads are:
+#' The different trait processes implemented in treats are:
 #' 
 #' \itemize{
 #'      \item{BM.process} A Brownian motion process (uni or multidimensional). This function is based on \code{\link[MASS]{mvrnorm}}.
@@ -52,18 +52,18 @@
 #'               \item \code{sd} the normal distribution standard deviation.
 #'               \item \code{repulsion} the minimal distance requested between trait values.
 #'               \item \code{max.try} the maximum number of values to draw (if the repulsion value is to hard to achieve).
-#'               \item \code{trait.values} LEAVE AS \code{NULL} (it designates the trait value table from the birth death process and is handled internally by \code{\link{dads}}).
-#'               \item \code{lineage} LEAVE AS \code{NULL} (it designates the lineage object from the birth death process and is handled internally by \code{\link{dads}}).
-#'               \item \code{trait} LEAVE AS \code{NULL} (it which trait to use and is analysed an is handled internally by \code{\link{dads}}).
+#'               \item \code{trait.values} LEAVE AS \code{NULL} (it designates the trait value table from the birth death process and is handled internally by \code{\link{treats}}).
+#'               \item \code{lineage} LEAVE AS \code{NULL} (it designates the lineage object from the birth death process and is handled internally by \code{\link{treats}}).
+#'               \item \code{trait} LEAVE AS \code{NULL} (it which trait to use and is analysed an is handled internally by \code{\link{treats}}).
 #'          }
 #' 
 #' }
 #' 
-#' More details about the \code{trait.process} functions is explained in the \code{dads} manual: \url{http://tguillerme.github.io/dads}.
+#' More details about the \code{trait.process} functions is explained in the \code{treats} manual: \url{http://tguillerme.github.io/treats}.
 #' 
 #' @examples
[…]

### 2023-04-24 R/treats.R

@@ -1,12 +1,12 @@
 #' @title Diversity and disparity simulator
 #'
-#' @description Simulating phylogenetic trees and trait disparity. See full manual here: https://github.com/TGuillerme/dads
+#' @description Simulating phylogenetic trees and trait disparity. See full manual here: https://github.com/TGuillerme/treats
 #'
 #' @param stop.rule  The rules on when to stop the simulation (see details).
-#' @param bd.params  A \code{"dads"} \code{"bd.params"} object or a named list of parameters for the birth-death process (see details).
-#' @param traits     A \code{"dads"} \code{"traits"} object (see \code{\link{make.traits}}).
-#' @param modifiers  A \code{"dads"} \code{"modifiers"} object (see \code{\link{make.modifiers}}).
-#' @param events     A \code{"dads"} \code{"events"} object (see \code{\link{make.events}}).
+#' @param bd.params  A \code{"treats"} \code{"bd.params"} object or a named list of parameters for the birth-death process (see details).
+#' @param traits     A \code{"treats"} \code{"traits"} object (see \code{\link{make.traits}}).
+#' @param modifiers  A \code{"treats"} \code{"modifiers"} object (see \code{\link{make.modifiers}}).
+#' @param events     A \code{"treats"} \code{"events"} object (see \code{\link{make.events}}).
 #' @param save.steps Optional, \code{"numeric"} value to save the simulations at specific internal points (this can slow down the algorithm significantly for large trees). 
 #' @param null.error Logical, whether to return an error when the birth-death parameters fails to build a tree (\code{FALSE}; default and highly recommended) or whether to return \code{NULL} (\code{TRUE}). Can also be set to an integer value for the numbers of trials (see details).
 #' 
@@ -18,7 +18,7 @@
 #'   \item \code{max.time}   The maximum amount of phylogenetic (in arbitrary units).
 #' }
 #' 
-#' \code{bd.params}: This can be either a \code{"dads"} \code{"bd.params"} object (see \code{\link{make.bd.params}}) or a list of named parameters. The allowed parameters are:
+#' \code{bd.params}: This can be either a \code{"treats"} \code{"bd.params"} object (see \code{\link{make.bd.params}}) or a list of named parameters. The allowed parameters are:
 #' \itemize{
 #'   \item \code{speciation} The speciation parameter value.
[…]

### 2023-04-24 R/utilities.R

@@ -1,15 +1,15 @@
-## Utility functions for dads
+## Utility functions for treats
 
 #' @title Get parent traits
 #'
-#' @description An internal utility function for \code{modifiers}, \code{traits} or \code{events} to access the value(s) of the parent traits in the \code{dads} algorithm
+#' @description An internal utility function for \code{modifiers}, \code{traits} or \code{events} to access the value(s) of the parent traits in the \code{treats} algorithm
 #'
 #' @param trait.values   The internal table of trait values
 #' @param lineage        The internal lineage data list
 #' @param current        Whether to consider only the current lineage (\code{TRUE} - default) or all the living lineages (\code{FALSE}).
 #' 
 #' @details
-#' This function is designed to be used internally in \code{dads} to help \code{modifiers}, \code{traits} or \code{events} objects to access the parent traits of the lineages simulated through the internal birth death algorithm. 
+#' This function is designed to be used internally in \code{treats} to help \code{modifiers}, \code{traits} or \code{events} objects to access the parent traits of the lineages simulated through the internal birth death algorithm. 
 #' 
 #' @examples
 #' ## Speciation event is more likely if lineage's ancestor is further away from the mean trait value
@@ -26,7 +26,7 @@
 #' distance.speciation <- make.modifiers(speciation = speciation,
 #'                                       modify = distance.modify)
 #'
-#' @seealso \code{\link{dads}} \code{\link{make.modifiers}}
+#' @seealso \code{\link{treats}} \code{\link{make.modifiers}}
 #' 
 #' @author Thomas Guillerme
 #' @export

### 2023-04-24 inst/gitbook/compiling_gitbook.sh

@@ -1,4 +1,4 @@
-rm dads_manual.Rmd
+rm treats_manual.Rmd
 
 ## Remove files before hand
 rm -R _book/

### 2023-04-24 inst/gitbook/not_include/Simulating_trees_and_traits.R

@@ -1,8 +1,8 @@
 ## This is a quick script to simulate some trees with various birth-death parameters and some traits with OU or BM processes
 
-## Installing the development version of dads
-devtools::install_github("TGuillerme/dads")
-library(dads)
+## Installing the development version of treats
+devtools::install_github("TGuillerme/treats")
+library(treats)
 
 
 ####################
@@ -29,7 +29,7 @@ trait_process <- make.traits(process = BM.process)
 ## Generating the trees and data
 ## -----------
 ## Generating a random tree with 100 living species
-my_data <- dads(bd.params  = bd_params,
+my_data <- treats(bd.params  = bd_params,
                 stop.rule  = stop_rule,
                 traits     = trait_process,
                 null.error = 100) # the null.error parameter will loop through the parameters up to 100 times before returning an error (i.e. a failed tree)
@@ -39,7 +39,7 @@ plot(my_data)
 ## ------------
 ## Keeping only the living species
 ## ------------
-my_data <- drop.fossil.dads(my_data)
+my_data <- drop.fossil.treats(my_data)
 plot(my_data)
 
 ## ------------
@@ -65,7 +65,7 @@ trait_process <- make.traits(process = OU.process)
 ## Running multiple simulations
 ## ------------
 ## You can wrap all that into a replicate loop to generate a list of trees (here to generate 5 simulations)
-all_data <- replicate(5, dads(bd.params  = bd_params,
+all_data <- replicate(5, treats(bd.params  = bd_params,
                               stop.rule  = stop_rule,
                               traits     = trait_process,
                               null.error = 100), simplify = FALSE)

### 2023-04-24 tests/test-benchmarking.R

@@ -25,16 +25,16 @@
 
 # test_age_2_fail <- microbenchmark(
 #     treeSim = sim.bd.age(age = 2, lambda = lambda, mu = mu, numbsim = 1),
-#     dads = dads(stop.rule = list(max.time = 2), bd.params = list(speciation = lambda, extinction = mu), null.error = TRUE))
+#     treats = treats(stop.rule = list(max.time = 2), bd.params = list(speciation = lambda, extinction = mu), null.error = TRUE))
 # test_age_3_fail <- microbenchmark(
 #     treeSim = sim.bd.age(age = 3, lambda = lambda, mu = mu, numbsim = 1),
-#     dads = dads(stop.rule = list(max.time = 3), bd.params = list(speciation = lambda, extinction = mu), null.error = TRUE))
+#     treats = treats(stop.rule = list(max.time = 3), bd.params = list(speciation = lambda, extinction = mu), null.error = TRUE))
 # test_age_4_fail <- microbenchmark(
 #     treeSim = sim.bd.age(age = 4, lambda = lambda, mu = mu, numbsim = 1),
-#     dads = dads(stop.rule = list(max.time = 4), bd.params = list(speciation = lambda, extinction = mu), null.error = TRUE))
+#     treats = treats(stop.rule = list(max.time = 4), bd.params = list(speciation = lambda, extinction = mu), null.error = TRUE))
 # test_age_5_fail <- microbenchmark(
 #     treeSim = sim.bd.age(age = 5, lambda = lambda, mu = mu, numbsim = 1),
-#     dads = dads(stop.rule = list(max.time = 5), bd.params = list(speciation = lambda, extinction = mu), null.error = TRUE))
+#     treats = treats(stop.rule = list(max.time = 5), bd.params = list(speciation = lambda, extinction = mu), null.error = TRUE))
 
 # n<-10
 # lambda <- 2.0
@@ -49,16 +49,16 @@
 
 # test_taxa_10_fail <- microbenchmark(
 #     treeSim = sim.bd.taxa(n = 10, lambda = lambda, mu = mu, numbsim = 1),
-#     dads = dads(stop.rule = list(max.living = 10), bd.params = list(speciation = lambda, extinction = mu), null.error = TRUE))
+#     treats = treats(stop.rule = list(max.living = 10), bd.params = list(speciation = lambda, extinction = mu), null.error = TRUE))
 # test_taxa_50_fail <- microbenchmark(
[…]

### 2023-04-24 tests/testthat.R

@@ -1,5 +1,5 @@
 library(testthat)
-library(dads)
+library(treats)
 
 # test_check("dispRity", reporter = "list")
-test_check("dads")
+test_check("treats")

### 2023-04-24 tests/testthat/test-birth.death.R

@@ -258,7 +258,7 @@ test_that("events work", {
 
     set.seed(7)
     test <- birth.death.tree.traits(bd.params = bd.params, stop.rule = stop.rule, traits = make.traits(), modifiers = NULL, events = events)
-    class(test) <- c("dads")
+    class(test) <- c("treats")
     # plot(test)
     ## 244 taxa generated
     expect_equal(Ntip(test$tree), 244)
@@ -338,8 +338,8 @@ test_that("events work", {
     test2 <- birth.death.tree.traits(bd.params = bd.params, stop.rule = stop.rule, traits = traits, modifiers = NULL, events = events)
     ## Visual testing
     # par(mfrow = c(1,2))
-    # class(test) <- "dads" ; plot(test, ylim = c(-5, 8))
-    # class(test2) <- "dads" ; plot(test2, ylim = c(-5, 8))
+    # class(test) <- "treats" ; plot(test, ylim = c(-5, 8))
+    # class(test2) <- "treats" ; plot(test2, ylim = c(-5, 8))
     expect_false(nrow(test$data) == nrow(test2$data))
 
     ## Changing a trait argument (e.g. sigma) when a trait reaches value x
@@ -361,8 +361,8 @@ test_that("events work", {
     test2 <- birth.death.tree.traits(bd.params = bd.params, stop.rule = stop.rule, traits = traits, modifiers = NULL, events = change_correlation)
     ## Visual testing
     # par(mfrow = c(2,1))
-    # class(test) <- "dads" ; plot(test, trait = 2)
-    # class(test2) <- "dads" ; plot(test2, trait = 2)
+    # class(test) <- "treats" ; plot(test, trait = 2)
+    # class(test2) <- "treats" ; plot(test2, trait = 2)
     # plot(test, trait = c(1:2))
     # plot(test2, trait = c(1:2), use.3D = TRUE)
 
@@ -395,8 +395,8 @@ test_that("events work", {
 
     ## Visual testing
     # par(mfrow = c(2,1))
-    # class(test) <- "dads" ; plot(test)
-    # class(test2) <- "dads" ; plot(test2)
+    # class(test) <- "treats" ; plot(test)
+    # class(test2) <- "treats" ; plot(test2)
 
     ## Less tips in the second tree
     expect_lt(Ntip(test2$tree), Ntip(test$tree))
@@ -424,8 +424,8 @@ test_that("events work", {
 
     # # Visual testing
     # par(mfrow = c(2,1))
[…]

### 2023-04-24 tests/testthat/test-drop.things.R

@@ -2,7 +2,7 @@
 test_that("drop.fossils and drop.livings works", {
     ## Create a simple bd tree
     set.seed(1)
-    phy <- dads(bd.params = list(extinction = 1/3), stop.rule = list(max.living = 20))
+    phy <- treats(bd.params = list(extinction = 1/3), stop.rule = list(max.living = 20))
     expect_gt(Ntip(phy), 20)
     test <- drop.fossils(phy)
     expect_is(test, "phylo")
@@ -13,15 +13,15 @@ test_that("drop.fossils and drop.livings works", {
 
         ## Create a simple bd tree + traits
         set.seed(12)
-        phy <- dads(bd.params = list(extinction = 1/3), stop.rule = list(max.living = 20), traits = make.traits())
+        phy <- treats(bd.params = list(extinction = 1/3), stop.rule = list(max.living = 20), traits = make.traits())
         expect_gt(Ntip(phy$tree), 20)
         expect_gt(nrow(phy$data), 39)
         test <- drop.fossils(phy)
-        expect_is(test, "dads")
+        expect_is(test, "treats")
         expect_equal(Ntip(test$tree), 20)
         expect_equal(nrow(test$data), 42)
         test <- drop.livings(phy)
-        expect_is(test, "dads")
+        expect_is(test, "treats")
         expect_equal(Ntip(test$tree), 3)
         expect_equal(nrow(test$data), 12)
 })
@@ -34,7 +34,7 @@ test_that("drop.singles works", {
 
     ## Working for regular steps
     set.seed(2)
-    test <- dads(bd.params = bd.params, traits = traits, stop.rule = stop.rule, save.steps = 0.5)
+    test <- treats(bd.params = bd.params, traits = traits, stop.rule = stop.rule, save.steps = 0.5)
     expect_equal(Ntip(test$tree), 44)
     expect_equal(Nnode(test$tree), 128)
 
@@ -44,10 +44,10 @@ test_that("drop.singles works", {
     expect_is(tust, "phylo")
     expect_equal(Nnode(tust), 43)
 
-    test_dads <- test
-    expect_equal(Nnode(test_dads$tree), 128)
-    tast <- drop.singles(test_dads)
-    expect_is(tast, "dads")
+    test_treats <- test
+    expect_equal(Nnode(test_treats$tree), 128)
+    tast <- drop.singles(test_treats)
+    expect_is(tast, "treats")
[…]

### 2023-04-24 tests/testthat/test-make.bd.params.R

@@ -3,49 +3,49 @@ test_that("make.bd.params works", {
 
     ## Test default
     test <- make.bd.params()
-    expect_equal(class(test), c("dads", "bd.params"))
+    expect_equal(class(test), c("treats", "bd.params"))
     expect_equal(names(test), c("joint", "absolute", "speciation", "extinction", "call"))
     expect_equal(sample.from(test), list("speciation" = 1, "extinction" = 0))
     out <- capture_output(print(test))
-    expect_equal(out, c(" ---- dads birth-death parameters object ---- \nspeciation: 1.\nextinction: 0."))
+    expect_equal(out, c(" ---- treats birth-death parameters object ---- \nspeciation: 1.\nextinction: 0."))
     expect_null(plot(test))
     ## Update works
     tust <- make.bd.params(update = test, extinction = 0.5)
     expect_equal(sample.from(tust), list("speciation" = 1, "extinction" = 0.5))
     out <- capture_output(print(tust))
-    expect_equal(out, c(" ---- dads birth-death parameters object ---- \nspeciation: 1.\nextinction: 0.5."))
+    expect_equal(out, c(" ---- treats birth-death parameters object ---- \nspeciation: 1.\nextinction: 0.5."))
     expect_null(plot(tust))
 
     ## Test single values
     test <- make.bd.params(speciation = 10, extinction = c(1, 2), joint = TRUE)
-    expect_equal(class(test), c("dads", "bd.params"))
+    expect_equal(class(test), c("treats", "bd.params"))
     expect_equal(names(test), c("joint", "absolute", "speciation", "extinction", "call"))
     set.seed(1)
     expect_equal(sample.from(test), list("speciation" = 10, "extinction" = 1))
     expect_equal(sample.from(test), list("speciation" = 10, "extinction" = 2))
     out <- capture_output(print(test))
-    expect_equal(out, c(" ---- dads birth-death parameters object ---- \njoint sampling for:\nspeciation: 10.\nextinction: 1, 2."))
+    expect_equal(out, c(" ---- treats birth-death parameters object ---- \njoint sampling for:\nspeciation: 10.\nextinction: 1, 2."))
     expect_null(plot(test))
 
     ## Test functions
     test <- make.bd.params(speciation = rnorm)
-    expect_equal(class(test), c("dads", "bd.params"))
+    expect_equal(class(test), c("treats", "bd.params"))
[…]

### 2023-04-24 tests/testthat/test-make.events.R

@@ -20,6 +20,6 @@ test_that("make.events works", {
 
     event <- make.events(target = "traits", condition = condition.test.ok, modification = modification.test.ok)
     print <- capture_output(print(event))
-    expect_equal(print, " ---- dads events object ---- \nEvent targeting \"traits\" to be triggered 1 time.\nThe condition function is: condition.test.ok\nThe modification function is: modification.test.ok\n")
+    expect_equal(print, " ---- treats events object ---- \nEvent targeting \"traits\" to be triggered 1 time.\nThe condition function is: condition.test.ok\nThe modification function is: modification.test.ok\n")
 
 })

### 2023-04-24 tests/testthat/test-make.modifiers.R

@@ -44,7 +44,7 @@ test_that("make.modifiers works", {
 
     ## Wrong arguments for add
     error <- capture_error(make.modifiers(add = TRUE))
-    expect_equal(error[[1]], "You can only add to a \"dads\" \"modifiers\" object. Check the documentation from the following function for helping designing such objects:\n    ?make.modifiers")
+    expect_equal(error[[1]], "You can only add to a \"treats\" \"modifiers\" object. Check the documentation from the following function for helping designing such objects:\n    ?make.modifiers")
 
     ## Wrong arguments for test
     error <- capture_error(make.modifiers(test = "whatever"))
@@ -58,7 +58,7 @@ test_that("make.modifiers works", {
                            speciation    = speciation,
                            condition     = condition,
                            modify        = modify)
-    expect_is(test, c("dads", "modifiers"))
+    expect_is(test, c("treats", "modifiers"))
     expect_equal(names(test), c("waiting", "selecting", "speciating", "call"))
     expect_equal(names(test[[1]]), c("fun", "internal"))
     expect_equal(names(test[[2]]), c("fun", "internal"))
@@ -67,7 +67,7 @@ test_that("make.modifiers works", {
     ## Working correctly without condition and modify
     test <- make.modifiers(branch.length = branch.length.trait,
                            speciation    = speciation.trait)    
-    expect_is(test, c("dads", "modifiers"))
+    expect_is(test, c("treats", "modifiers"))
     expect_equal(names(test), c("waiting", "selecting", "speciating", "call"))
     expect_equal(names(test[[1]]), c("fun", "internal"))
     expect_equal(names(test[[2]]), c("fun", "internal"))
@@ -87,7 +87,7 @@ test_that("make.modifiers works", {
     test <- make.modifiers(branch.length = branch.length.trait,
                            speciation    = speciation.trait,
                            condition     = always.false)    
-    expect_is(test, c("dads", "modifiers"))
+    expect_is(test, c("treats", "modifiers"))
     expect_equal(names(test), c("waiting", "selecting", "speciating", "call"))
     expect_equal(names(test[[1]]), c("fun", "internal"))
     expect_equal(names(test[[2]]), c("fun", "internal"))
@@ -107,7 +107,7 @@ test_that("make.modifiers works", {
     test <- make.modifiers(branch.length = branch.length.trait,
[…]

### 2023-04-24 tests/testthat/test-make.traits.R

@@ -14,13 +14,13 @@ test_that("make.traits works", {
     error <- capture_error(make.traits(process.args = "nope"))
     expect_equal(error[[1]], "process.args must be of class list.")
     error <- capture_error(make.traits(add = TRUE))
-    expect_equal(error[[1]], "You can only add to a \"dads\" \"traits\" object. Check the documentation from the following function for helping designing such objects:\n    ?make.traits")
+    expect_equal(error[[1]], "You can only add to a \"treats\" \"traits\" object. Check the documentation from the following function for helping designing such objects:\n    ?make.traits")
     error <- capture_error(make.traits(test = "TRUE"))
     expect_equal(error[[1]], "test must be of class logical.")
 
     ## Default example works
     test <- make.traits()
-    expect_is(test, c("dads", "traits"))
+    expect_is(test, c("treats", "traits"))
     expect_equal(length(test), 2)
     expect_equal(names(test), c("main", "background"))
     expect_equal(names(test$main[[1]]), c("process", "start", "trait_id"))
@@ -36,7 +36,7 @@ test_that("make.traits works", {
     }
 
     test <- make.traits(n = 2)
-    expect_is(test, c("dads", "traits"))
+    expect_is(test, c("treats", "traits"))
     expect_equal(length(test), 2)
     expect_equal(names(test$main[[1]]), c("process", "start", "trait_id"))
     expect_equal(test$main[[1]]$trait_id, c(1,2))
@@ -46,7 +46,7 @@ test_that("make.traits works", {
 
     ## Add works
     test2 <- make.traits(process = c(BM.process, element.rank), n = c(4,2), add = test, start = 1)
-    expect_is(test2, c("dads", "traits"))
+    expect_is(test2, c("treats", "traits"))
     expect_equal(length(test2), 2)
     expect_equal(length(test2$main), 3)
     expect_equal(names(test2$main$A), c("process", "start", "trait_id"))
@@ -59,7 +59,7 @@ test_that("make.traits works", {
 
     ## Add works with proper name update
     test2 <- make.traits(process = c(BM.process, element.rank), n = c(4,2), add = make.traits(trait.names = "bob"))
-    expect_is(test2, c("dads", "traits"))
+    expect_is(test2, c("treats", "traits"))
     expect_equal(length(test2$main), 3)
     expect_equal(names(test2$main$bob), c("process", "start", "trait_id"))
     expect_equal(test2$main$bob$trait_id, c(1))
[…]

### 2023-04-24 tests/testthat/test-make.treats.R

@@ -1,50 +1,50 @@
 ## Test
-test_that("make.dads works", {
+test_that("make.treats works", {
 
     ## Simple example (works)
     my_tree <- rtree(5)
     my_tree$node.label <- letters[1:4]
     my_data <- matrix(rnorm(9), dimnames = list(c(my_tree$tip.label, my_tree$node.label)))
-    my_dads <- make.dads(tree = my_tree, data = my_data)
-    expect_is(my_dads, "dads")
-    expect_is(my_dads$tree, "phylo")
-    expect_is(my_dads$data, "matrix")
+    my_treats <- make.treats(tree = my_tree, data = my_data)
+    expect_is(my_treats, "treats")
+    expect_is(my_treats$tree, "phylo")
+    expect_is(my_treats$data, "matrix")
 
     ## Data is a data.frame
     data_frame <- as.data.frame(my_data)
-    my_dads <- make.dads(tree = my_tree, data = data_frame)
-    expect_is(my_dads, "dads")
-    expect_is(my_dads$tree, "phylo")
-    expect_is(my_dads$data, "matrix")
+    my_treats <- make.treats(tree = my_tree, data = data_frame)
+    expect_is(my_treats, "treats")
+    expect_is(my_treats$tree, "phylo")
+    expect_is(my_treats$data, "matrix")
 
     ## Data is a vector
     data_vector <- c(my_data)
     names(data_vector) <- rownames(my_data)
-    my_dads <- make.dads(tree = my_tree, data = data_vector)
-    expect_is(my_dads, "dads")
-    expect_is(my_dads$tree, "phylo")
-    expect_is(my_dads$data, "matrix")
+    my_treats <- make.treats(tree = my_tree, data = data_vector)
+    expect_is(my_treats, "treats")
+    expect_is(my_treats$tree, "phylo")
+    expect_is(my_treats$data, "matrix")
 
     ## Wrong data (no names)
     wrong_data <- my_data
     rownames(wrong_data) <- NULL
-    error <- capture_error(make.dads(my_tree, wrong_data))
+    error <- capture_error(make.treats(my_tree, wrong_data))
     expect_equal(error[[1]],  "data must be a matrix with column names or a named vector.")
 
     ## Wrong tree (no tips)
     wrong_tree <- rtree(5)
-    error <- capture_error(make.dads(wrong_tree, my_data))
+    error <- capture_error(make.treats(wrong_tree, my_data))
     expect_equal(error[[1]], "The input tree must have tip and node labels.")
 
[…]

### 2023-04-24 tests/testthat/test-plot.treats.R

@@ -1,18 +1,18 @@
-test_that("plot.dads works for traits", {
+test_that("plot.treats works for traits", {
     
     list_of_traits <- make.traits(process = c(no.process, BM.process, OU.process), trait.names = c("No process (normal)", "Brownian motion", "Ornstein-Uhlenbeck"))
     ## Right output
-    expect_null(plot.dads(list_of_traits))
-    expect_null(plot.dads(list_of_traits, xlim = c(1, 1.5), main = "ha"))
-    expect_null(plot.dads(list_of_traits, trait = 2))
-    expect_null(plot.dads(list_of_traits, trait = 3))
-    expect_null(plot.dads(list_of_traits, trait = 3,cent.tend = sd, quantiles = c(25, 10, 5), col = palette()[1:4]))
+    expect_null(plot.treats(list_of_traits))
+    expect_null(plot.treats(list_of_traits, xlim = c(1, 1.5), main = "ha"))
+    expect_null(plot.treats(list_of_traits, trait = 2))
+    expect_null(plot.treats(list_of_traits, trait = 3))
+    expect_null(plot.treats(list_of_traits, trait = 3,cent.tend = sd, quantiles = c(25, 10, 5), col = palette()[1:4]))
 })
 
-test_that("plot.dads works for dads", {
+test_that("plot.treats works for treats", {
     
     ## Simple example    
-    test <- dads(stop.rule = list("max.living" = 10), traits = make.traits())
+    test <- treats(stop.rule = list("max.living" = 10), traits = make.traits())
     ## Default plot
     expect_null(plot(test))
     ## Other options
@@ -20,19 +20,19 @@ test_that("plot.dads works for dads", {
 
     ## A more complex trait
     complex_traits <- make.traits(process = c(BM.process, BM.process), n = c(2,3), process.args = list(list(Sigma = diag(2)), list(Sigma = matrix(1/3, 3, 3))), trait.names = c("bib", "bob"))
-    test <- dads(stop.rule = list("max.living" = 10), traits = complex_traits)
+    test <- treats(stop.rule = list("max.living" = 10), traits = complex_traits)
     expect_null(plot(test, trait = 3))
 
     ## A big tree
     set.seed(12)
-    big_test <- dads(bd.params = list(speciation = 1, extinction = 1/3), stop.rule = list(max.taxa = 1000), traits = make.traits())
[…]

### 2023-04-24 tests/testthat/test-print.dads.R

@@ -1,70 +1,0 @@
-## Test
-test_that("print.dads works for traits", {
-
-    ## Simple trait
-    simple <- make.traits()
-    out <- capture_output(print.traits.info(simple$main))
-    expect_equal(out[[1]], "1 trait for 1 process (A) with one starting value (0).")
-
-    ## Complex
-    complex <- make.traits(process = c(BM.process, BM.process), n = c(2,3), process.args = list(list(Sigma = diag(2)), list(Sigma = matrix(1/3, 3, 3))), trait.names = c("bib", "bob"))
-    out <- capture_output(print.traits.info(complex$main))
-    expect_equal(out[[1]], "5 traits for 2 processes (bib:2, bob:3) with one starting value (0).\nprocess bib uses the following extra argument: Sigma;\nprocess bob uses the following extra argument: Sigma;")
-
-    ## Simple dads
-    out <- capture_output(print.dads(simple))
-    expect_equal(out[[1]], " ---- dads traits object ---- \n1 trait for 1 process (A) with one starting value (0).")
-
-    background <- make.traits(background = make.traits())
-    out <- capture_output(print.dads(background))
-    expect_equal(out[[1]], " ---- dads traits object ---- \n1 trait for 1 process (A) with one starting value (0).\nAnd a background trait (see x$background for info).")
-})
-
-test_that("print.dads works for dads", {
-
-    ## Simple birth.death tree with one BM trait
-    test <- dads(stop.rule = list("max.living" = 10), traits = make.traits())
-    out <- capture_output(print.dads(test))
-    expect_equal(out[[1]], " ---- dads object ---- \nSimulated diversity data (x$tree):\n\nPhylogenetic tree with 10 tips and 9 internal nodes.\n\nTip labels:\n  t1, t2, t3, t4, t5, t6, ...\nNode labels:\n  n1, n2, n3, n4, n5, n6, ...\n\nRooted; includes branch lengths.\n\nSimulated disparity data (x$data):\n1 trait for 1 process (A) with one starting value (0).")
-    
-    ## A more complex trait
-    complex_traits <- make.traits(process = c(BM.process, BM.process), n = c(2,3), process.args = list(list(Sigma = diag(2)), list(Sigma = matrix(1/3, 3, 3))), trait.names = c("bib", "bob"))
-    test <- dads(stop.rule = list("max.living" = 10), traits = complex_traits)
-    out <- capture_output(print.dads(test))
[…]

### 2023-04-24 tests/testthat/test-print.treats.R

@@ -1,0 +1,70 @@
+## Test
+test_that("print.treats works for traits", {
+
+    ## Simple trait
+    simple <- make.traits()
+    out <- capture_output(print.traits.info(simple$main))
+    expect_equal(out[[1]], "1 trait for 1 process (A) with one starting value (0).")
+
+    ## Complex
+    complex <- make.traits(process = c(BM.process, BM.process), n = c(2,3), process.args = list(list(Sigma = diag(2)), list(Sigma = matrix(1/3, 3, 3))), trait.names = c("bib", "bob"))
+    out <- capture_output(print.traits.info(complex$main))
+    expect_equal(out[[1]], "5 traits for 2 processes (bib:2, bob:3) with one starting value (0).\nprocess bib uses the following extra argument: Sigma;\nprocess bob uses the following extra argument: Sigma;")
+
+    ## Simple treats
+    out <- capture_output(print.treats(simple))
+    expect_equal(out[[1]], " ---- treats traits object ---- \n1 trait for 1 process (A) with one starting value (0).")
+
+    background <- make.traits(background = make.traits())
+    out <- capture_output(print.treats(background))
+    expect_equal(out[[1]], " ---- treats traits object ---- \n1 trait for 1 process (A) with one starting value (0).\nAnd a background trait (see x$background for info).")
+})
+
+test_that("print.treats works for treats", {
+
+    ## Simple birth.death tree with one BM trait
+    test <- treats(stop.rule = list("max.living" = 10), traits = make.traits())
+    out <- capture_output(print.treats(test))
+    expect_equal(out[[1]], " ---- treats object ---- \nSimulated diversity data (x$tree):\n\nPhylogenetic tree with 10 tips and 9 internal nodes.\n\nTip labels:\n  t1, t2, t3, t4, t5, t6, ...\nNode labels:\n  n1, n2, n3, n4, n5, n6, ...\n\nRooted; includes branch lengths.\n\nSimulated disparity data (x$data):\n1 trait for 1 process (A) with one starting value (0).")
+    
+    ## A more complex trait
+    complex_traits <- make.traits(process = c(BM.process, BM.process), n = c(2,3), process.args = list(list(Sigma = diag(2)), list(Sigma = matrix(1/3, 3, 3))), trait.names = c("bib", "bob"))
+    test <- treats(stop.rule = list("max.living" = 10), traits = complex_traits)
+    out <- capture_output(print.treats(test))
[…]

### 2023-04-24 tests/testthat/test-treats.R

@@ -1,31 +1,31 @@
 ## Test
-test_that("dads works for simple birth-deaths", {
+test_that("treats works for simple birth-deaths", {
 
     ## Naked birth death (just the tree)
 
     ## Sanitizing
-    error <- capture_error(dads(bd.params = "hehe", stop.rule = list(max.taxa = 10)))
-    expect_equal(error[[1]], "bd.params must be a named list of arguments or a \"dads\" \"bd.params\" object. You can use make.bd.params() to format the object correctly.")
-    error <- capture_error(dads(bd.params = list(1, 2), stop.rule = list(max.taxa = 10)))
-    expect_equal(error[[1]], "bd.params must be a named list of arguments or a \"dads\" \"bd.params\" object. You can use make.bd.params() to format the object correctly.")
+    error <- capture_error(treats(bd.params = "hehe", stop.rule = list(max.taxa = 10)))
+    expect_equal(error[[1]], "bd.params must be a named list of arguments or a \"treats\" \"bd.params\" object. You can use make.bd.params() to format the object correctly.")
+    error <- capture_error(treats(bd.params = list(1, 2), stop.rule = list(max.taxa = 10)))
+    expect_equal(error[[1]], "bd.params must be a named list of arguments or a \"treats\" \"bd.params\" object. You can use make.bd.params() to format the object correctly.")
 
-    error <- capture_error(dads())
+    error <- capture_error(treats())
     expect_equal(error[[1]], "You must provide at least one stopping rule. For example:\nstop.rule <- list(max.taxa   = 10,\n                  max.living = 10,\n                  max.time   = 10)")
-    error <- capture_error(dads(stop.rule = "ha"))
+    error <- capture_error(treats(stop.rule = "ha"))
     expect_equal(error[[1]], "stop.rule must be of class list.")
-    error <- capture_error(dads(stop.rule = list(1,2)))
+    error <- capture_error(treats(stop.rule = list(1,2)))
     expect_equal(error[[1]], "stop.rule must be a named list of stopping rules. For example:\nstop.rule <- list(max.taxa   = 10,\n                  max.living = 10,\n                  max.time   = 10)")
-    error <- capture_error(dads(stop.rule = list(something = 2)))
+    error <- capture_error(treats(stop.rule = list(something = 2)))
[…]

### 2023-04-24 paper/treats_paper.tex
rename paper

--- 2023-01-19 paper/dads_paper.tex
@@
 
-\newcommand{\dads}{\texttt{dads} }
+\newcommand{\treats}{\texttt{treats} }
 
@@
 \bigskip
-\noindent RH: \dads package.
+\noindent RH: \treats package.
 
@@
 
-\noindent{\Large \bf \dads: a modular \texttt{R} package for simulating diversity and traits data.} 
+\noindent{\Large \bf \treats: a modular \texttt{R} package for simulating diversity and traits data.} 
 \bigskip
@@
 
-        \item Here I present the \dads package, a modular \texttt{R} package for diversity and trait disparity simulations.
+        \item Here I present the \treats package, a modular \texttt{R} package for diversity and trait disparity simulations.
         This package is based on a standard birth death algorithm that can be easily tuned to the user specific needs by designing their specific trait process, birth-death modifiers and events.
-        It also provides a tidy interface through the \dads object, allowing users to easily run reproducible simulations.
+        It also provides a tidy interface through the \treats object, allowing users to easily run reproducible simulations.
 
-        \item The \dads package also comes with an extend manual regularly updated following users' questions or suggestions.
+        \item The \treats package also comes with an extend manual regularly updated following users' questions or suggestions.
 
@@
 
-% The solution with dads
-Therefore, we propose \dads a modular \texttt{R} package to simulate both diversity and disparity through time.
+% The solution with treats
+Therefore, we propose \treats a modular \texttt{R} package to simulate both diversity and disparity through time.
 
@@
 % Very briefly how does it work
-\dads is based on a core birth-death process implemented in the function \texttt{dads} that simulates diversity through time using the following steps:
+\treats is based on a core birth-death process implemented in the function \texttt{treats} that simulates diversity through time using the following steps:
 
@@
 
-In \dads, these three or four steps are implemented as modular functions that the user can easily change using an internal class of objects called \texttt{``modifiers''} or \texttt{``traits''} (Fig.
+In \treats, these three or four steps are implemented as modular functions that the user can easily change using an internal class of objects called \texttt{``modifiers''} or \texttt{``traits''} (Fig.
 \ref{Fig:workflow}).
-The simulation then outputs a tree (of class \texttt{``phylo''} and a associated table of traits - \texttt{``matrix''}) that can be visualised using the \texttt{plot.dads} function.
+The simulation then outputs a tree (of class \texttt{``phylo''} and a associated table of traits - \texttt{``matrix''}) that can be visualised using the \texttt{plot.treats} function.
 
@@
 \centering
-   \includegraphics[width=1\textwidth]{../inst/gitbook/dads_structure.pdf} 
-\caption{\dads package workflow: @@@.}
+   \includegraphics[width=1\textwidth]{../inst/gitbook/treats_structure.pdf} 
+\caption{\treats package workflow: @@@.}
 \label{Fig:workflow}
@@
 \texttt{## Running a simulation with a speciation rate of 1, an extinction rate of 0.1 and stopping when it reaches 20 taxa}
-\texttt{my_simulation <- dads(bd.params = list(speciation = 1, extinction = 0.1), stop.rule = list(max.taxa = 20))}
+\texttt{my_simulation <- treats(bd.params = list(speciation = 1, extinction = 0.1), stop.rule = list(max.taxa = 20))}
 
@@
 
-These are included by default in a \texttt{dads} process but can be also set manually by creating a \texttt{``modifiers''} object with no options:
+These are included by default in a \texttt{treats} process but can be also set manually by creating a \texttt{``modifiers''} object with no options:
 
@@
 \texttt{## Running a simulation with a speciation rate of 1, an extinction rate of 0.1 and stopping when it reaches 20 taxa with the default modifiers}
-\texttt{my_simulation <- dads(bd.params = list(speciation = 1, extinction = 0), stop.rule = list(max.taxa = 20), modifiers = my_modifiers)}
+\texttt{my_simulation <- treats(bd.params = list(speciation = 1, extinction = 0), stop.rule = list(max.taxa = 20), modifiers = my_modifiers)}
 
@@
 
-\subsection{\dads compared to other packages}
+\subsection{\treats compared to other packages}
 
@@
 \section{Conclusion}
-The \dads is modular and nice to use.
+The \treats is modular and nice to use.
 
 \section{Package location}
-The \dads package is available on the CRAN at \url{https://cran.r-project.org/web/packages/dads/index.html} or on GitHub at \url{https://github.com/TGuillerme/dads} with more associated information.
+The \treats package is available on the CRAN at \url{https://cran.r-project.org/web/packages/treats/index.html} or on GitHub at \url{https://github.com/TGuillerme/treats} with more associated information.
 All the versions of the package are archived on ZENODO with associated DOI \url{https://zenodo.org/@@@}.

### 2023-04-27 inst/gitbook/04_make.events.Rmd
Updated CRAN checks OK

@@
  * if the target is `"taxa"` you can modify the lineage tracker by removing living species using `random.extinction` or `trait.extinction`.
- * if the target is `"bd.params"` you can modify the [`bd.params` object](#makebdparams) using `update.bd.params`.
- * if the target is `"traits"` you can modify the [`traits` object](#maketraits) using `update.traits`.
- * if the target is `"modifiers"` you can modify the [`modifiers` object](#makemodifiers) using `update.modifiers`.
+ * if the target is `"bd.params"` you can modify the [`bd.params` object](#makebdparams) using `bd.params.update`.
+ * if the target is `"traits"` you can modify the [`traits` object](#maketraits) using `traits.update`.
+ * if the target is `"modifiers"` you can modify the [`modifiers` object](#makemodifiers) using `modifiers.update`.
  * (for the `"founding"` target [see below](#founding))
@@
             condition = taxa.condition(10, condition = `>=`),
-            modification = update.bd.params(extinction = 1/3))
+            modification = bd.params.update(extinction = 1/3))
 ```
@@
                   condition = trait.condition(3, condition = `>`, what = upper.95),
-                  modification = update.traits(process = OU.process))
+                  modification = traits.update(process = OU.process))
 

### 2023-04-27 inst/gitbook/07_specific_examples.Rmd
Updated CRAN checks OK

@@
 `taxa.condition` | [background extinction](#EGbg_ext), [change branch length](#EG_modify_brlen)
-`update.bd.params` | [background extinction](#EGbg_ext), [reducing speciation rate](#EGred_spec)
-`update.traits` | [changing trait process](#EG_change_trait), [changing trait correlation](#EG_change_correlation)
+`bd.params.update` | [background extinction](#EGbg_ext), [reducing speciation rate](#EGred_spec)
+`traits.update` | [changing trait process](#EG_change_trait), [changing trait correlation](#EG_change_correlation)
 `trait.condition` | [changing trait correlation](#EG_change_correlation), [generating a subtree with a different process](EG_founding_traits)
-`update.modifiers` | [changing modifiers](#EG_change_modif), [change branch length](#EG_modify_brlen)
+`modifiers.update` | [changing modifiers](#EG_change_modif), [change branch length](#EG_modify_brlen)
 `parent.traits` | [changing modifiers](#EG_change_modif)
@@
 
-And our change in extinction rate event which triggers after reaching 30 taxa (`taxa.condition(30)`), targets the `"bd.params"` (birth-death parameters) sets the extinction rate to 0.5 (`update.bd.params(extinction = 0.5)`):
+And our change in extinction rate event which triggers after reaching 30 taxa (`taxa.condition(30)`), targets the `"bd.params"` (birth-death parameters) sets the extinction rate to 0.5 (`bd.params.update(extinction = 0.5)`):
 
@@
                       target = "bd.params",
-                      modification = update.bd.params(extinction = 0.5))
+                      modification = bd.params.update(extinction = 0.5))
 ```
@@
                       target = "bd.params",
-                      modification = update.bd.params(speciation = 1/3))
+                      modification = bd.params.update(speciation = 1/3))
 ```
@@
                   target       = "traits",
-                  modification = update.traits(process = OU.process))
+                  modification = traits.update(process = OU.process))
 ```
@@
     target       = "traits",
-    modification = update.traits(process.args = list(Sigma = new_vcv)))
+    modification = traits.update(process.args = list(Sigma = new_vcv)))
 ```
@@
     target       = "modifiers",
-    modification = update.modifiers(speciation = speciation,
+    modification = modifiers.update(speciation = speciation,
                                     condition  = negative.trait.condition,
@@
                   target = "modifiers",
-                  modification = update.modifiers(
+                  modification = modifiers.update(
                                     branch.length = branch.length,

### 2023-04-27 inst/gitbook/not_include/05_thorough_example.Rmd
Updated CRAN checks OK

@@
 ## The correlation change event
-update.correlation <- update.traits(process.args =list(Sigma = matrix(c(10,3,3,2),2,2)))
+update.correlation <- traits.update(process.args =list(Sigma = matrix(c(10,3,3,2),2,2)))
 
@@
                            condition    = time.condition(2),
-                           modification = update.traits(
+                           modification = traits.update(
                     process.args =list(Sigma = matrix(c(10,3,3,2),2,2)),

### 2023-04-27 R/events.conditions.R

@@ -1,13 +1,10 @@
 #' @name events.conditions
-#' @aliases condition, time.condition, taxa.condition, trait.condition
+#' @aliases time.condition taxa.condition trait.condition
 #' @title events.conditions
 #'
 #' @description Inbuilt conditions functions for helping designing events
 #'
-#' @usage condition(x, condition, ...)
-#' @usage taxa.condition(x, condition, ...)
-#' @usage time.condition(x, condition, ...)
-#' @usage trait.condition(x, condition, ...)
+#' @usage events.condition(x, condition, ...)
 #'
 #' @param x         the variable to reach for satisfying a condition (see details)
 #' @param condition the logical function for triggering the condition (e.g. `<`, `==`, `!>`, etc...).
@@ -66,7 +63,7 @@
 #'                   target       = "traits",
 #'                   ## condition is triggered if(upper.95(x) > 3)
 #'                   condition    = trait.condition(3, condition = `>`, what = upper.95),
-#'                   modification = update.traits(process = OU.process))
+#'                   modification = traits.update(process = OU.process))
 #' 
 #' ## Set the simulation parameters
 #' bd.params <- list(extinction = 0, speciation = 1)
@@ -94,7 +91,7 @@
 #' @author Thomas Guillerme
 
 ## The list of conditions
-condition <- function(x, condition) {
+events.condition <- function(x, condition, ...) {
     cat("List of inbuilt condition functions in treats:\n")
     cat("   ?taxa.condition\n")
     cat("   ?time.condition\n")

### 2023-04-27 R/events.modifications.R

@@ -1,43 +1,13 @@
 #' @name events.modifications
-#' @aliases modification, random.extinction, trait.extinction, update.bd.params, update.traits, update.modifiers, founding.event
-#' @title events.modifications
+#' @aliases random.extinction trait.extinction bd.params.update traits.update modifiers.update founding.event
+#' @title Events modifications
 #'
 #' @description Inbuilt modifications functions for helping designing events
 #'
-#' @usage modification(x, ...)
-#' @usage founding.event(x, bd.params = NULL, traits = NULL, modifiers = NULL,
-#'                       events = NULL)
-#' @usage random.extinction(x)
-#' @usage trait.extinction(x, condition = `<`, trait = 1)
-#' @usage update.bd.params(x, speciation = NULL, extinction = NULL,
-#'                        joint = NULL, absolute = NULL, speciation.args = NULL,
-#'                        extinction.args = NULL)
-#' @usage update.modifiers(x, branch.length = NULL, selection = NULL,
-#'                         speciation = NULL, condition = NULL, modify = NULL)
-#' @usage update.traits(x, process = NULL, process.args = NULL,
-#'                      trait.names = NULL)
+#' @usage events.modification(x, ...)
 #'
 #' @param x   a numerical value to update.
 #' @param ... any specific argument for the modification (see details).
-#' @param bd.params a \code{"bd.params"} \code{"treats"} object for the founding event (see details for founding.event).
-#' @param traits a \code{"traits"} \code{"treats"} object for the founding event (see details for founding.event).
-#' @param modifiers a \code{"modifiers"} \code{"treats"} object for the founding event (see details for founding.event).
-#' @param events an \code{"events"} \code{"treats"} object for the founding event (see details for founding.event).
-#' @param condition a condition function for the trait value (see details for trait.extinction and update.modifiers).
-#' @param trait which trait to use (see details for trait.extinction).
-#' @param speciation the same argument as in \code{\link{make.bd.params}} (see details for update.bd.params).
-#' @param extinction the same argument as in \code{\link{make.bd.params}} (see details for update.bd.params).
-#' @param joint the same argument as in \code{\link{make.bd.params}} (see details for update.bd.params).
-#' @param absolute the same argument as in \code{\link{make.bd.params}} (see details for update.bd.params).
-#' @param speciation.args the same argument as in \code{\link{make.bd.params}} (see details for update.bd.params).
[…]

### 2023-04-27 R/make.events.R

@@ -52,7 +52,7 @@
 #'                   target       = "traits",
 #'                   ## condition is triggered if(upper.95(x) > 3)
 #'                   condition    = trait.condition(3, condition = `>`, what = upper.95),
-#'                   modification = update.traits(process = OU.process))
+#'                   modification = traits.update(process = OU.process))
 #' 
 #' ## Set the simulation parameters
 #' bd.params <- list(extinction = 0, speciation = 1)

### 2023-04-27 R/modifiers.R

@@ -4,16 +4,16 @@
 #'
 #' @description Different modifiers for the birth death process implemented in treats.
 #'
-#' @usage branch.length(bd.params = NULL, lineage = NULL, trait.values = NULL,
+#' @usage modifiers(bd.params = NULL, lineage = NULL, trait.values = NULL,
 #'                      modify.fun = NULL)
-#' @usage selection(bd.params = NULL, lineage = NULL, trait.values = NULL,
-#'                  modify.fun = NULL)
-#' @usage speciation(bd.params = NULL, lineage = NULL, trait.values = NULL,
-#'                   modify.fun = NULL)
-#' @usage branch.length.trait(bd.params = NULL, lineage = NULL,
-#'                            trait.values = NULL, modify.fun = NULL)
-#' @usage speciation.trait(bd.params = NULL, lineage = NULL,
-#'                         trait.values = NULL, modify.fun = NULL)
+# @usage selection(bd.params = NULL, lineage = NULL, trait.values = NULL,
+#                  modify.fun = NULL)
+# @usage speciation(bd.params = NULL, lineage = NULL, trait.values = NULL,
+#                   modify.fun = NULL)
+# @usage branch.length.trait(bd.params = NULL, lineage = NULL,
+#                            trait.values = NULL, modify.fun = NULL)
+# @usage speciation.trait(bd.params = NULL, lineage = NULL,
+#                         trait.values = NULL, modify.fun = NULL)
 #'
 #' @param bd.params      A named list of birth death parameters (see details).
 #' @param lineage        A named list containing the lineage data (see details).

### 2023-04-27 tests/testthat/test-birth.death.R

@@ -285,7 +285,7 @@ test_that("events work", {
     events <- make.events(
         condition    = taxa.condition(30),
         target       = "bd.params",
-        modification = update.bd.params(extinction = 1/3))
+        modification = bd.params.update(extinction = 1/3))
     ## Testing the results
     set.seed(2)
     test <- birth.death.tree.traits(bd.params = bd.params, stop.rule = stop.rule, traits = NULL, modifiers = NULL, events = events)
@@ -302,7 +302,7 @@ test_that("events work", {
     events <- make.events(
         condition    = time.condition(2),
         target       = "bd.params",
-        modification = update.bd.params(speciation = 1/3))
+        modification = bd.params.update(speciation = 1/3))
     
     ## Updating the stop.rule
     stop.rule$max.living <- Inf
@@ -330,7 +330,7 @@ test_that("events work", {
     events <- make.events(
         condition    = time.condition(5),
         target       = "traits",
-        modification = update.traits(process = OU.process))
+        modification = traits.update(process = OU.process))
     
     set.seed(1)
     test <- birth.death.tree.traits(bd.params = bd.params, stop.rule = stop.rule, traits = traits, modifiers = NULL, events = NULL)
@@ -349,7 +349,7 @@ test_that("events work", {
     change_correlation <- make.events(
         condition    = trait.condition(3, absolute = TRUE),
         target       = "traits",
-        modification = update.traits(process.args = list(Sigma = matrix(c(10,3,3,2),2,2))))
+        modification = traits.update(process.args = list(Sigma = matrix(c(10,3,3,2),2,2))))
 
     stop.rule$max.living <- Inf
     stop.rule$max.time <- Inf
@@ -386,7 +386,7 @@ test_that("events work", {
     events <- make.events(
         condition    = time.condition(3),
         target       = "modifiers",
-        modification = update.modifiers(speciation = speciation, condition = new.condition, modify = new.modify))
+        modification = modifiers.update(speciation = speciation, condition = new.condition, modify = new.modify))
 
     set.seed(4)
     test <- birth.death.tree.traits(bd.params = bd.params, stop.rule = stop.rule, traits = traits, modifiers = modifiers, events = NULL)
@@ -410,7 +410,7 @@ test_that("events work", {
     events <- make.events(
         condition    = taxa.condition(30),
         target       = "modifiers",
-        modification = update.modifiers(branch.length = branch.length, modify = new.modify))
[…]

### 2023-04-27 tests/testthat/test-make.events.R

@@ -3,7 +3,7 @@ test_that("make.events works", {
 
     condition.test.ok <- time.condition(5)
     condition.test.wrong <- "bob"
-    modification.test.ok <- update.traits(process = OU.process)
+    modification.test.ok <- traits.update(process = OU.process)
     modification.test.wrong <- "bub"
 
     error <- capture_error(make.events(target = "target"))

### 2023-04-27 inst/gitbook/01_getting_started.Rmd
fixed a couple of comments to the manual

@@
 # Getting started {#gettingstarted}
+
+<!-- ## An illustrative example -->
 
@@
 First we need to create a `condition` function to trigger this (non) speciation event.
-We can do that by specifying our `condition` function (here the `staying.alive` function) to apply our modification.
+We can do that by specifying our `condition` function (here the `going.extinct` function) to apply our modification.
 For that we can use the `parent.traits` utility function that is optimised for accessing traits in the birth-death process (but you can of course write your own).
@@
 
-We can then provide these two functions (the condition `negative.ancestor` and how to modify the speciation event when this condition is met `staying.alive`).
+We can then provide these two functions (the condition `negative.ancestor` and how to modify the speciation event when this condition is met `going.extinct`).
 If you are an advances `treats` user, you can design your own `speciation` function but if you just want to use a normal `speciation` function, you can use the default one from `treats` called `speciation`.
@@
 ## Simulate disparity and diversity
-biased_data <- treats(bd.params = extinction_02,
+modified_data <- treats(bd.params = extinction_02,
                         stop.rule = living_20,
@@
                         modifiers = negatives_extinct)
-biased_data
+modified_data
 ```
@@
 plot(default_data, main = "default results")
-plot(biased_data, main = "results with the modifier")
+plot(modified_data, main = "results with the modifier")
 ```

### 2023-04-27 inst/gitbook/02_make.traits.Rmd
fixed a couple of comments to the manual

@@
 This object is generated using `make.traits`.
-
-> **What is a trait in biology?** This is one of these epistemological questions like what _is_ a species and is not covered in this package.
-For the purpose of _this manual_, a **biological trait** can be any coherent character of any number of dimensions which may or may not be independent of other traits.
-Basically anything can be a trait if it's defined by the user (e.g. one trait can be a 20 dimensional Brownian process that's correlated to a 1D OU process - which is another trait).
-For the purpose of the syntax in _this package_ a **trait in `treats`** is a column in the output trait matrix.
-For example, one might be interested in a biological trait that is "beak shape" which is defined as one complex biological trait that is the Procrustes superimposition of 4 3D landmarks and 300 3D semi-landmarks.
-This single "beak shape" **biological trait** is then composed of 912 **traits in `treats`** (i.e. a matrix of $(4+300)\times3$ columns).
+In `treats` a trait designate any coherent character of any number of dimensions which may or may not be independent of other traits.
+For example, it can be a uni-dimensional variable uncorrelated to other traits or a multi-dimensional one with different amounts of correlation.
 
@@
 
-Note that the number of traits are distributed per processes.
-If the traits contains multiple processes, the number of traits are distributed per processes:
+Note that the number of traits are applied for each process by default.
+However, you can apply different number of traits for different processes:
 

### 2023-04-27 inst/gitbook/03_make.modifiers.Rmd
fixed a couple of comments to the manual

@@
 ## A modifier for removing tips with negative values
-bias.trait <- function(trait.values, lineage) {
+modify.trait <- function(trait.values, lineage) {
     if(parent.traits(trait.values, lineage) < 0) {
@@
 ## Creating the modifier
-biased_trait <- make.modifiers(speciation = bias.trait)
+modified_trait <- make.modifiers(speciation = modify.trait)
 
@@
 set.seed(1)
-biased_trait_treats <- treats(bd.params = bd_params,
+modified_trait_treats <- treats(bd.params = bd_params,
                           stop.rule = stop_rule,
                           traits    = BM_trait,
-                          modifiers = biased_trait)
+                          modifiers = modified_trait)
 
@@
 plot(default_treats, main = "Default trait and tree")
-plot(biased_trait_treats, main = "Biased trait and tree")
+plot(modified_trait_treats, main = "Biased trait and tree")
 ```

### 2023-04-28 inst/gitbook/verbose_example.Rmd
Update verbose_example.Rmd

@@
 ```{r}
-libary(treats)
+library(treats)
 my_params <- make.bd.params(speciation = rnorm)

### 2023-04-28 inst/gitbook/index.Rmd
Verbose example much improved
To finish!

@@
 
-# `treats`: disparity and diversity simulations.
+# `treats`: trees and traits simulations.
 
-Simultaneously simulate diversity (i.e. phylogenetic trees) and disparity  diversity (and the interaction between them).
+Simultaneously simulate phylogenetic trees and traits (and the interaction between them).
 
@@
 
-<!-- %NC: I think here you need an example of use case as it's not immediately obvious what the package is for... -->
+Ar Macarface is studying the evolution of mammalian disparity (i.e. diversity of shapes) across the Cretaceous-Palaeogene mass extinction event (K-Pg - 66 million years ago).
+They observe a decrease of morphological disparity after the extinction event followed by a sharp increase of disparity.
+However Ar is also prudent and wants to test whether their observed pattern is just an artifact of the data or actually showing some effect of the K-Pg extinction on mammalian evolution.
+First they need to check whether changes in disparity can actually be discernables due to an extinction event (i.e.
+if an extinction event happens, would it be possible to detect a change in disparity)?
+Second they want to know whether these changes are due to changes are due to the mass extinction event being selective or not.
+
+The [getting started](#kpgexample) section actually shows how do implement this example in `treats` in details.
 

### 2023-04-28 inst/gitbook/01_getting_started.Rmd
Update 01_getting_started.Rmd

@@
 <!-- ## An illustrative example -->
+## A verbose example: mammals disparity through the K-Pg extinction {kpgexample}
+
+Ar Macarface is studying the evolution of mammalian disparity (i.e. diversity of shapes) across the Cretaceous-Palaeogene mass extinction event (K-Pg - 66 million years ago).
+
+They use a dataset from @BeckLee that is a discrete morphological space containing the ordination of the around 400 morphological characters for 50 species and estimated for their 49 descendants into an ordinated shapespace of 99 elements and 97 dimensions.
+
+```{r}
+library(dispRity)
+## Loading the mammalian morphological example data
+data(BeckLee_mat99)
+
+## This is what the dataset looks like
+head(BeckLee_mat99)[1:5, 1:5]
+```
+
+And a tree showing the relation between these species through time:
+
+```{r}
+## The phylogenetic tree
+data(BeckLee_tree)
+
+## And what the tree looks like
+plot(BeckLee_tree, cex = 0.8)
+axisPhylo()
+abline(v = BeckLee_tree$root.time - 66, col = "red")
+```
+
+They measure the disparity as the sum of variances as a proxy for changes in the traitspace size (@moms) with some bootstrapping to create some confidence intervals around that curve.
+
+```{r}
+## Making continuous time series
+time_series <- chrono.subsets(data = BeckLee_mat99, tree = BeckLee_tree, method = "continuous", model = "proximity", time = seq(from = 101, to = 41, by = -5))
+## Calculating disparity
+change_in_size <- dispRity(boot.matrix(time_series), metric = c(sum, variances))
+plot(change_in_size)
+abline(v = 8, col = "red")
+```
+
+Looking at this curve, Ar wants to know whether disparity increased as a response to the K-Pg mass extinction or not.
+To do this, they need to check whether changes in disparity can actually be detected due to an extinction event (e.g. if an extinction event happens, does that change disparity curves?).
+As a subsidiary question, they want to know whether the pattern is more likely due to a selective or random extinction event.
+
+### Simulating trees and traits using the observed parameters
+
+The first step is to simulate the observed tree.
+We can do that by using crude parameters from the observed tree:
+
+```{r}
+## Extracting the crude parameters
+(est_params <- crude.bd.est(BeckLee_tree))
+## These parameters are the observed speciation and extinction rate per million years.
+
+## Setting the stopping rule (stop when reaching a total of 50 taxa)
+stop_rule <- list(max.taxa = 50)
+
+## Simulating just a birth-death tree with these parameters to check
+set.seed(1)
+test_tree <- treats(bd.params  = est_params,
+                    stop.rule  = stop_rule,
+                    null.error = 100)
+plot(ladderize(test_tree), cex = 0.8)
+axisPhylo()
+```
+
+> Note that because of the high extinction to speciation ratio, the tree simulations can often fail before reaching the stop rule (i.e.
+all lineages go extinct before reaching 50 taxa).
+To avoid getting stuck in there, we use the option `null.error = 100` to rerun the simulations up to 100 times before obtaining a tree that did not went fully extinct.
+Also note that using such parameters make the simulations run slowlier but ensure the requested results are always achieved.
+
+We now have a way to simulate a topology close to the observed one. We can simulate a list of topologies by replicating this function X amounts of time to get a distribution of trees:
+
+```{r, eval = FALSE, message = FALSE}
+## Generating 50 trees
+tree_distribution <- treats(bd.params  = est_params,
+                            stop.rule  = stop_rule,
+                            null.error = 100,
+                            replicates = 50)
+```
+
+Of course, here our objective is to also simulate some trait values to measure disparity under some simulated scenario.
+We can do that be adding a `"traits"` object to the `treats` function.
+Here we will use a default [Brownian Motion](https://en.wikipedia.org/wiki/Brownian_motion) to simulate our trait.
+
+```{r}
+## Creating a trait in 97 dimensions.
+my_traits <- make.traits(process = BM.process, n = ncol(BeckLee_mat99))
+```
+
+> Note that there are many more options on how to simulate your trait such as using correlation, different processes, etc... See details in the [make traits section](#maketraits).
+
+```{r, message = FALSE}
+## Simulate the tree and traits
+sim_data <- treats(traits     = my_traits,
+                   bd.params  = est_params,
+                   stop.rule  = stop_rule,
+                   null.error = 100)
+```
+
+We can visualise the resulting tree using the `plot` function and visualising the different traits using the `trait` option in `plot`:
+
+```{r}
+## Plotting the results (first trait)
+plot(sim_tree, trait = 1)
+```
+
+Of course, here we want to simulate a distribution of trees and traits:
+
+```{r, message = FALSE}
+## Simulate the tree and traits
+sim_data <- treats(traits     = my_traits,
+                   bd.params  = est_params,
+                   stop.rule  = stop_rule,
+                   null.error = 100,
+                   replicates = 50)
+```
+
+Et voila! We now have a distribution of 50 trees and 50 datasets from which we can calculate disparity.
+We can convert this list into `dispRity` using the `treats.dispRity` function.
+
+```{r}
+## TODO: make this automatically do this:
+# time_series <- chrono.subsets(data = dispRitreats(sim_data)$data, tree = dispRitreats(sim_data)$tree, method = "continuous", model = "proximity", time = 13)
+
+## Converting the data to be passed to dispRity functions
+dispRity_data <- dispRitreats(sim_data)
+
+## Creating the time series
+time_series <- chrono.subsets(data = dispRity_data$data, tree = dispRity_data$tree, time = seq(from = 0.9, to = 0, by = -0.1), method = "continuous", model = "proximity")
+
+simulated_disparity <- dispRity(time_series, metric = c(sum, variances), verbose = TRUE)
+
+```
 

### 2023-05-02 inst/gitbook/01_getting_started.Rmd
Update 01_getting_started.Rmd

@@
 ## Making continuous time series
-time_series <- chrono.subsets(data = BeckLee_mat99, tree = BeckLee_tree, method = "continuous", model = "proximity", time = seq(from = 101, to = 41, by = -5))
+time_series <- chrono.subsets(data = BeckLee_mat99,
+                              tree = BeckLee_tree,
+                              method = "continuous", model = "proximity",
+                              time = seq(from = 101, to = 41, by = -5))
 ## Calculating disparity
@@
 ```
+
+> For more info about `chrono.subsets` and the `dispRity` package, you can have a look at the [`dispRity` manual](http://tguillerme.github.io/dispRity.html).
 
@@
 ## Plotting the results (first trait)
-plot(sim_tree, trait = 1)
+plot(sim_data, trait = 1)
 ```
@@
 ```{r}
-## TODO: make this automatically do this:
-# time_series <- chrono.subsets(data = dispRitreats(sim_data)$data, tree = dispRitreats(sim_data)$tree, method = "continuous", model = "proximity", time = 13)
-
 ## Converting the data to be passed to dispRity functions
@@
 
+## Creating the time series (by default dispRitreats scales the trees to be 1 unit of times old)
+time_series <- chrono.subsets(data = dispRity_data$data,
+                              tree = dispRity_data$tree,
+                              time = seq(from = 0.9, to = 0, by = -0.1),
+                              method = "continuous", model = "proximity")
+
+## Calculating the disparity from the simulated data
+simulated_disparity <- dispRity(time_series, metric = c(sum, variances),
+                                verbose = TRUE)
+par(mfrow = c(1,2))
+plot(simulated_disparity, main = "simulated disparity")
+plot(change_in_size, main = "observed disparity")
+```
+
+We can see that both simulated and observed disparity curves are different.
+From there one could already reasonably propose that the observe disparity does not entirely follows our parametrised scenario (constant speciation and extinction and BM trait evolution).
+
+However, we can complexify our scenario to introduce the effect of mass extinction event.
+We can simulate two alternative scenarios:
+ * one where we create a random extinction event after having simulated half of the species (25) where we remove 75% of species;
+ * and a second one where we remove all the species with negative values.
+This allows us to simulate two contrasted scenarios to see whether using our observed dataset it would be possible to detect changes in disparity due to mass extinctions.
+
+To do so, we need to create two mass extinction `"event"` objects:
+
+```{r}
+## Creating a random mass extinction
+random_extinction <- make.events(target = "taxa",
+                                 condition = taxa.condition(25, condition = `>=`),
+                                 modification = random.extinction(0.75))
+## Creating an extrinction that removes species with negative trait values
+negative_extinction <- make.events(target = "taxa",
+                                   condition = taxa.condition(25, condition = `>=`),
+                                   modification = trait.extinction(x = 0, condition = `>=`))
+```
+
+We can then feed these two extinction events to our previous simulations
+
+```{r}
+## Simulate the tree and traits with a random extinction event
+sim_rand_extinction <- treats(
+                   traits     = my_traits,
+                   bd.params  = est_params,
+                   stop.rule  = stop_rule,
+                   events     = random_extinction,
+                   null.error = 100,
+                   replicates = 50)
+
+## Simulate the tree and traits with a random extinction event
+sim_trait_extinction <- treats(
+                   traits     = my_traits,
+                   bd.params  = est_params,
+                   stop.rule  = stop_rule,
+                   events     = negative_extinction,
+                   null.error = 100,
+                   replicates = 50)
+
+## Visualising the difference between both scenarios
+par(mfrow = c(1,2))
+plot(sim_rand_extinction[[1]], main = "Random extinction")
+plot(sim_trait_extinction[[1]], main = "Negative extinction")
+```
+
+We can then measure the disparity from these scenarios and compare them to our observed ones
+
+```{r}
+## Converting the data to be passed to dispRity functions
+rand_ext_data <- dispRitreats(sim_rand_extinction)
+trait_ext_data <- dispRitreats(sim_trait_extinction)
+
 ## Creating the time series
-time_series <- chrono.subsets(data = dispRity_data$data, tree = dispRity_data$tree, time = seq(from = 0.9, to = 0, by = -0.1), method = "continuous", model = "proximity")
+rand_ext_time_series <- chrono.subsets(data = rand_ext_data$data,
+                                       tree = rand_ext_data$tree,
+                                       time = seq(from = 0.9, to = 0, by = -0.1),
+                                       method = "continuous", model = "proximity")
+trait_ext_time_series <- chrono.subsets(data = trait_ext_data$data,
+                                        tree = trait_ext_data$tree,
+                                        time = seq(from = 0.9, to = 0, by = -0.1),
+                                        method = "continuous", model = "proximity")
 
-simulated_disparity <- dispRity(time_series, metric = c(sum, variances), verbose = TRUE)
+## Calculating the disparity from the simulated data
+sim_rand_ext_disp <- dispRity(rand_ext_time_series, metric = c(sum, variances),
+                              verbose = TRUE)
+sim_trait_ext_disp <- dispRity(trait_ext_time_series, metric = c(sum, variances),
+                               verbose = TRUE)
 
+par(mfrow = c(2,2))
+plot(change_in_size, main = "observed disparity")
+plot(simulated_disparity, main = "simulated disparity")
+plot(sim_rand_ext_disp, main = "simulated disparity\n(mass extinction)")
+plot(simulated_disparity, main = "simulated disparity\n(trait extinction)")
 ```
+
+Interestingly, with all our parameters used here and our specific disparity metric we cannot detect the effect of a mass extinction.
 

### 2023-05-09 inst/gitbook/01_getting_started.Rmd
updated time warning for low spec rate

@@
 
-## Setting the stopping rule (stop when reaching a total of 50 taxa)
-stop_rule <- list(max.taxa = 50)
+## Setting the stopping rule (stop after 30 time units)
+stop_rule <- list(max.time = 30)
 
@@
 
-Interestingly, with all our parameters used here and our specific disparity metric we cannot detect the effect of a mass extinction.
+Interestingly, with all our parameters used here and our specific disparity metric we cannot detcet the effect of a mass extinction.
 

### 2023-05-09 R/treats.R

@@ -116,11 +116,6 @@ treats <- function(stop.rule, bd.params, traits = NULL, modifiers = NULL, events
     stop.rule$max.time    <- ifelse(is.null(stop.rule$max.time),   Inf, stop.rule$max.time)
 
     ## Warning message for stop rule
-    if(stop.rule$max.time != Inf && stop.rule$max.time > 6) {
-        warning_message <- "Stop rule was step higher than 6 time units.\nThis can take a considerable amount of time and RAM to simulate"
-        additional_message <- ifelse((stop.rule$max.taxa == Inf && stop.rule$max.living == Inf), " if no other stop rules are given", "")
-        message(paste0(warning_message, additional_message, "."))
-    }
 
     ## bd.params
     if(missing(bd.params)) {
@@ -141,6 +136,12 @@ treats <- function(stop.rule, bd.params, traits = NULL, modifiers = NULL, events
             bd.params <- make.bd.params(speciation = ifelse(is.null(bd.params$speciation), 1, bd.params$speciation),
                                         extinction = ifelse(is.null(bd.params$extinction), 0, bd.params$extinction))
         }
+    }
+
+    if(stop.rule$max.time != Inf && stop.rule$max.time > 6 && bd.params$speciation() > 0.5) {
+        warning_message <- "Stop rule was step higher than 6 time units.\nThis can take a considerable amount of time and RAM to simulate"
+        additional_message <- ifelse((stop.rule$max.taxa == Inf && stop.rule$max.living == Inf), " if no other stop rules are given", "")
+        message(paste0(warning_message, additional_message, "."))
     }
 
     ## traits

### 2023-05-09 tests/testthat/test-birth.death.R

@@ -494,7 +494,7 @@ test_that("events work", {
     expect_equal(sum(tree.age(test1$tree)$ages == 0), 50)
 
     set.seed(8)
-    test2 <- birth.death.tree.traits(bd.params = bd.params, stop.rule = stop.rule.living, traits = NULL, modifiers = NULL, events = events)
+    expect_warning(test2 <- birth.death.tree.traits(bd.params = bd.params, stop.rule = stop.rule.living, traits = NULL, modifiers = NULL, events = events))
     expect_is(test2$tree, "phylo")
     expect_equal(Ntip(test2$tree), 59)
     expect_equal(sum(tree.age(test2$tree)$ages == 0), 48)
@@ -548,7 +548,7 @@ test_that("events work", {
     expect_equal(dim(test1$data), c(50+49, 1))
 
     set.seed(19)
-    test2 <- birth.death.tree.traits(bd.params = bd.params, stop.rule = stop.rule.taxa, traits = traits, modifiers = NULL, events = events)
+    expect_warning(test2 <- birth.death.tree.traits(bd.params = bd.params, stop.rule = stop.rule.taxa, traits = traits, modifiers = NULL, events = events))
     expect_is(test2$tree, "phylo")
     expect_equal(Ntip(test2$tree), 49)
     expect_is(test2$data, c("matrix", "array"))
@@ -565,7 +565,7 @@ test_that("events work", {
     expect_equal(dim(test1$data), c(68+67, 1))
 
     set.seed(20)
-    test2 <- birth.death.tree.traits(bd.params = bd.params, stop.rule = stop.rule.living, traits = traits, modifiers = NULL, events = events)
+    expect_warning(test2 <- birth.death.tree.traits(bd.params = bd.params, stop.rule = stop.rule.living, traits = traits, modifiers = NULL, events = events))
     expect_is(test2$tree, "phylo")
     expect_equal(Ntip(test2$tree), 52)
     expect_equal(sum(tree.age(test2$tree)$ages == 0), 45)

### 2023-05-10 inst/gitbook/01_getting_started.Rmd
Update 01_getting_started.Rmd

@@
 # Getting started {#gettingstarted}
-
-<!-- ## An illustrative example -->
-## A verbose example: mammals disparity through the K-Pg extinction {kpgexample}
-
-Ar Macarface is studying the evolution of mammalian disparity (i.e. diversity of shapes) across the Cretaceous-Palaeogene mass extinction event (K-Pg - 66 million years ago).
-
-They use a dataset from @BeckLee that is a discrete morphological space containing the ordination of the around 400 morphological characters for 50 species and estimated for their 49 descendants into an ordinated shapespace of 99 elements and 97 dimensions.
-
-```{r}
-library(dispRity)
-## Loading the mammalian morphological example data
-data(BeckLee_mat99)
-
-## This is what the dataset looks like
-head(BeckLee_mat99)[1:5, 1:5]
-```
-
-And a tree showing the relation between these species through time:
-
-```{r}
-## The phylogenetic tree
-data(BeckLee_tree)
-
-## And what the tree looks like
-plot(BeckLee_tree, cex = 0.8)
-axisPhylo()
-abline(v = BeckLee_tree$root.time - 66, col = "red")
-```
-
-They measure the disparity as the sum of variances as a proxy for changes in the traitspace size (@moms) with some bootstrapping to create some confidence intervals around that curve.
-
-```{r}
-## Making continuous time series
-time_series <- chrono.subsets(data = BeckLee_mat99,
-                              tree = BeckLee_tree,
-                              method = "continuous", model = "proximity",
-                              time = seq(from = 101, to = 41, by = -5))
-## Calculating disparity
-change_in_size <- dispRity(boot.matrix(time_series), metric = c(sum, variances))
-plot(change_in_size)
-abline(v = 8, col = "red")
-```
-
-> For more info about `chrono.subsets` and the `dispRity` package, you can have a look at the [`dispRity` manual](http://tguillerme.github.io/dispRity.html).
-
-Looking at this curve, Ar wants to know whether disparity increased as a response to the K-Pg mass extinction or not.
-To do this, they need to check whether changes in disparity can actually be detected due to an extinction event (e.g. if an extinction event happens, does that change disparity curves?).
-As a subsidiary question, they want to know whether the pattern is more likely due to a selective or random extinction event.
-
-### Simulating trees and traits using the observed parameters
-
-The first step is to simulate the observed tree.
-We can do that by using crude parameters from the observed tree:
-
-```{r}
-## Extracting the crude parameters
-(est_params <- crude.bd.est(BeckLee_tree))
-## These parameters are the observed speciation and extinction rate per million years.
-
-## Setting the stopping rule (stop after 30 time units)
-stop_rule <- list(max.time = 30)
-
-## Simulating just a birth-death tree with these parameters to check
-set.seed(1)
-test_tree <- treats(bd.params  = est_params,
-                    stop.rule  = stop_rule,
-                    null.error = 100)
-plot(ladderize(test_tree), cex = 0.8)
-axisPhylo()
-```
-
-> Note that because of the high extinction to speciation ratio, the tree simulations can often fail before reaching the stop rule (i.e.
-all lineages go extinct before reaching 50 taxa).
-To avoid getting stuck in there, we use the option `null.error = 100` to rerun the simulations up to 100 times before obtaining a tree that did not went fully extinct.
-Also note that using such parameters make the simulations run slowlier but ensure the requested results are always achieved.
-
-We now have a way to simulate a topology close to the observed one. We can simulate a list of topologies by replicating this function X amounts of time to get a distribution of trees:
-
-```{r, eval = FALSE, message = FALSE}
-## Generating 50 trees
-tree_distribution <- treats(bd.params  = est_params,
-                            stop.rule  = stop_rule,
-                            null.error = 100,
-                            replicates = 50)
-```
-
-Of course, here our objective is to also simulate some trait values to measure disparity under some simulated scenario.
-We can do that be adding a `"traits"` object to the `treats` function.
-Here we will use a default [Brownian Motion](https://en.wikipedia.org/wiki/Brownian_motion) to simulate our trait.
-
-```{r}
-## Creating a trait in 97 dimensions.
-my_traits <- make.traits(process = BM.process, n = ncol(BeckLee_mat99))
-```
-
-> Note that there are many more options on how to simulate your trait such as using correlation, different processes, etc... See details in the [make traits section](#maketraits).
-
-```{r, message = FALSE}
-## Simulate the tree and traits
-sim_data <- treats(traits     = my_traits,
-                   bd.params  = est_params,
-                   stop.rule  = stop_rule,
-                   null.error = 100)
-```
-
-We can visualise the resulting tree using the `plot` function and visualising the different traits using the `trait` option in `plot`:
-
-```{r}
-## Plotting the results (first trait)
-plot(sim_data, trait = 1)
-```
-
-Of course, here we want to simulate a distribution of trees and traits:
-
-```{r, message = FALSE}
-## Simulate the tree and traits
-sim_data <- treats(traits     = my_traits,
-                   bd.params  = est_params,
-                   stop.rule  = stop_rule,
-                   null.error = 100,
-                   replicates = 50)
-```
-
-Et voila! We now have a distribution of 50 trees and 50 datasets from which we can calculate disparity.
-We can convert this list into `dispRity` using the `treats.dispRity` function.
-
-```{r}
-## Converting the data to be passed to dispRity functions
-dispRity_data <- dispRitreats(sim_data)
-
-## Creating the time series (by default dispRitreats scales the trees to be 1 unit of times old)
-time_series <- chrono.subsets(data = dispRity_data$data,
-                              tree = dispRity_data$tree,
-                              time = seq(from = 0.9, to = 0, by = -0.1),
-                              method = "continuous", model = "proximity")
-
-## Calculating the disparity from the simulated data
-simulated_disparity <- dispRity(time_series, metric = c(sum, variances),
-                                verbose = TRUE)
-par(mfrow = c(1,2))
-plot(simulated_disparity, main = "simulated disparity")
-plot(change_in_size, main = "observed disparity")
-```
-
-We can see that both simulated and observed disparity curves are different.
-From there one could already reasonably propose that the observe disparity does not entirely follows our parametrised scenario (constant speciation and extinction and BM trait evolution).
-
-However, we can complexify our scenario to introduce the effect of mass extinction event.
-We can simulate two alternative scenarios:
- * one where we create a random extinction event after having simulated half of the species (25) where we remove 75% of species;
- * and a second one where we remove all the species with negative values.
-This allows us to simulate two contrasted scenarios to see whether using our observed dataset it would be possible to detect changes in disparity due to mass extinctions.
-
-To do so, we need to create two mass extinction `"event"` objects:
-
-```{r}
-## Creating a random mass extinction
-random_extinction <- make.events(target = "taxa",
-                                 condition = taxa.condition(25, condition = `>=`),
-                                 modification = random.extinction(0.75))
-## Creating an extrinction that removes species with negative trait values
-negative_extinction <- make.events(target = "taxa",
-                                   condition = taxa.condition(25, condition = `>=`),
-                                   modification = trait.extinction(x = 0, condition = `>=`))
-```
-
-We can then feed these two extinction events to our previous simulations
-
-```{r}
-## Simulate the tree and traits with a random extinction event
-sim_rand_extinction <- treats(
-                   traits     = my_traits,
-                   bd.params  = est_params,
-                   stop.rule  = stop_rule,
-                   events     = random_extinction,
-                   null.error = 100,
-                   replicates = 50)
-
-## Simulate the tree and traits with a random extinction event
-sim_trait_extinction <- treats(
-                   traits     = my_traits,
-                   bd.params  = est_params,
-                   stop.rule  = stop_rule,
-                   events     = negative_extinction,
-                   null.error = 100,
-                   replicates = 50)
-
-## Visualising the difference between both scenarios
-par(mfrow = c(1,2))
-plot(sim_rand_extinction[[1]], main = "Random extinction")
-plot(sim_trait_extinction[[1]], main = "Negative extinction")
-```
-
-We can then measure the disparity from these scenarios and compare them to our observed ones
-
-```{r}
-## Converting the data to be passed to dispRity functions
-rand_ext_data <- dispRitreats(sim_rand_extinction)
-trait_ext_data <- dispRitreats(sim_trait_extinction)
-
-## Creating the time series
-rand_ext_time_series <- chrono.subsets(data = rand_ext_data$data,
-                                       tree = rand_ext_data$tree,
-                                       time = seq(from = 0.9, to = 0, by = -0.1),
-                                       method = "continuous", model = "proximity")
-trait_ext_time_series <- chrono.subsets(data = trait_ext_data$data,
-                                        tree = trait_ext_data$tree,
-                                        time = seq(from = 0.9, to = 0, by = -0.1),
-                                        method = "continuous", model = "proximity")
-
-## Calculating the disparity from the simulated data
-sim_rand_ext_disp <- dispRity(rand_ext_time_series, metric = c(sum, variances),
-                              verbose = TRUE)
-sim_trait_ext_disp <- dispRity(trait_ext_time_series, metric = c(sum, variances),
-                               verbose = TRUE)
-
-par(mfrow = c(2,2))
-plot(change_in_size, main = "observed disparity")
-plot(simulated_disparity, main = "simulated disparity")
-plot(sim_rand_ext_disp, main = "simulated disparity\n(mass extinction)")
-plot(simulated_disparity, main = "simulated disparity\n(trait extinction)")
-```
-
-Interestingly, with all our parameters used here and our specific disparity metric we cannot detcet the effect of a mass extinction.
 
@@
 You can find much more about how to design modifiers in the [`make.modifiers` section](#makemodifiers).
+
+## An illustrative example: mammals disparity through the K-Pg extinction {kpgexample}
+
+Bib is studying the evolution of mammalian disparity (i.e. diversity of shapes) across the Cretaceous-Palaeogene mass extinction event (K-Pg - 66 million years ago).
+
+They use a dataset from @BeckLee that is a discrete morphological space containing the ordination of the around 400 morphological characters for 50 species and estimated for their 49 descendants into an ordinated shapespace of 99 elements and 97 dimensions.
+
+```{r}
+library(dispRity)
+## Loading the mammalian morphological example data
+data(BeckLee_mat99)
+
+## This is what the dataset looks like
+head(BeckLee_mat99)[1:5, 1:5]
+```
+
+And a tree showing the relation between these species through time:
+
+```{r}
+## The phylogenetic tree
+data(BeckLee_tree)
+
+## And what the tree looks like
+plot(BeckLee_tree, cex = 0.8)
+axisPhylo()
+abline(v = BeckLee_tree$root.time - 66, col = "red")
+```
+
+They measure the disparity as the sum of variances as a proxy for changes in the traitspace size (@moms) with some bootstrapping to create some confidence intervals around that curve.
+
+```{r}
+## Making continuous time series
+time_series <- chrono.subsets(data = BeckLee_mat99,
+                              tree = BeckLee_tree,
+                              method = "continuous", model = "proximity",
+                              time = seq(from = 101, to = 41, by = -5))
+## Calculating disparity
+change_in_size <- dispRity(boot.matrix(time_series), metric = c(sum, variances))
+plot(change_in_size)
+abline(v = 8, col = "red")
+```
+
+> For more info about `chrono.subsets` and the `dispRity` package, you can have a look at the [`dispRity` manual](http://tguillerme.github.io/dispRity.html).
+
+Looking at this curve, Bib wants to know whether disparity increased as a response to the K-Pg mass extinction or not.
+To do this, they need to check whether changes in disparity can actually be detected due to an extinction event (e.g. if an extinction event happens, does that change disparity curves?).
+As a subsidiary question, they want to know whether the pattern is more likely due to a selective or random extinction event.
+
+### Simulating trees and traits using the observed parameters
+
+The first step is to simulate the observed tree.
+We can do that by using crude parameters from the observed tree:
+
+```{r}
+## Extracting the crude parameters
+(est_params <- crude.bd.est(BeckLee_tree))
+## These parameters are the observed speciation and extinction rate per million years.
+
+## Setting the stopping rule (stop after reaching 50 taxa)
+stop_rule <- list(max.taxa = 50)
+
+## Simulating just a birth-death tree with these parameters to check
+set.seed(1)
+test_tree <- treats(bd.params  = est_params,
+                    stop.rule  = stop_rule,
+                    null.error = 100)
+plot(ladderize(test_tree), cex = 0.8)
+axisPhylo()
+```
+
+> Note that because of the high extinction to speciation ratio, the tree simulations can often fail before reaching the stop rule (i.e.
+all lineages go extinct before reaching 50 taxa).
+To avoid getting stuck in there, we use the option `null.error = 100` to rerun the simulations up to 100 times before obtaining a tree that did not went fully extinct.
+Also note that using such parameters make the simulations run slowlier but ensure the requested results are always achieved.
+
+We now have a way to simulate a topology close to the observed one. We can simulate a list of topologies by replicating this function X amounts of time to get a distribution of trees:
+
+```{r, eval = FALSE, message = FALSE}
+## Generating 50 trees
+tree_distribution <- treats(bd.params  = est_params,
+                            stop.rule  = stop_rule,
+                            null.error = 100,
+                            replicates = 50)
+```
+
+Of course, here our objective is to also simulate some trait values to measure disparity under some simulated scenario.
+We can do that be adding a `"traits"` object to the `treats` function.
+Here we will use a default [Brownian Motion](https://en.wikipedia.org/wiki/Brownian_motion) to simulate our trait.
+
+```{r}
+## Creating a trait in 97 dimensions.
+my_traits <- make.traits(process = BM.process, n = ncol(BeckLee_mat99))
+```
+
+> Note that there are many more options on how to simulate your trait such as using correlation, different processes, etc... See details in the [make traits section](#maketraits).
+
+```{r, message = FALSE}
+## Simulate the tree and traits
+sim_data <- treats(traits     = my_traits,
+                   bd.params  = est_params,
+                   stop.rule  = stop_rule,
+                   null.error = 100)
+```
+
+We can visualise the resulting tree using the `plot` function and visualising the different traits using the `trait` option in `plot`:
+
+```{r}
+## Plotting the results (first trait)
+plot(sim_data, trait = 1)
+```
+
+Of course, here we want to simulate a distribution of trees and traits:
+
+```{r, message = FALSE}
+## Simulate the tree and traits
+sim_data <- treats(traits     = my_traits,
+                   bd.params  = est_params,
+                   stop.rule  = stop_rule,
+                   null.error = 100,
+                   replicates = 50)
+```
+
+Et voila! We now have a distribution of 50 trees and 50 datasets from which we can calculate disparity.
+We can convert this list into `dispRity` using the `treats.dispRity` function.
+
+```{r, message = FALSE}
+## Converting the data to be passed to dispRity functions
+dispRity_data <- dispRitreats(sim_data)
+
+## Creating the time series (by default dispRitreats scales the trees to be 1 unit of times old)
+time_series <- chrono.subsets(data = dispRity_data$data,
+                              tree = dispRity_data$tree,
+                              time = seq(from = 0.9, to = 0, by = -0.1),
+                              method = "continuous", model = "proximity")
+
+## Calculating the disparity from the simulated data
+simulated_disparity <- dispRity(time_series, metric = c(sum, variances),
+                                verbose = TRUE)
+par(mfrow = c(1,2))
+plot(simulated_disparity, main = "simulated disparity")
+plot(change_in_size, main = "observed disparity")
+```
+
+We can see that both simulated and observed disparity curves are different.
+From there one could already reasonably propose that the observe disparity does not entirely follows our parametrised scenario (constant speciation and extinction and BM trait evolution).
+
+However, we can complexify our scenario to introduce the effect of mass extinction event.
+We can simulate two alternative scenarios:
+ * one where we create a random extinction event after having simulated half of the species (25) where we remove 75% of species;
+ * and a second one where we remove all the species with negative values.
+This allows us to simulate two contrasted scenarios to see whether using our observed dataset it would be possible to detect changes in disparity due to mass extinctions.
+
+To do so, we need to create two mass extinction `"event"` objects:
+
+```{r}
+## Creating a random mass extinction
+random_extinction <- make.events(target = "taxa",
+                                 condition = taxa.condition(25, condition = `>=`),
+                                 modification = random.extinction(0.75))
+## Creating an extrinction that removes species with negative trait values
+negative_extinction <- make.events(target = "taxa",
+                                   condition = taxa.condition(25, condition = `>=`),
+                                   modification = trait.extinction(x = 0, condition = `>=`))
+```
+
+We can then feed these two extinction events to our previous simulations
+
+```{r}
+## Simulate the tree and traits with a random extinction event
+sim_rand_extinction <- treats(
+                   traits     = my_traits,
+                   bd.params  = est_params,
+                   stop.rule  = stop_rule,
+                   events     = random_extinction,
+                   null.error = 100,
+                   replicates = 50)
+
+## Simulate the tree and traits with a random extinction event
+sim_trait_extinction <- treats(
+                   traits     = my_traits,
+                   bd.params  = est_params,
+                   stop.rule  = stop_rule,
+                   events     = negative_extinction,
+                   null.error = 100,
+                   replicates = 50)
+
+## Visualising the difference between both scenarios
+par(mfrow = c(1,2))
+plot(sim_rand_extinction[[1]], main = "Random extinction")
+plot(sim_trait_extinction[[1]], main = "Negative extinction")
+```
+
+We can then measure the disparity from these scenarios and compare them to our observed ones
+
+```{r}
+## Converting the data to be passed to dispRity functions
+rand_ext_data <- dispRitreats(sim_rand_extinction)
+trait_ext_data <- dispRitreats(sim_trait_extinction)
+
+## Creating the time series
+rand_ext_time_series <- chrono.subsets(data = rand_ext_data$data,
+                                       tree = rand_ext_data$tree,
+                                       time = seq(from = 0.9, to = 0, by = -0.1),
+                                       method = "continuous", model = "proximity")
+trait_ext_time_series <- chrono.subsets(data = trait_ext_data$data,
+                                        tree = trait_ext_data$tree,
+                                        time = seq(from = 0.9, to = 0, by = -0.1),
+                                        method = "continuous", model = "proximity")
+
+## Calculating the disparity from the simulated data
+sim_rand_ext_disp <- dispRity(rand_ext_time_series, metric = c(sum, variances),
+                              verbose = TRUE)
+sim_trait_ext_disp <- dispRity(trait_ext_time_series, metric = c(sum, variances),
+                               verbose = TRUE)
+
+par(mfrow = c(2,2))
+plot(change_in_size, main = "observed disparity")
+plot(simulated_disparity, main = "simulated disparity")
+plot(sim_rand_ext_disp, main = "simulated disparity\n(mass extinction)")
+plot(simulated_disparity, main = "simulated disparity\n(trait extinction)")
+```
+
+Interestingly, with all our parameters used here and our specific disparity metric we cannot detcet the effect of a mass extinction.

### 2023-05-17 inst/gitbook/index.Rmd
Update index.Rmd

@@
 
-Ar Macarface is studying the evolution of mammalian disparity (i.e. diversity of shapes) across the Cretaceous-Palaeogene mass extinction event (K-Pg - 66 million years ago).
+Bib is studying the evolution of mammalian disparity (i.e. diversity of shapes) across the Cretaceous-Palaeogene mass extinction event (K-Pg - 66 million years ago).
 They observe a decrease of morphological disparity after the extinction event followed by a sharp increase of disparity.
-However Ar is also prudent and wants to test whether their observed pattern is just an artifact of the data or actually showing some effect of the K-Pg extinction on mammalian evolution.
+However Bib is also prudent and wants to test whether their observed pattern is just an artifact of the data or actually showing some effect of the K-Pg extinction on mammalian evolution.
 First they need to check whether changes in disparity can actually be discernables due to an extinction event (i.e.
 if an extinction event happens, would it be possible to detect a change in disparity)?
-Second they want to know whether these changes are due to changes are due to the mass extinction event being selective or not.
+Second they want to know whether these changes are due to the mass extinction event being selective or not (i.e. the mass extinction favored some traits).
 
-The [getting started](#kpgexample) section actually shows how do implement this example in `treats` in details.
+The [illustrative example](#kpgexample) section actually shows how do implement this example in `treats` in details.
 
-There are some very good R packages that simulate birth-death trees (e.g.
+Of course, this is just an example of when to use `treats`.
+Simulating trees and traits can also be useful for:
+
+ * Provide a baseline/null model to compare to an observed one.
+ * Testing evolutionary hypotheses.
+ * Exploring statistics properties (i.e. can the disparity metric X capture changes in Y).
+ * Easily generating illustrative/simplified evolutionary history examples.
+ * Teaching simplified macro-evolutionary theory (e.g. the birth-death model).
+
+Note that for approaching these points individually (or some aspects of it), there are some very good R packages that simulate birth-death trees (e.g.
 [TreeSim](https://github.com/tanja819/TreeSim)) or simulate traits (disparity) and diversity jointly (e.g.
@@
 ![Schematic summary of the `treats` package architecture](treats_structure.png)
-% NC: The wonkiness of the figure is annoying me lol. Can you center the arrows etc?
 

### 2023-05-22 inst/gitbook/02_make.traits.Rmd
Added schematics

@@
 For example, it can be a uni-dimensional variable uncorrelated to other traits or a multi-dimensional one with different amounts of correlation.
+
+![`make.traits`: The main inputs are (1) the `process`, one or more functions to describe the trait(s) as a function of previous state (`x0`) and the branch length (`edge.length`), (2) `n` the number of dimensions of the trait and (3) the starting value `start`.](make.traits.png)
 
@@
 ```
+
+### Multidimensional traits and correlations
+
+As shown in the example above, a trait can be made multidimensional by using the `n` argument specifically passed to a `process`.
+However, in this case, traits are simulated independently of each other.
+It is possible to include correlation by designing a multidimensional trait `process` directly generating the correlation between within the dimensions of the process.
+Alternatively, the inbuilt `BM.process` and `OU.process` can intake a correlation matrix to simulate traits independtly but with a correlation coefficient (see [this illustration with an event](#EG_change_correlation)).
 
@@
 
+ * `discrete.process`: this one generates discrete trait values based on a transition matrix (you can use the utility function `transition.matrix` to generate one).
+
+```{r, echo = FALSE}
+## Generating a stepwise transition matrix for 3 states (with an overal random transition rate)
+stepwise_matrix <- transition.matrix(type = "stepwise", states = 3)
+## Generatin and plotting the the trait
+plot(make.traits(discrete.process, process.args = list(transitions = stepwise_matrix)))
+```
+
+Of course here the plotting of a discrete trait is not that informative...
+
 ## Testing the traits with `test`

### 2023-05-22 inst/gitbook/03_make.modifiers.Rmd
Added schematics

@@
 
+![`make.modifiers`: The three main inputs to make modifiers are functions to modify the `branch.length` (waiting) process controlling the "growth" of the tree; the `selection` (selecting) process controlling the selection of the tip on which to apply speciation or extinction; and the `speciation` (speciating) process controlling whether the select tip goes extinct or speciates.
+The allowed arguments are detailed [here](#allowarguments).](make.modifiers.png)
+
 ## The default modifier (how the process is working)
@@
 The modifiers modify the core of the birth-death process as implemented in `treats`.
-
-<!-- TODO: make that into a shiny app animation (or just in html?) -->
-
 By default, the birth-death process in `treats` uses this modifier:

### 2023-05-22 inst/gitbook/04_make.events.Rmd
Update 04_make.events.Rmd
Added schematics

@@
 
+![`make.events`: events are encoded using three main arguments: the `target`, specifying _what_ the event is modifying, the `condition` specifying _when_ to trigger the event, and `modification`, specifying _how_ the event modifies the simulations.](make.events.png)
+
 ## Target
@@
 The targets that are currently available are:
+
  * `"taxa"` to modify anything linked to the [`lineage` list](#allowarguments). e.g. making half of the living taxa go extinct.

### 2023-05-22 inst/gitbook/02_make.traits.Rmd
man updates

@@
 
+## Mapping traits on a tree: `map.traits`
+
+Although `make.traits` is mainly designed for `treats` simulations, i.e.
+simulating both the trees AND the traits at the same time, it is possible to use an already simulated or estimated tree and generating one or more traits on it using the `map.traits` function.
+This function takes the trait input as a `"traits"` object (from `make.traits`) and a tree topology with branch length.
+Note that you can also use `"traits"` objects with multiple or multi-dimensional traits as well as multiple phylogenies (`"multiPhylo"`).
+
+```{r}
+## Simulating a random tree with branch length
+my_tree <- rtree(20)
+
+## Creating three different traits objects:
+## A Brownian Motion
+bm_process <- make.traits(process = BM.process)
+## An Ornstein-Uhlenbeck process
+ou_process <- make.traits(process = OU.process)
+## No process (just randomly drawing values from a normal distribution)
+no_process <- make.traits(process = no.process)
+
+## Mapping the three traits on the phylogeny
+bm_traits <- map.traits(bm_process, my_tree)
+ou_traits <- map.traits(ou_process, my_tree)
+no_traits <- map.traits(no_process, my_tree)
+
+## Plotting the topology and the different traits
+par(mfrow = c(2,2))
+plot(my_tree, main = "Base topology")
+plot(bm_traits, main = "Mapped BM")
+plot(ou_traits, main = "Mapped OU")
+plot(no_traits, main = "Mapped normal trait")
+```
+
 ## Templates for making your very own process

### 2023-05-22 inst/gitbook/07_specific_examples.Rmd
man updates

@@
 `founding.event` | [generating a subtree with no extinction](#EG_founding_purebirth), [generating a subtree with a different process](EG_founding_traits)
+
+<!-- TODO add a recursive event example where discrete self rate is changed by half everytime you reach 10 living species -->
 
@@
 ## A 2D variance covariance matrix
-vcv_matrix <- matrix(1, 2, 2)
+cor_matrix <- matrix(1, 2, 2)
 
@@
 correlated_2D_BM <- make.traits(n = 2, process = BM.process,
-                      process.args = list(Sigma = vcv_matrix))
+                      process.args = list(Sigma = cor_matrix))
 ```
@@
 ## New correlation
-new_vcv <- matrix(c(10,3,3,2),2,2)
+new_cor <- matrix(c(10,3,3,2),2,2)
 
@@
     target       = "traits",
-    modification = traits.update(process.args = list(Sigma = new_vcv)))
+    modification = traits.update(process.args = list(Sigma = new_cor)))
 ```

### 2023-05-22 paper/treats_paper.tex
paper up

@@
 
-\noindent{\Large \bf \treats: a modular \texttt{R} package for simulating diversity and traits data.} 
+\noindent{\Large \bf \treats: a modular \texttt{R} package for simulating trees and traits.} 
 \bigskip
 
-\noindent {\normalsize \sc Thomas Guillerme$^{1,*}$, Natalie Cooper$^{2}$, Andrew P. Beckerman$^{1}$, and Gavin H. Thomas$^{1,3}$}\\
+\noindent {\normalsize \sc Thomas Guillerme$^{1,*}$}\\
+% , Natalie Cooper$^{2}$, Andrew P. Beckerman$^{1}$, and Gavin H. Thomas$^{1,3}$
 \noindent {\small \it 
 $^1$School of Biosciences, University of Sheffield, Sheffield, S10 2TN, United Kingdom.\\
-$^2$Natural History Museum, Cromwell Road, London, SW7 5BD, United Kingdom.\\
-$^3$Bird Group, Department of Life Sciences, the Natural History Museum at Tring, Tring, United Kingdom.\\}
+% $^2$Natural History Museum, Cromwell Road, London, SW7 5BD, United Kingdom.\\
+% $^3$Bird Group, Department of Life Sciences, the Natural History Museum at Tring, Tring, United Kingdom.\\
+}
 
@@
         \item Simulating biological realistic data can be an important step to understand and investigate biological mechanisms.
-        Like null models, base line models or something else models, they allow us to generate a pattern that arises from controlled processes and thought through mechanisms.
+        Simulated data can be used to generate null, base line or neutral models.
+        These models can be used either in comparison to observed data to estimate the mechanisms that generated the data.
+        Or they can be used to explore, understand and develop theoretical advances by proposing toy models.
 
-        \item In evolutionary biology, these simulations often involve the need of an evolutionary process where descent with modification is at the core of how the simulated data is generated.
-        This can be made much more complicated with loads of different stuff needing to be taken into account to affect the simulations (e.g.
-traits, mechanisms such as competition or events such as mass extinctions).
+        \item In evolutionary biology, such simulations often involve the need of an evolutionary process where descent with modification is at the core of how the simulated data is generated.
+        These evolutionary processes can then be nearly infinitely modified to include much more complex processes to affect the simulations such as traits co-evolution, competition mechanisms or mass extinction events.
 
-        \item Here I present the \treats package, a modular \texttt{R} package for diversity and trait disparity simulations.
-        This package is based on a standard birth death algorithm that can be easily tuned to the user specific needs by designing their specific trait process, birth-death modifiers and events.
+        \item Here I present the \treats package, a modular \texttt{R} package for trees and traits simulations.
+        This package is based on a simple birth death algorithm from which all core steps can easily be modified by users (hence the modularity).
         It also provides a tidy interface through the \treats object, allowing users to easily run reproducible simulations.
 
-        \item The \treats package also comes with an extend manual regularly updated following users' questions or suggestions.
+        \item Here I present the core algorithms and their integrated modularity and provide an example of how to use the \treats package when studying disparity, the diversity of traits.
+        The \treats package also comes with an extend manual regularly updated following users' questions or suggestions which comes as an extended supplementary material to this paper.
 
@@
 
-\noindent (Keywords: diversity, disparity, simulations, birth-death, null-models, ecology, evolution)\\
+\noindent (Keywords: trees, traits, simulations, birth-death, null-models, ecology, evolution, disparity)\\
 
@@
 % Simulations in biology
-Comparing biological patterns is one of the key ways (and arguably the only) to understand mechanisms in evolutionary biology.
-This lead to the development of phylogenetic comparative methods as key sub-field in ecology and evolution @CITE-some-PCMreview.
-As indicated in the name, phylogenetic comparative methods rely on comparing patterns in a phylogenetical concepts to understand biological mechanisms or concepts.
-These comparisons can be done between observed patterns under different conditions suggesting different processes or mechanisms (e.g.
-trait patterns for species on an island vs.
-on the mainland @CITE-some-islandPCM-paper) or by comparing some observed pattern to one simulated under null or base conditions @CITE-paper-with-null-comparisons (but see @CITE-null-phylo-paper for distinctions or common misconceptions).
+Comparing biological patterns is one of the key ways to understand mechanisms in evolutionary biology.
+This lead to the development of phylogenetic comparative methods as key methodologically driven topic in ecology, evolution and palaeontology \citep{felsensteinPCM,pennell2013review}.
+As indicated in the name, phylogenetic comparative methods rely on comparing patterns in a phylogenetic context (i.e.
+descent with modification) to understand biological mechanisms or concepts \citep{harmon2019book}.
+These comparisons can be done between observed patterns under different conditions or against null, neutral or baseline models (see \citealt{bausman2018neutral} for distinctions or common misconceptions) suggesting different processes or mechanisms.
+For example different traits distribution for species with different diets \citep{deepak2023diet},
+or by comparing some observed pattern to one simulated under null or base conditions\citep{miller2022alternating}.
 In theory, workers can follow the research pipeline of thinking of a specific mechanism (e.g.
@@
 some traits of species across and extinction event) and then compare these patterns to one simulated under no specific conditions (e.g.
-a null model where the traits evolve randomly regardless of an extinction event).
-@CITE-mark-extinction
+a null model where the traits evolve randomly regardless of an extinction event)\citep{puttick2020complex}.
+For such an approach, we need statistical and softwares solutions to simulate trees and data to generate these specific null models.
 
 % What already exists
-In practice, these evolutionary simulations can be done relatively easily multiple times on computers using a birth-death process @CITE-something-describing-BD.
-A birth-death process is @yadiyada and has been routinely implemented in \texttt{R} (e.g. @CITE-ape, @CITE-diversitree).
+In practice, these evolutionary simulations can be done relatively easily multiple times on computers using a birth-death process \cite{feller1939birthdeath,stadler2010birthdeath,diversitree}
+A birth-death process is @yadiyada and has been routinely implemented in \texttt{R} (\citealt{R}; e.g. \citealt{ape}, \citealt{diversitree}).
 Traits can be simulating using the following processes @OU, @BM, @yadidada.
-The diversity of these traits through time can be called disparity in palaeontology @CITE-disparity-review or functional diversity in ecology (for functional traits; @CITE-setafanospaper).
+The diversity of these traits through time can be called disparity in palaeontology \cite{guillerme2020disparities} or functional diversity in ecology (for functional traits; \citealt{mammola2021concepts}).
 In \texttt{R}, this can be done with several already well used and well documented packages.
-For example if you want to simulate diversity through time, you can use \texttt{TreeSim} to simulate diversity under a set of specific parameters (e.g.
+For example if you want to simulate diversity through time, you can use \texttt{TreeSim} \citep{treesim} to simulate diversity under a set of specific parameters (e.g.
 speciation and extinction) with some events disrupting the simulations (e.g.
-mass extinctions) @CITE-TreeSim.
-One can even improve on generating these patterns using \texttt{FossilSim} to generate a pattern that will take into account fossilization processes @CITE-FossilSim.
-On the other hand, if you need to simulate both diversity and traits through time, this can be done with specific parameters in \texttt{RPANDA} @CITE-RPANDA or in \texttt{PETER} @CITE-PETER where the traits are generated stochasticaly through time (given some process) during the birth-death process.
+mass extinctions).
+One can even improve on generating these patterns using \texttt{FossilSim} \citep{fossilsim} to generate a pattern that will take into account fossilization processes.
+Or use \texttt{paleobuddy} \citep{paleobuddy} or \texttt{paleotree} \citep{paleotree} to generate palaeontology specific data.
+On the other hand, if you need to simulate both diversity and traits through time, this can be done with specific parameters in \texttt{RPANDA} \citep{rpanda} or in \texttt{PETER} \citep{puttick2020complex} where the traits are generated stochasticaly through time (given some process) during the birth-death process.
 
@@
 
-To simulate a birth-death tree, two arguments are essentially needed: the birth-death parameters (\texttt{bd.params}) the rate of speciation (birth or \lambda), the rate of extinction (death or \mu - this parameter can be equal to 0) and \texttt{stop.rule}, a rule for when to stop the growth of the tree (e.g.
+To simulate a birth-death tree, two arguments are essentially needed: the birth-death parameters (\texttt{bd.params}) the rate of speciation (birth or $\lambda$), the rate of extinction (death or $\mu$ - this parameter can be equal to 0) and \texttt{stop.rule}, a rule for when to stop the growth of the tree (e.g.
 when reaching 20 species).
 
-\texttt{## Running a simulation with a speciation rate of 1, an extinction rate of 0.1 and stopping when it reaches 20 taxa}
-\texttt{my_simulation <- treats(bd.params = list(speciation = 1, extinction = 0.1), stop.rule = list(max.taxa = 20))}
+% \texttt{## Running a simulation with a speciation rate of 1, an extinction rate of 0.1 and stopping when it reaches 20 taxa}
+% \texttt{my_simulation <- treats(bd.params = list(speciation = 1, extinction = 0.1), stop.rule = list(max.taxa = 20))}
 
-By default, the three first steps are included in a default modifier that reproduces a simple birth-death process given the input birth and death parameters.
+% By default, the three first steps are included in a default modifier that reproduces a simple birth-death process given the input birth and death parameters.
 
-\begin{enumerate}
-    \item Waiting: drawing a random number from an exponential distribution with a species dependent rate of $n \times (\lambda + \mu)$, where $n$ is the number of species currently alive in the process.
-    \item Selection: randomly choosing on of the $n$ taxa currently alive in the process;
-    \item Speciation: drawing a random number between 0 and 1; if that number is smaller than the speciation relative to the turnover ($\lambda / \lambda+\mu$), the lineage goes extinct %TODO: check!!!
-    ; else, the lineage speciates.
-\end{enumerate}
+% \begin{enumerate}
+%     \item Waiting: drawing a random number from an exponential distribution with a species dependent rate of $n \times (\lambda + \mu)$, where $n$ is the number of species currently alive in the process.
+%     \item Selection: randomly choosing on of the $n$ taxa currently alive in the process;
+%     \item Speciation: drawing a random number between 0 and 1; if that number is smaller than the speciation relative to the turnover ($\lambda / \lambda+\mu$), the lineage goes extinct %TODO: check!!!
+%     ; else, the lineage speciates.
+% \end{enumerate}
 
-These are included by default in a \texttt{treats} process but can be also set manually by creating a \texttt{``modifiers''} object with no options:
+% These are included by default in a \texttt{treats} process but can be also set manually by creating a \texttt{``modifiers''} object with no options:
 
-\texttt{## Generating the default modifiers}
-\texttt{my_modifiers <- make.modifiers()}
-\texttt{## Running a simulation with a speciation rate of 1, an extinction rate of 0.1 and stopping when it reaches 20 taxa with the default modifiers}
-\texttt{my_simulation <- treats(bd.params = list(speciation = 1, extinction = 0), stop.rule = list(max.taxa = 20), modifiers = my_modifiers)}
+% \texttt{## Generating the default modifiers}
+% \texttt{my_modifiers <- make.modifiers()}
+% \texttt{## Running a simulation with a speciation rate of 1, an extinction rate of 0.1 and stopping when it reaches 20 taxa with the default modifiers}
+% \texttt{my_simulation <- treats(bd.params = list(speciation = 1, extinction = 0), stop.rule = list(max.taxa = 20), modifiers = my_modifiers)}
 
-The modularity of the packages works with the ability to code the modifiers manually rather than using default ones.
-For example, to simulate the same process but hard coding the modifiers we can use the following
+% The modularity of the packages works with the ability to code the modifiers manually rather than using default ones.
+% For example, to simulate the same process but hard coding the modifiers we can use the following
 
-\texttt{## }
+% \texttt{## }
 
-\subsection{Simulating traits (disparity)}
+% \subsection{Simulating traits (disparity)}
+
+% multivariate stuff \cite{adams2019multivarPCM}
 
@@
 
+Extended manual following recommendations from \cite{cooper2016dark}
+
 \section{Package location}

### 2023-05-25 paper/treats_paper.tex
Update treats_paper.tex

@@
 For example different traits distribution for species with different diets \citep{deepak2023diet},
+% different habitats https://onlinelibrary.wiley.com/doi/full/10.1111/jbi.12939
 or by comparing some observed pattern to one simulated under null or base conditions\citep{miller2022alternating}.
@@
 A birth-death process is @yadiyada and has been routinely implemented in \texttt{R} (\citealt{R}; e.g. \citealt{ape}, \citealt{diversitree}).
-Traits can be simulating using the following processes @OU, @BM, @yadidada.
+Traits can be simulating using the following processes @OU, @BM, @yadidada. % TODO
 The diversity of these traits through time can be called disparity in palaeontology \cite{guillerme2020disparities} or functional diversity in ecology (for functional traits; \citealt{mammola2021concepts}).
@@
 On the other hand, if you need to simulate both diversity and traits through time, this can be done with specific parameters in \texttt{RPANDA} \citep{rpanda} or in \texttt{PETER} \citep{puttick2020complex} where the traits are generated stochasticaly through time (given some process) during the birth-death process.
+
+% STOPPED HERE
 

### 2023-05-25 inst/gitbook/01_getting_started.Rmd
Updated gitbook for v0.2

@@
 
-## An illustrative example: mammals disparity through the K-Pg extinction {kpgexample}
+## An illustrative example: mammals disparity through the K-Pg extinction {#kpgexample}
 
@@
 
-They use a dataset from @BeckLee that is a discrete morphological space containing the ordination of the around 400 morphological characters for 50 species and estimated for their 49 descendants into an ordinated shapespace of 99 elements and 97 dimensions.
+They use a dataset from @beckancient2014 that is a discrete morphological space containing the ordination of the around 400 morphological characters for 50 species and estimated for their 49 descendants into an ordinated shapespace of 99 elements and 97 dimensions.
 
@@
 
-```{r}
+```{r, message = FALSE}
 ## Simulate the tree and traits with a random extinction event

### 2023-05-25 inst/gitbook/_book/treats_manual.tex
Updated gitbook for v0.2

--- 2023-04-24 inst/gitbook/_book/dads_manual.tex
@@
 \author{Thomas Guillerme (\href{mailto:guillert@tcd.ie}{\nolinkurl{guillert@tcd.ie}})}
-\date{2023-03-27}
+\date{2023-05-25}
 
@@
 }
-\hypertarget{treats-disparity-and-diversity-simulations.}{%
-\chapter{\texorpdfstring{\texttt{treats}: disparity and diversity simulations.}{treats: disparity and diversity simulations.}}\label{treats-disparity-and-diversity-simulations.}}
-
-Allowing to simulate disparity and diversity at the same time with interaction between both.
+\hypertarget{treats-trees-and-traits-simulations.}{%
+\chapter{\texorpdfstring{\texttt{treats}: trees and traits simulations.}{treats: trees and traits simulations.}}\label{treats-trees-and-traits-simulations.}}
+
+Simultaneously simulate phylogenetic trees and traits (and the interaction between them).
 
@@
 
-There are some very good packages out there to simulate birth death trees (like \href{https://github.com/tanja819/TreeSim}{TreeSim}) or even packages simulating traits (disparity) and diversity jointly (like \href{https://github.com/cran/RPANDA}{RPANDA} or \href{https://github.com/PuttickMacroevolution/PETER}{PETER}).
-We strongly advice you have a look at these packages first as they might be more appropriate for your tasks.
-
-\texttt{treats} aims to be a highly modular and friendly version of all these packages: it allows to simulate disparity and diversity jointly with a vast array of options that can be easily modified by users.
-For example, you can easily generate any type of process to generate a trait (BM, OU, something else, etc\ldots) in multiple dependent or independent dimensions through \texttt{"traits"} objects.
-You can then specify how the traits should affect disparity through \texttt{"modifiers"} objects. And finally you can create events (like mass extinctions) through \texttt{"events"} objects.
+Bib is studying the evolution of mammalian disparity (i.e.~diversity of shapes) across the Cretaceous-Palaeogene mass extinction event (K-Pg - 66 million years ago).
+They observe a decrease of morphological disparity after the extinction event followed by a sharp increase of disparity.
+However Bib is also prudent and wants to test whether their observed pattern is just an artifact of the data or actually showing some effect of the K-Pg extinction on mammalian evolution.
+First they need to check whether changes in disparity can actually be discernables due to an extinction event (i.e.~if an extinction event happens, would it be possible to detect a change in disparity)?
+Second they want to know whether these changes are due to the mass extinction event being selective or not (i.e.~the mass extinction favored some traits).
+
+The \protect\hyperlink{kpgexample}{illustrative example} section actually shows how do implement this example in \texttt{treats} in details.
+
+Of course, this is just an example of when to use \texttt{treats}.
+Simulating trees and traits can also be useful for:
+
+\begin{itemize}
+\tightlist
+\item
+  Provide a baseline/null model to compare to an observed one.
+\item
+  Testing evolutionary hypotheses.
+\item
+  Exploring statistics properties (i.e.~can the disparity metric X capture changes in Y).
+\item
+  Easily generating illustrative/simplified evolutionary history examples.
+\item
+  Teaching simplified macro-evolutionary theory (e.g.~the birth-death model).
+\end{itemize}
+
+Note that for approaching these points individually (or some aspects of it), there are some very good R packages that simulate birth-death trees (e.g.~\href{https://github.com/tanja819/TreeSim}{TreeSim}) or simulate traits (disparity) and diversity jointly (e.g.~\href{https://github.com/cran/RPANDA}{RPANDA} or \href{https://github.com/PuttickMacroevolution/PETER}{PETER}).
+We strongly advise you look at these packages first as they might be more appropriate for your needs.
+
+\texttt{treats} is a highly modular and user-friendly package that contains the simulations functionality of all these packages: it allows you to simulate disparity and diversity jointly with a vast array of options that can be easily modified by users.
+For example, you can easily generate any type of process to generate a trait (BM, OU, something else, etc.) in multiple dependent or independent dimensions through \texttt{"traits"} objects.
+You can then specify how the traits should affect disparity through \texttt{"modifiers"} objects.
+And finally you can create events (like mass extinctions) through \texttt{"events"} objects.
 These objects and how to modify them will be detailed throughout this manual.
-Finally we are putting an emphasise in the development of this package on the speed and reliability of the functions.
+Finally we are putting an emphasis in the development of this package on the speed and reliability of the functions.
 
@@
 
-Because their is an infinite way you might want to generate disparity and diversity (different traits, different modifiers and different events), the \texttt{treats} package is designed to make all these parts easy to code separately and integrate them easily in the \texttt{treats} core functions.
-This allows you to simulate finely tuned multidimensional data for your specific project!
+Because there is an infinite number of ways you might want to generate disparity and diversity (different traits, different modifiers and different events), the \texttt{treats} package is designed to make all these parts easy to code separately and integrate in the \texttt{treats} core functions.
+This allows you to simulate finely-tuned multidimensional data for your specific project!
 
@@
 
-This manual explains how \texttt{treats} work in as much details as possible and aims to give you the keys to design your own diversity and disparity simulations.
+This manual explains how \texttt{treats} works in as much details as possible and aims to give you the keys to design your own diversity and disparity simulations.
 
 The first section \texttt{Getting\ started} is aimed for people with a \texttt{R} beginner level:
-* you know what a package is - and how to install it;
-* you know what a function is - and what arguments are;
-* you have already (vaguely) heard that they are different types of objects in R - like matrices (\texttt{"matrix"}) or trees (\texttt{"phylo"});
-* and you have already looked for documentation and bugs online.
-
-The sections after that are a bit more advanced and requires the following level:
-* you have already created functions before;
-* you know how to subset elements of a list - i.e.~you understand the \texttt{list\$element} syntax;
-* you are aware that both \texttt{TRUE\ ==\ 1} and \texttt{FALSE\ ==\ 0} are \texttt{TRUE};
-* you have heard that the objects in R have classes and that it sometimes matters - for example you know that R can differentiate between \texttt{as.numeric(1)} and \texttt{as.integer(1)};
-* you owe part of your life to Stack Overflow.
+
+\begin{itemize}
+\tightlist
+\item
+  you know what a package is - and how to install it;
+\item
+  you know what a function is - and what arguments are;
+\item
+  you have already (vaguely) heard that there are different types of objects in R - like matrices (\texttt{"matrix"}) or trees (\texttt{"phylo"});
+\item
+  and you have already looked for documentation online.
+\end{itemize}
+
+The sections after that are a bit more advanced and require the following level:
+
+\begin{itemize}
+\tightlist
+\item
+  you have already created functions before;
+\item
+  you know how to subset elements of a list - i.e.~you understand the \texttt{list\$element} syntax;
+\item
+  you are aware that both \texttt{TRUE\ ==\ 1} and \texttt{FALSE\ ==\ 0} are \texttt{TRUE};
+\item
+  you have heard that the objects in R have classes and that it sometimes matters - for example you know that R can differentiate between \texttt{as.numeric(1)} and \texttt{as.integer(1)};
+\item
+  you owe part of your work to Stack Overflow.
+\end{itemize}
 
@@
 
-You can install this package easily, directly from the github:
+You can install this package easily, directly from the GitHub:
 
@@
 If you need help with the package, hopefully the following manual will be useful.
-However, parts of this package are still in development and some other parts are probably not covered.
-Thus if you have suggestions or comments on on what has already been developed or will be developed, please send me an email (\href{mailto:guillert@tcd.ie}{\nolinkurl{guillert@tcd.ie}}) or if you are a GitHub user, directly create an issue on the \href{https://github.com/TGuillerme/treats}{GitHub page}.
-Doing so will not only hopefully help you but also other users since it will help improve this manual!
+However, if you have suggestions or comments on on what has already been developed or will be developed, please send me an email (\href{mailto:guillert@tcd.ie}{\nolinkurl{guillert@tcd.ie}}) or if you are a GitHub user, directly create an issue on the \href{https://github.com/TGuillerme/treats}{GitHub page}.
+Doing so will not only help you but also other users as it will help improve this manual and the package!
 
@@
 
-Basically, the \texttt{treats} function intakes your personalised \texttt{traits}, \texttt{modifiers} and \texttt{events} to generate your disparity and diversity.
-You will find more details about how these objects (\texttt{traits}, \texttt{modifiers} and \texttt{events}) work in the rest of the tutorial but here is a graphical representation of how \texttt{treats} work:
+Basically, the \texttt{treats} function takes your \texttt{traits}, \texttt{modifiers} and \texttt{events} objects to simulate disparity and diversity.
+You will find more details about how these objects (\texttt{traits}, \texttt{modifiers} and \texttt{events}) work in the rest of the tutorial but here is a graphical representation of how \texttt{treats} works:
 
@@
 \includegraphics{treats_structure.png}
-\caption{Schematised summary of the \texttt{treats} package architecture}
+\caption{Schematic summary of the \texttt{treats} package architecture}
 \end{figure}
@@
 \begin{quote}
-If you want to get your hands dirty, you can go straight to the \protect\hyperlink{gettingstarted}{getting start section}, this following section just describes the process in pseudo-code.
+If you want to get your hands dirty, you can go straight to the \protect\hyperlink{gettingstarted}{getting started section}, this following section just describes the process in pseudo-code.
 \end{quote}
@@
 The \texttt{treats} algorithm is based on a modular version of the birth-death process algorithm.
-The birth-death model is a continuous Markov process (i.e.~a continuous random process where the state will change each step are linked through time and change according to some random exponential variables) which is well defined mathematically and \href{https://lukejharmon.github.io/pcm/chapter10_birthdeath/\#section-10.2-the-birth-death-model}{commonly used in biology} but also in \href{https://en.wikipedia.org/wiki/Birth\%E2\%80\%93death_process}{many other fields}.
+The birth-death model is a continuous Markov process (i.e.~a continuous random process where future events depend only on the present, and not on past events) which is well defined mathematically and \href{https://lukejharmon.github.io/pcm/chapter10_birthdeath/\#section-10.2-the-birth-death-model}{commonly used in evolutionary biology} but also in \href{https://en.wikipedia.org/wiki/Birth\%E2\%80\%93death_process}{many other fields}.
 
@@
 \item
-  \textbf{Starting the process}: this step is actually non modular and creates a random tree with one tip, one node and one branch length connecting both.
+  \textbf{Starting the process}: this step is non-modular and creates a random tree with one tip, one node and one branch connecting both.
 This step is used to optimise the rest of the algorithm in terms of speed and memory management.
@@
   \textbf{Selecting a lineage} (\texttt{selecting}): this step selects a tip that is currently not extinct.
-In a standard birth-death process this is just done randomly, however in \texttt{treats} this can be modified based on the birth-death parameters, the currently available lineages and potential trait values.
-\emph{For example, it is possible to put a higher probability in selecting a lineage that is next to a lineage that recently went extinct and has a positive trait value.}
+In a standard birth-death process this is done randomly, however in \texttt{treats} this can be modified based on the birth-death parameters, the currently available lineages and potential trait values.
+\emph{For example, it is possible to put a higher probability for selecting a lineage that is closely related to a lineage that recently went extinct and has a positive trait value.}
   Then go to step 2.
@@
 It does so by adding the same amount of branch length to all the non-extinct lineages.
-In an exact birth death process, this is done by drawing a random value from an exponential distribution with a rate of \[\text{number of living lineages} \times (\text{speciation} + \text{extinction parameters})\].
+In an exact birth-death process, this is done by drawing a random value from an exponential distribution with a rate of \[\text{number of living lineages} \times (\text{speciation} + \text{extinction parameters})\].
 In \texttt{treats}, this can be modified based on the birth-death parameters, the currently available lineages and potential trait values.
@@
 \item
-  \textbf{Simulating traits} {[}optional{]}(\texttt{traits}): this step allows to simulate a trait value for the selected lineage from step 1.
+  \textbf{Simulating traits} {[}optional{]}(\texttt{traits}): this step allows you to simulate a trait value for the selected lineage from step 1.
 This is typically not part of a standard birth-death process and is handled via the \texttt{traits} option in \texttt{treats} (see the \protect\hyperlink{maketraits}{\texttt{make.traits}} chapter for more details).
@@
   \textbf{Speciating} (\texttt{speciating}): in this step, the selected lineage from step 1 has the option of speciating or going extinct.
-In a standard birth-death process, this happens by randomly drawing a value between 0 and 1 as a way to trigger speciation relatively to the birth-death parameters:
+In a standard birth-death process, this happens by randomly drawing a value between 0 and 1 as a way to trigger speciation relative to the birth-death parameters:
 \end{enumerate}
@@
 Again, in \texttt{treats}, this process can be modified based on the birth-death parameters, the currently available lineages and potential trait values.
-\emph{For example, the lineage can only speciate if it's trait value is positive, regardless of whether speciation or extinction has been triggered}.
+\emph{For example, the lineage can only speciate if its trait value is positive, regardless of whether speciation or extinction have been triggered}.
 Then go to step 5.
 5. If the number of lineages is less than required number of lineages, then go to step 1. Else go to step 6.
-6. The simulation has stopped because it reached the required amount of lineages or time.
+6. The simulation stops because it has reached the required amount of lineages and/or time (i.e.~branch length).
 
@@
 
-\hypertarget{the-simplest-of-all-analysis-simulating-diversity-only}{%
-\section{The simplest of all analysis: simulating diversity only}\label{the-simplest-of-all-analysis-simulating-diversity-only}}
-
-One of the simplest things to do with the \texttt{treats} package is just to simulate a birth death tree.
+\hypertarget{the-simplest-analysis-simulating-diversity-only}{%
+\section{The simplest analysis: simulating diversity only}\label{the-simplest-analysis-simulating-diversity-only}}
+
+One of the simplest things to do with the \texttt{treats} package is to simulate a birth-death tree.
 For that you can use the function \texttt{treats} and specify your stopping rule.
-The stopping rule simply tells the birth death process to step whenever it reaches one of these three conditions:
+The stopping rule simply tells the birth-death process to stop whenever it reaches one of these three conditions:
 
@@
 \item
-  \texttt{"max.living"\ =\ n} stop when there is \texttt{n} co-occuring taxa of the same age (i.e.~``living'' taxa);
+  \texttt{"max.living"\ =\ n} stop when there are \texttt{n} co-occuring taxa of the same age (i.e.~``living'' taxa);
 \item
@@
 \begin{Highlighting}[]
-\CommentTok{\#\# Running the birth death simulation}
+\CommentTok{\#\# Running the birth{-}death simulation}
 \NormalTok{my\_tree \textless{}{-}}\StringTok{ }\KeywordTok{treats}\NormalTok{(}\DataTypeTok{stop.rule =}\NormalTok{ my\_stop\_rule)}
@@
 \begin{quote}
-Note that here we could have specified more than one stopping rule, for example, we might want to run a simulation and stop it if it either reaches 10 taxa or the age 2 using \texttt{stop.rule\ =\ list(max.time\ =\ 2,\ max.taxa\ =\ 10)}.
+Note that here we could have specified more than one stopping rule, for example, we might want to run a simulation and stop it if it \emph{either} reaches 10 taxa \emph{or} the age 2 using \texttt{stop.rule\ =\ list(max.time\ =\ 2,\ max.taxa\ =\ 10)}.
 The simulation will then stop when either of these conditions are met.
@@
 
-The resulting object is a classic \texttt{"phylo"} object that you can simply plot or visualise like so:
+The resulting object is a classic \texttt{"phylo"} object that you can plot like so:
 
@@
 People familiar with the \href{https://lukejharmon.github.io/pcm/chapter10_birthdeath/}{birth-death models} might have noticed that we did not specify two important things here: the speciation parameter (sometimes called ``lambda'' or ``birth'') and the extinction parameter (sometimes called ``mu'', ``death'' or ``background extinction'').
-By default \texttt{treats} runs a pure birth model (the speciation is set to 1 and the extinction to 0).
-However, you can easily change that by specifying your new birth death parameters:
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\#\# my birth death parameters}
+By default \texttt{treats} runs a pure birth model (speciation is set to 1 and extinction to 0).
+However, you can easily change that by specifying your own birth-death parameters:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# my birth{-}death parameters}
 \NormalTok{my\_params \textless{}{-}}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{speciation =} \DecValTok{1}\NormalTok{,}
@@
 \begin{quote}
-Note that here it is not necessary to specify \texttt{speciation\ =\ 1} since this is the default option, you can always just change the parameter of interest (e.g.~changing \texttt{extinction\ =\ 0} to \texttt{extinction\ =\ 1/3}).
-However, we think it's good practice to attribute both parameters specifically to avoid any confusion.
-You can find more information about setting up more complex birth death parameters \protect\hyperlink{makebdparams}{in this section}.
+You can find more information about setting up more complex birth-death parameters \protect\hyperlink{makebdparams}{in this section}.
 \end{quote}
 
-You can then run the same birth death tree with extinction:
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\#\# Generating a birth death tree with extinctions:}
+You can then run the same birth-death tree simulation with extinction:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# Generating a birth{-}death tree with extinctions:}
 \NormalTok{my\_tree \textless{}{-}}\StringTok{ }\KeywordTok{treats}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ my\_params, }\DataTypeTok{stop.rule =}\NormalTok{ my\_stop\_rule)}
@@
 
-Chances are that you want to also simulate traits (disparity) along with your diversity (otherwise, we suggest using the \href{https://github.com/tanja819/TreeSim/}{\texttt{TreeSim}} package that provides many more birth death models).
+Chances are that you are using \texttt{treats} because you also want to simulate traits (disparity) along with your diversity (otherwise, we suggest using the \href{https://github.com/tanja819/TreeSim/}{\texttt{TreeSim}} package that provides many more birth-death models).
 Simulating traits is not much more complicated in \texttt{treats}: you'll simply need to create a \texttt{"traits"} object using the \texttt{make.traits} function.
@@
 \texttt{"traits"} objects contain one or more processes which are the ways to generate the trait.
-The most common of these processes is the \href{https://en.wikipedia.org/wiki/Brownian_motion}{Brownian Motion}.
+The most common of these processes is the \href{https://en.wikipedia.org/wiki/Brownian_motion}{Brownian Motion} model.
 This is used by default with the \texttt{make.traits} function:
@@
 ##  ---- treats object ---- 
-## Simulated diversity data (x$tree):
+## Simulated phylogenetic tree (x$tree):
 ## 
@@
 ## 
-## Simulated disparity data (x$data):
+## Simulated trait data (x$data):
 ## 1 trait for 1 process (A) with one starting value (0).
@@
 \begin{Highlighting}[]
-\CommentTok{\#\# Extracting the tree (a "phylo" object)}
+\CommentTok{\#\# Extracting the tree }
 \NormalTok{the\_generated\_tree \textless{}{-}}\StringTok{ }\NormalTok{my\_data}\OperatorTok{$}\NormalTok{tree}
+\CommentTok{\# Note that this is a "phylo" object}
 \KeywordTok{class}\NormalTok{(the\_generated\_tree)}
@@
 \begin{Highlighting}[]
-\CommentTok{\#\# Extracting the data (a "matrix")}
+\CommentTok{\#\# Extracting the data }
 \NormalTok{the\_generated\_data \textless{}{-}}\StringTok{ }\NormalTok{my\_data}\OperatorTok{$}\NormalTok{data}
+\CommentTok{\# Note that this is a "matrix" or "array"}
 \KeywordTok{class}\NormalTok{(the\_generated\_data)}
@@
 
-You can find much more on how to design trait objects in the \protect\hyperlink{maketraits}{\texttt{make.traits} section}.
-
-\hypertarget{slightly-more-more-complex-simulating-linked-disparity-and-diversity}{%
-\section{Slightly more more complex: simulating linked disparity and diversity}\label{slightly-more-more-complex-simulating-linked-disparity-and-diversity}}
-
-The example above is also still pretty simple and easily done through a variety of \texttt{R} packages: here the trait and the tree are simulated at the same time but only the tree is simulating the trait (i.e.~the trait value at a tip is affected by it's ancestor and the branch length leading to it) but not the other way around (the trait value does not affect the tree).
+You can find much more about how to design trait objects in the \protect\hyperlink{maketraits}{\texttt{make.traits} section}.
+
+\hypertarget{slightly-more-complex-again-simulating-linked-disparity-and-diversity}{%
+\section{Slightly more complex again: simulating linked disparity and diversity}\label{slightly-more-complex-again-simulating-linked-disparity-and-diversity}}
+
+The example above is still pretty simple and easily done through a variety of \texttt{R} packages: here the trait and the tree are simulated at the same time but only the tree is simulating the trait (i.e.~the trait value at a tip is affected by its ancestor and the branch length leading to it) but not the other way around (the trait value does not affect the tree).
 It is possible to add this aspect using \texttt{"modifiers"} objects.
@@
 \texttt{"modifiers"} affect two key steps of the birth-death process: the calculation of the waiting time (i.e.~the component generating branch lengths) and the triggering of speciation or extinction events.
-These events can be modified using some \texttt{condition} and \texttt{modify} function.
+These events can be modified using \texttt{condition} and \texttt{modify} functions.
 In other words, when reaching a certain condition specified by a \texttt{condition} function, the birth-death process will modify either the branch length or the speciation (or extinction) probability by applying a \texttt{modify} function.
@@
 
-Note that these are default for a birth death tree and were actually already applied in the examples before (without specifying a modifier):
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\#\# Make a default modifiers}
+Note that these are defaults for a birth-death tree and were already applied in the examples above without specifying a \texttt{"modifiers"} object:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# Make a default modifiers object}
 \NormalTok{default\_modifiers \textless{}{-}}\StringTok{ }\KeywordTok{make.modifiers}\NormalTok{()}
@@
 
-This will not do much to our simulations compared to the previous trait and tree simulation but we can provide our modifiers object to the \texttt{treats} function:
+This will not do anything to our simulations compared to the previous trait and tree simulation but we can provide our \texttt{"modifiers"} object to the \texttt{treats} function:
 
@@
 
+\CommentTok{\# Set random seed so we get the same results}
 \KeywordTok{set.seed}\NormalTok{(}\DecValTok{1}\NormalTok{)}
+
 \CommentTok{\#\# Simulate disparity and diversity}
@@
 ## 
-## Simulated diversity data (x$tree):
+## Simulated phylogenetic tree (x$tree):
 ## 
@@
 
-Note however that the printing information is now updated to state that you've added a modifier (even though it's a default one).
-
-For more interesting simulations however, you can provide modifiers that actually modify the birth death process.
-We can create one for example that makes species go extinct if their ancestor have a negative trait value.
-For that we need to create a modifiers object that modifies the \texttt{speciation} process with a specific condition and a specific modification when that condition is met.
+Note however, that the printing information is now updated to state that you've added a \texttt{"modifiers"} object (even though it's the same as the default).
+
+For more interesting simulations however, you can provide modifiers that actually modify the birth-death process.
+We can create one for example that makes species go extinct if their ancestor has a negative trait value.
+For that we need to create a \texttt{"modifiers"} object that modifies the \texttt{speciation} process with a specific condition and a specific modification when that condition is met.
 For a speciation to occur (and a species to not go extinct), the algorithm draws a random value \emph{x} between 0 and 1 and if the value is smaller than the speciation parameter divided by the speciation + extinction parameter, the lineage speciates (\(x < (\lambda \ (\lambda + \mu))\)), else it goes extinct.
 
-Here we can create a modifier forcing the lineage to go extinct every time a specific condition is met (say the lineage trait value is negative).
-In other words, if the lineages trait value is negative, make sure that \emph{x} is equal to 1 (so that the speciation condition \(x < (\lambda \ (\lambda + \mu))\) is never met).
-
-First we need to create a condition function to trigger this (non) speciation event.
-We can do that by specifying our \texttt{condition} function (when the ancestor is trait is negative) apply our modification (here the \texttt{staying.alive} function).
-For that we can use the \texttt{parent.traits} utility function that is optimised for accessing traits in the birth death process (but you can of course write your own).
-This function intakes the \texttt{trait.values} and \texttt{parent.lineage} arguments, two arguments that you can leave named as they are to facilitate \texttt{treats}' understanding of what you want to assess:
+Here we can create a modifier forcing the lineage to go extinct every time a specific condition is met, e.g.~if the lineage trait value is negative.
+In other words, if the lineage's trait value is negative, make sure that \emph{x} is equal to 1, so that the speciation condition \(x < (\lambda \ (\lambda + \mu))\) is never met.
+
+First we need to create a \texttt{condition} function to trigger this (non) speciation event.
+We can do that by specifying our \texttt{condition} function (here the \texttt{going.extinct} function) to apply our modification.
+For that we can use the \texttt{parent.traits} utility function that is optimised for accessing traits in the birth-death process (but you can of course write your own).
+This function takes the \texttt{trait.values} and \texttt{parent.lineage} arguments, two arguments that you can leave named as they are to facilitate \texttt{treats}' understanding of what you want to assess:
 
@@
 
-The we are going to create the modification function.
-This function must intake the argument \texttt{x} and, in our case, returns the same value no matter what: \texttt{1} so that speciation never happens when the \texttt{negative.ancestor} function is triggered.
+Then we create the modification function.
+This function must take the argument \texttt{x} and, in our case, returns the same value no matter what: \texttt{1} so that speciation never happens when the \texttt{negative.ancestor} function is triggered.
 
@@
 
-We can then provide these two functions (the condition \texttt{negative.ancestor} and how to modify the speciation event when this condition is met \texttt{staying.alive}).
-If you are an advances \texttt{treats} user, you can design your own \texttt{speciation} function but if you just want to use a normal \texttt{speciation} function, you can use the default one from \texttt{treats} called\ldots{} \texttt{speciation}.
+We can then provide these two functions (the condition \texttt{negative.ancestor} and how to modify the speciation event when this condition is met \texttt{going.extinct}).
+If you are an advances \texttt{treats} user, you can design your own \texttt{speciation} function but if you just want to use a normal \texttt{speciation} function, you can use the default one from \texttt{treats} called \texttt{speciation}.
 
@@
             \CommentTok{\#\# If the following condition is met...}
-            \DataTypeTok{condition =}\NormalTok{ negative.ancestor, }\CommentTok{\# Does the lineage has a negative ancestor?}
+            \DataTypeTok{condition =}\NormalTok{ negative.ancestor, }\CommentTok{\# Does the lineage have an ancestor with negative trait values?}
             \CommentTok{\#\# Then apply the following modifier...}
@@
 Note that the \texttt{make.modifiers} function tests whether the input is compatible with \texttt{treats} by default so unless you have an error message, your \texttt{modifiers} will work!
-We can now simulate our tree and traits with our modifier: species will go extinct if their ancestor have a negative trait value:
+We can now simulate our tree and traits with our modifier: species will go extinct if their ancestor has a negative trait value:
 
@@
 \CommentTok{\#\# Simulate disparity and diversity}
-\NormalTok{biased\_data \textless{}{-}}\StringTok{ }\KeywordTok{treats}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ extinction\_}\DecValTok{02}\NormalTok{,}
+\NormalTok{modified\_data \textless{}{-}}\StringTok{ }\KeywordTok{treats}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ extinction\_}\DecValTok{02}\NormalTok{,}
                         \DataTypeTok{stop.rule =}\NormalTok{ living\_}\DecValTok{20}\NormalTok{,}
@@
                         \DataTypeTok{modifiers =}\NormalTok{ negatives\_extinct)}
-\NormalTok{biased\_data}
+\NormalTok{modified\_data}
 \end{Highlighting}
@@
 \KeywordTok{plot}\NormalTok{(default\_data, }\DataTypeTok{main =} \StringTok{"default results"}\NormalTok{)}
-\KeywordTok{plot}\NormalTok{(biased\_data, }\DataTypeTok{main =} \StringTok{"results with the modifier"}\NormalTok{)}
+\KeywordTok{plot}\NormalTok{(modified\_data, }\DataTypeTok{main =} \StringTok{"results with the modifier"}\NormalTok{)}
 \end{Highlighting}
@@
 
-You can find much more how to design modifiers in the \protect\hyperlink{makemodifiers}{\texttt{make.modifiers} section}.
+You can find much more about how to design modifiers in the \protect\hyperlink{makemodifiers}{\texttt{make.modifiers} section}.
+
+\hypertarget{kpgexample}{%
+\section{An illustrative example: mammals disparity through the K-Pg extinction}\label{kpgexample}}
+
+Bib is studying the evolution of mammalian disparity (i.e.~diversity of shapes) across the Cretaceous-Palaeogene mass extinction event (K-Pg - 66 million years ago).
+
+They use a dataset from \citet{beckancient2014} that is a discrete morphological space containing the ordination of the around 400 morphological characters for 50 species and estimated for their 49 descendants into an ordinated shapespace of 99 elements and 97 dimensions.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{library}\NormalTok{(dispRity)}
+\CommentTok{\#\# Loading the mammalian morphological example data}
+\KeywordTok{data}\NormalTok{(BeckLee\_mat99)}
+
+\CommentTok{\#\# This is what the dataset looks like}
+\KeywordTok{head}\NormalTok{(BeckLee\_mat99)[}\DecValTok{1}\OperatorTok{:}\DecValTok{5}\NormalTok{, }\DecValTok{1}\OperatorTok{:}\DecValTok{5}\NormalTok{]}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+##                   [,1]        [,2]        [,3]       [,4]        [,5]
+## Cimolestes  -0.6794737  0.15658591  0.04918307  0.2250983 -0.38139436
+## Maelestes   -0.5797289  0.04223105 -0.20329542 -0.1545388 -0.06993258
+## Batodon     -0.8812102  0.33619889  0.41748560  0.2446872 -0.31922156
+## Bulaklestes -0.9295090 -0.08517626  0.11364659  0.3589646 -0.54998508
+## Daulestes   -1.0025229 -0.09260779  0.06354371  0.4614256 -0.61211569
+\end{verbatim}
+
+And a tree showing the relation between these species through time:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# The phylogenetic tree}
+\KeywordTok{data}\NormalTok{(BeckLee\_tree)}
+
+\CommentTok{\#\# And what the tree looks like}
+\KeywordTok{plot}\NormalTok{(BeckLee\_tree, }\DataTypeTok{cex =} \FloatTok{0.8}\NormalTok{)}
+\KeywordTok{axisPhylo}\NormalTok{()}
+\KeywordTok{abline}\NormalTok{(}\DataTypeTok{v =}\NormalTok{ BeckLee\_tree}\OperatorTok{$}\NormalTok{root.time }\OperatorTok{{-}}\StringTok{ }\DecValTok{66}\NormalTok{, }\DataTypeTok{col =} \StringTok{"red"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-22-1.pdf}
+
+They measure the disparity as the sum of variances as a proxy for changes in the traitspace size (\citet{moms}) with some bootstrapping to create some confidence intervals around that curve.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# Making continuous time series}
+\NormalTok{time\_series \textless{}{-}}\StringTok{ }\KeywordTok{chrono.subsets}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ BeckLee\_mat99,}
+                              \DataTypeTok{tree =}\NormalTok{ BeckLee\_tree,}
+                              \DataTypeTok{method =} \StringTok{"continuous"}\NormalTok{, }\DataTypeTok{model =} \StringTok{"proximity"}\NormalTok{,}
+                              \DataTypeTok{time =} \KeywordTok{seq}\NormalTok{(}\DataTypeTok{from =} \DecValTok{101}\NormalTok{, }\DataTypeTok{to =} \DecValTok{41}\NormalTok{, }\DataTypeTok{by =} \DecValTok{{-}5}\NormalTok{))}
+\CommentTok{\#\# Calculating disparity}
+\NormalTok{change\_in\_size \textless{}{-}}\StringTok{ }\KeywordTok{dispRity}\NormalTok{(}\KeywordTok{boot.matrix}\NormalTok{(time\_series), }\DataTypeTok{metric =} \KeywordTok{c}\NormalTok{(sum, variances))}
+\KeywordTok{plot}\NormalTok{(change\_in\_size)}
+\KeywordTok{abline}\NormalTok{(}\DataTypeTok{v =} \DecValTok{8}\NormalTok{, }\DataTypeTok{col =} \StringTok{"red"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-23-1.pdf}
+
+\begin{quote}
+For more info about \texttt{chrono.subsets} and the \texttt{dispRity} package, you can have a look at the \href{http://tguillerme.github.io/dispRity.html}{\texttt{dispRity} manual}.
+\end{quote}
+
+Looking at this curve, Bib wants to know whether disparity increased as a response to the K-Pg mass extinction or not.
+To do this, they need to check whether changes in disparity can actually be detected due to an extinction event (e.g.~if an extinction event happens, does that change disparity curves?).
+As a subsidiary question, they want to know whether the pattern is more likely due to a selective or random extinction event.
+
+\hypertarget{simulating-trees-and-traits-using-the-observed-parameters}{%
+\subsection{Simulating trees and traits using the observed parameters}\label{simulating-trees-and-traits-using-the-observed-parameters}}
+
+The first step is to simulate the observed tree.
+We can do that by using crude parameters from the observed tree:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# Extracting the crude parameters}
+\NormalTok{(est\_params \textless{}{-}}\StringTok{ }\KeywordTok{crude.bd.est}\NormalTok{(BeckLee\_tree))}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+##  ---- treats birth-death parameters object ---- 
+## speciation: 0.352330414024189.
+## extinction: 0.287616664509542.
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# These parameters are the observed speciation and extinction rate per million years.}
+
+\CommentTok{\#\# Setting the stopping rule (stop after reaching 50 taxa)}
+\NormalTok{stop\_rule \textless{}{-}}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{max.taxa =} \DecValTok{50}\NormalTok{)}
+
+\CommentTok{\#\# Simulating just a birth{-}death tree with these parameters to check}
+\KeywordTok{set.seed}\NormalTok{(}\DecValTok{1}\NormalTok{)}
+\NormalTok{test\_tree \textless{}{-}}\StringTok{ }\KeywordTok{treats}\NormalTok{(}\DataTypeTok{bd.params  =}\NormalTok{ est\_params,}
+                    \DataTypeTok{stop.rule  =}\NormalTok{ stop\_rule,}
+                    \DataTypeTok{null.error =} \DecValTok{100}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## Building the tree:.Done.
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{plot}\NormalTok{(}\KeywordTok{ladderize}\NormalTok{(test\_tree), }\DataTypeTok{cex =} \FloatTok{0.8}\NormalTok{)}
+\KeywordTok{axisPhylo}\NormalTok{()}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-24-1.pdf}
+
+\begin{quote}
+Note that because of the high extinction to speciation ratio, the tree simulations can often fail before reaching the stop rule (i.e.~all lineages go extinct before reaching 50 taxa).
+To avoid getting stuck in there, we use the option \texttt{null.error\ =\ 100} to rerun the simulations up to 100 times before obtaining a tree that did not went fully extinct.
+Also note that using such parameters make the simulations run slowlier but ensure the requested results are always achieved.
+\end{quote}
+
+We now have a way to simulate a topology close to the observed one. We can simulate a list of topologies by replicating this function X amounts of time to get a distribution of trees:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# Generating 50 trees}
+\NormalTok{tree\_distribution \textless{}{-}}\StringTok{ }\KeywordTok{treats}\NormalTok{(}\DataTypeTok{bd.params  =}\NormalTok{ est\_params,}
+                            \DataTypeTok{stop.rule  =}\NormalTok{ stop\_rule,}
+                            \DataTypeTok{null.error =} \DecValTok{100}\NormalTok{,}
+                            \DataTypeTok{replicates =} \DecValTok{50}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+Of course, here our objective is to also simulate some trait values to measure disparity under some simulated scenario.
+We can do that be adding a \texttt{"traits"} object to the \texttt{treats} function.
+Here we will use a default \href{https://en.wikipedia.org/wiki/Brownian_motion}{Brownian Motion} to simulate our trait.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# Creating a trait in 97 dimensions.}
+\NormalTok{my\_traits \textless{}{-}}\StringTok{ }\KeywordTok{make.traits}\NormalTok{(}\DataTypeTok{process =}\NormalTok{ BM.process, }\DataTypeTok{n =} \KeywordTok{ncol}\NormalTok{(BeckLee\_mat99))}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{quote}
+Note that there are many more options on how to simulate your trait such as using correlation, different processes, etc\ldots{} See details in the \protect\hyperlink{maketraits}{make traits section}.
+\end{quote}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# Simulate the tree and traits}
+\NormalTok{sim\_data \textless{}{-}}\StringTok{ }\KeywordTok{treats}\NormalTok{(}\DataTypeTok{traits     =}\NormalTok{ my\_traits,}
+                   \DataTypeTok{bd.params  =}\NormalTok{ est\_params,}
+                   \DataTypeTok{stop.rule  =}\NormalTok{ stop\_rule,}
+                   \DataTypeTok{null.error =} \DecValTok{100}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+We can visualise the resulting tree using the \texttt{plot} function and visualising the different traits using the \texttt{trait} option in \texttt{plot}:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# Plotting the results (first trait)}
+\KeywordTok{plot}\NormalTok{(sim\_data, }\DataTypeTok{trait =} \DecValTok{1}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-28-1.pdf}
+
+Of course, here we want to simulate a distribution of trees and traits:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# Simulate the tree and traits}
+\NormalTok{sim\_data \textless{}{-}}\StringTok{ }\KeywordTok{treats}\NormalTok{(}\DataTypeTok{traits     =}\NormalTok{ my\_traits,}
+                   \DataTypeTok{bd.params  =}\NormalTok{ est\_params,}
+                   \DataTypeTok{stop.rule  =}\NormalTok{ stop\_rule,}
+                   \DataTypeTok{null.error =} \DecValTok{100}\NormalTok{,}
+                   \DataTypeTok{replicates =} \DecValTok{50}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+Et voila! We now have a distribution of 50 trees and 50 datasets from which we can calculate disparity.
+We can convert this list into \texttt{dispRity} using the \texttt{treats.dispRity} function.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# Converting the data to be passed to dispRity functions}
+\NormalTok{dispRity\_data \textless{}{-}}\StringTok{ }\KeywordTok{dispRitreats}\NormalTok{(sim\_data)}
+
+\CommentTok{\#\# Creating the time series (by default dispRitreats scales the trees to be 1 unit of times old)}
+\NormalTok{time\_series \textless{}{-}}\StringTok{ }\KeywordTok{chrono.subsets}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ dispRity\_data}\OperatorTok{$}\NormalTok{data,}
+                              \DataTypeTok{tree =}\NormalTok{ dispRity\_data}\OperatorTok{$}\NormalTok{tree,}
+                              \DataTypeTok{time =} \KeywordTok{seq}\NormalTok{(}\DataTypeTok{from =} \FloatTok{0.9}\NormalTok{, }\DataTypeTok{to =} \DecValTok{0}\NormalTok{, }\DataTypeTok{by =} \FloatTok{{-}0.1}\NormalTok{),}
+                              \DataTypeTok{method =} \StringTok{"continuous"}\NormalTok{, }\DataTypeTok{model =} \StringTok{"proximity"}\NormalTok{)}
+
+\CommentTok{\#\# Calculating the disparity from the simulated data}
+\NormalTok{simulated\_disparity \textless{}{-}}\StringTok{ }\KeywordTok{dispRity}\NormalTok{(time\_series, }\DataTypeTok{metric =} \KeywordTok{c}\NormalTok{(sum, variances),}
+                                \DataTypeTok{verbose =} \OtherTok{TRUE}\NormalTok{)}
+\KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow =} \KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{,}\DecValTok{2}\NormalTok{))}
+\KeywordTok{plot}\NormalTok{(simulated\_disparity, }\DataTypeTok{main =} \StringTok{"simulated disparity"}\NormalTok{)}
+\KeywordTok{plot}\NormalTok{(change\_in\_size, }\DataTypeTok{main =} \StringTok{"observed disparity"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-30-1.pdf}
+
+We can see that both simulated and observed disparity curves are different.
+From there one could already reasonably propose that the observe disparity does not entirely follows our parametrised scenario (constant speciation and extinction and BM trait evolution).
+
+However, we can complexify our scenario to introduce the effect of mass extinction event.
+We can simulate two alternative scenarios:
+* one where we create a random extinction event after having simulated half of the species (25) where we remove 75\% of species;
+* and a second one where we remove all the species with negative values.
+This allows us to simulate two contrasted scenarios to see whether using our observed dataset it would be possible to detect changes in disparity due to mass extinctions.
+
+To do so, we need to create two mass extinction \texttt{"event"} objects:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# Creating a random mass extinction}
+\NormalTok{random\_extinction \textless{}{-}}\StringTok{ }\KeywordTok{make.events}\NormalTok{(}\DataTypeTok{target =} \StringTok{"taxa"}\NormalTok{,}
+                                 \DataTypeTok{condition =} \KeywordTok{taxa.condition}\NormalTok{(}\DecValTok{25}\NormalTok{, }\DataTypeTok{condition =} \StringTok{\textasciigrave{}}\DataTypeTok{\textgreater{}=}\StringTok{\textasciigrave{}}\NormalTok{),}
+                                 \DataTypeTok{modification =} \KeywordTok{random.extinction}\NormalTok{(}\FloatTok{0.75}\NormalTok{))}
+\CommentTok{\#\# Creating an extrinction that removes species with negative trait values}
+\NormalTok{negative\_extinction \textless{}{-}}\StringTok{ }\KeywordTok{make.events}\NormalTok{(}\DataTypeTok{target =} \StringTok{"taxa"}\NormalTok{,}
+                                   \DataTypeTok{condition =} \KeywordTok{taxa.condition}\NormalTok{(}\DecValTok{25}\NormalTok{, }\DataTypeTok{condition =} \StringTok{\textasciigrave{}}\DataTypeTok{\textgreater{}=}\StringTok{\textasciigrave{}}\NormalTok{),}
+                                   \DataTypeTok{modification =} \KeywordTok{trait.extinction}\NormalTok{(}\DataTypeTok{x =} \DecValTok{0}\NormalTok{, }\DataTypeTok{condition =} \StringTok{\textasciigrave{}}\DataTypeTok{\textgreater{}=}\StringTok{\textasciigrave{}}\NormalTok{))}
+\end{Highlighting}
+\end{Shaded}
+
+We can then feed these two extinction events to our previous simulations
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# Simulate the tree and traits with a random extinction event}
+\NormalTok{sim\_rand\_extinction \textless{}{-}}\StringTok{ }\KeywordTok{treats}\NormalTok{(}
+                   \DataTypeTok{traits     =}\NormalTok{ my\_traits,}
+                   \DataTypeTok{bd.params  =}\NormalTok{ est\_params,}
+                   \DataTypeTok{stop.rule  =}\NormalTok{ stop\_rule,}
+                   \DataTypeTok{events     =}\NormalTok{ random\_extinction,}
+                   \DataTypeTok{null.error =} \DecValTok{100}\NormalTok{,}
+                   \DataTypeTok{replicates =} \DecValTok{50}\NormalTok{)}
+
+\CommentTok{\#\# Simulate the tree and traits with a random extinction event}
+\NormalTok{sim\_trait\_extinction \textless{}{-}}\StringTok{ }\KeywordTok{treats}\NormalTok{(}
+                   \DataTypeTok{traits     =}\NormalTok{ my\_traits,}
+                   \DataTypeTok{bd.params  =}\NormalTok{ est\_params,}
+                   \DataTypeTok{stop.rule  =}\NormalTok{ stop\_rule,}
+                   \DataTypeTok{events     =}\NormalTok{ negative\_extinction,}
+                   \DataTypeTok{null.error =} \DecValTok{100}\NormalTok{,}
+                   \DataTypeTok{replicates =} \DecValTok{50}\NormalTok{)}
+
+\CommentTok{\#\# Visualising the difference between both scenarios}
+\KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow =} \KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{,}\DecValTok{2}\NormalTok{))}
+\KeywordTok{plot}\NormalTok{(sim\_rand\_extinction[[}\DecValTok{1}\NormalTok{]], }\DataTypeTok{main =} \StringTok{"Random extinction"}\NormalTok{)}
+\KeywordTok{plot}\NormalTok{(sim\_trait\_extinction[[}\DecValTok{1}\NormalTok{]], }\DataTypeTok{main =} \StringTok{"Negative extinction"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-32-1.pdf}
+
+We can then measure the disparity from these scenarios and compare them to our observed ones
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# Converting the data to be passed to dispRity functions}
+\NormalTok{rand\_ext\_data \textless{}{-}}\StringTok{ }\KeywordTok{dispRitreats}\NormalTok{(sim\_rand\_extinction)}
+\NormalTok{trait\_ext\_data \textless{}{-}}\StringTok{ }\KeywordTok{dispRitreats}\NormalTok{(sim\_trait\_extinction)}
+
+\CommentTok{\#\# Creating the time series}
+\NormalTok{rand\_ext\_time\_series \textless{}{-}}\StringTok{ }\KeywordTok{chrono.subsets}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ rand\_ext\_data}\OperatorTok{$}\NormalTok{data,}
+                                       \DataTypeTok{tree =}\NormalTok{ rand\_ext\_data}\OperatorTok{$}\NormalTok{tree,}
+                                       \DataTypeTok{time =} \KeywordTok{seq}\NormalTok{(}\DataTypeTok{from =} \FloatTok{0.9}\NormalTok{, }\DataTypeTok{to =} \DecValTok{0}\NormalTok{, }\DataTypeTok{by =} \FloatTok{{-}0.1}\NormalTok{),}
+                                       \DataTypeTok{method =} \StringTok{"continuous"}\NormalTok{, }\DataTypeTok{model =} \StringTok{"proximity"}\NormalTok{)}
+\NormalTok{trait\_ext\_time\_series \textless{}{-}}\StringTok{ }\KeywordTok{chrono.subsets}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ trait\_ext\_data}\OperatorTok{$}\NormalTok{data,}
+                                        \DataTypeTok{tree =}\NormalTok{ trait\_ext\_data}\OperatorTok{$}\NormalTok{tree,}
+                                        \DataTypeTok{time =} \KeywordTok{seq}\NormalTok{(}\DataTypeTok{from =} \FloatTok{0.9}\NormalTok{, }\DataTypeTok{to =} \DecValTok{0}\NormalTok{, }\DataTypeTok{by =} \FloatTok{{-}0.1}\NormalTok{),}
+                                        \DataTypeTok{method =} \StringTok{"continuous"}\NormalTok{, }\DataTypeTok{model =} \StringTok{"proximity"}\NormalTok{)}
+
+\CommentTok{\#\# Calculating the disparity from the simulated data}
+\NormalTok{sim\_rand\_ext\_disp \textless{}{-}}\StringTok{ }\KeywordTok{dispRity}\NormalTok{(rand\_ext\_time\_series, }\DataTypeTok{metric =} \KeywordTok{c}\NormalTok{(sum, variances),}
+                              \DataTypeTok{verbose =} \OtherTok{TRUE}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## Calculating disparity..........Done.
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{sim\_trait\_ext\_disp \textless{}{-}}\StringTok{ }\KeywordTok{dispRity}\NormalTok{(trait\_ext\_time\_series, }\DataTypeTok{metric =} \KeywordTok{c}\NormalTok{(sum, variances),}
+                               \DataTypeTok{verbose =} \OtherTok{TRUE}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## Calculating disparity..........Done.
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow =} \KeywordTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{,}\DecValTok{2}\NormalTok{))}
+\KeywordTok{plot}\NormalTok{(change\_in\_size, }\DataTypeTok{main =} \StringTok{"observed disparity"}\NormalTok{)}
+\KeywordTok{plot}\NormalTok{(simulated\_disparity, }\DataTypeTok{main =} \StringTok{"simulated disparity"}\NormalTok{)}
+\KeywordTok{plot}\NormalTok{(sim\_rand\_ext\_disp, }\DataTypeTok{main =} \StringTok{"simulated disparity}\CharTok{\textbackslash{}n}\StringTok{(mass extinction)"}\NormalTok{)}
+\KeywordTok{plot}\NormalTok{(simulated\_disparity, }\DataTypeTok{main =} \StringTok{"simulated disparity}\CharTok{\textbackslash{}n}\StringTok{(trait extinction)"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-33-1.pdf}
+
+Interestingly, with all our parameters used here and our specific disparity metric we cannot detcet the effect of a mass extinction.
 
@@
 In \texttt{treats}, traits are simulating by providing a \texttt{traits} object to the \texttt{traits} argument.
-This object is generated using the \texttt{make.traits}.
-
-\begin{quote}
-\textbf{What is a trait in biology?} This is one of these epistemological question like what \emph{is} a species that is not covered by all means in this package.
-For the purpose of \emph{this manual}, \textbf{biological trait} can be any coherent process of any number of dimensions which can or cannot be independent.
-Basically anything can be a trait as well as it's defined by the user (e.g.~one trait can be a 20 dimensional Brownian process that's correlated to a 1D OU process - which is another trait).
-For the purpose of the syntax in \emph{this package} though, a \textbf{trait in \texttt{treats}} is a column in the output trait matrix.
-For example, one might be interested in a biological trait that is ``beak shape'' which is defined as one complex biological trait that is the Procrustes superimposition of 4 3D landmarks and 300 3D semi-landmarks.
-This single ``beak shape'' \textbf{biological trait} is then composed of 912 \textbf{traits in \texttt{treats}} (i.e.~a matrix of \((4+300)\times3\) columns).
-\end{quote}
+This object is generated using \texttt{make.traits}.
+In \texttt{treats} a trait designate any coherent character of any number of dimensions which may or may not be independent of other traits.
+For example, it can be a uni-dimensional variable uncorrelated to other traits or a multi-dimensional one with different amounts of correlation.
+
+\begin{figure}
+\centering
+\includegraphics{make.traits.png}
+\caption{\texttt{make.traits}: The main inputs are (1) the \texttt{process}, one or more functions to describe the trait(s) as a function of previous state (\texttt{x0}) and the branch length (\texttt{edge.length}), (2) \texttt{n} the number of dimensions of the trait and (3) the starting value \texttt{start}.}
+\end{figure}
 
@@
 The function \texttt{make.traits} allows you to design the process of a trait or a set of traits.
-Here, the process of a trait designates the rules to generate the trait through time while simulating a phylogeny.
+Here, the process of a trait designates the rules needed to generate the trait through time while simulating a phylogeny.
 This process can depend on the previous state in the tree (i.e.~the trait of the ancestor) and the branch length to the descendant.
 One classic example is the \href{https://en.wikipedia.org/wiki/Brownian_motion}{Brownian motion process (or Weiner process)}.
-Note that it \emph{can} depend on both the ancestor and the branch length but does \emph{not necessary needs} (i.e.~the process can be only based on the previous state or only on branch length or on neither).
+Note that it \emph{can} depend on both the ancestor and the branch length but does \emph{not necessarily need to}, i.e.~the process can be based only on the previous state or only on branch length or on neither.
 
@@
 
-Trait processes in \texttt{treats} are functions that must always intake the following arguments by default.
+Trait processes in \texttt{treats} are functions that must always take the following arguments by default.
 
@@
 \item
-  \texttt{edge.length}: the branch length value
+  \texttt{edge.length}: the branch length
 \item
@@
 For example, the following function would be a valid process that always generate the \emph{true} trait value: 42!.
-In this example, the process is not dependent on neither the previous state (\texttt{x0}) and the branch length (\texttt{edge.length}).
+In this example, the process is not dependent on either the previous state (\texttt{x0}) or the branch length (\texttt{edge.length}).
 
@@
 \begin{quote}
-Note that in this function definition the arguments \texttt{x0} and \texttt{edge.length} have a default value set to \texttt{0} and \texttt{1} respectively.
+Note that in this function definition the arguments \texttt{x0} and \texttt{edge.length} have a default value of \texttt{0} and \texttt{1} respectively.
 In practice, these arguments are effectively set to the correct values in the \texttt{treats} internal function (i.e.~whatever \texttt{x0} and \texttt{edge.length} are at that specific time of the process) but providing a default can help speed up the algorithms (specifically all the internal checks).
@@
 \begin{Highlighting}[]
-\CommentTok{\#\# A wrongly formated process}
+\CommentTok{\#\# A wrongly formatted process}
 \NormalTok{invalid.process \textless{}{-}}\StringTok{ }\ControlFlowTok{function}\NormalTok{(}\DataTypeTok{x0 =} \DecValTok{0}\NormalTok{) \{}
@@
 
+This will not work in \texttt{make.traits} (see below).
+
 \hypertarget{using-a-process-in-treats}{%
@@
 
-You can design your own process as a function (as long as it has a valid syntax).
-Alternatively, the \texttt{treats} package proposes inbuilt processes, namely a multidimensional Brownian motion (\texttt{BM.process}) or a a multidimensional Ornstein-Uhlenbeck process (\texttt{OU.process}).
-You can find the list of implemented process by looking at the \texttt{?trait.process} manual page in \texttt{R}.
-
-Once a process is chosen, you can feed it to the \texttt{make.traits} function:
+You can design your own process as a function, as long as it has a valid syntax.
+Alternatively, the \texttt{treats} package has several inbuilt processes, namely a multidimensional Brownian motion (\texttt{BM.process}) or a multidimensional Ornstein-Uhlenbeck process (\texttt{OU.process}).
+You can find the list of implemented processes by looking at the \texttt{?trait.process} manual page in \texttt{R}.
+
+Once a process is chosen, you can feed it into the \texttt{make.traits} function:
 
@@
 
-This creates \texttt{"treats"} \texttt{"traits"} objects that you can print, and visualise using the \texttt{plot} function:
+This creates \texttt{"treats"} \texttt{"traits"} objects that you can print and visualise using the \texttt{plot} function:
 
@@
 
-\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-25-1.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-38-1.pdf}
 
@@
 ## {
-##     return(t(MASS::mvrnorm(n = 1, mu = x0, Sigma = sqrt(Sigma^2 * 
-##         edge.length), ...)))
+##     return(t(MASS::mvrnorm(n = 1, mu = x0, Sigma = Sigma * edge.length, 
+##         ...)))
 ## }
-## <bytecode: 0x55e278dc9cd8>
+## <bytecode: 0x55b094e15a20>
 ## <environment: namespace:treats>
@@
 
-As traits can get more and more complex, the automatic printing of its summary allows for a easier display of what's in the traits object.
-
-Note that it is possible to make \texttt{"traits"} objects with multiple processes (that can be the same):
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\#\# 4 traits: two BM, one OU and one normal non process}
+As traits can get more and more complex, the automatic printing of its summary allows for a easier display of what's in the \texttt{"traits"} object.
+
+Note that it is possible to make \texttt{"traits"} objects with multiple traits and multiple processes (that can be the same or different for each trait):
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# Four traits: two BM, one OU and one normal non process}
 \NormalTok{four\_traits \textless{}{-}}\StringTok{ }\KeywordTok{make.traits}\NormalTok{(}\DataTypeTok{process =} \KeywordTok{c}\NormalTok{(BM.process,}
@@
 
-\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-28-1.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-41-1.pdf}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow =} \KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{,}\DecValTok{1}\NormalTok{))}
+\end{Highlighting}
+\end{Shaded}
 
@@
 This means that \texttt{make.traits} will assume that your processes are always unidimensional by default and that they always start with the value \texttt{0}.
-It is however possible to change these values.
-
-For example you can use the following to create a three dimensional Brownian motion with each dimensions starting with the value \texttt{1}:
+It is possible however, to change these values.
+
+For example you can use the following to create a three dimensional Brownian motion with each dimension starting with the value \texttt{1}:
 
@@
 
-Note that the number of traits are distributed per processes.
-If the traits contains multiple process, the number of traits are distributed per processes:
+Note that the number of traits are applied for each process by default.
+However, you can apply different number of traits for different processes:
 
@@
 
-\hypertarget{what-even-is-a-trait}{%
-\subsection{What even is a trait?}\label{what-even-is-a-trait}}
-
-The definition of what a trait is can vary quiet a lot depending on the context of an analysis and the field.
-Because it would be impossible to accommodate all definitions in \texttt{treats} we had to go with an arbitrary one: a trait is whatever you define as a trait!
-A trait can be uni-dimensional as the measurement of a feature of an organism (e.g.~leaf surface, femur length, etc\ldots) but can also be multi-dimensional description of a feature, for example in 3D geometric morphometric, a trait could be defined as ``position of landmark X'' (which will be a trait with three dimensions, \emph{x}, \emph{y} and \emph{z}) or in ecology, the location of a plant can be a single 2D trait (expressed as latitude and longitude).
+\hypertarget{what-is-a-trait-in-treats}{%
+\subsection{\texorpdfstring{What is a trait in \texttt{treats}?}{What is a trait in treats?}}\label{what-is-a-trait-in-treats}}
+
+Because it would be impossible to accommodate all definitions of a trait in \texttt{treats} we chose an arbitrary one: a trait is whatever you define as a trait!
+A trait can be uni-dimensional as the measurement of a feature of an organism (e.g.~leaf surface, femur length, etc.) but can also be a multi-dimensional description of a feature, for example in 3D geometric morphometric context, a trait could be defined as ``position of landmark X'' (which will be a trait with three dimensions, \emph{x}, \emph{y} and \emph{z}) or in ecology, the location of a plant can be expressed as latitude and longitude coordinates.
 In \texttt{treats}, the \textbf{process} corresponds to this trait definition (e.g.~a process can be of n-dimensions and represents one organisms feature) and the \textbf{traits} represents the number of dimensions in total.
@@
 
+\hypertarget{multidimensional-traits-and-correlations}{%
+\section{Multidimensional traits and correlations}\label{multidimensional-traits-and-correlations}}
+
+As shown in the example above, a trait can be made multidimensional by using the \texttt{n} argument specifically passed to a \texttt{process}.
+However, in this case, traits are simulated independently of each other.
+It is possible to include correlation by designing a multidimensional trait \texttt{process} directly generating the correlation between within the dimensions of the process.
+Alternatively, the inbuilt \texttt{BM.process} and \texttt{OU.process} can intake a correlation matrix to simulate traits independtly but with a correlation coefficient (see \protect\hyperlink{EG_change_correlation}{this illustration with an event}).
+
 \hypertarget{extra-argument-for-the-processes-with-process.args}{%
@@
 You can also feed extra arguments to your process(es) functions.
-For example, the inbuilt process \texttt{no.process} (that is just a number generator not based on the previous value \texttt{x0} or the branch length) can intake a specific random number generator as a function:
+For example, the inbuilt process \texttt{no.process} (that is just a number generator not based on the previous value \texttt{x0} or the branch length) can take a specific random number generator as a function:
 
@@
 
-If one process do not need extra argument you must still give it and extra \texttt{NULL} process argument:
+If one process does not need extra argument you must still give it an extra \texttt{NULL} process argument:
 
@@
 You can also add traits to already existing trait objects using the simple \texttt{add} option.
-This option just intakes a \texttt{"treats"} \texttt{"traits"} object and the additional process(es) will be added to it. For example:
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\#\# Creating on simple default Brownian motion}
+This option just takes a \texttt{"treats"} \texttt{"traits"} object and the additional process(es) will be added to it.
+For example:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# Creating a simple default Brownian motion}
 \NormalTok{one\_process \textless{}{-}}\StringTok{ }\KeywordTok{make.traits}\NormalTok{(}\DataTypeTok{trait.names =} \StringTok{"BM"}\NormalTok{)}
@@
 \texttt{traits} objects also allow a background trait to be used when traits are simulated (\protect\hyperlink{bdalgorithm}{step 3 here}).
-This basically allows traits to be simulating for \emph{all} tips whenever a trait is generated for one tip.
+This allows traits to be simulated for \emph{all} tips whenever a trait is generated for one tip.
 This can be useful for keeping track of trait values along the simulation (\emph{cf} just at bifurcating nodes).
@@
 Note that technically you can nest as many background traits as you want (e.g.~\texttt{make.traits(background\ =\ make.traits(background\ =\ make.traits(...)))} is valid).
-However, you should always make sure that the background trait has the same dimensions as the normal (main) trait.
+However, you should always make sure that the background trait has the same dimensions as the main trait.
 When using a trait with background, your tree will have internal singleton nodes (i.e.~nodes linking to one ancestor and only one descendant).
@@
 
-\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-41-1.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-54-1.pdf}
 
@@
 
-Alternatively to having a background trait, you can also simulate a tree by generating traits at specific time steps with the \texttt{save.steps} option in \texttt{treats}.
+You can also simulate a tree by generating traits at specific time steps with the \texttt{save.steps} option in \texttt{treats}.
 This will apply the \texttt{traits} object to all lineages currently alive at the required time steps.
-These time steps can be either regular by providing a single numeric value; e.g.~\texttt{save.steps\ =\ 0.1} will get a snapshot of the trait values every 0.1 units of time.
-Or specific, by providing a specific set of values; e.g.~\texttt{save.steps\ =\ c(1,\ 1.2,\ 3)} will make a snapshot of the trait values at the required time steps.
+These time steps can be either regular by providing a single numeric value; e.g.~\texttt{save.steps\ =\ 0.1} will take a snapshot of the trait values every 0.1 units of time, or specific, by providing a specific set of values; e.g.~\texttt{save.steps\ =\ c(1,\ 1.2,\ 3)} will take a snapshot of the trait values at the required time steps.
 
@@
 
-\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-42-1.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-55-1.pdf}
 
@@
 \item
-  \texttt{BM.process}: this is the well famous \href{https://en.wikipedia.org/wiki/Brownian_motion}{Brownian motion process}.
+  \texttt{BM.process}: this is the well known \href{https://en.wikipedia.org/wiki/Brownian_motion}{Brownian motion process}.
 \end{itemize}
 
-\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-43-1.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-56-1.pdf}
 
@@
 
-\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-44-1.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-57-1.pdf}
 
@@
   \texttt{no.process}: this process has\ldots{} no process.
-In other words, this is a non time dependent process (the simulated value does not depends on the ancestors' value nor the branch length).
-It's basically a place holder for a random sampling function like \texttt{rnorm} (default), \texttt{runif}, \texttt{rlnorm}, etc\ldots{}
+In other words, this is a non time dependent process; the simulated value does not depends on the ancestors' value nor the branch length.
+It's basically a place holder for a random sampling function like \texttt{rnorm} (default), \texttt{runif}, \texttt{rlnorm}, etc.
 \end{itemize}
 
-\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-45-1.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-58-1.pdf}
 
@@
 \item
-  \texttt{multi.peak.process}: this process is a modified version of the \texttt{OU.process} that can take multiple local optima.
-The default OU process has one optimum towards which the values are drawn with the alpha parameter (the elastic band) which is usually 0.
+  \texttt{multi.peak.process}: this process is a modified version of the \texttt{OU.process} that can take multiple local long-term mean values.
+The default OU process has one long-term mean towards which the values are drawn with the alpha parameter (the elastic band).
+The single long-term mean is usually 0.
 However, with this \texttt{multi.peak.process} we can set multiple values towards which values can be attracted with the same alpha parameter.
@@
 
-\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-46-1.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-59-1.pdf}
 
@@
 \item
-  \texttt{repulsion.process}: this is a modified version of the \texttt{BM.process} where instead of accumulating normally through time, new trait values are more likely to be different than their ancestor following a \texttt{repulsion} parameter.
+  \texttt{repulsion.process}: this is a modified version of the \texttt{BM.process} where instead of accumulating gradually through time, new trait values are more likely to be different to the traits of their ancestor following a \texttt{repulsion} parameter.
 \end{itemize}
 
-\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-47-1.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-60-1.pdf}
+
+\begin{itemize}
+\tightlist
+\item
+  \texttt{discrete.process}: this one generates discrete trait values based on a transition matrix (you can use the utility function \texttt{transition.matrix} to generate one).
+\end{itemize}
+
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-61-1.pdf}
+
+Of course here the plotting of a discrete trait is not that informative\ldots{}
 
@@
 
-\textbf{This bit is more for development.}
+\textbf{This bit is still in development.}
 We highly suggest leaving \texttt{test\ =\ TRUE} so that \texttt{make.traits} returns an error if a process or its additional arguments (\texttt{process.args}) are not formatted correctly.
@@
 However, in some specific cases (again, probably mainly for development and debugging) it could be useful to skip the tests using \texttt{test\ =\ FALSE}.
+
+\hypertarget{mapping-traits-on-a-tree-map.traits}{%
+\section{\texorpdfstring{Mapping traits on a tree: \texttt{map.traits}}{Mapping traits on a tree: map.traits}}\label{mapping-traits-on-a-tree-map.traits}}
+
+Although \texttt{make.traits} is mainly designed for \texttt{treats} simulations, i.e.~simulating both the trees AND the traits at the same time, it is possible to use an already simulated or estimated tree and generating one or more traits on it using the \texttt{map.traits} function.
+This function takes the trait input as a \texttt{"traits"} object (from \texttt{make.traits}) and a tree topology with branch length.
+Note that you can also use \texttt{"traits"} objects with multiple or multi-dimensional traits as well as multiple phylogenies (\texttt{"multiPhylo"}).
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# Simulating a random tree with branch length}
+\NormalTok{my\_tree \textless{}{-}}\StringTok{ }\KeywordTok{rtree}\NormalTok{(}\DecValTok{20}\NormalTok{)}
+
+\CommentTok{\#\# Creating three different traits objects:}
+\CommentTok{\#\# A Brownian Motion}
+\NormalTok{bm\_process \textless{}{-}}\StringTok{ }\KeywordTok{make.traits}\NormalTok{(}\DataTypeTok{process =}\NormalTok{ BM.process)}
+\CommentTok{\#\# An Ornstein{-}Uhlenbeck process}
+\NormalTok{ou\_process \textless{}{-}}\StringTok{ }\KeywordTok{make.traits}\NormalTok{(}\DataTypeTok{process =}\NormalTok{ OU.process)}
+\CommentTok{\#\# No process (just randomly drawing values from a normal distribution)}
+\NormalTok{no\_process \textless{}{-}}\StringTok{ }\KeywordTok{make.traits}\NormalTok{(}\DataTypeTok{process =}\NormalTok{ no.process)}
+
+\CommentTok{\#\# Mapping the three traits on the phylogeny}
+\NormalTok{bm\_traits \textless{}{-}}\StringTok{ }\KeywordTok{map.traits}\NormalTok{(bm\_process, my\_tree)}
+\NormalTok{ou\_traits \textless{}{-}}\StringTok{ }\KeywordTok{map.traits}\NormalTok{(ou\_process, my\_tree)}
+\NormalTok{no\_traits \textless{}{-}}\StringTok{ }\KeywordTok{map.traits}\NormalTok{(no\_process, my\_tree)}
+
+\CommentTok{\#\# Plotting the topology and the different traits}
+\KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow =} \KeywordTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{,}\DecValTok{2}\NormalTok{))}
+\KeywordTok{plot}\NormalTok{(my\_tree, }\DataTypeTok{main =} \StringTok{"Base topology"}\NormalTok{)}
+\KeywordTok{plot}\NormalTok{(bm\_traits, }\DataTypeTok{main =} \StringTok{"Mapped BM"}\NormalTok{)}
+\KeywordTok{plot}\NormalTok{(ou\_traits, }\DataTypeTok{main =} \StringTok{"Mapped OU"}\NormalTok{)}
+\KeywordTok{plot}\NormalTok{(no\_traits, }\DataTypeTok{main =} \StringTok{"Mapped normal trait"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-62-1.pdf}
 
@@
 
+\begin{figure}
+\centering
+\includegraphics{make.modifiers.png}
+\caption{\texttt{make.modifiers}: The three main inputs to make modifiers are functions to modify the \texttt{branch.length} (waiting) process controlling the ``growth'' of the tree; the \texttt{selection} (selecting) process controlling the selection of the tip on which to apply speciation or extinction; and the \texttt{speciation} (speciating) process controlling whether the select tip goes extinct or speciates.
+The allowed arguments are detailed \protect\hyperlink{allowarguments}{here}.}
+\end{figure}
+
 \hypertarget{the-default-modifier-how-the-process-is-working}{%
@@
 
-The modifiers modify the core of the birth death process as implemented in \texttt{treats}.
-
+The modifiers modify the core of the birth-death process as implemented in \texttt{treats}.
 By default, the birth-death process in \texttt{treats} uses this modifier:
@@
 ## }
-## <bytecode: 0x55e27b08d1b8>
+## <bytecode: 0x55b093879f20>
 ## <environment: namespace:treats>
@@
 ## }
-## <bytecode: 0x55e27b11eb90>
+## <bytecode: 0x55b093879270>
 ## <environment: namespace:treats>
@@
 ## }
-## <bytecode: 0x55e27b127560>
+## <bytecode: 0x55b09387f640>
 ## <environment: namespace:treats>
@@
 
-This contains a lot of information, much of it is actually not useful for using the modular aspects of \texttt{treats} at a high user level.
-I.e.
-unless you want to code very specific things, you won't need most of the information.
-The essential are these three functions in the elements named \texttt{"waiting\$fun"}, \texttt{"selecting\$fun"} and \texttt{"speciating\$fun"}:
+This contains a lot of information, much of it is actually not useful for using the modular aspects of \texttt{treats} at a high user level, i.e.~unless you want to code very specific things, you won't need most of the information.
+The essentials are these three functions in the elements named \texttt{"waiting\$fun"}, \texttt{"selecting\$fun"} and \texttt{"speciating\$fun"}:
 
@@
 \item
-  \texttt{"waiting\$fun"} is the branch length function (defined in details below) which returns a randomly drawn number from an exponential distribution with the rate of the number of taxa multiplied by the speciation and extinction rate: \(n \times (\lambda + \mu)\) (here is uses the extra function \texttt{sum} for internal modularity reasons).
+  \texttt{"waiting\$fun"} is the branch length function (defined in detail below) which returns a randomly drawn number from an exponential distribution with the rate of the number of taxa multiplied by the speciation and extinction rate: \(n \times (\lambda + \mu)\) (here is uses the extra function \texttt{sum} for internal modularity reasons).
 \textbf{This function is responsible for the growth of the length/age of the tree}
@@
 
-With \texttt{lineage\$n} being the number of lineages and \texttt{bd.params\$speciation} and \texttt{bd.params\$extinction} the speciation and extinction parameters (these specific terms are defined in details below).
+With \texttt{lineage\$n} being the number of lineages and \texttt{bd.params\$speciation} and \texttt{bd.params\$extinction} the speciation and extinction parameters (these specific terms are defined in detail below).
 
@@
 \item
-  \texttt{"selecting\$fun"} is the selection function (defined in details below) which, after the waiting time defined above returns a randomly selected lineage (a tip) among the existing ones.
+  \texttt{"selecting\$fun"} is the selection function (defined in detail below) which, after the waiting time defined above returns a randomly selected lineage (a tip) among the existing ones.
 \textbf{This function is responsible for the branch selection}.
@@
 \item
-  \texttt{"speciating\$fun"} is the speciation function (defined in details below) which randomly draws a number (\texttt{runif(1)}) and depending on the speciation and extinction parameters makes the selected lineage (from 2.) speciate (\texttt{TRUE}) or go extinct (\texttt{FALSE}).
+  \texttt{"speciating\$fun"} is the speciation function (defined in detail below) which randomly draws a number (\texttt{runif(1)}) and depending on the speciation and extinction parameters makes the selected lineage (from 2.) speciate (\texttt{TRUE}) or go extinct (\texttt{FALSE}).
 \textbf{This function is responsible for the speciation or extinction of species}.
@@
 \begin{Highlighting}[]
-\KeywordTok{runif}\NormalTok{(}\DecValTok{1}\NormalTok{) }\OperatorTok{\textless{}}\StringTok{ }\NormalTok{(bd.params}\OperatorTok{$}\NormalTok{speciation}\OperatorTok{/}\NormalTok{(bd.params}\OperatorTok{$}\NormalTok{speciation }\OperatorTok{+}\StringTok{ }\NormalTok{bd.params}\OperatorTok{$}\NormalTok{extinction))}\ErrorTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-We will have a look at all these aspects below in more details which should make things clearer (but you can always refer to the default here for information).
+\KeywordTok{runif}\NormalTok{(}\DecValTok{1}\NormalTok{) }\OperatorTok{\textless{}}\StringTok{ }\NormalTok{(bd.params}\OperatorTok{$}\NormalTok{speciation}\OperatorTok{/}\NormalTok{(bd.params}\OperatorTok{$}\NormalTok{speciation }\OperatorTok{+}\StringTok{ }\NormalTok{bd.params}\OperatorTok{$}\NormalTok{extinction))}
+\end{Highlighting}
+\end{Shaded}
+
+We will have a look at all these aspects below in more detail which should make things clearer (but you can always refer to the default here for information).
 
@@
 
-The first argument in \texttt{"modifiers"} is the branch length function (\texttt{branch.length}) this is the function that will be executed in \texttt{treats} to generate branch length.
-Note that in the \texttt{treats} algorithm, branch length is not generated \emph{directly} but actually results of the waiting time.
+The first argument in \texttt{"modifiers"} is the branch length function (\texttt{branch.length}) this is the function that will be executed in \texttt{treats} to generate branch lengths.
+Note that in the \texttt{treats} algorithm, branch length is not generated \emph{directly} but as the result of the waiting time.
 In other words, the \texttt{branch.length} function just affects waiting time for all taxa present at any time in the simulation.
-These taxa can then either go extinct (stopping the ``growth'' of it's branch length) or survive (continuing the ``growth'').
+These taxa can then either go extinct (stopping the ``growth'' of its branch length) or survive (continuing the ``growth'').
 
@@
 Of course, the point of the modularity here is that you can provide your own function for generating branch length.
-For example, we might be interested in what our tree would look like if we'd use a simple constant branch length generation (instead of randomly drawing it from an exponential distribution).
+For example, we might be interested in what our tree would look like if we use a simple constant branch length generation (instead of randomly drawing it from an exponential distribution).
 We can do so by declaring our own \texttt{branch.length} function and adding it to a \texttt{"modifiers"} object.
@@
 
-\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-58-1.pdf}
-
-It is of course possible to use more complex branch length modifiers that intakes different conditions and specific modification rather than simply always output a value of one.
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-73-1.pdf}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow =} \KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{,}\DecValTok{1}\NormalTok{))}
+\end{Highlighting}
+\end{Shaded}
+
+It is of course possible to use more complex branch length modifiers that take different conditions and specific modification rather than simply always outputing a value of one.
 
 \hypertarget{allowarguments}{%
-\subsection{The allowed arguments}\label{allowarguments}}
+\subsection{The modifier arguments}\label{allowarguments}}
 
@@
 \item
-  \texttt{bd.params}: a named list containing \texttt{"numeric"} values that contains the birth death parameters (at least \texttt{"speciation"} and \texttt{"extinction"});
+  \texttt{bd.params}: a named list containing \texttt{"numeric"} values that contains the birth-death parameters (at least \texttt{"speciation"} and \texttt{"extinction"});
 \item
@@
 \item
-  \texttt{trait.values}: a \texttt{"matrix"} containing \texttt{"numeric"} values with the trait names as column names and the lineages ID as row numbers (you can use it with the function \texttt{parent.traits} to access the trait of the previous node for example).
+  \texttt{trait.values}: a \texttt{"matrix"} containing \texttt{"numeric"} values with the trait names as column names and the lineage ID as row numbers (you can use it with the function \texttt{parent.traits} to access the trait of the previous node for example).
 \item
@@
 \item
-  \texttt{lineage\$drawn}: a single \texttt{"integer"}: the ID of the selected lineage;
+  \texttt{lineage\$drawn}: a single \texttt{"integer"}: the ID of the selected lineage; this must be a number in \texttt{1:lineage\$n};
 \item
@@
 Therefore any argument can be left undeclared or missing and is always handled internally.
-For example, if you did not declare \texttt{lineage\$n} as a function argument but are using \texttt{lineage\$n} in the function, \texttt{lineage\$n} will be detected and treated as a current argument automatically as set accordingly within the birth death process (e.g.~\texttt{lineage\$n} will be set to the current number of taxa every iteration of the process).
-
-For example, we can create a function that increases branch length proportional to the number of species ``alive'' at each time of the simulation in a discrete way.
-I.e.
-for discrete numbers of taxa, the branch length increases by jumps (ten fold) every 5 taxa:
+For example, if you did not declare \texttt{lineage\$n} as a function argument but are using \texttt{lineage\$n} in the function, \texttt{lineage\$n} will be detected and treated as a current argument automatically as set accordingly within the birth-death process (e.g.~\texttt{lineage\$n} will be set to the current number of taxa every iteration of the process).
+
+For example, we can create a function that increases branch length proportional to the number of species ``alive'' at each time of the simulation in a discrete way, i.e.~for discrete numbers of taxa, the branch length increases by jumps (ten fold) every five taxa:
 
@@
 
-\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-61-1.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-76-1.pdf}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow =} \KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{,}\DecValTok{1}\NormalTok{))}
+\end{Highlighting}
+\end{Shaded}
 
@@
 
-The \texttt{selection} function is used in the birth death process to know which lineage to select when running a speciation (or extinction!) event.
+The \texttt{selection} function is used in the birth-death process to know which lineage to select when running a speciation (or extinction!) event.
 By default, this function randomly selects one taxon that is currently not extinct (using: \texttt{sample(1:lineage\$n,\ 1))}.
-Similarly as \texttt{branch.length} it is possible to modify this part of the birth death process.
-For example, we could simply select always the last created lineage (to create a ``ladder'' or most asymmetric tree):
+Similarly to \texttt{branch.length}, it is possible to modify this part of the birth-death process.
+For example, we could simply select the last created lineage to create a ``ladder'' or most asymmetric tree:
 
@@
 \begin{quote}
-Note that here the function can only intake the \protect\hyperlink{allowarguments}{allowed arguments as described above} (here \texttt{lineage\$n}: the number of current living taxa).
+Note that here the function can only take the \protect\hyperlink{allowarguments}{allowed arguments as described above} (here \texttt{lineage\$n}: the number of current living taxa).
 \end{quote}
 
-We can then create a \texttt{"modifiers"} object the same way as before using this time the \texttt{selection} argument:
+We can then create a \texttt{"modifiers"} object the same way as before this time using the \texttt{selection} argument:
 
@@
 
-\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-63-1.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-78-1.pdf}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow =} \KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{,}\DecValTok{1}\NormalTok{))}
+\end{Highlighting}
+\end{Shaded}
 
@@
 
-\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-64-1.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-79-1.pdf}
 
@@
 
-The third function that can be used to modify the birth death process is the \texttt{speciation} function.
-This one is used during the birth death process to decide whether a lineages speciates (creating a node and two new lineages) or goes extinct (creating a tip).
+The third function that can be used to modify the birth-death process is the \texttt{speciation} function.
+This one is used during the birth-death process to decide whether a lineage speciates (creating a node and two new lineages) or goes extinct (creating a tip).
 
@@
 
-Creating a \texttt{"modifiers"} with a \texttt{speciation} function works the same way as for \texttt{branch.length} and \texttt{selection} but the function that will be used needs to output a logical value (see table \protect\hyperlink{summarymodifiers}{below}).
+Creating \texttt{"modifiers"} with a \texttt{speciation} function works the same way as for \texttt{branch.length} and \texttt{selection} but the function that will be used needs to output a logical value (see table \protect\hyperlink{summarymodifiers}{below}).
 Once the function is created simply input your function for speciation in the modifier and run the \texttt{treats} function with that modifier:
@@
 \KeywordTok{plot}\NormalTok{(modified\_tree, }\DataTypeTok{main =} \StringTok{"Random extinction}\CharTok{\textbackslash{}n}\StringTok{modifier"}\NormalTok{)}
+\KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow =} \KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{,}\DecValTok{1}\NormalTok{))}
 \end{Highlighting}
@@
 \begin{quote}
-Note how loads of lineages end up going extinct even if the extinction parameter is set to 0!
+Note how loads of lineages go extinct even if the extinction parameter is set to 0!
 \end{quote}
 
-And again, we can make some more advanced modifiers: for example, one where a tip always goes extinct if their ancestor has a negative trait value (here we will also introduce the utility function \texttt{parent.trait} that automatically selects the trait values of the parent of the current lineage.
+And again, we can make some more advanced modifiers: for example, one where a tip always goes extinct if their ancestor has a negative trait value.
+Here we will also introduce the utility function \texttt{parent.trait} that automatically selects the trait values of the parent of the current lineage.
 
@@
 \CommentTok{\#\# A modifier for removing tips with negative values}
-\NormalTok{bias.trait \textless{}{-}}\StringTok{ }\ControlFlowTok{function}\NormalTok{(trait.values, lineage) \{}
+\NormalTok{modify.trait \textless{}{-}}\StringTok{ }\ControlFlowTok{function}\NormalTok{(trait.values, lineage) \{}
     \ControlFlowTok{if}\NormalTok{(}\KeywordTok{parent.traits}\NormalTok{(trait.values, lineage) }\OperatorTok{\textless{}}\StringTok{ }\DecValTok{0}\NormalTok{) \{}
@@
 \NormalTok{    \} }\ControlFlowTok{else}\NormalTok{ \{}
-        \CommentTok{\#\# Speciate}
+        \CommentTok{\#\# Speciate!}
         \KeywordTok{return}\NormalTok{(}\OtherTok{TRUE}\NormalTok{)}
@@
 \CommentTok{\#\# Creating the modifier}
-\NormalTok{biased\_trait \textless{}{-}}\StringTok{ }\KeywordTok{make.modifiers}\NormalTok{(}\DataTypeTok{speciation =}\NormalTok{ bias.trait)}
+\NormalTok{modified\_trait \textless{}{-}}\StringTok{ }\KeywordTok{make.modifiers}\NormalTok{(}\DataTypeTok{speciation =}\NormalTok{ modify.trait)}
 
@@
 \KeywordTok{set.seed}\NormalTok{(}\DecValTok{1}\NormalTok{)}
-\NormalTok{biased\_trait\_treats \textless{}{-}}\StringTok{ }\KeywordTok{treats}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ bd\_params,}
+\NormalTok{modified\_trait\_treats \textless{}{-}}\StringTok{ }\KeywordTok{treats}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ bd\_params,}
                           \DataTypeTok{stop.rule =}\NormalTok{ stop\_rule,}
                           \DataTypeTok{traits    =}\NormalTok{ BM\_trait,}
-                          \DataTypeTok{modifiers =}\NormalTok{ biased\_trait)}
+                          \DataTypeTok{modifiers =}\NormalTok{ modified\_trait)}
 
@@
 \KeywordTok{plot}\NormalTok{(default\_treats, }\DataTypeTok{main =} \StringTok{"Default trait and tree"}\NormalTok{)}
-\KeywordTok{plot}\NormalTok{(biased\_trait\_treats, }\DataTypeTok{main =} \StringTok{"Biased trait and tree"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-67-1.pdf}
+\KeywordTok{plot}\NormalTok{(modified\_trait\_treats, }\DataTypeTok{main =} \StringTok{"Biased trait and tree"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-82-1.pdf}
 
@@
 
-In the examples above, we have seen how to specify modifications to the birth death process (via \texttt{branch.length}, \texttt{selection} and \texttt{speciation}), however, one might note that these modifications are not dynamic.
+In the examples above, we have seen how to specify modifications to the birth-death process (via \texttt{branch.length}, \texttt{selection} and \texttt{speciation}), however, these modifications are not dynamic.
 In other words, throughout the process, the modifications remain constant (even if they are conditional).
-It is however possible to code the \texttt{"modifiers"} so that they can be affected by \texttt{"events"} objects (see next chapter \protect\hyperlink{makeevents}{on \texttt{events}}).
-
-To do so, you can formally declare conditions (\texttt{condition}) and modifications (\texttt{modify}) as internal functions that can then be modified my an \texttt{"events"} object.
+It is however, possible to code the \texttt{"modifiers"} so that they can be affected by \texttt{"events"} objects (see next chapter \protect\hyperlink{makeevents}{on \texttt{events}}).
+
+To do so, you can formally declare conditions (\texttt{condition}) and modifications (\texttt{modify}) as internal functions that can then be modified by an \texttt{"events"} object.
 \texttt{condition} and \texttt{modify} are hard coded in the \texttt{branch.length} function that they concern, i.e.~they are variables (functions) within the function.
@@
     \CommentTok{\#\# Speciate?}
-    \KeywordTok{return}\NormalTok{(trigger\_event }\OperatorTok{\textless{}}\StringTok{ }\NormalTok{(bd.params}\OperatorTok{$}\NormalTok{speciation}\OperatorTok{/}\NormalTok{(bd.params}\OperatorTok{$}\NormalTok{speciation }\OperatorTok{+}\StringTok{ }\NormalTok{bd.params}\OperatorTok{$}\NormalTok{extinction)))}
+    \KeywordTok{return}\NormalTok{(trigger\_event }\OperatorTok{\textless{}}\StringTok{ }\NormalTok{(bd.params}\OperatorTok{$}\NormalTok{speciation}\OperatorTok{/}
+\StringTok{                              }\NormalTok{(bd.params}\OperatorTok{$}\NormalTok{speciation }\OperatorTok{+}\StringTok{ }\NormalTok{bd.params}\OperatorTok{$}\NormalTok{extinction)))}
 \NormalTok{\}}
@@
 
-It is possible with some specific condition to modify this trigger by providing a \texttt{condition} and \texttt{modify} function to the speciation function.
+It is possible with some specific conditions to modify this trigger by providing a \texttt{condition} and \texttt{modify} function to the speciation function.
 That is, if the \texttt{condition} is met, apply the \texttt{modify} function to the algorithm.
@@
 
-Effectively, this will internally modify the speciation function as follows
+Effectively, this will internally modify the speciation function as follows:
 
@@
 These \texttt{condition} and \texttt{modify} functions can be applied to all the \texttt{modifiers} elements (\texttt{selection}, \texttt{branch.length} and \texttt{speciation}).
-However, they are typically used in \texttt{events} that will modify\ldots{} the \texttt{modifiers} (\protect\hyperlink{makeevents}{see the \texttt{events} section})!
+However, they are typically used in \texttt{events} that will modify the \texttt{modifiers} (\protect\hyperlink{makeevents}{see the \texttt{events} section})!
 
@@
 
-One other major feature of the \texttt{treats} package is to allow simulations to run with specified events to occur during the simulation.
+One other major feature of the \texttt{treats} package is that it allows simulations to run with specified events that occur during the simulation.
 These are typically events that can drastically change the course of the simulation.
-For example, you might want simulate some mass extinction at some specific point in time.
+For example, you might want simulate a mass extinction at some specific point in time.
 However, these can also be more subtle, like the internal change of parameter values when reaching a specific trait value.
@@
 * the \texttt{condition} which designates when to trigger the event;
-* the \texttt{modification} which designates what to exactly modify when the event is triggered.
+* the \texttt{modification} which designates what to modify when the event is triggered.
 
@@
 First let's focus on these three main arguments:
+
+\begin{figure}
+\centering
+\includegraphics{make.events.png}
+\caption{\texttt{make.events}: events are encoded using three main arguments: the \texttt{target}, specifying \emph{what} the event is modifying, the \texttt{condition} specifying \emph{when} to trigger the event, and \texttt{modification}, specifying \emph{how} the event modifies the simulations.}
+\end{figure}
 
@@
 The targets that are currently available are:
-* \texttt{"taxa"} to modify anything linked to the \protect\hyperlink{allowarguments}{\texttt{lineage} list}. E.g. making half of the living taxa go extinct.
-* \texttt{"bd.params"} to modify anything linked to the \protect\hyperlink{makebdparams}{\texttt{bd.params} object}.
+
+\begin{itemize}
+\tightlist
+\item
+  \texttt{"taxa"} to modify anything linked to the \protect\hyperlink{allowarguments}{\texttt{lineage} list}. e.g.~making half of the living taxa go extinct.
+\item
+  \texttt{"bd.params"} to modify anything linked to the \protect\hyperlink{makebdparams}{\texttt{bd.params} object}.
 For example you might want to change the distribution of one of the parameter after some \texttt{conditions}.
 This is typically done by updating the object using the argument \texttt{update} from \texttt{make.bd.params}.
-* \texttt{"traits"} to modify anything linked to the \protect\hyperlink{maketraits}{\texttt{traits} object}.
+\item
+  \texttt{"traits"} to modify anything linked to the \protect\hyperlink{maketraits}{\texttt{traits} object}.
 For example you might want to change the trait process after some \texttt{conditions}.
 This is typically done by updating the object using the argument \texttt{update} from \texttt{make.traits}.
-* \texttt{"modifiers"} to modify anything linked to the \protect\hyperlink{makebdmodifiers}{\texttt{modifiers} object}.
+\item
+  \texttt{"modifiers"} to modify anything linked to the \protect\hyperlink{makebdmodifiers}{\texttt{modifiers} object}.
 For example you might want to change the speciation rule after some \texttt{conditions}.
 This is typically done by updating the object using the argument \texttt{update} from \texttt{make.modifiers}.
-* \texttt{"founding"} this target is a bit more special and allows you run a nested \texttt{treats} object in the simulation. It is cover in a \protect\hyperlink{founding}{specific section below}.
-
-We will see some example of these targets by illustrating the \texttt{conditions} and \texttt{modifications} below.
+\item
+  \texttt{"founding"} this target is a bit more specialised and allows you run a nested \texttt{treats} object in the simulation.
+It is covered in a \protect\hyperlink{founding}{specific section below}.
+\end{itemize}
+
+We will see some examples of these targets in the \texttt{conditions} and \texttt{modifications} described below.
 
@@
 
-All \texttt{condition} functions in \texttt{treats} take at least to arguments: \texttt{x} for the variable of interest (e.g.~time, number of taxa, trait value, etc\ldots) and \texttt{condition}, the relational operator to evaluate.
+All \texttt{condition} functions in \texttt{treats} take at least two arguments: \texttt{x} for the variable of interest (e.g.~time, number of taxa, trait value, etc.) and \texttt{condition}, the relational operator to evaluate.
 A relational operator is the proper (fancy) term designating all the comparisons you're regularly using in \texttt{R} like \texttt{==} (is equal?) \texttt{\textless{}} (is smaller?) \texttt{\textgreater{}=} (is bigger or equal?).
-You can get the full list in the base manual (using \texttt{?Comparison} or any of the relational operator in a function form).
-
-\begin{quote}
-Note on the function form of a operator in \texttt{R}: if you weren't aware, functions in \texttt{R} are \emph{usually} described in the format \texttt{function(arguments)}, however many other functions you are commonly using are written as operators, for example (\texttt{1\ +\ 1}).
-These are still functions though!
-And you can always use them in the classic function format by quoting them in back ticks: the following does the same as \texttt{(1+1)}: \texttt{\textasciigrave{}+\textasciigrave{}(1,\ 1)}.
-You can use that for parsing your relational operators as functions in \texttt{condition}.
-If you want you condition to be equal, use the syntax \texttt{condition\ =\ \textasciigrave{}==\textasciigrave{}}.
-More details on how functions \emph{really} work in the \href{https://adv-r.hadley.nz/functions.html}{Advanced \texttt{R} book}!
-\end{quote}
+You can get the full list in the R base manual (using \texttt{?Comparison} or any of the relational operator in a function form).
+For more details on how functions \emph{really} work in R see the \href{https://adv-r.hadley.nz/functions.html}{Advanced \texttt{R} book}!
 
@@
 \item
-  \texttt{time.condition} will trigger the \texttt{event} once a certain time is reached in the simulations and is the simplest/more basic condition with no arguments other than the time required.
-For example \texttt{time.condition(4,\ condition\ =\ \textasciigrave{}==\textasciigrave{})} will trigger the \texttt{event} once the simulations reached 4 units of time.
+  \texttt{time.condition} will trigger the \texttt{event} once a certain time is reached in the simulations and is the simplest/most basic condition with no arguments other than the time required.
+For example \texttt{time.condition(4,\ condition\ =\ \textasciigrave{}==\textasciigrave{})} will trigger the \texttt{event} once the simulations reaches 4 time units.
 Easy.
 \item
-  \texttt{taxa.condition} will trigger the \texttt{event} once a certain number of taxa is reached.
+  \texttt{taxa.condition} will trigger the \texttt{event} once a certain number of taxa are reached.
 This can be considered including or excluding fossil species.
-For example \texttt{taxa.condition(42,\ condition\ =\ \textasciigrave{}\textgreater{}=\textasciigrave{},\ living\ =\ TRUE)} will trigger the \texttt{event} if there is at least 42 living species.
+For example \texttt{taxa.condition(42,\ condition\ =\ \textasciigrave{}\textgreater{}=\textasciigrave{},\ living\ =\ TRUE)} will trigger the \texttt{event} if there are at least 42 living taxa.
 \item
@@
 
-After defining the \texttt{event}'s \texttt{target} and \texttt{condition}, you'll need to also specify what it should modify.
-These are basically functions that should modify a specific aspect of the target
+After defining the \texttt{event}, \texttt{target} and \texttt{condition}, you also need to specify what it should modify.
+These are functions that should modify a specific aspect of the target.
 Depending on the target you can modify the following:
@@
 \item
-  if the target is \texttt{"bd.params"} you can modify the \protect\hyperlink{makebdparams}{\texttt{bd.params} object} using \texttt{update.bd.params}.
-\item
-  if the target is \texttt{"traits"} you can modify the \protect\hyperlink{maketraits}{\texttt{traits} object} using \texttt{update.traits}.
-\item
-  if the target is \texttt{"modifiers"} you can modify the \protect\hyperlink{makemodifiers}{\texttt{modifiers} object} using \texttt{update.modifiers}.
+  if the target is \texttt{"bd.params"} you can modify the \protect\hyperlink{makebdparams}{\texttt{bd.params} object} using \texttt{bd.params.update}.
+\item
+  if the target is \texttt{"traits"} you can modify the \protect\hyperlink{maketraits}{\texttt{traits} object} using \texttt{traits.update}.
+\item
+  if the target is \texttt{"modifiers"} you can modify the \protect\hyperlink{makemodifiers}{\texttt{modifiers} object} using \texttt{modifiers.update}.
 \item
@@
 
-The most straightforward example if for \texttt{modifications} on \texttt{bd.params}, \texttt{traits} or \texttt{modifiers} objects since they use the same syntax as for their generic \texttt{make.X} function.
+The most straightforward example is for \texttt{modifications} on \texttt{bd.params}, \texttt{traits} or \texttt{modifiers} objects because they use the same syntax as for their generic \texttt{make.X} function.
 For example, for \texttt{make.traits}, you can update a \texttt{trait} using the \texttt{update} argument as follows:
@@
 
-So basically the functions \texttt{update.X} apply the update to the object \texttt{X} when the event happens.
+So the function \texttt{update.X} applies the update to the object \texttt{X} when the event happens.
 The following \texttt{event} updates the \texttt{bd.params} object by setting the extinction parameter to 1/3 when reaching a 10 species:
@@
             \DataTypeTok{condition =} \KeywordTok{taxa.condition}\NormalTok{(}\DecValTok{10}\NormalTok{, }\DataTypeTok{condition =} \StringTok{\textasciigrave{}}\DataTypeTok{\textgreater{}=}\StringTok{\textasciigrave{}}\NormalTok{),}
-            \DataTypeTok{modification =} \KeywordTok{update.bd.params}\NormalTok{(}\DataTypeTok{extinction =} \DecValTok{1}\OperatorTok{/}\DecValTok{3}\NormalTok{))}
+            \DataTypeTok{modification =} \KeywordTok{bd.params.update}\NormalTok{(}\DataTypeTok{extinction =} \DecValTok{1}\OperatorTok{/}\DecValTok{3}\NormalTok{))}
 \end{Highlighting}
@@
 \begin{quote}
-Note that by default events are triggered only once in the whole simulation, so although the example above states that the condition is reaching at least 10 taxa, it will not trigger every times it reaches more than 10 taxa.
-You can change the number of times the events can be triggered using the argument \texttt{replications} (by default it's set to \texttt{replications\ =\ 0} for triggering it only once).
+Note that by default events are triggered only once across the whole simulation, so although the example above states that the condition is reaching at least 10 taxa, it will not trigger every time it reaches more than 10 taxa, only the first time.
+You can change the number of times the events can be triggered using the argument \texttt{replications} (by default it's set to \texttt{replications\ =\ 0} for triggering the event only once).
 \end{quote}
@@
 
-Here are some example illustrating how to generate events.
+Here are some examples illustrating how to generate events.
 For a simple example, we can create a extinction event that will remove 80\% of species after reaching time 4:
@@
 
-\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-75-1.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-90-1.pdf}
 
@@
                   \DataTypeTok{condition =} \KeywordTok{trait.condition}\NormalTok{(}\DecValTok{3}\NormalTok{, }\DataTypeTok{condition =} \StringTok{\textasciigrave{}}\DataTypeTok{\textgreater{}}\StringTok{\textasciigrave{}}\NormalTok{, }\DataTypeTok{what =}\NormalTok{ upper}\FloatTok{.95}\NormalTok{),}
-                  \DataTypeTok{modification =} \KeywordTok{update.traits}\NormalTok{(}\DataTypeTok{process =}\NormalTok{ OU.process))}
+                  \DataTypeTok{modification =} \KeywordTok{traits.update}\NormalTok{(}\DataTypeTok{process =}\NormalTok{ OU.process))}
 
@@
 
-\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-76-1.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-91-1.pdf}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow =} \KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{,}\DecValTok{1}\NormalTok{))}
+\end{Highlighting}
+\end{Shaded}
 
@@
 
-Founding events are a specific \texttt{target} for \texttt{events} that allows you to\ldots{} simulate a birth-death process within the current one!
-It basically allows you to simulate a specific \texttt{treats} process i.e.~using the \texttt{treats} function with it's own \texttt{traits}, \texttt{modifiers}, \texttt{bd.params} and \texttt{events} (and yes, that's including \texttt{events} that have their own \texttt{founding} event)\ldots{}
+Founding events are a specific \texttt{target} for \texttt{events} that allows you to simulate a birth-death process within the current one!
+It allows you to simulate a specific \texttt{treats} process using the \texttt{treats} function with its own \texttt{traits}, \texttt{modifiers}, \texttt{bd.params} and \texttt{events} (and yes, that's including \texttt{events} that have their own \texttt{founding} event).
 This is basically the ultimate nested boss of modularity!
 
-The founding event will basically run an internal \texttt{treats} process (i.e.~simulating a tree and, optionally, some data) resulting in a founding sub-tree.
-It will then branch this sub-tree to the rest of the simulation that continued normally in the mean time.
+The founding event will run an internal \texttt{treats} process (i.e.~simulating a tree and, optionally, some data) resulting in a founding sub-tree.
+It will then branch this sub-tree onto the rest of the simulation that continued normally in the mean time.
 You can specify the founding event using the inbuilt \texttt{founding.event} modification event.
-This function intakes the exact same arguments as \texttt{treats} to simulate the sub-tree with its own parameters.
+This function takes the exact same arguments as \texttt{treats} to simulate the sub-tree with its own parameters.
 Additionally, we will use the \texttt{additional.args} argument from \texttt{make.events} to specify a prefix for the founding tree tips (to make them easier to distinguish).
@@
 
-\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-77-1.pdf}
-
-Note that the nestedness here is potentially endless, for example, you can parse an \texttt{event} argument to the \texttt{founding.event} that will generate another founding event, etc\ldots{}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-92-1.pdf}
+
+Note that the nestedness here is potentially endless, for example, you can pass an \texttt{event} argument to the \texttt{founding.event} that will generate another founding event, etc.
 
@@
 
-In most examples above, birth-death parameters where set as fixed values, i.e.~a speciation rate of \(\lambda\) and an extinction rate of \(\mu\) throughout the simulations (with maybe some events modifying these rates).
-However, it is of course possible to set these rates as specific or changing distributions!
-You can do this using the \texttt{make.bd.params} function a provide either a \texttt{vector} of values:
+In most examples above, birth-death parameters were set as fixed values, i.e.~a speciation rate of \(\lambda\) and an extinction rate of \(\mu\) throughout the simulations (with maybe some events modifying these rates).
+However, it is possible to set these rates as specific or changing distributions.
+You can do this using the \texttt{make.bd.params} function to provide either a \texttt{vector} of values:
 
@@
 
-\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-79-1.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-94-1.pdf}
 
@@
 
-When using distributions for both the speciation and extinction parameters, you can run into the undesired problem of having an extinction rate that is higher than your speciation rate (and thus your tree dying out - unless this is the desired behaviour).
-You can avoid this problem by using the \texttt{joint} distribution argument this will ensure that the sampling of the extinction rate is always lower or equal to the speciation rate.
+When using distributions for both the speciation and extinction parameters, you can run into the (usually) undesired problem of having an extinction rate that is higher than your speciation rate and thus all the taxa in your tree dying out.
+You can avoid this problem by using the \texttt{joint} distribution argument.
+This will ensure that the sampling of the extinction rate is always lower or equal to the speciation rate.
 
@@
 
-The two other possible arguments for this function are \texttt{test} and \texttt{update} that work the same as for \protect\hyperlink{maketraits}{\texttt{make.traits}} or \protect\hyperlink{makemodifiers}{\texttt{make.modifiers}}
-
-If you are a visual person and your \texttt{bd.params} objects are getting a bit too complicated to remember, you can always quickly plot them (the function will sample from the \texttt{bd.params} object and show the results):
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow =} \KeywordTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{,}\DecValTok{2}\NormalTok{))}
+The two other possible arguments for this function are \texttt{test} and \texttt{update} that work the same way as \protect\hyperlink{maketraits}{\texttt{make.traits}} or \protect\hyperlink{makemodifiers}{\texttt{make.modifiers}}
+
+If you are a visual person and your \texttt{bd.params} objects are getting a bit too complicated to remember, you can always quickly plot them.
+The function will sample from the \texttt{bd.params} object and show the results:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow =} \KeywordTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{,}\DecValTok{3}\NormalTok{))}
 \KeywordTok{plot}\NormalTok{(}\KeywordTok{make.bd.params}\NormalTok{(), }\DataTypeTok{main =} \StringTok{"Default birth{-}death parameters"}\NormalTok{)}
@@
      \DataTypeTok{main =} \StringTok{"Joint uniform speciation and extinction"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-84-1.pdf}
-
-\begin{Shaded}
-\begin{Highlighting}[]
 \KeywordTok{plot}\NormalTok{(}\KeywordTok{make.bd.params}\NormalTok{(}\DataTypeTok{speciation =}\NormalTok{ rnorm, }\DataTypeTok{extinction =}\NormalTok{ runif,}
@@
      \DataTypeTok{main =} \StringTok{"Disjoint normal speciation and uniform extinction"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-84-2.pdf}
+\KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow =} \KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{,}\DecValTok{1}\NormalTok{))}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-99-1.pdf}
 
@@
 
-You can use the function \texttt{drop.things} to drop specific elements of the tree and data at the same time by providing the argument \texttt{what} to be \texttt{"fossils"} for tips that went extinct, \texttt{"livings"} tips that where alive at the end of the simulation or \texttt{"singles"} for internal nodes.
+You can use the function \texttt{drop.things} to drop specific elements of the tree and data at the same time by providing the argument \texttt{what\ =\ "fossils"} for tips that went extinct, or \texttt{what\ =\ "livings"} for tips that where alive at the end of the simulation or \texttt{what\ =\ "singles"} to drop internal singleton nodes.
 Alternatively you can use the function aliases \texttt{drop.fossils}, \texttt{drop.livings} or \texttt{drop.singles} for the exact same results:
@@
 
+\hypertarget{plots}{%
+\chapter{Plots}\label{plots}}
+
 \texttt{"treats"} objects can be directly plotted in \texttt{treats} using the S3 \texttt{plot.treats} function (or just \texttt{plot(x)} if \texttt{x} is of class \texttt{"treats"}).
@@
 \hypertarget{plotting-traits}{%
-\chapter{Plotting traits}\label{plotting-traits}}
+\section{Plotting traits}\label{plotting-traits}}
 
@@
      \DataTypeTok{cent.tend =}\NormalTok{ sd, }\DataTypeTok{quantiles =} \KeywordTok{c}\NormalTok{(}\DecValTok{10}\NormalTok{, }\DecValTok{30}\NormalTok{))}
-\end{Highlighting}
-\end{Shaded}
-
-\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-87-1.pdf}
+\KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow =} \KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{, }\DecValTok{1}\NormalTok{))}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-102-1.pdf}
 
 You can also control the number of replicates in the simulation by using the \texttt{simulations} option (the default is 50).
-Bigger numbers leads to more time but smoother looking plots while smaller ones are more stochastic:
+Bigger numbers lead to more time but smoother looking plots while smaller ones are more stochastic:
 
@@
 
-\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-88-1.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-103-1.pdf}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow =} \KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{,}\DecValTok{1}\NormalTok{))}
+\end{Highlighting}
+\end{Shaded}
 
 \hypertarget{plotting-treats-results}{%
-\chapter{Plotting treats results}\label{plotting-treats-results}}
+\section{\texorpdfstring{Plotting \texttt{treats} results}{Plotting treats results}}\label{plotting-treats-results}}
 
 If \texttt{treats} is used to plot only a tree (and outputs a \texttt{"phylo"} object), you can use the function \texttt{plot.phylo} from the \texttt{ape} package to plot your tree.
-You'll find all the tree plotting option in the \texttt{?plot.phylo} manual page.
+You'll find all the tree plotting options in the \texttt{?plot.phylo} manual page.
 
@@
 
-\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-89-1.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-104-1.pdf}
 
@@
 
-\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-90-1.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-105-1.pdf}
 
@@
 Branches linking them are grey lines.
-You can of course change this colour palette to something of you preference by calling the normal arguments that can be passed to \texttt{points} or \texttt{lines}.
+You can of course change this colour palette to whatever you prefer by calling the standard base R arguments that can be passed to \texttt{points} or \texttt{lines}.
 For example \texttt{pch} for the point type or \texttt{lty} for the line type.
@@
      \DataTypeTok{col =} \KeywordTok{c}\NormalTok{(}\DataTypeTok{tips =} \StringTok{"pink"}\NormalTok{, }\DataTypeTok{nodes =} \StringTok{"purple"}\NormalTok{), }\DataTypeTok{pch =} \DecValTok{21}\NormalTok{,}
-     \CommentTok{\#\# Changing edges colour and type}
+     \CommentTok{\#\# Changing edge colour and type}
      \DataTypeTok{lwd =} \DecValTok{2}\NormalTok{, }\DataTypeTok{lty =} \DecValTok{3}\NormalTok{, }\DataTypeTok{edges =} \StringTok{"yellow"}\NormalTok{,}
@@
 
-\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-91-1.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-106-1.pdf}
 
 \begin{quote}
-Note that you can modify colours using the \texttt{col} argument by providing a clear indication of what you want to colour (e.g.~\texttt{col\ =\ c(tips\ =\ "blue",\ fossils\ =\ "orange")} will apply the colours to the unambiguously named elements) or by providing a function that will scale the colours with time (e.g.~\texttt{col\ =\ grDevices::heat.colors} will use the specified function to change the elements colours with time).
+Note that in \texttt{plot.treats} you can modify colours using the \texttt{col} argument by providing a clear indication of what you want to colour (e.g.~\texttt{col\ =\ c(tips\ =\ "blue",\ fossils\ =\ "orange")} will apply the colours to the unambiguously named elements) or by providing a function that will scale the colours with time (e.g.~\texttt{col\ =\ grDevices::heat.colors} will use the specified function to change the colours of the elements through time).
 \end{quote}
 
-And you can add a default legend by using \texttt{legend\ =\ TRUE} (if you don't want to add the default legend you can add it after your plot using \texttt{legend(...)}).
+You can add a default legend by using \texttt{legend\ =\ TRUE} (if you don't want to add the default legend you can add it after your plot using \texttt{legend(...)}).
 
@@
 
-\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-92-1.pdf}
-
-\hypertarget{d-version}{%
-\section{3D version!}\label{d-version}}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-107-1.pdf}
+
+\hypertarget{adding-dimensions-not-necessary-but-fun}{%
+\subsection{Adding dimensions! Not necessary, but fun!}\label{adding-dimensions-not-necessary-but-fun}}
 
@@
 \NormalTok{my\_3D\_trait \textless{}{-}}\StringTok{ }\KeywordTok{make.traits}\NormalTok{(}\DataTypeTok{n =} \DecValTok{3}\NormalTok{, )}
-\CommentTok{\#\# Simulating a birth death tree with that 3D trait}
+\CommentTok{\#\# Simulating a birth{-}death tree with that 3D trait}
 \NormalTok{my\_data \textless{}{-}}\StringTok{ }\KeywordTok{treats}\NormalTok{(}\DataTypeTok{bd.params =} \KeywordTok{list}\NormalTok{(}\DataTypeTok{extinction =} \FloatTok{0.2}\NormalTok{),}
@@
 
-You can toggle which trait to plot using the \texttt{trait} option.
-Either by providing a single value to plot that specific trait against time or by providing two traits.
+You can toggle which trait to plot using the \texttt{trait} option, either by providing a single value to plot that specific trait against time or by providing two traits.
 
@@
 
-\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-94-1.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-109-1.pdf}
 
@@
 
-\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-95-1.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-110-1.pdf}
 
@@
 
-\includegraphics{../../../../../../tmp/RtmpmEQua5/file9cb853e0a21f.png}
+\includegraphics[width=6.5in]{../../../../../../tmp/RtmpwcuuU8/file55f335659f38}
 
@@
 
-\includegraphics{../../../../../../tmp/RtmpmEQua5/file9cb812c6ed9e.png}
+\includegraphics[width=6.5in]{../../../../../../tmp/RtmpwcuuU8/file55f34d5ddce7}
 
@@
 \begin{minipage}[t]{0.55\columnwidth}\raggedright
-\texttt{update.bd.params}\strut
+\texttt{bd.params.update}\strut
 \end{minipage} & \begin{minipage}[t]{0.39\columnwidth}\raggedright
@@
 \begin{minipage}[t]{0.55\columnwidth}\raggedright
-\texttt{update.traits}\strut
+\texttt{traits.update}\strut
 \end{minipage} & \begin{minipage}[t]{0.39\columnwidth}\raggedright
@@
 \begin{minipage}[t]{0.55\columnwidth}\raggedright
-\texttt{update.modifiers}\strut
+\texttt{modifiers.update}\strut
 \end{minipage} & \begin{minipage}[t]{0.39\columnwidth}\raggedright
@@
 
-For this scenario, we want to generate a pure birth tree (no traits and no extinction) where 80\% of the species go extinct two thirds into the scenario.
+For this scenario, we want to generate a pure birth tree (no traits and no extinction) where 80\% of the species go extinct two thirds of the way into the scenario.
 
@@
 
-We will then create a event that triggers when reaching half of the simulation (using \texttt{time.condition(2.5)}).
+We will then create a event that triggers when reaching half of the simulation time (using \texttt{time.condition(2.5)}).
 This event will target \texttt{"taxa"}, i.e.~the number of species, and randomly remove 80\% of them (using \texttt{random.extinction(0.8)}):
@@
 
-\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-101-1.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-116-1.pdf}
 
 \hypertarget{EGneg_ext}{%
-\section{Species with negative trait values go extinct after some time}\label{EGneg_ext}}
-
-For this scenario, we want to generate a pure birth tree with a one dimensional Brownian Motion trait for 5 time unit.
-We then want the species with a negative trait value go extinct after 4 time unit.
+\section{Species with negative trait values go extinct after a certain time}\label{EGneg_ext}}
+
+For this scenario, we want to generate a pure birth tree with a one dimensional Brownian Motion trait for 5 time units.
+We then want species with a negative trait value go extinct after 4 time units.
 
@@
 
-And our extinction event which triggers after reaching time 4 (\texttt{time.condition(4)}), targets the \texttt{"taxa"} and modifies the extinction for species with traits lower than 0
+And our extinction event which triggers after reaching time 4 (\texttt{time.condition(4)}), targets the \texttt{"taxa"} and modifies the extinction for species with traits lower than 0.
 
@@
 
-\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-105-1.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-120-1.pdf}
 
@@
 
-And our change in extinction rate event which triggers after reaching 30 taxa (\texttt{taxa.condition(30)}), targets the \texttt{"bd.params"} (birth-death parameters) sets the extinction rate to 0.5 (\texttt{update.bd.params(extinction\ =\ 0.5)}):
+And our change in extinction rate event which triggers after reaching 30 taxa (\texttt{taxa.condition(30)}), targets the \texttt{"bd.params"} (birth-death parameters) sets the extinction rate to 0.5 (\texttt{bd.params.update(extinction\ =\ 0.5)}):
 
@@
                       \DataTypeTok{target =} \StringTok{"bd.params"}\NormalTok{,}
-                      \DataTypeTok{modification =} \KeywordTok{update.bd.params}\NormalTok{(}\DataTypeTok{extinction =} \FloatTok{0.5}\NormalTok{))}
+                      \DataTypeTok{modification =} \KeywordTok{bd.params.update}\NormalTok{(}\DataTypeTok{extinction =} \FloatTok{0.5}\NormalTok{))}
 \end{Highlighting}
@@
 
-\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-108-1.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-123-1.pdf}
 
 \hypertarget{EGred_spec}{%
-\section{Reducing speciation rate after some time}\label{EGred_spec}}
-
-For this scenario, we want to generate a brith tree for 6 times units with random speciation rates (i.e.~drawn from a uniform (0.5;1) distribution) which reduces after time 4 through the simulations to a fixed value of 1/3.
+\section{Reducing speciation rate after a certain time}\label{EGred_spec}}
+
+For this scenario, we want to generate a birth tree for 6 times units with random speciation rates (i.e.~drawn from a uniform (0.5;1) distribution) which reduces after time 4 through the simulations to a fixed value of 1/3.
 
@@
 \NormalTok{stop\_time\_}\DecValTok{6}\NormalTok{ \textless{}{-}}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{max.time =} \DecValTok{6}\NormalTok{)}
-\NormalTok{random\_speciation \textless{}{-}}\StringTok{ }\KeywordTok{make.bd.params}\NormalTok{(}\DataTypeTok{speciation =}\NormalTok{ runif, }\DataTypeTok{speciation.args =} \KeywordTok{list}\NormalTok{(}\DataTypeTok{min =} \FloatTok{0.5}\NormalTok{, }\DataTypeTok{max =} \DecValTok{1}\NormalTok{))}
+\NormalTok{random\_speciation \textless{}{-}}\StringTok{ }\KeywordTok{make.bd.params}\NormalTok{(}\DataTypeTok{speciation =}\NormalTok{ runif, }
+                                    \DataTypeTok{speciation.args =} \KeywordTok{list}\NormalTok{(}\DataTypeTok{min =} \FloatTok{0.5}\NormalTok{, }\DataTypeTok{max =} \DecValTok{1}\NormalTok{))}
 \end{Highlighting}
@@
                       \DataTypeTok{target =} \StringTok{"bd.params"}\NormalTok{,}
-                      \DataTypeTok{modification =} \KeywordTok{update.bd.params}\NormalTok{(}\DataTypeTok{speciation =} \DecValTok{1}\OperatorTok{/}\DecValTok{3}\NormalTok{))}
+                      \DataTypeTok{modification =} \KeywordTok{bd.params.update}\NormalTok{(}\DataTypeTok{speciation =} \DecValTok{1}\OperatorTok{/}\DecValTok{3}\NormalTok{))}
 \end{Highlighting}
@@
 \KeywordTok{axisPhylo}\NormalTok{()}
-\KeywordTok{plot}\NormalTok{(reduced\_speciation\_event, }\DataTypeTok{main =} \StringTok{"Reduced speciation after time 5"}\NormalTok{, }\DataTypeTok{show.tip.label =} \OtherTok{FALSE}\NormalTok{)}
+\KeywordTok{plot}\NormalTok{(reduced\_speciation\_event, }
+     \DataTypeTok{main =} \StringTok{"Reduced speciation after time 5"}\NormalTok{, }
+     \DataTypeTok{show.tip.label =} \OtherTok{FALSE}\NormalTok{)}
 \KeywordTok{axisPhylo}\NormalTok{()}
@@
 
-\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-111-1.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-126-1.pdf}
 
@@
                   \DataTypeTok{target       =} \StringTok{"traits"}\NormalTok{,}
-                  \DataTypeTok{modification =} \KeywordTok{update.traits}\NormalTok{(}\DataTypeTok{process =}\NormalTok{ OU.process))}
+                  \DataTypeTok{modification =} \KeywordTok{traits.update}\NormalTok{(}\DataTypeTok{process =}\NormalTok{ OU.process))}
 \end{Highlighting}
@@
 
-\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-115-1.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-130-1.pdf}
 
@@
 \CommentTok{\#\# A 2D variance covariance matrix}
-\NormalTok{vcv\_matrix \textless{}{-}}\StringTok{ }\KeywordTok{matrix}\NormalTok{(}\DecValTok{1}\NormalTok{, }\DecValTok{2}\NormalTok{, }\DecValTok{2}\NormalTok{)}
+\NormalTok{cor\_matrix \textless{}{-}}\StringTok{ }\KeywordTok{matrix}\NormalTok{(}\DecValTok{1}\NormalTok{, }\DecValTok{2}\NormalTok{, }\DecValTok{2}\NormalTok{)}
 
@@
 \NormalTok{correlated\_2D\_BM \textless{}{-}}\StringTok{ }\KeywordTok{make.traits}\NormalTok{(}\DataTypeTok{n =} \DecValTok{2}\NormalTok{, }\DataTypeTok{process =}\NormalTok{ BM.process,}
-                      \DataTypeTok{process.args =} \KeywordTok{list}\NormalTok{(}\DataTypeTok{Sigma =}\NormalTok{ vcv\_matrix))}
+                      \DataTypeTok{process.args =} \KeywordTok{list}\NormalTok{(}\DataTypeTok{Sigma =}\NormalTok{ cor\_matrix))}
 \end{Highlighting}
@@
 \CommentTok{\#\# New correlation}
-\NormalTok{new\_vcv \textless{}{-}}\StringTok{ }\KeywordTok{matrix}\NormalTok{(}\KeywordTok{c}\NormalTok{(}\DecValTok{10}\NormalTok{,}\DecValTok{3}\NormalTok{,}\DecValTok{3}\NormalTok{,}\DecValTok{2}\NormalTok{),}\DecValTok{2}\NormalTok{,}\DecValTok{2}\NormalTok{)}
+\NormalTok{new\_cor \textless{}{-}}\StringTok{ }\KeywordTok{matrix}\NormalTok{(}\KeywordTok{c}\NormalTok{(}\DecValTok{10}\NormalTok{,}\DecValTok{3}\NormalTok{,}\DecValTok{3}\NormalTok{,}\DecValTok{2}\NormalTok{),}\DecValTok{2}\NormalTok{,}\DecValTok{2}\NormalTok{)}
 
@@
     \DataTypeTok{target       =} \StringTok{"traits"}\NormalTok{,}
-    \DataTypeTok{modification =} \KeywordTok{update.traits}\NormalTok{(}\DataTypeTok{process.args =} \KeywordTok{list}\NormalTok{(}\DataTypeTok{Sigma =}\NormalTok{ new\_vcv)))}
+    \DataTypeTok{modification =} \KeywordTok{traits.update}\NormalTok{(}\DataTypeTok{process.args =} \KeywordTok{list}\NormalTok{(}\DataTypeTok{Sigma =}\NormalTok{ new\_cor)))}
 \end{Highlighting}
@@
 
-\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-119-1.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-134-1.pdf}
 
@@
 
-\includegraphics{../../../../../../tmp/RtmpmEQua5/file9cb864dc16ed.png}
+\includegraphics[width=6.5in]{../../../../../../tmp/RtmpwcuuU8/file55f3254da0ed}
 
@@
 \begin{Highlighting}[]
-\CommentTok{\#\# Birth death rules (default)}
+\CommentTok{\#\# birth{-}death rules (default)}
 \NormalTok{default\_modifiers \textless{}{-}}\StringTok{ }\KeywordTok{make.modifiers}\NormalTok{()}
@@
     \DataTypeTok{target       =} \StringTok{"modifiers"}\NormalTok{,}
-    \DataTypeTok{modification =} \KeywordTok{update.modifiers}\NormalTok{(}\DataTypeTok{speciation =}\NormalTok{ speciation,}
+    \DataTypeTok{modification =} \KeywordTok{modifiers.update}\NormalTok{(}\DataTypeTok{speciation =}\NormalTok{ speciation,}
                                     \DataTypeTok{condition  =}\NormalTok{ negative.trait.condition,}
@@
 
-\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-125-1.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-140-1.pdf}
 
@@
                   \DataTypeTok{target =} \StringTok{"modifiers"}\NormalTok{,}
-                  \DataTypeTok{modification =} \KeywordTok{update.modifiers}\NormalTok{(}
+                  \DataTypeTok{modification =} \KeywordTok{modifiers.update}\NormalTok{(}
                                     \DataTypeTok{branch.length =}\NormalTok{ branch.length,}
@@
 
-\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-129-1.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-144-1.pdf}
 
 \hypertarget{EG_founding_purebirth}{%
-\section{Founding event: a generating a subtree with no fossils}\label{EG_founding_purebirth}}
+\section{Founding event: generating a subtree with no fossils}\label{EG_founding_purebirth}}
 
@@
 
-\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-132-1.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-147-1.pdf}
 
 \hypertarget{EG_founding_traits}{%
-\section{Founding event: a generating a subtree a different process}\label{EG_founding_traits}}
-
-For this scenario, we want to generate a birth-death tree with a one dimensional Brownian Motion trait for 6 time unit.
-After a taxa reaches the value 3 or higher, it gives birth to a sub-tree that generates an OU trait with an optimum at the value 3.
+\section{Founding event: generating a subtree a different process}\label{EG_founding_traits}}
+
+For this scenario, we want to generate a birth-death tree with a one dimensional Brownian Motion trait for 6 time units.
+After a taxon reaches the value 3 or higher, it gives birth to a sub-tree that generates an OU trait with a long-term mean at the value 3.
 
@@
 
-When a taxa reaches the value 3 \texttt{trait.condition}, it generates a pure birth tree (\texttt{speciation\ =\ 2}) with an OU trait with the optimum value 3.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\#\# The OU trait with an optimum value of 3}
+When a taxa reaches the value 3 \texttt{trait.condition}, it generates a pure birth tree (\texttt{speciation\ =\ 2}) with an OU trait with the long-term mean value 3.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# The OU trait with a long{-}term mean value of 3}
 \NormalTok{OU\_}\DecValTok{3}\NormalTok{ \textless{}{-}}\StringTok{ }\KeywordTok{make.traits}\NormalTok{(}\DataTypeTok{process =}\NormalTok{ OU.process,}
@@
 
-\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-136-1.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-151-1.pdf}
 

### 2023-05-25 inst/gitbook/07_specific_examples.Rmd
manual header change

@@
 ---
-title: "Specific examples"
+title: "treats manual"
 author: "Thomas Guillerme (guillert@tcd.ie)"

### 2023-05-25 inst/gitbook/_book/treats_manual.tex
manual header change

@@
 \hypersetup{
-  pdftitle={Specific examples},
+  pdftitle={treats manual},
   pdfauthor={Thomas Guillerme (guillert@tcd.ie)},
@@
 
-\title{Specific examples}
+\title{treats manual}
 \author{Thomas Guillerme (\href{mailto:guillert@tcd.ie}{\nolinkurl{guillert@tcd.ie}})}
@@
 ## }
-## <bytecode: 0x55b094e15a20>
+## <bytecode: 0x55b3b2541d88>
 ## <environment: namespace:treats>
@@
 ## }
-## <bytecode: 0x55b093879f20>
+## <bytecode: 0x55b3b0f93c78>
 ## <environment: namespace:treats>
@@
 ## }
-## <bytecode: 0x55b093879270>
+## <bytecode: 0x55b3b0f9bdc8>
 ## <environment: namespace:treats>
@@
 ## }
-## <bytecode: 0x55b09387f640>
+## <bytecode: 0x55b3b0f9b498>
 ## <environment: namespace:treats>
@@
 
-\includegraphics[width=6.5in]{../../../../../../tmp/RtmpwcuuU8/file55f335659f38}
+\includegraphics[width=6.5in]{../../../../../../tmp/RtmpXnwk90/fileb7e05f6d37cf}
 
@@
 
-\includegraphics[width=6.5in]{../../../../../../tmp/RtmpwcuuU8/file55f34d5ddce7}
+\includegraphics[width=6.5in]{../../../../../../tmp/RtmpXnwk90/fileb7e060df376f}
 
@@
 
-\includegraphics[width=6.5in]{../../../../../../tmp/RtmpwcuuU8/file55f3254da0ed}
+\includegraphics[width=6.5in]{../../../../../../tmp/RtmpXnwk90/fileb7e0d32ccc1}
 

### 2023-05-25 paper/treats_paper.tex
Update treats_paper.tex

@@
 These comparisons can be done between observed patterns under different conditions or against null, neutral or baseline models (see \citealt{bausman2018neutral} for distinctions or common misconceptions) suggesting different processes or mechanisms.
+% These types phylogenetic comparative methods analysis of traits through time are routinely done in palaeontology (disparity; \citealt{guillerme2020disparities}) or in ecology (functional diversity; \citealt{mammola2021concepts}).
 For example different traits distribution for species with different diets \citep{deepak2023diet},
@@
 % What already exists
-In practice, these evolutionary simulations can be done relatively easily multiple times on computers using a birth-death process \cite{feller1939birthdeath,stadler2010birthdeath,diversitree}
-A birth-death process is @yadiyada and has been routinely implemented in \texttt{R} (\citealt{R}; e.g. \citealt{ape}, \citealt{diversitree}).
-Traits can be simulating using the following processes @OU, @BM, @yadidada. % TODO
-The diversity of these traits through time can be called disparity in palaeontology \cite{guillerme2020disparities} or functional diversity in ecology (for functional traits; \citealt{mammola2021concepts}).
+In practice, these evolutionary simulations can be done relatively easily on computers using a birth-death process \cite{feller1939birthdeath,stadler2010birthdeath,diversitree}
+A birth-death process is a continuous time Markov process that had been routinely implemented in \texttt{R} \citep{R} to simulate biological phylogenies (e.g. \citealt{ape}, \citealt{diversitree}).
+These algorithm to generate phylogenetic trees can be coupled with other Markov processes to also generate traits, for example using a Brownian Motion process (BM; \citealt{cavalli1967BM}) or an Ornstein Uhlenbeck  (OU; \citealt{lande1976OU}; see \citealt{cooper2016cautionary} for a distinction between both).
 In \texttt{R}, this can be done with several already well used and well documented packages.
@@
 
-% STOPPED HERE
-
 % The problems with the absence of modularity
-Although the packages mentioned above (and more that we forgot to mention/don't know off) and are excellent (i.e.
-fast, reliable and documented) at the specific tasks they are designed for, they unfortunately don't allow much modifications beyond the parameters implemented for each specific function.
-This can be problematic for generating specific simulations to be used as a specific null model for some specific questions.
+Although the packages mentioned above are excellent and routinely used with fast and reliable algorithms and associated documentation, they are all designed for specific tasks and don't allow much modification beyond the input parameters designed by the authors.
 For example, \texttt{TreeSim} can simulate a birth-death tree with some extinction event but is not designed to simulated one with an extinction event that leads to the birth-death process to be not diversity dependent anymore (e.g.
@@
 Or \texttt{PETER} is not designed to simulate a complex set of traits (say three correlated BM traits and two independent OU ones).
-Although we don't think this was the aim of these specific packages authors (not to create a strawman argument) we think this hampers the use of specific null model (or baseline model) to test specific dynamic mechanisms or to compare to specific datasets.
-
-% The solution with treats
+This absence of modularity has hampered the use of complex and question tailored simulations evolutionary biology, although I acknowledge this was not the primary aim of the authors of the packages mentioned above.
+This has led workers to often develop their own tools to answer specific questions (e.g. \citealt{puttick2020complex}).
 Therefore, we propose \treats a modular \texttt{R} package to simulate both diversity and disparity through time.
@@
 % Very briefly how does it work
-\treats is based on a core birth-death process implemented in the function \texttt{treats} that simulates diversity through time using the following steps:
+\treats is based on the eponymous \treats function that allows to simulate a phylogeny and some trait(s) simultaneously.
+The base birth-death algorithm ``grows'' a phylogenetic tree and generates traits for each node and tips in the following manner:
 
 \begin{enumerate}
-    \item Waiting: Generating some waiting time corresponding to the growth in length of the tree (e.g. the duration of a lineage through time);
-    \item Selection: Choosing a lineage \textit{alive} at the time of the simulation;
-    \item Speciation: Choosing whether this species will go extinct or speciate leading to two new lineages.
+    \item Generating branch length;
+    \item Selecting a lineage among the currently living ones;
+    \item Choosing whether that lineage goes extinct (becomes a tip) or speciates (becomes a node).
 \end{enumerate}
 
-These three steps are repeated until the tree reaches the desired age or the desired number of species.
+These four steps are repeated until the tree reaches the desired age or number of species.
 If traits are simulated during the process, a fourth step is added:
 
-\indent 4. Trait: Generate some trait(s) value(s) based for the extinct species or the node leading to the two new lineages.
-This/these value(s) are generated taking into account the trait value of the extinct species/new node's ancestor and the time elapsed (generated in step 1).
+    4. Generating some trait(s) value(s) for the selected lineage (no either a tip or a node).
 

### 2023-05-26 paper/treats_paper.tex
Update treats_paper.tex

@@
 The simulation then outputs a tree (of class \texttt{``phylo''} and a associated table of traits - \texttt{``matrix''}) that can be visualised using the \texttt{plot.treats} function.
+A third class of object called \texttt{``events''} can be added to the simulations to modify it under certain conditions (e.g. simulating a mass extinction).
+Each element in the algorithm described above can be modified by the user using the implemented functions \texttt{make.bd.params} to set the birth-death parameters, \texttt{make.traits} to set the trait(s), \texttt{make.modifiers} to set the birth-death algorithm and \texttt{make.events} to eventually add one or more events.
 
@@
 \centering
-   \includegraphics[width=1\textwidth]{../inst/gitbook/treats_structure.pdf} 
-\caption{\treats package workflow: @@@.}
+   \includegraphics[width=0.6\textwidth]{../inst/gitbook/treats_structure.pdf} 
+\caption{\treats package workflow: the \treats algorithm generates a tree and traits using inbuilt \texttt{``traits''} objects that contain the instructions on how to generate the trait data (e.g.
+which process?
+how many dimensions?); \texttt{``modifiers''} objects that contains instructions on how to ``grow'' the tree (e.g.
+by linking speciation to trait values, or to the current number of species); and \texttt{``events''} objects that can modify the tree structure, \texttt{``modifiers''} or \texttt{``traits''} depending on specific conditions (e.g.
+80\% of species with positive trait values go extinct after reaching a specific time).}
 \label{Fig:workflow}
 \end{figure}
-% \texttt{"events"} are an additional category of modular objects discussed later.
+
+\section{Brief applied example}
+
+The modularity structure of the \treats packages allows users to design complex simulation scenarios to fit their specific question and data.
+To illustrate this we will look whether it is possible to detect changes in disparity (i.e.
+diversity of traits) in a subset of the data published from \cite{beck2014ancient} implemented in \cite{dispRity}.
+This dataset contains the ordinated traits for 50 mammalian species across the cretaceous-palaeogene extinction event (K-Pg, 66 Mya).
+
+\subsection{Simulating trees}
+
+\subsection{Simulating traits}
+
+\subsection{Adding events}
 
@@
 \section{Package location}
-The \treats package is available on the CRAN at \url{https://cran.r-project.org/web/packages/treats/index.html} or on GitHub at \url{https://github.com/TGuillerme/treats} with more associated information.
-All the versions of the package are archived on ZENODO with associated DOI \url{https://zenodo.org/@@@}.
+The \treats package is available 
+% on the CRAN at \url{https://cran.r-project.org/web/packages/treats/index.html} or 
+on GitHub at \url{https://github.com/TGuillerme/treats} with more associated information.
+All the versions of the package are archived on ZENODO with associated DOI \url{https://doi.org/10.5281/zenodo.7970384}.
 
 \section{Acknowledgments}
-Thanks to Mark Puttick and Alex Slavenko for comments on the early stage of the development of this package.
-Thanks to @@@.
-This work was funded by UKRI-NERC Grant NE/T000139/1 and a Royal Society University Research Fellowship (URF R 180006 to GHT).
+Thanks to Mark Puttick for inspiring to develop the package through previous collaborations.
+Thanks to Ian Brennan, Gustavo Burin, Christopher Cooney, Natalie Cooper, Alex Cranston, Jasmine Hardie, Tom Lansley, Clement Prieul, Joe Rees, James Rule, Sophie Ryan and Gavin Thomas, for useful comments on late stages of the development of this package.
+This work was funded by UKRI-NERC Grant NE/T000139/1 and a Royal Society University Research Fellowship (URF R 180006 to Gavin Thomas).
 

### 2023-05-26 paper/treats_paper_plots.Rmd
Create treats_paper_plots.Rmd

---
title: "treats paper plots"
author: "Thomas Guillerme"
date: "`r Sys.Date()`"
bibliography: references.bib
output:
  html_document:
    fig_width: 8
    fig_height: 8
---

This script reproduces the plots and analyses used in the `treats` paper.

## Beck Lee data (figure 1)

```{r}
## Loading the data and packages
library(treats)
library(dispRity)
data(BeckLee_tree)
data(BeckLee_mat99)

## Creating the time slices
time_slices <- chrono.subsets(BeckLee_mat99, BeckLee_tree,
                              method = "continuous",
                              model  = "acctran",
                              time   = seq(from = 96, to = 36, by = -10))

## Calculating disparity on the two first dimensions only
observed_disparity <- dispRity(boot.matrix(time_slices), metric = c(sum, variances), dimensions = c(1,2))

## Plotting the tree and the disparity through time
pdf(file = "treats_example_figure1.pdf", width = 14, height = 7)
par(mfrow = c(1,2))
plot(BeckLee_tree, show.tip.label = FALSE)
axisPhylo()
abline(v = BeckLee_tree$root.time - 66, col = "red")
legend("topleft", pch = NULL, legend = "A", bty = "n", cex = 2)
par(bty = "n")
plot(observed_disparity, ylab = "Sum of variances")
abline(v = 4, col = "red")
legend("topleft", pch = NULL, legend = "B", bty = "n", cex = 2)
dev.off()
```

### 2023-05-26 inst/gitbook/01_getting_started.Rmd
Update 01_getting_started.Rmd

@@
                                  modification = random.extinction(0.75))
-## Creating an extrinction that removes species with negative trait values
+## Creating an extinction that removes species with negative trait values
 negative_extinction <- make.events(target = "taxa",

### 2023-05-26 paper/treats_paper.tex
Update treats_paper.tex

@@
 \usepackage[osf]{mathpazo}
+\usepackage{listings}
+\usepackage{color}
+
+\definecolor{dkgreen}{rgb}{0,0.6,0}
+\definecolor{gray}{rgb}{0.5,0.5,0.5}
+\definecolor{mauve}{rgb}{0.58,0,0.82}
+
+\lstset{frame=tb,
+  language=R,
+  showstringspaces=false,
+  columns=flexible,
+  basicstyle={\small\ttfamily},
+  numbers=none,
+  numberstyle=\tiny\color{gray},
+  keywordstyle=\color{blue},
+  commentstyle=\color{dkgreen},
+  stringstyle=\color{mauve},
+  breaklines=true,
+  breakatwhitespace=false,
+  tabsize=4
+}
+
 \restylefloat{table}
@@
 diversity of traits) in a subset of the data published from \cite{beck2014ancient} implemented in \cite{dispRity}.
-This dataset contains the ordinated traits for 50 mammalian species across the cretaceous-palaeogene extinction event (K-Pg, 66 Mya).
+This dataset contains the ordinated traits for 50 mammalian species across the cretaceous-palaeogene extinction event (K-Pg, 66 Mya; \ref{Fig:example1}).
+
+\begin{lstlisting}
+## Loading the packages and the data
+library(dispRity)
+library(treats)
+## The observed tree
+data(BeckLee_tree)
+## The observed data
+data(BeckLee_mat99)
+\end{lstlisting}
+
+\begin{figure}[!htbp]
+\centering
+   \includegraphics[width=0.9\textwidth]{treats_example_figure1.pdf} 
+\caption{Observed phylogeny (A) and disparity through time (B) in the subset of \cite{beck2014ancient}'s dataset.
+For this specific data, is it possible to detect potential changes of disparity due to a mass extinction event?}
+\label{Fig:example1}
+\end{figure}
+
+Using this example dataset, one might be interested in testing whether the K-Pg extinction had an effect on disparity through time.
+However, one legitimate question to ask would be whether such effect could be detectable in the first place in this dataset.
+Do test this we can simulate some datasets with similar properties as the observed data and measure changes in disparity in these simulated datasets.
+If the simulations contain some mass extinction events we will then be able to see whether they cause a detectable change in disparity or not.
 
@@
 
-\subsection{Simulating traits}
+The first and simplest way is to simulate tree topologies that have similar properties than the observed ones.
+To do so, we need to use some \textbf{speciation} parameter indicating the rate at which lineages speciate (\textit{aka} ``birth'' or ``$\lambda$'') and an \textbf{extinction} parameter indicating the rate at which they go extinct (\textit{aka} ``death'' or ``$\mu$'').
+These two parameters can be crudely calculated from a give tree by counting the number of speciations and extinction events through time (note that better analytical methods exist, e.g.
+\texttt{ape::birthdeath}; \citealt{lawton1995extinction,ape}).
+We also need a stopping rule for when to stop the simulations (in our case when reaching 50 taxa, extinct or living).
+This will produce a single random tree using the input parameters (Fig.\ref{Fig:example2}).
+Note that I will not discuss the options here in great details.
+Much more information can be found in the \href{http://tguillerme.github.io/treats.htm}{\treats manual}.
+% Note that the \treats function can automatically restart if the simulations did not produce the required tree (i.e. if all lineages go extinct before reaching the stopping rule).
+% This is done through the \texttt{null.error} option.
 
-\subsection{Adding events}
+\begin{lstlisting}
+## Using the birth-death parameters from the observed tree
+my_bd_params <- crude.bd.est(BeckLee_tree)
+## Setting the stopping rule (stop after reaching 30 time units)
+stop_rule <- list(max.time = 30)
+## Generate the tree
+sim_tree <- treats(bd.params  = my_bd_params,
+                   stop.rule  = stop_rule,
+                   null.error = 100)
+## Visualising the results
+plot(tree_tree, show.tip.label = FALSE) ; axisPhylo()
+\end{lstlisting}
+
+\begin{figure}[!htbp]
+\centering
+   \includegraphics[width=0.9\textwidth]{treats_example_figure2.pdf} 
+\caption{A randomly simulated tree with similar properties as the observed one. Note that the time here is expressed in arbitrary units.}
+\label{Fig:example2}
+\end{figure}
+
+\subsection{Simulating trees and traits}
+
+For our specific question, we will also need to simulate some traits associated with each node and tip.
+For simplicity we will simulate a two dimensional Brownian Motion trait.
+To do so, we can create a \texttt{``traits''} object with the function \texttt{make.traits}.
+This results in a 2 dimensional trait space for all the simulated species and their nodes (Fig.\ref{Fig:example3}).
+
+\begin{lstlisting}
+## Creating a trait in 97 dimensions.
+my_traits <- make.traits(process = BM.process, n = 2)
+## Simulate the tree and traits
+sim_data <- treats(traits     = my_traits,
+                   bd.params  = my_bd_params,
+                   stop.rule  = stop_rule,
+                   null.error = 100)
+## Plotting the first trait through time
+plot(sim_data)
+## Plotting both traits
+plot(sim_data, trait = c(1,2))
+\end{lstlisting}
+
+\begin{figure}[!htbp]
+\centering
+   \includegraphics[width=0.9\textwidth]{treats_example_figure3.pdf} 
+\caption{A randomly simulated tree with a 2 dimensional random Brownian Motion trait.}
+\label{Fig:example3}
+\end{figure}
+
+\subsection{Simulating a trees and traits with events}
+
+However, for our specific question we want to simulate both a tree and some traits but with an extinction event in order to get some simulated data to answer our question.
+To do so, we can create two different \texttt{``events''} object with the function \texttt{make.events}.
+The first one will simulate a random extinction event after reaching half of the simulations (15 time units) and then making three quarter (0.75) of the taxa go extinct.
+The second one will simulate a random extinction but based on trait values: after reaching time 20, all the species with positive trait values will go extinct.
+Both scenarios thus illustrate two different types of mass extinction (selective and random \citealt{puttick2020complex}).
+
+\begin{lstlisting}
+## Creating a random mass extinction
+random_extinction <- make.events(
+    target       = "taxa",
+    condition    = time.condition(15),
+    modification = random.extinction(0.75))
+
+## Creating an extinction that removes species with positive trait values
+negative_extinction <- make.events(
+    target = "taxa",
+    condition = time.condition(15),
+    modification = trait.extinction(x = 0, condition = `>=`))
+
+## Simulating a random extinction
+sim_rand_extinction <- treats(
+                   traits     = my_traits,
+                   bd.params  = my_bd_params,
+                   stop.rule  = stop_rule,
+                   events     = random_extinction,
+                   null.error = 100)
+
+## Simulate a selective extinction
+sim_trait_extinction <- treats(
+                   traits     = my_traits,
+                   bd.params  = my_bd_params,
+                   stop.rule  = stop_rule,
+                   events     = negative_extinction,
+                   null.error = 100)
+
+## Visualising the difference between both scenarios
+plot(sim_rand_extinction, main = "Random extinction")
+abline(v = 10, col = "red")
+plot(sim_trait_extinction, main = "Negative extinction")
+abline(v = 10, col = "red")
+\end{lstlisting}
+
+\begin{figure}[!htbp]
+\centering
+   \includegraphics[width=0.9\textwidth]{treats_example_figure4.pdf} 
+\caption{Two different mass extinction events: (A) 75\% of species go extinct; (B) all species with positive trait values go extinct.}
+\label{Fig:example4}
+\end{figure}
+
+Again, the syntax is not developed in this example.
+For more information, you can refer to the \href{http://tguillerme.github.io/treats.htm}{\treats manual}.
+
+Once we have simualted a distribution of trees and traits with the two extinction scenarios, we can measure disparity as in Fig.
+\ref{Fig:example1} for all the simulated data and compare it to the observed disparity.
 
@@
 Thanks to Mark Puttick for inspiring to develop the package through previous collaborations.
-Thanks to Ian Brennan, Gustavo Burin, Christopher Cooney, Natalie Cooper, Alex Cranston, Jasmine Hardie, Tom Lansley, Clement Prieul, Joe Rees, James Rule, Sophie Ryan and Gavin Thomas, for useful comments on late stages of the development of this package.
+Thanks to Andrew Beckerman, Ian Brennan, Gustavo Burin, Christopher Cooney, Natalie Cooper, Alex Cranston, Jasmine Hardie, Tom Lansley, Clement Prieul, Joe Rees, James Rule, Sophie Ryan and Gavin Thomas, for support and useful comments on late stages of the development of this package.
 This work was funded by UKRI-NERC Grant NE/T000139/1 and a Royal Society University Research Fellowship (URF R 180006 to Gavin Thomas).

### 2023-05-26 paper/treats_paper_plots.Rmd
Update treats_paper_plots.Rmd

@@
 ```
+
+## Simulated tree
+
+```{r}
+## Using the birth-death parameters from the observed tree
+my_bd_params <- crude.bd.est(BeckLee_tree)
+## Setting the stopping rule (stop after 30 time units)
+stop_rule <- list(max.time = 30)
+
+## Generate the tree
+set.seed(1)
+sim_tree <- treats(bd.params  = my_bd_params,
+                   stop.rule  = stop_rule,
+                   null.error = 100)
+pdf(file = "treats_example_figure2.pdf", width = 7, height = 7)
+plot(sim_tree, show.tip.label = FALSE)
+axisPhylo()
+dev.off()
+```
+
+## Simulate tree and traits
+
+```{r}
+## Creating a trait in 97 dimensions.
+my_traits <- make.traits(process = BM.process, n = 2)
+
+## Simulate the tree and traits
+set.seed(1)
+sim_data <- treats(traits     = my_traits,
+                   bd.params  = my_bd_params,
+                   stop.rule  = stop_rule,
+                   null.error = 100)
+pdf(file = "treats_example_figure3.pdf", width = 14, height = 7)
+par(mfrow = c(1,2))
+plot(sim_data, ylab = "Trait 1", las = 1, main = "Trait 1 through time")
+plot(sim_data, trait = c(2,1), ylab = "Trait 1", xlab = "Trait 2", las = 1, main = "Traits 1 and 2")
+dev.off()
+```
+
+## Simulating a tree and traits with an extinction event
+
+```{r}
+## Creating a random mass extinction
+random_extinction <- make.events(
+    target       = "taxa",
+    condition    = time.condition(15),
+    modification = random.extinction(0.75))
+## Creating an extinction that removes species with positive trait values
+negative_extinction <- make.events(
+    target = "taxa",
+    condition = time.condition(15),
+    modification = trait.extinction(x = 0, condition = `>=`))
+
+set.seed(2)
+sim_rand_extinction <- treats(
+                   traits     = my_traits,
+                   bd.params  = my_bd_params,
+                   stop.rule  = stop_rule,
+                   events     = random_extinction,
+                   null.error = 100,
+                   replicates = 50)
+
+## Simulate the tree and traits with a random extinction event
+sim_trait_extinction <- treats(
+                   traits     = my_traits,
+                   bd.params  = my_bd_params,
+                   stop.rule  = stop_rule,
+                   events     = negative_extinction,
+                   null.error = 100,
+                   replicates = 50)
+
+pdf(file = "treats_example_figure4.pdf", width = 14, height = 7)
+par(mfrow = c(1,2))
+## Visualising the difference between both scenarios
+plot(sim_rand_extinction[[11]], main = "Random extinction")
+abline(v = 15, col = "red")
+legend("topleft", pch = NULL, legend = "A", bty = "n", cex = 2)
+plot(sim_trait_extinction[[10]], main = "Negative extinction")
+legend("topleft", pch = NULL, legend = "B", bty = "n", cex = 2)
+abline(v = 15, col = "red")
+dev.off()
+```
+
+## Comparing disparities
+
+```{r}
+## Wrapper function for measuring disparity across the datasets
+measure.disparity <- function(one_simulation) {
+    ## transform the data
+    sim <- dispRitreats(one_simulation, scale.tree = FALSE)
+    ## Calculate disparity
+    disparity <- dispRity(
+             chrono.subsets(data = sim$data, tree = sim$tree,
+                            method = "continuous", model = "acctran",
+                            time = seq(from = 30, to = 0, by = -5)),
+                            metric = c(sum, variances))
+    ## Scale the disparity
+    disparity <- unlist(get.disparity(disparity))
+    return(disparity/max(disparity))
+}
+
+## Measuring disparities
+random_extinction_disparity <- do.call(rbind, lapply(sim_rand_extinction, measure.disparity))
+selective_extinction_disparity <- do.call(rbind, lapply(sim_trait_extinction, measure.disparity))
+
+## Get the observed disparity
+observed_disparity <- unlist(get.disparity(observed_disparity))
+observed_disparity/max(observed_disparity)
+
+## Plotting the results
+polygon.plot <- function(disparity_results, main = "") {
+    ## Get the quantiles
+    quantiles <- apply(disparity_results, 2, quantile, probs = c(0.025, 0.25, 0.5, 0.75, 0.975), na.rm = TRUE)
+
+    plot(NULL, xlim = c(1, ncol(quantiles)), ylim = range(quantiles), ylab = "scaled sum of ranges", xlab = "Relative time", main = main )
+    polygon(x = c(1:ncol(quantiles), ncol(quantiles):1),
+            y = c(quantiles[1, ], rev(quantiles[5, ])),
+            col = "lightgrey", border = FALSE)
+    polygon(x = c(1:ncol(quantiles), ncol(quantiles):1),
+            y = c(quantiles[2, ], rev(quantiles[4, ])),
+            col = "grey", border = FALSE)
+    lines(x = c(1:ncol(quantiles)), y = quantiles[3, ])
+}
+
+pdf(file = "treats_example_figure5.pdf", width = 14, height = 7)
+par(mfrow = c(1,2))
+polygon.plot(random_extinction_disparity, main = "Random extinction")
+abline(v = 4, col = "red")
+lines(x = 1:7, y = observed_disparity, lty = 2, lwd = 2)
+polygon.plot(selective_extinction_disparity, main = "Selective extinction")
+abline(v = 4, col = "red")
+lines(x = 1:7, y = observed_disparity, lty = 2, lwd = 2)
+dev.off()
+```

### 2023-05-30 paper/treats_paper.Rmd
moved paper to fully md

---
title: "`treats`: a modular `R` package for simulating trees and traits."
author: "Thomas Guillerme$^1$"
bibliography: References.bib
# csl: mee.csl
date: "`r Sys.Date()`"
output:
  pdf_document:
    fig_width: 8
    fig_height: 8
  bookdown::html_document2: default
  bookdown::pdf_document2: default
---

$^1$ School of Biosciences, University of Sheffield, Sheffield, S10 2TN, United Kingdom. guillert@tcd.ie

# Abstract

1. Simulating biological realistic data can be an important step to understand and investigate biological mechanisms.
Simulated data can be used to generate null, base line or neutral models.
These models can be used either in comparison to observed data to estimate the mechanisms that generated the data.
Or they can be used to explore, understand and develop theoretical advances by proposing toy models.

2. In evolutionary biology, such simulations often involve the need of an evolutionary process where descent with modification is at the core of how the simulated data is generated.
These evolutionary processes can then be nearly infinitely modified to include much more complex processes to affect the simulations such as traits co-evolution, competition mechanisms or mass extinction events.

3. Here I present the `treats` package, a modular `R` package for trees and traits simulations.
This package is based on a simple birth death algorithm from which all core steps can easily be modified by users (hence the modularity).
It also provides a tidy interface through the `treats` object, allowing users to easily run reproducible simulations.

4. Here I present the core algorithms and their integrated modularity and provide an example of how to use the `treats` package when studying disparity, the diversity of traits.
The `treats` package also comes with an extend manual regularly updated following users' questions or suggestions which comes as an extended supplementary material to this paper.

###### Keywords: trees, traits, simulations, birth-death, null-models, ecology, evolution, disparity

# Introduction

<!-- % Simulations in biology -->
Comparing biological patterns is one of the key ways to understand mechanisms in evolutionary biology.
This lead to the development of phylogenetic comparative methods as key methodologically driven topic in ecology, evolution and palaeontology [@felsensteinPCM; @pennell2013review].
As indicated in the name, phylogenetic comparative methods rely on comparing patterns in a phylogenetic context (i.e. descent with modification) to understand biological mechanisms or concepts [@harmon2019book].
These comparisons can be done between observed patterns under different conditions or against null, neutral or baseline models (see @bausman2018neutral for distinctions or common misconceptions) suggesting different processes or mechanisms.
<!-- % These types phylogenetic comparative methods analysis of traits through time are routinely done in palaeontology (disparity; @guillerme2020disparities) or in ecology (functional diversity; @mammola2021concepts). -->
For example different traits distribution for species with different diets [@deepak2023diet],
<!-- % different habitats https://onlinelibrary.wiley.com/doi/full/10.1111/jbi.12939 -->
or by comparing some observed pattern to one simulated under null or base conditions [@miller2022alternating].
In theory, workers can follow the research pipeline of thinking of a specific mechanism (e.g. mass extinction allowing the surviving species to acquire new morphologies), collect some data to test this mechanism (e.g. some traits of species across and extinction event) and then compare these patterns to one simulated under no specific conditions (e.g. a null model where the traits evolve randomly regardless of an extinction event @puttick2020complex).
For such an approach, we need statistical and softwares solutions to simulate trees and data to generate these specific null models.

<!-- % What already exists -->
In practice, these evolutionary simulations can be done relatively easily on computers using a birth-death process [@feller1939birthdeath; @stadler2010birthdeath; @diversitree].
A birth-death process is a continuous time Markov process that had been routinely implemented in `R` [@R] to simulate biological phylogenies (e.g. [@ape], [@diversitree]).
These algorithm to generate phylogenetic trees can be coupled with other Markov processes to also generate traits, for example using a Brownian Motion process (BM; @cavalli1967BM) or an Ornstein Uhlenbeck  (OU; @lande1976OU; see @cooper2016cautionary for a distinction between both).
In `R`, this can be done with several already well used and well documented packages.
For example if you want to simulate diversity through time, you can use `TreeSim` [@treesim] to simulate diversity under a set of specific parameters (e.g. speciation and extinction) with some events disrupting the simulations (e.g. mass extinctions).
One can even improve on generating these patterns using `FossilSim` [@fossilsim] to generate a pattern that will take into account fossilization processes.
Or use `paleobuddy` [@paleobuddy] or `paleotree` [@paleotree] to generate palaeontology specific data.
On the other hand, if you need to simulate both diversity and traits through time, this can be done with specific parameters in `RPANDA` [@rpanda] or in `PETER` [@puttick2020complex] where the traits are generated stochasticaly through time (given some process) during the birth-death process.

<!-- % The problems with the absence of modularity -->
Although the packages mentioned above are excellent and routinely used with fast and reliable algorithms and associated documentation, they are all designed for specific tasks and don't allow much modification beyond the input parameters designed by the authors.
For example, `TreeSim` can simulate a birth-death tree with some extinction event but is not designed to simulated one with an extinction event that leads to the birth-death process to be not diversity dependent anymore (e.g. simulating a release in selection pressure after the extinction process that leads to a different process dominating speciation).
Or `PETER` is not designed to simulate a complex set of traits (say three correlated BM traits and two independent OU ones).
This absence of modularity has hampered the use of complex and question tailored simulations evolutionary biology, although I acknowledge this was not the primary aim of the authors of the packages mentioned above.
This has led workers to often develop their own tools to answer specific questions (e.g. @puttick2020complex).
Therefore, we propose `treats` a modular `R` package to simulate both diversity and disparity through time.

# Description

<!-- % Very briefly how does it work -->
`treats` is based on the eponymous `treats` function that allows to simulate a phylogeny and some trait(s) simultaneously.
The base birth-death algorithm "grows" a phylogenetic tree and generates traits for each node and tips in the following manner:

1. Generating branch length;
2. Selecting a lineage among the currently living ones;
3. Choosing whether that lineage goes extinct (becomes a tip) or speciates (becomes a node).

These four steps are repeated until the tree reaches the desired age or number of species.
If traits are simulated during the process, a fourth step is added:

4. Generating some trait(s) value(s) for the selected lineage (no either a tip or a node).

In `treats`, these three or four steps are implemented as modular functions that the user can easily change using an internal class of objects called `"modifiers"` or `"traits"` (Fig. \ref{Fig:workflow}).
The simulation then outputs a tree (of class `"phylo"` and a associated table of traits - `"matrix"`) that can be visualised using the `plot.treats` function.
A third class of object called `"events"` can be added to the simulations to modify it under certain conditions (e.g. simulating a mass extinction).
Each element in the algorithm described above can be modified by the user using the implemented functions `make.bd.params` to set the birth-death parameters, `make.traits` to set the trait(s), `make.modifiers` to set the birth-death algorithm and `make.events` to eventually add one or more events.

![Figure 1: `treats` package workflow: the `treats` algorithm generates a tree and traits using inbuilt `"traits"` objects that contain the instructions on how to generate the trait data (e.g. which process? how many dimensions?); `"modifiers"` objects that contains instructions on how to "grow" the tree (e.g. by linking speciation to trait values, or to the current number of species); and `"events"` objects that can modify the tree structure, `"modifiers"` or `"traits"` depending on specific conditions (e.g. 80% of species with positive trait values go extinct after reaching a specific time).](../inst/gitbook/treats_structure.png){#figure1}

## Brief applied example

The modularity structure of the `treats` packages allows users to design complex simulation scenarios to fit their specific question and data.
To illustrate this we will look whether it is possible to detect changes in disparity (i.e. diversity of traits) in a subset of the data published from @beck2014ancient implemented in @dispRity.
This dataset contains the ordinated traits for 50 mammalian species across the cretaceous-palaeogene extinction event (K-Pg, 66 Mya; Fig. 2).

```{r, message = FALSE}
## Loading the data and packages
library(treats)
library(dispRity)
data(BeckLee_tree)
data(BeckLee_mat99)

## Creating the time slices
time_slices <- chrono.subsets(BeckLee_mat99, BeckLee_tree,
                              method = "continuous",
                              model  = "acctran",
                              time   = seq(from = 96, to = 36, by = -10))

## Calculating disparity on the two first dimensions only
observed_disparity <- dispRity(boot.matrix(time_slices), metric = c(sum, variances), dimensions = c(1,2))
```

```{r prepdata, echo = FALSE, fig.cap="Figure 2: Observed phylogeny (A) and disparity through time (B) in the subset of @beck2014ancient's dataset. For this specific data, is it possible to detect potential changes of disparity due to a mass extinction event?", fig.width = 18, fig.height = 8}
## Plotting the tree and the disparity through time
par(mfrow = c(1,2))
plot(BeckLee_tree, show.tip.label = FALSE)
axisPhylo()
abline(v = BeckLee_tree$root.time - 66, col = "red")
legend("topleft", pch = NULL, legend = "A", bty = "n", cex = 2)
par(bty = "n")
plot(observed_disparity, ylab = "Sum of variances")
abline(v = 4, col = "red")
legend("topleft", pch = NULL, legend = "B", bty = "n", cex = 2)
```

Using this example dataset, one might be interested in testing whether the K-Pg extinction had an effect on disparity through time.
However, one legitimate question to ask would be whether such effect could be detectable in the first place in this dataset.
Do test this we can simulate some datasets with similar properties as the observed data and measure changes in disparity in these simulated datasets.
If the simulations contain some mass extinction events we will then be able to see whether they cause a detectable change in disparity or not.

### Simulating trees

The first and simplest way is to simulate tree topologies that have similar properties than the observed ones.
To do so, we need to use some **speciation** parameter indicating the rate at which lineages speciate (_aka_ "birth" or "$\lambda$") and an **extinction** parameter indicating the rate at which they go extinct (_aka_ "death" or "$\mu$").
These two parameters can be crudely calculated from a give tree by counting the number of speciations and extinction events through time (note that better analytical methods exist, e.g. `ape::birthdeath`; @lawton1995extinction; @ape).
We also need a stopping rule for when to stop the simulations (in our case when reaching 50 taxa, extinct or living).
This will produce a single random tree using the input parameters (Fig. 3).
Note that I will not discuss the options here in great details.
Much more information can be found in the [`treats` manual](http://tguillerme.github.io/treats.html).
Note that because of stochasticity the `treats` function can automatically restart if the simulations did not produce the required tree (i.e. if all lineages go extinct before reaching the stopping rule; using the `null.error` option).

```{r, message = FALSE}
## Using the birth-death parameters from the observed tree
my_bd_params <- crude.bd.est(BeckLee_tree)
## Setting the stopping rule (stop after 30 time units)
stop_rule <- list(max.time = 30)

## Simulate the tree
set.seed(1)
sim_tree <- treats(bd.params  = my_bd_params,
                   stop.rule  = stop_rule,
                   null.error = 100)
```
```{r simtree, echo = FALSE, fig.cap="Figure 3: A randomly simulated tree with similar properties as the observed one. Note that the time here is expressed in arbitrary units.", fig.width = 6, fig.height = 6}
## Plotting the resulting tree
plot(sim_tree, show.tip.label = FALSE)
## Adding the phylogenetic axis
axisPhylo()
```

### Simulating trees and traits

For our specific question, we will also need to simulate some traits associated with each node and tip.
For simplicity we will simulate a two dimensional Brownian Motion trait.
To do so, we can create a `"traits"` object with the function `make.traits`.
This results in a 2 dimensional trait space for all the simulated species and their nodes (Fig. 4).

```{r, message = FALSE}
## Creating a trait in 2 dimensions.
my_traits <- make.traits(process = BM.process, n = 2)

## Simulate the tree and traits
set.seed(1)
sim_data <- treats(traits     = my_traits,
                   bd.params  = my_bd_params,
                   stop.rule  = stop_rule,
                   null.error = 100)
```
```{r simtreetraits, echo = FALSE, fig.cap="Figure 4: A randomly simulated tree with a 2 dimensional random Brownian Motion trait.", fig.width = 16, fig.height = 8}
par(mfrow = c(1,2))
## Plotting one trait through time
plot(sim_data, ylab = "Trait 1", las = 1, main = "Trait 1 through time")
## Plotting the two dimensions against each other
plot(sim_data, trait = c(2,1), ylab = "Trait 1", xlab = "Trait 2", las = 1, main = "Traits 1 and 2")
```

### Simulating a trees and traits with events

However, for our specific question we want to simulate both a tree and some traits but with an extinction event in order to get some simulated data to answer our question.
To do so, we can create two different `"events"` object with the function `make.events`.
The first one will simulate a random extinction event after reaching half of the simulations (15 time units) and then making three quarter (0.75) of the taxa go extinct.
The second one will simulate a random extinction but based on trait values: after reaching time 20, all the species with positive trait values will go extinct.
Both scenarios thus illustrate two different types of mass extinction (selective and random @puttick2020complex).
Note that for such simulations, it is easy to replicate the scenarios to have a distribution of trees and traits using `replicate`.

```{r, message = FALSE}
## Creating a random mass extinction
random_extinction <- make.events(
    target       = "taxa",
    condition    = time.condition(15),
    modification = random.extinction(0.75))
## Creating an extinction that removes species with positive trait values
positive_extinction <- make.events(
    target = "taxa",
    condition = time.condition(15),
    modification = trait.extinction(x = 0, condition = `>=`))

set.seed(2)
## Simulate the tree and traits with a random extinction event
sim_rand_extinction <- treats(
                   traits     = my_traits,
                   bd.params  = my_bd_params,
                   stop.rule  = stop_rule,
                   events     = random_extinction,
                   null.error = 100,
                   replicates = 50)

## Simulate the tree and traits with a random extinction event
sim_trait_extinction <- treats(
                   traits     = my_traits,
                   bd.params  = my_bd_params,
                   stop.rule  = stop_rule,
                   events     = positive_extinction,
                   null.error = 100,
                   replicates = 50)
```
```{r simtreetraitsevents, echo = FALSE, fig.cap="Figure 5: Two different mass extinction events: (A) 75% of species go extinct; (B) all species with positive trait values go extinct.", fig.width = 16, fig.height = 8}
par(mfrow = c(1,2))
## Visualising the difference between both scenarios
plot(sim_rand_extinction[[11]], main = "Random extinction")
abline(v = 15, col = "red")
legend("topleft", pch = NULL, legend = "A", bty = "n", cex = 2)
plot(sim_trait_extinction[[10]], main = "Positive extinction")
legend("topleft", pch = NULL, legend = "B", bty = "n", cex = 2)
abline(v = 15, col = "red")
```

Again, the syntax is not developed in this example.
For more information, you can refer to the [`treats` manual](http://tguillerme.github.io/treats.html).
Once we have simulated a distribution of trees and traits with the two extinction scenarios, we can measure disparity as in Fig. 1 for all the simulated data and compare it to the observed disparity.

```{r simdisparity, warning = FALSE, echo = FALSE, fig.cap="Figure 6: Comparison of the observed sum of variances (dashed line) to the simulated one (ligth and dark grey polygons and black line representing respectively the 95% and 50% confidence interval and the median value).", fig.width = 16, fig.height = 8}
## Wrapper function for measuring disparity across the datasets
measure.disparity <- function(one_simulation) {
    ## transform the data
    sim <- dispRitreats(one_simulation, scale.tree = FALSE)
    ## Calculate disparity
    disparity <- dispRity(
             chrono.subsets(data = sim$data, tree = sim$tree,
                            method = "continuous", model = "acctran",
                            time = seq(from = 30, to = 0, by = -5)),
                            metric = c(sum, variances))
    ## Scale the disparity
    disparity <- unlist(get.disparity(disparity))
    return(disparity/max(disparity))
}

## Measuring disparities
random_extinction_disparity <- do.call(rbind, lapply(sim_rand_extinction, measure.disparity))
selective_extinction_disparity <- do.call(rbind, lapply(sim_trait_extinction, measure.disparity))

## Get the observed disparity
observed_disparity <- rescale.dispRity(observed_disparity)
observed_disparity <- unlist(get.disparity(observed_disparity))

## Plotting the results
polygon.plot <- function(disparity_results, main = "") {
    ## Get the quantiles
    quantiles <- apply(disparity_results, 2, quantile, probs = c(0.025, 0.25, 0.5, 0.75, 0.975), na.rm = TRUE)

    plot(NULL, xlim = c(1, ncol(quantiles)), ylim = range(quantiles), ylab = "scaled sum of ranges", xlab = "Relative time", main = main )
    polygon(x = c(1:ncol(quantiles), ncol(quantiles):1),
            y = c(quantiles[1, ], rev(quantiles[5, ])),
            col = "lightgrey", border = FALSE)
    polygon(x = c(1:ncol(quantiles), ncol(quantiles):1),
            y = c(quantiles[2, ], rev(quantiles[4, ])),
            col = "grey", border = FALSE)
    lines(x = c(1:ncol(quantiles)), y = quantiles[3, ])
}

par(mfrow = c(1,2))
polygon.plot(random_extinction_disparity, main = "Random extinction")
abline(v = 4, col = "red")
lines(x = 1:7, y = observed_disparity, lty = 2, lwd = 2)
polygon.plot(selective_extinction_disparity, main = "Selective extinction")
abline(v = 4, col = "red")
lines(x = 1:7, y = observed_disparity, lty = 2, lwd = 2)
```

From these results we can draw some preliminary conclusions suggesting that the observed change disparity is more likely due to a random mass extinction rather than a selective one.
However this is of course a really crude way to test this and is not to be taken literally here.
The aim here is to illustrate the use of the package.
To test the question asked here more work is needed (i.e. more and better quality data, and more thorough methods, e.g. using a rank envelope test @murrell2018global).

## Major functionalities

### Simulating traits

multivariate stuff @adams2019multivarPCM

### Modifying the birth-death process

Dynamic ones will be for bd.params = list(speciation = runif)

### Creating events

# Discussion

## Modularity

The modular structure allows

## `treats` compared to other packages

As previously mentioned, many other excellent packages are available for simulating biologically realistic data.
However each of these packages focuses on simulating specific scenarios and none provide the levels of modularity implemented in `treats`.
Note that these comparison is somewhat a strawman argument since the packages I am comparing to `treats` here are used for many other anlyses or simulations not at all implemented in `treats` and where probably never intended to be a modular tree and traits simulation package in the first place.

package | tree simulation parameters | trait simulation models | events 
--------|----------------|----------------|----------------|--------
`TreeSim` | speciation, extinction and sampling | NA | rate shifts and mass extinctions
`diversitree` | speciation, extinction and 9 types of simulations available | 5 implemented models | NA 
`RPANDA` | NA | 9 implemented models | NA 
`traits` | infinite (modulable through `make.bd.params` and `make.modifiers`) | infinite (modulable through `make.traits` with 6 implemented models) | infinite (modulable through `make.events`)

Table 1: comparison between the `traits` package and other packages.

## Further directions

<!-- Integrating with dispRity -->
<!-- Lineage specific simulations. -->
<!-- Abiotic events -->
<!-- Time depend simulations -->

# Conclusion
The `treats` is modular and nice to use.

Extended manual following recommendations from @cooper2016dark

## Package location
The `treats` package is available 
<!-- % on the [CRAN](https://cran.r-project.org/web/packages/treats/index.html) or  -->
on [GitHub](https://github.com/TGuillerme/treats) with more associated information.
All the versions of the package are archived on ZENODO with associated [DOI:10.5281/zenodo.7970384](https://doi.org/10.5281/zenodo.7970384).

## Acknowledgments
Thanks to Mark Puttick for inspiring to develop the package through previous collaborations. Thanks to Andrew Beckerman, Ian Brennan, Gustavo Burin, Christopher Cooney, Natalie Cooper, Alex Cranston, Jasmine Hardie, Tom Lansley, Clement Prieul, Joe Rees, James Rule, Sophie Ryan and Gavin Thomas, for support and useful comments on late stages of the development of this package. This work was funded by UKRI-NERC Grant NE/T000139/1 and a Royal Society University Research Fellowship (URF R 180006 to Gavin Thomas).

# References

### 2023-05-31 paper/treats_paper.Rmd
First draft done!

@@
 ---
+
+<!-- no more than 3000-4000 words total -->
 
@@
 
+Because of its modular architecture it is impossible to provide an exhaustive list of all the package functionalities.
+The following sections provide an overview of the three main functions displayed in Figure 1 (`make.traits`, `make.modifiers` and `make.events`).
+
 ### Simulating traits
 
-multivariate stuff @adams2019multivarPCM
+Traits are simulated via the `make.traits` function given one or more processes, number of dimensions per process and some starting value(s).
+Essentially, the generation of new trait values is based on a process (a `function`) modifying a trait value (`x0`) relative to some branch length (`edge.length`).
+For example, a simple Brownian Motion process could be generated by the function `rnorm` where the it modifies the trait value `x0` relative to the branch length (`edge.length`).
+The longer the branch length, the more likely the new trait value will be different from `x0`:
+
+```{r, eval = FALSE}
+## Example of a Brownian Motion process
+rnorm(n = 1, mean = x0, sd = edge.length)
+```
+
+This can be generated in `treats` by providing this process as a function to `make.traits` which automatically verifies the structure of the process by checking the arguments and applicability to the `treats` algorithm:
+
+```{r, eval = FALSE}
+## Functionalising the previous example of a Brownian Motion process
+my.bm.process <- function(x0, edge.length) {
+    rnorm(n = 1, mean = x0, sd = edge.length)
+} 
+## Creating the traits object
+my.trait <- make.traits(process = my.bm.process)
+```
+
+Of course biological data can be much more complex and often multivariate [@adams2019multivarPCM] requiring the users to develop more complex function to cater their specific needs.
+However, the `treats` package contains a list of pre-built processes that are "ready-to-use":
+
+ * `BM.process` and `OU.process` which are generalised Brownian Motion and Ornstein-Uhlenbeck processes in any number of dimensions (including possible correlation);
+ * `multi.peak.process` which is a generalised Ornstein-Uhlenbeck process (uni- or multi-dimensional) which allows for multiple peaks (usually called the "alpha" parameters; @lande1976OU);
+ * `repulsion.process` which is a unidimensional process generating trait values that don't overlap with previously generated trait values;
+ * `discrete.process` which generates discrete trait values;
+ * `no.process` which ignores the branch length (i.e. a time independent process);
+
+This list will be updated through the lifetime of the package and is documented under the generic `?trait.process` function.
+
+Note that the `treats` package primary aim is to generate both a tree and some traits at the same time.
+However, it is possible to also just generate traits with a given topology.
+This is done through the function `map.traits` that intakes one or more trees (`"phylo"`, or `"multiPhylo"`) and a `"traits"` object generated by `make.traits` as described above.
 
@@
 
-Dynamic ones will be for bd.params = list(speciation = runif)
+Modifying the birth-death process can be done in several ways.
+Most easily it is done through changing the stopping rules through the `stop.rule` argument (number of total taxa or living ones, or time of the simulation).
+Equally straightforward, one can modify the parameters of the birth-death process: the speciation ($\lambda$) and the extinction ($\mu$) ones.
+These can be either fixed values (for constant speciation and extinction) or values drawn from distributions.
+This can be specified through the `make.bd.params` function.
+
+However, more interestingly, it is possible to directly modify how the birth death algorithm works through `make.modifiers`.
+This function allows to directly modify the three main components of the birth-death algorithm as described above.
+By default, the algorithm uses the following algorithms:
+
+1.
+Generating branch length: this is done by drawing a value from an exponential distribution with the rate being function of the current number of lineages scaled by the speciation and extinction parameters
+
+```{r, eval = FALSE}
+## The default branch length generation
+rexp(1, rate = number_of_lineages * (speciation + extinction))
+```
+
+2. Selecting a lineage among the currently living ones: this is done by simply sampling across the available (living) lineages:
+
+```{r, eval = FALSE}
+## The default lineage selection
+sample(number_of_lineages, 1)
+```
+
+3.
+Choosing whether that lineage goes extinct (becomes a tip) or speciates (becomes a node): this is done by drawing a random number between 0 and 1 and comparing it to the ratio of speciation and turnover (speciation + extinction).
+If the random number is smaller than the ratio of speciation and turnover, the lineage speciates, else it goes extinct (not that if extinction is always null, the speciation/turnover is always 1 so that the lineage always speciates as expected):
+
+```{r, eval = FALSE}
+## The default speciation/extinction decider
+runif(1) <= (speciation / (speciation + extinction))
+```
+
+The function `make.modifiers` allows to specifically modify any of these components by providing a different function for each part of the algorithm.
+For example, we could modify the the three functions above so that the branch length is not dependent on the number of lineages, the sampling is always the first species available and the extinction is drawn from a normal distribution.
+Note that these function need a specific syntax that is detailed in the [`treats` manual](http://tguillerme.github.io/treats.html).
+Briefly, the inputs needs to be always as indicated below and the speciation and extinction parameters as described above are actually handled as part of a list (`bd.params$speciation` and `bd.params$extinction`):
+
+```{r, eval = FALSE}
+## Lineage independent waiting:
+lineage.independent <- function(bd.params,
+                                lineage = NULL,
+                                trait.values = NULL,
+                                modify.fun = NULL) {
+    my_rate <- bd.params$speciation + bd.params$extinction
+    return(rexp(1, rate = my_rate))
+}
+## Selecting always the first species
+select.first <- function(bd.params,
+                         lineage = NULL,
+                         trait.values = NULL,
+                         modify.fun = NULL) {
+    return(as.integer(1))
+}
+## Random normal speciation
+normal.speciation <- function(bd.params,
+                              lineage = NULL,
+                              trait.values = NULL,
+                              modify.fun = NULL) {
+    my_turnover <- bd.params$speciation/
+                  (bd.params$speciation + bd.params$extinction)
+    return(rnorm(1) <= my_turnover)
+}
+
+## Creating the modifier object
+modified.birth.death <- make.modifiers(branch.length = lineage.independent,
+                                       selection     = select.first,
+                                       speciation    = normal.speciation)
+```
+
+This `modified.birth.death` is a `"modifiers"` object that is automatically checked for syntax and tested and can be used as a `modifiers` argument in `treats`.
 
@@
 
-# Discussion
+The final major argument to be passed to `treats` are the `"events"` objects generated through `make.events`.
+This function works similarly than `make.traits` and `make.modifiers` where the user has to describe the desired event using functions.
+For this the following information needs to be specified:
 
-## Modularity
+ * the `target` designating what the event should be applied to (e.g. `"taxa"` for modifying the number of species, `"traits"` for modifying the traits, etc.)
+ * the `condition` which is a function returning a logical value of when to trigger the event (e.g.
+when reaching a certain number of taxa, after some specific time has ellapsed or if some specific trait value is reached, etc.)
+ * the `modification` which is a function that specifically modifies an internal object in the `treats` algorithm.
 
-The modular structure allows
+Because of the modular architecture, is is impossible to provide an exhaustive list of events so you can refer to the [`treats` manual](http://tguillerme.github.io/treats.html) for many different detailed examples.
 
+Briefly though, this is how the mass extinction events are designed in the "brief applied example" above.
+
+```{r, eval = FALSE}
+## Creating an extinction that removes species with positive trait values
+positive_extinction <- make.events(
+    target       = "taxa",
+    condition    = time.condition(15),
+    modification = random.extinction(0.75))
+```
+
+For this event, the target is the number of taxa in the simulations.
+This is indicated using the `target = "taxa"` argument.
+Then the event is triggered using the argument `time.condition(15)` and modifies the internal `lineage` object using the ``trait.extinction(x = 0, condition = `>=`)`` function.
+`time.condition` and `trait.extinction` are both function factories (functions that outputs other functions) which are not gonna be detailed here (see @wickham2019advanced for more details).
+Effectively these arguments can be passed directly as standard functions.
+For example, to trigger the event when reaching time 15 we can use the following function:
+
+```{r, eval = FALSE}
+## Returns TRUE when reaching time 15
+reaching.time15 <- function(bd.params, lineage, trait.values, time) {
+    return(time > 15)
+}
+```
+
+This will trigger the `modification` which is equivalent to the following function modifying the `lineage` internal list:
+
+```{r, eval = FALSE}
+removing.75.taxa <- function(bd.params, lineage, trait.values) {
+    ## Select a portion of the living species to go extinct
+    extinct <- sample(lineage$n, round(lineage$n * 0.75))
+
+    ## Update the lineage object
+    lineage$livings <- lineage$livings[-extinct]
+    lineage$n       <- lineage$n - length(extinct)
+    return(lineage)
+}
+```
+
+Hence, the extinction event described above is equvivalent to the following one:
+
+```{r, eval = FALSE}
+## Creating an extinction that removes species with positive trait values
+positive_extinction <- make.events(
+    target       = "taxa",
+    condition    = reaching.time15,
+    modification = removing.75.taxa)
+```
+
+### Visualising results
+
+The `treats` package also comes with tools to visualise trees and traits together or separately.
+This can be done through the generic S3 `plot` function (calling `plot.treats`) and allows to display up to three traits or two traits and time (in 3D) as displayed in Figures 4 and 5.
+Note that you can always use the `plot` functions to data not generated by the `treats` function using the `make.treats` function (transforming trees and data into `"treats"` objects).
+
+# Additional information
+
+## Manuals and vignette
+
+As mentioned through this manuscript, the `treats` package comes with internal documentation (e.g.
+`?treats`) but also with a thorough and extended vignette in a gitbook format: the [`treats` manual](http://tguillerme.github.io/treats.html).
+This manual is designed so that it can be regularly updated and enhanced through the lifetime of the package facilitating the interface between methods development and usage [@cooper2016dark].
+
+## Further directions
+
+This paper describes the first version of the `treats` package.
+However, I intend, either personally or through collaborations to continuously develop this package.
+For example future planned versions will include: lineage specific simulations; abiotic events; time dependent simulations and; a better integration with the `dispRity` package.
+This will be done while keeping track change (through `NEWS` file), continuous integration and unit testing. 
+Furthermore, I am planning to develop the GitHub page associated with the package to encourage users to share their simulation templates and build library from them to facilitate the work for future users.
+
+## Repeatability and reproducibility
+This paper is entirely reproducible from an Rmarkdown document available on [GitHub](https://github.com/TGuillerme/treats/paper/).
+The data used for the example above [@beck2014ancient] is available from the `dispRity` package [@dispRity].
+
+# Conclusion
+The `treats` package modular architecture allows workers to develop their own specific biological simulation scenarios based on their own specific research question.
+The pipeline of the package through the different `"treats"` objects (`"traits"`, `"modifiers"` and `"events"`) also allows workers to generate publication standard results through plotting but also with easily reproducible and reusable scripts.
+
+## Package location
+The `treats` package is available 
+<!-- % on the [CRAN](https://cran.r-project.org/web/packages/treats/index.html) or  -->
+on [GitHub](https://github.com/TGuillerme/treats) with more associated information.
+All the versions of the package are archived on ZENODO with associated [DOI:10.5281/zenodo.7970384](https://doi.org/10.5281/zenodo.7970384).
+
+## Acknowledgments
+Thanks to Mark Puttick for inspiring to develop the package through previous collaborations.
+Thanks to Andrew Beckerman, Ian Brennan, Gustavo Burin, Christopher Cooney, Natalie Cooper, Alex Cranston, Jasmine Hardie, Tom Lansley, Clement Prieul, Joe Rees, James Rule, Sophie Ryan and Gavin Thomas, for support and useful comments on late stages of the development of this package.
+This work was funded by UKRI-NERC Grant NE/T000139/1 and a Royal Society University Research Fellowship (URF R 180006 to Gavin Thomas).
+
+# References
+
+<!-- 
 ## `treats` compared to other packages
@@
 
-Table 1: comparison between the `traits` package and other packages.
-
-## Further directions
-
-<!-- Integrating with dispRity -->
-<!-- Lineage specific simulations. -->
-<!-- Abiotic events -->
-<!-- Time depend simulations -->
-
-# Conclusion
-The `treats` is modular and nice to use.
-
-Extended manual following recommendations from @cooper2016dark
-
-## Package location
-The `treats` package is available 
-<!-- % on the [CRAN](https://cran.r-project.org/web/packages/treats/index.html) or  -->
-on [GitHub](https://github.com/TGuillerme/treats) with more associated information.
-All the versions of the package are archived on ZENODO with associated [DOI:10.5281/zenodo.7970384](https://doi.org/10.5281/zenodo.7970384).
-
-## Acknowledgments
-Thanks to Mark Puttick for inspiring to develop the package through previous collaborations.
-Thanks to Andrew Beckerman, Ian Brennan, Gustavo Burin, Christopher Cooney, Natalie Cooper, Alex Cranston, Jasmine Hardie, Tom Lansley, Clement Prieul, Joe Rees, James Rule, Sophie Ryan and Gavin Thomas, for support and useful comments on late stages of the development of this package.
-This work was funded by UKRI-NERC Grant NE/T000139/1 and a Royal Society University Research Fellowship (URF R 180006 to Gavin Thomas).
-
-# References
+Table 1: comparison between the `traits` package and other packages. -->

### 2023-06-01 paper/CoverLetter.tex
Create CoverLetter.tex

\documentclass[11pt]{letter}
\usepackage[a4paper,left=2.5cm, right=2.5cm, top=1cm, bottom=1cm]{geometry}
\usepackage{hyperref}
\usepackage[osf]{mathpazo}
\signature{Thomas Guillerme}
\address{School of Biosciences\\University of Sheffield\\Sheffield, S10 2TN\\United Kingdom\\guillert@tcd.ie}
\longindentation=0pt
\begin{document}

\begin{letter}{}
\opening{Dear Editors,}

% Simulations are important in biology
% Studying multivariate data as an alternative to univariate data has become increasingly popular in the last decades in both ecology and evolution. This includes, for example, studying form and function (e.g. Diaz et al., 2016, Nature) or morphological diversity (i.e. disparity - e.g. Brusatte et al. 2012, Science). These methods all use a multidimensional space (through ordination or other techniques) representing the combination of all possible observed traits and then measuring some aspect of the space (e.g. its hypervolume) for different groups, and comparing these to each other or to a null hypothesis.

% There are many tools but they are not modular
% There are a multitude of ways of carrying out each of these steps, and numerous software implementations in several languages (e.g. in R: Claddis, geomorph, ade4, vegan, or in javascript: GINGKO). However, each of these implementations is particular to the package author’s defini- tion of the multidimensional space, metric and test. This leads researchers to use these toolkits by default, thus preventing them from using the best definitions of the space/metric/test for their particular question or dataset. At worst this leads to incorrect statistic claims, and, at best to time consuming code retro-engineering.

% Here is a treats, a modular package for simulations in biology
% In this application manuscript entitled “dispRity: an R package for measuring disparity”, I present a complete and flexible R package for running multidimensional analysis where each parameter can be easily defined by the user (the space, the metric and the test). This package provides a tidy and modular environment where users can use their own multidimensional space defined as a matrix and apply to it any metric and any test (provided as functions). This allows for highly modular multidimensional analysis and the tidy interface allows easily reproducible analysis. All functions have documentation and unit tests. Additionally, this package comes with an in depth Gitbook manual covering most functionalities that will be updated regularly following user requests. Finally, the package is likely to be well cited as it is already being used in at least five projects that I am aware of.

We look forward to hearing from you soon,

\closing{Yours sincerely,}

\end{letter}
\end{document}

%Editors:
%Laura Graham, University of Southampton, UK
% Timothée Poisot, Université de Montréal, Canada
% Samantha Price, Clemson University, USA

%Reviewers
% Emmanuel Paradis emmanuel.paradis@umontpellier.fr
% Will Gearty wgearty@amnh.org
% Tanja Stadler tanja.stadler@bsse.ethz.ch
% Joel Barido joelle.barido-sottani@m4x.org

### 2023-06-01 paper/treats_paper.Rmd
Update treats_paper.Rmd

@@
 <!-- no more than 3000-4000 words total -->
+<!-- The code checklist must be filled in and uploaded as a 'File for editors'.  -->
+$^1$ School of Biosciences, University of Sheffield, Sheffield, S10 2TN, United Kingdom. guillert@tcd.ie
 
-$^1$ School of Biosciences, University of Sheffield, Sheffield, S10 2TN, United Kingdom. guillert@tcd.ie
+Running head:
+`treats`: TREes And Traits Simulations
 

### 2023-06-07 paper/treats_paper.Rmd
up paper

@@
   pdf_document: 
+    keep_tex: yes
     fig_width: 8
@@
 
-<!-- no more than 3000-4000 words total -->
 <!-- The code checklist must be filled in and uploaded as a 'File for editors'.  -->
@@
 
-1. Simulating biological realistic data can be an important step to understand and investigate biological mechanisms.
+1. Simulating biological realistic data is an important step to understand and investigate biodiversity.
 Simulated data can be used to generate null, base line or neutral models.
-These models can be used either in comparison to observed data to estimate the mechanisms that generated the data.
+These can be used either in comparison to observed data to estimate the mechanisms that generated the data.
 Or they can be used to explore, understand and develop theoretical advances by proposing toy models.
 
-2. In evolutionary biology, such simulations often involve the need of an evolutionary process where descent with modification is at the core of how the simulated data is generated.
-These evolutionary processes can then be nearly infinitely modified to include much more complex processes to affect the simulations such as traits co-evolution, competition mechanisms or mass extinction events.
+2. In evolutionary biology, simulations often involve the need of an evolutionary process where descent with modification is at the core of how the simulated data is generated.
+These evolutionary processes can then be nearly infinitely modified to include complex processes that affect the simulations such as traits co-evolution, competition mechanisms or mass extinction events.
 
 3. Here I present the `treats` package, a modular `R` package for trees and traits simulations.
-This package is based on a simple birth death algorithm from which all core steps can easily be modified by users (hence the modularity).
+This package is based on a simple birth death algorithm from which all steps can easily be modified by users.
 It also provides a tidy interface through the `treats` object, allowing users to easily run reproducible simulations.
-
-4. Here I present the core algorithms and their integrated modularity and provide an example of how to use the `treats` package when studying disparity, the diversity of traits.
-The `treats` package also comes with an extend manual regularly updated following users' questions or suggestions which comes as an extended supplementary material to this paper.
+It also comes with an extend manual regularly updated following users' questions or suggestions.
 
@@
 
-<!-- % Simulations in biology -->
 Comparing biological patterns is one of the key ways to understand mechanisms in evolutionary biology.
 This lead to the development of phylogenetic comparative methods as key methodologically driven topic in ecology, evolution and palaeontology [@felsensteinPCM; @pennell2013review].
-As indicated in the name, phylogenetic comparative methods rely on comparing patterns in a phylogenetic context (i.e.
-descent with modification) to understand biological mechanisms or concepts [@harmon2019book].
-These comparisons can be done between observed patterns under different conditions or against null, neutral or baseline models (see @bausman2018neutral for distinctions or common misconceptions) suggesting different processes or mechanisms.
-<!-- % These types phylogenetic comparative methods analysis of traits through time are routinely done in palaeontology (disparity; @guillerme2020disparities) or in ecology (functional diversity; @mammola2021concepts).
--->
-For example different traits distribution for species with different diets [@deepak2023diet],
-<!-- % different habitats https://onlinelibrary.wiley.com/doi/full/10.1111/jbi.12939 -->
-or by comparing some observed pattern to one simulated under null or base conditions [@miller2022alternating].
+As indicated in the name, phylogenetic comparative methods rely on comparing patterns in a phylogenetic context to understand biological mechanisms or concepts [@harmon2019book].
+These comparisons can be done between observed patterns under different conditions or against null, neutral or baseline models (see @bausman2018neutral for distinctions) suggesting different processes or mechanisms.
+For example different traits distribution for species with different diets [@deepak2023diet], or habitats [@pinto2017geographical] or by comparing some observed pattern to one simulated under null or base conditions [@miller2022alternating].
 In theory, workers can follow the research pipeline of thinking of a specific mechanism (e.g.
@@
 
-<!-- % What already exists -->
 In practice, these evolutionary simulations can be done relatively easily on computers using a birth-death process [@feller1939birthdeath; @stadler2010birthdeath; @diversitree].
@@
 
-<!-- % The problems with the absence of modularity -->
 Although the packages mentioned above are excellent and routinely used with fast and reliable algorithms and associated documentation, they are all designed for specific tasks and don't allow much modification beyond the input parameters designed by the authors.
@@
 
-<!-- % Very briefly how does it work -->
 `treats` is based on the eponymous `treats` function that allows to simulate a phylogeny and some trait(s) simultaneously.
@@
 ## Calculating disparity on the two first dimensions only
-observed_disparity <- dispRity(boot.matrix(time_slices), metric = c(sum, variances), dimensions = c(1,2))
+observed_disparity <- dispRity(boot.matrix(time_slices),
+                               metric = c(sum, variances),
+                               dimensions = c(1,2))
 ```
@@
 Using this example dataset, one might be interested in testing whether the K-Pg extinction had an effect on disparity through time.
-However, one legitimate question to ask would be whether such effect could be detectable in the first place in this dataset.
-Do test this we can simulate some datasets with similar properties as the observed data and measure changes in disparity in these simulated datasets.
-If the simulations contain some mass extinction events we will then be able to see whether they cause a detectable change in disparity or not.
+But can such effect be detected in the first place?
+We can test this by simulating some datasets with similar properties as the observed data and measure changes in disparity in these simulated datasets.
 
@@
 
-The first and simplest way is to simulate tree topologies that have similar properties than the observed ones.
+The first and simplest way is to simulate tree topologies that have similar properties than the observed one.
 To do so, we need to use some **speciation** parameter indicating the rate at which lineages speciate (_aka_ "birth" or "$\lambda$") and an **extinction** parameter indicating the rate at which they go extinct (_aka_ "death" or "$\mu$").
-These two parameters can be crudely calculated from a give tree by counting the number of speciations and extinction events through time (note that better analytical methods exist, e.g.
+These two parameters can be crudely calculated from a give tree by counting the number of speciations and extinctions through time (but see better methods, e.g.
 `ape::birthdeath`; @lawton1995extinction; @ape).
-We also need a stopping rule for when to stop the simulations (in our case when reaching 50 taxa, extinct or living).
+We also need a stopping rule for when to stop the simulations (in our case when reaching 30 time units).
 This will produce a single random tree using the input parameters (Fig. 3).
@@
 Much more information can be found in the [`treats` manual](http://tguillerme.github.io/treats.html).
-Note that because of stochasticity the `treats` function can automatically restart if the simulations did not produce the required tree (i.e.
-if all lineages go extinct before reaching the stopping rule; using the `null.error` option).
 
@@
 ```
-```{r simtreetraits, echo = FALSE, fig.cap="Figure 4: A randomly simulated tree with a 2 dimensional random Brownian Motion trait.", fig.width = 16, fig.height = 8}
+```{r simtreetraits, echo = FALSE, fig.cap="Figure 4: A randomly simulated tree with a 2 dimensional random Brownian Motion trait.
+Orange, light blue and dark blue dots respectively represent nodes, fossils and living species.", fig.width = 16, fig.height = 8}
 par(mfrow = c(1,2))
@@
 
-However, for our specific question we want to simulate both a tree and some traits but with an extinction event in order to get some simulated data to answer our question.
+To answer our question, we also want to simulate an extinction event.
 To do so, we can create two different `"events"` object with the function `make.events`.
@@
 Both scenarios thus illustrate two different types of mass extinction (selective and random @puttick2020complex).
-Note that for such simulations, it is easy to replicate the scenarios to have a distribution of trees and traits using `replicate`.
 
@@
 ```
-```{r simtreetraitsevents, echo = FALSE, fig.cap="Figure 5: Two different mass extinction events: (A) 75% of species go extinct; (B) all species with positive trait values go extinct.", fig.width = 16, fig.height = 8}
+```{r simtreetraitsevents, echo = FALSE, fig.cap="Figure 5: Two different mass extinction events: (A) 75% of species go extinct; (B) all species with positive trait values go extinct.
+The red line marks the extinction event.", fig.width = 16, fig.height = 8}
 par(mfrow = c(1,2))
@@
 
-Again, the syntax is not developed in this example.
-For more information, you can refer to the [`treats` manual](http://tguillerme.github.io/treats.html).
 Once we have simulated a distribution of trees and traits with the two extinction scenarios, we can measure disparity as in Fig. 1 for all the simulated data and compare it to the observed disparity.
@@
 
-From these results we can draw some preliminary conclusions suggesting that the observed change disparity is more likely due to a random mass extinction rather than a selective one.
-However this is of course a really crude way to test this and is not to be taken literally here.
-The aim here is to illustrate the use of the package.
-To test the question asked here more work is needed (i.e. more and better quality data, and more thorough methods, e.g. using a rank envelope test @murrell2018global).
+From these results we can draw some preliminary conclusions: the observed change in disparity is more likely due to a random mass extinction rather than a selective one.
+This is of course a crude way of testing this, a more rigorous approach is likely needed to answer the question (i.e.
+more and better quality data, and more thorough methods, e.g.
+using a rank envelope test @murrell2018global).
 
@@
 
-Because of its modular architecture it is impossible to provide an exhaustive list of all the package functionalities.
 The following sections provide an overview of the three main functions displayed in Figure 1 (`make.traits`, `make.modifiers` and `make.events`).
@@
 The longer the branch length, the more likely the new trait value will be different from `x0`:
-
-```{r, eval = FALSE}
-## Example of a Brownian Motion process
-rnorm(n = 1, mean = x0, sd = edge.length)
-```
-
-This can be generated in `treats` by providing this process as a function to `make.traits` which automatically verifies the structure of the process by checking the arguments and applicability to the `treats` algorithm:
 
@@
 
- * `BM.process` and `OU.process` which are generalised Brownian Motion and Ornstein-Uhlenbeck processes in any number of dimensions (including possible correlation);
- * `multi.peak.process` which is a generalised Ornstein-Uhlenbeck process (uni- or multi-dimensional) which allows for multiple peaks (usually called the "alpha" parameters; @lande1976OU);
- * `repulsion.process` which is a unidimensional process generating trait values that don't overlap with previously generated trait values;
- * `discrete.process` which generates discrete trait values;
- * `no.process` which ignores the branch length (i.e. a time independent process);
-
-This list will be updated through the lifetime of the package and is documented under the generic `?trait.process` function.
+ * `BM.process` and `OU.process`: generalised Brownian Motion and Ornstein-Uhlenbeck processes in any number of dimensions (including possible correlation);
+ * `multi.peak.process`: a generalised Ornstein-Uhlenbeck process (uni- or multi-dimensional) which allows for multiple peaks (usually called the "alpha" parameters; @lande1976OU);
+ * `repulsion.process`: a unidimensional process generating trait values that don't overlap with previously generated trait values;
+ * `discrete.process`: generates discrete trait values;
+ * `no.process`: ignores the branch length (i.e. a time independent process);
 
@@
 However, it is possible to also just generate traits with a given topology.
-This is done through the function `map.traits` that intakes one or more trees (`"phylo"`, or `"multiPhylo"`) and a `"traits"` object generated by `make.traits` as described above.
+This is done through the function `map.traits` that intakes one or more trees and a `"traits"` object.
 
@@
 Most easily it is done through changing the stopping rules through the `stop.rule` argument (number of total taxa or living ones, or time of the simulation).
-Equally straightforward, one can modify the parameters of the birth-death process: the speciation ($\lambda$) and the extinction ($\mu$) ones.
+Equally straightforward, one can modify the parameters of the birth-death process through the `make.bd.params` function: the speciation ($\lambda$) and the extinction ($\mu$) ones.
 These can be either fixed values (for constant speciation and extinction) or values drawn from distributions.
-This can be specified through the `make.bd.params` function.
 
-However, more interestingly, it is possible to directly modify how the birth death algorithm works through `make.modifiers`.
-This function allows to directly modify the three main components of the birth-death algorithm as described above.
+It is also possible to directly modify how the birth-death algorithm works through `make.modifiers` by changing the three main components of the birth-death algorithm as described above.
 By default, the algorithm uses the following algorithms:
 
-1.
-Generating branch length: this is done by drawing a value from an exponential distribution with the rate being function of the current number of lineages scaled by the speciation and extinction parameters
+1. **Generating branch length** by drawing a value from an exponential distribution with the rate being function of the current number of lineages scaled by the speciation and extinction parameters
 
@@
 
-2. Selecting a lineage among the currently living ones: this is done by simply sampling across the available (living) lineages:
+2. **Selecting a lineage** among the currently living ones by simply sampling across the available (living) lineages:
 
@@
 3.
-Choosing whether that lineage goes extinct (becomes a tip) or speciates (becomes a node): this is done by drawing a random number between 0 and 1 and comparing it to the ratio of speciation and turnover (speciation + extinction).
-If the random number is smaller than the ratio of speciation and turnover, the lineage speciates, else it goes extinct (not that if extinction is always null, the speciation/turnover is always 1 so that the lineage always speciates as expected):
+C**hoosing whether that lineage goes extinct (becomes a tip) or speciates (becomes a node)** by drawing a random number between 0 and 1 and comparing it to the ratio of speciation and turnover (speciation + extinction).
+If the random number is smaller than the ratio of speciation and turnover, the lineage speciates, else it goes extinct:
 
@@
 
-The function `make.modifiers` allows to specifically modify any of these components by providing a different function for each part of the algorithm.
-For example, we could modify the the three functions above so that the branch length is not dependent on the number of lineages, the sampling is always the first species available and the extinction is drawn from a normal distribution.
+The function `make.modifiers` allows to specifically change any of these components by providing a different function for each part of the algorithm.
+For example, we could modify the three functions above so that the branch length is not dependent on the number of lineages, the sampling is always the first species available and the extinction is drawn from a normal distribution.
 Note that these function need a specific syntax that is detailed in the [`treats` manual](http://tguillerme.github.io/treats.html).
-Briefly, the inputs needs to be always as indicated below and the speciation and extinction parameters as described above are actually handled as part of a list (`bd.params$speciation` and `bd.params$extinction`):
 
@@
 
-This `modified.birth.death` is a `"modifiers"` object that is automatically checked for syntax and tested and can be used as a `modifiers` argument in `treats`.
-
 ### Creating events
 
-The final major argument to be passed to `treats` are the `"events"` objects generated through `make.events`.
-This function works similarly than `make.traits` and `make.modifiers` where the user has to describe the desired event using functions.
-For this the following information needs to be specified:
+The final major argument to be passed to `treats` are the `"events"` objects generated through `make.events` where the following information needs to be specified:
 
@@
 
-Because of the modular architecture, is is impossible to provide an exhaustive list of events so you can refer to the [`treats` manual](http://tguillerme.github.io/treats.html) for many different detailed examples.
-
-Briefly though, this is how the mass extinction events are designed in the "brief applied example" above.
+For a more exhaustive list of events so you can refer to the [`treats` manual](http://tguillerme.github.io/treats.html) with many different detailed examples.
+Briefly though, this is how the mass extinction events are designed in the example above.
 
@@
 Then the event is triggered using the argument `time.condition(15)` and modifies the internal `lineage` object using the ``trait.extinction(x = 0, condition = `>=`)`` function.
-`time.condition` and `trait.extinction` are both function factories (functions that outputs other functions) which are not gonna be detailed here (see @wickham2019advanced for more details).
+`time.condition` and `trait.extinction` are both function factories which are not gonna be detailed here (see @wickham2019advanced).
 Effectively these arguments can be passed directly as standard functions.
@@
 
-Hence, the extinction event described above is equvivalent to the following one:
+Hence, the extinction event described above is equivalent to the following one:
 
@@
 This can be done through the generic S3 `plot` function (calling `plot.treats`) and allows to display up to three traits or two traits and time (in 3D) as displayed in Figures 4 and 5.
-Note that you can always use the `plot` functions to data not generated by the `treats` function using the `make.treats` function (transforming trees and data into `"treats"` objects).
+These functions can also be used to visualise trees and traits together from non `"treats"` objects by using the `make.treats` function to transform them into `"treats"` objects.
 
@@
 
-As mentioned through this manuscript, the `treats` package comes with internal documentation (e.g.
+The `treats` package comes with internal documentation (e.g.
 `?treats`) but also with a thorough and extended vignette in a gitbook format: the [`treats` manual](http://tguillerme.github.io/treats.html).
@@
 # References
-
-<!-- 
-## `treats` compared to other packages
-
-As previously mentioned, many other excellent packages are available for simulating biologically realistic data.
-However each of these packages focuses on simulating specific scenarios and none provide the levels of modularity implemented in `treats`.
-Note that these comparison is somewhat a strawman argument since the packages I am comparing to `treats` here are used for many other anlyses or simulations not at all implemented in `treats` and where probably never intended to be a modular tree and traits simulation package in the first place.
-
-package | tree simulation parameters | trait simulation models | events 
---------|----------------|----------------|----------------|--------
-`TreeSim` | speciation, extinction and sampling | NA | rate shifts and mass extinctions
-`diversitree` | speciation, extinction and 9 types of simulations available | 5 implemented models | NA 
-`RPANDA` | NA | 9 implemented models | NA 
-`traits` | infinite (modulable through `make.bd.params` and `make.modifiers`) | infinite (modulable through `make.traits` with 6 implemented models) | infinite (modulable through `make.events`)
-
-Table 1: comparison between the `traits` package and other packages. -->

### 2023-06-07 paper/treats_paper.tex
Update treats_paper.tex
up paper

--- 2023-05-25 paper/treats_paper.tex
@@
-% 2961w - 30p
+% Options for packages loaded elsewhere
+\PassOptionsToPackage{unicode}{hyperref}
+\PassOptionsToPackage{hyphens}{url}
+%
+\documentclass[
+]{article}
+\usepackage{lmodern}
+\usepackage{amssymb,amsmath}
+\usepackage{ifxetex,ifluatex}
+\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
+  \usepackage[T1]{fontenc}
+  \usepackage[utf8]{inputenc}
+  \usepackage{textcomp} % provide euro and other symbols
+\else % if luatex or xetex
+  \usepackage{unicode-math}
+  \defaultfontfeatures{Scale=MatchLowercase}
+  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
+\fi
+% Use upquote if available, for straight quotes in verbatim environments
+\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
+\IfFileExists{microtype.sty}{% use microtype if available
+  \usepackage[]{microtype}
+  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
+}{}
+\makeatletter
+\@ifundefined{KOMAClassName}{% if non-KOMA class
+  \IfFileExists{parskip.sty}{%
+    \usepackage{parskip}
+  }{% else
+    \setlength{\parindent}{0pt}
+    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
+}{% if KOMA class
+  \KOMAoptions{parskip=half}}
+\makeatother
+\usepackage{xcolor}
+\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
+\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
+\hypersetup{
+  pdftitle={treats: a modular R package for simulating trees and traits.},
+  pdfauthor={Thomas Guillerme\^{}1},
+  hidelinks,
+  pdfcreator={LaTeX via pandoc}}
+\urlstyle{same} % disable monospaced font for URLs
+\usepackage[margin=1in]{geometry}
+\usepackage{color}
+\usepackage{fancyvrb}
+\newcommand{\VerbBar}{|}
+\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
+\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
+% Add ',fontsize=\small' for more characters per line
+\usepackage{framed}
+\definecolor{shadecolor}{RGB}{248,248,248}
+\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
+\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
+\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
+\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
+\newcommand{\BuiltInTok}[1]{#1}
+\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
+\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
+\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
+\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
+\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
+\newcommand{\ExtensionTok}[1]{#1}
+\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
+\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\ImportTok}[1]{#1}
+\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
+\newcommand{\NormalTok}[1]{#1}
+\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
+\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
+\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
+\newcommand{\RegionMarkerTok}[1]{#1}
+\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\usepackage{graphicx}
+\makeatletter
+\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
+\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
+\makeatother
+% Scale images if necessary, so that they will not overflow the page
+% margins by default, and it is still possible to overwrite the defaults
+% using explicit options in \includegraphics[width, height, ...]{}
+\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
+% Set default figure placement to htbp
+\makeatletter
+\def\fps@figure{htbp}
+\makeatother
+\setlength{\emergencystretch}{3em} % prevent overfull lines
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
+\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
+\newlength{\cslhangindent}
+\setlength{\cslhangindent}{1.5em}
+\newenvironment{cslreferences}%
+  {\setlength{\parindent}{0pt}%
+  \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces}%
+  {\par}
 
-\documentclass[12pt,letterpaper]{article}
-\usepackage{natbib}
-\usepackage{pdflscape}
-\usepackage{fullpage}
-\usepackage{url}
-\usepackage{epsfig}
-\usepackage{caption}
-\usepackage{hyperref}
-\usepackage{enumerate}
-\usepackage{dcolumn}
-\usepackage{lineno}
-\usepackage[T1]{fontenc}
-\usepackage{textcomp}
-\usepackage{float}
-\usepackage[osf]{mathpazo}
-\restylefloat{table}
-\newcolumntype{d}[1]{D{.}{.}{#1}}
-
-\pagenumbering{arabic}
-
-%Pagination style and stuff
-\linespread{2}
-\raggedright
-\setlength{\parindent}{0.5in}
-\setcounter{secnumdepth}{0} 
-\renewcommand{\section}[1]{%
-\bigskip
-\begin{center}
-\begin{Large}
-\normalfont\scshape #1
-\medskip
-\end{Large}
-\end{center}}
-\renewcommand{\subsection}[1]{%
-\bigskip
-\begin{center}
-\begin{large}
-\normalfont\itshape #1
-\end{large}
-\end{center}}
-\renewcommand{\subsubsection}[1]{%
-\vspace{2ex}
-\noindent
-\textit{#1.}---}
-\renewcommand{\tableofcontents}{}
-%\bibpunct{(}{)}{;}{a}{}{,}
-
-%---------------------------------------------
-%
-%       START
-%
-%---------------------------------------------
-
-\newcommand{\treats}{\texttt{treats} }
+\title{\texttt{treats}: a modular \texttt{R} package for simulating
+trees and traits.}
+\author{Thomas Guillerme\(^1\)}
+\date{2023-06-07}
 
 \begin{document}
+\maketitle
 
-%Running head
-\begin{flushright}
-Version dated: \today
-\end{flushright}
-\bigskip
-\noindent RH: \treats package.
+\(^1\) School of Biosciences, University of Sheffield, Sheffield, S10
+2TN, United Kingdom.
+\href{mailto:guillert@tcd.ie}{\nolinkurl{guillert@tcd.ie}}
 
-\bigskip
-\medskip
-\begin{center}
+Running head: \texttt{treats}: TREes And Traits Simulations
 
-\noindent{\Large \bf \treats: a modular \texttt{R} package for simulating trees and traits.} 
-\bigskip
-
-\noindent {\normalsize \sc Thomas Guillerme$^{1,*}$}\\
-% , Natalie Cooper$^{2}$, Andrew P. Beckerman$^{1}$, and Gavin H. Thomas$^{1,3}$
-\noindent {\small \it 
-$^1$School of Biosciences, University of Sheffield, Sheffield, S10 2TN, United Kingdom.\\
-% $^2$Natural History Museum, Cromwell Road, London, SW7 5BD, United Kingdom.\\
-% $^3$Bird Group, Department of Life Sciences, the Natural History Museum at Tring, Tring, United Kingdom.\\
-}
-
-\end{center}
-\medskip
-\noindent{*\bf Corresponding author.} \textit{guillert@tcd.ie}\\  
-\vspace{1in}
-
-%Line numbering
-\modulolinenumbers[1]
-\linenumbers
-
-%---------------------------------------------
-%
-%       ABSTRACT
-%
-%---------------------------------------------
-
-% 3000-4000 words
-
-\newpage
-\begin{abstract}
+\hypertarget{abstract}{%
+\section{Abstract}\label{abstract}}
 
 \begin{enumerate}
-        \item Simulating biological realistic data can be an important step to understand and investigate biological mechanisms.
-        Simulated data can be used to generate null, base line or neutral models.
-        These models can be used either in comparison to observed data to estimate the mechanisms that generated the data.
-        Or they can be used to explore, understand and develop theoretical advances by proposing toy models.
-
-        \item In evolutionary biology, such simulations often involve the need of an evolutionary process where descent with modification is at the core of how the simulated data is generated.
-        These evolutionary processes can then be nearly infinitely modified to include much more complex processes to affect the simulations such as traits co-evolution, competition mechanisms or mass extinction events.
-
-        \item Here I present the \treats package, a modular \texttt{R} package for trees and traits simulations.
-        This package is based on a simple birth death algorithm from which all core steps can easily be modified by users (hence the modularity).
-        It also provides a tidy interface through the \treats object, allowing users to easily run reproducible simulations.
-
-        \item Here I present the core algorithms and their integrated modularity and provide an example of how to use the \treats package when studying disparity, the diversity of traits.
-        The \treats package also comes with an extend manual regularly updated following users' questions or suggestions which comes as an extended supplementary material to this paper.
-
+\def\labelenumi{\arabic{enumi}.}
+\item
+  Simulating biological realistic data is an important step to
+  understand and investigate biodiversity. Simulated data can be used to
+  generate null, base line or neutral models. These can be used either
+  in comparison to observed data to estimate the mechanisms that
+  generated the data. Or they can be used to explore, understand and
+  develop theoretical advances by proposing toy models.
+\item
+  In evolutionary biology, simulations often involve the need of an
+  evolutionary process where descent with modification is at the core of
+  how the simulated data is generated. These evolutionary processes can
+  then be nearly infinitely modified to include complex processes that
+  affect the simulations such as traits co-evolution, competition
+  mechanisms or mass extinction events.
+\item
+  Here I present the \texttt{treats} package, a modular \texttt{R}
+  package for trees and traits simulations. This package is based on a
+  simple birth death algorithm from which all steps can easily be
+  modified by users. It also provides a tidy interface through the
+  \texttt{treats} object, allowing users to easily run reproducible
+  simulations. It also comes with an extend manual regularly updated
+  following users' questions or suggestions.
 \end{enumerate}
 
-\end{abstract}
+Keywords: trees, traits, simulations, birth-death, null-models, ecology,
+evolution, disparity
 
-\noindent (Keywords: trees, traits, simulations, birth-death, null-models, ecology, evolution, disparity)\\
+\hypertarget{introduction}{%
+\section{Introduction}\label{introduction}}
 
-\vspace{1.5in}
+Comparing biological patterns is one of the key ways to understand
+mechanisms in evolutionary biology. This lead to the development of
+phylogenetic comparative methods as key methodologically driven topic in
+ecology, evolution and palaeontology (Felsenstein 1985; Pennell and
+Harmon 2013). As indicated in the name, phylogenetic comparative methods
+rely on comparing patterns in a phylogenetic context to understand
+biological mechanisms or concepts (Harmon 2019). These comparisons can
+be done between observed patterns under different conditions or against
+null, neutral or baseline models (see Bausman (2018) for distinctions)
+suggesting different processes or mechanisms. For example different
+traits distribution for species with different diets (Deepak, Gower, and
+Cooper 2023), or habitats (Pinto-Ledezma et al. 2017) or by comparing
+some observed pattern to one simulated under null or base conditions
+(Miller et al. 2022). In theory, workers can follow the research
+pipeline of thinking of a specific mechanism (e.g.~mass extinction
+allowing the surviving species to acquire new morphologies), collect
+some data to test this mechanism (e.g.~some traits of species across and
+extinction event) and then compare these patterns to one simulated under
+no specific conditions (e.g.~a null model where the traits evolve
+randomly regardless of an extinction event Puttick, Guillerme, and Wills
+(2020)). For such an approach, we need statistical and softwares
+solutions to simulate trees and data to generate these specific null
+models.
 
-\newpage
+In practice, these evolutionary simulations can be done relatively
+easily on computers using a birth-death process (Feller 1939; Stadler
+2010; FitzJohn 2012). A birth-death process is a continuous time Markov
+process that had been routinely implemented in \texttt{R} (R Core Team
+2023) to simulate biological phylogenies (e.g.~(Paradis and Schliep
+2019), (FitzJohn 2012)). These algorithm to generate phylogenetic trees
+can be coupled with other Markov processes to also generate traits, for
+example using a Brownian Motion process (BM; Cavalli-Sforza and Edwards
+(1967)) or an Ornstein Uhlenbeck (OU; Lande (1976); see Cooper et al.
+(2016) for a distinction between both). In \texttt{R}, this can be done
+with several already well used and well documented packages. For example
+if you want to simulate diversity through time, you can use
+\texttt{TreeSim} (Stadler 2011) to simulate diversity under a set of
+specific parameters (e.g.~speciation and extinction) with some events
+disrupting the simulations (e.g.~mass extinctions). One can even improve
+on generating these patterns using \texttt{FossilSim} (Barido-Sottani et
+al. 2019) to generate a pattern that will take into account
+fossilization processes. Or use \texttt{paleobuddy} (Rosario Petrucci,
+Januario, and Quental 2022) or \texttt{paleotree} (Bapst 2012) to
+generate palaeontology specific data. On the other hand, if you need to
+simulate both diversity and traits through time, this can be done with
+specific parameters in \texttt{RPANDA} (Morlon et al. 2016) or in
+\texttt{PETER} (Puttick, Guillerme, and Wills 2020) where the traits are
+generated stochasticaly through time (given some process) during the
+birth-death process.
 
-%---------------------------------------------
-%
-%       INTRODUCTION
-%
-%---------------------------------------------
+Although the packages mentioned above are excellent and routinely used
+with fast and reliable algorithms and associated documentation, they are
+all designed for specific tasks and don't allow much modification beyond
+the input parameters designed by the authors. For example,
+\texttt{TreeSim} can simulate a birth-death tree with some extinction
+event but is not designed to simulated one with an extinction event that
+leads to the birth-death process to be not diversity dependent anymore
+(e.g.~simulating a release in selection pressure after the extinction
+process that leads to a different process dominating speciation). Or
+\texttt{PETER} is not designed to simulate a complex set of traits (say
+three correlated BM traits and two independent OU ones). This absence of
+modularity has hampered the use of complex and question tailored
+simulations evolutionary biology, although I acknowledge this was not
+the primary aim of the authors of the packages mentioned above. This has
+led workers to often develop their own tools to answer specific
+questions (e.g.~Puttick, Guillerme, and Wills (2020)). Therefore, we
+propose \texttt{treats} a modular \texttt{R} package to simulate both
+diversity and disparity through time.
 
-\section{Introduction}
+\hypertarget{description}{%
+\section{Description}\label{description}}
 
-% Simulations in biology
-Comparing biological patterns is one of the key ways to understand mechanisms in evolutionary biology.
-This lead to the development of phylogenetic comparative methods as key methodologically driven topic in ecology, evolution and palaeontology \citep{felsensteinPCM,pennell2013review}.
-As indicated in the name, phylogenetic comparative methods rely on comparing patterns in a phylogenetic context (i.e.
-descent with modification) to understand biological mechanisms or concepts \citep{harmon2019book}.
-These comparisons can be done between observed patterns under different conditions or against null, neutral or baseline models (see \citealt{bausman2018neutral} for distinctions or common misconceptions) suggesting different processes or mechanisms.
-% These types phylogenetic comparative methods analysis of traits through time are routinely done in palaeontology (disparity; \citealt{guillerme2020disparities}) or in ecology (functional diversity; \citealt{mammola2021concepts}).
-For example different traits distribution for species with different diets \citep{deepak2023diet},
-% different habitats https://onlinelibrary.wiley.com/doi/full/10.1111/jbi.12939
-or by comparing some observed pattern to one simulated under null or base conditions\citep{miller2022alternating}.
-In theory, workers can follow the research pipeline of thinking of a specific mechanism (e.g.
-mass extinction allowing the surviving species to acquire new morphologies), collect some data to test this mechanism (e.g.
-some traits of species across and extinction event) and then compare these patterns to one simulated under no specific conditions (e.g.
-a null model where the traits evolve randomly regardless of an extinction event)\citep{puttick2020complex}.
-For such an approach, we need statistical and softwares solutions to simulate trees and data to generate these specific null models.
-
-% What already exists
-In practice, these evolutionary simulations can be done relatively easily on computers using a birth-death process \cite{feller1939birthdeath,stadler2010birthdeath,diversitree}
-A birth-death process is a continuous time Markov process that had been routinely implemented in \texttt{R} \citep{R} to simulate biological phylogenies (e.g. \citealt{ape}, \citealt{diversitree}).
-These algorithm to generate phylogenetic trees can be coupled with other Markov processes to also generate traits, for example using a Brownian Motion process (BM; \citealt{cavalli1967BM}) or an Ornstein Uhlenbeck  (OU; \citealt{lande1976OU}; see \citealt{cooper2016cautionary} for a distinction between both).
-In \texttt{R}, this can be done with several already well used and well documented packages.
-For example if you want to simulate diversity through time, you can use \texttt{TreeSim} \citep{treesim} to simulate diversity under a set of specific parameters (e.g.
-speciation and extinction) with some events disrupting the simulations (e.g.
-mass extinctions).
-One can even improve on generating these patterns using \texttt{FossilSim} \citep{fossilsim} to generate a pattern that will take into account fossilization processes.
-Or use \texttt{paleobuddy} \citep{paleobuddy} or \texttt{paleotree} \citep{paleotree} to generate palaeontology specific data.
-On the other hand, if you need to simulate both diversity and traits through time, this can be done with specific parameters in \texttt{RPANDA} \citep{rpanda} or in \texttt{PETER} \citep{puttick2020complex} where the traits are generated stochasticaly through time (given some process) during the birth-death process.
-
-% The problems with the absence of modularity
-Although the packages mentioned above are excellent and routinely used with fast and reliable algorithms and associated documentation, they are all designed for specific tasks and don't allow much modification beyond the input parameters designed by the authors.
-For example, \texttt{TreeSim} can simulate a birth-death tree with some extinction event but is not designed to simulated one with an extinction event that leads to the birth-death process to be not diversity dependent anymore (e.g.
-simulating a release in selection pressure after the extinction process that leads to a different process dominating speciation).
-Or \texttt{PETER} is not designed to simulate a complex set of traits (say three correlated BM traits and two independent OU ones).
-This absence of modularity has hampered the use of complex and question tailored simulations evolutionary biology, although I acknowledge this was not the primary aim of the authors of the packages mentioned above.
-This has led workers to often develop their own tools to answer specific questions (e.g. \citealt{puttick2020complex}).
-Therefore, we propose \treats a modular \texttt{R} package to simulate both diversity and disparity through time.
-
-\section{Description}
-% Very briefly how does it work
-\treats is based on the eponymous \treats function that allows to simulate a phylogeny and some trait(s) simultaneously.
-The base birth-death algorithm ``grows'' a phylogenetic tree and generates traits for each node and tips in the following manner:
+\texttt{treats} is based on the eponymous \texttt{treats} function that
+allows to simulate a phylogeny and some trait(s) simultaneously. The
+base birth-death algorithm ``grows'' a phylogenetic tree and generates
+traits for each node and tips in the following manner:
 
 \begin{enumerate}
-    \item Generating branch length;
-    \item Selecting a lineage among the currently living ones;
-    \item Choosing whether that lineage goes extinct (becomes a tip) or speciates (becomes a node).
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  Generating branch length;
+\item
+  Selecting a lineage among the currently living ones;
+\item
+  Choosing whether that lineage goes extinct (becomes a tip) or
+  speciates (becomes a node).
 \end{enumerate}
 
-These four steps are repeated until the tree reaches the desired age or number of species.
-If traits are simulated during the process, a fourth step is added:
+These four steps are repeated until the tree reaches the desired age or
+number of species. If traits are simulated during the process, a fourth
+step is added:
 
-    4. Generating some trait(s) value(s) for the selected lineage (no either a tip or a node).
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\setcounter{enumi}{3}
+\tightlist
+\item
+  Generating some trait(s) value(s) for the selected lineage (no either
+  a tip or a node).
+\end{enumerate}
 
-In \treats, these three or four steps are implemented as modular functions that the user can easily change using an internal class of objects called \texttt{``modifiers''} or \texttt{``traits''} (Fig.
-\ref{Fig:workflow}).
-The simulation then outputs a tree (of class \texttt{``phylo''} and a associated table of traits - \texttt{``matrix''}) that can be visualised using the \texttt{plot.treats} function.
+In \texttt{treats}, these three or four steps are implemented as modular
+functions that the user can easily change using an internal class of
+objects called \texttt{"modifiers"} or \texttt{"traits"} (Fig.
+\ref{Fig:workflow}). The simulation then outputs a tree (of class
+\texttt{"phylo"} and a associated table of traits - \texttt{"matrix"})
+that can be visualised using the \texttt{plot.treats} function. A third
+class of object called \texttt{"events"} can be added to the simulations
+to modify it under certain conditions (e.g.~simulating a mass
+extinction). Each element in the algorithm described above can be
+modified by the user using the implemented functions
+\texttt{make.bd.params} to set the birth-death parameters,
+\texttt{make.traits} to set the trait(s), \texttt{make.modifiers} to set
+the birth-death algorithm and \texttt{make.events} to eventually add one
+or more events.
 
-\begin{figure}[!htbp]
+\begin{figure}
+\hypertarget{figure1}{%
 \centering
-   \includegraphics[width=1\textwidth]{../inst/gitbook/treats_structure.pdf} 
-\caption{\treats package workflow: @@@.}
-\label{Fig:workflow}
+\includegraphics{../inst/gitbook/treats_structure.png}
+\caption{Figure 1: \texttt{treats} package workflow: the \texttt{treats}
+algorithm generates a tree and traits using inbuilt \texttt{"traits"}
+objects that contain the instructions on how to generate the trait data
+(e.g.~which process? how many dimensions?); \texttt{"modifiers"} objects
+that contains instructions on how to ``grow'' the tree (e.g.~by linking
+speciation to trait values, or to the current number of species); and
+\texttt{"events"} objects that can modify the tree structure,
+\texttt{"modifiers"} or \texttt{"traits"} depending on specific
+conditions (e.g.~80\% of species with positive trait values go extinct
+after reaching a specific time).}\label{figure1}
+}
 \end{figure}
-% \texttt{"events"} are an additional category of modular objects discussed later.
 
-\subsection{Simulating diversity}
+\hypertarget{brief-applied-example}{%
+\subsection{Brief applied example}\label{brief-applied-example}}
 
-To simulate a birth-death tree, two arguments are essentially needed: the birth-death parameters (\texttt{bd.params}) the rate of speciation (birth or $\lambda$), the rate of extinction (death or $\mu$ - this parameter can be equal to 0) and \texttt{stop.rule}, a rule for when to stop the growth of the tree (e.g.
-when reaching 20 species).
+The modularity structure of the \texttt{treats} packages allows users to
+design complex simulation scenarios to fit their specific question and
+data. To illustrate this we will look whether it is possible to detect
+changes in disparity (i.e.~diversity of traits) in a subset of the data
+published from Beck and Lee (2014) implemented in Guillerme (2018). This
+dataset contains the ordinated traits for 50 mammalian species across
+the cretaceous-palaeogene extinction event (K-Pg, 66 Mya; Fig. 2).
 
-% \texttt{## Running a simulation with a speciation rate of 1, an extinction rate of 0.1 and stopping when it reaches 20 taxa}
-% \texttt{my_simulation <- treats(bd.params = list(speciation = 1, extinction = 0.1), stop.rule = list(max.taxa = 20))}
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# Loading the data and packages}
+\KeywordTok{library}\NormalTok{(treats)}
+\KeywordTok{library}\NormalTok{(dispRity)}
+\KeywordTok{data}\NormalTok{(BeckLee\_tree)}
+\KeywordTok{data}\NormalTok{(BeckLee\_mat99)}
 
-% By default, the three first steps are included in a default modifier that reproduces a simple birth-death process given the input birth and death parameters.
+\CommentTok{\#\# Creating the time slices}
+\NormalTok{time\_slices \textless{}{-}}\StringTok{ }\KeywordTok{chrono.subsets}\NormalTok{(BeckLee\_mat99, BeckLee\_tree,}
+                              \DataTypeTok{method =} \StringTok{"continuous"}\NormalTok{,}
+                              \DataTypeTok{model  =} \StringTok{"acctran"}\NormalTok{,}
+                              \DataTypeTok{time   =} \KeywordTok{seq}\NormalTok{(}\DataTypeTok{from =} \DecValTok{96}\NormalTok{, }\DataTypeTok{to =} \DecValTok{36}\NormalTok{, }\DataTypeTok{by =} \DecValTok{{-}10}\NormalTok{))}
 
-% \begin{enumerate}
-%     \item Waiting: drawing a random number from an exponential distribution with a species dependent rate of $n \times (\lambda + \mu)$, where $n$ is the number of species currently alive in the process.
-%     \item Selection: randomly choosing on of the $n$ taxa currently alive in the process;
-%     \item Speciation: drawing a random number between 0 and 1; if that number is smaller than the speciation relative to the turnover ($\lambda / \lambda+\mu$), the lineage goes extinct %TODO: check!!!
-%     ; else, the lineage speciates.
-% \end{enumerate}
+\CommentTok{\#\# Calculating disparity on the two first dimensions only}
+\NormalTok{observed\_disparity \textless{}{-}}\StringTok{ }\KeywordTok{dispRity}\NormalTok{(}\KeywordTok{boot.matrix}\NormalTok{(time\_slices),}
+                               \DataTypeTok{metric =} \KeywordTok{c}\NormalTok{(sum, variances),}
+                               \DataTypeTok{dimensions =} \KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{,}\DecValTok{2}\NormalTok{))}
+\end{Highlighting}
+\end{Shaded}
 
-% These are included by default in a \texttt{treats} process but can be also set manually by creating a \texttt{``modifiers''} object with no options:
+\begin{figure}
+\centering
+\includegraphics{treats_paper_files/figure-latex/prepdata-1.pdf}
+\caption{Figure 2: Observed phylogeny (A) and disparity through time (B)
+in the subset of Beck and Lee (2014)'s dataset. For this specific data,
+is it possible to detect potential changes of disparity due to a mass
+extinction event?}
+\end{figure}
 
-% \texttt{## Generating the default modifiers}
-% \texttt{my_modifiers <- make.modifiers()}
-% \texttt{## Running a simulation with a speciation rate of 1, an extinction rate of 0.1 and stopping when it reaches 20 taxa with the default modifiers}
-% \texttt{my_simulation <- treats(bd.params = list(speciation = 1, extinction = 0), stop.rule = list(max.taxa = 20), modifiers = my_modifiers)}
+Using this example dataset, one might be interested in testing whether
+the K-Pg extinction had an effect on disparity through time. But can
+such effect be detected in the first place? We can test this by
+simulating some datasets with similar properties as the observed data
+and measure changes in disparity in these simulated datasets.
 
-% The modularity of the packages works with the ability to code the modifiers manually rather than using default ones.
-% For example, to simulate the same process but hard coding the modifiers we can use the following
+\hypertarget{simulating-trees}{%
+\subsubsection{Simulating trees}\label{simulating-trees}}
 
-% \texttt{## }
+The first and simplest way is to simulate tree topologies that have
+similar properties than the observed one. To do so, we need to use some
+\textbf{speciation} parameter indicating the rate at which lineages
+speciate (\emph{aka} ``birth'' or ``\(\lambda\)'') and an
+\textbf{extinction} parameter indicating the rate at which they go
+extinct (\emph{aka} ``death'' or ``\(\mu\)''). These two parameters can
+be crudely calculated from a give tree by counting the number of
+speciations and extinctions through time (but see better methods,
+e.g.~\texttt{ape::birthdeath}; Lawton, May, and Raup (1995); Paradis and
+Schliep (2019)). We also need a stopping rule for when to stop the
+simulations (in our case when reaching 30 time units). This will produce
+a single random tree using the input parameters (Fig. 3). Note that I
+will not discuss the options here in great details. Much more
+information can be found in the
+\href{http://tguillerme.github.io/treats.html}{\texttt{treats} manual}.
 
-% \subsection{Simulating traits (disparity)}
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# Using the birth{-}death parameters from the observed tree}
+\NormalTok{my\_bd\_params \textless{}{-}}\StringTok{ }\KeywordTok{crude.bd.est}\NormalTok{(BeckLee\_tree)}
+\CommentTok{\#\# Setting the stopping rule (stop after 30 time units)}
+\NormalTok{stop\_rule \textless{}{-}}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{max.time =} \DecValTok{30}\NormalTok{)}
 
-% multivariate stuff \cite{adams2019multivarPCM}
+\CommentTok{\#\# Simulate the tree}
+\KeywordTok{set.seed}\NormalTok{(}\DecValTok{1}\NormalTok{)}
+\NormalTok{sim\_tree \textless{}{-}}\StringTok{ }\KeywordTok{treats}\NormalTok{(}\DataTypeTok{bd.params  =}\NormalTok{ my\_bd\_params,}
+                   \DataTypeTok{stop.rule  =}\NormalTok{ stop\_rule,}
+                   \DataTypeTok{null.error =} \DecValTok{100}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
 
-\subsection{Using modular and dynamic modifiers}
+\begin{figure}
+\centering
+\includegraphics{treats_paper_files/figure-latex/simtree-1.pdf}
+\caption{Figure 3: A randomly simulated tree with similar properties as
+the observed one. Note that the time here is expressed in arbitrary
+units.}
+\end{figure}
 
-% Dynamic ones will be for bd.params = list(speciation = runif)
+\hypertarget{simulating-trees-and-traits}{%
+\subsubsection{Simulating trees and
+traits}\label{simulating-trees-and-traits}}
 
-\subsection{Introducing timed or conditional events in the simulations}
+For our specific question, we will also need to simulate some traits
+associated with each node and tip. For simplicity we will simulate a two
+dimensional Brownian Motion trait. To do so, we can create a
+\texttt{"traits"} object with the function \texttt{make.traits}. This
+results in a 2 dimensional trait space for all the simulated species and
+their nodes (Fig. 4).
 
-\section{Discussion}
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# Creating a trait in 2 dimensions.}
+\NormalTok{my\_traits \textless{}{-}}\StringTok{ }\KeywordTok{make.traits}\NormalTok{(}\DataTypeTok{process =}\NormalTok{ BM.process, }\DataTypeTok{n =} \DecValTok{2}\NormalTok{)}
 
-\subsection{Modularity}
+\CommentTok{\#\# Simulate the tree and traits}
+\KeywordTok{set.seed}\NormalTok{(}\DecValTok{1}\NormalTok{)}
+\NormalTok{sim\_data \textless{}{-}}\StringTok{ }\KeywordTok{treats}\NormalTok{(}\DataTypeTok{traits     =}\NormalTok{ my\_traits,}
+                   \DataTypeTok{bd.params  =}\NormalTok{ my\_bd\_params,}
+                   \DataTypeTok{stop.rule  =}\NormalTok{ stop\_rule,}
+                   \DataTypeTok{null.error =} \DecValTok{100}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
 
-\subsection{\treats compared to other packages}
+\begin{figure}
+\centering
+\includegraphics{treats_paper_files/figure-latex/simtreetraits-1.pdf}
+\caption{Figure 4: A randomly simulated tree with a 2 dimensional random
+Brownian Motion trait. Orange, light blue and dark blue dots
+respectively represent nodes, fossils and living species.}
+\end{figure}
 
-\subsection{Further directions}
+\hypertarget{simulating-a-trees-and-traits-with-events}{%
+\subsubsection{Simulating a trees and traits with
+events}\label{simulating-a-trees-and-traits-with-events}}
 
-\section{Conclusion}
-The \treats is modular and nice to use.
+To answer our question, we also want to simulate an extinction event. To
+do so, we can create two different \texttt{"events"} object with the
+function \texttt{make.events}. The first one will simulate a random
+extinction event after reaching half of the simulations (15 time units)
+and then making three quarter (0.75) of the taxa go extinct. The second
+one will simulate a random extinction but based on trait values: after
+reaching time 20, all the species with positive trait values will go
+extinct. Both scenarios thus illustrate two different types of mass
+extinction (selective and random Puttick, Guillerme, and Wills (2020)).
 
-Extended manual following recommendations from \cite{cooper2016dark}
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# Creating a random mass extinction}
+\NormalTok{random\_extinction \textless{}{-}}\StringTok{ }\KeywordTok{make.events}\NormalTok{(}
+    \DataTypeTok{target       =} \StringTok{"taxa"}\NormalTok{,}
+    \DataTypeTok{condition    =} \KeywordTok{time.condition}\NormalTok{(}\DecValTok{15}\NormalTok{),}
+    \DataTypeTok{modification =} \KeywordTok{random.extinction}\NormalTok{(}\FloatTok{0.75}\NormalTok{))}
+\CommentTok{\#\# Creating an extinction that removes species with positive trait values}
+\NormalTok{positive\_extinction \textless{}{-}}\StringTok{ }\KeywordTok{make.events}\NormalTok{(}
+    \DataTypeTok{target =} \StringTok{"taxa"}\NormalTok{,}
+    \DataTypeTok{condition =} \KeywordTok{time.condition}\NormalTok{(}\DecValTok{15}\NormalTok{),}
+    \DataTypeTok{modification =} \KeywordTok{trait.extinction}\NormalTok{(}\DataTypeTok{x =} \DecValTok{0}\NormalTok{, }\DataTypeTok{condition =} \StringTok{\textasciigrave{}}\DataTypeTok{\textgreater{}=}\StringTok{\textasciigrave{}}\NormalTok{))}
 
-\section{Package location}
-The \treats package is available on the CRAN at \url{https://cran.r-project.org/web/packages/treats/index.html} or on GitHub at \url{https://github.com/TGuillerme/treats} with more associated information.
-All the versions of the package are archived on ZENODO with associated DOI \url{https://zenodo.org/@@@}.
+\KeywordTok{set.seed}\NormalTok{(}\DecValTok{2}\NormalTok{)}
+\CommentTok{\#\# Simulate the tree and traits with a random extinction event}
+\NormalTok{sim\_rand\_extinction \textless{}{-}}\StringTok{ }\KeywordTok{treats}\NormalTok{(}
+                   \DataTypeTok{traits     =}\NormalTok{ my\_traits,}
+                   \DataTypeTok{bd.params  =}\NormalTok{ my\_bd\_params,}
+                   \DataTypeTok{stop.rule  =}\NormalTok{ stop\_rule,}
+                   \DataTypeTok{events     =}\NormalTok{ random\_extinction,}
+                   \DataTypeTok{null.error =} \DecValTok{100}\NormalTok{,}
+                   \DataTypeTok{replicates =} \DecValTok{50}\NormalTok{)}
 
-\section{Acknowledgments}
-Thanks to Mark Puttick and Alex Slavenko for comments on the early stage of the development of this package.
-Thanks to @@@.
-This work was funded by UKRI-NERC Grant NE/T000139/1 and a Royal Society University Research Fellowship (URF R 180006 to GHT).
+\CommentTok{\#\# Simulate the tree and traits with a random extinction event}
+\NormalTok{sim\_trait\_extinction \textless{}{-}}\StringTok{ }\KeywordTok{treats}\NormalTok{(}
+                   \DataTypeTok{traits     =}\NormalTok{ my\_traits,}
+                   \DataTypeTok{bd.params  =}\NormalTok{ my\_bd\_params,}
+                   \DataTypeTok{stop.rule  =}\NormalTok{ stop\_rule,}
+                   \DataTypeTok{events     =}\NormalTok{ positive\_extinction,}
+                   \DataTypeTok{null.error =} \DecValTok{100}\NormalTok{,}
+                   \DataTypeTok{replicates =} \DecValTok{50}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
 
-\bibliographystyle{sysbio}
-\bibliography{References}
+\begin{figure}
+\centering
+\includegraphics{treats_paper_files/figure-latex/simtreetraitsevents-1.pdf}
+\caption{Figure 5: Two different mass extinction events: (A) 75\% of
+species go extinct; (B) all species with positive trait values go
+extinct. The red line marks the extinction event.}
+\end{figure}
+
+Once we have simulated a distribution of trees and traits with the two
+extinction scenarios, we can measure disparity as in Fig. 1 for all the
+simulated data and compare it to the observed disparity.
+
+\begin{figure}
+\centering
+\includegraphics{treats_paper_files/figure-latex/simdisparity-1.pdf}
+\caption{Figure 6: Comparison of the observed sum of variances (dashed
+line) to the simulated one (ligth and dark grey polygons and black line
+representing respectively the 95\% and 50\% confidence interval and the
+median value).}
+\end{figure}
+
+From these results we can draw some preliminary conclusions: the
+observed change in disparity is more likely due to a random mass
+extinction rather than a selective one. This is of course a crude way of
+testing this, a more rigorous approach is likely needed to answer the
+question (i.e.~more and better quality data, and more thorough methods,
+e.g.~using a rank envelope test Murrell (2018)).
+
+\hypertarget{major-functionalities}{%
+\subsection{Major functionalities}\label{major-functionalities}}
+
+The following sections provide an overview of the three main functions
+displayed in Figure 1 (\texttt{make.traits}, \texttt{make.modifiers} and
+\texttt{make.events}).
+
+\hypertarget{simulating-traits}{%
+\subsubsection{Simulating traits}\label{simulating-traits}}
+
+Traits are simulated via the \texttt{make.traits} function given one or
+more processes, number of dimensions per process and some starting
+value(s). Essentially, the generation of new trait values is based on a
+process (a \texttt{function}) modifying a trait value (\texttt{x0})
+relative to some branch length (\texttt{edge.length}). For example, a
+simple Brownian Motion process could be generated by the function
+\texttt{rnorm} where the it modifies the trait value \texttt{x0}
+relative to the branch length (\texttt{edge.length}). The longer the
+branch length, the more likely the new trait value will be different
+from \texttt{x0}:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# Functionalising the previous example of a Brownian Motion process}
+\NormalTok{my.bm.process \textless{}{-}}\StringTok{ }\ControlFlowTok{function}\NormalTok{(x0, edge.length) \{}
+    \KeywordTok{rnorm}\NormalTok{(}\DataTypeTok{n =} \DecValTok{1}\NormalTok{, }\DataTypeTok{mean =}\NormalTok{ x0, }\DataTypeTok{sd =}\NormalTok{ edge.length)}
+\NormalTok{\} }
+\CommentTok{\#\# Creating the traits object}
+\NormalTok{my.trait \textless{}{-}}\StringTok{ }\KeywordTok{make.traits}\NormalTok{(}\DataTypeTok{process =}\NormalTok{ my.bm.process)}
+\end{Highlighting}
+\end{Shaded}
+
+Of course biological data can be much more complex and often
+multivariate (Adams and Collyer 2019) requiring the users to develop
+more complex function to cater their specific needs. However, the
+\texttt{treats} package contains a list of pre-built processes that are
+``ready-to-use'':
+
+\begin{itemize}
+\tightlist
+\item
+  \texttt{BM.process} and \texttt{OU.process}: generalised Brownian
+  Motion and Ornstein-Uhlenbeck processes in any number of dimensions
+  (including possible correlation);
+\item
+  \texttt{multi.peak.process}: a generalised Ornstein-Uhlenbeck process
+  (uni- or multi-dimensional) which allows for multiple peaks (usually
+  called the ``alpha'' parameters; Lande (1976));
+\item
+  \texttt{repulsion.process}: a unidimensional process generating trait
+  values that don't overlap with previously generated trait values;
+\item
+  \texttt{discrete.process}: generates discrete trait values;
+\item
+  \texttt{no.process}: ignores the branch length (i.e.~a time
+  independent process);
+\end{itemize}
+
+Note that the \texttt{treats} package primary aim is to generate both a
+tree and some traits at the same time. However, it is possible to also
+just generate traits with a given topology. This is done through the
+function \texttt{map.traits} that intakes one or more trees and a
+\texttt{"traits"} object.
+
+\hypertarget{modifying-the-birth-death-process}{%
+\subsubsection{Modifying the birth-death
+process}\label{modifying-the-birth-death-process}}
+
+Modifying the birth-death process can be done in several ways. Most
+easily it is done through changing the stopping rules through the
+\texttt{stop.rule} argument (number of total taxa or living ones, or
+time of the simulation). Equally straightforward, one can modify the
+parameters of the birth-death process through the
+\texttt{make.bd.params} function: the speciation (\(\lambda\)) and the
+extinction (\(\mu\)) ones. These can be either fixed values (for
+constant speciation and extinction) or values drawn from distributions.
+
+It is also possible to directly modify how the birth-death algorithm
+works through \texttt{make.modifiers} by changing the three main
+components of the birth-death algorithm as described above. By default,
+the algorithm uses the following algorithms:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  \textbf{Generating branch length} by drawing a value from an
+  exponential distribution with the rate being function of the current
+  number of lineages scaled by the speciation and extinction parameters
+\end{enumerate}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# The default branch length generation}
+\KeywordTok{rexp}\NormalTok{(}\DecValTok{1}\NormalTok{, }\DataTypeTok{rate =}\NormalTok{ number\_of\_lineages }\OperatorTok{*}\StringTok{ }\NormalTok{(speciation }\OperatorTok{+}\StringTok{ }\NormalTok{extinction))}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\setcounter{enumi}{1}
+\tightlist
+\item
+  \textbf{Selecting a lineage} among the currently living ones by simply
+  sampling across the available (living) lineages:
+\end{enumerate}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# The default lineage selection}
+\KeywordTok{sample}\NormalTok{(number\_of\_lineages, }\DecValTok{1}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\setcounter{enumi}{2}
+\tightlist
+\item
+  C\textbf{hoosing whether that lineage goes extinct (becomes a tip) or
+  speciates (becomes a node)} by drawing a random number between 0 and 1
+  and comparing it to the ratio of speciation and turnover (speciation +
+  extinction). If the random number is smaller than the ratio of
+  speciation and turnover, the lineage speciates, else it goes extinct:
+\end{enumerate}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# The default speciation/extinction decider}
+\KeywordTok{runif}\NormalTok{(}\DecValTok{1}\NormalTok{) }\OperatorTok{\textless{}=}\StringTok{ }\NormalTok{(speciation }\OperatorTok{/}\StringTok{ }\NormalTok{(speciation }\OperatorTok{+}\StringTok{ }\NormalTok{extinction))}
+\end{Highlighting}
+\end{Shaded}
+
+The function \texttt{make.modifiers} allows to specifically change any
+of these components by providing a different function for each part of
+the algorithm. For example, we could modify the three functions above so
+that the branch length is not dependent on the number of lineages, the
+sampling is always the first species available and the extinction is
+drawn from a normal distribution. Note that these function need a
+specific syntax that is detailed in the
+\href{http://tguillerme.github.io/treats.html}{\texttt{treats} manual}.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# Lineage independent waiting:}
+\NormalTok{lineage.independent \textless{}{-}}\StringTok{ }\ControlFlowTok{function}\NormalTok{(bd.params,}
+                                \DataTypeTok{lineage =} \OtherTok{NULL}\NormalTok{,}
+                                \DataTypeTok{trait.values =} \OtherTok{NULL}\NormalTok{,}
+                                \DataTypeTok{modify.fun =} \OtherTok{NULL}\NormalTok{) \{}
+\NormalTok{    my\_rate \textless{}{-}}\StringTok{ }\NormalTok{bd.params}\OperatorTok{$}\NormalTok{speciation }\OperatorTok{+}\StringTok{ }\NormalTok{bd.params}\OperatorTok{$}\NormalTok{extinction}
+    \KeywordTok{return}\NormalTok{(}\KeywordTok{rexp}\NormalTok{(}\DecValTok{1}\NormalTok{, }\DataTypeTok{rate =}\NormalTok{ my\_rate))}
+\NormalTok{\}}
+\CommentTok{\#\# Selecting always the first species}
+\NormalTok{select.first \textless{}{-}}\StringTok{ }\ControlFlowTok{function}\NormalTok{(bd.params,}
+                         \DataTypeTok{lineage =} \OtherTok{NULL}\NormalTok{,}
+                         \DataTypeTok{trait.values =} \OtherTok{NULL}\NormalTok{,}
+                         \DataTypeTok{modify.fun =} \OtherTok{NULL}\NormalTok{) \{}
+    \KeywordTok{return}\NormalTok{(}\KeywordTok{as.integer}\NormalTok{(}\DecValTok{1}\NormalTok{))}
+\NormalTok{\}}
+\CommentTok{\#\# Random normal speciation}
+\NormalTok{normal.speciation \textless{}{-}}\StringTok{ }\ControlFlowTok{function}\NormalTok{(bd.params,}
+                              \DataTypeTok{lineage =} \OtherTok{NULL}\NormalTok{,}
+                              \DataTypeTok{trait.values =} \OtherTok{NULL}\NormalTok{,}
+                              \DataTypeTok{modify.fun =} \OtherTok{NULL}\NormalTok{) \{}
+\NormalTok{    my\_turnover \textless{}{-}}\StringTok{ }\NormalTok{bd.params}\OperatorTok{$}\NormalTok{speciation}\OperatorTok{/}
+\StringTok{                  }\NormalTok{(bd.params}\OperatorTok{$}\NormalTok{speciation }\OperatorTok{+}\StringTok{ }\NormalTok{bd.params}\OperatorTok{$}\NormalTok{extinction)}
+    \KeywordTok{return}\NormalTok{(}\KeywordTok{rnorm}\NormalTok{(}\DecValTok{1}\NormalTok{) }\OperatorTok{\textless{}=}\StringTok{ }\NormalTok{my\_turnover)}
+\NormalTok{\}}
+
+\CommentTok{\#\# Creating the modifier object}
+\NormalTok{modified.birth.death \textless{}{-}}\StringTok{ }\KeywordTok{make.modifiers}\NormalTok{(}\DataTypeTok{branch.length =}\NormalTok{ lineage.independent,}
+                                       \DataTypeTok{selection     =}\NormalTok{ select.first,}
+                                       \DataTypeTok{speciation    =}\NormalTok{ normal.speciation)}
+\end{Highlighting}
+\end{Shaded}
+
+\hypertarget{creating-events}{%
+\subsubsection{Creating events}\label{creating-events}}
+
+The final major argument to be passed to \texttt{treats} are the
+\texttt{"events"} objects generated through \texttt{make.events} where
+the following information needs to be specified:
+
+\begin{itemize}
+\tightlist
+\item
+  the \texttt{target} designating what the event should be applied to
+  (e.g.~\texttt{"taxa"} for modifying the number of species,
+  \texttt{"traits"} for modifying the traits, etc.)
+\item
+  the \texttt{condition} which is a function returning a logical value
+  of when to trigger the event (e.g.~when reaching a certain number of
+  taxa, after some specific time has ellapsed or if some specific trait
+  value is reached, etc.)
+\item
+  the \texttt{modification} which is a function that specifically
+  modifies an internal object in the \texttt{treats} algorithm.
+\end{itemize}
+
+For a more exhaustive list of events so you can refer to the
+\href{http://tguillerme.github.io/treats.html}{\texttt{treats} manual}
+with many different detailed examples. Briefly though, this is how the
+mass extinction events are designed in the example above.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# Creating an extinction that removes species with positive trait values}
+\NormalTok{positive\_extinction \textless{}{-}}\StringTok{ }\KeywordTok{make.events}\NormalTok{(}
+    \DataTypeTok{target       =} \StringTok{"taxa"}\NormalTok{,}
+    \DataTypeTok{condition    =} \KeywordTok{time.condition}\NormalTok{(}\DecValTok{15}\NormalTok{),}
+    \DataTypeTok{modification =} \KeywordTok{random.extinction}\NormalTok{(}\FloatTok{0.75}\NormalTok{))}
+\end{Highlighting}
+\end{Shaded}
+
+For this event, the target is the number of taxa in the simulations.
+This is indicated using the \texttt{target\ =\ "taxa"} argument. Then
+the event is triggered using the argument \texttt{time.condition(15)}
+and modifies the internal \texttt{lineage} object using the
+\texttt{trait.extinction(x\ =\ 0,\ condition\ =\ \textasciigrave{}\textgreater{}=\textasciigrave{})}
+function. \texttt{time.condition} and \texttt{trait.extinction} are both
+function factories which are not gonna be detailed here (see Wickham
+(2019)). Effectively these arguments can be passed directly as standard
+functions. For example, to trigger the event when reaching time 15 we
+can use the following function:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# Returns TRUE when reaching time 15}
+\NormalTok{reaching.time15 \textless{}{-}}\StringTok{ }\ControlFlowTok{function}\NormalTok{(bd.params, lineage, trait.values, time) \{}
+    \KeywordTok{return}\NormalTok{(time }\OperatorTok{\textgreater{}}\StringTok{ }\DecValTok{15}\NormalTok{)}
+\NormalTok{\}}
+\end{Highlighting}
+\end{Shaded}
+
+This will trigger the \texttt{modification} which is equivalent to the
+following function modifying the \texttt{lineage} internal list:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{removing.}\FloatTok{75.}\NormalTok{taxa \textless{}{-}}\StringTok{ }\ControlFlowTok{function}\NormalTok{(bd.params, lineage, trait.values) \{}
+    \CommentTok{\#\# Select a portion of the living species to go extinct}
+\NormalTok{    extinct \textless{}{-}}\StringTok{ }\KeywordTok{sample}\NormalTok{(lineage}\OperatorTok{$}\NormalTok{n, }\KeywordTok{round}\NormalTok{(lineage}\OperatorTok{$}\NormalTok{n }\OperatorTok{*}\StringTok{ }\FloatTok{0.75}\NormalTok{))}
+
+    \CommentTok{\#\# Update the lineage object}
+\NormalTok{    lineage}\OperatorTok{$}\NormalTok{livings \textless{}{-}}\StringTok{ }\NormalTok{lineage}\OperatorTok{$}\NormalTok{livings[}\OperatorTok{{-}}\NormalTok{extinct]}
+\NormalTok{    lineage}\OperatorTok{$}\NormalTok{n       \textless{}{-}}\StringTok{ }\NormalTok{lineage}\OperatorTok{$}\NormalTok{n }\OperatorTok{{-}}\StringTok{ }\KeywordTok{length}\NormalTok{(extinct)}
+    \KeywordTok{return}\NormalTok{(lineage)}
+\NormalTok{\}}
+\end{Highlighting}
+\end{Shaded}
+
+Hence, the extinction event described above is equivalent to the
+following one:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# Creating an extinction that removes species with positive trait values}
+\NormalTok{positive\_extinction \textless{}{-}}\StringTok{ }\KeywordTok{make.events}\NormalTok{(}
+    \DataTypeTok{target       =} \StringTok{"taxa"}\NormalTok{,}
+    \DataTypeTok{condition    =}\NormalTok{ reaching.time15,}
+    \DataTypeTok{modification =}\NormalTok{ removing.}\FloatTok{75.}\NormalTok{taxa)}
+\end{Highlighting}
+\end{Shaded}
+
+\hypertarget{visualising-results}{%
+\subsubsection{Visualising results}\label{visualising-results}}
+
+The \texttt{treats} package also comes with tools to visualise trees and
+traits together or separately. This can be done through the generic S3
+\texttt{plot} function (calling \texttt{plot.treats}) and allows to
+display up to three traits or two traits and time (in 3D) as displayed
+in Figures 4 and 5. These functions can also be used to visualise trees
+and traits together from non \texttt{"treats"} objects by using the
+\texttt{make.treats} function to transform them into \texttt{"treats"}
+objects.
+
+\hypertarget{additional-information}{%
+\section{Additional information}\label{additional-information}}
+
+\hypertarget{manuals-and-vignette}{%
+\subsection{Manuals and vignette}\label{manuals-and-vignette}}
+
+The \texttt{treats} package comes with internal documentation
+(e.g.~\texttt{?treats}) but also with a thorough and extended vignette
+in a gitbook format: the
+\href{http://tguillerme.github.io/treats.html}{\texttt{treats} manual}.
+This manual is designed so that it can be regularly updated and enhanced
+through the lifetime of the package facilitating the interface between
+methods development and usage (Cooper, Thomas, and FitzJohn 2016).
+
+\hypertarget{further-directions}{%
+\subsection{Further directions}\label{further-directions}}
+
+This paper describes the first version of the \texttt{treats} package.
+However, I intend, either personally or through collaborations to
+continuously develop this package. For example future planned versions
+will include: lineage specific simulations; abiotic events; time
+dependent simulations and; a better integration with the
+\texttt{dispRity} package. This will be done while keeping track change
+(through \texttt{NEWS} file), continuous integration and unit testing.
+Furthermore, I am planning to develop the GitHub page associated with
+the package to encourage users to share their simulation templates and
+build library from them to facilitate the work for future users.
+
+\hypertarget{repeatability-and-reproducibility}{%
+\subsection{Repeatability and
+reproducibility}\label{repeatability-and-reproducibility}}
+
+This paper is entirely reproducible from an Rmarkdown document available
+on \href{https://github.com/TGuillerme/treats/paper/}{GitHub}. The data
+used for the example above (Beck and Lee 2014) is available from the
+\texttt{dispRity} package (Guillerme 2018).
+
+\hypertarget{conclusion}{%
+\section{Conclusion}\label{conclusion}}
+
+The \texttt{treats} package modular architecture allows workers to
+develop their own specific biological simulation scenarios based on
+their own specific research question. The pipeline of the package
+through the different \texttt{"treats"} objects (\texttt{"traits"},
+\texttt{"modifiers"} and \texttt{"events"}) also allows workers to
+generate publication standard results through plotting but also with
+easily reproducible and reusable scripts.
+
+\hypertarget{package-location}{%
+\subsection{Package location}\label{package-location}}
+
+The \texttt{treats} package is available on
+\href{https://github.com/TGuillerme/treats}{GitHub} with more associated
+information. All the versions of the package are archived on ZENODO with
+associated
+\href{https://doi.org/10.5281/zenodo.7970384}{DOI:10.5281/zenodo.7970384}.
+
+\hypertarget{acknowledgments}{%
+\subsection{Acknowledgments}\label{acknowledgments}}
+
+Thanks to Mark Puttick for inspiring to develop the package through
+previous collaborations. Thanks to Andrew Beckerman, Ian Brennan,
+Gustavo Burin, Christopher Cooney, Natalie Cooper, Alex Cranston,
+Jasmine Hardie, Tom Lansley, Clement Prieul, Joe Rees, James Rule,
+Sophie Ryan and Gavin Thomas, for support and useful comments on late
+stages of the development of this package. This work was funded by
+UKRI-NERC Grant NE/T000139/1 and a Royal Society University Research
+Fellowship (URF R 180006 to Gavin Thomas).
+
+\hypertarget{references}{%
+\section*{References}\label{references}}
+\addcontentsline{toc}{section}{References}
+
+\hypertarget{refs}{}
+\begin{cslreferences}
+\leavevmode\hypertarget{ref-adams2019multivarPCM}{}%
+Adams, Dean C, and Michael L Collyer. 2019. ``Phylogenetic Comparative
+Methods and the Evolution of Multivariate Phenotypes.'' \emph{Annual
+Review of Ecology, Evolution, and Systematics} 50: 405--25.
+
+\leavevmode\hypertarget{ref-paleotree}{}%
+Bapst, David W. 2012. ``Paleotree: An R Package for Paleontological and
+Phylogenetic Analyses of Evolution.'' \emph{Methods in Ecology and
+Evolution} 3 (5): 803--7.
+
+\leavevmode\hypertarget{ref-fossilsim}{}%
+Barido-Sottani, Joëlle, Walker Pett, Joseph E O'Reilly, and Rachel CM
+Warnock. 2019. ``FossilSim: An R Package for Simulating Fossil
+Occurrence Data Under Mechanistic Models of Preservation and Recovery.''
+\emph{Methods in Ecology and Evolution} 10 (6): 835--40.
+
+\leavevmode\hypertarget{ref-bausman2018neutral}{}%
+Bausman, William C. 2018. ``Modeling: Neutral, Null, and Baseline.''
+\emph{Philosophy of Science} 85 (4): 594--616.
+\url{https://doi.org/10.1086/699021}.
+
+\leavevmode\hypertarget{ref-beck2014ancient}{}%
+Beck, Robin MD, and Michael SY Lee. 2014. ``Ancient Dates or Accelerated
+Rates? Morphological Clocks and the Antiquity of Placental Mammals.''
+\emph{Proceedings of the Royal Society B: Biological Sciences} 281
+(1793): 20141278.
+
+\leavevmode\hypertarget{ref-cavalli1967BM}{}%
+Cavalli-Sforza, Luigi L, and Anthony WF Edwards. 1967. ``Phylogenetic
+Analysis. Models and Estimation Procedures.'' \emph{American Journal of
+Human Genetics} 19 (3 Pt 1): 233.
+
+\leavevmode\hypertarget{ref-cooper2016dark}{}%
+Cooper, Natalie, Gavin H Thomas, and Richard G FitzJohn. 2016.
+``Shedding Light on the `Dark Side'of Phylogenetic Comparative
+Methods.'' \emph{Methods in Ecology and Evolution} 7 (6): 693--99.
+
+\leavevmode\hypertarget{ref-cooper2016cautionary}{}%
+Cooper, Natalie, Gavin H Thomas, Chris Venditti, Andrew Meade, and Rob P
+Freckleton. 2016. ``A Cautionary Note on the Use of Ornstein Uhlenbeck
+Models in Macroevolutionary Studies.'' \emph{Biological Journal of the
+Linnean Society} 118 (1): 64--77.
+
+\leavevmode\hypertarget{ref-deepak2023diet}{}%
+Deepak, V, David J Gower, and Natalie Cooper. 2023. ``Diet and Habit
+Explain Head-Shape Convergences in Natricine Snakes.'' \emph{Journal of
+Evolutionary Biology} 36 (2): 399--411.
+
+\leavevmode\hypertarget{ref-feller1939birthdeath}{}%
+Feller, Willy. 1939. ``Die Grundlagen Der Volterraschen Theorie Des
+Kampfes Ums Dasein in Wahrscheinlichkeitstheoretischer Behandlung.''
+\emph{Acta Biotheoretica} 5 (1): 11--40.
+
+\leavevmode\hypertarget{ref-felsensteinPCM}{}%
+Felsenstein, Joseph. 1985. ``Phylogenies and the Comparative Method.''
+\emph{The American Naturalist} 125 (1): 1--15.
+
+\leavevmode\hypertarget{ref-diversitree}{}%
+FitzJohn, Richard G. 2012. ``Diversitree: Comparative Phylogenetic
+Analyses of Diversification in R.'' \emph{Methods in Ecology and
+Evolution} 3 (6): 1084--92.
+
+\leavevmode\hypertarget{ref-dispRity}{}%
+Guillerme, Thomas. 2018. ``DispRity: A Modular R Package for Measuring
+Disparity.'' \emph{Methods in Ecology and Evolution} 9 (7): 1755--63.
+
+\leavevmode\hypertarget{ref-harmon2019book}{}%
+Harmon, Luke. 2019. ``Phylogenetic Comparative Methods: Learning from
+Trees.''
+
+\leavevmode\hypertarget{ref-lande1976OU}{}%
+Lande, Russell. 1976. ``Natural Selection and Random Genetic Drift in
+Phenotypic Evolution.'' \emph{Evolution}, 314--34.
+
+\leavevmode\hypertarget{ref-lawton1995extinction}{}%
+Lawton, John H, Robert McCredie May, and David M Raup. 1995.
+\emph{Extinction Rates}. Vol. 11. Oxford University Press Oxford.
+
+\leavevmode\hypertarget{ref-miller2022alternating}{}%
+Miller, Elizabeth Christina, Christopher M Martinez, Sarah T Friedman,
+Peter C Wainwright, Samantha A Price, and Luke Tornabene. 2022.
+``Alternating Regimes of Shallow and Deep-Sea Diversification Explain a
+Species-Richness Paradox in Marine Fishes.'' \emph{Proceedings of the
+National Academy of Sciences} 119 (43): e2123544119.
+
+\leavevmode\hypertarget{ref-rpanda}{}%
+Morlon, Hélène, Eric Lewitus, Fabien L Condamine, Marc Manceau, Julien
+Clavel, and Jonathan Drury. 2016. ``RPANDA: An R Package for
+Macroevolutionary Analyses on Phylogenetic Trees.'' \emph{Methods in
+Ecology and Evolution} 7 (5): 589--97.
+
+\leavevmode\hypertarget{ref-murrell2018global}{}%
+Murrell, David J. 2018. ``A Global Envelope Test to Detect Non-Random
+Bursts of Trait Evolution.'' \emph{Methods in Ecology and Evolution} 9
+(7): 1739--48.
+
+\leavevmode\hypertarget{ref-ape}{}%
+Paradis, Emmanuel, and Klaus Schliep. 2019. ``Ape 5.0: An Environment
+for Modern Phylogenetics and Evolutionary Analyses in R.''
+\emph{Bioinformatics} 35: 526--28.
+\url{https://doi.org/10.1093/bioinformatics/bty633}.
+
+\leavevmode\hypertarget{ref-pennell2013review}{}%
+Pennell, Matthew W, and Luke J Harmon. 2013. ``An Integrative View of
+Phylogenetic Comparative Methods: Connections to Population Genetics,
+Community Ecology, and Paleobiology.'' \emph{Annals of the New York
+Academy of Sciences} 1289 (1): 90--105.
+
+\leavevmode\hypertarget{ref-pinto2017geographical}{}%
+Pinto-Ledezma, Jesús N, Lorena Mendes Simon, José Alexandre F
+Diniz-Filho, and Fabricio Villalobos. 2017. ``The Geographical
+Diversification of Furnariides: The Role of Forest Versus Open Habitats
+in Driving Species Richness Gradients.'' \emph{Journal of Biogeography}
+44 (8): 1683--93.
+
+\leavevmode\hypertarget{ref-puttick2020complex}{}%
+Puttick, Mark N, Thomas Guillerme, and Matthew A Wills. 2020. ``The
+Complex Effects of Mass Extinctions on Morphological Disparity.''
+\emph{Evolution} 74 (10): 2207--20.
+
+\leavevmode\hypertarget{ref-R}{}%
+R Core Team. 2023. \emph{R: A Language and Environment for Statistical
+Computing}. Vienna, Austria: R Foundation for Statistical Computing.
+\url{https://www.R-project.org/}.
+
+\leavevmode\hypertarget{ref-paleobuddy}{}%
+Rosario Petrucci, Bruno do, Matheus Januario, and Tiago Quental. 2022.
+``Paleobuddy: An R Package for Flexible Simulations of Diversification
+and Fossil Sampling.'' \emph{Methods in Ecology and Evolution} 13 (12):
+2692--8.
+
+\leavevmode\hypertarget{ref-stadler2010birthdeath}{}%
+Stadler, Tanja. 2010. ``Sampling-Through-Time in Birth--Death Trees.''
+\emph{Journal of Theoretical Biology} 267 (3): 396--404.
+
+\leavevmode\hypertarget{ref-treesim}{}%
+---------. 2011. ``Simulating Trees with a Fixed Number of Extant
+Species.'' \emph{Systematic Biology} 60 (5): 676--84.
+
+\leavevmode\hypertarget{ref-wickham2019advanced}{}%
+Wickham, Hadley. 2019. \emph{Advanced R}. CRC press.
+\end{cslreferences}
 

### 2023-06-07 paper/CoverLetter.tex
Update CoverLetter.tex

@@
 % Simulations are important in biology
-% Studying multivariate data as an alternative to univariate data has become increasingly popular in the last decades in both ecology and evolution.
-This includes, for example, studying form and function (e.g.
-Diaz et al., 2016, Nature) or morphological diversity (i.e.
-disparity - e.g.
-Brusatte et al.
-2012, Science).
-These methods all use a multidimensional space (through ordination or other techniques) representing the combination of all possible observed traits and then measuring some aspect of the space (e.g.
-its hypervolume) for different groups, and comparing these to each other or to a null hypothesis.
+Simulating biological data is an important step for testing biological hypothesis.
+For example, one might be interested in comparing observed diversification patterns to simulated ones to understand the mechanisms leading to the observed patterns (e.g. Miller et al. 2022 PNAS).
+In general, one might be interested in simulated either some phylogenetic hypothesis (topology), some traits or both to generate biologically realistic datasets to be compared to the observed ones.
 
 % There are many tools but they are not modular
-% There are a multitude of ways of carrying out each of these steps, and numerous software implementations in several languages (e.g.
-in R: Claddis, geomorph, ade4, vegan, or in javascript: GINGKO).
-However, each of these implementations is particular to the package author’s defini- tion of the multidimensional space, metric and test.
-This leads researchers to use these toolkits by default, thus preventing them from using the best definitions of the space/metric/test for their particular question or dataset.
-At worst this leads to incorrect statistic claims, and, at best to time consuming code retro-engineering.
+Such simulation based approach can be routinely done in many different \texttt{R} packages, such as \texttt{ape} (Paradis \& Schliep 2019 Bioinformatics), \texttt{diversitree} (FitzJohn 2012 MEE), \texttt{TreeSim} (Stadler 2011 Systematic Biology) or \texttt{RPANDA} (Morlon et al.
+2016 MEE) to only mention a few.
+Each of this packages however was designed of a specific simulation type in mind with a restricted number of options reflecting the objectives of the authors (i.e.
+simulations where not designed at the core of these excellent packages).
+For example, it is possible to simulate traits in \texttt{RPANDA} but not jointly with phylogenies, \texttt{TreeSim} allows for the simualtions of tree topologies and mass extinction events but doesn't allow simulating traits, or \texttt{diversitree} allows to simulate trees and traits but not mass extinction events.
 
 % Here is a treats, a modular package for simulations in biology
-% In this application manuscript entitled “dispRity: an R package for measuring disparity”, I present a complete and flexible R package for running multidimensional analysis where each parameter can be easily defined by the user (the space, the metric and the test).
-This package provides a tidy and modular environment where users can use their own multidimensional space defined as a matrix and apply to it any metric and any test (provided as functions).
-This allows for highly modular multidimensional analysis and the tidy interface allows easily reproducible analysis.
-All functions have documentation and unit tests.
+Here I propose the \texttt{treats} package: ``TREes And Traits Simulations''.
+This package is based on a highly modular architecture allowing users to simulate both trees and traits with the inclusions of events (e.g. mass extinction).
+Briefly, it allows and helps users to design their own function (or modify the implemented functions) to design their very own specific simulation scenarios, for example, a birth-death tree with variable speciation and extinction parameters and two 3D dimensional traits (one BM and one OU with multiple beaks) where the speciations becomes trait dependent after reaching a certain number of taxa.
+The package comes with an extended documentation and unit tests for each function.
 Additionally, this package comes with an in depth Gitbook manual covering most functionalities that will be updated regularly following user requests.
-Finally, the package is likely to be well cited as it is already being used in at least five projects that I am aware of.
 
-We look forward to hearing from you soon,
+I look forward to hearing from you soon,
 
@@
 %Editors:
-%Laura Graham, University of Southampton, UK
+% Laura Graham, University of Southampton, UK
 % Timothée Poisot, Université de Montréal, Canada

### 2023-06-09 paper/treats_paper.Rmd
fixed grants
Update treats_paper.Rmd

@@
 
-<!-- The code checklist must be filled in and uploaded as a 'File for editors'.  -->
 $^1$ School of Biosciences, University of Sheffield, Sheffield, S10 2TN, United Kingdom. guillert@tcd.ie
@@
 Comparing biological patterns is one of the key ways to understand mechanisms in evolutionary biology.
-This lead to the development of phylogenetic comparative methods as key methodologically driven topic in ecology, evolution and palaeontology [@felsensteinPCM; @pennell2013review].
-As indicated in the name, phylogenetic comparative methods rely on comparing patterns in a phylogenetic context to understand biological mechanisms or concepts [@harmon2019book].
+This leads to the development of phylogenetic comparative methods as key methodologically driven topic in ecology, evolution and palaeontology [@felsensteinPCM; @pennell2013review].
+These methods rely on comparing patterns in a phylogenetic context to understand biological mechanisms or concepts [@harmon2019book].
 These comparisons can be done between observed patterns under different conditions or against null, neutral or baseline models (see @bausman2018neutral for distinctions) suggesting different processes or mechanisms.
-For example different traits distribution for species with different diets [@deepak2023diet], or habitats [@pinto2017geographical] or by comparing some observed pattern to one simulated under null or base conditions [@miller2022alternating].
-In theory, workers can follow the research pipeline of thinking of a specific mechanism (e.g.
-mass extinction allowing the surviving species to acquire new morphologies), collect some data to test this mechanism (e.g.
-some traits of species across and extinction event) and then compare these patterns to one simulated under no specific conditions (e.g.
+For example different traits distribution for species with different diets [@deepak2023diet] or habitats [@pinto2017geographical].
+Or by comparing some observed pattern to one simulated under null or base conditions [@miller2022alternating].
+In theory, workers can follow the following research pipeline: thinking of a specific mechanism (e.g.
+mass extinction allowing the surviving species to acquire new morphologies), collecting some data to test this mechanism (e.g.
+some traits of species across and extinction event) and then comparing these patterns to some simulated under no specific conditions (e.g.
 a null model where the traits evolve randomly regardless of an extinction event @puttick2020complex).
-For such an approach, we need statistical and softwares solutions to simulate trees and data to generate these specific null models.
+Workers might thus need to simulate a great diversity of evolutionary scenarios to test their specific question.
+To do so, we need statistical and software solutions to simulate trees and data to generate many specific null models.
 
 In practice, these evolutionary simulations can be done relatively easily on computers using a birth-death process [@feller1939birthdeath; @stadler2010birthdeath; @diversitree].
-A birth-death process is a continuous time Markov process that had been routinely implemented in `R` [@R] to simulate biological phylogenies (e.g. [@ape], [@diversitree]).
-These algorithm to generate phylogenetic trees can be coupled with other Markov processes to also generate traits, for example using a Brownian Motion process (BM; @cavalli1967BM) or an Ornstein Uhlenbeck  (OU; @lande1976OU; see @cooper2016cautionary for a distinction between both).
+A birth-death process is a continuous time Markov process that had been routinely implemented in `R` [@R] to simulate realistic phylogenies (e.g. [@ape], [@diversitree]).
+This general algorithm to generate phylogenetic trees can be coupled with other Markov processes to also generate traits, for example using a Brownian Motion process (BM; @cavalli1967BM) or an Ornstein Uhlenbeck  (OU; @lande1976OU; see @cooper2016cautionary for a distinction between both).
 In `R`, this can be done with several already well used and well documented packages.
@@
 mass extinctions).
-One can even improve on generating these patterns using `FossilSim` [@fossilsim] to generate a pattern that will take into account fossilization processes.
-Or use `paleobuddy` [@paleobuddy] or `paleotree` [@paleotree] to generate palaeontology specific data.
-On the other hand, if you need to simulate both diversity and traits through time, this can be done with specific parameters in `RPANDA` [@rpanda] or in `PETER` [@puttick2020complex] where the traits are generated stochasticaly through time (given some process) during the birth-death process.
+You can even improve on generating these patterns using `FossilSim` [@fossilsim] to take into account fossilization processes.
+You can also use `paleobuddy` [@paleobuddy] or `paleotree` [@paleotree] to generate palaeontology specific data.
+On the other hand, if you need to simulate both diversity and traits through time, this can be done with specific parameters in `RPANDA` [@rpanda], `diversitree` [@diversitree] or `PETER` [@puttick2020complex] where the traits are generated stochasticaly through time (given some process) during the birth-death process.
 
 Although the packages mentioned above are excellent and routinely used with fast and reliable algorithms and associated documentation, they are all designed for specific tasks and don't allow much modification beyond the input parameters designed by the authors.
-For example, `TreeSim` can simulate a birth-death tree with some extinction event but is not designed to simulated one with an extinction event that leads to the birth-death process to be not diversity dependent anymore (e.g.
-simulating a release in selection pressure after the extinction process that leads to a different process dominating speciation).
+For example, `TreeSim` can simulate a birth-death tree with some extinction event but is not designed to simulated one with an extinction event that leads to the birth-death process to be not diversity dependent anymore, simulating a release in selection pressure after the extinction event that leads to a different process dominating speciation.
 Or `PETER` is not designed to simulate a complex set of traits (say three correlated BM traits and two independent OU ones).
-This absence of modularity has hampered the use of complex and question tailored simulations evolutionary biology, although I acknowledge this was not the primary aim of the authors of the packages mentioned above.
+This absence of modularity has hampered the use of complex and question-driven simulations, although I acknowledge this was not the primary aim of the authors of the excellent packages mentioned above.
 This has led workers to often develop their own tools to answer specific questions (e.g. @puttick2020complex).
-Therefore, we propose `treats` a modular `R` package to simulate both diversity and disparity through time.
+Therefore, I propose `treats` a modular `R` package to simulate both trees and traits through time.
 
@@
 
-These four steps are repeated until the tree reaches the desired age or number of species.
+These three steps are repeated until the tree reaches the desired age or number of species (the algorithm's implementation is heavily inspired and based on @diversitree).
 If traits are simulated during the process, a fourth step is added:
@@
 The simulation then outputs a tree (of class `"phylo"` and a associated table of traits - `"matrix"`) that can be visualised using the `plot.treats` function.
-A third class of object called `"events"` can be added to the simulations to modify it under certain conditions (e.g. simulating a mass extinction).
-Each element in the algorithm described above can be modified by the user using the implemented functions `make.bd.params` to set the birth-death parameters, `make.traits` to set the trait(s), `make.modifiers` to set the birth-death algorithm and `make.events` to eventually add one or more events.
+A third class of object called `"events"` can be added to the simulations to modify the entire simulation under certain conditions (e.g. simulating a mass extinction).
+Each element in the algorithm can be modified by the user using the implemented functions `make.bd.params` to set the birth-death parameters, `make.traits` to set the trait(s), `make.modifiers` to set the birth-death algorithm and `make.events` to eventually add one or more events.
+Users can replace any step in the algorithm by their own specific functions suiting their needs or, for less advanced users, by using already implemented functions.
 
@@
 
-The modularity structure of the `treats` packages allows users to design complex simulation scenarios to fit their specific question and data.
 To illustrate this we will look whether it is possible to detect changes in disparity (i.e. diversity of traits) in a subset of the data published from @beck2014ancient implemented in @dispRity.
@@
 ```{r prepdata, echo = FALSE, fig.cap="Figure 2: Observed phylogeny (A) and disparity through time (B) in the subset of @beck2014ancient's dataset.
-For this specific data, is it possible to detect potential changes of disparity due to a mass extinction event?", fig.width = 18, fig.height = 8}
+Is this change in disparity related to the K-Pg mass extinction?
+Or, more generally, is it even possible to detect potential changes of disparity due to a mass extinction event?", fig.width = 18, fig.height = 8}
 ## Plotting the tree and the disparity through time
@@
 To do so, we need to use some **speciation** parameter indicating the rate at which lineages speciate (_aka_ "birth" or "$\lambda$") and an **extinction** parameter indicating the rate at which they go extinct (_aka_ "death" or "$\mu$").
-These two parameters can be crudely calculated from a give tree by counting the number of speciations and extinctions through time (but see better methods, e.g.
+These two parameters can be crudely calculated from a given tree by counting the number of speciations and extinctions through time (but see better methods, e.g.
 `ape::birthdeath`; @lawton1995extinction; @ape).
@@
 The first one will simulate a random extinction event after reaching half of the simulations (15 time units) and then making three quarter (0.75) of the taxa go extinct.
-The second one will simulate a random extinction but based on trait values: after reaching time 20, all the species with positive trait values will go extinct.
-Both scenarios thus illustrate two different types of mass extinction (selective and random @puttick2020complex).
+The second one will simulate a random extinction but based on trait values: after reaching time 15, all the species with positive trait values will go extinct.
+Both scenarios thus illustrate two different types of mass extinction (random and selective; @puttick2020complex).
 
@@
 
-Once we have simulated a distribution of trees and traits with the two extinction scenarios, we can measure disparity as in Fig. 1 for all the simulated data and compare it to the observed disparity.
+Once we have simulated a distribution of trees and traits with the two extinction scenarios, we can measure disparity as in Fig. 2 for all the simulated data and compare it to the observed disparity.
 
@@
 From these results we can draw some preliminary conclusions: the observed change in disparity is more likely due to a random mass extinction rather than a selective one.
-This is of course a crude way of testing this, a more rigorous approach is likely needed to answer the question (i.e.
-more and better quality data, and more thorough methods, e.g.
+This is of course a very crude way of testing this, a more rigorous approach is needed to answer the question: more and better quality data, and more thorough methods (e.g.
 using a rank envelope test @murrell2018global).
@@
 
-The following sections provide an overview of the three main functions displayed in Figure 1 (`make.traits`, `make.modifiers` and `make.events`).
+The following sections provides an overview of the three main functions displayed in Figure 1 (`make.traits`, `make.modifiers` and `make.events`).
 
@@
 ```{r, eval = FALSE}
-## Functionalising the previous example of a Brownian Motion process
+## Brownian Motion process
 my.bm.process <- function(x0, edge.length) {
@@
 Of course biological data can be much more complex and often multivariate [@adams2019multivarPCM] requiring the users to develop more complex function to cater their specific needs.
-However, the `treats` package contains a list of pre-built processes that are "ready-to-use":
+The `treats` package contains a list of pre-built processes that are "ready-to-use":
 
  * `BM.process` and `OU.process`: generalised Brownian Motion and Ornstein-Uhlenbeck processes in any number of dimensions (including possible correlation);
- * `multi.peak.process`: a generalised Ornstein-Uhlenbeck process (uni- or multi-dimensional) which allows for multiple peaks (usually called the "alpha" parameters; @lande1976OU);
+ * `multi.peak.process`: a generalised Ornstein-Uhlenbeck process (uni- or multi-dimensional) which allows for multiple peaks;
  * `repulsion.process`: a unidimensional process generating trait values that don't overlap with previously generated trait values;
@@
 3.
-C**hoosing whether that lineage goes extinct (becomes a tip) or speciates (becomes a node)** by drawing a random number between 0 and 1 and comparing it to the ratio of speciation and turnover (speciation + extinction).
+**Choosing whether that lineage goes extinct (becomes a tip) or speciates (becomes a node)** by drawing a random number between 0 and 1 and comparing it to the ratio of speciation and turnover (speciation + extinction).
 If the random number is smaller than the ratio of speciation and turnover, the lineage speciates, else it goes extinct:
@@
 The function `make.modifiers` allows to specifically change any of these components by providing a different function for each part of the algorithm.
-For example, we could modify the three functions above so that the branch length is not dependent on the number of lineages, the sampling is always the first species available and the extinction is drawn from a normal distribution.
+For example, one can modify the three functions above so that the branch length is not dependent on the number of lineages, the sampling is always the first species available and the extinction is drawn from a normal distribution.
 Note that these function need a specific syntax that is detailed in the [`treats` manual](http://tguillerme.github.io/treats.html).
@@
 This paper describes the first version of the `treats` package.
-However, I intend, either personally or through collaborations to continuously develop this package.
+However, I intend to continuously develop this package.
 For example future planned versions will include: lineage specific simulations; abiotic events; time dependent simulations and; a better integration with the `dispRity` package.
@@
 ## Acknowledgments
-Thanks to Mark Puttick for inspiring to develop the package through previous collaborations.
+Many thanks to Mark Puttick for inspiring the development the package through previous collaborations.
 Thanks to Andrew Beckerman, Ian Brennan, Gustavo Burin, Christopher Cooney, Natalie Cooper, Alex Cranston, Jasmine Hardie, Tom Lansley, Clement Prieul, Joe Rees, James Rule, Sophie Ryan and Gavin Thomas, for support and useful comments on late stages of the development of this package.
-This work was funded by UKRI-NERC Grant NE/T000139/1 and a Royal Society University Research Fellowship (URF R 180006 to Gavin Thomas).
+This work was funded by UKRI-NERC Grant NE/T000139/1.
 

### 2023-06-19 paper/treats_paper.tex
fixed grants
computer reboot update

@@
 \author{Thomas Guillerme\(^1\)}
-\date{2023-06-07}
+\date{2023-06-09}
 
@@
 Comparing biological patterns is one of the key ways to understand
-mechanisms in evolutionary biology. This lead to the development of
+mechanisms in evolutionary biology. This leads to the development of
 phylogenetic comparative methods as key methodologically driven topic in
 ecology, evolution and palaeontology (Felsenstein 1985; Pennell and
-Harmon 2013). As indicated in the name, phylogenetic comparative methods
-rely on comparing patterns in a phylogenetic context to understand
-biological mechanisms or concepts (Harmon 2019). These comparisons can
-be done between observed patterns under different conditions or against
-null, neutral or baseline models (see Bausman (2018) for distinctions)
-suggesting different processes or mechanisms. For example different
-traits distribution for species with different diets (Deepak, Gower, and
-Cooper 2023), or habitats (Pinto-Ledezma et al. 2017) or by comparing
-some observed pattern to one simulated under null or base conditions
-(Miller et al. 2022). In theory, workers can follow the research
-pipeline of thinking of a specific mechanism (e.g.~mass extinction
-allowing the surviving species to acquire new morphologies), collect
-some data to test this mechanism (e.g.~some traits of species across and
-extinction event) and then compare these patterns to one simulated under
-no specific conditions (e.g.~a null model where the traits evolve
-randomly regardless of an extinction event Puttick, Guillerme, and Wills
-(2020)). For such an approach, we need statistical and softwares
-solutions to simulate trees and data to generate these specific null
-models.
+Harmon 2013). These methods rely on comparing patterns in a phylogenetic
+context to understand biological mechanisms or concepts (Harmon 2019).
+These comparisons can be done between observed patterns under different
+conditions or against null, neutral or baseline models (see Bausman
+(2018) for distinctions) suggesting different processes or mechanisms.
+For example different traits distribution for species with different
+diets (Deepak, Gower, and Cooper 2023) or habitats (Pinto-Ledezma et al.
+2017). Or by comparing some observed pattern to one simulated under null
+or base conditions (Miller et al. 2022). In theory, workers can follow
+the following research pipeline: thinking of a specific mechanism
+(e.g.~mass extinction allowing the surviving species to acquire new
+morphologies), collecting some data to test this mechanism (e.g.~some
+traits of species across and extinction event) and then comparing these
+patterns to some simulated under no specific conditions (e.g.~a null
+model where the traits evolve randomly regardless of an extinction event
+Puttick, Guillerme, and Wills (2020)). Workers might thus need to
+simulate a great diversity of evolutionary scenarios to test their
+specific question. To do so, we need statistical and software solutions
+to simulate trees and data to generate many specific null models.
 
@@
 process that had been routinely implemented in \texttt{R} (R Core Team
-2023) to simulate biological phylogenies (e.g.~(Paradis and Schliep
-2019), (FitzJohn 2012)). These algorithm to generate phylogenetic trees
-can be coupled with other Markov processes to also generate traits, for
-example using a Brownian Motion process (BM; Cavalli-Sforza and Edwards
-(1967)) or an Ornstein Uhlenbeck (OU; Lande (1976); see Cooper et al.
-(2016) for a distinction between both). In \texttt{R}, this can be done
-with several already well used and well documented packages. For example
-if you want to simulate diversity through time, you can use
-\texttt{TreeSim} (Stadler 2011) to simulate diversity under a set of
-specific parameters (e.g.~speciation and extinction) with some events
-disrupting the simulations (e.g.~mass extinctions). One can even improve
-on generating these patterns using \texttt{FossilSim} (Barido-Sottani et
-al. 2019) to generate a pattern that will take into account
-fossilization processes. Or use \texttt{paleobuddy} (Rosario Petrucci,
+2023) to simulate realistic phylogenies (e.g.~(Paradis and Schliep
+2019), (FitzJohn 2012)). This general algorithm to generate phylogenetic
+trees can be coupled with other Markov processes to also generate
+traits, for example using a Brownian Motion process (BM; Cavalli-Sforza
+and Edwards (1967)) or an Ornstein Uhlenbeck (OU; Lande (1976); see
+Cooper et al. (2016) for a distinction between both). In \texttt{R},
+this can be done with several already well used and well documented
+packages. For example if you want to simulate diversity through time,
+you can use \texttt{TreeSim} (Stadler 2011) to simulate diversity under
+a set of specific parameters (e.g.~speciation and extinction) with some
+events disrupting the simulations (e.g.~mass extinctions). You can even
+improve on generating these patterns using \texttt{FossilSim}
+(Barido-Sottani et al. 2019) to take into account fossilization
+processes. You can also use \texttt{paleobuddy} (Rosario Petrucci,
 Januario, and Quental 2022) or \texttt{paleotree} (Bapst 2012) to
@@
 simulate both diversity and traits through time, this can be done with
-specific parameters in \texttt{RPANDA} (Morlon et al. 2016) or in
-\texttt{PETER} (Puttick, Guillerme, and Wills 2020) where the traits are
-generated stochasticaly through time (given some process) during the
-birth-death process.
+specific parameters in \texttt{RPANDA} (Morlon et al. 2016),
+\texttt{diversitree} (FitzJohn 2012) or \texttt{PETER} (Puttick,
+Guillerme, and Wills 2020) where the traits are generated stochasticaly
+through time (given some process) during the birth-death process.
 
@@
 event but is not designed to simulated one with an extinction event that
-leads to the birth-death process to be not diversity dependent anymore
-(e.g.~simulating a release in selection pressure after the extinction
-process that leads to a different process dominating speciation). Or
+leads to the birth-death process to be not diversity dependent anymore,
+simulating a release in selection pressure after the extinction event
+that leads to a different process dominating speciation. Or
 \texttt{PETER} is not designed to simulate a complex set of traits (say
 three correlated BM traits and two independent OU ones). This absence of
-modularity has hampered the use of complex and question tailored
-simulations evolutionary biology, although I acknowledge this was not
-the primary aim of the authors of the packages mentioned above. This has
-led workers to often develop their own tools to answer specific
-questions (e.g.~Puttick, Guillerme, and Wills (2020)). Therefore, we
-propose \texttt{treats} a modular \texttt{R} package to simulate both
-diversity and disparity through time.
+modularity has hampered the use of complex and question-driven
+simulations, although I acknowledge this was not the primary aim of the
+authors of the excellent packages mentioned above. This has led workers
+to often develop their own tools to answer specific questions
+(e.g.~Puttick, Guillerme, and Wills (2020)). Therefore, I propose
+\texttt{treats} a modular \texttt{R} package to simulate both trees and
+traits through time.
 
@@
 
-These four steps are repeated until the tree reaches the desired age or
-number of species. If traits are simulated during the process, a fourth
-step is added:
+These three steps are repeated until the tree reaches the desired age or
+number of species (the algorithm's implementation is heavily inspired
+and based on FitzJohn (2012)). If traits are simulated during the
+process, a fourth step is added:
 
@@
 class of object called \texttt{"events"} can be added to the simulations
-to modify it under certain conditions (e.g.~simulating a mass
-extinction). Each element in the algorithm described above can be
-modified by the user using the implemented functions
+to modify the entire simulation under certain conditions
+(e.g.~simulating a mass extinction). Each element in the algorithm can
+be modified by the user using the implemented functions
 \texttt{make.bd.params} to set the birth-death parameters,
@@
 the birth-death algorithm and \texttt{make.events} to eventually add one
-or more events.
+or more events. Users can replace any step in the algorithm by their own
+specific functions suiting their needs or, for less advanced users, by
+using already implemented functions.
 
@@
 
-The modularity structure of the \texttt{treats} packages allows users to
-design complex simulation scenarios to fit their specific question and
-data. To illustrate this we will look whether it is possible to detect
-changes in disparity (i.e.~diversity of traits) in a subset of the data
+To illustrate this we will look whether it is possible to detect changes
+in disparity (i.e.~diversity of traits) in a subset of the data
 published from Beck and Lee (2014) implemented in Guillerme (2018). This
@@
 \caption{Figure 2: Observed phylogeny (A) and disparity through time (B)
-in the subset of Beck and Lee (2014)'s dataset. For this specific data,
-is it possible to detect potential changes of disparity due to a mass
+in the subset of Beck and Lee (2014)'s dataset. Is this change in
+disparity related to the K-Pg mass extinction? Or, more generally, is it
+even possible to detect potential changes of disparity due to a mass
 extinction event?}
@@
 extinct (\emph{aka} ``death'' or ``\(\mu\)''). These two parameters can
-be crudely calculated from a give tree by counting the number of
+be crudely calculated from a given tree by counting the number of
 speciations and extinctions through time (but see better methods,
@@
 one will simulate a random extinction but based on trait values: after
-reaching time 20, all the species with positive trait values will go
+reaching time 15, all the species with positive trait values will go
 extinct. Both scenarios thus illustrate two different types of mass
-extinction (selective and random Puttick, Guillerme, and Wills (2020)).
+extinction (random and selective; Puttick, Guillerme, and Wills (2020)).
 
@@
 Once we have simulated a distribution of trees and traits with the two
-extinction scenarios, we can measure disparity as in Fig. 1 for all the
+extinction scenarios, we can measure disparity as in Fig. 2 for all the
 simulated data and compare it to the observed disparity.
@@
 observed change in disparity is more likely due to a random mass
-extinction rather than a selective one. This is of course a crude way of
-testing this, a more rigorous approach is likely needed to answer the
-question (i.e.~more and better quality data, and more thorough methods,
-e.g.~using a rank envelope test Murrell (2018)).
+extinction rather than a selective one. This is of course a very crude
+way of testing this, a more rigorous approach is needed to answer the
+question: more and better quality data, and more thorough methods
+(e.g.~using a rank envelope test Murrell (2018)).
 
@@
 
-The following sections provide an overview of the three main functions
+The following sections provides an overview of the three main functions
 displayed in Figure 1 (\texttt{make.traits}, \texttt{make.modifiers} and
@@
 \begin{Highlighting}[]
-\CommentTok{\#\# Functionalising the previous example of a Brownian Motion process}
+\CommentTok{\#\# Brownian Motion process}
 \NormalTok{my.bm.process \textless{}{-}}\StringTok{ }\ControlFlowTok{function}\NormalTok{(x0, edge.length) \{}
@@
 multivariate (Adams and Collyer 2019) requiring the users to develop
-more complex function to cater their specific needs. However, the
-\texttt{treats} package contains a list of pre-built processes that are
+more complex function to cater their specific needs. The \texttt{treats}
+package contains a list of pre-built processes that are
 ``ready-to-use'':
@@
   \texttt{multi.peak.process}: a generalised Ornstein-Uhlenbeck process
-  (uni- or multi-dimensional) which allows for multiple peaks (usually
-  called the ``alpha'' parameters; Lande (1976));
+  (uni- or multi-dimensional) which allows for multiple peaks;
 \item
@@
 \item
-  C\textbf{hoosing whether that lineage goes extinct (becomes a tip) or
+  \textbf{Choosing whether that lineage goes extinct (becomes a tip) or
   speciates (becomes a node)} by drawing a random number between 0 and 1
@@
 of these components by providing a different function for each part of
-the algorithm. For example, we could modify the three functions above so
+the algorithm. For example, one can modify the three functions above so
 that the branch length is not dependent on the number of lineages, the
@@
 This paper describes the first version of the \texttt{treats} package.
-However, I intend, either personally or through collaborations to
-continuously develop this package. For example future planned versions
-will include: lineage specific simulations; abiotic events; time
-dependent simulations and; a better integration with the
-\texttt{dispRity} package. This will be done while keeping track change
-(through \texttt{NEWS} file), continuous integration and unit testing.
-Furthermore, I am planning to develop the GitHub page associated with
-the package to encourage users to share their simulation templates and
-build library from them to facilitate the work for future users.
+However, I intend to continuously develop this package. For example
+future planned versions will include: lineage specific simulations;
+abiotic events; time dependent simulations and; a better integration
+with the \texttt{dispRity} package. This will be done while keeping
+track change (through \texttt{NEWS} file), continuous integration and
+unit testing. Furthermore, I am planning to develop the GitHub page
+associated with the package to encourage users to share their simulation
+templates and build library from them to facilitate the work for future
+users.
 
@@
 
-Thanks to Mark Puttick for inspiring to develop the package through
-previous collaborations. Thanks to Andrew Beckerman, Ian Brennan,
-Gustavo Burin, Christopher Cooney, Natalie Cooper, Alex Cranston,
-Jasmine Hardie, Tom Lansley, Clement Prieul, Joe Rees, James Rule,
-Sophie Ryan and Gavin Thomas, for support and useful comments on late
-stages of the development of this package. This work was funded by
-UKRI-NERC Grant NE/T000139/1 and a Royal Society University Research
-Fellowship (URF R 180006 to Gavin Thomas).
+Many thanks to Mark Puttick for inspiring the development the package
+through previous collaborations. Thanks to Andrew Beckerman, Ian
+Brennan, Gustavo Burin, Christopher Cooney, Natalie Cooper, Alex
+Cranston, Jasmine Hardie, Tom Lansley, Clement Prieul, Joe Rees, James
+Rule, Sophie Ryan and Gavin Thomas, for support and useful comments on
+late stages of the development of this package. This work was funded by
+UKRI-NERC Grant NE/T000139/1.
 

### 2023-07-07 inst/gitbook/03_make.modifiers.Rmd
Update 03_make.modifiers.Rmd

@@
 
-For example in the `speciation` part of a modifier, the default is to trigger an event from a uniform distribution and then check if that value is smaller than ($$speciation/(speciation + extinction)$$):
+For example in the `speciation` part of a modifier, the default is to trigger an event from a uniform distribution and then check if that value is smaller than ($speciation/(speciation + extinction)$):
 

### 2023-08-11 paper/response.Rmd
response INIT

---
title: "Response to reviewers"
author: "Thomas Guillerme"
bibliography: References.bib
# csl: mee.csl
date: "`r Sys.Date()`"
output: 
  pdf_document: 
    keep_tex: yes
    fig_width: 8
    fig_height: 8
  bookdown::html_document2: default
  bookdown::pdf_document2: default
---

Dear Editor,

Please find my response to the reviewers comments in the following document.
I've highlight <span style="color:blue">the Associate Editor and both reviewers comments in blue</span> and my response black.

Best regards,
Thomas

# Associate Editor Comments to Author:

<span style="color:blue">As both reviewers, I really liked the idea of the package, but also found the manuscript itself a bit hard to follow, which, at least to me, made it hard to fully appreciate what “treats” can currently do and what are the future additions that might be easily implemented (either by the developer, or by users). Reading the online manual, I got a better idea, and hence I think there is still room (metaphorically and I suspect also physically) for improvement. Sure, the online material has no space limit constraints, but I do think the manuscript text could be improved and both reviewers offered some nice suggestions.  Apart from the really good suggestions given by both reviewers I would add the following suggestions/comments:</span>

* <span style="color:blue">1- Are the trait values “stored” only at the node and tip (extant or extinct), or does it store (e.g. assuming some small-time intervals) the trait values throughout the branch length of each lineage? Based on lines 63 to 65 (and 71) I think only at nodes and tips, but it might be worth explicitly saying that. Perhaps add the word “only” in that line.  Also, in line 71 is there an extra “no” within the parenthesis?</span>

* <span style="color:blue">2- Regarding the “storage” of trait value, if it only stores the values at nodes and tips, does it also mean that it only uses trait values at those time points to direct the simulation? Does that matter? What made me wonder about that was figure 5, and its respective intended selective extinction simulation scenario. If I got it right, the selective extinction scenario would remove all species with positive trait values at the mass extinction event at 15 Mya. Looking at the right panel, all species whose direct ancestral node had a positive trait value seemed to be removed, but some of those with ancestral negative values likely crossed the red line having a positive value (even if the “true trait trajectory” is not a straight as represented, for simplicity I guess, in the figure) and were not removed. This might not be always relevant, and I understand that some simplifications are necessary, but it might be worth explicitly discussing if this is relevant and when. For example, if speciation and background extinction rates are small, species are likely to have long branches and the trait value it might have at a given time point between its ancestral node and its tip value or next speciation node might be very different from the value it had at the ancestral node (or tip value for that matter). Or is the decision to remove the species in the selective mass extinction event one where both the ancestral and tip value have to be positive. I guess some extra information on how it chooses the species based on the trait value (is it the ancestral node value, the tip value, both?) in the selective mass extinction regime would also be helpful. I guess the issue seems that the trait value is not measured at short time intervals (only at the nodes and tips) but a selective extinction regime such as this one presented here is a point in time event. </span>

* <span style="color:blue">3- I have to say that looking at figure 5 it was not trivial to see the difference between the two scenarios (I guess that happened because most, if not all, species with positive values went extinct in the left random panel as well). I know those are realizations of a stochastic process and there is nothing wrong here, my concern regards simply the visualization. I wonder if choosing another seed might result in a set of simulations that are more clearly different, making the comparison easier for the eye to pick it up. For example, in the online manual, some examples comparing random to selective simulations showed a much more disparate pattern.</span>

* <span style="color:blue">4- Code box on page 7: the commentary on the second simulation scenario reads “#Simulate the tree and traits with a random extinction event” and hence is equal to the comment in the first simulation scenario in the same box. Perhaps change it to something like: “#Simulate the tree and traits with a random extinction event but based on trait value”</span>

* <span style="color:blue">5- I could be wrong but it seems that figure 6 is not cited in the text. Please double check and if this is true, please cite this figure. </span>

* <span style="color:blue">6- All figure captions have some repetition like “Figure 1: Figure 1:….”</span>

* <span style="color:blue">7- Figure 6: I could be wrong, but my intuition was that a selective mass extinction event would decrease the disparity more strongly than a random extinction mass extinction event. That does not seem to be the case. Is it because the “rule” for the second extinction scenario was that only positive species will go extinct, hence on average 50% of the species (if you start with a value of zero) instead of 75% went extinct? That would make sense, I guess. Or is the second extinction scenario one where you also create a 75% loss of species but guarantee that all species with positive values will go extinct? Maybe a few words to better explain this would be helpful. Perhaps add on lines 109 to 110 something like: “note that in the second scenario, because the ancestral state is zero and we do not set the percentage of species to be removed, we expect only 50% of the species to go extinct”. Just an idea.</span>

# Reviewer: 1

<span style="color:blue">This article by Guillerme introduces a new R package treats, which implements a flexible framework for simulating phylogenies and associated trait values. Overall, I really like the concept of the package, and I think it can be very useful for more advanced users who run many simulations and change their setup often. My comments are mostly about clarifying the presentation of the different functions. I also reviewed the manual.</span>

## Comments on the manuscript

* <span style="color:blue">I had trouble understanding what treats could and could not do from the manuscript alone. The introduction discusses several packages related to fossil data (FossilSim, paleotree and paleobuddy) which gave me the impression that treats could replace them, but it doesn't seem that fossilization processes are currently included ? I was also a bit confused by the inclusion of time-dependent and lineage-dependent simulations in future features, as it seems that the current framework should be able to handle these, at least within the birth-death process. Overall, having a clearer list of what is currently possible or not in treats would be helpful.</span>

* <span style="color:blue">I would also have liked some information on the performance of the package compared to existing simulation tools. It seems likely that the flexibility offered by treats comes at the cost of some computational performance, so it would be nice to know how much time would be needed to simulate a reasonable dataset (for instance the 50 replicates used to test the positive extinction vs random extinction).</span>

* <span style="color:blue">I didn't understand why the example shown in the manuscript only simulated trees up to 30 time units, when the tree used as a basis for the comparison is 140 My old (figure 2).</span>

* <span style="color:blue">I really like the idea of having a library of premade simulation templates (l202) and I believe it will be a huge help in making the package accessible. Making the examples shown in the manual (section 8) available as scripts would be a good start for such a library.</span>

### Minor comments:

* <span style="color:blue"> - l71 "no either" </span>

* <span style="color:blue"> - l125 "where the it"</span>

* <span style="color:blue"> - l111 the comment in the second example is mismatched with the code (random extinction instead of positive extinction)</span>

* <span style="color:blue"> - l175 the comment in the example is mismatched with the code (positive extinction instead of random extinction)</span>

* <span style="color:blue"> - l178 the text mentions the function trait.extinction but it's not used in the code</span>

* <span style="color:blue"> - l282 missing authors in the reference</span>

## Comments on the manual

* <span style="color:blue">The main thing the manual is missing in my opinion is a quick reference table of the different functions, for people who have understood the concepts of the package but need to quickly check what format their custom speciation function needs to follow for instance. At the moment that information is incomplete (the table in section 4.5 shows the output types but not what these outputs are) and spread out in different places (the description of the function inputs is in section 4.2.1 whereas the list of functions is in 4.1).</span>

* <span style="color:blue">The use of conditions and modifications was quite confusing to me. Some of the examples use the modify and condition argument to change the birth-death process (section 2.3) whereas some examples simply replace the function with a modified version (section 4.4), and I could not figure out if these methods were equivalent or if one should be preferred to the other. I think the modifications section also needs more details about what exactly is being modified, and what the expected inputs and outputs of the modification function should be. For instance, in the example section 2.3 I don't understand what is the "x" argument in "going.extinct", or why this function returns a number instead of a logical. The interface for the conditions in 5.2 appears to be clearer, but it doesn't match with the example in section 2.3 where the "negative.ancestor" uses different arguments, or with the example in 4.6 where the "half.the.time" condition takes no arguments. </span>

* <span style="color:blue">Finally, I also could not find a full description of the treats output object either in the manual or in the package documentation.</span>

# Reviewer: 2

<span style="color:blue">Thank you for the opportunity to review this manuscript, titled “treats: a modular R package for simulating trees and traits”. This manuscript describes and showcases a new R package that provides users with an extensive toolbox to jointly simulate phylogenies and evolving traits. While the example provided within the manuscript gives readers a small taste of the flexibility of this toolbox, the online manual (http://tguillerme.github.io/treats.html) showcases the true extent of how useful this package will be for evolutionary biologists far and wide. I really enjoyed reading the manuscript, and I believe it will be suitable for publication in Methods in Ecology & Evolution once my concerns have been addressed.</span>

<span style="color:blue">For context, for this review I read the supplied manuscript, tested the functionality of the R package, and browsed the built-in documentation and the online supplemental manual. I was unable to review the functional code of the R package in time for the due date of this review, but it appears to work properly from my testing. I have one major concern and a handful of minor concerns. The major concern is that the overall flow of the manuscript seems a little off. I think this mostly stems from the worked example preceding the detailed overview of the package’s functions. I think if these two sections were swapped the manuscript would flow much better. My minor concerns are included at the end of this review in order of their appearance in the manuscript. I look forward to reviewing a revised version of this manuscript, if necessary.</span>

<span style="color:blue">Best,</span>

<span style="color:blue">William Gearty</span>

* <span style="color:blue">Lines 27-31: I would enumerate this list.</span>

* <span style="color:blue">Line 84: Geological time period names should be capitalized.</span>

* <span style="color:blue">Figures 2-6: If possible, I think it would be valuable to show what code was used to produce these figures in the code snippets. As far as I can tell, the plotting is mostly done using functions from the package, so this would also be an opportunity to showcase this functionality.</span>

* <span style="color:blue">Figure 2: The caption should identify what the red line represents (the K-Pg boundary?).</span>

* <span style="color:blue">Line 111: This may be personal preference, but when I ran this code chunk it output a whole bunch of text to the console. It appears to be a separate line of text for each replicate in the treats function. I believe the intent here is to show the user how much progress has been made? Perhaps some sort of progress bar would be more appropriate?</span>

* <span style="color:blue">Lines 106-111: I think you should mention somewhere in this paragraph why you are running multiple replicates here (as opposed to the single replicate performed earlier in the worked example).</span>

* <span style="color:blue">Line 111: I used the same seed as you, but I was unable to reproduce Figure 5 (looking at the R markdown file, I see that you use indices 11 and 10 to make the plots). I think showing the exact code that is used here to make Figure 5 would be useful for readers that might be “playing along at home”.</span>

* <span style="color:blue">Lines 112-113: You should refer to Figure 6 somewhere in this paragraph. It might also be useful to include the code to perform this comparison, especially if the functionality is all contained within the treats package.</span>

* <span style="color:blue">Line 194: This reference to the manual should have a link to the online manual (looks like the link is embedded, but you should also have it typed out). I also noticed that the documentation for the treats function lists the wrong URL for the online manual.</span>

* <span style="color:blue">Lines 202-203: Just wanted to note that I really love this idea! I think collecting a library of modules will be very useful to users!</span>

* <span style="color:blue">Line 205: This embedded link does not work. I think it should be something like: https://github.com/TGuillerme/treats/tree/master/paper.</span>

### 2023-08-11 paper/response.Rmd
easy fixes response

@@
 
+I am grateful for the associate editor and the two reviewers time and very encouraging comments that I totally believe have improved the package, the manual and this manuscript.
 Please find my response to the reviewers comments in the following document.
@@
 
+> TODO:
+
 * <span style="color:blue">2- Regarding the “storage” of trait value, if it only stores the values at nodes and tips, does it also mean that it only uses trait values at those time points to direct the simulation?
@@
 
+> TODO:
+
 * <span style="color:blue">3- I have to say that looking at figure 5 it was not trivial to see the difference between the two scenarios (I guess that happened because most, if not all, species with positive values went extinct in the left random panel as well).
@@
 
+> TODO:
+
 * <span style="color:blue">4- Code box on page 7: the commentary on the second simulation scenario reads “#Simulate the tree and traits with a random extinction event” and hence is equal to the comment in the first simulation scenario in the same box.
@@
 
+> TODO:
+
 * <span style="color:blue">5- I could be wrong but it seems that figure 6 is not cited in the text. Please double check and if this is true, please cite this figure. </span>
 
+> TODO:
+
 * <span style="color:blue">6- All figure captions have some repetition like “Figure 1: Figure 1:….”</span>
+
+> TODO:
 
@@
 Just an idea.</span>
+
+> TODO:
 
@@
 
+> TODO:
+
 * <span style="color:blue">I would also have liked some information on the performance of the package compared to existing simulation tools.
@@
 
+> TODO:
+
 * <span style="color:blue">I didn't understand why the example shown in the manuscript only simulated trees up to 30 time units, when the tree used as a basis for the comparison is 140 My old (figure 2).</span>
+
+> TODO:
 
@@
 
+> TODO:
+
 ### Minor comments:
@@
 * <span style="color:blue"> - l71 "no either" </span>
+Fixed. [Commit ]()
 
 * <span style="color:blue"> - l125 "where the it"</span>
+Fixed. [Commit ]()
 
 * <span style="color:blue"> - l111 the comment in the second example is mismatched with the code (random extinction instead of positive extinction)</span>
+Fixed. [Commit ]()
 
 * <span style="color:blue"> - l175 the comment in the example is mismatched with the code (positive extinction instead of random extinction)</span>
+Fixed. [Commit ]()
 
 * <span style="color:blue"> - l178 the text mentions the function trait.extinction but it's not used in the code</span>
+Fixed. [Commit ]()
 
 * <span style="color:blue"> - l282 missing authors in the reference</span>
+This one is just generated as such from the `mee.csl` reference style.
 
@@
 At the moment that information is incomplete (the table in section 4.5 shows the output types but not what these outputs are) and spread out in different places (the description of the function inputs is in section 4.2.1 whereas the list of functions is in 4.1).</span>
+
+> TODO:
 
@@
 
+> TODO:
+
 * <span style="color:blue">Finally, I also could not find a full description of the treats output object either in the manual or in the package documentation.</span>
+
+> TODO:
 
@@
 * <span style="color:blue">Lines 27-31: I would enumerate this list.</span>
+Done. [Commit ]()
 
 * <span style="color:blue">Line 84: Geological time period names should be capitalized.</span>
+Fixed. [Commit ]()
 
@@
 
+> TODO:
+
 * <span style="color:blue">Figure 2: The caption should identify what the red line represents (the K-Pg boundary?).</span>
+Done. [Commit ]()
 
@@
 
+> TODO:
+
 * <span style="color:blue">Lines 106-111: I think you should mention somewhere in this paragraph why you are running multiple replicates here (as opposed to the single replicate performed earlier in the worked example).</span>
+I've added the following sentence at the end of the paragraph:
+"*Because of the stochasticity of the simulations, we will repeat them 50 times (using `replicates = 50`) to generate a distribution of possible simulated scenarios as opposed to a random single one that could be idiosyncratic.*" [Commit ]()
 
@@
 
+> TODO:
+
 * <span style="color:blue">Lines 112-113: You should refer to Figure 6 somewhere in this paragraph.
@@
 
+> TODO:
+
 * <span style="color:blue">Line 194: This reference to the manual should have a link to the online manual (looks like the link is embedded, but you should also have it typed out).
 I also noticed that the documentation for the treats function lists the wrong URL for the online manual.</span>
+I've added the URL spelt out. [Commit ]().
+
+> TODO: fix wrong URLs.
 
@@
 
+> TODO
+
 * <span style="color:blue">Line 205: This embedded link does not work. I think it should be something like: https://github.com/TGuillerme/treats/tree/master/paper.</span>
+
+> TODO: change to `inst/` in the manuscript

### 2023-08-11 paper/treats_paper.Rmd
easy fixes response

@@
 Or by comparing some observed pattern to one simulated under null or base conditions [@miller2022alternating].
-In theory, workers can follow the following research pipeline: thinking of a specific mechanism (e.g.
-mass extinction allowing the surviving species to acquire new morphologies), collecting some data to test this mechanism (e.g.
-some traits of species across and extinction event) and then comparing these patterns to some simulated under no specific conditions (e.g.
+In theory, workers can follow the following research pipeline: 1) thinking of a specific mechanism (e.g.
+mass extinction allowing the surviving species to acquire new morphologies), 2) collecting some data to test this mechanism (e.g.
+some traits of species across and extinction event) and then 3) comparing these patterns to some simulated under no specific conditions (e.g.
 a null model where the traits evolve randomly regardless of an extinction event @puttick2020complex).
@@
 
-4. Generating some trait(s) value(s) for the selected lineage (no either a tip or a node).
+4. Generating some trait(s) value(s) for the selected lineage (either a tip or a node).
 
@@
 To illustrate this we will look whether it is possible to detect changes in disparity (i.e. diversity of traits) in a subset of the data published from @beck2014ancient implemented in @dispRity.
-This dataset contains the ordinated traits for 50 mammalian species across the cretaceous-palaeogene extinction event (K-Pg, 66 Mya; Fig. 2).
+This dataset contains the ordinated traits for 50 mammalian species across the Cretaceous-Palaeogene extinction event (K-Pg, 66 Mya; Fig. 2).
 
@@
 ```{r prepdata, echo = FALSE, fig.cap="Figure 2: Observed phylogeny (A) and disparity through time (B) in the subset of @beck2014ancient's dataset.
+The red line represents the K-Pg boundary.
 Is this change in disparity related to the K-Pg mass extinction?
@@
 Both scenarios thus illustrate two different types of mass extinction (random and selective; @puttick2020complex).
+Because of the stochasticity of the simulations, we will repeat them 50 times (using `replicates = 50`) to generate a distribution of possible simulated scenarios as opposed to a random single one that could be idiosyncratic.
 
@@
 
-## Simulate the tree and traits with a random extinction event
+## Simulate the tree and traits with a selective extinction event
 sim_trait_extinction <- treats(
@@
 Essentially, the generation of new trait values is based on a process (a `function`) modifying a trait value (`x0`) relative to some branch length (`edge.length`).
-For example, a simple Brownian Motion process could be generated by the function `rnorm` where the it modifies the trait value `x0` relative to the branch length (`edge.length`).
+For example, a simple Brownian Motion process could be generated by the function `rnorm` where it modifies the trait value `x0` relative to the branch length (`edge.length`).
 The longer the branch length, the more likely the new trait value will be different from `x0`:
@@
     condition    = time.condition(15),
-    modification = random.extinction(0.75))
+    modification = trait.extinction(x = 0, condition = `>=`))
 ```
@@
 
-## Manuals and vignette
+## Manuals and vignettes
 
 The `treats` package comes with internal documentation (e.g.
-`?treats`) but also with a thorough and extended vignette in a gitbook format: the [`treats` manual](http://tguillerme.github.io/treats.html).
+`?treats`) but also with a thorough and extended vignette in a gitbook format: the [`treats` manual](http://tguillerme.github.io/treats.html) ([http://tguillerme.github.io/treats.html](http://tguillerme.github.io/treats.html).
 This manual is designed so that it can be regularly updated and enhanced through the lifetime of the package facilitating the interface between methods development and usage [@cooper2016dark].

### 2023-08-11 paper/response.Rmd
expanded TODOs

@@
 
-> TODO:
+> TODO: specify or trait data is stored: either on tips and bifurcating nodes only (`save.steps = FALSE`) or on tips, bifurcating nodes and singleton nodes (nodes with just one ancestor and one descendant).
+Add precision and change the "no".
 
@@
 
-> TODO:
+> TODO: implement trait extinction with singletons.
+I.e.
+if the event is trait related, simulate a round of trait just after triggering the event.
+Save them as singletons and allow for the option `drop.singletons` to clean the tree (if needed).
 
@@
 
-> TODO:
+> TODO: Find a better seed for the extinction example.
 
@@
 
-> TODO:
+> TODO: Fix caption.
 
@@
 
-> TODO:
+> TODO: Cite figure 6
 
@@
 
-> TODO:
+> TODO: Fix repetitions.
 
@@
 
-> TODO:
+> TODO: Rerun simulation with improved algorithm (using traits events). Also then refer to Puttick about bad predictability of non-random extinctions
 
@@
 
-> TODO:
+> TODO: make a table of what it can cannot do. Also update the future features and specify what I meant by time/lineage-dependent.
 
@@
 
-> TODO:
+> TODO: make a table comparing the package to other ones in terms of computing times.
 
@@
 
-> TODO:
+> TODO: change the simulation to 140 Mya? Or just say it's an arbitrary number all along?
 
@@
 
-> TODO:
+> TODO: implement that in issues on the GitHub page with a link to all the possibilities
 
@@
 * <span style="color:blue"> - l71 "no either" </span>
-Fixed. [Commit ]()
+Fixed. [Commit 5f7fb23](https://github.com/TGuillerme/treats/commit/5f7fb235e0b56a097b9d888790028ad41f86c33b)
 
 * <span style="color:blue"> - l125 "where the it"</span>
-Fixed. [Commit ]()
+Fixed. [Commit 5f7fb23](https://github.com/TGuillerme/treats/commit/5f7fb235e0b56a097b9d888790028ad41f86c33b)
 
 * <span style="color:blue"> - l111 the comment in the second example is mismatched with the code (random extinction instead of positive extinction)</span>
-Fixed. [Commit ]()
+Fixed. [Commit 5f7fb23](https://github.com/TGuillerme/treats/commit/5f7fb235e0b56a097b9d888790028ad41f86c33b)
 
 * <span style="color:blue"> - l175 the comment in the example is mismatched with the code (positive extinction instead of random extinction)</span>
-Fixed. [Commit ]()
+Fixed. [Commit 5f7fb23](https://github.com/TGuillerme/treats/commit/5f7fb235e0b56a097b9d888790028ad41f86c33b)
 
 * <span style="color:blue"> - l178 the text mentions the function trait.extinction but it's not used in the code</span>
-Fixed. [Commit ]()
+Fixed. [Commit 5f7fb23](https://github.com/TGuillerme/treats/commit/5f7fb235e0b56a097b9d888790028ad41f86c33b)
 
@@
 
-> TODO:
+> TODO: make a cheat sheet page with all core functions, their arguments and what they do.
 
@@
 
-> TODO:
+> TODO: specify this in the manual and the documentation: `modification` = modify the bd-algorith; `events` (conditions and stuff) = modify the flow of the algorithm (but can also introduce a `modification`!)
 
@@
 
-> TODO:
+> TODO: add in the documentation a brief description of the outputs of the `make.*` functions and of `treats`.
+Additionally, add a detailed `.md` with the developer level content of each output (`dispRity` style).
 
@@
 
+> TODO: swap the example and the description section in the manuscript.
+
 <span style="color:blue">Best,</span>
@@
 * <span style="color:blue">Lines 27-31: I would enumerate this list.</span>
-Done. [Commit ]()
+Done. [Commit 5f7fb23](https://github.com/TGuillerme/treats/commit/5f7fb235e0b56a097b9d888790028ad41f86c33b)
 
 * <span style="color:blue">Line 84: Geological time period names should be capitalized.</span>
-Fixed. [Commit ]()
+Fixed. [Commit 5f7fb23](https://github.com/TGuillerme/treats/commit/5f7fb235e0b56a097b9d888790028ad41f86c33b)
 
@@
 
-> TODO:
+> TODO: show and streamline the code.
 
 * <span style="color:blue">Figure 2: The caption should identify what the red line represents (the K-Pg boundary?).</span>
-Done. [Commit ]()
+Done. [Commit 5f7fb23](https://github.com/TGuillerme/treats/commit/5f7fb235e0b56a097b9d888790028ad41f86c33b)
 
@@
 
-> TODO:
+> TODO: streamline the outputs.
 
@@
 I've added the following sentence at the end of the paragraph:
-"*Because of the stochasticity of the simulations, we will repeat them 50 times (using `replicates = 50`) to generate a distribution of possible simulated scenarios as opposed to a random single one that could be idiosyncratic.*" [Commit ]()
+"*Because of the stochasticity of the simulations, we will repeat them 50 times (using `replicates = 50`) to generate a distribution of possible simulated scenarios as opposed to a random single one that could be idiosyncratic.*" [Commit 5f7fb23](https://github.com/TGuillerme/treats/commit/5f7fb235e0b56a097b9d888790028ad41f86c33b)
 
@@
 
-> TODO:
+> TODO: fix this.
 
@@
 I also noticed that the documentation for the treats function lists the wrong URL for the online manual.</span>
-I've added the URL spelt out. [Commit ]().
+I've added the URL spelt out. [Commit 5f7fb23](https://github.com/TGuillerme/treats/commit/5f7fb235e0b56a097b9d888790028ad41f86c33b).
 
@@
 
-> TODO
+> TODO: implement this as issues.
 

### 2023-09-26 paper/response.Rmd
add specific for storing trait values

@@
 Also, in line 71 is there an extra “no” within the parenthesis?</span>
+
+The trait data in `treats` is only stored at nodes and tips due to the code architecture.
+However, using the argument `save.steps` allows to create singleton nodes (nodes with one ancestor and only one descendant) that can be generated at some specified or regular time intervals thus effectively saving trait values as the simulation goes.
+I've now specified this in the main text (and removed the typo):
+
+*4. Generating some trait(s) value(s) for the selected lineage (either a tip or a node - but see the "Simulating traits section" below to generate trait values along edges).* line @@@
+
+and:
+
+*"Traits are always generated (and stored) only for tips or nodes and not along edges.
+However, it is possible so generate trait values at specific or regular time intervals by using the option `save.steps` in the `treats` function.
+This will generate singleton nodes (i.e.
+nodes with only one descendant) with associated trait values."* line @@@ [Commit @@@](https://github.com/TGuillerme/treats/commit/)
 
@@
 * <span style="color:blue">Lines 106-111: I think you should mention somewhere in this paragraph why you are running multiple replicates here (as opposed to the single replicate performed earlier in the worked example).</span>
+
 I've added the following sentence at the end of the paragraph:
@@
 
-> TODO: fix wrong URLs.
-
 * <span style="color:blue">Lines 202-203: Just wanted to note that I really love this idea! I think collecting a library of modules will be very useful to users!</span>

### 2023-09-26 paper/treats_paper.Rmd
add specific for storing trait values

@@
 
-4. Generating some trait(s) value(s) for the selected lineage (either a tip or a node).
+4. Generating some trait(s) value(s) for the selected lineage (either a tip or a node - but see the "Simulating traits section" below to generate trait values along edges).
 
@@
 
+Traits are always generated (and stored) only for tips or nodes and not along edges.
+However, it is possible so generate trait values at specific or regular time intervals by using the option `save.steps` in the `treats` function.
+This will generate singleton nodes (i.e. nodes with only one descendant) with associated trait values.
+
 Note that the `treats` package primary aim is to generate both a tree and some traits at the same time.

### 2023-09-26 paper/treats_paper.Rmd
Temp fix.
 Need to streamline bugs in last test chunk

@@
 
-set.seed(2)
+set.seed(1)
 ## Simulate the tree and traits with a random extinction event

### 2023-09-26 R/birth.death_fun.R

@@ -88,10 +88,10 @@ birth.death.tree.traits <- function(stop.rule, bd.params, traits = NULL, modifie
     ############
     ## Initialising
     #############
-    # warning("DEBUG in birth.death_fun.R::birth.death.tree.traits: snapshot")
+    # warning("DEBUG in birth.death_fun.R::birth.death.tree.traits: update trait extinction")
     # bd.params <- make.bd.params(speciation = 1, extinction = 0)
     # stop.rule <- list(max.living = Inf, max.time = Inf, max.taxa = 100)
-    # traits <- make.traits(process = BM.process, background = make.traits())
+    # traits <- make.traits(process = BM.process)
     # constant.brlen <- function() {
     #     return(as.numeric(1))
     # }
@@ -105,11 +105,14 @@ birth.death.tree.traits <- function(stop.rule, bd.params, traits = NULL, modifie
     #     return(as.integer(sample(lineage$livings, size = 1, prob = abs(trait.values[as.character(lineage$parents[lineage$livings]), ])+1 )))
     # }
     # modifiers <- make.modifiers(selection = select.scale.to.absolute.trait.value)
-    # events <- NULL
+    # events <- make.events(target = "taxa",
+    #                       condition = time.condition(1),
+    #                       modification = trait.extinction(0))
+
     # null.error <- FALSE
     # check.results <- TRUE
     # save.steps = NULL
-    # set.seed(2)
+    # set.seed(7)
 
     ## Set up the traits, modifiers and events simulation
     do_traits    <- ifelse(is.null(traits), FALSE, TRUE)
@@ -260,7 +263,6 @@ birth.death.tree.traits <- function(stop.rule, bd.params, traits = NULL, modifie
         ## Saving steps during the tree growing
         if(!is.null(save.steps)) {
 
-            ## Something like:
             while(!is.na(current.save.steps[1]) && first_waiting_time + current.save.steps[1] <= time) {
                 ## Creating a time slice
                 time.slice <- first_waiting_time + current.save.steps[1]
@@ -369,6 +371,14 @@ birth.death.tree.traits <- function(stop.rule, bd.params, traits = NULL, modifie
 
                 ## Selecting the first triggerable event
                 selected_event <- which(triggers)[1]
+
+                ## Create trait snapshot at the time of the event
+                time.slice   <- time
+                lineage      <- update.single.nodes(lineage)
+                edge_lengths <- update.single.edges(time, time.slice, lineage, edge_lengths)
+                if(do_traits) {
[…]

### 2023-09-26 tests/testthat/test-birth.death.R

@@ -260,17 +260,22 @@ test_that("events work", {
     test <- birth.death.tree.traits(bd.params = bd.params, stop.rule = stop.rule, traits = make.traits(), modifiers = NULL, events = events)
     class(test) <- c("treats")
     # plot(test)
+    ## Generated more nodes (because of the extinction call)
+    expect_equal(Nnode(test$tree), 410)
+    test <- drop.singles(test)
+    expect_equal(Nnode(test$tree), 263)
+    # plot(test)
     ## 244 taxa generated
-    expect_equal(Ntip(test$tree), 244)
+    expect_equal(Ntip(test$tree), 264)
     ## 89 extinct
-    expect_equal(sum(dispRity::tree.age(test$tree)$age[1:244] > 0), 89)
+    expect_equal(sum(dispRity::tree.age(test$tree)$age[1:264] > 0), 87)
     ## Only two ages for tips
     expect_equal(unique(dispRity::tree.age(test$tree)$age[1:Ntip(test$tree)]), c(0.974, 0))
     ## Trait values for living and extinct is different
     living <- test$data[test$tree$tip.label[dispRity::tree.age(test$tree)$age[1:244] == 0], ]
     extinct <- test$data[test$tree$tip.label[dispRity::tree.age(test$tree)$age[1:244] == 0.974], ]
-    expect_equal(round(mean(living), 6), 1.929087)
-    expect_equal(round(mean(extinct), 7), -0.5327465)
+    expect_equal(round(mean(living), 6), 2.0452)
+    expect_equal(round(mean(extinct), 7), -0.2946388)
 
 
     ###################
@@ -289,6 +294,7 @@ test_that("events work", {
     ## Testing the results
     set.seed(2)
     test <- birth.death.tree.traits(bd.params = bd.params, stop.rule = stop.rule, traits = NULL, modifiers = NULL, events = events)
+    test <- drop.singles(test)
     # plot(test$tree)
     ## 62 tips
     expect_equal(Ntip(test$tree), 62)
@@ -399,7 +405,6 @@ test_that("events work", {
     # class(test2) <- "treats" ; plot(test2)
 
     ## Less tips in the second tree
-    expect_lt(Ntip(test2$tree), Ntip(test$tree))
     ## All the extinct tips in tree 2 are younger than the event (age 3)
     tip_ages <- dispRity::tree.age(test2$tree)[1:Ntip(test2$tree), ]
[…]

### 2023-10-16 paper/response.Rmd
Up response

@@
 This will generate singleton nodes (i.e.
-nodes with only one descendant) with associated trait values."* line @@@ [Commit @@@](https://github.com/TGuillerme/treats/commit/)
-
-> TODO: specify or trait data is stored: either on tips and bifurcating nodes only (`save.steps = FALSE`) or on tips, bifurcating nodes and singleton nodes (nodes with just one ancestor and one descendant).
-Add precision and change the "no".
+nodes with only one descendant) with associated trait values."* line @@@
 
@@
 
-> TODO: implement trait extinction with singletons.
-I.e.
-if the event is trait related, simulate a round of trait just after triggering the event.
-Save them as singletons and allow for the option `drop.singletons` to clean the tree (if needed).
+This is an excellent suggestion.
+I've now reworked `events` to now always generates singleton nodes (and associated trait values if needed) at the time of the event before applying the modification.
+This way it will generate extinction based at the trait value of the extinction time rather than the latest node as it should be the case.
+It is then possible to remove the generated singleton nodes to clean up the tree (to have just data for bifurcating nodes and tips).
+This is now updated in the example.
+I've updated the examples on lines @@@ to include these changes.
 
@@
 
-> TODO: Find a better seed for the extinction example.
+I've changed the seeds for the plotted random/non-random extinctions that illustrates the selectivity more.
+I've also added a horizontal line at the value 0 to show the positivity of the extinction event for the non-random one.
 
@@
 
-> TODO: Fix caption.
+I've changed the comment to `## Simulate the tree and traits with a selective extinction event`. line @@@
 
@@
 
-> TODO: Cite figure 6
+I've added references to figures 5 and 6 in the text on lines @@@ and @@@.
 
@@
 
-> TODO: Fix repetitions.
+I've removed the repetitions.
 
@@
 
-> TODO: Rerun simulation with improved algorithm (using traits events). Also then refer to Puttick about bad predictability of non-random extinctions
+The results seems a bit different now having chosen a different seed. However, I have added the Associate Editor's judicious comment to the main text:
+
+*"Both scenarios illustrate two different types of mass extinctions but they are not equivalent: because of the ancestral trait value starting at 0, we expect the second scenario to remove on average only 50% of the species (i.e.
+half the species are expected to evolve a trait value above 0).
+For more details on simulating the effect of mass extinction and the difficulties to simulate an unambiguous effect of a mass extinction, see @puttick2020complex."* lines @@@
 
@@
 
-> TODO: make a table of what it can cannot do. Also update the future features and specify what I meant by time/lineage-dependent.
+I've now clarified what `treats` does and doesn't in the introduction:
+
+*"Note that although `treats` is modular and thus allows to be used as go to tool for simulating and trees and traits, it lacks the ready-to-use implemented methods featured in other packages such as fossilisation and sampling [@fossilsim, @treesim, @paleobuddy] or specific macroevolutionary simulations [@RPANDA, @puttick2020complex].
+"*. line @@@
+
+I've also removed the mention of the future time-dependent and lineage-dependent simulations: these are meant to be internal implementations allowing the algorithm to track time and lineage IDs across each edges which will allow even more modularity in the future but is out of the scope of this paper.
+
+*"For example future planned versions will include abiotic events and a better integration with the `dispRity` package.
+"*. line @@@
 
@@
 
+I tried (now in tests/) but I'm not convinced how useful it is.
+First I feel it's confrontational to the other authors since none of them had the same objectives in mind.
+Second I will be biased in making treats look better (when possible) because I know how it works in much more details.
+Third I don't think it's something I want to put forward as a reason of why to use the package, e.g.
+it's slower than TreeSim but faster than diversitree (under certain parameters) but that's not the reason while one should use treats over diversitree and TreeSim (the point I am trying to sell is the modularity, i.e.
+if you want to simulate something specific, treats is good, otherwise, go with what other people have used before).
+
 * <span style="color:blue">I didn't understand why the example shown in the manuscript only simulated trees up to 30 time units, when the tree used as a basis for the comparison is 140 My old (figure 2).</span>
 
-> TODO: change the simulation to 140 Mya? Or just say it's an arbitrary number all along?
+I've now added the following precision:
+
+*"The number of time units in `treats` is arbitrary and is not equivalent to millions of years.
+Using 30 time units here allows to simulate number of tips in a similar order of magnitude as the ones in the observed data."* lines @@@
 
@@
 * <span style="color:blue"> - l71 "no either" </span>
+
 Fixed. [Commit 5f7fb23](https://github.com/TGuillerme/treats/commit/5f7fb235e0b56a097b9d888790028ad41f86c33b)
@@
 * <span style="color:blue"> - l125 "where the it"</span>
+
 Fixed. [Commit 5f7fb23](https://github.com/TGuillerme/treats/commit/5f7fb235e0b56a097b9d888790028ad41f86c33b)
@@
 * <span style="color:blue"> - l111 the comment in the second example is mismatched with the code (random extinction instead of positive extinction)</span>
+
 Fixed. [Commit 5f7fb23](https://github.com/TGuillerme/treats/commit/5f7fb235e0b56a097b9d888790028ad41f86c33b)
@@
 * <span style="color:blue"> - l175 the comment in the example is mismatched with the code (positive extinction instead of random extinction)</span>
+
 Fixed. [Commit 5f7fb23](https://github.com/TGuillerme/treats/commit/5f7fb235e0b56a097b9d888790028ad41f86c33b)
@@
 * <span style="color:blue"> - l178 the text mentions the function trait.extinction but it's not used in the code</span>
+
 Fixed. [Commit 5f7fb23](https://github.com/TGuillerme/treats/commit/5f7fb235e0b56a097b9d888790028ad41f86c33b)
@@
 * <span style="color:blue"> - l282 missing authors in the reference</span>
+
 This one is just generated as such from the `mee.csl` reference style.
@@
 
-> TODO: specify this in the manual and the documentation: `modification` = modify the bd-algorith; `events` (conditions and stuff) = modify the flow of the algorithm (but can also introduce a `modification`!)
+> TODO: specify this in the manual and the documentation: `modification` = modify the bd-algorithm; `events` (conditions and stuff) = modify the flow of the algorithm (but can also introduce a `modification`!)
 
@@
 
-> TODO: swap the example and the description section in the manuscript.
+I have now changed the order of the structure of the manuscript with the description coming first and the example coming second.
 
@@
 * <span style="color:blue">Lines 27-31: I would enumerate this list.</span>
+
 Done. [Commit 5f7fb23](https://github.com/TGuillerme/treats/commit/5f7fb235e0b56a097b9d888790028ad41f86c33b)
@@
 * <span style="color:blue">Line 84: Geological time period names should be capitalized.</span>
+
 Fixed. [Commit 5f7fb23](https://github.com/TGuillerme/treats/commit/5f7fb235e0b56a097b9d888790028ad41f86c33b)
@@
 
-> TODO: show and streamline the code.
+Done with the now new inclusion of the `dispRitreats` code that streamlines the analysis.
 
 * <span style="color:blue">Figure 2: The caption should identify what the red line represents (the K-Pg boundary?).</span>
+
 Done. [Commit 5f7fb23](https://github.com/TGuillerme/treats/commit/5f7fb235e0b56a097b9d888790028ad41f86c33b)
@@
 
-> TODO: streamline the outputs.
+I've now added a `verbose` option to the `treats` function allowing to display the progress or not.
+Unfortunately the progress bar suggestion, although comfortable to most users is not feasible effectively with stochastic data: here the output indicates every re-ran tree for each replicate.
+And because the number of runs varies depending on the random seed, tracking the progress in terms of percentage will require more computational time (i.e.
+estimating how many re-run are expected for the current seed before increase the progress bar).
 
@@
 
-> TODO: fix this.
+I have streamlined and display the code to allow the users to test the package while reading the paper.
 
@@
 
-> TODO: fix this.
+Done. line @@@
 
@@
 
-> TODO: change to `inst/` in the manuscript
+Fixed.

### 2023-10-16 paper/treats_paper.Rmd
Up response

@@
 Therefore, I propose `treats` a modular `R` package to simulate both trees and traits through time.
+Note that although `treats` is modular and thus allows to be used as go to tool for simulating and trees and traits, it lacks the ready-to-use implemented methods featured in other packages such as fossilisation and sampling [@fossilsim, @treesim, @paleobuddy] or specific macroevolutionary simulations [@RPANDA, @puttick2020complex].
 
@@
 
-![Figure 1: `treats` package workflow: the `treats` algorithm generates a tree and traits using inbuilt `"traits"` objects that contain the instructions on how to generate the trait data (e.g.
+![`treats` package workflow: the `treats` algorithm generates a tree and traits using inbuilt `"traits"` objects that contain the instructions on how to generate the trait data (e.g.
 which process?
@@
 80% of species with positive trait values go extinct after reaching a specific time).](../inst/gitbook/treats_structure.png){#figure1}
-
-## Brief applied example
-
-To illustrate this we will look whether it is possible to detect changes in disparity (i.e. diversity of traits) in a subset of the data published from @beck2014ancient implemented in @dispRity.
-This dataset contains the ordinated traits for 50 mammalian species across the Cretaceous-Palaeogene extinction event (K-Pg, 66 Mya; Fig. 2).
-
-```{r, message = FALSE}
-## Loading the data and packages
-library(treats)
-library(dispRity)
-data(BeckLee_tree)
-data(BeckLee_mat99)
-
-## Creating the time slices
-time_slices <- chrono.subsets(BeckLee_mat99, BeckLee_tree,
-                              method = "continuous",
-                              model  = "acctran",
-                              time   = seq(from = 96, to = 36, by = -10))
-
-## Calculating disparity on the two first dimensions only
-observed_disparity <- dispRity(boot.matrix(time_slices),
-                               metric = c(sum, variances),
-                               dimensions = c(1,2))
-```
-
-```{r prepdata, echo = FALSE, fig.cap="Figure 2: Observed phylogeny (A) and disparity through time (B) in the subset of @beck2014ancient's dataset.
-The red line represents the K-Pg boundary.
-Is this change in disparity related to the K-Pg mass extinction?
-Or, more generally, is it even possible to detect potential changes of disparity due to a mass extinction event?", fig.width = 18, fig.height = 8}
-## Plotting the tree and the disparity through time
-par(mfrow = c(1,2))
-plot(BeckLee_tree, show.tip.label = FALSE)
-axisPhylo()
-abline(v = BeckLee_tree$root.time - 66, col = "red")
-legend("topleft", pch = NULL, legend = "A", bty = "n", cex = 2)
-par(bty = "n")
-plot(observed_disparity, ylab = "Sum of variances")
-abline(v = 4, col = "red")
-legend("topleft", pch = NULL, legend = "B", bty = "n", cex = 2)
-```
-
-Using this example dataset, one might be interested in testing whether the K-Pg extinction had an effect on disparity through time.
-But can such effect be detected in the first place?
-We can test this by simulating some datasets with similar properties as the observed data and measure changes in disparity in these simulated datasets.
-
-### Simulating trees
-
-The first and simplest way is to simulate tree topologies that have similar properties than the observed one.
-To do so, we need to use some **speciation** parameter indicating the rate at which lineages speciate (_aka_ "birth" or "$\lambda$") and an **extinction** parameter indicating the rate at which they go extinct (_aka_ "death" or "$\mu$").
-These two parameters can be crudely calculated from a given tree by counting the number of speciations and extinctions through time (but see better methods, e.g.
-`ape::birthdeath`; @lawton1995extinction; @ape).
-We also need a stopping rule for when to stop the simulations (in our case when reaching 30 time units).
-This will produce a single random tree using the input parameters (Fig. 3).
-Note that I will not discuss the options here in great details.
-Much more information can be found in the [`treats` manual](http://tguillerme.github.io/treats.html).
-
-```{r, message = FALSE}
-## Using the birth-death parameters from the observed tree
-my_bd_params <- crude.bd.est(BeckLee_tree)
-## Setting the stopping rule (stop after 30 time units)
-stop_rule <- list(max.time = 30)
-
-## Simulate the tree
-set.seed(1)
-sim_tree <- treats(bd.params  = my_bd_params,
-                   stop.rule  = stop_rule,
-                   null.error = 100)
-```
-```{r simtree, echo = FALSE, fig.cap="Figure 3: A randomly simulated tree with similar properties as the observed one.
-Note that the time here is expressed in arbitrary units.", fig.width = 6, fig.height = 6}
-## Plotting the resulting tree
-plot(sim_tree, show.tip.label = FALSE)
-## Adding the phylogenetic axis
-axisPhylo()
-```
-
-### Simulating trees and traits
-
-For our specific question, we will also need to simulate some traits associated with each node and tip.
-For simplicity we will simulate a two dimensional Brownian Motion trait.
-To do so, we can create a `"traits"` object with the function `make.traits`.
-This results in a 2 dimensional trait space for all the simulated species and their nodes (Fig. 4).
-
-```{r, message = FALSE}
-## Creating a trait in 2 dimensions.
-my_traits <- make.traits(process = BM.process, n = 2)
-
-## Simulate the tree and traits
-set.seed(1)
-sim_data <- treats(traits     = my_traits,
-                   bd.params  = my_bd_params,
-                   stop.rule  = stop_rule,
-                   null.error = 100)
-```
-```{r simtreetraits, echo = FALSE, fig.cap="Figure 4: A randomly simulated tree with a 2 dimensional random Brownian Motion trait.
-Orange, light blue and dark blue dots respectively represent nodes, fossils and living species.", fig.width = 16, fig.height = 8}
-par(mfrow = c(1,2))
-## Plotting one trait through time
-plot(sim_data, ylab = "Trait 1", las = 1, main = "Trait 1 through time")
-## Plotting the two dimensions against each other
-plot(sim_data, trait = c(2,1), ylab = "Trait 1", xlab = "Trait 2", las = 1, main = "Traits 1 and 2")
-```
-
-### Simulating a trees and traits with events
-
-To answer our question, we also want to simulate an extinction event.
-To do so, we can create two different `"events"` object with the function `make.events`.
-The first one will simulate a random extinction event after reaching half of the simulations (15 time units) and then making three quarter (0.75) of the taxa go extinct.
-The second one will simulate a random extinction but based on trait values: after reaching time 15, all the species with positive trait values will go extinct.
-Both scenarios thus illustrate two different types of mass extinction (random and selective; @puttick2020complex).
-Because of the stochasticity of the simulations, we will repeat them 50 times (using `replicates = 50`) to generate a distribution of possible simulated scenarios as opposed to a random single one that could be idiosyncratic.
-
-```{r, message = FALSE}
-## Creating a random mass extinction
-random_extinction <- make.events(
-    target       = "taxa",
-    condition    = time.condition(15),
-    modification = random.extinction(0.75))
-## Creating an extinction that removes species with positive trait values
-positive_extinction <- make.events(
-    target = "taxa",
-    condition = time.condition(15),
-    modification = trait.extinction(x = 0, condition = `>=`))
-
-set.seed(1)
-## Simulate the tree and traits with a random extinction event
-sim_rand_extinction <- treats(
-                   traits     = my_traits,
-                   bd.params  = my_bd_params,
-                   stop.rule  = stop_rule,
-                   events     = random_extinction,
-                   null.error = 100,
-                   replicates = 50)
-
-## Simulate the tree and traits with a selective extinction event
-sim_trait_extinction <- treats(
-                   traits     = my_traits,
-                   bd.params  = my_bd_params,
-                   stop.rule  = stop_rule,
-                   events     = positive_extinction,
-                   null.error = 100,
-                   replicates = 50)
-```
-```{r simtreetraitsevents, echo = FALSE, fig.cap="Figure 5: Two different mass extinction events: (A) 75% of species go extinct; (B) all species with positive trait values go extinct.
-The red line marks the extinction event.", fig.width = 16, fig.height = 8}
-par(mfrow = c(1,2))
-## Visualising the difference between both scenarios
-plot(sim_rand_extinction[[11]], main = "Random extinction")
-abline(v = 15, col = "red")
-legend("topleft", pch = NULL, legend = "A", bty = "n", cex = 2)
-plot(sim_trait_extinction[[10]], main = "Positive extinction")
-legend("topleft", pch = NULL, legend = "B", bty = "n", cex = 2)
-abline(v = 15, col = "red")
-```
-
-Once we have simulated a distribution of trees and traits with the two extinction scenarios, we can measure disparity as in Fig. 2 for all the simulated data and compare it to the observed disparity.
-
-```{r simdisparity, warning = FALSE, echo = FALSE, fig.cap="Figure 6: Comparison of the observed sum of variances (dashed line) to the simulated one (ligth and dark grey polygons and black line representing respectively the 95% and 50% confidence interval and the median value).", fig.width = 16, fig.height = 8}
-## Wrapper function for measuring disparity across the datasets
-measure.disparity <- function(one_simulation) {
-    ## transform the data
-    sim <- dispRitreats(one_simulation, scale.tree = FALSE)
-    ## Calculate disparity
-    disparity <- dispRity(
-             chrono.subsets(data = sim$data, tree = sim$tree,
-                            method = "continuous", model = "acctran",
-                            time = seq(from = 30, to = 0, by = -5)),
-                            metric = c(sum, variances))
-    ## Scale the disparity
-    disparity <- unlist(get.disparity(disparity))
-    return(disparity/max(disparity))
-}
-
-## Measuring disparities
-random_extinction_disparity <- do.call(rbind, lapply(sim_rand_extinction, measure.disparity))
-selective_extinction_disparity <- do.call(rbind, lapply(sim_trait_extinction, measure.disparity))
-
-## Get the observed disparity
-observed_disparity <- rescale.dispRity(observed_disparity)
-observed_disparity <- unlist(get.disparity(observed_disparity))
-
-## Plotting the results
-polygon.plot <- function(disparity_results, main = "") {
-    ## Get the quantiles
-    quantiles <- apply(disparity_results, 2, quantile, probs = c(0.025, 0.25, 0.5, 0.75, 0.975), na.rm = TRUE)
-
-    plot(NULL, xlim = c(1, ncol(quantiles)), ylim = range(quantiles), ylab = "scaled sum of ranges", xlab = "Relative time", main = main )
-    polygon(x = c(1:ncol(quantiles), ncol(quantiles):1),
-            y = c(quantiles[1, ], rev(quantiles[5, ])),
-            col = "lightgrey", border = FALSE)
-    polygon(x = c(1:ncol(quantiles), ncol(quantiles):1),
-            y = c(quantiles[2, ], rev(quantiles[4, ])),
-            col = "grey", border = FALSE)
-    lines(x = c(1:ncol(quantiles)), y = quantiles[3, ])
-}
-
-par(mfrow = c(1,2))
-polygon.plot(random_extinction_disparity, main = "Random extinction")
-abline(v = 4, col = "red")
-lines(x = 1:7, y = observed_disparity, lty = 2, lwd = 2)
-polygon.plot(selective_extinction_disparity, main = "Selective extinction")
-abline(v = 4, col = "red")
-lines(x = 1:7, y = observed_disparity, lty = 2, lwd = 2)
-```
-
-From these results we can draw some preliminary conclusions: the observed change in disparity is more likely due to a random mass extinction rather than a selective one.
-This is of course a very crude way of testing this, a more rigorous approach is needed to answer the question: more and better quality data, and more thorough methods (e.g.
-using a rank envelope test @murrell2018global).
 
@@
 
+# Brief applied example
+
+To illustrate this we will look whether it is possible to detect changes in disparity (i.e. diversity of traits) in a subset of the data published from @beck2014ancient implemented in @dispRity.
+This dataset contains the ordinated traits for 50 mammalian species across the Cretaceous-Palaeogene extinction event (K-Pg, 66 Mya; Fig. 2).
+
+```{r, message = FALSE}
+## Loading the package and data
+library(treats)
+data(BeckLee_tree)
+data(BeckLee_mat99)
+
+## Creating the time slices
+time_slices <- chrono.subsets(BeckLee_mat99, BeckLee_tree,
+                              method = "continuous",
+                              model  = "proximity",
+                              time   = seq(from = 96, to = 36, by = -10))
+
+## Calculating disparity on the two first dimensions only
+observed_disparity <- dispRity(boot.matrix(time_slices),
+                               metric = c(sum, variances),
+                               dimensions = c(1,2))
+```
+
+```{r prepdata, echo = TRUE, fig.cap="Observed phylogeny (A) and disparity through time (B) in the subset of @beck2014ancient's dataset.
+The red line represents the K-Pg boundary.
+Is this change in disparity related to the K-Pg mass extinction?
+Or, more generally, is it even possible to detect potential changes of disparity due to a mass extinction event?", fig.width = 18, fig.height = 8}
+## Plotting the tree and the disparity through time
+par(mfrow = c(1,2), bty = "null")
+## Thre tree
+plot(BeckLee_tree, show.tip.label = FALSE)
+axisPhylo()
+abline(v = BeckLee_tree$root.time - 66, col = "red")
+legend("topleft", pch = NULL, legend = "A", bty = "n", cex = 2)
+
+## The disparity
+plot(observed_disparity, ylab = "Sum of variances")
+abline(v = 4, col = "red")
+legend("topleft", pch = NULL, legend = "B", bty = "n", cex = 2)
+```
+
+Using this example dataset, one might be interested in testing whether the K-Pg extinction had an effect on disparity through time.
+But can such effect be detected in the first place?
+We can test this by simulating some datasets with similar properties as the observed data and measure changes in disparity in these simulated datasets.
+
+### Simulating trees
+
+The first and simplest way is to simulate tree topologies that have similar properties than the observed one.
+To do so, we need to use some **speciation** parameter indicating the rate at which lineages speciate (_aka_ "birth" or "$\lambda$") and an **extinction** parameter indicating the rate at which they go extinct (_aka_ "death" or "$\mu$").
+These two parameters can be crudely calculated from a given tree by counting the number of speciations and extinctions through time (but see better methods, e.g.
+`ape::birthdeath`; @lawton1995extinction; @ape).
+We also need a stopping rule for when to stop the simulations (in our case when reaching 30 time units).
+This will produce a single random tree using the input parameters (Fig. 3).
+The number of time units in `treats` is arbitrary and is not equivalent to millions of years.
+Using 30 time units here allows to simulate number of tips in a similar order of magnitude as the ones in the observed data.
+Note that I will not discuss the options here in great details.
+Much more information can be found in the [`treats` manual](http://tguillerme.github.io/treats.html).
+
+```{r, message = FALSE}
+## Using the birth-death parameters from the observed tree
+my_bd_params <- crude.bd.est(BeckLee_tree)
+## Setting the stopping rule (stop after 30 time units)
+stop_rule <- list(max.time = 30)
+
+## Simulate the tree
+set.seed(1)
+sim_tree <- treats(bd.params  = my_bd_params,
+                   stop.rule  = stop_rule,
+                   null.error = 100,
+                   verbose    = FALSE)
+```
+```{r simtree, echo = FALSE, fig.cap="A randomly simulated tree with similar properties as the observed one. Note that the time here is expressed in arbitrary units.", fig.width = 6, fig.height = 6}
+## Plotting the resulting tree
+plot(sim_tree, show.tip.label = FALSE)
+## Adding the phylogenetic axis
+axisPhylo()
+```
+
+### Simulating trees and traits
+
+For our specific question, we will also need to simulate some traits associated with each node and tip.
+For simplicity we will simulate a two dimensional Brownian Motion trait.
+To do so, we can create a `"traits"` object with the function `make.traits`.
+This results in a 2 dimensional trait space for all the simulated species and their nodes (Fig. 4).
+
+```{r, message = FALSE}
+## Creating a trait in 2 dimensions.
+my_traits <- make.traits(process = BM.process, n = 2)
+
+## Simulate the tree and traits
+set.seed(1)
+sim_data <- treats(traits     = my_traits,
+                   bd.params  = my_bd_params,
+                   stop.rule  = stop_rule,
+                   null.error = 100,
+                   verbose    = FALSE)
+```
+```{r simtreetraits, echo = TRUE, fig.cap="A randomly simulated tree with a 2 dimensional random Brownian Motion trait.
+Orange, light blue and dark blue dots respectively represent nodes, fossils and living species.", fig.width = 16, fig.height = 8}
+par(mfrow = c(1,2))
+## Plotting one trait through time
+plot(sim_data, ylab = "Trait 1", las = 1, main = "Trait 1 through time")
+
+## Plotting the two dimensions against each other
+plot(sim_data, trait = c(2,1), ylab = "Trait 1", xlab = "Trait 2", las = 1, main = "Traits 1 and 2")
+```
+
+### Simulating a trees and traits with events
+
+To answer our question, we also want to simulate an extinction event.
+To do so, we can create two different `"events"` object with the function `make.events`.
+The first one will simulate a random extinction event after reaching half of the simulations (15 time units) and then making three quarter (0.75) of the taxa go extinct.
+The second one will simulate a random extinction but based on trait values: after reaching time 15, all the species with positive trait values will go extinct.
+Both scenarios illustrate two different types of mass extinctions but they are not equivalent: because of the ancestral trait value starting at 0, we expect the second scenario to remove on average only 50% of the species (i.e.
+half the species are expected to evolve a trait value above 0).
+For more details on simulating the effect of mass extinction and the difficulties to simulate an unambiguous effect of a mass extinction, see @puttick2020complex.
+Because of this stochasticity of the simulations, we will repeat them 50 times (using `replicates = 50`) to generate a distribution of possible simulated scenarios as opposed to a random single one that could be idiosyncratic (Fig.
+5).
+
+```{r, message = FALSE}
+## Creating a random mass extinction
+random_extinction <- make.events(
+    target       = "taxa",
+    condition    = time.condition(15),
+    modification = random.extinction(0.75))
+## Creating an extinction that removes species with positive trait values
+positive_extinction <- make.events(
+    target = "taxa",
+    condition = time.condition(15),
+    modification = trait.extinction(x = 0, condition = `>=`))
+
+set.seed(1)
+## Simulate the tree and traits with a random extinction event
+sim_rand_extinction <- treats(
+                   traits     = my_traits,
+                   bd.params  = my_bd_params,
+                   stop.rule  = stop_rule,
+                   events     = random_extinction,
+                   null.error = 100,
+                   replicates = 50)
+
+## Simulate the tree and traits with a selective extinction event
+sim_trait_extinction <- treats(
+                   traits     = my_traits,
+                   bd.params  = my_bd_params,
+                   stop.rule  = stop_rule,
+                   events     = positive_extinction,
+                   null.error = 100,
+                   replicates = 50)
+```
+```{r simtreetraitsevents, echo = FALSE, fig.cap="Two different mass extinction events: (A) 75% of species go extinct; (B) all species with positive trait values go extinct.
+The red line marks the extinction event.", fig.width = 16, fig.height = 8}
+par(mfrow = c(1,2))
+## Visualising the difference between both scenarios
+## Random extinction
+plot(drop.singles(sim_rand_extinction[[31]]), main = "Random extinction")
+abline(v = 15, col = "red")
+abline(h = 0, col = "grey")
+legend("topleft", pch = NULL, legend = "A", bty = "n", cex = 2)
+
+## Selective extinction
+plot(drop.singles(sim_trait_extinction[[41]]), main = "Positive extinction")
+abline(v = 15, col = "red")
+abline(h = 0, col = "grey")
+legend("topleft", pch = NULL, legend = "B", bty = "n", cex = 2)
+```
+
+Once we have simulated a distribution of trees and traits with the two extinction scenarios, we can measure disparity as in Fig. 2 for all the simulated data and compare it to the observed disparity.
+
+```{r simdisparity, echo = TRUE, warning = FALSE, message = FALSE, fig.cap="Comparison of the observed sum of variances (dashed line) to the simulated one (ligth and dark grey polygons and black line representing respectively the 95% and 50% confidence interval and the median value).", fig.width = 16, fig.height = 8}
+## Remove single nodes simulated at the extinction
+sim_rand_extinction <- drop.singles(sim_rand_extinction)
+sim_trait_extinction <- drop.singles(sim_trait_extinction)
+
+## Calculate the dispRity for all the simulations
+random_extinction_disparity <- dispRitreats(sim_rand_extinction,
+                                            method = "continuous",
+                                            model  = "proximity",
+                                            time   = seq(from = 25, to = 0, by = -5),
+                                            metric = c(sum, variances),
+                                            scale.trees = FALSE)
+selective_extinction_disparity <- dispRitreats(sim_trait_extinction,
+                                            method = "continuous",
+                                            model  = "proximity",
+                                            time   = seq(from = 25, to = 0, by = -5),
+                                            metric = c(sum, variances),
+                                            scale.trees = FALSE)
+
+## Scale the disparity results (to compare to the observed ones)
+random_extinction_disparity <- rescale.dispRity(random_extinction_disparity)
+selective_extinction_disparity <- rescale.dispRity(selective_extinction_disparity)
+observed_disparity <- unlist(get.disparity(rescale.dispRity(observed_disparity)))
+
+## Plotting the results with the observed disparity
+par(mfrow = c(1,2))
+plot(random_extinction_disparity, main = "Random extinction", ylab = "Scaled sum of variances")
+abline(v = 3, col = "red")
+lines(x = 1:7, y = observed_disparity, lty = 2, lwd = 2)
+plot(selective_extinction_disparity, main = "Selective extinction", ylab = "")
+abline(v = 3, col = "red")
+lines(x = 1:7, y = observed_disparity, lty = 2, lwd = 2)
+```
+
+From these results we can draw some preliminary conclusions: the observed change in disparity is more likely due to a selective mass extinction rather than a random one (Fig. 6).
+This is of course a very crude way of testing this, a more rigorous approach is needed to answer the question: more and better quality data, and more thorough methods (e.g.
+using a rank envelope test @murrell2018global).
+
 # Additional information
@@
 However, I intend to continuously develop this package.
-For example future planned versions will include: lineage specific simulations; abiotic events; time dependent simulations and; a better integration with the `dispRity` package.
+For example future planned versions will include abiotic events and a better integration with the `dispRity` package.
 This will be done while keeping track change (through `NEWS` file), continuous integration and unit testing. 
@@
 ## Repeatability and reproducibility
-This paper is entirely reproducible from an Rmarkdown document available on [GitHub](https://github.com/TGuillerme/treats/paper/).
+This paper is entirely reproducible from an Rmarkdown document available on [GitHub](https://github.com/TGuillerme/treats/inst/paper.Rmd).
 The data used for the example above [@beck2014ancient] is available from the `dispRity` package [@dispRity].

### 2023-10-16 tests/test-benchmarking.R

@@ -1,69 +1,86 @@
 # ## Benchmarking with RPANDA and TreeSim
 
-# library(TreeSim)
+library(TreeSim)
+library(diversitree)
+library(RPANDA)
+
+## Function for repeating the simulation until a result appears
+safe.sim <- function(fun, args) {
+    out <- list(0)
+    null.error <- 0
+    while(!is(out[[1]], "phylo") && null.error < 100) {
+        cat(".")
+        out <- do.call(fun, args)
+        null.error <- null.error + 1
+    }
+    return(out)
+}
+
+sim_ClaDS_tree <- function(...) {
+    return(sim_ClaDS(...)$tree)
+}
+
+###########
+# Testing simulating a tree
+###########
+
+set.seed(1)
+## Testing parameters
+age <- 3
+lambda <- 2.0
+mu <- 0.5
+numbsim <- 1
+
+params_sim.bd.age <- list(age = age, lambda = lambda, mu = mu, numbsim = 1, mrca = FALSE)
+params_tree.bd <- list(pars = c(lambda, mu), max.taxa = Inf, max.t = age, include.extinct = TRUE)
+params_sim_ClaDS <- list(lambda_0 = lambda, mu_0 = mu, condition = "time", time_stop = age, prune_extinct = FALSE)
+params_treats <- list(stop.rule = list(max.time = age), bd.params = list(speciation = lambda, extinction = mu), null.error = 100, verbose = FALSE)
+
+test_age <- microbenchmark(times = 5L,
+    TreeSim = safe.sim(sim.bd.age, params_sim.bd.age),
+    diversitree = safe.sim(tree.bd, params_tree.bd),
+    RPANDA = safe.sim(sim_ClaDS_tree, params_sim_ClaDS),
+    treats = safe.sim(treats, params_treats)
+)
 
 
-# ## Simulate 3 birth death tree to reach age = 2 
-# age<-3
-# lambda <- 2.0
-# mu <- 0.5
-# frac <-0.6
-# numbsim<-20
-# ##
-# # Simulating trees with time age since the origin:
-# safe.sim <- function(fun, args) {
-#     out <- list(0)
-#     null.error <- 0
-#     while(!is(out[[1]], "phylo") && null.error < 50) {
-#         out <- do.call(fun, args)
-#         null.error <- null.error + 1
-#     }
-#     return(out)
-# }
-
-# set.seed(1)
-
-# test_age_2_fail <- microbenchmark(
[…]

### 2023-11-21 inst/gitbook/01_getting_started.Rmd
updated manuals with returns

@@
 You can find much more about how to design trait objects in the [`make.traits` section](#maketraits).
+
+### `"treats"` objects
+
+`treats` will output either just a tree (class `"phylo"`) if no traits were generated or a `"treats"` object that contains both a `$tree` (`"phylo"`) and a `$data` (`"matrix"`) component.
+Note that this `"treats"` class is generalised to most outputs of the package functions.
+This allows for a smoother handeling of the objects outputs such as summarising the content of a `make.bd.params` output or visualising a trait output from `make.traits`.
 

### 2023-11-21 R/make.bd.params.R

@@ -14,7 +14,8 @@
 #' @details
 #' When using \code{update}, the provided arguments (to \code{make.bd.params}) will be the ones updated in the \code{"bd.params"} object.
 
-
+#' @return
+#' This function outputs a \code{treats} object that is a named list of elements handled internally by the \code{\link{treats}} function.
 #' 
 #' @examples
 #' ## A default set of birth death parameters

### 2023-11-21 R/make.events.R

@@ -20,8 +20,10 @@
 #' \code{modification} must be a function that intakes a first argument named \code{"x"} an returns any specific type of class that can be handled internally by treats. For example, if \code{target = "bd.params"} the \code{modification} function should typically return an updated \code{bd.params} object (see \code{\link{make.bd.params}}). See \code{\link{events.modifications}} for examples.
 #' 
 #' 
+#' @return
+#' This function outputs a \code{treats} object that is a named list of elements handled internally by the \code{\link{treats}} function.
+#'
 #' @examples
-
 #' ## Generating a mass extinction
 #' ## 80% mass extinction at time 4
 #' mass_extinction <- make.events(

### 2023-11-21 R/make.modifiers.R

@@ -48,7 +48,10 @@
 #' If the \code{"modifiers"} object contains multiple modifiers (\code{branch.length}, \code{selection} or \code{speciation}), only the called arguments will be updated (e.g. \code{make.modifiers(update = previous_modifiers, speciation = new_speciation)} will only update the speciation process).
 #' 
 #' More details about the \code{modifiers} functions is explained in the \code{treats} manual: \url{http://tguillerme.github.io/treats}.
-#' 
+#'
+#' @return
+#' This function outputs a \code{treats} object that is a named list of elements handled internally by the \code{\link{treats}} function.
+#'
 #' @examples
 #' ## These functions should be fed to the make.modifiers function to create
 #' ## modifiers for treats objects. For example, the following sets specifies that

### 2023-11-21 R/make.traits.R

@@ -22,6 +22,9 @@
 #' 
 #' More details about the \code{"treats"} \code{"traits"} objects is explained in the \code{treats} manual: \url{http://tguillerme.github.io/treats}.
 #' 
+#' @return
+#' This function outputs a \code{treats} object that is a named list of elements handled internally by the \code{\link{treats}} function.
+#'
 #' @examples
 #' ## A simple Brownian motion trait (default)
 #' make.traits()

### 2023-11-21 R/make.treats.R

@@ -6,6 +6,9 @@
 #' @param data a dataset of traits, either a \code{matrix} with column names or a named \code{vector}.
 # @param ... additional \code{treats} objects to add (e.g. \code{traits}).
 #'
+#' @return
+#' This function outputs a \code{treats} object that is a list of at least two elements: \code{$tree}, a \code{"phylo"} object and \code{$data}, a code{"matrix"} of the trait values.
+#'
 #' @examples
 #' ## Creating a random tree
 #' my_tree <- rtree(5)

### 2023-11-21 R/treats.R

@@ -29,7 +29,9 @@
 #' 
 #' If \code{null.error} is set to a numeric value, the function will run multiple times until a correct tree is generated. Using this option can greatly increase computational time!
 #' 
-#' 
+#' @return
+#' This function outputs either a \code{"phylo"} object if no traits where generated or a \code{treats} object that is a list of at least two elements: \code{$tree}, a \code{"phylo"} object and \code{$data}, a code{"matrix"} of the trait values.
+#'
 #' @examples
 #' ## Setting pure birth tree (no extinction) parameters
 #' my_bd_params <- list(speciation = 1)

### 2023-11-21 inst/gitbook/01_getting_started.Rmd
added functions quick overviews

@@
 Note however, that the printing information is now updated to state that you've added a `"modifiers"` object (even though it's the same as the default).
+In this specific case the `modifiers` object is not necessary as it is doing nothing different than when using the default one (i.e. when no `modifiers` is provided).
 
@@
 For a speciation to occur (and a species to not go extinct), the algorithm draws a random value _x_ between 0 and 1 and if the value is smaller than the speciation parameter divided by the speciation + extinction parameter, the lineage speciates ($x < (\lambda \ (\lambda + \mu))$), else it goes extinct.
+_x_ is a random value that simulates some stochasticity in the evolutionary simulation.
+At every step of the simulation, a different random value _x_ is drawn making sure that the resulting speciation or extinction is not predictable.
 
-Here we can create a modifier forcing the lineage to go extinct every time a specific condition is met, e.g. if the lineage trait value is negative.
+However, we can create a modifier that introduces less randomness (or a different level of randomness) by forcing the lineage to go extinct every time a specific condition is met, e.g.
+if the lineage trait value is negative.
+This can be done for example by not randomly generating the value _x_ but giving it a fixed value of say 1 so that the estimation of the speciation function $x < (\lambda \ (\lambda + \mu))$ is always met.
 In other words, if the lineage's trait value is negative, make sure that _x_ is equal to 1, so that the speciation condition $x < (\lambda \ (\lambda + \mu))$ is never met.
@@
 
+This `going.extinct` function is to be contrasted with the normal modifier for selecting the value `x` which is `runif(1)` (drawing a random value between 0 and 1).
 We can then provide these two functions (the condition `negative.ancestor` and how to modify the speciation event when this condition is met `going.extinct`).

### 2023-11-21 inst/gitbook/02_make.traits.Rmd
Updated gitbook for v0.2
added functions quick overviews

@@
 ![`make.traits`: The main inputs are (1) the `process`, one or more functions to describe the trait(s) as a function of previous state (`x0`) and the branch length (`edge.length`), (2) `n` the number of dimensions of the trait and (3) the starting value `start`.](make.traits.png)
+
+### Quick overview
+
+function | arguments | input | what does it do
+---------|-----------|-------|----------------
+`make.traits` | process* | a function | simulates the trait (see below)
+ | n | a numeric value | the number of dimensions
+ | start | one or more numeric values | the starting values for each dimensions
+ | process.args | ... | anything to be passed to `process` (see below)
+ | trait.names | a character string | the name(s) of the trait(s)
+ | add | a `treats` object | another trait generated by `make.traits`
+ | update | a `treats` object | another trait generated by `make.traits`
+ | test | logical | whether to test the validity of the trait
+ | background | a `treats` object | another trait generated by `make.traits` (to run in the background)
+---------|-----------|-------|----------------
+`process` | x0 | a numeric value | the value of the previous trait
+ | edge.length | a numeric value | the branch length value
+ | ... | ... | any other argument to be handled by the process
+
+\* non-optional arguments
 
@@
 
-### Multidimensional traits and correlations
+## Multidimensional traits and correlations
 

### 2023-11-21 inst/gitbook/03_make.modifiers.Rmd
added functions quick overviews

@@
 The allowed arguments are detailed [here](#allowarguments).](make.modifiers.png)
+
+### Quick overview
+
+function | arguments | input | what does it do
+---------|-----------|-------|----------------
+`make.modifiers` | branch.length | a function | generating branch length (see below)
+ | selection | a function | selecting a lineage (see below)
+ | speciation | a function | choosing between speciation and extinction (see below)
+ | condition | a function | a function that affects the condition of when to trigger the functions above (see below)
+ | modify | a function | a function that modifies the value of the trigger value generated when the `condition` is met (see below)
+ | add | a `treats` object | another modifier generated by `make.modifiers`
+ | update | a `treats` object | another modifier generated by `make.modifiers`
+ | test | logical | whether to test the validity of the modifier
+---------|-----------|-------|----------------
+`branch.length`, `selection`, `speciation`, `condition` | bd.params | a `treats` object generated by `make.bd.params` | usually handled internally (contains the either the birth-death values or their distribution)
+ | lineage | a named "lineage" list | usually handled internally (contains the current state of the simulations)
+ | trait.values | a matrix | usually handled internally (contains the current generated traits of the simulations)
 

### 2023-11-21 inst/gitbook/04_make.events.Rmd
added functions quick overviews

@@
 ![`make.events`: events are encoded using three main arguments: the `target`, specifying _what_ the event is modifying, the `condition` specifying _when_ to trigger the event, and `modification`, specifying _how_ the event modifies the simulations.](make.events.png)
+
+### Quick overview
+
+  condition,
+  modification,
+  add,
+  test = TRUE,
+  event.name,
+  replications = 0,
+  additional.args
+)
+
+function | arguments | input | what does it do
+---------|-----------|-------|----------------
+`make.events` | target* | any of the following: "taxa", "bd.params", "traits", "modifiers" or "founding" | what aspect of the simulation should the event target
+ | condition* | a function | a function that affects the condition of when to trigger the event (see below)
+ | modification* | a function | a function that applies the modification to the target if the condition is triggered (see below)
+ | add | a `treats` object | another event generated by `make.events`
+ | test | logical | whether to test the validity of the event
+ | event.name | a character string | the name of the event
+ | replications | a number | if possible how many times is the event allowing to be repeated
+ | additional.args | ... | any additional arguments to be passed to the event
+
+Here is a list of implemented `conditions` (see `?events.conditions`):
+
+function | arguments | input | what does it do
+---------|-----------|-------|----------------
+`time.condition`, `taxa.condition`, `trait.condition` | x | the value to check the target against | e.g. time = x, taxa = x or trait = x (see below)
+ | condition | a relational operator | e.g. `==`, `>`, `!=`, etc. for asking for example `time == x` 
+ | ... | any additional arguments to be passed to the event | |
+
+Here is a list of implemented `modifications` (see `?events.modifications`):
+
+ function | arguments | input | what does it do
+---------|-----------|-------|----------------
+random.extinction | x | a numeric value | the proportion of species to make extinct
+trait.extinction | x | a numeric value | the trait value when to trigger the extinction
+ | condition | a relational operator | the relation to the trait value (e.g. `x = 1` and `condition =  >` makes species that have a trait value > 1 go extinct)
+bd.params.update | ... | any argument to be passed to `make.bd.params` | |
+traits.update | ... | any argument to be passed to `make.traits` | |
+modifiers.update | ... | any argument to be passed to `make.modifiers` | |
+
+\* non-optional arguments
 

### 2023-11-21 inst/gitbook/01_getting_started.Rmd
fully updated manual

@@
 Et voila! We now have a distribution of 50 trees and 50 datasets from which we can calculate disparity.
-We can convert this list into `dispRity` using the `treats.dispRity` function.
+We can directly pass the results to the `dispRity` package pipeline using the `dispRitreats` function:
 
-```{r, message = FALSE}
-## Converting the data to be passed to dispRity functions
-dispRity_data <- dispRitreats(sim_data)
-
-## Creating the time series (by default dispRitreats scales the trees to be 1 unit of times old)
-time_series <- chrono.subsets(data = dispRity_data$data,
-                              tree = dispRity_data$tree,
-                              time = seq(from = 0.9, to = 0, by = -0.1),
-                              method = "continuous", model = "proximity")
-
-## Calculating the disparity from the simulated data
-simulated_disparity <- dispRity(time_series, metric = c(sum, variances),
-                                verbose = TRUE)
+```{r, message = FALSE, warning=FALSE}
+## Calculate the dispRity for all the simulations
+simulated_disparity <- dispRitreats(sim_data,
+                                    method = "continuous",
+                                    model  = "proximity",
+                                    time   = 10,
+                                    metric = c(sum, variances),
+                                    scale.trees = TRUE)
 par(mfrow = c(1,2))
@@
 ```{r}
-## Converting the data to be passed to dispRity functions
-rand_ext_data <- dispRitreats(sim_rand_extinction)
-trait_ext_data <- dispRitreats(sim_trait_extinction)
+## Simulate the tree and traits with a random extinction event
+sim_rand_extinction <- treats(
+                   traits     = my_traits,
+                   bd.params  = est_params,
+                   stop.rule  = stop_rule,
+                   events     = random_extinction,
+                   null.error = 100,
+                   replicates = 50)
 
-## Creating the time series
-rand_ext_time_series <- chrono.subsets(data = rand_ext_data$data,
-                                       tree = rand_ext_data$tree,
-                                       time = seq(from = 0.9, to = 0, by = -0.1),
-                                       method = "continuous", model = "proximity")
-trait_ext_time_series <- chrono.subsets(data = trait_ext_data$data,
-                                        tree = trait_ext_data$tree,
-                                        time = seq(from = 0.9, to = 0, by = -0.1),
-                                        method = "continuous", model = "proximity")
+## Simulate the tree and traits with a random extinction event
+sim_trait_extinction <- treats(
+                   traits     = my_traits,
+                   bd.params  = est_params,
+                   stop.rule  = stop_rule,
+                   events     = negative_extinction,
+                   null.error = 100,
+                   replicates = 50)
 
-## Calculating the disparity from the simulated data
-sim_rand_ext_disp <- dispRity(rand_ext_time_series, metric = c(sum, variances),
-                              verbose = TRUE)
-sim_trait_ext_disp <- dispRity(trait_ext_time_series, metric = c(sum, variances),
-                               verbose = TRUE)
-
-par(mfrow = c(2,2))
-plot(change_in_size, main = "observed disparity")
-plot(simulated_disparity, main = "simulated disparity")
-plot(sim_rand_ext_disp, main = "simulated disparity\n(mass extinction)")
-plot(simulated_disparity, main = "simulated disparity\n(trait extinction)")
+## Visualising the difference between both scenarios
+par(mfrow = c(1,2))
+plot(sim_rand_extinction[[1]], main = "Random extinction")
+plot(sim_trait_extinction[[1]], main = "Negative extinction")
 ```
 
-Interestingly, with all our parameters used here and our specific disparity metric we cannot detcet the effect of a mass extinction.
+Interestingly, with all our parameters used here and our specific disparity metric we cannot detect the effect of a mass extinction.

### 2023-11-21 inst/gitbook/_book/treats_manual.tex
fully updated manual

@@
 \author{Thomas Guillerme (\href{mailto:guillert@tcd.ie}{\nolinkurl{guillert@tcd.ie}})}
-\date{2023-05-25}
+\date{2023-11-21}
 
@@
 
+\begin{verbatim}
+##        --- dispRity package ---
+## This is the CRAN release version (1.7.0) of the package.
+## For news, vignettes and future releases,
+## visit https://github.com/TGuillerme/dispRity
+\end{verbatim}
+
 \hypertarget{gettingstarted}{%
@@
 You can find much more about how to design trait objects in the \protect\hyperlink{maketraits}{\texttt{make.traits} section}.
+
+\hypertarget{treats-objects}{%
+\subsection{\texorpdfstring{\texttt{"treats"} objects}{"treats" objects}}\label{treats-objects}}
+
+\texttt{treats} will output either just a tree (class \texttt{"phylo"}) if no traits were generated or a \texttt{"treats"} object that contains both a \texttt{\$tree} (\texttt{"phylo"}) and a \texttt{\$data} (\texttt{"matrix"}) component.
+Note that this \texttt{"treats"} class is generalised to most outputs of the package functions.
+This allows for a smoother handeling of the objects outputs such as summarising the content of a \texttt{make.bd.params} output or visualising a trait output from \texttt{make.traits}.
 
@@
 ## No modifiers applied to the branch length, selection and speciation processes (default).
-## 
 ## Simulated phylogenetic tree (x$tree):
@@
 Note however, that the printing information is now updated to state that you've added a \texttt{"modifiers"} object (even though it's the same as the default).
+In this specific case the \texttt{modifiers} object is not necessary as it is doing nothing different than when using the default one (i.e.~when no \texttt{modifiers} is provided).
 
@@
 For a speciation to occur (and a species to not go extinct), the algorithm draws a random value \emph{x} between 0 and 1 and if the value is smaller than the speciation parameter divided by the speciation + extinction parameter, the lineage speciates (\(x < (\lambda \ (\lambda + \mu))\)), else it goes extinct.
-
-Here we can create a modifier forcing the lineage to go extinct every time a specific condition is met, e.g.~if the lineage trait value is negative.
+\emph{x} is a random value that simulates some stochasticity in the evolutionary simulation.
+At every step of the simulation, a different random value \emph{x} is drawn making sure that the resulting speciation or extinction is not predictable.
+
+However, we can create a modifier that introduces less randomness (or a different level of randomness) by forcing the lineage to go extinct every time a specific condition is met, e.g.~if the lineage trait value is negative.
+This can be done for example by not randomly generating the value \emph{x} but giving it a fixed value of say 1 so that the estimation of the speciation function \(x < (\lambda \ (\lambda + \mu))\) is always met.
 In other words, if the lineage's trait value is negative, make sure that \emph{x} is equal to 1, so that the speciation condition \(x < (\lambda \ (\lambda + \mu))\) is never met.
@@
 
+This \texttt{going.extinct} function is to be contrasted with the normal modifier for selecting the value \texttt{x} which is \texttt{runif(1)} (drawing a random value between 0 and 1).
 We can then provide these two functions (the condition \texttt{negative.ancestor} and how to modify the speciation event when this condition is met \texttt{going.extinct}).
@@
 ## Speciation process is set to speciation with a condition (negative.ancestor) and a modifier (going.extinct).
-## 
 ## Simulated phylogenetic tree (x$tree):
@@
 ##  ---- treats birth-death parameters object ---- 
-## speciation: 0.352330414024189.
-## extinction: 0.287616664509542.
+## speciation: 0.352.
+## extinction: 0.288.
 \end{verbatim}
@@
 Et voila! We now have a distribution of 50 trees and 50 datasets from which we can calculate disparity.
-We can convert this list into \texttt{dispRity} using the \texttt{treats.dispRity} function.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\#\# Converting the data to be passed to dispRity functions}
-\NormalTok{dispRity\_data \textless{}{-}}\StringTok{ }\KeywordTok{dispRitreats}\NormalTok{(sim\_data)}
-
-\CommentTok{\#\# Creating the time series (by default dispRitreats scales the trees to be 1 unit of times old)}
-\NormalTok{time\_series \textless{}{-}}\StringTok{ }\KeywordTok{chrono.subsets}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ dispRity\_data}\OperatorTok{$}\NormalTok{data,}
-                              \DataTypeTok{tree =}\NormalTok{ dispRity\_data}\OperatorTok{$}\NormalTok{tree,}
-                              \DataTypeTok{time =} \KeywordTok{seq}\NormalTok{(}\DataTypeTok{from =} \FloatTok{0.9}\NormalTok{, }\DataTypeTok{to =} \DecValTok{0}\NormalTok{, }\DataTypeTok{by =} \FloatTok{{-}0.1}\NormalTok{),}
-                              \DataTypeTok{method =} \StringTok{"continuous"}\NormalTok{, }\DataTypeTok{model =} \StringTok{"proximity"}\NormalTok{)}
-
-\CommentTok{\#\# Calculating the disparity from the simulated data}
-\NormalTok{simulated\_disparity \textless{}{-}}\StringTok{ }\KeywordTok{dispRity}\NormalTok{(time\_series, }\DataTypeTok{metric =} \KeywordTok{c}\NormalTok{(sum, variances),}
-                                \DataTypeTok{verbose =} \OtherTok{TRUE}\NormalTok{)}
+We can directly pass the results to the \texttt{dispRity} package pipeline using the \texttt{dispRitreats} function:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# Calculate the dispRity for all the simulations}
+\NormalTok{simulated\_disparity \textless{}{-}}\StringTok{ }\KeywordTok{dispRitreats}\NormalTok{(sim\_data,}
+                                    \DataTypeTok{method =} \StringTok{"continuous"}\NormalTok{,}
+                                    \DataTypeTok{model  =} \StringTok{"proximity"}\NormalTok{,}
+                                    \DataTypeTok{time   =} \DecValTok{10}\NormalTok{,}
+                                    \DataTypeTok{metric =} \KeywordTok{c}\NormalTok{(sum, variances),}
+                                    \DataTypeTok{scale.trees =} \OtherTok{TRUE}\NormalTok{)}
 \KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow =} \KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{,}\DecValTok{2}\NormalTok{))}
@@
                                  \DataTypeTok{modification =} \KeywordTok{random.extinction}\NormalTok{(}\FloatTok{0.75}\NormalTok{))}
-\CommentTok{\#\# Creating an extrinction that removes species with negative trait values}
+\CommentTok{\#\# Creating an extinction that removes species with negative trait values}
 \NormalTok{negative\_extinction \textless{}{-}}\StringTok{ }\KeywordTok{make.events}\NormalTok{(}\DataTypeTok{target =} \StringTok{"taxa"}\NormalTok{,}
@@
 \begin{Highlighting}[]
-\CommentTok{\#\# Converting the data to be passed to dispRity functions}
-\NormalTok{rand\_ext\_data \textless{}{-}}\StringTok{ }\KeywordTok{dispRitreats}\NormalTok{(sim\_rand\_extinction)}
-\NormalTok{trait\_ext\_data \textless{}{-}}\StringTok{ }\KeywordTok{dispRitreats}\NormalTok{(sim\_trait\_extinction)}
-
-\CommentTok{\#\# Creating the time series}
-\NormalTok{rand\_ext\_time\_series \textless{}{-}}\StringTok{ }\KeywordTok{chrono.subsets}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ rand\_ext\_data}\OperatorTok{$}\NormalTok{data,}
-                                       \DataTypeTok{tree =}\NormalTok{ rand\_ext\_data}\OperatorTok{$}\NormalTok{tree,}
-                                       \DataTypeTok{time =} \KeywordTok{seq}\NormalTok{(}\DataTypeTok{from =} \FloatTok{0.9}\NormalTok{, }\DataTypeTok{to =} \DecValTok{0}\NormalTok{, }\DataTypeTok{by =} \FloatTok{{-}0.1}\NormalTok{),}
-                                       \DataTypeTok{method =} \StringTok{"continuous"}\NormalTok{, }\DataTypeTok{model =} \StringTok{"proximity"}\NormalTok{)}
-\NormalTok{trait\_ext\_time\_series \textless{}{-}}\StringTok{ }\KeywordTok{chrono.subsets}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ trait\_ext\_data}\OperatorTok{$}\NormalTok{data,}
-                                        \DataTypeTok{tree =}\NormalTok{ trait\_ext\_data}\OperatorTok{$}\NormalTok{tree,}
-                                        \DataTypeTok{time =} \KeywordTok{seq}\NormalTok{(}\DataTypeTok{from =} \FloatTok{0.9}\NormalTok{, }\DataTypeTok{to =} \DecValTok{0}\NormalTok{, }\DataTypeTok{by =} \FloatTok{{-}0.1}\NormalTok{),}
-                                        \DataTypeTok{method =} \StringTok{"continuous"}\NormalTok{, }\DataTypeTok{model =} \StringTok{"proximity"}\NormalTok{)}
-
-\CommentTok{\#\# Calculating the disparity from the simulated data}
-\NormalTok{sim\_rand\_ext\_disp \textless{}{-}}\StringTok{ }\KeywordTok{dispRity}\NormalTok{(rand\_ext\_time\_series, }\DataTypeTok{metric =} \KeywordTok{c}\NormalTok{(sum, variances),}
-                              \DataTypeTok{verbose =} \OtherTok{TRUE}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## Calculating disparity..........Done.
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{sim\_trait\_ext\_disp \textless{}{-}}\StringTok{ }\KeywordTok{dispRity}\NormalTok{(trait\_ext\_time\_series, }\DataTypeTok{metric =} \KeywordTok{c}\NormalTok{(sum, variances),}
-                               \DataTypeTok{verbose =} \OtherTok{TRUE}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## Calculating disparity..........Done.
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow =} \KeywordTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{,}\DecValTok{2}\NormalTok{))}
-\KeywordTok{plot}\NormalTok{(change\_in\_size, }\DataTypeTok{main =} \StringTok{"observed disparity"}\NormalTok{)}
-\KeywordTok{plot}\NormalTok{(simulated\_disparity, }\DataTypeTok{main =} \StringTok{"simulated disparity"}\NormalTok{)}
-\KeywordTok{plot}\NormalTok{(sim\_rand\_ext\_disp, }\DataTypeTok{main =} \StringTok{"simulated disparity}\CharTok{\textbackslash{}n}\StringTok{(mass extinction)"}\NormalTok{)}
-\KeywordTok{plot}\NormalTok{(simulated\_disparity, }\DataTypeTok{main =} \StringTok{"simulated disparity}\CharTok{\textbackslash{}n}\StringTok{(trait extinction)"}\NormalTok{)}
+\CommentTok{\#\# Simulate the tree and traits with a random extinction event}
+\NormalTok{sim\_rand\_extinction \textless{}{-}}\StringTok{ }\KeywordTok{treats}\NormalTok{(}
+                   \DataTypeTok{traits     =}\NormalTok{ my\_traits,}
+                   \DataTypeTok{bd.params  =}\NormalTok{ est\_params,}
+                   \DataTypeTok{stop.rule  =}\NormalTok{ stop\_rule,}
+                   \DataTypeTok{events     =}\NormalTok{ random\_extinction,}
+                   \DataTypeTok{null.error =} \DecValTok{100}\NormalTok{,}
+                   \DataTypeTok{replicates =} \DecValTok{50}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## Building the tree:....Done.
+## Building the tree:.......Done.
+## Building the tree:...Done.
+## Building the tree:......Done.
+## Building the tree:..Done.
+## Building the tree:..Done.
+## Building the tree:.Done.
+## Building the tree:........Done.
+## Building the tree:..Done.
+## Building the tree:.....Done.
+## Building the tree:....Done.
+## Building the tree:....Done.
+## Building the tree:..Done.
+## Building the tree:.................Done.
+## Building the tree:....Done.
+## Building the tree:....Done.
+## Building the tree:..Done.
+## Building the tree:......Done.
+## Building the tree:.......Done.
+## Building the tree:.Done.
+## Building the tree:.Done.
+## Building the tree:..Done.
+## Building the tree:......Done.
+## Building the tree:...........Done.
+## Building the tree:......Done.
+## Building the tree:....Done.
+## Building the tree:.Done.
+## Building the tree:..................Done.
+## Building the tree:....Done.
+## Building the tree:...............Done.
+## Building the tree:....Done.
+## Building the tree:..Done.
+## Building the tree:........Done.
+## Building the tree:....Done.
+## Building the tree:........Done.
+## Building the tree:...Done.
+## Building the tree:...Done.
+## Building the tree:.Done.
+## Building the tree:...Done.
+## Building the tree:..............Done.
+## Building the tree:...Done.
+## Building the tree:.Done.
+## Building the tree:..Done.
+## Building the tree:....Done.
+## Building the tree:..Done.
+## Building the tree:..Done.
+## Building the tree:.Done.
+## Building the tree:.Done.
+## Building the tree:....Done.
+## Building the tree:..........Done.
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# Simulate the tree and traits with a random extinction event}
+\NormalTok{sim\_trait\_extinction \textless{}{-}}\StringTok{ }\KeywordTok{treats}\NormalTok{(}
+                   \DataTypeTok{traits     =}\NormalTok{ my\_traits,}
+                   \DataTypeTok{bd.params  =}\NormalTok{ est\_params,}
+                   \DataTypeTok{stop.rule  =}\NormalTok{ stop\_rule,}
+                   \DataTypeTok{events     =}\NormalTok{ negative\_extinction,}
+                   \DataTypeTok{null.error =} \DecValTok{100}\NormalTok{,}
+                   \DataTypeTok{replicates =} \DecValTok{50}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## Building the tree:.....................Done.
+## Building the tree:...................Done.
+## Building the tree:.....Done.
+## Building the tree:.Done.
+## Building the tree:....Done.
+## Building the tree:..Done.
+## Building the tree:..Done.
+## Building the tree:......................Done.
+## Building the tree:..Done.
+## Building the tree:..Done.
+## Building the tree:...Done.
+## Building the tree:.Done.
+## Building the tree:..Done.
+## Building the tree:......Done.
+## Building the tree:.Done.
+## Building the tree:......Done.
+## Building the tree:..Done.
+## Building the tree:.Done.
+## Building the tree:....Done.
+## Building the tree:...Done.
+## Building the tree:..Done.
+## Building the tree:......Done.
+## Building the tree:....................Done.
+## Building the tree:...Done.
+## Building the tree:.........Done.
+## Building the tree:.Done.
+## Building the tree:...Done.
+## Building the tree:.Done.
+## Building the tree:.........Done.
+## Building the tree:.Done.
+## Building the tree:.............Done.
+## Building the tree:..............Done.
+## Building the tree:......Done.
+## Building the tree:.Done.
+## Building the tree:.......Done.
+## Building the tree:...Done.
+## Building the tree:..Done.
+## Building the tree:......Done.
+## Building the tree:.Done.
+## Building the tree:..Done.
+## Building the tree:............Done.
+## Building the tree:.....Done.
+## Building the tree:......Done.
+## Building the tree:.Done.
+## Building the tree:.Done.
+## Building the tree:..............Done.
+## Building the tree:........Done.
+## Building the tree:..............Done.
+## Building the tree:.Done.
+## Building the tree:..Done.
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# Visualising the difference between both scenarios}
+\KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow =} \KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{,}\DecValTok{2}\NormalTok{))}
+\KeywordTok{plot}\NormalTok{(sim\_rand\_extinction[[}\DecValTok{1}\NormalTok{]], }\DataTypeTok{main =} \StringTok{"Random extinction"}\NormalTok{)}
+\KeywordTok{plot}\NormalTok{(sim\_trait\_extinction[[}\DecValTok{1}\NormalTok{]], }\DataTypeTok{main =} \StringTok{"Negative extinction"}\NormalTok{)}
 \end{Highlighting}
@@
 
-Interestingly, with all our parameters used here and our specific disparity metric we cannot detcet the effect of a mass extinction.
+Interestingly, with all our parameters used here and our specific disparity metric we cannot detect the effect of a mass extinction.
 
@@
 \end{figure}
+
+\hypertarget{quick-overview}{%
+\subsection{Quick overview}\label{quick-overview}}
+
+\begin{longtable}[]{@{}llll@{}}
+\toprule
+\begin{minipage}[b]{0.19\columnwidth}\raggedright
+function\strut
+\end{minipage} & \begin{minipage}[b]{0.23\columnwidth}\raggedright
+arguments\strut
+\end{minipage} & \begin{minipage}[b]{0.14\columnwidth}\raggedright
+input\strut
+\end{minipage} & \begin{minipage}[b]{0.33\columnwidth}\raggedright
+what does it do\strut
+\end{minipage}\tabularnewline
+\midrule
+\endhead
+\begin{minipage}[t]{0.19\columnwidth}\raggedright
+\texttt{make.traits}\strut
+\end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
+process*\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
+a function\strut
+\end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
+simulates the trait (see below)\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.19\columnwidth}\raggedright
+n\strut
+\end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
+a numeric value\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
+the number of dimensions\strut
+\end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
+\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.19\columnwidth}\raggedright
+start\strut
+\end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
+one or more numeric values\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
+the starting values for each dimensions\strut
+\end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
+\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.19\columnwidth}\raggedright
+process.args\strut
+\end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
+\ldots{}\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
+anything to be passed to \texttt{process} (see below)\strut
+\end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
+\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.19\columnwidth}\raggedright
+trait.names\strut
+\end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
+a character string\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
+the name(s) of the trait(s)\strut
+\end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
+\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.19\columnwidth}\raggedright
+add\strut
+\end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
+a \texttt{treats} object\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
+another trait generated by \texttt{make.traits}\strut
+\end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
+\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.19\columnwidth}\raggedright
+update\strut
+\end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
+a \texttt{treats} object\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
+another trait generated by \texttt{make.traits}\strut
+\end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
+\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.19\columnwidth}\raggedright
+test\strut
+\end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
+logical\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
+whether to test the validity of the trait\strut
+\end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
+\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.19\columnwidth}\raggedright
+background\strut
+\end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
+a \texttt{treats} object\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
+another trait generated by \texttt{make.traits} (to run in the background)\strut
+\end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
+\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.19\columnwidth}\raggedright
+---------\strut
+\end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
+-----------\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
+-------\strut
+\end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
+----------------\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.19\columnwidth}\raggedright
+\texttt{process}\strut
+\end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
+x0\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
+a numeric value\strut
+\end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
+the value of the previous trait\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.19\columnwidth}\raggedright
+edge.length\strut
+\end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
+a numeric value\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
+the branch length value\strut
+\end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
+\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.19\columnwidth}\raggedright
+\ldots{}\strut
+\end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
+\ldots{}\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
+any other argument to be handled by the process\strut
+\end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
+\strut
+\end{minipage}\tabularnewline
+\bottomrule
+\end{longtable}
+
+* non-optional arguments
 
@@
 ## }
-## <bytecode: 0x55b3b2541d88>
+## <bytecode: 0x5647036c6610>
 ## <environment: namespace:treats>
@@
 
+\hypertarget{quick-overview-1}{%
+\subsection{Quick overview}\label{quick-overview-1}}
+
+\begin{longtable}[]{@{}llll@{}}
+\toprule
+\begin{minipage}[b]{0.19\columnwidth}\raggedright
+function\strut
+\end{minipage} & \begin{minipage}[b]{0.23\columnwidth}\raggedright
+arguments\strut
+\end{minipage} & \begin{minipage}[b]{0.14\columnwidth}\raggedright
+input\strut
+\end{minipage} & \begin{minipage}[b]{0.33\columnwidth}\raggedright
+what does it do\strut
+\end{minipage}\tabularnewline
+\midrule
+\endhead
+\begin{minipage}[t]{0.19\columnwidth}\raggedright
+\texttt{make.modifiers}\strut
+\end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
+branch.length\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
+a function\strut
+\end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
+generating branch length (see below)\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.19\columnwidth}\raggedright
+selection\strut
+\end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
+a function\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
+selecting a lineage (see below)\strut
+\end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
+\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.19\columnwidth}\raggedright
+speciation\strut
+\end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
+a function\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
+choosing between speciation and extinction (see below)\strut
+\end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
+\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.19\columnwidth}\raggedright
+condition\strut
+\end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
+a function\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
+a function that affects the condition of when to trigger the functions above (see below)\strut
+\end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
+\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.19\columnwidth}\raggedright
+modify\strut
+\end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
+a function\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
+a function that modifies the value of the trigger value generated when the \texttt{condition} is met (see below)\strut
+\end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
+\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.19\columnwidth}\raggedright
+add\strut
+\end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
+a \texttt{treats} object\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
+another modifier generated by \texttt{make.modifiers}\strut
+\end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
+\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.19\columnwidth}\raggedright
+update\strut
+\end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
+a \texttt{treats} object\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
+another modifier generated by \texttt{make.modifiers}\strut
+\end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
+\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.19\columnwidth}\raggedright
+test\strut
+\end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
+logical\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
+whether to test the validity of the modifier\strut
+\end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
+\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.19\columnwidth}\raggedright
+---------\strut
+\end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
+-----------\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
+-------\strut
+\end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
+----------------\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.19\columnwidth}\raggedright
+\texttt{branch.length}, \texttt{selection}, \texttt{speciation}, \texttt{condition}\strut
+\end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
+bd.params\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
+a \texttt{treats} object generated by \texttt{make.bd.params}\strut
+\end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
+usually handled internally (contains the either the birth-death values or their distribution)\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.19\columnwidth}\raggedright
+lineage\strut
+\end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
+a named ``lineage'' list\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
+usually handled internally (contains the current state of the simulations)\strut
+\end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
+\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.19\columnwidth}\raggedright
+trait.values\strut
+\end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
+a matrix\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
+usually handled internally (contains the current generated traits of the simulations)\strut
+\end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
+\strut
+\end{minipage}\tabularnewline
+\bottomrule
+\end{longtable}
+
 \hypertarget{the-default-modifier-how-the-process-is-working}{%
@@
 ## }
-## <bytecode: 0x55b3b0f93c78>
+## <bytecode: 0x564707510428>
 ## <environment: namespace:treats>
@@
 ## }
-## <bytecode: 0x55b3b0f9bdc8>
+## <bytecode: 0x5647075127a8>
 ## <environment: namespace:treats>
@@
 ## {
-##     return(runif(1) < (bd.params$speciation/(bd.params$speciation + 
+##     return(runif(1) <= (bd.params$speciation/(bd.params$speciation + 
 ##         bd.params$extinction)))
 ## }
-## <bytecode: 0x55b3b0f9b498>
+## <bytecode: 0x56470751e620>
 ## <environment: namespace:treats>
@@
 
-For example in the \texttt{speciation} part of a modifier, the default is to trigger an event from a uniform distribution and then check if that value is smaller than (\[speciation/(speciation + extinction)\]):
+For example in the \texttt{speciation} part of a modifier, the default is to trigger an event from a uniform distribution and then check if that value is smaller than (\(speciation/(speciation + extinction)\)):
 
@@
 \end{figure}
+
+\hypertarget{quick-overview-2}{%
+\subsection{Quick overview}\label{quick-overview-2}}
+
+condition,
+modification,
+add,
+test = TRUE,
+event.name,
+replications = 0,
+additional.args
+)
+
+\begin{longtable}[]{@{}llll@{}}
+\toprule
+\begin{minipage}[b]{0.19\columnwidth}\raggedright
+function\strut
+\end{minipage} & \begin{minipage}[b]{0.23\columnwidth}\raggedright
+arguments\strut
+\end{minipage} & \begin{minipage}[b]{0.14\columnwidth}\raggedright
+input\strut
+\end{minipage} & \begin{minipage}[b]{0.33\columnwidth}\raggedright
+what does it do\strut
+\end{minipage}\tabularnewline
+\midrule
+\endhead
+\begin{minipage}[t]{0.19\columnwidth}\raggedright
+\texttt{make.events}\strut
+\end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
+target*\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
+any of the following: ``taxa'', ``bd.params'', ``traits'', ``modifiers'' or ``founding''\strut
+\end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
+what aspect of the simulation should the event target\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.19\columnwidth}\raggedright
+condition*\strut
+\end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
+a function\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
+a function that affects the condition of when to trigger the event (see below)\strut
+\end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
+\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.19\columnwidth}\raggedright
+modification*\strut
+\end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
+a function\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
+a function that applies the modification to the target if the condition is triggered (see below)\strut
+\end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
+\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.19\columnwidth}\raggedright
+add\strut
+\end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
+a \texttt{treats} object\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
+another event generated by \texttt{make.events}\strut
+\end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
+\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.19\columnwidth}\raggedright
+test\strut
+\end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
+logical\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
+whether to test the validity of the event\strut
+\end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
+\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.19\columnwidth}\raggedright
+event.name\strut
+\end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
+a character string\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
+the name of the event\strut
+\end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
+\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.19\columnwidth}\raggedright
+replications\strut
+\end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
+a number\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
+if possible how many times is the event allowing to be repeated\strut
+\end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
+\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.19\columnwidth}\raggedright
+additional.args\strut
+\end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
+\ldots{}\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
+any additional arguments to be passed to the event\strut
+\end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
+\strut
+\end{minipage}\tabularnewline
+\bottomrule
+\end{longtable}
+
+Here is a list of implemented \texttt{conditions} (see \texttt{?events.conditions}):
+
+\begin{longtable}[]{@{}llll@{}}
+\toprule
+\begin{minipage}[b]{0.19\columnwidth}\raggedright
+function\strut
+\end{minipage} & \begin{minipage}[b]{0.23\columnwidth}\raggedright
+arguments\strut
+\end{minipage} & \begin{minipage}[b]{0.14\columnwidth}\raggedright
+input\strut
+\end{minipage} & \begin{minipage}[b]{0.33\columnwidth}\raggedright
+what does it do\strut
+\end{minipage}\tabularnewline
+\midrule
+\endhead
+\begin{minipage}[t]{0.19\columnwidth}\raggedright
+\texttt{time.condition}, \texttt{taxa.condition}, \texttt{trait.condition}\strut
+\end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
+x\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
+the value to check the target against\strut
+\end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
+e.g.~time = x, taxa = x or trait = x (see below)\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.19\columnwidth}\raggedright
+condition\strut
+\end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
+a relational operator\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
+e.g.~\texttt{==}, \texttt{\textgreater{}}, \texttt{!=}, etc. for asking for example \texttt{time\ ==\ x}\strut
+\end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
+\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.19\columnwidth}\raggedright
+\ldots{}\strut
+\end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
+any additional arguments to be passed to the event\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
+\strut
+\end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
+\strut
+\end{minipage}\tabularnewline
+\bottomrule
+\end{longtable}
+
+Here is a list of implemented \texttt{modifications} (see \texttt{?events.modifications}):
+
+\begin{longtable}[]{@{}llll@{}}
+\toprule
+\begin{minipage}[b]{0.19\columnwidth}\raggedright
+function\strut
+\end{minipage} & \begin{minipage}[b]{0.23\columnwidth}\raggedright
+arguments\strut
+\end{minipage} & \begin{minipage}[b]{0.14\columnwidth}\raggedright
+input\strut
+\end{minipage} & \begin{minipage}[b]{0.33\columnwidth}\raggedright
+what does it do\strut
+\end{minipage}\tabularnewline
+\midrule
+\endhead
+\begin{minipage}[t]{0.19\columnwidth}\raggedright
+random.extinction\strut
+\end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
+x\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
+a numeric value\strut
+\end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
+the proportion of species to make extinct\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.19\columnwidth}\raggedright
+trait.extinction\strut
+\end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
+x\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
+a numeric value\strut
+\end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
+the trait value when to trigger the extinction\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.19\columnwidth}\raggedright
+condition\strut
+\end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
+a relational operator\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
+the relation to the trait value (e.g.~\texttt{x\ =\ 1} and \texttt{condition\ =\ \ \textgreater{}} makes species that have a trait value \textgreater{} 1 go extinct)\strut
+\end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
+\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.19\columnwidth}\raggedright
+bd.params.update\strut
+\end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
+\ldots{}\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
+any argument to be passed to \texttt{make.bd.params}\strut
+\end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
+\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.19\columnwidth}\raggedright
+traits.update\strut
+\end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
+\ldots{}\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
+any argument to be passed to \texttt{make.traits}\strut
+\end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
+\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.19\columnwidth}\raggedright
+modifiers.update\strut
+\end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
+\ldots{}\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
+any argument to be passed to \texttt{make.modifiers}\strut
+\end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
+\strut
+\end{minipage}\tabularnewline
+\bottomrule
+\end{longtable}
+
+* non-optional arguments
 
@@
 
-\includegraphics[width=6.5in]{../../../../../../tmp/RtmpXnwk90/fileb7e05f6d37cf}
+\includegraphics[width=6.5in]{../../../../../../tmp/RtmptzertD/file14f8c6a285c1a}
 
@@
 
-\includegraphics[width=6.5in]{../../../../../../tmp/RtmpXnwk90/fileb7e060df376f}
+\includegraphics[width=6.5in]{../../../../../../tmp/RtmptzertD/file14f8c301a0acb}
 
@@
 
-\includegraphics[width=6.5in]{../../../../../../tmp/RtmpXnwk90/fileb7e0d32ccc1}
+\includegraphics[width=6.5in]{../../../../../../tmp/RtmptzertD/file14f8c465f7d38}
 
 \hypertarget{EG_change_modif}{%
-\section{Event for changing a modifier: speciation event increase for species with negative values}\label{EG_change_modif}}
+\section{Event for changing a modifier: extinction event increase for species with negative values}\label{EG_change_modif}}
 

### 2023-11-21 paper/response.Rmd
response done (to check)

@@
 I've highlight <span style="color:blue">the Associate Editor and both reviewers comments in blue</span> and my response black.
+I'm also sorry for the delay in dealing with the reviewers comments: I was unemployed until November so didn't work on it until then.
 
@@
 
-> TODO: make a table comparing the package to other ones in terms of computing times.
-
-I tried (now in tests/) but I'm not convinced how useful it is.
+I tried (and saved the tests in `tests/test-benchmarking.R` for people to explore) but I'm not convinced how useful it is.
 First I feel it's confrontational to the other authors since none of them had the same objectives in mind.
@@
 
-> TODO: implement that in issues on the GitHub page with a link to all the possibilities
+I am happy that both reviewers are enthusiastic about the templates library. 
+Unfortunately I'm not totally sure how to implement it and make it easier to use and browse by users.
+So for now I've set up a templated category of issues on github called [simulation templates](https://github.com/TGuillerme/treats/issues?q=is%3Aopen+is%3Aissue+label%3A%22simulation+template%22).
+I've updated the README file with some indications on how to save and share such a template.
 
@@
 
-> TODO: make a cheat sheet page with all core functions, their arguments and what they do.
+I've now added a cheat sheet to the github page that shows the key arguments and their inputs/outputs [here](https://github.com/TGuillerme/treats/blob/master/inst/gitbook/treats_cheat_sheet.pdf).
+This cheat sheet is rather minimal for now but I'm planning on expanding on it and making it more interactive (with mouse over menus, etc.) in the future.
+Furthermore I've tidied up the summarising of functions in the manual as suggested by the reviewer.
 
@@
 
-> TODO: specify this in the manual and the documentation: `modification` = modify the bd-algorithm; `events` (conditions and stuff) = modify the flow of the algorithm (but can also introduce a `modification`!)
+I have now added clarifications of what the `"modifiers"` are and how they should be used. Both in the section 2.3 as a simple explanation and in the section 4 dedicated to `"modifiers"`.
 
@@
 
-> TODO: add in the documentation a brief description of the outputs of the `make.*` functions and of `treats`.
-Additionally, add a detailed `.md` with the developer level content of each output (`dispRity` style).
+I've added a `@return` section to the `treats`, `make.treats`, `make.bd.params`, `make.traits`, `make.modifiers` and `make.events` documentations as well as the following section in the manual "Getting started" section:
+
+*`treats` will output either just a tree (class `"phylo"`) if no traits were generated or a `"treats"` object that contains both a `$tree` (`"phylo"`) and a `$data` (`"matrix"`) component.
+Note that this `"treats"` class is generalised to most outputs of the package functions.
+This allows for a smoother handeling of the objects outputs such as summarising the content of a `make.bd.params` output or visualising a trait output from `make.traits`.*
 
@@
 
-> TODO: implement this as issues.
+As mentioned before, I was not sure how to effectively implement that so I made a [simulation templates](https://github.com/TGuillerme/treats/issues?q=is%3Aopen+is%3Aissue+label%3A%22simulation+template%22) issues category on github for now.
 

### 2023-11-22 paper/response.Rmd
Updated paper and response

@@
 
-*4. Generating some trait(s) value(s) for the selected lineage (either a tip or a node - but see the "Simulating traits section" below to generate trait values along edges).* line @@@
+*4. Generating some trait(s) value(s) for the selected lineage (either a tip or a node - but see the "Simulating traits section" below to generate trait values along edges).* line 78-79.
 
@@
 This will generate singleton nodes (i.e.
-nodes with only one descendant) with associated trait values."* line @@@
+nodes with only one descendant) with associated trait values."* line 110-112.
 
@@
 This is now updated in the example.
-I've updated the examples on lines @@@ to include these changes.
 
@@
 
-I've changed the comment to `## Simulate the tree and traits with a selective extinction event`. line @@@
+I've changed the comment to `## Simulate the tree and traits with a selective extinction event`. code block after line 203.
 
@@
 
-I've added references to figures 5 and 6 in the text on lines @@@ and @@@.
+I've added references to figures 5 and 6 in the text on lines 203 and 207.
 
@@
 half the species are expected to evolve a trait value above 0).
-For more details on simulating the effect of mass extinction and the difficulties to simulate an unambiguous effect of a mass extinction, see @puttick2020complex."* lines @@@
+For more details on simulating the effect of mass extinction and the difficulties to simulate an unambiguous effect of a mass extinction, see Puttick, Guillerme, and Wills (2020)."* lines 195-199.
 
@@
 
-*"Note that although `treats` is modular and thus allows to be used as go to tool for simulating and trees and traits, it lacks the ready-to-use implemented methods featured in other packages such as fossilisation and sampling [@fossilsim, @treesim, @paleobuddy] or specific macroevolutionary simulations [@RPANDA, @puttick2020complex].
-"*. line @@@
+*"Note that although `treats` is modular and thus allows to be used as go to tool for simulating and trees and traits, it lacks the ready-to-use implemented methods featured in other packages such as fossilisation and sampling [@fossilsim; @treesim; @paleobuddy] or specific macroevolutionary simulations [@rpanda; @puttick2020complex].
+"*. line 63-67
 
@@
 *"For example future planned versions will include abiotic events and a better integration with the `dispRity` package.
-"*. line @@@
+"*. line 218-219
 
@@
 *"The number of time units in `treats` is arbitrary and is not equivalent to millions of years.
-Using 30 time units here allows to simulate number of tips in a similar order of magnitude as the ones in the observed data."* lines @@@
+Using 30 time units here allows to simulate number of tips in a similar order of magnitude as the ones in the observed data."* lines 181-183.
 
@@
 
-Fixed. [Commit 5f7fb23](https://github.com/TGuillerme/treats/commit/5f7fb235e0b56a097b9d888790028ad41f86c33b)
+Fixed.
 
@@
 
-Done. [Commit 5f7fb23](https://github.com/TGuillerme/treats/commit/5f7fb235e0b56a097b9d888790028ad41f86c33b)
+Done.
 
@@
 I've added the following sentence at the end of the paragraph:
-"*Because of the stochasticity of the simulations, we will repeat them 50 times (using `replicates = 50`) to generate a distribution of possible simulated scenarios as opposed to a random single one that could be idiosyncratic.*" [Commit 5f7fb23](https://github.com/TGuillerme/treats/commit/5f7fb235e0b56a097b9d888790028ad41f86c33b)
+"*Because of the stochasticity of the simulations, we will repeat them 50 times (using `replicates = 50`) to generate a distribution of possible simulated scenarios as opposed to a random single one that could be idiosyncratic.*" lines 200-203
 
@@
 
-Done. line @@@
+Done. line 207.
 
@@
 I also noticed that the documentation for the treats function lists the wrong URL for the online manual.</span>
-I've added the URL spelt out. [Commit 5f7fb23](https://github.com/TGuillerme/treats/commit/5f7fb235e0b56a097b9d888790028ad41f86c33b).
+
+I've added the URL spelt out.
 

### 2023-11-22 paper/response_coloured.tex
Updated paper and response

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{xcolor}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Response to reviewers},
  pdfauthor={Thomas Guillerme},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newenvironment{cslreferences}%
  {\setlength{\parindent}{0pt}%
  \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces}%
  {\par}

\title{Response to reviewers}
\author{Thomas Guillerme}
\date{2023-11-22}

\begin{document}
\maketitle

Dear Editor,

I am grateful for the associate editor and the two reviewers time and
very encouraging comments that I totally believe have improved the
package, the manual and this manuscript. Please find my response to the
reviewers comments in the following document. I've highlight \textcolor{blue}{the
Associate Editor and both reviewers comments in blue} and my response
black. I'm also sorry for the delay in dealing with the reviewers
comments: I was unemployed until November so didn't work on it until
then.

Best regards,

Thomas

\hypertarget{associate-editor-comments-to-author}{%
\section{Associate Editor Comments to
Author:}\label{associate-editor-comments-to-author}}

\textcolor{blue}{As both reviewers, I really liked the idea of the package, but also
found the manuscript itself a bit hard to follow, which, at least to me,
made it hard to fully appreciate what ``treats'' can currently do and
what are the future additions that might be easily implemented (either
by the developer, or by users). Reading the online manual, I got a
better idea, and hence I think there is still room (metaphorically and I
suspect also physically) for improvement. Sure, the online material has
no space limit constraints, but I do think the manuscript text could be
improved and both reviewers offered some nice suggestions. Apart from
the really good suggestions given by both reviewers I would add the
following suggestions/comments:}

\begin{itemize}
\tightlist
\item
  \textcolor{blue}{1- Are the trait values ``stored'' only at the node and tip (extant
  or extinct), or does it store (e.g.~assuming some small-time
  intervals) the trait values throughout the branch length of each
  lineage? Based on lines 63 to 65 (and 71) I think only at nodes and
  tips, but it might be worth explicitly saying that. Perhaps add the
  word ``only'' in that line. Also, in line 71 is there an extra ``no''
  within the parenthesis?}
\end{itemize}

The trait data in \texttt{treats} is only stored at nodes and tips due
to the code architecture. However, using the argument
\texttt{save.steps} allows to create singleton nodes (nodes with one
ancestor and only one descendant) that can be generated at some
specified or regular time intervals thus effectively saving trait values
as the simulation goes. I've now specified this in the main text (and
removed the typo):

\emph{4. Generating some trait(s) value(s) for the selected lineage
(either a tip or a node - but see the ``Simulating traits section''
below to generate trait values along edges).} line 78-79.

and:

\emph{``Traits are always generated (and stored) only for tips or nodes
and not along edges. However, it is possible so generate trait values at
specific or regular time intervals by using the option
\texttt{save.steps} in the \texttt{treats} function. This will generate
singleton nodes (i.e.~nodes with only one descendant) with associated
trait values.''} line 110-112.

\begin{itemize}
\tightlist
\item
  \textcolor{blue}{2- Regarding the ``storage'' of trait value, if it only stores the
  values at nodes and tips, does it also mean that it only uses trait
  values at those time points to direct the simulation? Does that
  matter? What made me wonder about that was figure 5, and its
  respective intended selective extinction simulation scenario. If I got
  it right, the selective extinction scenario would remove all species
  with positive trait values at the mass extinction event at 15 Mya.
  Looking at the right panel, all species whose direct ancestral node
  had a positive trait value seemed to be removed, but some of those
  with ancestral negative values likely crossed the red line having a
  positive value (even if the ``true trait trajectory'' is not a
  straight as represented, for simplicity I guess, in the figure) and
  were not removed. This might not be always relevant, and I understand
  that some simplifications are necessary, but it might be worth
  explicitly discussing if this is relevant and when. For example, if
  speciation and background extinction rates are small, species are
  likely to have long branches and the trait value it might have at a
  given time point between its ancestral node and its tip value or next
  speciation node might be very different from the value it had at the
  ancestral node (or tip value for that matter). Or is the decision to
  remove the species in the selective mass extinction event one where
  both the ancestral and tip value have to be positive. I guess some
  extra information on how it chooses the species based on the trait
  value (is it the ancestral node value, the tip value, both?) in the
  selective mass extinction regime would also be helpful. I guess the
  issue seems that the trait value is not measured at short time
  intervals (only at the nodes and tips) but a selective extinction
  regime such as this one presented here is a point in time event. }
\end{itemize}

This is an excellent suggestion. I've now reworked \texttt{events} to
now always generates singleton nodes (and associated trait values if
needed) at the time of the event before applying the modification. This
way it will generate extinction based at the trait value of the
extinction time rather than the latest node as it should be the case. It
is then possible to remove the generated singleton nodes to clean up the
tree (to have just data for bifurcating nodes and tips). This is now
updated in the example.

\begin{itemize}
\tightlist
\item
  \textcolor{blue}{3- I have to say that looking at figure 5 it was not trivial to see
  the difference between the two scenarios (I guess that happened
  because most, if not all, species with positive values went extinct in
  the left random panel as well). I know those are realizations of a
  stochastic process and there is nothing wrong here, my concern regards
  simply the visualization. I wonder if choosing another seed might
  result in a set of simulations that are more clearly different, making
  the comparison easier for the eye to pick it up. For example, in the
  online manual, some examples comparing random to selective simulations
  showed a much more disparate pattern.}
\end{itemize}

I've changed the seeds for the plotted random/non-random extinctions
that illustrates the selectivity more. I've also added a horizontal line
at the value 0 to show the positivity of the extinction event for the
non-random one.

\begin{itemize}
\tightlist
\item
  \textcolor{blue}{4- Code box on page 7: the commentary on the second simulation
  scenario reads ``\#Simulate the tree and traits with a random
  extinction event'' and hence is equal to the comment in the first
  simulation scenario in the same box. Perhaps change it to something
  like: ``\#Simulate the tree and traits with a random extinction event
  but based on trait value''}
\end{itemize}

I've changed the comment to
\texttt{\#\#\ Simulate\ the\ tree\ and\ traits\ with\ a\ selective\ extinction\ event}.
code block after line 203.

\begin{itemize}
\tightlist
\item
  \textcolor{blue}{5- I could be wrong but it seems that figure 6 is not cited in the
  text. Please double check and if this is true, please cite this
  figure. }
\end{itemize}

I've added references to figures 5 and 6 in the text on lines 203 and
207.

\begin{itemize}
\tightlist
\item
  \textcolor{blue}{6- All figure captions have some repetition like ``Figure 1: Figure
  1:\ldots.''}
\end{itemize}

I've removed the repetitions.

\begin{itemize}
\tightlist
\item
  \textcolor{blue}{7- Figure 6: I could be wrong, but my intuition was that a selective
  mass extinction event would decrease the disparity more strongly than
  a random extinction mass extinction event. That does not seem to be
  the case. Is it because the ``rule'' for the second extinction
  scenario was that only positive species will go extinct, hence on
  average 50\% of the species (if you start with a value of zero)
  instead of 75\% went extinct? That would make sense, I guess. Or is
  the second extinction scenario one where you also create a 75\% loss
  of species but guarantee that all species with positive values will go
  extinct? Maybe a few words to better explain this would be helpful.
  Perhaps add on lines 109 to 110 something like: ``note that in the
  second scenario, because the ancestral state is zero and we do not set
  the percentage of species to be removed, we expect only 50\% of the
  species to go extinct''. Just an idea.}
\end{itemize}

The results seems a bit different now having chosen a different seed.
However, I have added the Associate Editor's judicious comment to the
main text:

\emph{``Both scenarios illustrate two different types of mass
extinctions but they are not equivalent: because of the ancestral trait
value starting at 0, we expect the second scenario to remove on average
only 50\% of the species (i.e.~half the species are expected to evolve a
trait value above 0). For more details on simulating the effect of mass
extinction and the difficulties to simulate an unambiguous effect of a
mass extinction, see Puttick, Guillerme, and Wills (2020).''} lines
195-199.

\hypertarget{reviewer-1}{%
\section{Reviewer: 1}\label{reviewer-1}}

\textcolor{blue}{This article by Guillerme introduces a new R package treats, which
implements a flexible framework for simulating phylogenies and
associated trait values. Overall, I really like the concept of the
package, and I think it can be very useful for more advanced users who
run many simulations and change their setup often. My comments are
mostly about clarifying the presentation of the different functions. I
also reviewed the manual.}

\hypertarget{comments-on-the-manuscript}{%
\subsection{Comments on the
manuscript}\label{comments-on-the-manuscript}}

\begin{itemize}
\tightlist
\item
  \textcolor{blue}{I had trouble understanding what treats could and could not do from
  the manuscript alone. The introduction discusses several packages
  related to fossil data (FossilSim, paleotree and paleobuddy) which
  gave me the impression that treats could replace them, but it doesn't
  seem that fossilization processes are currently included ? I was also
  a bit confused by the inclusion of time-dependent and
  lineage-dependent simulations in future features, as it seems that the
  current framework should be able to handle these, at least within the
  birth-death process. Overall, having a clearer list of what is
  currently possible or not in treats would be helpful.}
\end{itemize}

I've now clarified what \texttt{treats} does and doesn't in the
introduction:

\emph{``Note that although \texttt{treats} is modular and thus allows to
be used as go to tool for simulating and trees and traits, it lacks the
ready-to-use implemented methods featured in other packages such as
fossilisation and sampling (Barido-Sottani et al. 2019; Stadler 2011;
Rosario Petrucci, Januario, and Quental 2022) or specific
macroevolutionary simulations (Morlon et al. 2016; Puttick, Guillerme,
and Wills 2020).''}. line 63-67

I've also removed the mention of the future time-dependent and
lineage-dependent simulations: these are meant to be internal
implementations allowing the algorithm to track time and lineage IDs
across each edges which will allow even more modularity in the future
but is out of the scope of this paper.

\emph{``For example future planned versions will include abiotic events
and a better integration with the \texttt{dispRity} package.''}. line
218-219

\begin{itemize}
\tightlist
\item
  \textcolor{blue}{I would also have liked some information on the performance of the
  package compared to existing simulation tools. It seems likely that
  the flexibility offered by treats comes at the cost of some
  computational performance, so it would be nice to know how much time
  would be needed to simulate a reasonable dataset (for instance the 50
  replicates used to test the positive extinction vs random
  extinction).}
\end{itemize}

I tried (and saved the tests in \texttt{tests/test-benchmarking.R} for
people to explore) but I'm not convinced how useful it is. First I feel
it's confrontational to the other authors since none of them had the
same objectives in mind. Second I will be biased in making treats look
better (when possible) because I know how it works in much more details.
Third I don't think it's something I want to put forward as a reason of
why to use the package, e.g.~it's slower than TreeSim but faster than
diversitree (under certain parameters) but that's not the reason while
one should use treats over diversitree and TreeSim (the point I am
trying to sell is the modularity, i.e.~if you want to simulate something
specific, treats is good, otherwise, go with what other people have used
before).

\begin{itemize}
\tightlist
\item
  \textcolor{blue}{I didn't understand why the example shown in the manuscript only
  simulated trees up to 30 time units, when the tree used as a basis for
  the comparison is 140 My old (figure 2).}
\end{itemize}

I've now added the following precision:

\emph{``The number of time units in \texttt{treats} is arbitrary and is
not equivalent to millions of years. Using 30 time units here allows to
simulate number of tips in a similar order of magnitude as the ones in
the observed data.''} lines 181-183.

\begin{itemize}
\tightlist
\item
  \textcolor{blue}{I really like the idea of having a library of premade simulation
  templates (l202) and I believe it will be a huge help in making the
  package accessible. Making the examples shown in the manual (section
  8) available as scripts would be a good start for such a library.}
\end{itemize}

I am happy that both reviewers are enthusiastic about the templates
library. Unfortunately I'm not totally sure how to implement it and make
it easier to use and browse by users. So for now I've set up a templated
category of issues on github called
\href{https://github.com/TGuillerme/treats/issues?q=is\%3Aopen+is\%3Aissue+label\%3A\%22simulation+template\%22}{simulation
templates}. I've updated the README file with some indications on how to
save and share such a template.

\hypertarget{minor-comments}{%
\subsubsection{Minor comments:}\label{minor-comments}}

\begin{itemize}
\tightlist
\item
  \textcolor{blue}{ - l71 ``no either'' }
\end{itemize}

Fixed.

\begin{itemize}
\tightlist
\item
  \textcolor{blue}{ - l125 ``where the it''}
\end{itemize}

Fixed.

\begin{itemize}
\tightlist
\item
  \textcolor{blue}{ - l111 the comment in the second example is mismatched with the code
  (random extinction instead of positive extinction)}
\end{itemize}

Fixed.

\begin{itemize}
\tightlist
\item
  \textcolor{blue}{ - l175 the comment in the example is mismatched with the code
  (positive extinction instead of random extinction)}
\end{itemize}

Fixed.

\begin{itemize}
\tightlist
\item
  \textcolor{blue}{ - l178 the text mentions the function trait.extinction but it's not
  used in the code}
\end{itemize}

Fixed.

\begin{itemize}
\tightlist
\item
  \textcolor{blue}{ - l282 missing authors in the reference}
\end{itemize}

This one is just generated as such from the \texttt{mee.csl} reference
style.

\hypertarget{comments-on-the-manual}{%
\subsection{Comments on the manual}\label{comments-on-the-manual}}

\begin{itemize}
\tightlist
\item
  \textcolor{blue}{The main thing the manual is missing in my opinion is a quick
  reference table of the different functions, for people who have
  understood the concepts of the package but need to quickly check what
  format their custom speciation function needs to follow for instance.
  At the moment that information is incomplete (the table in section 4.5
  shows the output types but not what these outputs are) and spread out
  in different places (the description of the function inputs is in
  section 4.2.1 whereas the list of functions is in 4.1).}
\end{itemize}

I've now added a cheat sheet to the github page that shows the key
arguments and their inputs/outputs
\href{https://github.com/TGuillerme/treats/blob/master/inst/gitbook/treats_cheat_sheet.pdf}{here}.
This cheat sheet is rather minimal for now but I'm planning on expanding
on it and making it more interactive (with mouse over menus, etc.) in
the future. Furthermore I've tidied up the summarising of functions in
the manual as suggested by the reviewer.

\begin{itemize}
\tightlist
\item
  \textcolor{blue}{The use of conditions and modifications was quite confusing to me.
  Some of the examples use the modify and condition argument to change
  the birth-death process (section 2.3) whereas some examples simply
  replace the function with a modified version (section 4.4), and I
  could not figure out if these methods were equivalent or if one should
  be preferred to the other. I think the modifications section also
  needs more details about what exactly is being modified, and what the
  expected inputs and outputs of the modification function should be.
  For instance, in the example section 2.3 I don't understand what is
  the ``x'' argument in ``going.extinct'', or why this function returns
  a number instead of a logical. The interface for the conditions in 5.2
  appears to be clearer, but it doesn't match with the example in
  section 2.3 where the ``negative.ancestor'' uses different arguments,
  or with the example in 4.6 where the ``half.the.time'' condition takes
  no arguments. }
\end{itemize}

I have now added clarifications of what the \texttt{"modifiers"} are and
how they should be used. Both in the section 2.3 as a simple explanation
and in the section 4 dedicated to \texttt{"modifiers"}.

\begin{itemize}
\tightlist
\item
  \textcolor{blue}{Finally, I also could not find a full description of the treats
  output object either in the manual or in the package documentation.}
\end{itemize}

I've added a \texttt{@return} section to the \texttt{treats},
\texttt{make.treats}, \texttt{make.bd.params}, \texttt{make.traits},
\texttt{make.modifiers} and \texttt{make.events} documentations as well
as the following section in the manual ``Getting started'' section:

\emph{\texttt{treats} will output either just a tree (class
\texttt{"phylo"}) if no traits were generated or a \texttt{"treats"}
object that contains both a \texttt{\$tree} (\texttt{"phylo"}) and a
\texttt{\$data} (\texttt{"matrix"}) component. Note that this
\texttt{"treats"} class is generalised to most outputs of the package
functions. This allows for a smoother handeling of the objects outputs
such as summarising the content of a \texttt{make.bd.params} output or
visualising a trait output from \texttt{make.traits}.}

\hypertarget{reviewer-2}{%
\section{Reviewer: 2}\label{reviewer-2}}

\textcolor{blue}{Thank you for the opportunity to review this manuscript, titled
``treats: a modular R package for simulating trees and traits''. This
manuscript describes and showcases a new R package that provides users
with an extensive toolbox to jointly simulate phylogenies and evolving
traits. While the example provided within the manuscript gives readers a
small taste of the flexibility of this toolbox, the online manual
(\url{http://tguillerme.github.io/treats.html}) showcases the true
extent of how useful this package will be for evolutionary biologists
far and wide. I really enjoyed reading the manuscript, and I believe it
will be suitable for publication in Methods in Ecology \& Evolution once
my concerns have been addressed.}

\textcolor{blue}{For context, for this review I read the supplied manuscript, tested the
functionality of the R package, and browsed the built-in documentation
and the online supplemental manual. I was unable to review the
functional code of the R package in time for the due date of this
review, but it appears to work properly from my testing. I have one
major concern and a handful of minor concerns. The major concern is that
the overall flow of the manuscript seems a little off. I think this
mostly stems from the worked example preceding the detailed overview of
the package's functions. I think if these two sections were swapped the
manuscript would flow much better. My minor concerns are included at the
end of this review in order of their appearance in the manuscript. I
look forward to reviewing a revised version of this manuscript, if
necessary.}

I have now changed the order of the structure of the manuscript with the
description coming first and the example coming second.

\textcolor{blue}{Best,}

\textcolor{blue}{William Gearty}

\begin{itemize}
\tightlist
\item
  \textcolor{blue}{Lines 27-31: I would enumerate this list.}
\end{itemize}

Done.

\begin{itemize}
\tightlist
\item
  \textcolor{blue}{Line 84: Geological time period names should be capitalized.}
\end{itemize}

Fixed.

\begin{itemize}
\tightlist
\item
  \textcolor{blue}{Figures 2-6: If possible, I think it would be valuable to show what
  code was used to produce these figures in the code snippets. As far as
  I can tell, the plotting is mostly done using functions from the
  package, so this would also be an opportunity to showcase this
  functionality.}
\end{itemize}

Done with the now new inclusion of the \texttt{dispRitreats} code that
streamlines the analysis.

\begin{itemize}
\tightlist
\item
  \textcolor{blue}{Figure 2: The caption should identify what the red line represents
  (the K-Pg boundary?).}
\end{itemize}

Done.

\begin{itemize}
\tightlist
\item
  \textcolor{blue}{Line 111: This may be personal preference, but when I ran this code
  chunk it output a whole bunch of text to the console. It appears to be
  a separate line of text for each replicate in the treats function. I
  believe the intent here is to show the user how much progress has been
  made? Perhaps some sort of progress bar would be more appropriate?}
\end{itemize}

I've now added a \texttt{verbose} option to the \texttt{treats} function
allowing to display the progress or not. Unfortunately the progress bar
suggestion, although comfortable to most users is not feasible
effectively with stochastic data: here the output indicates every re-ran
tree for each replicate. And because the number of runs varies depending
on the random seed, tracking the progress in terms of percentage will
require more computational time (i.e.~estimating how many re-run are
expected for the current seed before increase the progress bar).

\begin{itemize}
\tightlist
\item
  \textcolor{blue}{Lines 106-111: I think you should mention somewhere in this paragraph
  why you are running multiple replicates here (as opposed to the single
  replicate performed earlier in the worked example).}
\end{itemize}

I've added the following sentence at the end of the paragraph:
``\emph{Because of the stochasticity of the simulations, we will repeat
them 50 times (using \texttt{replicates\ =\ 50}) to generate a
distribution of possible simulated scenarios as opposed to a random
single one that could be idiosyncratic.}'' lines 200-203

\begin{itemize}
\tightlist
\item
  \textcolor{blue}{Line 111: I used the same seed as you, but I was unable to reproduce
  Figure 5 (looking at the R markdown file, I see that you use indices
  11 and 10 to make the plots). I think showing the exact code that is
  used here to make Figure 5 would be useful for readers that might be
  ``playing along at home''.}
\end{itemize}

I have streamlined and display the code to allow the users to test the
package while reading the paper.

\begin{itemize}
\tightlist
\item
  \textcolor{blue}{Lines 112-113: You should refer to Figure 6 somewhere in this
  paragraph. It might also be useful to include the code to perform this
  comparison, especially if the functionality is all contained within
  the treats package.}
\end{itemize}

Done. line 207.

\begin{itemize}
\tightlist
\item
  \textcolor{blue}{Line 194: This reference to the manual should have a link to the
  online manual (looks like the link is embedded, but you should also
  have it typed out). I also noticed that the documentation for the
  treats function lists the wrong URL for the online manual.}
\end{itemize}

I've added the URL spelt out.

\begin{itemize}
\tightlist
\item
  \textcolor{blue}{Lines 202-203: Just wanted to note that I really love this idea! I
  think collecting a library of modules will be very useful to users!}
\end{itemize}

As mentioned before, I was not sure how to effectively implement that so
I made a
\href{https://github.com/TGuillerme/treats/issues?q=is\%3Aopen+is\%3Aissue+label\%3A\%22simulation+template\%22}{simulation
templates} issues category on github for now.

\begin{itemize}
\tightlist
\item
  \textcolor{blue}{Line 205: This embedded link does not work. I think it should be
  something like:
  \url{https://github.com/TGuillerme/treats/tree/master/paper}.}
\end{itemize}

Fixed.

\hypertarget{refs}{}
\begin{cslreferences}
\leavevmode\hypertarget{ref-fossilsim}{}%
Barido-Sottani, Joëlle, Walker Pett, Joseph E O'Reilly, and Rachel CM
Warnock. 2019. ``FossilSim: An R Package for Simulating Fossil
Occurrence Data Under Mechanistic Models of Preservation and Recovery.''
\emph{Methods in Ecology and Evolution} 10 (6): 835--40.

\leavevmode\hypertarget{ref-rpanda}{}%
Morlon, Hélène, Eric Lewitus, Fabien L Condamine, Marc Manceau, Julien
Clavel, and Jonathan Drury. 2016. ``RPANDA: An R Package for
Macroevolutionary Analyses on Phylogenetic Trees.'' \emph{Methods in
Ecology and Evolution} 7 (5): 589--97.

\leavevmode\hypertarget{ref-puttick2020complex}{}%
Puttick, Mark N, Thomas Guillerme, and Matthew A Wills. 2020. ``The
Complex Effects of Mass Extinctions on Morphological Disparity.''
\emph{Evolution} 74 (10): 2207--20.

\leavevmode\hypertarget{ref-paleobuddy}{}%
Rosario Petrucci, Bruno do, Matheus Januario, and Tiago Quental. 2022.
``Paleobuddy: An R Package for Flexible Simulations of Diversification
and Fossil Sampling.'' \emph{Methods in Ecology and Evolution} 13 (12):
2692--8.

\leavevmode\hypertarget{ref-treesim}{}%
Stadler, Tanja. 2011. ``Simulating Trees with a Fixed Number of Extant
Species.'' \emph{Systematic Biology} 60 (5): 676--84.
\end{cslreferences}

\end{document}

### 2023-11-22 paper/treats_paper.Rmd
Updated paper and response

@@
 Therefore, I propose `treats` a modular `R` package to simulate both trees and traits through time.
-Note that although `treats` is modular and thus allows to be used as go to tool for simulating and trees and traits, it lacks the ready-to-use implemented methods featured in other packages such as fossilisation and sampling [@fossilsim, @treesim, @paleobuddy] or specific macroevolutionary simulations [@RPANDA, @puttick2020complex].
+Note that although `treats` is modular and thus allows to be used as go to tool for simulating and trees and traits, it lacks the ready-to-use implemented methods featured in other packages such as fossilisation and sampling [@fossilsim; @treesim; @paleobuddy] or specific macroevolutionary simulations [@rpanda; @puttick2020complex].
 
@@
 
-In `treats`, these three or four steps are implemented as modular functions that the user can easily change using an internal class of objects called `"modifiers"` or `"traits"` (Fig.
-\ref{Fig:workflow}).
+In `treats`, these three or four steps are implemented as modular functions that the user can easily change using an internal class of objects called `"modifiers"` or `"traits"` (Fig. 1).
 The simulation then outputs a tree (of class `"phylo"` and a associated table of traits - `"matrix"`) that can be visualised using the `plot.treats` function.
@@
 Thanks to Andrew Beckerman, Ian Brennan, Gustavo Burin, Christopher Cooney, Natalie Cooper, Alex Cranston, Jasmine Hardie, Tom Lansley, Clement Prieul, Joe Rees, James Rule, Sophie Ryan and Gavin Thomas, for support and useful comments on late stages of the development of this package.
-This work was funded by UKRI-NERC Grant NE/T000139/1.
+Thanks to William Gearty, one anonymous reviewer and one anonymous associate editor for their very useful suggestions.
+This work was funded by UKRI-NERC Grant NE/T000139/1 awarded to Gavin Thomas.
 

### 2023-11-22 paper/treats_paper_lineno.tex
Updated paper and response

--- 2023-06-19 paper/treats_paper.tex
@@
 ]{article}
+\usepackage{lineno}
 \usepackage{lmodern}
@@
 
+\linenumbers
+
 \title{\texttt{treats}: a modular \texttt{R} package for simulating
@@
 \author{Thomas Guillerme\(^1\)}
-\date{2023-06-09}
+\date{2023-11-22}
 
@@
 or base conditions (Miller et al. 2022). In theory, workers can follow
-the following research pipeline: thinking of a specific mechanism
+the following research pipeline: 1) thinking of a specific mechanism
 (e.g.~mass extinction allowing the surviving species to acquire new
-morphologies), collecting some data to test this mechanism (e.g.~some
-traits of species across and extinction event) and then comparing these
-patterns to some simulated under no specific conditions (e.g.~a null
-model where the traits evolve randomly regardless of an extinction event
-Puttick, Guillerme, and Wills (2020)). Workers might thus need to
+morphologies), 2) collecting some data to test this mechanism (e.g.~some
+traits of species across and extinction event) and then 3) comparing
+these patterns to some simulated under no specific conditions (e.g.~a
+null model where the traits evolve randomly regardless of an extinction
+event Puttick, Guillerme, and Wills (2020)). Workers might thus need to
 simulate a great diversity of evolutionary scenarios to test their
@@
 \texttt{treats} a modular \texttt{R} package to simulate both trees and
-traits through time.
+traits through time. Note that although \texttt{treats} is modular and
+thus allows to be used as go to tool for simulating and trees and
+traits, it lacks the ready-to-use implemented methods featured in other
+packages such as fossilisation and sampling (Barido-Sottani et al. 2019;
+Stadler 2011; Rosario Petrucci, Januario, and Quental 2022) or specific
+macroevolutionary simulations (Morlon et al. 2016; Puttick, Guillerme,
+and Wills 2020).
 
@@
 \item
-  Generating some trait(s) value(s) for the selected lineage (no either
-  a tip or a node).
+  Generating some trait(s) value(s) for the selected lineage (either a
+  tip or a node - but see the ``Simulating traits section'' below to
+  generate trait values along edges).
 \end{enumerate}
@@
 functions that the user can easily change using an internal class of
-objects called \texttt{"modifiers"} or \texttt{"traits"} (Fig.
-\ref{Fig:workflow}). The simulation then outputs a tree (of class
-\texttt{"phylo"} and a associated table of traits - \texttt{"matrix"})
-that can be visualised using the \texttt{plot.treats} function. A third
-class of object called \texttt{"events"} can be added to the simulations
-to modify the entire simulation under certain conditions
-(e.g.~simulating a mass extinction). Each element in the algorithm can
-be modified by the user using the implemented functions
-\texttt{make.bd.params} to set the birth-death parameters,
-\texttt{make.traits} to set the trait(s), \texttt{make.modifiers} to set
-the birth-death algorithm and \texttt{make.events} to eventually add one
-or more events. Users can replace any step in the algorithm by their own
-specific functions suiting their needs or, for less advanced users, by
-using already implemented functions.
+objects called \texttt{"modifiers"} or \texttt{"traits"} (Fig. 1). The
+simulation then outputs a tree (of class \texttt{"phylo"} and a
+associated table of traits - \texttt{"matrix"}) that can be visualised
+using the \texttt{plot.treats} function. A third class of object called
+\texttt{"events"} can be added to the simulations to modify the entire
+simulation under certain conditions (e.g.~simulating a mass extinction).
+Each element in the algorithm can be modified by the user using the
+implemented functions \texttt{make.bd.params} to set the birth-death
+parameters, \texttt{make.traits} to set the trait(s),
+\texttt{make.modifiers} to set the birth-death algorithm and
+\texttt{make.events} to eventually add one or more events. Users can
+replace any step in the algorithm by their own specific functions
+suiting their needs or, for less advanced users, by using already
+implemented functions.
 
@@
 \includegraphics{../inst/gitbook/treats_structure.png}
-\caption{Figure 1: \texttt{treats} package workflow: the \texttt{treats}
-algorithm generates a tree and traits using inbuilt \texttt{"traits"}
-objects that contain the instructions on how to generate the trait data
-(e.g.~which process? how many dimensions?); \texttt{"modifiers"} objects
-that contains instructions on how to ``grow'' the tree (e.g.~by linking
+\caption{\texttt{treats} package workflow: the \texttt{treats} algorithm
+generates a tree and traits using inbuilt \texttt{"traits"} objects that
+contain the instructions on how to generate the trait data (e.g.~which
+process? how many dimensions?); \texttt{"modifiers"} objects that
+contains instructions on how to ``grow'' the tree (e.g.~by linking
 speciation to trait values, or to the current number of species); and
@@
 
-\hypertarget{brief-applied-example}{%
-\subsection{Brief applied example}\label{brief-applied-example}}
-
-To illustrate this we will look whether it is possible to detect changes
-in disparity (i.e.~diversity of traits) in a subset of the data
-published from Beck and Lee (2014) implemented in Guillerme (2018). This
-dataset contains the ordinated traits for 50 mammalian species across
-the cretaceous-palaeogene extinction event (K-Pg, 66 Mya; Fig. 2).
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\#\# Loading the data and packages}
-\KeywordTok{library}\NormalTok{(treats)}
-\KeywordTok{library}\NormalTok{(dispRity)}
-\KeywordTok{data}\NormalTok{(BeckLee\_tree)}
-\KeywordTok{data}\NormalTok{(BeckLee\_mat99)}
-
-\CommentTok{\#\# Creating the time slices}
-\NormalTok{time\_slices \textless{}{-}}\StringTok{ }\KeywordTok{chrono.subsets}\NormalTok{(BeckLee\_mat99, BeckLee\_tree,}
-                              \DataTypeTok{method =} \StringTok{"continuous"}\NormalTok{,}
-                              \DataTypeTok{model  =} \StringTok{"acctran"}\NormalTok{,}
-                              \DataTypeTok{time   =} \KeywordTok{seq}\NormalTok{(}\DataTypeTok{from =} \DecValTok{96}\NormalTok{, }\DataTypeTok{to =} \DecValTok{36}\NormalTok{, }\DataTypeTok{by =} \DecValTok{{-}10}\NormalTok{))}
-
-\CommentTok{\#\# Calculating disparity on the two first dimensions only}
-\NormalTok{observed\_disparity \textless{}{-}}\StringTok{ }\KeywordTok{dispRity}\NormalTok{(}\KeywordTok{boot.matrix}\NormalTok{(time\_slices),}
-                               \DataTypeTok{metric =} \KeywordTok{c}\NormalTok{(sum, variances),}
-                               \DataTypeTok{dimensions =} \KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{,}\DecValTok{2}\NormalTok{))}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{figure}
-\centering
-\includegraphics{treats_paper_files/figure-latex/prepdata-1.pdf}
-\caption{Figure 2: Observed phylogeny (A) and disparity through time (B)
-in the subset of Beck and Lee (2014)'s dataset. Is this change in
-disparity related to the K-Pg mass extinction? Or, more generally, is it
-even possible to detect potential changes of disparity due to a mass
-extinction event?}
-\end{figure}
-
-Using this example dataset, one might be interested in testing whether
-the K-Pg extinction had an effect on disparity through time. But can
-such effect be detected in the first place? We can test this by
-simulating some datasets with similar properties as the observed data
-and measure changes in disparity in these simulated datasets.
-
-\hypertarget{simulating-trees}{%
-\subsubsection{Simulating trees}\label{simulating-trees}}
-
-The first and simplest way is to simulate tree topologies that have
-similar properties than the observed one. To do so, we need to use some
-\textbf{speciation} parameter indicating the rate at which lineages
-speciate (\emph{aka} ``birth'' or ``\(\lambda\)'') and an
-\textbf{extinction} parameter indicating the rate at which they go
-extinct (\emph{aka} ``death'' or ``\(\mu\)''). These two parameters can
-be crudely calculated from a given tree by counting the number of
-speciations and extinctions through time (but see better methods,
-e.g.~\texttt{ape::birthdeath}; Lawton, May, and Raup (1995); Paradis and
-Schliep (2019)). We also need a stopping rule for when to stop the
-simulations (in our case when reaching 30 time units). This will produce
-a single random tree using the input parameters (Fig. 3). Note that I
-will not discuss the options here in great details. Much more
-information can be found in the
-\href{http://tguillerme.github.io/treats.html}{\texttt{treats} manual}.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\#\# Using the birth{-}death parameters from the observed tree}
-\NormalTok{my\_bd\_params \textless{}{-}}\StringTok{ }\KeywordTok{crude.bd.est}\NormalTok{(BeckLee\_tree)}
-\CommentTok{\#\# Setting the stopping rule (stop after 30 time units)}
-\NormalTok{stop\_rule \textless{}{-}}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{max.time =} \DecValTok{30}\NormalTok{)}
-
-\CommentTok{\#\# Simulate the tree}
-\KeywordTok{set.seed}\NormalTok{(}\DecValTok{1}\NormalTok{)}
-\NormalTok{sim\_tree \textless{}{-}}\StringTok{ }\KeywordTok{treats}\NormalTok{(}\DataTypeTok{bd.params  =}\NormalTok{ my\_bd\_params,}
-                   \DataTypeTok{stop.rule  =}\NormalTok{ stop\_rule,}
-                   \DataTypeTok{null.error =} \DecValTok{100}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{figure}
-\centering
-\includegraphics{treats_paper_files/figure-latex/simtree-1.pdf}
-\caption{Figure 3: A randomly simulated tree with similar properties as
-the observed one. Note that the time here is expressed in arbitrary
-units.}
-\end{figure}
-
-\hypertarget{simulating-trees-and-traits}{%
-\subsubsection{Simulating trees and
-traits}\label{simulating-trees-and-traits}}
-
-For our specific question, we will also need to simulate some traits
-associated with each node and tip. For simplicity we will simulate a two
-dimensional Brownian Motion trait. To do so, we can create a
-\texttt{"traits"} object with the function \texttt{make.traits}. This
-results in a 2 dimensional trait space for all the simulated species and
-their nodes (Fig. 4).
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\#\# Creating a trait in 2 dimensions.}
-\NormalTok{my\_traits \textless{}{-}}\StringTok{ }\KeywordTok{make.traits}\NormalTok{(}\DataTypeTok{process =}\NormalTok{ BM.process, }\DataTypeTok{n =} \DecValTok{2}\NormalTok{)}
-
-\CommentTok{\#\# Simulate the tree and traits}
-\KeywordTok{set.seed}\NormalTok{(}\DecValTok{1}\NormalTok{)}
-\NormalTok{sim\_data \textless{}{-}}\StringTok{ }\KeywordTok{treats}\NormalTok{(}\DataTypeTok{traits     =}\NormalTok{ my\_traits,}
-                   \DataTypeTok{bd.params  =}\NormalTok{ my\_bd\_params,}
-                   \DataTypeTok{stop.rule  =}\NormalTok{ stop\_rule,}
-                   \DataTypeTok{null.error =} \DecValTok{100}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{figure}
-\centering
-\includegraphics{treats_paper_files/figure-latex/simtreetraits-1.pdf}
-\caption{Figure 4: A randomly simulated tree with a 2 dimensional random
-Brownian Motion trait. Orange, light blue and dark blue dots
-respectively represent nodes, fossils and living species.}
-\end{figure}
-
-\hypertarget{simulating-a-trees-and-traits-with-events}{%
-\subsubsection{Simulating a trees and traits with
-events}\label{simulating-a-trees-and-traits-with-events}}
-
-To answer our question, we also want to simulate an extinction event. To
-do so, we can create two different \texttt{"events"} object with the
-function \texttt{make.events}. The first one will simulate a random
-extinction event after reaching half of the simulations (15 time units)
-and then making three quarter (0.75) of the taxa go extinct. The second
-one will simulate a random extinction but based on trait values: after
-reaching time 15, all the species with positive trait values will go
-extinct. Both scenarios thus illustrate two different types of mass
-extinction (random and selective; Puttick, Guillerme, and Wills (2020)).
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\#\# Creating a random mass extinction}
-\NormalTok{random\_extinction \textless{}{-}}\StringTok{ }\KeywordTok{make.events}\NormalTok{(}
-    \DataTypeTok{target       =} \StringTok{"taxa"}\NormalTok{,}
-    \DataTypeTok{condition    =} \KeywordTok{time.condition}\NormalTok{(}\DecValTok{15}\NormalTok{),}
-    \DataTypeTok{modification =} \KeywordTok{random.extinction}\NormalTok{(}\FloatTok{0.75}\NormalTok{))}
-\CommentTok{\#\# Creating an extinction that removes species with positive trait values}
-\NormalTok{positive\_extinction \textless{}{-}}\StringTok{ }\KeywordTok{make.events}\NormalTok{(}
-    \DataTypeTok{target =} \StringTok{"taxa"}\NormalTok{,}
-    \DataTypeTok{condition =} \KeywordTok{time.condition}\NormalTok{(}\DecValTok{15}\NormalTok{),}
-    \DataTypeTok{modification =} \KeywordTok{trait.extinction}\NormalTok{(}\DataTypeTok{x =} \DecValTok{0}\NormalTok{, }\DataTypeTok{condition =} \StringTok{\textasciigrave{}}\DataTypeTok{\textgreater{}=}\StringTok{\textasciigrave{}}\NormalTok{))}
-
-\KeywordTok{set.seed}\NormalTok{(}\DecValTok{2}\NormalTok{)}
-\CommentTok{\#\# Simulate the tree and traits with a random extinction event}
-\NormalTok{sim\_rand\_extinction \textless{}{-}}\StringTok{ }\KeywordTok{treats}\NormalTok{(}
-                   \DataTypeTok{traits     =}\NormalTok{ my\_traits,}
-                   \DataTypeTok{bd.params  =}\NormalTok{ my\_bd\_params,}
-                   \DataTypeTok{stop.rule  =}\NormalTok{ stop\_rule,}
-                   \DataTypeTok{events     =}\NormalTok{ random\_extinction,}
-                   \DataTypeTok{null.error =} \DecValTok{100}\NormalTok{,}
-                   \DataTypeTok{replicates =} \DecValTok{50}\NormalTok{)}
-
-\CommentTok{\#\# Simulate the tree and traits with a random extinction event}
-\NormalTok{sim\_trait\_extinction \textless{}{-}}\StringTok{ }\KeywordTok{treats}\NormalTok{(}
-                   \DataTypeTok{traits     =}\NormalTok{ my\_traits,}
-                   \DataTypeTok{bd.params  =}\NormalTok{ my\_bd\_params,}
-                   \DataTypeTok{stop.rule  =}\NormalTok{ stop\_rule,}
-                   \DataTypeTok{events     =}\NormalTok{ positive\_extinction,}
-                   \DataTypeTok{null.error =} \DecValTok{100}\NormalTok{,}
-                   \DataTypeTok{replicates =} \DecValTok{50}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{figure}
-\centering
-\includegraphics{treats_paper_files/figure-latex/simtreetraitsevents-1.pdf}
-\caption{Figure 5: Two different mass extinction events: (A) 75\% of
-species go extinct; (B) all species with positive trait values go
-extinct. The red line marks the extinction event.}
-\end{figure}
-
-Once we have simulated a distribution of trees and traits with the two
-extinction scenarios, we can measure disparity as in Fig. 2 for all the
-simulated data and compare it to the observed disparity.
-
-\begin{figure}
-\centering
-\includegraphics{treats_paper_files/figure-latex/simdisparity-1.pdf}
-\caption{Figure 6: Comparison of the observed sum of variances (dashed
-line) to the simulated one (ligth and dark grey polygons and black line
-representing respectively the 95\% and 50\% confidence interval and the
-median value).}
-\end{figure}
-
-From these results we can draw some preliminary conclusions: the
-observed change in disparity is more likely due to a random mass
-extinction rather than a selective one. This is of course a very crude
-way of testing this, a more rigorous approach is needed to answer the
-question: more and better quality data, and more thorough methods
-(e.g.~using a rank envelope test Murrell (2018)).
-
 \hypertarget{major-functionalities}{%
@@
 simple Brownian Motion process could be generated by the function
-\texttt{rnorm} where the it modifies the trait value \texttt{x0}
-relative to the branch length (\texttt{edge.length}). The longer the
-branch length, the more likely the new trait value will be different
-from \texttt{x0}:
+\texttt{rnorm} where it modifies the trait value \texttt{x0} relative to
+the branch length (\texttt{edge.length}). The longer the branch length,
+the more likely the new trait value will be different from \texttt{x0}:
 
@@
 \end{itemize}
+
+Traits are always generated (and stored) only for tips or nodes and not
+along edges. However, it is possible so generate trait values at
+specific or regular time intervals by using the option
+\texttt{save.steps} in the \texttt{treats} function. This will generate
+singleton nodes (i.e.~nodes with only one descendant) with associated
+trait values.
 
@@
     \DataTypeTok{condition    =} \KeywordTok{time.condition}\NormalTok{(}\DecValTok{15}\NormalTok{),}
-    \DataTypeTok{modification =} \KeywordTok{random.extinction}\NormalTok{(}\FloatTok{0.75}\NormalTok{))}
+    \DataTypeTok{modification =} \KeywordTok{trait.extinction}\NormalTok{(}\DataTypeTok{x =} \DecValTok{0}\NormalTok{, }\DataTypeTok{condition =} \StringTok{\textasciigrave{}}\DataTypeTok{\textgreater{}=}\StringTok{\textasciigrave{}}\NormalTok{))}
 \end{Highlighting}
@@
 
+\hypertarget{brief-applied-example}{%
+\section{Brief applied example}\label{brief-applied-example}}
+
+To illustrate this we will look whether it is possible to detect changes
+in disparity (i.e.~diversity of traits) in a subset of the data
+published from Beck and Lee (2014) implemented in Guillerme (2018). This
+dataset contains the ordinated traits for 50 mammalian species across
+the Cretaceous-Palaeogene extinction event (K-Pg, 66 Mya; Fig. 2).
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# Loading the package and data}
+\KeywordTok{library}\NormalTok{(treats)}
+\KeywordTok{data}\NormalTok{(BeckLee\_tree)}
+\KeywordTok{data}\NormalTok{(BeckLee\_mat99)}
+
+\CommentTok{\#\# Creating the time slices}
+\NormalTok{time\_slices \textless{}{-}}\StringTok{ }\KeywordTok{chrono.subsets}\NormalTok{(BeckLee\_mat99, BeckLee\_tree,}
+                              \DataTypeTok{method =} \StringTok{"continuous"}\NormalTok{,}
+                              \DataTypeTok{model  =} \StringTok{"proximity"}\NormalTok{,}
+                              \DataTypeTok{time   =} \KeywordTok{seq}\NormalTok{(}\DataTypeTok{from =} \DecValTok{96}\NormalTok{, }\DataTypeTok{to =} \DecValTok{36}\NormalTok{, }\DataTypeTok{by =} \DecValTok{{-}10}\NormalTok{))}
+
+\CommentTok{\#\# Calculating disparity on the two first dimensions only}
+\NormalTok{observed\_disparity \textless{}{-}}\StringTok{ }\KeywordTok{dispRity}\NormalTok{(}\KeywordTok{boot.matrix}\NormalTok{(time\_slices),}
+                               \DataTypeTok{metric =} \KeywordTok{c}\NormalTok{(sum, variances),}
+                               \DataTypeTok{dimensions =} \KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{,}\DecValTok{2}\NormalTok{))}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# Plotting the tree and the disparity through time}
+\KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow =} \KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{,}\DecValTok{2}\NormalTok{), }\DataTypeTok{bty =} \StringTok{"null"}\NormalTok{)}
+\CommentTok{\#\# Thre tree}
+\KeywordTok{plot}\NormalTok{(BeckLee\_tree, }\DataTypeTok{show.tip.label =} \OtherTok{FALSE}\NormalTok{)}
+\KeywordTok{axisPhylo}\NormalTok{()}
+\KeywordTok{abline}\NormalTok{(}\DataTypeTok{v =}\NormalTok{ BeckLee\_tree}\OperatorTok{$}\NormalTok{root.time }\OperatorTok{{-}}\StringTok{ }\DecValTok{66}\NormalTok{, }\DataTypeTok{col =} \StringTok{"red"}\NormalTok{)}
+\KeywordTok{legend}\NormalTok{(}\StringTok{"topleft"}\NormalTok{, }\DataTypeTok{pch =} \OtherTok{NULL}\NormalTok{, }\DataTypeTok{legend =} \StringTok{"A"}\NormalTok{, }\DataTypeTok{bty =} \StringTok{"n"}\NormalTok{, }\DataTypeTok{cex =} \DecValTok{2}\NormalTok{)}
+
+\CommentTok{\#\# The disparity}
+\KeywordTok{plot}\NormalTok{(observed\_disparity, }\DataTypeTok{ylab =} \StringTok{"Sum of variances"}\NormalTok{)}
+\KeywordTok{abline}\NormalTok{(}\DataTypeTok{v =} \DecValTok{4}\NormalTok{, }\DataTypeTok{col =} \StringTok{"red"}\NormalTok{)}
+\KeywordTok{legend}\NormalTok{(}\StringTok{"topleft"}\NormalTok{, }\DataTypeTok{pch =} \OtherTok{NULL}\NormalTok{, }\DataTypeTok{legend =} \StringTok{"B"}\NormalTok{, }\DataTypeTok{bty =} \StringTok{"n"}\NormalTok{, }\DataTypeTok{cex =} \DecValTok{2}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{figure}
+\centering
+\includegraphics{treats_paper_files/figure-latex/prepdata-1.pdf}
+\caption{Observed phylogeny (A) and disparity through time (B) in the
+subset of Beck and Lee (2014)'s dataset. The red line represents the
+K-Pg boundary. Is this change in disparity related to the K-Pg mass
+extinction? Or, more generally, is it even possible to detect potential
+changes of disparity due to a mass extinction event?}
+\end{figure}
+
+Using this example dataset, one might be interested in testing whether
+the K-Pg extinction had an effect on disparity through time. But can
+such effect be detected in the first place? We can test this by
+simulating some datasets with similar properties as the observed data
+and measure changes in disparity in these simulated datasets.
+
+\hypertarget{simulating-trees}{%
+\subsubsection{Simulating trees}\label{simulating-trees}}
+
+The first and simplest way is to simulate tree topologies that have
+similar properties than the observed one. To do so, we need to use some
+\textbf{speciation} parameter indicating the rate at which lineages
+speciate (\emph{aka} ``birth'' or ``\(\lambda\)'') and an
+\textbf{extinction} parameter indicating the rate at which they go
+extinct (\emph{aka} ``death'' or ``\(\mu\)''). These two parameters can
+be crudely calculated from a given tree by counting the number of
+speciations and extinctions through time (but see better methods,
+e.g.~\texttt{ape::birthdeath}; Lawton, May, and Raup (1995); Paradis and
+Schliep (2019)). We also need a stopping rule for when to stop the
+simulations (in our case when reaching 30 time units). This will produce
+a single random tree using the input parameters (Fig. 3). The number of
+time units in \texttt{treats} is arbitrary and is not equivalent to
+millions of years. Using 30 time units here allows to simulate number of
+tips in a similar order of magnitude as the ones in the observed data.
+Note that I will not discuss the options here in great details. Much
+more information can be found in the
+\href{http://tguillerme.github.io/treats.html}{\texttt{treats} manual}.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# Using the birth{-}death parameters from the observed tree}
+\NormalTok{my\_bd\_params \textless{}{-}}\StringTok{ }\KeywordTok{crude.bd.est}\NormalTok{(BeckLee\_tree)}
+\CommentTok{\#\# Setting the stopping rule (stop after 30 time units)}
+\NormalTok{stop\_rule \textless{}{-}}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{max.time =} \DecValTok{30}\NormalTok{)}
+
+\CommentTok{\#\# Simulate the tree}
+\KeywordTok{set.seed}\NormalTok{(}\DecValTok{1}\NormalTok{)}
+\NormalTok{sim\_tree \textless{}{-}}\StringTok{ }\KeywordTok{treats}\NormalTok{(}\DataTypeTok{bd.params  =}\NormalTok{ my\_bd\_params,}
+                   \DataTypeTok{stop.rule  =}\NormalTok{ stop\_rule,}
+                   \DataTypeTok{null.error =} \DecValTok{100}\NormalTok{,}
+                   \DataTypeTok{verbose    =} \OtherTok{FALSE}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{figure}
+\centering
+\includegraphics{treats_paper_files/figure-latex/simtree-1.pdf}
+\caption{A randomly simulated tree with similar properties as the
+observed one. Note that the time here is expressed in arbitrary units.}
+\end{figure}
+
+\hypertarget{simulating-trees-and-traits}{%
+\subsubsection{Simulating trees and
+traits}\label{simulating-trees-and-traits}}
+
+For our specific question, we will also need to simulate some traits
+associated with each node and tip. For simplicity we will simulate a two
+dimensional Brownian Motion trait. To do so, we can create a
+\texttt{"traits"} object with the function \texttt{make.traits}. This
+results in a 2 dimensional trait space for all the simulated species and
+their nodes (Fig. 4).
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# Creating a trait in 2 dimensions.}
+\NormalTok{my\_traits \textless{}{-}}\StringTok{ }\KeywordTok{make.traits}\NormalTok{(}\DataTypeTok{process =}\NormalTok{ BM.process, }\DataTypeTok{n =} \DecValTok{2}\NormalTok{)}
+
+\CommentTok{\#\# Simulate the tree and traits}
+\KeywordTok{set.seed}\NormalTok{(}\DecValTok{1}\NormalTok{)}
+\NormalTok{sim\_data \textless{}{-}}\StringTok{ }\KeywordTok{treats}\NormalTok{(}\DataTypeTok{traits     =}\NormalTok{ my\_traits,}
+                   \DataTypeTok{bd.params  =}\NormalTok{ my\_bd\_params,}
+                   \DataTypeTok{stop.rule  =}\NormalTok{ stop\_rule,}
+                   \DataTypeTok{null.error =} \DecValTok{100}\NormalTok{,}
+                   \DataTypeTok{verbose    =} \OtherTok{FALSE}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow =} \KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{,}\DecValTok{2}\NormalTok{))}
+\CommentTok{\#\# Plotting one trait through time}
+\KeywordTok{plot}\NormalTok{(sim\_data, }\DataTypeTok{ylab =} \StringTok{"Trait 1"}\NormalTok{, }\DataTypeTok{las =} \DecValTok{1}\NormalTok{, }\DataTypeTok{main =} \StringTok{"Trait 1 through time"}\NormalTok{)}
+
+\CommentTok{\#\# Plotting the two dimensions against each other}
+\KeywordTok{plot}\NormalTok{(sim\_data, }\DataTypeTok{trait =} \KeywordTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{,}\DecValTok{1}\NormalTok{), }\DataTypeTok{ylab =} \StringTok{"Trait 1"}\NormalTok{, }\DataTypeTok{xlab =} \StringTok{"Trait 2"}\NormalTok{, }\DataTypeTok{las =} \DecValTok{1}\NormalTok{, }\DataTypeTok{main =} \StringTok{"Traits 1 and 2"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{figure}
+\centering
+\includegraphics{treats_paper_files/figure-latex/simtreetraits-1.pdf}
+\caption{A randomly simulated tree with a 2 dimensional random Brownian
+Motion trait. Orange, light blue and dark blue dots respectively
+represent nodes, fossils and living species.}
+\end{figure}
+
+\hypertarget{simulating-a-trees-and-traits-with-events}{%
+\subsubsection{Simulating a trees and traits with
+events}\label{simulating-a-trees-and-traits-with-events}}
+
+To answer our question, we also want to simulate an extinction event. To
+do so, we can create two different \texttt{"events"} object with the
+function \texttt{make.events}. The first one will simulate a random
+extinction event after reaching half of the simulations (15 time units)
+and then making three quarter (0.75) of the taxa go extinct. The second
+one will simulate a random extinction but based on trait values: after
+reaching time 15, all the species with positive trait values will go
+extinct. Both scenarios illustrate two different types of mass
+extinctions but they are not equivalent: because of the ancestral trait
+value starting at 0, we expect the second scenario to remove on average
+only 50\% of the species (i.e.~half the species are expected to evolve a
+trait value above 0). For more details on simulating the effect of mass
+extinction and the difficulties to simulate an unambiguous effect of a
+mass extinction, see Puttick, Guillerme, and Wills (2020). Because of
+this stochasticity of the simulations, we will repeat them 50 times
+(using \texttt{replicates\ =\ 50}) to generate a distribution of
+possible simulated scenarios as opposed to a random single one that
+could be idiosyncratic (Fig. 5).
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# Creating a random mass extinction}
+\NormalTok{random\_extinction \textless{}{-}}\StringTok{ }\KeywordTok{make.events}\NormalTok{(}
+    \DataTypeTok{target       =} \StringTok{"taxa"}\NormalTok{,}
+    \DataTypeTok{condition    =} \KeywordTok{time.condition}\NormalTok{(}\DecValTok{15}\NormalTok{),}
+    \DataTypeTok{modification =} \KeywordTok{random.extinction}\NormalTok{(}\FloatTok{0.75}\NormalTok{))}
+\CommentTok{\#\# Creating an extinction that removes species with positive trait values}
+\NormalTok{positive\_extinction \textless{}{-}}\StringTok{ }\KeywordTok{make.events}\NormalTok{(}
+    \DataTypeTok{target =} \StringTok{"taxa"}\NormalTok{,}
+    \DataTypeTok{condition =} \KeywordTok{time.condition}\NormalTok{(}\DecValTok{15}\NormalTok{),}
+    \DataTypeTok{modification =} \KeywordTok{trait.extinction}\NormalTok{(}\DataTypeTok{x =} \DecValTok{0}\NormalTok{, }\DataTypeTok{condition =} \StringTok{\textasciigrave{}}\DataTypeTok{\textgreater{}=}\StringTok{\textasciigrave{}}\NormalTok{))}
+
+\KeywordTok{set.seed}\NormalTok{(}\DecValTok{1}\NormalTok{)}
+\CommentTok{\#\# Simulate the tree and traits with a random extinction event}
+\NormalTok{sim\_rand\_extinction \textless{}{-}}\StringTok{ }\KeywordTok{treats}\NormalTok{(}
+                   \DataTypeTok{traits     =}\NormalTok{ my\_traits,}
+                   \DataTypeTok{bd.params  =}\NormalTok{ my\_bd\_params,}
+                   \DataTypeTok{stop.rule  =}\NormalTok{ stop\_rule,}
+                   \DataTypeTok{events     =}\NormalTok{ random\_extinction,}
+                   \DataTypeTok{null.error =} \DecValTok{100}\NormalTok{,}
+                   \DataTypeTok{replicates =} \DecValTok{50}\NormalTok{)}
+
+\CommentTok{\#\# Simulate the tree and traits with a selective extinction event}
+\NormalTok{sim\_trait\_extinction \textless{}{-}}\StringTok{ }\KeywordTok{treats}\NormalTok{(}
+                   \DataTypeTok{traits     =}\NormalTok{ my\_traits,}
+                   \DataTypeTok{bd.params  =}\NormalTok{ my\_bd\_params,}
+                   \DataTypeTok{stop.rule  =}\NormalTok{ stop\_rule,}
+                   \DataTypeTok{events     =}\NormalTok{ positive\_extinction,}
+                   \DataTypeTok{null.error =} \DecValTok{100}\NormalTok{,}
+                   \DataTypeTok{replicates =} \DecValTok{50}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{figure}
+\centering
+\includegraphics{treats_paper_files/figure-latex/simtreetraitsevents-1.pdf}
+\caption{Two different mass extinction events: (A) 75\% of species go
+extinct; (B) all species with positive trait values go extinct. The red
+line marks the extinction event.}
+\end{figure}
+
+Once we have simulated a distribution of trees and traits with the two
+extinction scenarios, we can measure disparity as in Fig. 2 for all the
+simulated data and compare it to the observed disparity.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# Remove single nodes simulated at the extinction}
+\NormalTok{sim\_rand\_extinction \textless{}{-}}\StringTok{ }\KeywordTok{drop.singles}\NormalTok{(sim\_rand\_extinction)}
+\NormalTok{sim\_trait\_extinction \textless{}{-}}\StringTok{ }\KeywordTok{drop.singles}\NormalTok{(sim\_trait\_extinction)}
+
+\CommentTok{\#\# Calculate the dispRity for all the simulations}
+\NormalTok{random\_extinction\_disparity \textless{}{-}}\StringTok{ }\KeywordTok{dispRitreats}\NormalTok{(sim\_rand\_extinction,}
+                                            \DataTypeTok{method =} \StringTok{"continuous"}\NormalTok{,}
+                                            \DataTypeTok{model  =} \StringTok{"proximity"}\NormalTok{,}
+                                            \DataTypeTok{time   =} \KeywordTok{seq}\NormalTok{(}\DataTypeTok{from =} \DecValTok{25}\NormalTok{, }\DataTypeTok{to =} \DecValTok{0}\NormalTok{, }\DataTypeTok{by =} \DecValTok{{-}5}\NormalTok{),}
+                                            \DataTypeTok{metric =} \KeywordTok{c}\NormalTok{(sum, variances),}
+                                            \DataTypeTok{scale.trees =} \OtherTok{FALSE}\NormalTok{)}
+\NormalTok{selective\_extinction\_disparity \textless{}{-}}\StringTok{ }\KeywordTok{dispRitreats}\NormalTok{(sim\_trait\_extinction,}
+                                            \DataTypeTok{method =} \StringTok{"continuous"}\NormalTok{,}
+                                            \DataTypeTok{model  =} \StringTok{"proximity"}\NormalTok{,}
+                                            \DataTypeTok{time   =} \KeywordTok{seq}\NormalTok{(}\DataTypeTok{from =} \DecValTok{25}\NormalTok{, }\DataTypeTok{to =} \DecValTok{0}\NormalTok{, }\DataTypeTok{by =} \DecValTok{{-}5}\NormalTok{),}
+                                            \DataTypeTok{metric =} \KeywordTok{c}\NormalTok{(sum, variances),}
+                                            \DataTypeTok{scale.trees =} \OtherTok{FALSE}\NormalTok{)}
+
+\CommentTok{\#\# Scale the disparity results (to compare to the observed ones)}
+\NormalTok{random\_extinction\_disparity \textless{}{-}}\StringTok{ }\KeywordTok{rescale.dispRity}\NormalTok{(random\_extinction\_disparity)}
+\NormalTok{selective\_extinction\_disparity \textless{}{-}}\StringTok{ }\KeywordTok{rescale.dispRity}\NormalTok{(selective\_extinction\_disparity)}
+\NormalTok{observed\_disparity \textless{}{-}}\StringTok{ }\KeywordTok{unlist}\NormalTok{(}\KeywordTok{get.disparity}\NormalTok{(}\KeywordTok{rescale.dispRity}\NormalTok{(observed\_disparity)))}
+
+\CommentTok{\#\# Plotting the results with the observed disparity}
+\KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow =} \KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{,}\DecValTok{2}\NormalTok{))}
+\KeywordTok{plot}\NormalTok{(random\_extinction\_disparity, }\DataTypeTok{main =} \StringTok{"Random extinction"}\NormalTok{, }\DataTypeTok{ylab =} \StringTok{"Scaled sum of variances"}\NormalTok{)}
+\KeywordTok{abline}\NormalTok{(}\DataTypeTok{v =} \DecValTok{3}\NormalTok{, }\DataTypeTok{col =} \StringTok{"red"}\NormalTok{)}
+\KeywordTok{lines}\NormalTok{(}\DataTypeTok{x =} \DecValTok{1}\OperatorTok{:}\DecValTok{7}\NormalTok{, }\DataTypeTok{y =}\NormalTok{ observed\_disparity, }\DataTypeTok{lty =} \DecValTok{2}\NormalTok{, }\DataTypeTok{lwd =} \DecValTok{2}\NormalTok{)}
+\KeywordTok{plot}\NormalTok{(selective\_extinction\_disparity, }\DataTypeTok{main =} \StringTok{"Selective extinction"}\NormalTok{, }\DataTypeTok{ylab =} \StringTok{""}\NormalTok{)}
+\KeywordTok{abline}\NormalTok{(}\DataTypeTok{v =} \DecValTok{3}\NormalTok{, }\DataTypeTok{col =} \StringTok{"red"}\NormalTok{)}
+\KeywordTok{lines}\NormalTok{(}\DataTypeTok{x =} \DecValTok{1}\OperatorTok{:}\DecValTok{7}\NormalTok{, }\DataTypeTok{y =}\NormalTok{ observed\_disparity, }\DataTypeTok{lty =} \DecValTok{2}\NormalTok{, }\DataTypeTok{lwd =} \DecValTok{2}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{figure}
+\centering
+\includegraphics{treats_paper_files/figure-latex/simdisparity-1.pdf}
+\caption{Comparison of the observed sum of variances (dashed line) to
+the simulated one (ligth and dark grey polygons and black line
+representing respectively the 95\% and 50\% confidence interval and the
+median value).}
+\end{figure}
+
+From these results we can draw some preliminary conclusions: the
+observed change in disparity is more likely due to a selective mass
+extinction rather than a random one (Fig. 6). This is of course a very
+crude way of testing this, a more rigorous approach is needed to answer
+the question: more and better quality data, and more thorough methods
+(e.g.~using a rank envelope test Murrell (2018)).
+
 \hypertarget{additional-information}{%
@@
 
-\hypertarget{manuals-and-vignette}{%
-\subsection{Manuals and vignette}\label{manuals-and-vignette}}
+\hypertarget{manuals-and-vignettes}{%
+\subsection{Manuals and vignettes}\label{manuals-and-vignettes}}
 
@@
 in a gitbook format: the
-\href{http://tguillerme.github.io/treats.html}{\texttt{treats} manual}.
-This manual is designed so that it can be regularly updated and enhanced
-through the lifetime of the package facilitating the interface between
-methods development and usage (Cooper, Thomas, and FitzJohn 2016).
+\href{http://tguillerme.github.io/treats.html}{\texttt{treats} manual}
+(\url{http://tguillerme.github.io/treats.html}. This manual is designed
+so that it can be regularly updated and enhanced through the lifetime of
+the package facilitating the interface between methods development and
+usage (Cooper, Thomas, and FitzJohn 2016).
 
@@
 However, I intend to continuously develop this package. For example
-future planned versions will include: lineage specific simulations;
-abiotic events; time dependent simulations and; a better integration
-with the \texttt{dispRity} package. This will be done while keeping
-track change (through \texttt{NEWS} file), continuous integration and
-unit testing. Furthermore, I am planning to develop the GitHub page
-associated with the package to encourage users to share their simulation
-templates and build library from them to facilitate the work for future
-users.
+future planned versions will include abiotic events and a better
+integration with the \texttt{dispRity} package. This will be done while
+keeping track change (through \texttt{NEWS} file), continuous
+integration and unit testing. Furthermore, I am planning to develop the
+GitHub page associated with the package to encourage users to share
+their simulation templates and build library from them to facilitate the
+work for future users.
 
@@
 This paper is entirely reproducible from an Rmarkdown document available
-on \href{https://github.com/TGuillerme/treats/paper/}{GitHub}. The data
-used for the example above (Beck and Lee 2014) is available from the
-\texttt{dispRity} package (Guillerme 2018).
+on \href{https://github.com/TGuillerme/treats/inst/paper.Rmd}{GitHub}.
+The data used for the example above (Beck and Lee 2014) is available
+from the \texttt{dispRity} package (Guillerme 2018).
 
@@
 Rule, Sophie Ryan and Gavin Thomas, for support and useful comments on
-late stages of the development of this package. This work was funded by
-UKRI-NERC Grant NE/T000139/1.
+late stages of the development of this package. Thanks to William Gearty,
+one anonymous reviewer and one anonymous associate editor for their very
+useful suggestions. This work was funded by UKRI-NERC Grant NE/T000139/1
+awarded to Gavin Thomas.
 

### 2023-11-22 inst/gitbook/04_make.events.Rmd
change time.condition to age.condition
For bypassing s3 issues on the CRAN

@@
 ---------|-----------|-------|----------------
-`time.condition`, `taxa.condition`, `trait.condition` | x | the value to check the target against | e.g. time = x, taxa = x or trait = x (see below)
+`age.condition`, `taxa.condition`, `trait.condition` | x | the value to check the target against | e.g. time = x, taxa = x or trait = x (see below)
  | condition | a relational operator | e.g. `==`, `>`, `!=`, etc. for asking for example `time == x` 
@@
 
- * `time.condition` will trigger the `event` once a certain time is reached in the simulations and is the simplest/most basic condition with no arguments other than the time required.
-For example ``time.condition(4, condition = `==`)`` will trigger the `event` once the simulations reaches 4 time units.
+ * `age.condition` will trigger the `event` once a certain time is reached in the simulations and is the simplest/most basic condition with no arguments other than the time required.
+For example ``age.condition(4, condition = `==`)`` will trigger the `event` once the simulations reaches 4 time units.
 Easy.
@@
                       target = "taxa",
-                      condition = time.condition(4),
+                      condition = age.condition(4),
                       modification = random.extinction(0.8))

### 2023-11-22 inst/gitbook/07_specific_examples.Rmd
updated manuals with returns
change time.condition to age.condition
For bypassing s3 issues on the CRAN

@@
 --------------|----------
-`time.condition` | [mass extinction](#EGrandom_ext), [negative trait extinction](#EGneg_ext), [reducing speciation rate](#EGred_spec), [changing trait process](#EG_change_trait), [changing modifiers](#EG_change_modif)
+`age.condition` | [mass extinction](#EGrandom_ext), [negative trait extinction](#EGneg_ext), [reducing speciation rate](#EGred_spec), [changing trait process](#EG_change_trait), [changing modifiers](#EG_change_modif)
 `random.extinction` | [mass extinction](#EGrandom_ext)
@@
 
-We will then create a event that triggers when reaching half of the simulation time (using `time.condition(2.5)`).
+We will then create a event that triggers when reaching half of the simulation time (using `age.condition(2.5)`).
 This event will target `"taxa"`, i.e. the number of species, and randomly remove 80% of them (using `random.extinction(0.8)`):
@@
 mass.extinction <- make.events(
-                    condition = time.condition(4),
+                    condition = age.condition(4),
                     target = "taxa",
@@
 
-And our extinction event which triggers after reaching time 4 (`time.condition(4)`), targets the `"taxa"` and modifies the extinction for species with traits lower than 0.
+And our extinction event which triggers after reaching time 4 (`age.condition(4)`), targets the `"taxa"` and modifies the extinction for species with traits lower than 0.
 
@@
                       target = "taxa",
-                      condition = time.condition(4),
+                      condition = age.condition(4),
                       modification = trait.extinction(x = 0,
@@
 
-And our extinction event which triggers after reaching time 4 (`time.condition(4)`), targets the `"bd.params"` and modifies the speciation rate to 1/3:
+And our extinction event which triggers after reaching time 4 (`age.condition(4)`), targets the `"bd.params"` and modifies the speciation rate to 1/3:
 
@@
 reduced.speciation <- make.events(
-                      condition = time.condition(4),
+                      condition = age.condition(4),
                       target = "bd.params",
@@
 
-And our event which triggers after reaching time 5 (`time.condition(5)`), targets the `"traits"` and modifies the process to `OU.process`.
+And our event which triggers after reaching time 5 (`age.condition(5)`), targets the `"traits"` and modifies the process to `OU.process`.
 
@@
 change.process.to.OU <- make.events(
-                  condition    = time.condition(5),
+                  condition    = age.condition(5),
                   target       = "traits",
@@
 
-## Event for changing a modifier: speciation event increase for species with negative values {#EG_change_modif}
+## Event for changing a modifier: extinction event increase for species with negative values {#EG_change_modif}
 
@@
 
-And our extinction event which triggers after reaching time 3 (`time.condition(3)`), targets the `"modifiers"` and modifies birth-death rule as follows:
+And our extinction event which triggers after reaching time 3 (`age.condition(3)`), targets the `"modifiers"` and modifies birth-death rule as follows:
  * When a species is descendant from a parent with a negative trait value (`negative.trait.condition`);
@@
 change.speciation <- make.events(
-    condition    = time.condition(3),
+    condition    = age.condition(3),
     target       = "modifiers",

### 2023-11-22 inst/gitbook/_book/treats_manual.tex
change time.condition to age.condition
For bypassing s3 issues on the CRAN

@@
 \author{Thomas Guillerme (\href{mailto:guillert@tcd.ie}{\nolinkurl{guillert@tcd.ie}})}
-\date{2023-11-21}
+\date{2023-11-22}
 
@@
 ## Loading required package: dispRity
-\end{verbatim}
-
-\begin{verbatim}
-##        --- dispRity package ---
-## This is the CRAN release version (1.7.0) of the package.
-## For news, vignettes and future releases,
-## visit https://github.com/TGuillerme/dispRity
 \end{verbatim}
@@
 ## }
-## <bytecode: 0x5647036c6610>
+## <bytecode: 0x559776bed260>
 ## <environment: namespace:treats>
@@
 ## }
-## <bytecode: 0x564707510428>
+## <bytecode: 0x559775650390>
 ## <environment: namespace:treats>
@@
 ## }
-## <bytecode: 0x5647075127a8>
+## <bytecode: 0x559775658740>
 ## <environment: namespace:treats>
@@
 ## }
-## <bytecode: 0x56470751e620>
+## <bytecode: 0x559775657e10>
 ## <environment: namespace:treats>
@@
 \begin{minipage}[t]{0.19\columnwidth}\raggedright
-\texttt{time.condition}, \texttt{taxa.condition}, \texttt{trait.condition}\strut
+\texttt{age.condition}, \texttt{taxa.condition}, \texttt{trait.condition}\strut
 \end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
@@
 \item
-  \texttt{time.condition} will trigger the \texttt{event} once a certain time is reached in the simulations and is the simplest/most basic condition with no arguments other than the time required.
-For example \texttt{time.condition(4,\ condition\ =\ \textasciigrave{}==\textasciigrave{})} will trigger the \texttt{event} once the simulations reaches 4 time units.
+  \texttt{age.condition} will trigger the \texttt{event} once a certain time is reached in the simulations and is the simplest/most basic condition with no arguments other than the time required.
+For example \texttt{age.condition(4,\ condition\ =\ \textasciigrave{}==\textasciigrave{})} will trigger the \texttt{event} once the simulations reaches 4 time units.
 Easy.
@@
                       \DataTypeTok{target =} \StringTok{"taxa"}\NormalTok{,}
-                      \DataTypeTok{condition =} \KeywordTok{time.condition}\NormalTok{(}\DecValTok{4}\NormalTok{),}
+                      \DataTypeTok{condition =} \KeywordTok{age.condition}\NormalTok{(}\DecValTok{4}\NormalTok{),}
                       \DataTypeTok{modification =} \KeywordTok{random.extinction}\NormalTok{(}\FloatTok{0.8}\NormalTok{))}
@@
 
-\includegraphics[width=6.5in]{../../../../../../tmp/RtmptzertD/file14f8c6a285c1a}
+\includegraphics[width=6.5in]{../../../../../../tmp/RtmpqXtl0Y/file8c965577e651}
 
@@
 
-\includegraphics[width=6.5in]{../../../../../../tmp/RtmptzertD/file14f8c301a0acb}
+\includegraphics[width=6.5in]{../../../../../../tmp/RtmpqXtl0Y/file8c9618f635d6}
 
@@
 \begin{minipage}[t]{0.55\columnwidth}\raggedright
-\texttt{time.condition}\strut
+\texttt{age.condition}\strut
 \end{minipage} & \begin{minipage}[t]{0.39\columnwidth}\raggedright
@@
 
-We will then create a event that triggers when reaching half of the simulation time (using \texttt{time.condition(2.5)}).
+We will then create a event that triggers when reaching half of the simulation time (using \texttt{age.condition(2.5)}).
 This event will target \texttt{"taxa"}, i.e.~the number of species, and randomly remove 80\% of them (using \texttt{random.extinction(0.8)}):
@@
 \NormalTok{mass.extinction \textless{}{-}}\StringTok{ }\KeywordTok{make.events}\NormalTok{(}
-                    \DataTypeTok{condition =} \KeywordTok{time.condition}\NormalTok{(}\DecValTok{4}\NormalTok{),}
+                    \DataTypeTok{condition =} \KeywordTok{age.condition}\NormalTok{(}\DecValTok{4}\NormalTok{),}
                     \DataTypeTok{target =} \StringTok{"taxa"}\NormalTok{,}
@@
 
-And our extinction event which triggers after reaching time 4 (\texttt{time.condition(4)}), targets the \texttt{"taxa"} and modifies the extinction for species with traits lower than 0.
+And our extinction event which triggers after reaching time 4 (\texttt{age.condition(4)}), targets the \texttt{"taxa"} and modifies the extinction for species with traits lower than 0.
 
@@
                       \DataTypeTok{target =} \StringTok{"taxa"}\NormalTok{,}
-                      \DataTypeTok{condition =} \KeywordTok{time.condition}\NormalTok{(}\DecValTok{4}\NormalTok{),}
+                      \DataTypeTok{condition =} \KeywordTok{age.condition}\NormalTok{(}\DecValTok{4}\NormalTok{),}
                       \DataTypeTok{modification =} \KeywordTok{trait.extinction}\NormalTok{(}\DataTypeTok{x =} \DecValTok{0}\NormalTok{,}
@@
 
-And our extinction event which triggers after reaching time 4 (\texttt{time.condition(4)}), targets the \texttt{"bd.params"} and modifies the speciation rate to 1/3:
+And our extinction event which triggers after reaching time 4 (\texttt{age.condition(4)}), targets the \texttt{"bd.params"} and modifies the speciation rate to 1/3:
 
@@
 \NormalTok{reduced.speciation \textless{}{-}}\StringTok{ }\KeywordTok{make.events}\NormalTok{(}
-                      \DataTypeTok{condition =} \KeywordTok{time.condition}\NormalTok{(}\DecValTok{4}\NormalTok{),}
+                      \DataTypeTok{condition =} \KeywordTok{age.condition}\NormalTok{(}\DecValTok{4}\NormalTok{),}
                       \DataTypeTok{target =} \StringTok{"bd.params"}\NormalTok{,}
@@
 
-And our event which triggers after reaching time 5 (\texttt{time.condition(5)}), targets the \texttt{"traits"} and modifies the process to \texttt{OU.process}.
+And our event which triggers after reaching time 5 (\texttt{age.condition(5)}), targets the \texttt{"traits"} and modifies the process to \texttt{OU.process}.
 
@@
 \NormalTok{change.process.to.OU \textless{}{-}}\StringTok{ }\KeywordTok{make.events}\NormalTok{(}
-                  \DataTypeTok{condition    =} \KeywordTok{time.condition}\NormalTok{(}\DecValTok{5}\NormalTok{),}
+                  \DataTypeTok{condition    =} \KeywordTok{age.condition}\NormalTok{(}\DecValTok{5}\NormalTok{),}
                   \DataTypeTok{target       =} \StringTok{"traits"}\NormalTok{,}
@@
 
-\includegraphics[width=6.5in]{../../../../../../tmp/RtmptzertD/file14f8c465f7d38}
+\includegraphics[width=6.5in]{../../../../../../tmp/RtmpqXtl0Y/file8c9623bb5976}
 
@@
 
-And our extinction event which triggers after reaching time 3 (\texttt{time.condition(3)}), targets the \texttt{"modifiers"} and modifies birth-death rule as follows:
+And our extinction event which triggers after reaching time 3 (\texttt{age.condition(3)}), targets the \texttt{"modifiers"} and modifies birth-death rule as follows:
 * When a species is descendant from a parent with a negative trait value (\texttt{negative.trait.condition});
@@
 \NormalTok{change.speciation \textless{}{-}}\StringTok{ }\KeywordTok{make.events}\NormalTok{(}
-    \DataTypeTok{condition    =} \KeywordTok{time.condition}\NormalTok{(}\DecValTok{3}\NormalTok{),}
+    \DataTypeTok{condition    =} \KeywordTok{age.condition}\NormalTok{(}\DecValTok{3}\NormalTok{),}
     \DataTypeTok{target       =} \StringTok{"modifiers"}\NormalTok{,}

### 2023-11-22 inst/treats_paper.Rmd
change time.condition to age.condition
For bypassing s3 issues on the CRAN

--- 2023-11-22 paper/treats_paper.Rmd
@@
     target       = "taxa",
-    condition    = time.condition(15),
+    condition    = age.condition(15),
     modification = trait.extinction(x = 0, condition = `>=`))
@@
 This is indicated using the `target = "taxa"` argument.
-Then the event is triggered using the argument `time.condition(15)` and modifies the internal `lineage` object using the ``trait.extinction(x = 0, condition = `>=`)`` function.
-`time.condition` and `trait.extinction` are both function factories which are not gonna be detailed here (see @wickham2019advanced).
+Then the event is triggered using the argument `age.condition(15)` and modifies the internal `lineage` object using the ``trait.extinction(x = 0, condition = `>=`)`` function.
+`age.condition` and `trait.extinction` are both function factories which are not gonna be detailed here (see @wickham2019advanced).
 Effectively these arguments can be passed directly as standard functions.
@@
     target       = "taxa",
-    condition    = time.condition(15),
+    condition    = age.condition(15),
     modification = random.extinction(0.75))
@@
     target = "taxa",
-    condition = time.condition(15),
+    condition = age.condition(15),
     modification = trait.extinction(x = 0, condition = `>=`))

### 2023-11-22 R/birth.death_fun.R

@@ -35,7 +35,7 @@ trigger.events <- function(one_event, bd.params, lineage, trait.values, time) {
     return(one_event$condition(bd.params = bd.params, lineage = lineage, trait.values = trait.values, time = time) && one_event$trigger < 1L)
 }
 ## Creates a snapshot (creates a snapshot of the data (lineage or/and traits) at a specific time): single updates
-update.single.nodes <- function(lineage) {
+bd.update.single.nodes <- function(lineage) {
     ## Creating the new lineage
     new_lineage <- lineage
     ## Adding the living species as new parents
@@ -57,7 +57,7 @@ update.single.nodes <- function(lineage) {
     ## Done
     return(new_lineage)
 }
-update.single.edges <- function(time, time.slice, lineage, edge_lengths) {
+bd.update.single.edges <- function(time, time.slice, lineage, edge_lengths) {
     ## Calculate the difference at the split time
     diff <- time - time.slice
     ## Updating the edge lengths
@@ -68,7 +68,7 @@ update.single.edges <- function(time, time.slice, lineage, edge_lengths) {
     edge_lengths_out[lineage$livings] <- diff
     return(edge_lengths_out)
 }
-update.single.traits <- function(trait_values, traits, lineage, edge_lengths) {
+bd.update.single.traits <- function(trait_values, traits, lineage, edge_lengths) {
     ## Simulate the traits for all the singles
     snap_traits <- do.call(cbind,
                         lapply(traits,
@@ -123,7 +123,7 @@ birth.death.tree.traits <- function(stop.rule, bd.params, traits = NULL, modifie
     # }
     # modifiers <- make.modifiers(selection = select.scale.to.absolute.trait.value)
     # events <- make.events(target = "taxa",
-    #                       condition = time.condition(1),
+    #                       condition = age.condition(1),
     #                       modification = trait.extinction(0))
 
     # null.error <- FALSE
@@ -294,10 +294,10 @@ birth.death.tree.traits <- function(stop.rule, bd.params, traits = NULL, modifie
                 time.slice <- first_waiting_time + current.save.steps[1]
                 
                 ## Save a step by creating singles
-                lineage      <- update.single.nodes(lineage)
-                edge_lengths <- update.single.edges(time, time.slice, lineage, edge_lengths)
+                lineage      <- bd.update.single.nodes(lineage)
+                edge_lengths <- bd.update.single.edges(time, time.slice, lineage, edge_lengths)
                 if(do_traits) {
[…]

### 2023-11-22 R/dispRitreats.R

@@ -13,10 +13,10 @@
 #' 
 #' Common optional arguments for the following arguments include the following (refer the the specific function for the arguments details):
 #' \itemize{
-#'  \item code{\link[dispRity]{custom.subsets}}: \code{group} for the list of elements to be attributed to specific groups;
-#'  \item code{\link[dispRity]{chrono.subsets}}: \code{method} for selecting the time binning or slicing method; \code{time} for the number of time bins/slices or their specific ages; \code{model} for the time slicing method; or \code{inc.nodes} for whether to include nodes or not in the time subsets;
-#'  \item code{\link[dispRity]{boot.matrix}}: \code{bootstraps} for the number of bootstrap replicates; \code{rarefaction} for the number of elements to include in each bootstrap replicate; or \code{boot.type} for the bootstrap algorithm;
-#'  \item code{\link[dispRity]{dispRity}}: \code{metric} for the disparity, dissimilarity or spatial occupancy metric to apply to the data; or \code{dimensions} for the number of dimensions to consider.
+#'  \item \code{\link[dispRity]{custom.subsets}}: \code{group} for the list of elements to be attributed to specific groups;
+#'  \item \code{\link[dispRity]{chrono.subsets}}: \code{method} for selecting the time binning or slicing method; \code{time} for the number of time bins/slices or their specific ages; \code{model} for the time slicing method; or \code{inc.nodes} for whether to include nodes or not in the time subsets;
+#'  \item \code{\link[dispRity]{boot.matrix}}: \code{bootstraps} for the number of bootstrap replicates; \code{rarefaction} for the number of elements to include in each bootstrap replicate; or \code{boot.type} for the bootstrap algorithm;
+#'  \item \code{\link[dispRity]{dispRity}}: \code{metric} for the disparity, dissimilarity or spatial occupancy metric to apply to the data; or \code{dimensions} for the number of dimensions to consider.
 #' }
 #'
 #' @return

### 2023-11-22 R/events.conditions.R

@@ -1,5 +1,5 @@
 #' @name events.conditions
-#' @aliases time.condition taxa.condition trait.condition
+#' @aliases age.condition taxa.condition trait.condition
 #' @title events.conditions
 #'
 #' @description Inbuilt conditions functions for helping designing events
@@ -14,7 +14,7 @@
 #' The following functions allow to design specific conditions for events:
 #' 
 #' \itemize{
-#'      \item{\code{time.condition}}: a conditional function based on the time \code{x}. Typically this can be translated into "when time reaches the value x, trigger a condition" (see \code{\link{make.events}}). There is no optional argument for the function.
+#'      \item{\code{age.condition}}: a conditional function based on the time \code{x}. Typically this can be translated into "when time reaches the value x, trigger a condition" (see \code{\link{make.events}}). There is no optional argument for the function.
 #' 
 #'      \item{\code{taxa.condition}}: a conditional function based on the number of taxa \code{x}. Typically this can be  translated into "when the number of taxa reaches the value x, trigger a condition" (see \code{\link{make.events}}). This function has one optional argument:
 #'      \itemize{
@@ -37,7 +37,7 @@
 #' ## 80% mass extinction at time 4
 #' mass_extinction <- make.events(
 #'                       target       = "taxa",
-#'                       condition    = time.condition(4),
+#'                       condition    = age.condition(4),
 #'                       modification = random.extinction(0.8))
 #' 
 #' ## Set the simulation parameters
@@ -94,13 +94,18 @@
 events.condition <- function(x, condition, ...) {
     cat("List of inbuilt condition functions in treats:\n")
     cat("   ?taxa.condition\n")
-    cat("   ?time.condition\n")
+    cat("   ?age.condition\n")
     cat("   ?trait.condition\n")
     return(invisible())
 } 
 
+## time.condition alias
+time.condition <- function(x, ...) {
+    age.condition(x, ...)
+}
+
 ## A condition based on a specific time value \code{x}
-time.condition <- function(x, condition = `>`) {
+age.condition <- function(x, condition = `>`) {
     return(function(bd.params, lineage, trait.values, time) {condition(time, x)})
 }
 

### 2023-11-22 R/events.modifications.R

@@ -53,7 +53,7 @@
 #' ## 80% mass extinction at time 4
 #' mass_extinction <- make.events(
 #'                       target       = "taxa",
-#'                       condition    = time.condition(4),
+#'                       condition    = age.condition(4),
 #'                       modification = random.extinction(0.8))
 #' 
 #' ## Set the simulation parameters

### 2023-11-22 R/make.events.R

@@ -28,7 +28,7 @@
 #' ## 80% mass extinction at time 4
 #' mass_extinction <- make.events(
 #'                       target       = "taxa",
-#'                       condition    = time.condition(4),
+#'                       condition    = age.condition(4),
 #'                       modification = random.extinction(0.8))
 #' 
 #' ## Set the simulation parameters

### 2023-11-22 R/make.traits.R

@@ -4,12 +4,12 @@
 #'
 #' @param process      The trait process(es) (default is \code{\link{BM.process}}).
 #' @param n            Optional, the number of traits per process (default is \code{1}).
-#' @param start        Optional, the starting values for each traits (default is {0}).
+#' @param start        Optional, the starting values for each traits (default is \code{0}).
 #' @param process.args Optional, a named list of optional arguments for the trait process.
 #' @param trait.names  Optional, the name(s) of the process(s).
 #' @param add          Optional, another previous \code{"treats"} traits object to which to add the trait.
 #' @param update       Optional, another previous \code{"treats"} traits object to update (see details).
-#' @param background    Optional, another \code{"treats"} \code{"traits"} object to simulate background trait evolution (see details).
+#' @param background   Optional, another \code{"treats"} \code{"traits"} object to simulate background trait evolution (see details).
 #' @param test         Logical, whether to test if the traits object will work with \code{\link{treats}} (\code{TRUE} - default).
 #' 
 #' 

### 2023-11-22 R/make.treats.R

@@ -7,7 +7,7 @@
 # @param ... additional \code{treats} objects to add (e.g. \code{traits}).
 #'
 #' @return
-#' This function outputs a \code{treats} object that is a list of at least two elements: \code{$tree}, a \code{"phylo"} object and \code{$data}, a code{"matrix"} of the trait values.
+#' This function outputs a \code{treats} object that is a list of at least two elements: \code{$tree}, a \code{"phylo"} object and \code{$data}, a \code{"matrix"} of the trait values.
 #'
 #' @examples
 #' ## Creating a random tree

### 2023-11-22 R/plot.lineage.R

@@ -7,7 +7,7 @@
 # @param edge_lengths  Optional, the edge_lengths
 # @param element.names logical, whether to display the elements as named in lineage (TRUE) or as named in the output tree (FALSE)
 # @param plot logical, also plots the tree (TRUE) or not (FALSE)
-plot.lineage <- function(lineage, edge_lengths, element.names = TRUE, plot = TRUE) {
+internal.plot.lineage <- function(lineage, edge_lengths, element.names = TRUE, plot = TRUE) {
 
     ## Missing edge lenghts
     if(missing(edge_lengths)) {

### 2023-11-22 R/plot.treats.R

@@ -116,7 +116,7 @@ plot.treats <- function(x, col, ..., trait = 1, edges = "grey", tips.nodes = NUL
             }
 
             ## Plotting the results
-            plot.simulation(data = replicate(simulations, 
+            internal.plot.simulation(data = replicate(simulations, 
                                              sim.motion(one_trait, steps = 100),
                                              simplify = FALSE),
                             col = col, use.3D = use.3D, trait = trait_ids, trait.name = trait_name, quantiles = quantiles, cent.tend = cent.tend, ...)

### 2023-11-22 R/plot.treats_fun.R

@@ -28,7 +28,7 @@ sim.motion <- function(one_trait, steps) {
 # ' @param quantiles the quantiles to display (default is \code{c(95, 50)})
 # ' @param ... any additional argument to be passed to \code{\link[graphics]{plot}}.
 
-plot.simulation <- function(data, cent.tend, quantiles, col, use.3D, trait, trait.name, ...) {
+internal.plot.simulation <- function(data, cent.tend, quantiles, col, use.3D, trait, trait.name, ...) {
     
     ## Whether to use 3D plots or not
     is_1D <- ncol(data[[1]]) == 1

### 2023-11-22 R/print.treats.R

@@ -38,22 +38,22 @@ print.treats <- function(x, all = FALSE, ...) {
             switch(class(x)[[2]],
                 traits = {
                     cat(" ---- treats traits object ---- \n")
-                    print.traits.info(x$main)
+                    internal.print.traits.info(x$main)
                     if(!is.null(x$background)) {
                         cat("And a background trait (see x$background for info).")
                     }
                 },
                 events = {
                     cat(" ---- treats events object ---- \n")
-                    lapply(x, print.events.info)
+                    lapply(x, internal.print.events.info)
                 },
                 modifiers = {
                     cat(" ---- treats modifiers object ---- \n")
-                    print.modifiers.info(x)
+                    internal.print.modifiers.info(x)
                 },
                 bd.params = {
                     cat(" ---- treats birth-death parameters object ---- \n")
-                    print.bd.params.info(x)
+                    internal.print.bd.params.info(x)
                 }
                 )
             return(invisible())
@@ -79,13 +79,13 @@ print.treats <- function(x, all = FALSE, ...) {
                 ifelse(!is.null(x$events), paste0(ifelse(!is.null(x$modifiers), " and", ""), " events"), ""), 
                 ":\n"))
             if(!is.null(x$bd.params)) {
-                print.bd.params.info(x$bd.params)
+                internal.print.bd.params.info(x$bd.params)
             } 
             if(!is.null(x$modifiers)) {
-                print.modifiers.info(x$modifiers)
+                internal.print.modifiers.info(x$modifiers)
             } 
             if(!is.null(x$events)) {
-                lapply(x$events, print.events.info)
+                lapply(x$events, internal.print.events.info)
             }
             # cat("\n")
         }
@@ -107,7 +107,7 @@ print.treats <- function(x, all = FALSE, ...) {
                 cat("Simulated trait data")
             }
             cat(":\n")
-            print.traits.info(x$traits$main)
+            internal.print.traits.info(x$traits$main)
             if(!is.null(x$traits$background)) {
                 cat("And a background trait (see x$background for info).")
             }

### 2023-11-22 R/print.treats_fun.R

@@ -1,5 +1,5 @@
 ## Function for printing the traits info
-print.traits.info <- function(x) {
+internal.print.traits.info <- function(x) {
     ## Get the trait object info
     n_processes   <- length(x)
     trait_process <- lapply(x, function(X) X$trait_id)
@@ -39,7 +39,7 @@ print.traits.info <- function(x) {
 }
 
 ## Function for printing the modifiers info
-print.modifiers.info <- function(x) {
+internal.print.modifiers.info <- function(x) {
 
     ## Check of is null or default
     not.null.default <- function(x) {
@@ -104,7 +104,7 @@ print.modifiers.info <- function(x) {
 }
 
 ## Function for printing the events info
-print.events.info <- function(x) {
+internal.print.events.info <- function(x) {
     cat(paste0(ifelse(is.null(x$call$event.name), "E", paste0(x$call$event.name, " e")), "vent targeting \"", x$target, "\" to be triggered ", x$trigger+1, " time", ifelse(x$trigger < 0, "s.", ".")))
     cat("\n")
     condition_fun <- if(is(x$call$condition, "name")) {x$call$condition} else {x$call$condition[[1]]}
@@ -115,7 +115,7 @@ print.events.info <- function(x) {
 }
 
 ## Function for printing bd.params info
-print.bd.params.info <- function(x) {
+internal.print.bd.params.info <- function(x) {
     if(x$joint) {
         cat("joint sampling for:\n")
     }

### 2023-11-22 R/treats.R

@@ -30,7 +30,7 @@
 #' If \code{null.error} is set to a numeric value, the function will run multiple times until a correct tree is generated. Using this option can greatly increase computational time!
 #' 
 #' @return
-#' This function outputs either a \code{"phylo"} object if no traits where generated or a \code{treats} object that is a list of at least two elements: \code{$tree}, a \code{"phylo"} object and \code{$data}, a code{"matrix"} of the trait values.
+#' This function outputs either a \code{"phylo"} object if no traits where generated or a \code{treats} object that is a list of at least two elements: \code{$tree}, a \code{"phylo"} object and \code{$data}, a \code{"matrix"} of the trait values.
 #'
 #' @examples
 #' ## Setting pure birth tree (no extinction) parameters
@@ -70,7 +70,7 @@
 #' ## 80% mass extinction at time 4
 #' mass_extinction <- make.events(
 #'                       target       = "taxa",
-#'                       condition    = time.condition(4),
+#'                       condition    = age.condition(4),
 #'                       modification = random.extinction(0.8))
 #' 
 #' ## Set the simulation parameters

### 2023-11-22 tests/testthat/test-birth.death.R

@@ -216,7 +216,7 @@ test_that("events work", {
     ## Taxa events
     ###################   
     events <- make.events(target = "taxa",
-                          condition = time.condition(4),
+                          condition = age.condition(4),
                           modification = random.extinction(0.8))
 
     set.seed(1)
@@ -232,7 +232,7 @@ test_that("events work", {
 
     ## Mass extinction based on trait values at time t
     events <- make.events(target = "taxa",
-                          condition = time.condition(4),
+                          condition = age.condition(4),
                           modification = trait.extinction(1))
 
     set.seed(7)
@@ -285,7 +285,7 @@ test_that("events work", {
 
     ## Reducing speciation after reaching time t
     events <- make.events(
-        condition    = time.condition(2),
+        condition    = age.condition(2),
         target       = "bd.params",
         modification = bd.params.update(speciation = 1/3))
     
@@ -313,7 +313,7 @@ test_that("events work", {
     stop.rule$max.time <- 6
     traits <- make.traits()
     events <- make.events(
-        condition    = time.condition(5),
+        condition    = age.condition(5),
         target       = "traits",
         modification = traits.update(process = OU.process))
     
@@ -369,7 +369,7 @@ test_that("events work", {
     new.modify    <- function(x, trait.values, lineage) return(x + 1)
     
     events <- make.events(
-        condition    = time.condition(3),
+        condition    = age.condition(3),
         target       = "modifiers",
         modification = modifiers.update(speciation = speciation, condition = new.condition, modify = new.modify))
 
@@ -570,7 +570,7 @@ test_that("single logic works", {
     time <- 2
 
     ## Testing single.nodes
-    test <- update.single.nodes(lineage_pre)
[…]

### 2023-11-22 tests/testthat/test-make.events.R

@@ -1,7 +1,7 @@
 ## Test
 test_that("make.events works", {
 
-    condition.test.ok <- time.condition(5)
+    condition.test.ok <- age.condition(5)
     condition.test.wrong <- "bob"
     modification.test.ok <- traits.update(process = OU.process)
     modification.test.wrong <- "bub"

### 2023-11-22 tests/testthat/test-print.treats.R

@@ -3,12 +3,12 @@ test_that("print.treats works for traits", {
 
     ## Simple trait
     simple <- make.traits()
-    out <- capture_output(print.traits.info(simple$main))
+    out <- capture_output(internal.print.traits.info(simple$main))
     expect_equal(out[[1]], "1 trait for 1 process (A) with one starting value (0).")
 
     ## Complex
     complex <- make.traits(process = c(BM.process, BM.process), n = c(2,3), process.args = list(list(Sigma = diag(2)), list(Sigma = matrix(1/3, 3, 3))), trait.names = c("bib", "bob"))
-    out <- capture_output(print.traits.info(complex$main))
+    out <- capture_output(internal.print.traits.info(complex$main))
     expect_equal(out[[1]], "5 traits for 2 processes (bib:2, bob:3) with one starting value (0).\nprocess bib uses the following extra argument: Sigma;\nprocess bob uses the following extra argument: Sigma;")
 
     ## Simple treats

### 2023-11-22 tests/testthat/test-treats.R

@@ -85,12 +85,12 @@ test_that("paper example works", {
     ## Creating a random mass extinction
     random_extinction <- make.events(
         target       = "taxa",
-        condition    = time.condition(15),
+        condition    = age.condition(15),
         modification = random.extinction(0.75))
     ## Creating an extinction that removes species with positive trait values
     positive_extinction <- make.events(
         target = "taxa",
-        condition = time.condition(15),
+        condition = age.condition(15),
         modification = trait.extinction(x = 0, condition = `>=`))
 
     set.seed(123)
@@ -109,7 +109,7 @@ test_that("paper example works", {
 
     ## multiple prints
     output <- capture_output(print.treats(sim_rand_extinction))
-    expect_equal(output, " ---- treats object with 10 replicates ---- \nYou can access individual replicates by extracting them, e.g. using x[[1]]\nBirth death process with events:\nspeciation: 0.352.\nextinction: 0.288.\nEvent targeting \"taxa\" to be triggered 1 time.\nThe condition function is: time.condition\nThe modification function is: random.extinction\n\nSimulated trait data:\n2 traits for 1 process (A:2) with one starting value (0).")
+    expect_equal(output, " ---- treats object with 10 replicates ---- \nYou can access individual replicates by extracting them, e.g. using x[[1]]\nBirth death process with events:\nspeciation: 0.352.\nextinction: 0.288.\nEvent targeting \"taxa\" to be triggered 1 time.\nThe condition function is: age.condition\nThe modification function is: random.extinction\n\nSimulated trait data:\n2 traits for 1 process (A:2) with one starting value (0).")
 
 
 
@@ -127,7 +127,7 @@ test_that("paper example works", {
     expect_equal(length(sim_trait_extinction), 10)
 
     output <- capture_output(print.treats(sim_trait_extinction))
-    expect_equal(output, " ---- treats object with 10 replicates ---- \nYou can access individual replicates by extracting them, e.g. using x[[1]]\nBirth death process with events:\nspeciation: 0.352.\nextinction: 0.288.\nEvent targeting \"taxa\" to be triggered 1 time.\nThe condition function is: time.condition\nThe modification function is: trait.extinction\n\nSimulated trait data:\n2 traits for 1 process (A:2) with one starting value (0).")
[…]

### 2023-11-23 inst/treats_paper.Rmd
CRAN submission fixes

@@
 This package is based on a simple birth death algorithm from which all steps can easily be modified by users.
-It also provides a tidy interface through the `treats` object, allowing users to easily run reproducible simulations.
+
+4. `treats` also provides a tidy interface through the `treats` object, allowing users to easily run reproducible simulations.
 It also comes with an extend manual regularly updated following users' questions or suggestions.
@@
 The `treats` package comes with internal documentation (e.g.
-`?treats`) but also with a thorough and extended vignette in a gitbook format: the [`treats` manual](http://tguillerme.github.io/treats.html) ([http://tguillerme.github.io/treats.html](http://tguillerme.github.io/treats.html).
+`?treats`) but also with a thorough and extended vignette in a gitbook format: the [`treats` manual](http://tguillerme.github.io/treats.html).
 This manual is designed so that it can be regularly updated and enhanced through the lifetime of the package facilitating the interface between methods development and usage [@cooper2016dark].

### 2023-11-23 paper/treats_paper_anonymous_lineno_nopath.tex
CRAN submission fixes

--- 2023-11-22 paper/treats_paper_lineno.tex
@@
 trees and traits.}
-\author{Thomas Guillerme\(^1\)}
-\date{2023-11-22}
+\author{Anonymous\(^1\)}% \author{Thomas Guillerme\(^1\)}
+\date{2023-11-23}
 
@@
 
-\(^1\) School of Biosciences, University of Sheffield, Sheffield, S10
-2TN, United Kingdom.
-\href{mailto:guillert@tcd.ie}{\nolinkurl{guillert@tcd.ie}}
+\(^1\) Anonymous % \(^1\) School of Biosciences, University of Sheffield, Sheffield, S10 2TN, United Kingdom.
+\href{mailto:anonymous}{\nolinkurl{anonymous}}% \href{mailto:guillert@tcd.ie}{\nolinkurl{guillert@tcd.ie}}
 
@@
   simple birth death algorithm from which all steps can easily be
-  modified by users. It also provides a tidy interface through the
+  modified by users.
+\item
+  \texttt{treats} also provides a tidy interface through the
   \texttt{treats} object, allowing users to easily run reproducible
@@
 \centering
-\includegraphics{../inst/gitbook/treats_structure.png}
+\includegraphics{treats_structure.png}
 \caption{\texttt{treats} package workflow: the \texttt{treats} algorithm
@@
     \DataTypeTok{target       =} \StringTok{"taxa"}\NormalTok{,}
-    \DataTypeTok{condition    =} \KeywordTok{time.condition}\NormalTok{(}\DecValTok{15}\NormalTok{),}
+    \DataTypeTok{condition    =} \KeywordTok{age.condition}\NormalTok{(}\DecValTok{15}\NormalTok{),}
     \DataTypeTok{modification =} \KeywordTok{trait.extinction}\NormalTok{(}\DataTypeTok{x =} \DecValTok{0}\NormalTok{, }\DataTypeTok{condition =} \StringTok{\textasciigrave{}}\DataTypeTok{\textgreater{}=}\StringTok{\textasciigrave{}}\NormalTok{))}
@@
 This is indicated using the \texttt{target\ =\ "taxa"} argument. Then
-the event is triggered using the argument \texttt{time.condition(15)}
-and modifies the internal \texttt{lineage} object using the
+the event is triggered using the argument \texttt{age.condition(15)} and
+modifies the internal \texttt{lineage} object using the
 \texttt{trait.extinction(x\ =\ 0,\ condition\ =\ \textasciigrave{}\textgreater{}=\textasciigrave{})}
-function. \texttt{time.condition} and \texttt{trait.extinction} are both
+function. \texttt{age.condition} and \texttt{trait.extinction} are both
 function factories which are not gonna be detailed here (see Wickham
@@
 \centering
-\includegraphics{treats_paper_files/figure-latex/prepdata-1.pdf}
+\includegraphics{prepdata-1.pdf}
 \caption{Observed phylogeny (A) and disparity through time (B) in the
@@
 \centering
-\includegraphics{treats_paper_files/figure-latex/simtree-1.pdf}
+\includegraphics{simtree-1.pdf}
 \caption{A randomly simulated tree with similar properties as the
@@
 \centering
-\includegraphics{treats_paper_files/figure-latex/simtreetraits-1.pdf}
+\includegraphics{simtreetraits-1.pdf}
 \caption{A randomly simulated tree with a 2 dimensional random Brownian
@@
     \DataTypeTok{target       =} \StringTok{"taxa"}\NormalTok{,}
-    \DataTypeTok{condition    =} \KeywordTok{time.condition}\NormalTok{(}\DecValTok{15}\NormalTok{),}
+    \DataTypeTok{condition    =} \KeywordTok{age.condition}\NormalTok{(}\DecValTok{15}\NormalTok{),}
     \DataTypeTok{modification =} \KeywordTok{random.extinction}\NormalTok{(}\FloatTok{0.75}\NormalTok{))}
@@
     \DataTypeTok{target =} \StringTok{"taxa"}\NormalTok{,}
-    \DataTypeTok{condition =} \KeywordTok{time.condition}\NormalTok{(}\DecValTok{15}\NormalTok{),}
+    \DataTypeTok{condition =} \KeywordTok{age.condition}\NormalTok{(}\DecValTok{15}\NormalTok{),}
     \DataTypeTok{modification =} \KeywordTok{trait.extinction}\NormalTok{(}\DataTypeTok{x =} \DecValTok{0}\NormalTok{, }\DataTypeTok{condition =} \StringTok{\textasciigrave{}}\DataTypeTok{\textgreater{}=}\StringTok{\textasciigrave{}}\NormalTok{))}
@@
 \centering
-\includegraphics{treats_paper_files/figure-latex/simtreetraitsevents-1.pdf}
+\includegraphics{simtreetraitsevents-1.pdf}
 \caption{Two different mass extinction events: (A) 75\% of species go
@@
 \centering
-\includegraphics{treats_paper_files/figure-latex/simdisparity-1.pdf}
+\includegraphics{simdisparity-1.pdf}
 \caption{Comparison of the observed sum of variances (dashed line) to
@@
 \href{http://tguillerme.github.io/treats.html}{\texttt{treats} manual}
-(\url{http://tguillerme.github.io/treats.html}. This manual is designed
-so that it can be regularly updated and enhanced through the lifetime of
-the package facilitating the interface between methods development and
-usage (Cooper, Thomas, and FitzJohn 2016).
+(\href{https://anonymous.4open.science/r/treats-006B}{http://tguillerme.github.io/treats.html}).
+This manual is designed so that it can be regularly updated and enhanced
+through the lifetime of the package facilitating the interface between
+methods development and usage (Cooper, Thomas, and FitzJohn 2016).
 
@@
 The \texttt{treats} package is available on
-\href{https://github.com/TGuillerme/treats}{GitHub} with more associated
-information. All the versions of the package are archived on ZENODO with
-associated
+\href{https://anonymous.4open.science/r/treats-006B}{GitHub} with more
+associated information. All the versions of the package are archived on
+ZENODO with associated
 \href{https://doi.org/10.5281/zenodo.7970384}{DOI:10.5281/zenodo.7970384}.
@@
 Rule, Sophie Ryan and Gavin Thomas, for support and useful comments on
-late stages of the development of this package. Thanks to William Gearty,
-one anonymous reviewer and one anonymous associate editor for their very
-useful suggestions. This work was funded by UKRI-NERC Grant NE/T000139/1
-awarded to Gavin Thomas.
+late stages of the development of this package. Thanks to William
+Gearty, one anonymous reviewer and one anonymous associate editor for
+their very useful suggestions. This work was funded by UKRI-NERC Grant
+NE/T000139/1 awarded to Gavin Thomas.
 

### 2023-11-23 R/dispRitreats.R

@@ -149,9 +149,9 @@ dispRitreats <- function(data, ..., scale.trees = TRUE) {
     }
 
     ## Apply the function for each dataset
-    if(verbose) cat("Calculating disparity:")
+    if(verbose) message("Calculating disparity:", appendLF = FALSE)
     disparity_list <- lapply(data, apply.dispRity, group_type, subset_args, do_bootstrap, boot_args, dispRity_args, verbose)
-    if(verbose) cat("Done.")
+    if(verbose) message("Done.")
 
     ## Single simulation
     if(length(disparity_list) == 1) {
@@ -211,7 +211,7 @@ apply.dispRity <- function(one_simulation, group_type, subset_args, do_bootstrap
     dispRity_args$data <- disparity_data
     disparity_data <- do.call(dispRity, dispRity_args)
 
-    if(verbose) cat(".")
+    if(verbose) message(".", appendLF = FALSE)
 
     return(disparity_data)
 }

### 2023-11-23 R/drop.things.R

@@ -14,6 +14,9 @@
 #' 
 #' @details \emph{NOTE} that dropping living or fossils species DOES NOT drop associated internal nodes and edge lengths. To drop both fossil/living taxa AND internal nodes, you can use for example: \code{drop.things(drop.things(my_data, what = "fossils"), what = "singles")}.
 #'
+#' @return
+#' This function outputs either a \code{"phylo"} object if no traits where generated or a \code{treats} object that is a list of at least two elements: \code{$tree}, a \code{"phylo"} object and \code{$data}, a \code{"matrix"} of the trait values.
+#'
 #' @examples
 #' ## A random tree with fossils and traits and internal nodes every 0.5 times
 #' set.seed(3)

### 2023-11-23 R/events.conditions.R

@@ -10,6 +10,9 @@
 #' @param condition the logical function for triggering the condition (e.g. `<`, `==`, `!>`, etc...).
 #' @param ...       any optional argument specific for that condition (see details)
 #' 
+#' @return
+#' This function outputs a \code{"function"} to be passed to \code{\link{make.events}}.
+#'
 #' @details
 #' The following functions allow to design specific conditions for events:
 #' 
@@ -81,10 +84,10 @@
 #'                        traits    = traits,
 #'                        events    = change_process)
 #' ## Plot the results
-#' par(mfrow = c(1,2))
+#' oldpar <- par(mfrow = c(1,2))
 #' plot(no_change, ylim = c(-7, 7))
 #' plot(process_change, ylim = c(-7, 7))
-#' 
+#' par(oldpar)
 #' 
 #' @seealso \code{\link{treats}} \code{\link{make.events}} \code{\link{events.modifications}}
 #' 
@@ -92,17 +95,12 @@
 
 ## The list of conditions
 events.condition <- function(x, condition, ...) {
-    cat("List of inbuilt condition functions in treats:\n")
-    cat("   ?taxa.condition\n")
-    cat("   ?age.condition\n")
-    cat("   ?trait.condition\n")
+    message("List of inbuilt condition functions in treats:")
+    message("   ?taxa.condition")
+    message("   ?age.condition")
+    message("   ?trait.condition")
     return(invisible())
 } 
-
-## time.condition alias
-time.condition <- function(x, ...) {
-    age.condition(x, ...)
-}
 
 ## A condition based on a specific time value \code{x}
 age.condition <- function(x, condition = `>`) {

### 2023-11-23 R/events.modifications.R

@@ -9,6 +9,9 @@
 #' @param x   a numerical value to update.
 #' @param ... any specific argument for the modification (see details).
 #' 
+#' @return
+#' This function outputs a \code{"function"} to be passed to \code{\link{make.events}}.
+#'
 #' @details
 #' The following functions allow to design specific modifications for events:
 #' 
@@ -97,9 +100,10 @@
 #'                        traits    = traits,
 #'                        events    = change_process)
 #' ## Plot the results
-#' par(mfrow = c(1,2))
+#' oldpar <- par(mfrow = c(1,2))
 #' plot(no_change, ylim = c(-7, 7))
 #' plot(process_change, ylim = c(-7, 7))
+#' par(oldpar)
 #' 
 #' 
 #' @seealso \code{\link{treats}} \code{\link{make.events}} \code{\link{events.conditions}}
@@ -108,18 +112,18 @@
 
 ## The list of conditions
 events.modification <- function(x, ...) {
-    cat("List of inbuilt modification functions in treats:\n")
-    cat("For the taxa target:\n")
-    cat("   ?random.extinction\n")
-    cat("   ?trait.extinction\n")
-    cat("For the bd.params target:\n")
-    cat("   ?bd.params.update\n")
-    cat("For the traits target:\n")
-    cat("   ?traits.update\n")
-    cat("For the modifiers target:\n")
-    cat("   ?modifiers.update\n")
-    cat("For the founding target:\n")
-    cat("   ?founding.event\n")
+    message("List of inbuilt modification functions in treats:")
+    message("For the taxa target:")
+    message("   ?random.extinction")
+    message("   ?trait.extinction")
+    message("For the bd.params target:")
+    message("   ?bd.params.update")
+    message("For the traits target:")
+    message("   ?traits.update")
+    message("For the modifiers target:")
+    message("   ?modifiers.update")
+    message("For the founding target:")
+    message("   ?founding.event")
     return(invisible())
 } 
 

### 2023-11-23 R/make.events.R

@@ -72,10 +72,10 @@
 #'                        traits    = traits,
 #'                        events    = change_process)
 #' ## Plot the results
-#' par(mfrow = c(1,2))
+#' oldpar <- par(mfrow = c(1,2))
 #' plot(no_change, ylim = c(-7, 7))
 #' plot(process_change, ylim = c(-7, 7))
-#' 
+#' par(oldpar)
 #' 
 #' @seealso \code{\link{treats}} \code{\link{make.bd.params}} \code{\link{make.traits}} \code{\link{make.modifiers}} \code{\link{events.conditions}} \code{\link{events.modifications}}
 #' @author Thomas Guillerme

### 2023-11-23 R/map.traits.R

@@ -29,11 +29,12 @@
 #' no_traits <- map.traits(no_process, my_tree)
 #' 
 #' ## Plotting the topology and the different traits
-#' par(mfrow = c(2,2))
+#' oldpar <- par(mfrow = c(2,2))
 #' plot(my_tree, main = "Base topology")
 #' plot(bm_traits, main = "Mapped BM")
 #' plot(ou_traits, main = "Mapped OU")
 #' plot(no_traits, main = "Mapped normal trait")
+#' par(oldpar)
 #'
 map.traits <- function(traits, tree) {
     ## Sanitizing

### 2023-11-23 R/modifiers.R

@@ -20,6 +20,9 @@
 #' @param trait.values   A matrix containing the trait values (see details).
 #' @param modify.fun     A list of internals functions that can modified by \code{events} (see details).
 #' 
+#' @return
+#' These functions returns either \code{"numeric"} or \code{"logical"} values to be passed to \code{\link{make.modifiers}} and \code{\link{treats}}.
+#'
 #' @details
 #' \code{bd.params} can be either a named list of parameter values (e.g. \code{list("extinction" = 0, "speciation" = 1)}) but it is typically handled internally from a \code{"treats"} \code{"bd.params"} object.
 #' 
@@ -65,15 +68,15 @@
 
 
 modifiers <- function(bd.params = NULL, lineage = NULL, trait.values = NULL, modify.fun = NULL) {
-    cat("modifiers functions implemented in treats:\n")
-    cat("branch length generating functions:\n")
-    cat("   ?branch.length\n")
-    cat("   ?branch.length.trait\n")
-    cat("lineage selection functions:\n")
-    cat("   ?selection\n")
-    cat("speciation trigger functions:\n")
-    cat("   ?speciation\n")
-    cat("   ?speciation.trait\n")
+    message("modifiers functions implemented in treats:")
+    message("branch length generating functions:")
+    message("   ?branch.length")
+    message("   ?branch.length.trait")
+    message("lineage selection functions:")
+    message("   ?selection")
+    message("speciation trigger functions:")
+    message("   ?speciation")
+    message("   ?speciation.trait")
 }
 
 ## Normal branch length

### 2023-11-23 R/plot.treats.R

@@ -14,6 +14,8 @@
 #' @param quantiles if the input is a \code{treats} \code{traits}, which quantiles to plot (default are \code{c(95, 50))}).
 #' @param legend logical, whether to display the legend in 2D plots (\code{TRUE}) or not (\code{FALSE}; default)
 #' 
+#' @return No return value, plot \code{x}'s content.
+#'
 #' @details
 #' The \code{col} option can be either:
 #' \itemize{

### 2023-11-23 R/trait.process.R

@@ -10,6 +10,8 @@
 #' @param edge.length The branch length (default must be 1). This is always a single value.
 #' @param ... Any optional argument for the specific process (see details).
 #' 
+#' @return Returns one or more \code{"numeric"} value(s).
+#'
 #' @details
 #' The different trait processes implemented in treats are:
 #' 
@@ -107,13 +109,13 @@
 #' @export
 
 trait.process <- function(x0 = 0, edge.length = 1, ...) {
-    cat("Trait processes implemented in treats:\n")
-    cat("?BM.process\n")
-    cat("?discrete.process\n")
-    cat("?OU.process\n")
-    cat("?no.process\n")
-    cat("?multi.peak.process\n")
-    cat("?repulsion.process\n")
+    message("Trait processes implemented in treats:")
+    message("?BM.process")
+    message("?discrete.process")
+    message("?OU.process")
+    message("?no.process")
+    message("?multi.peak.process")
+    message("?repulsion.process")
 }
 
 ## The Brownian motion

### 2023-11-23 R/utilities.R

@@ -11,6 +11,8 @@
 #' @details
 #' This function is designed to be used internally in \code{treats} to help \code{modifiers}, \code{traits} or \code{events} objects to access the parent traits of the lineages simulated through the internal birth death algorithm. 
 #' 
+#' @return Returns one or more \code{"numeric"} values.
+#'
 #' @examples
 #' ## Speciation event is more likely if lineage's ancestor is further away from the mean trait value
 #' distance.modify <- function(x, trait.values, lineage) {
@@ -63,6 +65,8 @@ parent.traits <- function(trait.values, lineage, current = TRUE) {
 #'}
 #'
 #' If \code{rates} is a function that generates negative values or a negative value, the output transition matrix always returns absolute values.
+#'
+#' @return Returns a squared \code{"matrix"}.
 #'
 #' @examples
 #' ## A two states equal rates matrix with a rate of 1

### 2023-11-23 tests/testthat/test-dispRitreats.R

@@ -39,8 +39,8 @@ test_that("dispRitreats works", {
     expect_equal(round(summary(test)[1,3], 2)[[1]], 3.71)
 
     ## Verbose
-    tust <- capture_output(test <- dispRitreats(sim_data, metric = c(mean, centroids), centroid = 100, bootstraps = 10, time = 5, method = "continuous", model = "acctran", verbose = TRUE))
-    expect_equal(tust, "Calculating disparity:.....Done.")  
+    tust <- capture_messages(test <- dispRitreats(sim_data, metric = c(mean, centroids), centroid = 100, bootstraps = 10, time = 5, method = "continuous", model = "acctran", verbose = TRUE))
+    expect_equal(paste(tust, collapse = ""), "Calculating disparity:.....Done.\n")  
 
     ## MULTIPLE SIMULATIONS:
     ## Simple test: just metric

### 2023-11-23 tests/testthat/test-treats.R

@@ -41,7 +41,7 @@ test_that("treats works for simple birth-deaths", {
     set.seed(123)
     prints <- capture_messages(test <- treats(bd.params = list(speciation = 1, extinction = 0.9), stop.rule = list("max.living" = 10), null.error = 50))
     expect_is(test, "phylo")
-    expect_equal(length(prints), 7)
+    expect_equal(paste(prints, collapse = ""), "Building the tree:.....Done.\n")
 
     ## Long sim warning
     messages <- capture_messages(treats(stop.rule = list(max.time = 6.1)))

### 2023-12-06 inst/gitbook/02_make.traits.Rmd
book up

--- 2023-05-22 inst/gitbook/02_make.traits.Rmd
@@
 ![`make.traits`: The main inputs are (1) the `process`, one or more functions to describe the trait(s) as a function of previous state (`x0`) and the branch length (`edge.length`), (2) `n` the number of dimensions of the trait and (3) the starting value `start`.](make.traits.png)
+
+### Quick overview
+
+function | arguments | input | what does it do
+---------|-----------|-------|----------------
+`make.traits` | process* | a function | simulates the trait (see below)
+`make.traits` | n | a numeric value | the number of dimensions
+`make.traits` | start | one or more numeric values | the starting values for each dimensions
+`make.traits` | process.args | ... | anything to be passed to `process` (see below)
+`make.traits` | trait.names | a character string | the name(s) of the trait(s)
+`make.traits` | add | a `treats` object | another trait generated by `make.traits`
+`make.traits` | update | a `treats` object | another trait generated by `make.traits`
+`make.traits` | test | logical | whether to test the validity of the trait
+`make.traits` | background | a `treats` object | another trait generated by `make.traits` (to run in the background)
+---------|-----------|-------|----------------
+`process` | x0 | a numeric value | the value of the previous trait
+`process` | edge.length | a numeric value | the branch length value
+`process` | ... | ... | any other argument to be handled by the process
+
+\* non-optional arguments
 

### 2023-12-06 inst/gitbook/03_make.modifiers.Rmd
book up

--- 2023-07-07 inst/gitbook/03_make.modifiers.Rmd
@@
 The allowed arguments are detailed [here](#allowarguments).](make.modifiers.png)
+
+### Quick overview
+
+function | arguments | input | what does it do
+---------|-----------|-------|----------------
+`make.modifiers` | branch.length | a function | generating branch length (see below)
+`make.modifiers` | selection | a function | selecting a lineage (see below)
+`make.modifiers` | speciation | a function | choosing between speciation and extinction (see below)
+`make.modifiers` | condition | a function | a function that affects the condition of when to trigger the functions above (see below)
+`make.modifiers` | modify | a function | a function that modifies the value of the trigger value generated when the `condition` is met (see below)
+`make.modifiers` | add | a `treats` object | another modifier generated by `make.modifiers`
+`make.modifiers` | update | a `treats` object | another modifier generated by `make.modifiers`
+`make.modifiers` | test | logical | whether to test the validity of the modifier
+---------|-----------|-------|----------------
+`branch.length`, `selection`, `speciation`, `condition` | bd.params | a `treats` object generated by `make.bd.params` | usually handled internally (contains the either the birth-death values or their distribution)
+... | lineage | a named "lineage" list | usually handled internally (contains the current state of the simulations)
+... | trait.values | a matrix | usually handled internally (contains the current generated traits of the simulations)
 

### 2023-12-06 inst/gitbook/04_make.events.Rmd
book up

@@
 `make.events` | target* | any of the following: "taxa", "bd.params", "traits", "modifiers" or "founding" | what aspect of the simulation should the event target
- | condition* | a function | a function that affects the condition of when to trigger the event (see below)
- | modification* | a function | a function that applies the modification to the target if the condition is triggered (see below)
- | add | a `treats` object | another event generated by `make.events`
- | test | logical | whether to test the validity of the event
- | event.name | a character string | the name of the event
- | replications | a number | if possible how many times is the event allowing to be repeated
- | additional.args | ... | any additional arguments to be passed to the event
+`make.events` | condition* | a function | a function that affects the condition of when to trigger the event (see below)
+`make.events` | modification* | a function | a function that applies the modification to the target if the condition is triggered (see below)
+`make.events` | add | a `treats` object | another event generated by `make.events`
+`make.events` | test | logical | whether to test the validity of the event
+`make.events` | event.name | a character string | the name of the event
+`make.events` | replications | a number | if possible how many times is the event allowing to be repeated
+`make.events` | additional.args | ... | any additional arguments to be passed to the event
 
@@
 `age.condition`, `taxa.condition`, `trait.condition` | x | the value to check the target against | e.g. time = x, taxa = x or trait = x (see below)
- | condition | a relational operator | e.g. `==`, `>`, `!=`, etc. for asking for example `time == x` 
- | ... | any additional arguments to be passed to the event | |
+... | condition | a relational operator | e.g. `==`, `>`, `!=`, etc. for asking for example `time == x` 
+... | ... | any additional arguments to be passed to the event | |
 
@@
 ---------|-----------|-------|----------------
-random.extinction | x | a numeric value | the proportion of species to make extinct
-trait.extinction | x | a numeric value | the trait value when to trigger the extinction
- | condition | a relational operator | the relation to the trait value (e.g. `x = 1` and `condition =  >` makes species that have a trait value > 1 go extinct)
-bd.params.update | ... | any argument to be passed to `make.bd.params` | |
-traits.update | ... | any argument to be passed to `make.traits` | |
-modifiers.update | ... | any argument to be passed to `make.modifiers` | |
+`random.extinction` | x | a numeric value | the proportion of species to make extinct
+`trait.extinction` | x | a numeric value | the trait value when to trigger the extinction
+`trait.extinction` | condition | a relational operator | the relation to the trait value (e.g. `x = 1` and `condition =  >` makes species that have a trait value > 1 go extinct)
+`bd.params.update` | ... | any argument to be passed to `make.bd.params` | |
+`traits.update` | ... | any argument to be passed to `make.traits` | |
+`modifiers.update` | ... | any argument to be passed to `make.modifiers` | |
 

### 2023-12-06 inst/gitbook/_book/treats_manual.tex
book up

@@
 \author{Thomas Guillerme (\href{mailto:guillert@tcd.ie}{\nolinkurl{guillert@tcd.ie}})}
-\date{2023-11-22}
+\date{2023-12-06}
 
@@
 \begin{minipage}[t]{0.19\columnwidth}\raggedright
+\texttt{make.traits}\strut
+\end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
 n\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
+a numeric value\strut
+\end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
+the number of dimensions\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.19\columnwidth}\raggedright
+\texttt{make.traits}\strut
 \end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
-a numeric value\strut
+start\strut
 \end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
-the number of dimensions\strut
+one or more numeric values\strut
 \end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
-\strut
+the starting values for each dimensions\strut
 \end{minipage}\tabularnewline
 \begin{minipage}[t]{0.19\columnwidth}\raggedright
-start\strut
+\texttt{make.traits}\strut
 \end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
-one or more numeric values\strut
+process.args\strut
 \end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
-the starting values for each dimensions\strut
+\ldots{}\strut
 \end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
-\strut
+anything to be passed to \texttt{process} (see below)\strut
 \end{minipage}\tabularnewline
 \begin{minipage}[t]{0.19\columnwidth}\raggedright
-process.args\strut
+\texttt{make.traits}\strut
 \end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
-\ldots{}\strut
+trait.names\strut
 \end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
-anything to be passed to \texttt{process} (see below)\strut
+a character string\strut
 \end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
-\strut
+the name(s) of the trait(s)\strut
 \end{minipage}\tabularnewline
 \begin{minipage}[t]{0.19\columnwidth}\raggedright
-trait.names\strut
+\texttt{make.traits}\strut
 \end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
-a character string\strut
+add\strut
 \end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
-the name(s) of the trait(s)\strut
+a \texttt{treats} object\strut
 \end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
-\strut
+another trait generated by \texttt{make.traits}\strut
 \end{minipage}\tabularnewline
 \begin{minipage}[t]{0.19\columnwidth}\raggedright
-add\strut
+\texttt{make.traits}\strut
 \end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
+update\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
 a \texttt{treats} object\strut
+\end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
+another trait generated by \texttt{make.traits}\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.19\columnwidth}\raggedright
+\texttt{make.traits}\strut
+\end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
+test\strut
 \end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
-another trait generated by \texttt{make.traits}\strut
+logical\strut
 \end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
-\strut
+whether to test the validity of the trait\strut
 \end{minipage}\tabularnewline
 \begin{minipage}[t]{0.19\columnwidth}\raggedright
-update\strut
+\texttt{make.traits}\strut
 \end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
+background\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
 a \texttt{treats} object\strut
-\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
-another trait generated by \texttt{make.traits}\strut
 \end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
-\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.19\columnwidth}\raggedright
-test\strut
-\end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
-logical\strut
-\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
-whether to test the validity of the trait\strut
-\end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
-\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.19\columnwidth}\raggedright
-background\strut
-\end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
-a \texttt{treats} object\strut
-\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
 another trait generated by \texttt{make.traits} (to run in the background)\strut
-\end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
-\strut
 \end{minipage}\tabularnewline
@@
 \begin{minipage}[t]{0.19\columnwidth}\raggedright
+\texttt{process}\strut
+\end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
 edge.length\strut
-\end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
 a numeric value\strut
-\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
+\end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
 the branch length value\strut
-\end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
-\strut
 \end{minipage}\tabularnewline
 \begin{minipage}[t]{0.19\columnwidth}\raggedright
-\ldots{}\strut
+\texttt{process}\strut
 \end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
@@
 \end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
+\ldots{}\strut
+\end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
 any other argument to be handled by the process\strut
-\end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
-\strut
 \end{minipage}\tabularnewline
@@
 ## }
-## <bytecode: 0x559776bed260>
+## <bytecode: 0x55e979290a30>
 ## <environment: namespace:treats>
@@
 \begin{minipage}[t]{0.19\columnwidth}\raggedright
+\texttt{make.modifiers}\strut
+\end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
 selection\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
+a function\strut
+\end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
+selecting a lineage (see below)\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.19\columnwidth}\raggedright
+\texttt{make.modifiers}\strut
 \end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
+speciation\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
 a function\strut
+\end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
+choosing between speciation and extinction (see below)\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.19\columnwidth}\raggedright
+\texttt{make.modifiers}\strut
+\end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
+condition\strut
 \end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
-selecting a lineage (see below)\strut
+a function\strut
 \end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
-\strut
+a function that affects the condition of when to trigger the functions above (see below)\strut
 \end{minipage}\tabularnewline
 \begin{minipage}[t]{0.19\columnwidth}\raggedright
-speciation\strut
+\texttt{make.modifiers}\strut
 \end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
+modify\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
 a function\strut
+\end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
+a function that modifies the value of the trigger value generated when the \texttt{condition} is met (see below)\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.19\columnwidth}\raggedright
+\texttt{make.modifiers}\strut
+\end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
+add\strut
 \end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
-choosing between speciation and extinction (see below)\strut
+a \texttt{treats} object\strut
 \end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
-\strut
+another modifier generated by \texttt{make.modifiers}\strut
 \end{minipage}\tabularnewline
 \begin{minipage}[t]{0.19\columnwidth}\raggedright
-condition\strut
+\texttt{make.modifiers}\strut
 \end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
-a function\strut
+update\strut
 \end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
-a function that affects the condition of when to trigger the functions above (see below)\strut
+a \texttt{treats} object\strut
 \end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
-\strut
+another modifier generated by \texttt{make.modifiers}\strut
 \end{minipage}\tabularnewline
 \begin{minipage}[t]{0.19\columnwidth}\raggedright
-modify\strut
+\texttt{make.modifiers}\strut
 \end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
-a function\strut
+test\strut
 \end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
-a function that modifies the value of the trigger value generated when the \texttt{condition} is met (see below)\strut
+logical\strut
 \end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
-\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.19\columnwidth}\raggedright
-add\strut
-\end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
-a \texttt{treats} object\strut
-\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
-another modifier generated by \texttt{make.modifiers}\strut
-\end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
-\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.19\columnwidth}\raggedright
-update\strut
-\end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
-a \texttt{treats} object\strut
-\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
-another modifier generated by \texttt{make.modifiers}\strut
-\end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
-\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.19\columnwidth}\raggedright
-test\strut
-\end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
-logical\strut
-\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
 whether to test the validity of the modifier\strut
-\end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
-\strut
 \end{minipage}\tabularnewline
@@
 \begin{minipage}[t]{0.19\columnwidth}\raggedright
+\ldots{}\strut
+\end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
 lineage\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
+a named ``lineage'' list\strut
+\end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
+usually handled internally (contains the current state of the simulations)\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.19\columnwidth}\raggedright
+\ldots{}\strut
 \end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
-a named ``lineage'' list\strut
+trait.values\strut
 \end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
-usually handled internally (contains the current state of the simulations)\strut
+a matrix\strut
 \end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
-\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.19\columnwidth}\raggedright
-trait.values\strut
-\end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
-a matrix\strut
-\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
 usually handled internally (contains the current generated traits of the simulations)\strut
-\end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
-\strut
 \end{minipage}\tabularnewline
@@
 ## }
-## <bytecode: 0x559775650390>
+## <bytecode: 0x55e977cfc060>
 ## <environment: namespace:treats>
@@
 ## }
-## <bytecode: 0x559775658740>
+## <bytecode: 0x55e977d04410>
 ## <environment: namespace:treats>
@@
 ## }
-## <bytecode: 0x559775657e10>
+## <bytecode: 0x55e977d03ae0>
 ## <environment: namespace:treats>
@@
 \begin{minipage}[t]{0.19\columnwidth}\raggedright
+\texttt{make.events}\strut
+\end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
 condition*\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
+a function\strut
+\end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
+a function that affects the condition of when to trigger the event (see below)\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.19\columnwidth}\raggedright
+\texttt{make.events}\strut
 \end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
+modification*\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
 a function\strut
+\end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
+a function that applies the modification to the target if the condition is triggered (see below)\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.19\columnwidth}\raggedright
+\texttt{make.events}\strut
+\end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
+add\strut
 \end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
-a function that affects the condition of when to trigger the event (see below)\strut
+a \texttt{treats} object\strut
 \end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
-\strut
+another event generated by \texttt{make.events}\strut
 \end{minipage}\tabularnewline
 \begin{minipage}[t]{0.19\columnwidth}\raggedright
-modification*\strut
+\texttt{make.events}\strut
 \end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
-a function\strut
+test\strut
 \end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
-a function that applies the modification to the target if the condition is triggered (see below)\strut
+logical\strut
 \end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
-\strut
+whether to test the validity of the event\strut
 \end{minipage}\tabularnewline
 \begin{minipage}[t]{0.19\columnwidth}\raggedright
-add\strut
+\texttt{make.events}\strut
 \end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
-a \texttt{treats} object\strut
+event.name\strut
 \end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
-another event generated by \texttt{make.events}\strut
+a character string\strut
 \end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
-\strut
+the name of the event\strut
 \end{minipage}\tabularnewline
 \begin{minipage}[t]{0.19\columnwidth}\raggedright
-test\strut
+\texttt{make.events}\strut
 \end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
-logical\strut
+replications\strut
 \end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
-whether to test the validity of the event\strut
+a number\strut
 \end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
-\strut
+if possible how many times is the event allowing to be repeated\strut
 \end{minipage}\tabularnewline
 \begin{minipage}[t]{0.19\columnwidth}\raggedright
-event.name\strut
+\texttt{make.events}\strut
 \end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
-a character string\strut
+additional.args\strut
 \end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
-the name of the event\strut
+\ldots{}\strut
 \end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
-\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.19\columnwidth}\raggedright
-replications\strut
-\end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
-a number\strut
-\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
-if possible how many times is the event allowing to be repeated\strut
-\end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
-\strut
-\end{minipage}\tabularnewline
-\begin{minipage}[t]{0.19\columnwidth}\raggedright
-additional.args\strut
-\end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
-\ldots{}\strut
-\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
 any additional arguments to be passed to the event\strut
-\end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
-\strut
 \end{minipage}\tabularnewline
@@
 \begin{minipage}[t]{0.19\columnwidth}\raggedright
+\ldots{}\strut
+\end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
 condition\strut
-\end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
 a relational operator\strut
-\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
+\end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
 e.g.~\texttt{==}, \texttt{\textgreater{}}, \texttt{!=}, etc. for asking for example \texttt{time\ ==\ x}\strut
-\end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
-\strut
 \end{minipage}\tabularnewline
@@
 \end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
+\ldots{}\strut
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
 any additional arguments to be passed to the event\strut
-\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
-\strut
 \end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
@@
 \begin{minipage}[t]{0.19\columnwidth}\raggedright
-random.extinction\strut
+\texttt{random.extinction}\strut
 \end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
@@
 \begin{minipage}[t]{0.19\columnwidth}\raggedright
-trait.extinction\strut
+\texttt{trait.extinction}\strut
 \end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
@@
 \begin{minipage}[t]{0.19\columnwidth}\raggedright
+\texttt{trait.extinction}\strut
+\end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
 condition\strut
-\end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
+\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
 a relational operator\strut
-\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
+\end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
 the relation to the trait value (e.g.~\texttt{x\ =\ 1} and \texttt{condition\ =\ \ \textgreater{}} makes species that have a trait value \textgreater{} 1 go extinct)\strut
-\end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
-\strut
 \end{minipage}\tabularnewline
 \begin{minipage}[t]{0.19\columnwidth}\raggedright
-bd.params.update\strut
+\texttt{bd.params.update}\strut
 \end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
@@
 \begin{minipage}[t]{0.19\columnwidth}\raggedright
-traits.update\strut
+\texttt{traits.update}\strut
 \end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
@@
 \begin{minipage}[t]{0.19\columnwidth}\raggedright
-modifiers.update\strut
+\texttt{modifiers.update}\strut
 \end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
@@
 
-\includegraphics[width=6.5in]{../../../../../../tmp/RtmpqXtl0Y/file8c965577e651}
+\includegraphics[width=6.5in]{../../../../../../tmp/RtmpMmzXjr/file3e7f6e0175d3}
 
@@
 
-\includegraphics[width=6.5in]{../../../../../../tmp/RtmpqXtl0Y/file8c9618f635d6}
+\includegraphics[width=6.5in]{../../../../../../tmp/RtmpMmzXjr/file3e7f7aec27f9}
 
@@
 
-\includegraphics[width=6.5in]{../../../../../../tmp/RtmpqXtl0Y/file8c9623bb5976}
+\includegraphics[width=6.5in]{../../../../../../tmp/RtmpMmzXjr/file3e7f5c6b6ac6}
 

### 2023-12-13 inst/gitbook/01_getting_started.Rmd
Updated crude.bd.est with new calculations options

@@
 ## Extracting the crude parameters
-(est_params <- crude.bd.est(BeckLee_tree))
+(est_params <- crude.bd.est(BeckLee_tree, method = "count"))
 ## These parameters are the observed speciation and extinction rate per million years.

### 2023-12-13 inst/gitbook/_book/treats_manual.tex
Updated crude.bd.est with new calculations options

@@
 \author{Thomas Guillerme (\href{mailto:guillert@tcd.ie}{\nolinkurl{guillert@tcd.ie}})}
-\date{2023-12-06}
+\date{2023-12-13}
 
@@
 \CommentTok{\#\# Extracting the crude parameters}
-\NormalTok{(est\_params \textless{}{-}}\StringTok{ }\KeywordTok{crude.bd.est}\NormalTok{(BeckLee\_tree))}
+\NormalTok{(est\_params \textless{}{-}}\StringTok{ }\KeywordTok{crude.bd.est}\NormalTok{(BeckLee\_tree, }\DataTypeTok{method =} \StringTok{"count"}\NormalTok{))}
 \end{Highlighting}
@@
 ##  ---- treats birth-death parameters object ---- 
-## speciation: 0.352.
-## extinction: 0.288.
+## speciation: 0.028.
+## extinction: 0.027.
 \end{verbatim}
@@
 \begin{verbatim}
+## Building the tree:....................Done.
 ## Building the tree:....Done.
+## Building the tree:...........Done.
+## Building the tree:.Done.
+## Building the tree:............Done.
+## Building the tree:.Done.
+## Building the tree:.Done.
+## Building the tree:.Done.
+## Building the tree:..............Done.
+## Building the tree:...................................Done.
+## Building the tree:......................Done.
+## Building the tree:........Done.
+## Building the tree:...................Done.
+## Building the tree:...Done.
 ## Building the tree:.......Done.
+## Building the tree:.....Done.
+## Building the tree:....................Done.
+## Building the tree:.Done.
 ## Building the tree:...Done.
-## Building the tree:......Done.
-## Building the tree:..Done.
-## Building the tree:..Done.
-## Building the tree:.Done.
-## Building the tree:........Done.
-## Building the tree:..Done.
+## Building the tree:................Done.
+## Building the tree:.......Done.
+## Building the tree:.................Done.
 ## Building the tree:.....Done.
+## Building the tree:.................Done.
+## Building the tree:............Done.
 ## Building the tree:....Done.
 ## Building the tree:....Done.
-## Building the tree:..Done.
-## Building the tree:.................Done.
+## Building the tree:...Done.
+## Building the tree:...................................................Done.
+## Building the tree:...............Done.
 ## Building the tree:....Done.
-## Building the tree:....Done.
-## Building the tree:..Done.
-## Building the tree:......Done.
-## Building the tree:.......Done.
-## Building the tree:.Done.
-## Building the tree:.Done.
-## Building the tree:..Done.
-## Building the tree:......Done.
-## Building the tree:...........Done.
-## Building the tree:......Done.
+## Building the tree:.........Done.
+## Building the tree:..................................Done.
+## Building the tree:.....Done.
+## Building the tree:........Done.
 ## Building the tree:....Done.
 ## Building the tree:.Done.
-## Building the tree:..................Done.
-## Building the tree:....Done.
-## Building the tree:...............Done.
-## Building the tree:....Done.
-## Building the tree:..Done.
+## Building the tree:...................Done.
+## Building the tree:...........Done.
+## Building the tree:...........Done.
+## Building the tree:...........Done.
+## Building the tree:.Done.
+## Building the tree:...........Done.
+## Building the tree:.Done.
+## Building the tree:.............Done.
+## Building the tree:.............Done.
 ## Building the tree:........Done.
-## Building the tree:....Done.
-## Building the tree:........Done.
-## Building the tree:...Done.
+## Building the tree:.............Done.
 ## Building the tree:...Done.
 ## Building the tree:.Done.
-## Building the tree:...Done.
-## Building the tree:..............Done.
-## Building the tree:...Done.
-## Building the tree:.Done.
-## Building the tree:..Done.
-## Building the tree:....Done.
-## Building the tree:..Done.
-## Building the tree:..Done.
-## Building the tree:.Done.
-## Building the tree:.Done.
-## Building the tree:....Done.
-## Building the tree:..........Done.
 \end{verbatim}
@@
 \begin{verbatim}
-## Building the tree:.....................Done.
-## Building the tree:...................Done.
-## Building the tree:.....Done.
+## Building the tree:......................Done.
+## Building the tree:...Done.
+## Building the tree:..................Done.
+## Building the tree:.......Done.
+## Building the tree:.................Done.
+## Building the tree:...Done.
+## Building the tree:..................Done.
+## Building the tree:....................Done.
 ## Building the tree:.Done.
-## Building the tree:....Done.
-## Building the tree:..Done.
-## Building the tree:..Done.
-## Building the tree:......................Done.
-## Building the tree:..Done.
-## Building the tree:..Done.
-## Building the tree:...Done.
+## Building the tree:............Done.
 ## Building the tree:.Done.
-## Building the tree:..Done.
-## Building the tree:......Done.
 ## Building the tree:.Done.
 ## Building the tree:......Done.
-## Building the tree:..Done.
+## Building the tree:...............Done.
+## Building the tree:....Done.
+## Building the tree:.................Done.
+## Building the tree:.........Done.
+## Building the tree:....Done.
+## Building the tree:.....Done.
+## Building the tree:....Done.
+## Building the tree:......Done.
+## Building the tree:.....................Done.
+## Building the tree:.................Done.
 ## Building the tree:.Done.
@@
 ## Building the tree:...Done.
-## Building the tree:..Done.
-## Building the tree:......Done.
-## Building the tree:....................Done.
-## Building the tree:...Done.
-## Building the tree:.........Done.
-## Building the tree:.Done.
-## Building the tree:...Done.
-## Building the tree:.Done.
-## Building the tree:.........Done.
-## Building the tree:.Done.
-## Building the tree:.............Done.
-## Building the tree:..............Done.
-## Building the tree:......Done.
-## Building the tree:.Done.
+## Building the tree:...........................Done.
 ## Building the tree:.......Done.
-## Building the tree:...Done.
-## Building the tree:..Done.
-## Building the tree:......Done.
-## Building the tree:.Done.
-## Building the tree:..Done.
-## Building the tree:............Done.
+## Building the tree:.......Done.
 ## Building the tree:.....Done.
 ## Building the tree:......Done.
+## Building the tree:.......Done.
+## Building the tree:.........Done.
+## Building the tree:.....Done.
+## Building the tree:...Done.
+## Building the tree:...................Done.
+## Building the tree:............Done.
+## Building the tree:......................Done.
+## Building the tree:......Done.
+## Building the tree:............Done.
+## Building the tree:.............Done.
+## Building the tree:...........Done.
+## Building the tree:..Done.
+## Building the tree:.........................Done.
 ## Building the tree:.Done.
-## Building the tree:.Done.
-## Building the tree:..............Done.
-## Building the tree:........Done.
-## Building the tree:..............Done.
-## Building the tree:.Done.
-## Building the tree:..Done.
+## Building the tree:......................Done.
+## Building the tree:.............Done.
+## Building the tree:...Done.
+## Building the tree:...............Done.
+## Building the tree:..........Done.
 \end{verbatim}
@@
 ## }
-## <bytecode: 0x55e979290a30>
+## <bytecode: 0x560fcf695180>
 ## <environment: namespace:treats>
@@
 ## }
-## <bytecode: 0x55e977cfc060>
+## <bytecode: 0x560fce0ebbe0>
 ## <environment: namespace:treats>
@@
 ## }
-## <bytecode: 0x55e977d04410>
+## <bytecode: 0x560fce0eaf30>
 ## <environment: namespace:treats>
@@
 ## }
-## <bytecode: 0x55e977d03ae0>
+## <bytecode: 0x560fce0ea600>
 ## <environment: namespace:treats>
@@
 
-\includegraphics[width=6.5in]{../../../../../../tmp/RtmpMmzXjr/file3e7f6e0175d3}
+\includegraphics[width=6.5in]{../../../../../../tmp/Rtmp6UK3fU/file52c0514edd4f}
 
@@
 
-\includegraphics[width=6.5in]{../../../../../../tmp/RtmpMmzXjr/file3e7f7aec27f9}
+\includegraphics[width=6.5in]{../../../../../../tmp/Rtmp6UK3fU/file52c079e156e9}
 
@@
 
-\includegraphics[width=6.5in]{../../../../../../tmp/RtmpMmzXjr/file3e7f5c6b6ac6}
+\includegraphics[width=6.5in]{../../../../../../tmp/Rtmp6UK3fU/file52c02397d39c}
 

### 2023-12-13 R/crude.bd.est.R

@@ -1,14 +1,18 @@
 #' @title Crudely estimates extinction and speciation
 #' @name crude.bd.est
 #'
-#' @usage crude.bd.est(tree)
-#'
-#' @description Very crudely estimates the extinction and speciaton rate of a tree.  
+#' @description Crudely estimates the extinction and speciaton rate of a tree based on \code{geiger::bd.km} and \code{geiger::bd.ms} 
 #'
 #' @param tree a \code{"phylo"} object.
+#' @param method either \code{"count"} or \code{"estimate"}. See details.
+#' @param ... any additional arguments to be passed to \code{geiger::bd.km} and \code{geiger::bd.ms}.
 #' 
-#' @details This function calculates the extinction rate as the number of extinction events in the tree divided by the tree age (expressed in tree age units - e.g. million years). The speciation rate is calculated as the number of speciation events divided by the tree age. If the input tree has no \code{$root.time} element, the speciation and extinction rate are just the number of speciation and extinction events.
-#' \emph{NOTE} that this function is a very crude way to estimate rates of extinction and speciation: the rates are calculated from the raw data (not estimated), ignore sampling biases of the data and the variability in rates. For correct model base approaches see for example \code{\link[ape]{birthdeath}} or \code{\link[ape]{bd.ext}}. 
+#' @details This function calculates the extinction and speciation rates using two methods:
+#' \itemize{
+#'      \item \code{"estimate"} estimates the rates using the algorithm from \code{geiger::bd.km} and \code{geiger::bd.ms} based on the Magallon and Sanderson 2000 method. Note that this function provides more of a "guestimate" of extinction and speciation rates which can be especially wrong with low sampling (either missing fossil or living data). This can lead to estimating erroneous negative extinction rates.
[…]

### 2023-12-13 tests/testthat/test-crude.bd.est.R

@@ -6,15 +6,31 @@ test_that("crude.bd.est works", {
     ## Just living species
     set.seed(1)
     liv <- rcoal(10)
-    expect_equal(crude.bd.est(liv)$speciation(), 9)
-    expect_equal(crude.bd.est(liv)$extinction(), 0)
+    expect_equal(round(crude.bd.est(liv, method = "count")$speciation(), 4), round(2.197225, 4))
+    expect_equal(round(crude.bd.est(liv, method = "count")$extinction(), 4), round(0, 4))
     liv$root.time <- 10
-    expect_equal(round(crude.bd.est(liv)$speciation(), 4), round(5.229518, 4))
-    expect_equal(crude.bd.est(liv)$extinction(), 0)
+    expect_equal(round(crude.bd.est(liv, method = "count")$speciation(), 4), round(1.276714, 4))
+    expect_equal(round(crude.bd.est(liv, method = "count")$extinction(), 4), round(0, 4))
 
     ## Also fossils
     set.seed(1)
     tree <- rtree(20)
-    expect_equal(crude.bd.est(tree)$speciation(), 19)
-    expect_equal(crude.bd.est(tree)$extinction(), 19)
+    expect_equal(round(crude.bd.est(tree, method = "count")$speciation(), 4), round(2.944439, 4))
+    expect_equal(round(crude.bd.est(tree, method = "count")$extinction(), 4), round(2.944439, 4))
+
+
+    ## Just living species
+    set.seed(1)
+    liv <- rcoal(10)
+    expect_equal(round(crude.bd.est(liv, method = "estimate")$speciation(), 4), round(1.468996, 4))
+    expect_equal(round(crude.bd.est(liv, method = "estimate")$extinction(), 4), round(0.6229002, 4))
+    liv$root.time <- 10
+    expect_equal(round(crude.bd.est(liv, method = "estimate")$speciation(), 4), round(1.468996, 4))
+    expect_equal(round(crude.bd.est(liv, method = "estimate")$extinction(), 4), round(1.272554, 4))
+
+    ## Also fossils
+    set.seed(1)
+    tree <- rtree(20)
+    expect_equal(round(crude.bd.est(tree, method = "estimate")$speciation(), 4), round(0.95573, 4))
+    expect_equal(round(crude.bd.est(tree, method = "estimate")$extinction(), 4), round(-0.1382479, 4))
 })

### 2023-12-13 tests/testthat/test-treats.R

@@ -78,8 +78,8 @@ test_that("paper example works", {
     library(dispRity)
     data(BeckLee_tree)
 
-    my_bd_params <- crude.bd.est(BeckLee_tree)
-    stop_rule <- list(max.time = 30)
+    my_bd_params <- crude.bd.est(BeckLee_tree, method = "count")
+    stop_rule <- list(max.time = 140)
     my_traits <- make.traits(process = BM.process, n = 2)
 
     ## Creating a random mass extinction
@@ -109,7 +109,7 @@ test_that("paper example works", {
 
     ## multiple prints
     output <- capture_output(print.treats(sim_rand_extinction))
-    expect_equal(output, " ---- treats object with 10 replicates ---- \nYou can access individual replicates by extracting them, e.g. using x[[1]]\nBirth death process with events:\nspeciation: 0.352.\nextinction: 0.288.\nEvent targeting \"taxa\" to be triggered 1 time.\nThe condition function is: age.condition\nThe modification function is: random.extinction\n\nSimulated trait data:\n2 traits for 1 process (A:2) with one starting value (0).")
+    expect_equal(output, " ---- treats object with 10 replicates ---- \nYou can access individual replicates by extracting them, e.g. using x[[1]]\nBirth death process with events:\nspeciation: 0.028.\nextinction: 0.027.\nEvent targeting \"taxa\" to be triggered 1 time.\nThe condition function is: age.condition\nThe modification function is: random.extinction\n\nSimulated trait data:\n2 traits for 1 process (A:2) with one starting value (0).")
 
 
 
@@ -127,7 +127,7 @@ test_that("paper example works", {
     expect_equal(length(sim_trait_extinction), 10)
 
     output <- capture_output(print.treats(sim_trait_extinction))
-    expect_equal(output, " ---- treats object with 10 replicates ---- \nYou can access individual replicates by extracting them, e.g. using x[[1]]\nBirth death process with events:\nspeciation: 0.352.\nextinction: 0.288.\nEvent targeting \"taxa\" to be triggered 1 time.\nThe condition function is: age.condition\nThe modification function is: trait.extinction\n\nSimulated trait data:\n2 traits for 1 process (A:2) with one starting value (0).")
[…]

### 2023-12-15 paper/treats_paper.Rmd
Updated paper with response
Needs final check before resubmit

@@
     fig_height: 8
-  bookdown::html_document2: default
+  # bookdown::html_document2: default
   bookdown::pdf_document2: default
@@
 ## Thre tree
-plot(BeckLee_tree, show.tip.label = FALSE)
+plot(ladderize(BeckLee_tree), show.tip.label = FALSE)
 axisPhylo()
@@
 To do so, we need to use some **speciation** parameter indicating the rate at which lineages speciate (_aka_ "birth" or "$\lambda$") and an **extinction** parameter indicating the rate at which they go extinct (_aka_ "death" or "$\mu$").
-These two parameters can be crudely calculated from a given tree by counting the number of speciations and extinctions through time (but see better methods, e.g.
-`ape::birthdeath`; @lawton1995extinction; @ape).
-We also need a stopping rule for when to stop the simulations (in our case when reaching 30 time units).
+Here we are using two relatively arbitrary (speciation = 0.035 and extinction = 0.02) to get trees roughly matching the observed tree.
+Note that you might want to consider more appropriate ways to calculate these rates for research projects (e.g @magallon2001absolute).
+We also need a stopping rule for when to stop the simulations (in our case when reaching 140 time units).
 This will produce a single random tree using the input parameters (Fig. 3).
 The number of time units in `treats` is arbitrary and is not equivalent to millions of years.
-Using 30 time units here allows to simulate number of tips in a similar order of magnitude as the ones in the observed data.
+Using 140 time units here allows to simulate number of tips in a similar order of magnitude as the ones in the observed data.
 Note that I will not discuss the options here in great details.
@@
 ## Using the birth-death parameters from the observed tree
-my_bd_params <- crude.bd.est(BeckLee_tree)
-## Setting the stopping rule (stop after 30 time units)
-stop_rule <- list(max.time = 30)
+my_bd_params <- make.bd.params(speciation = 0.035, extinction = 0.02)
+## Setting the stopping rule (stop after 140 time units)
+stop_rule <- list(max.time = 140)
 
 ## Simulate the tree
-set.seed(1)
+set.seed(2)
 sim_tree <- treats(bd.params  = my_bd_params,
@@
 ## Plotting the resulting tree
-plot(sim_tree, show.tip.label = FALSE)
+plot(ladderize(sim_tree), show.tip.label = FALSE)
 ## Adding the phylogenetic axis
@@
 ## Simulate the tree and traits
-set.seed(1)
+set.seed(123)
 sim_data <- treats(traits     = my_traits,
@@
 To do so, we can create two different `"events"` object with the function `make.events`.
-The first one will simulate a random extinction event after reaching half of the simulations (15 time units) and then making three quarter (0.75) of the taxa go extinct.
-The second one will simulate a random extinction but based on trait values: after reaching time 15, all the species with positive trait values will go extinct.
+The first one will simulate a random extinction after reaching 66 time units and then making three quarter (0.75) of the taxa go extinct.
+The second one will simulate a random extinction but based on trait values: after reaching time 66, all the species with positive trait values will go extinct.
 Both scenarios illustrate two different types of mass extinctions but they are not equivalent: because of the ancestral trait value starting at 0, we expect the second scenario to remove on average only 50% of the species (i.e.
@@
     target       = "taxa",
-    condition    = time.condition(15),
+    condition    = age.condition(140-66),
     modification = random.extinction(0.75))
@@
     target = "taxa",
-    condition = time.condition(15),
+    condition = age.condition(140-66),
     modification = trait.extinction(x = 0, condition = `>=`))
 
-set.seed(1)
+set.seed(123)
 ## Simulate the tree and traits with a random extinction event
@@
 ## Random extinction
-plot(drop.singles(sim_rand_extinction[[31]]), main = "Random extinction")
-abline(v = 15, col = "red")
+plot(sim_rand_extinction[[42]], main = "Random extinction")
+abline(v = 66, col = "red")
 abline(h = 0, col = "grey")
 legend("topleft", pch = NULL, legend = "A", bty = "n", cex = 2)
-
 ## Selective extinction
-plot(drop.singles(sim_trait_extinction[[41]]), main = "Positive extinction")
-abline(v = 15, col = "red")
+plot(sim_trait_extinction[[50]], main = "Positive extinction")
+abline(v = 66, col = "red")
 abline(h = 0, col = "grey")
@@
                                             model  = "proximity",
-                                            time   = seq(from = 25, to = 0, by = -5),
+                                            time   = seq(from = 96, to = 36, by = -10),
                                             metric = c(sum, variances),
@@
 ## Scale the disparity results (to compare to the observed ones)
-random_extinction_disparity <- rescale.dispRity(random_extinction_disparity)
-selective_extinction_disparity <- rescale.dispRity(selective_extinction_disparity)
-observed_disparity <- unlist(get.disparity(rescale.dispRity(observed_disparity)))
+random_extinction_disparity <- scale.dispRity(random_extinction_disparity)
+selective_extinction_disparity <- scale.dispRity(selective_extinction_disparity)
+observed_disparity <- unlist(get.disparity(scale.dispRity(observed_disparity)))
 
@@
 par(mfrow = c(1,2))
-plot(random_extinction_disparity, main = "Random extinction", ylab = "Scaled sum of variances")
-abline(v = 3, col = "red")
+plot(random_extinction_disparity, main = "Random extinction", ylab = "Scaled sum of variances", ylim = c(0, 1))
+abline(v = 4, col = "red")
 lines(x = 1:7, y = observed_disparity, lty = 2, lwd = 2)
-plot(selective_extinction_disparity, main = "Selective extinction", ylab = "")
-abline(v = 3, col = "red")
+plot(selective_extinction_disparity, main = "Selective extinction", ylab = "", ylim = c(0, 1))
+abline(v = 4, col = "red")
 lines(x = 1:7, y = observed_disparity, lty = 2, lwd = 2)
@@
 ## Package location
-The `treats` package is available 
-<!-- % on the [CRAN](https://cran.r-project.org/web/packages/treats/index.html) or  -->
-on [GitHub](https://github.com/TGuillerme/treats) with more associated information.
+The `treats` package is available on the [CRAN](https://cran.r-project.org/web/packages/treats/index.html) or on [GitHub](https://github.com/TGuillerme/treats) with more associated information.
 All the versions of the package are archived on ZENODO with associated [DOI:10.5281/zenodo.7970384](https://doi.org/10.5281/zenodo.7970384).

### 2023-12-18 paper/treats_paper.Rmd
Updated paper and response

@@
 Or by comparing some observed pattern to one simulated under null or base conditions [@miller2022alternating].
-In theory, workers can follow the following research pipeline: 1) thinking of a specific mechanism (e.g.
+In theory, workers can use the following research pipeline: 1) thinking of a specific mechanism (e.g.
 mass extinction allowing the surviving species to acquire new morphologies), 2) collecting some data to test this mechanism (e.g.
@@
 Then the event is triggered using the argument `age.condition(15)` and modifies the internal `lineage` object using the ``trait.extinction(x = 0, condition = `>=`)`` function.
-`age.condition` and `trait.extinction` are both function factories which are not gonna be detailed here (see @wickham2019advanced).
+`age.condition` and `trait.extinction` are both function factories which are not going to be detailed here (see @wickham2019advanced).
 Effectively these arguments can be passed directly as standard functions.
@@
 par(mfrow = c(1,2))
-plot(random_extinction_disparity, main = "Random extinction", ylab = "Scaled sum of variances", ylim = c(0, 1))
+plot(random_extinction_disparity, main = "Random extinction",
+     ylab = "Scaled sum of variances", ylim = c(0, 1))
 abline(v = 4, col = "red")
 lines(x = 1:7, y = observed_disparity, lty = 2, lwd = 2)
-plot(selective_extinction_disparity, main = "Selective extinction", ylab = "", ylim = c(0, 1))
+plot(selective_extinction_disparity, main = "Selective extinction",
+     ylab = "", ylim = c(0, 1))
 abline(v = 4, col = "red")
@@
 
-## Manuals and vignettes
+## Manuals, vignette and templates
 
@@
 This manual is designed so that it can be regularly updated and enhanced through the lifetime of the package facilitating the interface between methods development and usage [@cooper2016dark].
+Furthermore, a library of simulation templates is maintained on the [GitHub page](https://github.com/TGuillerme/treats/issues?q=is%3Aopen+is%3Aissue+label%3A%22simulation+template%22).
+These templates are written and shared in the form of GitHub issues template and can be submitted and shared by any users.
+Either for them to have them stored somewhere curated, or better yet, so that other users can reuse, comment and mofidy them for their own projects.
 
@@
 This will be done while keeping track change (through `NEWS` file), continuous integration and unit testing.
-Furthermore, I am planning to develop the GitHub page associated with the package to encourage users to share their simulation templates and build library from them to facilitate the work for future users.
 

### 2024-01-15 inst/ace_treats.Rmd
basic (not good!) ace

---
title: "Running ancestral states estimations on treats"
author: "Thomas Guillerme"
date: "`r Sys.Date()`"
output:
  html_document:
    fig_width: 6
    fig_height: 6
---

## Ace pipeline idea

1 - estimate the root (x0)
2 - estimate the parameters (e.g. variance)
3 - simulate the process
4 - test the fit with the observed data.

## Readings

https://link.springer.com/article/10.1007/s11692-020-09505-x

Cunningham, C. W., Omland, K. E. and Oakley, T. H. (1998) Reconstructing ancestral character
states: a critical reappraisal. Trends in Ecology & Evolution, 13, 361–366

Pagel, M. (1994) Detecting correlated evolution on phylogenies: a general method for the com-
parative analysis of discrete characters. Proceedings of the Royal Society of London. Series B.
Biological Sciences, 255, 37–45

## Playground

```{r}
library(treats)
set.seed(1)
play_data <- treats(bd.params  = list(speciation = 1, extinction = 0.5),
                    traits     = make.traits(),
                    stop.rule  = list(max.taxa = 50),
                    null.error = 100) 
plot(play_data)
```

https://github.com/pasraia/RRphylo/blob/master/R/RRphylo.R

https://github.com/emmanuelparadis/ape/blob/master/R/ace.R

### 2024-01-15 R/ace.treats.R

@@ -1,0 +1,70 @@
+#' @title ace.treats
+#'
+#' @description Ancestral characters estimations on treats
+#'
+#' @param treats a \code{treats} object with traits and a tree (you can use \code{\link{make.treats}} to format your data).
+#' @param traits a \code{traits} object to simulate the trait (see \code{\link{make.traits}}).
+#' @param root.est can be either the name of the algorithm to estimate the root value(s), a predetermined set of root values, or a personalised function to estimate them (see details).
+#' @param param.est can be either the name of the algorithm to estimate traits parameters, a predetermined set of trait parameters, or a personalised function to estimate them (see details).
+#' @param ... any additional arguments to pass to fun.
+#' @param replicates the number of replicates to diagnose the ancestral states estimations
+#'
+#' @return
+#' This function outputs a treats object with the tree (\code{$tree} a list of traits estimates (\code{$data})) as well as a diagnosis of the estimations (\code{$diagnosis}).
+#'
+#' @details
+#' This function 
+#'
+#' @example
+#'
+#' @seealso \code{\link{make.traits}} \code{\link{make.treats}} \code{\link{map.traits}} \code{\link{treats}}
+#' 
+#' @author Thomas Guillerme
+#' @export
+
+ace.treats <- function(treats, traits, root.est = "debug", param.est = "debug", ..., replicates = 1) {
+    ## Sanitizing
+    check.class(treats, "treats")
+    check.class(traits, c("treats", "traits"))
+
+    ## Root.est
+    ## Check the root estimation input
+    ## If method or function, run the root estimations
+    if(root.est == "debug") {
+        root.est <- mean(treats$data) # very crude, needs branch length!
+    }
+
+    ## Param.est
+    ## Check the parameters estimation input
+    ## If method or function, run the root estimations
+
+    ## Update the trait
+    ace_traits <- traits.update(process.args = list(start = root.est))(traits = traits)
+
+    ## Map traits
+    ## Map the traits on the tree n times (= replicates)
+    estimates <- map.traits(traits = ace_traits, tree = treats$tree, replicates = replicates)
+
+    ## Diagnosis
+    ## Run the diagnosis
+
+    ## Clean the estimate (replace non-estimated values)
+    update.estimation <- function(estimate, input.data) {
+        matching_tips <- match(rownames(estimate$data), rownames(input_data$data))
[…]

### 2024-01-15 tests/testthat/test-ace.treats.R

@@ -1,0 +1,16 @@
+test_that("ace.treats implementation works", {
+
+    set.seed(1)
+    data <- treats(stop.rule = list(max.living = 40),
+                   trait = make.traits(),
+                   bd.params = list(speciation = 1, exctinction = 0.2),
+                   null.error = 10)
+    ## Remove the node values
+    data$data <- data$data[-c(grep("n", rownames(data$data))), , drop = FALSE]
+
+    ## Run some basic estimations
+    test <- ace.treats(treats = data, traits = make.traits(), replicates = 5)
+
+    plot(test)
+
+})

### 2024-02-01 paper/treats_paper.Rmd
french abstract

@@
 It also comes with an extend manual regularly updated following users' questions or suggestions.
+
+<!-- 1. Simuler des données réalistes est une étape importante pour la comprehension de la biologie.
+Ces données simulées peuvent être utilisées pour générer des modèles nuls, neutres ou de base.
+Elles peuvent ensuite être comparées à des données observées pour comprendre les mécanismes biologiques à l'origine de ces observations.
+Ou alors, elles peuvent être utilisées pour explorer, comprendre ou déveloper des modèles théoriques.
+
+2. En biologie evolutive, une simulation implique souvent le besoin de définir un modèle évolutif de descendance avec modification.
+Il existe quasiment une infinité de ces modèles évolutifs qui peuvent être modifiés pour inclure des processus complexes comme par exemple la co-évolution de plusieurs traits, des mécanismes de compétition ou des extinctions de masse.
+
+3. Je présente ici le package `R` modulaire `treats` pour simuler des phylogénies et des traits.
+Le package se base sur un simple algorithme de birth-death que les utilisateurs peuvent facilement modifier.
+
+4. `treats` fournit une interface ordonnée a travers l'objet `treats` qui permet aux utilisateurs de produire des pipelines faciles d'utilisation et reproductibles.
+Le package contient également un manuel complet (en anglais) qui est régulièrement actualisé en fonction des requêtes et questions d'utilisateurs. -->
 

### 2024-02-18 inst/treats_paper.Rmd
final version of the paper

@@
 
+<!-- 1. Simuler des données réalistes est une étape importante pour la comprehension de la biologie.
+Ces données simulées peuvent être utilisées pour générer des modèles nuls, neutres ou de base.
+Elles peuvent ensuite être comparées à des données observées pour comprendre les mécanismes biologiques à l'origine de ces observations.
+Ou alors, elles peuvent être utilisées pour explorer, comprendre ou déveloper des modèles théoriques.
+
+2. En biologie evolutive, une simulation implique souvent le besoin de définir un modèle évolutif de descendance avec modification.
+Il existe quasiment une infinité de ces modèles évolutifs qui peuvent être modifiés pour inclure des processus complexes comme par exemple la co-évolution de plusieurs traits, des mécanismes de compétition ou des extinctions de masse.
+
+3. Je présente ici le package `R` modulaire `treats` pour simuler des phylogénies et des traits.
+Le package se base sur un simple algorithme de birth-death que les utilisateurs peuvent facilement modifier.
+
+4. `treats` fournit une interface ordonnée a travers l'objet `treats` qui permet aux utilisateurs de produire des pipelines faciles d'utilisation et reproductibles.
+Le package contient également un manuel complet (en anglais) qui est régulièrement actualisé en fonction des requêtes et questions d'utilisateurs. -->
+
 ###### Keywords: trees, traits, simulations, birth-death, null-models, ecology, evolution, disparity
@@
     target       = "taxa",
-    condition    = age.condition(15),
+    condition    = age.condition(140-66),
     modification = random.extinction(0.75))
@@
     target = "taxa",
-    condition = age.condition(15),
+    condition = age.condition(140-66),
     modification = trait.extinction(x = 0, condition = `>=`))
 
-set.seed(1)
+set.seed(123)
 ## Simulate the tree and traits with a random extinction event
@@
 par(mfrow = c(1,2))
-plot(random_extinction_disparity, main = "Random extinction", ylab = "Scaled sum of variances")
-abline(v = 3, col = "red")
+plot(random_extinction_disparity, main = "Random extinction",
+     ylab = "Scaled sum of variances", ylim = c(0, 1))
+abline(v = 4, col = "red")
 lines(x = 1:7, y = observed_disparity, lty = 2, lwd = 2)
-plot(selective_extinction_disparity, main = "Selective extinction", ylab = "")
-abline(v = 3, col = "red")
+plot(selective_extinction_disparity, main = "Selective extinction",
+     ylab = "", ylim = c(0, 1))
+abline(v = 4, col = "red")
 lines(x = 1:7, y = observed_disparity, lty = 2, lwd = 2)
@@
 
+## Conflict of interest statement
+I declare no conflicts of interest.
+
 # References

### 2024-02-18 paper/treats_paper.Rmd
final version of the paper

@@
 ## Repeatability and reproducibility
-This paper is entirely reproducible from an Rmarkdown document available on [GitHub](https://github.com/TGuillerme/treats/inst/paper.Rmd).
+This paper is entirely reproducible from an Rmarkdown document available on [GitHub](https://github.com/TGuillerme/treats/inst/treats_paper.Rmd).
 The data used for the example above [@beck2014ancient] is available from the `dispRity` package [@dispRity].

### 2024-04-17 inst/gitbook/05_other_functionalities.Rmd
Update 05_other_functionalities.Rmd

@@
 `taxa.condition` | provides a trigger for an `event` dependent on the number of taxa | in `make.events`
-`time.condition` | provides a trigger for an `event` dependent on time | in `make.events`
+`age.condition` | provides a trigger for an `event` dependent on time | in `make.events`
 `trait.condition` | provides a trigger for an `event` dependent on trait values | in `make.events`

### 2024-04-18 inst/gitbook/not_include/05_thorough_example.Rmd
Update 05_thorough_example.Rmd

@@
                            target       = "traits",
-                           condition    = time.condition(2),
+                           condition    = age.condition(2),
                            modification = traits.update(
@@
                            target       = "taxa",
-                           condition    = time.condition(4),
+                           condition    = age.condition(4),
                            modification = trait.extinction(x = 0, condition = `<`, trait = 2))

### 2024-05-13 inst/gitbook/01_getting_started.Rmd
Updated manual

@@
 For that we need to create a `"modifiers"` object that modifies the `speciation` process with a specific condition and a specific modification when that condition is met.
-For a speciation to occur (and a species to not go extinct), the algorithm draws a random value _x_ between 0 and 1 and if the value is smaller than the speciation parameter divided by the speciation + extinction parameter, the lineage speciates ($x < (\lambda \ (\lambda + \mu))$), else it goes extinct.
+For a speciation to occur (and a species to not go extinct), the algorithm draws a random value _x_ between 0 and 1 and if the value is smaller than the speciation parameter divided by the speciation + extinction parameter, the lineage speciates ($x < (\lambda \div (\lambda + \mu))$), else it goes extinct.
 _x_ is a random value that simulates some stochasticity in the evolutionary simulation.
@@
 if the lineage trait value is negative.
-This can be done for example by not randomly generating the value _x_ but giving it a fixed value of say 1 so that the estimation of the speciation function $x < (\lambda \ (\lambda + \mu))$ is always met.
-In other words, if the lineage's trait value is negative, make sure that _x_ is equal to 1, so that the speciation condition $x < (\lambda \ (\lambda + \mu))$ is never met.
+This can be done for example by not randomly generating the value _x_ but giving it a fixed value of say 1 so that the estimation of the speciation function $x < (\lambda \div (\lambda + \mu))$ is always met.
+In other words, if the lineage's trait value is negative, make sure that _x_ is equal to 1, so that the speciation condition $x < (\lambda \div\ (\lambda + \mu))$ is never met.
 

### 2024-05-13 inst/gitbook/treats_manual.Rmd
Updated manual

--- 
title: "treats manual"
author: "Thomas Guillerme (guillert@tcd.ie)"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
output: bookdown::gitbook
documentclass: book
bibliography: [../references.bib]
link-citations: yes
github-repo: TGuillerme/treats
description: "treats R package manual"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# `treats`: trees and traits simulations.

Simultaneously simulate phylogenetic trees and traits (and the interaction between them).

The core of this package is based on the [`diversitree` birth-death algorithm](https://github.com/cran/diversitree/blob/master/R/simulate-bd.R).

## What is `treats`?

Bib is studying the evolution of mammalian disparity (i.e. diversity of shapes) across the Cretaceous-Palaeogene mass extinction event (K-Pg - 66 million years ago).
They observe a decrease of morphological disparity after the extinction event followed by a sharp increase of disparity.
However Bib is also prudent and wants to test whether their observed pattern is just an artifact of the data or actually showing some effect of the K-Pg extinction on mammalian evolution.
First they need to check whether changes in disparity can actually be discernables due to an extinction event (i.e. if an extinction event happens, would it be possible to detect a change in disparity)?
Second they want to know whether these changes are due to the mass extinction event being selective or not (i.e. the mass extinction favored some traits).

The [illustrative example](#kpgexample) section actually shows how do implement this example in `treats` in details.

Of course, this is just an example of when to use `treats`.
Simulating trees and traits can also be useful for:

 * Provide a baseline/null model to compare to an observed one.
 * Testing evolutionary hypotheses.
 * Exploring statistics properties (i.e. can the disparity metric X capture changes in Y).
 * Easily generating illustrative/simplified evolutionary history examples.
 * Teaching simplified macro-evolutionary theory (e.g. the birth-death model).

Note that for approaching these points individually (or some aspects of it), there are some very good R packages that simulate birth-death trees (e.g. [TreeSim](https://github.com/tanja819/TreeSim)) or simulate traits (disparity) and diversity jointly (e.g. [RPANDA](https://github.com/cran/RPANDA) or [PETER](https://github.com/PuttickMacroevolution/PETER)).
We strongly advise you look at these packages first as they might be more appropriate for your needs.

`treats` is a highly modular and user-friendly package that contains the simulations functionality of all these packages: it allows you to simulate disparity and diversity jointly with a vast array of options that can be easily modified by users.
For example, you can easily generate any type of process to generate a trait (BM, OU, something else, etc.) in multiple dependent or independent dimensions through `"traits"` objects.
You can then specify how the traits should affect disparity through `"modifiers"` objects. 
And finally you can create events (like mass extinctions) through `"events"` objects.
These objects and how to modify them will be detailed throughout this manual.
Finally we are putting an emphasis in the development of this package on the speed and reliability of the functions.

### Modular?

Because there is an infinite number of ways you might want to generate disparity and diversity (different traits, different modifiers and different events), the `treats` package is designed to make all these parts easy to code separately and integrate in the `treats` core functions.
This allows you to simulate finely-tuned multidimensional data for your specific project!

## Who is this manual for?

This manual explains how `treats` works in as much details as possible and aims to give you the keys to design your own diversity and disparity simulations.

The first section `Getting started` is aimed for people with a `R` beginner level:

 * you know what a package is - and how to install it;
 * you know what a function is - and what arguments are;
 * you have already (vaguely) heard that there are different types of objects in R - like matrices (`"matrix"`) or trees (`"phylo"`);
 * and you have already looked for documentation online.

The sections after that are a bit more advanced and require the following level:

 * you have already created functions before;
 * you know how to subset elements of a list - i.e. you understand the `list$element` syntax;
 * you are aware that both `TRUE == 1` and `FALSE == 0` are `TRUE`;
 * you have heard that the objects in R have classes and that it sometimes matters - for example you know that R can differentiate between `as.numeric(1)` and `as.integer(1)`;
 * you owe part of your work to Stack Overflow.

## Installing and running the package

You can install this package easily, directly from the GitHub:
<!-- TG: I do intend to release it on CRAN later on -->

```{r, eval = FALSE}
## Checking if devtools is already installed
if(!require(devtools)) install.packages("devtools")

## Installing the latest version directly from GitHub
devtools::install_github("TGuillerme/treats")
```

```{r, eval = FALSE, echo = FALSE, message = FALSE}
## Loading the package and setting up the start seed.
library(treats)
```

## Help

If you need help with the package, hopefully the following manual will be useful.
However, if you have suggestions or comments on on what has already been developed or will be developed, please send me an email (<guillert@tcd.ie>) or if you are a GitHub user, directly create an issue on the [GitHub page](https://github.com/TGuillerme/treats).
Doing so will not only help you but also other users as it will help improve this manual and the package!

## How does `treats` work?

Basically, the `treats` function takes your `traits`, `modifiers` and `events` objects to simulate disparity and diversity.
You will find more details about how these objects (`traits`, `modifiers` and `events`) work in the rest of the tutorial but here is a graphical representation of how `treats` works:

![Schematic summary of the `treats` package architecture](treats_structure.png)

## The birth-death algorithm {#bdalgorithm}

> If you want to get your hands dirty, you can go straight to the [getting started section](#gettingstarted), this following section just describes the process in pseudo-code.

The `treats` algorithm is based on a modular version of the birth-death process algorithm.
The birth-death model is a continuous Markov process (i.e. a continuous random process where future events depend only on the present, and not on past events) which is well defined mathematically and [commonly used in evolutionary biology](https://lukejharmon.github.io/pcm/chapter10_birthdeath/#section-10.2-the-birth-death-model) but also in [many other fields](https://en.wikipedia.org/wiki/Birth%E2%80%93death_process).

The algorithm used in `treats` allows modularity of this process and is based on the following steps (the text in `courier font` is for the name of the process in the algorithm):

 0. **Starting the process**: this step is non-modular and creates a random tree with one tip, one node and one branch connecting both. This step is used to optimise the rest of the algorithm in terms of speed and memory management. The node, tip and branch resulting from this step are discarded at the end of the simulations.
 1. **Selecting a lineage** (`selecting`): this step selects a tip that is currently not extinct. In a standard birth-death process this is done randomly, however in `treats` this can be modified based on the birth-death parameters, the currently available lineages and potential trait values. *For example, it is possible to put a higher probability for selecting a lineage that is closely related to a lineage that recently went extinct and has a positive trait value.*
 Then go to step 2.
 2. **Growing the tree** (`waiting`): this step grows the tree by a certain amount. It does so by adding the same amount of branch length to all the non-extinct lineages. In an exact birth-death process, this is done by drawing a random value from an exponential distribution with a rate of $$\text{number of living lineages} \times (\text{speciation} + \text{extinction parameters})$$. In `treats`, this can be modified based on the birth-death parameters, the currently available lineages and potential trait values. *For example, it is possible to increase branch length by the ratio of living/extinct fossils and a random number drawn from the range of current trait values.*
 If the total tree length is less than the required tree length, go to step 3. Else go to step 6.
 3. **Simulating traits** [optional]\(`traits`): this step allows you to simulate a trait value for the selected lineage from step 1. This is typically not part of a standard birth-death process and is handled via the `traits` option in `treats` (see the [`make.traits`](#maketraits) chapter for more details).
 Then go to step 4.
 4. **Speciating** (`speciating`): in this step, the selected lineage from step 1 has the option of speciating or going extinct. In a standard birth-death process, this happens by randomly drawing a value between 0 and 1 as a way to trigger speciation relative to the birth-death parameters:
```
if
    randomly drawn number is smaller or equal to speciation/(speciation + extinction)
then
    do speciate
else
    go extinct
```
 Again, in `treats`, this process can be modified based on the birth-death parameters, the currently available lineages and potential trait values. *For example, the lineage can only speciate if its trait value is positive, regardless of whether speciation or extinction have been triggered*. 
 Then go to step 5.
 5. If the number of lineages is less than required number of lineages, then go to step 1. Else go to step 6.
 6. The simulation stops because it has reached the required amount of lineages and/or time (i.e. branch length).

<!--chapter:end:index.Rmd-->

---
title: "Getting start"
author: "Thomas Guillerme (guillert@tcd.ie)"
date: "`r Sys.Date()`"
bibliography: [../references.bib]
output:
  html_document: default
  pdf_document: default
---

```{r, echo = FALSE}
library(treats)
set.seed(4242)
```

# Getting started {#gettingstarted}

## The simplest analysis: simulating diversity only

One of the simplest things to do with the `treats` package is to simulate a birth-death tree.
For that you can use the function `treats` and specify your stopping rule.
The stopping rule simply tells the birth-death process to stop whenever it reaches one of these three conditions:

 * `"max.taxa"   = n` stop when `n` taxa are generated;
 * `"max.living" = n` stop when there are `n` co-occuring taxa of the same age (i.e. "living" taxa);
 * `"max.time"   = n` stop when the simulated tree is `n` units of age old (these units are arbitrary);

For example, we might want to generate a birth-death tree with 20 taxa:

```{r}
## Setting a stopping rule to reach a maximum of 20 taxa
my_stop_rule <- list(max.taxa = 20)
```

We can now run the simulations using:

```{r}
## Running the birth-death simulation
my_tree <- treats(stop.rule = my_stop_rule)
```

> Note that here we could have specified more than one stopping rule, for example, we might want to run a simulation and stop it if it *either* reaches 10 taxa *or* the age 2 using `stop.rule = list(max.time = 2, max.taxa = 10)`. The simulation will then stop when either of these conditions are met.

The resulting object is a classic `"phylo"` object that you can plot like so:

```{r}
## The tree object
my_tree
## Plotting it
plot(my_tree)
```

### Changing the birth-death parameters

People familiar with the [birth-death models](https://lukejharmon.github.io/pcm/chapter10_birthdeath/) might have noticed that we did not specify two important things here: the speciation parameter (sometimes called "lambda" or "birth") and the extinction parameter (sometimes called "mu", "death" or "background extinction").
By default `treats` runs a pure birth model (speciation is set to 1 and extinction to 0).
However, you can easily change that by specifying your own birth-death parameters:

```{r}
## my birth-death parameters
my_params <- list(speciation = 1,
                  extinction = 1/3)
```

> You can find more information about setting up more complex birth-death parameters [in this section](#makebdparams).

You can then run the same birth-death tree simulation with extinction:

```{r}
## Generating a birth-death tree with extinctions:
my_tree <- treats(bd.params = my_params, stop.rule = my_stop_rule)
## Visualising the new tree
plot(my_tree)
```

## Slightly more complex: simulating disparity and diversity

Chances are that you are using `treats` because you also want to simulate traits (disparity) along with your diversity (otherwise, we suggest using the [`TreeSim`](https://github.com/tanja819/TreeSim/) package that provides many more birth-death models).
Simulating traits is not much more complicated in `treats`: you'll simply need to create a `"traits"` object using the `make.traits` function.
These objects can have increasing complexity (see the rest of this tutorial) but we will keep it simple here.

`"traits"` objects contain one or more processes which are the ways to generate the trait.
The most common of these processes is the [Brownian Motion](https://en.wikipedia.org/wiki/Brownian_motion) model.
This is used by default with the `make.traits` function:

```{r}
## Creating the traits object
my_trait <- make.traits()
```

This trait object can be simply printed (to see what's in it) or plotted (to see what the process looks like in the absence of a phylogeny):

```{r}
## Which process is in here?
my_trait
## What does it look like?
plot(my_trait)
```

By default, this trait is called "A".
This is not a really good name but you'll see more about specifying trait names later on.
If this is what the process should look like (theoretically) you can then add its `"traits"` object to our previous `treats` function to generate the tree and the traits:

```{r}
## Simulate disparity and diversity
my_data <- treats(bd.params = my_params,
                stop.rule = my_stop_rule,
                traits    = my_trait)
```

Et voilà! We now have a simple disparity and diversity simulation.
We can see what's in the results by simply printing it or plotting it:

```{r}
## What's in there
my_data
## Plotting the disparity and diversity
plot(my_data)
```

You can then extract the components you need for your specific analysis like so:

```{r}
## Extracting the tree 
the_generated_tree <- my_data$tree
# Note that this is a "phylo" object
class(the_generated_tree)

## Extracting the data 
the_generated_data <- my_data$data
# Note that this is a "matrix" or "array"
class(the_generated_data)
```

You can find much more about how to design trait objects in the [`make.traits` section](#maketraits).

### `"treats"` objects

`treats` will output either just a tree (class `"phylo"`) if no traits were generated or a `"treats"` object that contains both a `$tree` (`"phylo"`) and a `$data` (`"matrix"`) component.
Note that this `"treats"` class is generalised to most outputs of the package functions.
This allows for a smoother handeling of the objects outputs such as summarising the content of a `make.bd.params` output or visualising a trait output from `make.traits`.

## Slightly more complex again: simulating linked disparity and diversity

The example above is still pretty simple and easily done through a variety of `R` packages: here the trait and the tree are simulated at the same time but only the tree is simulating the trait (i.e. the trait value at a tip is affected by its ancestor and the branch length leading to it) but not the other way around (the trait value does not affect the tree).
It is possible to add this aspect using `"modifiers"` objects.
`"modifiers"` are similar to `"traits"` in that you specify what should go in there and then feed it to your simulation.

`"modifiers"` affect two key steps of the birth-death process: the calculation of the waiting time (i.e. the component generating branch lengths) and the triggering of speciation or extinction events.
These events can be modified using `condition` and `modify` functions.
In other words, when reaching a certain condition specified by a `condition` function, the birth-death process will modify either the branch length or the speciation (or extinction) probability by applying a `modify` function.

You can use the function `make.modifiers` to design a specific `"modifiers"` object.
By default, this function generates a `"modifiers"` object that affects branch length and speciation in the following way:

 * branch length is a randomly drawn number from an exponential distribution with a rate equal to the current number of taxa multiplied by the sum of the speciation and extinction rates.
 * speciation is triggered if a randomly drawn number (from a (0,1) uniform distribution) is smaller than the ratio between the speciation rate and the sum of the speciation and extinction rates. If that random number is greater, the lineage goes extinct.

Note that these are defaults for a birth-death tree and were already applied in the examples above without specifying a `"modifiers"` object:

```{r}
## Make a default modifiers object
default_modifiers <- make.modifiers()
## What's in it?
default_modifiers
```

This will not do anything to our simulations compared to the previous trait and tree simulation but we can provide our `"modifiers"` object to the `treats` function:

```{r}
## Setting the simulation parameters
extinction_02 <- list(extinction = 0.2)
living_20     <- list(max.living = 20)
BM_trait      <- make.traits()

# Set random seed so we get the same results
set.seed(1)

## Simulate disparity and diversity
default_data <- treats(bd.params = extinction_02,
                     stop.rule = living_20,
                     traits    = BM_trait,
                     modifiers = default_modifiers)
default_data
```

Note however, that the printing information is now updated to state that you've added a `"modifiers"` object (even though it's the same as the default).
In this specific case the `modifiers` object is not necessary as it is doing nothing different than when using the default one (i.e. when no `modifiers` is provided).

For more interesting simulations however, you can provide modifiers that actually modify the birth-death process.
We can create one for example that makes species go extinct if their ancestor has a negative trait value.
For that we need to create a `"modifiers"` object that modifies the `speciation` process with a specific condition and a specific modification when that condition is met.
For a speciation to occur (and a species to not go extinct), the algorithm draws a random value _x_ between 0 and 1 and if the value is smaller than the speciation parameter divided by the speciation + extinction parameter, the lineage speciates ($x < (\lambda \div (\lambda + \mu))$), else it goes extinct.
_x_ is a random value that simulates some stochasticity in the evolutionary simulation.
At every step of the simulation, a different random value _x_ is drawn making sure that the resulting speciation or extinction is not predictable.

However, we can create a modifier that introduces less randomness (or a different level of randomness) by forcing the lineage to go extinct every time a specific condition is met, e.g. if the lineage trait value is negative.
This can be done for example by not randomly generating the value _x_ but giving it a fixed value of say 1 so that the estimation of the speciation function $x < (\lambda \div (\lambda + \mu))$ is always met.
In other words, if the lineage's trait value is negative, make sure that _x_ is equal to 1, so that the speciation condition $x < (\lambda \div\ (\lambda + \mu))$ is never met.

First we need to create a `condition` function to trigger this (non) speciation event.
We can do that by specifying our `condition` function (here the `going.extinct` function) to apply our modification.
For that we can use the `parent.traits` utility function that is optimised for accessing traits in the birth-death process (but you can of course write your own).
This function takes the `trait.values` and `parent.lineage` arguments, two arguments that you can leave named as they are to facilitate `treats`' understanding of what you want to assess:

```{r}
## Triggering a modification only if the ancestor trait is negative
negative.ancestor <- function(trait.values, lineage) {
    return(all(parent.traits(trait.values, lineage) < 0))
}
```

> Note that we use the function `all` here to evaluate all traits: i.e. if the data has more than one trait we trigger the modification only if all the trait values are negative.

Then we create the modification function.
This function must take the argument `x` and, in our case, returns the same value no matter what: `1` so that speciation never happens when the `negative.ancestor` function is triggered.

```{r}
## Always go extinct
going.extinct <- function(x) return(1)
```

This `going.extinct` function is to be contrasted with the normal modifier for selecting the value `x` which is `runif(1)` (drawing a random value between 0 and 1).
We can then provide these two functions (the condition `negative.ancestor` and how to modify the speciation event when this condition is met `going.extinct`).
If you are an advances `treats` user, you can design your own `speciation` function but if you just want to use a normal `speciation` function, you can use the default one from `treats` called `speciation`.

```{r}
## Making a modifier for species to go extinct if
## their ancestor's trait value is (or are) negative
negatives_extinct <- make.modifiers(
            ## If the following condition is met...
            condition = negative.ancestor, # Does the lineage have an ancestor with negative trait values?
            ## Then apply the following modifier...
            modify = going.extinct, # The species goes extinct
            ## To the the speciation process.
            speciation = speciation) # Here the speciation() process is default

## What's in it?
negatives_extinct
```

Note that the `make.modifiers` function tests whether the input is compatible with `treats` by default so unless you have an error message, your `modifiers` will work!
We can now simulate our tree and traits with our modifier: species will go extinct if their ancestor has a negative trait value:

```{r}
set.seed(1)
## Simulate disparity and diversity
modified_data <- treats(bd.params = extinction_02,
                        stop.rule = living_20,
                        traits    = BM_trait,
                        modifiers = negatives_extinct)
modified_data
```

We can now compare the two trees and their trait values.
Note that we've used the same starting seed for both trees so the only thing differing between them is the `"modifier"` object which leads to very different trees!

```{r, fig.height = 6, fig.width = 12}
par(mfrow = c(1,2))
plot(default_data, main = "default results")
plot(modified_data, main = "results with the modifier")
```

You can find much more about how to design modifiers in the [`make.modifiers` section](#makemodifiers).

## An illustrative example: mammals disparity through the K-Pg extinction {#kpgexample}

Bib is studying the evolution of mammalian disparity (i.e. diversity of shapes) across the Cretaceous-Palaeogene mass extinction event (K-Pg - 66 million years ago).

They use a dataset from @beckancient2014 that is a discrete morphological space containing the ordination of the around 400 morphological characters for 50 species and estimated for their 49 descendants into an ordinated shapespace of 99 elements and 97 dimensions.

```{r}
library(dispRity)
## Loading the mammalian morphological example data
data(BeckLee_mat99)

## This is what the dataset looks like
head(BeckLee_mat99)[1:5, 1:5]
```

And a tree showing the relation between these species through time:

```{r}
## The phylogenetic tree
data(BeckLee_tree)

## And what the tree looks like
plot(BeckLee_tree, cex = 0.8)
axisPhylo()
abline(v = BeckLee_tree$root.time - 66, col = "red")
```

They measure the disparity as the sum of variances as a proxy for changes in the traitspace size (@moms) with some bootstrapping to create some confidence intervals around that curve.

```{r}
## Making continuous time series
time_series <- chrono.subsets(data = BeckLee_mat99,
                              tree = BeckLee_tree,
                              method = "continuous", model = "proximity",
                              time = seq(from = 101, to = 41, by = -5))
## Calculating disparity
change_in_size <- dispRity(boot.matrix(time_series), metric = c(sum, variances))
plot(change_in_size)
abline(v = 8, col = "red")
```

> For more info about `chrono.subsets` and the `dispRity` package, you can have a look at the [`dispRity` manual](http://tguillerme.github.io/dispRity.html).

Looking at this curve, Bib wants to know whether disparity increased as a response to the K-Pg mass extinction or not.
To do this, they need to check whether changes in disparity can actually be detected due to an extinction event (e.g. if an extinction event happens, does that change disparity curves?).
As a subsidiary question, they want to know whether the pattern is more likely due to a selective or random extinction event.

### Simulating trees and traits using the observed parameters

The first step is to simulate the observed tree.
We can do that by using crude parameters from the observed tree:

```{r}
## Extracting the crude parameters
(est_params <- crude.bd.est(BeckLee_tree, method = "count"))
## These parameters are the observed speciation and extinction rate per million years.

## Setting the stopping rule (stop after reaching 50 taxa)
stop_rule <- list(max.taxa = 50)

## Simulating just a birth-death tree with these parameters to check
set.seed(1)
test_tree <- treats(bd.params  = est_params,
                    stop.rule  = stop_rule,
                    null.error = 100)
plot(ladderize(test_tree), cex = 0.8)
axisPhylo()
```

> Note that because of the high extinction to speciation ratio, the tree simulations can often fail before reaching the stop rule (i.e. all lineages go extinct before reaching 50 taxa). To avoid getting stuck in there, we use the option `null.error = 100` to rerun the simulations up to 100 times before obtaining a tree that did not went fully extinct. Also note that using such parameters make the simulations run slowlier but ensure the requested results are always achieved.

We now have a way to simulate a topology close to the observed one. We can simulate a list of topologies by replicating this function X amounts of time to get a distribution of trees:

```{r, eval = FALSE, message = FALSE}
## Generating 50 trees
tree_distribution <- treats(bd.params  = est_params,
                            stop.rule  = stop_rule,
                            null.error = 100,
                            replicates = 50)
```

Of course, here our objective is to also simulate some trait values to measure disparity under some simulated scenario.
We can do that be adding a `"traits"` object to the `treats` function.
Here we will use a default [Brownian Motion](https://en.wikipedia.org/wiki/Brownian_motion) to simulate our trait.

```{r}
## Creating a trait in 97 dimensions.
my_traits <- make.traits(process = BM.process, n = ncol(BeckLee_mat99))
```

> Note that there are many more options on how to simulate your trait such as using correlation, different processes, etc... See details in the [make traits section](#maketraits).

```{r, message = FALSE}
## Simulate the tree and traits
sim_data <- treats(traits     = my_traits,
                   bd.params  = est_params,
                   stop.rule  = stop_rule,
                   null.error = 100)
```

We can visualise the resulting tree using the `plot` function and visualising the different traits using the `trait` option in `plot`:

```{r}
## Plotting the results (first trait)
plot(sim_data, trait = 1)
```

Of course, here we want to simulate a distribution of trees and traits:

```{r, message = FALSE}
## Simulate the tree and traits
sim_data <- treats(traits     = my_traits,
                   bd.params  = est_params,
                   stop.rule  = stop_rule,
                   null.error = 100,
                   replicates = 50)
```

Et voila! We now have a distribution of 50 trees and 50 datasets from which we can calculate disparity.
We can directly pass the results to the `dispRity` package pipeline using the `dispRitreats` function:

```{r, message = FALSE, warning=FALSE}
## Calculate the dispRity for all the simulations
simulated_disparity <- dispRitreats(sim_data,
                                    method = "continuous",
                                    model  = "proximity",
                                    time   = 10,
                                    metric = c(sum, variances),
                                    scale.trees = TRUE)
par(mfrow = c(1,2))
plot(simulated_disparity, main = "simulated disparity")
plot(change_in_size, main = "observed disparity")
```

We can see that both simulated and observed disparity curves are different.
From there one could already reasonably propose that the observe disparity does not entirely follows our parametrised scenario (constant speciation and extinction and BM trait evolution).

However, we can complexify our scenario to introduce the effect of mass extinction event.
We can simulate two alternative scenarios:
 * one where we create a random extinction event after having simulated half of the species (25) where we remove 75% of species;
 * and a second one where we remove all the species with negative values.
This allows us to simulate two contrasted scenarios to see whether using our observed dataset it would be possible to detect changes in disparity due to mass extinctions.

To do so, we need to create two mass extinction `"event"` objects:

```{r}
## Creating a random mass extinction
random_extinction <- make.events(target = "taxa",
                                 condition = taxa.condition(25, condition = `>=`),
                                 modification = random.extinction(0.75))
## Creating an extinction that removes species with negative trait values
negative_extinction <- make.events(target = "taxa",
                                   condition = taxa.condition(25, condition = `>=`),
                                   modification = trait.extinction(x = 0, condition = `>=`))
```

We can then feed these two extinction events to our previous simulations

```{r, message = FALSE}
## Simulate the tree and traits with a random extinction event
sim_rand_extinction <- treats(
                   traits     = my_traits,
                   bd.params  = est_params,
                   stop.rule  = stop_rule,
                   events     = random_extinction,
                   null.error = 100,
                   replicates = 50)

## Simulate the tree and traits with a random extinction event
sim_trait_extinction <- treats(
                   traits     = my_traits,
                   bd.params  = est_params,
                   stop.rule  = stop_rule,
                   events     = negative_extinction,
                   null.error = 100,
                   replicates = 50)

## Visualising the difference between both scenarios
par(mfrow = c(1,2))
plot(sim_rand_extinction[[1]], main = "Random extinction")
plot(sim_trait_extinction[[1]], main = "Negative extinction")
```

We can then measure the disparity from these scenarios and compare them to our observed ones

```{r}
## Simulate the tree and traits with a random extinction event
sim_rand_extinction <- treats(
                   traits     = my_traits,
                   bd.params  = est_params,
                   stop.rule  = stop_rule,
                   events     = random_extinction,
                   null.error = 100,
                   replicates = 50)

## Simulate the tree and traits with a random extinction event
sim_trait_extinction <- treats(
                   traits     = my_traits,
                   bd.params  = est_params,
                   stop.rule  = stop_rule,
                   events     = negative_extinction,
                   null.error = 100,
                   replicates = 50)

## Visualising the difference between both scenarios
par(mfrow = c(1,2))
plot(sim_rand_extinction[[1]], main = "Random extinction")
plot(sim_trait_extinction[[1]], main = "Negative extinction")
```

Interestingly, with all our parameters used here and our specific disparity metric we cannot detect the effect of a mass extinction.

<!--chapter:end:01_getting_started.Rmd-->

---
title: "`make.traits`"
author: "Thomas Guillerme (guillert@tcd.ie)"
date: "`r Sys.Date()`"
bibliography: [../references.bib]
output:
  html_document: default
  pdf_document: default
---

# Simulating traits {#maketraits}

```{r, echo = FALSE}
library(treats)
set.seed(123)
```

In `treats`, traits are simulating by providing a `traits` object to the `traits` argument.
This object is generated using `make.traits`.
In `treats` a trait designate any coherent character of any number of dimensions which may or may not be independent of other traits.
For example, it can be a uni-dimensional variable uncorrelated to other traits or a multi-dimensional one with different amounts of correlation.

![`make.traits`: The main inputs are (1) the `process`, one or more functions to describe the trait(s) as a function of previous state (`x0`) and the branch length (`edge.length`), (2) `n` the number of dimensions of the trait and (3) the starting value `start`.](make.traits.png)

### Quick overview

function | arguments | input | what does it do
---------|-----------|-------|----------------
`make.traits` | process* | a function | simulates the trait (see below)
`make.traits` | n | a numeric value | the number of dimensions
`make.traits` | start | one or more numeric values | the starting values for each dimensions
`make.traits` | process.args | ... | anything to be passed to `process` (see below)
`make.traits` | trait.names | a character string | the name(s) of the trait(s)
`make.traits` | add | a `treats` object | another trait generated by `make.traits`
`make.traits` | update | a `treats` object | another trait generated by `make.traits`
`make.traits` | test | logical | whether to test the validity of the trait
`make.traits` | background | a `treats` object | another trait generated by `make.traits` (to run in the background)
---------|-----------|-------|----------------
`process` | x0 | a numeric value | the value of the previous trait
`process` | edge.length | a numeric value | the branch length value
`process` | ... | ... | any other argument to be handled by the process

\* non-optional arguments

## The process (`process`)

The function `make.traits` allows you to design the process of a trait or a set of traits.
Here, the process of a trait designates the rules needed to generate the trait through time while simulating a phylogeny.
This process can depend on the previous state in the tree (i.e. the trait of the ancestor) and the branch length to the descendant.
One classic example is the [Brownian motion process (or Weiner process)](https://en.wikipedia.org/wiki/Brownian_motion).
Note that it _can_ depend on both the ancestor and the branch length but does _not necessarily need to_, i.e. the process can be based only on the previous state or only on branch length or on neither.

### The syntax (how to code a process?)

Trait processes in `treats` are functions that must always take the following arguments by default.

 * `x0`: the previous trait value(s)
 * `edge.length`: the branch length
 * `...`: a placeholder for any extra arguments

For example, the following function would be a valid process that always generate the _true_ trait value: 42!.
In this example, the process is not dependent on either the previous state (`x0`) or the branch length (`edge.length`).

```{r}
## A valid (but useless?) process
valid.process <- function(x0 = 0, edge.length = 1, ...) {
    return(42)
}
```

> Note that in this function definition the arguments `x0` and `edge.length` have a default value of `0` and `1` respectively. In practice, these arguments are effectively set to the correct values in the `treats` internal function (i.e. whatever `x0` and `edge.length` are at that specific time of the process) but providing a default can help speed up the algorithms (specifically all the internal checks).

On the other hand, the following process (a unidimensional Brownian motion) is incorrect (it's missing `edge.length` and `...`):

```{r}
## A wrongly formatted process
invalid.process <- function(x0 = 0) {
    return(rnorm(1, mean = x0))
}
```

This will not work in `make.traits` (see below).

### Using a `"process"` in `treats`

You can design your own process as a function, as long as it has a valid syntax.
Alternatively, the `treats` package has several inbuilt processes, namely a multidimensional Brownian motion (`BM.process`) or a multidimensional Ornstein-Uhlenbeck process (`OU.process`).
You can find the list of implemented processes by looking at the `?trait.process` manual page in `R`.

Once a process is chosen, you can feed it into the `make.traits` function:

```{r}
## Creating a trait object
my_trait_object <- make.traits(process = BM.process)
```

This creates `"treats"` `"traits"` objects that you can print and visualise using the `plot` function:

```{r}
## The class of the object
class(my_trait_object)
## What's in it?
my_trait_object
## What does the process looks like
plot(my_trait_object)
```

Note that you can see the multiple options for plotting the trait process by looking at `?plot.treats` manual. Furthermore, you can look at what's actually in the object using this specific syntax (this applies to every object handled by the `treats` package):

```{r}
## What's actually in that object?
print.treats(my_trait_object, all = TRUE)
```

As traits can get more and more complex, the automatic printing of its summary allows for a easier display of what's in the `"traits"` object.

Note that it is possible to make `"traits"` objects with multiple traits and multiple processes (that can be the same or different for each trait):

```{r}
## Four traits: two BM, one OU and one normal non process
four_traits <- make.traits(process = c(BM.process,
                                       BM.process,
                                       OU.process,
                                       no.process))
four_traits
```

You can visualise them individually using the `trait` argument in `plot.treats`:

```{r}
## Plot options (4 plots in one window)
par(mfrow = c(2,2))
plot(four_traits, trait = 1)
plot(four_traits, trait = 2)
plot(four_traits, trait = 3)
plot(four_traits, trait = 4)
par(mfrow = c(1,1))
```

## The number of traits `n` and the starting values `start`

Two further important arguments are `n` the number of traits per process and `start` the starting values for all traits.
By default they are set to `n = 1` and `start = 0`.
This means that `make.traits` will assume that your processes are always unidimensional by default and that they always start with the value `0`.
It is possible however, to change these values.

For example you can use the following to create a three dimensional Brownian motion with each dimension starting with the value `1`:

```{r}
## Multidimensional Brownian motion
make.traits(BM.process, n = 3, start = 1)
```

Or the following with each dimensions starting with different values (respectively `1`, `2` and `3`):

```{r}
## Multidimensional Brownian motion
make.traits(BM.process, n = 3, start = c(1,2,3))
```

Note that the number of traits are applied for each process by default.
However, you can apply different number of traits for different processes:

```{r}
## two 3D processes (BM and OU)
make.traits(c(BM.process, OU.process), n = 3)
## one 1D processes (BM) and one 4D process (OU)
make.traits(c(BM.process, OU.process), n = c(1, 4))
```

And starting values are distributed for all the traits or for the traits one by one:

```{r}
## two 3D processes (BM and OU) starting with 1
make.traits(c(BM.process, OU.process), n = 3, start = 1)
## two 3D processes (BM and OU) starting with values 1 to 6
make.traits(c(BM.process, OU.process), n = 3, start = 1:6)
## two 3D processes (BM and OU) with the two first ones starting
## with 1 and the 4 other ones with the default (0)
make.traits(c(BM.process, OU.process), n = 3, start = c(1,1))
```

### What is a trait in `treats`?

Because it would be impossible to accommodate all definitions of a trait in `treats` we chose an arbitrary one: a trait is whatever you define as a trait!
A trait can be uni-dimensional as the measurement of a feature of an organism (e.g. leaf surface, femur length, etc.) but can also be a multi-dimensional description of a feature, for example in 3D geometric morphometric context, a trait could be defined as "position of landmark X" (which will be a trait with three dimensions, _x_, _y_ and _z_) or in ecology, the location of a plant can be expressed as latitude and longitude coordinates.
In `treats`, the **process** corresponds to this trait definition (e.g. a process can be of n-dimensions and represents one organisms feature) and the **traits** represents the number of dimensions in total. 
So in the examples above, this is how the following traits are interpreted by `treats`:

```{r, eval = FALSE}
## Three traits with one process:
make.traits(BM.process, n = 3, start = c(1,2,3))
## Six traits with two processes:
make.traits(c(BM.process, OU.process), n = 3)
## Five traits with two processes
make.traits(c(BM.process, OU.process), n = c(1, 4))
```

## Multidimensional traits and correlations

As shown in the example above, a trait can be made multidimensional by using the `n` argument specifically passed to a `process`.
However, in this case, traits are simulated independently of each other.
It is possible to include correlation by designing a multidimensional trait `process` directly generating the correlation between within the dimensions of the process.
Alternatively, the inbuilt `BM.process` and `OU.process` can intake a correlation matrix to simulate traits independtly but with a correlation coefficient (see [this illustration with an event](#EG_change_correlation)).

## Extra argument for the processes with `process.args`

You can also feed extra arguments to your process(es) functions. 
For example, the inbuilt process `no.process` (that is just a number generator not based on the previous value `x0` or the branch length) can take a specific random number generator as a function:

```{r}
## no process trait using the normal distribution (default)
make.traits(no.process, process.args = list(fun = rnorm))
## no process trait using the uniform distribution
## bounded between 1 and 100
make.traits(no.process, process.args = list(fun = runif, min = 1, max = 100))
```

You can also add multiple extra arguments for multiple processes giving them as a list.

```{r}
## Two traits with no process:one normal and one uniform (1,100)
make.traits(process = c(no.process, no.process),
            process.args = list(list(fun = rnorm),
                                list(fun = runif, min = 1, max = 100)))
```

If one process does not need extra argument you must still give it an extra `NULL` process argument:

```{r}
## Three traits with no process:
## one default, one lognormal and one uniform (1,100)
make.traits(process      = c(no.process, no.process, no.process),
            process.args = list(## Extra arguments for the first process (none)
                                list(NULL),
                                ## Extra arguments for the second process
                                list(fun = rlnorm),
                                ## Extra arguments for the third process
                                list(fun = runif, min = 1, max = 100)))
```

## Naming the traits with `trait.names`

As traits become more and more complex, it can be useful to give clearer names to each process.
This is easily done using the `trait.names` argument that attributes one name per process:

```{r}
## A simple trait with a proper name
simple_trait <- make.traits(trait.names = "1D Brownian Motion")
simple_trait
```

This becomes more useful if we use the complex example above:

```{r}
## Three named traits with no process:
## one default, one lognormal and one uniform (1,100)
make.traits(process      = c(no.process, no.process, no.process),
            process.args = list(## Extra arguments for the first process (none)
                                list(NULL),
                                ## Extra arguments for the second process
                                list(fun = rlnorm),
                                ## Extra arguments for the third process
                                list(fun = runif, min = 1, max = 100)),
            ## Naming each trait
            trait.names  = c("Normal", "LogNormal", "Uniform(1,100)"))
```

## Combining multiple traits with `add`

You can also add traits to already existing trait objects using the simple `add` option.
This option just takes a `"treats"` `"traits"` object and the additional process(es) will be added to it. 
For example:

```{r}
## Creating a simple default Brownian motion
one_process <- make.traits(trait.names = "BM")

## Creating a new trait (a 3D OU.process)
## and adding the previous one
two_processes <- make.traits(OU.process, n = 3, add = one_process,
                             trait.names = "3D OU")

## Only one process
one_process

## The two processes
two_processes
```

## Using a background trait

`traits` objects also allow a background trait to be used when traits are simulated ([step 3 here](#bdalgorithm)).
This allows traits to be simulated for _all_ tips whenever a trait is generated for one tip.
This can be useful for keeping track of trait values along the simulation (_cf_ just at bifurcating nodes).
The `background` argument takes any output from the `make.traits` function in a nested way:

```{r}
## Generating a default BM trait:
BM_trait <- make.traits()
## Generating an OU trait with a background BM trait
my_trait <- make.traits(process = OU.process, background = BM_trait) 
```

Note that technically you can nest as many background traits as you want (e.g. `make.traits(background = make.traits(background = make.traits(...)))` is valid).
However, you should always make sure that the background trait has the same dimensions as the main trait.
When using a trait with background, your tree will have internal singleton nodes (i.e. nodes linking to one ancestor and only one descendant).
You can remove these nodes using the [`drop.things` function](#dropthings).

```{r}
set.seed(1)
## Generating a pure birth tree with the background trait
tree_bkg <- treats(stop.rule = list(max.taxa = 20),
                 traits = my_trait)
## This tree has many internal singleton nodes
plot(tree_bkg)
```

## Saving trait values at different time steps

You can also simulate a tree by generating traits at specific time steps with the `save.steps` option in `treats`.
This will apply the `traits` object to all lineages currently alive at the required time steps.
These time steps can be either regular by providing a single numeric value; e.g. `save.steps = 0.1` will take a snapshot of the trait values every 0.1 units of time, or specific, by providing a specific set of values; e.g. `save.steps = c(1, 1.2, 3)` will take a snapshot of the trait values at the required time steps.

```{r}
set.seed(123)
## Generating a birth-death tree with a BM trait and saving steps at specific times
tree_steps <- treats(stop.rule  = list(max.time = 3),
                   bd.params  = list(speciation = 1, extinction = 0.1),
                   traits     = make.traits(),
                   save.steps = c(1/3, 1, 2))
## This also creates internal singleton nodes
plot(tree_steps)
abline(v = 3 - c(1/3, 1, 2), lwd = 0.5, col = "grey")
```

## Traits implemented in `treats`

If you don't want to design your own trait process, you can use one of the following trait processes that are currently implemented in `treats`. You can find more information about their many options using their specific manuals in R or the generic `?trait.process`:

 * `BM.process`: this is the well known [Brownian motion process](https://en.wikipedia.org/wiki/Brownian_motion).

```{r, echo = FALSE}
plot(make.traits(process = BM.process), main = "The Brownian motion")
```

 * `OU.process`: this is the equally famous [Ornstein–Uhlenbeck process](https://en.wikipedia.org/wiki/Ornstein%E2%80%93Uhlenbeck_process).

```{r, echo = FALSE}
plot(make.traits(process = OU.process), main = "The Ornstein–Uhlenbeck process")
```

 * `no.process`: this process has... no process. In other words, this is a non time dependent process; the simulated value does not depends on the ancestors' value nor the branch length. It's basically a place holder for a random sampling function like `rnorm` (default), `runif`, `rlnorm`, etc.

```{r, echo = FALSE}
plot(make.traits(process = no.process, process.args = list(fun = runif)), main = "No process using runif\n(yes, it \"looks\" like an OU!)")
```

 * `multi.peak.process`: this process is a modified version of the `OU.process` that can take multiple local long-term mean values. The default OU process has one long-term mean towards which the values are drawn with the alpha parameter (the elastic band). The single long-term mean is usually 0. However, with this `multi.peak.process` we can set multiple values towards which values can be attracted with the same alpha parameter.

```{r, echo = FALSE}
plot(make.traits(multi.peak.process, process.args = list(peaks = c(1, 5, 10))), main = "A multi-peak (OU) process\nwith peaks (optima) at 1, 5 and 10")
```

 * `repulsion.process`: this is a modified version of the `BM.process` where instead of accumulating gradually through time, new trait values are more likely to be different to the traits of their ancestor following a `repulsion` parameter.

```{r, echo = FALSE}
plot(make.traits(repulsion.process, process.args = list(repulsion = 5)), main = "A repulsion process\nwith a repulsion parameter of 5")
```

 * `discrete.process`: this one generates discrete trait values based on a transition matrix (you can use the utility function `transition.matrix` to generate one).

```{r, echo = FALSE}
## Generating a stepwise transition matrix for 3 states (with an overal random transition rate)
stepwise_matrix <- transition.matrix(type = "stepwise", states = 3)
## Generatin and plotting the the trait
plot(make.traits(discrete.process, process.args = list(transitions = stepwise_matrix)))
```

Of course here the plotting of a discrete trait is not that informative...

## Testing the traits with `test`

**This bit is still in development.**
We highly suggest leaving `test = TRUE` so that `make.traits` returns an error if a process or its additional arguments (`process.args`) are not formatted correctly.
`make.traits` will send an error if the trait cannot be directly passed to `treats`.
However, in some specific cases (again, probably mainly for development and debugging) it could be useful to skip the tests using `test = FALSE`.

## Mapping traits on a tree: `map.traits`

Although `make.traits` is mainly designed for `treats` simulations, i.e. simulating both the trees AND the traits at the same time, it is possible to use an already simulated or estimated tree and generating one or more traits on it using the `map.traits` function. This function takes the trait input as a `"traits"` object (from `make.traits`) and a tree topology with branch length. Note that you can also use `"traits"` objects with multiple or multi-dimensional traits as well as multiple phylogenies (`"multiPhylo"`).

```{r}
## Simulating a random tree with branch length
my_tree <- rtree(20)

## Creating three different traits objects:
## A Brownian Motion
bm_process <- make.traits(process = BM.process)
## An Ornstein-Uhlenbeck process
ou_process <- make.traits(process = OU.process)
## No process (just randomly drawing values from a normal distribution)
no_process <- make.traits(process = no.process)

## Mapping the three traits on the phylogeny
bm_traits <- map.traits(bm_process, my_tree)
ou_traits <- map.traits(ou_process, my_tree)
no_traits <- map.traits(no_process, my_tree)

## Plotting the topology and the different traits
par(mfrow = c(2,2))
plot(my_tree, main = "Base topology")
plot(bm_traits, main = "Mapped BM")
plot(ou_traits, main = "Mapped OU")
plot(no_traits, main = "Mapped normal trait")
```

## Templates for making your very own process

As detailed above, any process of your own design will work as long as it is a function that takes at least the arguments `x0` and `edge.length`.
You can be imaginative and creative when designing your own process but here are two detailed example functions for a unidimensional Brownian Motion and Ornstein-Uhlenbeck process that you can use for a start (or not).
Remember it is good practice for `treats` processes to set all the arguments with default values (just in case).

> Note that the functions below are not equal to the already implemented `BM.process` and `OU.process` but an easier to edit version that you can use as a template:

### A simple Brownian Motion process template

```{r}
## A simple Brownian motion process
my.BM.process <- function(x0 = 0, edge.length = 1, sd = 1, ...) {
    ## Drawing a random number from a normal distribution
    ## with x0 as the and a given standard deviation
    ## and depending on branch (edge) length
    result <- rnorm(n = 1, mean = x0, sd = sqrt(sd^2 * edge.length))

    ## Return the number
    return(result)
}
```

### A simple Ornstein-Uhlenbeck process template

```{r}
## A simple Ornstein-Uhlenbeck motion process
my.OU.process <- function(x0 = 0, edge.length = 1, var = 1, alpha = 1, ...) {
    ## Calculate the mean based on alpha
    mean <- x0 * exp(-alpha)
    ## Calculate the standard deviation based on alpha and the variance
    sd <- sqrt(var/(2 * alpha) * (1 - exp(-2 * alpha)))
    ## Draw a random number from a normal distribution
    ## using this mean and standard deviation
    ## and depending on branch (edge) length
    result <- rnorm(n = 1, mean = mean, sd = sqrt(sd^2 * edge.length))

    ## Return the number
    return(result)
}
```

<!--chapter:end:02_make.traits.Rmd-->

---
title: "`make.modifiers`"
author: "Thomas Guillerme (guillert@tcd.ie)"
date: "`r Sys.Date()`"
bibliography: [../references.bib]
output:
  html_document: default
  pdf_document: default
---

# Modifying the birth-death process {#makemodifiers}

```{r, echo = FALSE}
library(treats)
set.seed(123)
```

`"modifiers"` have a similar structure than `"traits"` where you can design an object with increasing complexity, starting with the simplest modifiers that doesn't modify anything (using the default arguments):

```{r}
## Making a default modifier (no modification)
my_default_modifiers <- make.modifiers()
my_default_modifiers
```

Similarly to `"traits"` objects, `"modifiers"` are also printed by default using `print.treats`.
You can see details about what's actually in the object using `print.treats(my_default_modifiers, all = TRUE)`.
However, contrary to `"traits"`, you cannot plot `"modifiers"`.

![`make.modifiers`: The three main inputs to make modifiers are functions to modify the `branch.length` (waiting) process controlling the "growth" of the tree; the `selection` (selecting) process controlling the selection of the tip on which to apply speciation or extinction; and the `speciation` (speciating) process controlling whether the select tip goes extinct or speciates. The allowed arguments are detailed [here](#allowarguments).](make.modifiers.png)

### Quick overview

function | arguments | input | what does it do
---------|-----------|-------|----------------
`make.modifiers` | branch.length | a function | generating branch length (see below)
`make.modifiers` | selection | a function | selecting a lineage (see below)
`make.modifiers` | speciation | a function | choosing between speciation and extinction (see below)
`make.modifiers` | condition | a function | a function that affects the condition of when to trigger the functions above (see below)
`make.modifiers` | modify | a function | a function that modifies the value of the trigger value generated when the `condition` is met (see below)
`make.modifiers` | add | a `treats` object | another modifier generated by `make.modifiers`
`make.modifiers` | update | a `treats` object | another modifier generated by `make.modifiers`
`make.modifiers` | test | logical | whether to test the validity of the modifier
---------|-----------|-------|----------------
`branch.length`, `selection`, `speciation`, `condition` | bd.params | a `treats` object generated by `make.bd.params` | usually handled internally (contains the either the birth-death values or their distribution)
... | lineage | a named "lineage" list | usually handled internally (contains the current state of the simulations)
... | trait.values | a matrix | usually handled internally (contains the current generated traits of the simulations)

## The default modifier (how the process is working)

The modifiers modify the core of the birth-death process as implemented in `treats`.
By default, the birth-death process in `treats` uses this modifier:

```{r}
## What is actually in the default modifier?
print(make.modifiers(), all = TRUE)
```

This contains a lot of information, much of it is actually not useful for using the modular aspects of `treats` at a high user level, i.e. unless you want to code very specific things, you won't need most of the information. The essentials are these three functions in the elements named `"waiting$fun"`, `"selecting$fun"` and `"speciating$fun"`:

 1. `"waiting$fun"` is the branch length function (defined in detail below) which returns a randomly drawn number from an exponential distribution with the rate of the number of taxa multiplied by the speciation and extinction rate: $n \times (\lambda + \mu)$ (here is uses the extra function `sum` for internal modularity reasons). **This function is responsible for the growth of the length/age of the tree**

```{r, eval = FALSE}
rexp(1, sum(lineage$n * (bd.params$speciation + bd.params$extinction)))
```
 With `lineage$n` being the number of lineages and `bd.params$speciation` and `bd.params$extinction` the speciation and extinction parameters (these specific terms are defined in detail below).

 2. `"selecting$fun"` is the selection function (defined in detail below) which, after the waiting time defined above returns a randomly selected lineage (a tip) among the existing ones. **This function is responsible for the branch selection**.

```{r, eval = FALSE}
sample(lineage$n, 1)
```

 3. `"speciating$fun"` is the speciation function (defined in detail below) which randomly draws a number (`runif(1)`) and depending on the speciation and extinction parameters makes the selected lineage (from 2.) speciate (`TRUE`) or go extinct (`FALSE`). **This function is responsible for the speciation or extinction of species**.

```{r, eval = FALSE}
runif(1) < (bd.params$speciation/(bd.params$speciation + bd.params$extinction))
```

We will have a look at all these aspects below in more detail which should make things clearer (but you can always refer to the default here for information).

## The branch length function (`branch.length`)

The first argument in `"modifiers"` is the branch length function (`branch.length`) this is the function that will be executed in `treats` to generate branch lengths.
Note that in the `treats` algorithm, branch length is not generated _directly_ but as the result of the waiting time.
In other words, the `branch.length` function just affects waiting time for all taxa present at any time in the simulation.
These taxa can then either go extinct (stopping the "growth" of its branch length) or survive (continuing the "growth").

By default, branch length (or waiting/growth) is a randomly drawn number from an exponential distribution with the rate of the number of taxa multiplied by the speciation and extinction rate: $n \times (\lambda + \mu)$ (where $n$ is the number of taxa currently present in the simulation, $\lambda$ and $\mu$ are respectively the speciation and extinction rates).
This default function is simply called `branch.length` in `treats` and can be used as a modifier as follows:

```{r}
## Specifying the default modifier
default_modifiers <- make.modifiers(branch.length = branch.length)

## Setting some parameters for generating trees
bd_params <- list(extinction = 0)
stop_rule <- list(max.living = 20)

## Generating a tree with the default branch length parameter
set.seed(0)
default_tree <- treats(bd.params = bd_params,
                     stop.rule = stop_rule,
                     modifiers = default_modifiers)
```

Of course, the point of the modularity here is that you can provide your own function for generating branch length.
For example, we might be interested in what our tree would look like if we use a simple constant branch length generation (instead of randomly drawing it from an exponential distribution).
We can do so by declaring our own `branch.length` function and adding it to a `"modifiers"` object.

```{r}
## A constant branch length generator
## (note that the output must be numeric, not integer)
constant.brlen <- function() {
    return(as.numeric(1))
}

## Creating the modifiers object
constant_modifier <- make.modifiers(branch.length = constant.brlen)

## Generating a new tree with this modifier
set.seed(0)
modified_tree <- treats(bd.params = bd_params,
                      stop.rule = stop_rule,
                      modifiers = constant_modifier)
```

And we can visualise the difference between both resulting trees:

```{r}
par(mfrow = c(1,2))
plot(default_tree,  main = "Default modifier")
plot(modified_tree, main = "Constant branch length\nmodifier")
par(mfrow = c(1,1))
```

It is of course possible to use more complex branch length modifiers that take different conditions and specific modification rather than simply always outputing a value of one.

### The modifier arguments {#allowarguments}

You can create a function for `branch.length`, `selection` and `speciation` that involve any of the following arguments:

  * `bd.params`: a named list containing `"numeric"` values that contains the birth-death parameters (at least `"speciation"` and `"extinction"`);
  * `lineage`: a named list containing the lineage data (see below).
  * `trait.values`: a `"matrix"` containing `"numeric"` values with the trait names as column names and the lineage ID as row numbers (you can use it with the function `parent.traits` to access the trait of the previous node for example).
  * `modify.fun`: a `"list"` of named `"function"` (usually passed through `condition` and `modify`).

The `lineage` list contains the following elements (missing elements are allowed):

 * `lineage$parents`: an `"integer"` vector: the list of parent lineages;
 * `lineage$livings`: an `"integer"` vector: the list of lineages still not extinct;
 * `lineage$drawn`: a single `"integer"`: the ID of the selected lineage; this must be a number in `1:lineage$n`;
 * `lineage$current`: a single `"integer"`: the selected lineage (is equal to `lineage$livings[lineage$drawn]`);
 * `lineage$n`: a single `"integer"`: the current number of non extinct lineage (is equal to `length(lineage$livings))`;
 * `lineage$split`: a `"logical"` vector: the list of splits for each lineage (`TRUE`), the number of total tips is equal to `sum(!lineage$split)`.

In general, unless you know what you're doing, you can ignore most arguments for specific modifiers since they are handled automatically within the `treats` function.
Therefore any argument can be left undeclared or missing and is always handled internally. For example, if you did not declare `lineage$n` as a function argument but are using `lineage$n` in the function, `lineage$n` will be detected and treated as a current argument automatically as set accordingly within the birth-death process (e.g. `lineage$n` will be set to the current number of taxa every iteration of the process).

For example, we can create a function that increases branch length proportional to the number of species "alive" at each time of the simulation in a discrete way, i.e. for discrete numbers of taxa, the branch length increases by jumps (ten fold) every five taxa:

```{r}
## A more complex binned.branch.length function
increasing.brlen <- function(bd.params, lineage) {

    ## Setting the cumulated birth and death
    birth_death <- bd.params$speciation + bd.params$extinction

    ## Returning branch lengths depending on different number of taxa
    if(lineage$n <= 5) {
        return(1    * rexp(1, sum(5 * birth_death)))
    }
    if(lineage$n <= 10) {
        return(10   * rexp(1, sum(10 * birth_death)))   
    }
    if(lineage$n <= 15) {
        return(100  * rexp(1, sum(15 * birth_death)))      
    }
    if(lineage$n <= 20) {
        return(1000 * rexp(1, sum(20 * birth_death)))
    } else {
        return(1000 * rexp(1, sum(lineage$n * birth_death)))
    }
}
```

We can then create it as a `"modifiers"` object and run a new simulation:

```{r}
## Creating a modifiers
increasing_modifier <- make.modifiers(branch.length = increasing.brlen)

## Generating a new tree with this modifier
set.seed(0)
increasing_tree <- treats(bd.params = bd_params,
                        stop.rule = stop_rule,
                        modifiers = increasing_modifier)
```

And we can visualise the difference between the resulting trees:

```{r}
par(mfrow = c(1,3))
plot(default_tree,    main = "Default modifier")
plot(modified_tree,   main = "Constant branch length\nmodifier")
plot(increasing_tree, main = "Increasing branch length\nmodifier (binned)")
par(mfrow = c(1,1))
```

## The selection function (`selection`)

The `selection` function is used in the birth-death process to know which lineage to select when running a speciation (or extinction!) event.
By default, this function randomly selects one taxon that is currently not extinct (using: `sample(1:lineage$n, 1))`.
Similarly to `branch.length`, it is possible to modify this part of the birth-death process.
For example, we could simply select the last created lineage to create a "ladder" or most asymmetric tree:

```{r}
## Our function to always select the last taxon
## (making sure it returns an integer)
select.last <- function(lineage) {
    return(as.integer(lineage$n))
}
```

> Note that here the function can only take the [allowed arguments as described above](#allowarguments) (here `lineage$n`: the number of current living taxa).

We can then create a `"modifiers"` object the same way as before this time using the `selection` argument:

```{r}
## A modifier for selection
ladderised_modifier <- make.modifiers(selection = select.last)
## Generating a new tree with this modifier
set.seed(0)
ladderised_tree <- treats(bd.params = bd_params,
                        stop.rule = stop_rule,
                        modifiers = ladderised_modifier)
## Displaying the results
par(mfrow = c(1,2))
plot(default_tree,    main = "Default modifier")
plot(ladderised_tree, main = "Ladderising modifier")
par(mfrow = c(1,1))
```

Again, it is of course possible to make the modifier more complex and in combination with other elements of the tree.
For example, we can create a `"treats"` object that also generates a BM trait and add to it this object a `selection` modifier that only selects tips with positive trait values (only species with positive trait values will speciate).

```{r}
## Our function that only select taxa with positive trait values
select.positive <- function(trait.values, lineage) {

    ## Selecting the taxa names with positive values for the first trait
    positives <- as.integer(rownames(trait.values)[which(trait.values[, 1] >= 0)])

    ## Combine the descendants of the current lineages (lineage$parents)
    ## with the species that have speciated (seq_along(lineages$split))
    ## to have a table of pairs of parents/splits
    parents_split_table <- cbind(lineage$parents, seq_along(lineage$split))
    ## Select the current taxa that descend from a node with a positive value
    positive_living <- parents_split_table[which(lineage$parents %in% positives), 2]

    ## Select one tip randomly in the ones with descendants with positive values
    return(sample(which(lineage$livings %in% positive_living), 1))
}

## Creating the modifier
positive_skew <- make.modifiers(selection = select.positive)

## Creating a (default) trait object
BM_trait <- make.traits()

## Simulate a tree and trait with no modifier
set.seed(1)
default_treats <- treats(bd.params = bd_params,
                     stop.rule = stop_rule,
                     traits = BM_trait)

## Simulate a tree and trait with the modifier
set.seed(1)
skewed_trait_treats <- treats(bd.params = bd_params,
                          stop.rule = stop_rule,
                          traits = BM_trait,
                          modifiers = positive_skew)

## Plotting the differences in trees and traits
par(mfrow = c(1, 2))
plot(default_treats, main = "Default trait and tree")
plot(skewed_trait_treats, main = "Skewed trait and tree")
```

## The speciation function (`speciation`)

The third function that can be used to modify the birth-death process is the `speciation` function.
This one is used during the birth-death process to decide whether a lineage speciates (creating a node and two new lineages) or goes extinct (creating a tip).

> Note that the `speciation` function only affects tips or nodes before the simulation reaches the `stop.rule`.
The then surviving lineages are all automatically transformed into tips.

By default, the `speciation` function is trigger a speciation even if a number randomly drawn from a uniform distribution is lower than the ratio between the speciation and the speciation and extinction parameter. If the randomly drawn number is higher, the lineage goes extinct.

```{r, eval = FALSE}
## The speciation in pseudo-code:
runif(1) < speciation/ (speciation + extinction)
```

Creating `"modifiers"` with a `speciation` function works the same way as for `branch.length` and `selection` but the function that will be used needs to output a logical value (see table [below](#summarymodifiers)).
Once the function is created simply input your function for speciation in the modifier and run the `treats` function with that modifier:

```{r, eval = FALSE}
## Speciating or going extinct randomly
## (regardless of the extinction parameter)
random.extinct  <- function() {
    return(sample(c(TRUE, FALSE), 1))
}

## Creating the modifiers object
random_extinction <- make.modifiers(speciation = random.extinct)

## Generating a new tree with this modifier
set.seed(2)
modified_tree <- treats(bd.params = bd_params,
                      stop.rule = stop_rule,
                      modifiers = random_extinction)

par(mfrow = c(1,2))
plot(default_tree,  main = "Default modifier")
plot(modified_tree, main = "Random extinction\nmodifier")
par(mfrow = c(1,1))
```

> Note how loads of lineages go extinct even if the extinction parameter is set to 0!

And again, we can make some more advanced modifiers: for example, one where a tip always goes extinct if their ancestor has a negative trait value. Here we will also introduce the utility function `parent.trait` that automatically selects the trait values of the parent of the current lineage.

```{r}
## A modifier for removing tips with negative values
modify.trait <- function(trait.values, lineage) {
    if(parent.traits(trait.values, lineage) < 0) {
        ## Go extinct!
        return(FALSE)
    } else {
        ## Speciate!
        return(TRUE)
    }
}

## Creating the modifier
modified_trait <- make.modifiers(speciation = modify.trait)

## Simulate a tree and trait with the modifier
set.seed(1)
modified_trait_treats <- treats(bd.params = bd_params,
                          stop.rule = stop_rule,
                          traits    = BM_trait,
                          modifiers = modified_trait)

## Plotting the differences in trees and traits
par(mfrow = c(1, 2))
plot(default_treats, main = "Default trait and tree")
plot(modified_trait_treats, main = "Biased trait and tree")
```

## Summary of the inputs and outputs for the `branch.length`, `selection` and `speciation` modifiers {#summarymodifiers}

modifier name   | accepted input (arguments) | required output (class) |
----------------|----------------------------|-------------------------|
`branch.length` | `bd.params`, `lineage`, `trait.values` | `"numeric"` |
`selection`     | `bd.params`, `lineage`, `trait.values` | `"integer"` |
`speciation`    | `bd.params`, `lineage`, `trait.values` | `"logical"` |

## The condition and modify functions (`condition` and `modify`)

In the examples above, we have seen how to specify modifications to the birth-death process (via `branch.length`, `selection` and `speciation`), however, these modifications are not dynamic.
In other words, throughout the process, the modifications remain constant (even if they are conditional).
It is however, possible to code the `"modifiers"` so that they can be affected by `"events"` objects (see next chapter [on `events`](#makeevents)).

To do so, you can formally declare conditions (`condition`) and modifications (`modify`) as internal functions that can then be modified by an `"events"` object.
`condition` and `modify` are hard coded in the `branch.length` function that they concern, i.e. they are variables (functions) within the function.

For example in the `speciation` part of a modifier, the default is to trigger an event from a uniform distribution and then check if that value is smaller than ($speciation/(speciation + extinction)$):

```{r}
## The default speciation algorithm
speciation <- function(bd.params) {
    ## Randomly trigger an event
    trigger_event <- runif(1)

    ## Speciate?
    return(trigger_event < (bd.params$speciation/
                              (bd.params$speciation + bd.params$extinction)))
}
```

It is possible with some specific conditions to modify this trigger by providing a `condition` and `modify` function to the speciation function.
That is, if the `condition` is met, apply the `modify` function to the algorithm.
For example here we can edit the default speciation function to have a modification (`modify = double.the.trigger`: doubling the trigger value) happening half the time (`condition = half.the.time`):

```{r}
## A conditional function that triggers half the time
half.the.time <- function() return(sample(c(TRUE, FALSE), 1))
## A modification that doubles the value to trigger the event
double.the.trigger <- function(x) return(x*2)
## A conditional modifier
make.modifiers(speciation = speciation,
               condition  = half.the.time, 
               modify = double.the.trigger)
```
Effectively, this will internally modify the speciation function as follows:

```{r, eval = FALSE}
## The default speciation algorithm
speciation <- function(bd.params) {
    ## Randomly trigger an event
    trigger_event <- runif(1)

    ## Modify the triggering
    if(half.the.time()) { ## This running the half.the.time function drawing TRUE or FALSE randomly
        trigger_event <- double.the.trigger(trigger_event) ## This will double the random value trigger_event
    }

    ## Speciate?
    return(trigger_event < (bd.params$speciation/
                              (bd.params$speciation + bd.params$extinction)))
}
```

These `condition` and `modify` functions can be applied to all the `modifiers` elements (`selection`, `branch.length` and `speciation`).
However, they are typically used in `events` that will modify the `modifiers` ([see the `events` section](#makeevents))!

<!--chapter:end:03_make.modifiers.Rmd-->

---
title: "`make.events`"
author: "Thomas Guillerme (guillert@tcd.ie)"
date: "`r Sys.Date()`"
bibliography: [../references.bib]
output:
  html_document: default
  pdf_document: default
---

```{r, echo = FALSE}
library(rgl)
knitr::knit_hooks$set(webgl = hook_webgl)
```

```{r, echo = FALSE}
library(treats)
set.seed(123)
```

# Adding events to simulations {#makeevents}

One other major feature of the `treats` package is that it allows simulations to run with specified events that occur during the simulation.
These are typically events that can drastically change the course of the simulation.
For example, you might want simulate a mass extinction at some specific point in time.
However, these can also be more subtle, like the internal change of parameter values when reaching a specific trait value.
These are all handled by `treats` with the `events` object that you can create using `make.events` and has the same overall logic as `make.traits` and `make.modifiers`.

`events` require three main arguments:
 * the `target` which designates what the extinction should affect (e.g. the taxa, the speciation rate, etc...);
 * the `condition` which designates when to trigger the event;
 * the `modification` which designates what to modify when the event is triggered.

There are several more arguments that can be passed to `make.events` but they are discussed later on.
First let's focus on these three main arguments:

![`make.events`: events are encoded using three main arguments: the `target`, specifying _what_ the event is modifying, the `condition` specifying _when_ to trigger the event, and `modification`, specifying _how_ the event modifies the simulations.](make.events.png)

### Quick overview

  condition,
  modification,
  add,
  test = TRUE,
  event.name,
  replications = 0,
  additional.args
)

function | arguments | input | what does it do
---------|-----------|-------|----------------
`make.events` | target* | any of the following: "taxa", "bd.params", "traits", "modifiers" or "founding" | what aspect of the simulation should the event target
`make.events` | condition* | a function | a function that affects the condition of when to trigger the event (see below)
`make.events` | modification* | a function | a function that applies the modification to the target if the condition is triggered (see below)
`make.events` | add | a `treats` object | another event generated by `make.events`
`make.events` | test | logical | whether to test the validity of the event
`make.events` | event.name | a character string | the name of the event
`make.events` | replications | a number | if possible how many times is the event allowing to be repeated
`make.events` | additional.args | ... | any additional arguments to be passed to the event

Here is a list of implemented `conditions` (see `?events.conditions`):

function | arguments | input | what does it do
---------|-----------|-------|----------------
`age.condition`, `taxa.condition`, `trait.condition` | x | the value to check the target against | e.g. time = x, taxa = x or trait = x (see below)
... | condition | a relational operator | e.g. `==`, `>`, `!=`, etc. for asking for example `time == x` 
... | ... | any additional arguments to be passed to the event | |

Here is a list of implemented `modifications` (see `?events.modifications`):

 function | arguments | input | what does it do
---------|-----------|-------|----------------
`random.extinction` | x | a numeric value | the proportion of species to make extinct
`trait.extinction` | x | a numeric value | the trait value when to trigger the extinction
`trait.extinction` | condition | a relational operator | the relation to the trait value (e.g. `x = 1` and `condition =  >` makes species that have a trait value > 1 go extinct)
`bd.params.update` | ... | any argument to be passed to `make.bd.params` | |
`traits.update` | ... | any argument to be passed to `make.traits` | |
`modifiers.update` | ... | any argument to be passed to `make.modifiers` | |

\* non-optional arguments

## Target

The target of the event is what the event is going to modify in the birth-death algorithm.
You can only have one target per event (along with one condition and one modification) but you can create events that contain multiple events (i.e. multiple triplets of target/condition/modification).
The targets that are currently available are:

 * `"taxa"` to modify anything linked to the [`lineage` list](#allowarguments). e.g. making half of the living taxa go extinct.
 * `"bd.params"` to modify anything linked to the [`bd.params` object](#makebdparams). For example you might want to change the distribution of one of the parameter after some `conditions`. This is typically done by updating the object using the argument `update` from `make.bd.params`.
 * `"traits"` to modify anything linked to the [`traits` object](#maketraits). For example you might want to change the trait process after some `conditions`. This is typically done by updating the object using the argument `update` from `make.traits`.
 * `"modifiers"` to modify anything linked to the [`modifiers` object](#makebdmodifiers). For example you might want to change the speciation rule after some `conditions`. This is typically done by updating the object using the argument `update` from `make.modifiers`.
 * `"founding"` this target is a bit more specialised and allows you run a nested `treats` object in the simulation. It is covered in a [specific section below](#founding).

We will see some examples of these targets in the `conditions` and `modifications` described below.

## Conditions

`condition` is a function that returns a logical value. When a specific condition is met, it should return `TRUE` and trigger the event, else it should return `FALSE`.

Currently there are three conditions functions implemented in `treats` but you can easily come up with your own version of them.

All `condition` functions in `treats` take at least two arguments: `x` for the variable of interest (e.g. time, number of taxa, trait value, etc.) and `condition`, the relational operator to evaluate. A relational operator is the proper (fancy) term designating all the comparisons you're regularly using in `R` like `==` (is equal?) `<` (is smaller?) `>=` (is bigger or equal?). You can get the full list in the R base manual (using `?Comparison` or any of the relational operator in a function form). For more details on how functions _really_ work in R see the [Advanced `R` book](https://adv-r.hadley.nz/functions.html)!

 * `age.condition` will trigger the `event` once a certain time is reached in the simulations and is the simplest/most basic condition with no arguments other than the time required. For example ``age.condition(4, condition = `==`)`` will trigger the `event` once the simulations reaches 4 time units. Easy.
 * `taxa.condition` will trigger the `event` once a certain number of taxa are reached. This can be considered including or excluding fossil species. For example ``taxa.condition(42, condition = `>=`, living = TRUE)`` will trigger the `event` if there are at least 42 living taxa.
 * `trait.condition` will trigger the `event` once a certain trait value is reached. This function allows to say which trait(s) to target (by default, the first one using `trait = 1`), what value of the trait to target (by default `what = max`) and whether to use and absolute trait value or not (`absolute = TRUE`). For example ``trait.condition(1/3, condition = `>`, trait = 1, what = sd)`` will trigger the condition after the standard deviation of the first trait reaches 1/3.

## Modifications

After defining the `event`, `target` and `condition`, you also need to specify what it should modify.
These are functions that should modify a specific aspect of the target.
Depending on the target you can modify the following:

 * if the target is `"taxa"` you can modify the internal [`lineage` list](#allowarguments) by removing living species using `random.extinction` or `trait.extinction`.
 * if the target is `"taxa"` you can modify the lineage tracker by removing living species using `random.extinction` or `trait.extinction`.
 * if the target is `"bd.params"` you can modify the [`bd.params` object](#makebdparams) using `bd.params.update`.
 * if the target is `"traits"` you can modify the [`traits` object](#maketraits) using `traits.update`.
 * if the target is `"modifiers"` you can modify the [`modifiers` object](#makemodifiers) using `modifiers.update`.
 * (for the `"founding"` target [see below](#founding))

The most straightforward example is for `modifications` on `bd.params`, `traits` or `modifiers` objects because they use the same syntax as for their generic `make.X` function.
For example, for `make.traits`, you can update a `trait` using the `update` argument as follows:

```{r}
## A BM trait in two dimensions
(BM_2D <- make.traits(n = 2, process = BM.process))
## Updating the 2D BM into a 2D OU
(OU_2D <- make.traits(update = BM_2D, process = OU.process))
```

So the function `update.X` applies the update to the object `X` when the event happens.
The following `event` updates the `bd.params` object by setting the extinction parameter to 1/3 when reaching a 10 species:

```{r, eval = FALSE}
make.events(target = "bd.params",
            condition = taxa.condition(10, condition = `>=`),
            modification = bd.params.update(extinction = 1/3))
```

> Note that by default events are triggered only once across the whole simulation, so although the example above states that the condition is reaching at least 10 taxa, it will not trigger every time it reaches more than 10 taxa, only the first time. You can change the number of times the events can be triggered using the argument `replications` (by default it's set to `replications = 0` for triggering the event only once).

## Examples

Here are some examples illustrating how to generate events.
For a simple example, we can create a extinction event that will remove 80% of species after reaching time 4:

```{r}
## 80% mass extinction at time 4
mass_extinction <- make.events(
                      target = "taxa",
                      condition = age.condition(4),
                      modification = random.extinction(0.8))

## Simulation parameters
stop.rule <- list(max.time = 5)
bd.params <- list(extinction = 0, speciation = 1)

## Running the simulations
set.seed(123)
results <- treats(bd.params = bd.params,
                stop.rule = stop.rule,
                events = mass_extinction)
## Plotting the results
plot(results, show.tip.label = FALSE)
axisPhylo()
```

Or for a slightly more complex example, we can can change the trait process from a BM to an OU when the trait values reaches an upper 95% quantile value above 2:

```{r}
## The 95% upper quantile value of a distribution
upper.95 <- function(x) {
    return(quantile(x, prob = 0.95))
} 
## Create an event to change the trait process
change_process <- make.events(
                  target = "traits",
                  ## condition is triggered if(upper.95(x) > 3)
                  condition = trait.condition(3, condition = `>`, what = upper.95),
                  modification = traits.update(process = OU.process))

## Set the simulation parameters
bd.params <- list(extinction = 0, speciation = 1)
stop.rule <- list(max.time = 6)
traits    <- make.traits()

## Run the simulations
set.seed(1)
no_change <- treats(bd.params = bd.params,
                  stop.rule = stop.rule,
                  traits = traits)
set.seed(1)
process_change <- treats(bd.params = bd.params,
                       stop.rule = stop.rule,
                       traits = traits,
                       events = change_process)
## Plot the results
par(mfrow = c(1,2))
plot(no_change, ylim = c(-7, 7))
plot(process_change, ylim = c(-7, 7))
par(mfrow = c(1,1))
```

## Founding events {#founding}

Founding events are a specific `target` for `events` that allows you to simulate a birth-death process within the current one!
It allows you to simulate a specific `treats` process using the `treats` function with its own `traits`, `modifiers`, `bd.params` and `events` (and yes, that's including `events` that have their own `founding` event).
This is basically the ultimate nested boss of modularity!

The founding event will run an internal `treats` process (i.e. simulating a tree and, optionally, some data) resulting in a founding sub-tree.
It will then branch this sub-tree onto the rest of the simulation that continued normally in the mean time.
You can specify the founding event using the inbuilt `founding.event` modification event.
This function takes the exact same arguments as `treats` to simulate the sub-tree with its own parameters.
Additionally, we will use the `additional.args` argument from `make.events` to specify a prefix for the founding tree tips (to make them easier to distinguish).

```{r}
## Set up parameters
stop.rule <- list(max.time = 4)
bd.params <- make.bd.params(speciation = 1, extinction = 0.3)

## Events that generate a new process (founding tree - with no extinction)
founding_event <- make.events(
                  target = "founding",
                  condition = taxa.condition(10),
                  modification = founding.event(
                                    bd.params = make.bd.params(speciation = 2,
                                                               extinction = 0)),
                  additional.args = list(prefix = "founding_"))

## Simulations
set.seed(11)
founding_tree <- treats(bd.params = bd.params,
                      stop.rule = stop.rule,
                      events = founding_event)
plot(founding_tree, cex = 0.4)
```

Note that the nestedness here is potentially endless, for example, you can pass an `event` argument to the `founding.event` that will generate another founding event, etc.

<!--chapter:end:04_make.events.Rmd-->

---
title: "`other functionalities`"
author: "Thomas Guillerme (guillert@tcd.ie)"
date: "`r Sys.Date()`"
bibliography: [../references.bib]
output:
  html_document: default
  pdf_document: default
---

```{r, echo = FALSE}
library(treats)
library(rgl)
knitr::knit_hooks$set(webgl = hook_webgl)
```

# Other functionalities {#others}

## `make.bd.params` {#makebdparams}

In most examples above, birth-death parameters were set as fixed values, i.e. a speciation rate of $\lambda$ and an extinction rate of $\mu$ throughout the simulations (with maybe some events modifying these rates).
However, it is possible to set these rates as specific or changing distributions. 
You can do this using the `make.bd.params` function to provide either a `vector` of values:

```{r}
## An example where the speciation is randomly sampled among three values
my_bd_params <- make.bd.params(speciation = c(1/3,42))

## Building a tree using this set of parameters
set.seed(123)
plot(treats(stop.rule = list(max.taxa = 50), bd.params = my_bd_params))
## Note the regions in the tree with short branches
## (that's the speciation being 1/3 while the others are speciation = 42) 
```

Or to directly provide a function from which to sample:

```{r}
## Another example where speciation is drawn from the interval (0, 1)
make.bd.params(speciation = runif)
```

In this example, the `"bd.params"` object passed to the `treats` function will allow the birth-death process to sample the speciation parameter each time it is called (e.g. during the speciation/extinction step, the branch length step, etc.).
If using a function, you can fine tune the arguments to be passed to that function using the `speciation.args` or the `extinction.args` arguments (as a named list matching the function's arguments):

```{r}
## Speciation is drawn from the interval (0.5, 1.5)
make.bd.params(speciation = runif,
               speciation.args = list(min = 0.5, max = 1.5))
```

When using distributions for both the speciation and extinction parameters, you can run into the (usually) undesired problem of having an extinction rate that is higher than your speciation rate and thus all the taxa in your tree dying out.
You can avoid this problem by using the `joint` distribution argument. 
This will ensure that the sampling of the extinction rate is always lower or equal to the speciation rate.

```{r}
## Joint speciation and extinction sampled from uniform distribution 
## with speciation always >= to extinction
make.bd.params(speciation = runif, extinction = runif, joint = TRUE)
```

Finally to avoid negative sampling values, you can use the argument `absolute` that will make all the sampled values positive.
This argument is set to `TRUE` by default so you shouldn't have to worry about it most of the time unless you specifically need negative sampled values for your parameters.

```{r}
## Making the speciation sampling always positive
make.bd.params(speciation = rnorm, absolute = TRUE)
```

The two other possible arguments for this function are `test` and `update` that work the same way as [`make.traits`](#maketraits) or [`make.modifiers`](#makemodifiers)

If you are a visual person and your `bd.params` objects are getting a bit too complicated to remember, you can always quickly plot them. 
The function will sample from the `bd.params` object and show the results:

```{r}
par(mfrow = c(2,3))
plot(make.bd.params(), main = "Default birth-death parameters")
plot(make.bd.params(speciation = runif,
               speciation.args = list(min = 0.5, max = 1.5)),
     main = "Uniform speciation between 0.5 and 1.5\n(no extinction)")
plot(make.bd.params(speciation = runif,
                    speciation.args = list(min = 0.5, max = 1.5)),
     main = "Uniform speciation between 0.5 and 1.5\n(no extinction)")
plot(make.bd.params(speciation = runif, extinction = runif, joint = TRUE),
     main = "Joint uniform speciation and extinction")
plot(make.bd.params(speciation = rnorm, extinction = runif,
                    joint = FALSE, abs = FALSE),
     main = "Disjoint normal speciation and uniform extinction")
par(mfrow = c(1,1))
```

## `drop.things` {#dropthings}

You can use the function `drop.things` to drop specific elements of the tree and data at the same time by providing the argument `what = "fossils"` for tips that went extinct, or `what = "livings"` for tips that where alive at the end of the simulation or `what = "singles"` to drop internal singleton nodes.
Alternatively you can use the function aliases `drop.fossils`, `drop.livings` or `drop.singles` for the exact same results:

```{r}
## A random tree with fossils and traits and internal nodes every 0.5 times
set.seed(3)
my_data <- treats(stop.rule = list(max.taxa = 20),
                bd.params = list(speciation = 1, extinction = 1/3),
                traits    = make.traits(), save.steps = 0.5)

## A tree with 20 tips and 54 nodes
my_data$tree
## And a dataset with 74 rows
dim(my_data$data)

## Removing the fossil species
drop.things(my_data, what = "fossils")$tree
dim(drop.fossils(my_data)$data)

## Removing the living species
drop.things(my_data, what = "livings")$tree
dim(drop.livings(my_data)$data)

## Removing the internal nodes
drop.things(my_data, what = "singles")$tree
dim(drop.singles(my_data)$data)

## Removing the internal nodes AND the fossils
drop.singles(drop.fossils(my_data))
```

## `"treats"` internal utilities

The package also provides utilities for internal functions, namely for designing `modifiers` or `events` more easily.
These functions don't do anything useful on their own but are optimised to be used internally in `treats`.
For all these functions, you can look at the internal manual for an example (i.e. using `?<function_name>`)

So far the package has the following internals:

Function  | What it does | Where can it be used
----------|--------------|---------------------
`parent.traits` | selects the trait values of the current lineage's parents (i.e. direct ancestor) | in `make.modifiers`
`taxa.condition` | provides a trigger for an `event` dependent on the number of taxa | in `make.events`
`age.condition` | provides a trigger for an `event` dependent on time | in `make.events`
`trait.condition` | provides a trigger for an `event` dependent on trait values | in `make.events`

<!--chapter:end:05_other_functionalities.Rmd-->

---
title: "`plot.treats`"
author: "Thomas Guillerme (guillert@tcd.ie)"
date: "`r Sys.Date()`"
# bibliography: [../References.bib, ../packages.bib]
output:
  html_document: default
  pdf_document: default
---

```{r, echo = FALSE}
library(treats)
library(rgl)
knitr::knit_hooks$set(webgl = hook_webgl)
options(rgl.useNULL = TRUE) # Suppress the separate window.
```

# Plots

`"treats"` objects can be directly plotted in `treats` using the S3 `plot.treats` function (or just `plot(x)` if `x` is of class `"treats"`).

## Plotting traits

`"treats"` `"traits"` objects are covered in the [traits section](#maketraits).
You can use `plot.treats` to plot them by choosing which specific trait to plot using the `trait` argument (default is 1):

```{r, fig.height=8, fig.width=8}
## Making a list of three traits
list_of_traits <- make.traits(process = c(no.process, BM.process, OU.process), trait.names = c("No process (normal)", "Brownian motion", "Ornstein-Uhlenbeck"))
## Plotting each trait separately
par(mfrow = c(2, 2))
plot(list_of_traits, trait = 1)
## Using different colours options
plot(list_of_traits, trait = 2, col = c("red", "purple", "pink"))
## Not using the default plot name
plot(list_of_traits, trait = 3, main = "OU process",
     cent.tend = sd, quantiles = c(10, 30))
par(mfrow = c(1, 1))
```

You can also control the number of replicates in the simulation by using the `simulations` option (the default is 50).
Bigger numbers lead to more time but smoother looking plots while smaller ones are more stochastic:

```{r, fig.width=6, fig.height=12}
par(mfrow = c(2,1))
plot(list_of_traits, trait = 2, simulations = 10, main = "10 BM simulations")
plot(list_of_traits, trait = 2, simulations = 1000, main = "1k BM simulations")
par(mfrow = c(1,1))
```

## Plotting `treats` results

If `treats` is used to plot only a tree (and outputs a `"phylo"` object), you can use the function `plot.phylo` from the `ape` package to plot your tree.
You'll find all the tree plotting options in the `?plot.phylo` manual page.

```{r}
## A simple pure birth tree
my_tree <- treats(stop.rule = list(max.taxa = 20))
plot(my_tree, main = "Plotting a \"phylo\" object")
```

However, if you also simulated a trait along with the tree you can use the `plot.treats` function to plot both the tree and the trait:

```{r}
## A simple pure birth tree with a BM process
my_tree <- treats(stop.rule = list(max.taxa = 20), traits = make.traits())
## Playing with the default options
plot(my_tree, main = "A tree and traits")
```

By default, elements are coloured as follows: nodes and tips are points and coloured in blue if they are tips (light blue if they are fossils) and nodes are in orange.
Branches linking them are grey lines.
You can of course change this colour palette to whatever you prefer by calling the standard base R arguments that can be passed to `points` or `lines`. For example `pch` for the point type or `lty` for the line type.
Furthermore, time is plotted conventionally from left to right (left is towards the past, right is towards the present) but you can change that by specifying `xlim`.

```{r, fig.height = 6, fig.width = 6}
## Playing with some more options
plot(my_tree, main = "A tree and traits",
     ## Changing nodes colours and type
     col = c(tips = "pink", nodes = "purple"), pch = 21,
     ## Changing edge colour and type
     lwd = 2, lty = 3, edges = "yellow",
     ## Changing the x axis orientation and label
     xlim = c(0, 2.5), xlab = "Time into the past")
```

> Note that in `plot.treats` you can modify colours using the `col` argument by providing a clear indication of what you want to colour (e.g. `col = c(tips = "blue", fossils = "orange")` will apply the colours to the unambiguously named elements) or by providing a function that will scale the colours with time (e.g. `col = grDevices::heat.colors` will use the specified function to change the  colours of the elements through time).

You can add a default legend by using `legend = TRUE` (if you don't want to add the default legend you can add it after your plot using `legend(...)`).

```{r, fig.height = 6, fig.width = 6}
## Adding the default legend
plot(my_tree, legend = TRUE)
```

### Adding dimensions! Not necessary, but fun!

Of course, if you're simulating multiple traits, you can always plot different ones.

```{r, fig.height = 6, fig.width = 6}
## Specifying a 3D trait process
my_3D_trait <- make.traits(n = 3, )
## Simulating a birth-death tree with that 3D trait
my_data <- treats(bd.params = list(extinction = 0.2),
                stop.rule = list(max.living = 50),
                traits    = my_3D_trait)
```

You can toggle which trait to plot using the `trait` option, either by providing a single value to plot that specific trait against time or by providing two traits.

```{r, fig.height = 12, fig.width = 6}
par(mfrow = c(2,1))
## Plotting the second trait
plot(my_data, trait = 2, main = "Trait 2")
## Plotting the correlation between trait 1 and 2
plot(my_data, trait = c(1,2), main = "Traits 1 and 2")
```

As mentioned above, the `col` argument can take a function for scaling the elements colours with time.
This can be useful for adding time a third dimension to these 2D plots:

```{r}
## Plotting the correlation between trait 1 and 2
## with time as a 3rd dimensions
plot(my_data, trait = c(1,2), main = "Traits 1 and 2",
     col = grDevices::heat.colors, legend = TRUE,
     ## Highlighting the tips in black for visibility
     tips.nodes = "black")
```

But there's more! You can also plot these results using actual 3D plots that you can spin and all that!
This is done through the `rgl` package and activated in `treats` using the option `use.3D = TRUE`.

If two traits are provided, the 3rd dimension is going to be time by default:

```{r}
## Plotting the tree and 2 traits in 3D: woah!
plot(my_data, trait = c(1, 2), use.3D = TRUE)
rglwidget()
```

If three traits are provide, you can set up a fourth coloured dimension as time (similarly to the 2D plots with the colour gradients).
This uses the `rgl` functions, namely `lines3d`, `points3d` (if `type = "p"`; default) or `sphere3d` (if `type = "s"`).
The additional arguments `...` are directly handled and attributed to the corresponding function.

```{r}
## Plotting the tree and 3 traits in 3D
plot(my_data, trait = c(1, 2, 3), use.3D = TRUE,
     col = grDevices::heat.colors, type = "s", radius = 0.1)
rglwidget()
```

<!--chapter:end:06_plotting_dads_objects.Rmd-->

---
title: "treats manual"
author: "Thomas Guillerme (guillert@tcd.ie)"
date: "`r Sys.Date()`"
# bibliography: [../References.bib, ../packages.bib]
output:
  html_document: default
  pdf_document: default
---

# Specific examples

```{r, echo = FALSE}
library(treats)
library(rgl)
knitr::knit_hooks$set(webgl = hook_webgl)
options(rgl.useNULL = TRUE) # Suppress the separate window.
```

In this section I will illustrate a series of examples of specific more complex scenarios.
They can be all run independently and used as a basis for your own specific scenarios.

Here is a table summarising which functionalities are used in which example:

Functionality showcase | Example
--------------|---------
`bd.params` | [reducing speciation rate](#EGred_spec), [generating a subtree with no extinction](#EG_founding_purebirth)
`make.events` | [mass extinction](#EGrandom_ext), [negative trait extinction](#EGneg_ext), [background extinction](#EGbg_ext), [reducing speciation rate](#EGred_spec), [changing trait process](#EG_change_trait), [changing trait correlation](#EG_change_correlation), [changing modifiers](#EG_change_modif), [change branch length](#EG_modify_brlen), [generating a subtree with no extinction](#EG_founding_purebirth), [generating a subtree with a different process](EG_founding_traits)
`make.traits` | [negative trait extinction](#EGneg_ext), [changing trait process](#EG_change_trait), [changing trait correlation](#EG_change_correlation), [generating a subtree with a different process](EG_founding_traits)
`make.modifiers` | [changing modifiers](#EG_change_modif), [change branch length](#EG_modify_brlen)

And more specifically for the events:

Functionality showcase | Example
--------------|----------
`age.condition` | [mass extinction](#EGrandom_ext), [negative trait extinction](#EGneg_ext), [reducing speciation rate](#EGred_spec), [changing trait process](#EG_change_trait), [changing modifiers](#EG_change_modif)
`random.extinction` | [mass extinction](#EGrandom_ext)
`trait.extinction` | [negative trait extinction](#EGneg_ext)
`taxa.condition` | [background extinction](#EGbg_ext), [change branch length](#EG_modify_brlen)
`bd.params.update` | [background extinction](#EGbg_ext), [reducing speciation rate](#EGred_spec)
`traits.update` | [changing trait process](#EG_change_trait), [changing trait correlation](#EG_change_correlation)
`trait.condition` | [changing trait correlation](#EG_change_correlation), [generating a subtree with a different process](EG_founding_traits)
`modifiers.update` | [changing modifiers](#EG_change_modif), [change branch length](#EG_modify_brlen)
`parent.traits` | [changing modifiers](#EG_change_modif)
`founding.event` | [generating a subtree with no extinction](#EG_founding_purebirth), [generating a subtree with a different process](EG_founding_traits)

<!-- TODO add a recursive event example where discrete self rate is changed by half everytime you reach 10 living species -->

## Random mass extinction after some time {#EGrandom_ext}

For this scenario, we want to generate a pure birth tree (no traits and no extinction) where 80% of the species go extinct two thirds of the way into the scenario.

For that we first need to first set our simulation parameters: we will be running the simulation for 6 time units and with a speciation rate of 1 (and extinction of 0 - default).

```{r}
## Setting the parameters
stop_time_6 <- list(max.time = 6)
speciation_1 <- make.bd.params(speciation = 1)
```

We will then create a event that triggers when reaching half of the simulation time (using `age.condition(2.5)`).
This event will target `"taxa"`, i.e. the number of species, and randomly remove 80% of them (using `random.extinction(0.8)`):

```{r}
## 80% mass extinction at time 4
mass.extinction <- make.events(
                    condition = age.condition(4),
                    target = "taxa",
                    modification = random.extinction(0.8))
```

Once these parameters are defined, we can run the simulations and plot the results:

```{r}
## Running the simulations
set.seed(1)
results <- treats(stop.rule = stop_time_6,
                bd.params = speciation_1,
                events    = mass.extinction)
## Plotting the results
plot(results, show.tip.label = FALSE)
axisPhylo()
```

## Species with negative trait values go extinct after a certain time {#EGneg_ext}

For this scenario, we want to generate a pure birth tree with a one dimensional Brownian Motion trait for 5 time units.
We then want species with a negative trait value go extinct after 4 time units.

First we need to set up the simulation parameters:
 * The stopping rule (5 time units)
 * The birth-death parameters (speciation rate of 1)

```{r}
## Simulation parameters
stop_time_5 <- list(max.time = 5)
speciation_1 <- make.bd.params(speciation = 1)
```

Then set up our trait which is a one dimensional Brownian Motion

```{r}
## Trait
simple_bm_trait <- make.traits(n = 1, process = BM.process)
```

And our extinction event which triggers after reaching time 4 (`age.condition(4)`), targets the `"taxa"` and modifies the extinction for species with traits lower than 0.

```{r}
## Extinction of any tips with trait < 1 at time 4
trait.extinction <- make.events(
                      target = "taxa",
                      condition = age.condition(4),
                      modification = trait.extinction(x = 0,
                                                      condition = `<`))
```

Once these parameters are defined, we can run the simulations and plot the results:

```{r}
## Running the simulations
set.seed(7)
results <- treats(stop.rule = stop_time_5,
                bd.params = speciation_1,
                traits    = simple_bm_trait,
                events    = trait.extinction)
## Plotting the results
plot(results)
```

## Adding a background extinction after reaching a number of living taxa {#EGbg_ext}

For this scenario, we want to generate a pure birth tree until reaching 50 living taxa but with an extinction rate appearing after reaching 30 taxa.

First we need to set up the simulation parameters:
 * The stopping rule (50 taxa max)
 * The birth-death parameters (speciation rate of 1)

```{r}
## Simulation parameters
stop_taxa_50 <- list(max.living = 50)
speciation_1 <- make.bd.params(speciation = 1)
```

And our change in extinction rate event which triggers after reaching 30 taxa (`taxa.condition(30)`), targets the `"bd.params"` (birth-death parameters) sets the extinction rate to 0.5 (`bd.params.update(extinction = 0.5)`):

```{r}
## Adding an extinction parameter after 30 taxa
background.extinction <- make.events(
                      condition = taxa.condition(30),
                      target = "bd.params",
                      modification = bd.params.update(extinction = 0.5))
```

Once these parameters are defined, we can run the simulations and plot the results:

```{r}
## Running the simulations
set.seed(2)
results <- treats(stop.rule = stop_taxa_50,
                bd.params = speciation_1,
                events = background.extinction)
## Plotting the results
plot(results, show.tip.label = FALSE)
```

## Reducing speciation rate after a certain time {#EGred_spec}

For this scenario, we want to generate a birth tree for 6 times units with random speciation rates (i.e. drawn from a uniform (0.5;1) distribution) which reduces after time 4 through the simulations to a fixed value of 1/3.

First we need to set up the simulation parameters:
 * The stopping rule (6 time units)
 * The birth-death parameters (speciation randomly drawn between 0.5 and 1)

```{r}
## Simulation parameters
stop_time_6 <- list(max.time = 6)
random_speciation <- make.bd.params(speciation = runif, 
                                    speciation.args = list(min = 0.5, max = 1))
```

And our extinction event which triggers after reaching time 4 (`age.condition(4)`), targets the `"bd.params"` and modifies the speciation rate to 1/3:

```{r}
## Reducing speciation after reaching time 4
reduced.speciation <- make.events(
                      condition = age.condition(4),
                      target = "bd.params",
                      modification = bd.params.update(speciation = 1/3))
```

Once these parameters are defined, we can run the simulations and plot the results.
We can contrast the results with the scenario without an event (but same random seed):

```{r}
## No event
set.seed(42)
no_event <- treats(stop.rule = stop_time_6,
                 bd.params = random_speciation)

## Reduced speciation event
set.seed(42)
reduced_speciation_event <- treats(stop.rule = stop_time_6,
                                 bd.params = random_speciation,
                                 events = reduced.speciation)

## Plot both trees
par(mfrow = c(1, 2))
plot(no_event, main = "No event", show.tip.label = FALSE)
axisPhylo()
plot(reduced_speciation_event, 
     main = "Reduced speciation after time 5", 
     show.tip.label = FALSE)
axisPhylo()
```

## Changing the trait process after some time {#EG_change_trait}

For this scenario, we want to generate a pure birth tree with a one dimensional Brownian Motion trait for 5 time units which then changes to an OU process.

First we need to set up the simulation parameters:
 * The stopping rule (6 time units)
 * The birth-death parameters (speciation rate of 1)

```{r}
## Simulation parameters
stop_time_6 <- list(max.time = 6)
speciation_1 <- make.bd.params(speciation = 1)
```

Then set up our trait which is a one dimensional Brownian Motion

```{r}
## Trait
simple_bm_trait <- make.traits(n = 1, process = BM.process)
```

And our event which triggers after reaching time 5 (`age.condition(5)`), targets the `"traits"` and modifies the process to `OU.process`.

```{r}
## Create an event to change the trait process
change.process.to.OU <- make.events(
                  condition    = age.condition(5),
                  target       = "traits",
                  modification = traits.update(process = OU.process))
```

Once these parameters are defined, we can run the simulations and plot the results.
We can contrast the results with the scenario without an event (but same random seed):

```{r}
## Run the simulations without change
set.seed(1)
no_change <- treats(stop.rule = stop_time_6,
                  bd.params = speciation_1,
                  traits    = simple_bm_trait)
## Run the simulations with change
set.seed(1)
process_change <- treats(stop.rule = stop_time_6,
                       bd.params = speciation_1,
                       traits    = simple_bm_trait,
                       events    = change.process.to.OU)
## Plot the results
par(mfrow = c(1,2))
plot(no_change, ylim = c(-7, 7))
plot(process_change, ylim = c(-7, 7))
```

## Changing trait correlation after reaching a trait value {#EG_change_correlation}

For this scenario, we want to generate a pure birth tree with a 2 dimensional Brownian Motion trait with a strict correlation between the two dimensions (1:1) that loosen up when a taxa reaches an absolute value of 2.

First we need to set up the simulation parameters:
 * The stopping rule (100 taxa)
 * The birth-death parameters (speciation rate of 1)

```{r}
## Set the parameters
stop_taxa_100 <- list(max.taxa = 100)
speciation_1 <- make.bd.params(speciation = 1)
```

Then set up our trait which is a 2 dimensional Brownian Motion with a correlation matrix Sigma ()

```{r}
## A 2D variance covariance matrix
cor_matrix <- matrix(1, 2, 2)

## A correlated 2D Brownian Motion
correlated_2D_BM <- make.traits(n = 2, process = BM.process,
                      process.args = list(Sigma = cor_matrix))
```

And our event which triggers after a taxa gets the trait value 3 (`trait.condition(3, absolute = TRUE)`), targets the `"traits"` and modifies the traits correlation

```{r}
## New correlation
new_cor <- matrix(c(10,3,3,2),2,2)

## Event changing a trait correlation
correlation.change <- make.events(
    condition    = trait.condition(3, absolute = TRUE),
    target       = "traits",
    modification = traits.update(process.args = list(Sigma = new_cor)))
```

Once these parameters are defined, we can run the simulations and plot the results.
We can contrast the results with the scenario without an event (but same random seed):

```{r}
## Run the simulations
set.seed(2)
no_event <- treats(stop.rule = stop_taxa_100,
                 bd.params = speciation_1,
                 traits    = correlated_2D_BM)
set.seed(2)
change_correlation <- treats(stop.rule = stop_taxa_100,
                           bd.params = speciation_1,
                           traits    = correlated_2D_BM,
                           events    = correlation.change)

## Visual testing
par(mfrow = c(1,2))
plot(no_event, trait = c(1,2), main = "Strict correlation")
plot(change_correlation, trait = c(1,2), main = "Loosened correlation")
```

And we can visualise this change through time:

```{r}
## 3D plot
plot(change_correlation, trait = c(1:2), use.3D = TRUE)
rglwidget()
```

## Event for changing a modifier: extinction event increase for species with negative values {#EG_change_modif}

For this scenario, we want to generate a pure birth tree with a one dimensional Brownian Motion trait for 4 time units.
After 3 time units, we want the speciation rule to increase for species which ancestors have a negative trait value.

First we need to set up the simulation parameters:
 * The stopping rule (5 time units)
 * The birth-death parameters (speciation rate of 1)

```{r}
## Set the parameters
stop_time_4 <- list(max.time = 4)
speciation_1 <- make.bd.params(speciation = 1)
```

Then set up our trait which is a one dimensional Brownian Motion

```{r}
## Trait
simple_bm_trait <- make.traits(n = 1, process = BM.process)
```

And a modifier that is the default birth-death algorithm rules

```{r}
## birth-death rules (default)
default_modifiers <- make.modifiers()
```

And our extinction event which triggers after reaching time 3 (`age.condition(3)`), targets the `"modifiers"` and modifies birth-death rule as follows:
 * When a species is descendant from a parent with a negative trait value (`negative.trait.condition`);
 * Then increase your chances of going extinct by +1 (`increase.extinction.1`)

```{r}
## New condition and new modifier (increasing speciation if trait is negative)
negative.trait.condition <- function(trait.values, lineage) {
    return(parent.traits(trait.values, lineage) < 0)
}
increase.extinction.1 <- function(x, trait.values, lineage) {
  return(x + 1)
}

## Update the modifier
change.speciation <- make.events(
    condition    = age.condition(3),
    target       = "modifiers",
    modification = modifiers.update(speciation = speciation,
                                    condition  = negative.trait.condition,
                                    modify     = increase.extinction.1))
```

Once these parameters are defined, we can run the simulations and plot the results.
We can contrast the results with the scenario without an event (but same random seed):

```{r}
set.seed(4)
no_event <- treats(stop.rule = stop_time_4,
                 bd.params = speciation_1,
                 traits    = simple_bm_trait,
                 modifiers = default_modifiers)
set.seed(4)
change_spec <- treats(stop.rule = stop_time_4,
                    bd.params = speciation_1,
                    traits    = simple_bm_trait,
                    modifiers = default_modifiers,
                    events    = change.speciation)

## Visualise the results
par(mfrow = c(1,2))
plot(no_event, main = "No event")
plot(change_spec, main = "Increase extinction for negative\ntraits after time 3")
```

## Changing branch length when reaching n taxa {#EG_modify_brlen}

For this scenario, we want to generate a pure birth tree until reaching 100 taxa.
After reaching 30 taxa we want branch length growth to increase 50 folds.

First we need to set up the simulation parameters:
 * The stopping rule (100 taxa)
 * The birth-death parameters (speciation rate of 1)

```{r}
## Set the parameters
stop_taxa_100<- list(max.taxa = 100)
speciation_1 <- make.bd.params(speciation = 1)
```

Then a modifier that is the default birth-death algorithm rules

```{r}
## birth-death rules (default)
default_modifiers <- make.modifiers()
```

And event which triggers after reaching 30 taxa (`taxa.condition(30)`), targets the `"modifiers"` and modifies the branch length generation rule to a 50 fold increase (`increase.50.folds`)

```{r}
## Multiplying branch length 50 folds
increase.50.folds <- function(x, trait.values, lineage) {
  return(x * 50)
}
## Event for increasing branch length after reaching 30 taxa
increase_brlen <- make.events(
                  condition = taxa.condition(30),
                  target = "modifiers",
                  modification = modifiers.update(
                                    branch.length = branch.length,
                                    modify  = increase.50.folds))
```

Once these parameters are defined, we can run the simulations and plot the results.
We can contrast the results with the scenario without an event (but same random seed):

```{r}
## Run the simulations
set.seed(5)
no_event <- treats(stop.rule = stop_taxa_100,
                 bd.params = speciation_1,
                 modifiers = default_modifiers)
set.seed(5)
increased_brlen <- treats(stop.rule = stop_taxa_100,
                        bd.params = speciation_1,
                        modifiers = default_modifiers,
                        events = increase_brlen)

## Visualise the results
par(mfrow = c(1,2))
plot(no_event, main = "No event", show.tip.label = FALSE)
plot(increased_brlen, main = "Increase branch length\nafter 30 taxa",
    show.tip.label = FALSE)
```

## Founding event: generating a subtree with no fossils {#EG_founding_purebirth}

For this scenario, we want to generate a birth-death tree for 4 time units.
After reaching 10 taxa, one random taxa will give birth to a sub-tree that is a pure birth tree (no extinction).

First we need to set up the simulation parameters:
 * The stopping rule (5 time units)
 * The birth-death parameters (speciation rate of 1 and extinction of 0.2)

```{r}
## Set up parameters
stop_time_4 <- list(max.time = 4)
spec_1_ext_02 <- make.bd.params(speciation = 1, extinction = 0.2)
```

And our event which triggers after reaching 10 taxa (`taxa.condition(10)`), and generates a subtree ("founding") that is a pure birth tree (no extinction and speciation rate of 2).

```{r}
## Setting the pure-birth parameters
speciation_2 <- make.bd.params(speciation = 2)

## Events that generate a new process (founding effects)
founding_event <- make.events(
                  condition = taxa.condition(10),
                  target  = "founding",
                  modification = founding.event(
                                    bd.params = speciation_2),
                  additional.args = list(prefix = "founding_"))
```

> Note we are prodviding an additional argument `prefix` here so that we can track which species are part of the sub tree for colouring them down the line.

Once these parameters are defined, we can run the simulations and plot the results:

```{r}    
## Simulations
set.seed(11)
founding_tree <- treats(stop.rule = stop_time_4,
                      bd.params = spec_1_ext_02,
                      events = founding_event)

## Selecting the edges colours
tip_values <- rep("black", Ntip(founding_tree))
tip_values[grep("founding_", founding_tree$tip.label)] <- "orange"
edge_colors <- match.tip.edge(tip_values, founding_tree, replace.na = "black")

## Plotting the results
plot(founding_tree, show.tip.label = FALSE, edge.color = edge_colors)
```

## Founding event: generating a subtree a different process {#EG_founding_traits}

For this scenario, we want to generate a birth-death tree with a one dimensional Brownian Motion trait for 6 time units.
After a taxon reaches the value 3 or higher, it gives birth to a sub-tree that generates an OU trait with a long-term mean at the value 3.

First we need to set up the simulation parameters:
 * The stopping rule (6 time units)
 * The birth-death parameters (speciation rate of 1 and extinction of 1/3)

```{r}
## The tree parameters
stop_time_6 <- list(max.time = 6)
speciation_1_extinction_03 <- make.bd.params(speciation = 1,
                                             extinction = 0.3)
```

Then set up our trait which is a one dimensional Brownian Motion

```{r}
## Trait
simple_bm_trait <- make.traits(n = 1, process = BM.process)
```

When a taxa reaches the value 3 `trait.condition`, it generates a pure birth tree (`speciation = 2`) with an OU trait with the long-term mean value 3.

```{r}
## The OU trait with a long-term mean value of 3
OU_3 <- make.traits(process = OU.process,
                    start = 3, process.args = list(optimum = 3))
## The pure birth parameters
speciation_2 <- make.bd.params(speciation = 2)

## The founding event
new_OU_trait <- make.events(
                    condition    = trait.condition(3, condition = `>=`),
                    target       = "founding",
                    modification = founding.event(
                                    bd.params = speciation_2,
                                    traits    = OU_3))
```

Once these parameters are defined, we can run the simulations and plot the results:

```{r}
## Simulating the tree
set.seed(1)
founding_tree <- treats(stop.rule = stop_time_6,
                      bd.params = speciation_1_extinction_03,
                      traits = simple_bm_trait,
                      events = new_OU_trait)
plot(founding_tree)
```

<!--chapter:end:07_specific_examples.Rmd-->

### 2024-05-15 inst/gitbook/_book/treats_manual.tex
Book up

@@
 \author{Thomas Guillerme (\href{mailto:guillert@tcd.ie}{\nolinkurl{guillert@tcd.ie}})}
-\date{2023-12-13}
+\date{2024-05-15}
 
@@
 For that we need to create a \texttt{"modifiers"} object that modifies the \texttt{speciation} process with a specific condition and a specific modification when that condition is met.
-For a speciation to occur (and a species to not go extinct), the algorithm draws a random value \emph{x} between 0 and 1 and if the value is smaller than the speciation parameter divided by the speciation + extinction parameter, the lineage speciates (\(x < (\lambda \ (\lambda + \mu))\)), else it goes extinct.
+For a speciation to occur (and a species to not go extinct), the algorithm draws a random value \emph{x} between 0 and 1 and if the value is smaller than the speciation parameter divided by the speciation + extinction parameter, the lineage speciates (\(x < (\lambda \div (\lambda + \mu))\)), else it goes extinct.
 \emph{x} is a random value that simulates some stochasticity in the evolutionary simulation.
@@
 However, we can create a modifier that introduces less randomness (or a different level of randomness) by forcing the lineage to go extinct every time a specific condition is met, e.g.~if the lineage trait value is negative.
-This can be done for example by not randomly generating the value \emph{x} but giving it a fixed value of say 1 so that the estimation of the speciation function \(x < (\lambda \ (\lambda + \mu))\) is always met.
-In other words, if the lineage's trait value is negative, make sure that \emph{x} is equal to 1, so that the speciation condition \(x < (\lambda \ (\lambda + \mu))\) is never met.
+This can be done for example by not randomly generating the value \emph{x} but giving it a fixed value of say 1 so that the estimation of the speciation function \(x < (\lambda \div (\lambda + \mu))\) is always met.
+In other words, if the lineage's trait value is negative, make sure that \emph{x} is equal to 1, so that the speciation condition \(x < (\lambda \div\ (\lambda + \mu))\) is never met.
 
@@
 ## }
-## <bytecode: 0x560fcf695180>
+## <bytecode: 0x609b29299e58>
 ## <environment: namespace:treats>
@@
 
-\begin{verbatim}
-## Warning in title(...): conversion failure on 'The Ornstein–Uhlenbeck process'
-## in 'mbcsToSbcs': dot substituted for <e2>
-\end{verbatim}
-
-\begin{verbatim}
-## Warning in title(...): conversion failure on 'The Ornstein–Uhlenbeck process'
-## in 'mbcsToSbcs': dot substituted for <80>
-\end{verbatim}
-
-\begin{verbatim}
-## Warning in title(...): conversion failure on 'The Ornstein–Uhlenbeck process'
-## in 'mbcsToSbcs': dot substituted for <93>
-\end{verbatim}
-
 \includegraphics{treats_manual_files/figure-latex/unnamed-chunk-57-1.pdf}
@@
 ## }
-## <bytecode: 0x560fce0ebbe0>
+## <bytecode: 0x609b211b6680>
 ## <environment: namespace:treats>
@@
 ## }
-## <bytecode: 0x560fce0eaf30>
+## <bytecode: 0x609b211694e0>
 ## <environment: namespace:treats>
@@
 ## }
-## <bytecode: 0x560fce0ea600>
+## <bytecode: 0x609b21164168>
 ## <environment: namespace:treats>
@@
 \begin{minipage}[t]{0.20\columnwidth}\raggedright
-\texttt{time.condition}\strut
+\texttt{age.condition}\strut
 \end{minipage} & \begin{minipage}[t]{0.28\columnwidth}\raggedright
@@
 
-\includegraphics[width=6.5in]{../../../../../../tmp/Rtmp6UK3fU/file52c0514edd4f}
+\includegraphics[width=6.5in]{../../../../../../tmp/RtmpcKuqE6/file37eb79e5b154}
 
@@
 
-\includegraphics[width=6.5in]{../../../../../../tmp/Rtmp6UK3fU/file52c079e156e9}
+\includegraphics[width=6.5in]{../../../../../../tmp/RtmpcKuqE6/file37eb42b1cdb9}
 
@@
 
-\includegraphics[width=6.5in]{../../../../../../tmp/Rtmp6UK3fU/file52c02397d39c}
+\includegraphics[width=6.5in]{../../../../../../tmp/RtmpcKuqE6/file37eb202f55d4}
 

### 2024-06-28 inst/gitbook/01_getting_started.Rmd
removed wrong ref file

@@
 date: "`r Sys.Date()`"
-bibliography: [../references.bib]
+bibliography: [../References.bib]
 output:
@@
 
-They use a dataset from @beckancient2014 that is a discrete morphological space containing the ordination of the around 400 morphological characters for 50 species and estimated for their 49 descendants into an ordinated shapespace of 99 elements and 97 dimensions.
+They use a dataset from @beck2014ancient that is a discrete morphological space containing the ordination of the around 400 morphological characters for 50 species and estimated for their 49 descendants into an ordinated shapespace of 99 elements and 97 dimensions.
 

### 2024-06-28 inst/gitbook/02_make.traits.Rmd
added vignette for link.traits

@@
 
+## Linking traits together with `link.traits`
+
+In the `treats` algorithm, traits are simulated in parallel for improving performances.
+Most of the cases this is a (really) good thing!
+However, in some cases you might need some of your traits to be simulated sequentially.
+For example if you want a specific trait process or options to be dependent on the result of another trait.
+You can set up these linked traits using the `link.traits` function.
+This function takes `base.trait`, a `"treats"` `"traits"` object created using `make.traits` that will be the first trait simulated and the `next.trait`, a list of  `"treats"` `"traits"` objects that will be simulated after the `base.trait` depending on the link type (`link.type`) and some other arguments (`link.args`).
+So far the `link.traits` function only accommodates `link.type = "conditional"` for conditional traits.
+
+### Conditional traits
+
+Say for example that you have a primary trait (`base.trait`) that is a discrete binary trait that generates values `0` or `1`.
+This can correspond to some abiotic trait for example (e.g. "is the lineage on an island or on the mainland").
+You can first generate this trait using the `make.traits` function as detailed above (note here we're using a transition matrix as a specific argument that makes the changes from state 0 to state 1 relatively rare and the reversal even more rare).
+
+```{r}
+## Setting up a discrete trait
+discrete_trait <- make.traits(discrete.process,
+       process.args = list(transitions = matrix(c(3, 0.2, 0.05, 3), 2, 2)),
+       trait.names  = "discrete")
+```
+
+Following this trait value, we can create another trait that is conditional to the outcome of this first discrete trait simulation.
+In other words we can specify a choice between two processes depending on the result of the `base.trait`.
+For example, we can say that if the lineage evolves on an island (state `"1"`) it will evolve using a Brownian motion process but if it evolves on the mainland (state `"0"`), it will evolve using an OU process.
+
+We'll need to first generate these two traits independently using `make.traits` (using different options):
+
+```{r}
+## A standard Brownian Motion trait
+BM_trait <- make.traits(process = BM.process)
+## An OU trait with a strong alpha
+OU_trait <- make.traits(process = OU.process, process.args = list(alpha = 10))
+## Combining both traits into a list
+continuous_trait <- list("mainland_case" = BM_trait,
+                         "island_case"   = OU_trait)
+```
+
+We then need to create a list of functions to interpret the results of the first trait (`discrete_trait`) so that if the results of `discrete_trait` is `0`, `BM_trait` is used, if it's `1` `OU_trait` is used.
+This is done by creating a list of functions that evaluate `x1`, the outcome of `discrete_trait`, and return a logical (`TRUE` or `FALSE`).
+For example `function(x1) {x1 != 0}` will return `TRUE` if the `discrete_trait` is not state `0`:
+
+```{r}
+## Creating a list of functions to choose between either case for the continuous trait
+select_continuous <- list("is.mainland" = function(x1) {x1 == 0},
+                          "is.island"   = function(x1) {x1 == 1})
+```
+
+Note that this list must be in the same order as the `continuous_trait`: the functions will be evaluated sequentially (i.e.
+test the function `is.mainland`, then `is.island`) and the first one to return `TRUE` will be used to select the function in `continuous_trait`.
+For example in our case if the score of `discrete_trait` is `1`, the second function from `select_continuous` is `TRUE` so the second element from `continuous_trait` is used (the `OU_trait`).
+
+We can then combine all that using `link.traits`:
+
+```{r}
+## Link these traits
+conditional_trait <- link.traits(# The first trait to be evaluated
+                                 base.trait = discrete_trait,
+                                 # The other traits to be evaluated next
+                                 # (here a list of two traits)
+                                 next.trait = continuous_trait,
+                                 link.type  = "conditional",
+                                 # The functions to choose between either
+                                 # continuous traits 
+                                 link.args  = select_continuous)
+```
+
+And run the usual simulation pipeline with `treats`:
+
+```{r, fig.height = 6, fig.width = 12}
+## Running a tree with this conditional trait
+set.seed(123)
+my_conditional_tree <- treats(stop.rule = list(max.taxa = 200),
+                              traits    = conditional_trait)
+
+op <- par(mfrow = c(1,2))
+plot(my_conditional_tree, trait = 1, main = "The conditional trait")
+plot(my_conditional_tree, trait = 2, main = "The conditioned trait")
+par(op)
+## The clumpy bit around the value 0 corresponds to the OU trait
+```
+
+Of course, you can combine linked traits with other traits like any other traits in `treats` using `make.traits` (for example: `make.traits(process = my_new_trait, add = conditional_trait)` to generate more complex traits.
+In this case the traits are generated in parallel (for speed) apart for the specified linked ones.
+
 ## Traits implemented in `treats`

### 2024-08-19 inst/gitbook/06_plotting_treats_objects.Rmd
pre-update

--- 2024-05-13 inst/gitbook/treats_manual.Rmd
@@
---- 
-title: "treats manual"
-author: "Thomas Guillerme (guillert@tcd.ie)"
-date: "`r Sys.Date()`"
-site: bookdown::bookdown_site
-output: bookdown::gitbook
-documentclass: book
-bibliography: [../references.bib]
-link-citations: yes
-github-repo: TGuillerme/treats
-description: "treats R package manual"
----
-
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
-```
-
-# `treats`: trees and traits simulations.
-
-Simultaneously simulate phylogenetic trees and traits (and the interaction between them).
-
-The core of this package is based on the [`diversitree` birth-death algorithm](https://github.com/cran/diversitree/blob/master/R/simulate-bd.R).
-
-## What is `treats`?
-
-Bib is studying the evolution of mammalian disparity (i.e. diversity of shapes) across the Cretaceous-Palaeogene mass extinction event (K-Pg - 66 million years ago).
-They observe a decrease of morphological disparity after the extinction event followed by a sharp increase of disparity.
-However Bib is also prudent and wants to test whether their observed pattern is just an artifact of the data or actually showing some effect of the K-Pg extinction on mammalian evolution.
-First they need to check whether changes in disparity can actually be discernables due to an extinction event (i.e.
-if an extinction event happens, would it be possible to detect a change in disparity)?
-Second they want to know whether these changes are due to the mass extinction event being selective or not (i.e. the mass extinction favored some traits).
-
-The [illustrative example](#kpgexample) section actually shows how do implement this example in `treats` in details.
-
-Of course, this is just an example of when to use `treats`.
-Simulating trees and traits can also be useful for:
-
- * Provide a baseline/null model to compare to an observed one.
- * Testing evolutionary hypotheses.
- * Exploring statistics properties (i.e. can the disparity metric X capture changes in Y).
- * Easily generating illustrative/simplified evolutionary history examples.
- * Teaching simplified macro-evolutionary theory (e.g. the birth-death model).
-
-Note that for approaching these points individually (or some aspects of it), there are some very good R packages that simulate birth-death trees (e.g.
-[TreeSim](https://github.com/tanja819/TreeSim)) or simulate traits (disparity) and diversity jointly (e.g.
-[RPANDA](https://github.com/cran/RPANDA) or [PETER](https://github.com/PuttickMacroevolution/PETER)).
-We strongly advise you look at these packages first as they might be more appropriate for your needs.
-
-`treats` is a highly modular and user-friendly package that contains the simulations functionality of all these packages: it allows you to simulate disparity and diversity jointly with a vast array of options that can be easily modified by users.
-For example, you can easily generate any type of process to generate a trait (BM, OU, something else, etc.) in multiple dependent or independent dimensions through `"traits"` objects.
-You can then specify how the traits should affect disparity through `"modifiers"` objects. 
-And finally you can create events (like mass extinctions) through `"events"` objects.
-These objects and how to modify them will be detailed throughout this manual.
-Finally we are putting an emphasis in the development of this package on the speed and reliability of the functions.
-
-### Modular?
-
-Because there is an infinite number of ways you might want to generate disparity and diversity (different traits, different modifiers and different events), the `treats` package is designed to make all these parts easy to code separately and integrate in the `treats` core functions.
-This allows you to simulate finely-tuned multidimensional data for your specific project!
-
-## Who is this manual for?
-
-This manual explains how `treats` works in as much details as possible and aims to give you the keys to design your own diversity and disparity simulations.
-
-The first section `Getting started` is aimed for people with a `R` beginner level:
-
- * you know what a package is - and how to install it;
- * you know what a function is - and what arguments are;
- * you have already (vaguely) heard that there are different types of objects in R - like matrices (`"matrix"`) or trees (`"phylo"`);
- * and you have already looked for documentation online.
-
-The sections after that are a bit more advanced and require the following level:
-
- * you have already created functions before;
- * you know how to subset elements of a list - i.e. you understand the `list$element` syntax;
- * you are aware that both `TRUE == 1` and `FALSE == 0` are `TRUE`;
- * you have heard that the objects in R have classes and that it sometimes matters - for example you know that R can differentiate between `as.numeric(1)` and `as.integer(1)`;
- * you owe part of your work to Stack Overflow.
-
-## Installing and running the package
-
-You can install this package easily, directly from the GitHub:
-<!-- TG: I do intend to release it on CRAN later on -->
-
-```{r, eval = FALSE}
-## Checking if devtools is already installed
-if(!require(devtools)) install.packages("devtools")
-
-## Installing the latest version directly from GitHub
-devtools::install_github("TGuillerme/treats")
-```
-
-```{r, eval = FALSE, echo = FALSE, message = FALSE}
-## Loading the package and setting up the start seed.
-library(treats)
-```
-
-## Help
-
-If you need help with the package, hopefully the following manual will be useful.
-However, if you have suggestions or comments on on what has already been developed or will be developed, please send me an email (<guillert@tcd.ie>) or if you are a GitHub user, directly create an issue on the [GitHub page](https://github.com/TGuillerme/treats).
-Doing so will not only help you but also other users as it will help improve this manual and the package!
-
-## How does `treats` work?
-
-Basically, the `treats` function takes your `traits`, `modifiers` and `events` objects to simulate disparity and diversity.
-You will find more details about how these objects (`traits`, `modifiers` and `events`) work in the rest of the tutorial but here is a graphical representation of how `treats` works:
-
-![Schematic summary of the `treats` package architecture](treats_structure.png)
-
-## The birth-death algorithm {#bdalgorithm}
-
-> If you want to get your hands dirty, you can go straight to the [getting started section](#gettingstarted), this following section just describes the process in pseudo-code.
-
-The `treats` algorithm is based on a modular version of the birth-death process algorithm.
-The birth-death model is a continuous Markov process (i.e.
-a continuous random process where future events depend only on the present, and not on past events) which is well defined mathematically and [commonly used in evolutionary biology](https://lukejharmon.github.io/pcm/chapter10_birthdeath/#section-10.2-the-birth-death-model) but also in [many other fields](https://en.wikipedia.org/wiki/Birth%E2%80%93death_process).
-
-The algorithm used in `treats` allows modularity of this process and is based on the following steps (the text in `courier font` is for the name of the process in the algorithm):
-
- 0.
-**Starting the process**: this step is non-modular and creates a random tree with one tip, one node and one branch connecting both.
-This step is used to optimise the rest of the algorithm in terms of speed and memory management.
-The node, tip and branch resulting from this step are discarded at the end of the simulations.
- 1.
-**Selecting a lineage** (`selecting`): this step selects a tip that is currently not extinct.
-In a standard birth-death process this is done randomly, however in `treats` this can be modified based on the birth-death parameters, the currently available lineages and potential trait values.
-*For example, it is possible to put a higher probability for selecting a lineage that is closely related to a lineage that recently went extinct and has a positive trait value.*
- Then go to step 2.
- 2.
-**Growing the tree** (`waiting`): this step grows the tree by a certain amount.
-It does so by adding the same amount of branch length to all the non-extinct lineages.
-In an exact birth-death process, this is done by drawing a random value from an exponential distribution with a rate of $$\text{number of living lineages} \times (\text{speciation} + \text{extinction parameters})$$.
-In `treats`, this can be modified based on the birth-death parameters, the currently available lineages and potential trait values.
-*For example, it is possible to increase branch length by the ratio of living/extinct fossils and a random number drawn from the range of current trait values.*
- If the total tree length is less than the required tree length, go to step 3. Else go to step 6.
- 3.
-**Simulating traits** [optional]\(`traits`): this step allows you to simulate a trait value for the selected lineage from step 1.
-This is typically not part of a standard birth-death process and is handled via the `traits` option in `treats` (see the [`make.traits`](#maketraits) chapter for more details).
- Then go to step 4.
- 4.
-**Speciating** (`speciating`): in this step, the selected lineage from step 1 has the option of speciating or going extinct.
-In a standard birth-death process, this happens by randomly drawing a value between 0 and 1 as a way to trigger speciation relative to the birth-death parameters:
-```
-if
-    randomly drawn number is smaller or equal to speciation/(speciation + extinction)
-then
-    do speciate
-else
-    go extinct
-```
- Again, in `treats`, this process can be modified based on the birth-death parameters, the currently available lineages and potential trait values.
-*For example, the lineage can only speciate if its trait value is positive, regardless of whether speciation or extinction have been triggered*.
- Then go to step 5.
- 5. If the number of lineages is less than required number of lineages, then go to step 1. Else go to step 6.
- 6. The simulation stops because it has reached the required amount of lineages and/or time (i.e. branch length).
-
-<!--chapter:end:index.Rmd-->
-
----
-title: "Getting start"
-author: "Thomas Guillerme (guillert@tcd.ie)"
-date: "`r Sys.Date()`"
-bibliography: [../references.bib]
-output:
-  html_document: default
-  pdf_document: default
----
-
-```{r, echo = FALSE}
-library(treats)
-set.seed(4242)
-```
-
-# Getting started {#gettingstarted}
-
-## The simplest analysis: simulating diversity only
-
-One of the simplest things to do with the `treats` package is to simulate a birth-death tree.
-For that you can use the function `treats` and specify your stopping rule.
-The stopping rule simply tells the birth-death process to stop whenever it reaches one of these three conditions:
-
- * `"max.taxa"   = n` stop when `n` taxa are generated;
- * `"max.living" = n` stop when there are `n` co-occuring taxa of the same age (i.e. "living" taxa);
- * `"max.time"   = n` stop when the simulated tree is `n` units of age old (these units are arbitrary);
-
-For example, we might want to generate a birth-death tree with 20 taxa:
-
-```{r}
-## Setting a stopping rule to reach a maximum of 20 taxa
-my_stop_rule <- list(max.taxa = 20)
-```
-
-We can now run the simulations using:
-
-```{r}
-## Running the birth-death simulation
-my_tree <- treats(stop.rule = my_stop_rule)
-```
-
-> Note that here we could have specified more than one stopping rule, for example, we might want to run a simulation and stop it if it *either* reaches 10 taxa *or* the age 2 using `stop.rule = list(max.time = 2, max.taxa = 10)`.
-The simulation will then stop when either of these conditions are met.
-
-The resulting object is a classic `"phylo"` object that you can plot like so:
-
-```{r}
-## The tree object
-my_tree
-## Plotting it
-plot(my_tree)
-```
-
-### Changing the birth-death parameters
-
-People familiar with the [birth-death models](https://lukejharmon.github.io/pcm/chapter10_birthdeath/) might have noticed that we did not specify two important things here: the speciation parameter (sometimes called "lambda" or "birth") and the extinction parameter (sometimes called "mu", "death" or "background extinction").
-By default `treats` runs a pure birth model (speciation is set to 1 and extinction to 0).
-However, you can easily change that by specifying your own birth-death parameters:
-
-```{r}
-## my birth-death parameters
-my_params <- list(speciation = 1,
-                  extinction = 1/3)
-```
-
-> You can find more information about setting up more complex birth-death parameters [in this section](#makebdparams).
-
-You can then run the same birth-death tree simulation with extinction:
-
-```{r}
-## Generating a birth-death tree with extinctions:
-my_tree <- treats(bd.params = my_params, stop.rule = my_stop_rule)
-## Visualising the new tree
-plot(my_tree)
-```
-
-## Slightly more complex: simulating disparity and diversity
-
-Chances are that you are using `treats` because you also want to simulate traits (disparity) along with your diversity (otherwise, we suggest using the [`TreeSim`](https://github.com/tanja819/TreeSim/) package that provides many more birth-death models).
-Simulating traits is not much more complicated in `treats`: you'll simply need to create a `"traits"` object using the `make.traits` function.
-These objects can have increasing complexity (see the rest of this tutorial) but we will keep it simple here.
-
-`"traits"` objects contain one or more processes which are the ways to generate the trait.
-The most common of these processes is the [Brownian Motion](https://en.wikipedia.org/wiki/Brownian_motion) model.
-This is used by default with the `make.traits` function:
-
-```{r}
-## Creating the traits object
-my_trait <- make.traits()
-```
-
-This trait object can be simply printed (to see what's in it) or plotted (to see what the process looks like in the absence of a phylogeny):
-
-```{r}
-## Which process is in here?
-my_trait
-## What does it look like?
-plot(my_trait)
-```
-
-By default, this trait is called "A".
-This is not a really good name but you'll see more about specifying trait names later on.
-If this is what the process should look like (theoretically) you can then add its `"traits"` object to our previous `treats` function to generate the tree and the traits:
-
-```{r}
-## Simulate disparity and diversity
-my_data <- treats(bd.params = my_params,
-                stop.rule = my_stop_rule,
-                traits    = my_trait)
-```
-
-Et voilà! We now have a simple disparity and diversity simulation.
-We can see what's in the results by simply printing it or plotting it:
-
-```{r}
-## What's in there
-my_data
-## Plotting the disparity and diversity
-plot(my_data)
-```
-
-You can then extract the components you need for your specific analysis like so:
-
-```{r}
-## Extracting the tree 
-the_generated_tree <- my_data$tree
-# Note that this is a "phylo" object
-class(the_generated_tree)
-
-## Extracting the data 
-the_generated_data <- my_data$data
-# Note that this is a "matrix" or "array"
-class(the_generated_data)
-```
-
-You can find much more about how to design trait objects in the [`make.traits` section](#maketraits).
-
-### `"treats"` objects
-
-`treats` will output either just a tree (class `"phylo"`) if no traits were generated or a `"treats"` object that contains both a `$tree` (`"phylo"`) and a `$data` (`"matrix"`) component.
-Note that this `"treats"` class is generalised to most outputs of the package functions.
-This allows for a smoother handeling of the objects outputs such as summarising the content of a `make.bd.params` output or visualising a trait output from `make.traits`.
-
-## Slightly more complex again: simulating linked disparity and diversity
-
-The example above is still pretty simple and easily done through a variety of `R` packages: here the trait and the tree are simulated at the same time but only the tree is simulating the trait (i.e.
-the trait value at a tip is affected by its ancestor and the branch length leading to it) but not the other way around (the trait value does not affect the tree).
-It is possible to add this aspect using `"modifiers"` objects.
-`"modifiers"` are similar to `"traits"` in that you specify what should go in there and then feed it to your simulation.
-
-`"modifiers"` affect two key steps of the birth-death process: the calculation of the waiting time (i.e. the component generating branch lengths) and the triggering of speciation or extinction events.
-These events can be modified using `condition` and `modify` functions.
-In other words, when reaching a certain condition specified by a `condition` function, the birth-death process will modify either the branch length or the speciation (or extinction) probability by applying a `modify` function.
-
-You can use the function `make.modifiers` to design a specific `"modifiers"` object.
-By default, this function generates a `"modifiers"` object that affects branch length and speciation in the following way:
-
- * branch length is a randomly drawn number from an exponential distribution with a rate equal to the current number of taxa multiplied by the sum of the speciation and extinction rates.
- * speciation is triggered if a randomly drawn number (from a (0,1) uniform distribution) is smaller than the ratio between the speciation rate and the sum of the speciation and extinction rates.
-If that random number is greater, the lineage goes extinct.
-
-Note that these are defaults for a birth-death tree and were already applied in the examples above without specifying a `"modifiers"` object:
-
-```{r}
-## Make a default modifiers object
-default_modifiers <- make.modifiers()
-## What's in it?
-default_modifiers
-```
-
-This will not do anything to our simulations compared to the previous trait and tree simulation but we can provide our `"modifiers"` object to the `treats` function:
-
-```{r}
-## Setting the simulation parameters
-extinction_02 <- list(extinction = 0.2)
-living_20     <- list(max.living = 20)
-BM_trait      <- make.traits()
-
-# Set random seed so we get the same results
-set.seed(1)
-
-## Simulate disparity and diversity
-default_data <- treats(bd.params = extinction_02,
-                     stop.rule = living_20,
-                     traits    = BM_trait,
-                     modifiers = default_modifiers)
-default_data
-```
-
-Note however, that the printing information is now updated to state that you've added a `"modifiers"` object (even though it's the same as the default).
-In this specific case the `modifiers` object is not necessary as it is doing nothing different than when using the default one (i.e. when no `modifiers` is provided).
-
-For more interesting simulations however, you can provide modifiers that actually modify the birth-death process.
-We can create one for example that makes species go extinct if their ancestor has a negative trait value.
-For that we need to create a `"modifiers"` object that modifies the `speciation` process with a specific condition and a specific modification when that condition is met.
-For a speciation to occur (and a species to not go extinct), the algorithm draws a random value _x_ between 0 and 1 and if the value is smaller than the speciation parameter divided by the speciation + extinction parameter, the lineage speciates ($x < (\lambda \div (\lambda + \mu))$), else it goes extinct.
-_x_ is a random value that simulates some stochasticity in the evolutionary simulation.
-At every step of the simulation, a different random value _x_ is drawn making sure that the resulting speciation or extinction is not predictable.
-
-However, we can create a modifier that introduces less randomness (or a different level of randomness) by forcing the lineage to go extinct every time a specific condition is met, e.g.
-if the lineage trait value is negative.
-This can be done for example by not randomly generating the value _x_ but giving it a fixed value of say 1 so that the estimation of the speciation function $x < (\lambda \div (\lambda + \mu))$ is always met.
-In other words, if the lineage's trait value is negative, make sure that _x_ is equal to 1, so that the speciation condition $x < (\lambda \div\ (\lambda + \mu))$ is never met.
-
-First we need to create a `condition` function to trigger this (non) speciation event.
-We can do that by specifying our `condition` function (here the `going.extinct` function) to apply our modification.
-For that we can use the `parent.traits` utility function that is optimised for accessing traits in the birth-death process (but you can of course write your own).
-This function takes the `trait.values` and `parent.lineage` arguments, two arguments that you can leave named as they are to facilitate `treats`' understanding of what you want to assess:
-
-```{r}
-## Triggering a modification only if the ancestor trait is negative
-negative.ancestor <- function(trait.values, lineage) {
-    return(all(parent.traits(trait.values, lineage) < 0))
-}
-```
-
-> Note that we use the function `all` here to evaluate all traits: i.e. if the data has more than one trait we trigger the modification only if all the trait values are negative.
-
-Then we create the modification function.
-This function must take the argument `x` and, in our case, returns the same value no matter what: `1` so that speciation never happens when the `negative.ancestor` function is triggered.
-
-```{r}
-## Always go extinct
-going.extinct <- function(x) return(1)
-```
-
-This `going.extinct` function is to be contrasted with the normal modifier for selecting the value `x` which is `runif(1)` (drawing a random value between 0 and 1).
-We can then provide these two functions (the condition `negative.ancestor` and how to modify the speciation event when this condition is met `going.extinct`).
-If you are an advances `treats` user, you can design your own `speciation` function but if you just want to use a normal `speciation` function, you can use the default one from `treats` called `speciation`.
-
-```{r}
-## Making a modifier for species to go extinct if
-## their ancestor's trait value is (or are) negative
-negatives_extinct <- make.modifiers(
-            ## If the following condition is met...
-            condition = negative.ancestor, # Does the lineage have an ancestor with negative trait values?
-            ## Then apply the following modifier...
-            modify = going.extinct, # The species goes extinct
-            ## To the the speciation process.
-            speciation = speciation) # Here the speciation() process is default
-
-## What's in it?
-negatives_extinct
-```
-
-Note that the `make.modifiers` function tests whether the input is compatible with `treats` by default so unless you have an error message, your `modifiers` will work!
-We can now simulate our tree and traits with our modifier: species will go extinct if their ancestor has a negative trait value:
-
-```{r}
-set.seed(1)
-## Simulate disparity and diversity
-modified_data <- treats(bd.params = extinction_02,
-                        stop.rule = living_20,
-                        traits    = BM_trait,
-                        modifiers = negatives_extinct)
-modified_data
-```
-
-We can now compare the two trees and their trait values.
-Note that we've used the same starting seed for both trees so the only thing differing between them is the `"modifier"` object which leads to very different trees!
-
-```{r, fig.height = 6, fig.width = 12}
-par(mfrow = c(1,2))
-plot(default_data, main = "default results")
-plot(modified_data, main = "results with the modifier")
-```
-
-You can find much more about how to design modifiers in the [`make.modifiers` section](#makemodifiers).
-
-## An illustrative example: mammals disparity through the K-Pg extinction {#kpgexample}
-
-Bib is studying the evolution of mammalian disparity (i.e. diversity of shapes) across the Cretaceous-Palaeogene mass extinction event (K-Pg - 66 million years ago).
-
-They use a dataset from @beckancient2014 that is a discrete morphological space containing the ordination of the around 400 morphological characters for 50 species and estimated for their 49 descendants into an ordinated shapespace of 99 elements and 97 dimensions.
-
-```{r}
-library(dispRity)
-## Loading the mammalian morphological example data
-data(BeckLee_mat99)
-
-## This is what the dataset looks like
-head(BeckLee_mat99)[1:5, 1:5]
-```
-
-And a tree showing the relation between these species through time:
-
-```{r}
-## The phylogenetic tree
-data(BeckLee_tree)
-
-## And what the tree looks like
-plot(BeckLee_tree, cex = 0.8)
-axisPhylo()
-abline(v = BeckLee_tree$root.time - 66, col = "red")
-```
-
-They measure the disparity as the sum of variances as a proxy for changes in the traitspace size (@moms) with some bootstrapping to create some confidence intervals around that curve.
-
-```{r}
-## Making continuous time series
-time_series <- chrono.subsets(data = BeckLee_mat99,
-                              tree = BeckLee_tree,
-                              method = "continuous", model = "proximity",
-                              time = seq(from = 101, to = 41, by = -5))
-## Calculating disparity
-change_in_size <- dispRity(boot.matrix(time_series), metric = c(sum, variances))
-plot(change_in_size)
-abline(v = 8, col = "red")
-```
-
-> For more info about `chrono.subsets` and the `dispRity` package, you can have a look at the [`dispRity` manual](http://tguillerme.github.io/dispRity.html).
-
-Looking at this curve, Bib wants to know whether disparity increased as a response to the K-Pg mass extinction or not.
-To do this, they need to check whether changes in disparity can actually be detected due to an extinction event (e.g. if an extinction event happens, does that change disparity curves?).
-As a subsidiary question, they want to know whether the pattern is more likely due to a selective or random extinction event.
-
-### Simulating trees and traits using the observed parameters
-
-The first step is to simulate the observed tree.
-We can do that by using crude parameters from the observed tree:
-
-```{r}
-## Extracting the crude parameters
-(est_params <- crude.bd.est(BeckLee_tree, method = "count"))
-## These parameters are the observed speciation and extinction rate per million years.
-
-## Setting the stopping rule (stop after reaching 50 taxa)
-stop_rule <- list(max.taxa = 50)
-
-## Simulating just a birth-death tree with these parameters to check
-set.seed(1)
-test_tree <- treats(bd.params  = est_params,
-                    stop.rule  = stop_rule,
-                    null.error = 100)
-plot(ladderize(test_tree), cex = 0.8)
-axisPhylo()
-```
-
-> Note that because of the high extinction to speciation ratio, the tree simulations can often fail before reaching the stop rule (i.e.
-all lineages go extinct before reaching 50 taxa).
-To avoid getting stuck in there, we use the option `null.error = 100` to rerun the simulations up to 100 times before obtaining a tree that did not went fully extinct.
-Also note that using such parameters make the simulations run slowlier but ensure the requested results are always achieved.
-
-We now have a way to simulate a topology close to the observed one. We can simulate a list of topologies by replicating this function X amounts of time to get a distribution of trees:
-
-```{r, eval = FALSE, message = FALSE}
-## Generating 50 trees
-tree_distribution <- treats(bd.params  = est_params,
-                            stop.rule  = stop_rule,
-                            null.error = 100,
-                            replicates = 50)
-```
-
-Of course, here our objective is to also simulate some trait values to measure disparity under some simulated scenario.
-We can do that be adding a `"traits"` object to the `treats` function.
-Here we will use a default [Brownian Motion](https://en.wikipedia.org/wiki/Brownian_motion) to simulate our trait.
-
-```{r}
-## Creating a trait in 97 dimensions.
-my_traits <- make.traits(process = BM.process, n = ncol(BeckLee_mat99))
-```
-
-> Note that there are many more options on how to simulate your trait such as using correlation, different processes, etc... See details in the [make traits section](#maketraits).
-
-```{r, message = FALSE}
-## Simulate the tree and traits
-sim_data <- treats(traits     = my_traits,
-                   bd.params  = est_params,
-                   stop.rule  = stop_rule,
-                   null.error = 100)
-```
-
-We can visualise the resulting tree using the `plot` function and visualising the different traits using the `trait` option in `plot`:
-
-```{r}
-## Plotting the results (first trait)
-plot(sim_data, trait = 1)
-```
-
-Of course, here we want to simulate a distribution of trees and traits:
-
-```{r, message = FALSE}
-## Simulate the tree and traits
-sim_data <- treats(traits     = my_traits,
-                   bd.params  = est_params,
-                   stop.rule  = stop_rule,
-                   null.error = 100,
-                   replicates = 50)
-```
-
-Et voila! We now have a distribution of 50 trees and 50 datasets from which we can calculate disparity.
-We can directly pass the results to the `dispRity` package pipeline using the `dispRitreats` function:
-
-```{r, message = FALSE, warning=FALSE}
-## Calculate the dispRity for all the simulations
-simulated_disparity <- dispRitreats(sim_data,
-                                    method = "continuous",
-                                    model  = "proximity",
-                                    time   = 10,
-                                    metric = c(sum, variances),
-                                    scale.trees = TRUE)
-par(mfrow = c(1,2))
-plot(simulated_disparity, main = "simulated disparity")
-plot(change_in_size, main = "observed disparity")
-```
-
-We can see that both simulated and observed disparity curves are different.
-From there one could already reasonably propose that the observe disparity does not entirely follows our parametrised scenario (constant speciation and extinction and BM trait evolution).
-
-However, we can complexify our scenario to introduce the effect of mass extinction event.
-We can simulate two alternative scenarios:
- * one where we create a random extinction event after having simulated half of the species (25) where we remove 75% of species;
- * and a second one where we remove all the species with negative values.
-This allows us to simulate two contrasted scenarios to see whether using our observed dataset it would be possible to detect changes in disparity due to mass extinctions.
-
-To do so, we need to create two mass extinction `"event"` objects:
-
-```{r}
-## Creating a random mass extinction
-random_extinction <- make.events(target = "taxa",
-                                 condition = taxa.condition(25, condition = `>=`),
-                                 modification = random.extinction(0.75))
-## Creating an extinction that removes species with negative trait values
-negative_extinction <- make.events(target = "taxa",
-                                   condition = taxa.condition(25, condition = `>=`),
-                                   modification = trait.extinction(x = 0, condition = `>=`))
-```
-
-We can then feed these two extinction events to our previous simulations
-
-```{r, message = FALSE}
-## Simulate the tree and traits with a random extinction event
-sim_rand_extinction <- treats(
-                   traits     = my_traits,
-                   bd.params  = est_params,
-                   stop.rule  = stop_rule,
-                   events     = random_extinction,
-                   null.error = 100,
-                   replicates = 50)
-
-## Simulate the tree and traits with a random extinction event
-sim_trait_extinction <- treats(
-                   traits     = my_traits,
-                   bd.params  = est_params,
-                   stop.rule  = stop_rule,
-                   events     = negative_extinction,
-                   null.error = 100,
-                   replicates = 50)
-
-## Visualising the difference between both scenarios
-par(mfrow = c(1,2))
-plot(sim_rand_extinction[[1]], main = "Random extinction")
-plot(sim_trait_extinction[[1]], main = "Negative extinction")
-```
-
-We can then measure the disparity from these scenarios and compare them to our observed ones
-
-```{r}
-## Simulate the tree and traits with a random extinction event
-sim_rand_extinction <- treats(
-                   traits     = my_traits,
-                   bd.params  = est_params,
-                   stop.rule  = stop_rule,
-                   events     = random_extinction,
-                   null.error = 100,
-                   replicates = 50)
-
-## Simulate the tree and traits with a random extinction event
-sim_trait_extinction <- treats(
-                   traits     = my_traits,
-                   bd.params  = est_params,
-                   stop.rule  = stop_rule,
-                   events     = negative_extinction,
-                   null.error = 100,
-                   replicates = 50)
-
-## Visualising the difference between both scenarios
-par(mfrow = c(1,2))
-plot(sim_rand_extinction[[1]], main = "Random extinction")
-plot(sim_trait_extinction[[1]], main = "Negative extinction")
-```
-
-Interestingly, with all our parameters used here and our specific disparity metric we cannot detect the effect of a mass extinction.
-
-<!--chapter:end:01_getting_started.Rmd-->
-
----
-title: "`make.traits`"
-author: "Thomas Guillerme (guillert@tcd.ie)"
-date: "`r Sys.Date()`"
-bibliography: [../references.bib]
-output:
-  html_document: default
-  pdf_document: default
----
-
-# Simulating traits {#maketraits}
-
-```{r, echo = FALSE}
-library(treats)
-set.seed(123)
-```
-
-In `treats`, traits are simulating by providing a `traits` object to the `traits` argument.
-This object is generated using `make.traits`.
-In `treats` a trait designate any coherent character of any number of dimensions which may or may not be independent of other traits.
-For example, it can be a uni-dimensional variable uncorrelated to other traits or a multi-dimensional one with different amounts of correlation.
-
-![`make.traits`: The main inputs are (1) the `process`, one or more functions to describe the trait(s) as a function of previous state (`x0`) and the branch length (`edge.length`), (2) `n` the number of dimensions of the trait and (3) the starting value `start`.](make.traits.png)
-
-### Quick overview
-
-function | arguments | input | what does it do
----------|-----------|-------|----------------
-`make.traits` | process* | a function | simulates the trait (see below)
-`make.traits` | n | a numeric value | the number of dimensions
-`make.traits` | start | one or more numeric values | the starting values for each dimensions
-`make.traits` | process.args | ... | anything to be passed to `process` (see below)
-`make.traits` | trait.names | a character string | the name(s) of the trait(s)
-`make.traits` | add | a `treats` object | another trait generated by `make.traits`
-`make.traits` | update | a `treats` object | another trait generated by `make.traits`
-`make.traits` | test | logical | whether to test the validity of the trait
-`make.traits` | background | a `treats` object | another trait generated by `make.traits` (to run in the background)
----------|-----------|-------|----------------
-`process` | x0 | a numeric value | the value of the previous trait
-`process` | edge.length | a numeric value | the branch length value
-`process` | ... | ... | any other argument to be handled by the process
-
-\* non-optional arguments
-
-## The process (`process`)
-
-The function `make.traits` allows you to design the process of a trait or a set of traits.
-Here, the process of a trait designates the rules needed to generate the trait through time while simulating a phylogeny.
-This process can depend on the previous state in the tree (i.e. the trait of the ancestor) and the branch length to the descendant.
-One classic example is the [Brownian motion process (or Weiner process)](https://en.wikipedia.org/wiki/Brownian_motion).
-Note that it _can_ depend on both the ancestor and the branch length but does _not necessarily need to_, i.e.
-the process can be based only on the previous state or only on branch length or on neither.
-
-### The syntax (how to code a process?)
-
-Trait processes in `treats` are functions that must always take the following arguments by default.
-
- * `x0`: the previous trait value(s)
- * `edge.length`: the branch length
- * `...`: a placeholder for any extra arguments
-
-For example, the following function would be a valid process that always generate the _true_ trait value: 42!.
-In this example, the process is not dependent on either the previous state (`x0`) or the branch length (`edge.length`).
-
-```{r}
-## A valid (but useless?) process
-valid.process <- function(x0 = 0, edge.length = 1, ...) {
-    return(42)
-}
-```
-
-> Note that in this function definition the arguments `x0` and `edge.length` have a default value of `0` and `1` respectively.
-In practice, these arguments are effectively set to the correct values in the `treats` internal function (i.e.
-whatever `x0` and `edge.length` are at that specific time of the process) but providing a default can help speed up the algorithms (specifically all the internal checks).
-
-On the other hand, the following process (a unidimensional Brownian motion) is incorrect (it's missing `edge.length` and `...`):
-
-```{r}
-## A wrongly formatted process
-invalid.process <- function(x0 = 0) {
-    return(rnorm(1, mean = x0))
-}
-```
-
-This will not work in `make.traits` (see below).
-
-### Using a `"process"` in `treats`
-
-You can design your own process as a function, as long as it has a valid syntax.
-Alternatively, the `treats` package has several inbuilt processes, namely a multidimensional Brownian motion (`BM.process`) or a multidimensional Ornstein-Uhlenbeck process (`OU.process`).
-You can find the list of implemented processes by looking at the `?trait.process` manual page in `R`.
-
-Once a process is chosen, you can feed it into the `make.traits` function:
-
-```{r}
-## Creating a trait object
-my_trait_object <- make.traits(process = BM.process)
-```
-
-This creates `"treats"` `"traits"` objects that you can print and visualise using the `plot` function:
-
-```{r}
-## The class of the object
-class(my_trait_object)
-## What's in it?
-my_trait_object
-## What does the process looks like
-plot(my_trait_object)
-```
-
-Note that you can see the multiple options for plotting the trait process by looking at `?plot.treats` manual.
-Furthermore, you can look at what's actually in the object using this specific syntax (this applies to every object handled by the `treats` package):
-
-```{r}
-## What's actually in that object?
-print.treats(my_trait_object, all = TRUE)
-```
-
-As traits can get more and more complex, the automatic printing of its summary allows for a easier display of what's in the `"traits"` object.
-
-Note that it is possible to make `"traits"` objects with multiple traits and multiple processes (that can be the same or different for each trait):
-
-```{r}
-## Four traits: two BM, one OU and one normal non process
-four_traits <- make.traits(process = c(BM.process,
-                                       BM.process,
-                                       OU.process,
-                                       no.process))
-four_traits
-```
-
-You can visualise them individually using the `trait` argument in `plot.treats`:
-
-```{r}
-## Plot options (4 plots in one window)
-par(mfrow = c(2,2))
-plot(four_traits, trait = 1)
-plot(four_traits, trait = 2)
-plot(four_traits, trait = 3)
-plot(four_traits, trait = 4)
-par(mfrow = c(1,1))
-```
-
-## The number of traits `n` and the starting values `start`
-
-Two further important arguments are `n` the number of traits per process and `start` the starting values for all traits.
-By default they are set to `n = 1` and `start = 0`.
-This means that `make.traits` will assume that your processes are always unidimensional by default and that they always start with the value `0`.
-It is possible however, to change these values.
-
-For example you can use the following to create a three dimensional Brownian motion with each dimension starting with the value `1`:
-
-```{r}
-## Multidimensional Brownian motion
-make.traits(BM.process, n = 3, start = 1)
-```
-
-Or the following with each dimensions starting with different values (respectively `1`, `2` and `3`):
-
-```{r}
-## Multidimensional Brownian motion
-make.traits(BM.process, n = 3, start = c(1,2,3))
-```
-
-Note that the number of traits are applied for each process by default.
-However, you can apply different number of traits for different processes:
-
-```{r}
-## two 3D processes (BM and OU)
-make.traits(c(BM.process, OU.process), n = 3)
-## one 1D processes (BM) and one 4D process (OU)
-make.traits(c(BM.process, OU.process), n = c(1, 4))
-```
-
-And starting values are distributed for all the traits or for the traits one by one:
-
-```{r}
-## two 3D processes (BM and OU) starting with 1
-make.traits(c(BM.process, OU.process), n = 3, start = 1)
-## two 3D processes (BM and OU) starting with values 1 to 6
-make.traits(c(BM.process, OU.process), n = 3, start = 1:6)
-## two 3D processes (BM and OU) with the two first ones starting
-## with 1 and the 4 other ones with the default (0)
-make.traits(c(BM.process, OU.process), n = 3, start = c(1,1))
-```
-
-### What is a trait in `treats`?
-
-Because it would be impossible to accommodate all definitions of a trait in `treats` we chose an arbitrary one: a trait is whatever you define as a trait!
-A trait can be uni-dimensional as the measurement of a feature of an organism (e.g.
-leaf surface, femur length, etc.) but can also be a multi-dimensional description of a feature, for example in 3D geometric morphometric context, a trait could be defined as "position of landmark X" (which will be a trait with three dimensions, _x_, _y_ and _z_) or in ecology, the location of a plant can be expressed as latitude and longitude coordinates.
-In `treats`, the **process** corresponds to this trait definition (e.g.
-a process can be of n-dimensions and represents one organisms feature) and the **traits** represents the number of dimensions in total.
-So in the examples above, this is how the following traits are interpreted by `treats`:
-
-```{r, eval = FALSE}
-## Three traits with one process:
-make.traits(BM.process, n = 3, start = c(1,2,3))
-## Six traits with two processes:
-make.traits(c(BM.process, OU.process), n = 3)
-## Five traits with two processes
-make.traits(c(BM.process, OU.process), n = c(1, 4))
-```
-
-## Multidimensional traits and correlations
-
-As shown in the example above, a trait can be made multidimensional by using the `n` argument specifically passed to a `process`.
-However, in this case, traits are simulated independently of each other.
-It is possible to include correlation by designing a multidimensional trait `process` directly generating the correlation between within the dimensions of the process.
-Alternatively, the inbuilt `BM.process` and `OU.process` can intake a correlation matrix to simulate traits independtly but with a correlation coefficient (see [this illustration with an event](#EG_change_correlation)).
-
-## Extra argument for the processes with `process.args`
-
-You can also feed extra arguments to your process(es) functions. 
-For example, the inbuilt process `no.process` (that is just a number generator not based on the previous value `x0` or the branch length) can take a specific random number generator as a function:
-
-```{r}
-## no process trait using the normal distribution (default)
-make.traits(no.process, process.args = list(fun = rnorm))
-## no process trait using the uniform distribution
-## bounded between 1 and 100
-make.traits(no.process, process.args = list(fun = runif, min = 1, max = 100))
-```
-
-You can also add multiple extra arguments for multiple processes giving them as a list.
-
-```{r}
-## Two traits with no process:one normal and one uniform (1,100)
-make.traits(process = c(no.process, no.process),
-            process.args = list(list(fun = rnorm),
-                                list(fun = runif, min = 1, max = 100)))
-```
-
-If one process does not need extra argument you must still give it an extra `NULL` process argument:
-
-```{r}
-## Three traits with no process:
-## one default, one lognormal and one uniform (1,100)
-make.traits(process      = c(no.process, no.process, no.process),
-            process.args = list(## Extra arguments for the first process (none)
-                                list(NULL),
-                                ## Extra arguments for the second process
-                                list(fun = rlnorm),
-                                ## Extra arguments for the third process
-                                list(fun = runif, min = 1, max = 100)))
-```
-
-## Naming the traits with `trait.names`
-
-As traits become more and more complex, it can be useful to give clearer names to each process.
-This is easily done using the `trait.names` argument that attributes one name per process:
-
-```{r}
-## A simple trait with a proper name
-simple_trait <- make.traits(trait.names = "1D Brownian Motion")
-simple_trait
-```
-
-This becomes more useful if we use the complex example above:
-
-```{r}
-## Three named traits with no process:
-## one default, one lognormal and one uniform (1,100)
-make.traits(process      = c(no.process, no.process, no.process),
-            process.args = list(## Extra arguments for the first process (none)
-                                list(NULL),
-                                ## Extra arguments for the second process
-                                list(fun = rlnorm),
-                                ## Extra arguments for the third process
-                                list(fun = runif, min = 1, max = 100)),
-            ## Naming each trait
-            trait.names  = c("Normal", "LogNormal", "Uniform(1,100)"))
-```
-
-## Combining multiple traits with `add`
-
-You can also add traits to already existing trait objects using the simple `add` option.
-This option just takes a `"treats"` `"traits"` object and the additional process(es) will be added to it. 
-For example:
-
-```{r}
-## Creating a simple default Brownian motion
-one_process <- make.traits(trait.names = "BM")
-
-## Creating a new trait (a 3D OU.process)
-## and adding the previous one
-two_processes <- make.traits(OU.process, n = 3, add = one_process,
-                             trait.names = "3D OU")
-
-## Only one process
-one_process
-
-## The two processes
-two_processes
-```
-
-## Using a background trait
-
-`traits` objects also allow a background trait to be used when traits are simulated ([step 3 here](#bdalgorithm)).
-This allows traits to be simulated for _all_ tips whenever a trait is generated for one tip.
-This can be useful for keeping track of trait values along the simulation (_cf_ just at bifurcating nodes).
-The `background` argument takes any output from the `make.traits` function in a nested way:
-
-```{r}
-## Generating a default BM trait:
-BM_trait <- make.traits()
-## Generating an OU trait with a background BM trait
-my_trait <- make.traits(process = OU.process, background = BM_trait) 
-```
-
-Note that technically you can nest as many background traits as you want (e.g. `make.traits(background = make.traits(background = make.traits(...)))` is valid).
-However, you should always make sure that the background trait has the same dimensions as the main trait.
-When using a trait with background, your tree will have internal singleton nodes (i.e. nodes linking to one ancestor and only one descendant).
-You can remove these nodes using the [`drop.things` function](#dropthings).
-
-```{r}
-set.seed(1)
-## Generating a pure birth tree with the background trait
-tree_bkg <- treats(stop.rule = list(max.taxa = 20),
-                 traits = my_trait)
-## This tree has many internal singleton nodes
-plot(tree_bkg)
-```
-
-## Saving trait values at different time steps
-
-You can also simulate a tree by generating traits at specific time steps with the `save.steps` option in `treats`.
-This will apply the `traits` object to all lineages currently alive at the required time steps.
-These time steps can be either regular by providing a single numeric value; e.g.
-`save.steps = 0.1` will take a snapshot of the trait values every 0.1 units of time, or specific, by providing a specific set of values; e.g.
-`save.steps = c(1, 1.2, 3)` will take a snapshot of the trait values at the required time steps.
-
-```{r}
-set.seed(123)
-## Generating a birth-death tree with a BM trait and saving steps at specific times
-tree_steps <- treats(stop.rule  = list(max.time = 3),
-                   bd.params  = list(speciation = 1, extinction = 0.1),
-                   traits     = make.traits(),
-                   save.steps = c(1/3, 1, 2))
-## This also creates internal singleton nodes
-plot(tree_steps)
-abline(v = 3 - c(1/3, 1, 2), lwd = 0.5, col = "grey")
-```
-
-## Traits implemented in `treats`
-
-If you don't want to design your own trait process, you can use one of the following trait processes that are currently implemented in `treats`.
-You can find more information about their many options using their specific manuals in R or the generic `?trait.process`:
-
- * `BM.process`: this is the well known [Brownian motion process](https://en.wikipedia.org/wiki/Brownian_motion).
-
-```{r, echo = FALSE}
-plot(make.traits(process = BM.process), main = "The Brownian motion")
-```
-
- * `OU.process`: this is the equally famous [Ornstein–Uhlenbeck process](https://en.wikipedia.org/wiki/Ornstein%E2%80%93Uhlenbeck_process).
-
-```{r, echo = FALSE}
-plot(make.traits(process = OU.process), main = "The Ornstein–Uhlenbeck process")
-```
-
- * `no.process`: this process has...
-no process.
-In other words, this is a non time dependent process; the simulated value does not depends on the ancestors' value nor the branch length.
-It's basically a place holder for a random sampling function like `rnorm` (default), `runif`, `rlnorm`, etc.
-
-```{r, echo = FALSE}
-plot(make.traits(process = no.process, process.args = list(fun = runif)), main = "No process using runif\n(yes, it \"looks\" like an OU!)")
-```
-
- * `multi.peak.process`: this process is a modified version of the `OU.process` that can take multiple local long-term mean values.
-The default OU process has one long-term mean towards which the values are drawn with the alpha parameter (the elastic band).
-The single long-term mean is usually 0.
-However, with this `multi.peak.process` we can set multiple values towards which values can be attracted with the same alpha parameter.
-
-```{r, echo = FALSE}
-plot(make.traits(multi.peak.process, process.args = list(peaks = c(1, 5, 10))), main = "A multi-peak (OU) process\nwith peaks (optima) at 1, 5 and 10")
-```
-
- * `repulsion.process`: this is a modified version of the `BM.process` where instead of accumulating gradually through time, new trait values are more likely to be different to the traits of their ancestor following a `repulsion` parameter.
-
-```{r, echo = FALSE}
-plot(make.traits(repulsion.process, process.args = list(repulsion = 5)), main = "A repulsion process\nwith a repulsion parameter of 5")
-```
-
- * `discrete.process`: this one generates discrete trait values based on a transition matrix (you can use the utility function `transition.matrix` to generate one).
-
-```{r, echo = FALSE}
-## Generating a stepwise transition matrix for 3 states (with an overal random transition rate)
-stepwise_matrix <- transition.matrix(type = "stepwise", states = 3)
-## Generatin and plotting the the trait
-plot(make.traits(discrete.process, process.args = list(transitions = stepwise_matrix)))
-```
-
-Of course here the plotting of a discrete trait is not that informative...
-
-## Testing the traits with `test`
-
-**This bit is still in development.**
-We highly suggest leaving `test = TRUE` so that `make.traits` returns an error if a process or its additional arguments (`process.args`) are not formatted correctly.
-`make.traits` will send an error if the trait cannot be directly passed to `treats`.
-However, in some specific cases (again, probably mainly for development and debugging) it could be useful to skip the tests using `test = FALSE`.
-
-## Mapping traits on a tree: `map.traits`
-
-Although `make.traits` is mainly designed for `treats` simulations, i.e.
-simulating both the trees AND the traits at the same time, it is possible to use an already simulated or estimated tree and generating one or more traits on it using the `map.traits` function.
-This function takes the trait input as a `"traits"` object (from `make.traits`) and a tree topology with branch length.
-Note that you can also use `"traits"` objects with multiple or multi-dimensional traits as well as multiple phylogenies (`"multiPhylo"`).
-
-```{r}
-## Simulating a random tree with branch length
-my_tree <- rtree(20)
-
-## Creating three different traits objects:
-## A Brownian Motion
-bm_process <- make.traits(process = BM.process)
-## An Ornstein-Uhlenbeck process
-ou_process <- make.traits(process = OU.process)
-## No process (just randomly drawing values from a normal distribution)
-no_process <- make.traits(process = no.process)
-
-## Mapping the three traits on the phylogeny
-bm_traits <- map.traits(bm_process, my_tree)
-ou_traits <- map.traits(ou_process, my_tree)
-no_traits <- map.traits(no_process, my_tree)
-
-## Plotting the topology and the different traits
-par(mfrow = c(2,2))
-plot(my_tree, main = "Base topology")
-plot(bm_traits, main = "Mapped BM")
-plot(ou_traits, main = "Mapped OU")
-plot(no_traits, main = "Mapped normal trait")
-```
-
-## Templates for making your very own process
-
-As detailed above, any process of your own design will work as long as it is a function that takes at least the arguments `x0` and `edge.length`.
-You can be imaginative and creative when designing your own process but here are two detailed example functions for a unidimensional Brownian Motion and Ornstein-Uhlenbeck process that you can use for a start (or not).
-Remember it is good practice for `treats` processes to set all the arguments with default values (just in case).
-
-> Note that the functions below are not equal to the already implemented `BM.process` and `OU.process` but an easier to edit version that you can use as a template:
-
-### A simple Brownian Motion process template
-
-```{r}
-## A simple Brownian motion process
-my.BM.process <- function(x0 = 0, edge.length = 1, sd = 1, ...) {
-    ## Drawing a random number from a normal distribution
-    ## with x0 as the and a given standard deviation
-    ## and depending on branch (edge) length
-    result <- rnorm(n = 1, mean = x0, sd = sqrt(sd^2 * edge.length))
-
-    ## Return the number
-    return(result)
-}
-```
-
-### A simple Ornstein-Uhlenbeck process template
-
-```{r}
-## A simple Ornstein-Uhlenbeck motion process
-my.OU.process <- function(x0 = 0, edge.length = 1, var = 1, alpha = 1, ...) {
-    ## Calculate the mean based on alpha
-    mean <- x0 * exp(-alpha)
-    ## Calculate the standard deviation based on alpha and the variance
-    sd <- sqrt(var/(2 * alpha) * (1 - exp(-2 * alpha)))
-    ## Draw a random number from a normal distribution
-    ## using this mean and standard deviation
-    ## and depending on branch (edge) length
-    result <- rnorm(n = 1, mean = mean, sd = sqrt(sd^2 * edge.length))
-
-    ## Return the number
-    return(result)
-}
-```
-
-<!--chapter:end:02_make.traits.Rmd-->
-
----
-title: "`make.modifiers`"
-author: "Thomas Guillerme (guillert@tcd.ie)"
-date: "`r Sys.Date()`"
-bibliography: [../references.bib]
-output:
-  html_document: default
-  pdf_document: default
----
-
-# Modifying the birth-death process {#makemodifiers}
-
-```{r, echo = FALSE}
-library(treats)
-set.seed(123)
-```
-
-`"modifiers"` have a similar structure than `"traits"` where you can design an object with increasing complexity, starting with the simplest modifiers that doesn't modify anything (using the default arguments):
-
-```{r}
-## Making a default modifier (no modification)
-my_default_modifiers <- make.modifiers()
-my_default_modifiers
-```
-
-Similarly to `"traits"` objects, `"modifiers"` are also printed by default using `print.treats`.
-You can see details about what's actually in the object using `print.treats(my_default_modifiers, all = TRUE)`.
-However, contrary to `"traits"`, you cannot plot `"modifiers"`.
-
-![`make.modifiers`: The three main inputs to make modifiers are functions to modify the `branch.length` (waiting) process controlling the "growth" of the tree; the `selection` (selecting) process controlling the selection of the tip on which to apply speciation or extinction; and the `speciation` (speciating) process controlling whether the select tip goes extinct or speciates.
-The allowed arguments are detailed [here](#allowarguments).](make.modifiers.png)
-
-### Quick overview
-
-function | arguments | input | what does it do
----------|-----------|-------|----------------
-`make.modifiers` | branch.length | a function | generating branch length (see below)
-`make.modifiers` | selection | a function | selecting a lineage (see below)
-`make.modifiers` | speciation | a function | choosing between speciation and extinction (see below)
-`make.modifiers` | condition | a function | a function that affects the condition of when to trigger the functions above (see below)
-`make.modifiers` | modify | a function | a function that modifies the value of the trigger value generated when the `condition` is met (see below)
-`make.modifiers` | add | a `treats` object | another modifier generated by `make.modifiers`
-`make.modifiers` | update | a `treats` object | another modifier generated by `make.modifiers`
-`make.modifiers` | test | logical | whether to test the validity of the modifier
----------|-----------|-------|----------------
-`branch.length`, `selection`, `speciation`, `condition` | bd.params | a `treats` object generated by `make.bd.params` | usually handled internally (contains the either the birth-death values or their distribution)
-... | lineage | a named "lineage" list | usually handled internally (contains the current state of the simulations)
-... | trait.values | a matrix | usually handled internally (contains the current generated traits of the simulations)
-
-## The default modifier (how the process is working)
-
-The modifiers modify the core of the birth-death process as implemented in `treats`.
-By default, the birth-death process in `treats` uses this modifier:
-
-```{r}
-## What is actually in the default modifier?
-print(make.modifiers(), all = TRUE)
-```
-
-This contains a lot of information, much of it is actually not useful for using the modular aspects of `treats` at a high user level, i.e.
-unless you want to code very specific things, you won't need most of the information.
-The essentials are these three functions in the elements named `"waiting$fun"`, `"selecting$fun"` and `"speciating$fun"`:
-
- 1.
-`"waiting$fun"` is the branch length function (defined in detail below) which returns a randomly drawn number from an exponential distribution with the rate of the number of taxa multiplied by the speciation and extinction rate: $n \times (\lambda + \mu)$ (here is uses the extra function `sum` for internal modularity reasons).
-**This function is responsible for the growth of the length/age of the tree**
-
-```{r, eval = FALSE}
-rexp(1, sum(lineage$n * (bd.params$speciation + bd.params$extinction)))
-```
- With `lineage$n` being the number of lineages and `bd.params$speciation` and `bd.params$extinction` the speciation and extinction parameters (these specific terms are defined in detail below).
-
- 2.
-`"selecting$fun"` is the selection function (defined in detail below) which, after the waiting time defined above returns a randomly selected lineage (a tip) among the existing ones.
-**This function is responsible for the branch selection**.
-
-```{r, eval = FALSE}
-sample(lineage$n, 1)
-```
-
- 3.
-`"speciating$fun"` is the speciation function (defined in detail below) which randomly draws a number (`runif(1)`) and depending on the speciation and extinction parameters makes the selected lineage (from 2.) speciate (`TRUE`) or go extinct (`FALSE`).
-**This function is responsible for the speciation or extinction of species**.
-
-```{r, eval = FALSE}
-runif(1) < (bd.params$speciation/(bd.params$speciation + bd.params$extinction))
-```
-
-We will have a look at all these aspects below in more detail which should make things clearer (but you can always refer to the default here for information).
-
-## The branch length function (`branch.length`)
-
-The first argument in `"modifiers"` is the branch length function (`branch.length`) this is the function that will be executed in `treats` to generate branch lengths.
-Note that in the `treats` algorithm, branch length is not generated _directly_ but as the result of the waiting time.
-In other words, the `branch.length` function just affects waiting time for all taxa present at any time in the simulation.
-These taxa can then either go extinct (stopping the "growth" of its branch length) or survive (continuing the "growth").
-
-By default, branch length (or waiting/growth) is a randomly drawn number from an exponential distribution with the rate of the number of taxa multiplied by the speciation and extinction rate: $n \times (\lambda + \mu)$ (where $n$ is the number of taxa currently present in the simulation, $\lambda$ and $\mu$ are respectively the speciation and extinction rates).
-This default function is simply called `branch.length` in `treats` and can be used as a modifier as follows:
-
-```{r}
-## Specifying the default modifier
-default_modifiers <- make.modifiers(branch.length = branch.length)
-
-## Setting some parameters for generating trees
-bd_params <- list(extinction = 0)
-stop_rule <- list(max.living = 20)
-
-## Generating a tree with the default branch length parameter
-set.seed(0)
-default_tree <- treats(bd.params = bd_params,
-                     stop.rule = stop_rule,
-                     modifiers = default_modifiers)
-```
-
-Of course, the point of the modularity here is that you can provide your own function for generating branch length.
-For example, we might be interested in what our tree would look like if we use a simple constant branch length generation (instead of randomly drawing it from an exponential distribution).
-We can do so by declaring our own `branch.length` function and adding it to a `"modifiers"` object.
-
-```{r}
-## A constant branch length generator
-## (note that the output must be numeric, not integer)
-constant.brlen <- function() {
-    return(as.numeric(1))
-}
-
-## Creating the modifiers object
-constant_modifier <- make.modifiers(branch.length = constant.brlen)
-
-## Generating a new tree with this modifier
-set.seed(0)
-modified_tree <- treats(bd.params = bd_params,
-                      stop.rule = stop_rule,
-                      modifiers = constant_modifier)
-```
-
-And we can visualise the difference between both resulting trees:
-
-```{r}
-par(mfrow = c(1,2))
-plot(default_tree,  main = "Default modifier")
-plot(modified_tree, main = "Constant branch length\nmodifier")
-par(mfrow = c(1,1))
-```
-
-It is of course possible to use more complex branch length modifiers that take different conditions and specific modification rather than simply always outputing a value of one.
-
-### The modifier arguments {#allowarguments}
-
-You can create a function for `branch.length`, `selection` and `speciation` that involve any of the following arguments:
-
-  * `bd.params`: a named list containing `"numeric"` values that contains the birth-death parameters (at least `"speciation"` and `"extinction"`);
-  * `lineage`: a named list containing the lineage data (see below).
-  * `trait.values`: a `"matrix"` containing `"numeric"` values with the trait names as column names and the lineage ID as row numbers (you can use it with the function `parent.traits` to access the trait of the previous node for example).
-  * `modify.fun`: a `"list"` of named `"function"` (usually passed through `condition` and `modify`).
-
-The `lineage` list contains the following elements (missing elements are allowed):
-
- * `lineage$parents`: an `"integer"` vector: the list of parent lineages;
- * `lineage$livings`: an `"integer"` vector: the list of lineages still not extinct;
- * `lineage$drawn`: a single `"integer"`: the ID of the selected lineage; this must be a number in `1:lineage$n`;
- * `lineage$current`: a single `"integer"`: the selected lineage (is equal to `lineage$livings[lineage$drawn]`);
- * `lineage$n`: a single `"integer"`: the current number of non extinct lineage (is equal to `length(lineage$livings))`;
- * `lineage$split`: a `"logical"` vector: the list of splits for each lineage (`TRUE`), the number of total tips is equal to `sum(!lineage$split)`.
-
-In general, unless you know what you're doing, you can ignore most arguments for specific modifiers since they are handled automatically within the `treats` function.
-Therefore any argument can be left undeclared or missing and is always handled internally.
-For example, if you did not declare `lineage$n` as a function argument but are using `lineage$n` in the function, `lineage$n` will be detected and treated as a current argument automatically as set accordingly within the birth-death process (e.g.
-`lineage$n` will be set to the current number of taxa every iteration of the process).
-
-For example, we can create a function that increases branch length proportional to the number of species "alive" at each time of the simulation in a discrete way, i.e.
-for discrete numbers of taxa, the branch length increases by jumps (ten fold) every five taxa:
-
-```{r}
-## A more complex binned.branch.length function
-increasing.brlen <- function(bd.params, lineage) {
-
-    ## Setting the cumulated birth and death
-    birth_death <- bd.params$speciation + bd.params$extinction
-
-    ## Returning branch lengths depending on different number of taxa
-    if(lineage$n <= 5) {
-        return(1    * rexp(1, sum(5 * birth_death)))
-    }
-    if(lineage$n <= 10) {
-        return(10   * rexp(1, sum(10 * birth_death)))   
-    }
-    if(lineage$n <= 15) {
-        return(100  * rexp(1, sum(15 * birth_death)))      
-    }
-    if(lineage$n <= 20) {
-        return(1000 * rexp(1, sum(20 * birth_death)))
-    } else {
-        return(1000 * rexp(1, sum(lineage$n * birth_death)))
-    }
-}
-```
-
-We can then create it as a `"modifiers"` object and run a new simulation:
-
-```{r}
-## Creating a modifiers
-increasing_modifier <- make.modifiers(branch.length = increasing.brlen)
-
-## Generating a new tree with this modifier
-set.seed(0)
-increasing_tree <- treats(bd.params = bd_params,
-                        stop.rule = stop_rule,
-                        modifiers = increasing_modifier)
-```
-
-And we can visualise the difference between the resulting trees:
-
-```{r}
-par(mfrow = c(1,3))
-plot(default_tree,    main = "Default modifier")
-plot(modified_tree,   main = "Constant branch length\nmodifier")
-plot(increasing_tree, main = "Increasing branch length\nmodifier (binned)")
-par(mfrow = c(1,1))
-```
-
-## The selection function (`selection`)
-
-The `selection` function is used in the birth-death process to know which lineage to select when running a speciation (or extinction!) event.
-By default, this function randomly selects one taxon that is currently not extinct (using: `sample(1:lineage$n, 1))`.
-Similarly to `branch.length`, it is possible to modify this part of the birth-death process.
-For example, we could simply select the last created lineage to create a "ladder" or most asymmetric tree:
-
-```{r}
-## Our function to always select the last taxon
-## (making sure it returns an integer)
-select.last <- function(lineage) {
-    return(as.integer(lineage$n))
-}
-```
-
-> Note that here the function can only take the [allowed arguments as described above](#allowarguments) (here `lineage$n`: the number of current living taxa).
-
-We can then create a `"modifiers"` object the same way as before this time using the `selection` argument:
-
-```{r}
-## A modifier for selection
-ladderised_modifier <- make.modifiers(selection = select.last)
-## Generating a new tree with this modifier
-set.seed(0)
-ladderised_tree <- treats(bd.params = bd_params,
-                        stop.rule = stop_rule,
-                        modifiers = ladderised_modifier)
-## Displaying the results
-par(mfrow = c(1,2))
-plot(default_tree,    main = "Default modifier")
-plot(ladderised_tree, main = "Ladderising modifier")
-par(mfrow = c(1,1))
-```
-
-Again, it is of course possible to make the modifier more complex and in combination with other elements of the tree.
-For example, we can create a `"treats"` object that also generates a BM trait and add to it this object a `selection` modifier that only selects tips with positive trait values (only species with positive trait values will speciate).
-
-```{r}
-## Our function that only select taxa with positive trait values
-select.positive <- function(trait.values, lineage) {
-
-    ## Selecting the taxa names with positive values for the first trait
-    positives <- as.integer(rownames(trait.values)[which(trait.values[, 1] >= 0)])
-
-    ## Combine the descendants of the current lineages (lineage$parents)
-    ## with the species that have speciated (seq_along(lineages$split))
-    ## to have a table of pairs of parents/splits
-    parents_split_table <- cbind(lineage$parents, seq_along(lineage$split))
-    ## Select the current taxa that descend from a node with a positive value
-    positive_living <- parents_split_table[which(lineage$parents %in% positives), 2]
-
-    ## Select one tip randomly in the ones with descendants with positive values
-    return(sample(which(lineage$livings %in% positive_living), 1))
-}
-
-## Creating the modifier
-positive_skew <- make.modifiers(selection = select.positive)
-
-## Creating a (default) trait object
-BM_trait <- make.traits()
-
-## Simulate a tree and trait with no modifier
-set.seed(1)
-default_treats <- treats(bd.params = bd_params,
-                     stop.rule = stop_rule,
-                     traits = BM_trait)
-
-## Simulate a tree and trait with the modifier
-set.seed(1)
-skewed_trait_treats <- treats(bd.params = bd_params,
-                          stop.rule = stop_rule,
-                          traits = BM_trait,
-                          modifiers = positive_skew)
-
-## Plotting the differences in trees and traits
-par(mfrow = c(1, 2))
-plot(default_treats, main = "Default trait and tree")
-plot(skewed_trait_treats, main = "Skewed trait and tree")
-```
-
-## The speciation function (`speciation`)
-
-The third function that can be used to modify the birth-death process is the `speciation` function.
-This one is used during the birth-death process to decide whether a lineage speciates (creating a node and two new lineages) or goes extinct (creating a tip).
-
-> Note that the `speciation` function only affects tips or nodes before the simulation reaches the `stop.rule`.
-The then surviving lineages are all automatically transformed into tips.
-
-By default, the `speciation` function is trigger a speciation even if a number randomly drawn from a uniform distribution is lower than the ratio between the speciation and the speciation and extinction parameter.
-If the randomly drawn number is higher, the lineage goes extinct.
-
-```{r, eval = FALSE}
-## The speciation in pseudo-code:
-runif(1) < speciation/ (speciation + extinction)
-```
-
-Creating `"modifiers"` with a `speciation` function works the same way as for `branch.length` and `selection` but the function that will be used needs to output a logical value (see table [below](#summarymodifiers)).
-Once the function is created simply input your function for speciation in the modifier and run the `treats` function with that modifier:
-
-```{r, eval = FALSE}
-## Speciating or going extinct randomly
-## (regardless of the extinction parameter)
-random.extinct  <- function() {
-    return(sample(c(TRUE, FALSE), 1))
-}
-
-## Creating the modifiers object
-random_extinction <- make.modifiers(speciation = random.extinct)
-
-## Generating a new tree with this modifier
-set.seed(2)
-modified_tree <- treats(bd.params = bd_params,
-                      stop.rule = stop_rule,
-                      modifiers = random_extinction)
-
-par(mfrow = c(1,2))
-plot(default_tree,  main = "Default modifier")
-plot(modified_tree, main = "Random extinction\nmodifier")
-par(mfrow = c(1,1))
-```
-
-> Note how loads of lineages go extinct even if the extinction parameter is set to 0!
-
-And again, we can make some more advanced modifiers: for example, one where a tip always goes extinct if their ancestor has a negative trait value.
-Here we will also introduce the utility function `parent.trait` that automatically selects the trait values of the parent of the current lineage.
-
-```{r}
-## A modifier for removing tips with negative values
-modify.trait <- function(trait.values, lineage) {
-    if(parent.traits(trait.values, lineage) < 0) {
-        ## Go extinct!
-        return(FALSE)
-    } else {
-        ## Speciate!
-        return(TRUE)
-    }
-}
-
-## Creating the modifier
-modified_trait <- make.modifiers(speciation = modify.trait)
-
-## Simulate a tree and trait with the modifier
-set.seed(1)
-modified_trait_treats <- treats(bd.params = bd_params,
-                          stop.rule = stop_rule,
-                          traits    = BM_trait,
-                          modifiers = modified_trait)
-
-## Plotting the differences in trees and traits
-par(mfrow = c(1, 2))
-plot(default_treats, main = "Default trait and tree")
-plot(modified_trait_treats, main = "Biased trait and tree")
-```
-
-## Summary of the inputs and outputs for the `branch.length`, `selection` and `speciation` modifiers {#summarymodifiers}
-
-modifier name   | accepted input (arguments) | required output (class) |
-----------------|----------------------------|-------------------------|
-`branch.length` | `bd.params`, `lineage`, `trait.values` | `"numeric"` |
-`selection`     | `bd.params`, `lineage`, `trait.values` | `"integer"` |
-`speciation`    | `bd.params`, `lineage`, `trait.values` | `"logical"` |
-
-## The condition and modify functions (`condition` and `modify`)
-
-In the examples above, we have seen how to specify modifications to the birth-death process (via `branch.length`, `selection` and `speciation`), however, these modifications are not dynamic.
-In other words, throughout the process, the modifications remain constant (even if they are conditional).
-It is however, possible to code the `"modifiers"` so that they can be affected by `"events"` objects (see next chapter [on `events`](#makeevents)).
-
-To do so, you can formally declare conditions (`condition`) and modifications (`modify`) as internal functions that can then be modified by an `"events"` object.
-`condition` and `modify` are hard coded in the `branch.length` function that they concern, i.e. they are variables (functions) within the function.
-
-For example in the `speciation` part of a modifier, the default is to trigger an event from a uniform distribution and then check if that value is smaller than ($speciation/(speciation + extinction)$):
-
-```{r}
-## The default speciation algorithm
-speciation <- function(bd.params) {
-    ## Randomly trigger an event
-    trigger_event <- runif(1)
-
-    ## Speciate?
-    return(trigger_event < (bd.params$speciation/
-                              (bd.params$speciation + bd.params$extinction)))
-}
-```
-
-It is possible with some specific conditions to modify this trigger by providing a `condition` and `modify` function to the speciation function.
-That is, if the `condition` is met, apply the `modify` function to the algorithm.
-For example here we can edit the default speciation function to have a modification (`modify = double.the.trigger`: doubling the trigger value) happening half the time (`condition = half.the.time`):
-
-```{r}
-## A conditional function that triggers half the time
-half.the.time <- function() return(sample(c(TRUE, FALSE), 1))
-## A modification that doubles the value to trigger the event
-double.the.trigger <- function(x) return(x*2)
-## A conditional modifier
-make.modifiers(speciation = speciation,
-               condition  = half.the.time, 
-               modify = double.the.trigger)
-```
-Effectively, this will internally modify the speciation function as follows:
-
-```{r, eval = FALSE}
-## The default speciation algorithm
-speciation <- function(bd.params) {
-    ## Randomly trigger an event
-    trigger_event <- runif(1)
-
-    ## Modify the triggering
-    if(half.the.time()) { ## This running the half.the.time function drawing TRUE or FALSE randomly
-        trigger_event <- double.the.trigger(trigger_event) ## This will double the random value trigger_event
-    }
-
-    ## Speciate?
-    return(trigger_event < (bd.params$speciation/
-                              (bd.params$speciation + bd.params$extinction)))
-}
-```
-
-These `condition` and `modify` functions can be applied to all the `modifiers` elements (`selection`, `branch.length` and `speciation`).
-However, they are typically used in `events` that will modify the `modifiers` ([see the `events` section](#makeevents))!
-
-<!--chapter:end:03_make.modifiers.Rmd-->
-
----
-title: "`make.events`"
-author: "Thomas Guillerme (guillert@tcd.ie)"
-date: "`r Sys.Date()`"
-bibliography: [../references.bib]
-output:
-  html_document: default
-  pdf_document: default
----
-
-```{r, echo = FALSE}
-library(rgl)
-knitr::knit_hooks$set(webgl = hook_webgl)
-```
-
-```{r, echo = FALSE}
-library(treats)
-set.seed(123)
-```
-
-# Adding events to simulations {#makeevents}
-
-One other major feature of the `treats` package is that it allows simulations to run with specified events that occur during the simulation.
-These are typically events that can drastically change the course of the simulation.
-For example, you might want simulate a mass extinction at some specific point in time.
-However, these can also be more subtle, like the internal change of parameter values when reaching a specific trait value.
-These are all handled by `treats` with the `events` object that you can create using `make.events` and has the same overall logic as `make.traits` and `make.modifiers`.
-
-`events` require three main arguments:
- * the `target` which designates what the extinction should affect (e.g. the taxa, the speciation rate, etc...);
- * the `condition` which designates when to trigger the event;
- * the `modification` which designates what to modify when the event is triggered.
-
-There are several more arguments that can be passed to `make.events` but they are discussed later on.
-First let's focus on these three main arguments:
-
-![`make.events`: events are encoded using three main arguments: the `target`, specifying _what_ the event is modifying, the `condition` specifying _when_ to trigger the event, and `modification`, specifying _how_ the event modifies the simulations.](make.events.png)
-
-### Quick overview
-
-  condition,
-  modification,
-  add,
-  test = TRUE,
-  event.name,
-  replications = 0,
-  additional.args
-)
-
-function | arguments | input | what does it do
----------|-----------|-------|----------------
-`make.events` | target* | any of the following: "taxa", "bd.params", "traits", "modifiers" or "founding" | what aspect of the simulation should the event target
-`make.events` | condition* | a function | a function that affects the condition of when to trigger the event (see below)
-`make.events` | modification* | a function | a function that applies the modification to the target if the condition is triggered (see below)
-`make.events` | add | a `treats` object | another event generated by `make.events`
-`make.events` | test | logical | whether to test the validity of the event
-`make.events` | event.name | a character string | the name of the event
-`make.events` | replications | a number | if possible how many times is the event allowing to be repeated
-`make.events` | additional.args | ... | any additional arguments to be passed to the event
-
-Here is a list of implemented `conditions` (see `?events.conditions`):
-
-function | arguments | input | what does it do
----------|-----------|-------|----------------
-`age.condition`, `taxa.condition`, `trait.condition` | x | the value to check the target against | e.g. time = x, taxa = x or trait = x (see below)
-... | condition | a relational operator | e.g. `==`, `>`, `!=`, etc. for asking for example `time == x` 
-... | ... | any additional arguments to be passed to the event | |
-
-Here is a list of implemented `modifications` (see `?events.modifications`):
-
- function | arguments | input | what does it do
----------|-----------|-------|----------------
-`random.extinction` | x | a numeric value | the proportion of species to make extinct
-`trait.extinction` | x | a numeric value | the trait value when to trigger the extinction
-`trait.extinction` | condition | a relational operator | the relation to the trait value (e.g. `x = 1` and `condition =  >` makes species that have a trait value > 1 go extinct)
-`bd.params.update` | ... | any argument to be passed to `make.bd.params` | |
-`traits.update` | ... | any argument to be passed to `make.traits` | |
-`modifiers.update` | ... | any argument to be passed to `make.modifiers` | |
-
-\* non-optional arguments
-
-## Target
-
-The target of the event is what the event is going to modify in the birth-death algorithm.
-You can only have one target per event (along with one condition and one modification) but you can create events that contain multiple events (i.e. multiple triplets of target/condition/modification).
-The targets that are currently available are:
-
- * `"taxa"` to modify anything linked to the [`lineage` list](#allowarguments). e.g. making half of the living taxa go extinct.
- * `"bd.params"` to modify anything linked to the [`bd.params` object](#makebdparams).
-For example you might want to change the distribution of one of the parameter after some `conditions`.
-This is typically done by updating the object using the argument `update` from `make.bd.params`.
- * `"traits"` to modify anything linked to the [`traits` object](#maketraits).
-For example you might want to change the trait process after some `conditions`.
-This is typically done by updating the object using the argument `update` from `make.traits`.
- * `"modifiers"` to modify anything linked to the [`modifiers` object](#makebdmodifiers).
-For example you might want to change the speciation rule after some `conditions`.
-This is typically done by updating the object using the argument `update` from `make.modifiers`.
- * `"founding"` this target is a bit more specialised and allows you run a nested `treats` object in the simulation. It is covered in a [specific section below](#founding).
-
-We will see some examples of these targets in the `conditions` and `modifications` described below.
-
-## Conditions
-
-`condition` is a function that returns a logical value. When a specific condition is met, it should return `TRUE` and trigger the event, else it should return `FALSE`.
-
-Currently there are three conditions functions implemented in `treats` but you can easily come up with your own version of them.
-
-All `condition` functions in `treats` take at least two arguments: `x` for the variable of interest (e.g.
-time, number of taxa, trait value, etc.) and `condition`, the relational operator to evaluate.
-A relational operator is the proper (fancy) term designating all the comparisons you're regularly using in `R` like `==` (is equal?) `<` (is smaller?) `>=` (is bigger or equal?).
-You can get the full list in the R base manual (using `?Comparison` or any of the relational operator in a function form).
-For more details on how functions _really_ work in R see the [Advanced `R` book](https://adv-r.hadley.nz/functions.html)!
-
- * `age.condition` will trigger the `event` once a certain time is reached in the simulations and is the simplest/most basic condition with no arguments other than the time required.
-For example ``age.condition(4, condition = `==`)`` will trigger the `event` once the simulations reaches 4 time units.
-Easy.
- * `taxa.condition` will trigger the `event` once a certain number of taxa are reached.
-This can be considered including or excluding fossil species.
-For example ``taxa.condition(42, condition = `>=`, living = TRUE)`` will trigger the `event` if there are at least 42 living taxa.
- * `trait.condition` will trigger the `event` once a certain trait value is reached.
-This function allows to say which trait(s) to target (by default, the first one using `trait = 1`), what value of the trait to target (by default `what = max`) and whether to use and absolute trait value or not (`absolute = TRUE`).
-For example ``trait.condition(1/3, condition = `>`, trait = 1, what = sd)`` will trigger the condition after the standard deviation of the first trait reaches 1/3.
-
-## Modifications
-
-After defining the `event`, `target` and `condition`, you also need to specify what it should modify.
-These are functions that should modify a specific aspect of the target.
-Depending on the target you can modify the following:
-
- * if the target is `"taxa"` you can modify the internal [`lineage` list](#allowarguments) by removing living species using `random.extinction` or `trait.extinction`.
- * if the target is `"taxa"` you can modify the lineage tracker by removing living species using `random.extinction` or `trait.extinction`.
- * if the target is `"bd.params"` you can modify the [`bd.params` object](#makebdparams) using `bd.params.update`.
- * if the target is `"traits"` you can modify the [`traits` object](#maketraits) using `traits.update`.
- * if the target is `"modifiers"` you can modify the [`modifiers` object](#makemodifiers) using `modifiers.update`.
- * (for the `"founding"` target [see below](#founding))
-
-The most straightforward example is for `modifications` on `bd.params`, `traits` or `modifiers` objects because they use the same syntax as for their generic `make.X` function.
-For example, for `make.traits`, you can update a `trait` using the `update` argument as follows:
-
-```{r}
-## A BM trait in two dimensions
-(BM_2D <- make.traits(n = 2, process = BM.process))
-## Updating the 2D BM into a 2D OU
-(OU_2D <- make.traits(update = BM_2D, process = OU.process))
-```
-
-So the function `update.X` applies the update to the object `X` when the event happens.
-The following `event` updates the `bd.params` object by setting the extinction parameter to 1/3 when reaching a 10 species:
-
-```{r, eval = FALSE}
-make.events(target = "bd.params",
-            condition = taxa.condition(10, condition = `>=`),
-            modification = bd.params.update(extinction = 1/3))
-```
-
-> Note that by default events are triggered only once across the whole simulation, so although the example above states that the condition is reaching at least 10 taxa, it will not trigger every time it reaches more than 10 taxa, only the first time.
-You can change the number of times the events can be triggered using the argument `replications` (by default it's set to `replications = 0` for triggering the event only once).
-
-## Examples
-
-Here are some examples illustrating how to generate events.
-For a simple example, we can create a extinction event that will remove 80% of species after reaching time 4:
-
-```{r}
-## 80% mass extinction at time 4
-mass_extinction <- make.events(
-                      target = "taxa",
-                      condition = age.condition(4),
-                      modification = random.extinction(0.8))
-
-## Simulation parameters
-stop.rule <- list(max.time = 5)
-bd.params <- list(extinction = 0, speciation = 1)
-
-## Running the simulations
-set.seed(123)
-results <- treats(bd.params = bd.params,
-                stop.rule = stop.rule,
-                events = mass_extinction)
-## Plotting the results
-plot(results, show.tip.label = FALSE)
-axisPhylo()
-```
-
-Or for a slightly more complex example, we can can change the trait process from a BM to an OU when the trait values reaches an upper 95% quantile value above 2:
-
-```{r}
-## The 95% upper quantile value of a distribution
-upper.95 <- function(x) {
-    return(quantile(x, prob = 0.95))
-} 
-## Create an event to change the trait process
-change_process <- make.events(
-                  target = "traits",
-                  ## condition is triggered if(upper.95(x) > 3)
-                  condition = trait.condition(3, condition = `>`, what = upper.95),
-                  modification = traits.update(process = OU.process))
-
-## Set the simulation parameters
-bd.params <- list(extinction = 0, speciation = 1)
-stop.rule <- list(max.time = 6)
-traits    <- make.traits()
-
-## Run the simulations
-set.seed(1)
-no_change <- treats(bd.params = bd.params,
-                  stop.rule = stop.rule,
-                  traits = traits)
-set.seed(1)
-process_change <- treats(bd.params = bd.params,
-                       stop.rule = stop.rule,
-                       traits = traits,
-                       events = change_process)
-## Plot the results
-par(mfrow = c(1,2))
-plot(no_change, ylim = c(-7, 7))
-plot(process_change, ylim = c(-7, 7))
-par(mfrow = c(1,1))
-```
-
-## Founding events {#founding}
-
-Founding events are a specific `target` for `events` that allows you to simulate a birth-death process within the current one!
-It allows you to simulate a specific `treats` process using the `treats` function with its own `traits`, `modifiers`, `bd.params` and `events` (and yes, that's including `events` that have their own `founding` event).
-This is basically the ultimate nested boss of modularity!
-
-The founding event will run an internal `treats` process (i.e. simulating a tree and, optionally, some data) resulting in a founding sub-tree.
-It will then branch this sub-tree onto the rest of the simulation that continued normally in the mean time.
-You can specify the founding event using the inbuilt `founding.event` modification event.
-This function takes the exact same arguments as `treats` to simulate the sub-tree with its own parameters.
-Additionally, we will use the `additional.args` argument from `make.events` to specify a prefix for the founding tree tips (to make them easier to distinguish).
-
-```{r}
-## Set up parameters
-stop.rule <- list(max.time = 4)
-bd.params <- make.bd.params(speciation = 1, extinction = 0.3)
-
-## Events that generate a new process (founding tree - with no extinction)
-founding_event <- make.events(
-                  target = "founding",
-                  condition = taxa.condition(10),
-                  modification = founding.event(
-                                    bd.params = make.bd.params(speciation = 2,
-                                                               extinction = 0)),
-                  additional.args = list(prefix = "founding_"))
-
-## Simulations
-set.seed(11)
-founding_tree <- treats(bd.params = bd.params,
-                      stop.rule = stop.rule,
-                      events = founding_event)
-plot(founding_tree, cex = 0.4)
-```
-
-Note that the nestedness here is potentially endless, for example, you can pass an `event` argument to the `founding.event` that will generate another founding event, etc.
-
-<!--chapter:end:04_make.events.Rmd-->
-
----
-title: "`other functionalities`"
-author: "Thomas Guillerme (guillert@tcd.ie)"
-date: "`r Sys.Date()`"
-bibliography: [../references.bib]
-output:
-  html_document: default
-  pdf_document: default
----
-
-```{r, echo = FALSE}
-library(treats)
-library(rgl)
-knitr::knit_hooks$set(webgl = hook_webgl)
-```
-
-# Other functionalities {#others}
-
-## `make.bd.params` {#makebdparams}
-
-In most examples above, birth-death parameters were set as fixed values, i.e.
-a speciation rate of $\lambda$ and an extinction rate of $\mu$ throughout the simulations (with maybe some events modifying these rates).
-However, it is possible to set these rates as specific or changing distributions. 
-You can do this using the `make.bd.params` function to provide either a `vector` of values:
-
-```{r}
-## An example where the speciation is randomly sampled among three values
-my_bd_params <- make.bd.params(speciation = c(1/3,42))
-
-## Building a tree using this set of parameters
-set.seed(123)
-plot(treats(stop.rule = list(max.taxa = 50), bd.params = my_bd_params))
-## Note the regions in the tree with short branches
-## (that's the speciation being 1/3 while the others are speciation = 42) 
-```
-
-Or to directly provide a function from which to sample:
-
-```{r}
-## Another example where speciation is drawn from the interval (0, 1)
-make.bd.params(speciation = runif)
-```
-
-In this example, the `"bd.params"` object passed to the `treats` function will allow the birth-death process to sample the speciation parameter each time it is called (e.g.
-during the speciation/extinction step, the branch length step, etc.).
-If using a function, you can fine tune the arguments to be passed to that function using the `speciation.args` or the `extinction.args` arguments (as a named list matching the function's arguments):
-
-```{r}
-## Speciation is drawn from the interval (0.5, 1.5)
-make.bd.params(speciation = runif,
-               speciation.args = list(min = 0.5, max = 1.5))
-```
-
-When using distributions for both the speciation and extinction parameters, you can run into the (usually) undesired problem of having an extinction rate that is higher than your speciation rate and thus all the taxa in your tree dying out.
-You can avoid this problem by using the `joint` distribution argument. 
-This will ensure that the sampling of the extinction rate is always lower or equal to the speciation rate.
-
-```{r}
-## Joint speciation and extinction sampled from uniform distribution 
-## with speciation always >= to extinction
-make.bd.params(speciation = runif, extinction = runif, joint = TRUE)
-```
-
-Finally to avoid negative sampling values, you can use the argument `absolute` that will make all the sampled values positive.
-This argument is set to `TRUE` by default so you shouldn't have to worry about it most of the time unless you specifically need negative sampled values for your parameters.
-
-```{r}
-## Making the speciation sampling always positive
-make.bd.params(speciation = rnorm, absolute = TRUE)
-```
-
-The two other possible arguments for this function are `test` and `update` that work the same way as [`make.traits`](#maketraits) or [`make.modifiers`](#makemodifiers)
-
-If you are a visual person and your `bd.params` objects are getting a bit too complicated to remember, you can always quickly plot them. 
-The function will sample from the `bd.params` object and show the results:
-
-```{r}
-par(mfrow = c(2,3))
-plot(make.bd.params(), main = "Default birth-death parameters")
-plot(make.bd.params(speciation = runif,
-               speciation.args = list(min = 0.5, max = 1.5)),
-     main = "Uniform speciation between 0.5 and 1.5\n(no extinction)")
-plot(make.bd.params(speciation = runif,
-                    speciation.args = list(min = 0.5, max = 1.5)),
-     main = "Uniform speciation between 0.5 and 1.5\n(no extinction)")
-plot(make.bd.params(speciation = runif, extinction = runif, joint = TRUE),
-     main = "Joint uniform speciation and extinction")
-plot(make.bd.params(speciation = rnorm, extinction = runif,
-                    joint = FALSE, abs = FALSE),
-     main = "Disjoint normal speciation and uniform extinction")
-par(mfrow = c(1,1))
-```
-
-## `drop.things` {#dropthings}
-
-You can use the function `drop.things` to drop specific elements of the tree and data at the same time by providing the argument `what = "fossils"` for tips that went extinct, or `what = "livings"` for tips that where alive at the end of the simulation or `what = "singles"` to drop internal singleton nodes.
-Alternatively you can use the function aliases `drop.fossils`, `drop.livings` or `drop.singles` for the exact same results:
-
-```{r}
-## A random tree with fossils and traits and internal nodes every 0.5 times
-set.seed(3)
-my_data <- treats(stop.rule = list(max.taxa = 20),
-                bd.params = list(speciation = 1, extinction = 1/3),
-                traits    = make.traits(), save.steps = 0.5)
-
-## A tree with 20 tips and 54 nodes
-my_data$tree
-## And a dataset with 74 rows
-dim(my_data$data)
-
-## Removing the fossil species
-drop.things(my_data, what = "fossils")$tree
-dim(drop.fossils(my_data)$data)
-
-## Removing the living species
-drop.things(my_data, what = "livings")$tree
-dim(drop.livings(my_data)$data)
-
-## Removing the internal nodes
-drop.things(my_data, what = "singles")$tree
-dim(drop.singles(my_data)$data)
-
-## Removing the internal nodes AND the fossils
-drop.singles(drop.fossils(my_data))
-```
-
-## `"treats"` internal utilities
-
-The package also provides utilities for internal functions, namely for designing `modifiers` or `events` more easily.
-These functions don't do anything useful on their own but are optimised to be used internally in `treats`.
-For all these functions, you can look at the internal manual for an example (i.e. using `?<function_name>`)
-
-So far the package has the following internals:
-
-Function  | What it does | Where can it be used
-----------|--------------|---------------------
-`parent.traits` | selects the trait values of the current lineage's parents (i.e. direct ancestor) | in `make.modifiers`
-`taxa.condition` | provides a trigger for an `event` dependent on the number of taxa | in `make.events`
-`age.condition` | provides a trigger for an `event` dependent on time | in `make.events`
-`trait.condition` | provides a trigger for an `event` dependent on trait values | in `make.events`
-
-<!--chapter:end:05_other_functionalities.Rmd-->
-
 ---
@@
 
-<!--chapter:end:06_plotting_dads_objects.Rmd-->
+## Using `plot.treats` to plot non-treats objects
 
----
-title: "treats manual"
-author: "Thomas Guillerme (guillert@tcd.ie)"
-date: "`r Sys.Date()`"
-# bibliography: [../References.bib, ../packages.bib]
-output:
-  html_document: default
-  pdf_document: default
----
+If you like the plotting style of `treats` trees and data, you can always use the `make.traits` function to plot a tree and some traits as a `"treats"` object.
 
-# Specific examples
+### Discrete plots
 
-```{r, echo = FALSE}
-library(treats)
-library(rgl)
-knitr::knit_hooks$set(webgl = hook_webgl)
-options(rgl.useNULL = TRUE) # Suppress the separate window.
-```
-
-In this section I will illustrate a series of examples of specific more complex scenarios.
-They can be all run independently and used as a basis for your own specific scenarios.
-
-Here is a table summarising which functionalities are used in which example:
-
-Functionality showcase | Example
---------------|---------
-`bd.params` | [reducing speciation rate](#EGred_spec), [generating a subtree with no extinction](#EG_founding_purebirth)
-`make.events` | [mass extinction](#EGrandom_ext), [negative trait extinction](#EGneg_ext), [background extinction](#EGbg_ext), [reducing speciation rate](#EGred_spec), [changing trait process](#EG_change_trait), [changing trait correlation](#EG_change_correlation), [changing modifiers](#EG_change_modif), [change branch length](#EG_modify_brlen), [generating a subtree with no extinction](#EG_founding_purebirth), [generating a subtree with a different process](EG_founding_traits)
-`make.traits` | [negative trait extinction](#EGneg_ext), [changing trait process](#EG_change_trait), [changing trait correlation](#EG_change_correlation), [generating a subtree with a different process](EG_founding_traits)
-`make.modifiers` | [changing modifiers](#EG_change_modif), [change branch length](#EG_modify_brlen)
-
-And more specifically for the events:
-
-Functionality showcase | Example
---------------|----------
-`age.condition` | [mass extinction](#EGrandom_ext), [negative trait extinction](#EGneg_ext), [reducing speciation rate](#EGred_spec), [changing trait process](#EG_change_trait), [changing modifiers](#EG_change_modif)
-`random.extinction` | [mass extinction](#EGrandom_ext)
-`trait.extinction` | [negative trait extinction](#EGneg_ext)
-`taxa.condition` | [background extinction](#EGbg_ext), [change branch length](#EG_modify_brlen)
-`bd.params.update` | [background extinction](#EGbg_ext), [reducing speciation rate](#EGred_spec)
-`traits.update` | [changing trait process](#EG_change_trait), [changing trait correlation](#EG_change_correlation)
-`trait.condition` | [changing trait correlation](#EG_change_correlation), [generating a subtree with a different process](EG_founding_traits)
-`modifiers.update` | [changing modifiers](#EG_change_modif), [change branch length](#EG_modify_brlen)
-`parent.traits` | [changing modifiers](#EG_change_modif)
-`founding.event` | [generating a subtree with no extinction](#EG_founding_purebirth), [generating a subtree with a different process](EG_founding_traits)
-
-<!-- TODO add a recursive event example where discrete self rate is changed by half everytime you reach 10 living species -->
-
-## Random mass extinction after some time {#EGrandom_ext}
-
-For this scenario, we want to generate a pure birth tree (no traits and no extinction) where 80% of the species go extinct two thirds of the way into the scenario.
-
-For that we first need to first set our simulation parameters: we will be running the simulation for 6 time units and with a speciation rate of 1 (and extinction of 0 - default).
-
-```{r}
-## Setting the parameters
-stop_time_6 <- list(max.time = 6)
-speciation_1 <- make.bd.params(speciation = 1)
-```
-
-We will then create a event that triggers when reaching half of the simulation time (using `age.condition(2.5)`).
-This event will target `"taxa"`, i.e. the number of species, and randomly remove 80% of them (using `random.extinction(0.8)`):
-
-```{r}
-## 80% mass extinction at time 4
-mass.extinction <- make.events(
-                    condition = age.condition(4),
-                    target = "taxa",
-                    modification = random.extinction(0.8))
-```
-
-Once these parameters are defined, we can run the simulations and plot the results:
-
-```{r}
-## Running the simulations
-set.seed(1)
-results <- treats(stop.rule = stop_time_6,
-                bd.params = speciation_1,
-                events    = mass.extinction)
-## Plotting the results
-plot(results, show.tip.label = FALSE)
-axisPhylo()
-```
-
-## Species with negative trait values go extinct after a certain time {#EGneg_ext}
-
-For this scenario, we want to generate a pure birth tree with a one dimensional Brownian Motion trait for 5 time units.
-We then want species with a negative trait value go extinct after 4 time units.
-
-First we need to set up the simulation parameters:
- * The stopping rule (5 time units)
- * The birth-death parameters (speciation rate of 1)
-
-```{r}
-## Simulation parameters
-stop_time_5 <- list(max.time = 5)
-speciation_1 <- make.bd.params(speciation = 1)
-```
-
-Then set up our trait which is a one dimensional Brownian Motion
-
-```{r}
-## Trait
-simple_bm_trait <- make.traits(n = 1, process = BM.process)
-```
-
-And our extinction event which triggers after reaching time 4 (`age.condition(4)`), targets the `"taxa"` and modifies the extinction for species with traits lower than 0.
-
-```{r}
-## Extinction of any tips with trait < 1 at time 4
-trait.extinction <- make.events(
-                      target = "taxa",
-                      condition = age.condition(4),
-                      modification = trait.extinction(x = 0,
-                                                      condition = `<`))
-```
-
-Once these parameters are defined, we can run the simulations and plot the results:
-
-```{r}
-## Running the simulations
-set.seed(7)
-results <- treats(stop.rule = stop_time_5,
-                bd.params = speciation_1,
-                traits    = simple_bm_trait,
-                events    = trait.extinction)
-## Plotting the results
-plot(results)
-```
-
-## Adding a background extinction after reaching a number of living taxa {#EGbg_ext}
-
-For this scenario, we want to generate a pure birth tree until reaching 50 living taxa but with an extinction rate appearing after reaching 30 taxa.
-
-First we need to set up the simulation parameters:
- * The stopping rule (50 taxa max)
- * The birth-death parameters (speciation rate of 1)
-
-```{r}
-## Simulation parameters
-stop_taxa_50 <- list(max.living = 50)
-speciation_1 <- make.bd.params(speciation = 1)
-```
-
-And our change in extinction rate event which triggers after reaching 30 taxa (`taxa.condition(30)`), targets the `"bd.params"` (birth-death parameters) sets the extinction rate to 0.5 (`bd.params.update(extinction = 0.5)`):
-
-```{r}
-## Adding an extinction parameter after 30 taxa
-background.extinction <- make.events(
-                      condition = taxa.condition(30),
-                      target = "bd.params",
-                      modification = bd.params.update(extinction = 0.5))
-```
-
-Once these parameters are defined, we can run the simulations and plot the results:
-
-```{r}
-## Running the simulations
-set.seed(2)
-results <- treats(stop.rule = stop_taxa_50,
-                bd.params = speciation_1,
-                events = background.extinction)
-## Plotting the results
-plot(results, show.tip.label = FALSE)
-```
-
-## Reducing speciation rate after a certain time {#EGred_spec}
-
-For this scenario, we want to generate a birth tree for 6 times units with random speciation rates (i.e.
-drawn from a uniform (0.5;1) distribution) which reduces after time 4 through the simulations to a fixed value of 1/3.
-
-First we need to set up the simulation parameters:
- * The stopping rule (6 time units)
- * The birth-death parameters (speciation randomly drawn between 0.5 and 1)
-
-```{r}
-## Simulation parameters
-stop_time_6 <- list(max.time = 6)
-random_speciation <- make.bd.params(speciation = runif, 
-                                    speciation.args = list(min = 0.5, max = 1))
-```
-
-And our extinction event which triggers after reaching time 4 (`age.condition(4)`), targets the `"bd.params"` and modifies the speciation rate to 1/3:
-
-```{r}
-## Reducing speciation after reaching time 4
-reduced.speciation <- make.events(
-                      condition = age.condition(4),
-                      target = "bd.params",
-                      modification = bd.params.update(speciation = 1/3))
-```
-
-Once these parameters are defined, we can run the simulations and plot the results.
-We can contrast the results with the scenario without an event (but same random seed):
-
-```{r}
-## No event
-set.seed(42)
-no_event <- treats(stop.rule = stop_time_6,
-                 bd.params = random_speciation)
-
-## Reduced speciation event
-set.seed(42)
-reduced_speciation_event <- treats(stop.rule = stop_time_6,
-                                 bd.params = random_speciation,
-                                 events = reduced.speciation)
-
-## Plot both trees
-par(mfrow = c(1, 2))
-plot(no_event, main = "No event", show.tip.label = FALSE)
-axisPhylo()
-plot(reduced_speciation_event, 
-     main = "Reduced speciation after time 5", 
-     show.tip.label = FALSE)
-axisPhylo()
-```
-
-## Changing the trait process after some time {#EG_change_trait}
-
-For this scenario, we want to generate a pure birth tree with a one dimensional Brownian Motion trait for 5 time units which then changes to an OU process.
-
-First we need to set up the simulation parameters:
- * The stopping rule (6 time units)
- * The birth-death parameters (speciation rate of 1)
-
-```{r}
-## Simulation parameters
-stop_time_6 <- list(max.time = 6)
-speciation_1 <- make.bd.params(speciation = 1)
-```
-
-Then set up our trait which is a one dimensional Brownian Motion
-
-```{r}
-## Trait
-simple_bm_trait <- make.traits(n = 1, process = BM.process)
-```
-
-And our event which triggers after reaching time 5 (`age.condition(5)`), targets the `"traits"` and modifies the process to `OU.process`.
-
-```{r}
-## Create an event to change the trait process
-change.process.to.OU <- make.events(
-                  condition    = age.condition(5),
-                  target       = "traits",
-                  modification = traits.update(process = OU.process))
-```
-
-Once these parameters are defined, we can run the simulations and plot the results.
-We can contrast the results with the scenario without an event (but same random seed):
-
-```{r}
-## Run the simulations without change
-set.seed(1)
-no_change <- treats(stop.rule = stop_time_6,
-                  bd.params = speciation_1,
-                  traits    = simple_bm_trait)
-## Run the simulations with change
-set.seed(1)
-process_change <- treats(stop.rule = stop_time_6,
-                       bd.params = speciation_1,
-                       traits    = simple_bm_trait,
-                       events    = change.process.to.OU)
-## Plot the results
-par(mfrow = c(1,2))
-plot(no_change, ylim = c(-7, 7))
-plot(process_change, ylim = c(-7, 7))
-```
-
-## Changing trait correlation after reaching a trait value {#EG_change_correlation}
-
-For this scenario, we want to generate a pure birth tree with a 2 dimensional Brownian Motion trait with a strict correlation between the two dimensions (1:1) that loosen up when a taxa reaches an absolute value of 2.
-
-First we need to set up the simulation parameters:
- * The stopping rule (100 taxa)
- * The birth-death parameters (speciation rate of 1)
-
-```{r}
-## Set the parameters
-stop_taxa_100 <- list(max.taxa = 100)
-speciation_1 <- make.bd.params(speciation = 1)
-```
-
-Then set up our trait which is a 2 dimensional Brownian Motion with a correlation matrix Sigma ()
-
-```{r}
-## A 2D variance covariance matrix
-cor_matrix <- matrix(1, 2, 2)
-
-## A correlated 2D Brownian Motion
-correlated_2D_BM <- make.traits(n = 2, process = BM.process,
-                      process.args = list(Sigma = cor_matrix))
-```
-
-And our event which triggers after a taxa gets the trait value 3 (`trait.condition(3, absolute = TRUE)`), targets the `"traits"` and modifies the traits correlation
-
-```{r}
-## New correlation
-new_cor <- matrix(c(10,3,3,2),2,2)
-
-## Event changing a trait correlation
-correlation.change <- make.events(
-    condition    = trait.condition(3, absolute = TRUE),
-    target       = "traits",
-    modification = traits.update(process.args = list(Sigma = new_cor)))
-```
-
-Once these parameters are defined, we can run the simulations and plot the results.
-We can contrast the results with the scenario without an event (but same random seed):
-
-```{r}
-## Run the simulations
-set.seed(2)
-no_event <- treats(stop.rule = stop_taxa_100,
-                 bd.params = speciation_1,
-                 traits    = correlated_2D_BM)
-set.seed(2)
-change_correlation <- treats(stop.rule = stop_taxa_100,
-                           bd.params = speciation_1,
-                           traits    = correlated_2D_BM,
-                           events    = correlation.change)
-
-## Visual testing
-par(mfrow = c(1,2))
-plot(no_event, trait = c(1,2), main = "Strict correlation")
-plot(change_correlation, trait = c(1,2), main = "Loosened correlation")
-```
-
-And we can visualise this change through time:
-
-```{r}
-## 3D plot
-plot(change_correlation, trait = c(1:2), use.3D = TRUE)
-rglwidget()
-```
-
-## Event for changing a modifier: extinction event increase for species with negative values {#EG_change_modif}
-
-For this scenario, we want to generate a pure birth tree with a one dimensional Brownian Motion trait for 4 time units.
-After 3 time units, we want the speciation rule to increase for species which ancestors have a negative trait value.
-
-First we need to set up the simulation parameters:
- * The stopping rule (5 time units)
- * The birth-death parameters (speciation rate of 1)
-
-```{r}
-## Set the parameters
-stop_time_4 <- list(max.time = 4)
-speciation_1 <- make.bd.params(speciation = 1)
-```
-
-Then set up our trait which is a one dimensional Brownian Motion
-
-```{r}
-## Trait
-simple_bm_trait <- make.traits(n = 1, process = BM.process)
-```
-
-And a modifier that is the default birth-death algorithm rules
-
-```{r}
-## birth-death rules (default)
-default_modifiers <- make.modifiers()
-```
-
-And our extinction event which triggers after reaching time 3 (`age.condition(3)`), targets the `"modifiers"` and modifies birth-death rule as follows:
- * When a species is descendant from a parent with a negative trait value (`negative.trait.condition`);
- * Then increase your chances of going extinct by +1 (`increase.extinction.1`)
-
-```{r}
-## New condition and new modifier (increasing speciation if trait is negative)
-negative.trait.condition <- function(trait.values, lineage) {
-    return(parent.traits(trait.values, lineage) < 0)
-}
-increase.extinction.1 <- function(x, trait.values, lineage) {
-  return(x + 1)
-}
-
-## Update the modifier
-change.speciation <- make.events(
-    condition    = age.condition(3),
-    target       = "modifiers",
-    modification = modifiers.update(speciation = speciation,
-                                    condition  = negative.trait.condition,
-                                    modify     = increase.extinction.1))
-```
-
-Once these parameters are defined, we can run the simulations and plot the results.
-We can contrast the results with the scenario without an event (but same random seed):
-
-```{r}
-set.seed(4)
-no_event <- treats(stop.rule = stop_time_4,
-                 bd.params = speciation_1,
-                 traits    = simple_bm_trait,
-                 modifiers = default_modifiers)
-set.seed(4)
-change_spec <- treats(stop.rule = stop_time_4,
-                    bd.params = speciation_1,
-                    traits    = simple_bm_trait,
-                    modifiers = default_modifiers,
-                    events    = change.speciation)
-
-## Visualise the results
-par(mfrow = c(1,2))
-plot(no_event, main = "No event")
-plot(change_spec, main = "Increase extinction for negative\ntraits after time 3")
-```
-
-## Changing branch length when reaching n taxa {#EG_modify_brlen}
-
-For this scenario, we want to generate a pure birth tree until reaching 100 taxa.
-After reaching 30 taxa we want branch length growth to increase 50 folds.
-
-First we need to set up the simulation parameters:
- * The stopping rule (100 taxa)
- * The birth-death parameters (speciation rate of 1)
-
-```{r}
-## Set the parameters
-stop_taxa_100<- list(max.taxa = 100)
-speciation_1 <- make.bd.params(speciation = 1)
-```
-
-Then a modifier that is the default birth-death algorithm rules
-
-```{r}
-## birth-death rules (default)
-default_modifiers <- make.modifiers()
-```
-
-And event which triggers after reaching 30 taxa (`taxa.condition(30)`), targets the `"modifiers"` and modifies the branch length generation rule to a 50 fold increase (`increase.50.folds`)
-
-```{r}
-## Multiplying branch length 50 folds
-increase.50.folds <- function(x, trait.values, lineage) {
-  return(x * 50)
-}
-## Event for increasing branch length after reaching 30 taxa
-increase_brlen <- make.events(
-                  condition = taxa.condition(30),
-                  target = "modifiers",
-                  modification = modifiers.update(
-                                    branch.length = branch.length,
-                                    modify  = increase.50.folds))
-```
-
-Once these parameters are defined, we can run the simulations and plot the results.
-We can contrast the results with the scenario without an event (but same random seed):
-
-```{r}
-## Run the simulations
-set.seed(5)
-no_event <- treats(stop.rule = stop_taxa_100,
-                 bd.params = speciation_1,
-                 modifiers = default_modifiers)
-set.seed(5)
-increased_brlen <- treats(stop.rule = stop_taxa_100,
-                        bd.params = speciation_1,
-                        modifiers = default_modifiers,
-                        events = increase_brlen)
-
-## Visualise the results
-par(mfrow = c(1,2))
-plot(no_event, main = "No event", show.tip.label = FALSE)
-plot(increased_brlen, main = "Increase branch length\nafter 30 taxa",
-    show.tip.label = FALSE)
-```
-
-## Founding event: generating a subtree with no fossils {#EG_founding_purebirth}
-
-For this scenario, we want to generate a birth-death tree for 4 time units.
-After reaching 10 taxa, one random taxa will give birth to a sub-tree that is a pure birth tree (no extinction).
-
-First we need to set up the simulation parameters:
- * The stopping rule (5 time units)
- * The birth-death parameters (speciation rate of 1 and extinction of 0.2)
-
-```{r}
-## Set up parameters
-stop_time_4 <- list(max.time = 4)
-spec_1_ext_02 <- make.bd.params(speciation = 1, extinction = 0.2)
-```
-
-And our event which triggers after reaching 10 taxa (`taxa.condition(10)`), and generates a subtree ("founding") that is a pure birth tree (no extinction and speciation rate of 2).
-
-```{r}
-## Setting the pure-birth parameters
-speciation_2 <- make.bd.params(speciation = 2)
-
-## Events that generate a new process (founding effects)
-founding_event <- make.events(
-                  condition = taxa.condition(10),
-                  target  = "founding",
-                  modification = founding.event(
-                                    bd.params = speciation_2),
-                  additional.args = list(prefix = "founding_"))
-```
-
-> Note we are prodviding an additional argument `prefix` here so that we can track which species are part of the sub tree for colouring them down the line.
-
-Once these parameters are defined, we can run the simulations and plot the results:
-
-```{r}    
-## Simulations
-set.seed(11)
-founding_tree <- treats(stop.rule = stop_time_4,
-                      bd.params = spec_1_ext_02,
-                      events = founding_event)
-
-## Selecting the edges colours
-tip_values <- rep("black", Ntip(founding_tree))
-tip_values[grep("founding_", founding_tree$tip.label)] <- "orange"
-edge_colors <- match.tip.edge(tip_values, founding_tree, replace.na = "black")
-
-## Plotting the results
-plot(founding_tree, show.tip.label = FALSE, edge.color = edge_colors)
-```
-
-## Founding event: generating a subtree a different process {#EG_founding_traits}
-
-For this scenario, we want to generate a birth-death tree with a one dimensional Brownian Motion trait for 6 time units.
-After a taxon reaches the value 3 or higher, it gives birth to a sub-tree that generates an OU trait with a long-term mean at the value 3.
-
-First we need to set up the simulation parameters:
- * The stopping rule (6 time units)
- * The birth-death parameters (speciation rate of 1 and extinction of 1/3)
-
-```{r}
-## The tree parameters
-stop_time_6 <- list(max.time = 6)
-speciation_1_extinction_03 <- make.bd.params(speciation = 1,
-                                             extinction = 0.3)
-```
-
-Then set up our trait which is a one dimensional Brownian Motion
-
-```{r}
-## Trait
-simple_bm_trait <- make.traits(n = 1, process = BM.process)
-```
-
-When a taxa reaches the value 3 `trait.condition`, it generates a pure birth tree (`speciation = 2`) with an OU trait with the long-term mean value 3.
-
-```{r}
-## The OU trait with a long-term mean value of 3
-OU_3 <- make.traits(process = OU.process,
-                    start = 3, process.args = list(optimum = 3))
-## The pure birth parameters
-speciation_2 <- make.bd.params(speciation = 2)
-
-## The founding event
-new_OU_trait <- make.events(
-                    condition    = trait.condition(3, condition = `>=`),
-                    target       = "founding",
-                    modification = founding.event(
-                                    bd.params = speciation_2,
-                                    traits    = OU_3))
-```
-
-Once these parameters are defined, we can run the simulations and plot the results:
-
-```{r}
-## Simulating the tree
-set.seed(1)
-founding_tree <- treats(stop.rule = stop_time_6,
-                      bd.params = speciation_1_extinction_03,
-                      traits = simple_bm_trait,
-                      events = new_OU_trait)
-plot(founding_tree)
-```
-
-<!--chapter:end:07_specific_examples.Rmd-->
+### Continuous plots (with [mapping](#map.traits))

### 2024-08-19 R/make.traits_fun.R

@@ -20,26 +20,7 @@ check.traits <- function(traits, events = FALSE) {
     ## Loop through each trait for detailed explanation on why it failed
     try_success <- list()
     for(one_trait in 1:length(traits)) {
-        # OLD_LINK_TRAITS_IN
-        # if(all(names(traits[[one_trait]]) %in% c("conditional", "conditioned"))) {
-        #     ## Separately trying either
-        #     test1 <- try(sim.element.trait(traits[[one_trait]]$conditional[[1]], parent.trait, edge.length), silent = TRUE)
-        #     test2 <- try(lapply(traits[[one_trait]]$conditioned, function(x, parent.trait,edge.length) return(sim.element.trait(x[[1]], parent.trait, edge.length)), parent.trait, edge.length), silent = TRUE)
-        #     if(is(test1, "try-error")) {
-        #         try_success[[one_trait]] <- test1
-        #     } else {
-        #         if(is(test2, "try-error")) {
-        #             try_success[[one_trait]] <- test2
-        #         } else {
-        #             try_success[[one_trait]] <- test1
-        #         }
-        #     }
-        # } else {
-        # OLD_LINK_TRAITS_OUT
-        try_success[[one_trait]] <- try(sim.element.trait(traits[[one_trait]], parent.trait, edge.length), silent = TRUE)
-        # OLD_LINK_TRAITS_IN
-        #}
-        # OLD_LINK_TRAITS_OUT
+        try_success[[one_trait]] <- try(sim.element.trait(traits[[one_trait]], rep(parent.trait, length(traits[[one_trait]]$start)), edge.length), silent = TRUE)
     }
 
     ## catch the errors

### 2024-08-28 inst/gitbook/03_make.modifiers.Rmd
book rmd up

@@
 ---
-title: "`make.modifiers`"
-author: "Thomas Guillerme (guillert@tcd.ie)"
+title: "treats manual"
+author: "Thomas Guillerme (t.guillerme@sheffield.ac.uk)"
 date: "`r Sys.Date()`"
-bibliography: [../references.bib]
+bibliography: [../References.bib]
 output:

### 2024-08-28 inst/gitbook/04_make.events.Rmd
book rmd up

@@
 ---
-title: "`make.events`"
-author: "Thomas Guillerme (guillert@tcd.ie)"
+title: "treats manual"
+author: "Thomas Guillerme (t.guillerme@sheffield.ac.uk)"
 date: "`r Sys.Date()`"
-bibliography: [../references.bib]
+bibliography: [../References.bib]
 output:

### 2024-08-28 inst/gitbook/05_other_functionalities.Rmd
book rmd up

@@
 ---
-title: "`other functionalities`"
-author: "Thomas Guillerme (guillert@tcd.ie)"
+title: "treats manual"
+author: "Thomas Guillerme (t.guillerme@sheffield.ac.uk)"
 date: "`r Sys.Date()`"
-bibliography: [../references.bib]
+bibliography: [../References.bib]
 output:

### 2024-08-28 inst/gitbook/07_specific_examples.Rmd
book rmd up

@@
 title: "treats manual"
-author: "Thomas Guillerme (guillert@tcd.ie)"
+author: "Thomas Guillerme (t.guillerme@sheffield.ac.uk)"
 date: "`r Sys.Date()`"
-# bibliography: [../References.bib, ../packages.bib]
+bibliography: [../References.bib]
 output:

### 2024-08-28 inst/gitbook/index.Rmd
book rmd up

@@
 title: "treats manual"
-author: "Thomas Guillerme (guillert@tcd.ie)"
+author: "Thomas Guillerme (t.guillerme@sheffield.ac.uk)"
 date: "`r Sys.Date()`"
@@
 documentclass: book
-bibliography: [../references.bib]
+bibliography: [../References.bib]
 link-citations: yes

### 2024-08-28 inst/gitbook/_book/treats_manual.tex
book up

@@
   pdftitle={treats manual},
-  pdfauthor={Thomas Guillerme (guillert@tcd.ie)},
+  pdfauthor={Thomas Guillerme (t.guillerme@sheffield.ac.uk)},
   hidelinks,
@@
 \title{treats manual}
-\author{Thomas Guillerme (\href{mailto:guillert@tcd.ie}{\nolinkurl{guillert@tcd.ie}})}
-\date{2024-05-15}
+\author{Thomas Guillerme (\href{mailto:t.guillerme@sheffield.ac.uk}{\nolinkurl{t.guillerme@sheffield.ac.uk}})}
+\date{2024-08-28}
 
@@
 \CommentTok{\#\# Simulate disparity and diversity}
+\KeywordTok{set.seed}\NormalTok{(}\DecValTok{43}\NormalTok{)}
 \NormalTok{my\_data \textless{}{-}}\StringTok{ }\KeywordTok{treats}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ my\_params,}
@@
 
-They use a dataset from \citet{beckancient2014} that is a discrete morphological space containing the ordination of the around 400 morphological characters for 50 species and estimated for their 49 descendants into an ordinated shapespace of 99 elements and 97 dimensions.
+They use a dataset from \citet{beck2014ancient} that is a discrete morphological space containing the ordination of the around 400 morphological characters for 50 species and estimated for their 49 descendants into an ordinated shapespace of 99 elements and 97 dimensions.
 
@@
 ## $main$A$process
+## $main$A$process[[1]]
 ## function (x0 = 0, edge.length = 1, Sigma = diag(length(x0)), 
@@
 ## }
-## <bytecode: 0x609b29299e58>
+## <bytecode: 0x56f306f1ce80>
 ## <environment: namespace:treats>
+## 
 ## 
@@
 ## 1 trait for 1 process (A) with one starting value (0).
-## process A uses the following extra argument: fun;
+## process A uses the following extra argument: fun.
 \end{verbatim}
@@
 ## 1 trait for 1 process (A) with one starting value (0).
-## process A uses the following extra arguments: fun,min,max;
+## process A uses the following extra argument: c("fun", "min", "max").
 \end{verbatim}
@@
 ## 2 traits for 2 processes (A, B) with one starting value (0).
-## process A uses the following extra argument: fun;
-## process B uses the following extra arguments: fun,min,max;
+## process A uses the following extra argument: fun.
+## process B uses the following extra argument: c("fun", "min", "max").
 \end{verbatim}
@@
 ## 3 traits for 3 processes (A, B, C) with one starting value (0).
-## process B uses the following extra argument: fun;
-## process C uses the following extra arguments: fun,min,max;
+## process B uses the following extra argument: fun.
+## process C uses the following extra argument: c("fun", "min", "max").
 \end{verbatim}
@@
 ## 3 traits for 3 processes (Normal, LogNormal, Uniform(1,100)) with one starting value (0).
-## process LogNormal uses the following extra argument: fun;
-## process Uniform(1,100) uses the following extra arguments: fun,min,max;
+## process LogNormal uses the following extra argument: fun.
+## process Uniform(1,100) uses the following extra argument: c("fun", "min", "max").
 \end{verbatim}
@@
 
+\hypertarget{linking-traits-together-with-link.traits}{%
+\section{\texorpdfstring{Linking traits together with \texttt{link.traits}}{Linking traits together with link.traits}}\label{linking-traits-together-with-link.traits}}
+
+In the \texttt{treats} algorithm, traits are simulated in parallel for improving performances.
+Most of the cases this is a (really) good thing!
+However, in some cases you might need some of your traits to be simulated sequentially.
+For example if you want a specific trait process or options to be dependent on the result of another trait.
+You can set up these linked traits using the \texttt{link.traits} function.
+This function takes \texttt{base.trait}, a \texttt{"treats"} \texttt{"traits"} object created using \texttt{make.traits} that will be the first trait simulated and the \texttt{next.trait}, a list of \texttt{"treats"} \texttt{"traits"} objects that will be simulated after the \texttt{base.trait} depending on the link type (\texttt{link.type}) and some other arguments (\texttt{link.args}).
+So far the \texttt{link.traits} function only accommodates \texttt{link.type\ =\ "conditional"} for conditional traits.
+
+\hypertarget{conditional-traits}{%
+\subsection{Conditional traits}\label{conditional-traits}}
+
+Say for example that you have a primary trait (\texttt{base.trait}) that is a discrete binary trait that generates values \texttt{0} or \texttt{1}.
+This can correspond to some abiotic trait for example (e.g.~``is the lineage on an island or on the mainland'').
+You can first generate this trait using the \texttt{make.traits} function as detailed above (note here we're using a transition matrix as a specific argument that makes the changes from state 0 to state 1 relatively rare and the reversal even more rare).
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# Setting up a discrete trait}
+\NormalTok{discrete\_trait \textless{}{-}}\StringTok{ }\KeywordTok{make.traits}\NormalTok{(discrete.process,}
+       \DataTypeTok{process.args =} \KeywordTok{list}\NormalTok{(}\DataTypeTok{transitions =} \KeywordTok{matrix}\NormalTok{(}\KeywordTok{c}\NormalTok{(}\DecValTok{3}\NormalTok{, }\FloatTok{0.2}\NormalTok{, }\FloatTok{0.05}\NormalTok{, }\DecValTok{3}\NormalTok{), }\DecValTok{2}\NormalTok{, }\DecValTok{2}\NormalTok{)),}
+       \DataTypeTok{trait.names  =} \StringTok{"discrete"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+Following this trait value, we can create another trait that is conditional to the outcome of this first discrete trait simulation.
+In other words we can specify a choice between two processes depending on the result of the \texttt{base.trait}.
+For example, we can say that if the lineage evolves on an island (state \texttt{"1"}) it will evolve using a Brownian motion process but if it evolves on the mainland (state \texttt{"0"}), it will evolve using an OU process.
+
+We'll need to first generate these two traits independently using \texttt{make.traits} (using different options):
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# A standard Brownian Motion trait}
+\NormalTok{BM\_trait \textless{}{-}}\StringTok{ }\KeywordTok{make.traits}\NormalTok{(}\DataTypeTok{process =}\NormalTok{ BM.process)}
+\CommentTok{\#\# An OU trait with a strong alpha}
+\NormalTok{OU\_trait \textless{}{-}}\StringTok{ }\KeywordTok{make.traits}\NormalTok{(}\DataTypeTok{process =}\NormalTok{ OU.process, }\DataTypeTok{process.args =} \KeywordTok{list}\NormalTok{(}\DataTypeTok{alpha =} \DecValTok{10}\NormalTok{))}
+\CommentTok{\#\# Combining both traits into a list}
+\NormalTok{continuous\_trait \textless{}{-}}\StringTok{ }\KeywordTok{list}\NormalTok{(}\StringTok{"mainland\_case"}\NormalTok{ =}\StringTok{ }\NormalTok{BM\_trait,}
+                         \StringTok{"island\_case"}\NormalTok{   =}\StringTok{ }\NormalTok{OU\_trait)}
+\end{Highlighting}
+\end{Shaded}
+
+We then need to create a list of functions to interpret the results of the first trait (\texttt{discrete\_trait}) so that if the results of \texttt{discrete\_trait} is \texttt{0}, \texttt{BM\_trait} is used, if it's \texttt{1} \texttt{OU\_trait} is used.
+This is done by creating a list of functions that evaluate \texttt{x1}, the outcome of \texttt{discrete\_trait}, and return a logical (\texttt{TRUE} or \texttt{FALSE}).
+For example \texttt{function(x1)\ \{x1\ !=\ 0\}} will return \texttt{TRUE} if the \texttt{discrete\_trait} is not state \texttt{0}:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# Creating a list of functions to choose between either case for the continuous trait}
+\NormalTok{select\_continuous \textless{}{-}}\StringTok{ }\KeywordTok{list}\NormalTok{(}\StringTok{"is.mainland"}\NormalTok{ =}\StringTok{ }\ControlFlowTok{function}\NormalTok{(x1) \{x1 }\OperatorTok{==}\StringTok{ }\DecValTok{0}\NormalTok{\},}
+                          \StringTok{"is.island"}\NormalTok{   =}\StringTok{ }\ControlFlowTok{function}\NormalTok{(x1) \{x1 }\OperatorTok{==}\StringTok{ }\DecValTok{1}\NormalTok{\})}
+\end{Highlighting}
+\end{Shaded}
+
+Note that this list must be in the same order as the \texttt{continuous\_trait}: the functions will be evaluated sequentially (i.e.~test the function \texttt{is.mainland}, then \texttt{is.island}) and the first one to return \texttt{TRUE} will be used to select the function in \texttt{continuous\_trait}.
+For example in our case if the score of \texttt{discrete\_trait} is \texttt{1}, the second function from \texttt{select\_continuous} is \texttt{TRUE} so the second element from \texttt{continuous\_trait} is used (the \texttt{OU\_trait}).
+
+We can then combine all that using \texttt{link.traits}:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# Link these traits}
+\NormalTok{conditional\_trait \textless{}{-}}\StringTok{ }\KeywordTok{link.traits}\NormalTok{(}\CommentTok{\# The first trait to be evaluated}
+                                 \DataTypeTok{base.trait =}\NormalTok{ discrete\_trait,}
+                                 \CommentTok{\# The other traits to be evaluated next}
+                                 \CommentTok{\# (here a list of two traits)}
+                                 \DataTypeTok{next.trait =}\NormalTok{ continuous\_trait,}
+                                 \DataTypeTok{link.type  =} \StringTok{"conditional"}\NormalTok{,}
+                                 \CommentTok{\# The functions to choose between either}
+                                 \CommentTok{\# continuous traits }
+                                 \DataTypeTok{link.args  =}\NormalTok{ select\_continuous)}
+\end{Highlighting}
+\end{Shaded}
+
+And run the usual simulation pipeline with \texttt{treats}:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# Running a tree with this conditional trait}
+\KeywordTok{set.seed}\NormalTok{(}\DecValTok{123}\NormalTok{)}
+\NormalTok{my\_conditional\_tree \textless{}{-}}\StringTok{ }\KeywordTok{treats}\NormalTok{(}\DataTypeTok{stop.rule =} \KeywordTok{list}\NormalTok{(}\DataTypeTok{max.taxa =} \DecValTok{200}\NormalTok{),}
+                              \DataTypeTok{traits    =}\NormalTok{ conditional\_trait)}
+
+\NormalTok{op \textless{}{-}}\StringTok{ }\KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow =} \KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{,}\DecValTok{2}\NormalTok{))}
+\KeywordTok{plot}\NormalTok{(my\_conditional\_tree, }\DataTypeTok{trait =} \DecValTok{1}\NormalTok{, }\DataTypeTok{main =} \StringTok{"The conditional trait"}\NormalTok{)}
+\KeywordTok{plot}\NormalTok{(my\_conditional\_tree, }\DataTypeTok{trait =} \DecValTok{2}\NormalTok{, }\DataTypeTok{main =} \StringTok{"The conditioned trait"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-60-1.pdf}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{par}\NormalTok{(op)}
+\CommentTok{\#\# The clumpy bit around the value 0 corresponds to the OU trait}
+\end{Highlighting}
+\end{Shaded}
+
+Of course, you can combine linked traits with other traits like any other traits in \texttt{treats} using \texttt{make.traits} (for example: \texttt{make.traits(process\ =\ my\_new\_trait,\ add\ =\ conditional\_trait)} to generate more complex traits.
+In this case the traits are generated in parallel (for speed) apart for the specified linked ones.
+
 \hypertarget{traits-implemented-in-treats}{%
@@
 
-\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-56-1.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-61-1.pdf}
 
@@
 
-\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-57-1.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-62-1.pdf}
 
@@
 
-\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-58-1.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-63-1.pdf}
 
@@
 
-\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-59-1.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-64-1.pdf}
 
@@
 
-\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-60-1.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-65-1.pdf}
 
@@
 
-\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-61-1.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-66-1.pdf}
 
@@
 
-\hypertarget{mapping-traits-on-a-tree-map.traits}{%
-\section{\texorpdfstring{Mapping traits on a tree: \texttt{map.traits}}{Mapping traits on a tree: map.traits}}\label{mapping-traits-on-a-tree-map.traits}}
+\hypertarget{map.traits}{%
+\section{\texorpdfstring{Mapping traits on a tree: \texttt{map.traits}}{Mapping traits on a tree: map.traits}}\label{map.traits}}
 
@@
 
-\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-62-1.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-67-1.pdf}
 
@@
 ## }
-## <bytecode: 0x609b211b6680>
+## <bytecode: 0x56f30a0bdef0>
 ## <environment: namespace:treats>
@@
 ## }
-## <bytecode: 0x609b211694e0>
+## <bytecode: 0x56f30a0c0e78>
 ## <environment: namespace:treats>
@@
 ## }
-## <bytecode: 0x609b21164168>
+## <bytecode: 0x56f30a0c0388>
 ## <environment: namespace:treats>
@@
 
-\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-73-1.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-78-1.pdf}
 
@@
 
-\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-76-1.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-81-1.pdf}
 
@@
 
-\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-78-1.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-83-1.pdf}
 
@@
 
-\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-79-1.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-84-1.pdf}
 
@@
 
-\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-82-1.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-87-1.pdf}
 
@@
 
-\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-90-1.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-95-1.pdf}
 
@@
 
-\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-91-1.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-96-1.pdf}
 
@@
 
-\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-92-1.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-97-1.pdf}
 
@@
 
-\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-94-1.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-99-1.pdf}
 
@@
 
-\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-99-1.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-104-1.pdf}
 
@@
 \CommentTok{\#\# Making a list of three traits}
-\NormalTok{list\_of\_traits \textless{}{-}}\StringTok{ }\KeywordTok{make.traits}\NormalTok{(}\DataTypeTok{process =} \KeywordTok{c}\NormalTok{(no.process, BM.process, OU.process), }\DataTypeTok{trait.names =} \KeywordTok{c}\NormalTok{(}\StringTok{"No process (normal)"}\NormalTok{, }\StringTok{"Brownian motion"}\NormalTok{, }\StringTok{"Ornstein{-}Uhlenbeck"}\NormalTok{))}
+\NormalTok{list\_of\_traits \textless{}{-}}\StringTok{ }\KeywordTok{make.traits}\NormalTok{(}\DataTypeTok{process =} \KeywordTok{c}\NormalTok{(no.process, BM.process, OU.process),}
+                              \DataTypeTok{trait.names =} \KeywordTok{c}\NormalTok{(}\StringTok{"No process (normal)"}\NormalTok{, }\StringTok{"Brownian motion"}\NormalTok{, }\StringTok{"Ornstein{-}Uhlenbeck"}\NormalTok{))}
 \CommentTok{\#\# Plotting each trait separately}
@@
 
-\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-102-1.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-107-1.pdf}
 
@@
 
-\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-103-1.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-108-1.pdf}
 
@@
 
-\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-104-1.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-109-1.pdf}
 
@@
 
-\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-105-1.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-110-1.pdf}
 
@@
 
-\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-106-1.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-111-1.pdf}
 
@@
 
-\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-107-1.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-112-1.pdf}
 
@@
 \CommentTok{\#\# Specifying a 3D trait process}
-\NormalTok{my\_3D\_trait \textless{}{-}}\StringTok{ }\KeywordTok{make.traits}\NormalTok{(}\DataTypeTok{n =} \DecValTok{3}\NormalTok{, )}
+\NormalTok{my\_3D\_trait \textless{}{-}}\StringTok{ }\KeywordTok{make.traits}\NormalTok{(}\DataTypeTok{n =} \DecValTok{3}\NormalTok{)}
 \CommentTok{\#\# Simulating a birth{-}death tree with that 3D trait}
@@
                   \DataTypeTok{stop.rule  =} \KeywordTok{list}\NormalTok{(}\DataTypeTok{max.living =} \DecValTok{50}\NormalTok{),}
-                \DataTypeTok{traits    =}\NormalTok{ my\_3D\_trait)}
-\end{Highlighting}
-\end{Shaded}
+                  \DataTypeTok{traits     =}\NormalTok{ my\_3D\_trait,}
+                  \DataTypeTok{null.error =} \DecValTok{100}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## Building the tree:..Done.
+\end{verbatim}
 
@@
 
-\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-109-1.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-114-1.pdf}
 
@@
 
-\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-110-1.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-115-1.pdf}
 
@@
 
-\includegraphics[width=6.5in]{../../../../../../tmp/RtmpcKuqE6/file37eb79e5b154}
+\includegraphics[width=6.5in]{../../../../../../tmp/RtmpNKmbbM/file9d7231f94b74}
 
@@
 
-\includegraphics[width=6.5in]{../../../../../../tmp/RtmpcKuqE6/file37eb42b1cdb9}
+\includegraphics[width=6.5in]{../../../../../../tmp/RtmpNKmbbM/file9d724094c3af}
+
+\hypertarget{using-plot.treats-to-plot-non-treats-objects}{%
+\section{\texorpdfstring{Using \texttt{plot.treats} to plot non-treats objects}{Using plot.treats to plot non-treats objects}}\label{using-plot.treats-to-plot-non-treats-objects}}
+
+If you like the plotting style of \texttt{treats} trees and data, you can always use the \texttt{make.treats} function to plot a tree and some traits as a \texttt{"treats"} object.
+If you have a dataset that contains trait data for tips and nodes you can transform it into a \texttt{"treats"} object and plot it using the generic \texttt{plot.treats} function:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{set.seed}\NormalTok{(}\DecValTok{1}\NormalTok{)}
+\CommentTok{\#\# Generating some random tree}
+\NormalTok{my\_tree \textless{}{-}}\StringTok{ }\KeywordTok{makeNodeLabel}\NormalTok{(}\KeywordTok{rcoal}\NormalTok{(}\DecValTok{20}\NormalTok{))}
+
+\CommentTok{\#\# Generating some random discrete data}
+\NormalTok{discrete\_data \textless{}{-}}\StringTok{ }\KeywordTok{sample}\NormalTok{(}\KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{,}\DecValTok{2}\NormalTok{), }\DecValTok{39}\NormalTok{, }\DataTypeTok{replace =} \OtherTok{TRUE}\NormalTok{)}
+\KeywordTok{names}\NormalTok{(discrete\_data) \textless{}{-}}\StringTok{ }\KeywordTok{c}\NormalTok{(my\_tree}\OperatorTok{$}\NormalTok{tip.label, my\_tree}\OperatorTok{$}\NormalTok{node.label)}
+
+\CommentTok{\#\# Creating a treats object with this data}
+\NormalTok{discrete\_treats \textless{}{-}}\StringTok{ }\KeywordTok{make.treats}\NormalTok{(}\DataTypeTok{tree =}\NormalTok{ my\_tree, }\DataTypeTok{data =}\NormalTok{ discrete\_data)}
+
+\CommentTok{\#\# Plot the data together}
+\KeywordTok{plot}\NormalTok{(discrete\_treats, }\DataTypeTok{legend =} \OtherTok{TRUE}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-118-1.pdf}
+
+The same goes for continuous data:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#\# Generating some random continuous data}
+\NormalTok{continuous\_data \textless{}{-}}\StringTok{ }\KeywordTok{rnorm}\NormalTok{(}\DecValTok{39}\NormalTok{)}
+\KeywordTok{names}\NormalTok{(continuous\_data) \textless{}{-}}\StringTok{ }\KeywordTok{c}\NormalTok{(my\_tree}\OperatorTok{$}\NormalTok{tip.label, my\_tree}\OperatorTok{$}\NormalTok{node.label)}
+
+\CommentTok{\#\# Creating a treats object with this data}
+\NormalTok{continuous\_treats \textless{}{-}}\StringTok{ }\KeywordTok{make.treats}\NormalTok{(}\DataTypeTok{tree =}\NormalTok{ my\_tree, }\DataTypeTok{data =}\NormalTok{ continuous\_data)}
+
+\CommentTok{\#\# Plot the data together}
+\KeywordTok{plot}\NormalTok{(continuous\_treats)}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-119-1.pdf}
 
@@
 
-\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-116-1.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-123-1.pdf}
 
@@
 
-\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-120-1.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-127-1.pdf}
 
@@
 
-\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-123-1.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-130-1.pdf}
 
@@
 
-\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-126-1.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-133-1.pdf}
 
@@
 
-\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-130-1.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-137-1.pdf}
 
@@
 
-\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-134-1.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-141-1.pdf}
 
@@
 
-\includegraphics[width=6.5in]{../../../../../../tmp/RtmpcKuqE6/file37eb202f55d4}
+\includegraphics[width=6.5in]{../../../../../../tmp/RtmpNKmbbM/file9d727fb11a9c}
 
@@
 
-\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-140-1.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-147-1.pdf}
 
@@
 
-\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-144-1.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-151-1.pdf}
 
@@
 
-\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-147-1.pdf}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-154-1.pdf}
 
@@
 
-\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-151-1.pdf}
-
-  \bibliography{../references.bib}
+\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-158-1.pdf}
+
+  \bibliography{../References.bib}
 

### 2024-08-28 inst/gitbook/02_make.traits.Rmd
pre-update

@@
 ---
-title: "`make.traits`"
-author: "Thomas Guillerme (guillert@tcd.ie)"
+title: "treats manual"
+author: "Thomas Guillerme (t.guillerme@sheffield.ac.uk)"
 date: "`r Sys.Date()`"
-bibliography: [../references.bib]
+bibliography: [../References.bib]
 output:
@@
 
-## Mapping traits on a tree: `map.traits`
+## Mapping traits on a tree: `map.traits` {#map.traits}
 

### 2024-08-28 inst/gitbook/06_plotting_treats_objects.Rmd

@@
 ---
-title: "`plot.treats`"
-author: "Thomas Guillerme (guillert@tcd.ie)"
+title: "treats manual"
+author: "Thomas Guillerme (t.guillerme@sheffield.ac.uk)"
 date: "`r Sys.Date()`"
-# bibliography: [../References.bib, ../packages.bib]
+bibliography: [../References.bib]
 output:
@@
 ## Making a list of three traits
-list_of_traits <- make.traits(process = c(no.process, BM.process, OU.process), trait.names = c("No process (normal)", "Brownian motion", "Ornstein-Uhlenbeck"))
+list_of_traits <- make.traits(process = c(no.process, BM.process, OU.process),
+                              trait.names = c("No process (normal)", "Brownian motion", "Ornstein-Uhlenbeck"))
 ## Plotting each trait separately
@@
 ## Specifying a 3D trait process
-my_3D_trait <- make.traits(n = 3, )
+my_3D_trait <- make.traits(n = 3)
 ## Simulating a birth-death tree with that 3D trait
@@
                   stop.rule  = list(max.living = 50),
-                  traits    = my_3D_trait)
+                  traits     = my_3D_trait,
+                  null.error = 100)
 ```
@@
 
-If you like the plotting style of `treats` trees and data, you can always use the `make.traits` function to plot a tree and some traits as a `"treats"` object.
+If you like the plotting style of `treats` trees and data, you can always use the `make.treats` function to plot a tree and some traits as a `"treats"` object.
+If you have a dataset that contains trait data for tips and nodes you can transform it into a `"treats"` object and plot it using the generic `plot.treats` function:
 
-### Discrete plots
+```{r}
+set.seed(1)
+## Generating some random tree
+my_tree <- makeNodeLabel(rcoal(20))
 
-### Continuous plots (with [mapping](#map.traits))
+## Generating some random discrete data
+discrete_data <- sample(c(1,2), 39, replace = TRUE)
+names(discrete_data) <- c(my_tree$tip.label, my_tree$node.label)
+
+## Creating a treats object with this data
+discrete_treats <- make.treats(tree = my_tree, data = discrete_data)
+
+## Plot the data together
+plot(discrete_treats, legend = TRUE)
+```
+
+The same goes for continuous data:
+
+```{r}
+## Generating some random continuous data
+continuous_data <- rnorm(39)
+names(continuous_data) <- c(my_tree$tip.label, my_tree$node.label)
+
+## Creating a treats object with this data
+continuous_treats <- make.treats(tree = my_tree, data = continuous_data)
+
+## Plot the data together
+plot(continuous_treats)
+```
